### 2016-07-13 code/HumanCdGF_microbiota_predict.Rmd
Create predict.Rmd

---
title: "Predictions From Day 0 Community Structures"
author: "Nicholas Lesniak"
date: "July 13, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE}
if(!require(randomForest)) install.packages('randomForest', repos = 'http://cran.us.r-project.org')
library(randomForest)
if(!require(miscTools)) install.packages('miscTools', repos = 'http://cran.us.r-project.org')
library(miscTools)
if(!require(ggplot2)) install.packages('ggplot2', repos = 'http://cran.us.r-project.org')
library(ggplot2)
```

```{r, echo=FALSE}
setwd("~/Documents/Github/Schubert_humanCdGF_XXXX_2016")

#read in files
meta_file <- read.table('data/process/human_CdGF_metadata.txt', sep='\t',header = T, row.names = 'group')
shared_file <- read.table('data/process/human_CdGF.an.unique_list.0.03.subsample.shared', sep='\t',header = T, row.names='Group')
tax_file <- read.table('data/mothur/gf_cdiff.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy', sep='\t',header = T, row.names = 'OTU')

#### subset files to day 0 data
day_0_meta <- meta_file[which(meta_file$day==0),!is.na(meta_file['430-307-D0',])]
# subset meta to day 1 data
cdiff_effect <- c("cdiff_cfu","nextDayCFU_colon","nextDayCFU_any","weight_grams","percent_weightLoss.x","Number","Experiment_day",
                  "loss_below90","day_end","Recip_Dilution_from_DP2","Recip_Dilution_from_adding_to_cells",
                  "Recip_Dilution_from_Postion_on_PlateA","Recip_Dilution_from_Postion_on_PlateB","Average.",
                  "Log_repiricoal_dilution")
day_1_cdiff <- meta_file[which(meta_file$day==1),c(cdiff_effect, 'Mouse_ID')]
names(day_1_cdiff)[!names(day_1_cdiff) %in% 'Mouse_ID'] <- 
     paste('Day_1_',names(day_1_cdiff)[!names(day_1_cdiff) %in% 'Mouse_ID'], sep='')
#subset meta_file to endpoint data
stool_meta_file <- meta_file[!meta_file$sample_type=='cecum',]
day_euthanized_cdiff <- stool_meta_file[which(stool_meta_file$day == stool_meta_file$Euthanized), c(cdiff_effect,'Mouse_ID')]
names(day_euthanized_cdiff)[!names(day_euthanized_cdiff) %in% 'Mouse_ID'] <- 
     paste('Day_Euth_',names(day_euthanized_cdiff)[!names(day_euthanized_cdiff) %in% 'Mouse_ID'], sep='')

#### subset shared
#Create df with relative abundances
day_0_shared <- shared_file[rownames(day_0_meta),]
rel_abund_day_0 <- 100*day_0_shared/unique(apply(day_0_shared, 1, sum))
#Create vector of OTUs with median abundances >1%
day_0_shared_median <- aggregate(rel_abund_day_0, by=list(day_0_meta$cage_id),median)
#cage_IDs <- day_0_shared_median[,"Group.1"]
day_0_shared_median <- day_0_shared_median[,!colnames(day_0_shared_median) %in% 'Group.1']
OTUs_day_0_01 <- apply(day_0_shared_median, 2, max) > 1
#df of OTUs with abundances >1% - by cage and inoculum
rel_abund_day_0 <- rel_abund_day_0[,OTUs_day_0_01]

#Merge data by mouse
Disease_predict_df <- merge(
          merge(day_1_cdiff[,c('Mouse_ID','Day_1_cdiff_cfu','Day_1_percent_weightLoss.x','Day_1_Average.','Day_1_Log_repiricoal_dilution')],
                day_euthanized_cdiff[,c('Mouse_ID','Day_Euth_cdiff_cfu','Day_Euth_percent_weightLoss.x','Day_Euth_Average.','Day_Euth_Log_repiricoal_dilution')], by='Mouse_ID',all.x=T),
          merge(day_0_meta[,c('Mouse_ID','cage_id','Euthanized')],rel_abund_day_0, by='row.names'),
          by='Mouse_ID',all.y=T)
Predict_df <- Disease_predict_df[,!names(Disease_predict_df) %in% c('Mouse_ID','Row.names','cage_id')]
Predict_median_cage <- aggregate(Predict_df, by=list(Disease_predict_df$cage_id), median, na.rm=TRUE)
Predict_median_cage <- data.frame(Predict_median_cage[,!names(Predict_median_cage) %in% 'Group.1'], row.names = Predict_median_cage$Group.1)
```

##Predict Day 1 C. difficile CFU by Day 0 Community 
```{r,echo=FALSE}
#rf to predict 
Predict_day_1_df <- Predict_df[!Predict_df$Day_1_cdiff_cfu %in% NA,c(1,10:117)]
rf_day_1_cfu <- randomForest(Day_1_cdiff_cfu ~ ., data = Predict_day_1_df)
r2 <- rSquared(Predict_day_1_df$Day_1_cdiff_cfu, Predict_day_1_df$Day_1_cdiff_cfu - predict(rf_day_1_cfu,Predict_day_1_df))
#0.826
mse <- mean((Predict_day_1_df$Day_1_cdiff_cfu - predict(rf_day_1_cfu,Predict_day_1_df))^2)
#1.243e+15
ggplot(data = data.frame(actual=Predict_day_1_df$Day_1_cdiff_cfu,pred=predict(rf_day_1_cfu,Predict_day_1_df)),
       aes(x=actual, y=pred)) + geom_point() + geom_abline(color='red')
```

##Predict Endpoint C. difficile CFU by Day 0 Community 
```{r, echo=FALSE}
Predict_euth_cfu_df <- Predict_df[!Predict_df$Day_Euth_cdiff_cfu %in% NA,c(5,10:117)]
rf_day_euth_cfu <- randomForest(Day_Euth_cdiff_cfu ~ ., data = Predict_euth_cfu_df)
r2 <- rSquared(Predict_euth_cfu_df$Day_Euth_cdiff_cfu, Predict_euth_cfu_df$Day_Euth_cdiff_cfu - predict(rf_day_euth_cfu,Predict_euth_cfu_df))
#0.797
mse <- mean((Predict_euth_cfu_df$Day_Euth_cdiff_cfu - predict(rf_day_euth_cfu,Predict_euth_cfu_df))^2)
#3.663e+16
ggplot(data = data.frame(actual=Predict_euth_cfu_df$Day_Euth_cdiff_cfu,pred=predict(rf_day_euth_cfu,Predict_euth_cfu_df)),
       aes(x=actual, y=pred)) + geom_point() + geom_abline(color='red')
```

##Predict Moribund/Persistent Infection by Day 0 Community

```{r, echo=FALSE}
Predict_euth_df <- Predict_df[!Predict_df$Euthanized %in% NA,c(9:117)]
rf_day_euth <- randomForest(Euthanized ~ ., data = Predict_euth_df)
r2 <- rSquared(Predict_euth_df$Euthanized, Predict_euth_df$Euthanized - predict(rf_day_euth,Predict_euth_df))
#0.956
mse <- mean((Predict_euth_df$Euthanized - predict(rf_day_euth,Predict_euth_df))^2)
#0.504
ggplot(data = data.frame(actual=Predict_euth_df$Euthanized,pred=predict(rf_day_euth,Predict_euth_df)),
       aes(x=actual, y=pred)) + geom_point() + geom_abline(color='red')
```

### 2016-07-14 code/HumanCdGF_microbiota_predict.Rmd
added rf function

@@
 shared_file <- read.table('data/process/human_CdGF.an.unique_list.0.03.subsample.shared', sep='\t',header = T, row.names='Group')
-tax_file <- read.table('data/mothur/gf_cdiff.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy', sep='\t',header = T, row.names = 'OTU')
 
-#### subset files to day 0 data
+#### main subset of meta file with all mouse and day 0 info
 day_0_meta <- meta_file[which(meta_file$day==0),!is.na(meta_file['430-307-D0',])]
-# subset meta to day 1 data
-cdiff_effect <- c("cdiff_cfu","nextDayCFU_colon","nextDayCFU_any","weight_grams","percent_weightLoss.x","Number","Experiment_day",
-                  "loss_below90","day_end","Recip_Dilution_from_DP2","Recip_Dilution_from_adding_to_cells",
-                  "Recip_Dilution_from_Postion_on_PlateA","Recip_Dilution_from_Postion_on_PlateB","Average.",
-                  "Log_repiricoal_dilution")
-day_1_cdiff <- meta_file[which(meta_file$day==1),c(cdiff_effect, 'Mouse_ID')]
-names(day_1_cdiff)[!names(day_1_cdiff) %in% 'Mouse_ID'] <- 
-     paste('Day_1_',names(day_1_cdiff)[!names(day_1_cdiff) %in% 'Mouse_ID'], sep='')
-#subset meta_file to endpoint data
-stool_meta_file <- meta_file[!meta_file$sample_type=='cecum',]
-day_euthanized_cdiff <- stool_meta_file[which(stool_meta_file$day == stool_meta_file$Euthanized), c(cdiff_effect,'Mouse_ID')]
+day_0_meta$RowNames <- rownames(day_0_meta)
+day_0_meta$Euth_Early <-ifelse(day_0_meta$Euthanized %in% 10, 'No','Yes')
+####
+#    function to move mouse data from rows by mouse and days --> rows by mouse and columns by days
+####
+# select data to move
+cdiff_effect <- c("cdiff_cfu","percent_weightLoss.x","Log_repiricoal_dilution")
+# function input meta_file as DATAFRAME and mouse dataframe as INPUT
+combine_days_by_mouse <- function(DATAFRAME, INPUT){
+          OUTPUT <- INPUT
+     for(DAY in 1:10){
+          NEW_ROW <- DATAFRAME[which(DATAFRAME$day==DAY),c(cdiff_effect, 'Mouse_ID')]
+          names(NEW_ROW)[!names(NEW_ROW) %in% 'Mouse_ID'] <- 
+               paste('Day_',paste(DAY,paste('_',names(NEW_ROW)[!names(NEW_ROW) %in% 'Mouse_ID'], sep=''), sep=''),sep='')
+          OUTPUT <- merge(OUTPUT, NEW_ROW, by='Mouse_ID', all.x=TRUE)
+     }
+     return(OUTPUT)
+}
+
+test <- merge(day_0_meta,meta_file[which(meta_file$day==3),c(cdiff_effect, 'Mouse_ID')], by='Mouse_ID')
+
+rf_meta_df <- combine_days_by_mouse(meta_file[!meta_file$sample_type=='cecum',], day_0_meta)
+# remove all columns with no data - toxin missing from some days
+rf_meta_df <- rf_meta_df[,colSums(is.na(rf_meta_df)) != nrow(rf_meta_df)]
+# add columns for data from day euthanized
+day_euthanized_cdiff <- meta_file[which(meta_file$sample_type=='stool' & meta_file$day == meta_file$Euthanized), c(cdiff_effect,'Mouse_ID')]
 names(day_euthanized_cdiff)[!names(day_euthanized_cdiff) %in% 'Mouse_ID'] <- 
      paste('Day_Euth_',names(day_euthanized_cdiff)[!names(day_euthanized_cdiff) %in% 'Mouse_ID'], sep='')
+rf_meta_df <- merge(rf_meta_df,day_euthanized_cdiff, by='Mouse_ID', all.x=TRUE)
 
@@
 #Create vector of OTUs with median abundances >1%
-day_0_shared_median <- aggregate(rel_abund_day_0, by=list(day_0_meta$cage_id),median)
-#cage_IDs <- day_0_shared_median[,"Group.1"]
-day_0_shared_median <- day_0_shared_median[,!colnames(day_0_shared_median) %in% 'Group.1']
-OTUs_day_0_01 <- apply(day_0_shared_median, 2, max) > 1
+OTUs_day_0_01 <- apply(rel_abund_day_0, 2, max) > 1
 #df of OTUs with abundances >1% - by cage and inoculum
 rel_abund_day_0 <- rel_abund_day_0[,OTUs_day_0_01]
+OTU_list <- names(rel_abund_day_0)
 
-#Merge data by mouse
-Disease_predict_df <- merge(
-          merge(day_1_cdiff[,c('Mouse_ID','Day_1_cdiff_cfu','Day_1_percent_weightLoss.x','Day_1_Average.','Day_1_Log_repiricoal_dilution')],
-                day_euthanized_cdiff[,c('Mouse_ID','Day_Euth_cdiff_cfu','Day_Euth_percent_weightLoss.x','Day_Euth_Average.','Day_Euth_Log_repiricoal_dilution')], by='Mouse_ID',all.x=T),
-          merge(day_0_meta[,c('Mouse_ID','cage_id','Euthanized')],rel_abund_day_0, by='row.names'),
-          by='Mouse_ID',all.y=T)
-Predict_df <- Disease_predict_df[,!names(Disease_predict_df) %in% c('Mouse_ID','Row.names','cage_id')]
-Predict_median_cage <- aggregate(Predict_df, by=list(Disease_predict_df$cage_id), median, na.rm=TRUE)
-Predict_median_cage <- data.frame(Predict_median_cage[,!names(Predict_median_cage) %in% 'Group.1'], row.names = Predict_median_cage$Group.1)
+#Merge OTU data with mouse/exp data
+Predict_df <- merge(rf_meta_df, rel_abund_day_0, by.x='RowNames', by.y='row.names')
+Predict_df <- Predict_df[,!names(Predict_df) %in% c('Mouse_ID','RowNames','cage_id')]
+
+#create function for random forest
+set.seed(1)
+run_rf_on <- function(DATAFRAME, RESPONSE, OTUs = OTU_list){
+     Predictor_df <- as.data.frame(DATAFRAME[!DATAFRAME[,RESPONSE] %in% NA,c(RESPONSE,OTUs)])
+     RF <- randomForest(as.formula(paste(RESPONSE, '~ .')), data = Predictor_df, importance = TRUE)
+     R2 <- rSquared(Predictor_df[,RESPONSE], Predictor_df[,RESPONSE] - predict(RF,Predictor_df))
+     MSE <- mean((Predictor_df[,RESPONSE] - predict(RF,Predictor_df))^2)
+     IMPORT <- importance(RF)
+     FOREST <- RF
+     PLOT<- ggplot(data = data.frame(actual=Predictor_df[,RESPONSE],pred=predict(RF,Predictor_df)),
+       aes(x=actual, y=pred)) + geom_point() + geom_abline(color='red')
+     return(list(rf=FOREST,r2=R2,mse=MSE,import=IMPORT,plot=PLOT))
+}
 ```
 
-##Predict Day 1 C. difficile CFU by Day 0 Community 
+##Can we predict the endpoint C. difficile CFU with the community structure at Day 0?
+But endpoint has data from Day 2,3,10
+     Might be misleading since contains CFU from different data points
+
+```{r, echo=FALSE}
+Day_Euth_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_Euth_cdiff_cfu')
+Day_Euth_cdiff_cfu_rf$plot
+```
+% Variance explained `r abs(Day_Euth_cdiff_cfu_rf$rf$rsq[length(Day_Euth_cdiff_cfu_rf$rf$rsq)]*100)`
+
+```{r, echo=FALSE}
+Day_Euth_cdiff_cfu_rf$rf
+
+```
+
+##Can we predict moribundity with the community structure at Day 0?
+Need to convert day euthanized to binary categorical
+```{r, echo=FALSE}
+##Fix to be categorical, and not regression
+Predict_early_euth_df <- Predict_df[!Predict_df$Euth_Early %in% NA,c('Euth_Early',OTU_list)]
+rf_early_euth <- randomForest(as.factor(Euth_Early) ~ ., data = Predict_early_euth_df)
+rf_early_euth
+```
+
+##Can we predict the Day 10 C. difficile CFU with the community structure at Day 0?
+
+```{r, echo=FALSE}
+Day_10_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_10_cdiff_cfu')
+Day_10_cdiff_cfu_rf$plot
+
+```
+
+Why day 10? What is significant about day 10?
+
+##Does Day 0 community structure predict the each day C. difficile CFU at a similar?
+
+##Can we predict the Day 1 C. difficile CFU with the community structure at Day 0?
+
 ```{r,echo=FALSE}
 #rf to predict 
-Predict_day_1_df <- Predict_df[!Predict_df$Day_1_cdiff_cfu %in% NA,c(1,10:117)]
-rf_day_1_cfu <- randomForest(Day_1_cdiff_cfu ~ ., data = Predict_day_1_df)
-r2 <- rSquared(Predict_day_1_df$Day_1_cdiff_cfu, Predict_day_1_df$Day_1_cdiff_cfu - predict(rf_day_1_cfu,Predict_day_1_df))
-#0.826
-mse <- mean((Predict_day_1_df$Day_1_cdiff_cfu - predict(rf_day_1_cfu,Predict_day_1_df))^2)
-#1.243e+15
-ggplot(data = data.frame(actual=Predict_day_1_df$Day_1_cdiff_cfu,pred=predict(rf_day_1_cfu,Predict_day_1_df)),
-       aes(x=actual, y=pred)) + geom_point() + geom_abline(color='red')
+Day_1_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_1_cdiff_cfu')
+Day_1_cdiff_cfu_rf$plot
+```
+```{r, echo=FALSE}
+Day_1_cdiff_cfu_rf$rf
+````
+
+# Day 2 C. difficile CFU
+```{r, echo=FALSE}
+Day_2_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_2_cdiff_cfu')
+Day_2_cdiff_cfu_rf$plot
+````
+```{r, echo=FALSE}
+Day_2_cdiff_cfu_rf$rf
+````
+
+# Day 3 C. difficile CFU
+```{r, echo=FALSE}
+Day_3_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_3_cdiff_cfu')
+Day_3_cdiff_cfu_rf$plot
+````
+```{r, echo=FALSE}
+Day_3_cdiff_cfu_rf$rf
+````
+
+# Day 4 C. difficile CFU
+```{r, echo=FALSE}
+Day_4_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_4_cdiff_cfu')
+Day_4_cdiff_cfu_rf$plot
+````
+```{r, echo=FALSE}
+Day_4_cdiff_cfu_rf$rf
+````
+
+# Day 5 C. difficile CFU
+```{r, echo=FALSE}
+Day_5_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_5_cdiff_cfu')
+Day_5_cdiff_cfu_rf$plot
+````
+```{r, echo=FALSE}
+Day_5_cdiff_cfu_rf$rf
+````
+
+# Day 6 C. difficile CFU
+```{r, echo=FALSE}
+Day_6_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_6_cdiff_cfu')
+Day_6_cdiff_cfu_rf$plot
+````
+```{r, echo=FALSE}
+Day_6_cdiff_cfu_rf$rf
+````
+
+# Day 7 C. difficile CFU
+```{r, echo=FALSE}
+Day_7_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_7_cdiff_cfu')
+Day_7_cdiff_cfu_rf$plot
+```
+```{r, echo=FALSE}
+Day_7_cdiff_cfu_rf$rf
 ```
 
-##Predict Endpoint C. difficile CFU by Day 0 Community 
+# Day 8 C. difficile CFU
 ```{r, echo=FALSE}
-Predict_euth_cfu_df <- Predict_df[!Predict_df$Day_Euth_cdiff_cfu %in% NA,c(5,10:117)]
-rf_day_euth_cfu <- randomForest(Day_Euth_cdiff_cfu ~ ., data = Predict_euth_cfu_df)
-r2 <- rSquared(Predict_euth_cfu_df$Day_Euth_cdiff_cfu, Predict_euth_cfu_df$Day_Euth_cdiff_cfu - predict(rf_day_euth_cfu,Predict_euth_cfu_df))
-#0.797
-mse <- mean((Predict_euth_cfu_df$Day_Euth_cdiff_cfu - predict(rf_day_euth_cfu,Predict_euth_cfu_df))^2)
-#3.663e+16
-ggplot(data = data.frame(actual=Predict_euth_cfu_df$Day_Euth_cdiff_cfu,pred=predict(rf_day_euth_cfu,Predict_euth_cfu_df)),
-       aes(x=actual, y=pred)) + geom_point() + geom_abline(color='red')
+Day_8_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_8_cdiff_cfu')
+Day_8_cdiff_cfu_rf$plot
+```
+```{r, echo=FALSE}
+Day_8_cdiff_cfu_rf$rf
 ```
 
-##Predict Moribund/Persistent Infection by Day 0 Community
+# Day 9 C. difficile CFU
+```{r, echo=FALSE}
+Day_9_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_9_cdiff_cfu')
+Day_9_cdiff_cfu_rf$plot
+```
+```{r, echo=FALSE}
+Day_9_cdiff_cfu_rf$rf
+```
 
+# Day 10 C. difficile CFU
 ```{r, echo=FALSE}
-Predict_euth_df <- Predict_df[!Predict_df$Euthanized %in% NA,c(9:117)]
-rf_day_euth <- randomForest(Euthanized ~ ., data = Predict_euth_df)
-r2 <- rSquared(Predict_euth_df$Euthanized, Predict_euth_df$Euthanized - predict(rf_day_euth,Predict_euth_df))
-#0.956
-mse <- mean((Predict_euth_df$Euthanized - predict(rf_day_euth,Predict_euth_df))^2)
-#0.504
-ggplot(data = data.frame(actual=Predict_euth_df$Euthanized,pred=predict(rf_day_euth,Predict_euth_df)),
-       aes(x=actual, y=pred)) + geom_point() + geom_abline(color='red')
+Day_10_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_10_cdiff_cfu')
+Day_10_cdiff_cfu_rf$plot
 ```
+```{r, echo=FALSE}
+Day_10_cdiff_cfu_rf$rf
+```
+
+Predicting CFU from day 1 to 10 based on Day 0 community
+There seems a flucuation in the % variance explained, from none to > 20%
+There s
+
+```{r knitr::kable, echo=FALSE}
+var_by_day <- data.frame(Day=c(1:10),
+                         Percent_Variance_Explained=c(21.86,43.56,4.810,-4.12,2.740,67.63,9.390,33.29,61.19,-6.21),
+                         Mean_of_squared_residuals=c(5.59e15,3.3e15,6.87e16,5.62e16,4.23e15,7.43e15,2.95e16,3.84e15,9.65e15,1.5e17))
+knitr::kable(var_by_day)
+```
+
+##Can we predict day 1 weight change with the community structure at Day 0?
+```{r, echo=FALSE}
+
+```
+
+##Can we predict day 10 weight change with the community structure at Day 0?
+```{r, echo=FALSE}
+
+```
+
+##Can we predict ___________ with the community structure at Day 0?
+```{r, echo=FALSE}
+
+```

### 2016-07-15 code/HumanCdGF_microbiota_predict.Rmd
add variance and cfu by day plots

@@
 ```
-% Variance explained `r abs(Day_Euth_cdiff_cfu_rf$rf$rsq[length(Day_Euth_cdiff_cfu_rf$rf$rsq)]*100)`
 
-```{r, echo=FALSE}
-Day_Euth_cdiff_cfu_rf$rf
+% Variance explained `r Day_Euth_cdiff_cfu_rf$rf$rsq[length(Day_Euth_cdiff_cfu_rf$rf$rsq)]*100`
 
-```
+***
 
@@
 
+% Variance explained `r Day_10_cdiff_cfu_rf$rf$rsq[length(Day_10_cdiff_cfu_rf$rf$rsq)]*100`
+
 Why day 10? What is significant about day 10?
+
+***
 
@@
 ```
-```{r, echo=FALSE}
-Day_1_cdiff_cfu_rf$rf
-````
+
+% Variance explained `r Day_1_cdiff_cfu_rf$rf$rsq[length(Day_1_cdiff_cfu_rf$rf$rsq)]*100`
 
@@
 Day_2_cdiff_cfu_rf$plot
-````
-```{r, echo=FALSE}
-Day_2_cdiff_cfu_rf$rf
-````
+```
+
+% Variance explained `r Day_2_cdiff_cfu_rf$rf$rsq[length(Day_2_cdiff_cfu_rf$rf$rsq)]*100`
 
@@
 Day_3_cdiff_cfu_rf$plot
-````
-```{r, echo=FALSE}
-Day_3_cdiff_cfu_rf$rf
-````
+```
+
+% Variance explained `r Day_3_cdiff_cfu_rf$rf$rsq[length(Day_3_cdiff_cfu_rf$rf$rsq)]*100`
 
@@
 Day_4_cdiff_cfu_rf$plot
-````
-```{r, echo=FALSE}
-Day_4_cdiff_cfu_rf$rf
-````
+```
+
+% Variance explained `r Day_4_cdiff_cfu_rf$rf$rsq[length(Day_4_cdiff_cfu_rf$rf$rsq)]*100`
 
@@
 Day_5_cdiff_cfu_rf$plot
-````
-```{r, echo=FALSE}
-Day_5_cdiff_cfu_rf$rf
-````
+```
+
+% Variance explained `r Day_5_cdiff_cfu_rf$rf$rsq[length(Day_5_cdiff_cfu_rf$rf$rsq)]*100`
 
@@
 Day_6_cdiff_cfu_rf$plot
-````
-```{r, echo=FALSE}
-Day_6_cdiff_cfu_rf$rf
-````
+```
+
+% Variance explained `r Day_6_cdiff_cfu_rf$rf$rsq[length(Day_6_cdiff_cfu_rf$rf$rsq)]*100`
 
@@
 ```
-```{r, echo=FALSE}
-Day_7_cdiff_cfu_rf$rf
-```
+
+% Variance explained `r Day_7_cdiff_cfu_rf$rf$rsq[length(Day_7_cdiff_cfu_rf$rf$rsq)]*100`
 
@@
 ```
-```{r, echo=FALSE}
-Day_8_cdiff_cfu_rf$rf
-```
+
+% Variance explained `r Day_8_cdiff_cfu_rf$rf$rsq[length(Day_8_cdiff_cfu_rf$rf$rsq)]*100`
 
@@
 ```
-```{r, echo=FALSE}
-Day_9_cdiff_cfu_rf$rf
-```
+
+% Variance explained `r Day_9_cdiff_cfu_rf$rf$rsq[length(Day_9_cdiff_cfu_rf$rf$rsq)]*100`
 
@@
 ```
-```{r, echo=FALSE}
-Day_10_cdiff_cfu_rf$rf
-```
+
+% Variance explained `r Day_10_cdiff_cfu_rf$rf$rsq[length(Day_10_cdiff_cfu_rf$rf$rsq)]*100`
 
 Predicting CFU from day 1 to 10 based on Day 0 community
-There seems a flucuation in the % variance explained, from none to > 20%
-There s
 
-```{r knitr::kable, echo=FALSE}
-var_by_day <- data.frame(Day=c(1:10),
-                         Percent_Variance_Explained=c(21.86,43.56,4.810,-4.12,2.740,67.63,9.390,33.29,61.19,-6.21),
-                         Mean_of_squared_residuals=c(5.59e15,3.3e15,6.87e16,5.62e16,4.23e15,7.43e15,2.95e16,3.84e15,9.65e15,1.5e17))
+There seems a flucuation in the % variance explained, which seems to coincide with the flutuations in cdiff cfu
+
+```{r knitr::kable, echo=FALSE, warning=F, message=F}
+var_by_day <- data.frame(Day=c(1:10),Percent_Variance_Explained=NA)
+cdiff_cfu <- vector()
+for (a in 1:10){
+          var_by_day$Percent_Variance_Explained[a] <- eval(parse(text = paste('Day_', a, '_cdiff_cfu_rf$rf$rsq[length(Day_', a, '_cdiff_cfu_rf$rf$rsq)]*100', sep='')))
+          cdiff_cfu <- c(cdiff_cfu,paste('Day_', a, '_cdiff_cfu', sep=''))
+     }
+
 knitr::kable(var_by_day)
+plot(var_by_day, type='n', xlab='Day',ylab='Percent Var Explained')
+lines(var_by_day$Day, var_by_day$Percent_Variance_Explained)
+
+cfu_plot <- cbind(Mouse = paste(Predict_df$human_source,Predict_df$mouse_eartag, sep='_'),Predict_df[, cdiff_cfu])
+names(cfu_plot) <- c('Mouse', c(1:10))
+cfu_plot <- reshape2::melt(cfu_plot, by='Mouse')
+
+ggplot(data=cfu_plot, aes(x=variable, y=value, group=Mouse)) + geom_line(aes(color=Mouse)) + geom_point() + scale_y_log10() +
+     theme_classic() + theme(legend.position='none') + labs(title='C. Difficile CFU by mouse',x='Day',y='CFU')
 ```

### 2016-07-18 code/HumanCdGF_microbiota_predict.Rmd
Split data into test/train for randomforest

@@
 #create function for random forest
-set.seed(1)
-run_rf_on <- function(DATAFRAME, RESPONSE, OTUs = OTU_list){
+run_rf_on <- function(DATAFRAME, RESPONSE, SAMPLE_SPLIT = 2/3,  OTUs = OTU_list){
+set.seed(2)
+     #
      Predictor_df <- as.data.frame(DATAFRAME[!DATAFRAME[,RESPONSE] %in% NA,c(RESPONSE,OTUs)])
-     RF <- randomForest(as.formula(paste(RESPONSE, '~ .')), data = Predictor_df, importance = TRUE)
-     R2 <- rSquared(Predictor_df[,RESPONSE], Predictor_df[,RESPONSE] - predict(RF,Predictor_df))
-     MSE <- mean((Predictor_df[,RESPONSE] - predict(RF,Predictor_df))^2)
+     sample_set <- sample(seq_len(nrow(Predictor_df)), size = floor((SAMPLE_SPLIT)*nrow(Predictor_df)))
+     test <- Predictor_df[-sample_set,]
+     train <- Predictor_df[sample_set,] 
+          RF <- randomForest(as.formula(paste(RESPONSE, '~ .')), data = train, importance = TRUE)
+     R2 <- rSquared(test[,RESPONSE], test[,RESPONSE] - predict(RF,test))
+     MSE <- mean((test[,RESPONSE] - predict(RF,test))^2)
      IMPORT <- importance(RF)
      FOREST <- RF
-     PLOT<- ggplot(data = data.frame(actual=Predictor_df[,RESPONSE],pred=predict(RF,Predictor_df)),
+     PLOT<- ggplot(data = data.frame(actual=test[,RESPONSE],pred=predict(RF,test)),
        aes(x=actual, y=pred)) + geom_point() + geom_abline(color='red')
+     ##MEAN_DECREASE <- varImpPlot(RF)
      return(list(rf=FOREST,r2=R2,mse=MSE,import=IMPORT,plot=PLOT))
@@
 But endpoint has data from Day 2,3,10
+
 Might be misleading since contains CFU from different data points
@@
 % Variance explained `r Day_Euth_cdiff_cfu_rf$rf$rsq[length(Day_Euth_cdiff_cfu_rf$rf$rsq)]*100`  
+r^2^ = `r Day_Euth_cdiff_cfu_rf$r2`  
+MSE = `r Day_Euth_cdiff_cfu_rf$mse`
 
@@
 ##Can we predict moribundity with the community structure at Day 0?
-Need to convert day euthanized to binary categorical
 ```{r, echo=FALSE}
@@
 Day_10_cdiff_cfu_rf$plot
-
 ```
@@
 
-There seems a flucuation in the % variance explained, which seems to coincide with the flutuations in cdiff cfu
+There seems a flucuation in the % variance explained, which seems to coincide with the fluctuations in cdiff cfu
+
+##seed greatly affects variance, likely due to data split, sample size might be too small to make reliable predictions
 
@@
 
+***
+***
+
 ##Can we predict day 1 weight change with the community structure at Day 0?

### 2016-08-08 code/HumanCdGF_microbiota_predict.Rmd
Move echo=F to global opt

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE)
 ```
 
-```{r, echo=FALSE, message=FALSE}
+```{r, message=FALSE}
 if(!require(randomForest)) install.packages('randomForest', repos = 'http://cran.us.r-project.org')
@@
 
-```{r, echo=FALSE}
+```{r}
 setwd("~/Documents/Github/Schubert_humanCdGF_XXXX_2016")
@@
 }
+rf_meta_df <- combine_days_by_mouse(meta_file[!meta_file$sample_type=='cecum',], day_0_meta)
 
-test <- merge(day_0_meta,meta_file[which(meta_file$day==3),c(cdiff_effect, 'Mouse_ID')], by='Mouse_ID')
-
-rf_meta_df <- combine_days_by_mouse(meta_file[!meta_file$sample_type=='cecum',], day_0_meta)
 # remove all columns with no data - toxin missing from some days
@@
 
-```{r, echo=FALSE}
+```{r}
 Day_Euth_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_Euth_cdiff_cfu')
@@
 ##Can we predict moribundity with the community structure at Day 0?
-```{r, echo=FALSE}
+```{r}
 ##Fix to be categorical, and not regression
@@
 
-```{r, echo=FALSE}
+```{r}
 Day_10_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_10_cdiff_cfu')
@@
 
-```{r,echo=FALSE}
+```{r}
 #rf to predict 
@@
 # Day 2 C. difficile CFU
-```{r, echo=FALSE}
+```{r}
 Day_2_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_2_cdiff_cfu')
@@
 # Day 3 C. difficile CFU
-```{r, echo=FALSE}
+```{r}
 Day_3_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_3_cdiff_cfu')
@@
 # Day 4 C. difficile CFU
-```{r, echo=FALSE}
+```{r}
 Day_4_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_4_cdiff_cfu')
@@
 # Day 5 C. difficile CFU
-```{r, echo=FALSE}
+```{r}
 Day_5_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_5_cdiff_cfu')
@@
 # Day 6 C. difficile CFU
-```{r, echo=FALSE}
+```{r}
 Day_6_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_6_cdiff_cfu')
@@
 # Day 7 C. difficile CFU
-```{r, echo=FALSE}
+```{r}
 Day_7_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_7_cdiff_cfu')
@@
 # Day 8 C. difficile CFU
-```{r, echo=FALSE}
+```{r}
 Day_8_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_8_cdiff_cfu')
@@
 # Day 9 C. difficile CFU
-```{r, echo=FALSE}
+```{r}
 Day_9_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_9_cdiff_cfu')
@@
 # Day 10 C. difficile CFU
-```{r, echo=FALSE}
+```{r}
 Day_10_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_10_cdiff_cfu')
@@
 
-##seed greatly affects variance, likely due to data split, sample size might be too small to make reliable predictions
+## seed greatly affects variance, likely due to data split, sample size might be too small to make reliable predictions
 
-```{r knitr::kable, echo=FALSE, warning=F, message=F}
+```{r knitr::kable, warning=F, message=F}
 var_by_day <- data.frame(Day=c(1:10),Percent_Variance_Explained=NA)
@@
 ##Can we predict day 1 weight change with the community structure at Day 0?
-```{r, echo=FALSE}
+```{r}
 
@@
 ##Can we predict day 10 weight change with the community structure at Day 0?
-```{r, echo=FALSE}
+```{r}
 
@@
 ##Can we predict ___________ with the community structure at Day 0?
-```{r, echo=FALSE}
+```{r}
 

### 2016-08-08 code/HumanCdGF_microbiota_predict.Rmd
Add log transformed cfu

@@
 
-#read in files
+# read in files
 meta_file <- read.table('data/process/human_CdGF_metadata.txt', sep='\t',header = T, row.names = 'group')
@@
 
-#### main subset of meta file with all mouse and day 0 info
-day_0_meta <- meta_file[which(meta_file$day==0),!is.na(meta_file['430-307-D0',])]
-day_0_meta$RowNames <- rownames(day_0_meta)
-day_0_meta$Euth_Early <-ifelse(day_0_meta$Euthanized %in% 10, 'No','Yes')
-####
-#    function to move mouse data from rows by mouse and days --> rows by mouse and columns by days
-####
-# select data to move
-cdiff_effect <- c("cdiff_cfu","percent_weightLoss.x","Log_repiricoal_dilution")
-# function input meta_file as DATAFRAME and mouse dataframe as INPUT
-combine_days_by_mouse <- function(DATAFRAME, INPUT){
-          OUTPUT <- INPUT
-     for(DAY in 1:10){
-          NEW_ROW <- DATAFRAME[which(DATAFRAME$day==DAY),c(cdiff_effect, 'Mouse_ID')]
-          names(NEW_ROW)[!names(NEW_ROW) %in% 'Mouse_ID'] <- 
-               paste('Day_',paste(DAY,paste('_',names(NEW_ROW)[!names(NEW_ROW) %in% 'Mouse_ID'], sep=''), sep=''),sep='')
-          OUTPUT <- merge(OUTPUT, NEW_ROW, by='Mouse_ID', all.x=TRUE)
-     }
-     return(OUTPUT)
-}
-rf_meta_df <- combine_days_by_mouse(meta_file[!meta_file$sample_type=='cecum',], day_0_meta)
+# add log transformed CFU
+meta_file$log_cfu <- log10(meta_file$cdiff_cfu + 1)
 
-# remove all columns with no data - toxin missing from some days
-rf_meta_df <- rf_meta_df[,colSums(is.na(rf_meta_df)) != nrow(rf_meta_df)]
-# add columns for data from day euthanized
-day_euthanized_cdiff <- meta_file[which(meta_file$sample_type=='stool' & meta_file$day == meta_file$Euthanized), c(cdiff_effect,'Mouse_ID')]
-names(day_euthanized_cdiff)[!names(day_euthanized_cdiff) %in% 'Mouse_ID'] <- 
-     paste('Day_Euth_',names(day_euthanized_cdiff)[!names(day_euthanized_cdiff) %in% 'Mouse_ID'], sep='')
-rf_meta_df <- merge(rf_meta_df,day_euthanized_cdiff, by='Mouse_ID', all.x=TRUE)
+# remove cecum samples
+meta_file <- meta_file[! meta_file$sample_type == 'cecum', ]
 
-#### subset shared
+# subset shared
 #Create df with relative abundances
-day_0_shared <- shared_file[rownames(day_0_meta),]
-rel_abund_day_0 <- 100*day_0_shared/unique(apply(day_0_shared, 1, sum))
+stool_shared <- shared_file[rownames(meta_file[meta_file$day %in% 0, ]) ,]
+# use %in% instead of == because to ignore NA
+# rename day 0 communities by mouse id
+rownames(stool_shared) <- meta_file$Mouse_ID[rownames(meta_file) %in% rownames(stool_shared) ]
+rel_abund_stool <- 100*stool_shared/unique(apply(stool_shared, 1, sum))
 #Create vector of OTUs with median abundances >1%
-OTUs_day_0_01 <- apply(rel_abund_day_0, 2, max) > 1
+OTUs_stool_0_01 <- apply(rel_abund_stool, 2, max) > 1
 #df of OTUs with abundances >1% - by cage and inoculum
-rel_abund_day_0 <- rel_abund_day_0[,OTUs_day_0_01]
-OTU_list <- names(rel_abund_day_0)
+rel_abund_stool <- rel_abund_stool[,OTUs_stool_0_01]
+OTU_list <- colnames(rel_abund_stool)
 
 #Merge OTU data with mouse/exp data
-Predict_df <- merge(rf_meta_df, rel_abund_day_0, by.x='RowNames', by.y='row.names')
-Predict_df <- Predict_df[,!names(Predict_df) %in% c('Mouse_ID','RowNames','cage_id')]
+#Predict_df <- merge(rf_meta_df, rel_abund_stool, by.x='RowNames', by.y='row.names')
+#Predict_df <- Predict_df[,!names(Predict_df) %in% c('Mouse_ID','RowNames','cage_id')]
+
+#predict day cfu by day 0 community
+predict_df <- merge(meta_file[meta_file$day==2, c('Mouse_ID', 'log_cfu') ], stool_shared, by='Mouse_ID', by.y='row.names', all.y = T)
+predict_df <- predict_df[ ,!(colnames(predict_df) %in% "Mouse_ID")]
+
+RF <- randomForest(log_cfu ~ ., data = predict_df[!is.na(predict_df$log_cfu),], importance = TRUE)
+
+varImpPlot(RF)
+ggplot(data = data.frame(actual=predict_df[,'log_cfu'],pred=predict(RF,predict_df)),
+       aes(x=actual, y=pred)) + geom_point() + geom_abline(color='red')
+percent_var_expl <- RF$rsq[length(RF$rsq)]*100
 

### 2016-08-09 code/HumanCdGF_microbiota_inocula_family.R
adding more comparison plot code for donor inoculua

@@ -8,6 +8,9 @@
 #
 ###################
 
+install.packages("RColorBrewer")
+library(RColorBrewer)
+
 setwd("~/Documents/Github/Schubert_humanCdGF_XXXX_2016")
 
 #read in files
@@ -42,12 +45,17 @@ source('code/Sum_OTU_by_Tax.R')
 family_level_RA <- sum_OTU_by_tax_level(2,rel_abund_0.01,tax_file)
 
 #reorganize the file a bit 
+inoc <- meta_file$human_source[meta_file$cage_id=='inoculum']
+inoc <- as.vector(inoc)
 family_level_RA[25] <- inoc
 colnames(family_level_RA)[25] <- "donor"
 
-#remove donor 810 because we can't use it anymore
+#remove donor 810 and duplicates of others
 
 family_level_RA <- family_level_RA[-8,]
+family_level_RA <- family_level_RA[-1,]
+family_level_RA <- family_level_RA[-4,]
+family_level_RA <- family_level_RA[-17,]
 
 inoc_labels <- colnames(family_level_RA[1:24])
 inoc_list <- as.vector(inoc_labels)
@@ -55,8 +63,8 @@ inoc_list <- as.vector(inoc_labels)
 #make multiplot of family plots for each donor
 #run this entire code to the end to plot the multipanel plot for all donors
 #set up layout
-par(mar=c(2,1.5,1.1,1), oma=c(6,7,6,0.5))
-layout(matrix(1:30, nrow=5))
+par(mar=c(2,1.5,1.1,1), oma=c(7,7,7,0.5))
+layout(matrix(1:24, nrow=6))
 
 #loop to plot all donors 
 for(d in family_level_RA$donor){
@@ -65,20 +73,23 @@ for(d in family_level_RA$donor){
   #make barplot
   z <- barplot(t(one_d[,1:24]), beside = TRUE, xaxt='n', col="white", ylim=c(0,max(one_d[,1:24])+5), ylab = "Relative Abundance (%)", main=paste("Donor", d))
   axis(1, at=seq(1,24, by=1), labels=FALSE, tick=FALSE)
-  text(seq(1,24, by=1), par("usr")[3] - 11, labels=inoc_list, srt = 65, pos = 1, xpd = TRUE, cex=0.6)
[…]

### 2016-08-09 code/HumanCdGF_microbiota_predict.Rmd
Create for loop RF by log10cfu day

@@
 # read in files
-meta_file <- read.table('data/process/human_CdGF_metadata.txt', sep='\t',header = T, row.names = 'group')
-shared_file <- read.table('data/process/human_CdGF.an.unique_list.0.03.subsample.shared', sep='\t',header = T, row.names='Group')
+meta_file   <- read.table('data/process/human_CdGF_metadata.txt', sep = '\t',
+                        header = T, row.names = 'group')
+shared_file <- read.table('data/process/human_CdGF.an.unique_list.0.03.subsample.shared',
+                          sep='\t',header = T, row.names='Group')
 
@@
 meta_file$log_cfu <- log10(meta_file$cdiff_cfu + 1)
-
 # remove cecum samples
@@
 # rename day 0 communities by mouse id
-rownames(stool_shared) <- meta_file$Mouse_ID[rownames(meta_file) %in% rownames(stool_shared) ]
-rel_abund_stool <- 100*stool_shared/unique(apply(stool_shared, 1, sum))
+rownames(stool_shared) <- meta_file$Mouse_ID[rownames(meta_file) %in% 
+                                               rownames(stool_shared) ]
+rel_abund_stool <- 100 * stool_shared / unique(apply(stool_shared, 1, sum))
 #Create vector of OTUs with median abundances >1%
@@
 #df of OTUs with abundances >1% - by cage and inoculum
-rel_abund_stool <- rel_abund_stool[,OTUs_stool_0_01]
+rel_abund_stool <- rel_abund_stool[ ,OTUs_stool_0_01]
 OTU_list <- colnames(rel_abund_stool)
 
-#Merge OTU data with mouse/exp data
-#Predict_df <- merge(rf_meta_df, rel_abund_stool, by.x='RowNames', by.y='row.names')
-#Predict_df <- Predict_df[,!names(Predict_df) %in% c('Mouse_ID','RowNames','cage_id')]
+#random forest regression model to determine Day 0 OTU correlated with increased CFU
+seed <- 18
+n_trees <- 2000
+set.seed(seed)
 
-#predict day cfu by day 0 community
-predict_df <- merge(meta_file[meta_file$day==2, c('Mouse_ID', 'log_cfu') ], stool_shared, by='Mouse_ID', by.y='row.names', all.y = T)
+#log_cfu_obs <- meta_file[meta_file$day %in% c(0:10), c('Mouse_ID', 'day', 'log_cfu') ]
+#log_cfu_obs$predict <- NA
+#log_cfu_obs[log_cfu_obs$day==2, ] <- predict(RF, rel_abund_stool)
+
+importance_df <- data.frame(matrix(nrow = length(OTU_list), 
+                                   ncol = max(meta_file$day, na.rm = T)),
+                            row.names = OTU_list)
+rsq_df <- data.frame(matrix(nrow = 1, ncol = max(meta_file$day, na.rm = T)),
+                         row.names = 'RSQ')
+prediction_df <- data.frame(matrix(nrow = 0, ncol = 4))
+
+for (i in 1:10){
+  predict_df <- merge(meta_file[meta_file$day==i, c('Mouse_ID', 'log_cfu') ],
+                    rel_abund_stool, by='Mouse_ID', by.y='row.names', all.y = T)
   predict_df <- predict_df[ ,!(colnames(predict_df) %in% "Mouse_ID")]
+  RF <- randomForest(log_cfu ~ ., data = predict_df[!is.na(predict_df$log_cfu), ],
+                   importance = TRUE, ntree = n_trees)
+  rsq_df[i] <- RF$rsq[length(RF$rsq)]
+  # sort the features by their effect on % increase in the standard error
+  importance_df[ , i] <- importance(RF)[,1]
+  colnames(importance_df)[i] <- paste('day', i, 'importance', sep = '_')
+# fit the model back on the datar and tack it to the end of the counts file
+  random_forest_output <- merge(meta_file[meta_file$day==i, c('Mouse_ID', 'day', 'log_cfu') ],
+                              as.data.frame(predict(RF, rel_abund_stool)), 
+                              by='Mouse_ID', by.y='row.names')
+  colnames(random_forest_output)[3:4] <- c('observed', 'predicted')
+  prediction_df <- rbind(prediction_df, random_forest_output)  
+}
 
-RF <- randomForest(log_cfu ~ ., data = predict_df[!is.na(predict_df$log_cfu),], importance = TRUE)
+ggplot(data = prediction_df, aes(x=observed, y=predicted, color = day)) + geom_point()
 
-varImpPlot(RF)
-ggplot(data = data.frame(actual=predict_df[,'log_cfu'],pred=predict(RF,predict_df)),
-       aes(x=actual, y=pred)) + geom_point() + geom_abline(color='red')
-percent_var_expl <- RF$rsq[length(RF$rsq)]*100
+rsq_df <- t(rsq_df)
+rsq_df <- as.data.frame(cbind(rsq_df, day = c(1:10)))
+ggplot(data = rsq_df, aes(x = factor(day), y = RSQ)) + geom_bar(stat = 'identity')
 
 #create function for random forest
-run_rf_on <- function(DATAFRAME, RESPONSE, SAMPLE_SPLIT = 2/3,  OTUs = OTU_list){
+run_rf_on <- function(DATAFRAME, RESPONSE,  OTUs = OTU_list){
 set.seed(2)
@@
      Predictor_df <- as.data.frame(DATAFRAME[!DATAFRAME[,RESPONSE] %in% NA,c(RESPONSE,OTUs)])
-     sample_set <- sample(seq_len(nrow(Predictor_df)), size = floor((SAMPLE_SPLIT)*nrow(Predictor_df)))
-     test <- Predictor_df[-sample_set,]
-     train <- Predictor_df[sample_set,] 
     RF <- randomForest(as.formula(paste(RESPONSE, '~ .')), data = train, importance = TRUE)
@@
 }
+
+# let's plot the top 12 features
+#n_features <- 12
+#importance_subset <- importance_sorted[1:n_features]
+
+# let's get Rsq for the subset
+#rf_partial <- randomForest(log_cfu ~ ., data=predict_df[!is.na(predict_df$log_cfu),
+                            c( 'log_cfu', names(importance_sorted)[1:n_features])], 
+                           importance=TRUE, ntree=n_trees)
+# write(paste("partial", n_features, rf_partial$rsq[n_trees], "\t"), file="data/process/random_forest.data", append=T)
+
 ```
@@
 ##Fix to be categorical, and not regression
-Predict_early_euth_df <- Predict_df[!Predict_df$Euth_Early %in% NA,c('Euth_Early',OTU_list)]
-rf_early_euth <- randomForest(as.factor(Euth_Early) ~ ., data = Predict_early_euth_df)
+Predict_early_euth_df <- data.frame(Euth_Early = ifelse(meta_file[meta_file$day %in% 0,
+                                                                  'Euthanized'] %in% 10,
+                                                        'No','Yes'), 
+                                    row.names = meta_file[meta_file$day %in% 0, 'Mouse_ID'])
+Predict_early_euth_df <- merge(Predict_early_euth_df, rel_abund_stool, by='row.names')
+Predict_early_euth_df <- Predict_early_euth_df[,!(colnames(Predict_early_euth_df) %in% "Row.names")]
+rf_early_euth <- randomForest(as.factor(Euth_Early) ~ ., data = Predict_early_euth_df, 
+                              importance = TRUE, ntree = n_trees)
+# write(paste("binary", ncol(rel_abund_stool), rf_early_euth$err.rate[n_trees], "\t"), file="data/process/random_forest.data", append=T)
 rf_early_euth

### 2016-08-10 code/HumanCdGF_microbiota_inocula_family.R
lots of progress on figure 1 and 2 tweaking

@@ -89,83 +89,11 @@ dev.off()
 #barplot of donors for comparison of figure types 
 getPalette = colorRampPalette(brewer.pal(9, "Set1"))
 p = 17
-barplot(t(family_level_RA[,1:24]), ylim=c(0,100), col = getPalette(p), names.arg=family_level_RA$donor, cex.names = 0.8)
-
-
-
-
-
-# stacked barplot of tax comp (family) in inoculum samples
-colors <- primary.colors(length(names(phylum_level_RA)))
-legend_labels <- names(phylum_level_RA)
-legend_labels[names(phylum_level_RA)=="Bacteria_unclassified"] <- "Other"
-legend_labels[names(phylum_level_RA)=="Candidatus_Saccharibacteria"] <- "C. Saccharibacteria"
-legend_labels[names(phylum_level_RA)=="Deinococcus-Thermus"] <- "D. Thermus"
-
-#stacked bar plot with base
 par(mar=c(7.1, 4.1, 4.1, 8.1), xpd=TRUE)
-barplot(t(phylum_level_RA), ylab='Relative Abundance', las=2, cex.sub=0.9,
-        main="Taxonomic Composition of GF mice inocula",cex.lab=0.9, cex.axis = 0.7, cex.names = 0.7,
-        border= colors, col = colors, ylim=c(0,100))
-     mtext('Human Source', side=1, line=5, cex=0.9)
-     legend('right',fill=colors,legend_labels, inset=c(-0.16,0), bty='n', border=colors, y.intersp = 0.8, cex=0.9)
+x <- barplot(t(family_level_RA[,1:24]), ylab= "Relative Abundance %", ylim=c(0,100), xaxt='n', col = getPalette(p), names.arg=family_level_RA$donor, cex.names = 0.9, cex.axis=0.9, cex.lab=0.9)
+legend_labels <- colnames(family_level_RA[,1:24])
+legend_labels <- as.character(legend_labels)
+legend("right",fill=getPalette(p),legend_labels, bty='n', cex=0.6, y.intersp = 0.7, inset=c(-0.25,0))
+box()
+text(cex=0.8, x=x+0.5, y=-4, label=family_level_RA$donor, xpd=TRUE, srt=45, pos=2)
 
-# community structure from 454 data
[…]

### 2016-08-10 code/mouse_d0_community.R

@@ -70,6 +70,12 @@ cages <- na.omit(cages)
 rel_d0[82] <- cages
 colnames(rel_d0)[82] <- "cage"
 
+#add a column of donors to the df
+donors <- meta_file$human_source[meta_file$day==0]
+donors <- na.omit(donors)
+rel_d0[83] <- donors
+colnames(rel_d0)[83] <- "donor"
+ 
 d0_med <- aggregate(rel_d0[, 1:81], list(rel_d0$cage), median)
 d0_med_phy <- sum_OTU_by_tax_level(5, d0_med, tax_file)
 #removes inoculum cage
@@ -89,18 +95,38 @@ legend('right',fill=colors,legend_labels, bty='n', inset=c(-0.1,0), border=color
 #make dataframe of median family abundances and organize it 
 d0_med_fam <- sum_OTU_by_tax_level(2, d0_med, tax_file)
 rownames(d0_med_fam) <- cage_only_char
-colnames(d0_med_fam)[1] <- "donor"
+colnames(d0_med_fam)[1] <- "cage"
+
+#just make table of human source and cage unique and then merge by cage, take out unused ones after 
+
+dc <- meta_file[!meta_file$cage_id=='inoculum',2:3]
+uc <- unique(dc)
+d0_med_fam <- merge(d0_med_fam, uc, by.x='cage', by.y='cage_id')
+
+#get rid of samples/cages not used here: 
+d0_med_fam <- d0_med_fam[-23,]
+d0_med_fam <- d0_med_fam[-4,]
+d0_med_fam <- d0_med_fam[-3,]
+d0_med_fam <- d0_med_fam[-6,]
+d0_med_fam <- d0_med_fam[-4,]
+d0_med_fam <- d0_med_fam[-4,]
+d0_med_fam <- d0_med_fam[-12,]
+
+
 
 #colorbrewer
 getPalette = colorRampPalette(brewer.pal(9, "Set1"))
 
 #run this whole command to make figure of family by cage/donor - add colorbrewer!
-n=19
+#main="Taxonomic composition of mice on D0 by donor, family level"
+n=16
 par(mar=c(7.1, 4.1, 4.1, 8.1), xpd=TRUE)
-barplot(t(d0_med_fam[2:20]), ylab='Relative Abundance', main="Taxonomic composition of mice on D0 by cage, family", 
[…]

### 2016-08-10 code/mouse_d0_community.R
finished d0 plot by donor

@@ -137,8 +137,8 @@ d0_upper <- aggregate(rel_d0[1:81], list(rel_d0$cage), FUN= quantile, probs =0.7
 d0_lower <- aggregate(rel_d0[1:81], list(rel_d0$cage), FUN= quantile, probs =0.25)
 d0_upper_fam <- sum_OTU_by_tax_level(2, d0_upper, tax_file)
 d0_lower_fam <- sum_OTU_by_tax_level(2, d0_lower, tax_file)
-colnames(d0_upper_fam)[1] <- "donor"
-colnames(d0_lower_fam)[1] <- "donor"
+colnames(d0_upper_fam)[1] <- "cage"
+colnames(d0_lower_fam)[1] <- "cage"
 fam_list <- as.vector(fam_labels)
 
 #to save as pdf to results, can't get the size right to render but fine for now
@@ -147,18 +147,20 @@ pdf(file="results/figures/figure2.pdf", width=25, height=15)
 #run this entire code to the end to plot the multipanel plot for all donors
 #set up layout
 par(mar=c(2,1.5,1.1,1), oma=c(6,7,6,0.5))
-layout(matrix(1:30, nrow=5))
+layout(matrix(1:16, nrow=4))
 
 #loop to plot all donors 
-for(d in d0_med_fam$cage){
-  one_d <- subset(d0_med_fam, cage == d)
-  one_uci <- subset(d0_upper_fam, cage == d)
-  one_lci <- subset(d0_lower_fam, cage == d) 
+for(d in d0_med_fam$human_source){
+  one_d <- subset(d0_med_fam, human_source == d)
+  for(c in one_d$cage){
+  one_uci <- subset(d0_upper_fam, cage == c)
+  one_lci <- subset(d0_lower_fam, cage == c) 
   medians <- t(one_d[,2:20])
   uci <- t(one_uci[,2:20])
   lci <- t(one_lci[,2:20])
+  }
   #make barplot
-  z <- barplot(t(one_d[,2:20]), beside = TRUE, xaxt='n', col="white", ylim=c(0,max(uci)+5), ylab = "Relative Abundance (%)", main=paste("Cage", d))
+  z <- barplot(t(one_d[,2:20]), beside = TRUE, xaxt='n', col="white", ylim=c(0,max(uci)+5), ylab = "Relative Abundance (%)", main=paste('Donor', d))
[…]

### 2016-08-10 code/HumanCdGF_microbiota_predict.Rmd
Use RF pedictors to create plots

@@
 ```{r, message=FALSE}
-if(!require(randomForest)) install.packages('randomForest', repos = 'http://cran.us.r-project.org')
-library(randomForest)
-if(!require(miscTools)) install.packages('miscTools', repos = 'http://cran.us.r-project.org')
-library(miscTools)
-if(!require(ggplot2)) install.packages('ggplot2', repos = 'http://cran.us.r-project.org')
-library(ggplot2)
+pack_used <- c('randomForest','miscTools','ggplot2','reshape2', 'pROC')
+for (dep in pack_used){
+  if (dep %in% installed.packages()[,"Package"] == FALSE){
+    install.packages(as.character(dep), quiet=TRUE);
+  }
+  library(dep, verbose=FALSE, character.only=TRUE)
+}
 ```
@@
 for (i in 1:10){
-  predict_df <- merge(meta_file[meta_file$day==i, c('Mouse_ID', 'log_cfu') ],
+  predict_all_df <- merge(meta_file[meta_file$day==i, c('Mouse_ID', 'day' , 'log_cfu') ],
                     rel_abund_stool, by='Mouse_ID', by.y='row.names', all.y = T)
+  predict_df <- predict_all_df[,! colnames(predict_all_df) %in% 'day']
   predict_df <- predict_df[ ,!(colnames(predict_df) %in% "Mouse_ID")]
@@
 # fit the model back on the datar and tack it to the end of the counts file
-  random_forest_output <- merge(meta_file[meta_file$day==i, c('Mouse_ID', 'day', 'log_cfu') ],
-                              as.data.frame(predict(RF, rel_abund_stool)), 
-                              by='Mouse_ID', by.y='row.names')
-  colnames(random_forest_output)[3:4] <- c('observed', 'predicted')
+  random_forest_output <- cbind(predict_all_df[!is.na(predict_all_df$log_cfu), c('day','log_cfu')],
+                              RF$predicted)
+  colnames(random_forest_output) <- c('day', 'observed', 'predicted')
   prediction_df <- rbind(prediction_df, random_forest_output)  
@@
 
-ggplot(data = prediction_df, aes(x=observed, y=predicted, color = day)) + geom_point()
+ggplot(data = prediction_df, aes(x=observed, y=predicted)) + geom_point() + 
+        facet_wrap( ~ day, nrow = 2) + 
+        labs(title = 'Random Forest Prediction of C Diff CFU dpi from Day 0 Community',
+        x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)')
 
@@
 rsq_df <- as.data.frame(cbind(rsq_df, day = c(1:10)))
-ggplot(data = rsq_df, aes(x = factor(day), y = RSQ)) + geom_bar(stat = 'identity')
+ggplot(data = rsq_df, aes(x = factor(day), y = RSQ)) + geom_point() + theme_bw() +
+        labs(title = 'Variation Explained by Day of C Diff CFU dpi from Day 0 Community',
+        x= 'Day', y = 'Variance Explained')
 
-#create function for random forest
-run_rf_on <- function(DATAFRAME, RESPONSE,  OTUs = OTU_list){
-set.seed(2)
-     #
-     Predictor_df <- as.data.frame(DATAFRAME[!DATAFRAME[,RESPONSE] %in% NA,c(RESPONSE,OTUs)])
-    RF <- randomForest(as.formula(paste(RESPONSE, '~ .')), data = train, importance = TRUE)
-     R2 <- rSquared(test[,RESPONSE], test[,RESPONSE] - predict(RF,test))
-     MSE <- mean((test[,RESPONSE] - predict(RF,test))^2)
-     IMPORT <- importance(RF)
-     FOREST <- RF
-     PLOT<- ggplot(data = data.frame(actual=test[,RESPONSE],pred=predict(RF,test)),
-       aes(x=actual, y=pred)) + geom_point() + geom_abline(color='red')
-     ##MEAN_DECREASE <- varImpPlot(RF)
-     return(list(rf=FOREST,r2=R2,mse=MSE,import=IMPORT,plot=PLOT))
-}
+n_important <- 20
+median_importance <- names(head(sort((apply(importance_df,1, median)), decreasing = T), n_important))
+med_import_df <- melt(cbind(rownames(importance_df), importance_df)[median_importance, ])
+colnames(med_import_df) <- c('OTU', 'Day', 'Importance')
+ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + facet_wrap( ~ OTU, nrow = 2) +
+        geom_point() + scale_x_discrete(labels = c(1:10)) +
+        labs(title = 'OTU Importance in Prediction of C Diff CFU dpi from Day 0 Community\n (Top 20 Median OTUs)',
+        x= 'Day', y = '% Increase in MSE') + theme_bw()
+        coord_flip()
 
-# let's plot the top 12 features
+#let's plot the top 12 features
 #n_features <- 12
 #importance_subset <- importance_sorted[1:n_features]
-
 # let's get Rsq for the subset
 #rf_partial <- randomForest(log_cfu ~ ., data=predict_df[!is.na(predict_df$log_cfu),
-                            c( 'log_cfu', names(importance_sorted)[1:n_features])], 
-                           importance=TRUE, ntree=n_trees)
+#                            c( 'log_cfu', names(importance_sorted)[1:n_features])], 
+#                           importance=TRUE, ntree=n_trees)
 # write(paste("partial", n_features, rf_partial$rsq[n_trees], "\t"), file="data/process/random_forest.data", append=T)
-
 ```
@@
 ```{r}
-Day_Euth_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_Euth_cdiff_cfu')
-Day_Euth_cdiff_cfu_rf$plot
+
+CFU_euth <- merge(meta_file[meta_file$day==meta_file$Euthanized, c('Mouse_ID', 'log_cfu') ],
+                    rel_abund_stool, by='Mouse_ID', by.y='row.names', all.y = T)
+CFU_euth <- CFU_euth[ ,!(colnames(CFU_euth) %in% "Mouse_ID")]
+
+RF_CFU_day_euth <- randomForest(log_cfu ~ ., data = CFU_euth[!is.na(CFU_euth$log_cfu), ],
+                   importance = TRUE, ntree = n_trees)
+rsq_day_euth <- RF_CFU_day_euth$rsq[length(RF_CFU_day_euth$rsq)]
+# sort the features by their effect on % increase in the standard error
+importance_day_euth <- importance(RF_CFU_day_euth)[,1]
+# fit the model back on the datar and tack it to the end of the counts file
+rf_predict_cfu_output <- merge(meta_file[meta_file$day==meta_file$Euthanized, c('Mouse_ID', 'day', 'log_cfu') ],
+                              as.data.frame(predict(RF_CFU_day_euth, rel_abund_stool)), 
+                              by='Mouse_ID', by.y='row.names')
+colnames(rf_predict_cfu_output)[3:4] <- c('observed', 'predicted')
+
+str(RF_CFU_day_euth)
+TEST <- cbind(CFU_euth[!is.na(CFU_euth$log_cfu), 'log_cfu'], data.frame(RF_CFU_day_euth$predicted))
+colnames(TEST) <- c('observed', 'predicted')
+ggplot(data = TEST, aes(x=observed, y=predicted)) + geom_point()
+
+ggplot(data = rf_predict_cfu_output, aes(x=observed, y=predicted)) + geom_point() + 
+        labs(title = 'Random Forest Prediction of C Diff CFU on Day Euthanized from Day 0 Community',
+        x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)') + theme_bw()
+        ggtitle(expression(atop("Random Forest Prediction of C Diff CFU on Day Euthanized from Day 0 Community",
+                                atop(paste('% Var Explained = ', rsq_day_euth), ""))))
+
+top_important_OTU <- data.frame(head(sort(importance_day_euth, decreasing = T), n_important))
+colnames(top_important_OTU) <- 'Importance'
+top_important_OTU$OTU <- rownames(top_important_OTU)
+
+ggplot(data = top_important_OTU, aes(x = factor(OTU), y = Importance)) + 
+        geom_point() + scale_x_discrete(limits = rev(top_important_OTU$OTU)) +
+        labs(title = 'OTU Importance in Prediction of C Diff CFU dpi from Day 0 Community\n (Top 20 Median OTUs)',
+        x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
+
 ```
-
-% Variance explained `r Day_Euth_cdiff_cfu_rf$rf$rsq[length(Day_Euth_cdiff_cfu_rf$rf$rsq)]*100`  
-r^2^ = `r Day_Euth_cdiff_cfu_rf$r2`  
-MSE = `r Day_Euth_cdiff_cfu_rf$mse`
 
@@
 # write(paste("binary", ncol(rel_abund_stool), rf_early_euth$err.rate[n_trees], "\t"), file="data/process/random_forest.data", append=T)
-rf_early_euth
+#varImpPlot(rf_early_euth)
+#prediction_euth <- as.numeric(predict(rf_early_euth, Predict_early_euth_df[,-1]))
+#ROC_euth <- roc(as.numeric(Predict_early_euth_df$Euth_Early), prediction_euth)
+#plot(ROC_euth)
 ```

### 2016-08-12 code/HumanCdGF_microbiota_predict.Rmd
compare day 0 comm to cfu

@@
 ---
-title: "Predictions From Day 0 Community Structures"
+title: "Predict Disease Outcome by Community Structures"
 author: "Nicholas Lesniak"
@@
                          row.names = 'RSQ')
+mse_df <- data.frame(matrix(nrow = 1, ncol = max(meta_file$day, na.rm = T)),
+                         row.names = 'MSE')
 prediction_df <- data.frame(matrix(nrow = 0, ncol = 4))
@@
   rsq_df[i] <- RF$rsq[length(RF$rsq)]
+  mse_df[i] <- RF$mse[length(RF$mse)]
+
   # sort the features by their effect on % increase in the standard error
@@
 
+mse_df <- t(mse_df)
+mse_df <- as.data.frame(cbind(mse_df, day = c(1:10)))
+ggplot(data = mse_df, aes(x = factor(day), y = MSE)) + geom_point() + theme_bw() +
+        labs(title = 'Variation Explained by Day of C Diff CFU dpi from Day 0 Community',
+        x= 'Day', y = 'Variance Explained')
+
 n_important <- 20
@@
         x= 'Day', y = '% Increase in MSE') + theme_bw()
-        coord_flip()
 
@@
 ##Can we predict the endpoint C. difficile CFU with the community structure at Day 0?
-But endpoint has data from Day 2,3,10
-
-Might be misleading since contains CFU from different data points
+Endpoint has data from Day 2,3,10 - might be misleading since contains CFU from different data points
 
@@
                    importance = TRUE, ntree = n_trees)
-rsq_day_euth <- RF_CFU_day_euth$rsq[length(RF_CFU_day_euth$rsq)]
+rsq_day_euth <- round(RF_CFU_day_euth$rsq[length(RF_CFU_day_euth$rsq)], 2)
+mse_day_euth <- round(RF_CFU_day_euth$mse[length(RF_CFU_day_euth$mse)], 2)
 # sort the features by their effect on % increase in the standard error
@@
 # fit the model back on the datar and tack it to the end of the counts file
-rf_predict_cfu_output <- merge(meta_file[meta_file$day==meta_file$Euthanized, c('Mouse_ID', 'day', 'log_cfu') ],
-                              as.data.frame(predict(RF_CFU_day_euth, rel_abund_stool)), 
-                              by='Mouse_ID', by.y='row.names')
-colnames(rf_predict_cfu_output)[3:4] <- c('observed', 'predicted')
-
-str(RF_CFU_day_euth)
-TEST <- cbind(CFU_euth[!is.na(CFU_euth$log_cfu), 'log_cfu'], data.frame(RF_CFU_day_euth$predicted))
-colnames(TEST) <- c('observed', 'predicted')
-ggplot(data = TEST, aes(x=observed, y=predicted)) + geom_point()
+rf_predict_cfu_output <- cbind(CFU_euth[!is.na(CFU_euth$log_cfu), 'log_cfu'], data.frame(RF_CFU_day_euth$predicted))
+colnames(rf_predict_cfu_output) <- c('observed', 'predicted')
 
 ggplot(data = rf_predict_cfu_output, aes(x=observed, y=predicted)) + geom_point() + 
-        labs(title = 'Random Forest Prediction of C Diff CFU on Day Euthanized from Day 0 Community',
-        x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)') + theme_bw()
-        ggtitle(expression(atop("Random Forest Prediction of C Diff CFU on Day Euthanized from Day 0 Community",
-                                atop(paste('% Var Explained = ', rsq_day_euth), ""))))
+        labs(x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)') + theme_bw() + 
+        ggtitle("Random Forest Prediction of C Diff CFU on Day Euthanized from Day 0 Community") +
+        annotate('text', x=0.15*(max(rf_predict_cfu_output$observed)), y=0.95*(max(rf_predict_cfu_output$predicted)), 
+                 label=paste('R^2 = ', rsq_day_euth,'\nMSE = ', mse_day_euth, sep = ''))
 
@@
         geom_point() + scale_x_discrete(limits = rev(top_important_OTU$OTU)) +
-        labs(title = 'OTU Importance in Prediction of C Diff CFU dpi from Day 0 Community\n (Top 20 Median OTUs)',
+        labs(title = 'OTU Importance in Prediction of C Diff CFU dpi from Day 0 Community',
         x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
-
 ```
@@
 
-##Can we predict the Day 10 C. difficile CFU with the community structure at Day 0?
-
-```{r}
-Day_10_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_10_cdiff_cfu')
-Day_10_cdiff_cfu_rf$plot
-```
-
-% Variance explained `r Day_10_cdiff_cfu_rf$rf$rsq[length(Day_10_cdiff_cfu_rf$rf$rsq)]*100`
-
-Why day 10? What is significant about day 10?
-
 ***
 
-##Does Day 0 community structure predict the each day C. difficile CFU at a similar?
+##Does Day Community structure predict the each day C. difficile CFU?
 
-##Can we predict the Day 1 C. difficile CFU with the community structure at Day 0?
-
-```{r}
-#rf to predict 
-Day_1_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_1_cdiff_cfu')
-Day_1_cdiff_cfu_rf$plot
-```
-
-% Variance explained `r Day_1_cdiff_cfu_rf$rf$rsq[length(Day_1_cdiff_cfu_rf$rf$rsq)]*100`
-
-# Day 2 C. difficile CFU
-```{r}
-Day_2_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_2_cdiff_cfu')
-Day_2_cdiff_cfu_rf$plot
-```
-
-% Variance explained `r Day_2_cdiff_cfu_rf$rf$rsq[length(Day_2_cdiff_cfu_rf$rf$rsq)]*100`
-
-# Day 3 C. difficile CFU
-```{r}
-Day_3_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_3_cdiff_cfu')
-Day_3_cdiff_cfu_rf$plot
-```
-
-% Variance explained `r Day_3_cdiff_cfu_rf$rf$rsq[length(Day_3_cdiff_cfu_rf$rf$rsq)]*100`
-
-# Day 4 C. difficile CFU
-```{r}
-Day_4_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_4_cdiff_cfu')
-Day_4_cdiff_cfu_rf$plot
-```
-
-% Variance explained `r Day_4_cdiff_cfu_rf$rf$rsq[length(Day_4_cdiff_cfu_rf$rf$rsq)]*100`
-
-# Day 5 C. difficile CFU
-```{r}
-Day_5_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_5_cdiff_cfu')
-Day_5_cdiff_cfu_rf$plot
-```
-
-% Variance explained `r Day_5_cdiff_cfu_rf$rf$rsq[length(Day_5_cdiff_cfu_rf$rf$rsq)]*100`
-
-# Day 6 C. difficile CFU
-```{r}
-Day_6_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_6_cdiff_cfu')
-Day_6_cdiff_cfu_rf$plot
-```
-
-% Variance explained `r Day_6_cdiff_cfu_rf$rf$rsq[length(Day_6_cdiff_cfu_rf$rf$rsq)]*100`
-
-# Day 7 C. difficile CFU
-```{r}
-Day_7_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_7_cdiff_cfu')
-Day_7_cdiff_cfu_rf$plot
-```
-
-% Variance explained `r Day_7_cdiff_cfu_rf$rf$rsq[length(Day_7_cdiff_cfu_rf$rf$rsq)]*100`
-
-# Day 8 C. difficile CFU
-```{r}
-Day_8_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_8_cdiff_cfu')
-Day_8_cdiff_cfu_rf$plot
-```
-
-% Variance explained `r Day_8_cdiff_cfu_rf$rf$rsq[length(Day_8_cdiff_cfu_rf$rf$rsq)]*100`
-
-# Day 9 C. difficile CFU
-```{r}
-Day_9_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_9_cdiff_cfu')
-Day_9_cdiff_cfu_rf$plot
-```
-
-% Variance explained `r Day_9_cdiff_cfu_rf$rf$rsq[length(Day_9_cdiff_cfu_rf$rf$rsq)]*100`
-
-# Day 10 C. difficile CFU
-```{r}
-Day_10_cdiff_cfu_rf <- run_rf_on(Predict_df,'Day_10_cdiff_cfu')
-Day_10_cdiff_cfu_rf$plot
-```
-
-% Variance explained `r Day_10_cdiff_cfu_rf$rf$rsq[length(Day_10_cdiff_cfu_rf$rf$rsq)]*100`
-
-Predicting CFU from day 1 to 10 based on Day 0 community
-
-There seems a flucuation in the % variance explained, which seems to coincide with the fluctuations in cdiff cfu
-
-## seed greatly affects variance, likely due to data split, sample size might be too small to make reliable predictions
-
-```{r knitr::kable, warning=F, message=F}
-var_by_day <- data.frame(Day=c(1:10),Percent_Variance_Explained=NA)
-cdiff_cfu <- vector()
-for (a in 1:10){
-          var_by_day$Percent_Variance_Explained[a] <- eval(parse(text = paste('Day_', a, '_cdiff_cfu_rf$rf$rsq[length(Day_', a, '_cdiff_cfu_rf$rf$rsq)]*100', sep='')))
-          cdiff_cfu <- c(cdiff_cfu,paste('Day_', a, '_cdiff_cfu', sep=''))
-     }
-
-knitr::kable(var_by_day)
-plot(var_by_day, type='n', xlab='Day',ylab='Percent Var Explained')
-lines(var_by_day$Day, var_by_day$Percent_Variance_Explained)
-
-cfu_plot <- cbind(Mouse = paste(Predict_df$human_source,Predict_df$mouse_eartag, sep='_'),Predict_df[, cdiff_cfu])
-names(cfu_plot) <- c('Mouse', c(1:10))
-cfu_plot <- reshape2::melt(cfu_plot, by='Mouse')
-
-ggplot(data=cfu_plot, aes(x=variable, y=value, group=Mouse)) + geom_line(aes(color=Mouse)) + geom_point() + scale_y_log10() +
-     theme_classic() + theme(legend.position='none') + labs(title='C. Difficile CFU by mouse',x='Day',y='CFU')
-```
-
-***
-***
-
-##Can we predict day 1 weight change with the community structure at Day 0?
-```{r}
-
-```
-
-##Can we predict day 10 weight change with the community structure at Day 0?
-```{r}
-
-```
-
-##Can we predict ___________ with the community structure at Day 0?
-```{r}
-
-```
+RF all mice/days together
+RF cfu vs community by day

### 2016-08-12 code/HumanCdGF_microbiota_predict.Rmd
Add comparisons between day/s and CFU

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
 ```
@@
                           sep='\t',header = T, row.names='Group')
+tax_file <- read.table('data/mothur/gf_cdiff.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy', 
+                       sep='\t',header = T, row.names = 'OTU')
+#source('code/tax_level.R')
 
@@
 # use %in% instead of == because to ignore NA
-# rename day 0 communities by mouse id
+# rename day 0 community rows by mouse id - since meta rows have day in row names, use mouse id for both
 rownames(stool_shared) <- meta_file$Mouse_ID[rownames(meta_file) %in% 
@@
 prediction_df <- data.frame(matrix(nrow = 0, ncol = 4))
+rsq_features <- data.frame('Day'=NA, 'Features (n)'= NA, 'RSQ'=NA, 'MSE'=NA)
 
@@
   importance_df[ , i] <- importance(RF)[,1]
+  importance_sorted <- sort(importance(RF)[,1], decreasing = T)
   colnames(importance_df)[i] <- paste('day', i, 'importance', sep = '_')
@@
   prediction_df <- rbind(prediction_df, random_forest_output) 
+  for (k in 1:50){
+    n_features <- k
+    rf_partial <- randomForest(log_cfu ~ ., data=predict_df[!is.na(predict_df$log_cfu),
+                            c( 'log_cfu', names(importance_sorted)[1:n_features])], 
+                           importance=TRUE, ntree=n_trees)
+    rsq_features <- rbind(rsq_features, c(i, k, tail(rf_partial$rsq, 1), tail(rf_partial$mse, 1)))
   }
+}
+rsq_features <- rsq_features[-1,]
 
@@
         labs(title = 'Variation Explained by Day of C Diff CFU dpi from Day 0 Community',
-        x= 'Day', y = 'Variance Explained')
+        x= 'Day', y = 'R^2')
 
@@
 ggplot(data = mse_df, aes(x = factor(day), y = MSE)) + geom_point() + theme_bw() +
-        labs(title = 'Variation Explained by Day of C Diff CFU dpi from Day 0 Community',
-        x= 'Day', y = 'Variance Explained')
+        labs(title = 'MSE by Day of C Diff CFU dpi from Day 0 Community',
+        x= 'Day', y = 'MSE')
+
+ggplot(rsq_features, aes(x=Features..n., y=RSQ, group=(Day), color=factor(Day))) + geom_line() +
+  xlim(0,15) + labs(title= 'R^2 from RF with top n features\n
+                    Day of C Diff CFU dpi from Day 0 Community', x = 'Features (n)', y = 'R^2') + 
+  scale_color_discrete(name = 'Day') + theme_bw()
 
@@
 
-#let's plot the top 12 features
-#n_features <- 12
-#importance_subset <- importance_sorted[1:n_features]
-# let's get Rsq for the subset
-#rf_partial <- randomForest(log_cfu ~ ., data=predict_df[!is.na(predict_df$log_cfu),
-#                            c( 'log_cfu', names(importance_sorted)[1:n_features])], 
-#                           importance=TRUE, ntree=n_trees)
-# write(paste("partial", n_features, rf_partial$rsq[n_trees], "\t"), file="data/process/random_forest.data", append=T)
 ```
@@
 
+n_feat_day_euth <- data.frame('Features (n)'= NA, 'RSQ'=NA, 'MSE'=NA)
+importance_sorted_euth <- sort(importance(RF)[,1], decreasing = T)
+
+for (k in 1:50){
+    n_features <- k
+    rf_partial <- randomForest(log_cfu ~ ., data=CFU_euth[!is.na(CFU_euth$log_cfu),
+                            c( 'log_cfu', names(importance_sorted_euth)[1:n_features])], 
+                           importance=TRUE, ntree=n_trees)
+    n_feat_day_euth <- rbind(n_feat_day_euth, c(k, tail(rf_partial$rsq, 1), tail(rf_partial$mse, 1)))
+}
+n_feat_day_euth <- n_feat_day_euth[-1,]
+
+ggplot(n_feat_day_euth, aes(x=Features..n., y=RSQ, )) + geom_line() +
+  xlim(0,15) + labs(title= 'R^2 from RF with top n features\n
+                    Day of C Diff CFU t euthanization from Day 0 Community', x = 'Features (n)', y = 'R^2') + 
+  theme_bw()
+
 ggplot(data = top_important_OTU, aes(x = factor(OTU), y = Importance)) + 
@@
         x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
+
 ```
+
+## Does Day Community structure predict C. difficile CFU of that day?
+
+RF all mice/days together
+RF cfu vs community by day
+
+```{r}
+
+#remove c diff from shared
+cdiff_otu <- rownames(tax_file[grep('Clostridium_XI', tax_file$Taxonomy),])
+shared_X_cdiff <- shared_file[, ! colnames(shared_file) %in% cdiff_otu]
+shared_X_cdiff_ra <-  100 * shared_X_cdiff / unique(apply(shared_X_cdiff, 1, sum))
+OTUs_01 <- apply(shared_X_cdiff_ra, 2, max) > 1
+shared_X_cdiff_ra <- shared_X_cdiff_ra[, OTUs_01]
+# create data frame with shared and cfu and day
+CFU_community_by_day <- merge(meta_file[meta_file$day %in% c(1:10), c('day','log_cfu')], shared_X_cdiff_ra,
+                              by = 'row.names')
+
+###################################
+
+importance_df_day <- data.frame(matrix(nrow = sum(OTUs_01*1), 
+                                   ncol = max(meta_file$day, na.rm = T)),
+                            row.names = colnames(shared_X_cdiff_ra))
+rsq_df_day <- data.frame(matrix(nrow = 1, ncol = max(meta_file$day, na.rm = T)),
+                         row.names = 'RSQ')
+mse_df_day <- data.frame(matrix(nrow = 1, ncol = max(meta_file$day, na.rm = T)),
+                         row.names = 'MSE')
+prediction_df_day <- data.frame(matrix(nrow = 0, ncol = 4))
+rsq_features_day <- data.frame('Day'=NA, 'Features (n)'= NA, 'RSQ'=NA, 'MSE'=NA)
+
+for (i in 1:10){
+  predict_all_df <- CFU_community_by_day[CFU_community_by_day$day == i ,]
+  predict_df <- predict_all_df[!is.na(predict_all_df$log_cfu), 
+                           !colnames(predict_all_df) %in% c('Row.names', 'day')]
+  RF <- randomForest(log_cfu ~ ., data = predict_df,
+                   importance = TRUE, ntree = n_trees)
+  rsq_df_day[i] <- RF$rsq[length(RF$rsq)]
+  mse_df_day[i] <- RF$mse[length(RF$mse)]
+
+  # sort the features by their effect on % increase in the standard error
+  importance_df_day[ , i] <- importance(RF)[,1]
+  importance_sorted <- sort(importance(RF)[,1], decreasing = T)
+  colnames(importance_df_day)[i] <- paste('day', i, 'importance', sep = '_')
+# fit the model back on the datar and tack it to the end of the counts file
+  random_forest_output <- cbind(predict_all_df[!is.na(predict_all_df$log_cfu), c('day','log_cfu')],
+                              RF$predicted)
+  colnames(random_forest_output) <- c('day', 'observed', 'predicted')
+  prediction_df_day <- rbind(prediction_df_day, random_forest_output) 
+  for (k in 1:50){
+    n_features <- k
+    rf_partial <- randomForest(log_cfu ~ ., 
+                               data=predict_df[ , c( 'log_cfu', names(importance_sorted)[1:n_features])], 
+                               importance=TRUE, ntree=n_trees)
+    rsq_features_day <- rbind(rsq_features_day, c(i, k, tail(rf_partial$rsq, 1), tail(rf_partial$mse, 1)))
+  }
+}
+rsq_features_day <- rsq_features_day[-1,]
+
+ggplot(data = prediction_df_day, aes(x=observed, y=predicted)) + geom_point() + 
+        facet_wrap( ~ day, nrow = 2) + 
+        labs(title = 'Random Forest Prediction of C Diff CFU dpi from Community on same day',
+        x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)')
+
+rsq_df_day <- t(rsq_df_day)
+rsq_df_day <- as.data.frame(cbind(rsq_df_day, day = c(1:10)))
+ggplot(data = rsq_df_day, aes(x = factor(day), y = RSQ)) + geom_point() + theme_bw() +
+        labs(title = 'Variation Explained by Day of C Diff CFU dpi from Community on same day',
+        x= 'Day', y = 'R^2')
+
+mse_df_day <- t(mse_df_day)
+mse_df_day <- as.data.frame(cbind(mse_df_day, day = c(1:10)))
+ggplot(data = mse_df_day, aes(x = factor(day), y = MSE)) + geom_point() + theme_bw() +
+        labs(title = 'MSE by Day of C Diff CFU dpi from Community on same day',
+        x= 'Day', y = 'MSE')
+
+ggplot(rsq_features_day, aes(x=Features..n., y=RSQ, group=(Day), color=factor(Day))) + geom_line() +
+  xlim(0,15) + labs(title= 'R^2 from RF with top n features\n
+                    Day of C Diff CFU dpi from Community on same day', x = 'Features (n)', y = 'R^2') + 
+  scale_color_discrete(name = 'Day') + theme_bw()
+
+n_important <- 20
+median_importance <- names(head(sort((apply(importance_df_day,1, median)), decreasing = T), n_important))
+med_import_df <- melt(cbind(rownames(importance_df_day), importance_df_day)[median_importance, ])
+colnames(med_import_df) <- c('OTU', 'Day', 'Importance')
+
+ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + facet_wrap( ~ OTU, nrow = 2) +
+        geom_point() + scale_x_discrete(labels = c(1:10)) +
+        labs(title = 'OTU Importance in Prediction of C Diff CFU dpi from Community on same day\n (Top 20 Median OTUs)',
+        x= 'Day', y = '% Increase in MSE') + theme_bw()
+
+```
+
+## Does removing the early euthanized samples improve prediction in days 1 to 3?
+
+```{r}
+
+without_early_euth <- meta_file[meta_file$Euthanized %in% 10 ,c('day', 'Mouse_ID', 'log_cfu')]
+without_early_euth <- without_early_euth[without_early_euth$day %in% c(1:4), ]
+
+without_by_day <- merge(without_early_euth, shared_X_cdiff_ra, by  = 'row.names')
+without_day_0 <- merge(without_early_euth, rel_abund_stool, by='Mouse_ID', by.y='row.names' )
+
+###################################
+
+importance_persist <- data.frame(matrix(nrow = length(OTU_list), 
+                                   ncol = max(without_early_euth$day, na.rm = T)),
+                            row.names = OTU_list)
+rsq_persist <- data.frame(matrix(nrow = 1, ncol = max(without_early_euth$day, na.rm = T)),
+                         row.names = 'RSQ')
+mse_persist <- data.frame(matrix(nrow = 1, ncol = max(without_early_euth$day, na.rm = T)),
+                         row.names = 'MSE')
+prediction_persist <- data.frame(matrix(nrow = 0, ncol = 4))
+features_persist <- data.frame('Day'=NA, 'Features (n)'= NA, 'RSQ'=NA, 'MSE'=NA)
+
+for (i in 1:4){
+  predict_all_df <- without_day_0[without_day_0$day == i ,]
+  predict_df <- predict_all_df[!is.na(predict_all_df$log_cfu), 
+                           !colnames(predict_all_df) %in% c('Mouse_ID', 'day')]
+  RF <- randomForest(log_cfu ~ ., data = predict_df,
+                   importance = TRUE, ntree = n_trees)
+  rsq_persist[i] <- RF$rsq[length(RF$rsq)]
+  mse_persist[i] <- RF$mse[length(RF$mse)]
+
+  # sort the features by their effect on % increase in the standard error
+  importance_persist[ , i] <- importance(RF)[,1]
+  importance_sorted <- sort(importance(RF)[,1], decreasing = T)
+  colnames(importance_persist)[i] <- paste('day', i, 'importance', sep = '_')
+# fit the model back on the datar and tack it to the end of the counts file
+  random_forest_output <- cbind(predict_all_df[!is.na(predict_all_df$log_cfu), c('day','log_cfu')],
+                              RF$predicted)
+  colnames(random_forest_output) <- c('day', 'observed', 'predicted')
+  prediction_persist <- rbind(prediction_persist, random_forest_output) 
+  for (k in 1:50){
+    n_features <- k
+    rf_partial <- randomForest(log_cfu ~ ., 
+                               data=predict_df[ , c( 'log_cfu', names(importance_sorted)[1:n_features])], 
+                               importance=TRUE, ntree=n_trees)
+    features_persist <- rbind(features_persist, c(i, k, tail(rf_partial$rsq, 1), tail(rf_partial$mse, 1)))
+  }
+}
+features_persist <- features_persist[-1,]
+
+ggplot(data = prediction_persist, aes(x=observed, y=predicted)) + geom_point() + 
+        facet_grid( ~ day) + 
+        labs(title = 'Random Forest Prediction of C Diff CFU dpi from Community on Day 0\nwithout mice euthanized early',
+        x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)')
+
+rsq_persist <- t(rsq_persist)
+rsq_persist <- as.data.frame(cbind(rsq_persist, day = c(1:10)))
+ggplot(data = rsq_persist, aes(x = factor(day), y = RSQ)) + geom_point() + theme_bw() +
+        labs(title = 'Variation Explained by Day of C Diff CFU dpi from Community on Day 0\nwithout mice euthanized early',
+        x= 'Day', y = 'R^2')
+
+mse_persist <- t(mse_persist)
+mse_persist <- as.data.frame(cbind(mse_persist, day = c(1:10)))
+ggplot(data = mse_persist, aes(x = factor(day), y = MSE)) + geom_point() + theme_bw() +
+        labs(title = 'MSE by Day of C Diff CFU dpi from Community on Day 0\nwithout mice euthanized early',
+        x= 'Day', y = 'MSE')
+
+ggplot(features_persist, aes(x=Features..n., y=RSQ, group=(Day), color=factor(Day))) + geom_line() +
+  xlim(0,15) + labs(title= 'R^2 from RF with top n features\n
+                    Day of C Diff CFU dpi from Community on Day 0\nwithout mice euthanized early', x = 'Features (n)', y = 'R^2') + 
+  scale_color_discrete(name = 'Day') + theme_bw()
+
+n_important <- 20
+median_importance <- names(head(sort((apply(importance_persist,1, median)), decreasing = T), n_important))
+med_import_df <- melt(cbind(rownames(importance_persist), importance_persist)[median_importance, ])
+colnames(med_import_df) <- c('OTU', 'Day', 'Importance')
+
+ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + facet_wrap( ~ OTU, nrow = 2) +
+        geom_point() + scale_x_discrete(labels = c(1:10)) +
+        labs(title = 'OTU Importance in Prediction of C Diff CFU dpi from Community on Day 0\nwithout mice euthanized early - (Top 20 Median OTUs)',
+        x= 'Day', y = '% Increase in MSE') + theme_bw()
+
+importance_persist_day <- data.frame(matrix(nrow = sum(OTUs_01*1), 
+                                   ncol = max(without_early_euth$day, na.rm = T)),
+                            row.names = colnames(shared_X_cdiff_ra))
+rsq_persist_day <- data.frame(matrix(nrow = 1, ncol = max(without_early_euth$day, na.rm = T)),
+                         row.names = 'RSQ')
+mse_persist_day <- data.frame(matrix(nrow = 1, ncol = max(without_early_euth$day, na.rm = T)),
+                         row.names = 'MSE')
+prediction_persist_day <- data.frame(matrix(nrow = 0, ncol = 4))
+features_persist_day <- data.frame('Day'=NA, 'Features (n)'= NA, 'RSQ'=NA, 'MSE'=NA)
+
+for (i in 1:4){
+  predict_all_df <- without_by_day[without_by_day$day == i ,]
+  predict_df <- predict_all_df[!is.na(predict_all_df$log_cfu), 
+                           !colnames(predict_all_df) %in% c('Mouse_ID', 'Row.names', 'day')]
+  RF <- randomForest(log_cfu ~ ., data = predict_df,
+                   importance = TRUE, ntree = n_trees)
+  rsq_persist_day[i] <- RF$rsq[length(RF$rsq)]
+  mse_persist_day[i] <- RF$mse[length(RF$mse)]
+
+  # sort the features by their effect on % increase in the standard error
+  importance_persist_day[ , i] <- importance(RF)[,1]
+  importance_sorted <- sort(importance(RF)[,1], decreasing = T)
+  colnames(importance_persist_day)[i] <- paste('day', i, 'importance', sep = '_')
+# fit the model back on the datar and tack it to the end of the counts file
+  random_forest_output <- cbind(predict_all_df[!is.na(predict_all_df$log_cfu), c('day','log_cfu')],
+                              RF$predicted)
+  colnames(random_forest_output) <- c('day', 'observed', 'predicted')
+  prediction_persist_day <- rbind(prediction_persist_day, random_forest_output) 
+  for (k in 1:50){
+    n_features <- k
+    rf_partial <- randomForest(log_cfu ~ ., 
+                               data=predict_df[ , c( 'log_cfu', names(importance_sorted)[1:n_features])], 
+                               importance=TRUE, ntree=n_trees)
+    features_persist_day <- rbind(features_persist_day, c(i, k, tail(rf_partial$rsq, 1), tail(rf_partial$mse, 1)))
+  }
+}
+features_persist_day <- features_persist_day[-1,]
+
+ggplot(data = prediction_persist_day, aes(x=observed, y=predicted)) + geom_point() + 
+        facet_grid( ~ day) + 
+        labs(title = 'Random Forest Prediction of C Diff CFU dpi from Community on same day\nwithout mice euthanized early',
+        x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)')
+
+rsq_persist_day <- t(rsq_persist_day)
+rsq_persist_day <- as.data.frame(cbind(rsq_persist_day, day = c(1:4)))
+ggplot(data = rsq_persist_day, aes(x = factor(day), y = RSQ)) + geom_point() + theme_bw() +
+        labs(title = 'Variation Explained by Day of C Diff CFU dpi from Community on same day\nwithout mice euthanized early',
+        x= 'Day', y = 'R^2')
+
+mse_persist_day <- t(mse_persist_day)
+mse_persist_day <- as.data.frame(cbind(mse_persist_day, day = c(1:4)))
+ggplot(data = mse_persist_day, aes(x = factor(day), y = MSE)) + geom_point() + theme_bw() +
+        labs(title = 'MSE by Day of C Diff CFU dpi from Community on same day\nwithout mice euthanized early',
+        x= 'Day', y = 'MSE')
+
+ggplot(features_persist_day, aes(x=Features..n., y=RSQ, group=(Day), color=factor(Day))) + geom_line() +
+  xlim(0,15) + labs(title= 'R^2 from RF with top n features\n
+                    Day of C Diff CFU dpi from Community on same day\nwithout mice euthanized early', x = 'Features (n)', y = 'R^2') + 
+  scale_color_discrete(name = 'Day') + theme_bw()
+
+n_important <- 20
+median_importance <- names(head(sort((apply(importance_persist_day,1, median)), decreasing = T), n_important))
+med_import_df <- melt(cbind(rownames(importance_persist_day), importance_persist_day)[median_importance, ])
+colnames(med_import_df) <- c('OTU', 'Day', 'Importance')
+
+ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + facet_wrap( ~ OTU, nrow = 2) +
+        geom_point() + scale_x_discrete(labels = c(1:10)) +
+        labs(title = 'OTU Importance in Prediction of C Diff CFU dpi from Community on same day\nwithout mice euthanized early - (Top 20 Median OTUs)',
+        x= 'Day', y = '% Increase in MSE') + theme_bw()
+
+```
+
+Eliminating mice euthanized early from the RF model gives similar R^2 and MSE, bu there is a slight advantage to community OTU features of Day 0 to predict Day1 CFU.
+As well as the R^2 value increases with increasing features, whereas when all mice are used the day 1 R^2 only decreases with increasing features.
+Of note, accompaied by this is an increase in the % MSE attributed to OTU15 (Akkermansia), which has seemed to stand out in all other days/analysis.Interestingly, akkermansia does not appear to have the same relationship when compared to the same day cfu and community.
+This could suggest akkermansia is promoting the intial colonization of c difficile.
+At first glance, OTU135 (Coriobacteriaceae) seems to stand out for predicting cfu of the same day as well as from day 0.
 
@@
 ***
-
-##Does Day Community structure predict the each day C. difficile CFU?
-
-RF all mice/days together
-RF cfu vs community by day

### 2016-08-15 code/inocula_nmds.R
added donor nmds file

@@ -1,0 +1,52 @@
+###################
+#
+# Inocula_nmds.R
+#
+#Create shared file of inocula only and plot NMDS and color points by outcome
+#
+#    
+#
+###################
+
+#read in files
+meta_file <- read.table('data/process/human_CdGF_metadata.txt', sep='\t',header = T, row.names = 2)
+shared <- read.table('data/mothur/gf_cdiff.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.subsample.shared', sep='\t',header=T)
+
+#subset shared to be inoculum only
+inocula_df <- shared_file[meta_file$cage_id=='inoculum',]
+inocula_shared <- shared[shared$Group %in% rownames(inocula_df),]
+
+#take out samples we cant use
+inocula_shared <- inocula_shared[-1,]
+inocula_shared <- inocula_shared[-4,]
+inocula_shared <- inocula_shared[-6,]
+inocula_shared <- inocula_shared[-17,]
+
+#write file out so we can use it in mothur
+write.table(inocula_shared, file='data/process/inocula.shared', quote=F,sep='\t',row.names=F)
+
+#run dist.shared, nmds in mothur to get axes file
+
+nmds <- read.table(file='data/process/inocula.thetayc.0.03.lt.ave.nmds.axes', header = T)
+
+#need to color points by type 
+inocula_design <- read.table(file='data/raw/donor.design', header = F)
+inocula_nmds <- merge(nmds, inocula_design, by.x='group', by.y='V1')
+
+severe <- inocula_nmds[grep('Severe', inocula_nmds$V2), c(2,3)]
+asymptomatic <- inocula_nmds[grep('Asymptomatic', inocula_nmds$V2), c(2,3)]
+
+#make nmds inocula plot
+plot(inocula_nmds$axis1, inocula_nmds$axis2, main="Similarity of donor inocula")
+points(severe, pch=16, col = "red")
+points(asymptomatic, pch=16, col = "black")
+legend <- c("severe", "asymptomatic")
+legend(x="topright", legend, col = c("red", "black"), pch=16)
+
+
+
+
+
+
+
+

### 2016-08-15 code/HumanCdGF_microbiota_predict.Rmd
add mse by number of features

@@
   if (dep %in% installed.packages()[,"Package"] == FALSE){
-    install.packages(as.character(dep), quiet=TRUE);
+    install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
+                     quiet=TRUE);
   }
@@
   prediction_df <- rbind(prediction_df, random_forest_output) 
-  for (k in 1:50){
+  for (k in 1:25){
     n_features <- k
@@
 ggplot(rsq_features, aes(x=Features..n., y=RSQ, group=(Day), color=factor(Day))) + geom_line() +
-  xlim(0,15) + labs(title= 'R^2 from RF with top n features\n
+  labs(title= 'R^2 from RF with top n features\n
                     Day of C Diff CFU dpi from Day 0 Community', x = 'Features (n)', y = 'R^2') + 
@@
 
-for (k in 1:50){
+for (k in 1:25){
     n_features <- k
@@
 ggplot(n_feat_day_euth, aes(x=Features..n., y=RSQ, )) + geom_line() +
-  xlim(0,15) + labs(title= 'R^2 from RF with top n features\n
-                    Day of C Diff CFU t euthanization from Day 0 Community', x = 'Features (n)', y = 'R^2') + 
-  theme_bw()
+  labs(title= 'R^2 from RF with top n features\n
+               Day euthanized C Diff CFU from Day 0 Community', 
+       x = 'Features (n)', y = 'R^2') + theme_bw()
 
@@
         geom_point() + scale_x_discrete(limits = rev(top_important_OTU$OTU)) +
-        labs(title = 'OTU Importance in Prediction of C Diff CFU dpi from Day 0 Community',
+        labs(title = 'OTU Importance in Prediction of euthanized C Diff CFU from Day 0 Community',
         x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
@@
   prediction_df_day <- rbind(prediction_df_day, random_forest_output) 
-  for (k in 1:50){
+  for (k in 1:25){
     n_features <- k
@@
 ggplot(rsq_features_day, aes(x=Features..n., y=RSQ, group=(Day), color=factor(Day))) + geom_line() +
-  xlim(0,15) + labs(title= 'R^2 from RF with top n features\n
+  labs(title= 'R^2 from RF with top n features\n
                     Day of C Diff CFU dpi from Community on same day', x = 'Features (n)', y = 'R^2') + 
@@
   prediction_persist <- rbind(prediction_persist, random_forest_output) 
-  for (k in 1:50){
+  for (k in 1:25){
     n_features <- k
@@
         facet_grid( ~ day) + 
-        labs(title = 'Random Forest Prediction of C Diff CFU dpi from Community on Day 0\nwithout mice euthanized early',
-        x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)')
+        labs(title = 'Random Forest Prediction of C Diff CFU dpi from Community on Day 0
+                      \nwithout mice euthanized early', x= 'Observed CFU (log10)', 
+             y = 'Predicted CFU (log10)')
 
 rsq_persist <- t(rsq_persist)
-rsq_persist <- as.data.frame(cbind(rsq_persist, day = c(1:10)))
+rsq_persist <- as.data.frame(cbind(rsq_persist, day = c(1:4)))
 ggplot(data = rsq_persist, aes(x = factor(day), y = RSQ)) + geom_point() + theme_bw() +
-        labs(title = 'Variation Explained by Day of C Diff CFU dpi from Community on Day 0\nwithout mice euthanized early',
-        x= 'Day', y = 'R^2')
+        labs(title = 'Variation Explained by Day of C Diff CFU dpi from Community on Day 0
+                      \nwithout mice euthanized early', x= 'Day', y = 'R^2')
 
 mse_persist <- t(mse_persist)
-mse_persist <- as.data.frame(cbind(mse_persist, day = c(1:10)))
+mse_persist <- as.data.frame(cbind(mse_persist, day = c(1:4)))
 ggplot(data = mse_persist, aes(x = factor(day), y = MSE)) + geom_point() + theme_bw() +
-        labs(title = 'MSE by Day of C Diff CFU dpi from Community on Day 0\nwithout mice euthanized early',
-        x= 'Day', y = 'MSE')
+        labs(title = 'MSE by Day of C Diff CFU dpi from Community on Day 0\n
+             without mice euthanized early', x = 'Day', y = 'MSE')
 
-ggplot(features_persist, aes(x=Features..n., y=RSQ, group=(Day), color=factor(Day))) + geom_line() +
-  xlim(0,15) + labs(title= 'R^2 from RF with top n features\n
-                    Day of C Diff CFU dpi from Community on Day 0\nwithout mice euthanized early', x = 'Features (n)', y = 'R^2') + 
+ggplot(features_persist, aes(x=Features..n., y=RSQ, group=(Day), color=factor(Day))) + 
+  geom_line() + labs(title= 'R^2 from RF with top n features\n
+                              Day of C Diff CFU dpi from Community on Day 0\n
+                     without mice euthanized early', x = 'Features (n)', y = 'R^2') + 
   scale_color_discrete(name = 'Day') + theme_bw()
@@
 
-ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + facet_wrap( ~ OTU, nrow = 2) +
-        geom_point() + scale_x_discrete(labels = c(1:10)) +
-        labs(title = 'OTU Importance in Prediction of C Diff CFU dpi from Community on Day 0\nwithout mice euthanized early - (Top 20 Median OTUs)',
+ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + 
+  facet_wrap( ~ OTU, nrow = 2) + geom_point() + scale_x_discrete(labels = c(1:10)) +
+        labs(title = 'OTU Importance in Prediction of C Diff CFU dpi from Community
+             on Day 0\nwithout mice euthanized early - (Top 20 Median OTUs)',
         x= 'Day', y = '% Increase in MSE') + theme_bw()
@@
   prediction_persist_day <- rbind(prediction_persist_day, random_forest_output) 
-  for (k in 1:50){
+  for (k in 1:25){
     n_features <- k
@@
 ggplot(features_persist_day, aes(x=Features..n., y=RSQ, group=(Day), color=factor(Day))) + geom_line() +
-  xlim(0,15) + labs(title= 'R^2 from RF with top n features\n
+  labs(title= 'R^2 from RF with top n features\n
                Day of C Diff CFU dpi from Community on same day\nwithout mice euthanized early', x = 'Features (n)', y = 'R^2') + 

### 2016-08-16 code/HumanCdGF_microbiota_predict.Rmd
Add RF prediction of mobundity

@@
 ```{r, message=FALSE}
-pack_used <- c('randomForest','miscTools','ggplot2','reshape2', 'pROC')
+pack_used <- c('randomForest','miscTools','ggplot2','reshape2', 'pROC', 'knitr')
 for (dep in pack_used){
@@
 ```{r}
-setwd("~/Documents/Github/Schubert_humanCdGF_XXXX_2016")
+setwd("~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code")
 
 # read in files
-meta_file   <- read.table('data/process/human_CdGF_metadata.txt', sep = '\t',
+meta_file   <- read.table('../data/process/human_CdGF_metadata.txt', sep = '\t',
                         header = T, row.names = 'group')
-shared_file <- read.table('data/process/human_CdGF.an.unique_list.0.03.subsample.shared',
+shared_file <- read.table('../data/process/human_CdGF.an.unique_list.0.03.subsample.shared',
                           sep='\t',header = T, row.names='Group')
-tax_file <- read.table('data/mothur/gf_cdiff.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy', 
+tax_file <- read.table('../data/mothur/gf_cdiff.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy',
                        sep='\t',header = T, row.names = 'OTU')
-#source('code/tax_level.R')
+#source('tax_level.R')
 
@@
 meta_file <- meta_file[! meta_file$sample_type == 'cecum', ]
+# remove unused cage P2 and non 431 cdiff
+meta_file <- meta_file[! meta_file$cage_id %in% c('P2', 'CON', 'CON2', 'CON3'), ]
 
@@
 
-ggplot(n_feat_day_euth, aes(x=Features..n., y=RSQ, )) + geom_line() +
+ggplot(n_feat_day_euth, aes(x=Features..n., y=RSQ)) + geom_line() +
   labs(title= 'R^2 from RF with top n features\n
@@
 ```{r}
-##Fix to be categorical, and not regression
 Predict_early_euth_df <- data.frame(Euth_Early = ifelse(meta_file[meta_file$day %in% 0,
@@
 Predict_early_euth_df <- merge(Predict_early_euth_df, rel_abund_stool, by='row.names')
+mouse_list <- Predict_early_euth_df$Row.names
 Predict_early_euth_df <- Predict_early_euth_df[,!(colnames(Predict_early_euth_df) %in% "Row.names")]
@@
                               importance = TRUE, ntree = n_trees)
-# write(paste("binary", ncol(rel_abund_stool), rf_early_euth$err.rate[n_trees], "\t"), file="data/process/random_forest.data", append=T)
-#varImpPlot(rf_early_euth)
-#prediction_euth <- as.numeric(predict(rf_early_euth, Predict_early_euth_df[,-1]))
-#ROC_euth <- roc(as.numeric(Predict_early_euth_df$Euth_Early), prediction_euth)
-#plot(ROC_euth)
+
+unpredicted <- mouse_list[Predict_early_euth_df$Euth_Early != rf_early_euth$predicted]
+rf_euth_import <- data.frame(importance = head(
+  sort(rf_early_euth$importance[ , 'MeanDecreaseAccuracy'], decreasing = T), 30))
+ROC_euth <- roc(as.numeric(Predict_early_euth_df$Euth_Early), as.numeric(rf_early_euth$predicted))
+plot(ROC_euth)
+
+ggplot(data = rf_euth_import, aes(x = factor(rownames(rf_euth_import)), y = importance)) + 
+        geom_point() + scale_x_discrete(limits = rev(rownames(rf_euth_import))) +
+        labs(title = 'OTU Importance in Prediction of Day Euthanized from Day 0 Community',
+        x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
 ```
 
+Confusion Matrix  
+`r kable(rf_early_euth$confusion)`
+
+OOB error rate = `r 100*tail(rf_early_euth$err.rate[ , 1], 1)`
+
+Mice which were euthanized early, but the model predicts the will were not (Mouse Tag - Cage):
+`r unpredicted`
+
 ***

### 2016-08-17 code/HumanCdGF_microbiota_predict.Rmd
group plots together

@@
 ```{r, message=FALSE}
-pack_used <- c('randomForest','miscTools','ggplot2','reshape2', 'pROC', 'knitr')
+pack_used <- c('randomForest','miscTools','ggplot2','reshape2', 'pROC', 'knitr', 'cowplot')
 for (dep in pack_used){
@@
 
-```{r}
+```{r fig.height= 10, fig.width=10}
+
 setwd("~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code")
@@
 
-ggplot(data = prediction_df, aes(x=observed, y=predicted)) + geom_point() + 
+day_0_rf_plot <- ggplot(data = prediction_df, aes(x=observed, y=predicted)) + geom_point() + 
         facet_wrap( ~ day, nrow = 2) + 
@@
 rsq_df <- as.data.frame(cbind(rsq_df, day = c(1:10)))
-ggplot(data = rsq_df, aes(x = factor(day), y = RSQ)) + geom_point() + theme_bw() +
-        labs(title = 'Variation Explained by Day of C Diff CFU dpi from Day 0 Community',
-        x= 'Day', y = 'R^2')
+day_0_rsq_plot <- ggplot(data = rsq_df, aes(x = factor(day), y = RSQ)) + geom_point() + theme_bw() +
+        labs(x= 'Day', y = 'R^2') #title = 'Variation Explained by Day of C Diff CFU dpi from Day 0 Community',
 
@@
 mse_df <- as.data.frame(cbind(mse_df, day = c(1:10)))
-ggplot(data = mse_df, aes(x = factor(day), y = MSE)) + geom_point() + theme_bw() +
-        labs(title = 'MSE by Day of C Diff CFU dpi from Day 0 Community',
-        x= 'Day', y = 'MSE')
-
-ggplot(rsq_features, aes(x=Features..n., y=RSQ, group=(Day), color=factor(Day))) + geom_line() +
-  labs(title= 'R^2 from RF with top n features\n
-                    Day of C Diff CFU dpi from Day 0 Community', x = 'Features (n)', y = 'R^2') + 
-  scale_color_discrete(name = 'Day') + theme_bw()
+day_0_mse_plot <- ggplot(data = mse_df, aes(x = factor(day), y = MSE)) + geom_point() + theme_bw() +
+        labs(x= 'Day', y = 'MSE') # title = 'MSE by Day of C Diff CFU dpi from Day 0 Community',
+day_0_r2feat_plot <- ggplot(rsq_features, aes(x=Features..n., y=RSQ, group=(Day), color=factor(Day))) + geom_line() +
+  labs(x = 'Features (n)', y = 'R^2') + scale_color_discrete(guide=FALSE) + theme_bw()
+#  title= 'R^2 from RF with top n features\n
+#                    Day of C Diff CFU dpi from Day 0 Community', scale_color_discrete(name = 'Day')
 
@@
 colnames(med_import_df) <- c('OTU', 'Day', 'Importance')
-ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + facet_wrap( ~ OTU, nrow = 2) +
+day_0_feat_plot <- ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + facet_wrap( ~ OTU, nrow = 2) +
         geom_point() + scale_x_discrete(labels = c(1:10)) +
@@
 
+ggdraw()+
+  draw_plot(day_0_rf_plot, 0, 0.5, 0.75, 0.5) + 
+  draw_plot(day_0_rsq_plot, 0.875, 0.75, 0.125, 0.25) +
+  draw_plot(day_0_mse_plot, 0.75, 0.75, 0.125, 0.25) + 
+  draw_plot(day_0_r2feat_plot, 0.75, 0.5, 0.25, 0.25) + 
+  draw_plot(day_0_feat_plot, 0, 0, 1, 0.5)
 ```
@@
 
-```{r}
+```{r fig.height= 5, fig.width=10}
 
@@
 
-ggplot(data = rf_predict_cfu_output, aes(x=observed, y=predicted)) + geom_point() + 
+day_euth_rf_plot <- ggplot(data = rf_predict_cfu_output, aes(x=observed, y=predicted)) + geom_point() + 
         labs(x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)') + theme_bw() + 
-        ggtitle("Random Forest Prediction of C Diff CFU on Day Euthanized from Day 0 Community") +
+        ggtitle("Random Forest Prediction of C Diff CFU \n on Day Euthanized from Day 0 Community") +
         annotate('text', x=0.15*(max(rf_predict_cfu_output$observed)), y=0.95*(max(rf_predict_cfu_output$predicted)), 
@@
 
-ggplot(n_feat_day_euth, aes(x=Features..n., y=RSQ)) + geom_line() +
-  labs(title= 'R^2 from RF with top n features\n
-               Day euthanized C Diff CFU from Day 0 Community', 
-       x = 'Features (n)', y = 'R^2') + theme_bw()
+day_euth_r2feat_plot <- ggplot(n_feat_day_euth, aes(x=Features..n., y=RSQ)) + geom_line() +
+  labs(x = 'Features (n)', y = 'R^2') + theme_bw()
+#     title= 'R^2 from RF with top n features\n Day euthanized C Diff CFU from Day 0 Community',
 
-ggplot(data = top_important_OTU, aes(x = factor(OTU), y = Importance)) + 
+day_euth_feat_plot <- ggplot(data = top_important_OTU, aes(x = factor(OTU), y = Importance)) + 
         geom_point() + scale_x_discrete(limits = rev(top_important_OTU$OTU)) +
-        labs(title = 'OTU Importance in Prediction of euthanized C Diff CFU from Day 0 Community',
-        x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
+        labs(x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
+#        title = 'OTU Importance in Prediction of euthanized C Diff CFU from Day 0 Community',
+
+ggdraw()+
+  draw_plot(day_euth_rf_plot, 0, 0.5, 0.5, 0.5) + 
+  draw_plot(day_euth_r2feat_plot, 0, 0, 0.5, 0.5) +
+  draw_plot(day_euth_feat_plot, 0.5, 0, 0.5, 1)
 
@@
 
-```{r}
+```{r fig.height= 10, fig.width=10}
 
@@
 
-ggplot(data = prediction_df_day, aes(x=observed, y=predicted)) + geom_point() + 
+day_rf_plot <- ggplot(data = prediction_df_day, aes(x=observed, y=predicted)) + geom_point() + 
         facet_wrap( ~ day, nrow = 2) + 
@@
 rsq_df_day <- as.data.frame(cbind(rsq_df_day, day = c(1:10)))
-ggplot(data = rsq_df_day, aes(x = factor(day), y = RSQ)) + geom_point() + theme_bw() +
-        labs(title = 'Variation Explained by Day of C Diff CFU dpi from Community on same day',
-        x= 'Day', y = 'R^2')
+day_rsq_plot <- ggplot(data = rsq_df_day, aes(x = factor(day), y = RSQ)) + geom_point() + theme_bw() +
+        labs(x= 'Day', y = 'R^2') # title = 'Variation Explained by Day of C Diff CFU dpi from Community on same day',
 
@@
 mse_df_day <- as.data.frame(cbind(mse_df_day, day = c(1:10)))
-ggplot(data = mse_df_day, aes(x = factor(day), y = MSE)) + geom_point() + theme_bw() +
-        labs(title = 'MSE by Day of C Diff CFU dpi from Community on same day',
-        x= 'Day', y = 'MSE')
+day_mse_plot <- ggplot(data = mse_df_day, aes(x = factor(day), y = MSE)) + geom_point() + theme_bw() +
+        labs(x= 'Day', y = 'MSE') # title = 'MSE by Day of C Diff CFU dpi from Community on same day',
 
-ggplot(rsq_features_day, aes(x=Features..n., y=RSQ, group=(Day), color=factor(Day))) + geom_line() +
-  labs(title= 'R^2 from RF with top n features\n
-                    Day of C Diff CFU dpi from Community on same day', x = 'Features (n)', y = 'R^2') + 
-  scale_color_discrete(name = 'Day') + theme_bw()
+day_r2feat_plot <- ggplot(rsq_features_day, aes(x=Features..n., y=RSQ, group=(Day), color=factor(Day))) + geom_line() +
+  labs(x = 'Features (n)', y = 'R^2') + scale_color_discrete(guide = FALSE) + theme_bw()
+# title= 'R^2 from RF with top n features\n Day of C Diff CFU dpi from Community on same day',
 
@@
 
-ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + facet_wrap( ~ OTU, nrow = 2) +
+day_feat_plot <- ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + facet_wrap( ~ OTU, nrow = 2) +
         geom_point() + scale_x_discrete(labels = c(1:10)) +
@@
 
+ggdraw()+
+  draw_plot(day_rf_plot, 0, 0.5, 0.75, 0.5) + 
+  draw_plot(day_rsq_plot, 0.875, 0.75, 0.125, 0.25) +
+  draw_plot(day_mse_plot, 0.75, 0.75, 0.125, 0.25) + 
+  draw_plot(day_r2feat_plot, 0.75, 0.5, 0.25, 0.25) + 
+  draw_plot(day_feat_plot, 0, 0, 1, 0.5)
 ```
@@
 
-ggplot(data = prediction_persist, aes(x=observed, y=predicted)) + geom_point() + 
+day_0_persist_rf_plot <- ggplot(data = prediction_persist, aes(x=observed, y=predicted)) + geom_point() + 
         facet_grid( ~ day) + 
@@
 rsq_persist <- as.data.frame(cbind(rsq_persist, day = c(1:4)))
-ggplot(data = rsq_persist, aes(x = factor(day), y = RSQ)) + geom_point() + theme_bw() +
-        labs(title = 'Variation Explained by Day of C Diff CFU dpi from Community on Day 0
-                      \nwithout mice euthanized early', x= 'Day', y = 'R^2')
+day_0_persist_rsq_plot <- ggplot(data = rsq_persist, aes(x = factor(day), y = RSQ)) + geom_point() + theme_bw() +
+        labs(x= 'Day', y = 'R^2') 
+#title = 'Variation Explained by Day of C Diff CFU dpi from Community on Day 0\nwithout mice euthanized early',
 
@@
 mse_persist <- as.data.frame(cbind(mse_persist, day = c(1:4)))
-ggplot(data = mse_persist, aes(x = factor(day), y = MSE)) + geom_point() + theme_bw() +
-        labs(title = 'MSE by Day of C Diff CFU dpi from Community on Day 0\n
-             without mice euthanized early', x = 'Day', y = 'MSE')
+day_0_persist_mse_plot <- ggplot(data = mse_persist, aes(x = factor(day), y = MSE)) + geom_point() + theme_bw() +
+        labs(x = 'Day', y = 'MSE')
+#title = 'MSE by Day of C Diff CFU dpi from Community on Day 0\nwithout mice euthanized early',
 
-ggplot(features_persist, aes(x=Features..n., y=RSQ, group=(Day), color=factor(Day))) + 
-  geom_line() + labs(title= 'R^2 from RF with top n features\n
-                              Day of C Diff CFU dpi from Community on Day 0\n
-                     without mice euthanized early', x = 'Features (n)', y = 'R^2') + 
-  scale_color_discrete(name = 'Day') + theme_bw()
+day_0_persist_r2feat_plot <- ggplot(features_persist, aes(x=Features..n., y=RSQ, group=(Day), color=factor(Day))) + 
+  geom_line() + labs(x = 'Features (n)', y = 'R^2') + scale_color_discrete(guide = FALSE) + theme_bw()
+#title= 'R^2 from RF with top n features\nDay of C Diff CFU dpi from Community on Day 0\nwithout mice euthanized early',
 
@@
 
-ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + 
+day_0_persist_feat_plot <- ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + 
   facet_wrap( ~ OTU, nrow = 2) + geom_point() + scale_x_discrete(labels = c(1:10)) +
@@
         x= 'Day', y = '% Increase in MSE') + theme_bw()
+
+ggdraw()+
+  draw_plot(day_0_persist_rf_plot, 0, 0.5, 0.75, 0.5) + 
+  draw_plot(day_0_persist_rsq_plot, 0.875, 0.75, 0.125, 0.25) +
+  draw_plot(day_0_persist_mse_plot, 0.75, 0.75, 0.125, 0.25) + 
+  draw_plot(day_0_persist_r2feat_plot, 0.75, 0.5, 0.25, 0.25) + 
+  draw_plot(day_0_persist_feat_plot, 0, 0, 1, 0.5)
 
@@
 
-ggplot(data = prediction_persist_day, aes(x=observed, y=predicted)) + geom_point() + 
+day_persist_rf_plot <- ggplot(data = prediction_persist_day, aes(x=observed, y=predicted)) + geom_point() + 
         facet_grid( ~ day) + 
@@
 rsq_persist_day <- as.data.frame(cbind(rsq_persist_day, day = c(1:4)))
-ggplot(data = rsq_persist_day, aes(x = factor(day), y = RSQ)) + geom_point() + theme_bw() +
-        labs(title = 'Variation Explained by Day of C Diff CFU dpi from Community on same day\nwithout mice euthanized early',
-        x= 'Day', y = 'R^2')
+day_persist_rsq_plot <- ggplot(data = rsq_persist_day, aes(x = factor(day), y = RSQ)) + geom_point() + theme_bw() +
+        labs(x= 'Day', y = 'R^2')
+#title = 'Variation Explained by Day of C Diff CFU dpi from Community on same day\nwithout mice euthanized early',
 
@@
 mse_persist_day <- as.data.frame(cbind(mse_persist_day, day = c(1:4)))
-ggplot(data = mse_persist_day, aes(x = factor(day), y = MSE)) + geom_point() + theme_bw() +
-        labs(title = 'MSE by Day of C Diff CFU dpi from Community on same day\nwithout mice euthanized early',
-        x= 'Day', y = 'MSE')
+day_persist_mse_plot <- ggplot(data = mse_persist_day, aes(x = factor(day), y = MSE)) + geom_point() + theme_bw() +
+        labs(x= 'Day', y = 'MSE')
+#title = 'MSE by Day of C Diff CFU dpi from Community on same day\nwithout mice euthanized early',
 
-ggplot(features_persist_day, aes(x=Features..n., y=RSQ, group=(Day), color=factor(Day))) + geom_line() +
-  labs(title= 'R^2 from RF with top n features\n
-               Day of C Diff CFU dpi from Community on same day\nwithout mice euthanized early', x = 'Features (n)', y = 'R^2') + 
-  scale_color_discrete(name = 'Day') + theme_bw()
+day_persist_r2feat_plot <- ggplot(features_persist_day, aes(x=Features..n., y=RSQ, group=(Day), color=factor(Day))) + geom_line() +
+  labs(x = 'Features (n)', y = 'R^2') + scale_color_discrete(guide  = FALSE) + theme_bw()
+# title= 'R^2 from RF with top n features\n Day of C Diff CFU dpi from Community on same day\nwithout mice euthanized early',
 
@@
 
-ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + facet_wrap( ~ OTU, nrow = 2) +
+day_persist_feat_plot <- ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + facet_wrap( ~ OTU, nrow = 2) +
         geom_point() + scale_x_discrete(labels = c(1:10)) +
@@
         x= 'Day', y = '% Increase in MSE') + theme_bw()
+
+ggdraw()+
+  draw_plot(day_persist_rf_plot, 0, 0.5, 0.75, 0.5) + 
+  draw_plot(day_persist_rsq_plot, 0.875, 0.75, 0.125, 0.25) +
+  draw_plot(day_persist_mse_plot, 0.75, 0.75, 0.125, 0.25) + 
+  draw_plot(day_persist_r2feat_plot, 0.75, 0.5, 0.25, 0.25) + 
+  draw_plot(day_persist_feat_plot, 0, 0, 1, 0.5)
 
@@
 ##Can we predict moribundity with the community structure at Day 0?
-```{r}
+```{r fig.width=10, fig.height=5}
 Predict_early_euth_df <- data.frame(Euth_Early = ifelse(meta_file[meta_file$day %in% 0,
@@
 ROC_euth <- roc(as.numeric(Predict_early_euth_df$Euth_Early), as.numeric(rf_early_euth$predicted))
-plot(ROC_euth)
+roc_plot <- data.frame(Specificity = 1 - ROC_euth$specificities, Sensitivity = ROC_euth$sensitivities)
 
-ggplot(data = rf_euth_import, aes(x = factor(rownames(rf_euth_import)), y = importance)) + 
+euth_rf_plot <- ggplot(data = roc_plot, aes(x = Specificity, y = Sensitivity)) + geom_path() + 
+  geom_abline(intercept = 0, slope = 1, linetype = 'dashed', alpha = 0.5)
+euth_import_plot <- ggplot(data = rf_euth_import, aes(x = factor(rownames(rf_euth_import)), y = importance)) + 
         geom_point() + scale_x_discrete(limits = rev(rownames(rf_euth_import))) +
-        labs(title = 'OTU Importance in Prediction of Day Euthanized from Day 0 Community',
-        x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
+        labs(x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
+# title = 'OTU Importance in Prediction of Day Euthanized from Day 0 Community',
+
+ggdraw()+
+  draw_plot(euth_rf_plot, 0, 0, 0.5, 1) + 
+  draw_plot(euth_import_plot, 0.5, 0, 0.5, 1)
 ```
@@
 ***
+
+## Analysis By Day
+
+```{r fig.width=10, fig.height=5}
+layout(matrix(rep(c(1,1,1,1,2,2,2,2),4), ncol=8, nrow = 4, byrow = T))
+
+```

### 2016-08-17 code/HumanCdGF_microbiota_predict.Rmd
add rf of all cfu by day together

@@
 
+## Can Day 0 Community members predict C. difficile CFU?
 ```{r fig.height= 10, fig.width=10}
@@
 
-## Does Day Community structure predict C. difficile CFU of that day?
+## Does Community structure predict C. difficile CFU of same day?
 
-RF all mice/days together
-RF cfu vs community by day
+### RF cfu vs community by day
 
@@
   draw_plot(day_feat_plot, 0, 0, 1, 0.5)
+```
+
+### RF all mice/days together
+
+```{r }
+predict_cfu_all_df <- CFU_community_by_day[!is.na(CFU_community_by_day$log_cfu),
+                                           ! colnames(CFU_community_by_day) %in% c('Row.names','day')]
+RF_cfu_all_rf <- randomForest(log_cfu ~ ., data = predict_cfu_all_df,
+                   importance = TRUE, ntree = n_trees)
+RF_cfu_all_df <- data.frame(cbind(CFU_community_by_day[!is.na(CFU_community_by_day$log_cfu), 
+                                                       c('day','log_cfu')], 
+                                  predicted = RF_cfu_all_rf$predicted))
+
+mse_all <- tail(RF_cfu_all_rf$mse, 1)
+rsq_all <- tail(RF_cfu_all_rf$rsq, 1)
+
+all_rf_plot <- ggplot(data = RF_cfu_all_df, aes(x = log_cfu, y = predicted)) + geom_point() +
+        labs(x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)') + theme_bw() + 
+        ggtitle("Random Forest Prediction of C Diff CFU") +
+        annotate('text', x=0.15*(max(RF_cfu_all_df$log_cfu)), y=0.95*(max(RF_cfu_all_df$predicted)), 
+                 label=paste('R^2 = ', tail(RF_cfu_all_rf$rsq, 1),'\nMSE = ', tail(RF_cfu_all_rf$mse, 1), sep = ''))
+
+importance_all <- sort(importance(RF_cfu_all_rf)[,1], decreasing = T)
+
+rsq_features_all <- data.frame('Features'= NA, 'RSQ'=NA, 'MSE'=NA)
+for (k in 1:25){
+    rf_partial <- randomForest(log_cfu ~ ., 
+                               data=predict_cfu_all_df[ , c( 'log_cfu', names(importance_all)[1:k])], 
+                               importance=TRUE, ntree=n_trees)
+    rsq_features_all <- rbind(rsq_features_all, c(k, tail(rf_partial$rsq, 1), tail(rf_partial$mse, 1)))
+}
+rsq_features_all <- rsq_features_all[-1,]
+all_r2feat_plot <- ggplot(rsq_features_all, aes(x=Features, y=RSQ)) + geom_line() +
+  labs(x = 'Features (n)', y = 'R^2') + scale_color_discrete(guide = FALSE) + theme_bw()
+
+important_OTU_all <- data.frame(head(importance_all, n_important))
+colnames(important_OTU_all) <- 'Importance'
+important_OTU_all$OTU <- rownames(important_OTU_all)
+
+all_feat_plot <- ggplot(data = important_OTU_all, aes(x = factor(OTU), y = Importance)) + 
+        geom_point() + scale_x_discrete(limits = rev(important_OTU_all$OTU)) +
+        labs(x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
+
+ggdraw()+
+  draw_plot(all_rf_plot, 0, 0.5, 0.5, 0.5) + 
+  draw_plot(all_r2feat_plot, 0, 0, 0.5, 0.5) +
+  draw_plot(all_feat_plot, 0.5, 0, 0.5, 1)
+
+```
+
+```{r fig.width=10, fig.height=5}
+predict_cfu_all_df <- CFU_community_by_day[!is.na(CFU_community_by_day$log_cfu),
+                                           ! colnames(CFU_community_by_day) %in% c('Row.names','day')]
+RF_cfu_all_rf <- randomForest(log_cfu ~ ., data = predict_cfu_all_df,
+                   importance = TRUE, ntree = n_trees)
+RF_cfu_all_df <- data.frame(cbind(CFU_community_by_day[!is.na(CFU_community_by_day$log_cfu), 
+                                                       c('day','log_cfu')], 
+                                  predicted = RF_cfu_all_rf$predicted))
+
+mse_all <- tail(RF_cfu_all_rf$mse, 1)
+rsq_all <- tail(RF_cfu_all_rf$rsq, 1)
+
+all_rf_plot <- ggplot(data = RF_cfu_all_df, aes(x = log_cfu, y = predicted)) + geom_point() +
+        labs(x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)') + theme_bw() + 
+        ggtitle("Random Forest Prediction of C Diff CFU") +
+        annotate('text', x=0.15*(max(RF_cfu_all_df$log_cfu)), y=0.95*(max(RF_cfu_all_df$predicted)), 
+                 label=paste('R^2 = ', tail(RF_cfu_all_rf$rsq, 1),'\nMSE = ', tail(RF_cfu_all_rf$mse, 1), sep = ''))
+
+importance_all <- sort(importance(RF_cfu_all_rf)[,1], decreasing = T)
+
+rsq_features_all <- data.frame('Features'= NA, 'RSQ'=NA, 'MSE'=NA)
+for (k in 1:25){
+    rf_partial <- randomForest(log_cfu ~ ., 
+                               data=predict_cfu_all_df[ , c( 'log_cfu', names(importance_all)[1:k])], 
+                               importance=TRUE, ntree=n_trees)
+    rsq_features_all <- rbind(rsq_features_all, c(k, tail(rf_partial$rsq, 1), tail(rf_partial$mse, 1)))
+}
+rsq_features_all <- rsq_features_all[-1,]
+all_r2feat_plot <- ggplot(rsq_features_all, aes(x=Features, y=RSQ)) + geom_line() +
+  labs(x = 'Features (n)', y = 'R^2') + scale_color_discrete(guide = FALSE) + theme_bw()
+
+important_OTU_all <- data.frame(head(importance_all, n_important))
+colnames(important_OTU_all) <- 'Importance'
+important_OTU_all$OTU <- rownames(important_OTU_all)
+
+all_feat_plot <- ggplot(data = important_OTU_all, aes(x = factor(OTU), y = Importance)) + 
+        geom_point() + scale_x_discrete(limits = rev(important_OTU_all$OTU)) +
+        labs(x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
+
+ggdraw()+
+  draw_plot(all_rf_plot, 0, 0.5, 0.5, 0.5) + 
+  draw_plot(all_r2feat_plot, 0, 0, 0.5, 0.5) +
+  draw_plot(all_feat_plot, 0.5, 0, 0.5, 1)
+
 ```
@@
   draw_plot(day_persist_feat_plot, 0, 0, 1, 0.5)
+```
+
+#### All CFU compared to same day community without mice euthanized early 
+```{r fig.width=10, fig.height=5}
+
+predict_cfu_persist_df <- without_by_day[!is.na(without_by_day$log_cfu),
+                                           ! colnames(without_by_day) %in% c('Row.names','day','Mouse_ID')]
+RF_cfu_persist_rf <- randomForest(log_cfu ~ ., data = predict_cfu_persist_df,
+                   importance = TRUE, ntree = n_trees)
+RF_cfu_persist_df <- data.frame(cbind(without_by_day[!is.na(without_by_day$log_cfu), 
+                                                       c('day','log_cfu')], 
+                                  predicted = RF_cfu_persist_rf$predicted))
+
+mse_all <- tail(RF_cfu_persist_rf$mse, 1)
+rsq_all <- tail(RF_cfu_persist_rf$rsq, 1)
+
+all_persist_rf_plot <- ggplot(data = RF_cfu_persist_df, aes(x = log_cfu, y = predicted)) + geom_point() +
+        labs(x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)') + theme_bw() + 
+        ggtitle("Random Forest Prediction of C Diff CFU") +
+        annotate('text', x=0.15*(max(RF_cfu_persist_df$log_cfu)), y=0.95*(max(RF_cfu_persist_df$predicted)), 
+                 label=paste('R^2 = ', tail(RF_cfu_persist_rf$rsq, 1),'\nMSE = ', tail(RF_cfu_persist_rf$mse, 1), sep = ''))
+
+importance_all_persist <- sort(importance(RF_cfu_persist_rf)[,1], decreasing = T)
+
+rsq_features_all_persist <- data.frame('Features'= NA, 'RSQ'=NA, 'MSE'=NA)
+for (k in 1:25){
+    rf_partial <- randomForest(log_cfu ~ ., 
+                               data=predict_cfu_persist_df[ , c( 'log_cfu', names(importance_all_persist)[1:k])], 
+                               importance=TRUE, ntree=n_trees)
+    rsq_features_all_persist <- rbind(rsq_features_all_persist, c(k, tail(rf_partial$rsq, 1), tail(rf_partial$mse, 1)))
+}
+rsq_features_all_persist <- rsq_features_all_persist[-1,]
+all_persist_r2feat_plot <- ggplot(rsq_features_all_persist, aes(x=Features, y=RSQ)) + geom_line() +
+  labs(x = 'Features (n)', y = 'R^2') + scale_color_discrete(guide = FALSE) + theme_bw()
+
+important_OTU_all_persist <- data.frame(head(importance_all_persist, n_important))
+colnames(important_OTU_all_persist) <- 'Importance'
+important_OTU_all_persist$OTU <- rownames(important_OTU_all_persist)
+
+all_persist_feat_plot <- ggplot(data = important_OTU_all_persist, aes(x = factor(OTU), y = Importance)) + 
+        geom_point() + scale_x_discrete(limits = rev(important_OTU_all_persist$OTU)) +
+        labs(x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
+
+ggdraw()+
+  draw_plot(all_persist_rf_plot, 0, 0.5, 0.5, 0.5) + 
+  draw_plot(all_persist_r2feat_plot, 0, 0, 0.5, 0.5) +
+  draw_plot(all_persist_feat_plot, 0.5, 0, 0.5, 1)
 

### 2016-08-17 code/HumanCdGF_microbiota_predict.Rmd
fixed errors in all_cfu plots

@@
 
-### RF all mice/days together
+####
+RF all mice/days together
 
-```{r }
-predict_cfu_all_df <- CFU_community_by_day[!is.na(CFU_community_by_day$log_cfu),
-                                           ! colnames(CFU_community_by_day) %in% c('Row.names','day')]
-RF_cfu_all_rf <- randomForest(log_cfu ~ ., data = predict_cfu_all_df,
-                   importance = TRUE, ntree = n_trees)
-RF_cfu_all_df <- data.frame(cbind(CFU_community_by_day[!is.na(CFU_community_by_day$log_cfu), 
-                                                       c('day','log_cfu')], 
-                                  predicted = RF_cfu_all_rf$predicted))
-
-mse_all <- tail(RF_cfu_all_rf$mse, 1)
-rsq_all <- tail(RF_cfu_all_rf$rsq, 1)
-
-all_rf_plot <- ggplot(data = RF_cfu_all_df, aes(x = log_cfu, y = predicted)) + geom_point() +
-        labs(x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)') + theme_bw() + 
-        ggtitle("Random Forest Prediction of C Diff CFU") +
-        annotate('text', x=0.15*(max(RF_cfu_all_df$log_cfu)), y=0.95*(max(RF_cfu_all_df$predicted)), 
-                 label=paste('R^2 = ', tail(RF_cfu_all_rf$rsq, 1),'\nMSE = ', tail(RF_cfu_all_rf$mse, 1), sep = ''))
-
-importance_all <- sort(importance(RF_cfu_all_rf)[,1], decreasing = T)
-
-rsq_features_all <- data.frame('Features'= NA, 'RSQ'=NA, 'MSE'=NA)
-for (k in 1:25){
-    rf_partial <- randomForest(log_cfu ~ ., 
-                               data=predict_cfu_all_df[ , c( 'log_cfu', names(importance_all)[1:k])], 
-                               importance=TRUE, ntree=n_trees)
-    rsq_features_all <- rbind(rsq_features_all, c(k, tail(rf_partial$rsq, 1), tail(rf_partial$mse, 1)))
-}
-rsq_features_all <- rsq_features_all[-1,]
-all_r2feat_plot <- ggplot(rsq_features_all, aes(x=Features, y=RSQ)) + geom_line() +
-  labs(x = 'Features (n)', y = 'R^2') + scale_color_discrete(guide = FALSE) + theme_bw()
-
-important_OTU_all <- data.frame(head(importance_all, n_important))
-colnames(important_OTU_all) <- 'Importance'
-important_OTU_all$OTU <- rownames(important_OTU_all)
-
-all_feat_plot <- ggplot(data = important_OTU_all, aes(x = factor(OTU), y = Importance)) + 
-        geom_point() + scale_x_discrete(limits = rev(important_OTU_all$OTU)) +
-        labs(x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
-
-ggdraw()+
-  draw_plot(all_rf_plot, 0, 0.5, 0.5, 0.5) + 
-  draw_plot(all_r2feat_plot, 0, 0, 0.5, 0.5) +
-  draw_plot(all_feat_plot, 0.5, 0, 0.5, 1)
-
-```
-
-```{r fig.width=10, fig.height=5}
+```{r fig.height= 7, fig.width=10}
 predict_cfu_all_df <- CFU_community_by_day[!is.na(CFU_community_by_day$log_cfu),

### 2016-08-22 code/test_code.Rmd
intial test code rmd

---
title: "Test_code"
author: "Nick Lesniak"
date: "August 18, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
```

```{r, message=FALSE}
pack_used <- c('randomForest','miscTools','tidyr','ggplot2','dplyr', 'pROC', 'knitr', 'cowplot')
for (dep in pack_used){
  if (dep %in% installed.packages()[,"Package"] == FALSE){
    install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
                     quiet=TRUE);
  }
  library(dep, verbose=FALSE, character.only=TRUE)
}
```

```{r}
# setup environment

# read in files
meta_file   <- read.table('../data/process/human_CdGF_metadata.txt', sep = '\t',
                        header = T)
shared_file <- read.table('../data/process/human_CdGF.an.unique_list.0.03.subsample.shared',
                          sep='\t',header = T)
tax_file <- read.table('../data/mothur/gf_cdiff.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy',
                       sep='\t',header = T, row.names = 'OTU')
# function to label otu with tax
source('tax_level.R')

unused_samples <- c("P2", "CON", "CON2", "CON3")

# remove samples not used or with other cdiff strains
meta_file <- 
  meta_file %>% 
  filter(sample_type == 'stool',
         !cage_id %in% unused_samples) %>% 
  select(Mouse_ID,Group = group, day, cdiff_cfu, Euthanized) %>% 
  mutate(log_cfu = log10(cdiff_cfu + 1)) %>% 
  filter(!is.na(log_cfu))

# create df with day 0 communities with column of mouse_id
day_0_otu_by_mouse <- 
  meta_file %>% 
  filter(day == 0 ) %>% 
  select(Group, Mouse_ID) %>% 
  left_join(shared_file)
day_0_otu_over_01 <- 
  day_0_otu_by_mouse %>% 
  select(contains('Otu')) %>% 
  apply(. , 2, function(i){mean(i)/mean(rowSums(.))}) > 0.01
day_0_otu_over_01 <-  names(day_0_otu_over_01)[day_0_otu_over_01]
day_0_otu_by_mouse_01 <- 
  day_0_otu_by_mouse %>% 
  select(Mouse_ID, one_of(day_0_otu_over_01))

seed <- 18
n_trees <- 2000
set.seed(seed)

rf_plot_list <- list()
n_features <- 10

rf_plot_list <- 
lapply(1:10, function(i){
rf_data <- 
  meta_file %>%
  filter(day == i) %>% 
  select(log_cfu, Mouse_ID) %>%
  inner_join(day_0_otu_by_mouse_01) %>% 
  select(-Mouse_ID) %>% 
  tbl_df()

rf_all <- randomForest(log_cfu ~ ., data = rf_data,
                   importance = TRUE, ntree = n_trees)
importance_sorted <- sort(importance(rf_all)[,1], decreasing = T)
rf_n_features <- randomForest(log_cfu ~ ., data=rf_data[ ,c( 'log_cfu',
                              names(importance_sorted)[1:n_features])], 
                           importance=TRUE, ntree=n_trees)

predict_plot_day <- paste('prediction_day_', i, sep = '')
importance_plot_day <- paste('importance_day_', i, sep = '')

rf_prediction <- cbind(rf_data[, 'log_cfu'], 
                       data.frame(rf_n_features$predicted))
colnames(rf_prediction) <- c('observed', 'predicted')

predict_plot_day <-  
  ggplot(data = rf_prediction, aes(x=observed, y=predicted)) + geom_point() + 
        labs(x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)') + 
        theme_bw() + ggtitle(paste("Random Forest Prediction of C Diff CFU \n on Day ",
                                   i, " from Day 0 Community", sep = '')) +
        annotate('text', x=0.15*(max(rf_prediction$observed)), 
                 y=0.95*(max(rf_prediction$predicted)), 
                 label=paste('R^2 = ', round(tail(rf_n_features$rsq, 1), 2),'\nMSE = ', 
                             round(tail(rf_n_features$mse, 1), 2), sep = ''))

top_important_OTU <- data.frame(head(importance_sorted, n_features))
colnames(top_important_OTU) <- 'Importance'
top_important_OTU$OTU <- rownames(top_important_OTU)
otu_taxa <- get_tax(1, top_important_OTU$OTU, tax_file)

importance_plot_day <- 
  ggplot(data = top_important_OTU, aes(x = factor(OTU), y = Importance)) + 
        geom_point() + scale_x_discrete(limits = rev(top_important_OTU$OTU),
                                        labels = paste(otu_taxa[,1],' (',
                                                       rownames(otu_taxa),')',
                                                       sep='')) +
        labs(x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()

rf_plot_list[[i]] <- list(predict_plot_day,importance_plot_day)

})
```

### Predict Day 1 CFU from Day 0 Community

```{r fig.height= 5, fig.width=10}
ggdraw()+
  draw_plot(rf_plot_list[[1]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_plot_list[[1]][[2]], 0.5, 0, 0.5, 1)

```

***

### Predict Day 2 CFU from Day 0 Community

```{r fig.height= 5, fig.width=10}
ggdraw()+
  draw_plot(rf_plot_list[[2]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_plot_list[[2]][[2]], 0.5, 0, 0.5, 1)

```

***

### Predict Day 3 CFU from Day 0 Community

```{r fig.height= 5, fig.width=10}
ggdraw()+
  draw_plot(rf_plot_list[[3]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_plot_list[[3]][[2]], 0.5, 0, 0.5, 1)

```

***

### Predict Day 4 CFU from Day 0 Community

```{r fig.height= 5, fig.width=10}
ggdraw()+
  draw_plot(rf_plot_list[[4]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_plot_list[[4]][[2]], 0.5, 0, 0.5, 1)

```

***

### Predict Day 5 CFU from Day 0 Community

```{r fig.height= 5, fig.width=10}
ggdraw()+
  draw_plot(rf_plot_list[[5]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_plot_list[[5]][[2]], 0.5, 0, 0.5, 1)

```

***

### Predict Day 6 CFU from Day 0 Community

```{r fig.height= 5, fig.width=10}
ggdraw()+
  draw_plot(rf_plot_list[[6]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_plot_list[[6]][[2]], 0.5, 0, 0.5, 1)

```

***

### Predict Day 7 CFU from Day 0 Community

```{r fig.height= 5, fig.width=10}
ggdraw()+
  draw_plot(rf_plot_list[[7]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_plot_list[[7]][[2]], 0.5, 0, 0.5, 1)

```

***

### Predict Day 8 CFU from Day 0 Community

```{r fig.height= 5, fig.width=10}
ggdraw()+
  draw_plot(rf_plot_list[[8]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_plot_list[[8]][[2]], 0.5, 0, 0.5, 1)

```

***

### Predict Day 9 CFU from Day 0 Community

```{r fig.height= 5, fig.width=10}
ggdraw()+
  draw_plot(rf_plot_list[[9]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_plot_list[[9]][[2]], 0.5, 0, 0.5, 1)

```

***

### Predict Day 10 CFU from Day 0 Community

```{r fig.height= 5, fig.width=10}
ggdraw()+
  draw_plot(rf_plot_list[[10]][[1]], 0, 0, 0.5, 1) + 
  draw_plot(rf_plot_list[[10]][[2]], 0.5, 0, 0.5, 1)

```

### 2016-08-22 code/make_HumanCdGF_metadata_file.R
remove DA00810 from metafile

@@ -42,7 +42,8 @@ human_GF_clinical <- human_GF_clinical[,!colnames(human_GF_clinical) %in%
 # Remove file names
 human_GF_mouse <- human_GF_mouse[,!colnames(human_GF_mouse) %in% 'file']
 # Remove AMS samples
-human_GF_mouse <- human_GF_mouse[!human_GF_mouse$human_source=='AMS',]
+human_GF_mouse <- human_GF_mouse[!human_GF_mouse$human_source== c('AMS', 'DA00810'),]
+
 
 #Combine Toxin, Exp, and metadata
 #this next line is where the reordering happens - will correct before writing to new file (KF)

### 2016-08-22 code/HumanCdGF_microbiota_predict_by_day.Rmd

--- 2016-08-22 code/test_code.Rmd
@@
 shared_file <- read.table('../data/process/human_CdGF.an.unique_list.0.03.subsample.shared',
-                          sep='\t',header = T)
+                          sep='\t',header = T, row.names = 'Group')
 tax_file <- read.table('../data/mothur/gf_cdiff.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy',
@@
 
-unused_samples <- c("P2", "CON", "CON2", "CON3")
-
 # remove samples not used or with other cdiff strains
@@
   meta_file %>% 
-  filter(sample_type == 'stool',
-         !cage_id %in% unused_samples) %>% 
-  select(Mouse_ID,Group = group, day, cdiff_cfu, Euthanized) %>% 
+  filter(sample_type == 'stool', cdiff_strain == 431) %>% 
+  select(Mouse_ID, group, day, cdiff_cfu, Euthanized) %>% 
   mutate(log_cfu = log10(cdiff_cfu + 1)) %>% 
   filter(!is.na(log_cfu))
+
+# create shared with relative abundances over 1%
+mean_otu_counts <- unique(apply(shared_file, 1, sum))
+rel_abund_shared <- 100 * shared_file / mean_otu_counts
+rel_abund_shared_01 <- rel_abund_shared[ , apply(rel_abund_shared, 2, max) > 1]
+rel_abund_shared_01$group <- rownames(rel_abund_shared_01)
 
@@
   filter(day == 0 ) %>% 
-  select(Group, Mouse_ID) %>% 
-  left_join(shared_file)
-day_0_otu_over_01 <- 
-  day_0_otu_by_mouse %>% 
-  select(contains('Otu')) %>% 
-  apply(. , 2, function(i){mean(i)/mean(rowSums(.))}) > 0.01
-day_0_otu_over_01 <-  names(day_0_otu_over_01)[day_0_otu_over_01]
-day_0_otu_by_mouse_01 <- 
-  day_0_otu_by_mouse %>% 
-  select(Mouse_ID, one_of(day_0_otu_over_01))
+  select(group, Mouse_ID) %>% 
+  left_join(rel_abund_shared_01)
+
+cfu_by_day <-
+  meta_file %>% 
+  select(group)
 
@@
 
-rf_plot_list <- 
-lapply(1:10, function(i){
+RandomForest_by_day <- function(i, community){
+if(community == day_0){
   rf_data <- 
@@
   inner_join(day_0_otu_by_mouse_01) %>% 
-  select(-Mouse_ID) %>% 
-  tbl_df()
+  select(-Mouse_ID)
+} else if (community == by_day) {
+  rf_data <- 
+  meta_file %>%
+  filter(day == i) %>% 
+  select(log_cfu, Mouse_ID) %>%
+  inner_join(day_0_otu_by_mouse_01) %>% 
+  select(-Mouse_ID)
+
+} else {
+  print('Input for community needed (day_0 or by_day')
+}
 
@@
         geom_point() + scale_x_discrete(limits = rev(top_important_OTU$OTU),
-                                        labels = paste(otu_taxa[,1],' (',
+                                        labels = rev(paste(otu_taxa[,1],' (',
                                                        rownames(otu_taxa),')',
-                                                       sep='')) +
+                                                       sep=''))) +
         labs(x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
@@
 })
+
+rf_day_0_comm_plots <- lapply(1:10, day_0)
+rf_by_day_com_plots <- lapply(1:10, by_day)
+
 ```

### 2016-08-22 code/HumanCdGF_microbiota_predict_by_day.Rmd
predict cfu by same and day 0 comm

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE)
 ```
 
-```{r, message=FALSE}
+```{r}
 pack_used <- c('randomForest','miscTools','tidyr','ggplot2','dplyr', 'pROC', 'knitr', 'cowplot')
@@
 # read in files
-meta_file   <- read.table('../data/process/human_CdGF_metadata.txt', sep = '\t',
+meta_file   <- read.table('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt', sep = '\t',
                         header = T)
-shared_file <- read.table('../data/process/human_CdGF.an.unique_list.0.03.subsample.shared',
+shared_file <- read.table('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared',
                           sep='\t',header = T, row.names = 'Group')
-tax_file <- read.table('../data/mothur/gf_cdiff.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy',
+tax_file <- read.table('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_cdiff.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy',
                        sep='\t',header = T, row.names = 'OTU')
 # function to label otu with tax
-source('tax_level.R')
+source('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R')
 
+# identify c difficile otus
+cdiff_otu <- rownames(tax_file[grep('Clostridium_XI', tax_file$Taxonomy),])
 # remove samples not used or with other cdiff strains
@@
   select(group, Mouse_ID) %>% 
-  left_join(rel_abund_shared_01)
-
-cfu_by_day <-
+  left_join(rel_abund_shared_01) %>% 
+  select(-group)
+# create df with cfu merge to shared
+cfu_by_day_comm <-
   meta_file %>% 
-  select(group)
+  filter(day %in% 1:10) %>% 
+  select(group, day, log_cfu) %>% 
+  left_join(rel_abund_shared_01) %>% 
+  select(-group, -one_of(cdiff_otu))
 
@@
 
-RandomForest_by_day <- function(i, community){
-if(community == day_0){
-  rf_data <- 
+RandomForest_by_day <- function(i, community_day_0){
+  if(!is.logical(community_day_0)){
+    stop('Input for community  - Need community_day_0 = T/F')
+  } else {
+    rf_data <- if(community_day_0 == TRUE){
       meta_file %>%
-  filter(day == i) %>% 
+        filter(day == 1) %>% 
         select(log_cfu, Mouse_ID) %>%
-  inner_join(day_0_otu_by_mouse_01) %>% 
+        inner_join(day_0_otu_by_mouse) %>% 
         select(-Mouse_ID)
-} else if (community == by_day) {
-  rf_data <- 
-  meta_file %>%
-  filter(day == i) %>% 
-  select(log_cfu, Mouse_ID) %>%
-  inner_join(day_0_otu_by_mouse_01) %>% 
-  select(-Mouse_ID)
-
     } else {
-  print('Input for community needed (day_0 or by_day')
+      cfu_by_day_comm %>% 
+        filter(day == 1) %>% 
+        select(-day)
     }
@@
         labs(x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)') + 
-        theme_bw() + ggtitle(paste("Random Forest Prediction of C Diff CFU \n on Day ",
-                                   i, " from Day 0 Community", sep = '')) +
+        theme_bw() + ggtitle(paste("Random Forest Prediction of C Diff CFU on Day ",
+                                   i, sep = '')) +
         annotate('text', x=0.15*(max(rf_prediction$observed)), 
@@
     rf_plot_list[[i]] <- list(predict_plot_day,importance_plot_day)
-
-})
-
-rf_day_0_comm_plots <- lapply(1:10, day_0)
-rf_by_day_com_plots <- lapply(1:10, by_day)
+  }
+}
+rf_day_0_comm_plots <- lapply(1:10, RandomForest_by_day, community_day_0 = T)
+rf_by_day_com_plots <- lapply(1:10, RandomForest_by_day, community_day_0 = F)
 
@@
 ggdraw()+
-  draw_plot(rf_plot_list[[1]][[1]], 0, 0, 0.5, 1) + 
-  draw_plot(rf_plot_list[[1]][[2]], 0.5, 0, 0.5, 1)
+  draw_plot(rf_day_0_comm_plots[[1]][[1]], 0, 0, 0.5, 1) + 
+  draw_plot(rf_day_0_comm_plots[[1]][[2]], 0.5, 0, 0.5, 1)
+
+```
+
+### Predict Day 1 CFU from Day 1 Community
+
+```{r fig.height= 5, fig.width=10}
+ggdraw()+
+  draw_plot(rf_by_day_com_plots[[1]][[1]], 0, 0, 0.5, 1) + 
+  draw_plot(rf_by_day_com_plots[[1]][[2]], 0.5, 0, 0.5, 1)
 
@@
 ggdraw()+
-  draw_plot(rf_plot_list[[2]][[1]], 0, 0, 0.5, 1) + 
-  draw_plot(rf_plot_list[[2]][[2]], 0.5, 0, 0.5, 1)
+  draw_plot(rf_day_0_comm_plots[[2]][[1]], 0, 0, 0.5, 1) + 
+  draw_plot(rf_day_0_comm_plots[[2]][[2]], 0.5, 0, 0.5, 1)
+
+```
+
+### Predict Day 2 CFU from Day 2 Community
+
+```{r fig.height= 5, fig.width=10}
+ggdraw()+
+  draw_plot(rf_by_day_com_plots[[2]][[1]], 0, 0, 0.5, 1) + 
+  draw_plot(rf_by_day_com_plots[[2]][[2]], 0.5, 0, 0.5, 1)
 
@@
 ggdraw()+
-  draw_plot(rf_plot_list[[3]][[1]], 0, 0, 0.5, 1) + 
-  draw_plot(rf_plot_list[[3]][[2]], 0.5, 0, 0.5, 1)
+  draw_plot(rf_day_0_comm_plots[[3]][[1]], 0, 0, 0.5, 1) + 
+  draw_plot(rf_day_0_comm_plots[[3]][[2]], 0.5, 0, 0.5, 1)
+
+```
+
+### Predict Day 3 CFU from Day 3 Community
+
+```{r fig.height= 5, fig.width=10}
+ggdraw()+
+  draw_plot(rf_by_day_com_plots[[3]][[1]], 0, 0, 0.5, 1) + 
+  draw_plot(rf_by_day_com_plots[[3]][[2]], 0.5, 0, 0.5, 1)
 
@@
 ggdraw()+
-  draw_plot(rf_plot_list[[4]][[1]], 0, 0, 0.5, 1) + 
-  draw_plot(rf_plot_list[[4]][[2]], 0.5, 0, 0.5, 1)
+  draw_plot(rf_day_0_comm_plots[[4]][[1]], 0, 0, 0.5, 1) + 
+  draw_plot(rf_day_0_comm_plots[[4]][[2]], 0.5, 0, 0.5, 1)
+
+```
+
+### Predict Day 4 CFU from Day 4 Community
+
+```{r fig.height= 5, fig.width=10}
+ggdraw()+
+  draw_plot(rf_by_day_com_plots[[4]][[1]], 0, 0, 0.5, 1) + 
+  draw_plot(rf_by_day_com_plots[[4]][[2]], 0.5, 0, 0.5, 1)
 
@@
 ggdraw()+
-  draw_plot(rf_plot_list[[5]][[1]], 0, 0, 0.5, 1) + 
-  draw_plot(rf_plot_list[[5]][[2]], 0.5, 0, 0.5, 1)
+  draw_plot(rf_day_0_comm_plots[[5]][[1]], 0, 0, 0.5, 1) + 
+  draw_plot(rf_day_0_comm_plots[[5]][[2]], 0.5, 0, 0.5, 1)
+
+```
+
+### Predict Day 5 CFU from Day 5 Community
+
+```{r fig.height= 5, fig.width=10}
+ggdraw()+
+  draw_plot(rf_by_day_com_plots[[5]][[1]], 0, 0, 0.5, 1) + 
+  draw_plot(rf_by_day_com_plots[[5]][[2]], 0.5, 0, 0.5, 1)
 
@@
 ggdraw()+
-  draw_plot(rf_plot_list[[6]][[1]], 0, 0, 0.5, 1) + 
-  draw_plot(rf_plot_list[[6]][[2]], 0.5, 0, 0.5, 1)
+  draw_plot(rf_day_0_comm_plots[[1]][[1]], 0, 0, 0.5, 1) + 
+  draw_plot(rf_day_0_comm_plots[[1]][[2]], 0.5, 0, 0.5, 1)
+
+```
+
+### Predict Day 6 CFU from Day 6 Community
+
+```{r fig.height= 5, fig.width=10}
+ggdraw()+
+  draw_plot(rf_by_day_com_plots[[6]][[1]], 0, 0, 0.5, 1) + 
+  draw_plot(rf_by_day_com_plots[[6]][[2]], 0.5, 0, 0.5, 1)
 
@@
 ggdraw()+
-  draw_plot(rf_plot_list[[7]][[1]], 0, 0, 0.5, 1) + 
-  draw_plot(rf_plot_list[[7]][[2]], 0.5, 0, 0.5, 1)
+  draw_plot(rf_day_0_comm_plots[[7]][[1]], 0, 0, 0.5, 1) + 
+  draw_plot(rf_day_0_comm_plots[[7]][[2]], 0.5, 0, 0.5, 1)
+
+```
+
+### Predict Day 7 CFU from Day 7 Community
+
+```{r fig.height= 5, fig.width=10}
+ggdraw()+
+  draw_plot(rf_by_day_com_plots[[7]][[1]], 0, 0, 0.5, 1) + 
+  draw_plot(rf_by_day_com_plots[[7]][[2]], 0.5, 0, 0.5, 1)
 
@@
 ggdraw()+
-  draw_plot(rf_plot_list[[8]][[1]], 0, 0, 0.5, 1) + 
-  draw_plot(rf_plot_list[[8]][[2]], 0.5, 0, 0.5, 1)
+  draw_plot(rf_day_0_comm_plots[[8]][[1]], 0, 0, 0.5, 1) + 
+  draw_plot(rf_day_0_comm_plots[[8]][[2]], 0.5, 0, 0.5, 1)
+
+```
+
+### Predict Day 8 CFU from Day 8 Community
+
+```{r fig.height= 5, fig.width=10}
+ggdraw()+
+  draw_plot(rf_by_day_com_plots[[8]][[1]], 0, 0, 0.5, 1) + 
+  draw_plot(rf_by_day_com_plots[[8]][[2]], 0.5, 0, 0.5, 1)
 
@@
 ggdraw()+
-  draw_plot(rf_plot_list[[9]][[1]], 0, 0, 0.5, 1) + 
-  draw_plot(rf_plot_list[[9]][[2]], 0.5, 0, 0.5, 1)
+  draw_plot(rf_day_0_comm_plots[[9]][[1]], 0, 0, 0.5, 1) + 
+  draw_plot(rf_day_0_comm_plots[[9]][[2]], 0.5, 0, 0.5, 1)
+
+```
+
+### Predict Day 9 CFU from Day 9 Community
+
+```{r fig.height= 5, fig.width=10}
+ggdraw()+
+  draw_plot(rf_by_day_com_plots[[9]][[1]], 0, 0, 0.5, 1) + 
+  draw_plot(rf_by_day_com_plots[[9]][[2]], 0.5, 0, 0.5, 1)
 
@@
 ggdraw()+
-  draw_plot(rf_plot_list[[10]][[1]], 0, 0, 0.5, 1) + 
-  draw_plot(rf_plot_list[[10]][[2]], 0.5, 0, 0.5, 1)
+  draw_plot(rf_day_0_comm_plots[[10]][[1]], 0, 0, 0.5, 1) + 
+  draw_plot(rf_day_0_comm_plots[[10]][[2]], 0.5, 0, 0.5, 1)
 
 ```
+
+### Predict Day 10 CFU from Day 10 Community
+
+```{r fig.height= 5, fig.width=10}
+ggdraw()+
+  draw_plot(rf_by_day_com_plots[[10]][[1]], 0, 0, 0.5, 1) + 
+  draw_plot(rf_by_day_com_plots[[10]][[2]], 0.5, 0, 0.5, 1)
+
+```
+
+***
+
+***
+
+### Otus associated with CFU C difficile
+
+```{r fig.width=10, fig.height=10}
+predictive_otus <- paste('Otu00', c('199','026','050','015','028',
+                                    '033','135','059','044'),    # lost with max rel abund > 10
+                                    #'007', '014', '035', '005'), # gain with max rel abund > 10
+                         sep = '')
+cfu_by_day_comm %>% 
+  select(log_cfu, one_of(predictive_otus)) %>% 
+  filter(!is.na(log_cfu)) %>% 
+  gather('otu','abundance',contains('Otu')) %>% 
+  ggplot(aes(x = abundance, y = log_cfu)) + geom_point() + 
+    facet_wrap(~ otu, scales = 'free_x') +
+    labs( x = 'Relative Abundance', y = 'CFU (log10)')
+
+```
+
+```{r fig.width=10, fig.height=10}
+cfu_by_day_comm %>% 
+  select(log_cfu, one_of(predictive_otus)) %>% 
+  filter(!is.na(log_cfu)) %>% 
+  gather('otu','abundance',contains('Otu')) %>% 
+  ggplot(aes(x = log10(abundance), y = log_cfu)) + geom_point() + 
+    facet_wrap(~ otu, scales = 'free_x') +
+    labs( x = 'Relative Abundance (log10)', y = 'CFU (log10)')
+
+```

### 2016-08-29 code/HumanCdGF_microbiota_predict_by_day.Rmd

@@
 ---
-title: "Test_code"
+title: "HumanCdGF_microbiota_predict_by_day"
 author: "Nick Lesniak"
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE)
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE, fig.width=10, fig.height=5)
 ```
@@
 
-```{r fig.height= 5, fig.width=10}
+```{r}
 ggdraw()+
@@
 
+Modeled with OTUs with max rel abund > 1% (216 OTUs) 
+When modeling based on mean rel abund > 1% (22 OTUs), R^2 decreases ~0.1 and MSE increases ~ 1 and lose all coriobacteriaceae/Actinobacteria
+
 ***
 
-### Otus associated with CFU C difficile
+### Otus associated with CFU C difficile and same day OTU
 
@@
                          sep = '')
+predictive_taxa <- get_tax(1, predictive_otus, tax_file)
+predictive_taxa$otu <- rownames(predictive_taxa)
+predictive_taxa$tax_label <- paste(gsub('_unclassified', '', predictive_taxa$tax),
+                                   ' (', gsub('tu0*', 'TU ', predictive_taxa$otu), 
+                                   ')', sep = '')
+
 cfu_by_day_comm %>% 
-  select(log_cfu, one_of(predictive_otus)) %>% 
+  select(log_cfu, day, one_of(predictive_otus)) %>% 
   filter(!is.na(log_cfu)) %>% 
   gather('otu','abundance',contains('Otu')) %>% 
-  ggplot(aes(x = abundance, y = log_cfu)) + geom_point() + 
-    facet_wrap(~ otu, scales = 'free_x') +
+  full_join(predictive_taxa) %>% 
+  ggplot(aes(x = abundance, y = log_cfu, color = factor(day))) + geom_point() + 
+    facet_wrap(~ tax_label, scales = 'free_x') +
     labs( x = 'Relative Abundance', y = 'CFU (log10)')
@@
 cfu_by_day_comm %>% 
-  select(log_cfu, one_of(predictive_otus)) %>% 
+  select(log_cfu, day, one_of(predictive_otus)) %>% 
   filter(!is.na(log_cfu)) %>% 
   gather('otu','abundance',contains('Otu')) %>% 
-  ggplot(aes(x = log10(abundance), y = log_cfu)) + geom_point() + 
-    facet_wrap(~ otu, scales = 'free_x') +
+  full_join(predictive_taxa) %>% 
+  ggplot(aes(x = log10(abundance), y = log_cfu, color = factor(day))) + geom_point() + 
+    facet_wrap(~ tax_label, scales = 'free_x') +
     labs( x = 'Relative Abundance (log10)', y = 'CFU (log10)')
@@
 ```
+
+### Day 0 OTUs associated with C difficile CFU
+
+```{r fig.width=10, fig.height=10}
+meta_file %>% 
+  select(day, log_cfu, Mouse_ID) %>% 
+  full_join(day_0_otu_by_mouse) %>% 
+  select(day, log_cfu, one_of(predictive_otus)) %>% 
+  filter(!day == 0) %>% 
+  gather('otu', 'abundance', contains('Otu')) %>% 
+  filter(!is.na(abundance)) %>% 
+  full_join(predictive_taxa) %>% 
+  ggplot(aes(x = abundance, y = log_cfu, color = factor(day))) + geom_point() + 
+    facet_wrap(~ tax_label, scales = 'free_x') +
+    labs( x = 'Relative Abundance', y = 'CFU (log10)')
+
+```
+
+```{r fig.width=10, fig.height=10}
+meta_file %>% 
+  select(day, log_cfu, Mouse_ID) %>% 
+  full_join(day_0_otu_by_mouse) %>% 
+  select(day, log_cfu, one_of(predictive_otus)) %>% 
+  filter(!day == 0) %>% 
+  gather('otu', 'abundance', contains('Otu')) %>% 
+  filter(!is.na(abundance)) %>% 
+  full_join(predictive_taxa) %>% 
+  ggplot(aes(x = log10(abundance), y = log_cfu, color = factor(day))) + geom_point() + 
+    facet_wrap(~ tax_label, scales = 'free_x') +
+    labs( x = 'Relative Abundance (log10)', y = 'CFU (log10)')
+
+```
+
+```{r fig.width=10, fig.height=10}
+
+corr_df_day_0_comm <- meta_file %>% 
+  select(day, log_cfu, Mouse_ID) %>% 
+  full_join(day_0_otu_by_mouse) %>% 
+  filter(!day == 0) %>% 
+  select(-day, -Mouse_ID)
+
+get_CFU_cor <- function(dataframe, CFU, OTU){
+    cor.test(dataframe[ , CFU], dataframe[ , OTU], method="spearman")[c("estimate","p.value")]
+}
+
+otu_cfu_corrs <- sapply(2:ncol(corr_df_day_0_comm), get_CFU_cor, CFU = 1, dataframe = corr_df_day_0_comm)
+colnames(otu_cfu_corrs) <- colnames(corr_df_day_0_comm)[-1]
+
+# identify the significant correlations after applying a correction for multiple
+# comparisons
+sig <- which(p.adjust(otu_cfu_corrs["p.value",], method="BH") < 0.05)
+sigg_corr_otus <- data.frame(estimate = unlist(otu_cfu_corrs["estimate", sig]),
+                   p_value = unlist(otu_cfu_corrs["p.value", sig]))
+sigg_corr_otus$otus <- gsub('.rho', '', rownames(sigg_corr_otus))
+sig_corr_phyla <- get_tax(1, sigg_corr_otus$otus, tax_file)
+sigg_corr_otus$phylogeny <- paste(gsub('_unclassified', '', sig_corr_phyla[,'tax']),
+                                   ' (', gsub('tu0*', 'TU ', sigg_corr_otus$otu), 
+                                   ')', sep = '')
+
+sigg_corr_otus %>%
+  filter(estimate >= 0.3 | estimate <= -0.3) %>%
+  mutate(color = ifelse(estimate >= 0.32 | estimate <= -0.32, 'black', 'red')) %>% 
+  ggplot( aes(x = reorder(factor(phylogeny), estimate), y = estimate)) + 
+  geom_bar(stat = 'identity', aes(fill = color)) +
+  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
+  scale_fill_manual(values = c(black = 'black', red = 'red'), guide = FALSE) +
+  labs( x = '', y = 'rho', title = 'Significant Spearman Correlation Coefficient - CFU vs OTU\n(only rho > 0.3 or < -0.3 shown)')
+
+# '199','026','050','015','028','033','135','059','044'),    # lost with max rel abund > 10
+                                    #'007', '014', '035', '005')
+
+#50, 15, 33, 135, 7, 14, 35, 5
+
+```

### 2016-08-29 code/HumanCdGF_microbiota_predict.Rmd
reduce features in mor/pers RF class model

@@
 # remove cecum samples
-meta_file <- meta_file[! meta_file$sample_type == 'cecum', ]
-# remove unused cage P2 and non 431 cdiff
-meta_file <- meta_file[! meta_file$cage_id %in% c('P2', 'CON', 'CON2', 'CON3'), ]
+meta_file <- meta_file[meta_file$sample_type == 'stool', ]
+# remove unused cage P2 and 
+meta_file <- meta_file[! meta_file$cage_id %in% 'P2', ]
+# select only mice with 431 cdiff
+meta_file <- meta_file[meta_file$cdiff_strain == '431', ]
 
@@
                               importance = TRUE, ntree = n_trees)
+rf_euth_import <- data.frame(importance = head(sort(rf_early_euth$importance[ , 'MeanDecreaseAccuracy'], decreasing = T), 10))
 
-unpredicted <- mouse_list[Predict_early_euth_df$Euth_Early != rf_early_euth$predicted]
-rf_euth_import <- data.frame(importance = head(
-  sort(rf_early_euth$importance[ , 'MeanDecreaseAccuracy'], decreasing = T), 30))
-ROC_euth <- roc(as.numeric(Predict_early_euth_df$Euth_Early), as.numeric(rf_early_euth$predicted))
-roc_plot <- data.frame(Specificity = 1 - ROC_euth$specificities, Sensitivity = ROC_euth$sensitivities)
+rf_early_euth_top_otu <- randomForest(as.factor(Euth_Early) ~ ., 
+                                      data = Predict_early_euth_df[, c(rownames(rf_euth_import), 'Euth_Early')], 
+                              importance = TRUE, ntree = n_trees)
 
-euth_rf_plot <- ggplot(data = roc_plot, aes(x = Specificity, y = Sensitivity)) + geom_path() + 
-  geom_abline(intercept = 0, slope = 1, linetype = 'dashed', alpha = 0.5)
-euth_import_plot <- ggplot(data = rf_euth_import, aes(x = factor(rownames(rf_euth_import)), y = importance)) + 
-        geom_point() + scale_x_discrete(limits = rev(rownames(rf_euth_import))) +
+unpredicted_top_otu <- mouse_list[Predict_early_euth_df$Euth_Early != rf_early_euth_top_otu$predicted]
+rf_euth_import_top_otu <- data.frame(importance = sort(rf_early_euth_top_otu$importance[ , 'MeanDecreaseAccuracy'], 
+                                                       decreasing = T))
+ROC_euth_top_otu <- roc(as.numeric(Predict_early_euth_df$Euth_Early), as.numeric(rf_early_euth_top_otu$predicted))
+roc_plot_top_otu <- data.frame(Specificity = 1 - ROC_euth_top_otu$specificities, 
+                       Sensitivity = ROC_euth_top_otu$sensitivities)
+
+euth_rf_plot_top_otu <- ggplot(data = roc_plot_top_otu, aes(x = Specificity, y = Sensitivity)) + 
+                      geom_path() + geom_abline(intercept = 0, slope = 1, 
+                                                linetype = 'dashed', alpha = 0.5)
+euth_import_plot_top_otu <- ggplot(data = rf_euth_import_top_otu, 
+                           aes(x = factor(rownames(rf_euth_import_top_otu)), y = importance)) + 
+        geom_point() + scale_x_discrete(limits = rev(rownames(rf_euth_import_top_otu))) +
         labs(x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
@@
 ggdraw()+
-  draw_plot(euth_rf_plot, 0, 0, 0.5, 1) + 
-  draw_plot(euth_import_plot, 0.5, 0, 0.5, 1)
+  draw_plot(euth_rf_plot_top_otu, 0, 0, 0.5, 1) + 
+  draw_plot(euth_import_plot_top_otu, 0.5, 0, 0.5, 1)
 ```
@@
 Confusion Matrix  
-`r kable(rf_early_euth$confusion)`
+`r kable(rf_early_euth_top_otu$confusion)`
 
-OOB error rate = `r 100*tail(rf_early_euth$err.rate[ , 1], 1)`
+OOB error rate = `r 100*tail(rf_early_euth_top_otu$err.rate[ , 1], 1)`
 
 Mice which were euthanized early, but the model predicts the will were not (Mouse Tag - Cage):
-`r unpredicted`
+`r unpredicted_top_otu`
 

### 2016-08-31 code/HumanCdGF_microbiota_predict.Rmd
added cross validation of mor/pers RF model

@@
 ```{r, message=FALSE}
-pack_used <- c('randomForest','miscTools','ggplot2','reshape2', 'pROC', 'knitr', 'cowplot')
+pack_used <- c('randomForest','miscTools','ggplot2','reshape2', 'pROC', 'knitr', 'cowplot','dplyr','AUCRF')
 for (dep in pack_used){
@@
                        sep='\t',header = T, row.names = 'OTU')
-#source('tax_level.R')
+source('tax_level.R')
 
@@
         annotate('text', x=0.15*(max(RF_cfu_all_df$log_cfu)), y=0.95*(max(RF_cfu_all_df$predicted)), 
-                 label=paste('R^2 = ', tail(RF_cfu_all_rf$rsq, 1),'\nMSE = ', tail(RF_cfu_all_rf$mse, 1), sep = ''))
+                 label=paste('R^2 = ', round(tail(RF_cfu_all_rf$rsq, 1), 2),
+                             '\nMSE = ', round(tail(RF_cfu_all_rf$mse, 1), 2), sep = ''))
 
@@
         annotate('text', x=0.15*(max(RF_cfu_persist_df$log_cfu)), y=0.95*(max(RF_cfu_persist_df$predicted)), 
-                 label=paste('R^2 = ', tail(RF_cfu_persist_rf$rsq, 1),'\nMSE = ', tail(RF_cfu_persist_rf$mse, 1), sep = ''))
+                 label=paste('R^2 = ', round(tail(RF_cfu_persist_rf$rsq, 1), 2),
+                             '\nMSE = ', round(tail(RF_cfu_persist_rf$mse, 1), 2), sep = ''))
 
@@
 ##Can we predict moribundity with the community structure at Day 0?
+#### randomForest
 ```{r fig.width=10, fig.height=5}
-Predict_early_euth_df <- data.frame(Euth_Early = ifelse(meta_file[meta_file$day %in% 0,
-                                                                  'Euthanized'] %in% 10,
-                                                        'No','Yes'), 
-                                    row.names = meta_file[meta_file$day %in% 0, 'Mouse_ID'])
-Predict_early_euth_df <- merge(Predict_early_euth_df, rel_abund_stool, by='row.names')
-mouse_list <- Predict_early_euth_df$Row.names
-Predict_early_euth_df <- Predict_early_euth_df[,!(colnames(Predict_early_euth_df) %in% "Row.names")]
-rf_early_euth <- randomForest(as.factor(Euth_Early) ~ ., data = Predict_early_euth_df, 
+meta_file$Euth_Early <- factor(ifelse(meta_file[ , 'Euthanized'] %in% 10, 1, 0))
+
+Predict_early_euth_df <- merge(meta_file[meta_file$day == 0, c('Mouse_ID', 'Euth_Early')], 
+                               rel_abund_stool, by.x = 'Mouse_ID', by.y='row.names')
+Predict_early_euth_df_names <- Predict_early_euth_df$Mouse_ID
+Predict_early_euth_df <- select(Predict_early_euth_df, -Mouse_ID)
+
+rf_early_euth <- randomForest(Euth_Early ~ ., data = Predict_early_euth_df, 
                               importance = TRUE, ntree = n_trees)
@@
 
-rf_early_euth_top_otu <- randomForest(as.factor(Euth_Early) ~ ., 
+rf_early_euth_top_otu <- randomForest(Euth_Early ~ ., 
                                       data = Predict_early_euth_df[, c(rownames(rf_euth_import), 'Euth_Early')], 
@@
 
-unpredicted_top_otu <- mouse_list[Predict_early_euth_df$Euth_Early != rf_early_euth_top_otu$predicted]
+unpredicted_top_otu <- Predict_early_euth_df_names[Predict_early_euth_df$Euth_Early != 
+                                               rf_early_euth_top_otu$predicted]
 rf_euth_import_top_otu <- data.frame(importance = sort(rf_early_euth_top_otu$importance[ , 'MeanDecreaseAccuracy'], 
@@
 
-***
+#### AUCRF
 
-## Analysis By Day
+```{r fig.width=6, fig.height=5}
+
+rf_early_aucrf <- AUCRF(Euth_Early ~ ., data = Predict_early_euth_df,
+                              ntree = n_trees, pdel = 0.05, ranking = 'MDA')
+
+otu_euth_probs <- predict(rf_early_aucrf$RFopt, type = 'prob')
+otu_euth_roc <- roc(Predict_early_euth_df$Euth_Early ~ otu_euth_probs[ , 2])
+
+#otu_euth_model <- rf_early_aucrf$RFopt
+#rf_early_aucrf$Kopt
+#rf_early_aucrf$Xopt
+#rf_early_aucrf$RFopt$confusion
+#OptimalSet(rf_early_aucrf)
+#plot(otu_euth_model_cv)
+#plot(otu_euth_model, which="ranking") 
+#plot(otu_euth_model, which="psel")
+
+aucrf_data <- Predict_early_euth_df[,c('Euth_Early',rf_early_aucrf$Xopt)]
+
+#### 10-fold Cross validated - 10 repetitions
+start <- proc.time()[3]
+iters <- 100
+cv10f_aucs <- c()
+cv10f_all_resp <- c()
+cv10f_all_pred <- c()
+for(j in 1:iters){
+  sampling <- sample(1:nrow(aucrf_data),nrow(aucrf_data),replace=F)
+  cv10f_probs <- rep(NA,nrow(aucrf_data))
+  for(i in seq(1,nrow(aucrf_data),5)){
+    train <- aucrf_data[sampling[-(i:(i+4))],]
+    test <- aucrf_data[sampling[i:(i+4)],]
+    temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
+    cv10f_probs[sampling[i:(i+4)]] <- predict(temp_model$RFopt, test, type='prob')[,2]
+  }
+  cv10f_roc <- roc(aucrf_data$Euth_Early~cv10f_probs)
+  cv10f_all_pred <- c(cv10f_all_pred, cv10f_probs)
+  cv10f_all_resp <- c(cv10f_all_resp, aucrf_data$Euth_Early)
+  cv10f_aucs[j] <- cv10f_roc$auc
+}
+cv10f_roc <- roc(cv10f_all_resp~cv10f_all_pred)
+end <- proc.time()[3]
+time <- end-start
+
+#### leave-one-out
+LOO_probs <- NULL
+LOO_probs <- sapply(1:nrow(aucrf_data), function(i){
+ train <- aucrf_data[-i,]
+  test <- aucrf_data[i,]
+  temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
+  LOO_probs[i] <- predict(temp_model$RFopt, test, type='prob')[,2]
+})
+LOO_roc <- roc(aucrf_data$Euth_Early~LOO_probs)
+
+#### plot rocs
+par(mar=c(4,4,1,1))
+plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='')
+plot(otu_euth_roc, col='black', lwd=2, add=T, lty=1)
+plot(cv10f_roc, col='gray', lwd=2, add=T, lty=1)
+plot(LOO_roc, col='blue', lwd=2, add=T, lty=2)
+mtext(side=2, text="Sensitivity", line=2.5)
+mtext(side=1, text="Specificity", line=2.5)
+legend('bottomright', legend=c(sprintf('Out-Of-Bag (AUC: %.3g)',otu_euth_roc$auc),
+                               sprintf('Leave-1-out (AUC: %.3g)',LOO_roc$auc),
+                               sprintf('10-fold CV (AUC: %.3g)',cv10f_roc$auc),
+                               sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_euth_roc,LOO_roc)$p.value),
+                               sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
+),lty=c(1,2,1,0,0), lwd=2, col=c('black','blue','gray'), bty='n')
+```
 
 ```{r fig.width=10, fig.height=5}
-layout(matrix(rep(c(1,1,1,1,2,2,2,2),4), ncol=8, nrow = 4, byrow = T))
+layout(matrix(c(1,3,2,2), nrow=2))
+
+#AUCRF curve
+par(mar=c(5,5,1.5,1))
+plot(rf_early_aucrf)
+#mtext('A', at=-90, font=2, side=3)
+
+#Importance Plot
+euth_otus <- rf_early_aucrf$Xopt
+euth_tax <- get_tax(1, euth_otus, tax_file)
+euth_tax$tax_label <- paste(gsub('_unclassified', '', euth_tax$tax),
+                                   ' (', gsub('tu0*', 'TU ', rownames(euth_tax)), 
+                                   ')', sep = '')
+dotchart(rev(rf_early_aucrf$ranking[euth_otus]), labels=rev(euth_tax[euth_otus, 'tax_label']), xlab='Mean Decrease Accuracy')
+#mtext('B', at=-0.014, font=2, side=3)
+
+#Abundance stripchart or most predictive otus
+euth_otus <- rf_early_aucrf$Xopt
+euth_otus <- rev(euth_otus[1:5])
+early_abunds <- Predict_early_euth_df[Predict_early_euth_df$Euth_Early==0, euth_otus]/10000 + 1e-4
+late_abunds <- Predict_early_euth_df[Predict_early_euth_df$Euth_Early==1, euth_otus]/10000 + 1e-4
+
+par(mar=c(4, 9, 1, 1))
+plot(1, type="n", ylim=c(0,length(euth_otus)*2), xlim=c(1e-4,1e-2), log="x", ylab="", xlab="Relative Abundance (%)", xaxt="n", yaxt="n")
+index <- 1
+for(i in euth_otus){
+  stripchart(at=index-0.35, jitter(early_abunds[,i], amount=1e-5), pch=21, bg="royalblue1", method="jitter", jitter=0.2, add=T, cex=1, lwd=0.5)
+  stripchart(at=index+0.35, jitter(late_abunds[,i], amount=1e-5), pch=21, bg="orange", method="jitter", jitter=0.2, add=T, cex=1, lwd=0.5)
+  segments(mean(early_abunds[,i]),index-0.7,mean(early_abunds[,i]),index, lwd=3)
+  segments(mean(late_abunds[,i]),index+0.7,mean(late_abunds[,i]),index, lwd=3)
+  index <- index + 2
+}
+axis(2, at=seq(1,index-2,2), labels=euth_tax[euth_otus, 'tax_label'], las=1, line=-0.5, tick=F, cex.axis=0.8)
+axis(1, at=c(1e-4, 1e-3, 1e-2), label=c("0", "0.1", "1"))
+legend('topright', legend=c("Persistant", "Susceptible"), pch=c(21, 21), pt.bg=c("orange","royalblue1"), cex=0.7)
+#mtext('C', at=1e-7, font=2, side=3)
+dev.off()
 

### 2016-09-02 code/HumanCdGF_entero_predict.Rmd
initial entero-based prediction

---
title: "humanCdGF_entero_predict"
author: "Nicholas Lesniak"
date: "September 1, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE, fig.width=10, fig.height=5)
```

```{r}
pack_used <- c('randomForest','tidyr','ggplot2','dplyr','knitr', 'cowplot', 'vegan', 'cluster')
for (dep in pack_used){
  if (dep %in% installed.packages()[,"Package"] == FALSE){
    install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
                     quiet=TRUE);
  }
  library(dep, verbose=FALSE, character.only=TRUE)
}
```

```{r}
# setup environment

# read in files
meta_file   <- read.table('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt', sep = '\t',
                        header = T)
shared_file <- read.table('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared',
                          sep='\t',header = T, row.names = 'Group')
tax_file <- read.table('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_cdiff.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy',
                       sep='\t',header = T, row.names = 'OTU')
# function to label otu with tax
source('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R')

# identify c difficile otus
cdiff_otu <- rownames(tax_file[grep('Clostridium_XI', tax_file$Taxonomy),])
# remove samples not used or with other cdiff strains
meta_file <- 
  meta_file %>% 
  filter(sample_type == 'stool', cdiff_strain == 431) %>% 
  select(Mouse_ID, cage_id, group, day, cdiff_cfu, Euthanized) %>% 
  mutate(log_cfu = log10(cdiff_cfu + 1)) %>% 
  filter(!is.na(log_cfu))

# create enterotypes
# first generate distance matrices
dist.JSD <- function(inMatrix, pseudocount=0.000001, ...) {
	KLD <- function(x,y) sum(x *log(x/y))
	JSD<- function(x,y) sqrt(0.5 * KLD(x, (x+y)/2) + 0.5 * KLD(y, (x+y)/2))
	matrixColSize <- length(colnames(inMatrix))
	matrixRowSize <- length(rownames(inMatrix))
	colnames <- colnames(inMatrix)
	resultsMatrix <- matrix(0, matrixColSize, matrixColSize)

  inMatrix = apply(inMatrix,1:2,function(x) ifelse (x==0,pseudocount,x))

	for(i in 1:matrixColSize) {
		for(j in 1:matrixColSize) { 
			resultsMatrix[i,j]=JSD(as.vector(inMatrix[,i]),
			as.vector(inMatrix[,j]))
		}
	}
	colnames -> colnames(resultsMatrix) -> rownames(resultsMatrix)
	as.dist(resultsMatrix)->resultsMatrix
	attr(resultsMatrix, "method") <- "dist"
	return(resultsMatrix) 
}

shared_dist <- dist.JSD(shared_file)

### 2016-09-02 code/HumanCdGF_entero_predict.Rmd
added enterotype files to rmd

@@
 # read in files
-meta_file   <- read.table('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt', sep = '\t',
-                        header = T)
-shared_file <- read.table('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared',
-                          sep='\t',header = T, row.names = 'Group')
-tax_file <- read.table('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_cdiff.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy',
-                       sep='\t',header = T, row.names = 'OTU')
+meta_file   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
+shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
+tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_cdiff.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy'
+entero_list <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.0.03.subsample.shared.for.0.03.dmm.5.mix.posterior'
+entero_abund <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.0.03.subsample.shared.for.0.03.dmm.5.mix.relabund'
+
+meta_file   <- read.table(meta_file, sep = '\t', header = T)
+shared_file <- read.table(shared_file, sep='\t',header = T, row.names = 'Group')
+tax_file <- read.table(tax_file, sep='\t',header = T, row.names = 'OTU')
+entero_list <- read.table(entero_list, sep = '\t', header = T)
+entero_abund <- read.table(entero_abund, sep = '\t', header = T)
+
 # function to label otu with tax
@@
 # create enterotypes
-# first generate distance matrices
-dist.JSD <- function(inMatrix, pseudocount=0.000001, ...) {
-	KLD <- function(x,y) sum(x *log(x/y))
-	JSD<- function(x,y) sqrt(0.5 * KLD(x, (x+y)/2) + 0.5 * KLD(y, (x+y)/2))
-	matrixColSize <- length(colnames(inMatrix))
-	matrixRowSize <- length(rownames(inMatrix))
-	colnames <- colnames(inMatrix)
-	resultsMatrix <- matrix(0, matrixColSize, matrixColSize)
-
-  inMatrix = apply(inMatrix,1:2,function(x) ifelse (x==0,pseudocount,x))
-
-	for(i in 1:matrixColSize) {
-		for(j in 1:matrixColSize) { 
-			resultsMatrix[i,j]=JSD(as.vector(inMatrix[,i]),
-			as.vector(inMatrix[,j]))
-		}
-	}
-	colnames -> colnames(resultsMatrix) -> rownames(resultsMatrix)
-	as.dist(resultsMatrix)->resultsMatrix
-	attr(resultsMatrix, "method") <- "dist"
-	return(resultsMatrix) 
-}
-
-shared_dist <- dist.JSD(shared_file)
+entero_list <-
+  unique(meta_file[meta_file$cage_id != 'inoculum',c('human_source','cage_id')])  
+  gather(entero_list, 'Partition', 2:6)

### 2016-09-07 code/make_HumanCdGF_metadata_file.R
fix output file names

@@ -9,14 +9,14 @@
 #         data/mothur/gf_all.an.0.03.subsample.shared
 #
 #    Outputs files:
-#         data/process/human_GF_mouse.txt
+#         data/process/human_CdGF_metadata.txt
 #         data/process/human_CdGF_all.0.03.subsample.shared
 #
 ###################
 input_mouse <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/raw/humanGF_ids.xlsx'
 input_shared <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_all.an.0.03.subsample.shared'
-output_metadata <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_GF_mouse.txt'
-output_shared <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_all.0.03.subsample.shared'
+output_metadata <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
+output_shared <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
 
 # files file used for metadata file
 input_meta_sample_ids <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/raw/gf_cdiff.files'

### 2016-09-07 code/make_HumanCdGF_metadata_file.R
fix output filenames

@@ -9,14 +9,14 @@
 #         data/mothur/gf_all.an.0.03.subsample.shared
 #
 #    Outputs files:
-#         data/process/human_GF_mouse.txt
-#         data/process/human_CdGF_all.0.03.subsample.shared
+#         data/process/human_CdGF_metadata.txt
+#         data/process/human_CdGF.an.unique_list.0.03.subsample.shared
 #
 ###################
 input_mouse <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/raw/humanGF_ids.xlsx'
 input_shared <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_all.an.0.03.subsample.shared'
-output_metadata <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_GF_mouse.txt'
-output_shared <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_all.0.03.subsample.shared'
+output_metadata <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
+output_shared <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
 
 # files file used for metadata file
 input_meta_sample_ids <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/raw/gf_cdiff.files'

### 2016-09-07 code/make_HumanCdGF_metadata_file.R
remove merge tool text

@@ -10,11 +10,7 @@
 #
 #    Outputs files:
 #         data/process/human_CdGF_metadata.txt
-<<<<<<< HEAD
 #         data/process/human_CdGF.an.unique_list.0.03.subsample.shared
-=======
-#         data/process/human_CdGF_all.0.03.subsample.shared
->>>>>>> master
 #
 ###################
 input_mouse <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/raw/humanGF_ids.xlsx'

### 2016-09-07 code/HumanCdGF_microbiota_predict_by_day.Rmd

@@
   filter(sample_type == 'stool', cdiff_strain == 431) %>% 
-  select(Mouse_ID, group, day, cdiff_cfu, Euthanized) %>% 
+  select(Mouse_ID, cage_id, group, day, cdiff_cfu, Euthanized) %>% 
   mutate(log_cfu = log10(cdiff_cfu + 1)) %>% 
@@
 ```
+
+Removing mice which were euthanized early did not significantly change the R2 or MSE
 
@@
 # '199','026','050','015','028','033','135','059','044'),    # lost with max rel abund > 10
-                                    #'007', '014', '035', '005')
-
-#50, 15, 33, 135, 7, 14, 35, 5
 
 ```
+
+```{r}
+# function to compare rel abund of otu with cdiff cfu
+plot_otu_w_cfu <- function(OTU){
+  rel_abund_otu <- data.frame(rel_abund = rel_abund_shared_01[,OTU],
+                              group = rownames(rel_abund_shared_01))
+  plot <- meta_file %>% 
+  select(group, day, cage_id, log_cfu) %>% 
+  left_join(rel_abund_otu) %>% 
+  select(cage_id, day, log_cfu, rel_abund) %>% 
+  filter(day %in% 0:10) %>%
+  group_by(cage_id, day) %>% 
+  summarise(abund = mean(rel_abund), cfu = mean(log_cfu)) %>% 
+  gather(data_type, values, abund:cfu) %>% 
+  ggplot(aes(x = factor(day), y = values, group = data_type, color = data_type)) +
+            facet_wrap(~cage_id) + 
+            geom_line() + scale_y_log10()
+}
+
+```

### 2016-09-07 code/HumanCdGF_entero_predict.Rmd
update predict files to work with latest meta file

@@
 meta_file   <- read.table(meta_file, sep = '\t', header = T)
-shared_file <- read.table(shared_file, sep='\t',header = T, row.names = 'Group')
+shared_file <- read.table(shared_file, sep='\t',header = T, row.names = 'sample_id')
 tax_file <- read.table(tax_file, sep='\t',header = T, row.names = 'OTU')
@@
   meta_file %>% 
-  filter(sample_type == 'stool', cdiff_strain == 431) %>% 
-  select(Mouse_ID, cage_id, group, day, cdiff_cfu, Euthanized) %>% 
-  mutate(log_cfu = log10(cdiff_cfu + 1)) %>% 
+  filter(cdiff_strain == 431) %>% 
+  select(mouse_id, cage_id, group, day, cdiff_cfu, Euthanized) %>% 
   filter(!is.na(log_cfu))

### 2016-09-07 code/HumanCdGF_microbiota_predict.Rmd
update predict files to work with latest meta file

@@
 
-setwd("~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code")
+meta_file   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
+shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
+tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_all.an.cons.taxonomy'
+tax_function <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R'
 
 # read in files
-meta_file   <- read.table('../data/process/human_CdGF_metadata.txt', sep = '\t',
-                        header = T, row.names = 'group')
-shared_file <- read.table('../data/process/human_CdGF.an.unique_list.0.03.subsample.shared',
-                          sep='\t',header = T, row.names='Group')
-tax_file <- read.table('../data/mothur/gf_cdiff.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy',
-                       sep='\t',header = T, row.names = 'OTU')
-source('tax_level.R')
+meta_file   <- read.table(meta_file, sep = '\t', header = T, row.names = 'sample_id')
+shared_file <- read.table(shared_file, sep = '\t', header = T, row.names = 'sample_id')
+tax_file <- read.table(tax_file, sep = '\t', header = T, row.names = 'OTU')
+source(tax_function)
 
-# add log transformed CFU
-meta_file$log_cfu <- log10(meta_file$cdiff_cfu + 1)
-# remove cecum samples
-meta_file <- meta_file[meta_file$sample_type == 'stool', ]
-# remove unused cage P2 and 
-meta_file <- meta_file[! meta_file$cage_id %in% 'P2', ]
 # select only mice with 431 cdiff
@@
 # rename day 0 community rows by mouse id - since meta rows have day in row names, use mouse id for both
-rownames(stool_shared) <- meta_file$Mouse_ID[rownames(meta_file) %in% 
+rownames(stool_shared) <- meta_file$mouse_id[rownames(meta_file) %in% 
                                                rownames(stool_shared) ]
@@
 
-#log_cfu_obs <- meta_file[meta_file$day %in% c(0:10), c('Mouse_ID', 'day', 'log_cfu') ]
+#log_cfu_obs <- meta_file[meta_file$day %in% c(0:10), c('mouse_id', 'day', 'log_cfu') ]
 #log_cfu_obs$predict <- NA
@@
 for (i in 1:10){
-  predict_all_df <- merge(meta_file[meta_file$day==i, c('Mouse_ID', 'day' , 'log_cfu') ],
-                    rel_abund_stool, by='Mouse_ID', by.y='row.names', all.y = T)
+  predict_all_df <- merge(meta_file[meta_file$day==i, c('mouse_id', 'day' , 'log_cfu') ],
+                    rel_abund_stool, by='mouse_id', by.y='row.names', all.y = T)
   predict_df <- predict_all_df[,! colnames(predict_all_df) %in% 'day']
-  predict_df <- predict_df[ ,!(colnames(predict_df) %in% "Mouse_ID")]
+  predict_df <- predict_df[ ,!(colnames(predict_df) %in% "mouse_id")]
   RF <- randomForest(log_cfu ~ ., data = predict_df[!is.na(predict_df$log_cfu), ],
@@
 
-CFU_euth <- merge(meta_file[meta_file$day==meta_file$Euthanized, c('Mouse_ID', 'log_cfu') ],
-                    rel_abund_stool, by='Mouse_ID', by.y='row.names', all.y = T)
-CFU_euth <- CFU_euth[ ,!(colnames(CFU_euth) %in% "Mouse_ID")]
+CFU_euth <- merge(meta_file[meta_file$day==meta_file$last_day_by_mouse, c('mouse_id', 'log_cfu') ],
+                    rel_abund_stool, by='mouse_id', by.y='row.names', all.y = T)
+CFU_euth <- CFU_euth[ ,!(colnames(CFU_euth) %in% "mouse_id")]
 
@@
 
-without_early_euth <- meta_file[meta_file$Euthanized %in% 10 ,c('day', 'Mouse_ID', 'log_cfu')]
+without_early_euth <- meta_file[meta_file$Early_Euth == FALSE ,c('day', 'mouse_id', 'log_cfu')]
 without_early_euth <- without_early_euth[without_early_euth$day %in% c(1:4), ]
@@
 without_by_day <- merge(without_early_euth, shared_X_cdiff_ra, by  = 'row.names')
-without_day_0 <- merge(without_early_euth, rel_abund_stool, by='Mouse_ID', by.y='row.names' )
+without_day_0 <- merge(without_early_euth, rel_abund_stool, by='mouse_id', by.y='row.names' )
 
@@
   predict_df <- predict_all_df[!is.na(predict_all_df$log_cfu), 
-                           !colnames(predict_all_df) %in% c('Mouse_ID', 'day')]
+                           !colnames(predict_all_df) %in% c('mouse_id', 'day')]
   RF <- randomForest(log_cfu ~ ., data = predict_df,
@@
   predict_df <- predict_all_df[!is.na(predict_all_df$log_cfu), 
-                           !colnames(predict_all_df) %in% c('Mouse_ID', 'Row.names', 'day')]
+                           !colnames(predict_all_df) %in% c('mouse_id', 'Row.names', 'day')]
   RF <- randomForest(log_cfu ~ ., data = predict_df,
@@
 predict_cfu_persist_df <- without_by_day[!is.na(without_by_day$log_cfu),
-                                           ! colnames(without_by_day) %in% c('Row.names','day','Mouse_ID')]
+                                           ! colnames(without_by_day) %in% c('Row.names','day','mouse_id')]
 RF_cfu_persist_rf <- randomForest(log_cfu ~ ., data = predict_cfu_persist_df,
@@
 ```{r fig.width=10, fig.height=5}
-meta_file$Euth_Early <- factor(ifelse(meta_file[ , 'Euthanized'] %in% 10, 1, 0))
+meta_file$Euth_Early <- factor(ifelse(meta_file[ , 'Early_Euth'] == FALSE, 1, 0))
 
-Predict_early_euth_df <- merge(meta_file[meta_file$day == 0, c('Mouse_ID', 'Euth_Early')], 
-                               rel_abund_stool, by.x = 'Mouse_ID', by.y='row.names')
-Predict_early_euth_df_names <- Predict_early_euth_df$Mouse_ID
-Predict_early_euth_df <- select(Predict_early_euth_df, -Mouse_ID)
+Predict_early_euth_df <- merge(meta_file[meta_file$day == 0, c('mouse_id', 'Euth_Early')], 
+                               rel_abund_stool, by.x = 'mouse_id', by.y='row.names')
+Predict_early_euth_df_names <- Predict_early_euth_df$mouse_id
+Predict_early_euth_df <- select(Predict_early_euth_df, -mouse_id)
 

### 2016-09-09 code/tax_level.R
update tax_level to work with missing taxa levels

@@ -5,14 +5,17 @@
 get_tax <- function(tax_level=1,row_list=1:length(rownames(tax_file)),df=tax_file){
      if (tax_level %in% c(1:5)){
           taxonomy <- df[row_list,]
-          taxonomy <-  data.frame(do.call('rbind', strsplit(as.character(taxonomy$Taxonomy),';',fixed=TRUE)))
-          taxonomy <- data.frame(sapply(taxonomy,gsub,pattern="\\(\\d*\\)",replacement=""))
+          taxonomy <- strsplit(as.character(taxonomy$Taxonomy),';',fixed=TRUE)
+          n <- max(sapply(taxonomy, length))
+          taxonomy <- lapply(taxonomy, `[`, seq_len(n))
+          taxonomy <- data.frame(do.call('rbind', taxonomy))
+          taxonomy <- sapply(taxonomy,gsub,pattern="\\(\\d*\\)",replacement="")
           level <- 7-tax_level
           tax_out <- as.character(taxonomy[,level])
           for (i in level:2){
                next_level <- i-1
-               tax_out[tax_out=='unclassified'] <- 
-                    as.character(taxonomy[tax_out=='unclassified',next_level])
+               tax_out[is.na(tax_out)] <- 
+                 as.character(taxonomy[is.na(tax_out),next_level])
           }
           return(data.frame(tax= tax_out, row.names=row_list))
      } else {print(

### 2016-09-09 code/HumanCdGF_microbiota_predict.Rmd

@@
 seed <- 18
-n_trees <- 2000
+n_trees <- 2001
 set.seed(seed)
@@
 ##Can we predict moribundity with the community structure at Day 0?
-#### randomForest
-```{r fig.width=10, fig.height=5}
+
+```{r fig.width=6, fig.height=5}
 meta_file$Euth_Early <- factor(ifelse(meta_file[ , 'Early_Euth'] == FALSE, 1, 0))
@@
 Predict_early_euth_df <- select(Predict_early_euth_df, -mouse_id)
-
-rf_early_euth <- randomForest(Euth_Early ~ ., data = Predict_early_euth_df, 
-                              importance = TRUE, ntree = n_trees)
-rf_euth_import <- data.frame(importance = head(sort(rf_early_euth$importance[ , 'MeanDecreaseAccuracy'], decreasing = T), 10))
-
-rf_early_euth_top_otu <- randomForest(Euth_Early ~ ., 
-                                      data = Predict_early_euth_df[, c(rownames(rf_euth_import), 'Euth_Early')], 
-                              importance = TRUE, ntree = n_trees)
-
-unpredicted_top_otu <- Predict_early_euth_df_names[Predict_early_euth_df$Euth_Early != 
-                                               rf_early_euth_top_otu$predicted]
-rf_euth_import_top_otu <- data.frame(importance = sort(rf_early_euth_top_otu$importance[ , 'MeanDecreaseAccuracy'], 
-                                                       decreasing = T))
-ROC_euth_top_otu <- roc(as.numeric(Predict_early_euth_df$Euth_Early), as.numeric(rf_early_euth_top_otu$predicted))
-roc_plot_top_otu <- data.frame(Specificity = 1 - ROC_euth_top_otu$specificities, 
-                       Sensitivity = ROC_euth_top_otu$sensitivities)
-
-euth_rf_plot_top_otu <- ggplot(data = roc_plot_top_otu, aes(x = Specificity, y = Sensitivity)) + 
-                      geom_path() + geom_abline(intercept = 0, slope = 1, 
-                                                linetype = 'dashed', alpha = 0.5)
-euth_import_plot_top_otu <- ggplot(data = rf_euth_import_top_otu, 
-                           aes(x = factor(rownames(rf_euth_import_top_otu)), y = importance)) + 
-        geom_point() + scale_x_discrete(limits = rev(rownames(rf_euth_import_top_otu))) +
-        labs(x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
-# title = 'OTU Importance in Prediction of Day Euthanized from Day 0 Community',
-
-ggdraw()+
-  draw_plot(euth_rf_plot_top_otu, 0, 0, 0.5, 1) + 
-  draw_plot(euth_import_plot_top_otu, 0.5, 0, 0.5, 1)
-```
-
-Confusion Matrix  
-`r kable(rf_early_euth_top_otu$confusion)`
-
-OOB error rate = `r 100*tail(rf_early_euth_top_otu$err.rate[ , 1], 1)`
-
-Mice which were euthanized early, but the model predicts the will were not (Mouse Tag - Cage):
-`r unpredicted_top_otu`
-
-#### AUCRF
-
-```{r fig.width=6, fig.height=5}
 

### 2016-09-09 code/HumanCdGF_microbiota_predict_by_day.Rmd

--- 2016-08-29 code/HumanCdGF_microbiota_predict_by_day.Rmd
@@
 # read in files
-meta_file   <- read.table('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt', sep = '\t',
-                        header = T)
-shared_file <- read.table('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared',
-                          sep='\t',header = T, row.names = 'Group')
-tax_file <- read.table('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_cdiff.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy',
-                       sep='\t',header = T, row.names = 'OTU')
-# function to label otu with tax
-source('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R')
+meta_file   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
+shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
+tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_all.an.cons.taxonomy'
+tax_function <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R'
+
+# read in files
+meta_file   <- read.table(meta_file, sep = '\t', header = T)
+shared_file <- read.table(shared_file, sep = '\t', header = T, row.names = 'sample_id')
+tax_file <- read.table(tax_file, sep = '\t', header = T, row.names = 'OTU')
+source(tax_function)
 
@@
   meta_file %>% 
-  filter(sample_type == 'stool', cdiff_strain == 431) %>% 
-  select(Mouse_ID, group, day, cdiff_cfu, Euthanized) %>% 
-  mutate(log_cfu = log10(cdiff_cfu + 1)) %>% 
+  filter(cdiff_strain == 431) %>% 
+  select(mouse_id, cage_id, sample_id, day, cdiff_cfu, Early_Euth, log_cfu) %>% 
   filter(!is.na(log_cfu))
@@
 rel_abund_shared_01 <- rel_abund_shared[ , apply(rel_abund_shared, 2, max) > 1]
-rel_abund_shared_01$group <- rownames(rel_abund_shared_01)
+rel_abund_shared_01$sample_id <- rownames(rel_abund_shared_01)
 
@@
   filter(day == 0 ) %>% 
-  select(group, Mouse_ID) %>% 
+  select(sample_id, mouse_id) %>% 
   left_join(rel_abund_shared_01) %>% 
-  select(-group)
+  select(-sample_id)
 # create df with cfu merge to shared
@@
   filter(day %in% 1:10) %>% 
-  select(group, day, log_cfu) %>% 
+  select(sample_id, day, log_cfu) %>% 
   left_join(rel_abund_shared_01) %>% 
-  select(-group, -one_of(cdiff_otu))
+  select(-sample_id, -one_of(cdiff_otu))
 
@@
         filter(day == 1) %>% 
-        select(log_cfu, Mouse_ID) %>%
+        select(log_cfu, mouse_id) %>%
         inner_join(day_0_otu_by_mouse) %>% 
-        select(-Mouse_ID)
+        select(-mouse_id)
     } else {
@@
 meta_file %>% 
-  select(day, log_cfu, Mouse_ID) %>% 
+  select(day, log_cfu, mouse_id) %>% 
   full_join(day_0_otu_by_mouse) %>% 
@@
 corr_df_day_0_comm <- meta_file %>% 
-  select(day, log_cfu, Mouse_ID) %>% 
+  select(day, log_cfu, mouse_id) %>% 
   full_join(day_0_otu_by_mouse) %>% 
   filter(!day == 0) %>% 
-  select(-day, -Mouse_ID)
+  select(-day, -mouse_id)
 
@@
 # '199','026','050','015','028','033','135','059','044'),    # lost with max rel abund > 10
-                                    #'007', '014', '035', '005')
-
-#50, 15, 33, 135, 7, 14, 35, 5
 
 ```
+
+```{r}
+# function to compare rel abund of otu with cdiff cfu
+plot_otu_w_cfu <- function(OTU){
+  rel_abund_otu <- data.frame(rel_abund = rel_abund_shared_01[,OTU],
+                              sample_id = rownames(rel_abund_shared_01))
+  plot <- meta_file %>% 
+  select(sample_id, day, cage_id, log_cfu) %>% 
+  left_join(rel_abund_otu) %>% 
+  select(cage_id, day, log_cfu, rel_abund) %>% 
+  filter(day %in% 0:10) %>%
+  group_by(cage_id, day) %>% 
+  summarise(abund = mean(rel_abund), cfu = mean(log_cfu)) %>% 
+  gather(data_type, values, abund:cfu) %>% 
+  ggplot(aes(x = factor(day), y = values, group = data_type, color = data_type)) +
+            facet_wrap(~cage_id) + 
+            geom_line() + scale_y_log10()
+}
+
+```

### 2016-09-09 code/tax_level.R
remove '_unclassified' in tax_level.R

@@ -17,6 +17,7 @@ get_tax <- function(tax_level=1,row_list=1:length(rownames(tax_file)),df=tax_fil
                tax_out[is.na(tax_out)] <- 
                  as.character(taxonomy[is.na(tax_out),next_level])
           }
+          tax_out <- gsub('_unclassified', '', tax_out)
           return(data.frame(tax= tax_out, row.names=row_list))
      } else {print(
           'Error: Level must be 1 (genus), 2 (family), 3 (order), 4 (class), 5 (phylum)')

### 2016-09-09 code/HumanCdGF_entero_predict.Rmd
edit predict files to work with new tax_level

@@
 shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
-tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_cdiff.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy'
+tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_all.an.cons.taxonomy'
 entero_list <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.0.03.subsample.shared.for.0.03.dmm.5.mix.posterior'

### 2016-09-09 code/HumanCdGF_microbiota_predict.Rmd
edit predict files to work with new tax_level

@@
 euth_tax <- get_tax(1, euth_otus, tax_file)
-euth_tax$tax_label <- paste(gsub('_unclassified', '', euth_tax$tax),
-                                   ' (', gsub('tu0*', 'TU ', rownames(euth_tax)), 
+euth_tax$tax_label <- paste(euth_tax$tax, ' (', 
+                            gsub('tu0*', 'TU ', rownames(euth_tax)), 
                                    ')', sep = '')

### 2016-09-12 code/inocula_nmds.R
updated nmds code and files

@@ -20,17 +20,38 @@ shared <- read.table('data/process/human_CdGF.an.unique_list.0.03.subsample.shar
 inocula <- rownames(meta_file)[meta_file$cage_id=='inoculum']
 inocula_shared <- shared[shared$sample_id %in% inocula,]
 
-#add back label and numOTUs column to subsetted shared file-- need this to run dist.shared
+#subset shared to be day 0 only- will use later for comparison 
+day0 <- rownames(meta_file)[meta_file$day==0]
+day0_shared <- shared[shared$sample_id %in% day0,]
 
+#add back label and numOTUs column to subsetted shared file-- need this to run dist.shared
 inocula_shared <- cbind(label=0.03, inocula_shared[1], numOTUs=2255, inocula_shared[2:ncol(inocula_shared)])
 colnames(inocula_shared)[2] <- 'Group'
 
+#for day 0 also
+day0_shared <- cbind(label=0.03, day0_shared[1], numOTUs=2255, day0_shared[2:ncol(day0_shared)])
+colnames(day0_shared)[2] <- 'Group'
+
+#combine these all to run as one file duhhhh
+inputd0_shared <- rbind(inocula_shared, day0_shared)
+
 #write file out so we can use it in mothur
 write.table(inocula_shared, file='data/process/inocula.shared', quote=F,sep='\t',row.names=F)
+write.table(day0_shared, file='data/process/day0.shared', quote=F,sep='\t',row.names=F)
+write.table(inputd0_shared, file='data/process/inputd0.shared', quote=F, sep='\t', row.names=F)
 
 #run dist.shared, nmds in mothur to get axes file
+#mothur ouput stats for combined file: 
+#Number of dimensions:	2
+#Lowest stress :	0.362683
+#R-squared for configuration:	0.380491
 
-nmds <- read.table(file='data/process/inocula.thetayc.0.03.lt.ave.nmds.axes', header = T)
+combo_nmds <- read.table(file='data/process/inputd0.thetayc.0.03.lt.nmds.axes', header = T)
+inoc_nmds <- read.table(file='data/process/inocula.thetayc.0.03.lt.nmds.axes', header = T)
+day0_nmds <- read.table(file='data/process/day0.thetayc.0.03.lt.nmds.axes', header = T)
+
+
+
 
 #need to color points by type 
 inocula_design <- read.table(file='data/raw/donor.design', header = F)

### 2016-09-12 code/HumanCdGF_microbiota_strains.Rmd
initial Rmd for non-431 strain analysis

Strain independence
========================================================
``` {r, include = FALSE}
# title: "Strain independence"
# author: "Nicholas Lesniak"
# date: "September 12, 2016"
# output: html_document
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
```

```{r, message=FALSE}
pack_used <- c('ggplot2','knitr','dplyr')
for (dep in pack_used){
  if (dep %in% installed.packages()[,"Package"] == FALSE){
    install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
                     quiet=TRUE);
  }
  library(dep, verbose=FALSE, character.only=TRUE)
}
non_431   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'

# read in files
non_431   <- read.table(non_431, sep = '\t', header = T, row.names = 'sample_id')

# select mice infected with two strains of c diff
cdiff_strains <- unique(non_431$cdiff_strain)
non_431_strains <- cdiff_strains[!unique(cdiff_strains) %in% c('431', NA)]
non_431_inocula <- unique(non_431[non_431$cdiff_strain %in% non_431_strains, 'human_source'])
non_431 <- non_431[non_431$human_source %in% non_431_inocula, ]
```

```{r fig.height= 5, fig.width=10}
non_431 %>% 
  filter(day %in% c(0:10), !is.na(log_cfu) ) %>%
  select(mouse_id, day, human_source, cdiff_strain, cage_id, log_cfu) %>% 
  ggplot(aes(x = factor(day), y = log_cfu, color = factor(cdiff_strain))) + geom_point() +
            stat_summary(fun.y=mean, geom="line", aes(group = cage_id)) + 
            facet_grid(. ~ human_source) + theme_bw() + 
            labs(x = 'Day', y = 'C Diff CFU (log10)') +
            theme(legend.justification=c(1,0.5), legend.position=c(1,0.5), 
                  legend.title=element_blank())

```

### 2016-09-15 code/HumanCdGF_microbiota_predict_by_day.Rmd

@@
 
-```{r}
+```{r load packages}
 pack_used <- c('randomForest','miscTools','tidyr','ggplot2','dplyr', 'pROC', 'knitr', 'cowplot')
@@
 
-```{r}
-# setup environment
+```{r setup environment}
 
@@
 
-```{r}
+```{r Day_1_CFU_Day_0_Community}
 ggdraw()+
@@
 
-```{r}
+```{r Day_1_CFU__Community}
 ggdraw()+
@@
 
-```{r}
+```{r Day_2_CFU_Day_0_Community}
 ggdraw()+
@@
 
-```{r}
+```{r Day_2_CFU__Community}
 ggdraw()+
@@
 
-```{r}
+```{r Day_3_CFU_Day_0_Community}
 ggdraw()+
@@
 
-```{r}
+```{r Day_3_CFU__Community}
 ggdraw()+
@@
 
-```{r}
+```{r Day_4_CFU_Day_0_Community}
 ggdraw()+
@@
 
-```{r}
+```{r Day_4_CFU__Community}
 ggdraw()+
@@
 
-```{r}
+```{r Day_5_CFU_Day_0_Community}
 ggdraw()+
@@
 
-```{r}
+```{r Day_5_CFU__Community}
 ggdraw()+
@@
 
-```{r}
+```{r Day_6_CFU_Day_0_Community}
 ggdraw()+
@@
 
-```{r}
+```{r Day_6_CFU__Community}
 ggdraw()+
@@
 
-```{r}
+```{r Day_7_CFU_Day_0_Community}
 ggdraw()+
@@
 
-```{r}
+```{r Day_7_CFU__Community}
 ggdraw()+
@@
 
-```{r}
+```{r Day_8_CFU_Day_0_Community}
 ggdraw()+
@@
 
-```{r}
+```{r Day_8_CFU__Community}
 ggdraw()+
@@
 
-```{r}
+```{r Day_9_CFU_Day_0_Community}
 ggdraw()+
@@
 
-```{r}
+```{r Day_9_CFU__Community}
 ggdraw()+
@@
 
-```{r}
+```{r Day_10_CFU_Day_0_Community}
 ggdraw()+
@@
 
-```{r}
+```{r Day_10_CFU__Community}
 ggdraw()+
@@
 predictive_taxa$otu <- rownames(predictive_taxa)
-predictive_taxa$tax_label <- paste(gsub('_unclassified', '', predictive_taxa$tax),
-                                   ' (', gsub('tu0*', 'TU ', predictive_taxa$otu), 
+predictive_taxa$tax_label <- paste(predictive_taxa$tax, ' (', 
+                                   gsub('tu0*', 'TU ', predictive_taxa$otu), 
                                    ')', sep = '')
@@
 sig_corr_phyla <- get_tax(1, sigg_corr_otus$otus, tax_file)
-sigg_corr_otus$phylogeny <- paste(gsub('_unclassified', '', sig_corr_phyla[,'tax']),
-                                   ' (', gsub('tu0*', 'TU ', sigg_corr_otus$otu), 
+sigg_corr_otus$phylogeny <- paste(sig_corr_phyla[,'tax'], ' (', 
+                                  gsub('tu0*', 'TU ', sigg_corr_otus$otu), 
                                    ')', sep = '')

### 2016-09-19 code/CdGF_AUCRF.Rmd
update predict with shared w/o cdiff

--- 2016-09-09 code/HumanCdGF_microbiota_predict.Rmd
@@
 ---
-title: "Predict Disease Outcome by Community Structures"
+title: "CdGF_aucrf"
 author: "Nicholas Lesniak"
-date: "July 13, 2016"
+date: "September 15, 2016"
 output: html_document
@@
 #Create df with relative abundances
-stool_shared <- shared_file[rownames(meta_file[meta_file$day %in% 0, ]) ,]
+stool_shared <- shared_file[rownames(meta_file[meta_file$day %in% 0, ]), ]
 # use %in% instead of == because to ignore NA
@@
 
-#log_cfu_obs <- meta_file[meta_file$day %in% c(0:10), c('mouse_id', 'day', 'log_cfu') ]
-#log_cfu_obs$predict <- NA
-#log_cfu_obs[log_cfu_obs$day==2, ] <- predict(RF, rel_abund_stool)
-
-importance_df <- data.frame(matrix(nrow = length(OTU_list), 
-                                   ncol = max(meta_file$day, na.rm = T)),
-                            row.names = OTU_list)
-rsq_df <- data.frame(matrix(nrow = 1, ncol = max(meta_file$day, na.rm = T)),
-                         row.names = 'RSQ')
-mse_df <- data.frame(matrix(nrow = 1, ncol = max(meta_file$day, na.rm = T)),
-                         row.names = 'MSE')
-prediction_df <- data.frame(matrix(nrow = 0, ncol = 4))
-rsq_features <- data.frame('Day'=NA, 'Features (n)'= NA, 'RSQ'=NA, 'MSE'=NA)
-
-for (i in 1:10){
-  predict_all_df <- merge(meta_file[meta_file$day==i, c('mouse_id', 'day' , 'log_cfu') ],
-                    rel_abund_stool, by='mouse_id', by.y='row.names', all.y = T)
-  predict_df <- predict_all_df[,! colnames(predict_all_df) %in% 'day']
-  predict_df <- predict_df[ ,!(colnames(predict_df) %in% "mouse_id")]
-  RF <- randomForest(log_cfu ~ ., data = predict_df[!is.na(predict_df$log_cfu), ],
-                   importance = TRUE, ntree = n_trees)
-  rsq_df[i] <- RF$rsq[length(RF$rsq)]
-  mse_df[i] <- RF$mse[length(RF$mse)]
-
-  # sort the features by their effect on % increase in the standard error
-  importance_df[ , i] <- importance(RF)[,1]
-  importance_sorted <- sort(importance(RF)[,1], decreasing = T)
-  colnames(importance_df)[i] <- paste('day', i, 'importance', sep = '_')
-# fit the model back on the datar and tack it to the end of the counts file
-  random_forest_output <- cbind(predict_all_df[!is.na(predict_all_df$log_cfu), c('day','log_cfu')],
-                              RF$predicted)
-  colnames(random_forest_output) <- c('day', 'observed', 'predicted')
-  prediction_df <- rbind(prediction_df, random_forest_output) 
-  for (k in 1:25){
-    n_features <- k
-    rf_partial <- randomForest(log_cfu ~ ., data=predict_df[!is.na(predict_df$log_cfu),
-                            c( 'log_cfu', names(importance_sorted)[1:n_features])], 
-                           importance=TRUE, ntree=n_trees)
-    rsq_features <- rbind(rsq_features, c(i, k, tail(rf_partial$rsq, 1), tail(rf_partial$mse, 1)))
-  }
-}
-rsq_features <- rsq_features[-1,]
-
-day_0_rf_plot <- ggplot(data = prediction_df, aes(x=observed, y=predicted)) + geom_point() + 
-        facet_wrap( ~ day, nrow = 2) + 
-        labs(title = 'Random Forest Prediction of C Diff CFU dpi from Day 0 Community',
-        x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)')
-
-rsq_df <- t(rsq_df)
-rsq_df <- as.data.frame(cbind(rsq_df, day = c(1:10)))
-day_0_rsq_plot <- ggplot(data = rsq_df, aes(x = factor(day), y = RSQ)) + geom_point() + theme_bw() +
-        labs(x= 'Day', y = 'R^2') #title = 'Variation Explained by Day of C Diff CFU dpi from Day 0 Community',
-
-mse_df <- t(mse_df)
-mse_df <- as.data.frame(cbind(mse_df, day = c(1:10)))
-day_0_mse_plot <- ggplot(data = mse_df, aes(x = factor(day), y = MSE)) + geom_point() + theme_bw() +
-        labs(x= 'Day', y = 'MSE') # title = 'MSE by Day of C Diff CFU dpi from Day 0 Community',
-day_0_r2feat_plot <- ggplot(rsq_features, aes(x=Features..n., y=RSQ, group=(Day), color=factor(Day))) + geom_line() +
-  labs(x = 'Features (n)', y = 'R^2') + scale_color_discrete(guide=FALSE) + theme_bw()
-#  title= 'R^2 from RF with top n features\n
-#                    Day of C Diff CFU dpi from Day 0 Community', scale_color_discrete(name = 'Day')
-
-n_important <- 20
-median_importance <- names(head(sort((apply(importance_df,1, median)), decreasing = T), n_important))
-med_import_df <- melt(cbind(rownames(importance_df), importance_df)[median_importance, ])
-colnames(med_import_df) <- c('OTU', 'Day', 'Importance')
-day_0_feat_plot <- ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + facet_wrap( ~ OTU, nrow = 2) +
-        geom_point() + scale_x_discrete(labels = c(1:10)) +
-        labs(title = 'OTU Importance in Prediction of C Diff CFU dpi from Day 0 Community\n (Top 20 Median OTUs)',
-        x= 'Day', y = '% Increase in MSE') + theme_bw()
-
-ggdraw()+
-  draw_plot(day_0_rf_plot, 0, 0.5, 0.75, 0.5) + 
-  draw_plot(day_0_rsq_plot, 0.875, 0.75, 0.125, 0.25) +
-  draw_plot(day_0_mse_plot, 0.75, 0.75, 0.125, 0.25) + 
-  draw_plot(day_0_r2feat_plot, 0.75, 0.5, 0.25, 0.25) + 
-  draw_plot(day_0_feat_plot, 0, 0, 1, 0.5)
 ```
-
-##Can we predict the endpoint C. difficile CFU with the community structure at Day 0?
-Endpoint has data from Day 2,3,10 - might be misleading since contains CFU from different data points
-
-```{r fig.height= 5, fig.width=10}
-
-CFU_euth <- merge(meta_file[meta_file$day==meta_file$last_day_by_mouse, c('mouse_id', 'log_cfu') ],
-                    rel_abund_stool, by='mouse_id', by.y='row.names', all.y = T)
-CFU_euth <- CFU_euth[ ,!(colnames(CFU_euth) %in% "mouse_id")]
-
-RF_CFU_day_euth <- randomForest(log_cfu ~ ., data = CFU_euth[!is.na(CFU_euth$log_cfu), ],
-                   importance = TRUE, ntree = n_trees)
-rsq_day_euth <- round(RF_CFU_day_euth$rsq[length(RF_CFU_day_euth$rsq)], 2)
-mse_day_euth <- round(RF_CFU_day_euth$mse[length(RF_CFU_day_euth$mse)], 2)
-# sort the features by their effect on % increase in the standard error
-importance_day_euth <- importance(RF_CFU_day_euth)[,1]
-# fit the model back on the datar and tack it to the end of the counts file
-rf_predict_cfu_output <- cbind(CFU_euth[!is.na(CFU_euth$log_cfu), 'log_cfu'], data.frame(RF_CFU_day_euth$predicted))
-colnames(rf_predict_cfu_output) <- c('observed', 'predicted')
-
-day_euth_rf_plot <- ggplot(data = rf_predict_cfu_output, aes(x=observed, y=predicted)) + geom_point() + 
-        labs(x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)') + theme_bw() + 
-        ggtitle("Random Forest Prediction of C Diff CFU \n on Day Euthanized from Day 0 Community") +
-        annotate('text', x=0.15*(max(rf_predict_cfu_output$observed)), y=0.95*(max(rf_predict_cfu_output$predicted)), 
-                 label=paste('R^2 = ', rsq_day_euth,'\nMSE = ', mse_day_euth, sep = ''))
-
-top_important_OTU <- data.frame(head(sort(importance_day_euth, decreasing = T), n_important))
-colnames(top_important_OTU) <- 'Importance'
-top_important_OTU$OTU <- rownames(top_important_OTU)
-
-n_feat_day_euth <- data.frame('Features (n)'= NA, 'RSQ'=NA, 'MSE'=NA)
-importance_sorted_euth <- sort(importance(RF)[,1], decreasing = T)
-
-for (k in 1:25){
-    n_features <- k
-    rf_partial <- randomForest(log_cfu ~ ., data=CFU_euth[!is.na(CFU_euth$log_cfu),
-                            c( 'log_cfu', names(importance_sorted_euth)[1:n_features])], 
-                           importance=TRUE, ntree=n_trees)
-    n_feat_day_euth <- rbind(n_feat_day_euth, c(k, tail(rf_partial$rsq, 1), tail(rf_partial$mse, 1)))
-}
-n_feat_day_euth <- n_feat_day_euth[-1,]
-
-day_euth_r2feat_plot <- ggplot(n_feat_day_euth, aes(x=Features..n., y=RSQ)) + geom_line() +
-  labs(x = 'Features (n)', y = 'R^2') + theme_bw()
-#     title= 'R^2 from RF with top n features\n Day euthanized C Diff CFU from Day 0 Community',
-
-day_euth_feat_plot <- ggplot(data = top_important_OTU, aes(x = factor(OTU), y = Importance)) + 
-        geom_point() + scale_x_discrete(limits = rev(top_important_OTU$OTU)) +
-        labs(x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
-#        title = 'OTU Importance in Prediction of euthanized C Diff CFU from Day 0 Community',
-
-ggdraw()+
-  draw_plot(day_euth_rf_plot, 0, 0.5, 0.5, 0.5) + 
-  draw_plot(day_euth_r2feat_plot, 0, 0, 0.5, 0.5) +
-  draw_plot(day_euth_feat_plot, 0.5, 0, 0.5, 1)
-
-```
-
-## Does Community structure predict C. difficile CFU of same day?
-
-### RF cfu vs community by day
-
-```{r fig.height= 10, fig.width=10}
-
-#remove c diff from shared
-cdiff_otu <- rownames(tax_file[grep('Clostridium_XI', tax_file$Taxonomy),])
-shared_X_cdiff <- shared_file[, ! colnames(shared_file) %in% cdiff_otu]
-shared_X_cdiff_ra <-  100 * shared_X_cdiff / unique(apply(shared_X_cdiff, 1, sum))
-OTUs_01 <- apply(shared_X_cdiff_ra, 2, max) > 1
-shared_X_cdiff_ra <- shared_X_cdiff_ra[, OTUs_01]
-# create data frame with shared and cfu and day
-CFU_community_by_day <- merge(meta_file[meta_file$day %in% c(1:10), c('day','log_cfu')], shared_X_cdiff_ra,
-                              by = 'row.names')
-
-###################################
-
-importance_df_day <- data.frame(matrix(nrow = sum(OTUs_01*1), 
-                                   ncol = max(meta_file$day, na.rm = T)),
-                            row.names = colnames(shared_X_cdiff_ra))
-rsq_df_day <- data.frame(matrix(nrow = 1, ncol = max(meta_file$day, na.rm = T)),
-                         row.names = 'RSQ')
-mse_df_day <- data.frame(matrix(nrow = 1, ncol = max(meta_file$day, na.rm = T)),
-                         row.names = 'MSE')
-prediction_df_day <- data.frame(matrix(nrow = 0, ncol = 4))
-rsq_features_day <- data.frame('Day'=NA, 'Features (n)'= NA, 'RSQ'=NA, 'MSE'=NA)
-
-for (i in 1:10){
-  predict_all_df <- CFU_community_by_day[CFU_community_by_day$day == i ,]
-  predict_df <- predict_all_df[!is.na(predict_all_df$log_cfu), 
-                           !colnames(predict_all_df) %in% c('Row.names', 'day')]
-  RF <- randomForest(log_cfu ~ ., data = predict_df,
-                   importance = TRUE, ntree = n_trees)
-  rsq_df_day[i] <- RF$rsq[length(RF$rsq)]
-  mse_df_day[i] <- RF$mse[length(RF$mse)]
-
-  # sort the features by their effect on % increase in the standard error
-  importance_df_day[ , i] <- importance(RF)[,1]
-  importance_sorted <- sort(importance(RF)[,1], decreasing = T)
-  colnames(importance_df_day)[i] <- paste('day', i, 'importance', sep = '_')
-# fit the model back on the datar and tack it to the end of the counts file
-  random_forest_output <- cbind(predict_all_df[!is.na(predict_all_df$log_cfu), c('day','log_cfu')],
-                              RF$predicted)
-  colnames(random_forest_output) <- c('day', 'observed', 'predicted')
-  prediction_df_day <- rbind(prediction_df_day, random_forest_output) 
-  for (k in 1:25){
-    n_features <- k
-    rf_partial <- randomForest(log_cfu ~ ., 
-                               data=predict_df[ , c( 'log_cfu', names(importance_sorted)[1:n_features])], 
-                               importance=TRUE, ntree=n_trees)
-    rsq_features_day <- rbind(rsq_features_day, c(i, k, tail(rf_partial$rsq, 1), tail(rf_partial$mse, 1)))
-  }
-}
-rsq_features_day <- rsq_features_day[-1,]
-
-day_rf_plot <- ggplot(data = prediction_df_day, aes(x=observed, y=predicted)) + geom_point() + 
-        facet_wrap( ~ day, nrow = 2) + 
-        labs(title = 'Random Forest Prediction of C Diff CFU dpi from Community on same day',
-        x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)')
-
-rsq_df_day <- t(rsq_df_day)
-rsq_df_day <- as.data.frame(cbind(rsq_df_day, day = c(1:10)))
-day_rsq_plot <- ggplot(data = rsq_df_day, aes(x = factor(day), y = RSQ)) + geom_point() + theme_bw() +
-        labs(x= 'Day', y = 'R^2') # title = 'Variation Explained by Day of C Diff CFU dpi from Community on same day',
-
-mse_df_day <- t(mse_df_day)
-mse_df_day <- as.data.frame(cbind(mse_df_day, day = c(1:10)))
-day_mse_plot <- ggplot(data = mse_df_day, aes(x = factor(day), y = MSE)) + geom_point() + theme_bw() +
-        labs(x= 'Day', y = 'MSE') # title = 'MSE by Day of C Diff CFU dpi from Community on same day',
-
-day_r2feat_plot <- ggplot(rsq_features_day, aes(x=Features..n., y=RSQ, group=(Day), color=factor(Day))) + geom_line() +
-  labs(x = 'Features (n)', y = 'R^2') + scale_color_discrete(guide = FALSE) + theme_bw()
-# title= 'R^2 from RF with top n features\n Day of C Diff CFU dpi from Community on same day',
-
-n_important <- 20
-median_importance <- names(head(sort((apply(importance_df_day,1, median)), decreasing = T), n_important))
-med_import_df <- melt(cbind(rownames(importance_df_day), importance_df_day)[median_importance, ])
-colnames(med_import_df) <- c('OTU', 'Day', 'Importance')
-
-day_feat_plot <- ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + facet_wrap( ~ OTU, nrow = 2) +
-        geom_point() + scale_x_discrete(labels = c(1:10)) +
-        labs(title = 'OTU Importance in Prediction of C Diff CFU dpi from Community on same day\n (Top 20 Median OTUs)',
-        x= 'Day', y = '% Increase in MSE') + theme_bw()
-
-ggdraw()+
-  draw_plot(day_rf_plot, 0, 0.5, 0.75, 0.5) + 
-  draw_plot(day_rsq_plot, 0.875, 0.75, 0.125, 0.25) +
-  draw_plot(day_mse_plot, 0.75, 0.75, 0.125, 0.25) + 
-  draw_plot(day_r2feat_plot, 0.75, 0.5, 0.25, 0.25) + 
-  draw_plot(day_feat_plot, 0, 0, 1, 0.5)
-```
-
-####
-RF all mice/days together
-
-```{r fig.height= 7, fig.width=10}
-predict_cfu_all_df <- CFU_community_by_day[!is.na(CFU_community_by_day$log_cfu),
-                                           ! colnames(CFU_community_by_day) %in% c('Row.names','day')]
-RF_cfu_all_rf <- randomForest(log_cfu ~ ., data = predict_cfu_all_df,
-                   importance = TRUE, ntree = n_trees)
-RF_cfu_all_df <- data.frame(cbind(CFU_community_by_day[!is.na(CFU_community_by_day$log_cfu), 
-                                                       c('day','log_cfu')], 
-                                  predicted = RF_cfu_all_rf$predicted))
-
-mse_all <- tail(RF_cfu_all_rf$mse, 1)
-rsq_all <- tail(RF_cfu_all_rf$rsq, 1)
-
-all_rf_plot <- ggplot(data = RF_cfu_all_df, aes(x = log_cfu, y = predicted)) + geom_point() +
-        labs(x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)') + theme_bw() + 
-        ggtitle("Random Forest Prediction of C Diff CFU") +
-        annotate('text', x=0.15*(max(RF_cfu_all_df$log_cfu)), y=0.95*(max(RF_cfu_all_df$predicted)), 
-                 label=paste('R^2 = ', round(tail(RF_cfu_all_rf$rsq, 1), 2),
-                             '\nMSE = ', round(tail(RF_cfu_all_rf$mse, 1), 2), sep = ''))
-
-importance_all <- sort(importance(RF_cfu_all_rf)[,1], decreasing = T)
-
-rsq_features_all <- data.frame('Features'= NA, 'RSQ'=NA, 'MSE'=NA)
-for (k in 1:25){
-    rf_partial <- randomForest(log_cfu ~ ., 
-                               data=predict_cfu_all_df[ , c( 'log_cfu', names(importance_all)[1:k])], 
-                               importance=TRUE, ntree=n_trees)
-    rsq_features_all <- rbind(rsq_features_all, c(k, tail(rf_partial$rsq, 1), tail(rf_partial$mse, 1)))
-}
-rsq_features_all <- rsq_features_all[-1,]
-all_r2feat_plot <- ggplot(rsq_features_all, aes(x=Features, y=RSQ)) + geom_line() +
-  labs(x = 'Features (n)', y = 'R^2') + scale_color_discrete(guide = FALSE) + theme_bw()
-
-important_OTU_all <- data.frame(head(importance_all, n_important))
-colnames(important_OTU_all) <- 'Importance'
-important_OTU_all$OTU <- rownames(important_OTU_all)
-
-all_feat_plot <- ggplot(data = important_OTU_all, aes(x = factor(OTU), y = Importance)) + 
-        geom_point() + scale_x_discrete(limits = rev(important_OTU_all$OTU)) +
-        labs(x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
-
-ggdraw()+
-  draw_plot(all_rf_plot, 0, 0.5, 0.5, 0.5) + 
-  draw_plot(all_r2feat_plot, 0, 0, 0.5, 0.5) +
-  draw_plot(all_feat_plot, 0.5, 0, 0.5, 1)
-
-```
-
-## Does removing the early euthanized samples improve prediction in days 1 to 3?
-
-```{r fig.height= 10, fig.width=10}
-
-without_early_euth <- meta_file[meta_file$Early_Euth == FALSE ,c('day', 'mouse_id', 'log_cfu')]
-without_early_euth <- without_early_euth[without_early_euth$day %in% c(1:4), ]
-
-without_by_day <- merge(without_early_euth, shared_X_cdiff_ra, by  = 'row.names')
-without_day_0 <- merge(without_early_euth, rel_abund_stool, by='mouse_id', by.y='row.names' )
-
-###################################
-
-importance_persist <- data.frame(matrix(nrow = length(OTU_list), 
-                                   ncol = max(without_early_euth$day, na.rm = T)),
-                            row.names = OTU_list)
-rsq_persist <- data.frame(matrix(nrow = 1, ncol = max(without_early_euth$day, na.rm = T)),
-                         row.names = 'RSQ')
-mse_persist <- data.frame(matrix(nrow = 1, ncol = max(without_early_euth$day, na.rm = T)),
-                         row.names = 'MSE')
-prediction_persist <- data.frame(matrix(nrow = 0, ncol = 4))
-features_persist <- data.frame('Day'=NA, 'Features (n)'= NA, 'RSQ'=NA, 'MSE'=NA)
-
-for (i in 1:4){
-  predict_all_df <- without_day_0[without_day_0$day == i ,]
-  predict_df <- predict_all_df[!is.na(predict_all_df$log_cfu), 
-                           !colnames(predict_all_df) %in% c('mouse_id', 'day')]
-  RF <- randomForest(log_cfu ~ ., data = predict_df,
-                   importance = TRUE, ntree = n_trees)
-  rsq_persist[i] <- RF$rsq[length(RF$rsq)]
-  mse_persist[i] <- RF$mse[length(RF$mse)]
-
-  # sort the features by their effect on % increase in the standard error
-  importance_persist[ , i] <- importance(RF)[,1]
-  importance_sorted <- sort(importance(RF)[,1], decreasing = T)
-  colnames(importance_persist)[i] <- paste('day', i, 'importance', sep = '_')
-# fit the model back on the datar and tack it to the end of the counts file
-  random_forest_output <- cbind(predict_all_df[!is.na(predict_all_df$log_cfu), c('day','log_cfu')],
-                              RF$predicted)
-  colnames(random_forest_output) <- c('day', 'observed', 'predicted')
-  prediction_persist <- rbind(prediction_persist, random_forest_output) 
-  for (k in 1:25){
-    n_features <- k
-    rf_partial <- randomForest(log_cfu ~ ., 
-                               data=predict_df[ , c( 'log_cfu', names(importance_sorted)[1:n_features])], 
-                               importance=TRUE, ntree=n_trees)
-    features_persist <- rbind(features_persist, c(i, k, tail(rf_partial$rsq, 1), tail(rf_partial$mse, 1)))
-  }
-}
-features_persist <- features_persist[-1,]
-
-day_0_persist_rf_plot <- ggplot(data = prediction_persist, aes(x=observed, y=predicted)) + geom_point() + 
-        facet_grid( ~ day) + 
-        labs(title = 'Random Forest Prediction of C Diff CFU dpi from Community on Day 0
-                      \nwithout mice euthanized early', x= 'Observed CFU (log10)', 
-             y = 'Predicted CFU (log10)')
-
-rsq_persist <- t(rsq_persist)
-rsq_persist <- as.data.frame(cbind(rsq_persist, day = c(1:4)))
-day_0_persist_rsq_plot <- ggplot(data = rsq_persist, aes(x = factor(day), y = RSQ)) + geom_point() + theme_bw() +
-        labs(x= 'Day', y = 'R^2') 
-#title = 'Variation Explained by Day of C Diff CFU dpi from Community on Day 0\nwithout mice euthanized early',
-
-mse_persist <- t(mse_persist)
-mse_persist <- as.data.frame(cbind(mse_persist, day = c(1:4)))
-day_0_persist_mse_plot <- ggplot(data = mse_persist, aes(x = factor(day), y = MSE)) + geom_point() + theme_bw() +
-        labs(x = 'Day', y = 'MSE')
-#title = 'MSE by Day of C Diff CFU dpi from Community on Day 0\nwithout mice euthanized early',
-
-day_0_persist_r2feat_plot <- ggplot(features_persist, aes(x=Features..n., y=RSQ, group=(Day), color=factor(Day))) + 
-  geom_line() + labs(x = 'Features (n)', y = 'R^2') + scale_color_discrete(guide = FALSE) + theme_bw()
-#title= 'R^2 from RF with top n features\nDay of C Diff CFU dpi from Community on Day 0\nwithout mice euthanized early',
-
-n_important <- 20
-median_importance <- names(head(sort((apply(importance_persist,1, median)), decreasing = T), n_important))
-med_import_df <- melt(cbind(rownames(importance_persist), importance_persist)[median_importance, ])
-colnames(med_import_df) <- c('OTU', 'Day', 'Importance')
-
-day_0_persist_feat_plot <- ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + 
-  facet_wrap( ~ OTU, nrow = 2) + geom_point() + scale_x_discrete(labels = c(1:10)) +
-        labs(title = 'OTU Importance in Prediction of C Diff CFU dpi from Community
-             on Day 0\nwithout mice euthanized early - (Top 20 Median OTUs)',
-        x= 'Day', y = '% Increase in MSE') + theme_bw()
-
-ggdraw()+
-  draw_plot(day_0_persist_rf_plot, 0, 0.5, 0.75, 0.5) + 
-  draw_plot(day_0_persist_rsq_plot, 0.875, 0.75, 0.125, 0.25) +
-  draw_plot(day_0_persist_mse_plot, 0.75, 0.75, 0.125, 0.25) + 
-  draw_plot(day_0_persist_r2feat_plot, 0.75, 0.5, 0.25, 0.25) + 
-  draw_plot(day_0_persist_feat_plot, 0, 0, 1, 0.5)
-
-importance_persist_day <- data.frame(matrix(nrow = sum(OTUs_01*1), 
-                                   ncol = max(without_early_euth$day, na.rm = T)),
-                            row.names = colnames(shared_X_cdiff_ra))
-rsq_persist_day <- data.frame(matrix(nrow = 1, ncol = max(without_early_euth$day, na.rm = T)),
-                         row.names = 'RSQ')
-mse_persist_day <- data.frame(matrix(nrow = 1, ncol = max(without_early_euth$day, na.rm = T)),
-                         row.names = 'MSE')
-prediction_persist_day <- data.frame(matrix(nrow = 0, ncol = 4))
-features_persist_day <- data.frame('Day'=NA, 'Features (n)'= NA, 'RSQ'=NA, 'MSE'=NA)
-
-for (i in 1:4){
-  predict_all_df <- without_by_day[without_by_day$day == i ,]
-  predict_df <- predict_all_df[!is.na(predict_all_df$log_cfu), 
-                           !colnames(predict_all_df) %in% c('mouse_id', 'Row.names', 'day')]
-  RF <- randomForest(log_cfu ~ ., data = predict_df,
-                   importance = TRUE, ntree = n_trees)
-  rsq_persist_day[i] <- RF$rsq[length(RF$rsq)]
-  mse_persist_day[i] <- RF$mse[length(RF$mse)]
-
-  # sort the features by their effect on % increase in the standard error
-  importance_persist_day[ , i] <- importance(RF)[,1]
-  importance_sorted <- sort(importance(RF)[,1], decreasing = T)
-  colnames(importance_persist_day)[i] <- paste('day', i, 'importance', sep = '_')
-# fit the model back on the datar and tack it to the end of the counts file
-  random_forest_output <- cbind(predict_all_df[!is.na(predict_all_df$log_cfu), c('day','log_cfu')],
-                              RF$predicted)
-  colnames(random_forest_output) <- c('day', 'observed', 'predicted')
-  prediction_persist_day <- rbind(prediction_persist_day, random_forest_output) 
-  for (k in 1:25){
-    n_features <- k
-    rf_partial <- randomForest(log_cfu ~ ., 
-                               data=predict_df[ , c( 'log_cfu', names(importance_sorted)[1:n_features])], 
-                               importance=TRUE, ntree=n_trees)
-    features_persist_day <- rbind(features_persist_day, c(i, k, tail(rf_partial$rsq, 1), tail(rf_partial$mse, 1)))
-  }
-}
-features_persist_day <- features_persist_day[-1,]
-
-day_persist_rf_plot <- ggplot(data = prediction_persist_day, aes(x=observed, y=predicted)) + geom_point() + 
-        facet_grid( ~ day) + 
-        labs(title = 'Random Forest Prediction of C Diff CFU dpi from Community on same day\nwithout mice euthanized early',
-        x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)')
-
-rsq_persist_day <- t(rsq_persist_day)
-rsq_persist_day <- as.data.frame(cbind(rsq_persist_day, day = c(1:4)))
-day_persist_rsq_plot <- ggplot(data = rsq_persist_day, aes(x = factor(day), y = RSQ)) + geom_point() + theme_bw() +
-        labs(x= 'Day', y = 'R^2')
-#title = 'Variation Explained by Day of C Diff CFU dpi from Community on same day\nwithout mice euthanized early',
-
-mse_persist_day <- t(mse_persist_day)
-mse_persist_day <- as.data.frame(cbind(mse_persist_day, day = c(1:4)))
-day_persist_mse_plot <- ggplot(data = mse_persist_day, aes(x = factor(day), y = MSE)) + geom_point() + theme_bw() +
-        labs(x= 'Day', y = 'MSE')
-#title = 'MSE by Day of C Diff CFU dpi from Community on same day\nwithout mice euthanized early',
-
-day_persist_r2feat_plot <- ggplot(features_persist_day, aes(x=Features..n., y=RSQ, group=(Day), color=factor(Day))) + geom_line() +
-  labs(x = 'Features (n)', y = 'R^2') + scale_color_discrete(guide  = FALSE) + theme_bw()
-# title= 'R^2 from RF with top n features\n Day of C Diff CFU dpi from Community on same day\nwithout mice euthanized early',
-
-n_important <- 20
-median_importance <- names(head(sort((apply(importance_persist_day,1, median)), decreasing = T), n_important))
-med_import_df <- melt(cbind(rownames(importance_persist_day), importance_persist_day)[median_importance, ])
-colnames(med_import_df) <- c('OTU', 'Day', 'Importance')
-
-day_persist_feat_plot <- ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + facet_wrap( ~ OTU, nrow = 2) +
-        geom_point() + scale_x_discrete(labels = c(1:10)) +
-        labs(title = 'OTU Importance in Prediction of C Diff CFU dpi from Community on same day\nwithout mice euthanized early - (Top 20 Median OTUs)',
-        x= 'Day', y = '% Increase in MSE') + theme_bw()
-
-ggdraw()+
-  draw_plot(day_persist_rf_plot, 0, 0.5, 0.75, 0.5) + 
-  draw_plot(day_persist_rsq_plot, 0.875, 0.75, 0.125, 0.25) +
-  draw_plot(day_persist_mse_plot, 0.75, 0.75, 0.125, 0.25) + 
-  draw_plot(day_persist_r2feat_plot, 0.75, 0.5, 0.25, 0.25) + 
-  draw_plot(day_persist_feat_plot, 0, 0, 1, 0.5)
-```
-
-#### All CFU compared to same day community without mice euthanized early 
-```{r fig.width=10, fig.height=5}
-
-predict_cfu_persist_df <- without_by_day[!is.na(without_by_day$log_cfu),
-                                           ! colnames(without_by_day) %in% c('Row.names','day','mouse_id')]
-RF_cfu_persist_rf <- randomForest(log_cfu ~ ., data = predict_cfu_persist_df,
-                   importance = TRUE, ntree = n_trees)
-RF_cfu_persist_df <- data.frame(cbind(without_by_day[!is.na(without_by_day$log_cfu), 
-                                                       c('day','log_cfu')], 
-                                  predicted = RF_cfu_persist_rf$predicted))
-
-mse_all <- tail(RF_cfu_persist_rf$mse, 1)
-rsq_all <- tail(RF_cfu_persist_rf$rsq, 1)
-
-all_persist_rf_plot <- ggplot(data = RF_cfu_persist_df, aes(x = log_cfu, y = predicted)) + geom_point() +
-        labs(x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)') + theme_bw() + 
-        ggtitle("Random Forest Prediction of C Diff CFU") +
-        annotate('text', x=0.15*(max(RF_cfu_persist_df$log_cfu)), y=0.95*(max(RF_cfu_persist_df$predicted)), 
-                 label=paste('R^2 = ', round(tail(RF_cfu_persist_rf$rsq, 1), 2),
-                             '\nMSE = ', round(tail(RF_cfu_persist_rf$mse, 1), 2), sep = ''))
-
-importance_all_persist <- sort(importance(RF_cfu_persist_rf)[,1], decreasing = T)
-
-rsq_features_all_persist <- data.frame('Features'= NA, 'RSQ'=NA, 'MSE'=NA)
-for (k in 1:25){
-    rf_partial <- randomForest(log_cfu ~ ., 
-                               data=predict_cfu_persist_df[ , c( 'log_cfu', names(importance_all_persist)[1:k])], 
-                               importance=TRUE, ntree=n_trees)
-    rsq_features_all_persist <- rbind(rsq_features_all_persist, c(k, tail(rf_partial$rsq, 1), tail(rf_partial$mse, 1)))
-}
-rsq_features_all_persist <- rsq_features_all_persist[-1,]
-all_persist_r2feat_plot <- ggplot(rsq_features_all_persist, aes(x=Features, y=RSQ)) + geom_line() +
-  labs(x = 'Features (n)', y = 'R^2') + scale_color_discrete(guide = FALSE) + theme_bw()
-
-important_OTU_all_persist <- data.frame(head(importance_all_persist, n_important))
-colnames(important_OTU_all_persist) <- 'Importance'
-important_OTU_all_persist$OTU <- rownames(important_OTU_all_persist)
-
-all_persist_feat_plot <- ggplot(data = important_OTU_all_persist, aes(x = factor(OTU), y = Importance)) + 
-        geom_point() + scale_x_discrete(limits = rev(important_OTU_all_persist$OTU)) +
-        labs(x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
-
-ggdraw()+
-  draw_plot(all_persist_rf_plot, 0, 0.5, 0.5, 0.5) + 
-  draw_plot(all_persist_r2feat_plot, 0, 0, 0.5, 0.5) +
-  draw_plot(all_persist_feat_plot, 0.5, 0, 0.5, 1)
-
-```
-
-Eliminating mice euthanized early from the RF model gives similar R^2 and MSE, bu there is a slight advantage to community OTU features of Day 0 to predict Day1 CFU.
-As well as the R^2 value increases with increasing features, whereas when all mice are used the day 1 R^2 only decreases with increasing features.
-Of note, accompaied by this is an increase in the % MSE attributed to OTU15 (Akkermansia), which has seemed to stand out in all other days/analysis.Interestingly, akkermansia does not appear to have the same relationship when compared to the same day cfu and community.
-This could suggest akkermansia is promoting the intial colonization of c difficile.
-At first glance, OTU135 (Coriobacteriaceae) seems to stand out for predicting cfu of the same day as well as from day 0.
-
-***
 
@@
 otu_euth_roc <- roc(Predict_early_euth_df$Euth_Early ~ otu_euth_probs[ , 2])
-
-#otu_euth_model <- rf_early_aucrf$RFopt
-#rf_early_aucrf$Kopt
-#rf_early_aucrf$Xopt
-#rf_early_aucrf$RFopt$confusion
-#OptimalSet(rf_early_aucrf)
-#plot(otu_euth_model_cv)
-#plot(otu_euth_model, which="ranking") 
-#plot(otu_euth_model, which="psel")
 

### 2016-09-19 code/HumanCdGF_microbiota_predict_by_day.Rmd

@@
 ```{r fig.width=10, fig.height=10}
-predictive_otus <- paste('Otu00', c('199','026','050','015','028',
-                                    '033','135','059','044'),    # lost with max rel abund > 10
+predictive_otus <- paste('Otu00', c('012','171','063','033','038',
+                                    '062','078','007','209'),    # lost with max rel abund > 10
                                     #'007', '014', '035', '005'), # gain with max rel abund > 10
@@
   ggplot( aes(x = reorder(factor(phylogeny), estimate), y = estimate)) + 
-  geom_bar(stat = 'identity', aes(fill = color)) +
+  geom_bar(stat = 'identity', aes(fill = color)) + coord_flip() + 
   theme(axis.text.x = element_text(angle = 90, hjust = 1)) +

### 2016-09-19 code/HumanCdGF_microbiota_predict.Rmd
remove AUCRF from OTU predict

@@
 ***
-
-##Can we predict moribundity with the community structure at Day 0?
-
-```{r fig.width=6, fig.height=5}
-meta_file$Euth_Early <- factor(ifelse(meta_file[ , 'Early_Euth'] == FALSE, 1, 0))
-
-Predict_early_euth_df <- merge(meta_file[meta_file$day == 0, c('mouse_id', 'Euth_Early')], 
-                               rel_abund_stool, by.x = 'mouse_id', by.y='row.names')
-Predict_early_euth_df_names <- Predict_early_euth_df$mouse_id
-Predict_early_euth_df <- select(Predict_early_euth_df, -mouse_id)
-
-rf_early_aucrf <- AUCRF(Euth_Early ~ ., data = Predict_early_euth_df,
-                              ntree = n_trees, pdel = 0.05, ranking = 'MDA')
-
-otu_euth_probs <- predict(rf_early_aucrf$RFopt, type = 'prob')
-otu_euth_roc <- roc(Predict_early_euth_df$Euth_Early ~ otu_euth_probs[ , 2])
-
-#otu_euth_model <- rf_early_aucrf$RFopt
-#rf_early_aucrf$Kopt
-#rf_early_aucrf$Xopt
-#rf_early_aucrf$RFopt$confusion
-#OptimalSet(rf_early_aucrf)
-#plot(otu_euth_model_cv)
-#plot(otu_euth_model, which="ranking") 
-#plot(otu_euth_model, which="psel")
-
-aucrf_data <- Predict_early_euth_df[,c('Euth_Early',rf_early_aucrf$Xopt)]
-
-#### 10-fold Cross validated - 10 repetitions
-start <- proc.time()[3]
-iters <- 100
-cv10f_aucs <- c()
-cv10f_all_resp <- c()
-cv10f_all_pred <- c()
-for(j in 1:iters){
-  sampling <- sample(1:nrow(aucrf_data),nrow(aucrf_data),replace=F)
-  cv10f_probs <- rep(NA,nrow(aucrf_data))
-  for(i in seq(1,nrow(aucrf_data),5)){
-    train <- aucrf_data[sampling[-(i:(i+4))],]
-    test <- aucrf_data[sampling[i:(i+4)],]
-    temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
-    cv10f_probs[sampling[i:(i+4)]] <- predict(temp_model$RFopt, test, type='prob')[,2]
-  }
-  cv10f_roc <- roc(aucrf_data$Euth_Early~cv10f_probs)
-  cv10f_all_pred <- c(cv10f_all_pred, cv10f_probs)
-  cv10f_all_resp <- c(cv10f_all_resp, aucrf_data$Euth_Early)
-  cv10f_aucs[j] <- cv10f_roc$auc
-}
-cv10f_roc <- roc(cv10f_all_resp~cv10f_all_pred)
-end <- proc.time()[3]
-time <- end-start
-
-#### leave-one-out
-LOO_probs <- NULL
-LOO_probs <- sapply(1:nrow(aucrf_data), function(i){
- train <- aucrf_data[-i,]
-  test <- aucrf_data[i,]
-  temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
-  LOO_probs[i] <- predict(temp_model$RFopt, test, type='prob')[,2]
-})
-LOO_roc <- roc(aucrf_data$Euth_Early~LOO_probs)
-
-#### plot rocs
-par(mar=c(4,4,1,1))
-plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='')
-plot(otu_euth_roc, col='black', lwd=2, add=T, lty=1)
-plot(cv10f_roc, col='gray', lwd=2, add=T, lty=1)
-plot(LOO_roc, col='blue', lwd=2, add=T, lty=2)
-mtext(side=2, text="Sensitivity", line=2.5)
-mtext(side=1, text="Specificity", line=2.5)
-legend('bottomright', legend=c(sprintf('Out-Of-Bag (AUC: %.3g)',otu_euth_roc$auc),
-                               sprintf('Leave-1-out (AUC: %.3g)',LOO_roc$auc),
-                               sprintf('10-fold CV (AUC: %.3g)',cv10f_roc$auc),
-                               sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_euth_roc,LOO_roc)$p.value),
-                               sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
-),lty=c(1,2,1,0,0), lwd=2, col=c('black','blue','gray'), bty='n')
-```
-
-```{r fig.width=10, fig.height=5}
-layout(matrix(c(1,3,2,2), nrow=2))
-
-#AUCRF curve
-par(mar=c(5,5,1.5,1))
-plot(rf_early_aucrf)
-#mtext('A', at=-90, font=2, side=3)
-
-#Importance Plot
-euth_otus <- rf_early_aucrf$Xopt
-euth_tax <- get_tax(1, euth_otus, tax_file)
-euth_tax$tax_label <- paste(euth_tax$tax, ' (', 
-                            gsub('tu0*', 'TU ', rownames(euth_tax)), 
-                                   ')', sep = '')
-dotchart(rev(rf_early_aucrf$ranking[euth_otus]), labels=rev(euth_tax[euth_otus, 'tax_label']), xlab='Mean Decrease Accuracy')
-#mtext('B', at=-0.014, font=2, side=3)
-
-#Abundance stripchart or most predictive otus
-euth_otus <- rf_early_aucrf$Xopt
-euth_otus <- rev(euth_otus[1:5])
-early_abunds <- Predict_early_euth_df[Predict_early_euth_df$Euth_Early==0, euth_otus]/10000 + 1e-4
-late_abunds <- Predict_early_euth_df[Predict_early_euth_df$Euth_Early==1, euth_otus]/10000 + 1e-4
-
-par(mar=c(4, 9, 1, 1))
-plot(1, type="n", ylim=c(0,length(euth_otus)*2), xlim=c(1e-4,1e-2), log="x", ylab="", xlab="Relative Abundance (%)", xaxt="n", yaxt="n")
-index <- 1
-for(i in euth_otus){
-  stripchart(at=index-0.35, jitter(early_abunds[,i], amount=1e-5), pch=21, bg="royalblue1", method="jitter", jitter=0.2, add=T, cex=1, lwd=0.5)
-  stripchart(at=index+0.35, jitter(late_abunds[,i], amount=1e-5), pch=21, bg="orange", method="jitter", jitter=0.2, add=T, cex=1, lwd=0.5)
-  segments(mean(early_abunds[,i]),index-0.7,mean(early_abunds[,i]),index, lwd=3)
-  segments(mean(late_abunds[,i]),index+0.7,mean(late_abunds[,i]),index, lwd=3)
-  index <- index + 2
-}
-axis(2, at=seq(1,index-2,2), labels=euth_tax[euth_otus, 'tax_label'], las=1, line=-0.5, tick=F, cex.axis=0.8)
-axis(1, at=c(1e-4, 1e-3, 1e-2), label=c("0", "0.1", "1"))
-legend('topright', legend=c("Persistant", "Susceptible"), pch=c(21, 21), pt.bg=c("orange","royalblue1"), cex=0.7)
-#mtext('C', at=1e-7, font=2, side=3)
-dev.off()
-
-```

### 2016-09-20 code/CdGF_AUCRF.Rmd
random validation of AUCRF

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
 ```
 
-```{r, message=FALSE}
+```{r}
 pack_used <- c('randomForest','miscTools','ggplot2','reshape2', 'pROC', 'knitr', 'cowplot','dplyr','AUCRF')
@@
 rel_abund_stool <- rel_abund_stool[ ,OTUs_stool_0_01]
-OTU_list <- colnames(rel_abund_stool)
 
@@
 
-```{r fig.width=6, fig.height=5}
+```{r fig.width=6, fig.height=5, results= 'hide'}
+#create dataframe for RF-classification
 meta_file$Euth_Early <- factor(ifelse(meta_file[ , 'Early_Euth'] == FALSE, 1, 0))
@@
 
+# create RF model
 rf_early_aucrf <- AUCRF(Euth_Early ~ ., data = Predict_early_euth_df,
                               ntree = n_trees, pdel = 0.05, ranking = 'MDA')
-
 otu_euth_probs <- predict(rf_early_aucrf$RFopt, type = 'prob')
 otu_euth_roc <- roc(Predict_early_euth_df$Euth_Early ~ otu_euth_probs[ , 2])
+random_roc <- roc(sample(c(1,0), length(otu_euth_probs[ , 2]), replace = T) ~ otu_euth_probs[ , 2])
 
@@
 
+#### leave-one-cage-out
+LOcO_df <- merge(meta_file[meta_file$day == 0, c('mouse_id', 'cage_id', 'Euth_Early')], 
+                               rel_abund_stool[ , rf_early_aucrf$Xopt], by.x = 'mouse_id', by.y='row.names')
+cage_list <- unique(LOcO_df$cage_id)
+
+LOcO_probs <- NULL
+LOcO_probs <- sapply(1:length(cage_list), function(i){
+  by_cage <- LOcO_df$cage_id == cage_list[i]
+  LOcO_predict <- select(LOcO_df, -mouse_id, -cage_id)
+  train <- LOcO_predict[-by_cage,]
+  test <- LOcO_predict[by_cage,]
+  temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
+  LOcO_probs[i] <- predict(temp_model$RFopt, test, type='prob')[,2]
+})
+LOcO_probs <- unlist(LOcO_probs)
+LOcO_roc <- roc(aucrf_data$Euth_Early~LOcO_probs)
+
+#### Random - 10-fold Cross validated - 10 repetitions
+ran_aucrf_data <- cbind(Euth_Early = factor(sample(Predict_early_euth_df$Euth_Early, 
+                                                     nrow(Predict_early_euth_df))),
+                             Predict_early_euth_df[,rf_early_aucrf$Xopt])
+ran_cv10f_aucs <- c()
+ran_cv10f_all_resp <- c()
+ran_cv10f_all_pred <- c()
+for(j in 1:iters){
+  sampling <- sample(1:nrow(ran_aucrf_data),nrow(ran_aucrf_data),replace=F)
+  ran_cv10f_probs <- rep(NA,nrow(ran_aucrf_data))
+  for(i in seq(1,nrow(ran_aucrf_data),5)){
+    train <- ran_aucrf_data[sampling[-(i:(i+4))],]
+    test <- ran_aucrf_data[sampling[i:(i+4)],]
+    temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
+    ran_cv10f_probs[sampling[i:(i+4)]] <- predict(temp_model$RFopt, test, type='prob')[,2]
+  }
+  ran_cv10f_roc <- roc(ran_aucrf_data$Euth_Early~ran_cv10f_probs)
+  ran_cv10f_all_pred <- c(ran_cv10f_all_pred, ran_cv10f_probs)
+  ran_cv10f_all_resp <- c(ran_cv10f_all_resp, ran_aucrf_data$Euth_Early)
+  ran_cv10f_aucs[j] <- ran_cv10f_roc$auc
+}
+ran_cv10f_roc <- roc(ran_cv10f_all_resp~ran_cv10f_all_pred)
+
+set.seed(seed)
+
 #### plot rocs
@@
 plot(LOO_roc, col='blue', lwd=2, add=T, lty=2)
+plot(LOcO_roc, col='red', lwd=2, add=T, lty=2)
+plot(ran_cv10f_roc, col='orange', lwd=3, add=T, lty=2)
 mtext(side=2, text="Sensitivity", line=2.5)
@@
 legend('bottomright', legend=c(sprintf('Out-Of-Bag (AUC: %.3g)',otu_euth_roc$auc),
-                               sprintf('Leave-1-out (AUC: %.3g)',LOO_roc$auc),
+                               sprintf('Leave-1 mouse-out (AUC: %.3g)',LOO_roc$auc),
+                               sprintf('Leave-1 cage-out (AUC: %.3g)',LOcO_roc$auc),
+                               sprintf('10-fold CV (AUC: %.3g)',cv10f_roc$auc),
+                               sprintf('Random (AUC: %.3g)',ran_cv10f_roc$auc),
+                               sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_euth_roc,LOO_roc)$p.value),
+                               sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
+),lty=c(1,2,2,1,3,0,0), lwd=2, col=c('black','blue','red','gray','orange'), bty='n')
+```
+
+```{r fig.width=10, fig.height=8}
+layout(matrix(c(1,1,4,4,
+                2,2,3,3,
+                2,2,3,3), nrow=3, byrow = TRUE))
+
+#AUCRF curve
+par(mar=c(5,5,1.5,1))
+plot(rf_early_aucrf, ylim = c(0.7, 1.1))
+#mtext('A', at=-90, font=2, side=3)
+
+#Importance Plot
+euth_otus <- rf_early_aucrf$Xopt
+euth_tax <- get_tax(1, euth_otus, tax_file)
+euth_tax$tax_label <- paste(euth_tax$tax, ' (', 
+                            gsub('tu0*', 'TU ', rownames(euth_tax)), 
+                                   ')', sep = '')
+dotchart(rev(rf_early_aucrf$ranking[euth_otus]), labels=rev(euth_tax[euth_otus, 'tax_label']), xlab='Mean Decrease Accuracy')
+#mtext('B', at=-0.014, font=2, side=3)
+
+#Abundance stripchart or most predictive otus
+euth_otus <- rev(rf_early_aucrf$Xopt)
+early_abunds <- Predict_early_euth_df[Predict_early_euth_df$Euth_Early==0, euth_otus]/10000 + 1e-4
+late_abunds <- Predict_early_euth_df[Predict_early_euth_df$Euth_Early==1, euth_otus]/10000 + 1e-4
+
+par(mar=c(4, 9, 1, 1))
+plot(1, type="n", ylim=c(0,length(euth_otus)*2), xlim=c(1e-4,1e-2), log="x", ylab="", xlab="Relative Abundance (%)", xaxt="n", yaxt="n")
+index <- 1
+for(i in euth_otus){
+  stripchart(at=index-0.35, jitter(early_abunds[,i], amount=1e-5), pch=21, bg="royalblue1", method="jitter", jitter=0.2, add=T, cex=1, lwd=0.5)
+  stripchart(at=index+0.35, jitter(late_abunds[,i], amount=1e-5), pch=21, bg="orange", method="jitter", jitter=0.2, add=T, cex=1, lwd=0.5)
+  segments(mean(early_abunds[,i]),index-0.7,mean(early_abunds[,i]),index, lwd=3)
+  segments(mean(late_abunds[,i]),index+0.7,mean(late_abunds[,i]),index, lwd=3)
+  index <- index + 2
+}
+axis(2, at=seq(1,index-2,2), labels=euth_tax[euth_otus, 'tax_label'], las=1, line=-0.5, tick=F, cex.axis=0.8)
+axis(1, at=c(1e-4, 1e-3, 1e-2), label=c("0", "0.1", "1"))
+legend('topright', legend=c("Mild", "Severe"), pch=c(21, 21), pt.bg=c("orange","royalblue1"), cex=0.7)
+#mtext('C', at=1e-7, font=2, side=3)
+```
+
+***  
+***
+
+### Validate AUCRF with Randomizing
+
+```{r Random_validation, fig.width=6, fig.height=5, results= 'hide'}
+#create dataframe for RF-classification
+Euth_Early <- factor(sample(c(1,0), size = nrow(rel_abund_stool), replace = T))
+Validate_early_euth_df <- cbind(Euth_Early, rel_abund_stool)
+
+# create RF model
+rf_early_aucrf <- AUCRF(Euth_Early ~ ., data = Validate_early_euth_df,
+                              ntree = n_trees, pdel = 0.05, ranking = 'MDA')
+
+otu_euth_probs <- predict(rf_early_aucrf$RFopt, type = 'prob')
+otu_euth_roc <- roc(Validate_early_euth_df$Euth_Early ~ otu_euth_probs[ , 2])
+
+aucrf_data <- Validate_early_euth_df[,c('Euth_Early',rf_early_aucrf$Xopt)]
+
+#### 10-fold Cross validated - 10 repetitions
+start <- proc.time()[3]
+iters <- 100
+cv10f_aucs <- c()
+cv10f_all_resp <- c()
+cv10f_all_pred <- c()
+for(j in 1:iters){
+  sampling <- sample(1:nrow(aucrf_data),nrow(aucrf_data),replace=F)
+  cv10f_probs <- rep(NA,nrow(aucrf_data))
+  for(i in seq(1,nrow(aucrf_data),5)){
+    train <- aucrf_data[sampling[-(i:(i+4))],]
+    test <- aucrf_data[sampling[i:(i+4)],]
+    temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
+    cv10f_probs[sampling[i:(i+4)]] <- predict(temp_model$RFopt, test, type='prob')[,2]
+  }
+  cv10f_roc <- roc(aucrf_data$Euth_Early~cv10f_probs)
+  cv10f_all_pred <- c(cv10f_all_pred, cv10f_probs)
+  cv10f_all_resp <- c(cv10f_all_resp, aucrf_data$Euth_Early)
+  cv10f_aucs[j] <- cv10f_roc$auc
+}
+cv10f_roc <- roc(cv10f_all_resp~cv10f_all_pred)
+end <- proc.time()[3]
+time <- end-start
+
+#### leave-one-out
+LOO_probs <- NULL
+LOO_probs <- sapply(1:nrow(aucrf_data), function(i){
+ train <- aucrf_data[-i,]
+  test <- aucrf_data[i,]
+  temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
+  LOO_probs[i] <- predict(temp_model$RFopt, test, type='prob')[,2]
+})
+LOO_roc <- roc(aucrf_data$Euth_Early~LOO_probs)
+
+#### leave-one-cage-out
+LOcO_df <- cbind(cage_id = sample(meta_file[meta_file$day %in% 0, 'cage_id']),
+                 cbind(Euth_Early, rel_abund_stool[ , rf_early_aucrf$Xopt]))
+cage_list <- unique(LOcO_df$cage_id)
+
+LOcO_probs <- NULL
+LOcO_probs <- sapply(1:length(cage_list), function(i){
+  by_cage <- LOcO_df$cage_id == cage_list[i]
+  LOcO_predict <- select(LOcO_df, -cage_id)
+  train <- LOcO_predict[-by_cage,]
+  test <- LOcO_predict[by_cage,]
+  temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
+  LOcO_probs[i] <- predict(temp_model$RFopt, test, type='prob')[,2]
+})
+LOcO_probs <- unlist(LOcO_probs)
+LOcO_roc <- roc(aucrf_data$Euth_Early~LOcO_probs)
+
+#### plot rocs
+par(mar=c(4,4,1,1))
+plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='')
+plot(otu_euth_roc, col='black', lwd=2, add=T, lty=1)
+plot(cv10f_roc, col='gray', lwd=2, add=T, lty=1)
+plot(LOO_roc, col='blue', lwd=2, add=T, lty=2)
+plot(LOcO_roc, col='red', lwd=2, add=T, lty=2)
+mtext(side=2, text="Sensitivity", line=2.5)
+mtext(side=1, text="Specificity", line=2.5)
+legend('bottomright', legend=c(sprintf('Out-Of-Bag (AUC: %.3g)',otu_euth_roc$auc),
+                               sprintf('Leave-1 mouse-out (AUC: %.3g)',LOO_roc$auc),
+                               sprintf('Leave-1 cage-out (AUC: %.3g)',LOcO_roc$auc),
                                sprintf('10-fold CV (AUC: %.3g)',cv10f_roc$auc),
@@
                                sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
-),lty=c(1,2,1,0,0), lwd=2, col=c('black','blue','gray'), bty='n')
+),lty=c(1,2,1,2,0,0), lwd=2, col=c('black','blue','gray','red'), bty='n')
 ```
 
-```{r fig.width=10, fig.height=5}
-layout(matrix(c(1,3,2,2), nrow=2))
+```{r Random_valid OTUs, fig.width=10, fig.height=8}
+layout(matrix(c(1,1,4,4,
+                2,2,3,3,
+                2,2,3,3), nrow=3, byrow = TRUE))
 
@@
 #Abundance stripchart or most predictive otus
-euth_otus <- rf_early_aucrf$Xopt
-euth_otus <- rev(euth_otus[1:5])
-early_abunds <- Predict_early_euth_df[Predict_early_euth_df$Euth_Early==0, euth_otus]/10000 + 1e-4
-late_abunds <- Predict_early_euth_df[Predict_early_euth_df$Euth_Early==1, euth_otus]/10000 + 1e-4
+euth_otus <- rev(rf_early_aucrf$Xopt)
+early_abunds <- Validate_early_euth_df[Validate_early_euth_df$Euth_Early==0, euth_otus]/10000 + 1e-4
+late_abunds <- Validate_early_euth_df[Validate_early_euth_df$Euth_Early==1, euth_otus]/10000 + 1e-4
 
@@
 axis(1, at=c(1e-4, 1e-3, 1e-2), label=c("0", "0.1", "1"))
-legend('topright', legend=c("Persistant", "Susceptible"), pch=c(21, 21), pt.bg=c("orange","royalblue1"), cex=0.7)
+legend('topright', legend=c("Mild", "Severe"), pch=c(21, 21), pt.bg=c("orange","royalblue1"), cex=0.7)
 #mtext('C', at=1e-7, font=2, side=3)
-dev.off()
 

### 2016-09-20 code/CdGF_AUCRF.Rmd
add cfu invsimp and entero to AUCRF

@@
 tax_function <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R'
+dist_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_all.an.0.03.subsample.groups.summary'
+entero_file <- '~/Desktop/dmm/human_CdGF.0.03.subsample.shared.for.0.03.dmm.2.mix.posterior'
 
@@
 tax_file <- read.table(tax_file, sep = '\t', header = T, row.names = 'OTU')
+dist_file <- read.table(dist_file, sep = '\t', header = T, row.names = 'group')
+entero_file <- read.table(entero_file, sep = '\t', header = T, row.names = 'X')
 source(tax_function)
@@
 
-Predict_early_euth_df <- merge(meta_file[meta_file$day == 0, c('mouse_id', 'Euth_Early')], 
-                               rel_abund_stool, by.x = 'mouse_id', by.y='row.names')
-Predict_early_euth_df_names <- Predict_early_euth_df$mouse_id
-Predict_early_euth_df <- select(Predict_early_euth_df, -mouse_id)
+rownames(entero_file) <- gsub('_','-', rownames(entero_file))
+dist_file <- dist_file[grep('D0', rownames(dist_file)), ]
+rownames(dist_file) <- gsub('D0', '-D0', rownames(dist_file))
+rownames(dist_file) <- gsub('--', '-', rownames(dist_file))
+
+Predict_df <- 
+  merge(
+    merge(
+      merge(
+        meta_file[meta_file$day %in% 2, c('mouse_id', 'log_cfu')],
+        merge(meta_file[meta_file$day %in% 0, c('cage_id', 'mouse_id', 'Euth_Early')],
+              entero_file, by = 'row.names'), by = 'mouse_id'), 
+      dist_file, by.x = 'Row.names', by.y = 'row.names'),
+    rel_abund_stool, by.x = 'mouse_id', by.y = 'row.names')
+
+Predict_early_euth_df <- select(Predict_df, -cage_id, 
+                                -mouse_id, -Row.names, -Partition_2,
+                                -label, -X, -invsimpson_lci, -invsimpson_hci)
 
@@
 #### 10-fold Cross validated - 10 repetitions
-start <- proc.time()[3]
 iters <- 100
@@
   cv10f_probs <- rep(NA,nrow(aucrf_data))
-  for(i in seq(1,nrow(aucrf_data),5)){
+  for(i in seq(1, 40,4)){
     train <- aucrf_data[sampling[-(i:(i+4))],]
@@
 cv10f_roc <- roc(cv10f_all_resp~cv10f_all_pred)
-end <- proc.time()[3]
-time <- end-start
 
@@
 #### leave-one-cage-out
-LOcO_df <- merge(meta_file[meta_file$day == 0, c('mouse_id', 'cage_id', 'Euth_Early')], 
-                               rel_abund_stool[ , rf_early_aucrf$Xopt], by.x = 'mouse_id', by.y='row.names')
-cage_list <- unique(LOcO_df$cage_id)
+LOcO_df <- Predict_df %>% 
+                select(Euth_Early, cage_id, one_of(rf_early_aucrf$Xopt))
+cage_list <- as.character(unique(LOcO_df$cage_id))
 
@@
   by_cage <- LOcO_df$cage_id == cage_list[i]
-  LOcO_predict <- select(LOcO_df, -mouse_id, -cage_id)
+  LOcO_predict <- select(LOcO_df, -cage_id)
   train <- LOcO_predict[-by_cage,]
@@
   ran_cv10f_probs <- rep(NA,nrow(ran_aucrf_data))
-  for(i in seq(1,nrow(ran_aucrf_data),5)){
+  for(i in seq(1,40,4)){
     train <- ran_aucrf_data[sampling[-(i:(i+4))],]
@@
 ran_cv10f_roc <- roc(ran_cv10f_all_resp~ran_cv10f_all_pred)
-
-set.seed(seed)
 
@@
 
+Would it help to add community information such as diversity or enterotype?  
+Would adding initial increase/slope of CFU or day 1 CFU help predict?
+But how to show since model is already approaching 1?  
+* Rerun AUCRF with added features and see if they replace any OTUs
+
 ***  
@@
 
-### Validate AUCRF with Randomizing
+### Validate AUCRF with Randomized Input
 
@@
 #create dataframe for RF-classification
-Euth_Early <- factor(sample(c(1,0), size = nrow(rel_abund_stool), replace = T))
-Validate_early_euth_df <- cbind(Euth_Early, rel_abund_stool)
+Validate_early_euth_df <- Predict_early_euth_df
+Validate_early_euth_df$Euth_Early <- sample(Validate_early_euth_df$Euth_Early)
 
@@
   cv10f_probs <- rep(NA,nrow(aucrf_data))
-  for(i in seq(1,nrow(aucrf_data),5)){
+  for(i in seq(1,40,4)){
     train <- aucrf_data[sampling[-(i:(i+4))],]
@@
 #### leave-one-cage-out
-LOcO_df <- cbind(cage_id = sample(meta_file[meta_file$day %in% 0, 'cage_id']),
-                 cbind(Euth_Early, rel_abund_stool[ , rf_early_aucrf$Xopt]))
-cage_list <- unique(LOcO_df$cage_id)
+LOcO_df <- cbind(cage_id = sample(meta_file[meta_file$day %in% 0, 'cage_id'],
+                                  nrow(Validate_early_euth_df)),
+                 Validate_early_euth_df[ , c('Euth_Early', rf_early_aucrf$Xopt)])
+cage_list <- as.character(unique(LOcO_df$cage_id))
 
@@
   temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
-  LOcO_probs[i] <- predict(temp_model$RFopt, test, type='prob')[,2]
+  LOcO_probs[[i]] <- predict(temp_model$RFopt, test, type='prob')[,2]
 })
@@
                                sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
-),lty=c(1,2,1,2,0,0), lwd=2, col=c('black','blue','gray','red'), bty='n')
+),lty=c(1,2,2,1,0,0), lwd=2, col=c('black','blue','red','gray'), bty='n')
 ```

### 2016-09-21 code/CdGF_AUCRF.Rmd
remove invsimp,cfu,entero from AUCRF

--- 2016-09-20 code/CdGF_AUCRF.Rmd
@@
 
-Predict_early_euth_df <- merge(meta_file[meta_file$day == 0, c('mouse_id', 'Euth_Early')], 
-                               rel_abund_stool, by.x = 'mouse_id', by.y='row.names')
-Predict_early_euth_df_names <- Predict_early_euth_df$mouse_id
-Predict_early_euth_df <- select(Predict_early_euth_df, -mouse_id)
+Predict_df <- merge(meta_file[meta_file$day %in% 0, 
+                              c('cage_id', 'mouse_id', 'Euth_Early')],
+                    rel_abund_stool, by.x = 'mouse_id', by.y = 'row.names')
+
+Predict_early_euth_df <- select(Predict_df, -cage_id, -mouse_id)
 
@@
 
-set.seed(seed)
-
 #### plot rocs
@@
 ```{r fig.width=10, fig.height=8}
-layout(matrix(c(1,1,4,4,
-                2,2,3,3,
-                2,2,3,3), nrow=3, byrow = TRUE))
+layout(matrix(c(1,1,1,1,4,4,4,
+                2,2,2,5,3,3,3,
+                2,2,2,5,3,3,3), nrow=3, byrow = TRUE))
 
@@
 
+Would adding initial increase/slope of CFU or day 1 CFU help predict?
+
+* Adding CFU decreases AUC and improves Random model, likely due to samples lost due to lack of cfu data which are biased towards the severe samples
+    - Day 1 CFU loses 2 samples  [AUCRF model with Day 1 CFU](~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/CdGF_AUCRF_withCFU_day1.html)
+    - Day 2 cfu loses 14 samples, 10 are severe  [AUCRF model with Day 2 CFU](~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/CdGF_AUCRF_withCFU_day2.html)
+
+Would it help to add community information such as diversity or enterotype?
+
+* When adding diversity and CFU, invsimpson and cfu appear as features, but when cfu is removed, invsimpson is no longer a predictor
+
+But how to show since model is already approaching 1?
+
+* Rerun AUCRF with added features and see if they replace any OTUs
+
 ***  
@@
 ```{r Random_valid OTUs, fig.width=10, fig.height=8}
-layout(matrix(c(1,1,4,4,
-                2,2,3,3,
-                2,2,3,3), nrow=3, byrow = TRUE))
+layout(matrix(c(1,1,1,1,4,4,4,
+                2,2,2,5,3,3,3,
+                2,2,2,5,3,3,3), nrow=3, byrow = TRUE))
 

### 2016-09-22 code/thetayc_time.R
updated theta YC graph code

@@ -3,11 +3,51 @@
 # thetayc_time.R
 #
 #Plot thetaYC over time distance from day 0 
-#
+#thetaYC distances cant have sample names that are just numbers, this script wont work 
 #    
 #
 ###################
 
+source('code/read.dist.R')
+
 #read in files
 meta_file <- read.table('data/process/human_CdGF_metadata.txt', sep='\t',header = T, row.names = 2)
 shared <- read.table('data/process/human_CdGF.an.unique_list.0.03.subsample.shared', sep='\t',header=T)
+theta <- read.dist(file='data/process/shared_group.thetayc.0.03.lt.dist', input="lt", make.square=T, diag=0)
+
+#then maybe write it as a function 
+
+#################THE RIGHT ONE!!!#######
+#compares day0 to every subsequent day. this asks, how different are the communities from the start? can this be linked to outcome?
+
+mouseIDVector <- as.character(unique(meta_file$mouse_id))
+mouseIDVector <- mouseIDVector[mouseIDVector != "inoculum_NA"]
+stability <- as.data.frame(matrix(ncol=10, nrow=length(mouseIDVector)))
+row.names(stability) <- mouseIDVector
+for (k in 1:10){
+  sampA <- rownames(meta_file)[meta_file$day %in% 0]
+  sampB <- rownames(meta_file)[meta_file$day %in% k]
+  dist <- theta[sampA, sampB]
+  row.names(dist) <- meta_file$mouse_id[rownames(meta_file) %in% rownames(dist)]
+  names(dist) <- meta_file$mouse_id[rownames(meta_file) %in% names(dist)]
+  for (i in mouseIDVector){
+    coord <- dist[i,i]
+    try(
+    stability[i,k] <- as.numeric(coord), silent = TRUE
+    )
+  }
+}
+
+#plot
+
+
+
+
+
+#build as function 
+
+
+#now write version where you compare each day to previous day. this asks question: how stable is community over time?
+
+
+

### 2016-09-22 code/CdGF_AUCRF.Rmd
improve random validation with boruta

@@
 #### Random - 10-fold Cross validated - 10 repetitions
-ran_aucrf_data <- cbind(Euth_Early = factor(sample(Predict_early_euth_df$Euth_Early, 
-                                                     nrow(Predict_early_euth_df))),
-                             Predict_early_euth_df[,rf_early_aucrf$Xopt])
+
 ran_cv10f_aucs <- c()
@@
 for(j in 1:iters){
+  set.seed(j)
+  ran_aucrf_data <- cbind(Euth_Early = factor(sample(Predict_early_euth_df$Euth_Early, 
+                                                     nrow(Predict_early_euth_df))),
+                             Predict_early_euth_df[,rf_early_aucrf$Xopt])
+  set.seed(seed)
   sampling <- sample(1:nrow(ran_aucrf_data),nrow(ran_aucrf_data),replace=F)
@@
 
+```{r}
+
+# boruta feature selection
+random_boruta <- read.table('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/random_rf_features_boruta.txt', sep = '\t')
+random_boruta <- random_boruta[, c('normHits', 'decision')]
+random_boruta$OTU <- rownames(random_boruta)
+```
+
+When randomly redistributing the Early Euthnization classification to the Day 0 commnunities, Boruta does not select any OTUs  
+Summing the OTU decisions from 100 iterations of randomly assigning classifications run in Boruta with 100 iterations produces:
+`r summary(random_boruta$decision)`
+Of all the Confirmed OTUs there are `r length(unique(random_boruta$OTU[random_boruta$decision == 'Confirmed']))` unique OTUs
+Of all the Tentative OTUs there are `r length(unique(random_boruta$OTU[random_boruta$decision == 'Tentative']))` unique OTUs
+Furthermore, there is no overlap in the identified OTUs, since there are `r length(unique(random_boruta$OTU[random_boruta$decision %in% c('Tentative', 'Confirmed')]))` unique OTUs.  
+Thus, no OTUs are able to create a model with out the specific pattern of observed disease severity.
+
 ```{r Random_validation, fig.width=6, fig.height=5, results= 'hide'}
@@
 #### leave-one-cage-out
-LOcO_df <- cbind(cage_id = sample(meta_file[meta_file$day %in% 0, 'cage_id'],
-                                  nrow(Validate_early_euth_df)),
+LOcO_df <- cbind(cage_id = Predict_df$cage_id,
                  Validate_early_euth_df[ , c('Euth_Early', rf_early_aucrf$Xopt)])

### 2016-09-22 code/random_rf_features_boruta.R

@@ -1,0 +1,61 @@
+pack_used <- c('dplyr','Boruta')
+for (dep in pack_used){
+  if (dep %in% installed.packages()[,"Package"] == FALSE){
+    install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
+                     quiet=TRUE);
+  }
+  library(dep, verbose=FALSE, character.only=TRUE)
+}
+
+meta_file   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
+shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
+
+# read in files
+meta_file   <- read.table(meta_file, sep = '\t', header = T, row.names = 'sample_id')
+shared_file <- read.table(shared_file, sep = '\t', header = T, row.names = 'sample_id')
+
+# select only mice with 431 cdiff
+meta_file <- meta_file[meta_file$cdiff_strain == '431', ]
+
+# subset shared
+#Create df with relative abundances
+stool_shared <- shared_file[rownames(meta_file[meta_file$day %in% 0, ]), ]
+# use %in% instead of == because to ignore NA
+# rename day 0 community rows by mouse id - since meta rows have day in row names, use mouse id for both
+rownames(stool_shared) <- meta_file$mouse_id[rownames(meta_file) %in% 
+                                               rownames(stool_shared) ]
+rel_abund_stool <- 100 * stool_shared / unique(apply(stool_shared, 1, sum))
+#Create vector of OTUs with median abundances >1%
+OTUs_stool_0_01 <- apply(rel_abund_stool, 2, max) > 1
+#df of OTUs with abundances >1% - by cage and inoculum
+rel_abund_stool <- rel_abund_stool[ ,OTUs_stool_0_01]
+
+#random forest regression model to determine Day 0 OTU correlated with increased CFU
+seed <- 18
+n_trees <- 2001
+set.seed(seed)
+
+meta_file$Euth_Early <- factor(ifelse(meta_file[ , 'Early_Euth'] == FALSE, 1, 0)) 
+
+Predict_df <- merge(meta_file[meta_file$day %in% 0, 
+                              c('cage_id', 'mouse_id', 'Euth_Early')],
+                    rel_abund_stool, by.x = 'mouse_id', by.y = 'row.names')
+
[…]

### 2016-09-23 code/random_rf_features_boruta.R
fix interative boruta

@@ -31,9 +31,9 @@ OTUs_stool_0_01 <- apply(rel_abund_stool, 2, max) > 1
 rel_abund_stool <- rel_abund_stool[ ,OTUs_stool_0_01]
 
 #random forest regression model to determine Day 0 OTU correlated with increased CFU
-seed <- 18
+seed <- -592166825
 n_trees <- 2001
-set.seed(seed)
+iters <- 100
 
 meta_file$Euth_Early <- factor(ifelse(meta_file[ , 'Early_Euth'] == FALSE, 1, 0)) 
 
@@ -45,16 +45,18 @@ Predict_early_euth_df <- select(Predict_df, -cage_id, -mouse_id)
 
 #create dataframe for RF-classification
 Validate_early_euth_df <- Predict_early_euth_df
+set.seed(seed)
 Validate_early_euth_df$Euth_Early <- sample(Validate_early_euth_df$Euth_Early)
 
 # boruta feature selection
-
 random_model_features <- c()
+
 for(i in 1:iters){
-  set.seed(i)
-  Validate_early_euth_df$Euth_Early <- sample(Validate_early_euth_df$Euth_Early)
-  random_bor <- Boruta(Euth_Early ~ ., data = Validate_early_euth_df)
-  random_model_features[[i]] <- attStats(random_bor)
+  set.seed(seed)
+  Boruta(Euth_Early ~ ., data = Validate_early_euth_df)
+  feature_df <- attStats(random_bor)
+  feature_df$OTU <- rownames(feature_df)
+  random_model_features[[i]] <- feature_df
 }
 random_model_features <- do.call('rbind', random_model_features)
 

### 2016-09-23 code/CdGF_AUCRF.Rmd
refine line plots for aucrf

--- 2016-09-21 code/CdGF_AUCRF.Rmd
@@
 #random forest regression model to determine Day 0 OTU correlated with increased CFU
-seed <- 18
+seed <- 1
 n_trees <- 2001
-set.seed(seed)
+iters <- 100
 
@@
 # create RF model
+set.seed(seed)
 rf_early_aucrf <- AUCRF(Euth_Early ~ ., data = Predict_early_euth_df,
@@
 otu_euth_roc <- roc(Predict_early_euth_df$Euth_Early ~ otu_euth_probs[ , 2])
+set.seed(seed)
 random_roc <- roc(sample(c(1,0), length(otu_euth_probs[ , 2]), replace = T) ~ otu_euth_probs[ , 2])
@@
 #### 10-fold Cross validated - 10 repetitions
-iters <- 100
 cv10f_aucs <- c()
@@
 for(j in 1:iters){
+  set.seed(j)
   sampling <- sample(1:nrow(aucrf_data),nrow(aucrf_data),replace=F)
@@
     test <- aucrf_data[sampling[i:(i+4)],]
+    set.seed(seed)
     temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
@@
   test <- aucrf_data[i,]
+  set.seed(seed)
   temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
@@
   test <- LOcO_predict[by_cage,]
+  set.seed(seed)
   temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
@@
 #### Random - 10-fold Cross validated - 10 repetitions
-ran_aucrf_data <- cbind(Euth_Early = factor(sample(Predict_early_euth_df$Euth_Early, 
+
+set.seed(seed)
+ran_outcome_data <- cbind(Euth_Early = factor(sample(Predict_early_euth_df$Euth_Early,  
                                                      nrow(Predict_early_euth_df))), 
                              Predict_early_euth_df[,rf_early_aucrf$Xopt]) 
+predict_random <-  predict(rf_early_aucrf$RFopt, ran_outcome_data, type = 'prob')[,2]
+predict_random_roc <- roc(ran_outcome_data$Euth_Early ~ predict_random)
+
 ran_cv10f_aucs <- c()
@@
 for(j in 1:iters){
-  sampling <- sample(1:nrow(ran_aucrf_data),nrow(ran_aucrf_data),replace=F)
-  ran_cv10f_probs <- rep(NA,nrow(ran_aucrf_data))
-  for(i in seq(1,nrow(ran_aucrf_data),5)){
-    train <- ran_aucrf_data[sampling[-(i:(i+4))],]
-    test <- ran_aucrf_data[sampling[i:(i+4)],]
+  set.seed(j)
+  sampling <- sample(1:nrow(aucrf_data),nrow(aucrf_data),replace=F)
+  ran_cv10f_probs <- rep(NA,nrow(aucrf_data))
+  for(i in seq(1,nrow(aucrf_data),5)){
+    train <- aucrf_data[sampling[-(i:(i+4))],]
+    test <- ran_outcome_data[sampling[i:(i+4)],]
+    set.seed(seed)
     temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
@@
   }
-  ran_cv10f_roc <- roc(ran_aucrf_data$Euth_Early~ran_cv10f_probs)
+  ran_cv10f_roc <- roc(ran_outcome_data$Euth_Early~ran_cv10f_probs)
   ran_cv10f_all_pred <- c(ran_cv10f_all_pred, ran_cv10f_probs)
-  ran_cv10f_all_resp <- c(ran_cv10f_all_resp, ran_aucrf_data$Euth_Early)
+  ran_cv10f_all_resp <- c(ran_cv10f_all_resp, ran_outcome_data$Euth_Early)
   ran_cv10f_aucs[j] <- ran_cv10f_roc$auc
@@
 plot(LOO_roc, col='blue', lwd=2, add=T, lty=2)
-plot(LOcO_roc, col='red', lwd=2, add=T, lty=2)
-plot(ran_cv10f_roc, col='orange', lwd=3, add=T, lty=2)
+plot(LOcO_roc, col='purple', lwd=2, add=T, lty=3)
+plot(ran_cv10f_roc, col='orange', lwd=2, add=T, lty=1)
+plot(predict_random_roc, col='red', lwd=2, add=T, lty=3)
 mtext(side=2, text="Sensitivity", line=2.5)
@@
 
-* When adding diversity and CFU, invsimpson and cfu appear as features, but when cfu is removed, invsimpson is no longer a predictor
+* When adding enterotype, diversity, and CFU, invsimpson and cfu appear as features, but when cfu is removed, invsimpson is no longer a predictor
 
@@
 ```{r Random_validation, fig.width=6, fig.height=5, results= 'hide'}
-#create dataframe for RF-classification
+
+# boruta feature selection
+#set.seed(seed)
+#Boruta(Euth_Early ~ ., data = Validate_early_euth_df, maxRuns = 200)
+
+#create dataframe for random RF-classification
 Validate_early_euth_df <- Predict_early_euth_df
-Validate_early_euth_df$Euth_Early <- sample(Validate_early_euth_df$Euth_Early)
+set.seed(seed)
+Validate_early_euth_df$Euth_Early <- sample(Predict_early_euth_df$Euth_Early)
 
 # create RF model
-rf_early_aucrf <- AUCRF(Euth_Early ~ ., data = Validate_early_euth_df,
+set.seed(seed)
+rf_random_aucrf <- AUCRF(Euth_Early ~ ., data = Validate_early_euth_df,
                               ntree = n_trees, pdel = 0.05, ranking = 'MDA')
+otu_random_probs <- predict(rf_random_aucrf$RFopt, type = 'prob')
+otu_random_roc <- roc(Validate_early_euth_df$Euth_Early ~ otu_random_probs[ , 2])
 
-otu_euth_probs <- predict(rf_early_aucrf$RFopt, type = 'prob')
-otu_euth_roc <- roc(Validate_early_euth_df$Euth_Early ~ otu_euth_probs[ , 2])
-
-aucrf_data <- Validate_early_euth_df[,c('Euth_Early',rf_early_aucrf$Xopt)]
+random_aucrf_data <- Validate_early_euth_df[,c('Euth_Early',rf_random_aucrf$Xopt)]
 
 #### 10-fold Cross validated - 10 repetitions
-iters <- 100
-cv10f_aucs <- c()
-cv10f_all_resp <- c()
-cv10f_all_pred <- c()
+cv10r_aucs <- c()
+cv10r_all_resp <- c()
+cv10r_all_pred <- c()
 for(j in 1:iters){
-  sampling <- sample(1:nrow(aucrf_data),nrow(aucrf_data),replace=F)
-  cv10f_probs <- rep(NA,nrow(aucrf_data))
-  for(i in seq(1,nrow(aucrf_data),5)){
-    train <- aucrf_data[sampling[-(i:(i+4))],]
-    test <- aucrf_data[sampling[i:(i+4)],]
+  set.seed(j)
+  sampling <- sample(1:nrow(random_aucrf_data),nrow(random_aucrf_data),replace=F)
+  cv10r_probs <- rep(NA,nrow(random_aucrf_data))
+  for(i in seq(1,nrow(random_aucrf_data),5)){
+    train <- random_aucrf_data[sampling[-(i:(i+4))],]
+    test <- random_aucrf_data[sampling[i:(i+4)],]
+    set.seed(seed)
     temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
-    cv10f_probs[sampling[i:(i+4)]] <- predict(temp_model$RFopt, test, type='prob')[,2]
+    cv10r_probs[sampling[i:(i+4)]] <- predict(temp_model$RFopt, test, type='prob')[,2]
   }
-  cv10f_roc <- roc(aucrf_data$Euth_Early~cv10f_probs)
-  cv10f_all_pred <- c(cv10f_all_pred, cv10f_probs)
-  cv10f_all_resp <- c(cv10f_all_resp, aucrf_data$Euth_Early)
-  cv10f_aucs[j] <- cv10f_roc$auc
+  cv10r_roc <- roc(random_aucrf_data$Euth_Early~cv10r_probs)
+  cv10r_all_pred <- c(cv10r_all_pred, cv10r_probs)
+  cv10r_all_resp <- c(cv10r_all_resp, random_aucrf_data$Euth_Early)
+  cv10r_aucs[j] <- cv10r_roc$auc
 }
-cv10f_roc <- roc(cv10f_all_resp~cv10f_all_pred)
+cv10r_roc <- roc(cv10r_all_resp~cv10r_all_pred)
 
 #### leave-one-out
-LOO_probs <- NULL
-LOO_probs <- sapply(1:nrow(aucrf_data), function(i){
- train <- aucrf_data[-i,]
-  test <- aucrf_data[i,]
+LOOr_probs <- NULL
+LOOr_probs <- sapply(1:nrow(random_aucrf_data), function(i){
+ train <- random_aucrf_data[-i,]
+  test <- random_aucrf_data[i,]
+  set.seed(seed)
   temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
-  LOO_probs[i] <- predict(temp_model$RFopt, test, type='prob')[,2]
+  LOOr_probs[i] <- predict(temp_model$RFopt, test, type='prob')[,2]
 })
-LOO_roc <- roc(aucrf_data$Euth_Early~LOO_probs)
+LOOr_roc <- roc(random_aucrf_data$Euth_Early~LOOr_probs)
 
 #### leave-one-cage-out
-LOcO_df <- cbind(cage_id = sample(meta_file[meta_file$day %in% 0, 'cage_id'],
-                                  nrow(Validate_early_euth_df)),
-                 Validate_early_euth_df[ , c('Euth_Early', rf_early_aucrf$Xopt)])
-cage_list <- as.character(unique(LOcO_df$cage_id))
+LOcOr_df <- cbind(cage_id = Predict_df$cage_id, random_aucrf_data)
+cage_list_random <- as.character(unique(LOcOr_df$cage_id))
 
-LOcO_probs <- NULL
-LOcO_probs <- sapply(1:length(cage_list), function(i){
-  by_cage <- LOcO_df$cage_id == cage_list[i]
-  LOcO_predict <- select(LOcO_df, -cage_id)
+LOcOr_probs <- NULL
+LOcOr_probs <- sapply(1:length(cage_list_random), function(i){
+  by_cage <- LOcOr_df$cage_id == cage_list_random[i]
+  LOcO_predict <- select(LOcOr_df, -cage_id)
   train <- LOcO_predict[-by_cage,]
   test <- LOcO_predict[by_cage,]
+  set.seed(seed)
   temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
-  LOcO_probs[[i]] <- predict(temp_model$RFopt, test, type='prob')[,2]
+  LOcOr_probs[[i]] <- predict(temp_model$RFopt, test, type='prob')[,2]
 })
-LOcO_probs <- unlist(LOcO_probs)
-LOcO_roc <- roc(aucrf_data$Euth_Early~LOcO_probs)
+LOcOr_probs <- unlist(LOcOr_probs)
+LOcOr_roc <- roc(random_aucrf_data$Euth_Early~LOcOr_probs)
 
@@
 plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='')
-plot(otu_euth_roc, col='black', lwd=2, add=T, lty=1)
-plot(cv10f_roc, col='gray', lwd=2, add=T, lty=1)
-plot(LOO_roc, col='blue', lwd=2, add=T, lty=2)
-plot(LOcO_roc, col='red', lwd=2, add=T, lty=2)
+plot(otu_random_roc, col='black', lwd=2, add=T, lty=1)
+plot(cv10r_roc, col='gray', lwd=2, add=T, lty=1)
+plot(LOOr_roc, col='blue', lwd=2, add=T, lty=2)
+plot(LOcOr_roc, col='red', lwd=2, add=T, lty=2)
 mtext(side=2, text="Sensitivity", line=2.5)
 mtext(side=1, text="Specificity", line=2.5)
-legend('bottomright', legend=c(sprintf('Out-Of-Bag (AUC: %.3g)',otu_euth_roc$auc),
-                               sprintf('Leave-1 mouse-out (AUC: %.3g)',LOO_roc$auc),
-                               sprintf('Leave-1 cage-out (AUC: %.3g)',LOcO_roc$auc),
-                               sprintf('10-fold CV (AUC: %.3g)',cv10f_roc$auc),
-                               sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_euth_roc,LOO_roc)$p.value),
-                               sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
-),lty=c(1,2,2,1,0,0), lwd=2, col=c('black','blue','red','gray'), bty='n')
+legend('bottomright', legend=c(sprintf('Out-Of-Bag (AUC: %.3g)',otu_random_roc$auc),
+                               sprintf('Leave-1 mouse-out (AUC: %.3g)',LOOr_roc$auc),
+                               sprintf('Leave-1 cage-out (AUC: %.3g)',LOcOr_roc$auc),
+                               sprintf('10-fold CV (AUC: %.3g)',cv10r_roc$auc)),
+#                               sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_random_roc,LOOr_roc)$p.value),
+#                               sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_random_roc,cv10r_roc)$p.value)),
+		lty=c(1,2,2,1,0,0), lwd=2, col=c('black','blue','red','gray'), bty='n')
 ```
-
-```{r Random_valid OTUs, fig.width=10, fig.height=8}
-layout(matrix(c(1,1,1,1,4,4,4,
-                2,2,2,5,3,3,3,
-                2,2,2,5,3,3,3), nrow=3, byrow = TRUE))
-
-#AUCRF curve
-par(mar=c(5,5,1.5,1))
-plot(rf_early_aucrf)
-#mtext('A', at=-90, font=2, side=3)
-
-#Importance Plot
-euth_otus <- rf_early_aucrf$Xopt
-euth_tax <- get_tax(1, euth_otus, tax_file)
-euth_tax$tax_label <- paste(euth_tax$tax, ' (', 
-                            gsub('tu0*', 'TU ', rownames(euth_tax)), 
-                                   ')', sep = '')
-dotchart(rev(rf_early_aucrf$ranking[euth_otus]), labels=rev(euth_tax[euth_otus, 'tax_label']), xlab='Mean Decrease Accuracy')
-#mtext('B', at=-0.014, font=2, side=3)
-
-#Abundance stripchart or most predictive otus
-euth_otus <- rev(rf_early_aucrf$Xopt)
-early_abunds <- Validate_early_euth_df[Validate_early_euth_df$Euth_Early==0, euth_otus]/10000 + 1e-4
-late_abunds <- Validate_early_euth_df[Validate_early_euth_df$Euth_Early==1, euth_otus]/10000 + 1e-4
-
-par(mar=c(4, 9, 1, 1))
-plot(1, type="n", ylim=c(0,length(euth_otus)*2), xlim=c(1e-4,1e-2), log="x", ylab="", xlab="Relative Abundance (%)", xaxt="n", yaxt="n")
-index <- 1
-for(i in euth_otus){
-  stripchart(at=index-0.35, jitter(early_abunds[,i], amount=1e-5), pch=21, bg="royalblue1", method="jitter", jitter=0.2, add=T, cex=1, lwd=0.5)
-  stripchart(at=index+0.35, jitter(late_abunds[,i], amount=1e-5), pch=21, bg="orange", method="jitter", jitter=0.2, add=T, cex=1, lwd=0.5)
-  segments(mean(early_abunds[,i]),index-0.7,mean(early_abunds[,i]),index, lwd=3)
-  segments(mean(late_abunds[,i]),index+0.7,mean(late_abunds[,i]),index, lwd=3)
-  index <- index + 2
-}
-axis(2, at=seq(1,index-2,2), labels=euth_tax[euth_otus, 'tax_label'], las=1, line=-0.5, tick=F, cex.axis=0.8)
-axis(1, at=c(1e-4, 1e-3, 1e-2), label=c("0", "0.1", "1"))
-legend('topright', legend=c("Mild", "Severe"), pch=c(21, 21), pt.bg=c("orange","royalblue1"), cex=0.7)
-#mtext('C', at=1e-7, font=2, side=3)
-
-```

### 2016-09-23 code/random_rf_features_boruta.R

@@ -53,7 +53,7 @@ random_model_features <- c()
 
 for(i in 1:iters){
   set.seed(seed)
-  Boruta(Euth_Early ~ ., data = Validate_early_euth_df)
+  random_bor <- Boruta(Euth_Early ~ ., data = Validate_early_euth_df)
   feature_df <- attStats(random_bor)
   feature_df$OTU <- rownames(feature_df)
   random_model_features[[i]] <- feature_df

### 2016-09-23 code/HumanCdGF_microbiota_predict.Rmd

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
+knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
 ```
@@
 
+rf_title <- expression(paste("Random Forest Prediction of ", italic("C. difficile"), " CFU dpi from Day 0 Community"))
+
 day_0_rf_plot <- ggplot(data = prediction_df, aes(x=observed, y=predicted)) + geom_point() + 
         facet_wrap( ~ day, nrow = 2) + 
-        labs(title = 'Random Forest Prediction of C Diff CFU dpi from Day 0 Community',
-        x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)')
+        labs(title = rf_title, x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)')
 
@@
 day_0_rsq_plot <- ggplot(data = rsq_df, aes(x = factor(day), y = RSQ)) + geom_point() + theme_bw() +
-        labs(x= 'Day', y = 'R^2') #title = 'Variation Explained by Day of C Diff CFU dpi from Day 0 Community',
+        labs(x= 'Day', y = 'R^2') + expand_limits(y=0) #title = 'Variation Explained by Day of C Diff CFU dpi from Day 0 Community',
 
@@
 day_0_mse_plot <- ggplot(data = mse_df, aes(x = factor(day), y = MSE)) + geom_point() + theme_bw() +
-        labs(x= 'Day', y = 'MSE') # title = 'MSE by Day of C Diff CFU dpi from Day 0 Community',
+        labs(x= 'Day', y = 'MSE') + expand_limits(y=0) # title = 'MSE by Day of C Diff CFU dpi from Day 0 Community',
 day_0_r2feat_plot <- ggplot(rsq_features, aes(x=Features..n., y=RSQ, group=(Day), color=factor(Day))) + geom_line() +
-  labs(x = 'Features (n)', y = 'R^2') + scale_color_discrete(guide=FALSE) + theme_bw()
+  labs(x = 'Features (n)', y = 'R^2') + scale_color_discrete(guide=FALSE) + theme_bw() + expand_limits(y=0)
 #  title= 'R^2 from RF with top n features\n
@@
 
-n_important <- 20
+n_important <- 10
 median_importance <- names(head(sort((apply(importance_df,1, median)), decreasing = T), n_important))
@@
 colnames(med_import_df) <- c('OTU', 'Day', 'Importance')
-day_0_feat_plot <- ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + facet_wrap( ~ OTU, nrow = 2) +
+otu_taxa <- get_tax(1, as.character(unique(med_import_df$OTU)), tax_file)
+med_import_df$OTU <- paste(otu_taxa[,1],' (', 
+                           gsub('tu0*', 'TU', rownames(otu_taxa)),')', sep='')
+
+day_0_feat_plot <- ggplot(data = med_import_df, aes(x = factor(Day), y = Importance)) + 
+        facet_wrap( ~ OTU, nrow = 2) +
         geom_point() + scale_x_discrete(labels = c(1:10)) +
-        labs(title = 'OTU Importance in Prediction of C Diff CFU dpi from Day 0 Community\n (Top 20 Median OTUs)',
+        labs(title = 'OTU Importance (Top 10 Median OTUs)',
         x= 'Day', y = '% Increase in MSE') + theme_bw()

### 2016-09-26 code/CdGF_AUCRF.Rmd
fix aucrf legend

@@
                                sprintf('10-fold CV (AUC: %.3g)',cv10f_roc$auc),
-                               sprintf('Random (AUC: %.3g)',ran_cv10f_roc$auc),
+                               sprintf('Random OOB (AUC: %.3g)',predict_random_roc$auc),
+                               sprintf('Random 10-fold CV (AUC: %.3g)',ran_cv10f_roc$auc),
                                sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_euth_roc,LOO_roc)$p.value),
                                sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
-),lty=c(1,2,2,1,3,0,0), lwd=2, col=c('black','blue','red','gray','orange'), bty='n')
+),lty=c(1,2,2,1,3,1,0,0), lwd=2, col=c('black','blue','purple','gray','red','orange'), bty='n')
 ```

### 2016-09-26 code/inocula_nmds.R
updated nmds and thetaYC code

@@ -16,6 +16,10 @@ shared <- read.table('data/process/human_CdGF.an.unique_list.0.03.subsample.shar
 #remove dashes from group names so the next stuff will work
 #orig_shared[2] <- gsub("-", "", orig_shared$Group)
 
+#make shared file look like a true shared:
+shared_group <- cbind(label=0.03, shared[1], numOTUs=2255, shared[2:ncol(shared)])
+colnames(shared_group)[2] <- 'Group'
+
 #subset shared to be inoculum only
 inocula <- rownames(meta_file)[meta_file$cage_id=='inoculum']
 inocula_shared <- shared[shared$sample_id %in% inocula,]
@@ -39,6 +43,7 @@ inputd0_shared <- rbind(inocula_shared, day0_shared)
 write.table(inocula_shared, file='data/process/inocula.shared', quote=F,sep='\t',row.names=F)
 write.table(day0_shared, file='data/process/day0.shared', quote=F,sep='\t',row.names=F)
 write.table(inputd0_shared, file='data/process/inputd0.shared', quote=F, sep='\t', row.names=F)
+write.table(shared_group, file='data/process/shared_group.shared', quote=F, sep='\t', row.names=F)
 
 #run dist.shared, nmds in mothur to get axes file
 #mothur ouput stats for combined file: 
@@ -89,7 +94,7 @@ legend(x="topright", legend, col = c("red", "black"), pch=16)
 
 severe4 <- day0_outcome_nmds[grep('Severe', day0_outcome_nmds$V2), c(2,3)]
 asymp4 <- day0_outcome_nmds[grep('Asymptomatic', day0_outcome_nmds$V2), c(2,3)]
-plot(day0_outcome_nmds$axis1, day0_outcome_nmds$axis2, main="Similarity day0 by outcome")
+plot(day0_outcome_nmds$axis1, day0_outcome_nmds$axis2, main="Similarity day0 by outcome", col = day0_outcome_nmds$V2, pch=as.numeric(day0_outcome_nmds$V2))
 points(severe4, pch=16, col = "red")
 points(asymp4, pch=16, col = "black")
 legend <- c("severe", "asymptomatic")
[…]

### 2016-09-26 code/thetayc_time.R

@@ -3,11 +3,55 @@
 # thetayc_time.R
 #
 #Plot thetaYC over time distance from day 0 
-#
+#thetaYC distances cant have sample names that are just numbers, this script wont work 
 #    
 #
 ###################
 
+source('code/read.dist.R')
+
 #read in files
 meta_file <- read.table('data/process/human_CdGF_metadata.txt', sep='\t',header = T, row.names = 2)
 shared <- read.table('data/process/human_CdGF.an.unique_list.0.03.subsample.shared', sep='\t',header=T)
+theta <- read.dist(file='data/process/shared_group.thetayc.0.03.lt.dist', input="lt", make.square=T, diag=0)
+
+#then maybe write it as a function 
+
+#################THE RIGHT ONE!!!#######
+#compares day0 to every subsequent day. this asks, how different are the communities from the start? can this be linked to outcome?
+
+mouseIDVector <- as.character(unique(meta_file$mouse_id))
+mouseIDVector <- mouseIDVector[mouseIDVector != "inoculum_NA"]
+stability <- as.data.frame(matrix(ncol=10, nrow=length(mouseIDVector)))
+row.names(stability) <- mouseIDVector
+names(stability) <- c(1,2,3,4,5,6,7,8,9,10)
+for (k in 1:10){
+  sampA <- rownames(meta_file)[meta_file$day %in% 0]
+  sampB <- rownames(meta_file)[meta_file$day %in% k]
+  dist <- theta[sampA, sampB]
+  row.names(dist) <- meta_file$mouse_id[rownames(meta_file) %in% rownames(dist)]
+  names(dist) <- meta_file$mouse_id[rownames(meta_file) %in% names(dist)]
+  for (i in mouseIDVector){
+    coord <- dist[i,i]
+    try(
+    stability[i,k] <- as.numeric(coord), silent = TRUE
+    )
+  }
+}
+
+test <- t(stability)
+#plot
+#tidyr or reshape from ggverse to melt 
+#or draw on blank plot and connect
+
+
+
+
+
+#build as function 
+
+
+#now write version where you compare each day to previous day. this asks question: how stable is community over time?
+
+
+

### 2016-09-26 code/make_HumanCdGF_metadata_file.R
updated make_metadata with new filepaths

@@ -14,7 +14,7 @@
 #
 ###################
 input_mouse <- 'data/raw/humanGF_ids.xlsx'
-input_shared <- 'data/mothur/gf_new.an.0.03.subsample.shared'
+input_shared <- 'data/process/gf_new.subsample.shared'
 output_metadata <- 'data/process/human_CdGF_metadata.txt'
 output_shared <- 'data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
 

### 2016-09-26 code/mouse_d0_community.R

@@ -12,9 +12,9 @@ install.packages("RColorBrewer")
 library(RColorBrewer)
 
 #import Nick's awesome combined metadata file
-meta_file <- read.table("data/process/human_CdGF_metadata.txt", header = TRUE, sep='\t', fill = TRUE, row.names=3)
+meta_file <- read.table("data/process/human_CdGF_metadata.txt", header = TRUE, sep='\t', fill = TRUE, row.names=2)
 shared_file <- read.table("data/process/human_CdGF.an.unique_list.0.03.subsample.shared", sep='\t', header = TRUE, row.names=1)
-tax_file <- read.table('data/mothur/gf_cdiff.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy', sep='\t',header = T, row.names = 1)
+tax_file <- read.table('data/process/gf_new.an.taxonomy', sep='\t',header = T, row.names = 1)
 
 #make OTU abundance file
 

### 2016-09-26 code/CdGF_AUCRF.Rmd
added other model code to aucrf

@@
 ```
+
+### Other prediction models
+
+##### C50
+
+```{r c50, eval=F}
+install.packages('C50')
+library(C50)
+
+set.seed(seed)
+shuffle <- runif(nrow(Predict_early_euth_df))
+Predict_early_euth_df_r <- Predict_early_euth_df[order(shuffle), ]
+
+c50_euth_model <- C5.0(Predict_early_euth_df_r[1:50 , -1], Predict_early_euth_df_r[1:50 , 1])
+summary(c50_euth_model)
+plot(c50_euth_model)
+c50_euth_predict <- predict(c50_euth_model, Predict_early_euth_df_r[51:55, ])
+table(Predict_early_euth_df_r[51:55, 1],c50_euth_predict)
+
+mylibrary<-c("SL.glm","SL.randomForest","SL.svm","SL.glmnet") 
+SL_euth_model <- SuperLearner(as.numeric(Predict_early_euth_df_r[ , 1]), Predict_early_euth_df_r[ , -1], SL.library = mylibrary)
+predict(SL_euth_model, newdata = as.numeric(Predict_early_euth_df_r[ , 1]))
+
+SL_euth_model$SL.predict
+sum=0
+tt<-length(Predict_early_euth_df_r)
+for(i in 1:tt) {
+sum<-sum+(Predict_early_euth_df_r[i,1]-SL_euth_model$SL.predict[i])^2
+  } 
+sumg <-sum/tt
+sumg
+
+listWrappers()
+
+Remove_otu_df <- Predict_early_euth_df_r %>% 
+  select(-Otu00074, -Otu00114, -Otu00050, -Otu00124, -Otu00120, -Otu00100)
+
+set.seed(seed)
+rf_rm_otu_early_aucrf <- AUCRF(Euth_Early ~ ., data = Remove_otu_df,
+                              ntree = n_trees, pdel = 0.05, ranking = 'MDA')
+rm_otu_euth_probs <- predict(rf_rm_otu_early_aucrf$RFopt, type = 'prob')
+rm_otu_euth_roc <- roc(Remove_otu_df$Euth_Early ~ rm_otu_euth_probs[ , 2])
+
+plot(rm_otu_euth_roc, col='gray', lwd=2, add=T, lty=1)
+```

### 2016-09-29 code/CdGF_AUCRF.Rmd
fixed leave one cage out

@@
                               ntree = n_trees, pdel = 0.05, ranking = 'MDA')
+summary(rf_early_aucrf)
 otu_euth_probs <- predict(rf_early_aucrf$RFopt, type = 'prob')
@@
 
-LOcO_probs <- NULL
-LOcO_probs <- sapply(1:length(cage_list), function(i){
+LOcO_probs <- c()
+for(i in 1:length(cage_list)){
   by_cage <- LOcO_df$cage_id == cage_list[i]
+  LOcO_predict <- select(LOcO_df, -cage_id)
+  train <- LOcO_predict[!by_cage,]
+  test <- LOcO_predict[by_cage,]
+  set.seed(seed)
+  temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
+  LOcO_probs[LOcO_df$cage_id == cage_list[i]] <- predict(temp_model$RFopt, test, type='prob')[,2]
+}
+LOcO_roc <- roc(LOcO_df$Euth_Early~LOcO_probs)
+
+LOcO_cv10f_aucs <- c()
+LOcO_cv10f_all_resp <- c()
+LOcO_cv10f_all_pred <- c()
+for(j in 1:iters){
+  set.seed(j)
+  shuffle <- runif(nrow(LOcO_df))
+  LOcO_df_r <- LOcO_df[order(shuffle), ]
+#  sampling <- sample(1:nrow(LOcO_df),nrow(LOcO_df),replace=F)
+  LOcO_cv10f_probs <- rep(NA,nrow(LOcO_df_r))
+  for(i in 1:length(cage_list)){
+    by_cage <- LOcO_df_r$cage_id == cage_list[i]
     LOcO_predict <- select(LOcO_df, -cage_id)
@@
     temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
-  LOcO_probs[i] <- predict(temp_model$RFopt, test, type='prob')[,2]
-})
-LOcO_probs <- unlist(LOcO_probs)
-LOcO_roc <- roc(aucrf_data$Euth_Early~LOcO_probs)
+    LOcO_cv10f_probs[by_cage] <- predict(temp_model$RFopt, test, type='prob')[,2]
+  }
+  LOcO_cv10f_roc <- roc(LOcO_df_r$Euth_Early~LOcO_cv10f_probs)
+  LOcO_cv10f_all_pred <- c(LOcO_cv10f_all_pred, LOcO_cv10f_probs)
+  LOcO_cv10f_all_resp <- c(LOcO_cv10f_all_resp, LOcO_df_r$Euth_Early)
+  LOcO_cv10f_aucs[j] <- LOcO_cv10f_roc$auc
+}
+LOcO_cv10f_roc <- roc(LOcO_cv10f_all_resp~LOcO_cv10f_all_pred)
 
@@
 
+### Add ranges to RF ROC
+
+```{r error_plot, eval=F}
+
+caretmodel <- function(x) {
+  fitControl <- trainControl(method = "repeatedcv",
+    number = 10,
+    repeats = 10,
+    classProbs = TRUE,
+    summaryFunction = twoClassSummary,
+    savePredictions = TRUE)
+  model <- train(Euth_Early ~ ., data=x, 
+                 trControl=fitControl, method="rf", metric="ROC")
+  return(model)
+}
+test <- aucrf_data
+test$Euth_Early <- factor(ifelse(test[ , 'Euth_Early'] == 1, 'Late', 'Early'))
+random <- test
+set.seed(seed)
+random$Euth_Early <- sample(random$Euth_Early, nrow(random))
+caret_aucrf <- caretmodel(test)
+caret_random <- caretmodel(random)
+caret_prediction <- caretmodel(Predict_early_euth_df)
+
+str(aucrf_data)
+head(caret_aucrf$pred)
+
+roclobster <- ggplot(caret_aucrf$pred, aes(d = obs, m = Late)) +
+  geom_roc(n.cuts = 0) +
+  theme_classic() +
+  theme(
+    axis.line.x = element_line(colour = "black"),
+    axis.line.y = element_line(colour = "black")
+  ) +
+  geom_segment(aes(x = 0, y = 0, xend = 1, yend = 1), linetype=2) +
+  ylab("Sensitivity") +
+xlab(paste("Inverse Specificity"))
+
+confusionMatrix(data = caret_aucrf$pred[,'pred'], reference = caret_aucrf$pred[,'obs'])
+confusionMatrix(data = caret_random$pred[,'pred'], reference = caret_random$pred[,'obs'])
+
+n <- nrow(Predict_early_euth_df)
+indextrain <- sample(1:n,round(0.6*n),replace=FALSE)
+Xtrain     <- Predict_early_euth_df[ indextrain,-1]
+Xtest      <- Predict_early_euth_df[-indextrain,-1]
+Ytrain     <- Predict_early_euth_df[ indextrain,1]
+Ytest      <- Predict_early_euth_df[-indextrain,1]
+
+qrf <- quantregForest(x=Xtrain, y=Ytrain, nodesize=10,sampsize=30)
+conditionalQuantiles  <- predict(qrf, Xtest, what=sd)
+print(conditionalQuantiles[1:4,])
+
+```
+
 ### Other prediction models

### 2016-09-30 code/CdGF_AUCRF.Rmd
remove randomized validation from aucrf

@@
 
-ran_cv10f_aucs <- c()
-ran_cv10f_all_resp <- c()
-ran_cv10f_all_pred <- c()
-for(j in 1:iters){
-  set.seed(j)
-  sampling <- sample(1:nrow(aucrf_data),nrow(aucrf_data),replace=F)
-  ran_cv10f_probs <- rep(NA,nrow(aucrf_data))
-  for(i in seq(1,nrow(aucrf_data),5)){
-    train <- aucrf_data[sampling[-(i:(i+4))],]
-    test <- ran_outcome_data[sampling[i:(i+4)],]
-    set.seed(seed)
-    temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
-    ran_cv10f_probs[sampling[i:(i+4)]] <- predict(temp_model$RFopt, test, type='prob')[,2]
-  }
-  ran_cv10f_roc <- roc(ran_outcome_data$Euth_Early~ran_cv10f_probs)
-  ran_cv10f_all_pred <- c(ran_cv10f_all_pred, ran_cv10f_probs)
-  ran_cv10f_all_resp <- c(ran_cv10f_all_resp, ran_outcome_data$Euth_Early)
-  ran_cv10f_aucs[j] <- ran_cv10f_roc$auc
-}
-ran_cv10f_roc <- roc(ran_cv10f_all_resp~ran_cv10f_all_pred)
-
 #### plot rocs
@@
 plot(LOcO_roc, col='purple', lwd=2, add=T, lty=3)
-plot(ran_cv10f_roc, col='orange', lwd=2, add=T, lty=1)
-plot(predict_random_roc, col='red', lwd=2, add=T, lty=3)
 mtext(side=2, text="Sensitivity", line=2.5)
@@
                                sprintf('10-fold CV (AUC: %.3g)',cv10f_roc$auc),
-                               sprintf('Random OOB (AUC: %.3g)',predict_random_roc$auc),
-                               sprintf('Random 10-fold CV (AUC: %.3g)',ran_cv10f_roc$auc),
                                sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_euth_roc,LOO_roc)$p.value),
                                sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
-),lty=c(1,2,2,1,3,1,0,0), lwd=2, col=c('black','blue','purple','gray','red','orange'), bty='n')
+),lty=c(1,2,2,1,0,0), lwd=2, col=c('black','blue','purple','gray','red'), bty='n')
 ```
@@
 #Boruta(Euth_Early ~ ., data = Validate_early_euth_df, maxRuns = 200)
-
-#create dataframe for random RF-classification
-Validate_early_euth_df <- Predict_early_euth_df
-set.seed(seed)
-Validate_early_euth_df$Euth_Early <- sample(Predict_early_euth_df$Euth_Early)
-
-# create RF model
-set.seed(seed)
-rf_random_aucrf <- AUCRF(Euth_Early ~ ., data = Validate_early_euth_df,
-                              ntree = n_trees, pdel = 0.05, ranking = 'MDA')
-otu_random_probs <- predict(rf_random_aucrf$RFopt, type = 'prob')
-otu_random_roc <- roc(Validate_early_euth_df$Euth_Early ~ otu_random_probs[ , 2])
-
-random_aucrf_data <- Validate_early_euth_df[,c('Euth_Early',rf_random_aucrf$Xopt)]
-
-#### 10-fold Cross validated - 10 repetitions
-cv10r_aucs <- c()
-cv10r_all_resp <- c()
-cv10r_all_pred <- c()
-for(j in 1:iters){
-  set.seed(j)
-  sampling <- sample(1:nrow(random_aucrf_data),nrow(random_aucrf_data),replace=F)
-  cv10r_probs <- rep(NA,nrow(random_aucrf_data))
-  for(i in seq(1,nrow(random_aucrf_data),5)){
-    train <- random_aucrf_data[sampling[-(i:(i+4))],]
-    test <- random_aucrf_data[sampling[i:(i+4)],]
-    set.seed(seed)
-    temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
-    cv10r_probs[sampling[i:(i+4)]] <- predict(temp_model$RFopt, test, type='prob')[,2]
-  }
-  cv10r_roc <- roc(random_aucrf_data$Euth_Early~cv10r_probs)
-  cv10r_all_pred <- c(cv10r_all_pred, cv10r_probs)
-  cv10r_all_resp <- c(cv10r_all_resp, random_aucrf_data$Euth_Early)
-  cv10r_aucs[j] <- cv10r_roc$auc
-}
-cv10r_roc <- roc(cv10r_all_resp~cv10r_all_pred)
-
-#### leave-one-out
-LOOr_probs <- NULL
-LOOr_probs <- sapply(1:nrow(random_aucrf_data), function(i){
- train <- random_aucrf_data[-i,]
-  test <- random_aucrf_data[i,]
-  set.seed(seed)
-  temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
-  LOOr_probs[i] <- predict(temp_model$RFopt, test, type='prob')[,2]
-})
-LOOr_roc <- roc(random_aucrf_data$Euth_Early~LOOr_probs)
-
-#### leave-one-cage-out
-LOcOr_df <- cbind(cage_id = Predict_df$cage_id, random_aucrf_data)
-cage_list_random <- as.character(unique(LOcOr_df$cage_id))
-
-LOcOr_probs <- NULL
-LOcOr_probs <- sapply(1:length(cage_list_random), function(i){
-  by_cage <- LOcOr_df$cage_id == cage_list_random[i]
-  LOcO_predict <- select(LOcOr_df, -cage_id)
-  train <- LOcO_predict[-by_cage,]
-  test <- LOcO_predict[by_cage,]
-  set.seed(seed)
-  temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
-  LOcOr_probs[[i]] <- predict(temp_model$RFopt, test, type='prob')[,2]
-})
-LOcOr_probs <- unlist(LOcOr_probs)
-LOcOr_roc <- roc(random_aucrf_data$Euth_Early~LOcOr_probs)
-
-#### plot rocs
-par(mar=c(4,4,1,1))
-plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='')
-plot(otu_random_roc, col='black', lwd=2, add=T, lty=1)
-plot(cv10r_roc, col='gray', lwd=2, add=T, lty=1)
-plot(LOOr_roc, col='blue', lwd=2, add=T, lty=2)
-plot(LOcOr_roc, col='red', lwd=2, add=T, lty=2)
-mtext(side=2, text="Sensitivity", line=2.5)
-mtext(side=1, text="Specificity", line=2.5)
-legend('bottomright', legend=c(sprintf('Out-Of-Bag (AUC: %.3g)',otu_random_roc$auc),
-                               sprintf('Leave-1 mouse-out (AUC: %.3g)',LOOr_roc$auc),
-                               sprintf('Leave-1 cage-out (AUC: %.3g)',LOcOr_roc$auc),
-                               sprintf('10-fold CV (AUC: %.3g)',cv10r_roc$auc)),
-#                               sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_random_roc,LOOr_roc)$p.value),
-#                               sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_random_roc,cv10r_roc)$p.value)),
-		lty=c(1,2,2,1,0,0), lwd=2, col=c('black','blue','red','gray'), bty='n')
 ```
 
-### Add ranges to RF ROC
+### Add ranges to RF ROC ??
 

### 2016-09-30 code/HumanCdGF_microbiota_predict_by_day.Rmd
color cfu otu dependence by outcome

@@
 
-cfu_by_day_comm %>% 
-  select(log_cfu, day, one_of(predictive_otus)) %>% 
+cfu_by_day <-
+  meta_file %>% 
+  filter(day %in% 1:10) %>% 
+  select(sample_id, day, log_cfu, Early_Euth) %>% 
+  left_join(rel_abund_shared_01) %>% 
+  select(-sample_id, -one_of(cdiff_otu))
+
+cfu_by_day %>% 
+  select(log_cfu, day, one_of(predictive_otus), Early_Euth) %>% 
   filter(!is.na(log_cfu)) %>% 
@@
   full_join(predictive_taxa) %>% 
-  ggplot(aes(x = abundance, y = log_cfu, color = factor(day))) + geom_point() + 
+  ggplot(aes(x = abundance, y = log_cfu, color = factor(Early_Euth))) + geom_point() + 
     facet_wrap(~ tax_label, scales = 'free_x') +
@@
 ```{r fig.width=10, fig.height=10}
-cfu_by_day_comm %>% 
-  select(log_cfu, day, one_of(predictive_otus)) %>% 
+cfu_by_day %>% 
+  select(log_cfu, day, one_of(predictive_otus), Early_Euth) %>% 
   filter(!is.na(log_cfu)) %>% 
@@
   full_join(predictive_taxa) %>% 
-  ggplot(aes(x = log10(abundance), y = log_cfu, color = factor(day))) + geom_point() + 
+  ggplot(aes(x = log10(abundance), y = log_cfu, color = factor(Early_Euth))) + geom_point() + 
     facet_wrap(~ tax_label, scales = 'free_x') +
@@
 meta_file %>% 
-  select(day, log_cfu, mouse_id) %>% 
+  select(day, log_cfu, mouse_id, Early_Euth) %>% 
   full_join(day_0_otu_by_mouse) %>% 
-  select(day, log_cfu, one_of(predictive_otus)) %>% 
+  select(day, log_cfu, one_of(predictive_otus), Early_Euth) %>% 
   filter(!day == 0) %>% 

### 2016-09-30 code/CdGF_AUCRF.Rmd
remove extra models from ROC plot

@@
 ```{r}
-pack_used <- c('randomForest','miscTools','ggplot2','reshape2', 'pROC', 'knitr', 'cowplot','dplyr','AUCRF')
+pack_used <- c('randomForest','miscTools','ggplot2','reshape2', 'pROC', 'knitr', 'cowplot','dplyr','AUCRF','Boruta')
 for (dep in pack_used){
@@
                               ntree = n_trees, pdel = 0.05, ranking = 'MDA')
-summary(rf_early_aucrf)
+
 otu_euth_probs <- predict(rf_early_aucrf$RFopt, type = 'prob')
 otu_euth_roc <- roc(Predict_early_euth_df$Euth_Early ~ otu_euth_probs[ , 2])
-set.seed(seed)
-random_roc <- roc(sample(c(1,0), length(otu_euth_probs[ , 2]), replace = T) ~ otu_euth_probs[ , 2])
 
@@
   test <- LOcO_predict[by_cage,]
-  set.seed(seed)
+  set.seed(j)
   temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
@@
 LOcO_roc <- roc(LOcO_df$Euth_Early~LOcO_probs)
-
-LOcO_cv10f_aucs <- c()
-LOcO_cv10f_all_resp <- c()
-LOcO_cv10f_all_pred <- c()
-for(j in 1:iters){
-  set.seed(j)
-  shuffle <- runif(nrow(LOcO_df))
-  LOcO_df_r <- LOcO_df[order(shuffle), ]
-#  sampling <- sample(1:nrow(LOcO_df),nrow(LOcO_df),replace=F)
-  LOcO_cv10f_probs <- rep(NA,nrow(LOcO_df_r))
-  for(i in 1:length(cage_list)){
-    by_cage <- LOcO_df_r$cage_id == cage_list[i]
-    LOcO_predict <- select(LOcO_df, -cage_id)
-    train <- LOcO_predict[-by_cage,]
-    test <- LOcO_predict[by_cage,]
-    set.seed(seed)
-    temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
-    LOcO_cv10f_probs[by_cage] <- predict(temp_model$RFopt, test, type='prob')[,2]
-  }
-  LOcO_cv10f_roc <- roc(LOcO_df_r$Euth_Early~LOcO_cv10f_probs)
-  LOcO_cv10f_all_pred <- c(LOcO_cv10f_all_pred, LOcO_cv10f_probs)
-  LOcO_cv10f_all_resp <- c(LOcO_cv10f_all_resp, LOcO_df_r$Euth_Early)
-  LOcO_cv10f_aucs[j] <- LOcO_cv10f_roc$auc
-}
-LOcO_cv10f_roc <- roc(LOcO_cv10f_all_resp~LOcO_cv10f_all_pred)
-
-#### Random - 10-fold Cross validated - 10 repetitions
-
-set.seed(seed)
-ran_outcome_data <- cbind(Euth_Early = factor(sample(Predict_early_euth_df$Euth_Early,  
-                                                     nrow(Predict_early_euth_df))), 
-                             Predict_early_euth_df[,rf_early_aucrf$Xopt]) 
-predict_random <-  predict(rf_early_aucrf$RFopt, ran_outcome_data, type = 'prob')[,2]
-predict_random_roc <- roc(ran_outcome_data$Euth_Early ~ predict_random)
 
@@
 
+Leave 1 cage out RF with all otus gives and AUC of 0.63-0.68, but using the reduced features selected by AUCRF on the whole dataset gives an AUC of 0.88-0.90
+
 ***  
 ***
-
-### Validate AUCRF with Randomized Input
-
-```{r Random_validation, fig.width=6, fig.height=5, results= 'hide'}
-
-# boruta feature selection
-#set.seed(seed)
-#Boruta(Euth_Early ~ ., data = Validate_early_euth_df, maxRuns = 200)
-```
-
-### Add ranges to RF ROC ??
-
-```{r error_plot, eval=F}
-
-caretmodel <- function(x) {
-  fitControl <- trainControl(method = "repeatedcv",
-    number = 10,
-    repeats = 10,
-    classProbs = TRUE,
-    summaryFunction = twoClassSummary,
-    savePredictions = TRUE)
-  model <- train(Euth_Early ~ ., data=x, 
-                 trControl=fitControl, method="rf", metric="ROC")
-  return(model)
-}
-test <- aucrf_data
-test$Euth_Early <- factor(ifelse(test[ , 'Euth_Early'] == 1, 'Late', 'Early'))
-random <- test
-set.seed(seed)
-random$Euth_Early <- sample(random$Euth_Early, nrow(random))
-caret_aucrf <- caretmodel(test)
-caret_random <- caretmodel(random)
-caret_prediction <- caretmodel(Predict_early_euth_df)
-
-str(aucrf_data)
-head(caret_aucrf$pred)
-
-roclobster <- ggplot(caret_aucrf$pred, aes(d = obs, m = Late)) +
-  geom_roc(n.cuts = 0) +
-  theme_classic() +
-  theme(
-    axis.line.x = element_line(colour = "black"),
-    axis.line.y = element_line(colour = "black")
-  ) +
-  geom_segment(aes(x = 0, y = 0, xend = 1, yend = 1), linetype=2) +
-  ylab("Sensitivity") +
-xlab(paste("Inverse Specificity"))
-
-confusionMatrix(data = caret_aucrf$pred[,'pred'], reference = caret_aucrf$pred[,'obs'])
-confusionMatrix(data = caret_random$pred[,'pred'], reference = caret_random$pred[,'obs'])
-
-n <- nrow(Predict_early_euth_df)
-indextrain <- sample(1:n,round(0.6*n),replace=FALSE)
-Xtrain     <- Predict_early_euth_df[ indextrain,-1]
-Xtest      <- Predict_early_euth_df[-indextrain,-1]
-Ytrain     <- Predict_early_euth_df[ indextrain,1]
-Ytest      <- Predict_early_euth_df[-indextrain,1]
-
-qrf <- quantregForest(x=Xtrain, y=Ytrain, nodesize=10,sampsize=30)
-conditionalQuantiles  <- predict(qrf, Xtest, what=sd)
-print(conditionalQuantiles[1:4,])
-
-```
-
-### Other prediction models
-
-##### C50
-
-```{r c50, eval=F}
-install.packages('C50')
-library(C50)
-
-set.seed(seed)
-shuffle <- runif(nrow(Predict_early_euth_df))
-Predict_early_euth_df_r <- Predict_early_euth_df[order(shuffle), ]
-
-c50_euth_model <- C5.0(Predict_early_euth_df_r[1:50 , -1], Predict_early_euth_df_r[1:50 , 1])
-summary(c50_euth_model)
-plot(c50_euth_model)
-c50_euth_predict <- predict(c50_euth_model, Predict_early_euth_df_r[51:55, ])
-table(Predict_early_euth_df_r[51:55, 1],c50_euth_predict)
-
-mylibrary<-c("SL.glm","SL.randomForest","SL.svm","SL.glmnet") 
-SL_euth_model <- SuperLearner(as.numeric(Predict_early_euth_df_r[ , 1]), Predict_early_euth_df_r[ , -1], SL.library = mylibrary)
-predict(SL_euth_model, newdata = as.numeric(Predict_early_euth_df_r[ , 1]))
-
-SL_euth_model$SL.predict
-sum=0
-tt<-length(Predict_early_euth_df_r)
-for(i in 1:tt) {
-sum<-sum+(Predict_early_euth_df_r[i,1]-SL_euth_model$SL.predict[i])^2
-  } 
-sumg <-sum/tt
-sumg
-
-listWrappers()
-
-Remove_otu_df <- Predict_early_euth_df_r %>% 
-  select(-Otu00074, -Otu00114, -Otu00050, -Otu00124, -Otu00120, -Otu00100)
-
-set.seed(seed)
-rf_rm_otu_early_aucrf <- AUCRF(Euth_Early ~ ., data = Remove_otu_df,
-                              ntree = n_trees, pdel = 0.05, ranking = 'MDA')
-rm_otu_euth_probs <- predict(rf_rm_otu_early_aucrf$RFopt, type = 'prob')
-rm_otu_euth_roc <- roc(Remove_otu_df$Euth_Early ~ rm_otu_euth_probs[ , 2])
-
-plot(rm_otu_euth_roc, col='gray', lwd=2, add=T, lty=1)
-```

### 2016-09-30 code/CdGF_AUCRF.Rmd

@@
 ```{r}
-pack_used <- c('randomForest','miscTools','ggplot2','reshape2', 'pROC', 'knitr', 'cowplot','dplyr','AUCRF','Boruta')
+pack_used <- c('randomForest','ggplot2', 'pROC', 'knitr','dplyr','AUCRF', 'tidyr')
 for (dep in pack_used){
@@
 
-```{r fig.width=6, fig.height=5, results= 'hide'}
+```{r fig.width=8, fig.height=8, results= 'hide'}
 #create dataframe for RF-classification
@@
 
-```{r fig.width=10, fig.height=8}
-layout(matrix(c(1,1,1,1,4,4,4,
-                2,2,2,5,3,3,3,
-                2,2,2,5,3,3,3), nrow=3, byrow = TRUE))
-
-#AUCRF curve
-par(mar=c(5,5,1.5,1))
-plot(rf_early_aucrf, ylim = c(0.7, 1.1))
-#mtext('A', at=-90, font=2, side=3)
+```{r fig.width=10, fig.height=10}
 
@@
                             gsub('tu0*', 'TU ', rownames(euth_tax)), 
-                                   ')', sep = '')
-dotchart(rev(rf_early_aucrf$ranking[euth_otus]), labels=rev(euth_tax[euth_otus, 'tax_label']), xlab='Mean Decrease Accuracy')
-#mtext('B', at=-0.014, font=2, side=3)
+                                   ') : ', round(rf_early_aucrf$ranking[euth_otus], 3), sep = '')
 
-#Abundance stripchart or most predictive otus
-euth_otus <- rev(rf_early_aucrf$Xopt)
-early_abunds <- Predict_early_euth_df[Predict_early_euth_df$Euth_Early==0, euth_otus]/10000 + 1e-4
-late_abunds <- Predict_early_euth_df[Predict_early_euth_df$Euth_Early==1, euth_otus]/10000 + 1e-4
+RF_features_polt <- Predict_early_euth_df %>% 
+  select(Euth_Early, one_of(euth_otus)) %>% 
+  gather('OTU', 'Value', -Euth_Early) %>% 
+  full_join(data.frame(OTU = names(rf_early_aucrf$ranking[euth_otus]), 
+               MDA =  rf_early_aucrf$ranking[euth_otus])) %>% 
+  ggplot( aes(x = reorder(factor(OTU), MDA), y = Value, fill = Euth_Early, dodge = Euth_Early)) +
+  theme(legend.title=element_blank(), legend.position = c(0.9,0.5),
+        panel.background = element_rect(fill = NA),
+        panel.grid.major = element_line(colour = NA),
+        panel.grid.minor = element_line(colour = NA)) +
+  geom_point(pch = 21, position = position_jitterdodge(), alpha = 0.4, size = 2) + coord_flip() +
+  scale_x_discrete(breaks = rownames(euth_tax), labels = euth_tax$tax_label) +
+  stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median,
+                 geom = "crossbar", width = 0.6, position=position_dodge(width=0.8)) + 
+  scale_fill_manual(values = c('red', 'black'), labels = c('Severe','Mild'),
+                    guide = guide_legend(reverse=TRUE))
 
-par(mar=c(4, 9, 1, 1))
-plot(1, type="n", ylim=c(0,length(euth_otus)*2), xlim=c(1e-4,1e-2), log="x", ylab="", xlab="Relative Abundance (%)", xaxt="n", yaxt="n")
-index <- 1
-for(i in euth_otus){
-  stripchart(at=index-0.35, jitter(early_abunds[,i], amount=1e-5), pch=21, bg="royalblue1", method="jitter", jitter=0.2, add=T, cex=1, lwd=0.5)
-  stripchart(at=index+0.35, jitter(late_abunds[,i], amount=1e-5), pch=21, bg="orange", method="jitter", jitter=0.2, add=T, cex=1, lwd=0.5)
-  segments(mean(early_abunds[,i]),index-0.7,mean(early_abunds[,i]),index, lwd=3)
-  segments(mean(late_abunds[,i]),index+0.7,mean(late_abunds[,i]),index, lwd=3)
-  index <- index + 2
-}
-axis(2, at=seq(1,index-2,2), labels=euth_tax[euth_otus, 'tax_label'], las=1, line=-0.5, tick=F, cex.axis=0.8)
-axis(1, at=c(1e-4, 1e-3, 1e-2), label=c("0", "0.1", "1"))
-legend('topright', legend=c("Mild", "Severe"), pch=c(21, 21), pt.bg=c("orange","royalblue1"), cex=0.7)
-#mtext('C', at=1e-7, font=2, side=3)
+RF_features_polt + labs( y = 'Relative Abundance (%)\nClassification (OTU) : MDA', x = '', title = '')
+
+RF_features_polt + labs( y = 'Relative Abundance (%) - log10\nClassification (OTU) : MDA', x = '', title = '') +
+  scale_y_log10()
 ```

### 2016-10-03 code/CdGF_AUCRF.Rmd
remove OTUs which select cage/inocula diffs

@@
 #random forest regression model to determine Day 0 OTU correlated with increased CFU
-seed <- 1
+seed <- 100316
 n_trees <- 2001
 iters <- 100
-
+n_features <- 30
 ```
@@
 Predict_df <- merge(meta_file[meta_file$day %in% 0, 
-                              c('cage_id', 'mouse_id', 'Euth_Early')],
+                              c('human_source', 'mouse_id', 'cage_id', 'Euth_Early')],
                     rel_abund_stool, by.x = 'mouse_id', by.y = 'row.names')
+Predict_cage_df <- select(Predict_df, -human_source, -mouse_id, -Euth_Early)
+Predict_source_df <- select(Predict_df, -cage_id, -mouse_id, -Euth_Early)
+Predict_cage_df$cage_id <- factor(Predict_cage_df$cage_id)
+Predict_source_df$human_source <- factor(Predict_source_df$human_source)
 
-Predict_early_euth_df <- select(Predict_df, -cage_id, -mouse_id)
+# identify OTUs that predict cage/inocula
+set.seed(seed)
+rf_cage <- randomForest(cage_id ~ ., data=Predict_cage_df, importance=TRUE, ntree=n_trees)
+set.seed(seed)
+rf_source <- randomForest(human_source ~ ., data=Predict_source_df, importance=TRUE, ntree=n_trees)
+features_source <- head(sort(rf_source$importance[ , 'MeanDecreaseAccuracy'], decreasing = T), n_features)
+features_cage <- head(sort(rf_cage$importance[ , 'MeanDecreaseAccuracy'], decreasing = T), n_features)
+features_all <- unique(c(names(features_cage), names(features_source)))
+
+# remove cage/inocula OTUs from prediction dataframes
+Predict_early_euth_df <- select(Predict_df, -cage_id, -mouse_id, -human_source, -one_of(features_all))
 
@@
   test <- LOcO_predict[by_cage,]
-  set.seed(j)
+  set.seed(seed)
   temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
@@
                                sprintf('Leave-1 cage-out (AUC: %.3g)',LOcO_roc$auc),
-                               sprintf('10-fold CV (AUC: %.3g)',cv10f_roc$auc),
-                               sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_euth_roc,LOO_roc)$p.value),
-                               sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
+                               sprintf('10-fold CV (AUC: %.3g)',cv10f_roc$auc) #,
+#                               sprintf('OOB vs Leave-1-out: p=%.2g', roc.test(otu_euth_roc,LOO_roc)$p.value),
+#                               sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
 ),lty=c(1,2,2,1,0,0), lwd=2, col=c('black','blue','purple','gray','red'), bty='n')
@@
 
-Leave 1 cage out RF with all otus gives and AUC of 0.63-0.68, but using the reduced features selected by AUCRF on the whole dataset gives an AUC of 0.88-0.90
+Leave 1 cage out RF with all otus gives and AUC of 0.70, but using the reduced features selected by AUCRF on the whole dataset gives an AUC of 0.80
+
+To remove any cage effect RF could be picking out instead of the actual outcome, I first run RF on all OTUs to predict cage and separately human source.
+Then select the OTUS that are necessary for the lowest error rate (I went slightly above) from both cage and human souce RFs.
+Remove those OTUs from the OTU data frame and predict Euthanization with AUCRF
 

### 2016-10-04 code/inocula_nmds.R
more figure updates

@@ -206,7 +206,7 @@ install.packages("ape")
 library(ape)
 
 inoc_tree <- read.tree(file='data/process/inocula_2194.thetayc.0.03.tre')
-day0_tree <- read.tree(file='data/process/day0.thetayc.0.03.tre')
+day0_tree <- read.tree(file='data/process/day0.thetayc.0.03.ave.tre')
 
 #for plotting donors by outcome 
 dead <- inoc_tree$tip.label[c(3,6,7,8,15)]
@@ -220,6 +220,16 @@ plot(inoc_tree, type='radial', tip.color=ifelse(inoc_tree$tip.label %in% diah, '
 plot(inoc_tree, tip.color=ifelse(inoc_tree$tip.label %in% diah, 'blue', ifelse(inoc_tree$tip.label %in% cdiff, 'red', 'black')))
 legend("bottomleft", c('diarrhea', 'cdiff infected', 'no diarrhea'), col = c('blue', 'red', 'black'), pch =16)
 
+#for plotting mice at day 0 by outcome 
 
+mild <- subset(d0_outcome, d0_outcome$V2 == 'Asymptomatic')
+milder <- mild[1]
+
+severe9 <- subset(d0_outcome, d0_outcome$V2 == 'Severe')
+severer <- severe9[1]
+
+#day0 mouse by outcome
+plot(day0_tree, tip.color=ifelse(day0_tree$tip.label %in% severer$V1, 'red', 'black'),  cex=0.7, main = "Day 0 dendrogram")
+legend("bottomleft", c("mild", "severe"), col = c('black', 'red'), pch = 16)
 
 

### 2016-10-04 code/mouse_weightloss_time.R

@@ -12,6 +12,48 @@
 #
 # Output: * results/figures/weight_loss_time.pdf
 
-meta_file <- read.table("data/process/human_CdGF_metadata.txt", header = TRUE, sep='\t', fill = TRUE, row.names=3)
+install.packages("cowplot")
+install.packages("dplyr")
+library(cowplot)
+library(dplyr)
+
+meta_file <- read.table("data/process/human_CdGF_metadata.txt", header = TRUE, sep='\t', fill = TRUE, row.names=2)
 shared_file <- read.table("data/process/human_CdGF.an.unique_list.0.03.subsample.shared", sep='\t', header = TRUE, row.names=1)
-tax_file <- read.table('data/mothur/gf_cdiff.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy', sep='\t',header = T, row.names = 1)
+tax_file <- read.table('data/process/gf_new.an.taxonomy', sep='\t',header = T, row.names = 1)
+
+
+#########All below is Nick's code that I am modifying ######
+
+wt_ls_plot <- meta_file %>% 
+  filter(day %in% c(0:10), !is.na(percent_weightLoss) ) %>%
+  select(mouse_id, day, Early_Euth, percent_weightLoss, cage_id) %>% 
+  ggplot(aes(x = factor(day), y = percent_weightLoss, color = factor(Early_Euth))) + 
+  geom_jitter(width = 0.5, alpha = 0.1) + 
+  scale_color_manual(values = c('black', 'red'), breaks=c("FALSE", "TRUE"),
+                     labels=c("Persistent", "Severe")) +
+  stat_summary(fun.y="median", geom="line", aes(group = Early_Euth)) +
+  labs(x = 'Day', y = ' Percent Relative to Day 0', title = 'Weight Loss') +
+  theme(legend.justification=c(1,0), legend.position=c(1,0.1), 
+        legend.title=element_blank()) +
+  stat_summary(fun.y= "median", geom = "point")
+
+
+cfu_plot <- meta_file %>% 
+  filter(day %in% c(0:10), !is.na(log_cfu) ) %>%
+  select(mouse_id, day, Early_Euth, log_cfu, cage_id) %>% 
+  ggplot(aes(x = factor(day), y = log_cfu, color = factor(Early_Euth))) + 
+  geom_jitter(width = 0.5, alpha = 0.2) + 
[…]

### 2016-10-04 code/CdGF_AUCRF.Rmd
add partial dependence plots

@@
 ## Can Day 0 Community members predict C. difficile CFU?
-```{r fig.height= 10, fig.width=10}
+```{r setup data, fig.height= 10, fig.width=10}
 
@@
 #random forest regression model to determine Day 0 OTU correlated with increased CFU
-seed <- 100316
+seed <- 1
 n_trees <- 2001
@@
 
-```{r fig.width=8, fig.height=8, results= 'hide'}
+```{r roc, fig.width=8, fig.height=8, results= 'hide'}
 #create dataframe for RF-classification
@@
 otu_euth_roc <- roc(Predict_early_euth_df$Euth_Early ~ otu_euth_probs[ , 2])
+euth_otus <- rf_early_aucrf$Xopt
 
-aucrf_data <- Predict_early_euth_df[,c('Euth_Early',rf_early_aucrf$Xopt)]
+aucrf_data <- Predict_early_euth_df[,c('Euth_Early',euth_otus)]
 
@@
 LOcO_df <- Predict_df %>% 
-                select(Euth_Early, cage_id, one_of(rf_early_aucrf$Xopt))
+                select(Euth_Early, cage_id, one_of(euth_otus))
 cage_list <- as.character(unique(LOcO_df$cage_id))
@@
 LOcO_probs <- c()
+LOcO_rf <- c()
 for(i in 1:length(cage_list)){
@@
   temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
+  LOcO_rf[[i]] <- temp_model$RFopt
   LOcO_probs[LOcO_df$cage_id == cage_list[i]] <- predict(temp_model$RFopt, test, type='prob')[,2]
@@
 
-```{r fig.width=10, fig.height=10}
+```{r importance, fig.width=10, fig.height=10}
 
 #Importance Plot
-euth_otus <- rf_early_aucrf$Xopt
 euth_tax <- get_tax(1, euth_otus, tax_file)
@@
 
+### Partial Dependency Plots
+###### Using optimal RF model generated from whole dataset
+
+```{r partial dependence, fig.width=10, fig.height=10}
+layout(matrix(c(1:20), nrow=5, byrow=TRUE))
+
+for (i in seq_along(euth_otus)) {
+  partialPlot(rf_early_aucrf$RFopt, aucrf_data, euth_otus[i], xlab='',
+                main=paste(euth_tax$tax_label[i]))
+}
+
+```
+
 Would adding initial increase/slope of CFU or day 1 CFU help predict?
@@
 
+* Of note, I tried a few different seeds, and one otu (Turicibacter OTU 120) was identified as a classifier of cage/inocula but wasnt in others and was identified as a classifier of Euthanization.
+Also we occasionally lose OTU 50 as a classifier
+
 ***  

### 2016-10-04 code/inocula_nmds.R
updated plots

@@ -89,11 +89,14 @@ combo_nmds <- merge(combo_nmds, combo_design, by.x='group', by.y='V1')
 
 #need 3D version of these 
 d0_donor <- read.table(file='data/raw/sample_donor.design', header = F)
+day0_donor_nmds <- merge(day0_nmds, d0_donor, by.x='group', by.y='V1')
 day0_3D_nmds <- merge(day0_3D, d0_donor, by.x='group', by.y='V1')
 
 d0_outcome <- read.table(file='data/raw/d0_only_outcome.design', header=F)
 day0_outcome_nmds <- merge(day0_donor_nmds, d0_outcome, by.x='group', by.y='V1')
 day0_outcome_nmds <- merge(day0_donor_nmds, d0_outcome, by.x='group', by.y='V1')
+
+day0_donor_nmds <- merge(day0_donor_nmds, d0_outcome, by.x='group', by.y='V1')
 
 #use this one
 day0_3D_outcome <- merge(day0_3D_nmds, d0_outcome, by.x='group', by.y='V1')

### 2016-10-04 code/mouse_weightloss_time.R

@@ -22,7 +22,16 @@ shared_file <- read.table("data/process/human_CdGF.an.unique_list.0.03.subsample
 tax_file <- read.table('data/process/gf_new.an.taxonomy', sep='\t',header = T, row.names = 1)
 
 
+
+
+
+
+
 #########All below is Nick's code that I am modifying ######
+
+meta_file <- meta_file[meta_file$cdiff_strain == '431', ]
+meta_file <- meta_file[meta_file$cage_id != 'NP1',]
+
 
 wt_ls_plot <- meta_file %>% 
   filter(day %in% c(0:10), !is.na(percent_weightLoss) ) %>%
@@ -30,7 +39,7 @@ wt_ls_plot <- meta_file %>%
   ggplot(aes(x = factor(day), y = percent_weightLoss, color = factor(Early_Euth))) + 
   geom_jitter(width = 0.5, alpha = 0.1) + 
   scale_color_manual(values = c('black', 'red'), breaks=c("FALSE", "TRUE"),
-                     labels=c("Persistent", "Severe")) +
+                     labels=c("Mild", "Severe")) +
   stat_summary(fun.y="median", geom="line", aes(group = Early_Euth)) +
   labs(x = 'Day', y = ' Percent Relative to Day 0', title = 'Weight Loss') +
   theme(legend.justification=c(1,0), legend.position=c(1,0.1), 
@@ -44,16 +53,137 @@ cfu_plot <- meta_file %>%
   ggplot(aes(x = factor(day), y = log_cfu, color = factor(Early_Euth))) + 
   geom_jitter(width = 0.5, alpha = 0.2) + 
   scale_color_manual(values = c('black', 'red'), breaks=c("FALSE", "TRUE"),
-                     labels=c("Persistent", "Severe")) +
+                     labels=c("Mild", "Severe")) +
   stat_summary(fun.y="median", geom="line", aes(group = Early_Euth)) +
   labs(x = 'Day', y = ' CFU (log10)', title = 'C. Difficile Colonization') +
   theme(legend.justification=c(1,0), legend.position=c(1,0.2), 
         legend.title=element_blank()) +
   stat_summary(fun.y= "median", geom = "point")
 
+cfu_cage_plot <- meta_file %>% 
+  filter(day %in% c(0:10), !is.na(log_cfu) ) %>%
[…]

### 2016-10-04 code/CdGF_AUCRF.Rmd
reduce features to decrease cage/inocula predictability

@@
 iters <- 100
-n_features <- 30
+n_features <- 65
 ```
@@
 otu_euth_roc <- roc(Predict_early_euth_df$Euth_Early ~ otu_euth_probs[ , 2])
-euth_otus <- rf_early_aucrf$Xopt
+euth_otu_feat <- rf_early_aucrf$Xopt
 
-aucrf_data <- Predict_early_euth_df[,c('Euth_Early',euth_otus)]
+aucrf_data <- Predict_early_euth_df[,c('Euth_Early',euth_otu_feat)]
 
@@
 LOcO_df <- Predict_df %>% 
-                select(Euth_Early, cage_id, one_of(euth_otus))
+                select(Euth_Early, cage_id, one_of(euth_otu_feat))
 cage_list <- as.character(unique(LOcO_df$cage_id))
@@
 
-```{r importance, fig.width=10, fig.height=10}
+```{r importance, fig.width=10, fig.height=5}
 
 #Importance Plot
+n_rf_features <- 5
+euth_otus <- euth_otu_feat[1:n_rf_features]
 euth_tax <- get_tax(1, euth_otus, tax_file)
@@
 
-```{r partial dependence, fig.width=10, fig.height=10}
-layout(matrix(c(1:20), nrow=5, byrow=TRUE))
+```{r partial dependence, fig.width=10, fig.height=5}
+layout(matrix(c(1:6), nrow=2, byrow=TRUE))
 

### 2016-10-05 code/CdGF_AUCRF.Rmd
add rel abund partial dependence overlay plots

@@
 #Importance Plot
-n_rf_features <- 5
+n_rf_features <- 9
 euth_otus <- euth_otu_feat[1:n_rf_features]
@@
 
-```{r partial dependence, fig.width=10, fig.height=5}
-layout(matrix(c(1:6), nrow=2, byrow=TRUE))
+```{r partial dependence, fig.width=10, fig.height=10}
+layout(matrix(c(1:9), nrow=3, byrow=TRUE))
 
-for (i in seq_along(euth_otus)) {
-  partialPlot(rf_early_aucrf$RFopt, aucrf_data, euth_otus[i], xlab='',
+partial_otu <- c()
+for (i in seq_along(euth_otu_feat)) {
+  partial_otu <- partialPlot(rf_early_aucrf$RFopt, aucrf_data, euth_otu_feat[i], xlab='',
                 main=paste(euth_tax$tax_label[i]))
+points(Predict_early_euth_df[Predict_early_euth_df$Euth_Early == 0 ,euth_otu_feat[i]], 
+       jitter(rep(0.75*(diff(range(partial_otu$y))) + min(partial_otu$y), 
+                  sum(Predict_early_euth_df$Euth_Early == 0)), 
+              amount = (0.1*(diff(range(partial_otu$y))))), 
+              pch = 21, bg = 'red')
+points(Predict_early_euth_df[Predict_early_euth_df$Euth_Early == 1 ,euth_otu_feat[i]], 
+       jitter(rep(0.25*(diff(range(partial_otu$y))) + min(partial_otu$y), 
+                  sum(Predict_early_euth_df$Euth_Early == 1)), 
+              amount = (0.1*(diff(range(partial_otu$y))))), 
+              pch = 21, bg = 'gray')
 }

### 2016-10-05 code/CdGF_AUCRF.Rmd
remove NP1 (unknown infection)

@@
 meta_file <- meta_file[meta_file$cdiff_strain == '431', ]
+
+# remove cage NP1 (no confirmed infection)
+meta_file <- meta_file[!meta_file$cage_id == 'NP1', ]
 
@@
   sampling <- sample(1:nrow(aucrf_data),nrow(aucrf_data),replace=F)
-  cv10f_probs <- rep(NA,nrow(aucrf_data))
-  for(i in seq(1,nrow(aucrf_data),5)){
+  cv10f_probs <- rep(NA,52)
+  for(i in seq(1,50,5)){
     train <- aucrf_data[sampling[-(i:(i+4))],]
@@
 #Importance Plot
-n_rf_features <- 9
-euth_otus <- euth_otu_feat[1:n_rf_features]
+#n_rf_features <- 9
+euth_otus <- euth_otu_feat#[1:n_rf_features]
 euth_tax <- get_tax(1, euth_otus, tax_file)

### 2016-10-05 code/CdGF_AUCRF.Rmd
clean up ROC plot file

@@
 #create dataframe for RF-classification
-meta_file$Euth_Early <- factor(ifelse(meta_file[ , 'Early_Euth'] == FALSE, 1, 0))
+meta_file$Euth_Early <- factor(ifelse(meta_file[ , 'Early_Euth'] == FALSE, 0, 1))
 
@@
 
-```{r importance, fig.width=10, fig.height=5}
+### Partial Dependency with Relative Abundance Overlayed Plots
+###### Using optimal RF model generated from whole dataset
+###### Plot labels contain MDA of each OTU
 
-#Importance Plot
+```{r rel abund partial depend, fig.width=10, fig.height=10}
+
 #n_rf_features <- 9
-euth_otus <- euth_otu_feat#[1:n_rf_features]
-euth_tax <- get_tax(1, euth_otus, tax_file)
+#euth_otus <- euth_otu_feat[1:n_rf_features]
+euth_tax <- get_tax(1, euth_otu_feat, tax_file)
 euth_tax$tax_label <- paste(euth_tax$tax, ' (', 
                             gsub('tu0*', 'TU ', rownames(euth_tax)), 
-                                   ') : ', round(rf_early_aucrf$ranking[euth_otus], 3), sep = '')
+                                   ') : ', round(rf_early_aucrf$ranking[euth_otu_feat], 3), sep = '')
 
-RF_features_polt <- Predict_early_euth_df %>% 
-  select(Euth_Early, one_of(euth_otus)) %>% 
-  gather('OTU', 'Value', -Euth_Early) %>% 
-  full_join(data.frame(OTU = names(rf_early_aucrf$ranking[euth_otus]), 
-               MDA =  rf_early_aucrf$ranking[euth_otus])) %>% 
-  ggplot( aes(x = reorder(factor(OTU), MDA), y = Value, fill = Euth_Early, dodge = Euth_Early)) +
-  theme(legend.title=element_blank(), legend.position = c(0.9,0.5),
-        panel.background = element_rect(fill = NA),
-        panel.grid.major = element_line(colour = NA),
-        panel.grid.minor = element_line(colour = NA)) +
-  geom_point(pch = 21, position = position_jitterdodge(), alpha = 0.4, size = 2) + coord_flip() +
-  scale_x_discrete(breaks = rownames(euth_tax), labels = euth_tax$tax_label) +
-  stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median,
-                 geom = "crossbar", width = 0.6, position=position_dodge(width=0.8)) + 
-  scale_fill_manual(values = c('red', 'black'), labels = c('Severe','Mild'),
-                    guide = guide_legend(reverse=TRUE))
+layout(matrix(c(1:9), ncol=3, byrow=TRUE))
 
-RF_features_polt + labs( y = 'Relative Abundance (%)\nClassification (OTU) : MDA', x = '', title = '')
-
-RF_features_polt + labs( y = 'Relative Abundance (%) - log10\nClassification (OTU) : MDA', x = '', title = '') +
-  scale_y_log10()
-```
-
-### Partial Dependency Plots
-###### Using optimal RF model generated from whole dataset
-
-```{r partial dependence, fig.width=10, fig.height=10}
-layout(matrix(c(1:9), nrow=3, byrow=TRUE))
-
-partial_otu <- c()
 for (i in seq_along(euth_otu_feat)) {
@@
                 main=paste(euth_tax$tax_label[i]))
+points(Predict_early_euth_df[Predict_early_euth_df$Euth_Early == 1 ,euth_otu_feat[i]], 
+       jitter(rep(0.25*(diff(range(partial_otu$y))) + min(partial_otu$y), 
+                  sum(Predict_early_euth_df$Euth_Early == 1)), 
+              amount = (0.1*(diff(range(partial_otu$y))))), 
+              pch = 21, bg = 'red')
 points(Predict_early_euth_df[Predict_early_euth_df$Euth_Early == 0 ,euth_otu_feat[i]], 
@@
               amount = (0.1*(diff(range(partial_otu$y))))), 
-              pch = 21, bg = 'red')
-points(Predict_early_euth_df[Predict_early_euth_df$Euth_Early == 1 ,euth_otu_feat[i]], 
-       jitter(rep(0.25*(diff(range(partial_otu$y))) + min(partial_otu$y), 
-                  sum(Predict_early_euth_df$Euth_Early == 1)), 
-              amount = (0.1*(diff(range(partial_otu$y))))), 
               pch = 21, bg = 'gray')
 }
-
+plot(NULL, xlim = c(0,1), ylim = c(0,1), xlab = '', ylab = '', axes = F)
+legend('topleft', c('Mild', 'Severe'), pch = 21, pt.bg = c('gray','red'), cex = 2)
 ```

### 2016-10-06 code/HumanCdGF_microbiota_predict_by_day.Rmd
remove code that removes cdiff otus

@@
 
-# identify c difficile otus
-cdiff_otu <- rownames(tax_file[grep('Clostridium_XI', tax_file$Taxonomy),])
 # remove samples not used or with other cdiff strains
@@
   left_join(rel_abund_shared_01) %>% 
-  select(-sample_id, -one_of(cdiff_otu))
+  select(-sample_id)
 

### 2016-10-06 code/HumanCdGF_microbiota_predict_by_day.Rmd
debug RF CFU error

@@
   meta_file %>% 
-  filter(cdiff_strain == 431) %>% 
+  filter(cdiff_strain == 431, !cage_id == 'NP1') %>% 
   select(mouse_id, cage_id, sample_id, day, cdiff_cfu, Early_Euth, log_cfu) %>% 
@@
 
-seed <- 18
+seed <- 1
 n_trees <- 2001
-set.seed(seed)
 
@@
       meta_file %>%
-        filter(day == 1) %>% 
+        filter(day == i) %>% 
         select(log_cfu, mouse_id) %>%
@@
       cfu_by_day_comm %>% 
-        filter(day == 1) %>% 
+        filter(day == i) %>% 
         select(-day)
     } 
-
+    set.seed(seed)
     rf_all <- randomForest(log_cfu ~ ., data = rf_data,
@@
     importance_sorted <- sort(importance(rf_all)[,1], decreasing = T)
+    set.seed(seed)
     rf_n_features <- randomForest(log_cfu ~ ., data=rf_data[ ,c( 'log_cfu',
@@
 rf_by_day_com_plots <- lapply(1:10, RandomForest_by_day, community_day_0 = F)
-
 ```

### 2016-10-07 code/HumanCdGF_predict_cfu.Rmd
initial all predictions plots (heatmap)

---
title: "humanCdGF_predict_cfu"
author: "Nicholas Lesniak"
date: "October 7, 2016"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE, fig.width=10, fig.height=5)
```

```{r load packages}
pack_used <- c('randomForest','miscTools','tidyr','ggplot2','dplyr', 'pROC', 'knitr', 'cowplot')
for (dep in pack_used){
  if (dep %in% installed.packages()[,"Package"] == FALSE){
    install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
                     quiet=TRUE);
  }
  library(dep, verbose=FALSE, character.only=TRUE)
}
```

```{r setup environment}
pack_used <- c('randomForest','miscTools','tidyr','ggplot2','dplyr', 'pROC', 'knitr', 'cowplot')
for (dep in pack_used){
  if (dep %in% installed.packages()[,"Package"] == FALSE){
    install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
                     quiet=TRUE);
  }
  library(dep, verbose=FALSE, character.only=TRUE)
}
```

```{r setup data}
# read in files
meta_file   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_all.an.cons.taxonomy'
tax_function <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R'

# read in files
meta_file   <- read.table(meta_file, sep = '\t', header = T)
shared_file <- read.table(shared_file, sep = '\t', header = T, row.names = 'sample_id')
tax_file <- read.table(tax_file, sep = '\t', header = T, row.names = 'OTU')
source(tax_function)

# remove samples not used or with other cdiff strains
meta_file <- 
  meta_file %>% 
  filter(cdiff_strain == 431, !cage_id == 'NP1') %>% 
  select(mouse_id, sample_id, day, log_cfu) %>% 
  filter(!is.na(log_cfu))

shared_list_days <- select(meta_file, -log_cfu)

# create shared with relative abundances over 1%
mean_otu_counts <- unique(apply(shared_file, 1, sum))
rel_abund_shared <- 100 * shared_file / mean_otu_counts
rel_abund_shared_01 <- rel_abund_shared[ , apply(rel_abund_shared, 2, max) > 1]
rel_abund_shared_01$sample_id <- rownames(rel_abund_shared_01)

seed <- 1
n_trees <- 2001
rf_plot_list <- list()
n_features <- 10
cutoff <- 30

heat.plot <- function(input_maxtrix, high_color = '#660000', low_color = '#FFFFFF', legend_title = NA) {
  long.df <- data.frame(input_maxtrix)
  colnames(long.df) <- c(1:10)
  long.df$community_day <- as.character(c(0:10))
  long.df %>% gather(cfu_day, variable, -community_day) %>% 
    ggplot(aes(cfu_day, community_day)) + 
    geom_tile(aes(fill = variable), color = 'white') +
    scale_fill_gradient(low = low_color, high = high_color, name = legend_title) +
    labs(x = 'Day (CFU)', y = 'Day (Community)') + 
    theme(axis.line.x = element_blank(), axis.line.y = element_blank(), axis.ticks = element_blank()) + 
    scale_x_discrete(expand = c(0, 0), limits = c('1','2','3','4','5','6','7','8','9','10')) + 
    scale_y_discrete(expand = c(0, 0), limits = c('0','1','2','3','4','5','6','7','8','9','10')) + 
    theme(legend.position = 'top', legend.key.width = unit(2, 'cm'))
}
```

```{r functions}
rsq <- matrix(NA, nrow = 11, ncol = 10)
mse <- matrix(NA, nrow = 11, ncol = 10)
otu_votes <- rep(0, ncol(rel_abund_shared_01))
names(otu_votes) <- colnames(rel_abund_shared_01)
importance.df <- c()
for(m in 0:10){
  sample_list <- shared_list_days %>% 
    filter(day == m) %>% 
    select(-day)
  community.df <- rel_abund_shared_01 %>% 
    right_join(sample_list) %>% 
    select(-sample_id)
    for(n in 1:10){
      cfu_shared.df <- meta_file %>% 
        filter(day == n) %>% 
        select(mouse_id, log_cfu) %>% 
        right_join(community.df) %>% 
        select(-mouse_id) %>% 
        filter(!is.na(log_cfu))

      set.seed(seed)
      rf_all <- randomForest(log_cfu ~ ., data = cfu_shared.df,
                           importance = TRUE, ntree = n_trees)
      importance_sorted <- sort(importance(rf_all)[,1], decreasing = T)
      top_important_OTU <- rownames(data.frame(head(importance_sorted, n_features)))
      set.seed(seed)
      rf_n_features <- randomForest(log_cfu ~ ., data=cfu_shared.df[ ,c( 'log_cfu',
                                                                 names(importance_sorted)[1:n_features])], 
                                  importance=TRUE, ntree=n_trees)
      rsq[m+1, n] <- round(tail(rf_n_features$rsq, 1), 2)
      mse[m+1, n] <- round(tail(rf_n_features$mse, 1), 2)
      otu_votes <- otu_votes + 1*(colnames(rel_abund_shared_01) %in% top_important_OTU)
      importance.df <- rbind(importance.df, c(RSQ = round(tail(rf_n_features$rsq, 1), 2), 
                      MSE = round(tail(rf_n_features$mse, 1), 2),
                      importance(rf_all)[,1]/max(importance(rf_all)[,1])))
    }
  }

importance.df <- data.frame(importance.df)
import_features <- importance.df[importance.df$RSQ >= 0.6 & importance.df$MSE <= 0.8, ]
import_features <- import_features[ , !colnames(import_features) %in% c('RSQ','MSE')]
med_import_feat <- apply(import_features, 2, median)
otus_above_20 <- import_features[,names(med_import_feat)[med_import_feat >= 0.2]]
top_n_overall <- names(otu_votes)[otu_votes >= cutoff]

rsq_n <- matrix(NA, nrow = 11, ncol = 10)
mse_n <- matrix(NA, nrow = 11, ncol = 10)
otu_votes_n <- rep(0, ncol(rel_abund_shared_01))
names(otu_votes_n) <- colnames(rel_abund_shared_01)

for(m in 0:10){
  sample_list <- shared_list_days %>% 
    filter(day == m) %>% 
    select(-day)
  community.df <- rel_abund_shared_01 %>% 
    right_join(sample_list) %>% 
    select(-sample_id)
  for(n in 1:10){
    cfu_shared.df <- meta_file %>% 
      filter(day == n) %>% 
      select(mouse_id, log_cfu) %>% 
      right_join(community.df) %>% 
      select(log_cfu, one_of(top_n_overall)) %>% 
      filter(!is.na(log_cfu))

    set.seed(seed)
    rf_all <- randomForest(log_cfu ~ ., data = cfu_shared.df,
                           importance = TRUE, ntree = n_trees)
    importance_sorted <- sort(importance(rf_all)[,1], decreasing = T)
    top_important_OTU <- rownames(data.frame(head(importance_sorted, n_features)))
    set.seed(seed)
    rf_n_features <- randomForest(log_cfu ~ ., data=cfu_shared.df[ ,c( 'log_cfu',
                                                                       names(importance_sorted)[1:n_features])], 
                                  importance=TRUE, ntree=n_trees)
    rsq_n[m+1, n] <- round(tail(rf_n_features$rsq, 1), 2)
    mse_n[m+1, n] <- round(tail(rf_n_features$mse, 1), 2)
    otu_votes_n <- otu_votes_n + 1*(colnames(rel_abund_shared_01) %in% top_important_OTU)
  }
}
taxa <- get_tax(1, names(otu_votes_n[otu_votes_n >= 1]), tax_file)
taxa_names <- paste0(taxa$tax, ' (', gsub('tu0*', 'TU ', rownames(taxa)),')')
```

### All OTUs input to Random Forest
```{r all otus, fig.width=10, fig.height=6}
ggdraw()+
  draw_plot(heat.plot(rsq, '#003300', legend_title = expression(paste('R'^'2'))), 0, 0, 0.5, 1) + 
  draw_plot(heat.plot(mse, 'white', '#000066', 'MSE'), 0.5, 0, 0.5, 1)
#plot(otu_votes[otu_votes >= 10], xlab = 'OTU', ylab = 'Freq')
```

### OTUs selected by >`r cutoff`% of models generated by all OTU models 
###### Using the following bacteria: `r taxa_names`

```{r top otus, fig.width=10, fig.height=6}

ggdraw()+
  draw_plot(heat.plot(rsq_n, '#003300', legend_title = expression(paste('R'^'2'))), 0, 0, 0.5, 1) + 
  draw_plot(heat.plot(mse_n, 'white', '#000066', 'MSE'), 0.5, 0, 0.5, 1)
#plot(otu_votes_n[otu_votes_n >= 1], xlab = 'OTU', ylab = 'Freq')
```

###### Re-runnning the program and only focusing on the OTUs predicting day 9/10 to see if there is consistence in the predictive features gives the following selection:  
Otu00004 Otu00005 Otu00015 Otu00019 Otu00030 Otu00199 Otu00200 Otu00250   
      14       17       13       15       13       16       17       19

```{r boruta feature selection, eval = F}
bor_all <- c()

for(m in 0:10){
  sample_list <- shared_list_days %>% 
    filter(day == m) %>% 
    select(-day)
  community.df <- rel_abund_shared_01 %>% 
    right_join(sample_list) %>% 
    select(-sample_id)
    for(n in 9:10){
      cfu_shared.df <- meta_file %>% 
        filter(day == n) %>% 
        select(mouse_id, log_cfu) %>% 
        right_join(community.df) %>% 
        select(-mouse_id) %>% 
        filter(!is.na(log_cfu))

      set.seed(seed)
      bor_all[[(m*2)+(n-8)]] <- Boruta(log_cfu ~ ., data = cfu_shared.df, maxRuns = 2000)
    }
  }

boruta.votes <- rep(0, ncol(rel_abund_shared_01))
names(boruta.votes) <- colnames(rel_abund_shared_01)
for(i in 1:22){
  boruta.votes <- boruta.votes + 1*(bor_all[[i]]$finalDecision   == 'Confirmed')
}
plot(boruta.votes)
boruta.votes[boruta.votes >= 16]

colnames(otus_above_20)
```

Boruta Confirmed the following OTUs as important for predicting day 9/10 cfu:   
Otu00004 Otu00005 Otu00015 Otu00030 Otu00081 Otu00199 Otu00200 Otu00250 Otu00297  
      21       22       21       21       21       22       22       22       20

Selecting OTUs through collecting the features from the most predictive community/cfu models (R^2 >= 0.6 and MSE <= 0.8), then converting all % Increase in MSE to relative values and taking the median value of of each OTU, then selecting OTUs that fall above the median value results in the following OTUs:  
"Otu00001" "Otu00002" "Otu00004" "Otu00005" "Otu00012" "Otu00014" "Otu00015" "Otu00016" "Otu00019" "Otu00028" "Otu00030" "Otu00048" "Otu00081" "Otu00101" "Otu00118" "Otu00199" "Otu00200" "Otu00250" "Otu00297"
```{r }

### 2016-10-10 code/HumanCdGF_predict_cfu.Rmd
remove early euth samples

@@
   meta_file %>% 
-  filter(cdiff_strain == 431, !cage_id == 'NP1') %>% 
+  filter(Early_Euth == FALSE, cdiff_strain == 431, !cage_id == 'NP1') %>% 
   select(mouse_id, sample_id, day, log_cfu) %>% 
@@
 
-### All OTUs input to Random Forest
+### All OTUs input to Random Forest - without mice euthanized early
 ```{r all otus, fig.width=10, fig.height=6}
@@
 
-###### Re-runnning the program and only focusing on the OTUs predicting day 9/10 to see if there is consistence in the predictive features gives the following selection:  
+###### Re-runnning the program and only focusing on the OTUs predicting day 9/10 to see if there is consistence in the predictive features gives the following selection  (including all samples):  
 Otu00004 Otu00005 Otu00015 Otu00019 Otu00030 Otu00199 Otu00200 Otu00250   

### 2016-10-10 code/HumanCdGF_predict_cfu.Rmd
only plot RSQ for reduced features

@@
     scale_x_discrete(expand = c(0, 0), limits = c('1','2','3','4','5','6','7','8','9','10')) + 
-    scale_y_discrete(expand = c(0, 0), limits = c('0','1','2','3','4','5','6','7','8','9','10')) + 
+    scale_y_discrete(expand = c(0, 0), limits = rev(c('0','1','2','3','4','5','6','7','8','9','10'))) + 
     theme(legend.position = 'top', legend.key.width = unit(2, 'cm'))
@@
 ```{r all otus, fig.width=10, fig.height=6}
-ggdraw()+
-  draw_plot(heat.plot(rsq, '#003300', legend_title = expression(paste('R'^'2'))), 0, 0, 0.5, 1) + 
-  draw_plot(heat.plot(mse, 'white', '#000066', 'MSE'), 0.5, 0, 0.5, 1)
+heat.plot(rsq, '#003300', legend_title = expression(paste('R'^'2')))
 #plot(otu_votes[otu_votes >= 10], xlab = 'OTU', ylab = 'Freq')
@@
 ```{r top otus, fig.width=10, fig.height=6}
-
-ggdraw()+
-  draw_plot(heat.plot(rsq_n, '#003300', legend_title = expression(paste('R'^'2'))), 0, 0, 0.5, 1) + 
-  draw_plot(heat.plot(mse_n, 'white', '#000066', 'MSE'), 0.5, 0, 0.5, 1)
+heat.plot(rsq_n, '#003300', legend_title = expression(paste('R'^'2')))
 #plot(otu_votes_n[otu_votes_n >= 1], xlab = 'OTU', ylab = 'Freq')

### 2016-10-10 code/HumanCdGF_predict_cfu.Rmd
reduce samples to check model improvement

@@
 # read in files
-meta_file   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
+meta_data   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
 shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
@@
 # read in files
-meta_file   <- read.table(meta_file, sep = '\t', header = T)
+meta_data   <- read.table(meta_data, sep = '\t', header = T)
 shared_file <- read.table(shared_file, sep = '\t', header = T, row.names = 'sample_id')
@@
 meta_file <- 
-  meta_file %>% 
+  meta_data %>% 
   filter(Early_Euth == FALSE, cdiff_strain == 431, !cage_id == 'NP1') %>% 
@@
 
-### All OTUs input to Random Forest - without mice euthanized early
-```{r all otus, fig.width=10, fig.height=6}
-heat.plot(rsq, '#003300', legend_title = expression(paste('R'^'2')))
-#plot(otu_votes[otu_votes >= 10], xlab = 'OTU', ylab = 'Freq')
-```
-
 ### OTUs selected by >`r cutoff`% of models generated by all OTU models 
@@
 
-```{r top otus, fig.width=10, fig.height=6}
+```{r top otus, fig.width=6, fig.height=6}
 heat.plot(rsq_n, '#003300', legend_title = expression(paste('R'^'2')))
@@
 ```
+
+### CFU and Rel abundance plots
+
+```{r fig.height=5, fig.width=8}
+meta_data %>% 
+  filter(!is.na(day), !cage_id == 'NP1', 
+         Early_Euth == FALSE, cdiff_strain == 431) %>% 
+  ggplot(aes(x = factor(day), y = log_cfu, color = cage_id)) + 
+    geom_line(aes(group = mouse_id), size=.25, alpha=0.4) + geom_point() +
+    stat_summary(fun.y=mean, geom="line", aes(group = cage_id)) + 
+    theme_bw() + labs(x = 'Day', y = 'C Diff CFU (log10)') +
+    theme(legend.justification=c(1,0), legend.position=c(1,0), 
+          legend.title=element_blank())
+```
+
+How does removing the low samples (Day 8 and CFUs around 0) affect prediction?
+
+```{r identify reason for lowered R2, fig.height=5, fig.width=8}
+log_cutoff <- 2.5
+
+meta_data %>% 
+  filter(!sample_id == 'OUT2-2533-D8', log_cfu >= log_cutoff) %>% 
+  filter(!is.na(day), !cage_id == 'NP1', 
+         Early_Euth == FALSE, cdiff_strain == 431) %>% 
+  ggplot(aes(x = factor(day), y = log_cfu, color = cage_id)) + 
+    geom_line(aes(group = mouse_id), size=.25, alpha=0.4) + geom_point() +
+    stat_summary(fun.y=mean, geom="line", aes(group = cage_id)) + 
+    theme_bw() + labs(x = 'Day', y = 'C Diff CFU (log10)') +
+    theme(legend.justification=c(1,0), legend.position=c(1,0), 
+          legend.title=element_blank()) + scale_y_continuous(limits = c(0,9))
+
+meta_reduced <- meta_data %>% 
+  filter(!sample_id == 'OUT2-2533-D8', log_cfu >= log_cutoff) %>% 
+  filter(Early_Euth == FALSE, cdiff_strain == 431, !cage_id == 'NP1') %>% 
+  select(mouse_id, sample_id, day, log_cfu) %>% 
+  filter(!is.na(log_cfu))
+
+rsq_r <- matrix(NA, nrow = 11, ncol = 10)
+otu_votes_r <- rep(0, ncol(rel_abund_shared_01))
+names(otu_votes_r) <- colnames(rel_abund_shared_01)
+
+for(m in 0:10){
+  sample_list <- shared_list_days %>% 
+    filter(day == m) %>% 
+    select(-day)
+  community.df <- rel_abund_shared_01 %>% 
+    right_join(sample_list) %>% 
+    select(-sample_id)
+  for(n in 1:10){
+    cfu_shared.df <- meta_reduced %>% 
+      filter(day == n) %>% 
+      select(mouse_id, log_cfu) %>% 
+      right_join(community.df) %>% 
+      select(log_cfu, one_of(top_n_overall)) %>% 
+      filter(!is.na(log_cfu))
+
+    set.seed(seed)
+    rf_top <- randomForest(log_cfu ~ ., data = cfu_shared.df,
+                           importance = TRUE, ntree = n_trees)
+    importance_sorted <- sort(importance(rf_top)[,1], decreasing = T)
+    top_important_OTU <- rownames(data.frame(head(importance_sorted, n_features)))
+    set.seed(seed)
+    rf_n_features <- randomForest(log_cfu ~ ., data=cfu_shared.df[ ,c( 'log_cfu',
+                                                                       names(importance_sorted)[1:n_features])], 
+                                  importance=TRUE, ntree=n_trees)
+    importance_final <- sort(importance(rf_n_features)[,1], decreasing = T)
+    top_OTU_final <- rownames(data.frame(head(importance_final, n_features)))
+    rsq_r[m+1, n] <- round(tail(rf_n_features$rsq, 1), 2)
+    otu_votes_r <- otu_votes_r + 1*(colnames(rel_abund_shared_01) %in% top_OTU_final)
+  }
+}
+taxa_r <- get_tax(1, names(otu_votes_r[otu_votes_r >= 1]), tax_file)
+taxa_names_r <- paste0(taxa$tax, ' (', gsub('tu0*', 'TU ', rownames(taxa)),')')
+```
+
+```{r fig.width=6,fig.height=6}
+heat.plot(rsq_r, '#003300', legend_title = expression(paste('R'^'2')))
+```
+
+Removing these data points improves predictability for day 6 and 8 but not the earlier days.
+If I increase the cutoff to remove all CFU below 4, there is still no improvement and even in some cases decreased performance.
+
+### Is there any abnormalities in the OTUs of any these days?
+
+```{r plot features,fig.width=8, fig.height=20}
+meta_data %>% 
+  filter(!sample_id == 'OUT2-2533-D8', log_cfu >= log_cutoff) %>% 
+  filter(!cage_id == 'NP1', Early_Euth == FALSE, cdiff_strain == 431) %>% 
+  full_join(rel_abund_shared_01) %>% 
+  select(day, sample_id, mouse_id, cage_id, 
+         one_of(names(otu_votes_r[otu_votes_r >= 1]))) %>% 
+  gather(otu, rel_abund, Otu00002:Otu00250) %>% 
+  filter(!is.na(day)) %>% 
+  ggplot(aes(x=factor(day),y=rel_abund, color = cage_id)) +
+    geom_line(aes(group = mouse_id), size=.25, alpha=0.4) + geom_point() +
+    stat_summary(fun.y=mean, geom="line", aes(group = cage_id)) + 
+    facet_grid(otu ~., scales='free_y') +
+    theme_bw() + labs(x = 'Day', y = 'Relative Abundance (%)')
+```
+
+***
+
+***
 

### 2016-10-10 code/HumanCdGF_predict_cfu.Rmd
add median CFU prediction

@@
     geom_tile(aes(fill = variable), color = 'white') +
-    scale_fill_gradient(low = low_color, high = high_color, name = legend_title) +
+    scale_fill_gradient(low = low_color, high = high_color, name = legend_title, limits = c(0,1)) +
     labs(x = 'Day (CFU)', y = 'Day (Community)') + 
@@
 Removing these data points improves predictability for day 6 and 8 but not the earlier days.
-If I increase the cutoff to remove all CFU below 4, there is still no improvement and even in some cases decreased performance.
+If I increase the cutoff to remove all CFU below 5, there is a slight increase in day 3 cfu classification at the cost of decrease day 1 cfu classification.
 
@@
     theme_bw() + labs(x = 'Day', y = 'Relative Abundance (%)')
+```
+
+### Does taking the median CFU improve predicatbility?
+
+```{r predict median CFU, fig.width=6,fig.height=6}
+
+rsq_med <- matrix(NA, nrow = 11, ncol = 10)
+otu_votes_med <- rep(0, ncol(rel_abund_shared_01))
+names(otu_votes_med) <- colnames(rel_abund_shared_01)
+for(n in 1:10){
+med_log_cfu <- meta_data %>% 
+  filter(day %in% c(n:10), !cage_id == 'NP1', 
+         Early_Euth == FALSE, cdiff_strain == 431) %>% 
+  select(mouse_id, cage_id, log_cfu) %>% 
+  group_by(mouse_id, cage_id) %>% 
+  summarise(log_cfu_med = median(log_cfu)) %>% ungroup()
+for(m in 0:10){
+  sample_list <- shared_list_days %>% 
+    filter(day == m) %>% 
+    select(-day)
+  community.df <- rel_abund_shared_01 %>% 
+    right_join(sample_list) %>% 
+    select(-sample_id)
+  cfu_shared.df <- med_log_cfu %>% 
+      select(mouse_id, log_cfu_med) %>% 
+      right_join(community.df) %>% 
+      select(-mouse_id) %>% 
+#      select(log_cfu_med, one_of(top_n_overall)) %>% 
+      filter(!is.na(log_cfu_med))
+    set.seed(seed)
+    rf_top <- randomForest(log_cfu_med ~ ., data = cfu_shared.df,
+                           importance = TRUE, ntree = n_trees)
+    importance_sorted <- sort(importance(rf_top)[,1], decreasing = T)
+    top_important_OTU <- rownames(data.frame(head(importance_sorted, n_features)))
+    set.seed(seed)
+    rf_n_features <- randomForest(log_cfu_med ~ ., data=cfu_shared.df[ ,c( 'log_cfu_med',
+                                                                       names(importance_sorted)[1:n_features])], 
+                                  importance=TRUE, ntree=n_trees)
+    importance_final <- sort(importance(rf_n_features)[,1], decreasing = T)
+    top_OTU_final <- rownames(data.frame(head(importance_final, n_features)))
+    rsq_med[m+1, n] <- round(tail(rf_n_features$rsq, 1), 2)
+    otu_votes_med <- otu_votes_med + 1*(colnames(rel_abund_shared_01) %in% top_OTU_final)
+}
+}
+
+heat.plot(rsq_med, '#003300', legend_title = expression(paste('R'^'2'))) + 
+  ggtitle('Day CFU is median value of Day n to Day 10')
+
 ```

### 2016-10-11 code/mouse_d0_community.R
added paper documents

@@ -4,7 +4,7 @@
 #
 # Create microbiota files for day 0 and plot different ways
 #
-#
+# ## this code is broken for some reason!!!! ack. the selection part.
 #
 #############
 
@@ -31,7 +31,12 @@ OTU_list <- colnames(rel_abund)[OTUs_1]
 
 #df of OTUs with abundances >1% - by cage and inoculum
 rel_abund_cage <- rel_abund[!meta_file$cage_id=='inoculum',OTUs_1]
+test_cage <- rel_abund[meta_file$mouse_id!='inoculum_NA', OTUs_1]
 inoculum_rel_abund <- rel_abund[meta_file$cage_id=='inoculum',OTUs_1]
+
+#hack for U01 meeting to remove donors
+rel_abund_cage <- rel_abund_cage[-(1:10),]
+rel_abund <- rel_abund[-(1:10),]
 
 #df of OTUs w abundances >1% 
 rel_abund_d0 <- rel_abund[meta_file$day == 0, OTUs_1]
@@ -68,8 +73,8 @@ barplot(t(taxonomy_phylum), ylab='Relative Abundance', main="Taxonomic compositi
 #add a column for cages to the df
 cages <- meta_file$cage_id[meta_file$day==0]
 cages <- na.omit(cages)
-rel_d0[82] <- cages
-colnames(rel_d0)[82] <- "cage"
+rel_d0[(ncol(rel_d0)+1)] <- cages
+colnames(rel_d0)[ncol(rel_d0)] <- "cage"
 
 #or could merge subsetted metafile for cage ids and donor and merge by mouse id. rel_d0$cage <- cage_iDs
 

### 2016-10-11 code/HumanCdGF_predict_cfu.Rmd
add plot to compare median colonization to day 0

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE, fig.width=10, fig.height=5)
-```
-
-```{r load packages}
-pack_used <- c('randomForest','miscTools','tidyr','ggplot2','dplyr', 'pROC', 'knitr', 'cowplot')
-for (dep in pack_used){
-  if (dep %in% installed.packages()[,"Package"] == FALSE){
-    install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
-                     quiet=TRUE);
-  }
-  library(dep, verbose=FALSE, character.only=TRUE)
-}
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE, fig.width=7, fig.height=8)
 ```
@@
 seed <- 1
-n_trees <- 2001
+n_trees <- 1001
 rf_plot_list <- list()
@@
 
+### Reframing the question for predicting CFU
+#### 1. Can Day 0 community predict the median steady-state-colonized CFU?
+#### 2. Can Day 0 community predict the rate of colonization?
+
+```{r predict median CFU, fig.width=6,fig.height=6}
+
+Cage_dep_df <- meta_data %>% 
+  filter(day == 0) %>% 
+  select(human_source, sample_id, cage_id) %>% 
+  left_join(rel_abund_shared_01)
+Predict_cage_df <- select(Cage_dep_df, -human_source, -sample_id)
+Predict_source_df <- select(Cage_dep_df, -cage_id, -sample_id)
+Predict_cage_df$cage_id <- factor(Predict_cage_df$cage_id)
+Predict_source_df$human_source <- factor(Predict_source_df$human_source)
+
+# identify OTUs that predict cage/inocula
+set.seed(seed)
+rf_cage <- randomForest(cage_id ~ ., data=Predict_cage_df, importance=TRUE, ntree=n_trees)
+set.seed(seed)
+rf_source <- randomForest(human_source ~ ., data=Predict_source_df, importance=TRUE, ntree=n_trees)
+features_source <- head(sort(rf_source$importance[ , 'MeanDecreaseAccuracy'], decreasing = T), 65)
+features_cage <- head(sort(rf_cage$importance[ , 'MeanDecreaseAccuracy'], decreasing = T), 65)
+features_all <- unique(c(names(features_cage), names(features_source)))
+
+colonized_cfu <- meta_data %>% 
+  filter(day %in% c(4:10), !cage_id == 'NP1', 
+         Early_Euth == FALSE, cdiff_strain == 431) %>% 
+  select(mouse_id, cage_id, log_cfu) %>% 
+  group_by(mouse_id, cage_id) %>% 
+  summarise(log_cfu_med = median(log_cfu)) %>% 
+  ungroup()
+sample_list_d0 <- shared_list_days %>% 
+    filter(day == 0) %>% 
+    select(-day)
+community_d0.df <- rel_abund_shared_01 %>% 
+    right_join(sample_list_d0) %>% 
+    select(-sample_id)
+cfu_shared_colonized.df <- colonized_cfu %>% 
+    right_join(community_d0.df) %>% 
+    select(-mouse_id, -cage_id, -one_of(features_all)) %>% 
+    filter(!is.na(log_cfu_med))
+
+test <- Boruta(log_cfu_med ~ ., data = cfu_shared_colonized.df, maxRuns = 5000)
+
+cfu_shared_colonized <- cfu_shared_colonized.df %>% 
+  select(log_cfu_med, one_of(names(test$finalDecision[test$finalDecision %in% c('Confirmed', 'Tentative')])))
+
+set.seed(seed)
+rf_all <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df,
+                           importance = TRUE, ntree = n_trees, mtry = 14)
+set.seed(seed)
+rf_bor <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized,
+                           importance = TRUE, ntree = n_trees, mtry = 2)
+
+plot(importance_sorted/max(importance_sorted))
+importance_sorted <- sort(importance(rf_bor)[,1], decreasing = T)
+top_important_OTU <- rownames(data.frame(importance_sorted[importance_sorted/max(importance_sorted) > 0.6]))
+
+set.seed(seed)
+rf_n_features <- randomForest(log_cfu_med ~ ., data=cfu_shared_colonized.df %>% 
+                                    select(log_cfu_med, one_of(top_important_OTU)),
+                                           importance=TRUE, ntree=n_trees, mtry = 2)
+
+importance_final <- sort(importance(rf_n_features)[,1], decreasing = T)
+
+rf_prediction <- data.frame(observed = cfu_shared_colonized.df$log_cfu_med, 
+                       predicted = rf_n_features$predicted)
+rf_prediction <- merge(rf_prediction, colonized_cfu, by.x = 'observed', by.y = 'log_cfu_med')
+predict_plot_day <- 
+  ggplot(data = rf_prediction, aes(x=observed, y=predicted, color = cage_id)) + geom_point() + 
+    labs(x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)',
+         title = "Random Forest Prediction of Colonized C Difficile CFU from Day 0 Community") + 
+    theme_bw() + xlim(0,9) + ylim(0,9) + 
+    annotate('text', x=0.15*(max(rf_prediction$observed)), 
+             y=0.95*(max(rf_prediction$predicted)), 
+             label=paste0('R^2 = ', round(tail(rf_n_features$rsq, 1), 2)))
+
+importance.df <- data.frame(otu = names(importance_final),
+                            importance = importance_final)
+otu_taxa <- get_tax(1, importance.df$otu, tax_file)
+
+importance_plot_day <- 
+  ggplot(data = importance.df, aes(x = factor(otu), y = importance)) + 
+        geom_point() + scale_x_discrete(limits = rev(importance.df$otu),
+                                        labels = rev(paste(otu_taxa[,1],' (',
+                                                       rownames(otu_taxa),')',
+                                                       sep=''))) +
+        labs(x= '', y = '% Increase in MSE') + theme_bw() + coord_flip()
+
+```
+
 ***

### 2016-10-11 code/HumanCdGF_predict_cfu.Rmd
remove boruta

@@
 
-test <- Boruta(log_cfu_med ~ ., data = cfu_shared_colonized.df, maxRuns = 5000)
-
-cfu_shared_colonized <- cfu_shared_colonized.df %>% 
-  select(log_cfu_med, one_of(names(test$finalDecision[test$finalDecision %in% c('Confirmed', 'Tentative')])))
-
 set.seed(seed)
@@
                            importance = TRUE, ntree = n_trees, mtry = 14)
-set.seed(seed)
-rf_bor <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized,
-                           importance = TRUE, ntree = n_trees, mtry = 2)
 
-plot(importance_sorted/max(importance_sorted))
-importance_sorted <- sort(importance(rf_bor)[,1], decreasing = T)
+importance_sorted <- sort(importance(rf_all)[,1], decreasing = T)
 top_important_OTU <- rownames(data.frame(importance_sorted[importance_sorted/max(importance_sorted) > 0.6]))

### 2016-10-11 code/inocula_nmds.R
paper updates

@@ -16,12 +16,8 @@ meta_file <- read.table('data/process/human_CdGF_metadata.txt', sep='\t',header 
 shared <- read.table('data/process/human_CdGF.an.unique_list.0.03.subsample.shared', sep='\t',header=T)
 full_shared <- read.table('data/process/humanCdGF_full.shared', sep='\t', header = T)
 
-
-#orig_shared <- read.table('data/mothur/gf_all.an.0.03.subsample.shared', sep='\t', header = T)
-#remove dashes from group names so the next stuff will work
-#orig_shared[2] <- gsub("-", "", orig_shared$Group)
-
-#make shared file look like a true shared:
+#shared files are processed in a make_meta script to remove erroneous samples 
+#so we need format processed shared to look like a true shared file:
 full_shared_group <- cbind(label=0.03, full_shared[1], numOtus=5671, full_shared[2:ncol(full_shared)])
 colnames(full_shared_group)[2] <- 'Group'
 
@@ -29,7 +25,7 @@ colnames(full_shared_group)[2] <- 'Group'
 inocula <- rownames(meta_file)[meta_file$cage_id=='inoculum']
 inocula_shared <- full_shared[full_shared$sample_id %in% inocula,]
 
-#subset shared to be day 0 only- will use later for comparison 
+#subset shared to be day 0 only 
 day0 <- rownames(meta_file)[meta_file$day==0]
 day0_shared <- full_shared[full_shared$sample_id %in% day0,]
 
@@ -37,7 +33,7 @@ day0_shared <- full_shared[full_shared$sample_id %in% day0,]
 inocula_shared <- cbind(label=0.03, inocula_shared[1], numOtus=5671, inocula_shared[2:ncol(inocula_shared)])
 colnames(inocula_shared)[2] <- 'Group'
 
-#for day 0 also
+#for day 0 also.. probably could make this a function
 day0_shared <- cbind(label=0.03, day0_shared[1], numOtus=5671, day0_shared[2:ncol(day0_shared)])
 colnames(day0_shared)[2] <- 'Group'
 

### 2016-10-11 code/thetayc_time.R

@@ -16,14 +16,9 @@ source('code/read.dist.R')
 
 #read in files
 meta_file <- read.table('data/process/human_CdGF_metadata.txt', sep='\t',header = T, row.names = 2)
-shared <- read.table('data/process/human_CdGF.an.unique_list.0.03.subsample.shared', sep='\t',header=T)
 theta <- read.dist(file='data/process/shared_subset_2194.thetayc.0.03.lt.dist', input="lt", make.square=T, diag=0)
 
 
-
-#then maybe write it as a function 
-
-#################THE RIGHT ONE!!!#######
 #compares day0 to every subsequent day. this asks, how different are the communities from the start? can this be linked to outcome?
 
 mouseIDVector <- as.character(unique(meta_file$mouse_id))
@@ -45,9 +40,7 @@ for (k in 1:10){
   }
 }
 
-#need to melt data to get it in plottable form. add mouse id and cage id column. WOO. 
-
-
+#need to melt data to get it in plottable form. add mouse id and cage id column. woo! 
 melted <- melt(stability) %>% mutate(mouse_id = rep(rownames(stability), 10))
 melted$cage_id <- sapply((strsplit(melted$mouse_id, "_")), '[', 1)
 
@@ -88,10 +81,7 @@ for(j in melted$cage_id){
   }
 }
 do.call(grid.arrange, p)
-#build as function 
 
-
-#now write version where you compare each day to previous day. this asks question: how stable is community over time?
 
 
 

### 2016-10-11 code/HumanCdGF_predict_cfu.Rmd
add prediction of steadystate CFU from day 0

--- 2016-10-10 code/HumanCdGF_predict_cfu.Rmd
@@
 
-```{r predict median CFU, fig.width=6,fig.height=6}
+```{r predict median CFU heatmap, fig.width=6,fig.height=6}
 
@@
 
+```
+
+### Reframing the question for predicting CFU
+#### 1. Can Day 0 community predict the median steady-state-colonized CFU?
+#### 2. Can Day 0 community predict the rate of colonization?
+
+```{r predict median CFU, fig.width=10,fig.height=10}
+
+Cage_dep_df <- meta_data %>% 
+  filter(day == 0) %>% 
+  select(human_source, sample_id, cage_id) %>% 
+  left_join(rel_abund_shared_01)
+Predict_cage_df <- select(Cage_dep_df, -human_source, -sample_id)
+Predict_source_df <- select(Cage_dep_df, -cage_id, -sample_id)
+Predict_cage_df$cage_id <- factor(Predict_cage_df$cage_id)
+Predict_source_df$human_source <- factor(Predict_source_df$human_source)
+
+# identify OTUs that predict cage/inocula
+set.seed(seed)
+rf_cage <- randomForest(cage_id ~ ., data=Predict_cage_df, importance=TRUE, ntree=n_trees)
+set.seed(seed)
+rf_source <- randomForest(human_source ~ ., data=Predict_source_df, importance=TRUE, ntree=n_trees)
+features_source <- head(sort(rf_source$importance[ , 'MeanDecreaseAccuracy'], decreasing = T), 65)
+features_cage <- head(sort(rf_cage$importance[ , 'MeanDecreaseAccuracy'], decreasing = T), 65)
+cage_source_features <- unique(c(names(features_cage), names(features_source)))
+
+colonized_cfu <- meta_data %>% 
+  filter(day %in% c(4:10), !cage_id == 'NP1', 
+         Early_Euth == FALSE, cdiff_strain == 431) %>% 
+  select(mouse_id, cage_id, log_cfu) %>% 
+  group_by(mouse_id, cage_id) %>% 
+  summarise(log_cfu_med = median(log_cfu)) %>% 
+  ungroup()
+sample_list_d0 <- shared_list_days %>% 
+    filter(day == 0) %>% 
+    select(-day)
+community_d0.df <- rel_abund_shared_01 %>% 
+    right_join(sample_list_d0) %>% 
+    select(-sample_id)
+cfu_shared_colonized.df <- colonized_cfu %>% 
+    right_join(community_d0.df) %>% 
+    select(-mouse_id, -cage_id, -one_of(cage_source_features)) %>% 
+    filter(!is.na(log_cfu_med))
+
+set.seed(seed)
+rf_all <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df,
+                           importance = TRUE, ntree = n_trees, mtry = 14)
+
+importance_sorted <- sort(importance(rf_all)[,1], decreasing = T)
+top_important_OTU <- rownames(data.frame(importance_sorted[importance_sorted/max(importance_sorted) > 0.6]))
+
+top_colonized_df <- data.frame(cfu_shared_colonized.df %>% 
+  select(log_cfu_med, one_of(top_important_OTU)))
+
+set.seed(seed)
+rf_n_features <- randomForest(log_cfu_med ~ ., data=top_colonized_df,
+                              importance=TRUE, ntree=n_trees, mtry = 2,
+                              keep.forest=TRUE)
+
+importance_final <- sort(importance(rf_n_features)[,1], decreasing = T)
+
+rf_prediction <- data.frame(observed = cfu_shared_colonized.df$log_cfu_med, 
+                       predicted = rf_n_features$predicted)
+rf_prediction <- merge(rf_prediction, colonized_cfu, by.x = 'observed', by.y = 'log_cfu_med')
+
+ggplot(data = rf_prediction, aes(x=observed, y=predicted, color = cage_id)) + geom_point() + 
+  labs(x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)',
+       title = "Random Forest Prediction of Colonized C Difficile CFU from Day 0 Community") + 
+  theme_bw() + xlim(0,9) + ylim(0,9) + 
+  annotate('text', x=0.15*(max(rf_prediction$observed)), 
+           y=0.95*(max(rf_prediction$predicted)), 
+           label=paste0('R^2 = ', round(tail(rf_n_features$rsq, 1), 2)))
+
+importance.df <- data.frame(otu = names(importance_final),
+                            importance = importance_final,
+                            tax = get_tax(1, names(importance_final), tax_file))
+tax_label <- paste0(importance.df$tax, ' (', 
+                    gsub('tu0*', 'TU ', importance.df$otu), 
+                    ') : ', 
+                    round(importance.df$importance, 3))
+
+layout(matrix(c(1:9), ncol=3, byrow=TRUE))
+
+cage_color <- rainbow(length(unique(rf_prediction$cage_id)))
+for (i in seq_along(importance_final)) {
+  partial_otu <- 
+    partialPlot(rf_n_features, top_colonized_df, names(importance_final)[i],
+                xlab='', main=paste(tax_label[i]), 
+                ylim = c(min(top_colonized_df$log_cfu_med),max(top_colonized_df$log_cfu_med)))
+  for(m in seq_along(unique(rf_prediction$cage_id))){
+    cage <- unique(rf_prediction$cage_id)[m]
+    points(y = top_colonized_df[rf_prediction$cage_id %in% cage, 'log_cfu_med'],
+         x = top_colonized_df[rf_prediction$cage_id %in% cage , names(importance_final)[i]], 
+         pch = 21, bg = cage_color[m])
+  }
+}
 ```

### 2016-10-12 code/HumanCdGF_predict_cfu.Rmd
colonization rate prediction

@@
 #### 1. Can Day 0 community predict the median steady-state-colonized CFU?
-#### 2. Can Day 0 community predict the rate of colonization?
 
@@
 
+#### 2. Can Day 0 community predict the rate of colonization?
+
+```{r predict colonization rate, fig.width=10,fig.height=10}
+
+meta_data %>% 
+  filter(!is.na(day), !cage_id == 'NP1', 
+         Early_Euth == FALSE, cdiff_strain == 431) %>% 
+  ggplot(aes(x = factor(day), y = log_cfu, color = cage_id)) + 
+    geom_line(aes(group = mouse_id), size=.25, alpha=0.4) + geom_point() +
+    stat_summary(fun.y=mean, geom="line", aes(group = cage_id)) + 
+    theme_bw() + labs(x = 'Day', y = 'C Diff CFU (log10)') +
+    theme(legend.justification=c(1,0), legend.position=c(1,0), 
+          legend.title=element_blank())
+
+mouse_list <- meta_data %>% 
+  filter(!is.na(day), !cage_id == 'NP1', 
+         Early_Euth == FALSE, cdiff_strain == 431) %>% 
+  select(mouse_id) %>% 
+  unique()
+mouse_list <- as.character(mouse_list$mouse_id)
+
+col_rate <- c()
+for(i in mouse_list){
+  test <- meta_data %>% 
+    filter(!is.na(day), mouse_id == i) %>% 
+    select(day, log_cfu) %>% 
+    filter(!is.na(log_cfu))
+  if(1 %in% test$day){
+    colonization_ratio <- test[test$day==1, 'log_cfu']/median(head(sort(test$log_cfu, decreasing = T),3))
+  } else {
+    colonization_ratio <- median(head(test[order(test$day), "log_cfu"],2))/
+                            median(head(sort(test$log_cfu, decreasing = T),3))
+  }
+  col_rate <- rbind(col_rate, c(mouse_id = i,  ratio = colonization_ratio))
+}
+
+col_rate <- data.frame(mouse_id = col_rate[ , 1], ratio = as.numeric(col_rate[ , 2]))
+
+col_rate <- meta_data %>% 
+  select(mouse_id, cage_id) %>% 
+  unique() %>% 
+  right_join(col_rate)
+
+shared_col_rate.df <- col_rate %>% 
+    right_join(community_d0.df) %>% 
+    select(-one_of(cage_source_features))
+shared_colonized.df <- select(shared_col_rate.df, -mouse_id, -cage_id)
+
+set.seed(seed)
+rf_col_rate <- randomForest(ratio ~ ., data = shared_colonized.df,
+                           importance = TRUE, ntree = n_trees, mtry = 14)
+
+importance_sorted <- sort(importance(rf_col_rate)[,1], decreasing = T)
+top_important_OTU <- rownames(data.frame(importance_sorted[importance_sorted/max(importance_sorted) > 0.6]))
+
+top_col_rate_df <- data.frame(shared_colonized.df %>% 
+  select(ratio, one_of(top_important_OTU)))
+
+set.seed(seed)
+rf_col_rate_features <- randomForest(ratio ~ ., data=top_col_rate_df,
+                              importance=TRUE, ntree=n_trees, mtry = 2,
+                              keep.forest=TRUE)
+
+importance_final <- sort(importance(rf_col_rate_features)[,1], decreasing = T)
+
+rf_col_rate_prediction <- data.frame(observed = shared_colonized.df$ratio, 
+                       predicted = rf_col_rate_features$predicted)
+rf_cr_prediction <- cbind(cage_id = shared_col_rate.df[,'cage_id'], rf_col_rate_prediction)
+
+ggplot(data = rf_cr_prediction, aes(x=observed, y=predicted, color = cage_id)) + geom_point() + 
+  labs(x= 'Observed', y = 'Predicted',
+       title = "Random Forest Prediction of Relative Colonization level of C Difficile from Day 0 Community") + 
+  theme_bw() + xlim(0,1) + ylim(0,1) + 
+  annotate('text', x=0.15*(max(rf_cr_prediction$observed)), 
+           y=0.95*(max(rf_cr_prediction$predicted)), 
+           label=paste0('R^2 = ', round(tail(rf_col_rate_features$rsq, 1), 2)))
+
+importance.df <- data.frame(otu = names(importance_final),
+                            importance = importance_final,
+                            tax = get_tax(1, names(importance_final), tax_file))
+tax_label <- paste0(importance.df$tax, ' (', 
+                    gsub('tu0*', 'TU ', importance.df$otu), 
+                    ') : ', 
+                    round(importance.df$importance, 3))
+
+layout(matrix(c(1:9), ncol=3, byrow=TRUE))
+
+cage_color <- rainbow(length(unique(rf_cr_prediction$cage_id)))
+for (i in seq_along(importance_final)) {
+  partial_otu <- 
+    partialPlot(rf_col_rate_features, top_col_rate_df, names(importance_final)[i],
+                xlab='', main=paste(tax_label[i]), 
+                ylim = c(min(top_col_rate_df$ratio),max(top_col_rate_df$ratio)))
+  for(m in seq_along(unique(rf_cr_prediction$cage_id))){
+    cage <- unique(rf_cr_prediction$cage_id)[m]
+    points(y = top_col_rate_df[rf_cr_prediction$cage_id %in% cage, 'ratio'],
+         x = top_col_rate_df[rf_cr_prediction$cage_id %in% cage , names(importance_final)[i]], 
+         pch = 21, bg = cage_color[m])
+  }
+}
+
+```
+
 ***

### 2016-10-12 code/HumanCdGF_predict_cfu.Rmd
colonization notes
update regression labels

@@
   annotate('text', x=0.15*(max(rf_prediction$observed)), 
-           y=0.95*(max(rf_prediction$predicted)), 
-           label=paste0('R^2 = ', round(tail(rf_n_features$rsq, 1), 2)))
+           y=0.95*(max(rf_prediction$predicted)), parse = T,
+           label=paste0('R^2 == ', round(tail(rf_n_features$rsq, 1), 2))) + 
+  theme(legend.position = 'none')
 
@@
 }
+
 ```
@@
   labs(x= 'Observed', y = 'Predicted',
-       title = "Random Forest Prediction of Relative Colonization level of C Difficile from Day 0 Community") + 
+       title = "Random Forest Prediction of Relative Day 1 Colonization Level of C. Difficile from Day 0 Community") + 
   theme_bw() + xlim(0,1) + ylim(0,1) + 
   annotate('text', x=0.15*(max(rf_cr_prediction$observed)), 
-           y=0.95*(max(rf_cr_prediction$predicted)), 
-           label=paste0('R^2 = ', round(tail(rf_col_rate_features$rsq, 1), 2)))
+           y=0.95*(max(rf_cr_prediction$predicted)), parse = T,
+           label=paste("R^2 == ", round(tail(rf_col_rate_features$rsq, 1), 2))) + 
+  theme(legend.position = 'none')
 
@@
 ```
+
+Does not predict colonization rate as well  
+  - could be due to removed (cage/inocula-sensitive) otus  
+  - improper measurement of colonization rate  
+  - colonization rate not microbiome dependent, may be dependent on host factors of acute infection
 

### 2016-10-14 code/HumanCdGF_predict_cfu.Rmd
dual y axis for partial dependence plots

@@
 
+```
+
+```{r functions, eval = F}
+
 heat.plot <- function(input_maxtrix, high_color = '#660000', low_color = '#FFFFFF', legend_title = NA) {
@@
 }
-```
 
-```{r functions}
 rsq <- matrix(NA, nrow = 11, ncol = 10)
@@
 
-```{r top otus, fig.width=6, fig.height=6}
+```{r top otus, fig.width=6, fig.height=6, eval = F}
 heat.plot(rsq_n, '#003300', legend_title = expression(paste('R'^'2')))
@@
 
-```{r fig.height=5, fig.width=8}
+```{r fig.height=5, fig.width=8, eval = F}
 meta_data %>% 
@@
 
-```{r identify reason for lowered R2, fig.height=5, fig.width=8}
+```{r identify reason for lowered R2, fig.height=5, fig.width=8, eval = F}
 log_cutoff <- 2.5
@@
 
-```{r fig.width=6,fig.height=6}
+```{r fig.width=6,fig.height=6, eval = F}
 heat.plot(rsq_r, '#003300', legend_title = expression(paste('R'^'2')))
@@
 
-```{r plot features,fig.width=8, fig.height=20}
+```{r plot features,fig.width=8, fig.height=20, eval = F}
 meta_data %>% 
@@
 
-```{r predict median CFU heatmap, fig.width=6,fig.height=6}
+```{r predict median CFU heatmap, fig.width=6,fig.height=6, eval = F}
 
@@
 for (i in seq_along(importance_final)) {
-  partial_otu <- 
+    par(mar = c(5,5,2,5))
     partialPlot(rf_n_features, top_colonized_df, names(importance_final)[i],
-                xlab='', main=paste(tax_label[i]), 
-                ylim = c(min(top_colonized_df$log_cfu_med),max(top_colonized_df$log_cfu_med)))
-  for(m in seq_along(unique(rf_prediction$cage_id))){
-    cage <- unique(rf_prediction$cage_id)[m]
-    points(y = top_colonized_df[rf_prediction$cage_id %in% cage, 'log_cfu_med'],
-         x = top_colonized_df[rf_prediction$cage_id %in% cage , names(importance_final)[i]], 
-         pch = 21, bg = cage_color[m])
-  }
+                xlab='', main=paste(tax_label[i]))
+    par(new = T)
+    plot(y = top_colonized_df[ , 'log_cfu_med'],
+         x = top_colonized_df[ , names(importance_final)[i]], 
+         pch = 21, bg = 'gray', xlab = NA, ylab = NA, axes = F)
+    axis(side = 4)
+    mtext(side = 4, line = 3, 'Log10 CFU')
+# To color points by cage - Change pch to 21 and un-comment following lines
+#  for(m in seq_along(unique(rf_prediction$cage_id))){
+#    cage <- unique(rf_prediction$cage_id)[m]
+#    points(y = top_colonized_df[rf_prediction$cage_id %in% cage, 'log_cfu_med'],
+#         x = top_colonized_df[rf_prediction$cage_id %in% cage , names(importance_final)[i]], 
+#         pch = 21, bg = cage_color[m], xlab = NA, ylab = NA, axes = F)
+#  }
 }
@@
 for (i in seq_along(importance_final)) {
-  partial_otu <- 
     partialPlot(rf_col_rate_features, top_col_rate_df, names(importance_final)[i],
-                xlab='', main=paste(tax_label[i]), 
-                ylim = c(min(top_col_rate_df$ratio),max(top_col_rate_df$ratio)))
-  for(m in seq_along(unique(rf_cr_prediction$cage_id))){
-    cage <- unique(rf_cr_prediction$cage_id)[m]
-    points(y = top_col_rate_df[rf_cr_prediction$cage_id %in% cage, 'ratio'],
-         x = top_col_rate_df[rf_cr_prediction$cage_id %in% cage , names(importance_final)[i]], 
-         pch = 21, bg = cage_color[m])
+                xlab='', main=paste(tax_label[i]))
+    par(new = T)
+    plot(y = top_col_rate_df[ , 'ratio'],
+         x = top_col_rate_df[ , names(importance_final)[i]], 
+         pch = 21, bg = 'gray', xlab = NA, ylab = NA, axes = F)
+    axis(side = 4)
+    mtext(side = 4, line = 3, 'Relative Colonization')
+# To color points by cage - Change pch to 21 and un-comment following lines
+#  for(m in seq_along(unique(rf_cr_prediction$cage_id))){
+#    cage <- unique(rf_cr_prediction$cage_id)[m]
+#    points(y = top_col_rate_df[rf_cr_prediction$cage_id %in% cage, 'ratio'],
+#         x = top_col_rate_df[rf_cr_prediction$cage_id %in% cage , names(importance_final)[i]], 
+#         pch = 21, bg = cage_color[m])
+#  }
 }
-}
-
 ```

### 2016-10-14 code/HumanCdGF_predict_cfu.Rmd
remove heatmaps from colonization prediction

@@
 ```
+
+### Reframing the question for predicting CFU
+#### 1. Can Day 0 community predict the median steady-state-colonized CFU?
+
+```{r predict median CFU, fig.width=10,fig.height=10}
+
+Cage_dep_df <- meta_data %>% 
+  filter(day == 0) %>% 
+  select(human_source, sample_id, cage_id) %>% 
+  left_join(rel_abund_shared_01)
+Predict_cage_df <- select(Cage_dep_df, -human_source, -sample_id)
+Predict_source_df <- select(Cage_dep_df, -cage_id, -sample_id)
+Predict_cage_df$cage_id <- factor(Predict_cage_df$cage_id)
+Predict_source_df$human_source <- factor(Predict_source_df$human_source)
+
+# identify OTUs that predict cage/inocula
+set.seed(seed)
+rf_cage <- randomForest(cage_id ~ ., data=Predict_cage_df, importance=TRUE, ntree=n_trees)
+set.seed(seed)
+rf_source <- randomForest(human_source ~ ., data=Predict_source_df, importance=TRUE, ntree=n_trees)
+features_source <- head(sort(rf_source$importance[ , 'MeanDecreaseAccuracy'], decreasing = T), 65)
+features_cage <- head(sort(rf_cage$importance[ , 'MeanDecreaseAccuracy'], decreasing = T), 65)
+cage_source_features <- unique(c(names(features_cage), names(features_source)))
+
+colonized_cfu <- meta_data %>% 
+  filter(day %in% c(4:10), !cage_id == 'NP1', 
+         Early_Euth == FALSE, cdiff_strain == 431) %>% 
+  select(mouse_id, cage_id, log_cfu) %>% 
+  group_by(mouse_id, cage_id) %>% 
+  summarise(log_cfu_med = median(log_cfu)) %>% 
+  ungroup()
+sample_list_d0 <- shared_list_days %>% 
+    filter(day == 0) %>% 
+    select(-day)
+community_d0.df <- rel_abund_shared_01 %>% 
+    right_join(sample_list_d0) %>% 
+    select(-sample_id)
+cfu_shared_colonized.df <- colonized_cfu %>% 
+    right_join(community_d0.df) %>% 
+    select(-mouse_id, -cage_id, -one_of(cage_source_features)) %>% 
+    filter(!is.na(log_cfu_med))
+
+set.seed(seed)
+rf_all <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df,
+                           importance = TRUE, ntree = n_trees, mtry = 14)
+
+importance_sorted <- sort(importance(rf_all)[,1], decreasing = T)
+top_important_OTU <- rownames(data.frame(importance_sorted[importance_sorted/max(importance_sorted) > 0.6]))
+
+top_colonized_df <- data.frame(cfu_shared_colonized.df %>% 
+  select(log_cfu_med, one_of(top_important_OTU)))
+
+set.seed(seed)
+rf_n_features <- randomForest(log_cfu_med ~ ., data=top_colonized_df,
+                              importance=TRUE, ntree=n_trees, mtry = 2,
+                              keep.forest=TRUE)
+
+importance_final <- sort(importance(rf_n_features)[,1], decreasing = T)
+
+rf_prediction <- data.frame(observed = cfu_shared_colonized.df$log_cfu_med, 
+                       predicted = rf_n_features$predicted)
+rf_prediction <- merge(rf_prediction, colonized_cfu, by.x = 'observed', by.y = 'log_cfu_med')
+
+ggplot(data = rf_prediction, aes(x=observed, y=predicted, color = cage_id)) + geom_point() + 
+  labs(x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)',
+       title = "Random Forest Prediction of Colonized C Difficile CFU from Day 0 Community") + 
+  theme_bw() + xlim(0,9) + ylim(0,9) + 
+  annotate('text', x=0.15*(max(rf_prediction$observed)), 
+           y=0.95*(max(rf_prediction$predicted)), parse = T,
+           label=paste0('R^2 == ', round(tail(rf_n_features$rsq, 1), 2))) + 
+  theme(legend.position = 'none')
+
+importance.df <- data.frame(otu = names(importance_final),
+                            importance = importance_final,
+                            tax = get_tax(1, names(importance_final), tax_file))
+tax_label <- paste0(importance.df$tax, ' (', 
+                    gsub('tu0*', 'TU ', importance.df$otu), 
+                    ') : ', 
+                    round(importance.df$importance, 3))
+
+layout(matrix(c(1:9), ncol=3, byrow=TRUE))
+
+cage_color <- rainbow(length(unique(rf_prediction$cage_id)))
+for (i in seq_along(importance_final)) {
+    partialPlot(rf_n_features, top_colonized_df, names(importance_final)[i],
+                xlab='', main=paste(tax_label[i]))
+    par(new = T)
+    plot(y = top_colonized_df[ , 'log_cfu_med'],
+         x = top_colonized_df[ , names(importance_final)[i]], 
+         pch = 21, bg = 'gray', xlab = NA, ylab = NA, axes = F)
+    axis(side = 4)
+    mtext(side = 4, line = 3, 'Log10 CFU')
+# To color points by cage - Change pch to 21 and un-comment following lines
+#  for(m in seq_along(unique(rf_prediction$cage_id))){
+#    cage <- unique(rf_prediction$cage_id)[m]
+#    points(y = top_colonized_df[rf_prediction$cage_id %in% cage, 'log_cfu_med'],
+#         x = top_colonized_df[rf_prediction$cage_id %in% cage , names(importance_final)[i]], 
+#         pch = 21, bg = cage_color[m], xlab = NA, ylab = NA, axes = F)
+#  }
+}
+
+```
+
+#### 2. Can Day 0 community predict the rate of colonization?
+
+```{r predict colonization rate, fig.width=10,fig.height=10}
+
+meta_data %>% 
+  filter(!is.na(day), !cage_id == 'NP1', 
+         Early_Euth == FALSE, cdiff_strain == 431) %>% 
+  ggplot(aes(x = factor(day), y = log_cfu, color = cage_id)) + 
+    geom_line(aes(group = mouse_id), size=.25, alpha=0.4) + geom_point() +
+    stat_summary(fun.y=mean, geom="line", aes(group = cage_id)) + 
+    theme_bw() + labs(x = 'Day', y = 'C Diff CFU (log10)') +
+    theme(legend.justification=c(1,0), legend.position=c(1,0), 
+          legend.title=element_blank())
+
+mouse_list <- meta_data %>% 
+  filter(!is.na(day), !cage_id == 'NP1', 
+         Early_Euth == FALSE, cdiff_strain == 431) %>% 
+  select(mouse_id) %>% 
+  unique()
+mouse_list <- as.character(mouse_list$mouse_id)
+
+col_rate <- c()
+for(i in mouse_list){
+  test <- meta_data %>% 
+    filter(!is.na(day), mouse_id == i) %>% 
+    select(day, log_cfu) %>% 
+    filter(!is.na(log_cfu))
+  if(1 %in% test$day){
+    colonization_ratio <- test[test$day==1, 'log_cfu']/median(head(sort(test$log_cfu, decreasing = T),3))
+  } else {
+    colonization_ratio <- median(head(test[order(test$day), "log_cfu"],2))/
+                            median(head(sort(test$log_cfu, decreasing = T),3))
+  }
+  col_rate <- rbind(col_rate, c(mouse_id = i,  ratio = colonization_ratio))
+}
+
+col_rate <- data.frame(mouse_id = col_rate[ , 1], ratio = as.numeric(col_rate[ , 2]))
+
+col_rate <- meta_data %>% 
+  select(mouse_id, cage_id) %>% 
+  unique() %>% 
+  right_join(col_rate)
+
+shared_col_rate.df <- col_rate %>% 
+    right_join(community_d0.df) %>% 
+    select(-one_of(cage_source_features))
+shared_colonized.df <- select(shared_col_rate.df, -mouse_id, -cage_id)
+
+set.seed(seed)
+rf_col_rate <- randomForest(ratio ~ ., data = shared_colonized.df,
+                           importance = TRUE, ntree = n_trees, mtry = 14)
+
+importance_sorted <- sort(importance(rf_col_rate)[,1], decreasing = T)
+top_important_OTU <- rownames(data.frame(importance_sorted[importance_sorted/max(importance_sorted) > 0.6]))
+
+top_col_rate_df <- data.frame(shared_colonized.df %>% 
+  select(ratio, one_of(top_important_OTU)))
+
+set.seed(seed)
+rf_col_rate_features <- randomForest(ratio ~ ., data=top_col_rate_df,
+                              importance=TRUE, ntree=n_trees, mtry = 2,
+                              keep.forest=TRUE)
+
+importance_final <- sort(importance(rf_col_rate_features)[,1], decreasing = T)
+
+rf_col_rate_prediction <- data.frame(observed = shared_colonized.df$ratio, 
+                       predicted = rf_col_rate_features$predicted)
+rf_cr_prediction <- cbind(cage_id = shared_col_rate.df[,'cage_id'], rf_col_rate_prediction)
+
+ggplot(data = rf_cr_prediction, aes(x=observed, y=predicted, color = cage_id)) + geom_point() + 
+  labs(x= 'Observed', y = 'Predicted',
+       title = "Random Forest Prediction of Relative Day 1 Colonization Level of C. Difficile from Day 0 Community") + 
+  theme_bw() + xlim(0,1) + ylim(0,1) + 
+  annotate('text', x=0.15*(max(rf_cr_prediction$observed)), 
+           y=0.95*(max(rf_cr_prediction$predicted)), parse = T,
+           label=paste("R^2 == ", round(tail(rf_col_rate_features$rsq, 1), 2))) + 
+  theme(legend.position = 'none')
+
+importance.df <- data.frame(otu = names(importance_final),
+                            importance = importance_final,
+                            tax = get_tax(1, names(importance_final), tax_file))
+tax_label <- paste0(importance.df$tax, ' (', 
+                    gsub('tu0*', 'TU ', importance.df$otu), 
+                    ') : ', 
+                    round(importance.df$importance, 3))
+
+layout(matrix(c(1:9), ncol=3, byrow=TRUE))
+
+cage_color <- rainbow(length(unique(rf_cr_prediction$cage_id)))
+for (i in seq_along(importance_final)) {
+    partialPlot(rf_col_rate_features, top_col_rate_df, names(importance_final)[i],
+                xlab='', main=paste(tax_label[i]))
+    par(new = T)
+    plot(y = top_col_rate_df[ , 'ratio'],
+         x = top_col_rate_df[ , names(importance_final)[i]], 
+         pch = 21, bg = 'gray', xlab = NA, ylab = NA, axes = F)
+    axis(side = 4)
+    mtext(side = 4, line = 3, 'Relative Colonization')
+# To color points by cage - Change pch to 21 and un-comment following lines
+#  for(m in seq_along(unique(rf_cr_prediction$cage_id))){
+#    cage <- unique(rf_cr_prediction$cage_id)[m]
+#    points(y = top_col_rate_df[rf_cr_prediction$cage_id %in% cage, 'ratio'],
+#         x = top_col_rate_df[rf_cr_prediction$cage_id %in% cage , names(importance_final)[i]], 
+#         pch = 21, bg = cage_color[m])
+#  }
+}
+```
+
+Does not predict colonization rate as well  
+  - could be due to removed (cage/inocula-sensitive) otus  
+  - improper measurement of colonization rate  
+  - colonization rate not microbiome dependent, may be dependent on host factors of acute infection
+
+***
+
+***
 
@@
 ### OTUs selected by >`r cutoff`% of models generated by all OTU models 
-###### Using the following bacteria: `r taxa_names`
+###### Using the following bacteria: - r taxa_names
 
@@
 
-### Reframing the question for predicting CFU
-#### 1. Can Day 0 community predict the median steady-state-colonized CFU?
-
-```{r predict median CFU, fig.width=10,fig.height=10}
-
-Cage_dep_df <- meta_data %>% 
-  filter(day == 0) %>% 
-  select(human_source, sample_id, cage_id) %>% 
-  left_join(rel_abund_shared_01)
-Predict_cage_df <- select(Cage_dep_df, -human_source, -sample_id)
-Predict_source_df <- select(Cage_dep_df, -cage_id, -sample_id)
-Predict_cage_df$cage_id <- factor(Predict_cage_df$cage_id)
-Predict_source_df$human_source <- factor(Predict_source_df$human_source)
-
-# identify OTUs that predict cage/inocula
-set.seed(seed)
-rf_cage <- randomForest(cage_id ~ ., data=Predict_cage_df, importance=TRUE, ntree=n_trees)
-set.seed(seed)
-rf_source <- randomForest(human_source ~ ., data=Predict_source_df, importance=TRUE, ntree=n_trees)
-features_source <- head(sort(rf_source$importance[ , 'MeanDecreaseAccuracy'], decreasing = T), 65)
-features_cage <- head(sort(rf_cage$importance[ , 'MeanDecreaseAccuracy'], decreasing = T), 65)
-cage_source_features <- unique(c(names(features_cage), names(features_source)))
-
-colonized_cfu <- meta_data %>% 
-  filter(day %in% c(4:10), !cage_id == 'NP1', 
-         Early_Euth == FALSE, cdiff_strain == 431) %>% 
-  select(mouse_id, cage_id, log_cfu) %>% 
-  group_by(mouse_id, cage_id) %>% 
-  summarise(log_cfu_med = median(log_cfu)) %>% 
-  ungroup()
-sample_list_d0 <- shared_list_days %>% 
-    filter(day == 0) %>% 
-    select(-day)
-community_d0.df <- rel_abund_shared_01 %>% 
-    right_join(sample_list_d0) %>% 
-    select(-sample_id)
-cfu_shared_colonized.df <- colonized_cfu %>% 
-    right_join(community_d0.df) %>% 
-    select(-mouse_id, -cage_id, -one_of(cage_source_features)) %>% 
-    filter(!is.na(log_cfu_med))
-
-set.seed(seed)
-rf_all <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df,
-                           importance = TRUE, ntree = n_trees, mtry = 14)
-
-importance_sorted <- sort(importance(rf_all)[,1], decreasing = T)
-top_important_OTU <- rownames(data.frame(importance_sorted[importance_sorted/max(importance_sorted) > 0.6]))
-
-top_colonized_df <- data.frame(cfu_shared_colonized.df %>% 
-  select(log_cfu_med, one_of(top_important_OTU)))
-
-set.seed(seed)
-rf_n_features <- randomForest(log_cfu_med ~ ., data=top_colonized_df,
-                              importance=TRUE, ntree=n_trees, mtry = 2,
-                              keep.forest=TRUE)
-
-importance_final <- sort(importance(rf_n_features)[,1], decreasing = T)
-
-rf_prediction <- data.frame(observed = cfu_shared_colonized.df$log_cfu_med, 
-                       predicted = rf_n_features$predicted)
-rf_prediction <- merge(rf_prediction, colonized_cfu, by.x = 'observed', by.y = 'log_cfu_med')
-
-ggplot(data = rf_prediction, aes(x=observed, y=predicted, color = cage_id)) + geom_point() + 
-  labs(x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)',
-       title = "Random Forest Prediction of Colonized C Difficile CFU from Day 0 Community") + 
-  theme_bw() + xlim(0,9) + ylim(0,9) + 
-  annotate('text', x=0.15*(max(rf_prediction$observed)), 
-           y=0.95*(max(rf_prediction$predicted)), parse = T,
-           label=paste0('R^2 == ', round(tail(rf_n_features$rsq, 1), 2))) + 
-  theme(legend.position = 'none')
-
-importance.df <- data.frame(otu = names(importance_final),
-                            importance = importance_final,
-                            tax = get_tax(1, names(importance_final), tax_file))
-tax_label <- paste0(importance.df$tax, ' (', 
-                    gsub('tu0*', 'TU ', importance.df$otu), 
-                    ') : ', 
-                    round(importance.df$importance, 3))
-
-layout(matrix(c(1:9), ncol=3, byrow=TRUE))
-
-cage_color <- rainbow(length(unique(rf_prediction$cage_id)))
-for (i in seq_along(importance_final)) {
-    par(mar = c(5,5,2,5))
-    partialPlot(rf_n_features, top_colonized_df, names(importance_final)[i],
-                xlab='', main=paste(tax_label[i]))
-    par(new = T)
-    plot(y = top_colonized_df[ , 'log_cfu_med'],
-         x = top_colonized_df[ , names(importance_final)[i]], 
-         pch = 21, bg = 'gray', xlab = NA, ylab = NA, axes = F)
-    axis(side = 4)
-    mtext(side = 4, line = 3, 'Log10 CFU')
-# To color points by cage - Change pch to 21 and un-comment following lines
-#  for(m in seq_along(unique(rf_prediction$cage_id))){
-#    cage <- unique(rf_prediction$cage_id)[m]
-#    points(y = top_colonized_df[rf_prediction$cage_id %in% cage, 'log_cfu_med'],
-#         x = top_colonized_df[rf_prediction$cage_id %in% cage , names(importance_final)[i]], 
-#         pch = 21, bg = cage_color[m], xlab = NA, ylab = NA, axes = F)
-#  }
-}
-
-```
-
-#### 2. Can Day 0 community predict the rate of colonization?
-
-```{r predict colonization rate, fig.width=10,fig.height=10}
-
-meta_data %>% 
-  filter(!is.na(day), !cage_id == 'NP1', 
-         Early_Euth == FALSE, cdiff_strain == 431) %>% 
-  ggplot(aes(x = factor(day), y = log_cfu, color = cage_id)) + 
-    geom_line(aes(group = mouse_id), size=.25, alpha=0.4) + geom_point() +
-    stat_summary(fun.y=mean, geom="line", aes(group = cage_id)) + 
-    theme_bw() + labs(x = 'Day', y = 'C Diff CFU (log10)') +
-    theme(legend.justification=c(1,0), legend.position=c(1,0), 
-          legend.title=element_blank())
-
-mouse_list <- meta_data %>% 
-  filter(!is.na(day), !cage_id == 'NP1', 
-         Early_Euth == FALSE, cdiff_strain == 431) %>% 
-  select(mouse_id) %>% 
-  unique()
-mouse_list <- as.character(mouse_list$mouse_id)
-
-col_rate <- c()
-for(i in mouse_list){
-  test <- meta_data %>% 
-    filter(!is.na(day), mouse_id == i) %>% 
-    select(day, log_cfu) %>% 
-    filter(!is.na(log_cfu))
-  if(1 %in% test$day){
-    colonization_ratio <- test[test$day==1, 'log_cfu']/median(head(sort(test$log_cfu, decreasing = T),3))
-  } else {
-    colonization_ratio <- median(head(test[order(test$day), "log_cfu"],2))/
-                            median(head(sort(test$log_cfu, decreasing = T),3))
-  }
-  col_rate <- rbind(col_rate, c(mouse_id = i,  ratio = colonization_ratio))
-}
-
-col_rate <- data.frame(mouse_id = col_rate[ , 1], ratio = as.numeric(col_rate[ , 2]))
-
-col_rate <- meta_data %>% 
-  select(mouse_id, cage_id) %>% 
-  unique() %>% 
-  right_join(col_rate)
-
-shared_col_rate.df <- col_rate %>% 
-    right_join(community_d0.df) %>% 
-    select(-one_of(cage_source_features))
-shared_colonized.df <- select(shared_col_rate.df, -mouse_id, -cage_id)
-
-set.seed(seed)
-rf_col_rate <- randomForest(ratio ~ ., data = shared_colonized.df,
-                           importance = TRUE, ntree = n_trees, mtry = 14)
-
-importance_sorted <- sort(importance(rf_col_rate)[,1], decreasing = T)
-top_important_OTU <- rownames(data.frame(importance_sorted[importance_sorted/max(importance_sorted) > 0.6]))
-
-top_col_rate_df <- data.frame(shared_colonized.df %>% 
-  select(ratio, one_of(top_important_OTU)))
-
-set.seed(seed)
-rf_col_rate_features <- randomForest(ratio ~ ., data=top_col_rate_df,
-                              importance=TRUE, ntree=n_trees, mtry = 2,
-                              keep.forest=TRUE)
-
-importance_final <- sort(importance(rf_col_rate_features)[,1], decreasing = T)
-
-rf_col_rate_prediction <- data.frame(observed = shared_colonized.df$ratio, 
-                       predicted = rf_col_rate_features$predicted)
-rf_cr_prediction <- cbind(cage_id = shared_col_rate.df[,'cage_id'], rf_col_rate_prediction)
-
-ggplot(data = rf_cr_prediction, aes(x=observed, y=predicted, color = cage_id)) + geom_point() + 
-  labs(x= 'Observed', y = 'Predicted',
-       title = "Random Forest Prediction of Relative Day 1 Colonization Level of C. Difficile from Day 0 Community") + 
-  theme_bw() + xlim(0,1) + ylim(0,1) + 
-  annotate('text', x=0.15*(max(rf_cr_prediction$observed)), 
-           y=0.95*(max(rf_cr_prediction$predicted)), parse = T,
-           label=paste("R^2 == ", round(tail(rf_col_rate_features$rsq, 1), 2))) + 
-  theme(legend.position = 'none')
-
-importance.df <- data.frame(otu = names(importance_final),
-                            importance = importance_final,
-                            tax = get_tax(1, names(importance_final), tax_file))
-tax_label <- paste0(importance.df$tax, ' (', 
-                    gsub('tu0*', 'TU ', importance.df$otu), 
-                    ') : ', 
-                    round(importance.df$importance, 3))
-
-layout(matrix(c(1:9), ncol=3, byrow=TRUE))
-
-cage_color <- rainbow(length(unique(rf_cr_prediction$cage_id)))
-for (i in seq_along(importance_final)) {
-    partialPlot(rf_col_rate_features, top_col_rate_df, names(importance_final)[i],
-                xlab='', main=paste(tax_label[i]))
-    par(new = T)
-    plot(y = top_col_rate_df[ , 'ratio'],
-         x = top_col_rate_df[ , names(importance_final)[i]], 
-         pch = 21, bg = 'gray', xlab = NA, ylab = NA, axes = F)
-    axis(side = 4)
-    mtext(side = 4, line = 3, 'Relative Colonization')
-# To color points by cage - Change pch to 21 and un-comment following lines
-#  for(m in seq_along(unique(rf_cr_prediction$cage_id))){
-#    cage <- unique(rf_cr_prediction$cage_id)[m]
-#    points(y = top_col_rate_df[rf_cr_prediction$cage_id %in% cage, 'ratio'],
-#         x = top_col_rate_df[rf_cr_prediction$cage_id %in% cage , names(importance_final)[i]], 
-#         pch = 21, bg = cage_color[m])
-#  }
-}
-```
-
-Does not predict colonization rate as well  
-  - could be due to removed (cage/inocula-sensitive) otus  
-  - improper measurement of colonization rate  
-  - colonization rate not microbiome dependent, may be dependent on host factors of acute infection
-
-***
-
-***
-
 ###### Re-runnning the program and only focusing on the OTUs predicting day 9/10 to see if there is consistence in the predictive features gives the following selection  (including all samples):  

### 2016-10-14 code/CdGF_AUCRF.Rmd
add y axis label to partial dependence

@@
               pch = 21, bg = 'gray')
+  axis(side = 4, labels = c('Severe', 'Mild'),
+       at = c(0.25*(diff(range(partial_otu$y))) + min(partial_otu$y),
+              0.75*(diff(range(partial_otu$y))) + min(partial_otu$y)))
 }
@@
 
+####### Model run with both relative and absolute abundances and no significant difference in their output
+
 ***  

### 2016-10-14 code/HumanCdGF_microbiota_strains.Rmd
fix median and axis label

@@
   ggplot(aes(x = factor(day), y = log_cfu, color = factor(cdiff_strain))) + geom_point() +
-            stat_summary(fun.y=mean, geom="line", aes(group = cage_id)) + 
+            stat_summary(fun.y=median, geom="line", aes(group = cage_id)) + 
             facet_grid(. ~ human_source) + theme_bw() + 
-            labs(x = 'Day', y = 'C Diff CFU (log10)') +
+            labs(x = 'Day', y = expression(paste(italic('C. Difficile'), ' Log'[10], ' CFU', sep = ''))) +
             theme(legend.justification=c(1,0.5), legend.position=c(1,0.5), 
                   legend.title=element_blank())
-
 ```

### 2016-10-19 code/CdGF_AUCRF.Rmd
remove code that remove cage/inocula otus

@@
                     rel_abund_stool, by.x = 'mouse_id', by.y = 'row.names')
-Predict_cage_df <- select(Predict_df, -human_source, -mouse_id, -Euth_Early)
-Predict_source_df <- select(Predict_df, -cage_id, -mouse_id, -Euth_Early)
-Predict_cage_df$cage_id <- factor(Predict_cage_df$cage_id)
-Predict_source_df$human_source <- factor(Predict_source_df$human_source)
-
-# identify OTUs that predict cage/inocula
-set.seed(seed)
-rf_cage <- randomForest(cage_id ~ ., data=Predict_cage_df, importance=TRUE, ntree=n_trees)
-set.seed(seed)
-rf_source <- randomForest(human_source ~ ., data=Predict_source_df, importance=TRUE, ntree=n_trees)
-features_source <- head(sort(rf_source$importance[ , 'MeanDecreaseAccuracy'], decreasing = T), n_features)
-features_cage <- head(sort(rf_cage$importance[ , 'MeanDecreaseAccuracy'], decreasing = T), n_features)
-features_all <- unique(c(names(features_cage), names(features_source)))
 
 # remove cage/inocula OTUs from prediction dataframes
-Predict_early_euth_df <- select(Predict_df, -cage_id, -mouse_id, -human_source, -one_of(features_all))
+Predict_early_euth_df <- select(Predict_df, -cage_id, -mouse_id, -human_source)
 

### 2016-10-19 code/HumanCdGF_predict_cfu.Rmd
remove code that remove cage/inocula otus

@@
 
-Cage_dep_df <- meta_data %>% 
-  filter(day == 0) %>% 
-  select(human_source, sample_id, cage_id) %>% 
-  left_join(rel_abund_shared_01)
-Predict_cage_df <- select(Cage_dep_df, -human_source, -sample_id)
-Predict_source_df <- select(Cage_dep_df, -cage_id, -sample_id)
-Predict_cage_df$cage_id <- factor(Predict_cage_df$cage_id)
-Predict_source_df$human_source <- factor(Predict_source_df$human_source)
-
-# identify OTUs that predict cage/inocula
-set.seed(seed)
-rf_cage <- randomForest(cage_id ~ ., data=Predict_cage_df, importance=TRUE, ntree=n_trees)
-set.seed(seed)
-rf_source <- randomForest(human_source ~ ., data=Predict_source_df, importance=TRUE, ntree=n_trees)
-features_source <- head(sort(rf_source$importance[ , 'MeanDecreaseAccuracy'], decreasing = T), 65)
-features_cage <- head(sort(rf_cage$importance[ , 'MeanDecreaseAccuracy'], decreasing = T), 65)
-cage_source_features <- unique(c(names(features_cage), names(features_source)))
-
 colonized_cfu <- meta_data %>% 
@@
     right_join(community_d0.df) %>% 
-    select(-mouse_id, -cage_id, -one_of(cage_source_features)) %>% 
+    select(-mouse_id, -cage_id) %>% 
     filter(!is.na(log_cfu_med))
@@
 shared_col_rate.df <- col_rate %>% 
-    right_join(community_d0.df) %>% 
-    select(-one_of(cage_source_features))
+    right_join(community_d0.df)
 shared_colonized.df <- select(shared_col_rate.df, -mouse_id, -cage_id)

### 2016-10-20 code/HumanCdGF_predict_cfu.Rmd
add leave 1 cage out for cfu

@@
 
+##### - leave one cage out cv
+LOcO_df <- data.frame(colonized_cfu %>% 
+    right_join(community_d0.df) %>% 
+    filter(!is.na(log_cfu_med)) %>% 
+    select(log_cfu_med, cage_id, one_of(top_important_OTU)))
+cage_list <- as.character(unique(LOcO_df$cage_id))
+
+LOcO_probs <- c()
+LOcO_rf <- c()
+rsq <- c()
+import_otus <- data.frame(otu = top_important_OTU)
+for(i in 1:length(cage_list)){
+  by_cage <- LOcO_df$cage_id == cage_list[i]
+  LOcO_predict <- select(LOcO_df, -cage_id)
+  train <- LOcO_predict[!by_cage,]
+  test <- LOcO_predict[by_cage,]
+  set.seed(seed)
+  temp_model <- randomForest(log_cfu_med~., data=train, pdel=0.99, ntree=500)
+  import_df <- importance(temp_model)
+  colnames(import_df) <- cage_list[i]
+  import_otus <- merge(import_otus, import_df, by.x = 'otu', by.y = 'row.names')
+  rsq[i] <- round(tail(temp_model$rsq, 1), 2)
+  LOcO_probs <- rbind(LOcO_probs, data.frame(cage=rep(cage_list[i], nrow(test)),
+                                             actual=test$log_cfu_med, 
+                                             pred=predict(temp_model, test)))
+}
+LOcO_roc <- roc(LOcO_df$Euth_Early~LOcO_probs)
+##### - leave one cage out cv
+
 ggplot(data = rf_prediction, aes(x=observed, y=predicted, color = cage_id)) + geom_point() + 
@@
        title = "Random Forest Prediction of Colonized C Difficile CFU from Day 0 Community") + 
-  theme_bw() + xlim(0,9) + ylim(0,9) + 
-  annotate('text', x=0.15*(max(rf_prediction$observed)), 
+  theme_bw() + annotate('text', x=0.15*(max(rf_prediction$observed)), 
                         y=0.95*(max(rf_prediction$predicted)), parse = T,

### 2016-10-22 code/HumanCdGF_predict_cfu.Rmd
add rf trained with one mouse per cage

--- 2016-10-19 code/HumanCdGF_predict_cfu.Rmd
@@
 # read in files
-meta_data   <- read.table(meta_data, sep = '\t', header = T)
+meta_data   <- read.table(meta_data, sep = '\t', header = T, stringsAsFactors = F)
 shared_file <- read.table(shared_file, sep = '\t', header = T, row.names = 'sample_id')
@@
 n_trees <- 1001
-rf_plot_list <- list()
 n_features <- 10
@@
 
+##### - leave one cage out cv
+LOcO_df <- data.frame(colonized_cfu %>% 
+    right_join(community_d0.df) %>% 
+    filter(!is.na(log_cfu_med)) %>% 
+    select(log_cfu_med, mouse_id, cage_id, one_of(top_important_OTU)))
+cage_list <- unique(LOcO_df$cage_id)
+
+LOcO_probs <- c()
+rsq <- c()
+import_otus <- data.frame(otu = top_important_OTU)
+for(i in 1:length(cage_list)){
+  by_cage <- LOcO_df$cage_id == cage_list[i]
+  LOcO_predict <- select(LOcO_df, -cage_id, -mouse_id)
+  train <- LOcO_predict[!by_cage,]
+  test <- LOcO_predict[by_cage,]
+  set.seed(seed)
+  temp_model <- randomForest(log_cfu_med~., data=train, pdel=0.99, ntree=500, importance=T)
+  import_df <- importance(temp_model)
+  colnames(import_df)[1] <- cage_list[i]
+  import_otus <- merge(import_otus, import_df, by.x = 'otu', by.y = 'row.names')
+  import_otus <- select(import_otus, -IncNodePurity)
+  rsq[i] <- round(tail(temp_model$rsq, 1), 2)
+  LOcO_probs <- rbind(LOcO_probs, data.frame(observed=test$log_cfu_med, 
+                                             predicted=predict(temp_model, test),
+                                             mouse_id=LOcO_df$mouse_id[by_cage],
+                                             cage_id=rep(cage_list[i], nrow(test))))
+}
+med_otu_importance <- data.frame(otu = import_otus$otu,
+                                 importance = apply(import_otus[, -1], 1, median))
+med_otu_importance <- med_otu_importance[order(med_otu_importance$importance, decreasing = T),]
+
+LOcO_rsq <- 1 - (sum((LOcO_probs$observed - LOcO_probs$predicted)^2)/
+                   sum((LOcO_probs$observed - mean(LOcO_probs$observed))^2))
+
+#One per cage
+one_mouse_df <- data.frame(colonized_cfu %>% 
+    right_join(community_d0.df) %>% 
+    filter(!is.na(log_cfu_med))  %>% 
+    select(log_cfu_med, mouse_id, cage_id, one_of(top_important_OTU)))
+
+one_mouse_probs <- c()
+om_rsq <- c()
+om_rsq_rf <- c()
+om_import_otus <- data.frame(otu = colnames(select(one_mouse_df, contains('Otu'))))
+one_df <- c()
+om_pred <- c()
+for(i in 1:20){
+    set.seed(i)
+    train <- one_mouse_df %>% 
+      group_by(cage_id) %>% 
+      sample_n(1) %>% ungroup() 
+    test <- one_mouse_df %>% 
+      filter(!mouse_id %in% train$mouse_id)
+    set.seed(seed)
+    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id, -cage_id),
+                             importance = TRUE, ntree = n_trees)
+    om_rsq_rf[i] <- tail(rf_om$rsq, 1)
+    om_pred <- rbind(om_pred,
+                     data.frame(select(test, mouse_id, cage_id, log_cfu_med),
+                          predicted=predict(rf_om, test)))
+    om_rsq[i] <- 1 - (sum((om_pred$log_cfu_med - om_pred$predicted)^2)/
+               sum((om_pred$log_cfu_med - mean(om_pred$log_cfu_med))^2))
+    import_df <- importance(rf_om)
+    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
+    om_import_otus <- merge(om_import_otus, import_df, by.x = 'otu', by.y = 'row.names')
+    om_import_otus <- select(om_import_otus, -IncNodePurity)
+}
+
+om_pred <- rename(om_pred, observed = log_cfu_med)
+
+med_otu_importance_om <- data.frame(otu = om_import_otus$otu,
+                                 importance = apply(import_otus[, -1], 1, median))
+med_otu_importance_om <- med_otu_importance_om[order(med_otu_importance_om$importance, decreasing = T),]
+
+#### merge OOB and leave 1 cage out dataframes
+all_rf <- rbind(cbind(rf_prediction, data = rep('OOB', nrow(rf_prediction))),
+                cbind(LOcO_probs, data = rep('LOO', nrow(LOcO_probs))),
+                cbind(om_pred, data = rep('1_mouse', nrow(om_pred))))
+
+ggplot(data = all_rf, aes(x=observed, y=predicted, color = data)) + geom_point() + 
+  geom_abline(slope = 1) + 
+  labs(x= 'Observed Log10 CFU', y = 'Predicted Log10 CFU',
+       title = "Random Forest Prediction of Colonized C Difficile CFU from Day 0 Community") + 
+  theme_bw() + annotate('text', x=1.05*(min(LOcO_probs$observed)), 
+                        y=0.95*(max(LOcO_probs$predicted)), # parse = T,
+                        label=paste0('OOB ', 'R^2 == ', round(tail(rf_n_features$rsq, 1), 2),
+                                     '\nLOcO ','R^2 == ', round(LOcO_rsq, 2),
+                                     '\n1 mouse', 'R^2 == ', round(tail(om_rsq, 1), 2))) + 
+  theme(legend.position = 'none')
+
+get_tax(1, med_otu_importance$otu, tax_file)
+##### - leave one cage out cv
+
 ggplot(data = rf_prediction, aes(x=observed, y=predicted, color = cage_id)) + geom_point() + 
@@
        title = "Random Forest Prediction of Colonized C Difficile CFU from Day 0 Community") + 
-  theme_bw() + xlim(0,9) + ylim(0,9) + 
-  annotate('text', x=0.15*(max(rf_prediction$observed)), 
+  theme_bw() + annotate('text', x=1.05*(min(rf_prediction$observed)), 
                         y=0.95*(max(rf_prediction$predicted)), parse = T,
@@
 
-importance.df <- data.frame(otu = names(importance_final),
-                            importance = importance_final,
-                            tax = get_tax(1, names(importance_final), tax_file))
-tax_label <- paste0(importance.df$tax, ' (', 
-                    gsub('tu0*', 'TU ', importance.df$otu), 
-                    ') : ', 
+importance.df <- data.frame(get_tax(1, names(importance_final), tax_file),
+                            importance = importance_final)
+tax_label <- paste0(importance.df$tax_label, ' : ', 
                     round(importance.df$importance, 3))

### 2016-10-24 code/HumanCdGF_predict_cfu.Rmd
analyze diff between main feature selections

@@
 n_trees <- 1001
-n_features <- 10
 cutoff <- 30
@@
 
+test <- data.frame(model = c('boruta',rep('reiterative_rf', 6),rep('one_mouse_per_cg',4)),
+                   numOtus = c(length(boruta_otus), length(otus_used[[1]]),length(otus_used[[2]]),length(otus_used[[3]]),
+                               length(otus_used[[4]]),length(otus_used[[5]]),length(otus_used[[6]]),30,27,6,2),
+                   Rsq = c(tail(rf_boruta$rsq, 1), rf_rsq, 0.72, 0.71, 0.75, 0.71)
+)
+##### feature reduction #####
+### boruta
 set.seed(seed)
-rf_all <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df,
-                           importance = TRUE, ntree = n_trees, mtry = 14)
+rf_features <- Boruta(log_cfu_med ~ ., data = cfu_shared_colonized.df, maxRuns = 2000)
+boruta_otus <- names(rf_features$finalDecision)[rf_features$finalDecision == 'Confirmed']
+set.seed(seed)
+rf_boruta <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df[, c('log_cfu_med', boruta_otus)],
+                           importance = TRUE, ntree = n_trees)
+### above mean features, reiteratively reduce to 2 otus 
+###### used mean because median results in including many low values in early OTU selection steps
+rf_rsq <- c()
+otus_used <- c()
+rf_otus <- colnames(cfu_shared_colonized.df)[-1]
+for(i in 1:100){
+  if(length(rf_otus) > 2){
+  set.seed(seed)
+  rf <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df[, c('log_cfu_med', rf_otus)],
+                     importance = TRUE, ntree = n_trees)
+  rf_import_otus <- importance(rf)[,1]
+  otus_used[[i]] <- rf_otus
+  rf_otus <- rownames(data.frame(rf_import_otus[rf_import_otus >= mean(rf_import_otus)]))
+  rf_rsq[i] <- tail(rf$rsq, 1)
+  } else {
+    set.seed(seed)
+    rf <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df[, c('log_cfu_med', rf_otus)],
+                       importance = TRUE, ntree = n_trees)
+    otus_used[[i]] <- rf_otus
+    rf_rsq[i] <- tail(rf$rsq, 1)
+    stop('Only one OTU remains')
+  }
+}
+#### select top otus from reiteration through RF
+optimum_otus <- otus_used[[which(rf_rsq == max(rf_rsq))]]
+set.seed(seed)
+rf_all <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df[, c('log_cfu_med', optimum_otus)],
+                           importance = TRUE, ntree = n_trees)
 
-importance_sorted <- sort(importance(rf_all)[,1], decreasing = T)
-top_important_OTU <- rownames(data.frame(importance_sorted[importance_sorted/max(importance_sorted) > 0.6]))
+reduced_otus <- otus_used[[3]]
+set.seed(seed)
+rf_all <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df[, c('log_cfu_med', reduced_otus)],
+                           importance = TRUE, ntree = n_trees)
 
+### test models with top subset of otus
+top_important_OTU <- optimum_otus
 top_colonized_df <- data.frame(cfu_shared_colonized.df %>% 
@@
 rf_n_features <- randomForest(log_cfu_med ~ ., data=top_colonized_df,
-                              importance=TRUE, ntree=n_trees, mtry = 2,
-                              keep.forest=TRUE)
+                              importance=TRUE, ntree=n_trees, keep.forest=TRUE)
 
@@
 
-one_mouse_probs <- c()
-om_rsq <- c()
-om_rsq_rf <- c()
+om_rsq_rf <- c() # rsq of each rf model
+om_rsq <- c() # cumulative rsq of all models
 om_import_otus <- data.frame(otu = colnames(select(one_mouse_df, contains('Otu'))))
@@
 
+#####
+test <- c() # test if any model has same train set
+  for(i in 2:21){
+  test[i-1] <- colnames(om_import_otus)[i] %in% colnames(om_import_otus)[-c(1,i)]
+}
+any(test)
+#####
+
 om_pred <- rename(om_pred, observed = log_cfu_med)
@@
 med_otu_importance_om <- data.frame(otu = om_import_otus$otu,
-                                 importance = apply(import_otus[, -1], 1, median))
+                                 importance = apply(om_import_otus[, -1], 1, median))
 med_otu_importance_om <- med_otu_importance_om[order(med_otu_importance_om$importance, decreasing = T),]
@@
   geom_abline(slope = 1) + 
+  xlim(c(4,9)) + ylim(c(4,9)) +
   labs(x= 'Observed Log10 CFU', y = 'Predicted Log10 CFU',
@@
 
+n_features <- 10
+
 rsq <- matrix(NA, nrow = 11, ncol = 10)
@@
       mse[m+1, n] <- round(tail(rf_n_features$mse, 1), 2)
-      otu_votes <- otu_votes + 1*(colnames(rel_abund_shared_01) %in% top_important_OTU)
       importance.df <- rbind(importance.df, c(RSQ = round(tail(rf_n_features$rsq, 1), 2), 

### 2016-10-24 code/HumanCdGF_predict_cfu.Rmd
include down to 1 otu

@@
 for(i in 1:100){
-  if(length(rf_otus) > 2){
+  if(length(rf_otus) > 1){
   set.seed(seed)
@@
                      importance = TRUE, ntree = n_trees)
-  rf_import_otus <- importance(rf)[,1]
+  rf_import_otus <- importance(rf, type = 1)
   otus_used[[i]] <- rf_otus
@@
     rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id, -cage_id),
-                             importance = TRUE, ntree = n_trees)
+                             importance = TRUE, keep.forest = T, ntree = n_trees)
     om_rsq_rf[i] <- tail(rf_om$rsq, 1)
@@
 }
-
+plot(om_rsq_rf)
 #####

### 2016-10-25 code/HumanCdGF_predict_cfu.Rmd
remove misc code

@@
 
-test <- data.frame(model = c('boruta',rep('reiterative_rf', 6),rep('one_mouse_per_cg',4)),
-                   numOtus = c(length(boruta_otus), length(otus_used[[1]]),length(otus_used[[2]]),length(otus_used[[3]]),
-                               length(otus_used[[4]]),length(otus_used[[5]]),length(otus_used[[6]]),30,27,6,2),
-                   Rsq = c(tail(rf_boruta$rsq, 1), rf_rsq, 0.72, 0.71, 0.75, 0.71)
-)
 ##### feature reduction #####
@@
   otus_used[[i]] <- rf_otus
-  rf_otus <- rownames(data.frame(rf_import_otus[rf_import_otus >= mean(rf_import_otus)]))
+  rf_otus <- rownames(rf_import_otus)[rf_import_otus >= mean(rf_import_otus)]
   rf_rsq[i] <- tail(rf$rsq, 1)
@@
 optimum_otus <- otus_used[[which(rf_rsq == max(rf_rsq))]]
-set.seed(seed)
-rf_all <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df[, c('log_cfu_med', optimum_otus)],
-                           importance = TRUE, ntree = n_trees)
-
 reduced_otus <- otus_used[[3]]
-set.seed(seed)
-rf_all <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df[, c('log_cfu_med', reduced_otus)],
-                           importance = TRUE, ntree = n_trees)
 

### 2016-10-25 code/HumanCdGF_predict_cfu.Rmd
add cumulative importance for one mouse model

@@
     filter(!is.na(log_cfu_med))  %>% 
-    select(log_cfu_med, mouse_id, cage_id, one_of(top_important_OTU)))
+    select(log_cfu_med, mouse_id, cage_id, contains('Otu')))
 
@@
 }
-plot(om_rsq_rf)
+test <- data.frame(om_import_otus[ , -1], row.names = om_import_otus$otu)
+colnames(test) <- seq(ncol(test))
+
+test_mean <- apply(test, 2, function(x) mean(x, na.rm=T)) # mean importance for each rf model
+test_select <- apply(test, 1, function(x) x >= test_mean) # T for each otu above mean by sample
+test_raw_mean <- apply(test_select, 2, function(x) sum(x[!is.na(x)])/length(x[!is.na(x)])) # portion of rf model otu is above mean imp
+test_import_mean <- sort(test_raw_mean[test_raw_mean > 0.5], decreasing = T)
+
+#### median with 
+test_med <- apply(test, 2, function(x) median(x, na.rm=T)) # median importance for each rf model
+test_select <- apply(test, 1, function(x) x >= test_med) # T for each otu above median by sample
+test_raw_med <- apply(test_select, 2, function(x) sum(x[!is.na(x)])/length(x[!is.na(x)])) # portion of rf model otu is above med imp
+test_import_med <- sort(test_raw_med[test_raw_med > 0.5], decreasing = T)
+
+test_rank <- apply(test, 2, function(x) rank(x))
+test_raw_rank <- apply(test_rank, 1, function(x) sum(x, na.rm = T))
+sort(test_import_rank[test_raw_rank >=  ncol(test_rank)*sum(test_rank[ , 1])/nrow(test_rank)], decreasing = T)
+
 #####
@@
 }
-any(test)
+any(test) # if true, then trained on the same set of mice multiple times
 #####

### 2016-10-25 code/HumanCdGF_predict_cfu.Rmd
remove old heatplot code (eval=F)

@@
 n_trees <- 1001
-cutoff <- 30
 
@@
 ***
-
-```{r functions, eval = F}
-
-heat.plot <- function(input_maxtrix, high_color = '#660000', low_color = '#FFFFFF', legend_title = NA) {
-  long.df <- data.frame(input_maxtrix)
-  colnames(long.df) <- c(1:10)
-  long.df$community_day <- as.character(c(0:10))
-  long.df %>% gather(cfu_day, variable, -community_day) %>% 
-    ggplot(aes(cfu_day, community_day)) + 
-    geom_tile(aes(fill = variable), color = 'white') +
-    scale_fill_gradient(low = low_color, high = high_color, name = legend_title, limits = c(0,1)) +
-    labs(x = 'Day (CFU)', y = 'Day (Community)') + 
-    theme(axis.line.x = element_blank(), axis.line.y = element_blank(), axis.ticks = element_blank()) + 
-    scale_x_discrete(expand = c(0, 0), limits = c('1','2','3','4','5','6','7','8','9','10')) + 
-    scale_y_discrete(expand = c(0, 0), limits = rev(c('0','1','2','3','4','5','6','7','8','9','10'))) + 
-    theme(legend.position = 'top', legend.key.width = unit(2, 'cm'))
-}
-
-n_features <- 10
-
-rsq <- matrix(NA, nrow = 11, ncol = 10)
-mse <- matrix(NA, nrow = 11, ncol = 10)
-otu_votes <- rep(0, ncol(rel_abund_shared_01))
-names(otu_votes) <- colnames(rel_abund_shared_01)
-importance.df <- c()
-for(m in 0:10){
-  sample_list <- shared_list_days %>% 
-    filter(day == m) %>% 
-    select(-day)
-  community.df <- rel_abund_shared_01 %>% 
-    right_join(sample_list) %>% 
-    select(-sample_id)
-    for(n in 1:10){
-      cfu_shared.df <- meta_file %>% 
-        filter(day == n) %>% 
-        select(mouse_id, log_cfu) %>% 
-        right_join(community.df) %>% 
-        select(-mouse_id) %>% 
-        filter(!is.na(log_cfu))
-
-      set.seed(seed)
-      rf_all <- randomForest(log_cfu ~ ., data = cfu_shared.df,
-                           importance = TRUE, ntree = n_trees)
-      importance_sorted <- sort(importance(rf_all)[,1], decreasing = T)
-      top_important_OTU <- rownames(data.frame(head(importance_sorted, n_features)))
-      set.seed(seed)
-      rf_n_features <- randomForest(log_cfu ~ ., data=cfu_shared.df[ ,c( 'log_cfu',
-                                                                 names(importance_sorted)[1:n_features])], 
-                                  importance=TRUE, ntree=n_trees)
-      rsq[m+1, n] <- round(tail(rf_n_features$rsq, 1), 2)
-      mse[m+1, n] <- round(tail(rf_n_features$mse, 1), 2)
-      importance.df <- rbind(importance.df, c(RSQ = round(tail(rf_n_features$rsq, 1), 2), 
-                      MSE = round(tail(rf_n_features$mse, 1), 2),
-                      importance(rf_all)[,1]/max(importance(rf_all)[,1])))
-    }
-  }
-
-importance.df <- data.frame(importance.df)
-import_features <- importance.df[importance.df$RSQ >= 0.6 & importance.df$MSE <= 0.8, ]
-import_features <- import_features[ , !colnames(import_features) %in% c('RSQ','MSE')]
-med_import_feat <- apply(import_features, 2, median)
-otus_above_20 <- import_features[,names(med_import_feat)[med_import_feat >= 0.2]]
-top_n_overall <- names(otu_votes)[otu_votes >= cutoff]
-
-rsq_n <- matrix(NA, nrow = 11, ncol = 10)
-mse_n <- matrix(NA, nrow = 11, ncol = 10)
-otu_votes_n <- rep(0, ncol(rel_abund_shared_01))
-names(otu_votes_n) <- colnames(rel_abund_shared_01)
-
-for(m in 0:10){
-  sample_list <- shared_list_days %>% 
-    filter(day == m) %>% 
-    select(-day)
-  community.df <- rel_abund_shared_01 %>% 
-    right_join(sample_list) %>% 
-    select(-sample_id)
-  for(n in 1:10){
-    cfu_shared.df <- meta_file %>% 
-      filter(day == n) %>% 
-      select(mouse_id, log_cfu) %>% 
-      right_join(community.df) %>% 
-      select(log_cfu, one_of(top_n_overall)) %>% 
-      filter(!is.na(log_cfu))
-
-    set.seed(seed)
-    rf_all <- randomForest(log_cfu ~ ., data = cfu_shared.df,
-                           importance = TRUE, ntree = n_trees)
-    importance_sorted <- sort(importance(rf_all)[,1], decreasing = T)
-    top_important_OTU <- rownames(data.frame(head(importance_sorted, n_features)))
-    set.seed(seed)
-    rf_n_features <- randomForest(log_cfu ~ ., data=cfu_shared.df[ ,c( 'log_cfu',
-                                                                       names(importance_sorted)[1:n_features])], 
-                                  importance=TRUE, ntree=n_trees)
-    rsq_n[m+1, n] <- round(tail(rf_n_features$rsq, 1), 2)
-    mse_n[m+1, n] <- round(tail(rf_n_features$mse, 1), 2)
-    otu_votes_n <- otu_votes_n + 1*(colnames(rel_abund_shared_01) %in% top_important_OTU)
-  }
-}
-taxa <- get_tax(1, names(otu_votes_n[otu_votes_n >= 1]), tax_file)
-taxa_names <- paste0(taxa$tax, ' (', gsub('tu0*', 'TU ', rownames(taxa)),')')
-```
-
-### OTUs selected by >`r cutoff`% of models generated by all OTU models 
-###### Using the following bacteria: - r taxa_names
-
-```{r top otus, fig.width=6, fig.height=6, eval = F}
-heat.plot(rsq_n, '#003300', legend_title = expression(paste('R'^'2')))
-#plot(otu_votes_n[otu_votes_n >= 1], xlab = 'OTU', ylab = 'Freq')
-```
-
-### CFU and Rel abundance plots
-
-```{r fig.height=5, fig.width=8, eval = F}
-meta_data %>% 
-  filter(!is.na(day), !cage_id == 'NP1', 
-         Early_Euth == FALSE, cdiff_strain == 431) %>% 
-  ggplot(aes(x = factor(day), y = log_cfu, color = cage_id)) + 
-    geom_line(aes(group = mouse_id), size=.25, alpha=0.4) + geom_point() +
-    stat_summary(fun.y=mean, geom="line", aes(group = cage_id)) + 
-    theme_bw() + labs(x = 'Day', y = 'C Diff CFU (log10)') +
-    theme(legend.justification=c(1,0), legend.position=c(1,0), 
-          legend.title=element_blank())
-```
-
-How does removing the low samples (Day 8 and CFUs around 0) affect prediction?
-
-```{r identify reason for lowered R2, fig.height=5, fig.width=8, eval = F}
-log_cutoff <- 2.5
-
-meta_data %>% 
-  filter(!sample_id == 'OUT2-2533-D8', log_cfu >= log_cutoff) %>% 
-  filter(!is.na(day), !cage_id == 'NP1', 
-         Early_Euth == FALSE, cdiff_strain == 431) %>% 
-  ggplot(aes(x = factor(day), y = log_cfu, color = cage_id)) + 
-    geom_line(aes(group = mouse_id), size=.25, alpha=0.4) + geom_point() +
-    stat_summary(fun.y=mean, geom="line", aes(group = cage_id)) + 
-    theme_bw() + labs(x = 'Day', y = 'C Diff CFU (log10)') +
-    theme(legend.justification=c(1,0), legend.position=c(1,0), 
-          legend.title=element_blank()) + scale_y_continuous(limits = c(0,9))
-
-meta_reduced <- meta_data %>% 
-  filter(!sample_id == 'OUT2-2533-D8', log_cfu >= log_cutoff) %>% 
-  filter(Early_Euth == FALSE, cdiff_strain == 431, !cage_id == 'NP1') %>% 
-  select(mouse_id, sample_id, day, log_cfu) %>% 
-  filter(!is.na(log_cfu))
-
-rsq_r <- matrix(NA, nrow = 11, ncol = 10)
-otu_votes_r <- rep(0, ncol(rel_abund_shared_01))
-names(otu_votes_r) <- colnames(rel_abund_shared_01)
-
-for(m in 0:10){
-  sample_list <- shared_list_days %>% 
-    filter(day == m) %>% 
-    select(-day)
-  community.df <- rel_abund_shared_01 %>% 
-    right_join(sample_list) %>% 
-    select(-sample_id)
-  for(n in 1:10){
-    cfu_shared.df <- meta_reduced %>% 
-      filter(day == n) %>% 
-      select(mouse_id, log_cfu) %>% 
-      right_join(community.df) %>% 
-      select(log_cfu, one_of(top_n_overall)) %>% 
-      filter(!is.na(log_cfu))
-
-    set.seed(seed)
-    rf_top <- randomForest(log_cfu ~ ., data = cfu_shared.df,
-                           importance = TRUE, ntree = n_trees)
-    importance_sorted <- sort(importance(rf_top)[,1], decreasing = T)
-    top_important_OTU <- rownames(data.frame(head(importance_sorted, n_features)))
-    set.seed(seed)
-    rf_n_features <- randomForest(log_cfu ~ ., data=cfu_shared.df[ ,c( 'log_cfu',
-                                                                       names(importance_sorted)[1:n_features])], 
-                                  importance=TRUE, ntree=n_trees)
-    importance_final <- sort(importance(rf_n_features)[,1], decreasing = T)
-    top_OTU_final <- rownames(data.frame(head(importance_final, n_features)))
-    rsq_r[m+1, n] <- round(tail(rf_n_features$rsq, 1), 2)
-    otu_votes_r <- otu_votes_r + 1*(colnames(rel_abund_shared_01) %in% top_OTU_final)
-  }
-}
-taxa_r <- get_tax(1, names(otu_votes_r[otu_votes_r >= 1]), tax_file)
-taxa_names_r <- paste0(taxa$tax, ' (', gsub('tu0*', 'TU ', rownames(taxa)),')')
-```
-
-```{r fig.width=6,fig.height=6, eval = F}
-heat.plot(rsq_r, '#003300', legend_title = expression(paste('R'^'2')))
-```
-
-Removing these data points improves predictability for day 6 and 8 but not the earlier days.
-If I increase the cutoff to remove all CFU below 5, there is a slight increase in day 3 cfu classification at the cost of decrease day 1 cfu classification.
-
-### Is there any abnormalities in the OTUs of any these days?
-
-```{r plot features,fig.width=8, fig.height=20, eval = F}
-meta_data %>% 
-  filter(!sample_id == 'OUT2-2533-D8', log_cfu >= log_cutoff) %>% 
-  filter(!cage_id == 'NP1', Early_Euth == FALSE, cdiff_strain == 431) %>% 
-  full_join(rel_abund_shared_01) %>% 
-  select(day, sample_id, mouse_id, cage_id, 
-         one_of(names(otu_votes_r[otu_votes_r >= 1]))) %>% 
-  gather(otu, rel_abund, Otu00002:Otu00250) %>% 
-  filter(!is.na(day)) %>% 
-  ggplot(aes(x=factor(day),y=rel_abund, color = cage_id)) +
-    geom_line(aes(group = mouse_id), size=.25, alpha=0.4) + geom_point() +
-    stat_summary(fun.y=mean, geom="line", aes(group = cage_id)) + 
-    facet_grid(otu ~., scales='free_y') +
-    theme_bw() + labs(x = 'Day', y = 'Relative Abundance (%)')
-```
-
-### Does taking the median CFU improve predicatbility?
-
-```{r predict median CFU heatmap, fig.width=6,fig.height=6, eval = F}
-
-rsq_med <- matrix(NA, nrow = 11, ncol = 10)
-otu_votes_med <- rep(0, ncol(rel_abund_shared_01))
-names(otu_votes_med) <- colnames(rel_abund_shared_01)
-for(n in 1:10){
-med_log_cfu <- meta_data %>% 
-  filter(day %in% c(n:10), !cage_id == 'NP1', 
-         Early_Euth == FALSE, cdiff_strain == 431) %>% 
-  select(mouse_id, cage_id, log_cfu) %>% 
-  group_by(mouse_id, cage_id) %>% 
-  summarise(log_cfu_med = median(log_cfu)) %>% ungroup()
-for(m in 0:10){
-  sample_list <- shared_list_days %>% 
-    filter(day == m) %>% 
-    select(-day)
-  community.df <- rel_abund_shared_01 %>% 
-    right_join(sample_list) %>% 
-    select(-sample_id)
-  cfu_shared.df <- med_log_cfu %>% 
-      select(mouse_id, log_cfu_med) %>% 
-      right_join(community.df) %>% 
-      select(-mouse_id) %>% 
-#      select(log_cfu_med, one_of(top_n_overall)) %>% 
-      filter(!is.na(log_cfu_med))
-    set.seed(seed)
-    rf_top <- randomForest(log_cfu_med ~ ., data = cfu_shared.df,
-                           importance = TRUE, ntree = n_trees)
-    importance_sorted <- sort(importance(rf_top)[,1], decreasing = T)
-    top_important_OTU <- rownames(data.frame(head(importance_sorted, n_features)))
-    set.seed(seed)
-    rf_n_features <- randomForest(log_cfu_med ~ ., data=cfu_shared.df[ ,c( 'log_cfu_med',
-                                                                       names(importance_sorted)[1:n_features])], 
-                                  importance=TRUE, ntree=n_trees)
-    importance_final <- sort(importance(rf_n_features)[,1], decreasing = T)
-    top_OTU_final <- rownames(data.frame(head(importance_final, n_features)))
-    rsq_med[m+1, n] <- round(tail(rf_n_features$rsq, 1), 2)
-    otu_votes_med <- otu_votes_med + 1*(colnames(rel_abund_shared_01) %in% top_OTU_final)
-}
-}
-
-heat.plot(rsq_med, '#003300', legend_title = expression(paste('R'^'2'))) + 
-  ggtitle('Day CFU is median value of Day n to Day 10')
-
-```
-
-###### Re-runnning the program and only focusing on the OTUs predicting day 9/10 to see if there is consistence in the predictive features gives the following selection  (including all samples):  
-Otu00004 Otu00005 Otu00015 Otu00019 Otu00030 Otu00199 Otu00200 Otu00250   
-      14       17       13       15       13       16       17       19
-
-```{r boruta feature selection, eval = F}
-bor_all <- c()
-
-for(m in 0:10){
-  sample_list <- shared_list_days %>% 
-    filter(day == m) %>% 
-    select(-day)
-  community.df <- rel_abund_shared_01 %>% 
-    right_join(sample_list) %>% 
-    select(-sample_id)
-    for(n in 9:10){
-      cfu_shared.df <- meta_file %>% 
-        filter(day == n) %>% 
-        select(mouse_id, log_cfu) %>% 
-        right_join(community.df) %>% 
-        select(-mouse_id) %>% 
-        filter(!is.na(log_cfu))
-
-      set.seed(seed)
-      bor_all[[(m*2)+(n-8)]] <- Boruta(log_cfu ~ ., data = cfu_shared.df, maxRuns = 2000)
-    }
-  }
-
-boruta.votes <- rep(0, ncol(rel_abund_shared_01))
-names(boruta.votes) <- colnames(rel_abund_shared_01)
-for(i in 1:22){
-  boruta.votes <- boruta.votes + 1*(bor_all[[i]]$finalDecision   == 'Confirmed')
-}
-plot(boruta.votes)
-boruta.votes[boruta.votes >= 16]
-
-colnames(otus_above_20)
-```
-
-Boruta Confirmed the following OTUs as important for predicting day 9/10 cfu:   
-Otu00004 Otu00005 Otu00015 Otu00030 Otu00081 Otu00199 Otu00200 Otu00250 Otu00297  
-      21       22       21       21       21       22       22       22       20
-
-Selecting OTUs through collecting the features from the most predictive community/cfu models (R^2 >= 0.6 and MSE <= 0.8), then converting all % Increase in MSE to relative values and taking the median value of of each OTU, then selecting OTUs that fall above the median value results in the following OTUs:
-"Otu00001" "Otu00002" "Otu00004" "Otu00005" "Otu00012" "Otu00014" "Otu00015" "Otu00016" "Otu00019" "Otu00028" "Otu00030" "Otu00048" "Otu00081" "Otu00101" "Otu00118" "Otu00199" "Otu00200" "Otu00250" "Otu00297"
-```{r }

### 2016-10-25 code/HumanCdGF_predict_cfu.Rmd
add multiple feature reduction methods

@@
 ### boruta
+library(Boruta)
 set.seed(seed)
@@
                            importance = TRUE, ntree = n_trees)
+rf_bor_import <- sort(importance(rf_boruta)[,1], decreasing = T)
+detach('package:Boruta', unload = T)
+detach('package:ranger', unload = T)
+detach('package:randomForest', unload = T)
+library(randomForest)
 ### above mean features, reiteratively reduce to 2 otus 
@@
 optimum_otus <- otus_used[[which(rf_rsq == max(rf_rsq))]]
-reduced_otus <- otus_used[[3]]
+reduced_otus <- otus_used[[6]]
 
 ### test models with top subset of otus
-top_important_OTU <- optimum_otus
+top_important_OTU <- reduced_otus
 top_colonized_df <- data.frame(cfu_shared_colonized.df %>% 
@@
     filter(!is.na(log_cfu_med))  %>% 
-    select(log_cfu_med, mouse_id, cage_id, contains('Otu')))
+    select(log_cfu_med, mouse_id, cage_id, Otu00250, Otu00005))
 
@@
 om_pred <- c()
-for(i in 1:20){
+for(i in 1:500){
     set.seed(i)
@@
 colnames(test) <- seq(ncol(test))
-
+### rank
+test_rank <- apply(test, 2, function(x) rank(x))
+test_raw_rank <- apply(test_rank, 1, function(x) sum(x, na.rm = T))
+test_import_rank <- sort(test_raw_rank[test_raw_rank >=  ncol(test_rank)*sum(test_rank[ , 1])/nrow(test_rank)], decreasing = T)
+### mean
 test_mean <- apply(test, 2, function(x) mean(x, na.rm=T)) # mean importance for each rf model
@@
 test_import_mean <- sort(test_raw_mean[test_raw_mean > 0.5], decreasing = T)
-
-#### median with 
+#### median
 test_med <- apply(test, 2, function(x) median(x, na.rm=T)) # median importance for each rf model
@@
 test_import_med <- sort(test_raw_med[test_raw_med > 0.5], decreasing = T)
-
-test_rank <- apply(test, 2, function(x) rank(x))
-test_raw_rank <- apply(test_rank, 1, function(x) sum(x, na.rm = T))
-sort(test_import_rank[test_raw_rank >=  ncol(test_rank)*sum(test_rank[ , 1])/nrow(test_rank)], decreasing = T)
 
@@
 ```
+
+## Feature reduction
+
+#### Boruta  
+Variance Explained = `r round(tail(rf_boruta$rsq, 1)*100, 2)`%  
+OTUs  
+`r rf_bor_import`
+
+#### Mean % Increase MSE  
+`r length(otus_used[[1]])` OTUs = Variance Explained = `r round(rf_rsq[[1]]*100, 2)`%  
+`r length(otus_used[[2]])` OTUs = Variance Explained = `r round(rf_rsq[[2]]*100, 2)`%  
+`r length(otus_used[[3]])` OTUs = Variance Explained = `r round(rf_rsq[[3]]*100, 2)`%  
+`r length(otus_used[[4]])` OTUs = Variance Explained = `r round(rf_rsq[[4]]*100, 2)`%  
+`r length(otus_used[[5]])` OTUs = Variance Explained = `r round(rf_rsq[[5]]*100, 2)`%  
+    - OTUs:   `r otus_used[[5]]`  
+`r length(otus_used[[6]])` OTUs = Variance Explained = `r round(rf_rsq[[6]]*100, 2)`%  
+    - OTUs:   `r otus_used[[6]]`  
+`r length(otus_used[[7]])` OTUs = Variance Explained = `r round(rf_rsq[[7]]*100, 2)`%  
+    - OTUs:   `r otus_used[[7]]`
+
+Either Enterococcus (Otu 5) or Clostridium_XlVa (OTU 19) alone allows for ~65% explained variance
+
+#### one mouse model beginning with all otus and removing OTUs ranked below half max
+Rsq 0.6505947 with all otus
+
+Rsq 0.6893368 with otus:
+Otu00001 Otu00002 Otu00003 Otu00004 Otu00005 Otu00006 Otu00009 Otu00011 Otu00012 Otu00014 Otu00015 Otu00016 Otu00017 Otu00018 Otu00019 Otu00021 Otu00022 
+ 19727.0   7883.0  10100.0  20087.0  19774.0  15910.0   6999.0  10213.5  13479.0  16050.0  19178.0  15146.0  11461.0   8879.0  21037.0   8616.0   8201.0 
+Otu00029 Otu00030 Otu00032 Otu00045 Otu00048 Otu00049 Otu00053 Otu00057 Otu00060 Otu00063 Otu00064 Otu00074 Otu00077 Otu00081 Otu00082 Otu00083 Otu00084 
+  9849.0  13750.0   8594.0   7568.0  14825.0  12015.0  12291.0  13443.0  10921.0   8801.0   7474.0   9516.0   8280.0   9914.0  11609.0  12119.0  12961.0 
+Otu00095 Otu00101 Otu00118 Otu00128 Otu00131 Otu00159 Otu00163 Otu00164 Otu00173 Otu00175 Otu00176 Otu00178 Otu00199 Otu00200 Otu00250 Otu00297 Otu00461 
+ 15665.0  15639.0  10566.0  10740.0   9099.0  14344.0   9217.0  18500.0  11910.0   8002.5  19911.0   8550.5  20788.0  21936.0  22797.0  20767.0   7897.5
+
+Rsq 0.7430718 with otus:
+Otu00001 Otu00004 Otu00005 Otu00006 Otu00012 Otu00014 Otu00015 Otu00016 Otu00019 Otu00030 Otu00048 Otu00057 Otu00095 Otu00101 Otu00159 Otu00164 Otu00176 
+    6855     6074     6119     5436     4061     4580     6285     4193     7725     3766     3350     3980     4305     4138     3282     5724     6645 
+Otu00199 Otu00200 Otu00250 Otu00297 
+    6845     7898     7511     6728
+
+Rsq 0.7447479 with otus:
+Otu00001 Otu00004 Otu00005 Otu00015 Otu00019 Otu00164 Otu00176 Otu00199 Otu00200 Otu00250 Otu00297 
+    3081     2423     3019     3236     3517     2546     2521     2830     3151     3739     2937
+
+Rsq 0.7572649 with otus:
+Otu00001 Otu00005 Otu00015 Otu00019 Otu00200 Otu00250 
+    1644     1580     1636     1589     2017     2034
+
+Rsq 0.7499362 with otus:
+Otu00005 Otu00019 Otu00200 Otu00250 
+     758     1121     1421     1700
+
+Rsq 0.7305611 with otus:
+Otu00005 Otu00019 
+     611      889
+
+Rsq 0.6887173 with otus:
+Otu00200 Otu00250 
+     550      950
+
+also tried:
+Otus	Rsq 
+15  5	0.59
+250 19	0.757
+250 5	0.712
+
+Otu		 Rsq
+Otu00001 0.5014594
+Otu00005 0.6353671
+Otu00015 0.4497236
+Otu00019 0.6319819
+Otu00200 0.6557830
+Otu00250 0.7009930
 

### 2016-10-25 code/HumanCdGF_predict_cfu.Rmd
refine feature selection

@@
 n_trees <- 1001
-
+n_otus <- 12
+iters <- 100
 ```
@@
 ##### feature reduction #####
-### boruta
-library(Boruta)
-set.seed(seed)
-rf_features <- Boruta(log_cfu_med ~ ., data = cfu_shared_colonized.df, maxRuns = 2000)
-boruta_otus <- names(rf_features$finalDecision)[rf_features$finalDecision == 'Confirmed']
-set.seed(seed)
-rf_boruta <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df[, c('log_cfu_med', boruta_otus)],
+###### select top n otus reiteratively and select otus that appear > 50% of iterations
+rf_n_otus <-lapply(1:iters, function(i){
+  set.seed(i)
+  rf <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df,
                      importance = TRUE, ntree = n_trees)
-rf_bor_import <- sort(importance(rf_boruta)[,1], decreasing = T)
-detach('package:Boruta', unload = T)
-detach('package:ranger', unload = T)
-detach('package:randomForest', unload = T)
-library(randomForest)
-### above mean features, reiteratively reduce to 2 otus 
-###### used mean because median results in including many low values in early OTU selection steps
-rf_rsq <- c()
-otus_used <- c()
-rf_otus <- colnames(cfu_shared_colonized.df)[-1]
-for(i in 1:100){
-  if(length(rf_otus) > 1){
-  set.seed(seed)
-  rf <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df[, c('log_cfu_med', rf_otus)],
-                     importance = TRUE, ntree = n_trees)
-  rf_import_otus <- importance(rf, type = 1)
-  otus_used[[i]] <- rf_otus
-  rf_otus <- rownames(rf_import_otus)[rf_import_otus >= mean(rf_import_otus)]
-  rf_rsq[i] <- tail(rf$rsq, 1)
-  } else {
-    set.seed(seed)
-    rf <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df[, c('log_cfu_med', rf_otus)],
-                       importance = TRUE, ntree = n_trees)
-    otus_used[[i]] <- rf_otus
-    rf_rsq[i] <- tail(rf$rsq, 1)
-    stop('Only one OTU remains')
-  }
-}
-#### select top otus from reiteration through RF
-optimum_otus <- otus_used[[which(rf_rsq == max(rf_rsq))]]
-reduced_otus <- otus_used[[6]]
+  names(sort(importance(rf)[,1], decreasing = T))[1:n_otus]
+})
+n_otu_freq <- table(unlist(rf_n_otus))
+top_important_OTU <- names(n_otu_freq)[n_otu_freq >= iters/2]
 
-### test models with top subset of otus
-top_important_OTU <- reduced_otus
 top_colonized_df <- data.frame(cfu_shared_colonized.df %>% 
@@
     filter(!is.na(log_cfu_med))  %>% 
-    select(log_cfu_med, mouse_id, cage_id, Otu00250, Otu00005))
+    select(log_cfu_med, mouse_id, cage_id, one_of(top_important_OTU)))
 
@@
 om_import_otus <- data.frame(otu = colnames(select(one_mouse_df, contains('Otu'))))
-one_df <- c()
 om_pred <- c()
-for(i in 1:500){
+for(i in 1:iters){
     set.seed(i)
@@
 ### mean
-test_mean <- apply(test, 2, function(x) mean(x, na.rm=T)) # mean importance for each rf model
-test_select <- apply(test, 1, function(x) x >= test_mean) # T for each otu above mean by sample
-test_raw_mean <- apply(test_select, 2, function(x) sum(x[!is.na(x)])/length(x[!is.na(x)])) # portion of rf model otu is above mean imp
-test_import_mean <- sort(test_raw_mean[test_raw_mean > 0.5], decreasing = T)
+#test_mean <- apply(test, 2, function(x) mean(x, na.rm=T)) # mean importance for each rf model
+#test_select <- apply(test, 1, function(x) x >= test_mean) # T for each otu above mean by sample
+#test_raw_mean <- apply(test_select, 2, function(x) sum(x[!is.na(x)])/length(x[!is.na(x)])) # portion of rf model otu is above mean imp
+#test_import_mean <- sort(test_raw_mean[test_raw_mean > 0.5], decreasing = T)
 #### median
-test_med <- apply(test, 2, function(x) median(x, na.rm=T)) # median importance for each rf model
-test_select <- apply(test, 1, function(x) x >= test_med) # T for each otu above median by sample
-test_raw_med <- apply(test_select, 2, function(x) sum(x[!is.na(x)])/length(x[!is.na(x)])) # portion of rf model otu is above med imp
-test_import_med <- sort(test_raw_med[test_raw_med > 0.5], decreasing = T)
+#test_med <- apply(test, 2, function(x) median(x, na.rm=T)) # median importance for each rf model
+#test_select <- apply(test, 1, function(x) x >= test_med) # T for each otu above median by sample
+#test_raw_med <- apply(test_select, 2, function(x) sum(x[!is.na(x)])/length(x[!is.na(x)])) # portion of rf model otu is above med imp
+#test_import_med <- sort(test_raw_med[test_raw_med > 0.5], decreasing = T)
 
 #####
-test <- c() # test if any model has same train set
-  for(i in 2:21){
-  test[i-1] <- colnames(om_import_otus)[i] %in% colnames(om_import_otus)[-c(1,i)]
-}
-any(test) # if true, then trained on the same set of mice multiple times
+#test <- c() # test if any model has same train set
+#  for(i in 2:21){
+#  test[i-1] <- colnames(om_import_otus)[i] %in% colnames(om_import_otus)[-c(1,i)]
+#}
+#any(test) # if true, then trained on the same set of mice multiple times
 #####
@@
 
-med_otu_importance_om <- data.frame(otu = om_import_otus$otu,
-                                 importance = apply(om_import_otus[, -1], 1, median))
-med_otu_importance_om <- med_otu_importance_om[order(med_otu_importance_om$importance, decreasing = T),]
+med_otu_importance_om <- data.frame(otu = names(test_import_rank),
+                                 importance = test_import_rank)
 
@@
 ggplot(data = all_rf, aes(x=observed, y=predicted, color = data)) + geom_point() + 
-  geom_abline(slope = 1) + 
+  # geom_abline(slope = 1) + 
   xlim(c(4,9)) + ylim(c(4,9)) +
@@
        title = "Random Forest Prediction of Colonized C Difficile CFU from Day 0 Community") + 
-  theme_bw() + annotate('text', x=1.05*(min(LOcO_probs$observed)), 
-                        y=0.95*(max(LOcO_probs$predicted)), # parse = T,
+  theme_bw() + annotate('text', x = 4.5, y = 8.5, # parse = T,
                         label=paste0('OOB ', 'R^2 == ', round(tail(rf_n_features$rsq, 1), 2),

### 2016-10-25 code/HumanCdGF_predict_cfu.Rmd
update reduced features for one mouse model

@@
 #One per cage
-one_mouse_df <- data.frame(colonized_cfu %>% 
+one_mouse_initial_df <- data.frame(colonized_cfu %>% 
     right_join(community_d0.df) %>% 
@@
 
-om_rsq_rf <- c() # rsq of each rf model
-om_rsq <- c() # cumulative rsq of all models
-om_import_otus <- data.frame(otu = colnames(select(one_mouse_df, contains('Otu'))))
-om_pred <- c()
+om_raw_otus <- data.frame(otu = colnames(select(one_mouse_initial_df, contains('Otu'))))
 for(i in 1:iters){
     set.seed(i)
-    train <- one_mouse_df %>% 
+    train <- one_mouse_initial_df %>% 
       group_by(cage_id) %>% 
       sample_n(1) %>% ungroup() 
-    test <- one_mouse_df %>% 
+    test <- one_mouse_initial_df %>% 
       filter(!mouse_id %in% train$mouse_id)
@@
                              importance = TRUE, keep.forest = T, ntree = n_trees)
-    om_rsq_rf[i] <- tail(rf_om$rsq, 1)
-    om_pred <- rbind(om_pred,
-                     data.frame(select(test, mouse_id, cage_id, log_cfu_med),
-                          predicted=predict(rf_om, test)))
-    om_rsq[i] <- 1 - (sum((om_pred$log_cfu_med - om_pred$predicted)^2)/
-               sum((om_pred$log_cfu_med - mean(om_pred$log_cfu_med))^2))
     import_df <- importance(rf_om)
     colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
-    om_import_otus <- merge(om_import_otus, import_df, by.x = 'otu', by.y = 'row.names')
-    om_import_otus <- select(om_import_otus, -IncNodePurity)
+    om_raw_otus <- merge(om_raw_otus, import_df, by.x = 'otu', by.y = 'row.names')
+    om_raw_otus <- select(om_raw_otus, -IncNodePurity)
 }
-test <- data.frame(om_import_otus[ , -1], row.names = om_import_otus$otu)
+test <- data.frame(om_raw_otus[ , -1], row.names = om_raw_otus$otu)
 colnames(test) <- seq(ncol(test))
 ### rank
-test_rank <- apply(test, 2, function(x) rank(x))
+test_rank <- apply(test, 2, function(x) rank(x, na.last = F))
 test_raw_rank <- apply(test_rank, 1, function(x) sum(x, na.rm = T))
@@
 #####
+one_mouse_df <- data.frame(colonized_cfu %>% 
+    right_join(community_d0.df) %>% 
+    filter(!is.na(log_cfu_med))  %>% 
+    select(log_cfu_med, mouse_id, cage_id, one_of(names(test_import_rank))))
+
+om_import_otus <- data.frame(otu = colnames(select(one_mouse_df, contains('Otu'))))
+om_pred <- c()
+for(i in 1:iters){
+    set.seed(i)
+    train <- one_mouse_df %>% 
+      group_by(cage_id) %>% 
+      sample_n(1) %>% ungroup() 
+    test <- one_mouse_df %>% 
+      filter(!mouse_id %in% train$mouse_id)
+    set.seed(seed)
+    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id, -cage_id),
+                             importance = TRUE, keep.forest = T, ntree = n_trees)
+    om_pred <- rbind(om_pred,
+                     data.frame(select(test, mouse_id, cage_id, log_cfu_med),
+                          predicted=predict(rf_om, test)))
+    import_df <- importance(rf_om)
+    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
+    om_import_otus <- merge(om_import_otus, import_df, by.x = 'otu', by.y = 'row.names')
+    om_import_otus <- select(om_import_otus, -IncNodePurity)
+}
+om_rsq <- 1 - (sum((om_pred$log_cfu_med - om_pred$predicted)^2)/
+               sum((om_pred$log_cfu_med - mean(om_pred$log_cfu_med))^2))
 
@@
 
-med_otu_importance_om <- data.frame(otu = names(test_import_rank),
-                                 importance = test_import_rank)
+om_import_otus <- data.frame(om_import_otus[ , -1], row.names = om_import_otus$otu)
+colnames(om_import_otus) <- seq(ncol(om_import_otus))
+om_otu_rank <- apply(om_import_otus, 2, function(x) rank(x, na.last = F))
+om_raw_rank <- apply(om_otu_rank, 1, function(x) sum(x, na.rm = T))
+med_otu_importance_om <- data.frame(otu = names(om_raw_rank),
+                                 importance = om_raw_rank)
 
@@
 ```
-
-## Feature reduction
-
-#### Boruta  
-Variance Explained = `r round(tail(rf_boruta$rsq, 1)*100, 2)`%  
-OTUs  
-`r rf_bor_import`
-
-#### Mean % Increase MSE  
-`r length(otus_used[[1]])` OTUs = Variance Explained = `r round(rf_rsq[[1]]*100, 2)`%  
-`r length(otus_used[[2]])` OTUs = Variance Explained = `r round(rf_rsq[[2]]*100, 2)`%  
-`r length(otus_used[[3]])` OTUs = Variance Explained = `r round(rf_rsq[[3]]*100, 2)`%  
-`r length(otus_used[[4]])` OTUs = Variance Explained = `r round(rf_rsq[[4]]*100, 2)`%  
-`r length(otus_used[[5]])` OTUs = Variance Explained = `r round(rf_rsq[[5]]*100, 2)`%  
-    - OTUs:   `r otus_used[[5]]`  
-`r length(otus_used[[6]])` OTUs = Variance Explained = `r round(rf_rsq[[6]]*100, 2)`%  
-    - OTUs:   `r otus_used[[6]]`  
-`r length(otus_used[[7]])` OTUs = Variance Explained = `r round(rf_rsq[[7]]*100, 2)`%  
-    - OTUs:   `r otus_used[[7]]`
-
-Either Enterococcus (Otu 5) or Clostridium_XlVa (OTU 19) alone allows for ~65% explained variance
-
-#### one mouse model beginning with all otus and removing OTUs ranked below half max
-Rsq 0.6505947 with all otus
-
-Rsq 0.6893368 with otus:
-Otu00001 Otu00002 Otu00003 Otu00004 Otu00005 Otu00006 Otu00009 Otu00011 Otu00012 Otu00014 Otu00015 Otu00016 Otu00017 Otu00018 Otu00019 Otu00021 Otu00022 
- 19727.0   7883.0  10100.0  20087.0  19774.0  15910.0   6999.0  10213.5  13479.0  16050.0  19178.0  15146.0  11461.0   8879.0  21037.0   8616.0   8201.0 
-Otu00029 Otu00030 Otu00032 Otu00045 Otu00048 Otu00049 Otu00053 Otu00057 Otu00060 Otu00063 Otu00064 Otu00074 Otu00077 Otu00081 Otu00082 Otu00083 Otu00084 
-  9849.0  13750.0   8594.0   7568.0  14825.0  12015.0  12291.0  13443.0  10921.0   8801.0   7474.0   9516.0   8280.0   9914.0  11609.0  12119.0  12961.0 
-Otu00095 Otu00101 Otu00118 Otu00128 Otu00131 Otu00159 Otu00163 Otu00164 Otu00173 Otu00175 Otu00176 Otu00178 Otu00199 Otu00200 Otu00250 Otu00297 Otu00461 
- 15665.0  15639.0  10566.0  10740.0   9099.0  14344.0   9217.0  18500.0  11910.0   8002.5  19911.0   8550.5  20788.0  21936.0  22797.0  20767.0   7897.5
-
-Rsq 0.7430718 with otus:
-Otu00001 Otu00004 Otu00005 Otu00006 Otu00012 Otu00014 Otu00015 Otu00016 Otu00019 Otu00030 Otu00048 Otu00057 Otu00095 Otu00101 Otu00159 Otu00164 Otu00176 
-    6855     6074     6119     5436     4061     4580     6285     4193     7725     3766     3350     3980     4305     4138     3282     5724     6645 
-Otu00199 Otu00200 Otu00250 Otu00297 
-    6845     7898     7511     6728
-
-Rsq 0.7447479 with otus:
-Otu00001 Otu00004 Otu00005 Otu00015 Otu00019 Otu00164 Otu00176 Otu00199 Otu00200 Otu00250 Otu00297 
-    3081     2423     3019     3236     3517     2546     2521     2830     3151     3739     2937
-
-Rsq 0.7572649 with otus:
-Otu00001 Otu00005 Otu00015 Otu00019 Otu00200 Otu00250 
-    1644     1580     1636     1589     2017     2034
-
-Rsq 0.7499362 with otus:
-Otu00005 Otu00019 Otu00200 Otu00250 
-     758     1121     1421     1700
-
-Rsq 0.7305611 with otus:
-Otu00005 Otu00019 
-     611      889
-
-Rsq 0.6887173 with otus:
-Otu00200 Otu00250 
-     550      950
-
-also tried:
-Otus	Rsq 
-15  5	0.59
-250 19	0.757
-250 5	0.712
-
-Otu		 Rsq
-Otu00001 0.5014594
-Otu00005 0.6353671
-Otu00015 0.4497236
-Otu00019 0.6319819
-Otu00200 0.6557830
-Otu00250 0.7009930
 

### 2016-10-26 code/HumanCdGF_predict_cfu.Rmd
update prediction/partial plots for colonize level

@@
 
-ggplot(data = all_rf, aes(x=observed, y=predicted, color = data)) + geom_point() + 
-  # geom_abline(slope = 1) + 
-  xlim(c(4,9)) + ylim(c(4,9)) +
+ggplot(data = all_rf, aes(x=observed, y=predicted, color = data)) + 
+  geom_abline(color = 'gray', size = 0.5, linetype = 2, slope = 1) + 
+  # geom_point() + 
+  geom_smooth(method = 'lm', se = F) +
+  stat_summary(fun.y=median, geom="point", aes(group = data)) + 
+  stat_summary(fun.data = 'median_hilow') +
+  xlim(c(5,9)) + ylim(c(5,9)) +
   labs(x= 'Observed Log10 CFU', y = 'Predicted Log10 CFU',
        title = "Random Forest Prediction of Colonized C Difficile CFU from Day 0 Community") + 
-  theme_bw() + annotate('text', x = 4.5, y = 8.5, # parse = T,
-                        label=paste0('OOB ', 'R^2 == ', round(tail(rf_n_features$rsq, 1), 2),
-                                     '\nLOcO ','R^2 == ', round(LOcO_rsq, 2),
-                                     '\n1 mouse', 'R^2 == ', round(tail(om_rsq, 1), 2))) + 
-  theme(legend.position = 'none')
+  theme_bw() + 
+  scale_color_discrete(name = 'Model',
+                       breaks = c('OOB', 'LOO', '1_mouse'),
+                       labels = c(bquote('OOB (R'^'2'*' = '*.(round(tail(rf_n_features$rsq, 1), 2))*')'),
+                                  bquote('LOO (R'^'2'*' = '*.(round(LOcO_rsq, 2))*')'),
+                                  bquote('OM (R'^'2'*' = '*.(round(om_rsq,2))*')'))) + 
+  theme(legend.justification=c(0,1), legend.position=c(0,1))
 
-get_tax(1, med_otu_importance$otu, tax_file)
-##### - leave one cage out cv
+importance_om <- get_tax(1, med_otu_importance_om$otu, tax_file)
 
-ggplot(data = rf_prediction, aes(x=observed, y=predicted, color = cage_id)) + geom_point() + 
-  labs(x= 'Observed CFU (log10)', y = 'Predicted CFU (log10)',
-       title = "Random Forest Prediction of Colonized C Difficile CFU from Day 0 Community") + 
-  theme_bw() + annotate('text', x=1.05*(min(rf_prediction$observed)), 
-                        y=0.95*(max(rf_prediction$predicted)), parse = T,
-                        label=paste0('R^2 == ', round(tail(rf_n_features$rsq, 1), 2))) + 
-  theme(legend.position = 'none')
+partial_df <- data.frame(colonized_cfu %>% 
+    right_join(community_d0.df) %>% 
+    select(mouse_id, cage_id, log_cfu_med, one_of(rownames(importance_om))) %>% 
+    filter(!is.na(log_cfu_med)))
 
-importance.df <- data.frame(get_tax(1, names(importance_final), tax_file),
-                            importance = importance_final)
-tax_label <- paste0(importance.df$tax_label, ' : ', 
-                    round(importance.df$importance, 3))
+color <- sapply(1:length(unique(partial_df$cage_id)), 
+                           function(n)rgb(sample(c(0:255), 1),sample(c(0:255), 1),
+                                          sample(c(0:255), 1), maxColorValue = 255))
 
-layout(matrix(c(1:9), ncol=3, byrow=TRUE))
-
-cage_color <- rainbow(length(unique(rf_prediction$cage_id)))
-for (i in seq_along(importance_final)) {
-    partialPlot(rf_n_features, top_colonized_df, names(importance_final)[i],
-                xlab='', main=paste(tax_label[i]))
+layout(matrix(c(1:4), ncol=2, byrow=TRUE))
+plot_loop_output <- lapply(seq_along(importance_om$tax), function(i){
+    partialPlot(rf_n_features, top_colonized_df, rownames(importance_om)[i],
+                xlab='', main=paste(importance_om$tax_label[i]))
     par(new = T)
     plot(y = top_colonized_df[ , 'log_cfu_med'],
-         x = top_colonized_df[ , names(importance_final)[i]], 
+         x = top_colonized_df[ , rownames(importance_om)[i]], 
          pch = 21, bg = 'gray', xlab = NA, ylab = NA, axes = F)
@@
     mtext(side = 4, line = 3, 'Log10 CFU')
-# To color points by cage - Change pch to 21 and un-comment following lines
-#  for(m in seq_along(unique(rf_prediction$cage_id))){
-#    cage <- unique(rf_prediction$cage_id)[m]
-#    points(y = top_colonized_df[rf_prediction$cage_id %in% cage, 'log_cfu_med'],
-#         x = top_colonized_df[rf_prediction$cage_id %in% cage , names(importance_final)[i]], 
-#         pch = 21, bg = cage_color[m], xlab = NA, ylab = NA, axes = F)
-#  }
-}
-
+# To color points by mouse - Change pch to 21 and un-comment following lines
+  lapply(seq_along(unique(partial_df$cage_id)), function(m){
+    mouse <- unique(partial_df$cage_id)[m]
+    points(y = partial_df[partial_df$cage_id %in% mouse, 'log_cfu_med'],
+           x = partial_df[partial_df$cage_id %in% mouse , rownames(importance_om)[i]], 
+           pch = 21, cex = 2, bg = color[m], xlab = NA, ylab = NA, axes = F)
+  })
+})
+rm(plot_loop_output)
 ```

### 2016-10-27 submission/study.tex
converting manuscript draft to Rmd

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Microbiota predict Clostridium difficile severity in germ-free mice colonized with human feces},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{\textbf{Microbiota predict \emph{Clostridium difficile} severity in
germ-free mice colonized with human feces}}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

\usepackage[none]{hyphenat}

\usepackage{setspace}
\doublespacing
\setlength{\parskip}{1em}

\usepackage{lineno}

\usepackage{pdfpages}

\begin{document}

\maketitle

\vspace{35mm}

Running title: Microbiota predict \emph{C. difficile} severity in
humanized mice

\vspace{35mm}

Kaitlin J. Flynn\textsuperscript{1}, Nicholas
Lesniak\textsuperscript{1}, Alyxandria M. Schubert\textsuperscript{2},
Hamide Sinani\textsuperscript{?}, and Patrick D.
Schloss\textsuperscript{1\(\dagger\)}

\vspace{40mm}

\(\dagger\) Corresponding author:
\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}

1. Department of Microbiology and Immunology, University of Michigan,
Ann Arbor, Michigan 48109

2. Food and Drug Administration?

3. Department for Hamide?

\newpage
\linenumbers

\subsubsection{Abstract}\label{abstract}

\emph{Clostridium difficile} causes diarrheal disease when it
successfully colonizes a dysbiotic gut microbial community. Current
mouse models to study \emph{C. difficile} infection (CDI) rely on
pre-treatment with antibiotics to disrupt the mouse microbiome prior to
inoculation. This model does not allow for analysis of human-associated
microbial community members that modulate \emph{C. difficile}
colonization and expansion. To study human-associated microbes in the
context of CDI, we inoculated germ-free C57BL/6 mice with one of 16
human fecal samples from diarrheal or healthy donors and challenged with
C. difficile 14 days later. Five unique donor-mice combinations resulted
in severe CDI while the remaining 11 only experienced mild disease. Both
healthy and diarrheal donors were susceptible to colonization and severe
symptoms of CDI. To determine if specific microbes were associated with
disease severity outcomes, we built a classification Random Forest
machine learning model based on relative abundance data of the
communities prior to infection. The model identified a number of
bacterial populations associated with the development of severe CDI,
including \emph{Bacilliales, Ruminococcaceae, Ruminococcus,
Staphylococcus, Streptococcus and Bacteriodetes}. Additionally, a
regression model accurately predicted colonization levels of \emph{C.
difficile} at one to ten days post-infection. This model explained 99\%
of the variance in the number of CFU isolated from mouse stool. Members
of \emph{Lachnospiraceae, Parabacteroides, Bacteroidales, Bacteroidetes,
Porphyromonadaceae} and unclassified \emph{Bacteria} families were
predictive of future \emph{C. difficile} colonization levels. Finally,
challenging these mice with different strains of \emph{C. difficile}
revealed that susceptible human-associated microbial communities were
prone to severe disease independent of strain type. Taken together these
results suggest that human-associated microbial communities can be
recapitulated in germ-free mice and used to characterize dynamics of
CDI. Because both healthy and diarrheal patients were susceptible to
severe CDI, machine-learning models are useful to identify bacterial
populations that allow colonization and contribute to the development of
\emph{C. difficile} associated disease in humans.

\newpage

\subsubsection{Introduction}\label{introduction}

Clostridium difficile is an opportunistic pathogen of the human lower
gastrointestinal tract. Disruption of the native microbial community of
the gut by antibiotics is the most common risk factor for development of
C. difficile infection (CDI) (1). C. difficile is a spore-forming
bacteria and can persist on abiotic surfaces and is not readily killed
by ethanol-based hand-sanitizers, putting hospital patients particularly
at risk. Indeed, \textasciitilde{}12\% of hospital acquired infections
in the United States are due to C. difficile and result in up to 15,000
deaths annually (2).

Murine models to study CDI typically rely on treating
conventionally-raised mice with antibiotics either in drinking water or
by injection to induce susceptibility (3, 4). This model provides a
convenient way to study C. difficile pathogenesis and virulence factors.
Numerous microbiome studies have been performed using this model to
determine the antibiotic classes (5), starting microbial community (6)
and metabolites (7) that impact development and severity of CDI. While
informative, these studies are somewhat removed from human disease
because they only examine mouse-associated microbial communities.

Gnotobiotic or germ-free mouse models have been used for a range of
studies of CDI, including assessment of species-specific interactions
between C. difficile and competing microbial community members (8),
analysis of nutrient restriction (9), in vivo transcriptomics of C.
difficile and examination of host immune response to CDI (10). Further,
CDI therapeutics such as antibiotics and fecal microbiota transplants
have been tested extensively in a gnotobiotic-piglet or
piglet-to-gnotobiotic-mouse model of disease (11), (12). Pigs have a
longer digestive tract with components more similar to humans than mice
and are typically infected by strains typical in human infection (13).
However, the murine and porcine microbiomes typically do not resemble
those of the human gut.

The power of the gnotobiotic models to study CDI has been further
realized by first inoculating germ-free mice and piglets with human
stool microbes. In one study, germ-free piglets were acutely colonized
with human feces for one week and then treated with tigecycline. After
challenge with C. difficile none of the antibiotic-treated piglets
succumbed to infection, while some of the untreated human-colonized pigs
did (11). Further, germ-free mice colonized with human feces were bred
over several generations to create a cohort of mice with identical
human-derived microbiomes (14). These mice were subsequently treated
with a five-antibiotic cocktail to induce dysbiosis and then were
successfully colonized by C. difficile (14). While informative, these
studies were limited in their use of only one human donor as input
inoculum. In order to best understand the impact of C. difficile
pathogenesis on human disease, we must have a laboratory model that
allows for study of a variety of human-derived microbiomes.

To test the impact of individual human microbiomes on CDI, we colonized
germ-free mice with 16 different human stool donors. We then
characterized human-associated microbiome response to C. difficile
challenge. Additionally, the use of machine-learning models allowed us
to build a predictive model that classified ``at-risk'' microbiomes
prior to infection with C. difficile. These findings show that
human-associated microbiomes can be at risk for CDI even in the absence
of antibiotics and that study of mice colonized with human feces
provides a range of clinical outcomes.

\subsubsection{Results}\label{results}

\subsubsection{Discussion}\label{discussion}

\subsubsection{Materials and Methods}\label{materials-and-methods}

\newpage

Insert figure legends with the first sentence in bold, for example:

\textbf{Figure 1. Number of OTUs sampled among bacterial and archaeal
16S rRNA gene sequences for different OTU definitions and level of
sequencing effort.} Rarefaction curves for different OTU definitions of
Bacteria (A) and Archaea (B). Rarefaction curves for the coarse
environments in Table 1 for Bacteria (C) and Archaea (D).

\newpage

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

\end{document}

### 2016-10-27 code/inocula_nmds.R

@@ -108,8 +108,13 @@ legend(x="topright", legend, col = c("red", "black"), pch=16)
 #amova in mothur output for asymptomatic vs severe DONORS: p-value: 0.508
 
 #nmds inocula plot with clinical metadata, use this one 
+
+pdf("fig1B_actual.pdf", width = 8, height = 5)
+
 ggplot(inocula_meta_nmds, aes(axis1, axis2, group = Disease_status, color = Disease_status)) + geom_point(size = 3) + 
-  theme_bw() + ggtitle("Similarity of donor inocula communities") + xlab("NMDS Axis 1") + ylab("NMDS Axis 2")
+  theme_bw() + xlab("NMDS Axis 1") + ylab("NMDS Axis 2") + theme(legend.justification = c(1, 0), legend.position = c(1, 0))
+
+dev.off()
 
 plot3d(inocula_meta_nmds$axis1, inocula_meta_nmds$axis2, inocula_meta_nmds$axis3, type = "p", col = as.numeric(inocula_meta_nmds$Disease_status), size = 5, xlab='NMDS axis 1', ylab= 'NMDS axis 2', zlab= 'NMDS axis 3')
 legend3d("topleft", legend= paste(c("C. difficile", "No diarrhea", "Diarrhea")), pch=16, col=c('black', 'green', 'red'), inset=c(0.08))

### 2016-10-27 code/tax_level.R
output as character

@@ -21,7 +21,8 @@ get_tax <- function(tax_level=1,row_list=1:length(rownames(tax_file)),df=tax_fil
           tax_out <- gsub('_unclassified', '', tax_out)
           label_out <- paste0(tax_out, ' (', 
                               gsub('tu0*', 'TU ', row_list),')')
-          return(data.frame(tax= tax_out, tax_label = label_out, row.names=row_list))
+          return(data.frame(tax= tax_out, tax_label = label_out, row.names=row_list,
+                            stringsAsFactors = FALSE))
      } else {print(
           'Error: Level must be 1 (genus), 2 (family), 3 (order), 4 (class), 5 (phylum)')
      }

### 2016-10-27 submission/study.tex

@@
 
+\textbf{Germ-free mice inoculated with human feces as model for C.
+difficile infection.} To generate mice with human-derived microbiomes,
+we inoculated one cage of gnotobiotic C57/BL6 mice with one of 16
+different human fecal donors. Five donors were patients that had
+diarrhea that was not attributable to \emph{C. difficile} infection
+while 11 donors were healthy at time of donation. Stool from a patient
+that was colonized with virulent \emph{C. difficile} was used as a
+positive control. After inoculation with human stool, mice were allowed
+to equilibrate for 14 days. Prior to infection, stool samples were taken
+from each mouse to establish baseline. Then, the \emph{C. difficile}
+strain isolated from the positive control patient's sample (strain 430)
+was used to infect each mouse with 100 spores. Mice were monitored for
+weight loss and clinical signs of disease. Fecal samples were taken to
+enumerate \emph{C. difficile} CFU and for microbiome analysis every day
+for up to 10 days post-infection (Fig 1A). To ensure that the donors we
+selected represented a diverse array of human microbiomes, we sequenced
+the 16S rRNA genes from donor fecal inocula. Ordination of the distances
+between donor communities showed that the donors each had distinctly
+different communities, independent of whether the sample came from a
+sick or healthy person (Fig 1B). Likewise, the starting microbial
+communities of the mice on day 0 were characterizing by sequencing of
+fecal pellets DNA prior to infection. Ordination of all of the mouse
+communities on day 0 shows that mice were similar to each other within
+each cage and donor, but distinct from other donors (Fig 1C). This
+result confirmed that human-associated microbes were able to colonize
+gnotobiotic mice and provide distinct initial communities to test
+\emph{C. difficile} dynamics.
+
+\textbf{C. difficile infection in mice with human-derived microbiota
+cause a range of outcomes.} \emph{C. difficile} colonization was
+monitored by daily plating of stool pellets for \emph{C. difficile} CFU.
+Nearly all of the mice were colonized to 10\textsuperscript{5} --
+10\textsuperscript{7} CFU by one day post-infection and remained
+colonized at that level until the end of the experiment (Fig 2A). As one
+indicator of disease, mouse weights were taken each day post-infection
+and weight-loss was monitored alongside clinical signs of disease. When
+mice were judged to be too ill to continue they were humanely
+euthanized. Overall, disease phenotypes fell into two classes. Mice that
+became severely ill and lost 20\% or more of their starting body weight
+within one to two days post-infection were classified as ``severe''
+whereas mice that were colonized with \emph{C. difficile} but did not
+show signs of disease or severe weight loss were considered to have
+``mild'' disease (Fig 2A, 2B). Interestingly, \emph{C. difficile} was
+able to cause severe disease in both mice that had been colonized with
+healthy stool and those colonized with diarrheal stool, suggesting
+susceptibility to CDI is dependent on the composition of the starting
+microbiome and not associated with donor clinical metadata.
+
+\textbf{Results to be written} 1. Microbes present in the gut prior to
+infection are predictive of C. difficile CFU and severity a. Figure 3:
+Random forest to predict CFU, predictive OTUs b. Figure 4: Random Forest
+predicts CDI severity, predictive OTUs 2. Propensity for severe CDI is
+community-dependent and strain-independent a. Figure 5: Infection of
+mice with different C. difficile strains.
+
 \subsubsection{Discussion}\label{discussion}
+
+• Restate results, • caveats about mouse weights • No donors were
+colonization resistant, discuss donor differences • Discuss prediction
+methods and outcomes • Discuss potential mechanisms for interesting OTUs
+• Discuss different strain results • Future work blah blah
 
@@
 
+• Mice ULAM number • Donor stool ERIN IRB shit • Bacteria/plating •
+Sequencing • Data analysis, code availability • Machine learning models
+
+\subsection{Acknowledgments}\label{acknowledgments}
+
+Lab, sequencing core, Jhansi
+
 \newpage
 
-Insert figure legends with the first sentence in bold, for example:
+\subsubsection{Figure Legends}\label{figure-legends}
 
-\textbf{Figure 1. Number of OTUs sampled among bacterial and archaeal
-16S rRNA gene sequences for different OTU definitions and level of
-sequencing effort.} Rarefaction curves for different OTU definitions of
-Bacteria (A) and Archaea (B). Rarefaction curves for the coarse
-environments in Table 1 for Bacteria (C) and Archaea (D).
+\textbf{Figure 1. Germ-free mice inoculated with human feces as a model
+for C. difficile infection.} A) Experimental design. Stool from 16
+healthy, diarrheal and CDI patients were independently inoculated into
+3-4 germ-free mice by oral gavage. 14 days later mice were orally
+infected with 100 spores of C. difficile strain 431. Weight and stool
+CFU were monitored for up to 10 days post infection. B) NMDS ordination
+of donor stool communities prior to inoculating mice. Each point
+represents one donor and are colored by clinical diagnosis. C) NDMS
+ordination of the stool communities on day 0. Each symbol represents one
+mouse and is colored by donor. Circles represent mice that experienced
+mild disease and triangles represent those that suffered severe disease.
+
+\textbf{Figure 2. \emph{C. difficile} infection results in mild or
+severe disease.} A) \emph{C. difficile CFU} was enumerated by plating of
+mouse stool pellets daily. Each point represents a mouse and the lines
+represent the median CFU in each group. Error bars are interquartile
+ranges. Red lines and points correspond to mice that succumbed to severe
+disease whereas black lines and points correspond to mice that had mild
+or no disease. B) Mouse weights were recorded and daily percent weight
+loss calculated for each mouse. Data is presented as the median of each
+group and interquartile ranges. Mice that succumbed to severe infection
+typically lost a significant amount of weight by day 1 or 2 post
+infection. Red lines correspond to severely ill mice, black to mice with
+mild disease.
+
+\textbf{Figure 3. Random Forest prediction of \emph{C. difficile}
+colonization level.} A) OTUs above 1\% relative abundance on day 0 were
+used to predict median log\textsubscript{10} CFU of \emph{C. difficile}
+after colonization. OTUs were chosen such that they were not predictive
+of cage or donor. Each point is a mouse colored by cage. B) Partial
+dependency plots of the top six predictive OTUs. Line displats the
+partial dependence of log\textsubscript{10} CFU on the relative
+abundance of each predictive OUT. Each median log\textsubscript{10} CFU
+is plotted against its relative abundance for each predictive OTU.
+
+\textbf{Figure 4. Random Forest prediction of CDI severity.} OTUs above
+1\% relative abundance on day 0 were used to predict disease severity.
+OTUs were chosen such that they were not predictive of cage or donor.
+Predictive classification tested via 10-fold (gray), leave-one-cage-out
+(purple dashed) or leave-one-mouse-out (blue dashed) models are
+displayed in (A). B) Partial dependency plots of most predictive OTUs.
+Line displays the partial dependence of log\textsubscript{10} CFU on OTU
+relative abundance. Points are the OTU relative abundance of each mouse
+colored by outcome (red, severe, black, mild).
 

### 2016-10-27 code/HumanCdGF_predict_cfu.Rmd
revise colonization level code

@@
 layout(matrix(c(1:4), ncol=2, byrow=TRUE))
-plot_loop_output <- lapply(seq_along(importance_om$tax), function(i){
-    partialPlot(rf_n_features, top_colonized_df, rownames(importance_om)[i],
-                xlab='', main=paste(importance_om$tax_label[i]))
+for(otu in seq_along(importance_om$tax)){
+    partialPlot(rf_n_features, top_colonized_df, x.var=rownames(importance_om)[otu],
+                xlab='', main=paste(importance_om$tax_label[otu]))
     par(new = T)
     plot(y = top_colonized_df[ , 'log_cfu_med'],
-         x = top_colonized_df[ , rownames(importance_om)[i]], 
+         x = top_colonized_df[ , rownames(importance_om)[otu]], 
          pch = 21, bg = 'gray', xlab = NA, ylab = NA, axes = F)
@@
     points(y = partial_df[partial_df$cage_id %in% mouse, 'log_cfu_med'],
-           x = partial_df[partial_df$cage_id %in% mouse , rownames(importance_om)[i]], 
+           x = partial_df[partial_df$cage_id %in% mouse , rownames(importance_om)[otu]], 
            pch = 21, cex = 2, bg = color[m], xlab = NA, ylab = NA, axes = F)
   })
-})
-rm(plot_loop_output)
+}
 ```
@@
 meta_data %>% 
-  filter(!is.na(day), !cage_id == 'NP1', 
+  filter(day %in% c(1,2,3,4), !cage_id == 'NP1', 
          Early_Euth == FALSE, cdiff_strain == 431) %>% 
@@
   select(mouse_id) %>% 
-  unique()
+  unique() %>% 
+  filter(!mouse_id == 'NP2_983') # remove NP2_983 no colonization data
+
 mouse_list <- as.character(mouse_list$mouse_id)
@@
     filter(!is.na(log_cfu))
-  if(1 %in% test$day){
-    colonization_ratio <- test[test$day==1, 'log_cfu']/median(head(sort(test$log_cfu, decreasing = T),3))
-  } else {
-    colonization_ratio <- median(head(test[order(test$day), "log_cfu"],2))/
-                            median(head(sort(test$log_cfu, decreasing = T),3))
-  }
+  test$rel_cr <- test$log_cfu/median(test$log_cfu[test$day %in% c(4:10)])
+  colonization_ratio <- sum(test$rel_cr[test$day %in% c(1:4)])/sum(test$day %in% c(1:4))
   col_rate <- rbind(col_rate, c(mouse_id = i,  ratio = colonization_ratio))
 }
-
 col_rate <- data.frame(mouse_id = col_rate[ , 1], ratio = as.numeric(col_rate[ , 2]))
@@
 shared_col_rate.df <- col_rate %>% 
-    right_join(community_d0.df)
+    inner_join(community_d0.df)
 shared_colonized.df <- select(shared_col_rate.df, -mouse_id, -cage_id)
@@
                             importance = importance_final,
-                            tax = get_tax(1, names(importance_final), tax_file))
-tax_label <- paste0(importance.df$tax, ' (', 
-                    gsub('tu0*', 'TU ', importance.df$otu), 
-                    ') : ', 
-                    round(importance.df$importance, 3))
+                            get_tax(1, names(importance_final), tax_file))
 
@@
     partialPlot(rf_col_rate_features, top_col_rate_df, names(importance_final)[i],
-                xlab='', main=paste(tax_label[i]))
+                xlab='', main=paste(importance.df$tax_label[i]), log = 'x')
     par(new = T)

### 2016-10-27 code/HumanCdGF_microbiota_predict_by_day.Rmd

@@
         theme_bw() + ggtitle(paste("Random Forest Prediction of C Diff CFU on Day ",
-                                   i, sep = '')) +
+                                   i, sep = '')) + xlim(0,9) + ylim(0,9) + 
         annotate('text', x=0.15*(max(rf_prediction$observed)), 

### 2016-11-02 code/HumanCdGF_predict_cfu.Rmd
change col_rate from for loop to apply function

@@
   filter(Early_Euth == FALSE, cdiff_strain == 431, !cage_id == 'NP1') %>% 
-  select(mouse_id, sample_id, day, log_cfu) %>% 
+  select(mouse_id, cage_id, sample_id, day, log_cfu) %>% 
   filter(!is.na(log_cfu))
@@
 
-col_rate <- c()
-for(i in mouse_list){
+col_rate <- sapply(mouse_list, function(i){
   test <- meta_data %>% 
@@
   test$rel_cr <- test$log_cfu/median(test$log_cfu[test$day %in% c(4:10)])
-  colonization_ratio <- sum(test$rel_cr[test$day %in% c(1:4)])/sum(test$day %in% c(1:4))
-  col_rate <- rbind(col_rate, c(mouse_id = i,  ratio = colonization_ratio))
-}
-col_rate <- data.frame(mouse_id = col_rate[ , 1], ratio = as.numeric(col_rate[ , 2]))
+  colonization_ratio <- sum(test$rel_cr[test$day %in% c(1:4)])/sum(test$day %in% c(1:4))})
+col_rate <- data.frame(mouse_id = names(col_rate), ratio = col_rate, 
+                       stringsAsFactors = F, row.names = NULL)
 
-col_rate <- meta_data %>% 
+col_rate_apply <- meta_data %>% 
   select(mouse_id, cage_id) %>% 
@@
            y=0.95*(max(rf_cr_prediction$predicted)), parse = T,
-           label=paste("R^2 == ", round(tail(rf_col_rate_features$rsq, 1), 2))) + 
-  theme(legend.position = 'none')
+           label=paste("R^2 == ", round(tail(rf_col_rate_features$rsq, 1), 2))) # + 
+#  theme(legend.position = 'none')
 
@@
     partialPlot(rf_col_rate_features, top_col_rate_df, names(importance_final)[i],
-                xlab='', main=paste(importance.df$tax_label[i]), log = 'x')
+                xlab='', main=paste(importance.df$tax_label[i]))#, log = 'x')
     par(new = T)

### 2016-11-03 code/tax_level.R
convert to character to ensure not selecting numeric

@@ -2,10 +2,10 @@
 # level - 1 (genus), 2 (family), 3 (order), 4 (class), 5 (phylum), 6 (kingdom)
 ## - convert taxonomy text list to df, remove percentages
 ## - subset df by desired tax level, then replace any unclassified with next level up
-get_tax <- function(tax_level=1,row_list=1:length(rownames(tax_file)),df=tax_file){
+get_tax <- function(tax_level=1,row_list=rownames(tax_file),df=tax_file){
      if (tax_level %in% c(1:5)){
           row_list <- as.character(row_list)
-          taxonomy <- df[row_list,]
+          taxonomy <- df[rownames(df) %in% row_list,]
           taxonomy <- strsplit(as.character(taxonomy$Taxonomy),';',fixed=TRUE)
           n <- max(sapply(taxonomy, length))
           taxonomy <- lapply(taxonomy, `[`, seq_len(n))

### 2016-11-03 code/HumanCdGF_predict_cfu.Rmd
remove correlative otus

@@
 ```{r setup environment}
-pack_used <- c('randomForest','miscTools','tidyr','ggplot2','dplyr', 'pROC', 'knitr', 'cowplot')
+pack_used <- c('randomForest','miscTools','tidyr','ggplot2','dplyr', 'pROC', 'knitr', 'cowplot', 'caret')
 for (dep in pack_used){
@@
 # read in files
-meta_data   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
-shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
-tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_all.an.cons.taxonomy'
+meta_data   <- 'data/process/human_CdGF_metadata.txt'
+shared_file <- 'data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
+tax_file <- 'data/process/gf_new.an.taxonomy'
 tax_function <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R'
@@
 
+# set parameters
+seed <- 1
+n_trees <- 1001
+n_otus <- 12
+iters <- 100
+cor_cutoff <- 0.75
+
 # remove samples not used or with other cdiff strains
@@
   filter(Early_Euth == FALSE, cdiff_strain == 431, !cage_id == 'NP1') %>% 
-  select(mouse_id, cage_id, sample_id, day, log_cfu) %>% 
+  select(mouse_id, sample_id, day, log_cfu) %>% 
   filter(!is.na(log_cfu))
@@
 rel_abund_shared_01 <- rel_abund_shared[ , apply(rel_abund_shared, 2, max) > 1]
-rel_abund_shared_01$sample_id <- rownames(rel_abund_shared_01)
-
-seed <- 1
-n_trees <- 1001
-n_otus <- 12
-iters <- 100
+# remove correlative otus
+rel_abund_cor <- cor(rel_abund_shared_01, method = 'spearman')
+rownames(rel_abund_cor) <- colnames(rel_abund_shared_01) # add row names to identify otus correlated with features
+cor_otus <- findCorrelation(rel_abund_cor, cutoff = cor_cutoff)
+rel_abund_01_filtr <- rel_ab_otu_df[,-cor_otus]
+rel_abund_01_filtr$sample_id <- rownames(rel_abund_01_filtr)
 ```
@@
     select(-day)
-community_d0.df <- rel_abund_shared_01 %>% 
+community_d0.df <- rel_abund_01_filtr %>% 
     right_join(sample_list_d0) %>% 
@@
 importance_om <- get_tax(1, med_otu_importance_om$otu, tax_file)
+features_cor <- sapply(as.character(med_otu_importance_om$otu), 
+                       function(i)rownames(rel_abund_cor)[rel_abund_cor[,i] > cor_cutoff])
+importance_om <- get_tax(row_list = unlist(features_cor))
 
-partial_df <- data.frame(colonized_cfu %>% 
-    right_join(community_d0.df) %>% 
-    select(mouse_id, cage_id, log_cfu_med, one_of(rownames(importance_om))) %>% 
+import_rel_ab <- rel_abund_shared_01[ , rownames(importance_om)]
+import_rel_ab$sample_id <- rownames(import_rel_ab)
+partial_df <- data.frame(import_rel_ab %>%
+    right_join(sample_list_d0) %>% 
+    select(-sample_id) %>%
+    left_join(colonized_cfu) %>% 
     filter(!is.na(log_cfu_med)))
@@
 color <- sapply(1:length(unique(partial_df$cage_id)), 
-                           function(n)rgb(sample(c(0:255), 1),sample(c(0:255), 1),
-                                          sample(c(0:255), 1), maxColorValue = 255))
+                           function(n){
+                             set.seed(n*n)
+                             rgb(sample(c(0:255), 1),sample(c(0:255), 1),
+                                 sample(c(0:255), 1), maxColorValue = 255)
+                             })
 
@@
     mtext(side = 4, line = 3, 'Log10 CFU')
-# To color points by mouse - Change pch to 21 and un-comment following lines
   lapply(seq_along(unique(partial_df$cage_id)), function(m){

### 2016-11-03 code/HumanCdGF_predict_cfu.Rmd
remove correlated otus from partial plots
no partial plot data for removed otus

@@
                        function(i)rownames(rel_abund_cor)[rel_abund_cor[,i] > cor_cutoff])
-importance_om <- get_tax(row_list = unlist(features_cor))
+get_tax(row_list = unlist(features_cor))
 
-import_rel_ab <- rel_abund_shared_01[ , rownames(importance_om)]
-import_rel_ab$sample_id <- rownames(import_rel_ab)
-partial_df <- data.frame(import_rel_ab %>%
-    right_join(sample_list_d0) %>% 
-    select(-sample_id) %>%
-    left_join(colonized_cfu) %>% 
+partial_df <- data.frame(colonized_cfu %>% 
+    right_join(community_d0.df) %>% 
+    select(mouse_id, cage_id, log_cfu_med, one_of(rownames(importance_om))) %>% 
     filter(!is.na(log_cfu_med)))
@@
     mtext(side = 4, line = 3, 'Log10 CFU')
+# To color points by mouse - Change pch to 21 and un-comment following lines
   lapply(seq_along(unique(partial_df$cage_id)), function(m){

### 2016-11-03 code/HumanCdGF_predict_cfu.Rmd
add correlated otu subtitle to partialplot

@@
                            function(n){
-                             set.seed(n*n)
+                             set.seed(n+10) # add modify to generate distinct colors
                              rgb(sample(c(0:255), 1),sample(c(0:255), 1),
@@
     mtext(side = 4, line = 3, 'Log10 CFU')
+    if(length(features_cor[[otu]]) > 1){
+    mtext(side = 1, line = 3, cex = 0.6,
+          paste('Bacteria removed in preprocessing due to correlation to this OTU:\n',
+                paste(get_tax(row_list = features_cor[[otu]])$tax_label, collapse = ',')))
+    }
 # To color points by mouse - Change pch to 21 and un-comment following lines

### 2016-11-04 code/HumanCdGF_predict_cfu.Rmd
fix corr df object call

@@
 # read in files
+setwd('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/')
 meta_data   <- 'data/process/human_CdGF_metadata.txt'
@@
 cor_otus <- findCorrelation(rel_abund_cor, cutoff = cor_cutoff)
-rel_abund_01_filtr <- rel_ab_otu_df[,-cor_otus]
+rel_abund_01_filtr <- rel_abund_shared_01[,-cor_otus]
 rel_abund_01_filtr$sample_id <- rownames(rel_abund_01_filtr)

### 2016-11-04 code/HumanCdGF_predict_cfu.Rmd
changed feature selection to caret:rfe
has some errors with partial plots now

@@
 ###### select top n otus reiteratively and select otus that appear > 50% of iterations
-rf_n_otus <-lapply(1:iters, function(i){
-  set.seed(i)
-  rf <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df,
-                     importance = TRUE, ntree = n_trees)
-  names(sort(importance(rf)[,1], decreasing = T))[1:n_otus]
-})
-n_otu_freq <- table(unlist(rf_n_otus))
-top_important_OTU <- names(n_otu_freq)[n_otu_freq >= iters/2]
+rfRFE <-  list(summary = defaultSummary,
+               fit = function(x, y, first, last, ...){
+                 library(randomForest)
+                 randomForest(x, y, importance = first, keep.forest = T, ntree = n_trees, ...)
+                 },
+               pred = function(object, x)  predict(object, x),
+               rank = function(object, x, y) {
+                 vimp <- varImp(object)
+                 vimp <- vimp[order(vimp$Overall,decreasing = TRUE),,drop = FALSE]
+                 vimp$var <- rownames(vimp)                  
+                 vimp
+                 },
+               selectSize = pickSizeTolerance,
+               selectVar = pickVars)
 
-top_colonized_df <- data.frame(cfu_shared_colonized.df %>% 
-  select(log_cfu_med, one_of(top_important_OTU)))
+rfe_ctrl <- rfeControl(functions = rfRFE,
+                   method = "repeatedcv",
+                   repeats = 10,
+                   verbose = FALSE)
+subsets <- c(1:5, c(seq(10, 70, 10)))
+set.seed(seed)
+rf_model <- rfe(cfu_shared_colonized.df[ , -1], cfu_shared_colonized.df$log_cfu_med,
+                 sizes = subsets,
+                 rfeControl = rfe_ctrl)
+rf_test <- train(cfu_shared_colonized.df[ , importance_final], cfu_shared_colonized.df$log_cfu_med,
+                 method = 'rf', importance = T)
+partialPlot(rf_test$finalModel, cfu_shared_colonized.df, 'Otu000004')
 
-set.seed(seed)
-rf_n_features <- randomForest(log_cfu_med ~ ., data=top_colonized_df,
-                              importance=TRUE, ntree=n_trees, keep.forest=TRUE)
+xrf_test <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df[ , 1:30],
+                     importance = TRUE, ntree = 100)
 
-importance_final <- sort(importance(rf_n_features)[,1], decreasing = T)
+importance_final <- predictors(rf_model)
 
+rf_n_features <- rf_model$fit
 rf_prediction <- data.frame(observed = cfu_shared_colonized.df$log_cfu_med, 
@@
                        function(i)rownames(rel_abund_cor)[rel_abund_cor[,i] > cor_cutoff])
-get_tax(row_list = unlist(features_cor))
 
@@
                              })
-
+partialPlot(rf_n_features, top_colonized_df, x.var=rownames(importance_om)[otu])
 layout(matrix(c(1:4), ncol=2, byrow=TRUE))

### 2016-11-07 code/tax_level.R
fix input order error
previously when inputting out of order, output OTU in same order but tax in numerical order, mismatching otu and tax

@@ -4,11 +4,11 @@
 ## - subset df by desired tax level, then replace any unclassified with next level up
 get_tax <- function(tax_level=1,row_list=rownames(tax_file),df=tax_file){
      if (tax_level %in% c(1:5)){
-          row_list <- as.character(row_list)
-          taxonomy <- df[rownames(df) %in% row_list,]
-          taxonomy <- strsplit(as.character(taxonomy$Taxonomy),';',fixed=TRUE)
-          n <- max(sapply(taxonomy, length))
-          taxonomy <- lapply(taxonomy, `[`, seq_len(n))
+          otu_list <- as.character(row_list)
+          otu_taxonomy <- df[rownames(df) %in% otu_list,]
+          otu_taxonomy <- otu_taxonomy[as.numeric(factor(otu_list)), ]
+          taxonomy <- strsplit(as.character(otu_taxonomy$Taxonomy),';',fixed=TRUE)
+          taxonomy <- lapply(taxonomy, `[`, seq_len(6))
           taxonomy <- data.frame(do.call('rbind', taxonomy))
           taxonomy <- sapply(taxonomy,gsub,pattern="\\(\\d*\\)",replacement="")
           level <- 7-tax_level
@@ -20,8 +20,8 @@ get_tax <- function(tax_level=1,row_list=rownames(tax_file),df=tax_file){
           }
           tax_out <- gsub('_unclassified', '', tax_out)
           label_out <- paste0(tax_out, ' (', 
-                              gsub('tu0*', 'TU ', row_list),')')
-          return(data.frame(tax= tax_out, tax_label = label_out, row.names=row_list,
+                              gsub('tu0*', 'TU ', rownames(otu_taxonomy)),')')
+          return(data.frame(tax= tax_out, tax_label = label_out, row.names=rownames(otu_taxonomy),
                             stringsAsFactors = FALSE))
      } else {print(
           'Error: Level must be 1 (genus), 2 (family), 3 (order), 4 (class), 5 (phylum)')

### 2016-11-07 code/HumanCdGF_predict_cfu.Rmd
fix partial plots to work with rfe output

@@
 ```{r setup environment}
-pack_used <- c('randomForest','miscTools','tidyr','ggplot2','dplyr', 'pROC', 'knitr', 'cowplot', 'caret')
+pack_used <- c('randomForest','tidyr','ggplot2','plyr', 'dplyr', 'caret')
 for (dep in pack_used){
@@
                  rfeControl = rfe_ctrl)
-rf_test <- train(cfu_shared_colonized.df[ , importance_final], cfu_shared_colonized.df$log_cfu_med,
-                 method = 'rf', importance = T)
-partialPlot(rf_test$finalModel, cfu_shared_colonized.df, 'Otu000004')
-
-xrf_test <- randomForest(log_cfu_med ~ ., data = cfu_shared_colonized.df[ , 1:30],
-                     importance = TRUE, ntree = 100)
-
-importance_final <- predictors(rf_model)
+top_important_OTU <- predictors(rf_model)
 
@@
 om_otu_rank <- apply(om_import_otus, 2, function(x) rank(x, na.last = F))
-om_raw_rank <- apply(om_otu_rank, 1, function(x) sum(x, na.rm = T))
+om_raw_rank <- sort(apply(om_otu_rank, 1, function(x) sum(x, na.rm = T)), decreasing = T)
 med_otu_importance_om <- data.frame(otu = names(om_raw_rank),
@@
                              })
-partialPlot(rf_n_features, top_colonized_df, x.var=rownames(importance_om)[otu])
-layout(matrix(c(1:4), ncol=2, byrow=TRUE))
+
+layout(matrix(c(1:16), ncol=4, byrow=TRUE))
 for(otu in seq_along(importance_om$tax)){
-    partialPlot(rf_n_features, top_colonized_df, x.var=rownames(importance_om)[otu],
+    partialPlot(rf_n_features, data.frame(cfu_shared_colonized.df), x.var=rownames(importance_om)[otu],
                 xlab='', main=paste(importance_om$tax_label[otu]))
     par(new = T)
-    plot(y = top_colonized_df[ , 'log_cfu_med'],
-         x = top_colonized_df[ , rownames(importance_om)[otu]], 
-         pch = 21, bg = 'gray', xlab = NA, ylab = NA, axes = F)
+    plot(y = partial_df[ , 'log_cfu_med'],
+         x = partial_df[ , rownames(importance_om)[otu]], 
+         pch = NA, bg = 'gray', xlab = NA, ylab = NA, axes = F)
     axis(side = 4)

### 2016-11-07 code/HumanCdGF_predict_cfu.Rmd
add caret::rfe to col rate

@@
 set.seed(seed)
-rf_model <- rfe(cfu_shared_colonized.df[ , -1], cfu_shared_colonized.df$log_cfu_med,
+rf_cfu_model <- rfe(cfu_shared_colonized.df[ , -1], cfu_shared_colonized.df$log_cfu_med,
                  sizes = subsets,
                  rfeControl = rfe_ctrl)
-top_important_OTU <- predictors(rf_model)
+top_important_OTU <- predictors(rf_cfu_model)
 
-rf_n_features <- rf_model$fit
+rf_n_features <- rf_cfu_model$fit
 rf_prediction <- data.frame(observed = cfu_shared_colonized.df$log_cfu_med, 
@@
 
-layout(matrix(c(1:16), ncol=4, byrow=TRUE))
+layout(matrix(c(1:9), ncol=3, byrow=TRUE))
 for(otu in seq_along(importance_om$tax)){
@@
 ```{r predict colonization rate, fig.width=10,fig.height=10}
-
-meta_data %>% 
-  filter(day %in% c(1,2,3,4), !cage_id == 'NP1', 
-         Early_Euth == FALSE, cdiff_strain == 431) %>% 
-  ggplot(aes(x = factor(day), y = log_cfu, color = cage_id)) + 
-    geom_line(aes(group = mouse_id), size=.25, alpha=0.4) + geom_point() +
-    stat_summary(fun.y=mean, geom="line", aes(group = cage_id)) + 
-    theme_bw() + labs(x = 'Day', y = 'C Diff CFU (log10)') +
-    theme(legend.justification=c(1,0), legend.position=c(1,0), 
-          legend.title=element_blank())
+# plot initial C. difficile colonization of mice 
+#meta_data %>% 
+#  filter(day %in% c(1,2,3,4), !cage_id == 'NP1', 
+#         Early_Euth == FALSE, cdiff_strain == 431) %>% 
+#  ggplot(aes(x = factor(day), y = log_cfu, color = cage_id)) + 
+#    geom_line(aes(group = mouse_id), size=.25, alpha=0.4) + geom_point() +
+#    stat_summary(fun.y=mean, geom="line", aes(group = cage_id)) + 
+#    theme_bw() + labs(x = 'Day', y = 'C Diff CFU (log10)') +
+#   theme(legend.justification=c(1,0), legend.position=c(1,0), 
+#          legend.title=element_blank())
 
@@
   colonization_ratio <- sum(test$rel_cr[test$day %in% c(1:4)])/sum(test$day %in% c(1:4))})
-col_rate <- data.frame(mouse_id = names(col_rate), ratio = col_rate, 
-                       stringsAsFactors = F, row.names = NULL)
-
-col_rate_apply <- meta_data %>% 
+col_rate <- meta_data %>% 
   select(mouse_id, cage_id) %>% 
   unique() %>% 
-  right_join(col_rate)
+  right_join(data.frame(mouse_id = names(col_rate), ratio = col_rate, 
+                       stringsAsFactors = F, row.names = NULL))
 
-shared_col_rate.df <- col_rate %>% 
+shared_col_rate_df <- col_rate %>% 
     inner_join(community_d0.df)
-shared_colonized.df <- select(shared_col_rate.df, -mouse_id, -cage_id)
+shared_cr_df <- select(shared_col_rate_df, -mouse_id, -cage_id)
 
 set.seed(seed)
-rf_col_rate <- randomForest(ratio ~ ., data = shared_colonized.df,
-                           importance = TRUE, ntree = n_trees, mtry = 14)
+rf_cr_model <- rfe(shared_cr_df[ , -1], shared_cr_df$ratio,
+                 sizes = subsets,
+                 rfeControl = rfe_ctrl)
+cr_features <- predictors(rf_cr_model)
+rf_col_rate <- rf_cr_model$fit
 
-importance_sorted <- sort(importance(rf_col_rate)[,1], decreasing = T)
-top_important_OTU <- rownames(data.frame(importance_sorted[importance_sorted/max(importance_sorted) > 0.6]))
+cr_prediction <- data.frame(observed = shared_cr_df$ratio, 
+                       predicted = rf_col_rate$predicted)
+cr_prediction <- merge(cr_prediction, col_rate, by.x = 'observed', by.y = 'ratio')
 
-top_col_rate_df <- data.frame(shared_colonized.df %>% 
-  select(ratio, one_of(top_important_OTU)))
+#top_col_rate_df <- data.frame(shared_colonized.df %>% 
+#  select(ratio, one_of(cr_features)) %>% 
+#  left_join(col_rate))
 
-set.seed(seed)
-rf_col_rate_features <- randomForest(ratio ~ ., data=top_col_rate_df,
-                              importance=TRUE, ntree=n_trees, mtry = 2,
-                              keep.forest=TRUE)
+#set.seed(seed)
+#rf_col_rate_features <- randomForest(ratio ~ ., data=top_col_rate_df,
+#                              importance=TRUE, ntree=n_trees, mtry = 2,
+#                              keep.forest=TRUE)
+#importance_final <- sort(importance(rf_col_rate_features)[,1], decreasing = T)
+#rf_col_rate_prediction <- data.frame(observed = shared_colonized.df$ratio, 
+#                       predicted = rf_col_rate_features$predicted)
+#rf_cr_prediction <- cbind(cage_id = shared_col_rate.df[,'cage_id'], rf_col_rate_prediction)
 
-importance_final <- sort(importance(rf_col_rate_features)[,1], decreasing = T)
-
-rf_col_rate_prediction <- data.frame(observed = shared_colonized.df$ratio, 
-                       predicted = rf_col_rate_features$predicted)
-rf_cr_prediction <- cbind(cage_id = shared_col_rate.df[,'cage_id'], rf_col_rate_prediction)
-
-ggplot(data = rf_cr_prediction, aes(x=observed, y=predicted, color = cage_id)) + geom_point() + 
+ggplot(data = cr_prediction, aes(x=observed, y=predicted, color = cage_id)) + geom_point() + 
   labs(x= 'Observed', y = 'Predicted',
@@
   theme_bw() + xlim(0,1) + ylim(0,1) + 
-  annotate('text', x=0.15*(max(rf_cr_prediction$observed)), 
-           y=0.95*(max(rf_cr_prediction$predicted)), parse = T,
-           label=paste("R^2 == ", round(tail(rf_col_rate_features$rsq, 1), 2))) # + 
+  annotate('text', x=0.15*(max(cr_prediction$observed)), 
+           y=0.95*(max(cr_prediction$predicted)), parse = T,
+           label=paste("R^2 == ", round(tail(rf_col_rate$rsq, 1), 2))) # + 
 #  theme(legend.position = 'none')
 
-importance.df <- data.frame(otu = names(importance_final),
-                            importance = importance_final,
-                            get_tax(1, names(importance_final), tax_file))
+#importance.df <- data.frame(otu = names(importance_final),
+#                            importance = importance_final,
+#                            get_tax(1, names(importance_final), tax_file))
+
+cr_features_cor <- sapply(as.character(cr_features), 
+                       function(i)rownames(rel_abund_cor)[rel_abund_cor[,i] > cor_cutoff])
 
@@
 
-cage_color <- rainbow(length(unique(rf_cr_prediction$cage_id)))
-for (i in seq_along(importance_final)) {
-    partialPlot(rf_col_rate_features, top_col_rate_df, names(importance_final)[i],
-                xlab='', main=paste(importance.df$tax_label[i]))#, log = 'x')
+cage_color <- rainbow(length(unique(cr_prediction$cage_id)))
+for (otu in seq_along(cr_features)) {
+    partialPlot(rf_col_rate, shared_cr_df, cr_features[otu],
+                xlab='', main=paste(get_tax(row_list = cr_features)$tax_label[otu]))#, log = 'x')
     par(new = T)
-    plot(y = top_col_rate_df[ , 'ratio'],
-         x = top_col_rate_df[ , names(importance_final)[i]], 
-         pch = 21, bg = 'gray', xlab = NA, ylab = NA, axes = F)
+    plot(y = shared_col_rate_df[ , 'ratio'],
+         x = shared_col_rate_df[ , cr_features[otu]], 
+         pch = NA, bg = 'gray', xlab = NA, ylab = NA, axes = F)
     axis(side = 4)
     mtext(side = 4, line = 3, 'Relative Colonization')
-# To color points by cage - Change pch to 21 and un-comment following lines
-#  for(m in seq_along(unique(rf_cr_prediction$cage_id))){
-#    cage <- unique(rf_cr_prediction$cage_id)[m]
-#    points(y = top_col_rate_df[rf_cr_prediction$cage_id %in% cage, 'ratio'],
-#         x = top_col_rate_df[rf_cr_prediction$cage_id %in% cage , names(importance_final)[i]], 
-#         pch = 21, bg = cage_color[m])
-#  }
+    if(length(cr_features_cor[[otu]]) > 1){
+      mtext(side = 1, line = 3, cex = 0.6,
+            paste('Bacteria removed in preprocessing due to correlation to this OTU:\n',
+                  paste(get_tax(row_list = cr_features_cor[[otu]])$tax_label, collapse = ',')))
+    }
+# To color points by mouse - Change pch to 21 and un-comment following lines
+  lapply(seq_along(unique(shared_col_rate_df$cage_id)), function(m){
+    mouse <- unique(shared_col_rate_df$cage_id)[m]
+    points(y = shared_col_rate_df[shared_col_rate_df$cage_id %in% mouse, 'ratio'],
+           x = shared_col_rate_df[shared_col_rate_df$cage_id %in% mouse , cr_features[otu]], 
+           pch = 21, cex = 2, bg = color[m], xlab = NA, ylab = NA, axes = F)
+  })
 }

### 2016-11-07 code/HumanCdGF_predict_cfu.Rmd
remove old col rate code

@@
 ```{r predict colonization rate, fig.width=10,fig.height=10}
-# plot initial C. difficile colonization of mice 
-#meta_data %>% 
-#  filter(day %in% c(1,2,3,4), !cage_id == 'NP1', 
-#         Early_Euth == FALSE, cdiff_strain == 431) %>% 
-#  ggplot(aes(x = factor(day), y = log_cfu, color = cage_id)) + 
-#    geom_line(aes(group = mouse_id), size=.25, alpha=0.4) + geom_point() +
-#    stat_summary(fun.y=mean, geom="line", aes(group = cage_id)) + 
-#    theme_bw() + labs(x = 'Day', y = 'C Diff CFU (log10)') +
-#   theme(legend.justification=c(1,0), legend.position=c(1,0), 
-#          legend.title=element_blank())
 
@@
 
-#top_col_rate_df <- data.frame(shared_colonized.df %>% 
-#  select(ratio, one_of(cr_features)) %>% 
-#  left_join(col_rate))
-
-#set.seed(seed)
-#rf_col_rate_features <- randomForest(ratio ~ ., data=top_col_rate_df,
-#                              importance=TRUE, ntree=n_trees, mtry = 2,
-#                              keep.forest=TRUE)
-#importance_final <- sort(importance(rf_col_rate_features)[,1], decreasing = T)
-#rf_col_rate_prediction <- data.frame(observed = shared_colonized.df$ratio, 
-#                       predicted = rf_col_rate_features$predicted)
-#rf_cr_prediction <- cbind(cage_id = shared_col_rate.df[,'cage_id'], rf_col_rate_prediction)
-
 ggplot(data = cr_prediction, aes(x=observed, y=predicted, color = cage_id)) + geom_point() + 
@@
 #  theme(legend.position = 'none')
-
-#importance.df <- data.frame(otu = names(importance_final),
-#                            importance = importance_final,
-#                            get_tax(1, names(importance_final), tax_file))
 

### 2016-11-07 code/HumanCdGF_predict_cfu.Rmd
add LOcO and OM models to col rate

@@
 }
-test <- data.frame(om_raw_otus[ , -1], row.names = om_raw_otus$otu)
-colnames(test) <- seq(ncol(test))
-### rank
-test_rank <- apply(test, 2, function(x) rank(x, na.last = F))
-test_raw_rank <- apply(test_rank, 1, function(x) sum(x, na.rm = T))
-test_import_rank <- sort(test_raw_rank[test_raw_rank >=  ncol(test_rank)*sum(test_rank[ , 1])/nrow(test_rank)], decreasing = T)
-### mean
-#test_mean <- apply(test, 2, function(x) mean(x, na.rm=T)) # mean importance for each rf model
-#test_select <- apply(test, 1, function(x) x >= test_mean) # T for each otu above mean by sample
-#test_raw_mean <- apply(test_select, 2, function(x) sum(x[!is.na(x)])/length(x[!is.na(x)])) # portion of rf model otu is above mean imp
-#test_import_mean <- sort(test_raw_mean[test_raw_mean > 0.5], decreasing = T)
-#### median
-#test_med <- apply(test, 2, function(x) median(x, na.rm=T)) # median importance for each rf model
-#test_select <- apply(test, 1, function(x) x >= test_med) # T for each otu above median by sample
-#test_raw_med <- apply(test_select, 2, function(x) sum(x[!is.na(x)])/length(x[!is.na(x)])) # portion of rf model otu is above med imp
-#test_import_med <- sort(test_raw_med[test_raw_med > 0.5], decreasing = T)
+raw_importance <- data.frame(om_raw_otus[ , -1], row.names = om_raw_otus$otu)
+### rank importance
+importance_rank <- apply(raw_importance, 2, function(x) rank(x, na.last = F))
+raw_rank <- apply(importance_rank, 1, function(x) sum(x, na.rm = T))
+final_rank <- sort(raw_rank[raw_rank >=  ncol(importance_rank)*sum(importance_rank[ , 1])/nrow(importance_rank)], decreasing = T)
 
-#####
-#test <- c() # test if any model has same train set
-#  for(i in 2:21){
-#  test[i-1] <- colnames(om_import_otus)[i] %in% colnames(om_import_otus)[-c(1,i)]
-#}
-#any(test) # if true, then trained on the same set of mice multiple times
-#####
 one_mouse_df <- data.frame(colonized_cfu %>% 
@@
     filter(!is.na(log_cfu_med))  %>% 
-    select(log_cfu_med, mouse_id, cage_id, one_of(names(test_import_rank))))
+    select(log_cfu_med, mouse_id, cage_id, one_of(names(final_rank))))
 
@@
 om_pred <- c()
+om_pp <- c()
 for(i in 1:iters){
@@
                              importance = TRUE, keep.forest = T, ntree = n_trees)
+    for(i in 4:ncol(train)){
+      pp_var <- data.frame(otu = colnames(train)[i],
+                           partialPlot(rf_om, data.frame(select(train, -mouse_id, -cage_id)), 
+                                       x.var= colnames(train)[i], plot = FALSE))
+      om_pp <- rbind(om_pp, pp_var)
+      }
     om_pred <- rbind(om_pred,
@@
   theme(legend.justification=c(0,1), legend.position=c(0,1))
+
+OTUS <- unique(as.character(om_pp$otu))
+part_plot_cfu <- c()
+for(i in OTUS){
+om_v_oob <- rbind(data.frame(model = 'OM',
+                             om_pp[om_pp$otu == i, ]),
+                  data.frame(model = 'OOB', otu = i,
+                             partialPlot(rf_n_features, data.frame(cfu_shared_colonized.df), 
+                                         OTUS[OTUS == i], plot = F)))
+part_plot_cfu[[i]] <- list(ggplot(om_v_oob, aes(x,y, group = model, color = model)) +geom_smooth() + geom_point(alpha = 0.5) + ggtitle(i))
+}
+part_plot_cfu[1]
+part_plot_cfu[2]
+part_plot_cfu[3]
+
+# partial plot for one_mouse model
+#ggplot(data=om_pp[om_pp$otu==names(final_rank[5]), ], aes(x=x, y=y)) + geom_smooth()
+#  stat_summary(fun.y=median, geom="line")
 
@@
 
-ggplot(data = cr_prediction, aes(x=observed, y=predicted, color = cage_id)) + geom_point() + 
-  labs(x= 'Observed', y = 'Predicted',
-       title = "Random Forest Prediction of Relative Day 1 Colonization Level of C. Difficile from Day 0 Community") + 
-  theme_bw() + xlim(0,1) + ylim(0,1) + 
-  annotate('text', x=0.15*(max(cr_prediction$observed)), 
-           y=0.95*(max(cr_prediction$predicted)), parse = T,
-           label=paste("R^2 == ", round(tail(rf_col_rate$rsq, 1), 2))) # + 
-#  theme(legend.position = 'none')
+##### - leave one cage out cv
+LOcO_cr_df <- data.frame(shared_col_rate_df %>% 
+    select(ratio, mouse_id, cage_id, one_of(cr_features)))
+cage_list <- unique(LOcO_cr_df$cage_id)
+
+LOcO_cr_probs <- c()
+rsq_cr <- c()
+import_otus_cr <- data.frame(otu = cr_features)
+for(i in 1:length(cage_list)){
+  by_cage <- LOcO_cr_df$cage_id == cage_list[i]
+  LOcO_predict_cr <- select(LOcO_cr_df, -cage_id, -mouse_id)
+  train <- LOcO_predict_cr[!by_cage,]
+  test <- LOcO_predict_cr[by_cage,]
+  set.seed(seed)
+  temp_model <- randomForest(ratio~., data=train, pdel=0.99, ntree=500, importance=T)
+  import_df <- importance(temp_model)
+  colnames(import_df)[1] <- cage_list[i]
+  import_otus_cr <- merge(import_otus_cr, import_df, by.x = 'otu', by.y = 'row.names')
+  import_otus_cr <- select(import_otus_cr, -IncNodePurity)
+  rsq_cr[i] <- round(tail(temp_model$rsq, 1), 2)
+  LOcO_cr_probs <- rbind(LOcO_cr_probs, data.frame(observed=test$ratio, 
+                                             predicted=predict(temp_model, test),
+                                             mouse_id=LOcO_cr_df$mouse_id[by_cage],
+                                             cage_id=rep(cage_list[i], nrow(test))))
+}
+med_otu_importance_cr <- data.frame(otu = import_otus_cr$otu,
+                                 importance = apply(import_otus_cr[, -1], 1, median))
+med_otu_importance_cr <- med_otu_importance_cr[order(med_otu_importance_cr$importance, decreasing = T),]
+
+LOcO_rsq_cr <- 1 - (sum((LOcO_cr_probs$observed - LOcO_cr_probs$predicted)^2)/
+                   sum((LOcO_cr_probs$observed - mean(LOcO_cr_probs$observed))^2))
+
+#One per cage
+one_mouse_cr_df <- LOcO_cr_df
+
+om_raw_otus_cr <- data.frame(otu = colnames(select(one_mouse_cr_df, contains('Otu'))))
+for(i in 1:iters){
+    set.seed(i)
+    train <- one_mouse_cr_df %>% 
+      group_by(cage_id) %>% 
+      sample_n(1) %>% ungroup() 
+    test <- one_mouse_cr_df %>% 
+      filter(!mouse_id %in% train$mouse_id)
+    set.seed(seed)
+    rf_om <- randomForest(ratio ~ ., data = select(train, -mouse_id, -cage_id),
+                             importance = TRUE, keep.forest = T, ntree = n_trees)
+    import_df <- importance(rf_om)
+    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
+    om_raw_otus_cr <- merge(om_raw_otus_cr, import_df, by.x = 'otu', by.y = 'row.names')
+    om_raw_otus_cr <- select(om_raw_otus_cr, -IncNodePurity)
+}
+raw_importance_cr <- data.frame(om_raw_otus_cr[ , -1], row.names = om_raw_otus_cr$otu)
+### rank importance
+importance_rank_cr <- apply(raw_importance_cr, 2, function(x) rank(x, na.last = F))
+raw_rank_cr <- apply(importance_rank_cr, 1, function(x) sum(x, na.rm = T))
+final_rank_cr <- sort(raw_rank_cr[raw_rank_cr >=  ncol(importance_rank_cr)*sum(importance_rank_cr[ , 1])/nrow(importance_rank_cr)], decreasing = T)
+
+om_cr_df <- data.frame(shared_col_rate_df %>% 
+    select(ratio, mouse_id, cage_id, one_of(names(final_rank_cr))))
+
+om_import_otus_cr <- data.frame(otu = colnames(select(om_cr_df, contains('Otu'))))
+om_pred_cr <- c()
+om_pp_cr <- c()
+for(i in 1:iters){
+    set.seed(i)
+    train <- om_cr_df %>% 
+      group_by(cage_id) %>% 
+      sample_n(1) %>% ungroup() 
+    test <- om_cr_df %>% 
+      filter(!mouse_id %in% train$mouse_id)
+    set.seed(seed)
+    rf_om <- randomForest(ratio ~ ., data = select(train, -mouse_id, -cage_id),
+                             importance = TRUE, keep.forest = T, ntree = n_trees)
+    for(i in 4:ncol(train)){
+      pp_var <- data.frame(otu = colnames(train)[i],
+                           partialPlot(rf_om, data.frame(select(train, -mouse_id, -cage_id)), 
+                                       x.var= colnames(train)[i], plot = FALSE))
+      om_pp_cr <- rbind(om_pp_cr, pp_var)
+      }
+    om_pred_cr <- rbind(om_pred_cr,
+                     data.frame(select(test, mouse_id, cage_id, ratio),
+                          predicted=predict(rf_om, test)))
+    import_df <- importance(rf_om)
+    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
+    om_import_otus_cr <- merge(om_import_otus_cr, import_df, by.x = 'otu', by.y = 'row.names')
+    om_import_otus_cr <- select(om_import_otus_cr, -IncNodePurity)
+}
+om_rsq_cr <- 1 - (sum((om_pred_cr$ratio - om_pred_cr$predicted)^2)/
+                  sum((om_pred_cr$ratio - mean(om_pred_cr$ratio))^2))
+
+om_pred_cr <- rename(om_pred_cr, observed = ratio)
+
+om_import_otus_cr <- data.frame(om_import_otus_cr[ , -1], row.names = om_import_otus_cr$otu)
+om_otu_rank_cr <- apply(om_import_otus_cr, 2, function(x) rank(x, na.last = F))
+om_raw_rank_cr <- sort(apply(om_otu_rank_cr, 1, function(x) sum(x, na.rm = T)), decreasing = T)
+med_otu_importance_om <- data.frame(otu = names(om_raw_rank_cr),
+                                 importance = om_raw_rank_cr)
+
+#### merge OOB and leave 1 cage out dataframes
+all_cr_rf <- rbind(cbind(cr_prediction, data = rep('OOB', nrow(cr_prediction))),
+                cbind(LOcO_cr_probs, data = rep('LOO', nrow(LOcO_cr_probs))),
+                cbind(om_pred_cr, data = rep('1_mouse', nrow(om_pred_cr))))
+
+ggplot(data = all_cr_rf, aes(x=observed, y=predicted, color = data)) + 
+  geom_abline(color = 'gray', size = 0.5, linetype = 2, slope = 1) + 
+  # geom_point() + 
+  geom_smooth(method = 'lm', se = F) +
+  stat_summary(fun.y=median, geom="point", aes(group = data)) + 
+  stat_summary(fun.data = 'median_hilow') +
+  xlim(c(min(all_cr_rf[, c('observed','predicted')]),
+         max(all_cr_rf[, c('observed','predicted')]))) + 
+  ylim(c(min(all_cr_rf[, c('observed','predicted')]),
+         max(all_cr_rf[, c('observed','predicted')]))) +
+  labs(x= 'Observed Log10 CFU', y = 'Predicted Log10 CFU',
+       title = "Random Forest Prediction of C. Difficile Colonization Rate from Day 0 Community") + 
+  theme_bw() + 
+  scale_color_discrete(name = 'Model',
+                       breaks = c('OOB', 'LOO', '1_mouse'),
+                       labels = c(bquote('OOB (R'^'2'*' = '*.(round(tail(rf_col_rate$rsq, 1), 2))*')'),
+                                  bquote('LOO (R'^'2'*' = '*.(round(LOcO_rsq_cr, 2))*')'),
+                                  bquote('OM (R'^'2'*' = '*.(round(om_rsq_cr,2))*')'))) + 
+  theme(legend.justification=c(0,1), legend.position=c(0,1))
+
+om_v_oob <- rbind(data.frame(model = 'OM',
+                             om_pp_cr[om_pp_cr$otu == 'Otu000004', ]),
+                  data.frame(model = 'OOB', otu = 'Otu000004',
+                             partialPlot(rf_col_rate, shared_cr_df, 'Otu000004',
+                                         plot = F)))
+ggplot(om_v_oob, aes(x,y, group = model, color = model)) +geom_smooth() + geom_point()
 

### 2016-11-08 code/HumanCdGF_predict_cfu.Rmd
add partial plot for one mouse model

@@
                                          OTUS[OTUS == i], plot = F)))
-part_plot_cfu[[i]] <- list(ggplot(om_v_oob, aes(x,y, group = model, color = model)) +geom_smooth() + geom_point(alpha = 0.5) + ggtitle(i))
+gg_pp <- ggplot(om_v_oob, aes(x,y, group = model, color = model)) +geom_smooth() + ggtitle(i)
+part_plot_cfu[[i]] <- gg_pp
 }
-part_plot_cfu[1]
-part_plot_cfu[2]
-part_plot_cfu[3]
+library(cowplot)
+
+ggdraw() + 
+  draw_plot(part_plot_cfu[[1]])
+plot_grid(part_plot_cfu[[1]], part_plot_cfu[[2]])
+
+do.call('plot_grid',part_plot_cfu)
+i <- 'Otu000004'
+plot(om_pp$x[om_pp$otu == i], om_pp$y[om_pp$otu == i])
+lines(lowess(om_pp$x[om_pp$otu == i], om_pp$y[om_pp$otu == i]))
+library(car)
+scatterplot(y~x, om_pp[om_pp$otu == i, ], spread = F, smoother = gamLine,
+            grid = F, lty = 0)
 

### 2016-11-08 code/HumanCdGF_predict_cfu.Rmd
remove one mouse partial plot

--- 2016-11-07 code/HumanCdGF_predict_cfu.Rmd
@@
 om_pred <- c()
-om_pp <- c()
 for(i in 1:iters){
@@
                              importance = TRUE, keep.forest = T, ntree = n_trees)
-    for(i in 4:ncol(train)){
-      pp_var <- data.frame(otu = colnames(train)[i],
-                           partialPlot(rf_om, data.frame(select(train, -mouse_id, -cage_id)), 
-                                       x.var= colnames(train)[i], plot = FALSE))
-      om_pp <- rbind(om_pp, pp_var)
-      }
     om_pred <- rbind(om_pred,
@@
   theme(legend.justification=c(0,1), legend.position=c(0,1))
-
-OTUS <- unique(as.character(om_pp$otu))
-part_plot_cfu <- c()
-for(i in OTUS){
-om_v_oob <- rbind(data.frame(model = 'OM',
-                             om_pp[om_pp$otu == i, ]),
-                  data.frame(model = 'OOB', otu = i,
-                             partialPlot(rf_n_features, data.frame(cfu_shared_colonized.df), 
-                                         OTUS[OTUS == i], plot = F)))
-part_plot_cfu[[i]] <- list(ggplot(om_v_oob, aes(x,y, group = model, color = model)) +geom_smooth() + geom_point(alpha = 0.5) + ggtitle(i))
-}
-part_plot_cfu[1]
-part_plot_cfu[2]
-part_plot_cfu[3]
-
-# partial plot for one_mouse model
-#ggplot(data=om_pp[om_pp$otu==names(final_rank[5]), ], aes(x=x, y=y)) + geom_smooth()
-#  stat_summary(fun.y=median, geom="line")
 
@@
 om_pred_cr <- c()
-om_pp_cr <- c()
 for(i in 1:iters){
@@
                              importance = TRUE, keep.forest = T, ntree = n_trees)
-    for(i in 4:ncol(train)){
-      pp_var <- data.frame(otu = colnames(train)[i],
-                           partialPlot(rf_om, data.frame(select(train, -mouse_id, -cage_id)), 
-                                       x.var= colnames(train)[i], plot = FALSE))
-      om_pp_cr <- rbind(om_pp_cr, pp_var)
-      }
     om_pred_cr <- rbind(om_pred_cr,
@@
 
-om_v_oob <- rbind(data.frame(model = 'OM',
-                             om_pp_cr[om_pp_cr$otu == 'Otu000004', ]),
-                  data.frame(model = 'OOB', otu = 'Otu000004',
-                             partialPlot(rf_col_rate, shared_cr_df, 'Otu000004',
-                                         plot = F)))
-ggplot(om_v_oob, aes(x,y, group = model, color = model)) +geom_smooth() + geom_point()
-
 cr_features_cor <- sapply(as.character(cr_features), 

### 2016-11-10 submission/study.Rmd
paper writing progress

--- 2016-10-27 submission/study.tex
@@
-\documentclass[11pt,]{article}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-    \usepackage{xltxtra,xunicode}
-  \else
-    \usepackage{fontspec}
-  \fi
-  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
-  \newcommand{\euro}{€}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\usepackage[margin=1.0in]{geometry}
-\ifxetex
-  \usepackage[setpagesize=false, % page size defined by xetex
-              unicode=false, % unicode breaks when used with xetex
-              xetex]{hyperref}
-\else
-  \usepackage[unicode=true]{hyperref}
-\fi
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={},
-            pdftitle={Microbiota predict Clostridium difficile severity in germ-free mice colonized with human feces},
-            colorlinks=true,
-            citecolor=blue,
-            urlcolor=blue,
-            linkcolor=magenta,
-            pdfborder={0 0 0}}
-\urlstyle{same}  % don't use monospace font for urls
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
-\setcounter{secnumdepth}{0}
-
-%%% Use protect on footnotes to avoid problems with footnotes in titles
-\let\rmarkdownfootnote\footnote%
-\def\footnote{\protect\rmarkdownfootnote}
-
-%%% Change title format to be more compact
-\usepackage{titling}
-
-% Create subtitle command for use in maketitle
-\newcommand{\subtitle}[1]{
-  \posttitle{
-    \begin{center}\large#1\end{center}
-    }
-}
-
-\setlength{\droptitle}{-2em}
-  \title{\textbf{Microbiota predict \emph{Clostridium difficile} severity in
-germ-free mice colonized with human feces}}
-  \pretitle{\vspace{\droptitle}\centering\huge}
-  \posttitle{\par}
-  \author{}
-  \preauthor{}\postauthor{}
-  \date{}
-  \predate{}\postdate{}
-
-\usepackage{helvet} % Helvetica font
-\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
-\usepackage[T1]{fontenc}
-
-\usepackage[none]{hyphenat}
-
-\usepackage{setspace}
-\doublespacing
-\setlength{\parskip}{1em}
-
-\usepackage{lineno}
-
-\usepackage{pdfpages}
-
-\begin{document}
-
-\maketitle
+---
+title: "**Microbiota predict _Clostridium difficile_ severity in germ-free mice colonized with human feces**"
+[//]: # (bibliography: references.bib)
+output:
+  pdf_document:
+    includes:
+      in_header: header.tex
+    keep_tex: yes
+csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
+fontsize: 11pt
+geometry: margin=1.0in
+---
 
@@
 
-Running title: Microbiota predict \emph{C. difficile} severity in
-humanized mice
+Running title: Microbiota predict _C. difficile_ severity in humanized mice
 
@@
 
-Kaitlin J. Flynn\textsuperscript{1}, Nicholas
-Lesniak\textsuperscript{1}, Alyxandria M. Schubert\textsuperscript{2},
-Hamide Sinani\textsuperscript{?}, and Patrick D.
-Schloss\textsuperscript{1\(\dagger\)}
+Kaitlin J. Flynn^1^, Nicholas Lesniak^1^, Alyxandria M. Schubert^2^, Hamide Sinani^?^, and Patrick D. Schloss^1$\dagger$^
 
@@
 
-\(\dagger\) Corresponding author:
-\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}
+$\dagger$ Corresponding author: pschloss@umich.edu
 
-1. Department of Microbiology and Immunology, University of Michigan,
-Ann Arbor, Michigan 48109
+1\. Department of Microbiology and Immunology, University of Michigan, Ann Arbor, Michigan 48109
 
-2. Food and Drug Administration?
+2\. Food and Drug Administration?
 
-3. Department for Hamide?
+3\. Department for Hamide?
 
@@
 
-\subsubsection{Abstract}\label{abstract}
-
-\emph{Clostridium difficile} causes diarrheal disease when it
-successfully colonizes a dysbiotic gut microbial community. Current
-mouse models to study \emph{C. difficile} infection (CDI) rely on
-pre-treatment with antibiotics to disrupt the mouse microbiome prior to
-inoculation. This model does not allow for analysis of human-associated
-microbial community members that modulate \emph{C. difficile}
-colonization and expansion. To study human-associated microbes in the
-context of CDI, we inoculated germ-free C57BL/6 mice with one of 16
-human fecal samples from diarrheal or healthy donors and challenged with
-C. difficile 14 days later. Five unique donor-mice combinations resulted
-in severe CDI while the remaining 11 only experienced mild disease. Both
-healthy and diarrheal donors were susceptible to colonization and severe
-symptoms of CDI. To determine if specific microbes were associated with
-disease severity outcomes, we built a classification Random Forest
-machine learning model based on relative abundance data of the
-communities prior to infection. The model identified a number of
-bacterial populations associated with the development of severe CDI,
-including \emph{Bacilliales, Ruminococcaceae, Ruminococcus,
-Staphylococcus, Streptococcus and Bacteriodetes}. Additionally, a
-regression model accurately predicted colonization levels of \emph{C.
-difficile} at one to ten days post-infection. This model explained 99\%
-of the variance in the number of CFU isolated from mouse stool. Members
-of \emph{Lachnospiraceae, Parabacteroides, Bacteroidales, Bacteroidetes,
-Porphyromonadaceae} and unclassified \emph{Bacteria} families were
-predictive of future \emph{C. difficile} colonization levels. Finally,
-challenging these mice with different strains of \emph{C. difficile}
-revealed that susceptible human-associated microbial communities were
-prone to severe disease independent of strain type. Taken together these
-results suggest that human-associated microbial communities can be
-recapitulated in germ-free mice and used to characterize dynamics of
-CDI. Because both healthy and diarrheal patients were susceptible to
-severe CDI, machine-learning models are useful to identify bacterial
-populations that allow colonization and contribute to the development of
-\emph{C. difficile} associated disease in humans.
+### Abstract
+_Clostridium difficile_ causes diarrheal disease when it successfully colonizes a dysbiotic gut microbial community.
+Current mouse models to study _C.
+difficile_ infection (CDI) rely on pre-treatment with antibiotics to disrupt the mouse microbiome prior to inoculation.
+This model does not allow for analysis of human-associated microbial community members that modulate _C.
+difficile_ colonization and expansion.
+To study human-associated microbes in the context of CDI, we inoculated germ-free C57BL/6 mice with one of 16 human fecal samples from diarrheal or healthy donors and challenged with C.
+difficile 14 days later.
+Five unique donor-mice combinations resulted in severe CDI while the remaining 11 only experienced mild disease.
+Both healthy and diarrheal donors were susceptible to colonization and severe symptoms of CDI.
+To determine if specific microbes were associated with disease severity outcomes, we built a classification Random Forest machine learning model based on relative abundance data of the communities prior to infection.
+The model identified a number of bacterial populations associated with the development of severe CDI, including _Bacilliales, Ruminococcaceae, Ruminococcus, Staphylococcus, Streptococcus and Bacteriodetes_.
+Additionally, a regression model accurately predicted colonization levels of _C.
+difficile_ at one to ten days post-infection.
+This model explained 99% of the variance in the number of CFU isolated from mouse stool.
+Members of _Lachnospiraceae, Parabacteroides, Bacteroidales, Bacteroidetes, Porphyromonadaceae_ and unclassified _Bacteria_ families were predictive of future _C.
+difficile_ colonization levels.
+Finally, challenging these mice with different strains of _C.
+difficile_ revealed that susceptible human-associated microbial communities were prone to severe disease independent of strain type.
+Taken together these results suggest that human-associated microbial communities can be recapitulated in germ-free mice and used to characterize dynamics of CDI.
+Because both healthy and diarrheal patients were susceptible to severe CDI, machine-learning models are useful to identify bacterial populations that allow colonization and contribute to the development of _C.
+difficile_ associated disease in humans.
 
@@
 
-\subsubsection{Introduction}\label{introduction}
+### Introduction
+Clostridium difficile is an opportunistic pathogen of the human lower gastrointestinal tract.
+Disruption of the native microbial community of the gut by antibiotics is the most common risk factor for development of C.
+difficile infection (CDI) (1).
+C.
+difficile is a spore-forming bacteria and can persist on abiotic surfaces and is not readily killed by ethanol-based hand-sanitizers, putting hospital patients particularly at risk.
+Indeed, ~12% of hospital acquired infections in the United States are due to C.
+difficile and result in up to 15,000 deaths annually (2).
 
-Clostridium difficile is an opportunistic pathogen of the human lower
-gastrointestinal tract. Disruption of the native microbial community of
-the gut by antibiotics is the most common risk factor for development of
-C. difficile infection (CDI) (1). C. difficile is a spore-forming
-bacteria and can persist on abiotic surfaces and is not readily killed
-by ethanol-based hand-sanitizers, putting hospital patients particularly
-at risk. Indeed, \textasciitilde{}12\% of hospital acquired infections
-in the United States are due to C. difficile and result in up to 15,000
-deaths annually (2).
+**need to add a paragraph about colonization resistence**
 
-Murine models to study CDI typically rely on treating
-conventionally-raised mice with antibiotics either in drinking water or
-by injection to induce susceptibility (3, 4). This model provides a
-convenient way to study C. difficile pathogenesis and virulence factors.
-Numerous microbiome studies have been performed using this model to
-determine the antibiotic classes (5), starting microbial community (6)
-and metabolites (7) that impact development and severity of CDI. While
-informative, these studies are somewhat removed from human disease
-because they only examine mouse-associated microbial communities.
+Murine models to study CDI typically rely on treating conventionally-raised mice with antibiotics either in drinking water or by injection to induce susceptibility (3, 4).
+This model provides a convenient way to study C.
+difficile pathogenesis and virulence factors.
+Numerous microbiome studies have been performed using this model to determine the antibiotic classes (5), starting microbial community (6) and metabolites (7) that impact development and severity of CDI.
+While informative, these studies are somewhat removed from human disease because they only examine mouse-associated microbial communities.
 
-Gnotobiotic or germ-free mouse models have been used for a range of
-studies of CDI, including assessment of species-specific interactions
-between C. difficile and competing microbial community members (8),
-analysis of nutrient restriction (9), in vivo transcriptomics of C.
-difficile and examination of host immune response to CDI (10). Further,
-CDI therapeutics such as antibiotics and fecal microbiota transplants
-have been tested extensively in a gnotobiotic-piglet or
-piglet-to-gnotobiotic-mouse model of disease (11), (12). Pigs have a
-longer digestive tract with components more similar to humans than mice
-and are typically infected by strains typical in human infection (13).
-However, the murine and porcine microbiomes typically do not resemble
-those of the human gut.
+Gnotobiotic or germ-free mouse models have been used for a range of studies of CDI, including assessment of species-specific interactions between C.
+difficile and competing microbial community members (8), analysis of nutrient restriction (9), _in vivo_ transcriptomics of C.
+difficile and examination of host immune response to CDI (10).
+Further, CDI therapeutics such as antibiotics and fecal microbiota transplants have been tested extensively in a gnotobiotic-piglet or piglet-to-gnotobiotic-mouse model of disease (11), (12).
+Pigs have a longer digestive tract with components more similar to humans than mice and are typically infected by strains typical in human infection (13).
+However, the murine and porcine microbiomes typically do not resemble those of the human gut.
 
-The power of the gnotobiotic models to study CDI has been further
-realized by first inoculating germ-free mice and piglets with human
-stool microbes. In one study, germ-free piglets were acutely colonized
-with human feces for one week and then treated with tigecycline. After
-challenge with C. difficile none of the antibiotic-treated piglets
-succumbed to infection, while some of the untreated human-colonized pigs
-did (11). Further, germ-free mice colonized with human feces were bred
-over several generations to create a cohort of mice with identical
-human-derived microbiomes (14). These mice were subsequently treated
-with a five-antibiotic cocktail to induce dysbiosis and then were
-successfully colonized by C. difficile (14). While informative, these
-studies were limited in their use of only one human donor as input
-inoculum. In order to best understand the impact of C. difficile
-pathogenesis on human disease, we must have a laboratory model that
-allows for study of a variety of human-derived microbiomes.
+The power of the gnotobiotic models to study CDI has been further realized by first inoculating germ-free mice and piglets with human stool microbes.
+In one study, germ-free piglets were acutely colonized with human feces for one week and then treated with tigecycline.
+After challenge with C.
+difficile none of the antibiotic-treated piglets succumbed to infection, while some of the untreated human-colonized pigs did (11).
+Further, germ-free mice colonized with human feces were bred over several generations to create a cohort of mice with identical human-derived microbiomes (14).
+These mice were subsequently treated with a five-antibiotic cocktail to induce dysbiosis and then were successfully colonized by C.
+difficile (14).
+While informative, these studies were limited in their use of only one human donor as input inoculum.
+In order to best understand the impact of C.
+difficile pathogenesis on human disease, we must have a laboratory model that allows for study of a variety of human-derived microbiomes.
 
-To test the impact of individual human microbiomes on CDI, we colonized
-germ-free mice with 16 different human stool donors. We then
-characterized human-associated microbiome response to C. difficile
-challenge. Additionally, the use of machine-learning models allowed us
-to build a predictive model that classified ``at-risk'' microbiomes
-prior to infection with C. difficile. These findings show that
-human-associated microbiomes can be at risk for CDI even in the absence
-of antibiotics and that study of mice colonized with human feces
-provides a range of clinical outcomes.
+To test the impact of individual human microbiomes on CDI, we colonized germ-free mice with 16 different human stool donors.
+We then characterized human-associated microbiome response to C.
+difficile challenge.
+Additionally, the use of machine-learning models allowed us to build a predictive model that classified “at-risk” microbiomes prior to infection with C.
+difficile.
+These findings show that human-associated microbiomes can be at risk for CDI even in the absence of antibiotics and that study of mice colonized with human feces provides a range of clinical outcomes.
 
-\subsubsection{Results}\label{results}
+### Results 
+**Germ-free mice inoculated with human feces as model for C.
+difficile infection.** To generate mice with human-derived microbiomes, we inoculated one cage of gnotobiotic C57/BL6 mice with one of 16 different human fecal donors.
+Five donors were patients that had diarrhea that was not attributable to _C.
+difficile_ infection while 11 donors were healthy at time of donation.
+Stool from a patient that was colonized with virulent _C.
+difficile_ was used as a positive control.
+After inoculation with human stool, mice were allowed to equilibrate for 14 days.
+Prior to infection, stool samples were taken from each mouse to establish baseline.
+Then, the _C.
+difficile_ strain isolated from the positive control patient’s sample (strain 430) was used to infect each mouse with 100 spores.
+Mice were monitored for weight loss and clinical signs of disease.
+Fecal samples were taken to enumerate _C.
+difficile_ CFU and for microbiome analysis every day for up to 10 days post-infection (Fig 1A).
+To ensure that the donors we selected represented a diverse array of human microbiomes, we sequenced the 16S rRNA genes from donor fecal inocula.
+Ordination of the distances between donor communities showed that the donors each had distinctly different communities, independent of whether the sample came from a sick or healthy person (Fig 1B).
+Likewise, the starting microbial communities of the mice on day 0 were characterizing by sequencing of fecal pellets DNA prior to infection.
+Ordination of all of the mouse communities on day 0 shows that mice were similar to each other within each cage and donor, but distinct from other donors (Fig 1C).
+This result confirmed that human-associated microbes were able to colonize gnotobiotic mice and provide distinct initial communities to test _C.
+difficile_ dynamics.
 
-\textbf{Germ-free mice inoculated with human feces as model for C.
-difficile infection.} To generate mice with human-derived microbiomes,
-we inoculated one cage of gnotobiotic C57/BL6 mice with one of 16
-different human fecal donors. Five donors were patients that had
-diarrhea that was not attributable to \emph{C. difficile} infection
-while 11 donors were healthy at time of donation. Stool from a patient
-that was colonized with virulent \emph{C. difficile} was used as a
-positive control. After inoculation with human stool, mice were allowed
-to equilibrate for 14 days. Prior to infection, stool samples were taken
-from each mouse to establish baseline. Then, the \emph{C. difficile}
-strain isolated from the positive control patient's sample (strain 430)
-was used to infect each mouse with 100 spores. Mice were monitored for
-weight loss and clinical signs of disease. Fecal samples were taken to
-enumerate \emph{C. difficile} CFU and for microbiome analysis every day
-for up to 10 days post-infection (Fig 1A). To ensure that the donors we
-selected represented a diverse array of human microbiomes, we sequenced
-the 16S rRNA genes from donor fecal inocula. Ordination of the distances
-between donor communities showed that the donors each had distinctly
-different communities, independent of whether the sample came from a
-sick or healthy person (Fig 1B). Likewise, the starting microbial
-communities of the mice on day 0 were characterizing by sequencing of
-fecal pellets DNA prior to infection. Ordination of all of the mouse
-communities on day 0 shows that mice were similar to each other within
-each cage and donor, but distinct from other donors (Fig 1C). This
-result confirmed that human-associated microbes were able to colonize
-gnotobiotic mice and provide distinct initial communities to test
-\emph{C. difficile} dynamics.
+**C.
+difficile infection in mice with human-derived microbiota cause a range of outcomes.** _C.
+difficile_ colonization was monitored by daily plating of stool pellets for _C.
+difficile_ CFU.
+Nearly all of the mice were colonized to 10^5^ – 10^7^ CFU by one day post-infection and remained colonized at that level until the end of the experiment (Fig 2A).
+As one indicator of disease, mouse weights were taken each day post-infection and weight-loss was monitored alongside clinical signs of disease.
+When mice were judged to be too ill to continue they were humanely euthanized.
+Overall, disease phenotypes fell into two classes.
+Mice that became severely ill and lost 20% or more of their starting body weight within one to two days post-infection were classified as “severe” whereas mice that were colonized with _C.
+difficile_ but did not show signs of disease or severe weight loss were considered to have “mild” disease (Fig 2A, 2B).
+Interestingly, _C.
+difficile_ was able to cause severe disease in both mice that had been colonized with healthy stool and those colonized with diarrheal stool, suggesting susceptibility to CDI is dependent on the composition of the starting microbiome and not associated with donor clinical metadata.
 
-\textbf{C. difficile infection in mice with human-derived microbiota
-cause a range of outcomes.} \emph{C. difficile} colonization was
-monitored by daily plating of stool pellets for \emph{C. difficile} CFU.
-Nearly all of the mice were colonized to 10\textsuperscript{5} --
-10\textsuperscript{7} CFU by one day post-infection and remained
-colonized at that level until the end of the experiment (Fig 2A). As one
-indicator of disease, mouse weights were taken each day post-infection
-and weight-loss was monitored alongside clinical signs of disease. When
-mice were judged to be too ill to continue they were humanely
-euthanized. Overall, disease phenotypes fell into two classes. Mice that
-became severely ill and lost 20\% or more of their starting body weight
-within one to two days post-infection were classified as ``severe''
-whereas mice that were colonized with \emph{C. difficile} but did not
-show signs of disease or severe weight loss were considered to have
-``mild'' disease (Fig 2A, 2B). Interestingly, \emph{C. difficile} was
-able to cause severe disease in both mice that had been colonized with
-healthy stool and those colonized with diarrheal stool, suggesting
-susceptibility to CDI is dependent on the composition of the starting
-microbiome and not associated with donor clinical metadata.
+**Microbes present in the gut prior to infection are predictive of C. difficile colonization levels**
 
-\textbf{Results to be written} 1. Microbes present in the gut prior to
-infection are predictive of C. difficile CFU and severity a. Figure 3:
-Random forest to predict CFU, predictive OTUs b. Figure 4: Random Forest
-predicts CDI severity, predictive OTUs 2. Propensity for severe CDI is
-community-dependent and strain-independent a. Figure 5: Infection of
-mice with different C. difficile strains.
+**Results to be written**
+1.	Microbes present in the gut prior to infection are predictive of C. difficile CFU and severity 
+a.	Figure 3: Random forest to predict CFU, predictive OTUs 
+b.	Figure 4: Random Forest predicts CDI severity, predictive OTUs
+2.	Propensity for severe CDI is community-dependent and strain-independent
+a.	Figure 5:  Infection of mice with different C. difficile strains.
 
-\subsubsection{Discussion}\label{discussion}
+### Discussion
+•	Restate results,
+•	caveats about mouse weights 
+•	No donors were colonization resistant, discuss donor differences 
+•	Discuss prediction methods and outcomes
+•	Discuss potential mechanisms for interesting OTUs
+•	Discuss different strain results
+•	Future work blah blah
 
-• Restate results, • caveats about mouse weights • No donors were
-colonization resistant, discuss donor differences • Discuss prediction
-methods and outcomes • Discuss potential mechanisms for interesting OTUs
-• Discuss different strain results • Future work blah blah
+### Materials and Methods
 
-\subsubsection{Materials and Methods}\label{materials-and-methods}
+**Mice** The experiments in this study used 6-8 week-old male and female gnotobiotic C57/BL6 mice.
+The mice were born and maintainted in a completely sterile environment in the Germ Free Mouse Facility at the University of Michigan.
+The University Committee on Use and Care of Animals approved all protocols and experiments conducted in this study.
 
-• Mice ULAM number • Donor stool ERIN IRB shit • Bacteria/plating •
-Sequencing • Data analysis, code availability • Machine learning models
+**Human samples** Donor stool samples were chosen from a sample bank collected from patients from October 2010 to November 2012 at the University of Michigan.
+Sample collection methods and study was approved by the Institutional Review Board at the University of Michigan.
+Stool samples from University of Michigan Health System patients were submitted to the microbiology lab for _C.
+difficile_ testing.
+_C.
+difficile_ presence was measured using WHATEVER- Cdiff quik check?
+toxin AB?.
+Samples from these patients were stored in Cary-Blair medium at -80C.
+All patients consented to participation in the study.
+Donor stool for this study was selected from one patient that tested positive for _C.
+difficile_ infection and four patients that presented with diarrhea that tested negative for _C.
+difficile_.
+Healthy donor stool was collected from volunteer donors living in the Ann Arbor, MI area.
 
-\subsection{Acknowledgments}\label{acknowledgments}
+•	Bacteria/plating 
+•	Sequencing
+•	Data analysis, code availability
+•	Machine learning models
 
+##Acknowledgments## 
 Lab, sequencing core, Jhansi
@@
 
-\subsubsection{Figure Legends}\label{figure-legends}
+###Figure Legends###
 
-\textbf{Figure 1. Germ-free mice inoculated with human feces as a model
-for C. difficile infection.} A) Experimental design. Stool from 16
-healthy, diarrheal and CDI patients were independently inoculated into
-3-4 germ-free mice by oral gavage. 14 days later mice were orally
-infected with 100 spores of C. difficile strain 431. Weight and stool
-CFU were monitored for up to 10 days post infection. B) NMDS ordination
-of donor stool communities prior to inoculating mice. Each point
-represents one donor and are colored by clinical diagnosis. C) NDMS
-ordination of the stool communities on day 0. Each symbol represents one
-mouse and is colored by donor. Circles represent mice that experienced
-mild disease and triangles represent those that suffered severe disease.
+**Figure 1.
+Germ-free mice inoculated with human feces as a model for _C.
+difficile infection._** A) Experimental design.
+Stool from 16 healthy, diarrheal and CDI patients were independently inoculated into 3-4 germ-free mice by oral gavage.
+14 days later mice were orally infected with 100 spores of C.
+difficile strain 431.
+Weight and stool CFU were monitored for up to 10 days post infection.
+B) NMDS ordination of donor stool communities prior to inoculating mice.
+Each point represents one donor and are colored by clinical diagnosis.
+C) NDMS ordination of the stool communities on day 0.
+Each symbol represents one mouse and is colored by donor.
+Circles represent mice that experienced mild disease and triangles represent those that suffered severe disease.
 
-\textbf{Figure 2. \emph{C. difficile} infection results in mild or
-severe disease.} A) \emph{C. difficile CFU} was enumerated by plating of
-mouse stool pellets daily. Each point represents a mouse and the lines
-represent the median CFU in each group. Error bars are interquartile
-ranges. Red lines and points correspond to mice that succumbed to severe
-disease whereas black lines and points correspond to mice that had mild
-or no disease. B) Mouse weights were recorded and daily percent weight
-loss calculated for each mouse. Data is presented as the median of each
-group and interquartile ranges. Mice that succumbed to severe infection
-typically lost a significant amount of weight by day 1 or 2 post
-infection. Red lines correspond to severely ill mice, black to mice with
-mild disease.
+**Figure 2.
+_C.
+difficile_ infection results in mild or severe disease.** A) _C.
+difficile CFU_ was enumerated by plating of mouse stool pellets daily.
+Each point represents a mouse and the lines represent the median CFU in each group.
+Error bars are interquartile ranges.
+Red lines and points correspond to mice that succumbed to severe disease whereas black lines and points correspond to mice that had mild or no disease.
+B) Mouse weights were recorded and daily percent weight loss calculated for each mouse.
+Data is presented as the median of each group and interquartile ranges.
+Mice that succumbed to severe infection typically lost a significant amount of weight by day 1 or 2 post infection.
+Red lines correspond to severely ill mice, black to mice with mild disease.
 
-\textbf{Figure 3. Random Forest prediction of \emph{C. difficile}
-colonization level.} A) OTUs above 1\% relative abundance on day 0 were
-used to predict median log\textsubscript{10} CFU of \emph{C. difficile}
-after colonization. OTUs were chosen such that they were not predictive
-of cage or donor. Each point is a mouse colored by cage. B) Partial
-dependency plots of the top six predictive OTUs. Line displats the
-partial dependence of log\textsubscript{10} CFU on the relative
-abundance of each predictive OUT. Each median log\textsubscript{10} CFU
-is plotted against its relative abundance for each predictive OTU.
+**Figure 3.
+Random Forest prediction of _C.
+difficile_ colonization level.** A) OTUs above 1% relative abundance on day 0 were used to predict median log~10~ CFU of _C.
+difficile_ after colonization.
+OTUs were chosen such that they were not predictive of cage or donor.
+Each point is a mouse colored by cage.
+B) Partial dependency plots of the top six predictive OTUs.
+Line displats the partial dependence of log~10~ CFU on the relative abundance of each predictive OUT.
+Each median log~10~ CFU is plotted against its relative abundance for each predictive OTU.
 
-\textbf{Figure 4. Random Forest prediction of CDI severity.} OTUs above
-1\% relative abundance on day 0 were used to predict disease severity.
+**Figure 4.
+Random Forest prediction of CDI severity.** OTUs above 1% relative abundance on day 0 were used to predict disease severity.
 OTUs were chosen such that they were not predictive of cage or donor.
-Predictive classification tested via 10-fold (gray), leave-one-cage-out
-(purple dashed) or leave-one-mouse-out (blue dashed) models are
-displayed in (A). B) Partial dependency plots of most predictive OTUs.
-Line displays the partial dependence of log\textsubscript{10} CFU on OTU
-relative abundance. Points are the OTU relative abundance of each mouse
-colored by outcome (red, severe, black, mild).
+Predictive classification tested via 10-fold (gray), leave-one-cage-out (purple dashed) or leave-one-mouse-out (blue dashed) models are displayed in (A).
+B) Partial dependency plots of most predictive OTUs.
+Line displays the partial dependence of log~10~ CFU on OTU relative abundance.
+Points are the OTU relative abundance of each mouse colored by outcome (red, severe, black, mild).
+
+**Figure 5.
+Infection of mice with different _C.
+difficile strains_.** 3 strains of _C.
+difficile_ were used to infect mice colonized with susceptible (DA00578) or resistant (DA00369, DA00430) human donor stool.
+A) _C.
+difficile_ stool CFU was enumerated over 10 days.
+B) Percent weight loss was calculated each day for each mouse.
+In both plots, each mouse is a point and lines represent the mean of each cage.
+
+####Supplement####
+**Table S1: Mouse day 0 communities by donor genera (avg + stdev of cage)**
 
@@
 
-\subsection*{References}\label{references}
-\addcontentsline{toc}{subsection}{References}
-
-\end{document}
+## References

### 2016-11-13 submission/Manuscript.Rmd
almost done with results

--- 2016-11-10 submission/study.Rmd
@@
 title: "**Microbiota predict _Clostridium difficile_ severity in germ-free mice colonized with human feces**"
-[//]: # (bibliography: references.bib)
 output:
@@
 ### Introduction
-Clostridium difficile is an opportunistic pathogen of the human lower gastrointestinal tract.
-Disruption of the native microbial community of the gut by antibiotics is the most common risk factor for development of C.
-difficile infection (CDI) (1).
-C.
-difficile is a spore-forming bacteria and can persist on abiotic surfaces and is not readily killed by ethanol-based hand-sanitizers, putting hospital patients particularly at risk.
-Indeed, ~12% of hospital acquired infections in the United States are due to C.
-difficile and result in up to 15,000 deaths annually (2).
-
-**need to add a paragraph about colonization resistence**
+_Clostridium difficile_ is an opportunistic pathogen of the human lower gastrointestinal tract.
+_C.
+difficile_ forms spores that can persist on abiotic surfaces and are not readily killed by ethanol-based hand-sanitizers, putting hospital patients particularly at risk.
+Indeed, ~12% of hospital acquired infections in the United States are due to _C.
+difficile_ and result in up to 15,000 deaths annually (2).
+Disruption of the native microbial community is the most common risk factor for development of _C.
+difficile_ infection (CDI) (1).
+Antibiotic use and inflammatory bowel diseases are associated with loss of colonization resistance to _C.
+difficile_ through the loss of potentially protective bacterial families such as _Barnesiella_ and _Lachnospiraceae_ in both mouse models and human association experiments (cite alyx 2014 and vincent 2013).
+The composition of the community is clearly important for the acquisition, resistance and treatment of _C.
+difficile_ infection, as giving patients a healthy fecal microbiome transplant is the most effective treatment for this disease (cite Anna 2014).
+The precise mechanisms of colonization resistance and restoration of a healthy community are yet to be discovered.
 
 Murine models to study CDI typically rely on treating conventionally-raised mice with antibiotics either in drinking water or by injection to induce susceptibility (3, 4).
-This model provides a convenient way to study C.
-difficile pathogenesis and virulence factors.
-Numerous microbiome studies have been performed using this model to determine the antibiotic classes (5), starting microbial community (6) and metabolites (7) that impact development and severity of CDI.
+This model provides a convenient way to study _C.
+difficile_ pathogenesis and virulence factors.
+Numerous microbiome studies have been performed using this model to determine the antibiotic classes (5), starting microbial community (6) and metabolites (7) that impact development and severity of _C.
+difficile_ infection.
 While informative, these studies are somewhat removed from human disease because they only examine mouse-associated microbial communities.
 
-Gnotobiotic or germ-free mouse models have been used for a range of studies of CDI, including assessment of species-specific interactions between C.
-difficile and competing microbial community members (8), analysis of nutrient restriction (9), _in vivo_ transcriptomics of C.
-difficile and examination of host immune response to CDI (10).
+Gnotobiotic or germ-free mouse models have been used for a range of studies of CDI, including assessment of species-specific interactions between _C.
+difficile_ and competing microbial community members (8), analysis of nutrient restriction (9), _in vivo_ transcriptomics of _C.
+difficile_ and examination of host immune response to CDI (10).
 Further, CDI therapeutics such as antibiotics and fecal microbiota transplants have been tested extensively in a gnotobiotic-piglet or piglet-to-gnotobiotic-mouse model of disease (11), (12).
@@
 difficile none of the antibiotic-treated piglets succumbed to infection, while some of the untreated human-colonized pigs did (11).
-Further, germ-free mice colonized with human feces were bred over several generations to create a cohort of mice with identical human-derived microbiomes (14).
-These mice were subsequently treated with a five-antibiotic cocktail to induce dysbiosis and then were successfully colonized by C.
-difficile (14).
+In another study, germ-free mice colonized with human feces were bred over several generations to create a cohort of mice with identical human-derived microbiomes (14).
+These mice were subsequently treated with a five-antibiotic cocktail to induce dysbiosis and then were successfully colonized by _C.
+difficile_ (14).
 While informative, these studies were limited in their use of only one human donor as input inoculum.
-In order to best understand the impact of C.
-difficile pathogenesis on human disease, we must have a laboratory model that allows for study of a variety of human-derived microbiomes.
+In order to best understand the impact of _C.
+difficile_ pathogenesis on human disease, we must have a laboratory model that allows for study of a variety of human-derived microbiomes.
 
+Here we designed a laboratory model that allows for studying and modeling microbial coumminity interactions with _C.
+difficile_ infection in mice with human derived microbiomes.
 To test the impact of individual human microbiomes on CDI, we colonized germ-free mice with 16 different human stool donors.
@@
 
+**need to put in a line to reference supplemental table with RA in it**
+
 **C.
@@
 
-**Microbes present in the gut prior to infection are predictive of C. difficile colonization levels**
+**Microbes present in the gut prior to infection are predictive of _C.
+difficile_ colonization levels** Previous work in our group has demonstrated that the microbes present in the mouse microbiome prior to _C.
+difficile_ challenge can predict future colonization levels (cite Alyx).
+To determine if the human-associated microbiome in the mouse could similarly predict _C.
+difficile_ colonization we employed a Random Forest machine learning algorithm similar to the one used in previous studies.
+Relative abundance of OTUs with greater than 1% abundance at day 0 were used as input to the model.
+We found that the model explained X% of the variance in colonization levels of _C.
+difficile_ (Fig 3A).
+By refining the model to include only the top X predictive OTUs, the model explained X% of the variance.
+Partial dependency plots of the top 6 OTUs predictive of _C.
+difficile_ colonization reveal that colonization is positively associated with _Bacteriodetes_, _Parabacteriodes_ and _Porphyromonadaceae_ (really?) (Fig 3B).
+Colonization was negatively associated with the presence of _Lachnospiraceae_ (Fig 3B) though none of the human-associated mouse communities were completely resistant to colonization by _C.
+difficile_ (Fig 2A, 3B).
+These results confirm that as in mouse microbiome studies, the human-associated gut community prior to challenge is predictive of _C.
+difficile_ colonization levels.
+
+**The microbiome predicts _C.
+difficile_ severity.** As _C.
+difficile_ colonization levels can be predicted by the microbiome composition on day 0, we postulated that severity of disease could also be predicted from day 0.
+Using weight loss as a proxy for severity, we built a Random Forest classification model based on relative abundance to determine which community members were associated with either mild or severe disease.
+We optimized the model using the AUC-RF algorithm to generate a Reciever Operating Curve (ROC) with maximal Area Under the Curve (AUC) (cite Calle 2011), as used previously in our group to classify colon cancer outcomes (cite Niel 2015 gen med).
+The optimized model used X number of OTUs and identified several that predicted increased weightloss and subsequent disease severity (Fig 4A, 4B).
+The populations that were most commonly found in mice that succumbed to _C.
+difficile_ infection were members of the _Clostridiales_, _Bacilliales_, _Streptococcus_ (Lactobacilles?) and _Bacteriodes_.
+(need more OTUs here, or number that were used for model and then group by rank).
+The AUC was calculated using the Out-Of-Bag (OOB) error for each sample (Fig 4A).
+Cross-validation using 10-fold, leave-one-mouse-out and leave-one-cage-out resulted in AUCs not significantly different from the OOB (Fig 4A).
+This model reveals that the levels of X OTUs on day 0 can predict disease outcome in mice with human-associated microbiomes.
 
 **Results to be written**
-1.	Microbes present in the gut prior to infection are predictive of C. difficile CFU and severity 
-a.	Figure 3: Random forest to predict CFU, predictive OTUs 
+
 b.	Figure 4: Random Forest predicts CDI severity, predictive OTUs

### 2016-11-14 code/CdGF_AUCRF.Rmd

@@
 iters <- 100
-n_features <- 65
 ```
@@
 
+#### leave-2-out
+LOdO_df <- Predict_df %>% 
+                select(Euth_Early, human_source, one_of(euth_otu_feat))
+donor_list <- as.character(unique(LOdO_df$human_source))
+
+LOdO_probs <- c()
+LOdO_rf <- c()
+for(i in seq(1,length(donor_list),2)){
+  set.seed(seed)
+  donors <- sample(donor_list)
+  by_donor <- LOdO_df$human_source %in% donors[i:(i+1)]
+  LOdO_predict <- select(LOdO_df, -human_source)
+  train <- LOdO_predict[!by_donor,]
+  test <- LOdO_predict[by_donor,]
+  set.seed(seed)
+  temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
+  LOdO_rf[[i]] <- temp_model$RFopt
+  LOdO_probs[LOdO_df$human_source %in% donors[i:(i+1)]] <- predict(temp_model$RFopt, test, type='prob')[,2]
+}
+LOdO_roc <- roc(LOdO_df$Euth_Early~LOdO_probs)
+
 #### plot rocs
@@
 plot(LOcO_roc, col='purple', lwd=2, add=T, lty=3)
+plot(LOdO_roc, col='red', lwd=2, add=T, lty=3)
 mtext(side=2, text="Sensitivity", line=2.5)
@@
                                sprintf('Leave-1 cage-out (AUC: %.3g)',LOcO_roc$auc),
+                               sprintf('Leave-2 cages-out (AUC: %.3g)',LOdO_roc$auc),
                                sprintf('10-fold CV (AUC: %.3g)',cv10f_roc$auc) #,
@@
 #                               sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
-),lty=c(1,2,2,1,0,0), lwd=2, col=c('black','blue','purple','gray','red'), bty='n')
+),lty=c(1,2,2,2,1,0,0), lwd=2, col=c('black','blue','purple','red','gray'), bty='n')
 ```

### 2016-11-14 code/CdGF_disease_severity.Rmd

---
title: "humanCdGF_disease_severity"
author: "Nicholas Lesniak"
date: "October 27, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE, fig.width=7, fig.height=8)
```

```{r setup environment}
pack_used <- c('tidyr','ggplot2','dplyr', 'knitr')
for (dep in pack_used){
  if (dep %in% installed.packages()[,"Package"] == FALSE){
    install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
                     quiet=TRUE);
  }
  library(dep, verbose=FALSE, character.only=TRUE)
}
```

```{r setup, include=FALSE}
meta_file   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_all.an.cons.taxonomy'
tax_function <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R'

# read in files
meta_file   <- read.table(meta_file, sep = '\t', header = T, row.names = 'sample_id')
shared_file <- read.table(shared_file, sep = '\t', header = T, row.names = 'sample_id')
tax_file <- read.table(tax_file, sep = '\t', header = T, row.names = 'OTU')
source(tax_function)

# select only mice with 431 cdiff
meta_file <- meta_file[meta_file$cdiff_strain == '431', ]

# subset shared
#Create df with relative abundances
stool_shared <- shared_file[rownames(meta_file[meta_file$day %in% 0, ]) ,]
# use %in% instead of == because to ignore NA
# rename day 0 community rows by mouse id - since meta rows have day in row names, use mouse id for both
rownames(stool_shared) <- meta_file$mouse_id[rownames(meta_file) %in% 
                                               rownames(stool_shared) ]
rel_abund_stool <- 100 * stool_shared / unique(apply(stool_shared, 1, sum))
#Create vector of OTUs with median abundances >1%
OTUs_stool_0_01 <- apply(rel_abund_stool, 2, max) > 1
#df of OTUs with abundances >1% - by cage and inoculum
rel_abund_stool <- rel_abund_stool[ ,OTUs_stool_0_01]
OTU_list <- colnames(rel_abund_stool)
```

```{r Fig_2, fig.width=13, fig.height=7}

# Weight loss  
meta_file %>% 
  filter(day %in% c(0:10), !is.na(percent_weightLoss) ) %>%
  filter(!cage_id == 'NP1') %>% 
  select(mouse_id, day, Early_Euth, percent_weightLoss, cage_id) %>% 
  ggplot(aes(x = factor(day), y = percent_weightLoss, color = factor(Early_Euth))) + 
        geom_jitter(width = 0.5, alpha = 0.1) + 
        scale_color_manual(values = c('black', 'red'), breaks=c("FALSE", "TRUE"),
                           labels=c("Persistent", "Severe")) +
        stat_summary(fun.y=median, geom="line", aes(group = Early_Euth)) +
        labs(x = 'Day', y = ' Percent Relative to Day 0', title = 'Weight Loss') +
        theme(legend.justification=c(1,0), legend.position=c(1,0.1), 
                  legend.title=element_blank()) +
          stat_summary(fun.data = 'median_hilow')

# C. Diff CFU
meta_file %>% 
  filter(day %in% c(0:10), !is.na(log_cfu) ) %>%
  filter(!cage_id == 'NP1') %>% 
  select(mouse_id, day, Early_Euth, log_cfu, cage_id) %>% 
  ggplot(aes(x = factor(day), y = log_cfu, color = factor(Early_Euth))) + 
        geom_jitter(width = 0.5, alpha = 0.2) + 
        scale_color_manual(values = c('black', 'red'), breaks=c("FALSE", "TRUE"),
                           labels=c("Persistent", "Severe")) +
        stat_summary(fun.y=median, geom="line", aes(group = Early_Euth)) +
        labs(x = 'Day', y = ' CFU (log10)', title = 'C. Difficile Colonization') +
        theme(legend.justification=c(1,0), legend.position=c(1,0.2), 
                  legend.title=element_blank()) +
        stat_summary(fun.data = 'median_hilow')
```

### 2016-11-14 submission/Manuscript.Rmd

@@
 
+**also want something in the intro about severity maybe, and asymptomatic colonization. or in discussion?**
+
 ### Results 
@@
 **The microbiome predicts _C.
-difficile_ severity.** As _C.
+difficile_ infection severity.** As _C.
 difficile_ colonization levels can be predicted by the microbiome composition on day 0, we postulated that severity of disease could also be predicted from day 0.
@@
 This model reveals that the levels of X OTUs on day 0 can predict disease outcome in mice with human-associated microbiomes.
+(probably also going to want a supplemental table for predictive OTUs for this and Fig 3)
+
+**Susceptibility to infection is community-dependent.** _C.
+difficile_ strains are known to vary in disease phenotypes based both on pathogenicity factors of the strain and host microbiome composition (cite someone).
+To test the susceptibility of human-derived microbiomes to _C.
+difficile_ infection, two donor-mouse microbiomes that were previously shown to be resistant to _C.
+difficile_ strain 431 induced weight loss (Fig 2B) (DA00396, DA00430) and one that was suscpetible to severe disease (DA00578) were independently challenged with one of three unique _C.
+difficile_ strains (table or reference for toxin prod, etc).
+As done previously, _C.
+difficile_ stool CFU was enumerated daily and mouse weight-loss tracked.
+Infection of mice with resistant communities with virulent strain 458 or the weaker strain 299 resulted in colonization to levels near that of 431 (Fig 5A) and these mice did not experience severe weight loss or show clinical signs of disease (Fig 5B).
+In contrast, infection of mice harboring microbiomes susceptible to severe disease (DA00578, Fig 1B) with strain 3
 
@@
 
-b.	Figure 4: Random Forest predicts CDI severity, predictive OTUs
 2.	Propensity for severe CDI is community-dependent and strain-independent
@@
 •	No donors were colonization resistant, discuss donor differences 
+severity, asymptomatic colonzation discussion? 
 •	Discuss prediction methods and outcomes

### 2016-11-14 submission/Manuscript.tex

--- 2016-10-27 submission/study.tex
@@
 
-Clostridium difficile is an opportunistic pathogen of the human lower
-gastrointestinal tract. Disruption of the native microbial community of
-the gut by antibiotics is the most common risk factor for development of
-C. difficile infection (CDI) (1). C. difficile is a spore-forming
-bacteria and can persist on abiotic surfaces and is not readily killed
-by ethanol-based hand-sanitizers, putting hospital patients particularly
-at risk. Indeed, \textasciitilde{}12\% of hospital acquired infections
-in the United States are due to C. difficile and result in up to 15,000
-deaths annually (2).
+\emph{Clostridium difficile} is an opportunistic pathogen of the human
+lower gastrointestinal tract. \emph{C. difficile} forms spores that can
+persist on abiotic surfaces and are not readily killed by ethanol-based
+hand-sanitizers, putting hospital patients particularly at risk. Indeed,
+\textasciitilde{}12\% of hospital acquired infections in the United
+States are due to \emph{C. difficile} and result in up to 15,000 deaths
+annually (2). Disruption of the native microbial community is the most
+common risk factor for development of \emph{C. difficile} infection
+(CDI) (1). Antibiotic use and inflammatory bowel diseases are associated
+with loss of colonization resistance to \emph{C. difficile} through the
+loss of potentially protective bacterial families such as
+\emph{Barnesiella} and \emph{Lachnospiraceae} in both mouse models and
+human association experiments (cite alyx 2014 and vincent 2013). The
+composition of the community is clearly important for the acquisition,
+resistance and treatment of \emph{C. difficile} infection, as giving
+patients a healthy fecal microbiome transplant is the most effective
+treatment for this disease (cite Anna 2014). The precise mechanisms of
+colonization resistance and restoration of a healthy community are yet
+to be discovered.
 
@@
 by injection to induce susceptibility (3, 4). This model provides a
-convenient way to study C. difficile pathogenesis and virulence factors.
-Numerous microbiome studies have been performed using this model to
-determine the antibiotic classes (5), starting microbial community (6)
-and metabolites (7) that impact development and severity of CDI. While
-informative, these studies are somewhat removed from human disease
-because they only examine mouse-associated microbial communities.
+convenient way to study \emph{C. difficile} pathogenesis and virulence
+factors. Numerous microbiome studies have been performed using this
+model to determine the antibiotic classes (5), starting microbial
+community (6) and metabolites (7) that impact development and severity
+of \emph{C. difficile} infection. While informative, these studies are
+somewhat removed from human disease because they only examine
+mouse-associated microbial communities.
 
@@
 studies of CDI, including assessment of species-specific interactions
-between C. difficile and competing microbial community members (8),
-analysis of nutrient restriction (9), in vivo transcriptomics of C.
-difficile and examination of host immune response to CDI (10). Further,
-CDI therapeutics such as antibiotics and fecal microbiota transplants
-have been tested extensively in a gnotobiotic-piglet or
-piglet-to-gnotobiotic-mouse model of disease (11), (12). Pigs have a
-longer digestive tract with components more similar to humans than mice
-and are typically infected by strains typical in human infection (13).
-However, the murine and porcine microbiomes typically do not resemble
-those of the human gut.
+between \emph{C. difficile} and competing microbial community members
+(8), analysis of nutrient restriction (9), \emph{in vivo}
+transcriptomics of \emph{C. difficile} and examination of host immune
+response to CDI (10). Further, CDI therapeutics such as antibiotics and
+fecal microbiota transplants have been tested extensively in a
+gnotobiotic-piglet or piglet-to-gnotobiotic-mouse model of disease (11),
+(12). Pigs have a longer digestive tract with components more similar to
+humans than mice and are typically infected by strains typical in human
+infection (13). However, the murine and porcine microbiomes typically do
+not resemble those of the human gut.
 
@@
 succumbed to infection, while some of the untreated human-colonized pigs
-did (11). Further, germ-free mice colonized with human feces were bred
-over several generations to create a cohort of mice with identical
-human-derived microbiomes (14). These mice were subsequently treated
-with a five-antibiotic cocktail to induce dysbiosis and then were
-successfully colonized by C. difficile (14). While informative, these
-studies were limited in their use of only one human donor as input
-inoculum. In order to best understand the impact of C. difficile
-pathogenesis on human disease, we must have a laboratory model that
-allows for study of a variety of human-derived microbiomes.
+did (11). In another study, germ-free mice colonized with human feces
+were bred over several generations to create a cohort of mice with
+identical human-derived microbiomes (14). These mice were subsequently
+treated with a five-antibiotic cocktail to induce dysbiosis and then
+were successfully colonized by \emph{C. difficile} (14). While
+informative, these studies were limited in their use of only one human
+donor as input inoculum. In order to best understand the impact of
+\emph{C. difficile} pathogenesis on human disease, we must have a
+laboratory model that allows for study of a variety of human-derived
+microbiomes.
 
-To test the impact of individual human microbiomes on CDI, we colonized
-germ-free mice with 16 different human stool donors. We then
-characterized human-associated microbiome response to C. difficile
-challenge. Additionally, the use of machine-learning models allowed us
-to build a predictive model that classified ``at-risk'' microbiomes
-prior to infection with C. difficile. These findings show that
-human-associated microbiomes can be at risk for CDI even in the absence
-of antibiotics and that study of mice colonized with human feces
-provides a range of clinical outcomes.
+Here we designed a laboratory model that allows for studying and
+modeling microbial coumminity interactions with \emph{C. difficile}
+infection in mice with human derived microbiomes. To test the impact of
+individual human microbiomes on CDI, we colonized germ-free mice with 16
+different human stool donors. We then characterized human-associated
+microbiome response to C. difficile challenge. Additionally, the use of
+machine-learning models allowed us to build a predictive model that
+classified ``at-risk'' microbiomes prior to infection with C. difficile.
+These findings show that human-associated microbiomes can be at risk for
+CDI even in the absence of antibiotics and that study of mice colonized
+with human feces provides a range of clinical outcomes.
 
@@
 
+\textbf{need to put in a line to reference supplemental table with RA in
+it}
+
 \textbf{C. difficile infection in mice with human-derived microbiota
@@
 
-\textbf{Results to be written} 1. Microbes present in the gut prior to
-infection are predictive of C. difficile CFU and severity a. Figure 3:
-Random forest to predict CFU, predictive OTUs b. Figure 4: Random Forest
-predicts CDI severity, predictive OTUs 2. Propensity for severe CDI is
-community-dependent and strain-independent a. Figure 5: Infection of
-mice with different C. difficile strains.
+\textbf{Microbes present in the gut prior to infection are predictive of
+\emph{C. difficile} colonization levels} Previous work in our group has
+demonstrated that the microbes present in the mouse microbiome prior to
+\emph{C. difficile} challenge can predict future colonization levels
+(cite Alyx). To determine if the human-associated microbiome in the
+mouse could similarly predict \emph{C. difficile} colonization we
+employed a Random Forest machine learning algorithm similar to the one
+used in previous studies. Relative abundance of OTUs with greater than
+1\% abundance at day 0 were used as input to the model. We found that
+the model explained X\% of the variance in colonization levels of
+\emph{C. difficile} (Fig 3A). By refining the model to include only the
+top X predictive OTUs, the model explained X\% of the variance. Partial
+dependency plots of the top 6 OTUs predictive of \emph{C. difficile}
+colonization reveal that colonization is positively associated with
+\emph{Bacteriodetes}, \emph{Parabacteriodes} and
+\emph{Porphyromonadaceae} (really?) (Fig 3B). Colonization was
+negatively associated with the presence of \emph{Lachnospiraceae} (Fig
+3B) though none of the human-associated mouse communities were
+completely resistant to colonization by \emph{C. difficile} (Fig 2A,
+3B). These results confirm that as in mouse microbiome studies, the
+human-associated gut community prior to challenge is predictive of
+\emph{C. difficile} colonization levels.
+
+\textbf{The microbiome predicts \emph{C. difficile} severity.}
+
+\textbf{Results to be written}
+
+\begin{enumerate}
+\def\labelenumi{\alph{enumi}.}
+\setcounter{enumi}{1}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  Figure 4: Random Forest predicts CDI severity, predictive OTUs
+\end{enumerate}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{1}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  Propensity for severe CDI is community-dependent and
+  strain-independent
+\end{enumerate}
+
+\begin{enumerate}
+\def\labelenumi{\alph{enumi}.}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  Figure 5: Infection of mice with different C. difficile strains.
+\end{enumerate}
 
@@
 
-• Mice ULAM number • Donor stool ERIN IRB shit • Bacteria/plating •
-Sequencing • Data analysis, code availability • Machine learning models
+\textbf{Mice} The experiments in this study used 6-8 week-old male and
+female gnotobiotic C57/BL6 mice. The mice were born and maintainted in a
+completely sterile environment in the Germ Free Mouse Facility at the
+University of Michigan. The University Committee on Use and Care of
+Animals approved all protocols and experiments conducted in this study.
+
+\textbf{Human samples} Donor stool samples were chosen from a sample
+bank collected from patients from October 2010 to November 2012 at the
+University of Michigan. Sample collection methods and study was approved
+by the Institutional Review Board at the University of Michigan. Stool
+samples from University of Michigan Health System patients were
+submitted to the microbiology lab for \emph{C. difficile} testing.
+\emph{C. difficile} presence was measured using WHATEVER- Cdiff quik
+check? toxin AB?. Samples from these patients were stored in Cary-Blair
+medium at -80C. All patients consented to participation in the study.
+Donor stool for this study was selected from one patient that tested
+positive for \emph{C. difficile} infection and four patients that
+presented with diarrhea that tested negative for \emph{C. difficile}.
+Healthy donor stool was collected from volunteer donors living in the
+Ann Arbor, MI area.
+
+• Bacteria/plating • Sequencing • Data analysis, code availability •
+Machine learning models
 
@@
 \textbf{Figure 1. Germ-free mice inoculated with human feces as a model
-for C. difficile infection.} A) Experimental design. Stool from 16
-healthy, diarrheal and CDI patients were independently inoculated into
-3-4 germ-free mice by oral gavage. 14 days later mice were orally
+for \emph{C. difficile infection.}} A) Experimental design. Stool from
+16 healthy, diarrheal and CDI patients were independently inoculated
+into 3-4 germ-free mice by oral gavage. 14 days later mice were orally
 infected with 100 spores of C. difficile strain 431. Weight and stool
@@
 
+\textbf{Figure 5. Infection of mice with different \emph{C. difficile
+strains}.} 3 strains of \emph{C. difficile} were used to infect mice
+colonized with susceptible (DA00578) or resistant (DA00369, DA00430)
+human donor stool. A) \emph{C. difficile} stool CFU was enumerated over
+10 days. B) Percent weight loss was calculated each day for each mouse.
+In both plots, each mouse is a point and lines represent the mean of
+each cage.
+
+\paragraph{Supplement}\label{supplement}
+
+\textbf{Table S1: Mouse day 0 communities by donor genera (avg + stdev
+of cage)}
+
 \newpage
 
-\subsection*{References}\label{references}
-\addcontentsline{toc}{subsection}{References}
+\subsection{References}\label{references}
 

### 2016-11-15 code/CdGF_AUCRF.Rmd
one mouse model leaving one source out

@@
 ```{r}
-pack_used <- c('randomForest','ggplot2', 'pROC', 'knitr','dplyr','AUCRF', 'tidyr')
+pack_used <- c('randomForest','ggplot2', 'pROC', 'knitr','dplyr','AUCRF', 'tidyr', 'caret')
 for (dep in pack_used){
@@
 shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
-tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_all.an.cons.taxonomy'
+tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/gf_new.an.taxonomy'
 tax_function <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R'
@@
 rel_abund_stool <- rel_abund_stool[ ,OTUs_stool_0_01]
+
+# remove correlative otus
+rel_abund_cor <- cor(rel_abund_stool, method = 'spearman')
+cor_cutoff <- 0.75
+cor_otus <- findCorrelation(rel_abund_cor, cutoff = cor_cutoff)
+rel_abund_stool <- select(rel_abund_stool, -cor_otus)
 
@@
 cv10f_roc <- roc(cv10f_all_resp~cv10f_all_pred)
+
+#### One_mouse model
+Predict_om_df <- select(Predict_df, 
+                        Euth_Early, human_source, mouse_id, contains('Otu'))
+om_import_otus <- data.frame(otu = colnames(select(Predict_om_df, contains("Otu"))),
+                             importance = NA)
+om_aucs <- c()
+om_all <- c()
+for(j in 1:iters){
+  set.seed(j)
+  mouse_set <- Predict_om_df %>% 
+    group_by(human_source) %>% 
+    sample_n(1) %>% ungroup() 
+  om_probs <- c()
+  om_cage_import <- c()
+  for(n in unique(as.character(mouse_set$human_source))){
+    train <- mouse_set[mouse_set$human_source != n, ]
+    test <- Predict_om_df %>% 
+      filter(human_source %in% n)
+    set.seed(seed)
+    temp_model <- AUCRF(Euth_Early~., data=select(train, -mouse_id, -human_source)
+                      , pdel=0.2, ntree=500, ranking = 'MDA')
+    om_probs <- rbind(om_probs,
+                      data.frame(obs = test$Euth_Early,
+                                 pred = predict(temp_model$RFopt, test, type='prob')[,2]))
+    om_cage_import <- c(om_cage_import, temp_model$Xopt)
+  }
+  om_roc <- roc(om_probs$obs~om_probs$pred)
+  om_import_otus <- merge(om_import_otus,
+                          data.frame(otu = names(table(om_cage_import)),
+                            importance = as.numeric(table(om_cage_import))*om_roc$auc),
+                          by = 'otu', all.x = T, suffixes = c('',round(om_roc$auc, 3)))
+  om_all <- rbind(om_all, om_probs)
+  om_aucs[j] <- om_roc$auc
+}
+om_roc <- roc(om_all$obs~om_all$pred)
+
+raw_importance <- data.frame(om_import_otus[ , -c(1,2)], row.names = om_import_otus$otu)
+
+importance_otus <-   apply(raw_importance, 1, sum, na.rm = T)
+importance_otus <- importance_otus[importance_otus>0]
+importance_otus <- importance_otus[importance_otus > median(om_aucs)*150/2]
+
+### rank importance
+importance_rank <- apply(raw_importance, 2, function(x) rank(x, na.last = T))
+raw_rank <- apply(importance_rank, 1, function(x) sum(x, na.rm = T))
+final_rank <- sort(raw_rank[raw_rank >=  ncol(importance_rank)*sum(importance_rank[ , 1])/nrow(importance_rank)], decreasing = T)
 
@@
 plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='')
-plot(otu_euth_roc, col='black', lwd=2, add=T, lty=1)
+plot(om_roc, col='red', lwd=2, add=T, lty=1)
 plot(cv10f_roc, col='gray', lwd=2, add=T, lty=1)
@@
 ```{r rel abund partial depend, fig.width=10, fig.height=10}
-
-#n_rf_features <- 9
-#euth_otus <- euth_otu_feat[1:n_rf_features]
-euth_tax <- get_tax(1, euth_otu_feat, tax_file)
-euth_tax$tax_label <- paste(euth_tax$tax, ' (', 
-                            gsub('tu0*', 'TU ', rownames(euth_tax)), 
-                                   ') : ', round(rf_early_aucrf$ranking[euth_otu_feat], 3), sep = '')
-
 layout(matrix(c(1:9), ncol=3, byrow=TRUE))
@@
   partial_otu <- partialPlot(rf_early_aucrf$RFopt, aucrf_data, euth_otu_feat[i], xlab='',
-                main=paste(euth_tax$tax_label[i]))
+                main=paste(get_tax(row_list = euth_otu_feat)$tax_label[i]))
 points(Predict_early_euth_df[Predict_early_euth_df$Euth_Early == 1 ,euth_otu_feat[i]], 

### 2016-11-15 submission/Manuscript.Rmd

@@
 Infection of mice with resistant communities with virulent strain 458 or the weaker strain 299 resulted in colonization to levels near that of 431 (Fig 5A) and these mice did not experience severe weight loss or show clinical signs of disease (Fig 5B).
-In contrast, infection of mice harboring microbiomes susceptible to severe disease (DA00578, Fig 1B) with strain 3
-
-**Results to be written**
-
-2.	Propensity for severe CDI is community-dependent and strain-independent
-a.	Figure 5:  Infection of mice with different C. difficile strains.
+In contrast, infection of mice harboring microbiomes susceptible to severe disease (DA00578, Fig 1B) with strain 395 resulted in colonization and severe weight loss by two days post-infection (Fig 5A, 5B).
+Strain 395 is similarly virulent as strain 431, and shows a similar infection phenotype in these mice.
+These preliminary results suggest that human-associated microbiomes may be more predictive of disease severity than the _C.
+difficile_ strain.
 
 ### Discussion
-•	Restate results,
-•	caveats about mouse weights 
+
+Study of human-associated microbiomes and their interactions with _C.
+difficile_ has been limited by the lack of an appropriate model and available samples.
+Here we colonized 16 groups of mice with human donor stool from patients with different clinical profiles or healthy volunteers.
+This created a diverse set of human-associated microbiomes in the mice that allowed for testing _C.
+difficile_ infection dynamics and community interactions.
+Our results showed that both healthy and diarrheal-associated human communities were susceptible to severe disease.
+Additionally, our results confirmed our and others previous studies that proved that resistance to _C.
+difficile_ disease is dependent on the entire microbial community and not just one or two species (cite alyx).
+Further, we were able to learn classification models that identified specific OTUs within susceptible communities that predispose mice to infection.
+By subsequently infecting susceptible or resistant communities with varying strains of _C.
+difficile_ we confirmed that the initial state of the human-associated microbiome is more predictive of disease severity than the strain used.
+
+Here we were able to colonize germ-free mice with a variety of human stool communities from diarrheal or healthy people.
+After allowing these communities to colonize the mice for 2 weeks, they were challenged with _C.
+difficile_.
+It is important to note that the mouse gut environment is different from that of the human, and as such exerts a selective pressure on the human stool microbial community that stabilizes over the 2 weeks.
+Thus these mice cannot be considered "humanized" but rather containing human-associated microbes.
+Interestingly, clinical status was not associated with proclivity to severe _C.
+difficile_ infection as some healthy donor mice were susceptible.
+A surprising finding was that none of the human-associated mouse communities were completely resistant to colonization by _C.
+difficile_.
+This is in contrast to numerous reports of mouse microbiome studies (cite) and the knowledge that in general healthy humans without gut perturbations are resistant to _C.
+difficile_ infection (cite).
+On the other hand, it is known that humans, especially health care employees, can be asymptomatic carriers of _C.
+difficile_ (though at what CFU?!).
+This may explain the fact that the majority of our groups of mice were colonized by _C.
+difficile_ but did not display outward signs of illness or infection-associated weight loss (Fig 2).
+
 •	No donors were colonization resistant, discuss donor differences 
 severity, asymptomatic colonzation discussion? 
-•	Discuss prediction methods and outcomes
-•	Discuss potential mechanisms for interesting OTUs
+•	Discuss prediction methods and outcomes - Nick
+•	Discuss potential mechanisms for interesting OTUs - Nick
 •	Discuss different strain results
@@
 Donor stool for this study was selected from one patient that tested positive for _C.
-difficile_ infection and four patients that presented with diarrhea that tested negative for _C.
+difficile_ infection and five patients that presented with diarrhea and tested negative for _C.
 difficile_.
-Healthy donor stool was collected from volunteer donors living in the Ann Arbor, MI area.
+Healthy donor stool was collected from adult volunteers living in the Ann Arbor, MI area.
+100ul of stool slurry was gavaged into each mouse.
+
+**_Clostridium difficile_ and sample processing** For all experiments, 100 spores of _Clostridium difficile_ were used to orally infect mice as described previously (cite alyx).
+The _C.
+difficile_ strains used were acquired from spore stocks of (WHERE) and comprised of strains 431, 395, 299 and 468 (cite?).
+After gavage, the remaining inoculum was diluted and plated to confirm infection dose.
+Bacteria were cultured anaerobically on TCCFA plates and incubated at 37C.
+During infection, two fresh stool pellets were collected from each mouse daily.
+One pellet was resuspended in anaerobic phosphate-buffered saline and diluted and plated on TCCFA plates to enumerate _C.
+difficile_ CFU.
+The other pellet was frozen at -80C until the end of the experiment.
+
+**DNA isolation and sequencing** Bacterial DNA was isolated from banked donor stool and each mouse stool pellet using the MO-BIO PowerSoil DNA isolation kit. This DNA was used as template for
 

### 2016-11-16 code/CdGF_AUCRF.Rmd
onemouse otus with 1/2sources out models

@@
 
+#### leave-one-out
+LOO_probs <- NULL
+LOO_probs <- sapply(1:nrow(aucrf_data), function(i){
+ train <- aucrf_data[-i,]
+  test <- aucrf_data[i,]
+  set.seed(seed)
+  temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
+  LOO_probs[i] <- predict(temp_model$RFopt, test, type='prob')[,2]
+})
+LOO_roc <- roc(aucrf_data$Euth_Early~LOO_probs)
+
 #### 10-fold Cross validated - 10 repetitions
@@
 
-#### One_mouse model
+#### One_mouse trainset
 Predict_om_df <- select(Predict_df, 
                         Euth_Early, human_source, mouse_id, contains('Otu'))
-om_import_otus <- data.frame(otu = colnames(select(Predict_om_df, contains("Otu"))),
-                             importance = NA)
+om_import_otus <- data.frame(otu = colnames(select(Predict_om_df, contains("Otu"))))
 om_aucs <- c()
 om_all <- c()
+om_probs <- c()
 for(j in 1:iters){
   set.seed(j)
-  mouse_set <- Predict_om_df %>% 
+  train <- Predict_om_df %>% 
     group_by(human_source) %>% 
     sample_n(1) %>% ungroup() 
-  om_probs <- c()
-  om_cage_import <- c()
-  for(n in unique(as.character(mouse_set$human_source))){
-    train <- mouse_set[mouse_set$human_source != n, ]
-    test <- Predict_om_df %>% 
-      filter(human_source %in% n)
+  test <- filter(Predict_om_df, !mouse_id %in% train$mouse_id)
   set.seed(seed)
-    temp_model <- AUCRF(Euth_Early~., data=select(train, -mouse_id, -human_source)
-                      , pdel=0.2, ntree=500, ranking = 'MDA')
+  temp_model <- AUCRF(Euth_Early~., data=select(train, -mouse_id, -human_source), 
+                      pdel=0.1, ntree=1000, ranking = 'MDA')
   om_probs <- rbind(om_probs,
@@
                                pred = predict(temp_model$RFopt, test, type='prob')[,2]))
-    om_cage_import <- c(om_cage_import, temp_model$Xopt)
-  }
   om_roc <- roc(om_probs$obs~om_probs$pred)
+  selected_var <- temp_model$ranking[temp_model$Xopt]
   om_import_otus <- merge(om_import_otus,
-                          data.frame(otu = names(table(om_cage_import)),
-                            importance = as.numeric(table(om_cage_import))*om_roc$auc),
-                          by = 'otu', all.x = T, suffixes = c('',round(om_roc$auc, 3)))
+                          data.frame(otu = names(selected_var), 
+                                     rank = rank(selected_var)/max(rank(selected_var))),
+                          by = 'otu', all.x = T, 
+                          suffixes = c('',paste0('_auc_', round(om_roc$auc, 3))))
   om_all <- rbind(om_all, om_probs)
@@
 
-raw_importance <- data.frame(om_import_otus[ , -c(1,2)], row.names = om_import_otus$otu)
-
+raw_importance <- data.frame(om_import_otus[ , -c(1)], row.names = om_import_otus$otu)
 importance_otus <-   apply(raw_importance, 1, sum, na.rm = T)
 importance_otus <- importance_otus[importance_otus>0]
-importance_otus <- importance_otus[importance_otus > median(om_aucs)*150/2]
+importance_otus <- importance_otus[importance_otus > mean(importance_otus)]
+om_import_otus <- names(sort(importance_otus, decreasing = T))
 
-### rank importance
-importance_rank <- apply(raw_importance, 2, function(x) rank(x, na.last = T))
-raw_rank <- apply(importance_rank, 1, function(x) sum(x, na.rm = T))
-final_rank <- sort(raw_rank[raw_rank >=  ncol(importance_rank)*sum(importance_rank[ , 1])/nrow(importance_rank)], decreasing = T)
-
-#### leave-one-out
-LOO_probs <- NULL
-LOO_probs <- sapply(1:nrow(aucrf_data), function(i){
- train <- aucrf_data[-i,]
-  test <- aucrf_data[i,]
-  set.seed(seed)
-  temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
-  LOO_probs[i] <- predict(temp_model$RFopt, test, type='prob')[,2]
-})
-LOO_roc <- roc(aucrf_data$Euth_Early~LOO_probs)
-
-#### leave-one-cage-out
+#### leave-one-source-out
 LOcO_df <- Predict_df %>% 
-                select(Euth_Early, cage_id, one_of(euth_otu_feat))
-cage_list <- as.character(unique(LOcO_df$cage_id))
+                select(Euth_Early, human_source, one_of(om_import_otus))
+source_list <- as.character(unique(LOcO_df$human_source))
 
@@
 LOcO_rf <- c()
-for(i in 1:length(cage_list)){
-  by_cage <- LOcO_df$cage_id == cage_list[i]
-  LOcO_predict <- select(LOcO_df, -cage_id)
-  train <- LOcO_predict[!by_cage,]
-  test <- LOcO_predict[by_cage,]
+for(i in 1:length(source_list)){
+  source <- LOcO_df$human_source == source_list[i]
+  LOcO_predict <- select(LOcO_df, -human_source)
+  train <- LOcO_predict[!source,]
+  test <- LOcO_predict[source,]
   set.seed(seed)
@@
   LOcO_rf[[i]] <- temp_model$RFopt
-  LOcO_probs[LOcO_df$cage_id == cage_list[i]] <- predict(temp_model$RFopt, test, type='prob')[,2]
+  LOcO_probs[LOcO_df$human_source == source_list[i]] <- predict(temp_model$RFopt, test, type='prob')[,2]
 }
@@
 LOdO_df <- Predict_df %>% 
-                select(Euth_Early, human_source, one_of(euth_otu_feat))
+                select(Euth_Early, human_source, one_of(om_import_otus))
 donor_list <- as.character(unique(LOdO_df$human_source))
@@
 #### plot rocs
+# all mice
 par(mar=c(4,4,1,1))
 plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='')
-plot(om_roc, col='red', lwd=2, add=T, lty=1)
-plot(cv10f_roc, col='gray', lwd=2, add=T, lty=1)
-plot(LOO_roc, col='blue', lwd=2, add=T, lty=2)
-plot(LOcO_roc, col='purple', lwd=2, add=T, lty=3)
-plot(LOdO_roc, col='red', lwd=2, add=T, lty=3)
+plot(otu_euth_roc, col='gray', lwd=2, add=T, lty=2)
+plot(cv10f_roc, col='black', lwd=2, add=T, lty=1)
+plot(LOO_roc, col='blue', lwd=2, add=T, lty=3)
 mtext(side=2, text="Sensitivity", line=2.5)
@@
                                sprintf('Leave-1 mouse-out (AUC: %.3g)',LOO_roc$auc),
-                               sprintf('Leave-1 cage-out (AUC: %.3g)',LOcO_roc$auc),
-                               sprintf('Leave-2 cages-out (AUC: %.3g)',LOdO_roc$auc),
                                sprintf('10-fold CV (AUC: %.3g)',cv10f_roc$auc) #,
@@
 #                               sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
-),lty=c(1,2,2,2,1,0,0), lwd=2, col=c('black','blue','purple','red','gray'), bty='n')
+),lty=c(2,1,3), lwd=2, col=c('gray','black','blue'), bty='n')
+
+#remove cage bias plots
+par(mar=c(4,4,1,1))
+plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='')
+plot(om_roc, col='black', lwd=2, add=T, lty=1)
+plot(LOcO_roc, col='purple', lwd=2, add=T, lty=3)
+plot(LOdO_roc, col='red', lwd=2, add=T, lty=3)
+plot(om_LOcO_roc, col='blue', lwd=2, add=T, lty=2)
+mtext(side=2, text="Sensitivity", line=2.5)
+mtext(side=1, text="Specificity", line=2.5)
+legend(0.3, 0.1, legend=c(sprintf('One-mouse (AUC: %.3g)',om_roc$auc),
+                           sprintf('Leave-1 source-out (AUC: %.3g)',LOcO_roc$auc),
+                           sprintf('Leave-2 sources-out (AUC: %.3g)',LOdO_roc$auc),
+                           sprintf('One-mouse-Leave-1 source-out (AUC: %.3g)',om_LOcO_roc$auc)),
+       lty=c(1,3,3,2), lwd=2, col=c('black','purple','red','blue'), bty='n')
 ```

### 2016-11-16 code/CdGF_AUCRF.Rmd
add one mouse/source-leave one source out

@@
 
+#### One_mouse-Leave one out model 
+om_loo_df <- select(Predict_df,  
+                    Euth_Early, human_source, mouse_id, one_of(om_import_otus)) 
+om_loo_otus <- data.frame(otu = om_import_otus)
+
+om_loo_aucs <- c() 
+om_loo_all <- c() 
+for(j in 1:iters){ 
+  set.seed(j) 
+  mouse_set <- om_loo_df %>%  
+    group_by(human_source) %>%  
+    sample_n(1) %>% ungroup()  
+  om_loo_probs <- c() 
+  om_source_import <- data.frame(otu = om_import_otus) 
+  for(n in unique(as.character(mouse_set$human_source))){ 
+    train <- mouse_set[mouse_set$human_source != n, ] 
+    test <- om_loo_df %>%  
+      filter(human_source %in% n) 
+    set.seed(seed) 
+    temp_model <- AUCRF(Euth_Early~., data=select(train, -mouse_id, -human_source), 
+                        pdel=0.99, ntree=500, ranking = 'MDA') 
+    om_loo_probs <- rbind(om_loo_probs, 
+                          data.frame(obs = test$Euth_Early, 
+                                     pred = predict(temp_model$RFopt, 
+                                                    test, type='prob')[,2])) 
+    om_source_import <- merge(om_source_import,
+                            data.frame(otu = names(temp_model$ranking), 
+                                       rank = rank(temp_model$ranking)/
+                                              max(rank(temp_model$ranking))),
+                            by = 'otu', all.x = T, 
+                            suffixes = c('',paste0('_', n)))
+  } 
+  om_loo_roc <- roc(om_loo_probs$obs~om_loo_probs$pred) 
+  om_loo_otus <- merge(om_loo_otus, om_source_import,
+                       by = 'otu', all.x = T, suffixes = c('',paste0('_',j)))
+  om_loo_all <- rbind(om_loo_all, om_loo_probs) 
+  om_loo_aucs[j] <- om_loo_roc$auc 
+} 
+om_loo_roc <- roc(om_loo_all$obs~om_loo_all$pred)
+
+raw_importance <- data.frame(om_source_import[ , -c(1,2)], row.names = om_source_import$otu)
+
+om_loo_import_otus <- apply(raw_importance, 1, sum, na.rm = T)
+om_loo_import_otus <- om_loo_import_otus[om_loo_import_otus > mean(om_loo_import_otus)]
+important_otus_om_loo <- names(sort(om_loo_import_otus, decreasing = T))
+
 #### plot rocs
@@
 plot(LOdO_roc, col='red', lwd=2, add=T, lty=3)
-plot(om_LOcO_roc, col='blue', lwd=2, add=T, lty=2)
+plot(om_loo_roc, col='blue', lwd=2, add=T, lty=2)
 mtext(side=2, text="Sensitivity", line=2.5)
 mtext(side=1, text="Specificity", line=2.5)
-legend(0.3, 0.1, legend=c(sprintf('One-mouse (AUC: %.3g)',om_roc$auc),
+legend(0.4, 0.125, legend=c(sprintf('One-mouse (AUC: %.3g)',om_roc$auc),
                            sprintf('Leave-1 source-out (AUC: %.3g)',LOcO_roc$auc),
                            sprintf('Leave-2 sources-out (AUC: %.3g)',LOdO_roc$auc),
-                           sprintf('One-mouse-Leave-1 source-out (AUC: %.3g)',om_LOcO_roc$auc)),
+                           sprintf('One-mouse-Leave-1 source-out (AUC: %.3g)',om_loo_roc$auc)),
        lty=c(1,3,3,2), lwd=2, col=c('black','purple','red','blue'), bty='n')
@@
 ```{r rel abund partial depend, fig.width=10, fig.height=10}
-layout(matrix(c(1:9), ncol=3, byrow=TRUE))
 
-for (i in seq_along(euth_otu_feat)) {
-  partial_otu <- partialPlot(rf_early_aucrf$RFopt, aucrf_data, euth_otu_feat[i], xlab='',
-                main=paste(get_tax(row_list = euth_otu_feat)$tax_label[i]))
-points(Predict_early_euth_df[Predict_early_euth_df$Euth_Early == 1 ,euth_otu_feat[i]], 
+partialplot_df <- Predict_early_euth_df[,c('Euth_Early',important_otus_om_loo)]
+pp_aucrf <- AUCRF(Euth_Early ~ ., data = partialplot_df,
+                  ntree = 500, pdel = 0.99)
+
+cor_features <- sapply(as.character(important_otus_om_loo), 
+                       function(i)rownames(rel_abund_cor)[rel_abund_cor[,i] > cor_cutoff])
+
+layout(matrix(c(1:12), ncol=3, byrow=TRUE))
+
+for (i in seq_along(important_otus_om_loo)) {
+  partial_otu <- partialPlot(pp_aucrf$RFopt, partialplot_df, important_otus_om_loo[i], xlab='',
+                main=NULL)
+  if(length(cor_features[[i]]) > 1){
+      title(main = paste0('Unidentifiable Feature (', 
+                          get_tax(row_list = important_otus_om_loo[i])$tax_label,
+                          ' Shown)'))
+      mtext(side = 1, line = 3, cex = 0.6,
+        paste('Correlated features: \n', 
+                  get_tax(row_list = cor_features[[i]])$tax_label, collapse = ','))
+  } else {
+      title(paste(get_tax(row_list = important_otus_om_loo[i])$tax_label))
+  }
+points(partialplot_df[partialplot_df$Euth_Early == 1 ,important_otus_om_loo[i]], 
        jitter(rep(0.25*(diff(range(partial_otu$y))) + min(partial_otu$y), 
-                  sum(Predict_early_euth_df$Euth_Early == 1)), 
+                  sum(partialplot_df$Euth_Early == 1)), 
               amount = (0.1*(diff(range(partial_otu$y))))), 
               pch = 21, bg = 'red')
-points(Predict_early_euth_df[Predict_early_euth_df$Euth_Early == 0 ,euth_otu_feat[i]], 
+points(partialplot_df[partialplot_df$Euth_Early == 0 ,important_otus_om_loo[i]], 
        jitter(rep(0.75*(diff(range(partial_otu$y))) + min(partial_otu$y), 
-                  sum(Predict_early_euth_df$Euth_Early == 0)), 
+                  sum(partialplot_df$Euth_Early == 0)), 
               amount = (0.1*(diff(range(partial_otu$y))))), 

### 2016-11-16 code/CdGF_AUCRF.Rmd
refine text of rmd

@@
 
-##Can we predict moribundity with the community structure at Day 0?
+### Can we predict moribundity with the community structure at Day 0?
 
@@
 important_otus_om_loo <- names(sort(om_loo_import_otus, decreasing = T))
+```
 
+#### RF with all mice input
+
+```{r All data plt, fig.width=8, fig.height=8, results= 'hide'}
 #### plot rocs
@@
 ),lty=c(2,1,3), lwd=2, col=c('gray','black','blue'), bty='n')
+```
 
+#### RF with features from One_mouse_per_source
+
+```{r one_mouse plots, fig.width=8, fig.height=8, results= 'hide'}
 #remove cage bias plots
@@
 mtext(side=1, text="Specificity", line=2.5)
-legend(0.4, 0.125, legend=c(sprintf('One-mouse (AUC: %.3g)',om_roc$auc),
+legend('bottomright', legend=c(sprintf('One-mouse (AUC: %.3g)',om_roc$auc),
                            sprintf('Leave-1 source-out (AUC: %.3g)',LOcO_roc$auc),
@@
 ### Partial Dependency with Relative Abundance Overlayed Plots
-###### Using optimal RF model generated from whole dataset
-###### Plot labels contain MDA of each OTU
+###### Using features selected by one-mouse-per-source model
 
@@
 
-Would adding initial increase/slope of CFU or day 1 CFU help predict?
-
-* Adding CFU decreases AUC and improves Random model, likely due to samples lost due to lack of cfu data which are biased towards the severe samples
+#### Notes
+* CFU data doesnt help model, likely due to samples lost due to lack of cfu data which are biased towards the severe samples  
     - Day 1 CFU loses 2 samples  [AUCRF model with Day 1 CFU](~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/CdGF_AUCRF_withCFU_day1.html)  
     - Day 2 cfu loses 14 samples, 10 are severe  [AUCRF model with Day 2   CFU](~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/CdGF_AUCRF_withCFU_day2.html)  
-
-Would it help to add community information such as diversity or enterotype?
-
-* When adding enterotype, diversity, and CFU, invsimpson and cfu appear as features, but when cfu is removed, invsimpson is no longer a predictor
-
-But how to show since model is already approaching 1?
-
-* Rerun AUCRF with added features and see if they replace any OTUs
-
-Leave 1 cage out RF with all otus gives and AUC of 0.70, but using the reduced features selected by AUCRF on the whole dataset gives an AUC of 0.80
-
-To remove any cage effect RF could be picking out instead of the actual outcome, I first run RF on all OTUs to predict cage and separately human source.
-Then select the OTUS that are necessary for the lowest error rate (I went slightly above) from both cage and human souce RFs.
-Remove those OTUs from the OTU data frame and predict Euthanization with AUCRF
-
-* Of note, I tried a few different seeds, and one otu (Turicibacter OTU 120) was identified as a classifier of cage/inocula but wasnt in others and was identified as a classifier of Euthanization.
-Also we occasionally lose OTU 50 as a classifier
-
-####### Model run with both relative and absolute abundances and no significant difference in their output
+* Diversity/Enterotype don't help model - When adding enterotype, diversity, and CFU, invsimpson and cfu appear as features, but when cfu is removed, invsimpson is no longer a predictor  
+* Model run with both relative and absolute abundances and no significant difference in their output
 

### 2016-11-17 code/CdGF_AUCRF.Rmd

--- 2016-11-16 code/CdGF_AUCRF.Rmd
@@
 
-#### leave-one-out
-LOO_probs <- NULL
-LOO_probs <- sapply(1:nrow(aucrf_data), function(i){
- train <- aucrf_data[-i,]
-  test <- aucrf_data[i,]
-  set.seed(seed)
-  temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
-  LOO_probs[i] <- predict(temp_model$RFopt, test, type='prob')[,2]
-})
-LOO_roc <- roc(aucrf_data$Euth_Early~LOO_probs)
-
 #### 10-fold Cross validated - 10 repetitions
@@
 
-#### One_mouse trainset
-Predict_om_df <- select(Predict_df, 
-                        Euth_Early, human_source, mouse_id, contains('Otu'))
-om_import_otus <- data.frame(otu = colnames(select(Predict_om_df, contains("Otu"))))
-om_aucs <- c()
-om_all <- c()
-om_probs <- c()
-for(j in 1:iters){
-  set.seed(j)
-  train <- Predict_om_df %>% 
-    group_by(human_source) %>% 
-    sample_n(1) %>% ungroup() 
-  test <- filter(Predict_om_df, !mouse_id %in% train$mouse_id)
+#### leave-one-out
+LOO_probs <- NULL
+LOO_probs <- sapply(1:nrow(aucrf_data), function(i){
+ train <- aucrf_data[-i,]
+  test <- aucrf_data[i,]
   set.seed(seed)
-  temp_model <- AUCRF(Euth_Early~., data=select(train, -mouse_id, -human_source), 
-                      pdel=0.1, ntree=1000, ranking = 'MDA')
-  om_probs <- rbind(om_probs,
-                    data.frame(obs = test$Euth_Early,
-                               pred = predict(temp_model$RFopt, test, type='prob')[,2]))
-  om_roc <- roc(om_probs$obs~om_probs$pred)
-  selected_var <- temp_model$ranking[temp_model$Xopt]
-  om_import_otus <- merge(om_import_otus,
-                          data.frame(otu = names(selected_var), 
-                                     rank = rank(selected_var)/max(rank(selected_var))),
-                          by = 'otu', all.x = T, 
-                          suffixes = c('',paste0('_auc_', round(om_roc$auc, 3))))
-  om_all <- rbind(om_all, om_probs)
-  om_aucs[j] <- om_roc$auc
-}
-om_roc <- roc(om_all$obs~om_all$pred)
+  temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
+  LOO_probs[i] <- predict(temp_model$RFopt, test, type='prob')[,2]
+})
+LOO_roc <- roc(aucrf_data$Euth_Early~LOO_probs)
 
-raw_importance <- data.frame(om_import_otus[ , -c(1)], row.names = om_import_otus$otu)
-importance_otus <-   apply(raw_importance, 1, sum, na.rm = T)
-importance_otus <- importance_otus[importance_otus>0]
-importance_otus <- importance_otus[importance_otus > mean(importance_otus)]
-om_import_otus <- names(sort(importance_otus, decreasing = T))
-
-#### leave-one-source-out
+#### leave-one-cage-out
 LOcO_df <- Predict_df %>% 
-                select(Euth_Early, human_source, one_of(om_import_otus))
-source_list <- as.character(unique(LOcO_df$human_source))
+                select(Euth_Early, cage_id, one_of(euth_otu_feat))
+cage_list <- as.character(unique(LOcO_df$cage_id))
 
@@
 LOcO_rf <- c()
-for(i in 1:length(source_list)){
-  source <- LOcO_df$human_source == source_list[i]
-  LOcO_predict <- select(LOcO_df, -human_source)
-  train <- LOcO_predict[!source,]
-  test <- LOcO_predict[source,]
+for(i in 1:length(cage_list)){
+  by_cage <- LOcO_df$cage_id == cage_list[i]
+  LOcO_predict <- select(LOcO_df, -cage_id)
+  train <- LOcO_predict[!by_cage,]
+  test <- LOcO_predict[by_cage,]
   set.seed(seed)
@@
   LOcO_rf[[i]] <- temp_model$RFopt
-  LOcO_probs[LOcO_df$human_source == source_list[i]] <- predict(temp_model$RFopt, test, type='prob')[,2]
+  LOcO_probs[LOcO_df$cage_id == cage_list[i]] <- predict(temp_model$RFopt, test, type='prob')[,2]
 }
@@
 LOdO_df <- Predict_df %>% 
-                select(Euth_Early, human_source, one_of(om_import_otus))
+                select(Euth_Early, human_source, one_of(euth_otu_feat))
 donor_list <- as.character(unique(LOdO_df$human_source))
@@
 #### plot rocs
-# all mice
 par(mar=c(4,4,1,1))
 plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='')
-plot(otu_euth_roc, col='gray', lwd=2, add=T, lty=2)
-plot(cv10f_roc, col='black', lwd=2, add=T, lty=1)
-plot(LOO_roc, col='blue', lwd=2, add=T, lty=3)
+plot(otu_euth_roc, col='black', lwd=2, add=T, lty=1)
+plot(cv10f_roc, col='gray', lwd=2, add=T, lty=1)
+plot(LOO_roc, col='blue', lwd=2, add=T, lty=2)
+plot(LOcO_roc, col='purple', lwd=2, add=T, lty=3)
+plot(LOdO_roc, col='red', lwd=2, add=T, lty=3)
 mtext(side=2, text="Sensitivity", line=2.5)
@@
                                sprintf('Leave-1 mouse-out (AUC: %.3g)',LOO_roc$auc),
+                               sprintf('Leave-1 cage-out (AUC: %.3g)',LOcO_roc$auc),
+                               sprintf('Leave-2 cages-out (AUC: %.3g)',LOdO_roc$auc),
                                sprintf('10-fold CV (AUC: %.3g)',cv10f_roc$auc) #,
@@
 #                               sprintf('OOB vs 10-fold CV: p=%.2g', roc.test(otu_euth_roc,cv10f_roc)$p.value)
-),lty=c(2,1,3), lwd=2, col=c('gray','black','blue'), bty='n')
-
-#remove cage bias plots
-par(mar=c(4,4,1,1))
-plot(c(1,0),c(0,1), type='l', lty=3, xlim=c(1.01,0), ylim=c(-0.01,1.01), xaxs='i', yaxs='i', ylab='', xlab='')
-plot(om_roc, col='black', lwd=2, add=T, lty=1)
-plot(LOcO_roc, col='purple', lwd=2, add=T, lty=3)
-plot(LOdO_roc, col='red', lwd=2, add=T, lty=3)
-plot(om_LOcO_roc, col='blue', lwd=2, add=T, lty=2)
-mtext(side=2, text="Sensitivity", line=2.5)
-mtext(side=1, text="Specificity", line=2.5)
-legend(0.3, 0.1, legend=c(sprintf('One-mouse (AUC: %.3g)',om_roc$auc),
-                           sprintf('Leave-1 source-out (AUC: %.3g)',LOcO_roc$auc),
-                           sprintf('Leave-2 sources-out (AUC: %.3g)',LOdO_roc$auc),
-                           sprintf('One-mouse-Leave-1 source-out (AUC: %.3g)',om_LOcO_roc$auc)),
-       lty=c(1,3,3,2), lwd=2, col=c('black','purple','red','blue'), bty='n')
+),lty=c(1,2,2,2,1,0,0), lwd=2, col=c('black','blue','purple','red','gray'), bty='n')
 ```

### 2016-11-18 submission/Manuscript.Rmd

@@
 
-Kaitlin J. Flynn^1^, Nicholas Lesniak^1^, Alyxandria M. Schubert^2^, Hamide Sinani^?^, and Patrick D. Schloss^1$\dagger$^
+Kaitlin J. Flynn^1+^, Nicholas Lesniak^1+^, Alyxandria M. Schubert^2^, Hamide Sinani^?^, and Patrick D. Schloss^1$\dagger$^
 
@@
 $\dagger$ Corresponding author: pschloss@umich.edu
+
+\+ These authors contributed equally to this work
 
@@
 
-**also want something in the intro about severity maybe, and asymptomatic colonization. or in discussion?**
-
 ### Results 
@@
 
+**Nick will be updating the next two sections**
 **Microbes present in the gut prior to infection are predictive of _C.
@@
 On the other hand, it is known that humans, especially health care employees, can be asymptomatic carriers of _C.
-difficile_ (though at what CFU?!).
+difficile_ (Furyura-Kanamari).
 This may explain the fact that the majority of our groups of mice were colonized by _C.
@@
 
-•	No donors were colonization resistant, discuss donor differences 
-severity, asymptomatic colonzation discussion? 
-•	Discuss prediction methods and outcomes - Nick
+•	Discuss prediction methods and outcomes - Nick - we have bugs that predict severity, but do we have bugs that predict mild/asymptomatic? probably, could discuss here
 •	Discuss potential mechanisms for interesting OTUs - Nick
-•	Discuss different strain results
-•	Future work blah blah
+
+Finally, our pilot experiments testing the efficiency of infection with different _C.
+difficile_ strains have allowed for testing community versus pathogen factors of infection.
+Our results are consistent with previous reports that the composition of the gut microbiome determines if an individual is successfully infected or not (cite Alyx and others).
+_C.
+difficile_ strains are known to vary in pathogenecity, primarily via the amount of toxin that is produced during infection (cite).
+In these pilot experiments, we challenged a susceptible community with another pathogenic strain of _C.
+difficile_ to confirm that the susceptibility lied in the community and not in the specific challenge strain.
+In contrast, we exposed a resistant/asymptomatic community to a relatively weak strain of _C.
+difficile_ and found that the community could be colonized by this one as well-- but not cause disease.
+This suggests that this colony is prone to colonization by any _C.
+difficile_ strain.
+Further experiments will test if infecting susceptible communities with weak strains allows the mice to survive.
+It has been recently reported that carriage of a non-toxigenic strain of _C.
+difficile_ can prevent additional colonization by a pathogenic strain, suggesting some role of niche-filling or immunity this provides (cite cite).
+It would be interesting to see if we could make a susceptible community resistant to infection and disease by a pathogenic _C.
+difficle_ strain in our human-associated microbiome mice.
+Further work with human-associated communities in mice and mathematical modeling will allow for uncovering of complex human-associated disease processes and dynamics as they relate to human disease.
 
@@
 
-**DNA isolation and sequencing** Bacterial DNA was isolated from banked donor stool and each mouse stool pellet using the MO-BIO PowerSoil DNA isolation kit. This DNA was used as template for
+**DNA isolation and sequencing** Bacterial DNA was isolated from banked donor stool and each mouse stool pellet using the MO-BIO PowerSoil DNA isolation kit.
+This DNA was used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described (cite Jim paper).
 
-•	Bacteria/plating 
-•	Sequencing
-•	Data analysis, code availability
-•	Machine learning models
+•	Sequence curation/mothur, data analysis, code + availability of data(SRA) and code (Github) - Nick
 
 ##Acknowledgments## 
-Lab, sequencing core, Jhansi
+We would like to thank members of the Schloss lab and Jhansi Leslie for analysis advice and the Microbial Systems Laboratory at the University of Michigan for sequencing assistance.
 

### 2016-11-18 submission/Manuscript.tex

@@
 
-Kaitlin J. Flynn\textsuperscript{1}, Nicholas
-Lesniak\textsuperscript{1}, Alyxandria M. Schubert\textsuperscript{2},
+Kaitlin J. Flynn\textsuperscript{1+}, Nicholas
+Lesniak\textsuperscript{1+}, Alyxandria M. Schubert\textsuperscript{2},
 Hamide Sinani\textsuperscript{?}, and Patrick D.
@@
 \href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}
+
++ These authors contributed equally to this work
 
@@
 
+\textbf{also want something in the intro about severity maybe, and
+asymptomatic colonization. or in discussion?}
+
 \subsubsection{Results}\label{results}
@@
 
-\textbf{The microbiome predicts \emph{C. difficile} severity.}
+\textbf{The microbiome predicts \emph{C. difficile} infection severity.}
+As \emph{C. difficile} colonization levels can be predicted by the
+microbiome composition on day 0, we postulated that severity of disease
+could also be predicted from day 0. Using weight loss as a proxy for
+severity, we built a Random Forest classification model based on
+relative abundance to determine which community members were associated
+with either mild or severe disease. We optimized the model using the
+AUC-RF algorithm to generate a Reciever Operating Curve (ROC) with
+maximal Area Under the Curve (AUC) (cite Calle 2011), as used previously
+in our group to classify colon cancer outcomes (cite Niel 2015 gen med).
+The optimized model used X number of OTUs and identified several that
+predicted increased weightloss and subsequent disease severity (Fig 4A,
+4B). The populations that were most commonly found in mice that
+succumbed to \emph{C. difficile} infection were members of the
+\emph{Clostridiales}, \emph{Bacilliales}, \emph{Streptococcus}
+(Lactobacilles?) and \emph{Bacteriodes}. (need more OTUs here, or number
+that were used for model and then group by rank). The AUC was calculated
+using the Out-Of-Bag (OOB) error for each sample (Fig 4A).
+Cross-validation using 10-fold, leave-one-mouse-out and
+leave-one-cage-out resulted in AUCs not significantly different from the
+OOB (Fig 4A). This model reveals that the levels of X OTUs on day 0 can
+predict disease outcome in mice with human-associated microbiomes.
+(probably also going to want a supplemental table for predictive OTUs
+for this and Fig 3)
 
-\textbf{Results to be written}
-
-\begin{enumerate}
-\def\labelenumi{\alph{enumi}.}
-\setcounter{enumi}{1}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  Figure 4: Random Forest predicts CDI severity, predictive OTUs
-\end{enumerate}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{1}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  Propensity for severe CDI is community-dependent and
-  strain-independent
-\end{enumerate}
-
-\begin{enumerate}
-\def\labelenumi{\alph{enumi}.}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  Figure 5: Infection of mice with different C. difficile strains.
-\end{enumerate}
+\textbf{Susceptibility to infection is community-dependent.} \emph{C.
+difficile} strains are known to vary in disease phenotypes based both on
+pathogenicity factors of the strain and host microbiome composition
+(cite someone). To test the susceptibility of human-derived microbiomes
+to \emph{C. difficile} infection, two donor-mouse microbiomes that were
+previously shown to be resistant to \emph{C. difficile} strain 431
+induced weight loss (Fig 2B) (DA00396, DA00430) and one that was
+suscpetible to severe disease (DA00578) were independently challenged
+with one of three unique \emph{C. difficile} strains (table or reference
+for toxin prod, etc). As done previously, \emph{C. difficile} stool CFU
+was enumerated daily and mouse weight-loss tracked. Infection of mice
+with resistant communities with virulent strain 458 or the weaker strain
+299 resulted in colonization to levels near that of 431 (Fig 5A) and
+these mice did not experience severe weight loss or show clinical signs
+of disease (Fig 5B). In contrast, infection of mice harboring
+microbiomes susceptible to severe disease (DA00578, Fig 1B) with strain
+395 resulted in colonization and severe weight loss by two days
+post-infection (Fig 5A, 5B). Strain 395 is similarly virulent as strain
+431, and shows a similar infection phenotype in these mice. These
+preliminary results suggest that human-associated microbiomes may be
+more predictive of disease severity than the \emph{C. difficile} strain.
 
@@
 
-• Restate results, • caveats about mouse weights • No donors were
-colonization resistant, discuss donor differences • Discuss prediction
-methods and outcomes • Discuss potential mechanisms for interesting OTUs
-• Discuss different strain results • Future work blah blah
+Study of human-associated microbiomes and their interactions with
+\emph{C. difficile} has been limited by the lack of an appropriate model
+and available samples. Here we colonized 16 groups of mice with human
+donor stool from patients with different clinical profiles or healthy
+volunteers. This created a diverse set of human-associated microbiomes
+in the mice that allowed for testing \emph{C. difficile} infection
+dynamics and community interactions. Our results showed that both
+healthy and diarrheal-associated human communities were susceptible to
+severe disease. Additionally, our results confirmed our and others
+previous studies that proved that resistance to \emph{C. difficile}
+disease is dependent on the entire microbial community and not just one
+or two species (cite alyx). Further, we were able to learn
+classification models that identified specific OTUs within susceptible
+communities that predispose mice to infection. By subsequently infecting
+susceptible or resistant communities with varying strains of \emph{C.
+difficile} we confirmed that the initial state of the human-associated
+microbiome is more predictive of disease severity than the strain used.
+
+Here we were able to colonize germ-free mice with a variety of human
+stool communities from diarrheal or healthy people. After allowing these
+communities to colonize the mice for 2 weeks, they were challenged with
+\emph{C. difficile}. It is important to note that the mouse gut
+environment is different from that of the human, and as such exerts a
+selective pressure on the human stool microbial community that
+stabilizes over the 2 weeks. Thus these mice cannot be considered
+``humanized'' but rather containing human-associated microbes.
+Interestingly, clinical status was not associated with proclivity to
+severe \emph{C. difficile} infection as some healthy donor mice were
+susceptible. A surprising finding was that none of the human-associated
+mouse communities were completely resistant to colonization by \emph{C.
+difficile}. This is in contrast to numerous reports of mouse microbiome
+studies (cite) and the knowledge that in general healthy humans without
+gut perturbations are resistant to \emph{C. difficile} infection (cite).
+On the other hand, it is known that humans, especially health care
+employees, can be asymptomatic carriers of \emph{C. difficile}
+(Furyura-Kanamari). This may explain the fact that the majority of our
+groups of mice were colonized by \emph{C. difficile} but did not display
+outward signs of illness or infection-associated weight loss (Fig 2).
+
+• Discuss prediction methods and outcomes - Nick - we have bugs that
+predict severity, but do we have bugs that predict mild/asymptomatic?
+probably, could discuss here • Discuss potential mechanisms for
+interesting OTUs - Nick
+
+Finally, our pilot experiments testing the efficiency of infection with
+different \emph{C. difficile} strains
+
+Discuss different strain results
+
+• Future work/conclusion section/sentence
 
@@
 Donor stool for this study was selected from one patient that tested
-positive for \emph{C. difficile} infection and four patients that
-presented with diarrhea that tested negative for \emph{C. difficile}.
-Healthy donor stool was collected from volunteer donors living in the
-Ann Arbor, MI area.
+positive for \emph{C. difficile} infection and five patients that
+presented with diarrhea and tested negative for \emph{C. difficile}.
+Healthy donor stool was collected from adult volunteers living in the
+Ann Arbor, MI area. 100ul of stool slurry was gavaged into each mouse.
+
+\textbf{\emph{Clostridium difficile} and sample processing} For all
+experiments, 100 spores of \emph{Clostridium difficile} were used to
+orally infect mice as described previously (cite alyx). The \emph{C.
+difficile} strains used were acquired from spore stocks of (WHERE) and
+comprised of strains 431, 395, 299 and 468 (cite?). After gavage, the
+remaining inoculum was diluted and plated to confirm infection dose.
+Bacteria were cultured anaerobically on TCCFA plates and incubated at
+37C. During infection, two fresh stool pellets were collected from each
+mouse daily. One pellet was resuspended in anaerobic phosphate-buffered
+saline and diluted and plated on TCCFA plates to enumerate \emph{C.
+difficile} CFU. The other pellet was frozen at -80C until the end of the
+experiment.
+
+\textbf{DNA isolation and sequencing} Bacterial DNA was isolated from
+banked donor stool and each mouse stool pellet using the MO-BIO
+PowerSoil DNA isolation kit. This DNA was used as template for
 

### 2016-11-21 code/HumanCdGF_predict_cfu.Rmd
add cfu om_loo validation

@@
                                  importance = om_raw_rank)
+
+# one mouse leave one out
+om_loo_pred <- c()
+for(i in 1:iters){
+    set.seed(i)
+    one_mouse <- one_mouse_df %>% 
+      group_by(cage_id) %>% 
+      sample_n(1) %>% ungroup() 
+    for (n in unique(as.character(one_mouse$cage_id))){
+      train <- filter(one_mouse, !cage_id == n)
+      test <- one_mouse_df %>% 
+        filter(cage_id == n)
+      set.seed(seed)
+      rf_om_loo <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id, -cage_id),
+                                ntree = 500)
+      om_loo_pred <- rbind(om_loo_pred,
+                           data.frame(select(test, mouse_id, cage_id, log_cfu_med),
+                           predicted=predict(rf_om_loo, test)))
+    }
+      #import_df <- importance(rf_om)
+      #colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
+      #om_import_otus <- merge(om_import_otus, import_df, by.x = 'otu', by.y = 'row.names')
+      #om_import_otus <- select(om_import_otus, -IncNodePurity)
+}
+om_loo_rsq <- 1 - (sum((om_loo_pred$log_cfu_med - om_loo_pred$predicted)^2)/
+               sum((om_loo_pred$log_cfu_med - mean(om_loo_pred$log_cfu_med))^2))
+
+om_loo_pred <- rename(om_loo_pred, observed = log_cfu_med)
+
+cor.test(om_loo_pred$observed, om_loo_pred$predicted, method = 'pearson')
 

### 2016-11-21 code/HumanCdGF_predict_cfu.Rmd
add om_loo to colonization rate
also updated plots with unidentifiable title

--- 2016-11-08 code/HumanCdGF_predict_cfu.Rmd
@@
 
+# one mouse leave one out
+om_loo_pred <- c()
+for(i in 1:iters){
+    set.seed(i)
+    one_mouse <- one_mouse_df %>% 
+      group_by(cage_id) %>% 
+      sample_n(1) %>% ungroup() 
+    for (n in unique(as.character(one_mouse$cage_id))){
+      train <- filter(one_mouse, !cage_id == n)
+      test <- one_mouse_df %>% 
+        filter(cage_id == n)
+      set.seed(seed)
+      rf_om_loo <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id, -cage_id),
+                                ntree = 500)
+      om_loo_pred <- rbind(om_loo_pred,
+                           data.frame(select(test, mouse_id, cage_id, log_cfu_med),
+                           predicted=predict(rf_om_loo, test)))
+    }
+}
+
+om_loo_rsq <- 1 - (sum((om_loo_pred$log_cfu_med - om_loo_pred$predicted)^2)/
+               sum((om_loo_pred$log_cfu_med - mean(om_loo_pred$log_cfu_med))^2))
+om_loo_pred <- rename(om_loo_pred, observed = log_cfu_med)
+
 #### merge OOB and leave 1 cage out dataframes
@@
                 cbind(LOcO_probs, data = rep('LOO', nrow(LOcO_probs))),
+                cbind(om_loo_pred, data = rep('om_loo', nrow(LOcO_probs))),
                 cbind(om_pred, data = rep('1_mouse', nrow(om_pred))))
@@
 ggplot(data = all_rf, aes(x=observed, y=predicted, color = data)) + 
-  geom_abline(color = 'gray', size = 0.5, linetype = 2, slope = 1) + 
+  #geom_abline(color = 'gray', size = 0.5, linetype = 2, slope = 1) + 
   # geom_point() + 
-  geom_smooth(method = 'lm', se = F) +
+  #geom_smooth(method = 'lm', se = F) +
   stat_summary(fun.y=median, geom="point", aes(group = data)) + 
@@
   theme_bw() + 
-  scale_color_discrete(name = 'Model',
-                       breaks = c('OOB', 'LOO', '1_mouse'),
+  scale_color_discrete(name = NULL,
+                       breaks = c('OOB', 'LOO', '1_mouse', 'om_loo'),
                        labels = c(bquote('OOB (R'^'2'*' = '*.(round(tail(rf_n_features$rsq, 1), 2))*')'),
                                   bquote('LOO (R'^'2'*' = '*.(round(LOcO_rsq, 2))*')'),
-                                  bquote('OM (R'^'2'*' = '*.(round(om_rsq,2))*')'))) + 
+                                  bquote('OM (R'^'2'*' = '*.(round(om_rsq,2))*')'),
+                                  bquote('OM-LOO (R'^'2'*' = '*.(round(om_loo_rsq,2))*')'))) + 
   theme(legend.justification=c(0,1), legend.position=c(0,1))
@@
     partialPlot(rf_n_features, data.frame(cfu_shared_colonized.df), x.var=rownames(importance_om)[otu],
-                xlab='', main=paste(importance_om$tax_label[otu]))
+                xlab='', main = NULL)
+    if(length(features_cor[[otu]]) > 1){
+      title(main = paste0('Unidentifiable Feature (', 
+                          get_tax(row_list = rownames(importance_om)[otu])$tax_label,
+                          ' Shown)'))
+      mtext(side = 1, line = 3, cex = 0.6,
+        paste('Correlated features: \n', 
+              paste(get_tax(row_list = features_cor[[otu]])$tax_label, collapse = ', ')))
+      } else {
+      title(paste(get_tax(row_list = rownames(importance_om)[otu])$tax_label))
+        }
     par(new = T)
@@
     mtext(side = 4, line = 3, 'Log10 CFU')
-    if(length(features_cor[[otu]]) > 1){
-    mtext(side = 1, line = 3, cex = 0.6,
-          paste('Bacteria removed in preprocessing due to correlation to this OTU:\n',
-                paste(get_tax(row_list = features_cor[[otu]])$tax_label, collapse = ',')))
-    }
 # To color points by mouse - Change pch to 21 and un-comment following lines
@@
 
+# one mouse leave one out
+om_loo_pred_cr <- c()
+for(i in 1:iters){
+    set.seed(i)
+    one_mouse <- om_cr_df %>% 
+      group_by(cage_id) %>% 
+      sample_n(1) %>% ungroup() 
+    for (n in unique(as.character(one_mouse$cage_id))){
+      train <- filter(one_mouse, !cage_id == n)
+      test <- om_cr_df %>% 
+        filter(cage_id == n)
+      set.seed(seed)
+      rf_om_loo_cr <- randomForest(ratio ~ ., data = select(train, -mouse_id, -cage_id),
+                                ntree = 500)
+      om_loo_pred_cr <- rbind(om_loo_pred_cr,
+                           data.frame(select(test, mouse_id, cage_id, ratio),
+                           predicted=predict(rf_om_loo_cr, test)))
+    }
+}
+
+om_loo_rsq_cr <- 1 - (sum((om_loo_pred_cr$ratio - om_loo_pred_cr$predicted)^2)/
+               sum((om_loo_pred_cr$ratio - mean(om_loo_pred_cr$ratio))^2))
+om_loo_pred_cr <- rename(om_loo_pred_cr, observed = ratio)
+
 #### merge OOB and leave 1 cage out dataframes
@@
                 cbind(LOcO_cr_probs, data = rep('LOO', nrow(LOcO_cr_probs))),
+                cbind(om_loo_pred_cr, data = rep('OM_LOO', nrow(om_loo_pred_cr))),
                 cbind(om_pred_cr, data = rep('1_mouse', nrow(om_pred_cr))))
@@
 ggplot(data = all_cr_rf, aes(x=observed, y=predicted, color = data)) + 
-  geom_abline(color = 'gray', size = 0.5, linetype = 2, slope = 1) + 
+  #geom_abline(color = 'gray', size = 0.5, linetype = 2, slope = 1) + 
   # geom_point() + 
-  geom_smooth(method = 'lm', se = F) +
+  #geom_smooth(method = 'lm', se = F) +
   stat_summary(fun.y=median, geom="point", aes(group = data)) + 
@@
   theme_bw() + 
-  scale_color_discrete(name = 'Model',
-                       breaks = c('OOB', 'LOO', '1_mouse'),
+  scale_color_discrete(name = NULL,
+                       breaks = c('OOB', 'LOO', '1_mouse', 'OM_LOO'),
                        labels = c(bquote('OOB (R'^'2'*' = '*.(round(tail(rf_col_rate$rsq, 1), 2))*')'),
                                   bquote('LOO (R'^'2'*' = '*.(round(LOcO_rsq_cr, 2))*')'),
-                                  bquote('OM (R'^'2'*' = '*.(round(om_rsq_cr,2))*')'))) + 
+                                  bquote('OM (R'^'2'*' = '*.(round(om_rsq_cr,2))*')'),
+                                  bquote('OM_LOO (R'^'2'*' = '*.(round(om_loo_rsq_cr,2))*')'))) + 
   theme(legend.justification=c(0,1), legend.position=c(0,1))
@@
     partialPlot(rf_col_rate, shared_cr_df, cr_features[otu],
-                xlab='', main=paste(get_tax(row_list = cr_features)$tax_label[otu]))#, log = 'x')
+                xlab='', main = NULL)
+    if(length(cr_features_cor[[otu]]) > 1){
+      title(main = paste0('Unidentifiable Feature (', 
+                          get_tax(row_list = cr_features[otu])$tax_label,
+                          ' Shown)'))
+      mtext(side = 1, line = 3, cex = 0.6,
+        paste('Correlated features: \n', 
+              paste(get_tax(row_list = cr_features_cor[[otu]])$tax_label, collapse = ', ')))
+      } else {
+      title(paste(get_tax(row_list = cr_features[otu])$tax_label))
+        }
   par(new = T)
@@
     mtext(side = 4, line = 3, 'Relative Colonization')
-    if(length(cr_features_cor[[otu]]) > 1){
-      mtext(side = 1, line = 3, cex = 0.6,
-            paste('Bacteria removed in preprocessing due to correlation to this OTU:\n',
-                  paste(get_tax(row_list = cr_features_cor[[otu]])$tax_label, collapse = ',')))
-    }
 # To color points by mouse - Change pch to 21 and un-comment following lines

### 2016-11-22 code/HumanCdGF_predict_cfu.Rmd
remove code reducing features of 1%  rel abun
remove initial model feature reductions

@@
 # remove samples not used or with other cdiff strains
-meta_file <- 
+meta_data <- 
   meta_data %>% 
   filter(Early_Euth == FALSE, cdiff_strain == 431, !cage_id == 'NP1') %>% 
-  select(mouse_id, sample_id, day, log_cfu) %>% 
+  select(mouse_id, sample_id, cage_id, day, log_cfu) %>% 
   filter(!is.na(log_cfu))
-
-shared_list_days <- select(meta_file, -log_cfu)
+shared_list_days <- meta_data %>% 
+  filter(day == 0) %>% 
+  select(-day, -log_cfu, -cage_id)
+shared_file <- shared_file[rownames(shared_file) %in% shared_list_days$sample_id, ]
 
@@
 mean_otu_counts <- unique(apply(shared_file, 1, sum))
-rel_abund_shared <- 100 * shared_file / mean_otu_counts
+rel_abund_shared <- shared_file
 rel_abund_shared_01 <- rel_abund_shared[ , apply(rel_abund_shared, 2, max) > 1]
@@
 rel_abund_01_filtr <- rel_abund_shared_01[,-cor_otus]
-rel_abund_01_filtr$sample_id <- rownames(rel_abund_01_filtr)
+community_d0.df <- merge(shared_list_days, rel_abund_01_filtr, 
+                         by.x = 'sample_id', by.y = 'row.names') %>% 
+                    select(-sample_id)
 ```
@@
 ```{r predict median CFU, fig.width=10,fig.height=10}
-
 colonized_cfu <- meta_data %>% 
-  filter(day %in% c(4:10), !cage_id == 'NP1', 
-         Early_Euth == FALSE, cdiff_strain == 431) %>% 
-  select(mouse_id, cage_id, log_cfu) %>% 
+  filter(day %in% c(4:10)) %>% 
   group_by(mouse_id, cage_id) %>% 
@@
   ungroup()
-sample_list_d0 <- shared_list_days %>% 
-    filter(day == 0) %>% 
-    select(-day)
-community_d0.df <- rel_abund_01_filtr %>% 
-    right_join(sample_list_d0) %>% 
-    select(-sample_id)
+
 cfu_shared_colonized.df <- colonized_cfu %>% 
@@
                    verbose = FALSE)
-subsets <- c(1:5, c(seq(10, 70, 10)))
+subsets <- c(1:20, c(seq(25, 85, 10)))
 set.seed(seed)
@@
 #One per cage
-one_mouse_initial_df <- data.frame(colonized_cfu %>% 
+one_mouse_df <- data.frame(colonized_cfu %>% 
     right_join(community_d0.df) %>% 
@@
     select(log_cfu_med, mouse_id, cage_id, one_of(top_important_OTU)))
-
-om_raw_otus <- data.frame(otu = colnames(select(one_mouse_initial_df, contains('Otu'))))
-for(i in 1:iters){
-    set.seed(i)
-    train <- one_mouse_initial_df %>% 
-      group_by(cage_id) %>% 
-      sample_n(1) %>% ungroup() 
-    test <- one_mouse_initial_df %>% 
-      filter(!mouse_id %in% train$mouse_id)
-    set.seed(seed)
-    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id, -cage_id),
-                             importance = TRUE, keep.forest = T, ntree = n_trees)
-    import_df <- importance(rf_om)
-    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
-    om_raw_otus <- merge(om_raw_otus, import_df, by.x = 'otu', by.y = 'row.names')
-    om_raw_otus <- select(om_raw_otus, -IncNodePurity)
-}
-raw_importance <- data.frame(om_raw_otus[ , -1], row.names = om_raw_otus$otu)
-### rank importance
-importance_rank <- apply(raw_importance, 2, function(x) rank(x, na.last = F))
-raw_rank <- apply(importance_rank, 1, function(x) sum(x, na.rm = T))
-final_rank <- sort(raw_rank[raw_rank >=  ncol(importance_rank)*sum(importance_rank[ , 1])/nrow(importance_rank)], decreasing = T)
-
-one_mouse_df <- data.frame(colonized_cfu %>% 
-    right_join(community_d0.df) %>% 
-    filter(!is.na(log_cfu_med))  %>% 
-    select(log_cfu_med, mouse_id, cage_id, one_of(names(final_rank))))
 
@@
     if(length(features_cor[[otu]]) > 1){
-      title(main = paste0('Unidentifiable Feature (', 
+      title(main = paste0('Unidentifiable Feature\n(', 
                           get_tax(row_list = rownames(importance_om)[otu])$tax_label,
@@
 mouse_list <- meta_data %>% 
-  filter(!is.na(day), !cage_id == 'NP1',  
-         Early_Euth == FALSE, cdiff_strain == 431) %>% 
   select(mouse_id) %>% 
@@
 #One per cage
-one_mouse_cr_df <- LOcO_cr_df
-
-om_raw_otus_cr <- data.frame(otu = colnames(select(one_mouse_cr_df, contains('Otu'))))
-for(i in 1:iters){
-    set.seed(i)
-    train <- one_mouse_cr_df %>% 
-      group_by(cage_id) %>% 
-      sample_n(1) %>% ungroup() 
-    test <- one_mouse_cr_df %>% 
-      filter(!mouse_id %in% train$mouse_id)
-    set.seed(seed)
-    rf_om <- randomForest(ratio ~ ., data = select(train, -mouse_id, -cage_id),
-                             importance = TRUE, keep.forest = T, ntree = n_trees)
-    import_df <- importance(rf_om)
-    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
-    om_raw_otus_cr <- merge(om_raw_otus_cr, import_df, by.x = 'otu', by.y = 'row.names')
-    om_raw_otus_cr <- select(om_raw_otus_cr, -IncNodePurity)
-}
-raw_importance_cr <- data.frame(om_raw_otus_cr[ , -1], row.names = om_raw_otus_cr$otu)
-### rank importance
-importance_rank_cr <- apply(raw_importance_cr, 2, function(x) rank(x, na.last = F))
-raw_rank_cr <- apply(importance_rank_cr, 1, function(x) sum(x, na.rm = T))
-final_rank_cr <- sort(raw_rank_cr[raw_rank_cr >=  ncol(importance_rank_cr)*sum(importance_rank_cr[ , 1])/nrow(importance_rank_cr)], decreasing = T)
-
-om_cr_df <- data.frame(shared_col_rate_df %>% 
-    select(ratio, mouse_id, cage_id, one_of(names(final_rank_cr))))
+om_cr_df <- LOcO_cr_df
 
@@
     if(length(cr_features_cor[[otu]]) > 1){
-      title(main = paste0('Unidentifiable Feature (', 
+      title(main = paste0('Unidentifiable Feature\n(', 
                           get_tax(row_list = cr_features[otu])$tax_label,

### 2016-11-23 code/CdGF_AUCRF.Rmd
AUCRF - delete otus less than 10% presence

--- 2016-11-16 code/CdGF_AUCRF.Rmd
@@
 
+seed <- 1
+n_trees <- 2001
+iters <- 100
+otu_presence_cutoff <- 0.1
+cor_cutoff <- 0.75
+
 # select only mice with 431 cdiff
@@
 stool_shared <- shared_file[rownames(meta_file[meta_file$day %in% 0, ]), ]
-# use %in% instead of == because to ignore NA
+# use %in% instead of == in order to ignore NA
 # rename day 0 community rows by mouse id - since meta rows have day in row names, use mouse id for both
@@
 #Create vector of OTUs with median abundances >1%
-OTUs_stool_0_01 <- apply(rel_abund_stool, 2, max) > 1
-#df of OTUs with abundances >1% - by cage and inoculum
-rel_abund_stool <- rel_abund_stool[ ,OTUs_stool_0_01]
+otu_presence <- ifelse(stool_shared > 0, T, F) 
+OTUs_present <- colnames(stool_shared[ , apply(otu_presence, 2, sum) > round(otu_presence_cutoff*nrow(stool_shared))])
+#df of OTUs with present in >10%
+rel_abund_stool <- rel_abund_stool[ ,OTUs_present]
 
@@
 rel_abund_cor <- cor(rel_abund_stool, method = 'spearman')
-cor_cutoff <- 0.75
 cor_otus <- findCorrelation(rel_abund_cor, cutoff = cor_cutoff)
 rel_abund_stool <- select(rel_abund_stool, -cor_otus)
-
-#random forest regression model to determine Day 0 OTU correlated with increased CFU
-seed <- 1
-n_trees <- 2001
-iters <- 100
 ```
@@
 layout(matrix(c(1:12), ncol=3, byrow=TRUE))
-
+par(mar = c(6,4,4,2)+0.1)
 for (i in seq_along(important_otus_om_loo)) {
@@
                           ' Shown)'))
-      mtext(side = 1, line = 3, cex = 0.6,
+      mtext(side = 1, line = 6, cex = 0.6,
         paste('Correlated features: \n', 
-                  get_tax(row_list = cor_features[[i]])$tax_label, collapse = ','))
+              paste(paste0(get_tax(row_list = cor_features[[i]])$tax_label, c(', ', ',\n ')), 
+                    collapse = '')))
   } else {

### 2016-11-23 submission/Manuscript.Rmd
minor text edits

@@
 difficile_ colonization and expansion.
-To study human-associated microbes in the context of CDI, we inoculated germ-free C57BL/6 mice with one of 16 human fecal samples from diarrheal or healthy donors and challenged with C.
-difficile 14 days later.
+To study human-associated microbes in the context of CDI, we inoculated germ-free C57BL/6 mice with one of 16 human fecal samples from diarrheal or healthy donors and challenged with _C.
+difficile_ 14 days later.
 Five unique donor-mice combinations resulted in severe CDI while the remaining 11 only experienced mild disease.
@@
 difficile_ at one to ten days post-infection.
-This model explained 99% of the variance in the number of CFU isolated from mouse stool.
+This model explained 86% of the variance in the number of CFU isolated from mouse stool.
 Members of _Lachnospiraceae, Parabacteroides, Bacteroidales, Bacteroidetes, Porphyromonadaceae_ and unclassified _Bacteria_ families were predictive of future _C.
@@
 In one study, germ-free piglets were acutely colonized with human feces for one week and then treated with tigecycline.
-After challenge with C.
-difficile none of the antibiotic-treated piglets succumbed to infection, while some of the untreated human-colonized pigs did (11).
+After challenge with _C.
+difficile_ none of the antibiotic-treated piglets succumbed to infection, while some of the untreated human-colonized pigs did (11).
 In another study, germ-free mice colonized with human feces were bred over several generations to create a cohort of mice with identical human-derived microbiomes (14).
@@
 Here we designed a laboratory model that allows for studying and modeling microbial coumminity interactions with _C.
-difficile_ infection in mice with human derived microbiomes.
+difficile_ infection in mice with human-derived microbiomes.
 To test the impact of individual human microbiomes on CDI, we colonized germ-free mice with 16 different human stool donors.
-We then characterized human-associated microbiome response to C.
-difficile challenge.
-Additionally, the use of machine-learning models allowed us to build a predictive model that classified “at-risk” microbiomes prior to infection with C.
-difficile.
+We then characterized human-associated microbiome response to _C.
+difficile_ challenge.
+Additionally, the use of machine-learning models allowed us to build a predictive model that classified “at-risk” microbiomes prior to infection with _C.
+difficile_.
 These findings show that human-associated microbiomes can be at risk for CDI even in the absence of antibiotics and that study of mice colonized with human feces provides a range of clinical outcomes.
@@
 ### Results 
-**Germ-free mice inoculated with human feces as model for C.
-difficile infection.** To generate mice with human-derived microbiomes, we inoculated one cage of gnotobiotic C57/BL6 mice with one of 16 different human fecal donors.
+**Germ-free mice inoculated with human feces as model for _C.
+difficile_ infection.** To generate mice with human-derived microbiomes, we inoculated one cage of gnotobiotic C57/BL6 mice with one of 16 different human fecal donors.
 Five donors were patients that had diarrhea that was not attributable to _C.
@@
 
-**C.
-difficile infection in mice with human-derived microbiota cause a range of outcomes.** _C.
+**_C.
+difficile_ infection in mice with human-derived microbiota cause a range of outcomes.** _C.
 difficile_ colonization was monitored by daily plating of stool pellets for _C.
@@
 
-**Nick will be updating the next two sections**
 **Microbes present in the gut prior to infection are predictive of _C.
@@
 Relative abundance of OTUs with greater than 1% abundance at day 0 were used as input to the model.
-We found that the model explained X% of the variance in colonization levels of _C.
+We found that the model explained 86% of the variance in colonization levels of _C.
 difficile_ (Fig 3A).
-By refining the model to include only the top X predictive OTUs, the model explained X% of the variance.
+By refining the model to include only the top predictive OTUs and training with one individual mouse per cage, the model explained 77%% of the variance.
 Partial dependency plots of the top 6 OTUs predictive of _C.
-difficile_ colonization reveal that colonization is positively associated with _Bacteriodetes_, _Parabacteriodes_ and _Porphyromonadaceae_ (really?) (Fig 3B).
-Colonization was negatively associated with the presence of _Lachnospiraceae_ (Fig 3B) though none of the human-associated mouse communities were completely resistant to colonization by _C.
+difficile_ colonization reveal that colonization is positively associated with _Enterococcus_, _Akkermansia_ (~2%) and an unidentifiable feature consisting of strongly corrlated OTUs of _Enterobacteriaceae_(13.5%), _Anaeroplasma_ (3.35%), _Prevotella_ (4.10%), _Alloprevotella_ (3.55%), and _Porphyromonadaceae_ (1.2%) (Fig 3B).
+Colonization was negatively associated with the presence of _Blautia_, _Ruminococcaceae_, _Lachnospiraceae_, _Clostridium XVII_, _Clostridium XIVb_, _Clostridium XIVa_, _Clostridiales_, _Oscilibacter_, and an unidentfiable feature representing highly correlated OTUs of _Clostridium XVIII_ (2.55%), _Holdemania_ (1.45%), _Robinsiella_ (2.20%), and _Clostridium IV_ (1.5%) (Fig 3B) though none of the human-associated mouse communities were completely resistant to colonization by _C.
 difficile_ (Fig 2A, 3B).
@@
 difficile_ colonization levels can be predicted by the microbiome composition on day 0, we postulated that severity of disease could also be predicted from day 0.
-Using weight loss as a proxy for severity, we built a Random Forest classification model based on relative abundance to determine which community members were associated with either mild or severe disease.
-We optimized the model using the AUC-RF algorithm to generate a Reciever Operating Curve (ROC) with maximal Area Under the Curve (AUC) (cite Calle 2011), as used previously in our group to classify colon cancer outcomes (cite Niel 2015 gen med).
-The optimized model used X number of OTUs and identified several that predicted increased weightloss and subsequent disease severity (Fig 4A, 4B).
+__Using weight loss as a proxy__(?
+and clinical signs of disease ?) for severity, we built a Random Forest classification model based on relative abundance to determine which community members were associated with either mild or severe disease.
+In order to rule out the potential cage effect was confounding the effect of the microbiome on disease severity, the model was built with a single mouse per human source.
+We optimized the model using the AUC-RF algorithm to generate a Reciever Operating Characteristic (ROC) curve with maximal Area Under the Curve (AUC) (cite Calle 2011), as used previously in our group to classify colon cancer outcomes (cite Niel 2015 gen med).
+The optimized model used 10 OTUs and identified several that predicted increased ~~weightloss and subsequent~~ disease severity (Fig 4A, 4B).
 The populations that were most commonly found in mice that succumbed to _C.
-difficile_ infection were members of the _Clostridiales_, _Bacilliales_, _Streptococcus_ (Lactobacilles?) and _Bacteriodes_.
-(need more OTUs here, or number that were used for model and then group by rank).
-The AUC was calculated using the Out-Of-Bag (OOB) error for each sample (Fig 4A).
-Cross-validation using 10-fold, leave-one-mouse-out and leave-one-cage-out resulted in AUCs not significantly different from the OOB (Fig 4A).
-This model reveals that the levels of X OTUs on day 0 can predict disease outcome in mice with human-associated microbiomes.
+difficile_ infection were members of the _Clostridiaceae_, _Akkermansia_, _Turicibacter_, _Odoribacter_, _Ruminococcaceae_ and _Parabacteriodes_.
+Whereas, _Clostridium XIVa_ and _Lachnospiraceae_ were in higher abundance in mice with a mild _C.
+difficle_ infection.
+The AUC was calculated using the Out-Of-Bag (OOB) probabilities for each sample (Fig 4A).
+Validation using leave-one-source-out, leave-two-sources-out, and 100-iterations of training with one-mouse per source leaving one source out resulted in AUCs not significantly different from the OOB (Fig 4A).
+This model reveals that the levels of 10 OTUs on day 0 can predict disease outcome in mice with human-associated microbiomes.
 (probably also going to want a supplemental table for predictive OTUs for this and Fig 3)
@@
 _C.
-difficile_ presence was measured using WHATEVER- Cdiff quik check?
-toxin AB?.
+difficile_ presence was measured using **WHATEVER- Cdiff quik check?
+toxin AB?**.
 Samples from these patients were stored in Cary-Blair medium at -80C.
@@
 
-•	Sequence curation/mothur, data analysis, code + availability of data(SRA) and code (Github) - Nick
+•	Sequence curation/mothur, 
+•	data analysis, 
+•	code + availability of data(SRA) and code (Github) - Nick
 
@@
 Germ-free mice inoculated with human feces as a model for _C.
-difficile infection._** A) Experimental design.
+difficile_ infection.** A) Experimental design.
 Stool from 16 healthy, diarrheal and CDI patients were independently inoculated into 3-4 germ-free mice by oral gavage.
-14 days later mice were orally infected with 100 spores of C.
-difficile strain 431.
+14 days later mice were orally infected with 100 spores of _C.
+difficile_ strain 431.
 Weight and stool CFU were monitored for up to 10 days post infection.
@@
 difficile_ infection results in mild or severe disease.** A) _C.
-difficile CFU_ was enumerated by plating of mouse stool pellets daily.
+difficile_ CFU was enumerated by plating of mouse stool pellets daily.
 Each point represents a mouse and the lines represent the median CFU in each group.
@@
 difficile_ after colonization.
-OTUs were chosen such that they were not predictive of cage or donor.
-Each point is a mouse colored by cage.
 B) Partial dependency plots of the top six predictive OTUs.
-Line displats the partial dependence of log~10~ CFU on the relative abundance of each predictive OUT.
+Line displays the partial dependence of log~10~ CFU on the relative abundance of each predictive OTU.
 Each median log~10~ CFU is plotted against its relative abundance for each predictive OTU.

### 2016-11-23 code/CdGF_AUCRF.Rmd
using full shared files

@@
 meta_file   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
-shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
+shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/humanCdGF_full.shared'
 tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/gf_new.an.taxonomy'
@@
                                                rownames(stool_shared) ]
-rel_abund_stool <- 100 * stool_shared / unique(apply(stool_shared, 1, sum))
+rel_abund_stool <- 100 * stool_shared / apply(stool_shared, 1, sum)
 #Create vector of OTUs with median abundances >1%
-otu_presence <- ifelse(stool_shared > 0, T, F) 
-OTUs_present <- colnames(stool_shared[ , apply(otu_presence, 2, sum) > round(otu_presence_cutoff*nrow(stool_shared))])
+otu_presence <- ifelse(rel_abund_stool >= (1/2000), T, F) 
+OTUs_present <- colnames(rel_abund_stool[ , apply(otu_presence, 2, sum) > round(otu_presence_cutoff*nrow(rel_abund_stool))])
 #df of OTUs with present in >10%

### 2016-11-23 code/HumanCdGF_predict_cfu.Rmd
using full shared files

@@
 meta_data   <- 'data/process/human_CdGF_metadata.txt'
-shared_file <- 'data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
+shared_file <- 'data/process/humanCdGF_full.shared'
 tax_file <- 'data/process/gf_new.an.taxonomy'
@@
 cor_cutoff <- 0.75
+otu_presence_cutoff <- 0.1
 
@@
 # create shared with relative abundances over 1%
-mean_otu_counts <- unique(apply(shared_file, 1, sum))
-rel_abund_shared <- shared_file
-rel_abund_shared_01 <- rel_abund_shared[ , apply(rel_abund_shared, 2, max) > 1]
+rel_abund_shared <- 100 * shared_file / apply(shared_file, 1, sum)
+otu_presence <- ifelse(rel_abund_shared >= (1/2000), T, F) 
+rel_abund_shared_01 <- rel_abund_shared[ , apply(otu_presence, 2, sum) > round(otu_presence_cutoff*nrow(rel_abund_shared))]
 # remove correlative otus
@@
                    verbose = FALSE)
-subsets <- c(1:20, c(seq(25, 85, 10)))
+subsets <- c(1:15, c(seq(20, 70, 10)))
 set.seed(seed)
@@
 layout(matrix(c(1:9), ncol=3, byrow=TRUE))
+par(mar = c(7,4,4,2)+0.1)
 for(otu in seq_along(importance_om$tax)){
@@
                           ' Shown)'))
-      mtext(side = 1, line = 3, cex = 0.6,
+      mtext(side = 1, line = 6, cex = 0.6,
         paste('Correlated features: \n', 
-              paste(get_tax(row_list = features_cor[[otu]])$tax_label, collapse = ', ')))
+              paste(paste0(get_tax(row_list = features_cor[[otu]])$tax_label, c(', ', ',\n ')), 
+                    collapse = '')))
       } else {
@@
 layout(matrix(c(1:9), ncol=3, byrow=TRUE))
-
+par(mar = c(7,4,4,2)+0.1)
 cage_color <- rainbow(length(unique(cr_prediction$cage_id)))
@@
                           ' Shown)'))
-      mtext(side = 1, line = 3, cex = 0.6,
+      mtext(side = 1, line = 6, cex = 0.6,
         paste('Correlated features: \n', 
-              paste(get_tax(row_list = cr_features_cor[[otu]])$tax_label, collapse = ', ')))
+              paste(paste0(get_tax(row_list = cr_features_cor[[otu]])$tax_label, c(', ', ',\n ')), 
+                    collapse = '')))
       } else {

### 2016-11-23 code/HumanCdGF_predict_cfu.Rmd
remove changes for full shared file

@@
 meta_data   <- 'data/process/human_CdGF_metadata.txt'
-shared_file <- 'data/process/humanCdGF_full.shared'
+shared_file <- 'data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
 tax_file <- 'data/process/gf_new.an.taxonomy'
@@
 # create shared with relative abundances over 1%
-rel_abund_shared <- 100 * shared_file / apply(shared_file, 1, sum)
-otu_presence <- ifelse(rel_abund_shared >= (1/2000), T, F) 
-rel_abund_shared_01 <- rel_abund_shared[ , apply(otu_presence, 2, sum) > round(otu_presence_cutoff*nrow(rel_abund_shared))]
+mean_otu_counts <- unique(apply(shared_file, 1, sum))
+otu_presence <- ifelse(shared_file > 0, T, F) 
+rel_abund_shared <- shared_file[ , apply(otu_presence, 2, sum) > round(otu_presence_cutoff*nrow(shared_file))]
+rel_abund_shared_01 <- 100*rel_abund_shared/mean_otu_counts
 # remove correlative otus

### 2016-12-01 submission/Manuscript.Rmd

--- 2016-11-18 submission/Manuscript.Rmd
@@
 
-\vspace{40mm}
+\vspace{35mm}
 
@@
 ### Abstract
-_Clostridium difficile_ causes diarrheal disease when it successfully colonizes a dysbiotic gut microbial community.
+_Clostridium difficile_ causes diarrheal disease when it successfully colonizes a disrupted gut microbial community.
 Current mouse models to study _C.
-difficile_ infection (CDI) rely on pre-treatment with antibiotics to disrupt the mouse microbiome prior to inoculation.
+difficile_ infection (CDI) rely on pre-treatment with antibiotics to alter the mouse microbiome prior to inoculation.
 This model does not allow for analysis of human-associated microbial community members that modulate _C.
@@
 While informative, these studies were limited in their use of only one human donor as input inoculum.
-In order to best understand the impact of _C.
-difficile_ pathogenesis on human disease, we must have a laboratory model that allows for study of a variety of human-derived microbiomes.
 
-Here we designed a laboratory model that allows for studying and modeling microbial coumminity interactions with _C.
+Here we designed a laboratory system that allows for studying and modeling microbial coumminity interactions with _C.
 difficile_ infection in mice with human derived microbiomes.
 To test the impact of individual human microbiomes on CDI, we colonized germ-free mice with 16 different human stool donors.
-We then characterized human-associated microbiome response to C.
-difficile challenge.
-Additionally, the use of machine-learning models allowed us to build a predictive model that classified “at-risk” microbiomes prior to infection with C.
-difficile.
-These findings show that human-associated microbiomes can be at risk for CDI even in the absence of antibiotics and that study of mice colonized with human feces provides a range of clinical outcomes.
+We then characterized human-associated microbiome response to _C.
+difficile_ challenge.
+Further, we built a predictive model based on bacterial composition that classified “at-risk” microbiomes prior to infection with _C.
+difficile_.
+These findings show that human-associated microbiomes can be at risk for CDI even in the absence of antibiotic-induced dysbiosis.
+Finally, our results indicate that study of mice colonized with human feces can replicate a range of clinical outcomes.
 
@@
 **Germ-free mice inoculated with human feces as model for C.
-difficile infection.** To generate mice with human-derived microbiomes, we inoculated one cage of gnotobiotic C57/BL6 mice with one of 16 different human fecal donors.
+difficile infection.** To generate mice with human-derived microbiomes, we inoculated one cage each of gnotobiotic C57/BL6 mice with one of 16 different human fecal donors.
 Five donors were patients that had diarrhea that was not attributable to _C.
@@
 difficile_ was used as a positive control.
-After inoculation with human stool, mice were allowed to equilibrate for 14 days.
-Prior to infection, stool samples were taken from each mouse to establish baseline.
-Then, the _C.
-difficile_ strain isolated from the positive control patient’s sample (strain 430) was used to infect each mouse with 100 spores.
+After inoculation with human stool, mice microbiomes were allowed to equilibrate for 14 days.
+Prior to infection, stool samples were taken from each mouse to establish baseline microbiome composition.
+Then, 100 spores of _C.
+difficile_ strain 430 (isolated from the positive control patient) were used to orally infect each mouse.
 Mice were monitored for weight loss and clinical signs of disease.
 Fecal samples were taken to enumerate _C.
-difficile_ CFU and for microbiome analysis every day for up to 10 days post-infection (Fig 1A).
+difficile_ CFU and for microbiome analysis each day up to 10 days post-infection (Fig 1A).
 To ensure that the donors we selected represented a diverse array of human microbiomes, we sequenced the 16S rRNA genes from donor fecal inocula.
 Ordination of the distances between donor communities showed that the donors each had distinctly different communities, independent of whether the sample came from a sick or healthy person (Fig 1B).
-Likewise, the starting microbial communities of the mice on day 0 were characterizing by sequencing of fecal pellets DNA prior to infection.
-Ordination of all of the mouse communities on day 0 shows that mice were similar to each other within each cage and donor, but distinct from other donors (Fig 1C).
-This result confirmed that human-associated microbes were able to colonize gnotobiotic mice and provide distinct initial communities to test _C.
+Likewise, the microbial communities of the mice on day 0 were characterized by sequencing of fecal pellet DNA prior to infection (Table S1).
+Ordination of day 0 communities shows that mice were similar to each other within each cage and donor, but distinct from other donors (Fig 1C).
+This result confirmed that human-associated microbes were able to colonize gnotobiotic mice and provide various initial communities to test _C.
 difficile_ dynamics.
-
-**need to put in a line to reference supplemental table with RA in it**
 
@@
 Nearly all of the mice were colonized to 10^5^ – 10^7^ CFU by one day post-infection and remained colonized at that level until the end of the experiment (Fig 2A).
-As one indicator of disease, mouse weights were taken each day post-infection and weight-loss was monitored alongside clinical signs of disease.
+As one indicator of disease, mouse weights were taken each day post-infection and weight-loss was monitored alongside clinical signs.
 When mice were judged to be too ill to continue they were humanely euthanized.
@@
 By refining the model to include only the top X predictive OTUs, the model explained X% of the variance.
-Partial dependency plots of the top 6 OTUs predictive of _C.
+Partial dependency plots of the top X OTUs predictive of _C.
 difficile_ colonization reveal that colonization is positively associated with _Bacteriodetes_, _Parabacteriodes_ and _Porphyromonadaceae_ (really?) (Fig 3B).
@@
 difficile_ infection, two donor-mouse microbiomes that were previously shown to be resistant to _C.
-difficile_ strain 431 induced weight loss (Fig 2B) (DA00396, DA00430) and one that was suscpetible to severe disease (DA00578) were independently challenged with one of three unique _C.
-difficile_ strains (table or reference for toxin prod, etc).
+difficile_ strain 430 induced weight loss (Fig 2B) (DA00396, DA00430-wait isnt this a control) and one that was susceptible to severe disease (DA00578) were independently challenged with one of three unique _C.
+difficile_ strains (*table or reference for toxin prod*, etc).
 As done previously, _C.
 difficile_ stool CFU was enumerated daily and mouse weight-loss tracked.
-Infection of mice with resistant communities with virulent strain 458 or the weaker strain 299 resulted in colonization to levels near that of 431 (Fig 5A) and these mice did not experience severe weight loss or show clinical signs of disease (Fig 5B).
-In contrast, infection of mice harboring microbiomes susceptible to severe disease (DA00578, Fig 1B) with strain 395 resulted in colonization and severe weight loss by two days post-infection (Fig 5A, 5B).
+Infection of resistant mice with virulent strain 458 or the weaker strain 299 resulted in colonization to levels near that of 431 (Fig 5A).
+Accordingly, these mice did not experience severe weight loss or show clinical signs of disease (Fig 5B).
+In contrast, infection of susceptible mice (DA00578, Fig 1B) with pathogenic strain 395 resulted in colonization and severe weight loss by two days post-infection (Fig 5A, 5B).
 Strain 395 is similarly virulent as strain 431, and shows a similar infection phenotype in these mice.
@@
 Here we colonized 16 groups of mice with human donor stool from patients with different clinical profiles or healthy volunteers.
-This created a diverse set of human-associated microbiomes in the mice that allowed for testing _C.
+This created a diverse set of human-associated mouse microbiomes that allowed for testing _C.
 difficile_ infection dynamics and community interactions.
 Our results showed that both healthy and diarrheal-associated human communities were susceptible to severe disease.
-Additionally, our results confirmed our and others previous studies that proved that resistance to _C.
-difficile_ disease is dependent on the entire microbial community and not just one or two species (cite alyx).
+Additionally, our results confirmed previous findings that resistance to _C.
+difficile_ disease is dependent on the entire microbial community and not just one or two taxa (cite alyx).
 Further, we were able to learn classification models that identified specific OTUs within susceptible communities that predispose mice to infection.
@@
 
-Here we were able to colonize germ-free mice with a variety of human stool communities from diarrheal or healthy people.
-After allowing these communities to colonize the mice for 2 weeks, they were challenged with _C.
-difficile_.
+Here we colonized germ-free mice with a variety of human stool communities and allowed these communities to colonize the mice for 2 weeks, prior to _C.
+difficile_ challenge.
 It is important to note that the mouse gut environment is different from that of the human, and as such exerts a selective pressure on the human stool microbial community that stabilizes over the 2 weeks.
-Thus these mice cannot be considered "humanized" but rather containing human-associated microbes.
+Thus these mice cannot be considered "humanized" but rather colonized by human-associated microbes.
 Interestingly, clinical status was not associated with proclivity to severe _C.
@@
 _C.
-difficile_ strains are known to vary in pathogenecity, primarily via the amount of toxin that is produced during infection (cite).
-In these pilot experiments, we challenged a susceptible community with another pathogenic strain of _C.
-difficile_ to confirm that the susceptibility lied in the community and not in the specific challenge strain.
-In contrast, we exposed a resistant/asymptomatic community to a relatively weak strain of _C.
+difficile_ strains are known to vary in pathogenicity, primarily via the amount of toxin that is produced during infection (cite).
+To confirm that disease risk is dependent on the microbiome, we challenged susceptibile mice with an additional pathogenic strain of _C.
+difficile_ and found that the mice indeed fell ill.
+In contrast, we exposed an asymptomatic community to a relatively weak strain of _C.
 difficile_ and found that the community could be colonized by this one as well-- but not cause disease.
-This suggests that this colony is prone to colonization by any _C.
+This suggests that this particular colony is prone to colonization by any _C.
 difficile_ strain.
@@
 It has been recently reported that carriage of a non-toxigenic strain of _C.
-difficile_ can prevent additional colonization by a pathogenic strain, suggesting some role of niche-filling or immunity this provides (cite cite).
-It would be interesting to see if we could make a susceptible community resistant to infection and disease by a pathogenic _C.
-difficle_ strain in our human-associated microbiome mice.
-Further work with human-associated communities in mice and mathematical modeling will allow for uncovering of complex human-associated disease processes and dynamics as they relate to human disease.
+difficile_ can prevent additional colonization by a pathogenic strain, suggesting a role of niche-filling or immunity this pre-colonization provides (cite cite).
+One interesting experiment would be to recapitulate this finding in human-associated mouse microbiomes by making a susceptible community resistant to infection.
+Further work with human-associated communities in mice and mathematical modeling will allow for uncovering of complex human-associated disease processes and dynamics as they relate to human _C.
+difficile_ infection.
 
@@
 After gavage, the remaining inoculum was diluted and plated to confirm infection dose.
-Bacteria were cultured anaerobically on TCCFA plates and incubated at 37C.
+Bacteria were cultured anaerobically on TCCFA plates and incubated at 37C (cite plate recipe).
 During infection, two fresh stool pellets were collected from each mouse daily.
@@
 ##Acknowledgments## 
-We would like to thank members of the Schloss lab and Jhansi Leslie for analysis advice and the Microbial Systems Laboratory at the University of Michigan for sequencing assistance.
+We would like to thank members of the Schloss lab and Jhansi Leslie for analysis advice.
+We would also like to thank the Microbial Systems Laboratory at the University of Michigan for sequencing assistance.
 

### 2016-12-01 submission/Manuscript.tex

@@
 \emph{Clostridium difficile} causes diarrheal disease when it
-successfully colonizes a dysbiotic gut microbial community. Current
+successfully colonizes a disrupted gut microbial community. Current
 mouse models to study \emph{C. difficile} infection (CDI) rely on
-pre-treatment with antibiotics to disrupt the mouse microbiome prior to
+pre-treatment with antibiotics to alter the mouse microbiome prior to
 inoculation. This model does not allow for analysis of human-associated
@@
 with human feces for one week and then treated with tigecycline. After
-challenge with C. difficile none of the antibiotic-treated piglets
-succumbed to infection, while some of the untreated human-colonized pigs
-did (11). In another study, germ-free mice colonized with human feces
-were bred over several generations to create a cohort of mice with
-identical human-derived microbiomes (14). These mice were subsequently
-treated with a five-antibiotic cocktail to induce dysbiosis and then
-were successfully colonized by \emph{C. difficile} (14). While
-informative, these studies were limited in their use of only one human
-donor as input inoculum. In order to best understand the impact of
-\emph{C. difficile} pathogenesis on human disease, we must have a
-laboratory model that allows for study of a variety of human-derived
-microbiomes.
+challenge with \emph{C. difficile} none of the antibiotic-treated
+piglets succumbed to infection, while some of the untreated
+human-colonized pigs did (11). In another study, germ-free mice
+colonized with human feces were bred over several generations to create
+a cohort of mice with identical human-derived microbiomes (14). These
+mice were subsequently treated with a five-antibiotic cocktail to induce
+dysbiosis and then were successfully colonized by \emph{C. difficile}
+(14). While informative, these studies were limited in their use of only
+one human donor as input inoculum.
 
-Here we designed a laboratory model that allows for studying and
+Here we designed a laboratory system that allows for studying and
 modeling microbial coumminity interactions with \emph{C. difficile}
@@
 different human stool donors. We then characterized human-associated
-microbiome response to C. difficile challenge. Additionally, the use of
-machine-learning models allowed us to build a predictive model that
-classified ``at-risk'' microbiomes prior to infection with C. difficile.
+microbiome response to \emph{C. difficile} challenge. Further, we built
+a predictive model based on bacterial composition that classified
+``at-risk'' microbiomes prior to infection with \emph{C. difficile}.
 These findings show that human-associated microbiomes can be at risk for
-CDI even in the absence of antibiotics and that study of mice colonized
-with human feces provides a range of clinical outcomes.
-
-\textbf{also want something in the intro about severity maybe, and
-asymptomatic colonization. or in discussion?}
+CDI even in the absence of antibiotic-induced dysbiosis. Finally, our
+results indicate that study of mice colonized with human feces can
+replicate a range of clinical outcomes.
 
@@
 difficile infection.} To generate mice with human-derived microbiomes,
-we inoculated one cage of gnotobiotic C57/BL6 mice with one of 16
+we inoculated one cage each of gnotobiotic C57/BL6 mice with one of 16
 different human fecal donors. Five donors were patients that had
@@
 that was colonized with virulent \emph{C. difficile} was used as a
-positive control. After inoculation with human stool, mice were allowed
-to equilibrate for 14 days. Prior to infection, stool samples were taken
-from each mouse to establish baseline. Then, the \emph{C. difficile}
-strain isolated from the positive control patient's sample (strain 430)
-was used to infect each mouse with 100 spores. Mice were monitored for
-weight loss and clinical signs of disease. Fecal samples were taken to
-enumerate \emph{C. difficile} CFU and for microbiome analysis every day
-for up to 10 days post-infection (Fig 1A). To ensure that the donors we
-selected represented a diverse array of human microbiomes, we sequenced
-the 16S rRNA genes from donor fecal inocula. Ordination of the distances
-between donor communities showed that the donors each had distinctly
-different communities, independent of whether the sample came from a
-sick or healthy person (Fig 1B). Likewise, the starting microbial
-communities of the mice on day 0 were characterizing by sequencing of
-fecal pellets DNA prior to infection. Ordination of all of the mouse
-communities on day 0 shows that mice were similar to each other within
-each cage and donor, but distinct from other donors (Fig 1C). This
-result confirmed that human-associated microbes were able to colonize
-gnotobiotic mice and provide distinct initial communities to test
-\emph{C. difficile} dynamics.
-
-\textbf{need to put in a line to reference supplemental table with RA in
-it}
+positive control. After inoculation with human stool, mice microbiomes
+were allowed to equilibrate for 14 days. Prior to infection, stool
+samples were taken from each mouse to establish baseline microbiome
+composition. Then, 100 spores of \emph{C. difficile} strain 430
+(isolated from the positive control patient) were used to orally infect
+each mouse. Mice were monitored for weight loss and clinical signs of
+disease. Fecal samples were taken to enumerate \emph{C. difficile} CFU
+and for microbiome analysis each day up to 10 days post-infection (Fig
+1A). To ensure that the donors we selected represented a diverse array
+of human microbiomes, we sequenced the 16S rRNA genes from donor fecal
+inocula. Ordination of the distances between donor communities showed
+that the donors each had distinctly different communities, independent
+of whether the sample came from a sick or healthy person (Fig 1B).
+Likewise, the microbial communities of the mice on day 0 were
+characterized by sequencing of fecal pellet DNA prior to infection
+(Table S1). Ordination of day 0 communities shows that mice were similar
+to each other within each cage and donor, but distinct from other donors
+(Fig 1C). This result confirmed that human-associated microbes were able
+to colonize gnotobiotic mice and provide various initial communities to
+test \emph{C. difficile} dynamics.
 
@@
 indicator of disease, mouse weights were taken each day post-infection
-and weight-loss was monitored alongside clinical signs of disease. When
-mice were judged to be too ill to continue they were humanely
-euthanized. Overall, disease phenotypes fell into two classes. Mice that
-became severely ill and lost 20\% or more of their starting body weight
-within one to two days post-infection were classified as ``severe''
-whereas mice that were colonized with \emph{C. difficile} but did not
-show signs of disease or severe weight loss were considered to have
-``mild'' disease (Fig 2A, 2B). Interestingly, \emph{C. difficile} was
-able to cause severe disease in both mice that had been colonized with
-healthy stool and those colonized with diarrheal stool, suggesting
-susceptibility to CDI is dependent on the composition of the starting
-microbiome and not associated with donor clinical metadata.
+and weight-loss was monitored alongside clinical signs. When mice were
+judged to be too ill to continue they were humanely euthanized. Overall,
+disease phenotypes fell into two classes. Mice that became severely ill
+and lost 20\% or more of their starting body weight within one to two
+days post-infection were classified as ``severe'' whereas mice that were
+colonized with \emph{C. difficile} but did not show signs of disease or
+severe weight loss were considered to have ``mild'' disease (Fig 2A,
+2B). Interestingly, \emph{C. difficile} was able to cause severe disease
+in both mice that had been colonized with healthy stool and those
+colonized with diarrheal stool, suggesting susceptibility to CDI is
+dependent on the composition of the starting microbiome and not
+associated with donor clinical metadata.
 
-\textbf{Microbes present in the gut prior to infection are predictive of
-\emph{C. difficile} colonization levels} Previous work in our group has
+\textbf{Nick will be updating the next two sections} \textbf{Microbes
+present in the gut prior to infection are predictive of \emph{C.
+difficile} colonization levels} Previous work in our group has
 demonstrated that the microbes present in the mouse microbiome prior to
@@
 Finally, our pilot experiments testing the efficiency of infection with
-different \emph{C. difficile} strains
-
-Discuss different strain results
-
-• Future work/conclusion section/sentence
+different \emph{C. difficile} strains have allowed for testing community
+versus pathogen factors of infection. Our results are consistent with
+previous reports that the composition of the gut microbiome determines
+if an individual is successfully infected or not (cite Alyx and others).
+\emph{C. difficile} strains are known to vary in pathogenecity,
+primarily via the amount of toxin that is produced during infection
+(cite). In these pilot experiments, we challenged a susceptible
+community with another pathogenic strain of \emph{C. difficile} to
+confirm that the susceptibility lied in the community and not in the
+specific challenge strain. In contrast, we exposed a
+resistant/asymptomatic community to a relatively weak strain of \emph{C.
+difficile} and found that the community could be colonized by this one
+as well-- but not cause disease. This suggests that this colony is prone
+to colonization by any \emph{C. difficile} strain. Further experiments
+will test if infecting susceptible communities with weak strains allows
+the mice to survive. It has been recently reported that carriage of a
+non-toxigenic strain of \emph{C. difficile} can prevent additional
+colonization by a pathogenic strain, suggesting some role of
+niche-filling or immunity this provides (cite cite). It would be
+interesting to see if we could make a susceptible community resistant to
+infection and disease by a pathogenic \emph{C. difficle} strain in our
+human-associated microbiome mice. Further work with human-associated
+communities in mice and mathematical modeling will allow for uncovering
+of complex human-associated disease processes and dynamics as they
+relate to human disease.
 
@@
 PowerSoil DNA isolation kit. This DNA was used as template for
+amplification of the V4 region of the 16S rRNA gene and fragments were
+sequenced on an Illumina MiSeq as previously described (cite Jim paper).
 
-• Bacteria/plating • Sequencing • Data analysis, code availability •
-Machine learning models
+• Sequence curation/mothur, data analysis, code + availability of
+data(SRA) and code (Github) - Nick
 
@@
 
-Lab, sequencing core, Jhansi
+We would like to thank members of the Schloss lab and Jhansi Leslie for
+analysis advice and the Microbial Systems Laboratory at the University
+of Michigan for sequencing assistance.
 

### 2016-12-05 submission/Manuscript.Rmd
got citations working, not done yet

@@
 title: "**Microbiota predict _Clostridium difficile_ severity in germ-free mice colonized with human feces**"
+bibliography: references.bib
 output:
@@
     keep_tex: yes
-csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
+csl: plos.csl #Get themes at https://github.com/citation-style-language/styles
 fontsize: 11pt
@@
 Indeed, ~12% of hospital acquired infections in the United States are due to _C.
-difficile_ and result in up to 15,000 deaths annually (2).
+difficile_ and result in up to 15,000 deaths annually [@Lessa2015].
 Disruption of the native microbial community is the most common risk factor for development of _C.
-difficile_ infection (CDI) (1).
+difficile_ infection (CDI) [@Britton2012].
 Antibiotic use and inflammatory bowel diseases are associated with loss of colonization resistance to _C.
-difficile_ through the loss of potentially protective bacterial families such as _Barnesiella_ and _Lachnospiraceae_ in both mouse models and human association experiments (cite alyx 2014 and vincent 2013).
+difficile_ through the loss of potentially protective bacterial families such as _Barnesiella_ and _Lachnospiraceae_ in both mouse models and human association experiments [@Schubert2014;@Vincent2013].
 The composition of the community is clearly important for the acquisition, resistance and treatment of _C.
-difficile_ infection, as giving patients a healthy fecal microbiome transplant is the most effective treatment for this disease (cite Anna 2014).
+difficile_ infection, as giving patients a healthy fecal microbiome transplant is the most effective treatment for this disease [@Seekatz2014].
 The precise mechanisms of colonization resistance and restoration of a healthy community are yet to be discovered.

### 2016-12-05 submission/Manuscript.tex
got citations working, not done yet

@@
 human fecal samples from diarrheal or healthy donors and challenged with
-C. difficile 14 days later. Five unique donor-mice combinations resulted
-in severe CDI while the remaining 11 only experienced mild disease. Both
-healthy and diarrheal donors were susceptible to colonization and severe
-symptoms of CDI. To determine if specific microbes were associated with
-disease severity outcomes, we built a classification Random Forest
-machine learning model based on relative abundance data of the
-communities prior to infection. The model identified a number of
-bacterial populations associated with the development of severe CDI,
-including \emph{Bacilliales, Ruminococcaceae, Ruminococcus,
-Staphylococcus, Streptococcus and Bacteriodetes}. Additionally, a
-regression model accurately predicted colonization levels of \emph{C.
-difficile} at one to ten days post-infection. This model explained 99\%
-of the variance in the number of CFU isolated from mouse stool. Members
-of \emph{Lachnospiraceae, Parabacteroides, Bacteroidales, Bacteroidetes,
-Porphyromonadaceae} and unclassified \emph{Bacteria} families were
-predictive of future \emph{C. difficile} colonization levels. Finally,
-challenging these mice with different strains of \emph{C. difficile}
-revealed that susceptible human-associated microbial communities were
-prone to severe disease independent of strain type. Taken together these
-results suggest that human-associated microbial communities can be
-recapitulated in germ-free mice and used to characterize dynamics of
-CDI. Because both healthy and diarrheal patients were susceptible to
-severe CDI, machine-learning models are useful to identify bacterial
-populations that allow colonization and contribute to the development of
-\emph{C. difficile} associated disease in humans.
+\emph{C. difficile} 14 days later. Five unique donor-mice combinations
+resulted in severe CDI while the remaining 11 only experienced mild
+disease. Both healthy and diarrheal donors were susceptible to
+colonization and severe symptoms of CDI. To determine if specific
+microbes were associated with disease severity outcomes, we built a
+classification Random Forest machine learning model based on relative
+abundance data of the communities prior to infection. The model
+identified a number of bacterial populations associated with the
+development of severe CDI, including \emph{Bacilliales, Ruminococcaceae,
+Ruminococcus, Staphylococcus, Streptococcus and Bacteriodetes}.
+Additionally, a regression model accurately predicted colonization
+levels of \emph{C. difficile} at one to ten days post-infection. This
+model explained 99\% of the variance in the number of CFU isolated from
+mouse stool. Members of \emph{Lachnospiraceae, Parabacteroides,
+Bacteroidales, Bacteroidetes, Porphyromonadaceae} and unclassified
+\emph{Bacteria} families were predictive of future \emph{C. difficile}
+colonization levels. Finally, challenging these mice with different
+strains of \emph{C. difficile} revealed that susceptible
+human-associated microbial communities were prone to severe disease
+independent of strain type. Taken together these results suggest that
+human-associated microbial communities can be recapitulated in germ-free
+mice and used to characterize dynamics of CDI. Because both healthy and
+diarrheal patients were susceptible to severe CDI, machine-learning
+models are useful to identify bacterial populations that allow
+colonization and contribute to the development of \emph{C. difficile}
+associated disease in humans.
 
@@
 States are due to \emph{C. difficile} and result in up to 15,000 deaths
-annually (2). Disruption of the native microbial community is the most
-common risk factor for development of \emph{C. difficile} infection
-(CDI) (1). Antibiotic use and inflammatory bowel diseases are associated
-with loss of colonization resistance to \emph{C. difficile} through the
-loss of potentially protective bacterial families such as
+annually {[}1{]}. Disruption of the native microbial community is the
+most common risk factor for development of \emph{C. difficile} infection
+(CDI) {[}2{]}. Antibiotic use and inflammatory bowel diseases are
+associated with loss of colonization resistance to \emph{C. difficile}
+through the loss of potentially protective bacterial families such as
 \emph{Barnesiella} and \emph{Lachnospiraceae} in both mouse models and
-human association experiments (cite alyx 2014 and vincent 2013). The
-composition of the community is clearly important for the acquisition,
-resistance and treatment of \emph{C. difficile} infection, as giving
-patients a healthy fecal microbiome transplant is the most effective
-treatment for this disease (cite Anna 2014). The precise mechanisms of
-colonization resistance and restoration of a healthy community are yet
-to be discovered.
+human association experiments {[}3,4{]}. The composition of the
+community is clearly important for the acquisition, resistance and
+treatment of \emph{C. difficile} infection, as giving patients a healthy
+fecal microbiome transplant is the most effective treatment for this
+disease (cite Anna 2014). The precise mechanisms of colonization
+resistance and restoration of a healthy community are yet to be
+discovered.
 
@@
 top X predictive OTUs, the model explained X\% of the variance. Partial
-dependency plots of the top 6 OTUs predictive of \emph{C. difficile}
+dependency plots of the top X OTUs predictive of \emph{C. difficile}
 colonization reveal that colonization is positively associated with
@@
 to \emph{C. difficile} infection, two donor-mouse microbiomes that were
-previously shown to be resistant to \emph{C. difficile} strain 431
-induced weight loss (Fig 2B) (DA00396, DA00430) and one that was
-suscpetible to severe disease (DA00578) were independently challenged
-with one of three unique \emph{C. difficile} strains (table or reference
-for toxin prod, etc). As done previously, \emph{C. difficile} stool CFU
-was enumerated daily and mouse weight-loss tracked. Infection of mice
-with resistant communities with virulent strain 458 or the weaker strain
-299 resulted in colonization to levels near that of 431 (Fig 5A) and
-these mice did not experience severe weight loss or show clinical signs
-of disease (Fig 5B). In contrast, infection of mice harboring
-microbiomes susceptible to severe disease (DA00578, Fig 1B) with strain
+previously shown to be resistant to \emph{C. difficile} strain 430
+induced weight loss (Fig 2B) (DA00396, DA00430-wait isnt this a control)
+and one that was susceptible to severe disease (DA00578) were
+independently challenged with one of three unique \emph{C. difficile}
+strains (\emph{table or reference for toxin prod}, etc). As done
+previously, \emph{C. difficile} stool CFU was enumerated daily and mouse
+weight-loss tracked. Infection of resistant mice with virulent strain
+458 or the weaker strain 299 resulted in colonization to levels near
+that of 431 (Fig 5A). Accordingly, these mice did not experience severe
+weight loss or show clinical signs of disease (Fig 5B). In contrast,
+infection of susceptible mice (DA00578, Fig 1B) with pathogenic strain
 395 resulted in colonization and severe weight loss by two days
@@
 donor stool from patients with different clinical profiles or healthy
-volunteers. This created a diverse set of human-associated microbiomes
-in the mice that allowed for testing \emph{C. difficile} infection
+volunteers. This created a diverse set of human-associated mouse
+microbiomes that allowed for testing \emph{C. difficile} infection
 dynamics and community interactions. Our results showed that both
 healthy and diarrheal-associated human communities were susceptible to
-severe disease. Additionally, our results confirmed our and others
-previous studies that proved that resistance to \emph{C. difficile}
-disease is dependent on the entire microbial community and not just one
-or two species (cite alyx). Further, we were able to learn
-classification models that identified specific OTUs within susceptible
-communities that predispose mice to infection. By subsequently infecting
-susceptible or resistant communities with varying strains of \emph{C.
-difficile} we confirmed that the initial state of the human-associated
-microbiome is more predictive of disease severity than the strain used.
+severe disease. Additionally, our results confirmed previous findings
+that resistance to \emph{C. difficile} disease is dependent on the
+entire microbial community and not just one or two taxa (cite alyx).
+Further, we were able to learn classification models that identified
+specific OTUs within susceptible communities that predispose mice to
+infection. By subsequently infecting susceptible or resistant
+communities with varying strains of \emph{C. difficile} we confirmed
+that the initial state of the human-associated microbiome is more
+predictive of disease severity than the strain used.
 
-Here we were able to colonize germ-free mice with a variety of human
-stool communities from diarrheal or healthy people. After allowing these
-communities to colonize the mice for 2 weeks, they were challenged with
-\emph{C. difficile}. It is important to note that the mouse gut
-environment is different from that of the human, and as such exerts a
-selective pressure on the human stool microbial community that
-stabilizes over the 2 weeks. Thus these mice cannot be considered
-``humanized'' but rather containing human-associated microbes.
-Interestingly, clinical status was not associated with proclivity to
-severe \emph{C. difficile} infection as some healthy donor mice were
-susceptible. A surprising finding was that none of the human-associated
-mouse communities were completely resistant to colonization by \emph{C.
-difficile}. This is in contrast to numerous reports of mouse microbiome
-studies (cite) and the knowledge that in general healthy humans without
-gut perturbations are resistant to \emph{C. difficile} infection (cite).
-On the other hand, it is known that humans, especially health care
-employees, can be asymptomatic carriers of \emph{C. difficile}
-(Furyura-Kanamari). This may explain the fact that the majority of our
-groups of mice were colonized by \emph{C. difficile} but did not display
-outward signs of illness or infection-associated weight loss (Fig 2).
+Here we colonized germ-free mice with a variety of human stool
+communities and allowed these communities to colonize the mice for 2
+weeks, prior to \emph{C. difficile} challenge. It is important to note
+that the mouse gut environment is different from that of the human, and
+as such exerts a selective pressure on the human stool microbial
+community that stabilizes over the 2 weeks. Thus these mice cannot be
+considered ``humanized'' but rather colonized by human-associated
+microbes. Interestingly, clinical status was not associated with
+proclivity to severe \emph{C. difficile} infection as some healthy donor
+mice were susceptible. A surprising finding was that none of the
+human-associated mouse communities were completely resistant to
+colonization by \emph{C. difficile}. This is in contrast to numerous
+reports of mouse microbiome studies (cite) and the knowledge that in
+general healthy humans without gut perturbations are resistant to
+\emph{C. difficile} infection (cite). On the other hand, it is known
+that humans, especially health care employees, can be asymptomatic
+carriers of \emph{C. difficile} (Furyura-Kanamari). This may explain the
+fact that the majority of our groups of mice were colonized by \emph{C.
+difficile} but did not display outward signs of illness or
+infection-associated weight loss (Fig 2).
 
@@
 if an individual is successfully infected or not (cite Alyx and others).
-\emph{C. difficile} strains are known to vary in pathogenecity,
+\emph{C. difficile} strains are known to vary in pathogenicity,
 primarily via the amount of toxin that is produced during infection
-(cite). In these pilot experiments, we challenged a susceptible
-community with another pathogenic strain of \emph{C. difficile} to
-confirm that the susceptibility lied in the community and not in the
-specific challenge strain. In contrast, we exposed a
-resistant/asymptomatic community to a relatively weak strain of \emph{C.
-difficile} and found that the community could be colonized by this one
-as well-- but not cause disease. This suggests that this colony is prone
-to colonization by any \emph{C. difficile} strain. Further experiments
-will test if infecting susceptible communities with weak strains allows
-the mice to survive. It has been recently reported that carriage of a
-non-toxigenic strain of \emph{C. difficile} can prevent additional
-colonization by a pathogenic strain, suggesting some role of
-niche-filling or immunity this provides (cite cite). It would be
-interesting to see if we could make a susceptible community resistant to
-infection and disease by a pathogenic \emph{C. difficle} strain in our
-human-associated microbiome mice. Further work with human-associated
-communities in mice and mathematical modeling will allow for uncovering
-of complex human-associated disease processes and dynamics as they
-relate to human disease.
+(cite). To confirm that disease risk is dependent on the microbiome, we
+challenged susceptibile mice with an additional pathogenic strain of
+\emph{C. difficile} and found that the mice indeed fell ill. In
+contrast, we exposed an asymptomatic community to a relatively weak
+strain of \emph{C. difficile} and found that the community could be
+colonized by this one as well-- but not cause disease. This suggests
+that this particular colony is prone to colonization by any \emph{C.
+difficile} strain. Further experiments will test if infecting
+susceptible communities with weak strains allows the mice to survive. It
+has been recently reported that carriage of a non-toxigenic strain of
+\emph{C. difficile} can prevent additional colonization by a pathogenic
+strain, suggesting a role of niche-filling or immunity this
+pre-colonization provides (cite cite). One interesting experiment would
+be to recapitulate this finding in human-associated mouse microbiomes by
+making a susceptible community resistant to infection. Further work with
+human-associated communities in mice and mathematical modeling will
+allow for uncovering of complex human-associated disease processes and
+dynamics as they relate to human \emph{C. difficile} infection.
 
@@
 Bacteria were cultured anaerobically on TCCFA plates and incubated at
-37C. During infection, two fresh stool pellets were collected from each
-mouse daily. One pellet was resuspended in anaerobic phosphate-buffered
-saline and diluted and plated on TCCFA plates to enumerate \emph{C.
-difficile} CFU. The other pellet was frozen at -80C until the end of the
-experiment.
+37C (cite plate recipe). During infection, two fresh stool pellets were
+collected from each mouse daily. One pellet was resuspended in anaerobic
+phosphate-buffered saline and diluted and plated on TCCFA plates to
+enumerate \emph{C. difficile} CFU. The other pellet was frozen at -80C
+until the end of the experiment.
 
@@
 We would like to thank members of the Schloss lab and Jhansi Leslie for
-analysis advice and the Microbial Systems Laboratory at the University
-of Michigan for sequencing assistance.
+analysis advice. We would also like to thank the Microbial Systems
+Laboratory at the University of Michigan for sequencing assistance.
 
@@
 
-\subsection{References}\label{references}
+\subsection*{References}\label{references}
+\addcontentsline{toc}{subsection}{References}
+
+1. Lessa FC, Mu Y, Bamberg WM, Beldavs ZG, Dumyati GK, Dunn JR, et al.
+Burden ofClostridium difficileInfection in the united states. New
+England Journal of Medicine. New England Journal of Medicine (NEJM/MMS);
+2015;372: 825--834.
+doi:\href{http://dx.doi.org/10.1056/nejmoa1408913}{10.1056/nejmoa1408913}
+
+2. Britton RA, Young VB. Interaction between the intestinal microbiota
+and host in clostridium difficile colonization resistance. Trends in
+Microbiology. Elsevier BV; 2012;20: 313--319.
+doi:\href{http://dx.doi.org/10.1016/j.tim.2012.04.001}{10.1016/j.tim.2012.04.001}
+
+3. Schubert AM, Rogers MAM, Ring C, Mogle J, Petrosino JP, Young VB, et
+al. Microbiome data distinguish patients with clostridium difficile
+infection and non-c. difficile-associated diarrhea from healthy
+controls. mBio. American Society for Microbiology; 2014;5:
+e01021--14--e01021--14.
+doi:\href{http://dx.doi.org/10.1128/mbio.01021-14}{10.1128/mbio.01021-14}
+
+4. Vincent C, Stephens DA, Loo VG, Edens TJ, Behr MA, Dewar K, et al.
+Reductions in intestinal clostridiales precede the development of
+nosocomial clostridium difficile infection. Microbiome. Springer Nature;
+2013;1: 18.
+doi:\href{http://dx.doi.org/10.1186/2049-2618-1-18}{10.1186/2049-2618-1-18}
 

### 2016-12-05 code/inocula_nmds.R

@@ -180,7 +180,7 @@ colors <- c("dodgerblue2","#E31A1C", # red
 #use this plot
 pdf("fig1B.pdf", width = 13, height = 5)
 
-ggplot(day0_donor_nmds, aes(axis1, axis2, group = Donor, color = Donor, shape = V2.x)) + geom_point(aes(fill=Donor), size = 3) +
+ggplot(day0_donor_nmds, aes(axis1, axis2, group = Donor, color = Donor, shape = Outcome)) + geom_point(aes(fill=Donor), size = 3) +
    theme_bw() + ggtitle("Similarity of day 0 communities") + scale_shape_discrete(name = "Outcome") + labs(x='NDMS Axis 1', y = 'NDMS axis 2') + scale_color_manual(values = colors)
 
 

### 2016-12-05 submission/Manuscript.Rmd
more citations

@@
 
-Murine models to study CDI typically rely on treating conventionally-raised mice with antibiotics either in drinking water or by injection to induce susceptibility (3, 4).
+Murine models to study CDI typically rely on treating conventionally-raised mice with antibiotics either in drinking water or by injection to induce susceptibility [@Buffie2011;@Theriot2011].
 This model provides a convenient way to study _C.

### 2016-12-05 code/HumanCdGF_model_cfu.Rmd
initalize model_cfu to compare rf models

---
title: "Model_CFU"
author: "Nicholas Lesniak"
date: "December 5, 2016"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE, fig.width=7, fig.height=8)
```

```{r setup environment}
pack_used <- c('randomForest','tidyr','ggplot2','plyr', 'dplyr', 'caret')
for (dep in pack_used){
  if (dep %in% installed.packages()[,"Package"] == FALSE){
    install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
                     quiet=TRUE);
  }
  library(dep, verbose=FALSE, character.only=TRUE)
}
```

```{r setup data}
# read in files
setwd('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/')
meta_data   <- 'data/process/human_CdGF_metadata.txt'
shared_file <- 'data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
tax_file <- 'data/process/gf_new.an.taxonomy'
tax_function <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R'

# read in files
meta_data   <- read.table(meta_data, sep = '\t', header = T, stringsAsFactors = F)
shared_file <- read.table(shared_file, sep = '\t', header = T, row.names = 'sample_id')
tax_file <- read.table(tax_file, sep = '\t', header = T, row.names = 'OTU')
source(tax_function)

# set parameters
seed <- 1
n_trees <- 1001
n_otus <- 12
iters <- 100
cor_cutoff <- 0.75
otu_presence_cutoff <- 0.1

# remove samples not used or with other cdiff strains
meta_data <- 
  meta_data %>% 
  filter(Early_Euth == FALSE, cdiff_strain == 431, !cage_id == 'NP1') %>% 
  select(mouse_id, sample_id, cage_id, human_source, day, log_cfu) %>% 
  filter(!is.na(log_cfu))
shared_list_days <- meta_data %>% 
  filter(day == 0) %>% 
  select(-day, -log_cfu, -cage_id, -human_source)
shared_file <- shared_file[rownames(shared_file) %in% shared_list_days$sample_id, ]

# create shared with relative abundances over 1%
mean_otu_counts <- unique(apply(shared_file, 1, sum))
otu_presence <- ifelse(shared_file > 0, T, F) 
rel_abund_shared <- shared_file[ , apply(otu_presence, 2, sum) > round(otu_presence_cutoff*nrow(shared_file))]
rel_abund_shared_01 <- 100*rel_abund_shared/mean_otu_counts
# remove correlative otus
rel_abund_cor <- cor(rel_abund_shared_01, method = 'spearman')
rownames(rel_abund_cor) <- colnames(rel_abund_shared_01) # add row names to identify otus correlated with features
cor_otus <- findCorrelation(rel_abund_cor, cutoff = cor_cutoff)
rel_abund_01_filtr <- rel_abund_shared_01[,-cor_otus]
community_d0.df <- merge(shared_list_days, rel_abund_01_filtr, 
                         by.x = 'sample_id', by.y = 'row.names') %>% 
                    select(-sample_id)

colonized_cfu <- meta_data %>% 
  filter(day %in% c(4:10)) %>% 
  group_by(mouse_id, cage_id, human_source) %>% 
  summarise(log_cfu_med = median(log_cfu)) %>% 
  ungroup()

##### feature reduction #####
rfRFE <-  list(summary = defaultSummary,
               fit = function(x, y, first, last, ...){
                 library(randomForest)
                 randomForest(x, y, importance = first, keep.forest = T, ntree = n_trees, ...)
                 },
               pred = function(object, x)  predict(object, x),
               rank = function(object, x, y) {
                 vimp <- varImp(object)
                 vimp <- vimp[order(vimp$Overall,decreasing = TRUE),,drop = FALSE]
                 vimp$var <- rownames(vimp)                  
                 vimp
                 },
               selectSize = pickSizeTolerance,
               selectVar = pickVars)

rfe_ctrl <- rfeControl(functions = rfRFE,
                   method = "repeatedcv",
                   repeats = 10,
                   verbose = FALSE)
subsets <- c(1:40, c(seq(45, 70, 5)))
```

### Predicting Colonization Level (Log10 CFU) with Day 0 Community

```{r predict median CFU, fig.width=10,fig.height=10}

### - Uncorrelated OTUs
cfu_uncor_shared_df <- colonized_cfu %>% 
    right_join(community_d0.df) %>% 
    select(-mouse_id, -cage_id, -human_source) %>% 
    filter(!is.na(log_cfu_med))
set.seed(seed)
cfu_uncor_rf <- rfe(cfu_uncor_shared_df[ , -1], cfu_uncor_shared_df$log_cfu_med,
                 sizes = subsets,
                 rfeControl = rfe_ctrl)
cfu_uncor_features <- predictors(cfu_uncor_rf)
cfu_uncor_model <- cfu_uncor_rf$fit
#rf_prediction <- data.frame(observed = cfu_uncor_shared_df$log_cfu_med, 
#                       predicted = rf_n_features$predicted)
#rf_prediction <- merge(rf_prediction, colonized_cfu, by.x = 'observed', by.y = 'log_cfu_med')

### - All OTUs
community_all_d0 <- merge(shared_list_days, rel_abund_shared_01, 
                          by.x = 'sample_id', by.y = 'row.names') %>% 
                     select(-sample_id)

cfu_all_shared_df <- colonized_cfu %>% 
    right_join(community_all_d0) %>% 
    select(-mouse_id, -cage_id, -human_source) %>% 
    filter(!is.na(log_cfu_med))
set.seed(seed)
cfu_all_rf <- rfe(cfu_all_shared_df[ , -1], cfu_all_shared_df$log_cfu_med,
                 sizes = subsets,
                 rfeControl = rfe_ctrl)
cfu_all_features <- predictors(cfu_all_rf)
cfu_all_model <- cfu_all_rf$fit

### - Uncorrelated OTUs + cage
cfu_uncor_cage_shared_df <- colonized_cfu %>% 
    right_join(community_d0.df) %>% 
    select(-mouse_id, -human_source) %>% 
    filter(!is.na(log_cfu_med))
cfu_uncor_cage_shared_df$cage_id <- as.factor(cfu_uncor_cage_shared_df$cage_id)
set.seed(seed)
cfu_uncor_cage_rf <- rfe(cfu_uncor_cage_shared_df[ , -2], cfu_uncor_cage_shared_df$log_cfu_med,
                 sizes = subsets,
                 rfeControl = rfe_ctrl)
cfu_uncor_cage_features <- predictors(cfu_uncor_cage_rf)
cfu_uncor_cage_model <- cfu_uncor_cage_rf$fit
#rf_prediction <- data.frame(observed = cfu_uncor_shared_df$log_cfu_med, 
#                       predicted = rf_n_features$predicted)
#rf_prediction <- merge(rf_prediction, colonized_cfu, by.x = 'observed', by.y = 'log_cfu_med')

### - All OTUs + cage
cfu_all_cage_shared_df <- colonized_cfu %>% 
    right_join(community_all_d0) %>% 
    select(-mouse_id, -human_source) %>% 
    filter(!is.na(log_cfu_med))
cfu_all_cage_shared_df$cage_id <- as.factor(cfu_all_cage_shared_df$cage_id)
set.seed(seed)
cfu_all_cage_rf <- rfe(cfu_all_cage_shared_df[ , -2], cfu_all_cage_shared_df$log_cfu_med,
                 sizes = subsets,
                 rfeControl = rfe_ctrl)
cfu_all_cage_features <- predictors(cfu_all_cage_rf)
cfu_all_cage_model <- cfu_all_cage_rf$fit

### - Uncorrelated OTUs + source
cfu_uncor_src_shared_df <- colonized_cfu %>% 
    right_join(community_d0.df) %>% 
    select(-mouse_id, -cage_id) %>% 
    filter(!is.na(log_cfu_med))
cfu_uncor_src_shared_df$cage_id <- as.factor(cfu_uncor_src_shared_df$cage_id)
set.seed(seed)
cfu_uncor_src_cage_rf <- rfe(cfu_uncor_src_shared_df[ , -2], cfu_uncor_src_shared_df$log_cfu_med,
                 sizes = subsets,
                 rfeControl = rfe_ctrl)
cfu_uncor_src_cage_features <- predictors(cfu_uncor_src_cage_rf)
cfu_uncor_src_cage_model <- cfu_uncor_src_cage_rf$fit
#rf_prediction <- data.frame(observed = cfu_uncor_shared_df$log_cfu_med, 
#                       predicted = rf_n_features$predicted)
#rf_prediction <- merge(rf_prediction, colonized_cfu, by.x = 'observed', by.y = 'log_cfu_med')

### - All OTUs + source
cfu_all_src_cage_shared_df <- colonized_cfu %>% 
    right_join(community_all_d0) %>% 
    select(-mouse_id, -cage_id) %>% 
    filter(!is.na(log_cfu_med))
cfu_all_src_cage_shared_df$cage_id <- as.factor(cfu_all_src_cage_shared_df$cage_id)
set.seed(seed)
cfu_all_src_cage_rf <- rfe(cfu_all_src_cage_shared_df[ , -2], cfu_all_src_cage_shared_df$log_cfu_med,
                 sizes = subsets,
                 rfeControl = rfe_ctrl)
cfu_all_src_cage_features <- predictors(cfu_all_src_cage_rf)
cfu_all_src_cage_model <- cfu_all_src_cage_rf$fit

### - Uncorrelated OTUs + source/cage
cfu_uncor_cg_src_shared_df <- colonized_cfu %>% 
    right_join(community_d0.df) %>% 
    select(-mouse_id) %>% 
    filter(!is.na(log_cfu_med))
cfu_uncor_cg_src_shared_df$cage_id <- as.factor(cfu_uncor_cg_src_shared_df$cage_id)
set.seed(seed)
cfu_uncor_cg_src_cage_rf <- rfe(cfu_uncor_cg_src_shared_df[ , -3], cfu_uncor_cg_src_shared_df$log_cfu_med,
                 sizes = subsets,
                 rfeControl = rfe_ctrl)
cfu_uncor_cg_src_cage_features <- predictors(cfu_uncor_cg_src_cage_rf)
cfu_uncor_cg_src_cage_model <- cfu_uncor_cg_src_cage_rf$fit
#rf_prediction <- data.frame(observed = cfu_uncor_shared_df$log_cfu_med, 
#                       predicted = rf_n_features$predicted)
#rf_prediction <- merge(rf_prediction, colonized_cfu, by.x = 'observed', by.y = 'log_cfu_med')

### - All OTUs + source/cage
cfu_all_cg_src_cage_shared_df <- colonized_cfu %>% 
    right_join(community_all_d0) %>% 
    select(-mouse_id) %>% 
    filter(!is.na(log_cfu_med))
cfu_all_cg_src_cage_shared_df$cage_id <- as.factor(cfu_all_cg_src_cage_shared_df$cage_id)
set.seed(seed)
cfu_all_cg_src_cage_rf <- rfe(cfu_all_cg_src_cage_shared_df[ , -3], cfu_all_cg_src_cage_shared_df$log_cfu_med,
                 sizes = subsets,
                 rfeControl = rfe_ctrl)
cfu_all_cg_src_cage_features <- predictors(cfu_all_cg_src_cage_rf)
cfu_all_cg_src_cage_model <- cfu_all_cg_src_cage_rf$fit

```

##### All OTUs  
Model:  
`r cfu_all_model`  
Features:  
`r cfu_all_features`

####  All OTUs + Cage
Model:  
`r cfu_all_cage_model`  
Features:  
`r cfu_all_cage_features`

####  Uncorrelated OTUs
Model:  
`r cfu_uncor_model`  
Features:  
`r cfu_uncor_features`

####  Uncorrelated OTUs + Cage
Model:  
`r cfu_uncor_cage_model`  
Features:  
`r cfu_uncor_cage_features`

####  All OTUs + Source
Model:  
`r cfu_all_src_model`  
Features:  
`r cfu_all_src_features`

####  Uncorrelated OTUs + Source
Model:  
`r cfu_uncor_src_model`  
Features:  
`r cfu_uncor_src_features`

####  All OTUs + Source/Cage
Model:  
`r cfu_all_cg_src_model`  
Features:  
`r cfu_all_cg_src_features`

####  Uncorrelated OTUs + Source/Cage
Model:  
`r cfu_uncor_cg_src_model`  
Features:  
`r cfu_uncor_cg_src_features`

### 2016-12-05 code/HumanCdGF_model_cfu.Rmd
include source/cage

@@
     filter(!is.na(log_cfu_med))
-cfu_uncor_src_shared_df$cage_id <- as.factor(cfu_uncor_src_shared_df$cage_id)
+cfu_uncor_src_shared_df$human_source <- as.factor(cfu_uncor_src_shared_df$human_source)
 set.seed(seed)
-cfu_uncor_src_cage_rf <- rfe(cfu_uncor_src_shared_df[ , -2], cfu_uncor_src_shared_df$log_cfu_med,
+cfu_uncor_src_rf <- rfe(cfu_uncor_src_shared_df[ , -2], cfu_uncor_src_shared_df$log_cfu_med,
                  sizes = subsets,
                  rfeControl = rfe_ctrl)
-cfu_uncor_src_cage_features <- predictors(cfu_uncor_src_cage_rf)
-cfu_uncor_src_cage_model <- cfu_uncor_src_cage_rf$fit
+cfu_uncor_src_features <- predictors(cfu_uncor_src_rf)
+cfu_uncor_src_model <- cfu_uncor_src_rf$fit
 #rf_prediction <- data.frame(observed = cfu_uncor_shared_df$log_cfu_med, 
@@
 ### - All OTUs + source
-cfu_all_src_cage_shared_df <- colonized_cfu %>% 
+cfu_all_src_shared_df <- colonized_cfu %>% 
     right_join(community_all_d0) %>% 
@@
     filter(!is.na(log_cfu_med))
-cfu_all_src_cage_shared_df$cage_id <- as.factor(cfu_all_src_cage_shared_df$cage_id)
+cfu_all_src_shared_df$human_source <- as.factor(cfu_all_src_shared_df$human_source)
 set.seed(seed)
-cfu_all_src_cage_rf <- rfe(cfu_all_src_cage_shared_df[ , -2], cfu_all_src_cage_shared_df$log_cfu_med,
+cfu_all_src_rf <- rfe(cfu_all_src_shared_df[ , -2], cfu_all_src_shared_df$log_cfu_med,
                  sizes = subsets,
                  rfeControl = rfe_ctrl)
-cfu_all_src_cage_features <- predictors(cfu_all_src_cage_rf)
-cfu_all_src_cage_model <- cfu_all_src_cage_rf$fit
+cfu_all_src_features <- predictors(cfu_all_src_rf)
+cfu_all_src_model <- cfu_all_src_rf$fit
 
@@
 cfu_uncor_cg_src_shared_df$cage_id <- as.factor(cfu_uncor_cg_src_shared_df$cage_id)
+cfu_uncor_cg_src_shared_df$human_source <- as.factor(cfu_uncor_cg_src_shared_df$human_source)
 set.seed(seed)
-cfu_uncor_cg_src_cage_rf <- rfe(cfu_uncor_cg_src_shared_df[ , -3], cfu_uncor_cg_src_shared_df$log_cfu_med,
+cfu_uncor_cg_src_rf <- rfe(cfu_uncor_cg_src_shared_df[ , -3], cfu_uncor_cg_src_shared_df$log_cfu_med,
                  sizes = subsets,
                  rfeControl = rfe_ctrl)
-cfu_uncor_cg_src_cage_features <- predictors(cfu_uncor_cg_src_cage_rf)
-cfu_uncor_cg_src_cage_model <- cfu_uncor_cg_src_cage_rf$fit
+cfu_uncor_cg_src_features <- predictors(cfu_uncor_cg_src_rf)
+cfu_uncor_cg_src_model <- cfu_uncor_cg_src_rf$fit
 #rf_prediction <- data.frame(observed = cfu_uncor_shared_df$log_cfu_med, 
@@
 ### - All OTUs + source/cage
-cfu_all_cg_src_cage_shared_df <- colonized_cfu %>% 
+cfu_all_cg_src_shared_df <- colonized_cfu %>% 
     right_join(community_all_d0) %>% 
@@
     filter(!is.na(log_cfu_med))
-cfu_all_cg_src_cage_shared_df$cage_id <- as.factor(cfu_all_cg_src_cage_shared_df$cage_id)
+cfu_all_cg_src_shared_df$cage_id <- as.factor(cfu_all_cg_src_shared_df$cage_id)
+cfu_all_cg_src_shared_df$human_source <- as.factor(cfu_all_cg_src_shared_df$human_source)
 set.seed(seed)
-cfu_all_cg_src_cage_rf <- rfe(cfu_all_cg_src_cage_shared_df[ , -3], cfu_all_cg_src_cage_shared_df$log_cfu_med,
+cfu_all_cg_src_rf <- rfe(cfu_all_cg_src_shared_df[ , -3], cfu_all_cg_src_shared_df$log_cfu_med,
                  sizes = subsets,
                  rfeControl = rfe_ctrl)
-cfu_all_cg_src_cage_features <- predictors(cfu_all_cg_src_cage_rf)
-cfu_all_cg_src_cage_model <- cfu_all_cg_src_cage_rf$fit
+cfu_all_cg_src_features <- predictors(cfu_all_cg_src_rf)
+cfu_all_cg_src_model <- cfu_all_cg_src_rf$fit
 
@@
 ##### All OTUs
-Model:  
-`r cfu_all_model`  
-Features:  
-`r cfu_all_features`
+
+```{r all_otus}
+print(cfu_all_model)
+print(cfu_all_features)
+```
 
 ####  All OTUs + Cage  
-Model:  
-`r cfu_all_cage_model`  
-Features:  
-`r cfu_all_cage_features`
+```{r all_cg}
+print(cfu_all_cage_model)
+print(cfu_all_cage_features)
+```
 
 ####  Uncorrelated OTUs
-Model:  
-`r cfu_uncor_model`  
-Features:  
-`r cfu_uncor_features`
+
+```{r uncor}
+print(cfu_uncor_model)
+print(cfu_uncor_features)
+```
 
 ####  Uncorrelated OTUs + Cage
-Model:  
-`r cfu_uncor_cage_model`  
-Features:  
-`r cfu_uncor_cage_features`
+```{r uncor_cg}
+print(cfu_uncor_cage_model)
+print(cfu_uncor_cage_features)
+```
 
 ####  All OTUs + Source
-Model:  
-`r cfu_all_src_model`  
-Features:  
-`r cfu_all_src_features`
+
+```{r all_src}
+print(cfu_all_src_model)
+print(cfu_all_src_features)
+```
 
 ####  Uncorrelated OTUs + Source
-Model:  
-`r cfu_uncor_src_model`  
-Features:  
-`r cfu_uncor_src_features`
+
+```{r uncor_src}
+print(cfu_uncor_src_model)
+print(cfu_uncor_src_features)
+```
 
 ####  All OTUs + Source/Cage
-Model:  
-`r cfu_all_cg_src_model`  
-Features:  
-`r cfu_all_cg_src_features`
+
+```{r all_cg_src}
+print(cfu_all_cg_src_model)
+print(cfu_all_cg_src_features)
+```
 
 ####  Uncorrelated OTUs + Source/Cage
-Model:  
-`r cfu_uncor_cg_src_model`  
-Features:  
-`r cfu_uncor_cg_src_features`
+
+```{r uncor_cg_src}
+print(cfu_uncor_cg_src_model)
+print(cfu_uncor_cg_src_features)
+```

### 2016-12-05 submission/Manuscript.Rmd

--- 2016-12-01 submission/Manuscript.Rmd
@@
 
-\vspace{35mm}
+\vspace{40mm}
 
@@
 ### Abstract
-_Clostridium difficile_ causes diarrheal disease when it successfully colonizes a disrupted gut microbial community.
+_Clostridium difficile_ causes diarrheal disease when it successfully colonizes a dysbiotic gut microbial community.
 Current mouse models to study _C.
-difficile_ infection (CDI) rely on pre-treatment with antibiotics to alter the mouse microbiome prior to inoculation.
+difficile_ infection (CDI) rely on pre-treatment with antibiotics to disrupt the mouse microbiome prior to inoculation.
 This model does not allow for analysis of human-associated microbial community members that modulate _C.
 difficile_ colonization and expansion.
-To study human-associated microbes in the context of CDI, we inoculated germ-free C57BL/6 mice with one of 16 human fecal samples from diarrheal or healthy donors and challenged with C.
-difficile 14 days later.
+To study human-associated microbes in the context of CDI, we inoculated germ-free C57BL/6 mice with one of 16 human fecal samples from diarrheal or healthy donors and challenged with _C.
+difficile_ 14 days later.
 Five unique donor-mice combinations resulted in severe CDI while the remaining 11 only experienced mild disease.
@@
 While informative, these studies were limited in their use of only one human donor as input inoculum.
+In order to best understand the impact of _C.
+difficile_ pathogenesis on human disease, we must have a laboratory model that allows for study of a variety of human-derived microbiomes.
 
-Here we designed a laboratory system that allows for studying and modeling microbial coumminity interactions with _C.
-difficile_ infection in mice with human derived microbiomes.
+Here we designed a laboratory model that allows for studying and modeling microbial coumminity interactions with _C.
+difficile_ infection in mice with human-derived microbiomes.
 To test the impact of individual human microbiomes on CDI, we colonized germ-free mice with 16 different human stool donors.
@@
 difficile_ challenge.
-Further, we built a predictive model based on bacterial composition that classified “at-risk” microbiomes prior to infection with _C.
+Additionally, the use of machine-learning models allowed us to build a predictive model that classified “at-risk” microbiomes prior to infection with _C.
 difficile_.
-These findings show that human-associated microbiomes can be at risk for CDI even in the absence of antibiotic-induced dysbiosis.
-Finally, our results indicate that study of mice colonized with human feces can replicate a range of clinical outcomes.
+These findings show that human-associated microbiomes can be at risk for CDI even in the absence of antibiotics and that study of mice colonized with human feces provides a range of clinical outcomes.
 
 ### Results 
-**Germ-free mice inoculated with human feces as model for C.
-difficile infection.** To generate mice with human-derived microbiomes, we inoculated one cage each of gnotobiotic C57/BL6 mice with one of 16 different human fecal donors.
+**Germ-free mice inoculated with human feces as model for _C.
+difficile_ infection.** To generate mice with human-derived microbiomes, we inoculated one cage of gnotobiotic C57/BL6 mice with one of 16 different human fecal donors.
 Five donors were patients that had diarrhea that was not attributable to _C.
@@
 difficile_ was used as a positive control.
-After inoculation with human stool, mice microbiomes were allowed to equilibrate for 14 days.
-Prior to infection, stool samples were taken from each mouse to establish baseline microbiome composition.
-Then, 100 spores of _C.
-difficile_ strain 430 (isolated from the positive control patient) were used to orally infect each mouse.
+After inoculation with human stool, mice were allowed to equilibrate for 14 days.
+Prior to infection, stool samples were taken from each mouse to establish baseline.
+Then, the _C.
+difficile_ strain isolated from the positive control patient’s sample (strain 430) was used to infect each mouse with 100 spores.
 Mice were monitored for weight loss and clinical signs of disease.
 Fecal samples were taken to enumerate _C.
-difficile_ CFU and for microbiome analysis each day up to 10 days post-infection (Fig 1A).
+difficile_ CFU and for microbiome analysis every day for up to 10 days post-infection (Fig 1A).
 To ensure that the donors we selected represented a diverse array of human microbiomes, we sequenced the 16S rRNA genes from donor fecal inocula.
 Ordination of the distances between donor communities showed that the donors each had distinctly different communities, independent of whether the sample came from a sick or healthy person (Fig 1B).
-Likewise, the microbial communities of the mice on day 0 were characterized by sequencing of fecal pellet DNA prior to infection (Table S1).
-Ordination of day 0 communities shows that mice were similar to each other within each cage and donor, but distinct from other donors (Fig 1C).
-This result confirmed that human-associated microbes were able to colonize gnotobiotic mice and provide various initial communities to test _C.
+Likewise, the starting microbial communities of the mice on day 0 were characterizing by sequencing of fecal pellets DNA prior to infection.
+Ordination of all of the mouse communities on day 0 shows that mice were similar to each other within each cage and donor, but distinct from other donors (Fig 1C).
+This result confirmed that human-associated microbes were able to colonize gnotobiotic mice and provide distinct initial communities to test _C.
 difficile_ dynamics.
 
-**C.
-difficile infection in mice with human-derived microbiota cause a range of outcomes.** _C.
+**need to put in a line to reference supplemental table with RA in it**
+
+**_C.
+difficile_ infection in mice with human-derived microbiota cause a range of outcomes.** _C.
 difficile_ colonization was monitored by daily plating of stool pellets for _C.
@@
 Nearly all of the mice were colonized to 10^5^ – 10^7^ CFU by one day post-infection and remained colonized at that level until the end of the experiment (Fig 2A).
-As one indicator of disease, mouse weights were taken each day post-infection and weight-loss was monitored alongside clinical signs.
+As one indicator of disease, mouse weights were taken each day post-infection and weight-loss was monitored alongside clinical signs of disease.
 When mice were judged to be too ill to continue they were humanely euthanized.
@@
 Relative abundance of OTUs with greater than 1% abundance at day 0 were used as input to the model.
-We found that the model explained X% of the variance in colonization levels of _C.
+We found that the model explained 86% of the variance in colonization levels of _C.
 difficile_ (Fig 3A).
-By refining the model to include only the top X predictive OTUs, the model explained X% of the variance.
-Partial dependency plots of the top X OTUs predictive of _C.
-difficile_ colonization reveal that colonization is positively associated with _Bacteriodetes_, _Parabacteriodes_ and _Porphyromonadaceae_ (really?) (Fig 3B).
-Colonization was negatively associated with the presence of _Lachnospiraceae_ (Fig 3B) though none of the human-associated mouse communities were completely resistant to colonization by _C.
+By refining the model to include only the top predictive OTUs and training with one individual mouse per cage, the model explained 77%% of the variance.
+Partial dependency plots of the top 6 OTUs predictive of _C.
+difficile_ colonization reveal that colonization is positively associated with _Enterococcus_, _Akkermansia_ (~2%) and an unidentifiable feature consisting of strongly corrlated OTUs of _Enterobacteriaceae_(13.5%), _Anaeroplasma_ (3.35%), _Prevotella_ (4.10%), _Alloprevotella_ (3.55%), and _Porphyromonadaceae_ (1.2%) (Fig 3B).
+Colonization was negatively associated with the presence of _Blautia_, _Ruminococcaceae_, _Lachnospiraceae_, _Clostridium XVII_, _Clostridium XIVb_, _Clostridium XIVa_, _Clostridiales_, _Oscilibacter_, and an unidentfiable feature representing highly correlated OTUs of _Clostridium XVIII_ (2.55%), _Holdemania_ (1.45%), _Robinsiella_ (2.20%), and _Clostridium IV_ (1.5%) (Fig 3B) though none of the human-associated mouse communities were completely resistant to colonization by _C.
 difficile_ (Fig 2A, 3B).
@@
 difficile_ infection, two donor-mouse microbiomes that were previously shown to be resistant to _C.
-difficile_ strain 430 induced weight loss (Fig 2B) (DA00396, DA00430-wait isnt this a control) and one that was susceptible to severe disease (DA00578) were independently challenged with one of three unique _C.
-difficile_ strains (*table or reference for toxin prod*, etc).
+difficile_ strain 431 induced weight loss (Fig 2B) (DA00396, DA00430) and one that was suscpetible to severe disease (DA00578) were independently challenged with one of three unique _C.
+difficile_ strains (table or reference for toxin prod, etc).
 As done previously, _C.
 difficile_ stool CFU was enumerated daily and mouse weight-loss tracked.
-Infection of resistant mice with virulent strain 458 or the weaker strain 299 resulted in colonization to levels near that of 431 (Fig 5A).
-Accordingly, these mice did not experience severe weight loss or show clinical signs of disease (Fig 5B).
-In contrast, infection of susceptible mice (DA00578, Fig 1B) with pathogenic strain 395 resulted in colonization and severe weight loss by two days post-infection (Fig 5A, 5B).
+Infection of mice with resistant communities with virulent strain 458 or the weaker strain 299 resulted in colonization to levels near that of 431 (Fig 5A) and these mice did not experience severe weight loss or show clinical signs of disease (Fig 5B).
+In contrast, infection of mice harboring microbiomes susceptible to severe disease (DA00578, Fig 1B) with strain 395 resulted in colonization and severe weight loss by two days post-infection (Fig 5A, 5B).
 Strain 395 is similarly virulent as strain 431, and shows a similar infection phenotype in these mice.
@@
 Here we colonized 16 groups of mice with human donor stool from patients with different clinical profiles or healthy volunteers.
-This created a diverse set of human-associated mouse microbiomes that allowed for testing _C.
+This created a diverse set of human-associated microbiomes in the mice that allowed for testing _C.
 difficile_ infection dynamics and community interactions.
 Our results showed that both healthy and diarrheal-associated human communities were susceptible to severe disease.
-Additionally, our results confirmed previous findings that resistance to _C.
-difficile_ disease is dependent on the entire microbial community and not just one or two taxa (cite alyx).
+Additionally, our results confirmed our and others previous studies that proved that resistance to _C.
+difficile_ disease is dependent on the entire microbial community and not just one or two species (cite alyx).
 Further, we were able to learn classification models that identified specific OTUs within susceptible communities that predispose mice to infection.
@@
 
-Here we colonized germ-free mice with a variety of human stool communities and allowed these communities to colonize the mice for 2 weeks, prior to _C.
-difficile_ challenge.
+Here we were able to colonize germ-free mice with a variety of human stool communities from diarrheal or healthy people.
+After allowing these communities to colonize the mice for 2 weeks, they were challenged with _C.
+difficile_.
 It is important to note that the mouse gut environment is different from that of the human, and as such exerts a selective pressure on the human stool microbial community that stabilizes over the 2 weeks.
-Thus these mice cannot be considered "humanized" but rather colonized by human-associated microbes.
+Thus these mice cannot be considered "humanized" but rather containing human-associated microbes.
 Interestingly, clinical status was not associated with proclivity to severe _C.
@@
 _C.
-difficile_ strains are known to vary in pathogenicity, primarily via the amount of toxin that is produced during infection (cite).
-To confirm that disease risk is dependent on the microbiome, we challenged susceptibile mice with an additional pathogenic strain of _C.
-difficile_ and found that the mice indeed fell ill.
-In contrast, we exposed an asymptomatic community to a relatively weak strain of _C.
+difficile_ strains are known to vary in pathogenecity, primarily via the amount of toxin that is produced during infection (cite).
+In these pilot experiments, we challenged a susceptible community with another pathogenic strain of _C.
+difficile_ to confirm that the susceptibility lied in the community and not in the specific challenge strain.
+In contrast, we exposed a resistant/asymptomatic community to a relatively weak strain of _C.
 difficile_ and found that the community could be colonized by this one as well-- but not cause disease.
-This suggests that this particular colony is prone to colonization by any _C.
+This suggests that this colony is prone to colonization by any _C.
 difficile_ strain.
@@
 It has been recently reported that carriage of a non-toxigenic strain of _C.
-difficile_ can prevent additional colonization by a pathogenic strain, suggesting a role of niche-filling or immunity this pre-colonization provides (cite cite).
-One interesting experiment would be to recapitulate this finding in human-associated mouse microbiomes by making a susceptible community resistant to infection.
-Further work with human-associated communities in mice and mathematical modeling will allow for uncovering of complex human-associated disease processes and dynamics as they relate to human _C.
-difficile_ infection.
+difficile_ can prevent additional colonization by a pathogenic strain, suggesting some role of niche-filling or immunity this provides (cite cite).
+It would be interesting to see if we could make a susceptible community resistant to infection and disease by a pathogenic _C.
+difficle_ strain in our human-associated microbiome mice.
+Further work with human-associated communities in mice and mathematical modeling will allow for uncovering of complex human-associated disease processes and dynamics as they relate to human disease.
 
@@
 After gavage, the remaining inoculum was diluted and plated to confirm infection dose.
-Bacteria were cultured anaerobically on TCCFA plates and incubated at 37C (cite plate recipe).
+Bacteria were cultured anaerobically on TCCFA plates and incubated at 37C.
 During infection, two fresh stool pellets were collected from each mouse daily.
@@
 
-•	Sequence curation/mothur, data analysis, code + availability of data(SRA) and code (Github) - Nick
+•	Sequence curation/mothur, 
+•	data analysis, 
+•	code + availability of data(SRA) and code (Github) - Nick
 
 ##Acknowledgments## 
-We would like to thank members of the Schloss lab and Jhansi Leslie for analysis advice.
-We would also like to thank the Microbial Systems Laboratory at the University of Michigan for sequencing assistance.
+We would like to thank members of the Schloss lab and Jhansi Leslie for analysis advice and the Microbial Systems Laboratory at the University of Michigan for sequencing assistance.
 
@@
 Random Forest prediction of _C.
-difficile_ colonization level.** A) OTUs above 1% relative abundance on day 0 were used to predict median log~10~ CFU of _C.
+difficile_ colonization level.** A) OTUs present in at least 10% of samples day 0 were used to predict median log~10~ CFU of _C.
 difficile_ after colonization.
-OTUs were chosen such that they were not predictive of cage or donor.
-Each point is a mouse colored by cage.
 B) Partial dependency plots of the top six predictive OTUs.
-Line displats the partial dependence of log~10~ CFU on the relative abundance of each predictive OUT.
+Line displays the partial dependence of log~10~ CFU on the relative abundance of each predictive OTU.
 Each median log~10~ CFU is plotted against its relative abundance for each predictive OTU.
@@
 **Figure 4.
-Random Forest prediction of CDI severity.** OTUs above 1% relative abundance on day 0 were used to predict disease severity.
+Random Forest prediction of CDI severity.** OTUs present in at least 10% of samples day 0 were used to predict disease severity.
 OTUs were chosen such that they were not predictive of cage or donor.

### 2016-12-06 submission/Manuscript.Rmd
citations complete

--- 2016-12-05 submission/Manuscript.Rmd
@@
 difficile_ pathogenesis and virulence factors.
-Numerous microbiome studies have been performed using this model to determine the antibiotic classes (5), starting microbial community (6) and metabolites (7) that impact development and severity of _C.
+Numerous microbiome studies have been performed using this model to determine the antibiotic classes [@Schubert2015], starting microbial community [@Reeves2011] and metabolites [@Theriot2013] that impact development and severity of _C.
 difficile_ infection.
@@
 Gnotobiotic or germ-free mouse models have been used for a range of studies of CDI, including assessment of species-specific interactions between _C.
-difficile_ and competing microbial community members (8), analysis of nutrient restriction (9), _in vivo_ transcriptomics of _C.
-difficile_ and examination of host immune response to CDI (10).
-Further, CDI therapeutics such as antibiotics and fecal microbiota transplants have been tested extensively in a gnotobiotic-piglet or piglet-to-gnotobiotic-mouse model of disease (11), (12).
-Pigs have a longer digestive tract with components more similar to humans than mice and are typically infected by strains typical in human infection (13).
+difficile_ and competing microbial community members [@Reeves2012], analysis of nutrient restriction [@Ng2013], _in vivo_ transcriptomics of _C.
+difficile_ and examination of host immune response to CDI [@Pawlowski2012].
+Further, CDI therapeutics such as antibiotics and fecal microbiota transplants have been tested extensively in a gnotobiotic-piglet or piglet-to-gnotobiotic-mouse model of disease [@Kim2014;@Diao2016].
+Pigs have a longer digestive tract with components more similar to humans than mice and are typically infected by strains typical in human infection [@Steele2010].
 However, the murine and porcine microbiomes typically do not resemble those of the human gut.
@@
 After challenge with _C.
-difficile_ none of the antibiotic-treated piglets succumbed to infection, while some of the untreated human-colonized pigs did (11).
-In another study, germ-free mice colonized with human feces were bred over several generations to create a cohort of mice with identical human-derived microbiomes (14).
+difficile_ none of the antibiotic-treated piglets succumbed to infection, while some of the untreated human-colonized pigs did [@Kim2014].
+In another study, germ-free mice colonized with human feces were bred over several generations to create a cohort of mice with identical human-derived microbiomes [@Collins2015].
 These mice were subsequently treated with a five-antibiotic cocktail to induce dysbiosis and then were successfully colonized by _C.
-difficile_ (14).
+difficile_ [@Collins2015].
 While informative, these studies were limited in their use of only one human donor as input inoculum.
@@
 difficile_ colonization levels** Previous work in our group has demonstrated that the microbes present in the mouse microbiome prior to _C.
-difficile_ challenge can predict future colonization levels (cite Alyx).
+difficile_ challenge can predict future colonization levels [@Schubert2015].
 To determine if the human-associated microbiome in the mouse could similarly predict _C.
@@
 Using weight loss as a proxy for severity, we built a Random Forest classification model based on relative abundance to determine which community members were associated with either mild or severe disease.
-We optimized the model using the AUC-RF algorithm to generate a Reciever Operating Curve (ROC) with maximal Area Under the Curve (AUC) (cite Calle 2011), as used previously in our group to classify colon cancer outcomes (cite Niel 2015 gen med).
+We optimized the model using the AUC-RF algorithm to generate a Reciever Operating Curve (ROC) with maximal Area Under the Curve (AUC) [@Calle2011], as used previously in our group to classify colon cancer outcomes [@Baxter2016].
 The optimized model used X number of OTUs and identified several that predicted increased weightloss and subsequent disease severity (Fig 4A, 4B).
@@
 (probably also going to want a supplemental table for predictive OTUs for this and Fig 3)
-
-**Susceptibility to infection is community-dependent.** _C.
-difficile_ strains are known to vary in disease phenotypes based both on pathogenicity factors of the strain and host microbiome composition (cite someone).
-To test the susceptibility of human-derived microbiomes to _C.
-difficile_ infection, two donor-mouse microbiomes that were previously shown to be resistant to _C.
-difficile_ strain 430 induced weight loss (Fig 2B) (DA00396, DA00430-wait isnt this a control) and one that was susceptible to severe disease (DA00578) were independently challenged with one of three unique _C.
-difficile_ strains (*table or reference for toxin prod*, etc).
-As done previously, _C.
-difficile_ stool CFU was enumerated daily and mouse weight-loss tracked.
-Infection of resistant mice with virulent strain 458 or the weaker strain 299 resulted in colonization to levels near that of 431 (Fig 5A).
-Accordingly, these mice did not experience severe weight loss or show clinical signs of disease (Fig 5B).
-In contrast, infection of susceptible mice (DA00578, Fig 1B) with pathogenic strain 395 resulted in colonization and severe weight loss by two days post-infection (Fig 5A, 5B).
-Strain 395 is similarly virulent as strain 431, and shows a similar infection phenotype in these mice.
-These preliminary results suggest that human-associated microbiomes may be more predictive of disease severity than the _C.
-difficile_ strain.
 
@@
 Additionally, our results confirmed previous findings that resistance to _C.
-difficile_ disease is dependent on the entire microbial community and not just one or two taxa (cite alyx).
+difficile_ disease is dependent on the entire microbial community and not just one or two taxa [@Schubert2015].
 Further, we were able to learn classification models that identified specific OTUs within susceptible communities that predispose mice to infection.
-By subsequently infecting susceptible or resistant communities with varying strains of _C.
-difficile_ we confirmed that the initial state of the human-associated microbiome is more predictive of disease severity than the strain used.
 
@@
 difficile_.
-This is in contrast to numerous reports of mouse microbiome studies (cite) and the knowledge that in general healthy humans without gut perturbations are resistant to _C.
-difficile_ infection (cite).
+This is in contrast to numerous reports of mouse microbiome studies [@Schubert2015] and the knowledge that in general healthy humans without gut perturbations are resistant to _C.
+difficile_ infection [Heinlen2010].
 On the other hand, it is known that humans, especially health care employees, can be asymptomatic carriers of _C.
-difficile_ (Furyura-Kanamari).
+difficile_ [@FuruyaKanamori2015].
 This may explain the fact that the majority of our groups of mice were colonized by _C.
@@
 
-Finally, our pilot experiments testing the efficiency of infection with different _C.
-difficile_ strains have allowed for testing community versus pathogen factors of infection.
-Our results are consistent with previous reports that the composition of the gut microbiome determines if an individual is successfully infected or not (cite Alyx and others).
-_C.
-difficile_ strains are known to vary in pathogenicity, primarily via the amount of toxin that is produced during infection (cite).
-To confirm that disease risk is dependent on the microbiome, we challenged susceptibile mice with an additional pathogenic strain of _C.
-difficile_ and found that the mice indeed fell ill.
-In contrast, we exposed an asymptomatic community to a relatively weak strain of _C.
-difficile_ and found that the community could be colonized by this one as well-- but not cause disease.
-This suggests that this particular colony is prone to colonization by any _C.
-difficile_ strain.
 Further experiments will test if infecting susceptible communities with weak strains allows the mice to survive.
 It has been recently reported that carriage of a non-toxigenic strain of _C.
-difficile_ can prevent additional colonization by a pathogenic strain, suggesting a role of niche-filling or immunity this pre-colonization provides (cite cite).
+difficile_ can prevent additional colonization or recurrence by a pathogenic strain, suggesting a role of niche-filling or immunity this pre-colonization provides [@Gerding2015].
 One interesting experiment would be to recapitulate this finding in human-associated mouse microbiomes by making a susceptible community resistant to infection.
@@
 _C.
-difficile_ presence was measured using WHATEVER- Cdiff quik check?
-toxin AB?.
+difficile_ presence was measured using C.Diff Quik Chek Complete Assay (Techlab).
 Samples from these patients were stored in Cary-Blair medium at -80C.
@@
 
-**_Clostridium difficile_ and sample processing** For all experiments, 100 spores of _Clostridium difficile_ were used to orally infect mice as described previously (cite alyx).
-The _C.
-difficile_ strains used were acquired from spore stocks of (WHERE) and comprised of strains 431, 395, 299 and 468 (cite?).
+**_Clostridium difficile_ and sample processing** For all experiments, 100 spores of _Clostridium difficile_ were used to orally infect mice as described previously [@Schubert2015].
+_C.
+difficile_ strain 431 is an 027 ribotype clinical isolate isolated from a patient with severe CDI [@Carlson2013].
 After gavage, the remaining inoculum was diluted and plated to confirm infection dose.
-Bacteria were cultured anaerobically on TCCFA plates and incubated at 37C (cite plate recipe).
+Bacteria were cultured anaerobically on TCCFA plates and incubated at 37C [@Bliss1997].
 During infection, two fresh stool pellets were collected from each mouse daily.
@@
 **DNA isolation and sequencing** Bacterial DNA was isolated from banked donor stool and each mouse stool pellet using the MO-BIO PowerSoil DNA isolation kit.
-This DNA was used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described (cite Jim paper).
+This DNA was used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described [@Kozich2013].
 

### 2016-12-07 submission/Manuscript.Rmd

@@
 difficile_ and competing microbial community members [@Reeves2012], analysis of nutrient restriction [@Ng2013], _in vivo_ transcriptomics of _C.
-difficile_ and examination of host immune response to CDI [@Pawlowski2012].
+difficile_ and examination of host immune response to CDI [@Pawlowski2010].
 Further, CDI therapeutics such as antibiotics and fecal microbiota transplants have been tested extensively in a gnotobiotic-piglet or piglet-to-gnotobiotic-mouse model of disease [@Kim2014;@Diao2016].
@@
 This is in contrast to numerous reports of mouse microbiome studies [@Schubert2015] and the knowledge that in general healthy humans without gut perturbations are resistant to _C.
-difficile_ infection [Heinlen2010].
+difficile_ infection [@Heinlen2010].
 On the other hand, it is known that humans, especially health care employees, can be asymptomatic carriers of _C.
@@
 
-Further experiments will test if infecting susceptible communities with weak strains allows the mice to survive.
 It has been recently reported that carriage of a non-toxigenic strain of _C.
@@
 
-**Human samples** Donor stool samples were chosen from a sample bank collected from patients from October 2010 to November 2012 at the University of Michigan.
-Sample collection methods and study was approved by the Institutional Review Board at the University of Michigan.
+**Human samples** Donor stool samples were chosen from a sample bank collected from patients from October 2010 to November 2012 at the University of Michigan Health System.
+Sample collection methods and the study were approved by the Institutional Review Board at the University of Michigan.
 Stool samples from University of Michigan Health System patients were submitted to the microbiology lab for _C.
@@
 difficile_ presence was measured using C.Diff Quik Chek Complete Assay (Techlab).
-Samples from these patients were stored in Cary-Blair medium at -80C.
+Fecal samples from these patients were stored in Cary-Blair medium at -80C.
 All patients consented to participation in the study.
-Donor stool for this study was selected from one patient that tested positive for _C.
+Donor stool was selected from one patient that tested positive for _C.
 difficile_ infection and five patients that presented with diarrhea and tested negative for _C.
@@
 ##Acknowledgments## 
-We would like to thank members of the Schloss lab and Jhansi Leslie for analysis advice.
+We would like to thank members of the Schloss lab and Jhansi Leslie for experimental and analytical advice.
 We would also like to thank the Microbial Systems Laboratory at the University of Michigan for sequencing assistance.
@@
 
-**Figure 5.
-Infection of mice with different _C.
-difficile strains_.** 3 strains of _C.
-difficile_ were used to infect mice colonized with susceptible (DA00578) or resistant (DA00369, DA00430) human donor stool.
-A) _C.
-difficile_ stool CFU was enumerated over 10 days.
-B) Percent weight loss was calculated each day for each mouse.
-In both plots, each mouse is a point and lines represent the mean of each cage.
-
 ####Supplement####

### 2016-12-07 submission/Manuscript.tex

@@
 fecal microbiome transplant is the most effective treatment for this
-disease (cite Anna 2014). The precise mechanisms of colonization
-resistance and restoration of a healthy community are yet to be
-discovered.
+disease {[}5{]}. The precise mechanisms of colonization resistance and
+restoration of a healthy community are yet to be discovered.
 
@@
 conventionally-raised mice with antibiotics either in drinking water or
-by injection to induce susceptibility (3, 4). This model provides a
+by injection to induce susceptibility {[}6,7{]}. This model provides a
 convenient way to study \emph{C. difficile} pathogenesis and virulence
 factors. Numerous microbiome studies have been performed using this
-model to determine the antibiotic classes (5), starting microbial
-community (6) and metabolites (7) that impact development and severity
-of \emph{C. difficile} infection. While informative, these studies are
-somewhat removed from human disease because they only examine
-mouse-associated microbial communities.
+model to determine the antibiotic classes {[}8{]}, starting microbial
+community {[}9{]} and metabolites {[}10{]} that impact development and
+severity of \emph{C. difficile} infection. While informative, these
+studies are somewhat removed from human disease because they only
+examine mouse-associated microbial communities.
 
@@
 between \emph{C. difficile} and competing microbial community members
-(8), analysis of nutrient restriction (9), \emph{in vivo}
+{[}11{]}, analysis of nutrient restriction {[}12{]}, \emph{in vivo}
 transcriptomics of \emph{C. difficile} and examination of host immune
-response to CDI (10). Further, CDI therapeutics such as antibiotics and
-fecal microbiota transplants have been tested extensively in a
-gnotobiotic-piglet or piglet-to-gnotobiotic-mouse model of disease (11),
-(12). Pigs have a longer digestive tract with components more similar to
-humans than mice and are typically infected by strains typical in human
-infection (13). However, the murine and porcine microbiomes typically do
-not resemble those of the human gut.
+response to CDI {[}\textbf{???}{]}. Further, CDI therapeutics such as
+antibiotics and fecal microbiota transplants have been tested
+extensively in a gnotobiotic-piglet or piglet-to-gnotobiotic-mouse model
+of disease {[}13,14{]}. Pigs have a longer digestive tract with
+components more similar to humans than mice and are typically infected
+by strains typical in human infection {[}15{]}. However, the murine and
+porcine microbiomes typically do not resemble those of the human gut.
 
@@
 piglets succumbed to infection, while some of the untreated
-human-colonized pigs did (11). In another study, germ-free mice
+human-colonized pigs did {[}13{]}. In another study, germ-free mice
 colonized with human feces were bred over several generations to create
-a cohort of mice with identical human-derived microbiomes (14). These
-mice were subsequently treated with a five-antibiotic cocktail to induce
-dysbiosis and then were successfully colonized by \emph{C. difficile}
-(14). While informative, these studies were limited in their use of only
-one human donor as input inoculum.
+a cohort of mice with identical human-derived microbiomes {[}16{]}.
+These mice were subsequently treated with a five-antibiotic cocktail to
+induce dysbiosis and then were successfully colonized by \emph{C.
+difficile} {[}16{]}. While informative, these studies were limited in
+their use of only one human donor as input inoculum.
 
@@
 \emph{C. difficile} challenge can predict future colonization levels
-(cite Alyx). To determine if the human-associated microbiome in the
-mouse could similarly predict \emph{C. difficile} colonization we
-employed a Random Forest machine learning algorithm similar to the one
-used in previous studies. Relative abundance of OTUs with greater than
-1\% abundance at day 0 were used as input to the model. We found that
-the model explained X\% of the variance in colonization levels of
-\emph{C. difficile} (Fig 3A). By refining the model to include only the
-top X predictive OTUs, the model explained X\% of the variance. Partial
+{[}8{]}. To determine if the human-associated microbiome in the mouse
+could similarly predict \emph{C. difficile} colonization we employed a
+Random Forest machine learning algorithm similar to the one used in
+previous studies. Relative abundance of OTUs with greater than 1\%
+abundance at day 0 were used as input to the model. We found that the
+model explained X\% of the variance in colonization levels of \emph{C.
+difficile} (Fig 3A). By refining the model to include only the top X
+predictive OTUs, the model explained X\% of the variance. Partial
 dependency plots of the top X OTUs predictive of \emph{C. difficile}
@@
 AUC-RF algorithm to generate a Reciever Operating Curve (ROC) with
-maximal Area Under the Curve (AUC) (cite Calle 2011), as used previously
-in our group to classify colon cancer outcomes (cite Niel 2015 gen med).
-The optimized model used X number of OTUs and identified several that
-predicted increased weightloss and subsequent disease severity (Fig 4A,
-4B). The populations that were most commonly found in mice that
-succumbed to \emph{C. difficile} infection were members of the
-\emph{Clostridiales}, \emph{Bacilliales}, \emph{Streptococcus}
-(Lactobacilles?) and \emph{Bacteriodes}. (need more OTUs here, or number
-that were used for model and then group by rank). The AUC was calculated
-using the Out-Of-Bag (OOB) error for each sample (Fig 4A).
-Cross-validation using 10-fold, leave-one-mouse-out and
-leave-one-cage-out resulted in AUCs not significantly different from the
-OOB (Fig 4A). This model reveals that the levels of X OTUs on day 0 can
-predict disease outcome in mice with human-associated microbiomes.
-(probably also going to want a supplemental table for predictive OTUs
-for this and Fig 3)
-
-\textbf{Susceptibility to infection is community-dependent.} \emph{C.
-difficile} strains are known to vary in disease phenotypes based both on
-pathogenicity factors of the strain and host microbiome composition
-(cite someone). To test the susceptibility of human-derived microbiomes
-to \emph{C. difficile} infection, two donor-mouse microbiomes that were
-previously shown to be resistant to \emph{C. difficile} strain 430
-induced weight loss (Fig 2B) (DA00396, DA00430-wait isnt this a control)
-and one that was susceptible to severe disease (DA00578) were
-independently challenged with one of three unique \emph{C. difficile}
-strains (\emph{table or reference for toxin prod}, etc). As done
-previously, \emph{C. difficile} stool CFU was enumerated daily and mouse
-weight-loss tracked. Infection of resistant mice with virulent strain
-458 or the weaker strain 299 resulted in colonization to levels near
-that of 431 (Fig 5A). Accordingly, these mice did not experience severe
-weight loss or show clinical signs of disease (Fig 5B). In contrast,
-infection of susceptible mice (DA00578, Fig 1B) with pathogenic strain
-395 resulted in colonization and severe weight loss by two days
-post-infection (Fig 5A, 5B). Strain 395 is similarly virulent as strain
-431, and shows a similar infection phenotype in these mice. These
-preliminary results suggest that human-associated microbiomes may be
-more predictive of disease severity than the \emph{C. difficile} strain.
+maximal Area Under the Curve (AUC) {[}17{]}, as used previously in our
+group to classify colon cancer outcomes {[}18{]}. The optimized model
+used X number of OTUs and identified several that predicted increased
+weightloss and subsequent disease severity (Fig 4A, 4B). The populations
+that were most commonly found in mice that succumbed to \emph{C.
+difficile} infection were members of the \emph{Clostridiales},
+\emph{Bacilliales}, \emph{Streptococcus} (Lactobacilles?) and
+\emph{Bacteriodes}. (need more OTUs here, or number that were used for
+model and then group by rank). The AUC was calculated using the
+Out-Of-Bag (OOB) error for each sample (Fig 4A). Cross-validation using
+10-fold, leave-one-mouse-out and leave-one-cage-out resulted in AUCs not
+significantly different from the OOB (Fig 4A). This model reveals that
+the levels of X OTUs on day 0 can predict disease outcome in mice with
+human-associated microbiomes. (probably also going to want a
+supplemental table for predictive OTUs for this and Fig 3)
 
@@
 that resistance to \emph{C. difficile} disease is dependent on the
-entire microbial community and not just one or two taxa (cite alyx).
+entire microbial community and not just one or two taxa {[}8{]}.
 Further, we were able to learn classification models that identified
 specific OTUs within susceptible communities that predispose mice to
-infection. By subsequently infecting susceptible or resistant
-communities with varying strains of \emph{C. difficile} we confirmed
-that the initial state of the human-associated microbiome is more
-predictive of disease severity than the strain used.
+infection.
 
@@
 colonization by \emph{C. difficile}. This is in contrast to numerous
-reports of mouse microbiome studies (cite) and the knowledge that in
+reports of mouse microbiome studies {[}8{]} and the knowledge that in
 general healthy humans without gut perturbations are resistant to
-\emph{C. difficile} infection (cite). On the other hand, it is known
-that humans, especially health care employees, can be asymptomatic
-carriers of \emph{C. difficile} (Furyura-Kanamari). This may explain the
-fact that the majority of our groups of mice were colonized by \emph{C.
-difficile} but did not display outward signs of illness or
+\emph{C. difficile} infection {[}Heinlen2010{]}. On the other hand, it
+is known that humans, especially health care employees, can be
+asymptomatic carriers of \emph{C. difficile} {[}19{]}. This may explain
+the fact that the majority of our groups of mice were colonized by
+\emph{C. difficile} but did not display outward signs of illness or
 infection-associated weight loss (Fig 2).
@@
 
-Finally, our pilot experiments testing the efficiency of infection with
-different \emph{C. difficile} strains have allowed for testing community
-versus pathogen factors of infection. Our results are consistent with
-previous reports that the composition of the gut microbiome determines
-if an individual is successfully infected or not (cite Alyx and others).
-\emph{C. difficile} strains are known to vary in pathogenicity,
-primarily via the amount of toxin that is produced during infection
-(cite). To confirm that disease risk is dependent on the microbiome, we
-challenged susceptibile mice with an additional pathogenic strain of
-\emph{C. difficile} and found that the mice indeed fell ill. In
-contrast, we exposed an asymptomatic community to a relatively weak
-strain of \emph{C. difficile} and found that the community could be
-colonized by this one as well-- but not cause disease. This suggests
-that this particular colony is prone to colonization by any \emph{C.
-difficile} strain. Further experiments will test if infecting
-susceptible communities with weak strains allows the mice to survive. It
-has been recently reported that carriage of a non-toxigenic strain of
-\emph{C. difficile} can prevent additional colonization by a pathogenic
-strain, suggesting a role of niche-filling or immunity this
-pre-colonization provides (cite cite). One interesting experiment would
-be to recapitulate this finding in human-associated mouse microbiomes by
-making a susceptible community resistant to infection. Further work with
+Further experiments will test if infecting susceptible communities with
+weak strains allows the mice to survive. It has been recently reported
+that carriage of a non-toxigenic strain of \emph{C. difficile} can
+prevent additional colonization or recurrence by a pathogenic strain,
+suggesting a role of niche-filling or immunity this pre-colonization
+provides {[}20{]}. One interesting experiment would be to recapitulate
+this finding in human-associated mouse microbiomes by making a
+susceptible community resistant to infection. Further work with
 human-associated communities in mice and mathematical modeling will
@@
 submitted to the microbiology lab for \emph{C. difficile} testing.
-\emph{C. difficile} presence was measured using WHATEVER- Cdiff quik
-check? toxin AB?. Samples from these patients were stored in Cary-Blair
-medium at -80C. All patients consented to participation in the study.
-Donor stool for this study was selected from one patient that tested
-positive for \emph{C. difficile} infection and five patients that
+\emph{C. difficile} presence was measured using C.Diff Quik Chek
+Complete Assay (Techlab). Samples from these patients were stored in
+Cary-Blair medium at -80C. All patients consented to participation in
+the study. Donor stool for this study was selected from one patient that
+tested positive for \emph{C. difficile} infection and five patients that
 presented with diarrhea and tested negative for \emph{C. difficile}.
@@
 experiments, 100 spores of \emph{Clostridium difficile} were used to
-orally infect mice as described previously (cite alyx). The \emph{C.
-difficile} strains used were acquired from spore stocks of (WHERE) and
-comprised of strains 431, 395, 299 and 468 (cite?). After gavage, the
-remaining inoculum was diluted and plated to confirm infection dose.
-Bacteria were cultured anaerobically on TCCFA plates and incubated at
-37C (cite plate recipe). During infection, two fresh stool pellets were
-collected from each mouse daily. One pellet was resuspended in anaerobic
-phosphate-buffered saline and diluted and plated on TCCFA plates to
-enumerate \emph{C. difficile} CFU. The other pellet was frozen at -80C
-until the end of the experiment.
+orally infect mice as described previously {[}8{]}. \emph{C. difficile}
+strain 431 is an 027 ribotype clinical isolate isolated from a patient
+with severe CDI {[}21{]}. After gavage, the remaining inoculum was
+diluted and plated to confirm infection dose. Bacteria were cultured
+anaerobically on TCCFA plates and incubated at 37C {[}22{]}. During
+infection, two fresh stool pellets were collected from each mouse daily.
+One pellet was resuspended in anaerobic phosphate-buffered saline and
+diluted and plated on TCCFA plates to enumerate \emph{C. difficile} CFU.
+The other pellet was frozen at -80C until the end of the experiment.
 
@@
 amplification of the V4 region of the 16S rRNA gene and fragments were
-sequenced on an Illumina MiSeq as previously described (cite Jim paper).
+sequenced on an Illumina MiSeq as previously described {[}23{]}.
 
@@
 We would like to thank members of the Schloss lab and Jhansi Leslie for
-analysis advice. We would also like to thank the Microbial Systems
-Laboratory at the University of Michigan for sequencing assistance.
+experimental and analytical advice. We would also like to thank the
+Microbial Systems Laboratory at the University of Michigan for
+sequencing assistance.
 
@@
 
-\textbf{Figure 5. Infection of mice with different \emph{C. difficile
-strains}.} 3 strains of \emph{C. difficile} were used to infect mice
-colonized with susceptible (DA00578) or resistant (DA00369, DA00430)
-human donor stool. A) \emph{C. difficile} stool CFU was enumerated over
-10 days. B) Percent weight loss was calculated each day for each mouse.
-In both plots, each mouse is a point and lines represent the mean of
-each cage.
-
 \paragraph{Supplement}\label{supplement}
@@
 
+5. Seekatz AM, Aas J, Gessert CE, Rubin TA, Saman DM, Bakken JS, et al.
+Recovery of the gut microbiome following fecal microbiota
+transplantation. mBio. American Society for Microbiology; 2014;5:
+e00893--14--e00893--14.
+doi:\href{http://dx.doi.org/10.1128/mbio.00893-14}{10.1128/mbio.00893-14}
+
+6. Buffie CG, Jarchum I, Equinda M, Lipuma L, Gobourne A, Viale A, et
+al. Profound alterations of intestinal microbiota following a single
+dose of clindamycin results in sustained susceptibility to clostridium
+difficile-induced colitis. Infection and Immunity. American Society for
+Microbiology; 2011;80: 62--73.
+doi:\href{http://dx.doi.org/10.1128/iai.05496-11}{10.1128/iai.05496-11}
+
+7. Theriot CM, Koumpouras CC, Carlson PE, Bergin II, Aronoff DM, Young
+VB. Cefoperazone-treated mice as an experimental platform to assess
+differential virulence of clostridium difficile strains. Gut Microbes.
+Informa UK Limited; 2011;2: 326--334.
+doi:\href{http://dx.doi.org/10.4161/gmic.19142}{10.4161/gmic.19142}
+
+8. Schubert AM, Sinani H, Schloss PD. Antibiotic-induced alterations of
+the murine gut microbiota and subsequent effects on colonization
+resistance against clostridium difficile. mBio. American Society for
+Microbiology; 2015;6: e00974--15.
+doi:\href{http://dx.doi.org/10.1128/mbio.00974-15}{10.1128/mbio.00974-15}
+
+9. Reeves AE, Theriot CM, Bergin IL, Huffnagle GB, Schloss PD, Young VB.
+The interplay between microbiome dynamics and pathogen dynamics in a
+murine model of clostridium difficile infection. Gut Microbes. Informa
+UK Limited; 2011;2: 145--158.
+doi:\href{http://dx.doi.org/10.4161/gmic.2.3.16333}{10.4161/gmic.2.3.16333}
+
+10. Theriot CM, Young VB. Microbial and metabolic interactions between
+the gastrointestinal tract and clostridium difficile infection. Gut
+Microbes. Informa UK Limited; 2013;5: 86--95.
+doi:\href{http://dx.doi.org/10.4161/gmic.27131}{10.4161/gmic.27131}
+
+11. Reeves AE, Koenigsknecht MJ, Bergin IL, Young VB. Suppression of
+clostridium difficile in the gastrointestinal tracts of germfree mice
+inoculated with a murine isolate from the family lachnospiraceae.
+Infection and Immunity. American Society for Microbiology; 2012;80:
+3786--3794.
+doi:\href{http://dx.doi.org/10.1128/iai.00647-12}{10.1128/iai.00647-12}
+
+12. Ng KM, Ferreyra JA, Higginbottom SK, Lynch JB, Kashyap PC, Gopinath
+S, et al. Microbiota-liberated host sugars facilitate post-antibiotic
+expansion of enteric pathogens. Nature. Springer Nature; 2013;502:
+96--99.
+doi:\href{http://dx.doi.org/10.1038/nature12503}{10.1038/nature12503}
+
+13. Kim HB, Zhang Q, Sun X, Beamer G, Wang Y, Tzipori S. Beneficial
+effect of oral tigecycline treatment on clostridium difficile infection
+in gnotobiotic piglets. Antimicrobial Agents and Chemotherapy. American
+Society for Microbiology; 2014;58: 7560--7564.
+doi:\href{http://dx.doi.org/10.1128/aac.03447-14}{10.1128/aac.03447-14}
+
+14. Diao H, Yan HL, Xiao Y, Yu B, Yu J, He J, et al. Intestinal
+microbiota could transfer host gut characteristics from pigs to mice.
+BMC Microbiology. Springer Nature; 2016;16.
+doi:\href{http://dx.doi.org/10.1186/s12866-016-0851-z}{10.1186/s12866-016-0851-z}
+
+15. Steele J, Feng H, Parry N, Tzipori S. Piglet models of acute or
+chronic clostridium difficile illness. The Journal of Infectious
+Diseases. Oxford University Press (OUP); 2010;201: 428--434.
+doi:\href{http://dx.doi.org/10.1086/649799}{10.1086/649799}
+
+16. Collins J, Auchtung JM, Schaefer L, Eaton KA, Britton RA. Humanized
+microbiota mice as a model of recurrent clostridium difficile disease.
+Microbiome. Springer Nature; 2015;3.
+doi:\href{http://dx.doi.org/10.1186/s40168-015-0097-2}{10.1186/s40168-015-0097-2}
+
+17. Calle ML, Urrea V, Boulesteix A-L, Malats N. AUC-RF: A new strategy
+for genomic profiling with random forest. Human Heredity. S. Karger AG;
+2011;72: 121--132.
+doi:\href{http://dx.doi.org/10.1159/000330778}{10.1159/000330778}
+
+18. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
+improves the sensitivity of fecal immunochemical test for detecting
+colonic lesions. Genome Medicine. Springer Nature; 2016;8.
+doi:\href{http://dx.doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}
+
+19. Furuya-Kanamori L, Marquess J, Yakob L, Riley TV, Paterson DL,
+Foster NF, et al. Asymptomatic clostridium difficile colonization:
+Epidemiology and clinical implications. BMC Infectious Diseases.
+Springer Nature; 2015;15.
+doi:\href{http://dx.doi.org/10.1186/s12879-015-1258-4}{10.1186/s12879-015-1258-4}
+
+20. Gerding DN, Meyer T, Lee C, Cohen SH, Murthy UK, Poirier A, et al.
+Administration of spores of nontoxigenic clostridium difficile strain m3
+for prevention of recurrent c difficile infection. JAMA. American
+Medical Association (AMA); 2015;313: 1719.
+doi:\href{http://dx.doi.org/10.1001/jama.2015.3725}{10.1001/jama.2015.3725}
+
+21. Carlson PE, Walk ST, Bourgis AE, Liu MW, Kopliku F, Lo E, et al. The
+relationship between phenotype, ribotype, and clinical disease in human
+clostridium difficile isolates. Anaerobe. Elsevier BV; 2013;24:
+109--116.
+doi:\href{http://dx.doi.org/10.1016/j.anaerobe.2013.04.003}{10.1016/j.anaerobe.2013.04.003}
+
+22. Bliss DZ, Johnson S, Clabots CR, Savik K, Gerding DN. Comparison of
+cycloserine-cefoxitin-fructose agar (CCFA) and taurocholate-CCFA for
+recovery of clostridium difficile during surveillance of hospitalized
+patients. Diagnostic Microbiology and Infectious Disease. Elsevier BV;
+1997;29: 1--4.
+doi:\href{http://dx.doi.org/10.1016/s0732-8893(97)00113-2}{10.1016/s0732-8893(97)00113-2}
+
+23. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
+Development of a dual-index sequencing strategy and curation pipeline
+for analyzing amplicon sequence data on the MiSeq illumina sequencing
+platform. Applied and Environmental Microbiology. American Society for
+Microbiology; 2013;79: 5112--5120.
+doi:\href{http://dx.doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}
+
 \end{document}

### 2016-12-07 Projects/HumanCdGF_cdiff_colonization.Rmd
use RF strictly as OTU selection

---
title: "Model_CFU"
author: "Nicholas Lesniak"
date: "December 5, 2016"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE, fig.width=7, fig.height=8)
```

```{r setup environment}
pack_used <- c('randomForest','tidyr','ggplot2','plyr', 'dplyr', 'caret')
for (dep in pack_used){
  if (dep %in% installed.packages()[,"Package"] == FALSE){
    install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
                     quiet=TRUE);
  }
  library(dep, verbose=FALSE, character.only=TRUE)
}
```

```{r setup data}
# read in files
setwd('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/')
meta_data   <- 'data/process/human_CdGF_metadata.txt'
shared_file <- 'data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
tax_file <- 'data/process/gf_new.an.taxonomy'
tax_function <- 'code/tax_level.R'

# read in files
meta_data   <- read.table(meta_data, sep = '\t', header = T, stringsAsFactors = F)
shared_file <- read.table(shared_file, sep = '\t', header = T, row.names = 'sample_id')
tax_file <- read.table(tax_file, sep = '\t', header = T, row.names = 'OTU')
source(tax_function)

# set parameters
seed <- 1
#n_trees <- 1001
#n_otus <- 12
#iters <- 100
#cor_cutoff <- 0.75
otu_presence_cutoff <- 0.1

# remove samples not used or with other cdiff strains
meta_data <- 
  meta_data %>% 
  filter(Early_Euth == FALSE, cdiff_strain == 431, !cage_id == 'NP1') %>% 
  select(mouse_id, sample_id, cage_id, human_source, day, log_cfu) %>% 
  filter(!is.na(log_cfu))
shared_list_days <- meta_data %>% 
  filter(day == 0) %>% 
  select(-day, -log_cfu, -cage_id, -human_source)
shared_file <- shared_file[rownames(shared_file) %in% shared_list_days$sample_id, ]

# create shared with relative abundances over 1%
mean_otu_counts <- unique(apply(shared_file, 1, sum))
otu_presence <- ifelse(shared_file > 0, T, F) 
rel_abund_shared <- shared_file[ , apply(otu_presence, 2, sum) > round(otu_presence_cutoff*nrow(shared_file))]
rel_abund_shared_01 <- 100*rel_abund_shared/mean_otu_counts

n_features <- round(0.1*ncol(rel_abund_shared))

## remove correlative otus
#rel_abund_cor <- cor(rel_abund_shared_01, method = 'spearman')
#rownames(rel_abund_cor) <- colnames(rel_abund_shared_01) # add row names to identify otus correlated with features
#cor_otus <- findCorrelation(rel_abund_cor, cutoff = cor_cutoff)
#rel_abund_01_filtr <- rel_abund_shared_01[,-cor_otus]
#community_d0.df <- merge(shared_list_days, rel_abund_01_filtr, 
#                         by.x = 'sample_id', by.y = 'row.names') %>% 
#                    select(-sample_id)

colonized_cfu <- meta_data %>% 
  filter(day %in% c(4:10)) %>% 
  group_by(mouse_id, cage_id, human_source) %>% 
  summarise(log_cfu_med = median(log_cfu)) %>% 
  ungroup()

community_all_d0 <- merge(shared_list_days, rel_abund_shared_01, 
                          by.x = 'sample_id', by.y = 'row.names') %>% 
                     select(-sample_id)
```

### Predicting Colonization Level (Log10 CFU) with Day 0 Community

```{r predict median CFU, fig.width=10,fig.height=10}
cfu_all_shared_df <- colonized_cfu %>% 
    select(mouse_id, log_cfu_med) %>% 
    right_join(community_all_d0) %>% 
    select(-mouse_id) %>% 
    filter(!is.na(log_cfu_med))

#output <- data.frame(ntree = NA, ntry = NA, rsq = NA)
#for(tree in seq(1, 1500, 100)){
#  for(try in 1:30){
#    set.seed(seed)
#    cfu_all_rf <- randomForest(cfu_all_shared_df[ , -1], cfu_all_shared_df$log_cfu_med,
#                           ntree = tree,
#                           mtry = try,
#                           importance = T)
#    output <- rbind(output, c(tree, try, tail(cfu_all_rf$rsq, 1)))
#  }
#}
#
#ggplot(output[output$ntree !=1,], aes(x=ntree, y=rsq, group = ntry, color = as.factor(ntry))) + geom_line()
#ggplot(output[output$ntree == 1001,], aes(x=ntry, y=rsq)) + geom_line()

set.seed(seed)
cfu_all_rf <- randomForest(cfu_all_shared_df[ , -1], cfu_all_shared_df$log_cfu_med,
                           ntree = 1001,
                           mtry = 15,
                           importance = T)
top_features <- head(sort(importance(cfu_all_rf)[,1],decreasing = T), n_features)
feat_cor_cfu <- sort(abs(cor(cfu_all_shared_df[ , c('log_cfu_med', 
                                                    names(top_features))],
                             method = 'spearman')[,'log_cfu_med']),
                     decreasing = T)
```

OTUs predictive of _Colonization level_:  
`r names(top_features)`

```{r predict inocula, fig.width=10,fig.height=10}
source_all_shared_df <- colonized_cfu %>% 
    select(mouse_id, human_source) %>% 
    right_join(community_all_d0) %>% 
    select(-mouse_id)
source_all_shared_df$human_source <- as.factor(source_all_shared_df$human_source)

#output <- data.frame(ntree = NA, ntry = NA, err = NA)
#for(tree in seq(1, 1500, 100)){
#  for(try in 1:30){
#    set.seed(seed)
#    source_all_rf <- randomForest(source_all_shared_df[ , -1], source_all_shared_df$human_source,
#                           ntree = tree,
#                           mtry = try,
#                           importance = T)
#    output <- rbind(output, c(tree, try, tail(source_all_rf$err.rate[,'OOB'], 1)))
#  }
#}
#
#ggplot(output, aes(x=ntry, y=err, group = ntree, color = as.factor(ntree))) + geom_line()
#ggplot(output[output$ntry == 15,], aes(x=ntree, y=err)) + geom_line()

set.seed(seed)
source_all_rf <- randomForest(source_all_shared_df[ , -1], source_all_shared_df$human_source,
                           ntree = 501,
                           mtry = 15,
                           importance = T)
top_features_src <- head(sort(importance(source_all_rf)[,1],decreasing = T), n_features)
#any(top_features %in% top_features_src)
src_feat_cor_cfu <- sort(abs(cor(cfu_all_shared_df[ , c('log_cfu_med', 
                                                        names(top_features_src))], 
                                 method = 'spearman')[,'log_cfu_med']), 
                         decreasing = T)
```

OTUs predictive of _Inocula Source_:  
`r names(top_features_src)`

### 2016-12-07 Projects/HumanCdGF_cdiff_colonization.Rmd
correlation plots for otus correlated to cfu

@@
 
-OTUs predictive of _Colonization level_:  
+#### Predictive OTUs  
+##### OTUs predictive of _Colonization level_:  
 `r names(top_features)`
@@
 top_features_src <- head(sort(importance(source_all_rf)[,1],decreasing = T), n_features)
-#any(top_features %in% top_features_src)
+#any(names(top_features) %in% names(top_features_src))
 src_feat_cor_cfu <- sort(abs(cor(cfu_all_shared_df[ , c('log_cfu_med', 
@@
 
-OTUs predictive of _Inocula Source_:  
+##### OTUs predictive of _Inocula Source_:  
 `r names(top_features_src)`
+
+There are `r sum(names(top_features) %in% names(top_features_src))` OTUs that overlap between predicting Colonization and Source:  
+`r names(top_features)[names(top_features) %in% names(top_features_src)]`
+
+#### Predictive OTUs that correlate with CFU  
+The following predictive OTUs correlate with CFU (Spearman's _rho_ > 0.5)
+
+Colonization Level  
+`r names(feat_cor_cfu)[feat_cor_cfu>0.5 & feat_cor_cfu < 1]`
+
+```{r plot colonization otu correlations, fig.width=10, fig.height=18}
+subset(cfu_all_shared_df, select = 
+       c("log_cfu_med", names(feat_cor_cfu)[feat_cor_cfu>0.5 & feat_cor_cfu < 1])) %>% 
+  gather(OTU, abundance, contains('Otu')) %>% 
+  ggplot(aes(x = abundance, y = log_cfu_med)) + geom_point() + 
+    facet_wrap(~OTU, ncol = 4, scales = 'free_x')
+```
+
+Inocula Source  
+`r names(src_feat_cor_cfu)[src_feat_cor_cfu>0.5 & src_feat_cor_cfu < 1]`
+
+```{r plot source otu correlations, fig.width=7.5, fig.height=3}
+subset(cfu_all_shared_df, select = 
+       c("log_cfu_med", names(src_feat_cor_cfu)[src_feat_cor_cfu>0.5 & src_feat_cor_cfu < 1])) %>% 
+  gather(OTU, abundance, contains('Otu')) %>% 
+  ggplot(aes(x = abundance, y = log_cfu_med)) + geom_point() + 
+    facet_wrap(~OTU, ncol = 3, scales = 'free_x')
+```

### 2016-12-07 Projects/HumanCdGF_cdiff_colonization.Rmd
add importance plots

@@
 #### Predictive OTUs  
-##### OTUs predictive of _Colonization level_:  
-`r names(top_features)`
+##### OTUs predictive of _Colonization level_ (Top 10%)
+
+```{r colonization otu importance, fig.width=6, fig.height=6}
+top_otus <- get_tax(row_list = names(top_features))$tax_label
+
+data.frame(OTU = factor(top_otus, 
+                        levels = top_otus[order(top_features)]), 
+           Importance = top_features/max(top_features)) %>% 
+  ggplot(aes(x = OTU, y = Importance)) + 
+    geom_point() + coord_flip() + 
+    labs(y = 'Importance (Relative MDA)', x = NULL) + 
+    theme(panel.background = element_rect(linetype = 1, size = 2, fill = 'white', color = 'black'),
+          panel.grid.major.x = element_blank(),
+          panel.grid.minor.x = element_blank(),
+          panel.grid.major.y = element_line(colour = "gray", linetype = "dotted"),
+          panel.grid.minor.y = element_line(colour = "gray", linetype = "dotted"),
+          axis.text.y = element_text(color = c(rep('black', length(top_features)-12),
+                                               rep('red', 12)), 
+                                     hjust = 0))
+```
 
@@
 
-##### OTUs predictive of _Inocula Source_:  
-`r names(top_features_src)`
+##### OTUs predictive of _Inocula Source_ (Top 10%)
+
+```{r source otu importance, fig.width=6, fig.height=6}
+top_otus_src <- get_tax(row_list = names(top_features_src))$tax_label
+
+data.frame(OTU = factor(top_otus_src, 
+                        levels = top_otus_src[order(top_features_src)]), 
+           Importance = top_features_src/max(top_features_src)) %>% 
+  ggplot(aes(x = OTU, y = Importance)) + 
+    geom_point() + coord_flip() + 
+    labs(y = 'Importance (Relative MDA)', x = NULL) + 
+    theme(panel.background = element_rect(linetype = 1, size = 2, fill = 'white', color = 'black'),
+          panel.grid.major.x = element_blank(),
+          panel.grid.minor.x = element_blank(),
+          panel.grid.major.y = element_line(colour = "gray", linetype = "dotted"),
+          panel.grid.minor.y = element_line(colour = "gray", linetype = "dotted"),
+          axis.text.y = element_text(color = c(rep('black', length(top_features)-12),
+                                               rep('red', 12)), 
+                                     hjust = 0))
+```
 
@@
 subset(cfu_all_shared_df, select = 
-       c("log_cfu_med", names(feat_cor_cfu)[feat_cor_cfu>0.5 & feat_cor_cfu < 1])) %>% 
+       c("log_cfu_med", names(head(top_features,12)))) %>% 
   gather(OTU, abundance, contains('Otu')) %>% 
   ggplot(aes(x = abundance, y = log_cfu_med)) + geom_point() + 
-    facet_wrap(~OTU, ncol = 4, scales = 'free_x')
+    facet_wrap(~OTU, ncol = 4, scales = 'free_x') + scale_x_log10()
 ```

### 2016-12-07 Projects/HumanCdGF_cdiff_colonization.Rmd
all correlation plots for top 12 otus

@@
 top_features <- head(sort(importance(cfu_all_rf)[,1],decreasing = T), n_features)
-feat_cor_cfu <- sort(abs(cor(cfu_all_shared_df[ , c('log_cfu_med', 
+feat_cor_cfu <- cor(cfu_all_shared_df[ , c('log_cfu_med', 
                                            names(top_features))],
-                             method = 'spearman')[,'log_cfu_med']),
-                     decreasing = T)
+                    method = 'spearman')[-1,'log_cfu_med']
 ```
@@
 ```
+
+```{r plot colonization otu correlations, fig.width=10, fig.height=10}
+top_n_otus <- get_tax(row_list = names(head(top_features,12)))
+top_n_otus$cor_label <- paste0(top_n_otus$tax, '(OTU',
+                           gsub('Otu0*', '', rownames(top_n_otus)),
+                           ';rho=', round(feat_cor_cfu[names(feat_cor_cfu) 
+                                                       %in% rownames(top_n_otus)], 
+                                          2), ')')
+top_n_otus$cor_label <- factor(top_n_otus$cor_label,
+                               levels = top_n_otus$cor_label[order(head(top_features,12),
+                                                                   decreasing = T)])
+top_n_otus$OTU <- rownames(top_n_otus)
+subset(cfu_all_shared_df, select = 
+       c("log_cfu_med", rownames(top_n_otus))) %>% 
+  gather(OTU, abundance, contains('Otu')) %>%
+  left_join(top_n_otus) %>% 
+  ggplot(aes(x = abundance, y = log_cfu_med)) + geom_point() + 
+    facet_wrap(~cor_label, ncol = 4, scales = 'free_x') + 
+    labs(x= 'Relative Abundance - Day 0 (log %)', y = 'Observed Colonization (log CFU)') + 
+    ylim(5, 9) + scale_x_log10(breaks = c(0, 0.01, 0.1, 1, 10, 100)) +
+    theme(strip.background = element_rect(colour=NULL, fill='white'),
+          panel.background = element_rect(linetype = 1, size = 1, fill = 'white', color = 'black'),
+          panel.grid.major.y = element_blank(),
+          panel.grid.minor.y = element_blank(),
+          panel.grid.major.x = element_line(linetype = 1, size = .1, color = 'gray'),
+          panel.grid.minor.x = element_line(linetype = 1, size = .1, color = 'gray'))
+```
+
+***
 
@@
 #any(names(top_features) %in% names(top_features_src))
-src_feat_cor_cfu <- sort(abs(cor(cfu_all_shared_df[ , c('log_cfu_med', 
+src_feat_cor_cfu <- cor(cfu_all_shared_df[ , c('log_cfu_med', 
                                                names(top_features_src))], 
-                                 method = 'spearman')[,'log_cfu_med']), 
-                         decreasing = T)
+                        method = 'spearman')[-1,'log_cfu_med']
 ```
@@
 
+```{r plot source otu correlations, fig.width=10, fig.height=10}
+top_n_otus_src <- get_tax(row_list = names(head(top_features_src,12)))
+top_n_otus_src$cor_label <- paste0(top_n_otus_src$tax, '(OTU',
+                           gsub('Otu0*', '', rownames(top_n_otus_src)),
+                           ';rho=', round(src_feat_cor_cfu[names(src_feat_cor_cfu) 
+                                                       %in% rownames(top_n_otus_src)], 
+                                          2), ')')
+top_n_otus_src$OTU <- rownames(top_n_otus_src)
+top_n_otus_src$cor_label <- factor(top_n_otus_src$cor_label,
+                               levels = top_n_otus_src$cor_label[order(head(top_features_src,12),
+                                                                   decreasing = T)])
+
+subset(cfu_all_shared_df, select = 
+       c("log_cfu_med", rownames(top_n_otus_src))) %>% 
+  gather(OTU, abundance, contains('Otu')) %>%
+  left_join(top_n_otus_src) %>% 
+  ggplot(aes(x = abundance, y = log_cfu_med)) + geom_point() + 
+    facet_wrap(~cor_label, ncol = 4, scales = 'free_x') +  
+    labs(x= 'Relative Abundance - Day 0 (log %)', y = 'Observed Colonization (log CFU)') + 
+    ylim(5, 9) + scale_x_log10(breaks = c(0, 0.01, 0.1, 1, 10, 100)) +
+    theme(strip.background = element_rect(colour=NULL, fill='white'),
+          panel.background = element_rect(linetype = 1, size = 1, fill = 'white', color = 'black'),
+          panel.grid.major.y = element_blank(),
+          panel.grid.minor.y = element_blank(),
+          panel.grid.major.x = element_line(linetype = 1, size = .1, color = 'gray'),
+          panel.grid.minor.x = element_line(linetype = 1, size = .1, color = 'gray'))
+```
+
+***
+
 There are `r sum(names(top_features) %in% names(top_features_src))` OTUs that overlap between predicting Colonization and Source:  
 `r names(top_features)[names(top_features) %in% names(top_features_src)]`
-
-#### Predictive OTUs that correlate with CFU  
-The following predictive OTUs correlate with CFU (Spearman's _rho_ > 0.5)
-
-Colonization Level  
-`r names(feat_cor_cfu)[feat_cor_cfu>0.5 & feat_cor_cfu < 1]`
-
-```{r plot colonization otu correlations, fig.width=10, fig.height=18}
-subset(cfu_all_shared_df, select = 
-       c("log_cfu_med", names(head(top_features,12)))) %>% 
-  gather(OTU, abundance, contains('Otu')) %>% 
-  ggplot(aes(x = abundance, y = log_cfu_med)) + geom_point() + 
-    facet_wrap(~OTU, ncol = 4, scales = 'free_x') + scale_x_log10()
-```
-
-Inocula Source  
-`r names(src_feat_cor_cfu)[src_feat_cor_cfu>0.5 & src_feat_cor_cfu < 1]`
-
-```{r plot source otu correlations, fig.width=7.5, fig.height=3}
-subset(cfu_all_shared_df, select = 
-       c("log_cfu_med", names(src_feat_cor_cfu)[src_feat_cor_cfu>0.5 & src_feat_cor_cfu < 1])) %>% 
-  gather(OTU, abundance, contains('Otu')) %>% 
-  ggplot(aes(x = abundance, y = log_cfu_med)) + geom_point() + 
-    facet_wrap(~OTU, ncol = 3, scales = 'free_x')
-```

### 2016-12-08 Projects/HumanCdGF_cdiff_colonization.Rmd
add 0 to log corr plots

@@
 
-```{r plot colonization otu correlations, fig.width=10, fig.height=10}
+```{r plot colonization otu correlations, fig.width=10, fig.height=8}
 top_n_otus <- get_tax(row_list = names(head(top_features,12)))
@@
                            gsub('Otu0*', '', rownames(top_n_otus)),
-                           ';rho=', round(feat_cor_cfu[names(feat_cor_cfu) 
+                           ';rho= ', round(feat_cor_cfu[names(feat_cor_cfu) 
                                                        %in% rownames(top_n_otus)], 
@@
   left_join(top_n_otus) %>% 
+  mutate(abundance = abundance + 0.001) %>% 
   ggplot(aes(x = abundance, y = log_cfu_med)) + geom_point() + 
@@
     labs(x= 'Relative Abundance - Day 0 (log %)', y = 'Observed Colonization (log CFU)') + 
-    ylim(5, 9) + scale_x_log10(breaks = c(0, 0.01, 0.1, 1, 10, 100)) +
+    ylim(5, 9) + scale_x_log10(breaks = c(0.001, 0.01, 0.1, 1, 10, 100),
+                               labels = c(0, 0.01, 0.1, 1, 10, 100)) +
+    geom_vline(xintercept = 0.0015, color = 'gray') + 
     theme(strip.background = element_rect(colour=NULL, fill='white'),
@@
           panel.grid.minor.y = element_blank(),
-          panel.grid.major.x = element_line(linetype = 1, size = .1, color = 'gray'),
-          panel.grid.minor.x = element_line(linetype = 1, size = .1, color = 'gray'))
+          panel.grid.major.x = #element_line(linetype = 1, size = .1, color = 'gray'),
+                               element_blank(),    
+          panel.grid.minor.x = element_blank())
 ```
@@
 
-```{r plot source otu correlations, fig.width=10, fig.height=10}
+```{r plot source otu correlations, fig.width=10, fig.height=8}
 top_n_otus_src <- get_tax(row_list = names(head(top_features_src,12)))
@@
                            gsub('Otu0*', '', rownames(top_n_otus_src)),
-                           ';rho=', round(src_feat_cor_cfu[names(src_feat_cor_cfu) 
+                           ';rho= ', round(src_feat_cor_cfu[names(src_feat_cor_cfu) 
                                                        %in% rownames(top_n_otus_src)], 
@@
   left_join(top_n_otus_src) %>%
+  mutate(abundance = abundance + 0.001) %>% 
   ggplot(aes(x = abundance, y = log_cfu_med)) + geom_point() + 

### 2016-12-09 Projects/HumanCdGF_cdiff_colonization.Rmd
all model performance plots

@@
 
+```{r prediction cfu plot, fig.width=6, fig.height=6}
+
+data.frame(Observed = cfu_all_shared_df$log_cfu_med, 
+           Predicted = cfu_all_rf$predicted) %>% 
+  ggplot(aes(x = Predicted, y = Observed)) + geom_point() +
+    xlim(5,9) + ylim(5,9) + 
+    annotate('text', label = paste0(round(tail(cfu_all_rf$rsq, 1), 4)*100, '% Var Explained'),
+             x = min(cfu_all_rf$predicted), y = max(cfu_all_shared_df$log_cfu_med)) + 
+    theme(panel.background = element_rect(linetype = 1, size = 1, fill = 'white', color = 'black'),
+          panel.grid.major.x = element_blank(),
+          panel.grid.minor.x = element_blank(),
+          panel.grid.major.y = element_blank(),
+          panel.grid.minor.y = element_blank()) +
+  labs(x = 'Predicted Colonization (log CFU)', y = 'Observed Colonization (log CFU)')
+
+```
+
+```{r colonization performance, fig.width=8, fig.height=6}
+top_otus <- get_tax(row_list = names(top_features))$tax_label
+```
+
 ```{r colonization otu importance, fig.width=6, fig.height=6}
-top_otus <- get_tax(row_list = names(top_features))$tax_label
-
 data.frame(OTU = factor(top_otus, 
@@
 ```{r source otu importance, fig.width=6, fig.height=6}
+confusionMatrix(source_all_rf$predicted, source_all_shared_df$human_source)$table
+
 top_otus_src <- get_tax(row_list = names(top_features_src))$tax_label

### 2016-12-09 code/build_schematic_figure.R
added figure 1 schematic code

@@ -1,0 +1,62 @@
+# Builds figure 1A, experimental timeline
+# Code from Matt Jenior 
+
+
+# Load dependencies 
+deps <- c('shape', 'plotrix', 'wesanderson');
+for (dep in deps){
+  if (dep %in% installed.packages()[,"Package"] == FALSE){
+    install.packages(as.character(dep), quiet=TRUE);
+  } 
+  library(dep, verbose=FALSE, character.only=TRUE)
+}
+
+#-------------------------------------------------------------------------------------------------------------------------------------#
+
+plot_file <- '~/Documents/Schloss_Lab/Schubert_humanCdGF_XXXX_2016/results/figures/figure_1.pdf'
+pdf(file=plot_file, width=9, height=9)
+
+# Create layout for multi-plot
+layout(mat=matrix(c(1,
+                    2,
+                    3,
+                    4,
+                    5), nrow=5, ncol=1, byrow=TRUE))
+
+#-------------------------------------------------------------------------------------------------------------------------------------#
+
+par(mar=c(0,0,0,0))
+plot(0, type='n', axes=F, xlab='', ylab='', xlim=c(-5,5), ylim=c(-2,2)) # Empty plot
+
+# Legend
+legend('center', legend=expression('Untreated water', paste(italic('C. difficile'), ' gavage'), 'Euthanize & Necropsy'), 
+       pt.bg=c(wes_palette("Chevalier")[3],'darkorchid2','black'), cex=1.5,  pch=c(22,25,25), pt.cex=c(4,3,3), bty='n', ncol=3)
+
+#----------------------------#
+
+plot(0, type='n', axes=F, xlab='', ylab='', xlim=c(-5,5), ylim=c(-2,2)) # Empty plot
+
+# Strep in drinking water timeline
+#rect(xleft=-4, ybottom=-0.4, xright=1, ytop=0.4, col=wes_palette("Darjeeling")[2], border='black')
+#rect(xleft=1, ybottom=-0.4, xright=3.75, ytop=0.4, col=wes_palette("Chevalier")[3], border='black')
+Arrows(x0=-4, y0=0, x1=4.8, y1=0, lwd=4, arr.type='triangle', arr.length=0.6, arr.width=0.3)
[…]

### 2016-12-09 Projects/HumanCdGF_cdiff_colonization.Rmd
add prediction of cfu with all otus and source

@@
 ##### OTUs predictive of _Colonization level_ (Top 10%)  
+Model built using all OTUs present in >10% of samples, only displaying top 10% of predictive features
 
@@
 ##### OTUs predictive of _Inocula Source_ (Top 10%)  
+Model built using all OTUs present in >10% of samples, only displaying top 10% of predictive features
 
@@
 `r names(top_features)[names(top_features) %in% names(top_features_src)]`
+
+***
+
+##### OTUs predictive of _Colonization level_ with Source included (Top 10%)  
+Model built using all OTUs present in >10% of samples, only displaying top 10% of predictive features
+
+```{r predict median CFU with source, fig.width=10,fig.height=10}
+cfu_all_src_shared_df <- colonized_cfu %>% 
+    select(mouse_id, log_cfu_med, human_source) %>% 
+    right_join(community_all_d0) %>% 
+    select(-mouse_id) %>% 
+    filter(!is.na(log_cfu_med))
+cfu_all_src_shared_df$human_source <- factor(cfu_all_src_shared_df$human_source)
+#output <- data.frame(ntree = NA, ntry = NA, rsq = NA)
+#for(tree in seq(1, 1500, 100)){
+#  for(try in 1:30){
+#    set.seed(seed)
+#    cfu_all_rf <- randomForest(cfu_all_shared_df[ , -1], cfu_all_shared_df$log_cfu_med,
+#                           ntree = tree,
+#                           mtry = try,
+#                           importance = T)
+#    output <- rbind(output, c(tree, try, tail(cfu_all_rf$rsq, 1)))
+#  }
+#}
+#
+#ggplot(output[output$ntree !=1,], aes(x=ntree, y=rsq, group = ntry, color = as.factor(ntry))) + geom_line()
+#ggplot(output[output$ntree == 1001,], aes(x=ntry, y=rsq)) + geom_line()
+
+set.seed(seed)
+cfu_all_src_rf <- randomForest(cfu_all_src_shared_df[ , -1], cfu_all_src_shared_df$log_cfu_med,
+                           ntree = 1001,
+                           mtry = 15,
+                           importance = T)
+top_features_src_cfu <- head(sort(importance(cfu_all_src_rf)[,1],decreasing = T), n_features)
+feature_df_src_numeric <- cfu_all_src_shared_df[ , c('log_cfu_med', names(top_features_src_cfu))]
+feature_df_src_numeric$human_source <- as.numeric(feature_df_src_numeric$human_source)
+feat_cor_cfu_src_all <- cor(feature_df_src_numeric, method = 'spearman')[-1,'log_cfu_med']
+```
+
+```{r prediction cfu plot with source, fig.width=6, fig.height=6}
+
+data.frame(Observed = cfu_all_src_shared_df$log_cfu_med, 
+           Predicted = cfu_all_src_rf$predicted) %>% 
+  ggplot(aes(x = Predicted, y = Observed)) + geom_point() +
+    xlim(5,9) + ylim(5,9) + 
+    annotate('text', label = paste0(round(tail(cfu_all_src_rf$rsq, 1), 4)*100, '% Var Explained'),
+             x = min(cfu_all_src_rf$predicted), y = max(cfu_all_src_shared_df$log_cfu_med)) + 
+    theme(panel.background = element_rect(linetype = 1, size = 1, fill = 'white', color = 'black'),
+          panel.grid.major.x = element_blank(),
+          panel.grid.minor.x = element_blank(),
+          panel.grid.major.y = element_blank(),
+          panel.grid.minor.y = element_blank()) +
+  labs(x = 'Predicted Colonization (log CFU)', y = 'Observed Colonization (log CFU)')
+
+```
+
+```{r colonization performance with source, fig.width=8, fig.height=6}
+top_otus_all_src <- get_tax(row_list = names(top_features_src_cfu)[-1])$tax_label
+top_otus_all_src <- c('Human Source', top_otus_all_src)
+```
+
+```{r colonization otu importance with source, fig.width=6, fig.height=6}
+data.frame(OTU = factor(top_otus_all_src, 
+                        levels = top_otus_all_src[order(top_features_src_cfu)]), 
+           Importance = top_features_src_cfu/max(top_features_src_cfu)) %>% 
+  ggplot(aes(x = OTU, y = Importance)) + 
+    geom_point() + coord_flip() + 
+    labs(y = 'Importance (Relative MDA)', x = NULL) + 
+    theme(panel.background = element_rect(linetype = 1, size = 2, fill = 'white', color = 'black'),
+          axis.ticks = element_line(size = 2),
+          panel.grid.major.x = element_blank(),
+          panel.grid.minor.x = element_blank(),
+          panel.grid.major.y = element_line(colour = "gray", linetype = "dotted"),
+          panel.grid.minor.y = element_line(colour = "gray", linetype = "dotted"),
+          axis.text.y = element_text(color = c(rep('black', length(top_features_src_cfu)-12),
+                                               rep('red', 12)), 
+                                     hjust = 0))
+```
+
+```{r plot colonization otu correlations with source, fig.width=10, fig.height=8}
+top_n_otus_all_src <- get_tax(row_list = names(head(top_features_src_cfu[-1],12)))
+top_n_otus_all_src$cor_label <- paste0(top_n_otus_all_src$tax, '(OTU',
+                           gsub('Otu0*', '', rownames(top_n_otus_all_src)),
+                           ';rho= ', round(feat_cor_cfu_src_all[names(feat_cor_cfu_src_all) 
+                                                       %in% rownames(top_n_otus_all_src)], 
+                                          2), ')')
+top_n_otus_all_src$cor_label <- factor(top_n_otus_all_src$cor_label,
+                               levels = top_n_otus_all_src$cor_label[order(head(top_features,12),
+                                                                   decreasing = T)])
+top_n_otus_all_src$OTU <- rownames(top_n_otus_all_src)
+subset(feature_df_src_numeric, select = 
+       c("log_cfu_med", rownames(top_n_otus_all_src))) %>% 
+  gather(OTU, abundance, -log_cfu_med) %>%
+  left_join(top_n_otus_all_src) %>% 
+  mutate(abundance = abundance + 0.001) %>% 
+  ggplot(aes(x = abundance, y = log_cfu_med)) + geom_point() + 
+    facet_wrap(~cor_label, ncol = 4, scales = 'free_x') +  
+    labs(x= 'Relative Abundance - Day 0 (log %)', y = 'Observed Colonization (log CFU)') + 
+    ylim(5, 9) + scale_x_log10(breaks = c(0.001, 0.01, 0.1, 1, 10, 100),
+                               labels = c(0, 0.01, 0.1, 1, 10, 100)) +
+    geom_vline(xintercept = 0.0015, color = 'gray') + 
+    theme(strip.background = element_rect(colour=NULL, fill='white'),
+          panel.background = element_rect(linetype = 1, size = 1, fill = 'white', color = 'black'),
+          panel.grid.major.y = element_blank(),
+          panel.grid.minor.y = element_blank(),
+          panel.grid.major.x = #element_line(linetype = 1, size = .1, color = 'gray'),
+                               element_blank(),    
+          panel.grid.minor.x = element_blank())
+```
+
+***
+
+#### OTUs predictive of CFU (One Mouse)

### 2016-12-09 Projects/HumanCdGF_cdiff_colonization.Rmd
adjust limit of detection on corr plots

@@
     labs(y = 'Importance (Relative MDA)', x = NULL) + 
-    theme(panel.background = element_rect(linetype = 1, size = 2, fill = 'white', color = 'black'),
+    theme(panel.background = element_rect(linetype = 1, size = 1, fill = 'white', color = 'black'),
           panel.grid.major.x = element_blank(),
@@
   left_join(top_n_otus) %>% 
-  mutate(abundance = abundance + 0.001) %>% 
+  mutate(abundance = abundance + 0.01) %>% 
   ggplot(aes(x = abundance, y = log_cfu_med)) + geom_point() + 
@@
     labs(x= 'Relative Abundance - Day 0 (log %)', y = 'Observed Colonization (log CFU)') + 
-    ylim(5, 9) + scale_x_log10(breaks = c(0.001, 0.01, 0.1, 1, 10, 100),
-                               labels = c(0, 0.01, 0.1, 1, 10, 100)) +
-    geom_vline(xintercept = 0.0015, color = 'gray') + 
+    ylim(5, 9) + scale_x_log10(breaks = c(0.01, 0.1, 1, 10, 100),
+                               labels = c(0, 0.1, 1, 10, 100)) +
+    geom_vline(xintercept = 0.06, color = 'gray') + 
     theme(strip.background = element_rect(colour=NULL, fill='white'),
@@
   left_join(top_n_otus_src) %>%
-  mutate(abundance = abundance + 0.001) %>% 
-  ggplot(aes(x = abundance, y = log_cfu_med)) + geom_point() + 
+  mutate(abundance = abundance + 0.01) %>% 
+  ggplot(aes(x = abundance, y = log_cfu_med)) + 
+    geom_vline(xintercept = 0.06, color = 'gray') + geom_point() + 
     facet_wrap(~cor_label, ncol = 4, scales = 'free_x') +  
     labs(x= 'Relative Abundance - Day 0 (log %)', y = 'Observed Colonization (log CFU)') + 
-    ylim(5, 9) + scale_x_log10(breaks = c(0.001, 0.01, 0.1, 1, 10, 100),
-                               labels = c(0, 0.01, 0.1, 1, 10, 100)) +
-    geom_vline(xintercept = 0.0015, color = 'gray') + 
+    ylim(5, 9) + scale_x_log10(breaks = c(0.01, 0.1, 1, 10, 100),
+                               labels = c(0, 0.1, 1, 10, 100)) +
     theme(strip.background = element_rect(colour=NULL, fill='white'),
@@
     labs(y = 'Importance (Relative MDA)', x = NULL) + 
-    theme(panel.background = element_rect(linetype = 1, size = 2, fill = 'white', color = 'black'),
+    theme(panel.background = element_rect(linetype = 1, size = 1, fill = 'white', color = 'black'),
           axis.ticks = element_line(size = 2),
@@
   left_join(top_n_otus_all_src) %>% 
-  mutate(abundance = abundance + 0.001) %>% 
-  ggplot(aes(x = abundance, y = log_cfu_med)) + geom_point() + 
+  mutate(abundance = abundance + 0.01) %>% 
+  ggplot(aes(x = abundance, y = log_cfu_med)) + 
+    geom_vline(xintercept = 0.06, color = 'gray') + geom_point() + 
     facet_wrap(~cor_label, ncol = 4, scales = 'free_x') +  
     labs(x= 'Relative Abundance - Day 0 (log %)', y = 'Observed Colonization (log CFU)') + 
-    ylim(5, 9) + scale_x_log10(breaks = c(0.001, 0.01, 0.1, 1, 10, 100),
-                               labels = c(0, 0.01, 0.1, 1, 10, 100)) +
-    geom_vline(xintercept = 0.0015, color = 'gray') + 
+    ylim(5, 9) + scale_x_log10(breaks = c(0.01, 0.1, 1, 10, 100),
+                               labels = c(0, 0.1, 1, 10, 100)) +
     theme(strip.background = element_rect(colour=NULL, fill='white'),

### 2016-12-09 Projects/HumanCdGF_cdiff_colonization.Rmd
setup one mouse model

@@
 #n_otus <- 12
-#iters <- 100
+iters <- 100
 #cor_cutoff <- 0.75
@@
 #### OTUs predictive of CFU (One Mouse)
+
+```{r one mouse model}
+
+#One per cage
+one_mouse_df <- data.frame(colonized_cfu %>% 
+    right_join(community_all_d0) %>% 
+    filter(!is.na(log_cfu_med)))
+one_mouse_df$human_source <- factor(one_mouse_df$human_source)
+
+om_import_otus <- data.frame(otu = colnames(select(one_mouse_df, contains('Otu'))))
+om_pred <- c()
+om_rsq <- c()
+for(i in 1:iters){
+    set.seed(i)
+    train <- one_mouse_df %>% 
+      group_by(cage_id) %>% 
+      sample_n(1) %>% ungroup() 
+    test <- one_mouse_df %>% 
+      filter(!mouse_id %in% train$mouse_id)
+    set.seed(seed)
+    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id, -cage_id),
+                             importance = TRUE, keep.forest = T, ntree = n_trees)
+    om_pred <- rbind(om_pred,
+                     data.frame(select(test, mouse_id, cage_id, log_cfu_med),
+                          predicted=predict(rf_om, test)))
+    om_rsq <- rbind(om_rsq, c(paste(train$mouse_id, collapse = ' '), 
+                              tail(rf_om$rsq,1),
+                              1 - (sum((test$log_cfu_med - predict(rf_om, test))^2)/
+                                sum((test$log_cfu_med - mean(test$log_cfu_med))^2))))
+    import_df <- importance(rf_om)
+    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
+    om_import_otus <- merge(om_import_otus, import_df, by.x = 'otu', by.y = 'row.names')
+    om_import_otus <- select(om_import_otus, -IncNodePurity)
+}
+om_rsq <- 1 - (sum((om_pred$log_cfu_med - om_pred$predicted)^2)/
+               sum((om_pred$log_cfu_med - mean(om_pred$log_cfu_med))^2))
+
+om_pred <- rename(om_pred, observed = log_cfu_med)
+
+om_import_otus <- data.frame(om_import_otus[ , -1], row.names = om_import_otus$otu)
+colnames(om_import_otus) <- seq(ncol(om_import_otus))
+om_otu_rank <- apply(om_import_otus, 2, function(x) rank(x, na.last = F))
+om_otu_rank <- om_otu_rank*as.numeric(om_rsq[,3])
+om_raw_rank <- sort(apply(om_otu_rank, 1, function(x) sum(x, na.rm = T)), decreasing = T)
+med_otu_importance_om <- data.frame(otu = names(om_raw_rank),
+                                 importance = om_raw_rank)

### 2016-12-12 code/build_schematic_figure.R
updated figure 1 schematic code

@@ -24,29 +24,14 @@ layout(mat=matrix(c(1,
                     5), nrow=5, ncol=1, byrow=TRUE))
 
 #-------------------------------------------------------------------------------------------------------------------------------------#
-
-par(mar=c(0,0,0,0))
 plot(0, type='n', axes=F, xlab='', ylab='', xlim=c(-5,5), ylim=c(-2,2)) # Empty plot
-
-# Legend
-legend('center', legend=expression('Untreated water', paste(italic('C. difficile'), ' gavage'), 'Euthanize & Necropsy'), 
-       pt.bg=c(wes_palette("Chevalier")[3],'darkorchid2','black'), cex=1.5,  pch=c(22,25,25), pt.cex=c(4,3,3), bty='n', ncol=3)
-
-#----------------------------#
-
-plot(0, type='n', axes=F, xlab='', ylab='', xlim=c(-5,5), ylim=c(-2,2)) # Empty plot
-
-# Strep in drinking water timeline
-#rect(xleft=-4, ybottom=-0.4, xright=1, ytop=0.4, col=wes_palette("Darjeeling")[2], border='black')
-#rect(xleft=1, ybottom=-0.4, xright=3.75, ytop=0.4, col=wes_palette("Chevalier")[3], border='black')
-Arrows(x0=-4, y0=0, x1=4.8, y1=0, lwd=4, arr.type='triangle', arr.length=0.6, arr.width=0.3)
-segments(x0=c(-4,1,3.75), y0=c(0.5,0.5,0.5), x1=c(-4,1,3.75), y1=c(-0.5,-0.5,-0.5), lwd=4)
-segments(x0=c(-3.5,-3,-2.5,-2,-1.5, -1, -0.5), y0=c(0.25,0.25,0.25,0.25,0.25,0.25), x1=c(-3.5,-3,-2.5,-2,-1.5, -1, -0.5), y1=c(-0.25,-0.25,-0.25,-0.25,-0.25,-0.25), lwd=2)
-text(x=c(-4,1), y=c(-0.8,-0.8), c('Day -14', 'Day 0'), cex=1.5)
-text(x=c(-3.35,1), y=c(0.7,0.7), c('Gavage human stool', 'Gavage C. difficile'), cex=1.5)
-text(x=-4.5, y=0, 'A', cex=1.9, font=2)
-
[…]

### 2016-12-12 Projects/HumanCdGF_cdiff_colonization.Rmd
add text comparing data

@@
 
-### Predicting Colonization Level (Log10 CFU) with Day 0 Community
+### Predicting __Colonization Level__ (Log10 CFU) with Day 0 Community
 
@@
 
-#### Predictive OTUs  
-##### OTUs predictive of _Colonization level_ (Top 10%)  
+##### Modeling _Colonization level_ (Top 10%)  
 Model built using all OTUs present in >10% of samples, only displaying top 10% of predictive features
@@
 
+##### OTUs predictive of __Colonization level__ (Top 10%)  
+Red OTUs are top 12 which are plotted against CFU below
+
 ```{r colonization otu importance, fig.width=6, fig.height=6}
@@
 
+##### Correlation of OTU and CFU (top 12 OTUs predictive of __Colonization level__)  
+Points colored by human donor
+
 ```{r plot colonization otu correlations, fig.width=10, fig.height=8}
@@
        c("log_cfu_med", rownames(top_n_otus))) %>% 
+  left_join(colonized_cfu[ , c('log_cfu_med', 'human_source')]) %>% 
   gather(OTU, abundance, contains('Otu')) %>%
@@
   mutate(abundance = abundance + 0.01) %>% 
-  ggplot(aes(x = abundance, y = log_cfu_med)) + geom_point() + 
+  ggplot(aes(x = abundance, y = log_cfu_med, color = human_source)) + geom_point() + 
     facet_wrap(~cor_label, ncol = 4, scales = 'free_x') +  
@@
                                element_blank(),    
-          panel.grid.minor.x = element_blank())
+          panel.grid.minor.x = element_blank(),
+          legend.position="none")
 ```
 
+Are these OTUs identifying susceptiblility to _C. Difficile_ of only correlated to _C. difficile_ due to OTUs unique to the donor, i.e. is the OTU predicting the cage grouping?
+
 ***
+
+### Predicting __Human Donor__ with Day 0 Community  
+Is there overlap in OTUs predictive of CFU and Donor?
 
@@
 
-##### OTUs predictive of _Inocula Source_ (Top 10%)  
+##### Modeling of __Human Donor__ (Top 10%)  
 Model built using all OTUs present in >10% of samples, only displaying top 10% of predictive features
 
+```{r source model, fig.width=6, fig.height=6}
+confusionMatrix(source_all_rf$predicted, source_all_shared_df$human_source)$table
+```
+
+##### OTUs predictive of __Human Donor__ (Top 10%)  
+Red OTUs are top 12 which are plotted against CFU below
+
 ```{r source otu importance, fig.width=6, fig.height=6}
-confusionMatrix(source_all_rf$predicted, source_all_shared_df$human_source)$table
-
 top_otus_src <- get_tax(row_list = names(top_features_src))$tax_label
@@
 
+##### Correlation of OTU and CFU (top 12 OTUs predictive of __Human Donor__)  
+Points colored by human donor
+
 ```{r plot source otu correlations, fig.width=10, fig.height=8}
@@
        c("log_cfu_med", rownames(top_n_otus_src))) %>% 
+  left_join(colonized_cfu[ , c('log_cfu_med', 'human_source')]) %>% 
   gather(OTU, abundance, contains('Otu')) %>%
@@
   mutate(abundance = abundance + 0.01) %>% 
-  ggplot(aes(x = abundance, y = log_cfu_med)) + 
+  ggplot(aes(x = abundance, y = log_cfu_med, color = human_source)) + 
     geom_vline(xintercept = 0.06, color = 'gray') + geom_point() + 
@@
                                element_blank(),    
-          panel.grid.minor.x = element_blank())
+          panel.grid.minor.x = element_blank(),
+          legend.position='none')
 ```
+
+There are `r sum(names(top_features) %in% names(top_features_src))` OTUs that overlap between predicting Colonization and Source:  
+`r names(top_features)[names(top_features) %in% names(top_features_src)]`  
+It is possible these OTUs are selected in predicting __Colonization__ because they are helping to seperate the groups of mice by cage.
+It is interesting though that the OTU 4 and OTU 14 were previously identified as being correlated (rho = 0.7586887) but only OTU 14 is predictive of Donor.
+Otu000004 is in the top `r print(paste0(round(which(names(sort(importance(source_all_rf)[,1],decreasing = T)) == 'Otu000004')/length(importance(source_all_rf)[,1]), 2)*100, '%'))` of OTUs predictive of Human Donor with a relative MDA of `r importance(source_all_rf)['Otu000004',1]/max(importance(source_all_rf)[,1])`.
+```{r plot otu correlations_entero, fig.width=3,fig.height=3}
+ggplot(cfu_all_shared_df, aes(x=Otu000004, y=Otu000014))+geom_point(alpha = 0.5)
+```
+
+As for OTU 283 and 5, those are correlated (rho = 0.8235480) and they are also correlated to Otu000171 (0.7891608), Otu000196 (0.8190665), Otu000202 (0.8147822), Otu000255 (0.8230815), Otu000262 (0.8141814), Otu000283 (0.8268792), Otu000394 (0.8321597), Otu000403 (0.8258554), which are all in the top 10% of OTUs predictive of __Colonization__.
+```{r plot otu correlations, fig.width=9,fig.height=9}
+cfu_all_shared_df %>% 
+  select(Otu000005, Otu000171, Otu000196, Otu000202, Otu000255, Otu000262, 
+         Otu000283, Otu000394, Otu000403) %>% 
+  gather(OTU, Abundance, -Otu000283) %>% 
+  ggplot(aes(x=Otu000283, y=Abundance))+geom_point(alpha = 0.5) + facet_wrap(~OTU, nrow =3, scales = 'free_y')
+```
+
+Do correlated OTUs matter?
 
@@
 
-There are `r sum(names(top_features) %in% names(top_features_src))` OTUs that overlap between predicting Colonization and Source:  
-`r names(top_features)[names(top_features) %in% names(top_features_src)]`
-
-***
-
-##### OTUs predictive of _Colonization level_ with Source included (Top 10%)  
+##### Model __Colonization level__ with Human Donor included
 Model built using all OTUs present in >10% of samples, only displaying top 10% of predictive features
@@
 
+##### OTUs predictive of _Colonization level_ with Source included (Top 10%)  
+Top 12 OTUs in red are plotted against CFU below
+
 ```{r colonization performance with source, fig.width=8, fig.height=6}
@@
           panel.grid.minor.y = element_line(colour = "gray", linetype = "dotted"),
-          axis.text.y = element_text(color = c(rep('black', length(top_features_src_cfu)-12),
-                                               rep('red', 12)), 
+          axis.text.y = element_text(color = c(rep('black', length(top_features_src_cfu)-13),
+                                               rep('red', 12), 'black'), 
                                      hjust = 0))
@@
 
+There are `r sum(names(top_features) %in% names(top_features_src_cfu))` OTUs that overlap between predicting Colonization and Source:  
+`r names(top_features)[names(top_features) %in% names(top_features_src_cfu)]`  
+Including Human Source includes all the OTUs predictive of __Colonization__ as well as Human Source as the top predictor.
+Relative importances of some OTUs are shifted, which could be due to the randomness of the RF algorithm or due to the effect of adding Human source, which may imply part of their predictive power is attributed to the human donor dependency.
+OTU 357 is added, however it is just below the 10% cutoff for the model without human source.
+Thus there appears to be no significant difference in predictive OTUs in the presence/absence of human donor data.
+
+There are `r sum(names(top_features_src) %in% names(top_features_src_cfu))` OTUs that overlap between predicting Colonization and Source:  
+`r names(top_features_src)[names(top_features_src) %in% names(top_features_src_cfu)]`  
+Comparing the OTUs predictive of __Colonization__ when including human source is similarlly different from the OTUs predictive of __Human Donor__.
+The same OTUs are shared, which is not suprising since both models predicting __Colonization__ use the same OTUs.
+Aside from those shared and even the correlated OTUs we still have a much different set of predictors, which seems to potentially reveal the ability of the microbiome to predict community members associated with _C.
+difficile_ colonization.
+
 ***
@@
 
-```{r one mouse model}
+```{r one mouse model, eval=F}
 

### 2016-12-14 Projects/HumanCdGF_cdiff_colonization.Rmd
model tuning mtry/ntree

@@
 
-#### OTUs predictive of CFU (One Mouse)
+#### Modeling CFU with One Mouse per Human Donor
+In an effort to remove any potential donor community dependency, the model is trained on a single mouse per human donor source, and then tested on all the remaining mice
 
-```{r one mouse model, eval=F}
+##### With all OTUs
+```{r one mouse model}
 
-#One per cage
+#One per source
 one_mouse_df <- data.frame(colonized_cfu %>% 
@@
 
-om_import_otus <- data.frame(otu = colnames(select(one_mouse_df, contains('Otu'))))
+om_import_otus <- data.frame(otu = colnames(select(one_mouse_df, -mouse_id, -cage_id, -log_cfu_med)))
 om_pred <- c()
@@
     train <- one_mouse_df %>% 
-      group_by(cage_id) %>% 
+      group_by(human_source) %>% 
       sample_n(1) %>% ungroup() 
@@
                      data.frame(select(test, mouse_id, cage_id, log_cfu_med),
-                          predicted=predict(rf_om, test)))
-    om_rsq <- rbind(om_rsq, c(paste(train$mouse_id, collapse = ' '), 
-                              tail(rf_om$rsq,1),
-                              1 - (sum((test$log_cfu_med - predict(rf_om, test))^2)/
+                          predicted=predict(rf_om, test),
+                          iter = i))
+    om_rsq <- rbind(om_rsq, c(iter = i,
+                              mice = paste(train$mouse_id, collapse = ' '), 
+                              train_rsq = tail(rf_om$rsq,1),
+                              test_rsq = 1 - (sum((test$log_cfu_med - predict(rf_om, test))^2)/
                                 sum((test$log_cfu_med - mean(test$log_cfu_med))^2))))
@@
     colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
-    om_import_otus <- merge(om_import_otus, import_df, by.x = 'otu', by.y = 'row.names')
-    om_import_otus <- select(om_import_otus, -IncNodePurity)
+    om_import_otus <- merge(om_import_otus, import_df, by.x = 'otu', by.y = 'row.names') %>% 
+      select(-IncNodePurity)
 }
-om_rsq <- 1 - (sum((om_pred$log_cfu_med - om_pred$predicted)^2)/
+om_rsq <- data.frame(iter = as.numeric(om_rsq[,1]),
+           mice = om_rsq[,2],
+           train_rsq = as.numeric(om_rsq[,3]),
+           test_rsq = as.numeric(om_rsq[,4]),
+           stringsAsFactors = F)
+om_rsq_overall <- 1 - (sum((om_pred$log_cfu_med - om_pred$predicted)^2)/
                sum((om_pred$log_cfu_med - mean(om_pred$log_cfu_med))^2))
@@
 
-om_import_otus <- data.frame(om_import_otus[ , -1], row.names = om_import_otus$otu)
-colnames(om_import_otus) <- seq(ncol(om_import_otus))
-om_otu_rank <- apply(om_import_otus, 2, function(x) rank(x, na.last = F))
-om_otu_rank <- om_otu_rank*as.numeric(om_rsq[,3])
-om_raw_rank <- sort(apply(om_otu_rank, 1, function(x) sum(x, na.rm = T)), decreasing = T)
-med_otu_importance_om <- data.frame(otu = names(om_raw_rank),
-                                 importance = om_raw_rank)
+all_otus_importance <- gather(om_import_otus,OTU, import_val,-otu)[,-2]
+all_otus_importance$votes <- ifelse(all_otus_importance$import_val > 0, 1, 0)
+all_otus_importance <- data.frame(table(all_otus_importance[,-2]))
+pos_otu <- all_otus_importance[(filter(all_otus_importance, votes == 1) %>% select(Freq) - 
+                      filter(all_otus_importance, votes == 0) %>% select(Freq)) > 0, ]
+pos_otu <- data.frame(predictor = pos_otu$otu[pos_otu$votes==1],
+                      filter(pos_otu, votes == 1) %>% select(Freq) - 
+                      filter(pos_otu, votes == 0) %>% select(Freq)) %>% 
+              arrange(desc(Freq))
+```
+
+When using all OTUs the model preforms poorly.
+The median Rsq of the trainset was `r median(om_rsq$train_rsq)`, this is likely do to the features.
+This is a summary of the importance scores for all predictors:
+`r summary(unlist(select(om_import_otus, -otu)))`  
+`r round(100*sum(om_import_otus[,-1]<=0, na.rm = T)/length(om_import_otus[,-1]<=0),2)`% of OTUs have an importance of 0 or less.
+The median Rsq of the test set was `r median(om_rsq$test_rsq)` and the Rsq of all predicted values was `r om_rsq_overall`.
+
+This one mouse per source model is improved through reducing the amount of negative features, either utilizing the feature set via the top X % of predictors from the model of all mice or take only the features that are more often have an importance > 0 (OTUs : `r as.character(pos_otu$predictor)`).
+
+```{r one mouse model - with OTUs positive in All OTU one mouse model}
+
+#One per source - > 0
+one_mouse_pos_otus_df <- data.frame(colonized_cfu %>% 
+    right_join(community_all_d0) %>% 
+    filter(!is.na(log_cfu_med))) %>% 
+    select(mouse_id, human_source, log_cfu_med,
+           one_of(as.character(pos_otu$predictor)))
+one_mouse_pos_otus_df$human_source <- factor(one_mouse_pos_otus_df$human_source)
+
+om_import_pos_otus <- data.frame(otu = colnames(select(one_mouse_pos_otus_df, -mouse_id, -log_cfu_med)))
+om_pred_pos <- c()
+om_rsq_pos <- c()
+for(i in 1:iters){
+    set.seed(i)
+    train <- one_mouse_pos_otus_df %>% 
+      group_by(human_source) %>% 
+      sample_n(1) %>% ungroup() 
+    test <- one_mouse_pos_otus_df %>% 
+      filter(!mouse_id %in% train$mouse_id)
+    for(m_try in 1:25){
+      for(n_tree in seq(200, 2000, 100)){
+    set.seed(seed)
+    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id),
+                             importance = TRUE, ntree = n_tree, mtry = m_try)
+    om_pred_pos <- rbind(om_pred_pos,
+                     data.frame(select(test, mouse_id, log_cfu_med),
+                          predicted=predict(rf_om, test),
+                          iter = i,
+                          rf_tree = n_tree,
+                          rf_try = m_try))
+    om_rsq_pos <- rbind(om_rsq_pos, c(iter = i,
+                                      try = m_try,
+                                      tree = n_tree,
+                              mice = paste0(paste(train$mouse_id, collapse = ' '), '_', m_try, '_', n_tree), 
+                              train_rsq = tail(rf_om$rsq,1),
+                              test_rsq = 1 - (sum((test$log_cfu_med - predict(rf_om, test))^2)/
+                                sum((test$log_cfu_med - mean(test$log_cfu_med))^2))))
+    import_df <- importance(rf_om)
+    colnames(import_df)[1] <- paste0(paste(train$mouse_id, collapse = ' '),'_', m_try, '_', n_tree)
+    om_import_pos_otus <- merge(om_import_pos_otus, import_df, by.x = 'otu', by.y = 'row.names') %>% 
+      select(-IncNodePurity)
+    }}
+}
+om_rsq_pos_df <- data.frame(iter = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='iter']),
+                            try = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='try']),
+                            tree = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='tree']),
+                            mice = om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='mice'],
+           train_rsq = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='train_rsq']),
+           test_rsq = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='test_rsq']),
+           stringsAsFactors = F)
+om_pf_df <- om_rsq_pos_df
+
+om_rsq_pos_overall <- 1 - (sum((om_pred_pos$log_cfu_med - om_pred_pos$predicted)^2)/
+               sum((om_pred_pos$log_cfu_med - mean(om_pred_pos$log_cfu_med))^2))
+
+om_pred_pos <- rename(om_pred_pos, observed = log_cfu_med)
+
+all_otus_importance <- gather(om_import_pos_otus,OTU, import_val,-otu)[,-2]
+all_otus_importance$votes <- ifelse(all_otus_importance$import_val > 0, 1, 0)
+all_otus_importance <- data.frame(table(all_otus_importance[,-2]))
+pos_otu <- all_otus_importance[(filter(all_otus_importance, votes == 1) %>% select(Freq) - 
+                      filter(all_otus_importance, votes == 0) %>% select(Freq)) > 0, ]
+pos_otu <- data.frame(predictor = pos_otu$otu[pos_otu$votes==1],
+                      filter(pos_otu, votes == 1) %>% select(Freq) - 
+                      filter(pos_otu, votes == 0) %>% select(Freq)) %>% 
+              arrange(desc(Freq))
+
+ggplot(om_rsq_pos_df, aes(x = try, y = test_rsq)) + 
+  stat_summary(fun.y=median, geom="line", aes(group = iter)) + 
+  stat_summary(fun.data = 'median_hilow') +
+geom_line() + facet_wrap(~tree, nrow=4)
+
+ggplot(filter(om_rsq_pos_df, try ==6), aes(x = tree, y = test_rsq)) + 
+  stat_summary(fun.y=median, geom="line", aes(group = iter)) + 
+  stat_summary(fun.data = 'median_hilow')
+```
+
+`r median(om_rsq_pos_df$train_rsq)`
+`r median(om_rsq_pos_df$test_rsq)`
+`r om_rsq_pos_overall`
+
+OM > 0
+> median(om_rsq_pos$train_rsq)
+[1] 0.08652421
+> median(om_rsq_pos$test_rsq)
+[1] 0.5331836
+> om_rsq_pos_overall
+[1] 0.5534567
+
+All mice top 10%
+> median(om_rsq_pos$train_rsq)
+[1] 0.07336261
+> median(om_rsq_pos$test_rsq)
+[1] 0.6070359
+> om_rsq_pos_overall
+[1] 0.633238
+
+> median(om_rsq_pos_df$train_rsq)
+[1] 0.05986871
+> median(om_rsq_pos_df$test_rsq)
+[1] 0.5898076
+> `r om_rsq_pos_overall`
+> om_rsq_pos_overall
+[1] 0.6150207
+
+one mouse - using top 10% of predictive features for all mice in the training set
+optimal at
+train 
+  try = 10, tree = 600
+test
+  try = 6, tree = 600
+but the test set has two groupings in performance. one around a Rsq 0.75, one at 0.5 and two iterations at 0.325
+
+       mouse         high       low          med MeanDecreaseAccuracy MeanDecreaseGini
+1      D_507 24.746833562  1.737270 20.102178345          24.73595607        8.4911043
+2      C_596 23.285539876  1.001002 20.447905674          24.43000483        7.6690227
+3      D_527 18.032458266  1.001002 15.500296401          18.71071088        4.9877513
+4      A_566 12.032722269  0.000000 12.067340706          13.81516364        3.0119082
+5      D_540 10.866644663  1.737270 12.253700494          13.61280464        2.2541893
+6      C_516 11.047770210  1.001002 10.744008798          12.43936737        2.2192996
+7      A_503 10.956589615  4.195038  6.774750778          10.79692514        2.2143605
+8      A_526  7.778933002  0.000000  6.659309475           8.50216395        1.1996857
+9    369_992 -0.875534077  3.932419  4.882427149           3.72993812        1.3625130
+10   369_942 -0.596862323  1.417051  3.309012343           2.37657639        0.7644884
+11  IN3_2318  1.043691934  2.664468  1.245180833           2.18825289        0.8340440
+12  MOUT_896  2.014382262 -2.247333  1.336156704           2.17645115        0.8413007
+13   430_920  1.549432277  0.000000  1.108030915           1.91835030        0.6586209
+14  MOUT_568  0.666044064 -1.417051  1.909190798           1.75633167        0.7946055
+15  IN3_2336 -0.217030290  0.000000  2.406333245           1.45691227        0.5817839
+16   NP2_979  1.078030444  0.000000  1.206469792           1.43939415        0.6973072
+17   OUTA_NT  0.697539079  1.417051  0.845326576           1.18172290        0.7653337
+18   NP2_983 -0.727654094  1.001002  1.931180990           1.05707031        0.8507964
+19 OUTA_2063  1.808105182 -1.417051 -0.047912339           0.89013348        0.7126412
+20   430_307  0.261169822 -2.464320  1.107393946           0.67386593        0.7143389
+21   430_964 -0.574499626  1.001002  1.123257626           0.56350353        0.4823411
+22   369_939 -0.619215217  0.000000  1.274820050           0.56227756        0.5328745
+23 OUT2_2510  0.009109248 -1.001002  0.721274725           0.37736881        0.6381838
+24   OUT2_NT -0.430535085  0.000000  0.460746344           0.14766696        0.6223585
+25  IN2_2613  0.000000000  0.000000  0.000000000           0.00000000        0.0000000
+26  IN2_2629  0.000000000  0.000000  0.000000000           0.00000000        0.0000000
+27  IN2_2673  0.000000000  0.000000  0.000000000           0.00000000        0.0000000
+28  NP2_2630  0.000000000  0.000000  0.000000000           0.00000000        0.0000000
+29   LINE_NT  1.730524126  0.000000 -1.610672526          -0.09130011        0.6774122
+30   430_304  0.153756741 -2.008048 -0.653527605          -0.51088419        0.6561920
+31 OUTA_2062 -0.244531023 -1.737270 -0.443149129          -0.70889274        0.7545123
+32  LINE_978 -1.613030982  1.001002 -0.003319811          -1.09823532        0.5058066
+33   369_331 -2.611303629  0.000000 -0.024601226          -1.37881023        0.6251534
+34  LINE_MNT -0.399645816 -1.737270 -1.503302377          -1.75744745        0.5979079
+35  LINE_188 -1.752110136 -1.001002 -1.104193541          -1.91010075        0.5634185
+36    IN3_NT -2.122086607  0.000000 -1.213308576          -2.28278894        0.7106271
+37 OUT2_2533 -1.443012131  0.000000 -2.212914984          -2.28741305        0.5829754
+
+```{r}
+om_tf_df$group_mean <- 'med'
+om_tf_df$group_mean[om_tf_df$test_rsq > 0.65] <- 'high'
+om_tf_df$group_mean[om_tf_df$test_rsq < 0.4] <- 'low'
+mice_group <- filter(om_tf_df, try == 5, tree == 500) %>% 
+  select(mice, group_mean)
+mice_group$mice <- gsub('_\\d*_\\d*$', '', mice_group$mice)
+
+mice_diff <- data.frame(matrix(rep(NA, nrow(mice_group)*length(unique(colonized_cfu$mouse_id))),
+                               nrow=nrow(mice_group)))
+colnames(mice_diff) <- unique(colonized_cfu$mouse_id)
+for(i in 1:nrow(mice_diff)){
+    mice_diff[i,] <- ifelse(colnames(mice_diff) %in% strsplit(mice_group$mice, ' ')[[i]], 0,1)
+}
+predict_rsq <- cbind(mice_group, mice_diff) %>% 
+  select(-mice)
+predict_rsq$group_mean <- factor(predict_rsq$group_mean)
+rsq_grp_rf <- randomForest(predict_rsq[,-1],predict_rsq$group_mean, importance = T)
+importance_rsq <- data.frame(mouse = dimnames(importance_rsq)[[1]],importance(rsq_grp_rf))
+arrange(importance_rsq, desc(MeanDecreaseAccuracy))
+str(importance_rsq)
+
+```

### 2016-12-15 Projects/HumanCdGF_cdiff_colonization.Rmd
investigate bimodality of one mouse rsq

--- 2016-12-12 Projects/HumanCdGF_cdiff_colonization.Rmd
@@
 
-om_import_otus <- data.frame(om_import_otus[ , -1], row.names = om_import_otus$otu)
-colnames(om_import_otus) <- seq(ncol(om_import_otus))
-om_otu_rank <- apply(om_import_otus, 2, function(x) rank(x, na.last = F))
-om_otu_rank <- om_otu_rank*as.numeric(om_rsq[,3])
-om_raw_rank <- sort(apply(om_otu_rank, 1, function(x) sum(x, na.rm = T)), decreasing = T)
-med_otu_importance_om <- data.frame(otu = names(om_raw_rank),
-                                 importance = om_raw_rank)
+all_otus_importance <- gather(om_import_otus,OTU, import_val,-otu)[,-2]
+all_otus_importance$votes <- ifelse(all_otus_importance$import_val > 0, 1, 0)
+all_otus_importance <- data.frame(table(all_otus_importance[,-2]))
+pos_otu <- all_otus_importance[(filter(all_otus_importance, votes == 1) %>% select(Freq) - 
+                      filter(all_otus_importance, votes == 0) %>% select(Freq)) > 0, ]
+pos_otu <- data.frame(predictor = pos_otu$otu[pos_otu$votes==1],
+                      filter(pos_otu, votes == 1) %>% select(Freq) - 
+                      filter(pos_otu, votes == 0) %>% select(Freq)) %>% 
+              arrange(desc(Freq))
+```
+
+When using all OTUs the model preforms poorly.
+The median Rsq of the trainset was `r median(om_rsq$train_rsq)`, this is likely do to the features.
+This is a summary of the importance scores for all predictors:
+`r summary(unlist(select(om_import_otus, -otu)))`  
+`r round(100*sum(om_import_otus[,-1]<=0, na.rm = T)/length(om_import_otus[,-1]<=0),2)`% of OTUs have an importance of 0 or less.
+The median Rsq of the test set was `r median(om_rsq$test_rsq)` and the Rsq of all predicted values was `r om_rsq_overall`.
+
+This one mouse per source model is improved through reducing the amount of negative features, either utilizing the feature set via the top X % of predictors from the model of all mice or take only the features that are more often have an importance > 0 (OTUs : `r as.character(pos_otu$predictor)`).
+
+```{r one mouse model - with OTUs positive in All OTU one mouse model}
+
+#One per source - > 0
+one_mouse_pos_otus_df <- data.frame(colonized_cfu %>% 
+    right_join(community_all_d0) %>% 
+    filter(!is.na(log_cfu_med))) %>% 
+    select(mouse_id, human_source, log_cfu_med,
+           one_of(as.character(pos_otu$predictor)))
+one_mouse_pos_otus_df$human_source <- factor(one_mouse_pos_otus_df$human_source)
+
+om_import_pos_otus <- data.frame(otu = colnames(select(one_mouse_pos_otus_df, -mouse_id, -log_cfu_med)))
+om_pred_pos <- c()
+om_rsq_pos <- c()
+for(i in 1:iters){
+    set.seed(i)
+    train <- one_mouse_pos_otus_df %>% 
+      group_by(human_source) %>% 
+      sample_n(1) %>% ungroup() 
+    test <- one_mouse_pos_otus_df %>% 
+      filter(!mouse_id %in% train$mouse_id)
+    set.seed(seed)
+    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id),
+                             importance = TRUE, ntree = n_tree, mtry = 10)
+    om_pred_pos <- rbind(om_pred_pos,
+                     data.frame(select(test, mouse_id, log_cfu_med),
+                          predicted=predict(rf_om, test),
+                          iter = i))
+    om_rsq_pos <- rbind(om_rsq_pos, c(iter = i,
+                              mice = paste(train$mouse_id, collapse = ' '), 
+                              train_rsq = tail(rf_om$rsq,1),
+                              test_rsq = 1 - (sum((test$log_cfu_med - predict(rf_om, test))^2)/
+                                sum((test$log_cfu_med - mean(test$log_cfu_med))^2))))
+    import_df <- importance(rf_om)
+    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
+    om_import_pos_otus <- merge(om_import_pos_otus, import_df, by.x = 'otu', by.y = 'row.names') %>% 
+      select(-IncNodePurity)
+}
+
+om_rsq_pos_df <- data.frame(iter = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='iter']),
+                            mice = om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='mice'],
+           train = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='train_rsq']),
+           test = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='test_rsq']),
+           stringsAsFactors = F) %>% 
+  gather(data_set, rsq, train, test)
+
+om_rsq_pos_overall <- 1 - (sum((om_pred_pos$log_cfu_med - om_pred_pos$predicted)^2)/
+               sum((om_pred_pos$log_cfu_med - mean(om_pred_pos$log_cfu_med))^2))
+
+om_pred_pos <- rename(om_pred_pos, observed = log_cfu_med)
+
+#all_otus_importance <- gather(om_import_pos_otus,OTU, import_val,-otu)[,-2]
+#all_otus_importance$votes <- ifelse(all_otus_importance$import_val > 0, 1, 0)
+#all_otus_importance <- data.frame(table(all_otus_importance[,-2]))
+#pos_otu <- all_otus_importance[(filter(all_otus_importance, votes == 1) %>% select(Freq) - 
+#                      filter(all_otus_importance, votes == 0) %>% select(Freq)) > 0, ]
+#pos_otu <- data.frame(predictor = pos_otu$otu[pos_otu$votes==1],
+#                      filter(pos_otu, votes == 1) %>% select(Freq) - 
+#                      filter(pos_otu, votes == 0) %>% select(Freq)) %>% 
+#              arrange(desc(Freq))
+```
+
+One Mouse per Human Donor with OTUs > 0 give:
+  *  median training set rsq of 0.08652421  
+  *  median testing set rsq of 0.5331836  
+  *  overall rsq of 0.5534567
+
+One Mouse per Human Donor with the top 10% predictive OTUs when using all mice  
+  *  median training set rsq of 0.07336261  
+  *  median testing set rsq of 0.6070359  
+  *  overall rsq of 0.633238
+
+Looking at the % var explained values from each individual iteration, there appears to be a bimodal distribution in the performance.
+
+```{r om_bimodality, fig.width= = 4, fig.height=4}
+ggplot(om_rsq_pos_df, aes(rsq, color = data_set)) + 
+  geom_density()
+```
+
+Using random forest to identify the mice best seperating the Rsq distribution
+
+```{r one_mouse_rsq_predictors}
+
+predict_rsq <- om_rsq_pos_df %>% 
+                 filter(data_set == 'test')
+predict_rsq$group_mean <- ifelse(predict_rsq$rsq > 0.62, T, F)
+
+mice_diff <- data.frame(matrix(rep(NA, nrow(predict_rsq)*length(unique(colonized_cfu$mouse_id))),
+                               nrow=nrow(predict_rsq)))
+colnames(mice_diff) <- unique(colonized_cfu$mouse_id)
+for(i in 1:nrow(mice_diff)){
+    mice_diff[i,] <- ifelse(colnames(mice_diff) %in% strsplit(predict_rsq$mice, ' ')[[i]], 0,1)
+}
+
+predict_rsq_df <- cbind(rsq = predict_rsq$group_mean, mice_diff)
+predict_rsq_df$rsq <- as.factor(predict_rsq_df$rsq)
+set.seed(seed)
+rsq_grp_rf <- randomForest(predict_rsq_df[,-1],predict_rsq_df$rsq, importance = T)
+importance_rsq <- data.frame(mouse = dimnames(importance_rsq)[[1]],importance(rsq_grp_rf))
+one_mouse_rsq_pred <- arrange(importance_rsq, desc(MeanDecreaseAccuracy)) %>% 
+  select(mouse,MeanDecreaseAccuracy) %>% 
+  mutate(Relative_MDA = MeanDecreaseAccuracy/max(MeanDecreaseAccuracy))
+
+data.frame(mouse = factor(one_mouse_rsq_pred$mouse, 
+                        levels = one_mouse_rsq_pred$mouse[order(one_mouse_rsq_pred$Relative_MDA)]), 
+           Importance = one_mouse_rsq_pred$Relative_MDA) %>% 
+ggplot(aes(x=mouse, y=Importance)) + 
+   geom_point() + coord_flip() + 
+    labs(y = 'Importance (Relative MDA)', x = NULL) + 
+    theme(panel.background = element_rect(linetype = 1, size = 1, fill = 'white', color = 'black'),
+          axis.ticks = element_line(size = 2),
+          panel.grid.major.x = element_blank(),
+          panel.grid.minor.x = element_blank(),
+          panel.grid.major.y = element_line(colour = "gray", linetype = "dotted"),
+          panel.grid.minor.y = element_line(colour = "gray", linetype = "dotted"),
+          axis.text.y = element_text(color = c(rep('black', length(one_mouse_rsq_pred$Relative_MDA)-
+                                                     sum(one_mouse_rsq_pred$MeanDecreaseAccuracy > 0)),
+                                               rep('red', sum(one_mouse_rsq_pred$MeanDecreaseAccuracy > 0))), 
+                                     hjust = 0))
+
+```
+
+Due to small n?  
+Due to incomplete source dependency?  
+Ratio of train to test?

### 2016-12-15 code/build_schematic_figure.R
figure 1 schematic fixed

@@ -1,52 +1,17 @@
 # Builds figure 1
 # Experimental timeline code from Matt Jenior
-
-# Load dependencies 
-deps <- c('shape', 'plotrix', 'wesanderson');
+# Load dependencies
+deps <- c('shape', 'wesanderson');
 for (dep in deps){
   if (dep %in% installed.packages()[,"Package"] == FALSE){
     install.packages(as.character(dep), quiet=TRUE);
   } 
   library(dep, verbose=FALSE, character.only=TRUE)
 }
+#------------------------------------------------------------------------------------------------------------------------------------#
+# Import, analyze and clean data for figures before plotting
 
-#-------------------------------------------------------------------------------------------------------------------------------------#
-
-plot_file <- '~/Documents/Schloss_Lab/Schubert_humanCdGF_XXXX_2016/results/figures/figure_1.pdf'
-pdf(file=plot_file, width=9, height=9)
-
-# Create layout for multi-plot
-layout(mat=matrix(c(1,
-                    2,
-                    3,
-                    4,
-                    5), nrow=5, ncol=1, byrow=TRUE))
-
-# Figure 1A - experimental timeline code 
-#-------------------------------------------------------------------------------------------------------------------------------------#
-
-#---------Option 1, text based timeline---------#
-plot(0, type='n', axes=F, xlab='', ylab='', xlim=c(-5,5), ylim=c(-2,2)) # Empty plot
-Arrows(x0=-4, y0=0, x1=3, y1=0, lwd=4, arr.type='triangle', arr.length=0.6, arr.width=0.3)
-segments(x0=c(-4,-0.5,2), y0=c(0.5,0.5,0.5), x1=c(-4,-0.5,2), y1=c(-0.5,-0.5,-0.5), lwd=4)
-segments(x0=c(-3.5,-3,-2.5,-2,-1.5, -1), y0=c(0.25,0.25,0.25,0.25,0.25), x1=c(-3.5,-3,-2.5,-2,-1.5, -1), y1=c(-0.25,-0.25,-0.25,-0.25,-0.25), lwd=2)
-segments(x0=c(0,0.5,1,1.5), y0=c(0.25,0.25,0.25,0.25,0.25), x1=c(0,0.5,1,1.5), y1=c(-0.25,-0.25,-0.25,-0.25,-0.25), lwd=2)
[…]

### 2016-12-15 Projects/HumanCdGF_cdiff_colonization.Rmd
rsq dist with out 581 cages

@@
   gather(data_set, rsq, train, test)
-
-om_rsq_pos_overall <- 1 - (sum((om_pred_pos$log_cfu_med - om_pred_pos$predicted)^2)/
-               sum((om_pred_pos$log_cfu_med - mean(om_pred_pos$log_cfu_med))^2))
-
-om_pred_pos <- rename(om_pred_pos, observed = log_cfu_med)
-
-#all_otus_importance <- gather(om_import_pos_otus,OTU, import_val,-otu)[,-2]
-#all_otus_importance$votes <- ifelse(all_otus_importance$import_val > 0, 1, 0)
-#all_otus_importance <- data.frame(table(all_otus_importance[,-2]))
-#pos_otu <- all_otus_importance[(filter(all_otus_importance, votes == 1) %>% select(Freq) - 
-#                      filter(all_otus_importance, votes == 0) %>% select(Freq)) > 0, ]
-#pos_otu <- data.frame(predictor = pos_otu$otu[pos_otu$votes==1],
-#                      filter(pos_otu, votes == 1) %>% select(Freq) - 
-#                      filter(pos_otu, votes == 0) %>% select(Freq)) %>% 
-#              arrange(desc(Freq))
 ```
@@
 ```{r om_bimodality, fig.width= = 4, fig.height=4}
-ggplot(om_rsq_pos_df, aes(rsq, color = data_set)) + 
-  geom_density()
+ggplot(om_rsq_pos_df, aes(rsq, fill = data_set)) + 
+  geom_density(alpha = 0.25) + ggtitle('Rsq Distribution for One Mouse per Human Donor')
 ```
 
-Using random forest to identify the mice best seperating the Rsq distribution
+Is this bimodal distribution due to :
+  * Due to small n?  
+  * Due to incomplete source dependency?  
+  * Ratio of train to test?
 
-```{r one_mouse_rsq_predictors}
+Using random forest to identify the mice driving bimodal Rsq distribution
+
+```{r one_mouse_rsq_predictors fig.width=4, fig.height=8}
 
@@
 
-Due to small n?  
-Due to incomplete source dependency?  
-Ratio of train to test?
+Is it possible one of the cages of the same donor are not similar, and causing the poor performance?
+
+```{r one mouse model CAGE - with OTUs positive in All OTU one mouse model}
+
+#One per source - > 0
+one_mouse_pos_otus_df <- data.frame(colonized_cfu %>% 
+    right_join(community_all_d0) %>% 
+    filter(!is.na(log_cfu_med))) %>% 
+    select(mouse_id, cage_id, log_cfu_med,
+           one_of(as.character(pos_otu$predictor)))
+one_mouse_pos_otus_df$cage_id <- factor(one_mouse_pos_otus_df$cage_id)
+
+om_import_pos_otus <- data.frame(otu = colnames(select(one_mouse_pos_otus_df, -mouse_id, -log_cfu_med)))
+om_pred_pos <- c()
+om_rsq_pos <- c()
+for(i in 1:iters){
+    set.seed(i)
+    train <- one_mouse_pos_otus_df %>% 
+      group_by(cage_id) %>% 
+      sample_n(1) %>% ungroup() 
+    test <- one_mouse_pos_otus_df %>% 
+      filter(!mouse_id %in% train$mouse_id)
+    set.seed(seed)
+    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id),
+                             importance = TRUE, ntree = n_tree, mtry = 10)
+    om_pred_pos <- rbind(om_pred_pos,
+                     data.frame(select(test, mouse_id, log_cfu_med),
+                          predicted=predict(rf_om, test),
+                          iter = i))
+    om_rsq_pos <- rbind(om_rsq_pos, c(iter = i,
+                              mice = paste(train$mouse_id, collapse = ' '), 
+                              train_rsq = tail(rf_om$rsq,1),
+                              test_rsq = 1 - (sum((test$log_cfu_med - predict(rf_om, test))^2)/
+                                sum((test$log_cfu_med - mean(test$log_cfu_med))^2))))
+    import_df <- importance(rf_om)
+    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
+    om_import_pos_otus <- merge(om_import_pos_otus, import_df, by.x = 'otu', by.y = 'row.names') %>% 
+      select(-IncNodePurity)
+}
+
+om_rsq_cage_df <- data.frame(iter = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='iter']),
+                            mice = om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='mice'],
+           train = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='train_rsq']),
+           test = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='test_rsq']),
+           stringsAsFactors = F) %>% 
+  gather(data_set, rsq, train, test)
+```
+
+One Mouse per __Cage__ with OTUs > 0 give:
+  *  median training set rsq of `r median()`  
+  *  median testing set rsq of `r `
+  *  overall rsq of `r
+
+Looking at the % var explained values from each individual iteration, there appears to be a bimodal distribution in the performance.
+
+```{r om_cage_rsq_dist, fig.width= = 4, fig.height=4}
+ggplot(om_rsq_cage_df, aes(rsq, fill = data_set)) + 
+  geom_density(alpha = 0.25) + ggtitle('Rsq Distribution for One Mouse per Cage')
+```
+
+By cage:  
+`r summary(om_rsq_cage_df$rsq[om_rsq_cage_df$data_set=='test'])`
+
+By donor:  
+`r summary(om_rsq_pos_df$rsq[om_rsq_pos_df$data_set=='test'])`
+
+```{r one mouse model - with OTUs positive in All OTU one mouse model}
+
+#One per source - > 0
+one_mouse_pos_otus_df <- data.frame(colonized_cfu %>% 
+    right_join(community_all_d0) %>% 
+    filter(!is.na(log_cfu_med))) %>% 
+    select(mouse_id, human_source, log_cfu_med, cage_id,
+           one_of(as.character(pos_otu$predictor))) %>% 
+    filter(!cage_id %in% c('A','D')) %>% 
+    select(-cage_id)
+one_mouse_pos_otus_df$human_source <- factor(one_mouse_pos_otus_df$human_source)
+
+om_rsq_pos <- c()
+for(i in 1:iters){
+    set.seed(i)
+    train <- one_mouse_pos_otus_df %>% 
+      group_by(human_source) %>% 
+      sample_n(1) %>% ungroup() 
+    test <- one_mouse_pos_otus_df %>% 
+      filter(!mouse_id %in% train$mouse_id)
+    set.seed(seed)
+    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id),
+                             importance = TRUE, ntree = n_tree, mtry = 10)
+    om_rsq_pos <- rbind(om_rsq_pos, c(iter = i,
+                              mice = paste(train$mouse_id, collapse = ' '), 
+                              train_rsq = tail(rf_om$rsq,1),
+                              test_rsq = 1 - (sum((test$log_cfu_med - predict(rf_om, test))^2)/
+                                sum((test$log_cfu_med - mean(test$log_cfu_med))^2))))
+    import_df <- importance(rf_om)
+    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
+}
+
+om_rsq_C_only_df <- data.frame(iter = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='iter']),
+                            mice = om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='mice'],
+           train = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='train_rsq']),
+           test = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='test_rsq']),
+           stringsAsFactors = F) %>% 
+  gather(data_set, rsq, train, test)
+
+one_mouse_pos_otus_df <- data.frame(colonized_cfu %>% 
+    right_join(community_all_d0) %>% 
+    filter(!is.na(log_cfu_med))) %>% 
+    select(mouse_id, human_source, log_cfu_med, cage_id,
+           one_of(as.character(pos_otu$predictor))) %>% 
+    filter(!cage_id == 'A') %>% 
+    select(-cage_id)
+one_mouse_pos_otus_df$human_source <- factor(one_mouse_pos_otus_df$human_source)
+
+om_rsq_pos <- c()
+for(i in 1:iters){
+    set.seed(i)
+    train <- one_mouse_pos_otus_df %>% 
+      group_by(human_source) %>% 
+      sample_n(1) %>% ungroup() 
+    test <- one_mouse_pos_otus_df %>% 
+      filter(!mouse_id %in% train$mouse_id)
+    set.seed(seed)
+    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id),
+                             importance = TRUE, ntree = n_tree, mtry = 10)
+    om_rsq_pos <- rbind(om_rsq_pos, c(iter = i,
+                              mice = paste(train$mouse_id, collapse = ' '), 
+                              train_rsq = tail(rf_om$rsq,1),
+                              test_rsq = 1 - (sum((test$log_cfu_med - predict(rf_om, test))^2)/
+                                sum((test$log_cfu_med - mean(test$log_cfu_med))^2))))
+    import_df <- importance(rf_om)
+    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
+}
+
+om_rsq_AD_df <- data.frame(iter = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='iter']),
+                            mice = om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='mice'],
+           train = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='train_rsq']),
+           test = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='test_rsq']),
+           stringsAsFactors = F) %>% 
+  gather(data_set, rsq, train, test)
+
+rbind(data.frame(rsq = om_rsq_pos_df$rsq[om_rsq_pos_df$data_set=='test'], data = 'all'),
+      rbind(data.frame(rsq = om_rsq_AD_df$rsq[om_rsq_AD_df$data_set=='test'], data = 'Two_only'),
+            data.frame(rsq = om_rsq_C_only_df$rsq[om_rsq_C_only_df$data_set=='test'], data = 'One_only'))) %>% 
+  ggplot(aes(rsq, color = data)) + 
+  geom_density() + ggtitle('Rsq Distribution when multiple cages from same Donor present')
+```
+
+It seemed possible the assumption all donors have the same community may be false.
+However, when running the analysis with only one cage from the donor with multiple cages it performs as we observe for the lower peak in the bimodal distribution and when a second cage from the same donor is added we observe Rsq at the higher peak in the bimodal distribution.
+Thus there appears to be a cage effect improving the performance and the bimodal distribution maybe due to a variation in the ability of these communities to be able to predict each other, ie some be more similar to each other but still as a group more different than the other cages.

### 2016-12-15 Projects/HumanCdGF_cdiff_colonization.Rmd
fix errors so file can knit

@@
 seed <- 1
-#n_trees <- 1001
+n_trees <- 1001
 #n_otus <- 12
@@
     rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id),
-                             importance = TRUE, ntree = n_tree, mtry = 10)
+                             importance = TRUE, ntree = n_trees, mtry = 10)
     om_pred_pos <- rbind(om_pred_pos,
@@
 
-```{r om_bimodality, fig.width= = 4, fig.height=4}
+```{r om_bimodality, fig.width= 4, fig.height=4}
 ggplot(om_rsq_pos_df, aes(rsq, fill = data_set)) + 
@@
 
-```{r one_mouse_rsq_predictors fig.width=4, fig.height=8}
+```{r one_mouse_rsq_predictors, fig.width=4, fig.height=8}
 
@@
 rsq_grp_rf <- randomForest(predict_rsq_df[,-1],predict_rsq_df$rsq, importance = T)
-importance_rsq <- data.frame(mouse = dimnames(importance_rsq)[[1]],importance(rsq_grp_rf))
+importance_rsq <- data.frame(mouse = dimnames(importance(rsq_grp_rf))[[1]],importance(rsq_grp_rf))
 one_mouse_rsq_pred <- arrange(importance_rsq, desc(MeanDecreaseAccuracy)) %>% 
@@
 
-One Mouse per __Cage__ with OTUs > 0 give:
-  *  median training set rsq of `r median()`  
-  *  median testing set rsq of `r `
-  *  overall rsq of `r
-
 Looking at the % var explained values from each individual iteration, there appears to be a bimodal distribution in the performance.
 
-```{r om_cage_rsq_dist, fig.width= = 4, fig.height=4}
+```{r om_cage_rsq_dist, fig.width= 4, fig.height=4}
 ggplot(om_rsq_cage_df, aes(rsq, fill = data_set)) + 
@@
 
-```{r one mouse model - with OTUs positive in All OTU one mouse model}
+```{r compare number of 581 donor cages, fig.height=5, fig.width=5}
 
@@
     rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id),
-                             importance = TRUE, ntree = n_tree, mtry = 10)
+                             importance = TRUE, ntree = n_trees, mtry = 10)
     om_rsq_pos <- rbind(om_rsq_pos, c(iter = i,

### 2016-12-16 code/build_schematic_figure.R
figure 1 assembled and done for now

@@ -1,13 +1,7 @@
 # Builds figure 1
 # Experimental timeline code from Matt Jenior
 # Load dependencies
-deps <- c('shape', 'wesanderson');
-for (dep in deps){
-  if (dep %in% installed.packages()[,"Package"] == FALSE){
-    install.packages(as.character(dep), quiet=TRUE);
-  } 
-  library(dep, verbose=FALSE, character.only=TRUE)
-}
+
 #------------------------------------------------------------------------------------------------------------------------------------#
 # Import, analyze and clean data for figures before plotting
 
@@ -78,12 +72,21 @@ names(day0_donor_nmds)[4] <- "Donor"
 names(day0_donor_nmds)[5] <- "Outcome"
 
 
-
+#-----------------------------------------------------------------------------------------------------------------------------------#
+#Build and assemble figure 1
+#Load dependencies
+deps <- c('shape', 'wesanderson');
+for (dep in deps){
+  if (dep %in% installed.packages()[,"Package"] == FALSE){
+    install.packages(as.character(dep), quiet=TRUE);
+  } 
+  library(dep, verbose=FALSE, character.only=TRUE)
+}
 
 #-----------------------------------------------------------------------------------------------------------------------------------#
 # Set up multi-panel figure
 plot_file <- '~/Documents/Schloss_Lab/Schubert_humanCdGF_XXXX_2016/results/figures/figure_test.pdf'
-pdf(file=plot_file, width=9, height=9)
+pdf(file=plot_file, width=7, height=9)
 layout(matrix(c(1,
                 2,
                 3), 
@@ -92,15 +95,16 @@ layout(matrix(c(1,
 # Figure 1A - experimental timeline code 
 #-------------------------------------------------------------------------------------------------------------------------------------#
 #---------Option 1, text based timeline---------#
[…]

### 2016-12-16 Projects/HumanCdGF_cdiff_colonization.Rmd
adding Leave one cage out on all

@@
 
+##### __Leave One Out__ - Model __Colonization level__ with Human Donor included
+Model built using all OTUs present in >10% of samples, only displaying top 10% of predictive features
+
+```{r LOCO}
+cfu_all_src_shared_loo_df <- data.frame(colonized_cfu %>% 
+    right_join(community_all_d0) %>% 
+    filter(!is.na(log_cfu_med)) %>% 
+    select(log_cfu_med, cage_id, mouse_id, one_of(names(top_features_src_cfu))))
+source_list <- unique(cfu_all_src_shared_loo_df$cage_id)
+cfu_all_src_shared_loo_df$human_source <- factor(cfu_all_src_shared_loo_df$human_source)
+
+LOcO_probs <- c()
+LOcO_rsq <- c()
+for(i in 1:length(source_list)){
+  by_source <- cfu_all_src_shared_loo_df$cage_id == source_list[i]
+  LOcO_predict <- select(cfu_all_src_shared_loo_df, -mouse_id, -cage_id)
+  train <- LOcO_predict[!by_source,]
+  test <- LOcO_predict[by_source,]
+  set.seed(seed)
+  temp_model <- randomForest(log_cfu_med~., data=train, ntree=1001)
+  LOcO_rsq[i] <- round(tail(temp_model$rsq, 1), 2)
+  LOcO_probs <- rbind(LOcO_probs, data.frame(observed=test$log_cfu_med, 
+                                             predicted=predict(temp_model, test),
+                                             mouse_id=cfu_all_src_shared_loo_df$mouse_id[by_source],
+                                             human_source=rep(source_list[i], nrow(test))))
+}
+LOcO_rsq_overall <- 1 - (sum((LOcO_probs$observed - LOcO_probs$predicted)^2)/
+                   sum((LOcO_probs$observed - mean(LOcO_probs$observed))^2))
+```
+
+##### Leave one Cage out:  
+Rsq of all predictions - `r LOcO_rsq_overall`  
+Rsq of each iteration - `r LOcO_rsq`  
+(cage order =  `r source_list`)
+
+```{r LOSO}
+cfu_all_src_shared_loo_df <- data.frame(colonized_cfu %>% 
+    right_join(community_all_d0) %>% 
+    filter(!is.na(log_cfu_med)) %>% 
+    select(log_cfu_med, mouse_id, one_of(names(top_features_src_cfu))))
+source_list <- unique(cfu_all_src_shared_loo_df$human_source)
+cfu_all_src_shared_loo_df$human_source <- factor(cfu_all_src_shared_loo_df$human_source)
+
+LOcO_probs <- c()
+LOcO_rsq <- c()
+for(i in 1:length(source_list)){
+  by_source <- cfu_all_src_shared_loo_df$human_source == source_list[i]
+  LOcO_predict <- select(cfu_all_src_shared_loo_df, -mouse_id)
+  train <- LOcO_predict[!by_source,]
+  test <- LOcO_predict[by_source,]
+  set.seed(seed)
+  temp_model <- randomForest(log_cfu_med~., data=train, ntree=1001)
+  LOcO_rsq[i] <- round(tail(temp_model$rsq, 1), 2)
+  LOcO_probs <- rbind(LOcO_probs, data.frame(observed=test$log_cfu_med, 
+                                             predicted=predict(temp_model, test),
+                                             mouse_id=cfu_all_src_shared_loo_df$mouse_id[by_source],
+                                             human_source=rep(source_list[i], nrow(test))))
+}
+LOcO_rsq_overall <- 1 - (sum((LOcO_probs$observed - LOcO_probs$predicted)^2)/
+                   sum((LOcO_probs$observed - mean(LOcO_probs$observed))^2))
+```
+
+#### Leave one Source out:  
+Rsq of all predictions - `r LOcO_rsq_overall`  
+Rsq of each iteration - `r LOcO_rsq`  
+(source order =  `r source_list`)
+
+Removing the repeated cages of DA00581 (A and C) causes the model to perform poorly leaving one out.
+So is the performance of the whole model appears to be dependent on two grouping of points (one around ~5-6 log and one around ~8 log ).
+So how well can random forest model the data with out the DA00581 source?
+
+```{r LOCO with out 581}
+cfu_all_src_shared_loo_df <- data.frame(colonized_cfu %>% 
+    right_join(community_all_d0) %>% 
+    filter(!is.na(log_cfu_med)) %>% 
+      filter(!human_source == 'DA00581') %>% 
+    select(log_cfu_med, mouse_id, one_of(names(top_features_src_cfu))))
+source_list <- unique(cfu_all_src_shared_loo_df$human_source)
+cfu_all_src_shared_loo_df$human_source <- factor(cfu_all_src_shared_loo_df$human_source)
+
+LOcO_probs <- c()
+LOcO_rsq <- c()
+for(i in 1:length(source_list)){
+  by_source <- cfu_all_src_shared_loo_df$human_source == source_list[i]
+  LOcO_predict <- select(cfu_all_src_shared_loo_df, -mouse_id)
+  train <- LOcO_predict[!by_source,]
+  test <- LOcO_predict[by_source,]
+  set.seed(seed)
+  temp_model <- randomForest(log_cfu_med~., data=train, ntree=1001)
+  LOcO_rsq[i] <- round(tail(temp_model$rsq, 1), 2)
+  LOcO_probs <- rbind(LOcO_probs, data.frame(observed=test$log_cfu_med, 
+                                             predicted=predict(temp_model, test),
+                                             mouse_id=cfu_all_src_shared_loo_df$mouse_id[by_source],
+                                             human_source=rep(source_list[i], nrow(test))))
+}
+LOcO_rsq_overall <- 1 - (sum((LOcO_probs$observed - LOcO_probs$predicted)^2)/
+                   sum((LOcO_probs$observed - mean(LOcO_probs$observed))^2))
+```
+
+#### Leaving one cage/source out (without Donor 581):  
+Rsq of all predictions - `r LOcO_rsq_overall`  
+Rsq of each iteration - `r LOcO_rsq`  
+(cage/source order =  `r source_list`)
+
+While this is removing 3 cages (8 mice) and decreasing the n, it is removing the smaller, lower grouping of points. So how well can the whole model predict CFU without the DA00581 samples?
+
+```{r all without 581}
+cfu_wo581_df <- colonized_cfu %>% 
+    select(mouse_id, human_source, log_cfu_med) %>% 
+    right_join(community_all_d0) %>% 
+    filter(!human_source == 'DA00581') %>% 
+    select(-mouse_id) %>% 
+    filter(!is.na(log_cfu_med))
+cfu_wo581_df$human_source <- factor(cfu_wo581_df$human_source)
+
+set.seed(seed)
+cfu_wo581_rf <- randomForest(cfu_wo581_df[ , -2], cfu_wo581_df$log_cfu_med,
+                           ntree = 1001,
+                           mtry = 15,
+                           importance = T)
+top_features_wo581 <- head(sort(importance(cfu_wo581_rf)[,1],decreasing = T), n_features)
+```
+
+```{r colonization performance without 581, fig.width=5, fig.height=4}
+data.frame(Observed = cfu_wo581_df$log_cfu_med, 
+           Predicted = cfu_wo581_rf$predicted) %>% 
+  ggplot(aes(x = Predicted, y = Observed)) + geom_point() +
+    xlim(5,9) + ylim(5,9) + 
+    annotate('text', label = paste0(round(tail(cfu_wo581_rf$rsq, 1), 4)*100, '% Var Explained'),
+             x = min(cfu_wo581_rf$predicted), y = max(cfu_wo581_df$log_cfu_med)) + 
+    theme(panel.background = element_rect(linetype = 1, size = 1, fill = 'white', color = 'black'),
+          panel.grid.major.x = element_blank(),
+          panel.grid.minor.x = element_blank(),
+          panel.grid.major.y = element_blank(),
+          panel.grid.minor.y = element_blank()) +
+  labs(x = 'Predicted Colonization (log CFU)', y = 'Observed Colonization (log CFU)')
+
+top_otus_no581 <- get_tax(row_list = names(top_features_wo581)[-1])$tax_label
+top_otus_no581 <- c('Human Source', top_otus_no581)
+```
+
+```{r colonization otu importance without 581, fig.width=6, fig.height=6}
+data.frame(OTU = factor(top_otus_no581, 
+                        levels = top_otus_no581[order(top_features_wo581)]), 
+           Importance = top_features_wo581/max(top_features_wo581)) %>% 
+  ggplot(aes(x = OTU, y = Importance)) + 
+    geom_point() + coord_flip() + 
+    labs(y = 'Importance (Relative MDA)', x = NULL) + 
+    theme(panel.background = element_rect(linetype = 1, size = 1, fill = 'white', color = 'black'),
+          panel.grid.major.x = element_blank(),
+          panel.grid.minor.x = element_blank(),
+          panel.grid.major.y = element_line(colour = "gray", linetype = "dotted"),
+          panel.grid.minor.y = element_line(colour = "gray", linetype = "dotted"),
+          axis.text.y = element_text(color = c(rep('black', length(top_features_src_cfu)-13),
+                                               rep('red', 12), 'black'), 
+                                     hjust = 0))
+```
+
+Without human donor, performance was slightly lower with similar top features
+
+__This decrease in performance when removing Human Donor 581 suggests that with this data set, random forest is unable to predict CFU with a high degree of accuracy.
+This model seems to potentially be identifying the split between the high and low CFU groups, although this is difficult to say with any weight since we only have one source in the low grouping.__
+
+***
+***
+
 #### Modeling CFU with One Mouse per Human Donor
@@
 
+Cages A, C, and D all have the same donor.  
 Is it possible one of the cages of the same donor are not similar, and causing the poor performance?
@@
 By cage:  
+Min. 1st Qu.  Median    Mean 3rd Qu.    Max.   
 `r summary(om_rsq_cage_df$rsq[om_rsq_cage_df$data_set=='test'])`
@@
 By donor:  
+Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
+
 `r summary(om_rsq_pos_df$rsq[om_rsq_pos_df$data_set=='test'])`

### 2016-12-16 code/build_schematic_figure.R
figure 1 assembled and done for now

@@ -30,13 +30,13 @@ write.table(inocula_shared, file='data/process/inocula.shared', quote=F,sep='\t'
 #run dist.shared with iters and subsamp in TYC form, nmds in mothur to get axes file
 #mothur commands
 #dist.shared(shared=inocula.shared, calc=thetayc, subsample=2000)
-#nmds(phylip=inocula.thetayc.0.03.lt.dist)
+#nmds(phylip=inocula.thetayc.0.03.lt.ave.dist)
 #mothur NMDS output:
-# Number of dimensions:	2
-# Lowest stress :	0.318153
-# R-squared for configuration:	0.458028
+#Number of dimensions:	2
+#Lowest stress :	0.3203
+#R-squared for configuration:	0.442409
 
-inoc_nmds <- read.table(file='data/process/inocula.thetayc.0.03.lt.nmds.axes', header = T)
+inoc_nmds <- read.table(file='data/process/inocula.thetayc.0.03.lt.ave.nmds.axes', header = T)
 #merge design and nmds files to prep for plotting 
 inocula_design <- read.table(file='data/raw/donor.design', header = F)
 inocula_meta_design <- read.table(file='data/raw/donor_meta.design', header = F)
@@ -55,12 +55,12 @@ write.table(day0_shared, file='data/process/day0.shared', quote=F,sep='\t',row.n
 
 #run in mothur
 #dist.shared(shared=day0.shared, calc=thetayc, subsample=2000)
-#nmds(phylip=day0.thetayc.0.03.lt.dist)
+#nmds(phylip=day0.thetayc.0.03.lt.ave.dist)
 #mothur NMDS output:
 #Number of dimensions:	2
-#Lowest stress :	0.35381
-#R-squared for configuration:	0.385726
-day0_nmds <- read.table(file='data/process/day0.thetayc.0.03.lt.nmds.axes', header = T)
+#Lowest stress :	0.349084
+#R-squared for configuration:	0.40841
+day0_nmds <- read.table(file='data/process/day0.thetayc.0.03.lt.ave.nmds.axes', header = T)
 #merge with design files to get ready to plot 
 d0_donor <- read.table(file='data/raw/sample_donor.design', header = F)
 day0_donor_nmds <- merge(day0_nmds, d0_donor, by.x='group', by.y='V1')
@@ -85,17 +85,18 @@ for (dep in deps){
 
[…]

### 2016-12-16 Projects/HumanCdGF_cdiff_colonization.Rmd
fix label for plot with leaving 581 out

@@
 
+#### Model __Colonization level__ with Human Donor included (without Donor 581):
+
 ```{r all without 581}
@@
 top_features_wo581 <- head(sort(importance(cfu_wo581_rf)[,1],decreasing = T), n_features)
+
+cfu_ACD_df <- colonized_cfu %>% 
+    select(mouse_id, human_source, cage_id, log_cfu_med) %>% 
+    right_join(community_all_d0) %>% 
+    select(-mouse_id) %>% 
+    filter(!is.na(log_cfu_med))
+cfu_ACD_df$human_source <- factor(cfu_ACD_df$human_source)
+
+cfu_single581_df <- filter(cfu_ACD_df, !cage_id %in% c('A','C')) %>% 
+                      select(-cage_id)
+set.seed(seed)
+cfu_D_rf <- randomForest(cfu_single581_df[ , -2], cfu_single581_df$log_cfu_med,
+                           ntree = 1001,
+                           mtry = 15,
+                           importance = T)
+cfu_single581_df <- filter(cfu_ACD_df, !cage_id %in% c('A','D')) %>% 
+                      select(-cage_id)
+set.seed(seed)
+cfu_C_rf <- randomForest(cfu_single581_df[ , -2], cfu_single581_df$log_cfu_med,
+                           ntree = 1001,
+                           mtry = 15,
+                           importance = T)
+cfu_single581_df <- filter(cfu_ACD_df, !cage_id %in% c('D','C')) %>% 
+                      select(-cage_id)
+set.seed(seed)
+cfu_A_rf <- randomForest(cfu_single581_df[ , -2], cfu_single581_df$log_cfu_med,
+                           ntree = 1001,
+                           mtry = 15,
+                           importance = T)
 ```
@@
 ```
+
+Including one Donor 581 cage increases performance:  
+w/Cage A: `r paste0(round(tail(cfu_A_rf$rsq, 1), 4)*100, '% Var Explained')`   
+w/Cage C: `r paste0(round(tail(cfu_C_rf$rsq, 1), 4)*100, '% Var Explained')`   
+w/Cage D: `r paste0(round(tail(cfu_D_rf$rsq, 1), 4)*100, '% Var Explained')`
 

### 2016-12-16 code/build_figure_2.R
figure 2 assembled

@@ -1,0 +1,95 @@
+# Builds figure 2
+# This script builds the mouse weight loss figure
+#
+#
+# Dependencies: 
+#     * data/process/human_CdGF_metadata.txt
+#     * data/mothur/gf_cdiff.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.subsample.shared
+#     * data/mothur/gf_cdiff.trim.contigs.good.unique.good.filter.unique.precluster.pick.pick.pick.an.unique_list.0.03.cons.taxonomy
+#
+#
+# Output: * results/figures/figure_2.pdf
+
+#load dependencies 
+pack_used <- c('tidyr','ggplot2','dplyr', 'knitr')
+for (dep in pack_used){
+  if (dep %in% installed.packages()[,"Package"] == FALSE){
+    install.packages(as.character(dep), repos = 'http://cran.us.r-project.org', 
+                     quiet=TRUE);
+  }
+  library(dep, verbose=FALSE, character.only=TRUE)
+}
+# need to change these file paths so they match yours duhhh
+
+meta_file   <- 'data/process/human_CdGF_metadata.txt'
+shared_file <- 'data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
+tax_file <- 'data/process/gf_new.an.taxonomy'
+tax_function <- 'code/tax_level.R'
+
+# read in files
+meta_file   <- read.table(meta_file, sep = '\t', header = T, row.names = 'sample_id')
+shared_file <- read.table(shared_file, sep = '\t', header = T, row.names = 'sample_id')
+tax_file <- read.table(tax_file, sep = '\t', header = T, row.names = 'OTU')
+source(tax_function)
+
+# select only mice with 431 cdiff
+meta_file <- meta_file[meta_file$cdiff_strain == '431', ]
+
+# subset shared
+#Create df with relative abundances
+stool_shared <- shared_file[rownames(meta_file[meta_file$day %in% 0, ]) ,]
+# use %in% instead of == because to ignore NA
+# rename day 0 community rows by mouse id - since meta rows have day in row names, use mouse id for both
+rownames(stool_shared) <- meta_file$mouse_id[rownames(meta_file) %in% 
+                                               rownames(stool_shared) ]
[…]

### 2016-12-16 Projects/HumanCdGF_cdiff_colonization.Rmd
rearrange Rmd and clean up language

@@
 ***
+***
+
+#### Modeling CFU with One Mouse per Human Donor
+In an effort to remove any potential donor community dependency, the model is trained on a single mouse per human donor source, and then tested on all the remaining mice
+
+##### With all OTUs
+```{r one mouse model}
+
+#One per source
+one_mouse_df <- data.frame(colonized_cfu %>% 
+    right_join(community_all_d0) %>% 
+    filter(!is.na(log_cfu_med)))
+one_mouse_df$human_source <- factor(one_mouse_df$human_source)
+
+om_import_otus <- data.frame(otu = colnames(select(one_mouse_df, -mouse_id, -cage_id, -log_cfu_med)))
+om_pred <- c()
+om_rsq <- c()
+for(i in 1:iters){
+    set.seed(i)
+    train <- one_mouse_df %>% 
+      group_by(human_source) %>% 
+      sample_n(1) %>% ungroup() 
+    test <- one_mouse_df %>% 
+      filter(!mouse_id %in% train$mouse_id)
+    set.seed(seed)
+    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id, -cage_id),
+                             importance = TRUE, keep.forest = T, ntree = n_trees)
+    om_pred <- rbind(om_pred,
+                     data.frame(select(test, mouse_id, cage_id, log_cfu_med),
+                          predicted=predict(rf_om, test),
+                          iter = i))
+    om_rsq <- rbind(om_rsq, c(iter = i,
+                              mice = paste(train$mouse_id, collapse = ' '), 
+                              train_rsq = tail(rf_om$rsq,1),
+                              test_rsq = 1 - (sum((test$log_cfu_med - predict(rf_om, test))^2)/
+                                sum((test$log_cfu_med - mean(test$log_cfu_med))^2))))
+    import_df <- importance(rf_om)
+    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
+    om_import_otus <- merge(om_import_otus, import_df, by.x = 'otu', by.y = 'row.names') %>% 
+      select(-IncNodePurity)
+}
+om_rsq <- data.frame(iter = as.numeric(om_rsq[,1]),
+           mice = om_rsq[,2],
+           train_rsq = as.numeric(om_rsq[,3]),
+           test_rsq = as.numeric(om_rsq[,4]),
+           stringsAsFactors = F)
+om_rsq_overall <- 1 - (sum((om_pred$log_cfu_med - om_pred$predicted)^2)/
+               sum((om_pred$log_cfu_med - mean(om_pred$log_cfu_med))^2))
+
+om_pred <- rename(om_pred, observed = log_cfu_med)
+
+all_otus_importance <- gather(om_import_otus,OTU, import_val,-otu)[,-2]
+all_otus_importance$votes <- ifelse(all_otus_importance$import_val > 0, 1, 0)
+all_otus_importance <- data.frame(table(all_otus_importance[,-2]))
+pos_otu <- all_otus_importance[(filter(all_otus_importance, votes == 1) %>% select(Freq) - 
+                      filter(all_otus_importance, votes == 0) %>% select(Freq)) > 0, ]
+pos_otu <- data.frame(predictor = pos_otu$otu[pos_otu$votes==1],
+                      filter(pos_otu, votes == 1) %>% select(Freq) - 
+                      filter(pos_otu, votes == 0) %>% select(Freq)) %>% 
+              arrange(desc(Freq))
+```
+
+When using all OTUs the model preforms poorly.
+The median Rsq of the trainset was `r median(om_rsq$train_rsq)`, this is likely do to the features.
+This is a summary of the importance scores for all predictors:
+Min.  1st Qu.   Median     Mean  3rd Qu.     Max.     NA's   
+`r summary(unlist(select(om_import_otus, -otu)))`  
+`r round(100*sum(om_import_otus[,-1]<=0, na.rm = T)/length(om_import_otus[,-1]<=0),2)`% of OTUs have an importance of 0 or less.
+The median Rsq of the test set was `r median(om_rsq$test_rsq)` and the Rsq of all predicted values was `r om_rsq_overall`.
+
+This one mouse per source model is improved through reducing the amount of negative features, either utilizing the feature set via the top X % of predictors from the model of all mice or take only the features that are more often have an importance > 0 (OTUs : `r as.character(pos_otu$predictor)`).
+
+```{r one mouse model - with OTUs positive in All OTU one mouse model}
+
+#One per source - > 0
+one_mouse_pos_otus_df <- data.frame(colonized_cfu %>% 
+    right_join(community_all_d0) %>% 
+    filter(!is.na(log_cfu_med))) %>% 
+    select(mouse_id, human_source, log_cfu_med,
+           one_of(as.character(pos_otu$predictor)))
+one_mouse_pos_otus_df$human_source <- factor(one_mouse_pos_otus_df$human_source)
+
+om_import_pos_otus <- data.frame(otu = colnames(select(one_mouse_pos_otus_df, -mouse_id, -log_cfu_med)))
+om_pred_pos <- c()
+om_rsq_pos <- c()
+for(i in 1:iters){
+    set.seed(i)
+    train <- one_mouse_pos_otus_df %>% 
+      group_by(human_source) %>% 
+      sample_n(1) %>% ungroup() 
+    test <- one_mouse_pos_otus_df %>% 
+      filter(!mouse_id %in% train$mouse_id)
+    set.seed(seed)
+    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id),
+                             importance = TRUE, ntree = n_trees, mtry = 10)
+    om_pred_pos <- rbind(om_pred_pos,
+                     data.frame(select(test, mouse_id, log_cfu_med),
+                          predicted=predict(rf_om, test),
+                          iter = i))
+    om_rsq_pos <- rbind(om_rsq_pos, c(iter = i,
+                              mice = paste(train$mouse_id, collapse = ' '), 
+                              train_rsq = tail(rf_om$rsq,1),
+                              test_rsq = 1 - (sum((test$log_cfu_med - predict(rf_om, test))^2)/
+                                sum((test$log_cfu_med - mean(test$log_cfu_med))^2))))
+    import_df <- importance(rf_om)
+    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
+    om_import_pos_otus <- merge(om_import_pos_otus, import_df, by.x = 'otu', by.y = 'row.names') %>% 
+      select(-IncNodePurity)
+}
+
+om_rsq_pos_df <- data.frame(iter = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='iter']),
+                            mice = om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='mice'],
+           train = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='train_rsq']),
+           test = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='test_rsq']),
+           stringsAsFactors = F) %>% 
+  gather(data_set, rsq, train, test)
+```
+
+One Mouse per Human Donor with OTUs > 0 give:  
+  *  median training set rsq of 0.08652421   
+  *  median testing set rsq of 0.5331836   
+  *  overall rsq of 0.5534567
+
+One Mouse per Human Donor with the top 10% predictive OTUs when using all mice  
+  *  median training set rsq of 0.07336261  
+  *  median testing set rsq of 0.6070359  
+  *  overall rsq of 0.633238
+
+The % var explained values from each individual iteration appears to have a bimodal distribution.
+
+```{r om_bimodality, fig.width= 4, fig.height=4}
+ggplot(om_rsq_pos_df, aes(rsq, fill = data_set)) + 
+  geom_density(alpha = 0.25) + ggtitle('Rsq Distribution for One Mouse per Human Donor')
+```
+
+Is this bimodal distribution due to :  
+  * Due to small n?  
+  * Due to incomplete source dependency?  
+  * Ratio of train to test?
+
+Using random forest to identify the mice driving bimodal Rsq distribution
+
+```{r one_mouse_rsq_predictors, fig.width=4, fig.height=8}
+
+predict_rsq <- om_rsq_pos_df %>% 
+                 filter(data_set == 'test')
+predict_rsq$group_mean <- ifelse(predict_rsq$rsq > 0.62, T, F)
+
+mice_diff <- data.frame(matrix(rep(NA, nrow(predict_rsq)*length(unique(colonized_cfu$mouse_id))),
+                               nrow=nrow(predict_rsq)))
+colnames(mice_diff) <- unique(colonized_cfu$mouse_id)
+for(i in 1:nrow(mice_diff)){
+    mice_diff[i,] <- ifelse(colnames(mice_diff) %in% strsplit(predict_rsq$mice, ' ')[[i]], 0,1)
+}
+
+predict_rsq_df <- cbind(rsq = predict_rsq$group_mean, mice_diff)
+predict_rsq_df$rsq <- as.factor(predict_rsq_df$rsq)
+set.seed(seed)
+rsq_grp_rf <- randomForest(predict_rsq_df[,-1],predict_rsq_df$rsq, importance = T)
+importance_rsq <- data.frame(mouse = dimnames(importance(rsq_grp_rf))[[1]],importance(rsq_grp_rf))
+one_mouse_rsq_pred <- arrange(importance_rsq, desc(MeanDecreaseAccuracy)) %>% 
+  select(mouse,MeanDecreaseAccuracy) %>% 
+  mutate(Relative_MDA = MeanDecreaseAccuracy/max(MeanDecreaseAccuracy))
+
+data.frame(mouse = factor(one_mouse_rsq_pred$mouse, 
+                        levels = one_mouse_rsq_pred$mouse[order(one_mouse_rsq_pred$Relative_MDA)]), 
+           Importance = one_mouse_rsq_pred$Relative_MDA) %>% 
+ggplot(aes(x=mouse, y=Importance)) + 
+   geom_point() + coord_flip() + 
+    labs(y = 'Importance (Relative MDA)', x = NULL) + 
+    theme(panel.background = element_rect(linetype = 1, size = 1, fill = 'white', color = 'black'),
+          axis.ticks = element_line(size = 2),
+          panel.grid.major.x = element_blank(),
+          panel.grid.minor.x = element_blank(),
+          panel.grid.major.y = element_line(colour = "gray", linetype = "dotted"),
+          panel.grid.minor.y = element_line(colour = "gray", linetype = "dotted"),
+          axis.text.y = element_text(color = c(rep('black', length(one_mouse_rsq_pred$Relative_MDA)-
+                                                     sum(one_mouse_rsq_pred$MeanDecreaseAccuracy > 0)),
+                                               rep('red', sum(one_mouse_rsq_pred$MeanDecreaseAccuracy > 0))), 
+                                     hjust = 0))
+
+```
+
+Cages A, C, and D all have the same donor.  
+Is it possible one of the cages of the same donor are not similar, and causing the poor performance?
+
+```{r one mouse model CAGE - with OTUs positive in All OTU one mouse model}
+
+#One per source - > 0
+one_mouse_pos_otus_df <- data.frame(colonized_cfu %>% 
+    right_join(community_all_d0) %>% 
+    filter(!is.na(log_cfu_med))) %>% 
+    select(mouse_id, cage_id, log_cfu_med,
+           one_of(as.character(pos_otu$predictor)))
+one_mouse_pos_otus_df$cage_id <- factor(one_mouse_pos_otus_df$cage_id)
+
+om_import_pos_otus <- data.frame(otu = colnames(select(one_mouse_pos_otus_df, -mouse_id, -log_cfu_med)))
+om_pred_pos <- c()
+om_rsq_pos <- c()
+for(i in 1:iters){
+    set.seed(i)
+    train <- one_mouse_pos_otus_df %>% 
+      group_by(cage_id) %>% 
+      sample_n(1) %>% ungroup() 
+    test <- one_mouse_pos_otus_df %>% 
+      filter(!mouse_id %in% train$mouse_id)
+    set.seed(seed)
+    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id),
+                             importance = TRUE, ntree = n_trees, mtry = 10)
+    om_pred_pos <- rbind(om_pred_pos,
+                     data.frame(select(test, mouse_id, log_cfu_med),
+                          predicted=predict(rf_om, test),
+                          iter = i))
+    om_rsq_pos <- rbind(om_rsq_pos, c(iter = i,
+                              mice = paste(train$mouse_id, collapse = ' '), 
+                              train_rsq = tail(rf_om$rsq,1),
+                              test_rsq = 1 - (sum((test$log_cfu_med - predict(rf_om, test))^2)/
+                                sum((test$log_cfu_med - mean(test$log_cfu_med))^2))))
+    import_df <- importance(rf_om)
+    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
+    om_import_pos_otus <- merge(om_import_pos_otus, import_df, by.x = 'otu', by.y = 'row.names') %>% 
+      select(-IncNodePurity)
+}
+
+om_rsq_cage_df <- data.frame(iter = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='iter']),
+                            mice = om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='mice'],
+           train = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='train_rsq']),
+           test = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='test_rsq']),
+           stringsAsFactors = F) %>% 
+  gather(data_set, rsq, train, test)
+```
+
+```{r om_cage_rsq_dist, fig.width= 4, fig.height=4}
+ggplot(om_rsq_cage_df, aes(rsq, fill = data_set)) + 
+  geom_density(alpha = 0.25) + ggtitle('Rsq Distribution for One Mouse per Cage')
+```
+
+The distribution of Rsq values when modeling by cage appears to be improved in comparison of by human donor
+
+By cage:  
+Min., 1st Qu.,  Median,    Mean, 3rd Qu.,    Max.,   
+`r summary(om_rsq_cage_df$rsq[om_rsq_cage_df$data_set=='test'])`
+
+By donor:  
+Min., 1st Qu.,  Median,    Mean, 3rd Qu.,    Max.,   
+`r summary(om_rsq_pos_df$rsq[om_rsq_pos_df$data_set=='test'])`
+
+```{r compare number of 581 donor cages, fig.height=5, fig.width=6}
+
+#One per source - > 0
+one_mouse_pos_otus_df <- data.frame(colonized_cfu %>% 
+    right_join(community_all_d0) %>% 
+    filter(!is.na(log_cfu_med))) %>% 
+    select(mouse_id, human_source, log_cfu_med, cage_id,
+           one_of(as.character(pos_otu$predictor))) %>% 
+    filter(!cage_id %in% c('A','D')) %>% 
+    select(-cage_id)
+one_mouse_pos_otus_df$human_source <- factor(one_mouse_pos_otus_df$human_source)
+
+om_rsq_pos <- c()
+for(i in 1:iters){
+    set.seed(i)
+    train <- one_mouse_pos_otus_df %>% 
+      group_by(human_source) %>% 
+      sample_n(1) %>% ungroup() 
+    test <- one_mouse_pos_otus_df %>% 
+      filter(!mouse_id %in% train$mouse_id)
+    set.seed(seed)
+    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id),
+                             importance = TRUE, ntree = n_trees, mtry = 10)
+    om_rsq_pos <- rbind(om_rsq_pos, c(iter = i,
+                              mice = paste(train$mouse_id, collapse = ' '), 
+                              train_rsq = tail(rf_om$rsq,1),
+                              test_rsq = 1 - (sum((test$log_cfu_med - predict(rf_om, test))^2)/
+                                sum((test$log_cfu_med - mean(test$log_cfu_med))^2))))
+    import_df <- importance(rf_om)
+    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
+}
+
+om_rsq_C_only_df <- data.frame(iter = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='iter']),
+                            mice = om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='mice'],
+           train = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='train_rsq']),
+           test = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='test_rsq']),
+           stringsAsFactors = F) %>% 
+  gather(data_set, rsq, train, test)
+
+one_mouse_pos_otus_df <- data.frame(colonized_cfu %>% 
+    right_join(community_all_d0) %>% 
+    filter(!is.na(log_cfu_med))) %>% 
+    select(mouse_id, human_source, log_cfu_med, cage_id,
+           one_of(as.character(pos_otu$predictor))) %>% 
+    filter(!cage_id == 'A') %>% 
+    select(-cage_id)
+one_mouse_pos_otus_df$human_source <- factor(one_mouse_pos_otus_df$human_source)
+
+om_rsq_pos <- c()
+for(i in 1:iters){
+    set.seed(i)
+    train <- one_mouse_pos_otus_df %>% 
+      group_by(human_source) %>% 
+      sample_n(1) %>% ungroup() 
+    test <- one_mouse_pos_otus_df %>% 
+      filter(!mouse_id %in% train$mouse_id)
+    set.seed(seed)
+    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id),
+                             importance = TRUE, ntree = n_trees, mtry = 10)
+    om_rsq_pos <- rbind(om_rsq_pos, c(iter = i,
+                              mice = paste(train$mouse_id, collapse = ' '), 
+                              train_rsq = tail(rf_om$rsq,1),
+                              test_rsq = 1 - (sum((test$log_cfu_med - predict(rf_om, test))^2)/
+                                sum((test$log_cfu_med - mean(test$log_cfu_med))^2))))
+    import_df <- importance(rf_om)
+    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
+}
+
+om_rsq_AD_df <- data.frame(iter = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='iter']),
+                            mice = om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='mice'],
+           train = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='train_rsq']),
+           test = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='test_rsq']),
+           stringsAsFactors = F) %>% 
+  gather(data_set, rsq, train, test)
+
+rbind(data.frame(rsq = om_rsq_pos_df$rsq[om_rsq_pos_df$data_set=='test'], data = '3'),
+      rbind(data.frame(rsq = om_rsq_AD_df$rsq[om_rsq_AD_df$data_set=='test'], data = '2'),
+            data.frame(rsq = om_rsq_C_only_df$rsq[om_rsq_C_only_df$data_set=='test'], data = '1'))) %>% 
+  ggplot(aes(rsq, color = data)) + 
+  geom_density() + ggtitle('Rsq Distribution varying number of Donor 581 cages')
+```
+
+It seems the assumption all donors have the same community may be false because the bimodal distribution appears to be attributed to which mice from that group are used.
+However, when running the analysis with only one cage from the donor with multiple cages (581) it performs as we observe for the lower peak in the bimodal distribution and when a second cage from the same donor is added we observe Rsq values at the higher peak in the bimodal distribution.
+Thus there appears to be a cage effect improving the performance and the bimodal distribution maybe due to a variation in the ability of these communities to be able to predict each other, ie some may be more similar to each other but still as a group more different than the other cages.
+
+***
+***
+
+## So how does this cage/donor affect the overall CFU prediction?
 
@@
 Removing the repeated cages of DA00581 (A and C) causes the model to perform poorly leaving one out.  
-So is the performance of the whole model appears to be dependent on two grouping of points (one around ~5-6 log and one around ~8 log ).
-So how well can random forest model the data with out the DA00581 source?
+The performance of the whole model appears to be dependent on two grouping of points (one around ~5-6 log (all Donor 581) and one around ~8 log ).  
+__How well can random forest model the data with out the DA00581 source?__
 
@@
 This model seems to potentially be identifying the split between the high and low CFU groups, although this is difficult to say with any weight since we only have one source in the low grouping.__
-
-***
-***
-
-#### Modeling CFU with One Mouse per Human Donor
-In an effort to remove any potential donor community dependency, the model is trained on a single mouse per human donor source, and then tested on all the remaining mice
-
-##### With all OTUs
-```{r one mouse model}
-
-#One per source
-one_mouse_df <- data.frame(colonized_cfu %>% 
-    right_join(community_all_d0) %>% 
-    filter(!is.na(log_cfu_med)))
-one_mouse_df$human_source <- factor(one_mouse_df$human_source)
-
-om_import_otus <- data.frame(otu = colnames(select(one_mouse_df, -mouse_id, -cage_id, -log_cfu_med)))
-om_pred <- c()
-om_rsq <- c()
-for(i in 1:iters){
-    set.seed(i)
-    train <- one_mouse_df %>% 
-      group_by(human_source) %>% 
-      sample_n(1) %>% ungroup() 
-    test <- one_mouse_df %>% 
-      filter(!mouse_id %in% train$mouse_id)
-    set.seed(seed)
-    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id, -cage_id),
-                             importance = TRUE, keep.forest = T, ntree = n_trees)
-    om_pred <- rbind(om_pred,
-                     data.frame(select(test, mouse_id, cage_id, log_cfu_med),
-                          predicted=predict(rf_om, test),
-                          iter = i))
-    om_rsq <- rbind(om_rsq, c(iter = i,
-                              mice = paste(train$mouse_id, collapse = ' '), 
-                              train_rsq = tail(rf_om$rsq,1),
-                              test_rsq = 1 - (sum((test$log_cfu_med - predict(rf_om, test))^2)/
-                                sum((test$log_cfu_med - mean(test$log_cfu_med))^2))))
-    import_df <- importance(rf_om)
-    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
-    om_import_otus <- merge(om_import_otus, import_df, by.x = 'otu', by.y = 'row.names') %>% 
-      select(-IncNodePurity)
-}
-om_rsq <- data.frame(iter = as.numeric(om_rsq[,1]),
-           mice = om_rsq[,2],
-           train_rsq = as.numeric(om_rsq[,3]),
-           test_rsq = as.numeric(om_rsq[,4]),
-           stringsAsFactors = F)
-om_rsq_overall <- 1 - (sum((om_pred$log_cfu_med - om_pred$predicted)^2)/
-               sum((om_pred$log_cfu_med - mean(om_pred$log_cfu_med))^2))
-
-om_pred <- rename(om_pred, observed = log_cfu_med)
-
-all_otus_importance <- gather(om_import_otus,OTU, import_val,-otu)[,-2]
-all_otus_importance$votes <- ifelse(all_otus_importance$import_val > 0, 1, 0)
-all_otus_importance <- data.frame(table(all_otus_importance[,-2]))
-pos_otu <- all_otus_importance[(filter(all_otus_importance, votes == 1) %>% select(Freq) - 
-                      filter(all_otus_importance, votes == 0) %>% select(Freq)) > 0, ]
-pos_otu <- data.frame(predictor = pos_otu$otu[pos_otu$votes==1],
-                      filter(pos_otu, votes == 1) %>% select(Freq) - 
-                      filter(pos_otu, votes == 0) %>% select(Freq)) %>% 
-              arrange(desc(Freq))
-```
-
-When using all OTUs the model preforms poorly.
-The median Rsq of the trainset was `r median(om_rsq$train_rsq)`, this is likely do to the features.
-This is a summary of the importance scores for all predictors:
-`r summary(unlist(select(om_import_otus, -otu)))`  
-`r round(100*sum(om_import_otus[,-1]<=0, na.rm = T)/length(om_import_otus[,-1]<=0),2)`% of OTUs have an importance of 0 or less.
-The median Rsq of the test set was `r median(om_rsq$test_rsq)` and the Rsq of all predicted values was `r om_rsq_overall`.
-
-This one mouse per source model is improved through reducing the amount of negative features, either utilizing the feature set via the top X % of predictors from the model of all mice or take only the features that are more often have an importance > 0 (OTUs : `r as.character(pos_otu$predictor)`).
-
-```{r one mouse model - with OTUs positive in All OTU one mouse model}
-
-#One per source - > 0
-one_mouse_pos_otus_df <- data.frame(colonized_cfu %>% 
-    right_join(community_all_d0) %>% 
-    filter(!is.na(log_cfu_med))) %>% 
-    select(mouse_id, human_source, log_cfu_med,
-           one_of(as.character(pos_otu$predictor)))
-one_mouse_pos_otus_df$human_source <- factor(one_mouse_pos_otus_df$human_source)
-
-om_import_pos_otus <- data.frame(otu = colnames(select(one_mouse_pos_otus_df, -mouse_id, -log_cfu_med)))
-om_pred_pos <- c()
-om_rsq_pos <- c()
-for(i in 1:iters){
-    set.seed(i)
-    train <- one_mouse_pos_otus_df %>% 
-      group_by(human_source) %>% 
-      sample_n(1) %>% ungroup() 
-    test <- one_mouse_pos_otus_df %>% 
-      filter(!mouse_id %in% train$mouse_id)
-    set.seed(seed)
-    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id),
-                             importance = TRUE, ntree = n_trees, mtry = 10)
-    om_pred_pos <- rbind(om_pred_pos,
-                     data.frame(select(test, mouse_id, log_cfu_med),
-                          predicted=predict(rf_om, test),
-                          iter = i))
-    om_rsq_pos <- rbind(om_rsq_pos, c(iter = i,
-                              mice = paste(train$mouse_id, collapse = ' '), 
-                              train_rsq = tail(rf_om$rsq,1),
-                              test_rsq = 1 - (sum((test$log_cfu_med - predict(rf_om, test))^2)/
-                                sum((test$log_cfu_med - mean(test$log_cfu_med))^2))))
-    import_df <- importance(rf_om)
-    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
-    om_import_pos_otus <- merge(om_import_pos_otus, import_df, by.x = 'otu', by.y = 'row.names') %>% 
-      select(-IncNodePurity)
-}
-
-om_rsq_pos_df <- data.frame(iter = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='iter']),
-                            mice = om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='mice'],
-           train = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='train_rsq']),
-           test = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='test_rsq']),
-           stringsAsFactors = F) %>% 
-  gather(data_set, rsq, train, test)
-```
-
-One Mouse per Human Donor with OTUs > 0 give:
-  *  median training set rsq of 0.08652421  
-  *  median testing set rsq of 0.5331836  
-  *  overall rsq of 0.5534567
-
-One Mouse per Human Donor with the top 10% predictive OTUs when using all mice  
-  *  median training set rsq of 0.07336261  
-  *  median testing set rsq of 0.6070359  
-  *  overall rsq of 0.633238
-
-Looking at the % var explained values from each individual iteration, there appears to be a bimodal distribution in the performance.
-
-```{r om_bimodality, fig.width= 4, fig.height=4}
-ggplot(om_rsq_pos_df, aes(rsq, fill = data_set)) + 
-  geom_density(alpha = 0.25) + ggtitle('Rsq Distribution for One Mouse per Human Donor')
-```
-
-Is this bimodal distribution due to :  
-  * Due to small n?  
-  * Due to incomplete source dependency?  
-  * Ratio of train to test?
-
-Using random forest to identify the mice driving bimodal Rsq distribution
-
-```{r one_mouse_rsq_predictors, fig.width=4, fig.height=8}
-
-predict_rsq <- om_rsq_pos_df %>% 
-                 filter(data_set == 'test')
-predict_rsq$group_mean <- ifelse(predict_rsq$rsq > 0.62, T, F)
-
-mice_diff <- data.frame(matrix(rep(NA, nrow(predict_rsq)*length(unique(colonized_cfu$mouse_id))),
-                               nrow=nrow(predict_rsq)))
-colnames(mice_diff) <- unique(colonized_cfu$mouse_id)
-for(i in 1:nrow(mice_diff)){
-    mice_diff[i,] <- ifelse(colnames(mice_diff) %in% strsplit(predict_rsq$mice, ' ')[[i]], 0,1)
-}
-
-predict_rsq_df <- cbind(rsq = predict_rsq$group_mean, mice_diff)
-predict_rsq_df$rsq <- as.factor(predict_rsq_df$rsq)
-set.seed(seed)
-rsq_grp_rf <- randomForest(predict_rsq_df[,-1],predict_rsq_df$rsq, importance = T)
-importance_rsq <- data.frame(mouse = dimnames(importance(rsq_grp_rf))[[1]],importance(rsq_grp_rf))
-one_mouse_rsq_pred <- arrange(importance_rsq, desc(MeanDecreaseAccuracy)) %>% 
-  select(mouse,MeanDecreaseAccuracy) %>% 
-  mutate(Relative_MDA = MeanDecreaseAccuracy/max(MeanDecreaseAccuracy))
-
-data.frame(mouse = factor(one_mouse_rsq_pred$mouse, 
-                        levels = one_mouse_rsq_pred$mouse[order(one_mouse_rsq_pred$Relative_MDA)]), 
-           Importance = one_mouse_rsq_pred$Relative_MDA) %>% 
-ggplot(aes(x=mouse, y=Importance)) + 
-   geom_point() + coord_flip() + 
-    labs(y = 'Importance (Relative MDA)', x = NULL) + 
-    theme(panel.background = element_rect(linetype = 1, size = 1, fill = 'white', color = 'black'),
-          axis.ticks = element_line(size = 2),
-          panel.grid.major.x = element_blank(),
-          panel.grid.minor.x = element_blank(),
-          panel.grid.major.y = element_line(colour = "gray", linetype = "dotted"),
-          panel.grid.minor.y = element_line(colour = "gray", linetype = "dotted"),
-          axis.text.y = element_text(color = c(rep('black', length(one_mouse_rsq_pred$Relative_MDA)-
-                                                     sum(one_mouse_rsq_pred$MeanDecreaseAccuracy > 0)),
-                                               rep('red', sum(one_mouse_rsq_pred$MeanDecreaseAccuracy > 0))), 
-                                     hjust = 0))
-
-```
-
-Cages A, C, and D all have the same donor.  
-Is it possible one of the cages of the same donor are not similar, and causing the poor performance?
-
-```{r one mouse model CAGE - with OTUs positive in All OTU one mouse model}
-
-#One per source - > 0
-one_mouse_pos_otus_df <- data.frame(colonized_cfu %>% 
-    right_join(community_all_d0) %>% 
-    filter(!is.na(log_cfu_med))) %>% 
-    select(mouse_id, cage_id, log_cfu_med,
-           one_of(as.character(pos_otu$predictor)))
-one_mouse_pos_otus_df$cage_id <- factor(one_mouse_pos_otus_df$cage_id)
-
-om_import_pos_otus <- data.frame(otu = colnames(select(one_mouse_pos_otus_df, -mouse_id, -log_cfu_med)))
-om_pred_pos <- c()
-om_rsq_pos <- c()
-for(i in 1:iters){
-    set.seed(i)
-    train <- one_mouse_pos_otus_df %>% 
-      group_by(cage_id) %>% 
-      sample_n(1) %>% ungroup() 
-    test <- one_mouse_pos_otus_df %>% 
-      filter(!mouse_id %in% train$mouse_id)
-    set.seed(seed)
-    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id),
-                             importance = TRUE, ntree = n_trees, mtry = 10)
-    om_pred_pos <- rbind(om_pred_pos,
-                     data.frame(select(test, mouse_id, log_cfu_med),
-                          predicted=predict(rf_om, test),
-                          iter = i))
-    om_rsq_pos <- rbind(om_rsq_pos, c(iter = i,
-                              mice = paste(train$mouse_id, collapse = ' '), 
-                              train_rsq = tail(rf_om$rsq,1),
-                              test_rsq = 1 - (sum((test$log_cfu_med - predict(rf_om, test))^2)/
-                                sum((test$log_cfu_med - mean(test$log_cfu_med))^2))))
-    import_df <- importance(rf_om)
-    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
-    om_import_pos_otus <- merge(om_import_pos_otus, import_df, by.x = 'otu', by.y = 'row.names') %>% 
-      select(-IncNodePurity)
-}
-
-om_rsq_cage_df <- data.frame(iter = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='iter']),
-                            mice = om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='mice'],
-           train = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='train_rsq']),
-           test = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='test_rsq']),
-           stringsAsFactors = F) %>% 
-  gather(data_set, rsq, train, test)
-```
-
-Looking at the % var explained values from each individual iteration, there appears to be a bimodal distribution in the performance.
-
-```{r om_cage_rsq_dist, fig.width= 4, fig.height=4}
-ggplot(om_rsq_cage_df, aes(rsq, fill = data_set)) + 
-  geom_density(alpha = 0.25) + ggtitle('Rsq Distribution for One Mouse per Cage')
-```
-
-By cage:  
-Min. 1st Qu.  Median    Mean 3rd Qu.    Max.   
-`r summary(om_rsq_cage_df$rsq[om_rsq_cage_df$data_set=='test'])`
-
-By donor:  
-Min. 1st Qu.  Median    Mean 3rd Qu.    Max.
-
-`r summary(om_rsq_pos_df$rsq[om_rsq_pos_df$data_set=='test'])`
-
-```{r compare number of 581 donor cages, fig.height=5, fig.width=5}
-
-#One per source - > 0
-one_mouse_pos_otus_df <- data.frame(colonized_cfu %>% 
-    right_join(community_all_d0) %>% 
-    filter(!is.na(log_cfu_med))) %>% 
-    select(mouse_id, human_source, log_cfu_med, cage_id,
-           one_of(as.character(pos_otu$predictor))) %>% 
-    filter(!cage_id %in% c('A','D')) %>% 
-    select(-cage_id)
-one_mouse_pos_otus_df$human_source <- factor(one_mouse_pos_otus_df$human_source)
-
-om_rsq_pos <- c()
-for(i in 1:iters){
-    set.seed(i)
-    train <- one_mouse_pos_otus_df %>% 
-      group_by(human_source) %>% 
-      sample_n(1) %>% ungroup() 
-    test <- one_mouse_pos_otus_df %>% 
-      filter(!mouse_id %in% train$mouse_id)
-    set.seed(seed)
-    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id),
-                             importance = TRUE, ntree = n_trees, mtry = 10)
-    om_rsq_pos <- rbind(om_rsq_pos, c(iter = i,
-                              mice = paste(train$mouse_id, collapse = ' '), 
-                              train_rsq = tail(rf_om$rsq,1),
-                              test_rsq = 1 - (sum((test$log_cfu_med - predict(rf_om, test))^2)/
-                                sum((test$log_cfu_med - mean(test$log_cfu_med))^2))))
-    import_df <- importance(rf_om)
-    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
-}
-
-om_rsq_C_only_df <- data.frame(iter = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='iter']),
-                            mice = om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='mice'],
-           train = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='train_rsq']),
-           test = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='test_rsq']),
-           stringsAsFactors = F) %>% 
-  gather(data_set, rsq, train, test)
-
-one_mouse_pos_otus_df <- data.frame(colonized_cfu %>% 
-    right_join(community_all_d0) %>% 
-    filter(!is.na(log_cfu_med))) %>% 
-    select(mouse_id, human_source, log_cfu_med, cage_id,
-           one_of(as.character(pos_otu$predictor))) %>% 
-    filter(!cage_id == 'A') %>% 
-    select(-cage_id)
-one_mouse_pos_otus_df$human_source <- factor(one_mouse_pos_otus_df$human_source)
-
-om_rsq_pos <- c()
-for(i in 1:iters){
-    set.seed(i)
-    train <- one_mouse_pos_otus_df %>% 
-      group_by(human_source) %>% 
-      sample_n(1) %>% ungroup() 
-    test <- one_mouse_pos_otus_df %>% 
-      filter(!mouse_id %in% train$mouse_id)
-    set.seed(seed)
-    rf_om <- randomForest(log_cfu_med ~ ., data = select(train, -mouse_id),
-                             importance = TRUE, ntree = n_trees, mtry = 10)
-    om_rsq_pos <- rbind(om_rsq_pos, c(iter = i,
-                              mice = paste(train$mouse_id, collapse = ' '), 
-                              train_rsq = tail(rf_om$rsq,1),
-                              test_rsq = 1 - (sum((test$log_cfu_med - predict(rf_om, test))^2)/
-                                sum((test$log_cfu_med - mean(test$log_cfu_med))^2))))
-    import_df <- importance(rf_om)
-    colnames(import_df)[1] <- paste(train$mouse_id, collapse = ' ')
-}
-
-om_rsq_AD_df <- data.frame(iter = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='iter']),
-                            mice = om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='mice'],
-           train = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='train_rsq']),
-           test = as.numeric(om_rsq_pos[,dimnames(om_rsq_pos)[[2]]=='test_rsq']),
-           stringsAsFactors = F) %>% 
-  gather(data_set, rsq, train, test)
-
-rbind(data.frame(rsq = om_rsq_pos_df$rsq[om_rsq_pos_df$data_set=='test'], data = 'all'),
-      rbind(data.frame(rsq = om_rsq_AD_df$rsq[om_rsq_AD_df$data_set=='test'], data = 'Two_only'),
-            data.frame(rsq = om_rsq_C_only_df$rsq[om_rsq_C_only_df$data_set=='test'], data = 'One_only'))) %>% 
-  ggplot(aes(rsq, color = data)) + 
-  geom_density() + ggtitle('Rsq Distribution when multiple cages from same Donor present')
-```
-
-It seemed possible the assumption all donors have the same community may be false.
-However, when running the analysis with only one cage from the donor with multiple cages it performs as we observe for the lower peak in the bimodal distribution and when a second cage from the same donor is added we observe Rsq at the higher peak in the bimodal distribution.
-Thus there appears to be a cage effect improving the performance and the bimodal distribution maybe due to a variation in the ability of these communities to be able to predict each other, ie some be more similar to each other but still as a group more different than the other cages.

### 2017-01-05 code/HumanCdGF_cdiff_colonization.Rmd
Add plot predicting CFU without High and Low

--- 2016-12-16 Projects/HumanCdGF_cdiff_colonization.Rmd
@@
 This model seems to potentially be identifying the split between the high and low CFU groups, although this is difficult to say with any weight since we only have one source in the low grouping.__
+
+#### Remove High and Low CFU sources
+
+```{r all without 581 or 1134/1146}
+cfu_wo_hilo_df <- colonized_cfu %>% 
+    select(mouse_id, human_source, log_cfu_med) %>% 
+    right_join(community_all_d0) %>% 
+    filter(!human_source %in% c('DA00581', 'DA01134','DA01146')) %>% 
+    select(-mouse_id) %>% 
+    filter(!is.na(log_cfu_med))
+cfu_wo_hilo_df$human_source <- factor(cfu_wo_hilo_df$human_source)
+
+set.seed(seed)
+cfu_wo_hilo_rf <- randomForest(cfu_wo_hilo_df[ , -2], cfu_wo_hilo_df$log_cfu_med,
+                           ntree = 1001,
+                           mtry = 15,
+                           importance = T)
+top_features_wo_hilo <- head(sort(importance(cfu_wo_hilo_rf)[,1],decreasing = T), n_features)
+
+data.frame(Observed = cfu_wo_hilo_df$log_cfu_med, 
+           Predicted = cfu_wo_hilo_rf$predicted) %>% 
+  ggplot(aes(x = Predicted, y = Observed)) + geom_point() +
+    xlim(5,9) + ylim(5,9) + 
+    annotate('text', label = paste0(round(tail(cfu_wo_hilo_rf$rsq, 1), 4)*100, '% Var Explained'),
+             x = 5.5, y = 8.9) + 
+    theme(panel.background = element_rect(linetype = 1, size = 1, fill = 'white', color = 'black'),
+          panel.grid.major.x = element_blank(),
+          panel.grid.minor.x = element_blank(),
+          panel.grid.major.y = element_blank(),
+          panel.grid.minor.y = element_blank()) +
+  labs(x = 'Predicted Colonization (log CFU)', y = 'Observed Colonization (log CFU)')
+```

### 2017-01-05 code/HumanCdGF_cdiff_colonization.Rmd
include IN2/DA01134

@@
   filter(Early_Euth == FALSE, cdiff_strain == 431, !cage_id == 'NP1') %>% 
-  select(mouse_id, sample_id, cage_id, human_source, day, log_cfu) %>% 
-  filter(!is.na(log_cfu))
+  select(mouse_id, sample_id, cage_id, human_source, day, log_cfu)
 shared_list_days <- meta_data %>% 
@@
 shared_file <- shared_file[rownames(shared_file) %in% shared_list_days$sample_id, ]
+meta_data <- filter(meta_data, !is.na(log_cfu))
 
@@
 ```
+
+*** 
+***
+I did notice that the way I was selecting samples (Day == 0) was removing a cage, since there was no CFU recorded for Day 0.
+This cage is higher in CFU and does have an effect on CFU however without any CFU data for day 0 it is not completely clear there was no CDI present at day 0.
+
+***
+***
 
@@
     annotate('text', label = paste0(round(tail(cfu_wo581_rf$rsq, 1), 4)*100, '% Var Explained'),
-             x = min(cfu_wo581_rf$predicted), y = max(cfu_wo581_df$log_cfu_med)) + 
+             x = 6, y = 8.9) + 
     theme(panel.background = element_rect(linetype = 1, size = 1, fill = 'white', color = 'black'),

### 2017-01-05 code/HumanCdGF_cdiff_colonization.Rmd
remove IN2/DA01134 since no Day 0 CFU data

@@
   filter(Early_Euth == FALSE, cdiff_strain == 431, !cage_id == 'NP1') %>% 
-  select(mouse_id, sample_id, cage_id, human_source, day, log_cfu)
+  select(mouse_id, sample_id, cage_id, human_source, day, log_cfu) %>% 
+  filter(!is.na(log_cfu))
 shared_list_days <- meta_data %>% 
@@
 shared_file <- shared_file[rownames(shared_file) %in% shared_list_days$sample_id, ]
-meta_data <- filter(meta_data, !is.na(log_cfu))
 
@@
 ```
-
-*** 
-***
-I did notice that the way I was selecting samples (Day == 0) was removing a cage, since there was no CFU recorded for Day 0.
-This cage is higher in CFU and does have an effect on CFU however without any CFU data for day 0 it is not completely clear there was no CDI present at day 0.
-
-***
-***
 

### 2017-01-05 code/HumanCdGF_cdiff_colonization.Rmd
add baseline/randomized models

@@
 
+baseline_df <- c()
+baseline_rsq <- c()
+for(i in 1:100){
+  set.seed(i)
+  random_cfu <- sample(cfu_all_shared_df$log_cfu_med, 
+                       length(cfu_all_shared_df$log_cfu_med), replace = F)
+  set.seed(seed)
+  baseline_cfu_all_rf <- randomForest(cfu_all_shared_df[ , -1], random_cfu,
+                           ntree = 501,
+                           mtry = 15)
+  baseline_df <- rbind(baseline_df, data.frame(Model = 'Baseline', Observed = random_cfu,
+                                               Predicted = baseline_cfu_all_rf$predicted))
+  baseline_rsq <- c(baseline_rsq, tail(baseline_cfu_all_rf$rsq,1))
+}
 set.seed(seed)
 cfu_all_rf <- randomForest(cfu_all_shared_df[ , -1], cfu_all_shared_df$log_cfu_med,
-                           ntree = 1001,
+                           ntree = 501,
                            mtry = 15,
@@
 
-data.frame(Observed = cfu_all_shared_df$log_cfu_med, 
-           Predicted = cfu_all_rf$predicted) %>% 
-  ggplot(aes(x = Predicted, y = Observed)) + geom_point() +
-    xlim(5,9) + ylim(5,9) + 
+rbind(baseline_df,
+      data.frame(Model = 'Prediction', Observed = cfu_all_shared_df$log_cfu_med, 
+           Predicted = cfu_all_rf$predicted)) %>% 
+  ggplot(aes(x = Predicted, y = Observed, color=Model)) + geom_point() +
+    xlim(5,9) + ylim(5,9) + scale_colour_manual(values = c('#CCCCCC','#000000')) +
     annotate('text', label = paste0(round(tail(cfu_all_rf$rsq, 1), 4)*100, '% Var Explained'),
@@
           panel.grid.major.y = element_blank(),
-          panel.grid.minor.y = element_blank()) +
+          panel.grid.minor.y = element_blank(),
+          legend.position = c(0.9,0.1)) +
   labs(x = 'Predicted Colonization (log CFU)', y = 'Observed Colonization (log CFU)')
@@
 ```
+
+The % Variation explained in the baseline model is:  
+Min.  1st Qu.   Median     Mean  3rd Qu.     Max.   
+`r summary(baseline_rsq)`
 
@@
 
+baseline_source_df <- c()
+for(i in 1:100){
+  set.seed(i)
+  random_src <- sample(source_all_shared_df$human_source, 
+                       length(source_all_shared_df$human_source), replace = F)
+  set.seed(seed)
+  baseline_src_all_rf <- randomForest(source_all_shared_df[ , -1], random_src,
+                           ntree = 501,
+                           mtry = 15)
+  baseline_source_df <- c(baseline_source_df, tail(baseline_src_all_rf$err.rate[,'OOB'],1))
+}
 set.seed(seed)
@@
 ```
+
+Models predicting randomized donor perform with a high error rate:   
+Mean Error Rate = `r summary(baseline_source_df)[4]` with an IQR of `r summary(baseline_source_df)[2]` - `r summary(baseline_source_df)[5]`.
 
@@
 
+baseline_cfu_src_df <- c()
+baseline_src_rsq <- c()
+for(i in 1:100){
+  set.seed(i)
+  random_cfu <- sample(cfu_all_src_shared_df$log_cfu_med, 
+                       length(cfu_all_src_shared_df$log_cfu_med), replace = F)
+  set.seed(seed)
+  baseline_cfu_src_rf <- randomForest(cfu_all_src_shared_df[ , -1], random_cfu,
+                           ntree = 501,
+                           mtry = 15)
+  baseline_cfu_src_df <- rbind(baseline_cfu_src_df, 
+                               data.frame(Model = 'Baseline', Observed = random_cfu,
+                                          Predicted = baseline_cfu_src_rf$predicted))
+  baseline_src_rsq <- c(baseline_src_rsq, tail(baseline_cfu_src_rf$rsq,1))
+
+}
 set.seed(seed)
@@
 
-data.frame(Observed = cfu_all_src_shared_df$log_cfu_med, 
-           Predicted = cfu_all_src_rf$predicted) %>% 
-  ggplot(aes(x = Predicted, y = Observed)) + geom_point() +
-    xlim(5,9) + ylim(5,9) + 
+rbind(baseline_cfu_src_df,
+      data.frame(Model = 'Prediction',
+                 Observed = cfu_all_src_shared_df$log_cfu_med, 
+                 Predicted = cfu_all_src_rf$predicted)) %>% 
+  ggplot(aes(x = Predicted, y = Observed, color = Model)) + geom_point() +
+    xlim(5,9) + ylim(5,9) + scale_colour_manual(values = c('#CCCCCC','#000000')) +
     annotate('text', label = paste0(round(tail(cfu_all_src_rf$rsq, 1), 4)*100, '% Var Explained'),
@@
 ```
+
+The % Variation explained in the baseline model is:  
+Min.  1st Qu.   Median     Mean  3rd Qu.     Max.   
+`r summary(baseline_src_rsq)`
 
@@
 
+baseline_wo581 <- c()
+base_wo581_rsq <- c()
+for(i in 1:100){
+  set.seed(i)
+  random_cfu <- sample(cfu_wo581_df$log_cfu_med, 
+                       length(cfu_wo581_df$log_cfu_med), replace = F)
+  set.seed(seed)
+  baseline_wo581_rf <- randomForest(cfu_wo581_df[ , -1], random_cfu,
+                           ntree = 501,
+                           mtry = 15)
+  baseline_wo581 <- rbind(baseline_wo581, 
+                               data.frame(Model = 'Baseline', Observed = random_cfu,
+                                          Predicted = baseline_wo581_rf$predicted))
+  base_wo581_rsq <- c(base_wo581_rsq, tail(baseline_wo581_rf$rsq,1))
+}
 set.seed(seed)
@@
 ```{r colonization performance without 581, fig.width=5, fig.height=4}
-data.frame(Observed = cfu_wo581_df$log_cfu_med, 
-           Predicted = cfu_wo581_rf$predicted) %>% 
-  ggplot(aes(x = Predicted, y = Observed)) + geom_point() +
-    xlim(5,9) + ylim(5,9) + 
+rbind(baseline_wo581,
+      data.frame(Model = 'Prediction', Observed = cfu_wo581_df$log_cfu_med, 
+           Predicted = cfu_wo581_rf$predicted)) %>% 
+  ggplot(aes(x = Predicted, y = Observed, color = Model)) + geom_point() +
+    xlim(5,9) + ylim(5,9) + scale_colour_manual(values = c('#CCCCCC','#000000')) +
     annotate('text', label = paste0(round(tail(cfu_wo581_rf$rsq, 1), 4)*100, '% Var Explained'),
@@
           panel.grid.major.y = element_blank(),
-          panel.grid.minor.y = element_blank()) +
+          panel.grid.minor.y = element_blank(),
+          legend.position = c(0.8,0.2)) +
   labs(x = 'Predicted Colonization (log CFU)', y = 'Observed Colonization (log CFU)')
@@
 ```
+
+The % Variation explained in the baseline model is:  
+Min.  1st Qu.   Median     Mean  3rd Qu.     Max.   
+`r summary(base_wo581_rsq)`
 
@@
 
-```{r all without 581 or 1134/1146}
+```{r all without 581 or 1134/1146, fig.width=5, fig.height=4}
 cfu_wo_hilo_df <- colonized_cfu %>% 
@@
 
+baseline_wo_hilo <- c()
+base_wo_hilo_rsq <- c()
+for(i in 1:100){
+  set.seed(i)
+  random_cfu <- sample(cfu_wo_hilo_df$log_cfu_med, 
+                       length(cfu_wo_hilo_df$log_cfu_med), replace = F)
+  set.seed(seed)
+  baseline_wo_hilo_rf <- randomForest(cfu_wo_hilo_df[ , -1], random_cfu,
+                           ntree = 501,
+                           mtry = 15)
+  baseline_wo_hilo <- rbind(baseline_wo_hilo, 
+                               data.frame(Model = 'Baseline', Observed = random_cfu,
+                                          Predicted = baseline_wo_hilo_rf$predicted))
+  base_wo_hilo_rsq <- c(base_wo_hilo_rsq, tail(baseline_wo_hilo_rf$rsq,1))
+}
 set.seed(seed)
@@
 
-data.frame(Observed = cfu_wo_hilo_df$log_cfu_med, 
-           Predicted = cfu_wo_hilo_rf$predicted) %>% 
-  ggplot(aes(x = Predicted, y = Observed)) + geom_point() +
-    xlim(5,9) + ylim(5,9) + 
+rbind(baseline_wo_hilo,
+      data.frame(Model = 'Prediction', Observed = cfu_wo_hilo_df$log_cfu_med, 
+           Predicted = cfu_wo_hilo_rf$predicted)) %>% 
+  ggplot(aes(x = Predicted, y = Observed, color = Model)) + geom_point() +
+    xlim(5,9) + ylim(5,9) + scale_colour_manual(values = c('#CCCCCC','#000000')) +
     annotate('text', label = paste0(round(tail(cfu_wo_hilo_rf$rsq, 1), 4)*100, '% Var Explained'),
-             x = 5.5, y = 8.9) + 
+             x = 6, y = 8.9) + 
     theme(panel.background = element_rect(linetype = 1, size = 1, fill = 'white', color = 'black'),
@@
           panel.grid.major.y = element_blank(),
-          panel.grid.minor.y = element_blank()) +
+          panel.grid.minor.y = element_blank(),
+          legend.position = c(0.8,0.2)) +
   labs(x = 'Predicted Colonization (log CFU)', y = 'Observed Colonization (log CFU)')
 ```
+
+The % Variation explained in the baseline model is:  
+Min.  1st Qu.   Median     Mean  3rd Qu.     Max.   
+`r summary(base_wo_hilo_rsq)`

### 2017-01-06 code/HumanCdGF_cdiff_colonization.Rmd
fixed "baseline" to read "randomized"

--- 2017-01-05 code/HumanCdGF_cdiff_colonization.Rmd
@@
 
+randomized_df <- c()
+randomized_rsq <- c()
+for(i in 1:100){
+  set.seed(i)
+  random_cfu <- sample(cfu_all_shared_df$log_cfu_med, 
+                       length(cfu_all_shared_df$log_cfu_med), replace = F)
+  set.seed(seed)
+  randomized_cfu_all_rf <- randomForest(cfu_all_shared_df[ , -1], random_cfu,
+                           ntree = 501,
+                           mtry = 15)
+  randomized_df <- rbind(randomized_df, data.frame(Model = 'Randomized', Observed = random_cfu,
+                                               Predicted = randomized_cfu_all_rf$predicted))
+  randomized_rsq <- c(randomized_rsq, tail(randomized_cfu_all_rf$rsq,1))
+}
 set.seed(seed)
 cfu_all_rf <- randomForest(cfu_all_shared_df[ , -1], cfu_all_shared_df$log_cfu_med,
-                           ntree = 1001,
+                           ntree = 501,
                            mtry = 15,
@@
 
-data.frame(Observed = cfu_all_shared_df$log_cfu_med, 
-           Predicted = cfu_all_rf$predicted) %>% 
-  ggplot(aes(x = Predicted, y = Observed)) + geom_point() +
-    xlim(5,9) + ylim(5,9) + 
+rbind(randomized_df,
+      data.frame(Model = 'Actual', Observed = cfu_all_shared_df$log_cfu_med, 
+           Predicted = cfu_all_rf$predicted)) %>% 
+  ggplot(aes(x = Predicted, y = Observed, color=Model)) + geom_point() +
+    xlim(5,9) + ylim(5,9) + scale_colour_manual(values = c('#CCCCCC','#000000')) +
     annotate('text', label = paste0(round(tail(cfu_all_rf$rsq, 1), 4)*100, '% Var Explained'),
@@
 ```
+
+The % Variation explained in the randomized model is:  
+Min.  1st Qu.   Median     Mean  3rd Qu.     Max.   
+`r summary(randomized_rsq)`
 
@@
 
+randomized_source_df <- c()
+for(i in 1:100){
+  set.seed(i)
+  random_src <- sample(source_all_shared_df$human_source, 
+                       length(source_all_shared_df$human_source), replace = F)
+  set.seed(seed)
+  randomized_src_all_rf <- randomForest(source_all_shared_df[ , -1], random_src,
+                           ntree = 501,
+                           mtry = 15)
+  randomized_source_df <- c(randomized_source_df, tail(randomized_src_all_rf$err.rate[,'OOB'],1))
+}
 set.seed(seed)
@@
 ```
+
+Models predicting randomized donor perform with a high error rate:   
+Mean Error Rate = `r summary(randomized_source_df)[4]` with an IQR of `r summary(randomized_source_df)[2]` - `r summary(randomized_source_df)[5]`.
 
@@
 
+randomized_cfu_src_df <- c()
+randomized_src_rsq <- c()
+for(i in 1:100){
+  set.seed(i)
+  random_cfu <- sample(cfu_all_src_shared_df$log_cfu_med, 
+                       length(cfu_all_src_shared_df$log_cfu_med), replace = F)
+  set.seed(seed)
+  randomized_cfu_src_rf <- randomForest(cfu_all_src_shared_df[ , -1], random_cfu,
+                           ntree = 501,
+                           mtry = 15)
+  randomized_cfu_src_df <- rbind(randomized_cfu_src_df, 
+                               data.frame(Model = 'Randomized', Observed = random_cfu,
+                                          Predicted = randomized_cfu_src_rf$predicted))
+  randomized_src_rsq <- c(randomized_src_rsq, tail(randomized_cfu_src_rf$rsq,1))
+
+}
 set.seed(seed)
@@
 
-data.frame(Observed = cfu_all_src_shared_df$log_cfu_med, 
-           Predicted = cfu_all_src_rf$predicted) %>% 
-  ggplot(aes(x = Predicted, y = Observed)) + geom_point() +
-    xlim(5,9) + ylim(5,9) + 
+rbind(randomized_cfu_src_df,
+      data.frame(Model = 'Actual',
+                 Observed = cfu_all_src_shared_df$log_cfu_med, 
+                 Predicted = cfu_all_src_rf$predicted)) %>% 
+  ggplot(aes(x = Predicted, y = Observed, color = Model)) + geom_point() +
+    xlim(5,9) + ylim(5,9) + scale_colour_manual(values = c('#CCCCCC','#000000')) +
     annotate('text', label = paste0(round(tail(cfu_all_src_rf$rsq, 1), 4)*100, '% Var Explained'),
@@
 ```
+
+The % Variation explained in the randomized model is:  
+Min.  1st Qu.   Median     Mean  3rd Qu.     Max.   
+`r summary(randomized_src_rsq)`
 
@@
 
+randomized_wo581 <- c()
+base_wo581_rsq <- c()
+for(i in 1:100){
+  set.seed(i)
+  random_cfu <- sample(cfu_wo581_df$log_cfu_med, 
+                       length(cfu_wo581_df$log_cfu_med), replace = F)
+  set.seed(seed)
+  randomized_wo581_rf <- randomForest(cfu_wo581_df[ , -1], random_cfu,
+                           ntree = 501,
+                           mtry = 15)
+  randomized_wo581 <- rbind(randomized_wo581, 
+                               data.frame(Model = 'Randomized', Observed = random_cfu,
+                                          Predicted = randomized_wo581_rf$predicted))
+  base_wo581_rsq <- c(base_wo581_rsq, tail(randomized_wo581_rf$rsq,1))
+}
 set.seed(seed)
@@
 ```{r colonization performance without 581, fig.width=5, fig.height=4}
-data.frame(Observed = cfu_wo581_df$log_cfu_med, 
-           Predicted = cfu_wo581_rf$predicted) %>% 
-  ggplot(aes(x = Predicted, y = Observed)) + geom_point() +
-    xlim(5,9) + ylim(5,9) + 
+rbind(randomized_wo581,
+      data.frame(Model = 'Actual', Observed = cfu_wo581_df$log_cfu_med, 
+           Predicted = cfu_wo581_rf$predicted)) %>% 
+  ggplot(aes(x = Predicted, y = Observed, color = Model)) + geom_point() +
+    xlim(5,9) + ylim(5,9) + scale_colour_manual(values = c('#CCCCCC','#000000')) +
     annotate('text', label = paste0(round(tail(cfu_wo581_rf$rsq, 1), 4)*100, '% Var Explained'),
@@
 ```
+
+The % Variation explained in the randomized model is:  
+Min.  1st Qu.   Median     Mean  3rd Qu.     Max.   
+`r summary(base_wo581_rsq)`
 
@@
 
+randomized_wo_hilo <- c()
+base_wo_hilo_rsq <- c()
+for(i in 1:100){
+  set.seed(i)
+  random_cfu <- sample(cfu_wo_hilo_df$log_cfu_med, 
+                       length(cfu_wo_hilo_df$log_cfu_med), replace = F)
+  set.seed(seed)
+  randomized_wo_hilo_rf <- randomForest(cfu_wo_hilo_df[ , -1], random_cfu,
+                           ntree = 501,
+                           mtry = 15)
+  randomized_wo_hilo <- rbind(randomized_wo_hilo, 
+                               data.frame(Model = 'Randomized', Observed = random_cfu,
+                                          Predicted = randomized_wo_hilo_rf$predicted))
+  base_wo_hilo_rsq <- c(base_wo_hilo_rsq, tail(randomized_wo_hilo_rf$rsq,1))
+}
 set.seed(seed)
@@
 
-data.frame(Observed = cfu_wo_hilo_df$log_cfu_med, 
-           Predicted = cfu_wo_hilo_rf$predicted) %>% 
-  ggplot(aes(x = Predicted, y = Observed)) + geom_point() +
-    xlim(5,9) + ylim(5,9) + 
+rbind(randomized_wo_hilo,
+      data.frame(Model = 'Actual', Observed = cfu_wo_hilo_df$log_cfu_med, 
+           Predicted = cfu_wo_hilo_rf$predicted)) %>% 
+  ggplot(aes(x = Predicted, y = Observed, color = Model)) + geom_point() +
+    xlim(5,9) + ylim(5,9) + scale_colour_manual(values = c('#CCCCCC','#000000')) +
     annotate('text', label = paste0(round(tail(cfu_wo_hilo_rf$rsq, 1), 4)*100, '% Var Explained'),
-             x = 5.5, y = 8.9) + 
+             x = 6, y = 8.9) + 
     theme(panel.background = element_rect(linetype = 1, size = 1, fill = 'white', color = 'black'),
@@
           panel.grid.major.y = element_blank(),
-          panel.grid.minor.y = element_blank()) +
+          panel.grid.minor.y = element_blank(),
+          legend.position = c(0.8,0.2)) +
   labs(x = 'Predicted Colonization (log CFU)', y = 'Observed Colonization (log CFU)')
 ```
+
+The % Variation explained in the randomized model is:  
+Min.  1st Qu.   Median     Mean  3rd Qu.     Max.   
+`r summary(base_wo_hilo_rsq)`

### 2017-01-06 submission/Manuscript.tex

@@
 transcriptomics of \emph{C. difficile} and examination of host immune
-response to CDI {[}\textbf{???}{]}. Further, CDI therapeutics such as
-antibiotics and fecal microbiota transplants have been tested
-extensively in a gnotobiotic-piglet or piglet-to-gnotobiotic-mouse model
-of disease {[}13,14{]}. Pigs have a longer digestive tract with
-components more similar to humans than mice and are typically infected
-by strains typical in human infection {[}15{]}. However, the murine and
-porcine microbiomes typically do not resemble those of the human gut.
+response to CDI {[}13{]}. Further, CDI therapeutics such as antibiotics
+and fecal microbiota transplants have been tested extensively in a
+gnotobiotic-piglet or piglet-to-gnotobiotic-mouse model of disease
+{[}14,15{]}. Pigs have a longer digestive tract with components more
+similar to humans than mice and are typically infected by strains
+typical in human infection {[}16{]}. However, the murine and porcine
+microbiomes typically do not resemble those of the human gut.
 
@@
 piglets succumbed to infection, while some of the untreated
-human-colonized pigs did {[}13{]}. In another study, germ-free mice
+human-colonized pigs did {[}14{]}. In another study, germ-free mice
 colonized with human feces were bred over several generations to create
-a cohort of mice with identical human-derived microbiomes {[}16{]}.
+a cohort of mice with identical human-derived microbiomes {[}17{]}.
 These mice were subsequently treated with a five-antibiotic cocktail to
 induce dysbiosis and then were successfully colonized by \emph{C.
-difficile} {[}16{]}. While informative, these studies were limited in
+difficile} {[}17{]}. While informative, these studies were limited in
 their use of only one human donor as input inoculum.
@@
 AUC-RF algorithm to generate a Reciever Operating Curve (ROC) with
-maximal Area Under the Curve (AUC) {[}17{]}, as used previously in our
-group to classify colon cancer outcomes {[}18{]}. The optimized model
+maximal Area Under the Curve (AUC) {[}18{]}, as used previously in our
+group to classify colon cancer outcomes {[}19{]}. The optimized model
 used X number of OTUs and identified several that predicted increased
@@
 general healthy humans without gut perturbations are resistant to
-\emph{C. difficile} infection {[}Heinlen2010{]}. On the other hand, it
-is known that humans, especially health care employees, can be
-asymptomatic carriers of \emph{C. difficile} {[}19{]}. This may explain
-the fact that the majority of our groups of mice were colonized by
-\emph{C. difficile} but did not display outward signs of illness or
-infection-associated weight loss (Fig 2).
+\emph{C. difficile} infection {[}20{]}. On the other hand, it is known
+that humans, especially health care employees, can be asymptomatic
+carriers of \emph{C. difficile} {[}21{]}. This may explain the fact that
+the majority of our groups of mice were colonized by \emph{C. difficile}
+but did not display outward signs of illness or infection-associated
+weight loss (Fig 2).
 
@@
 
-Further experiments will test if infecting susceptible communities with
-weak strains allows the mice to survive. It has been recently reported
-that carriage of a non-toxigenic strain of \emph{C. difficile} can
-prevent additional colonization or recurrence by a pathogenic strain,
-suggesting a role of niche-filling or immunity this pre-colonization
-provides {[}20{]}. One interesting experiment would be to recapitulate
-this finding in human-associated mouse microbiomes by making a
-susceptible community resistant to infection. Further work with
+It has been recently reported that carriage of a non-toxigenic strain of
+\emph{C. difficile} can prevent additional colonization or recurrence by
+a pathogenic strain, suggesting a role of niche-filling or immunity this
+pre-colonization provides {[}22{]}. One interesting experiment would be
+to recapitulate this finding in human-associated mouse microbiomes by
+making a susceptible community resistant to infection. Further work with
 human-associated communities in mice and mathematical modeling will
@@
 bank collected from patients from October 2010 to November 2012 at the
-University of Michigan. Sample collection methods and study was approved
-by the Institutional Review Board at the University of Michigan. Stool
-samples from University of Michigan Health System patients were
-submitted to the microbiology lab for \emph{C. difficile} testing.
-\emph{C. difficile} presence was measured using C.Diff Quik Chek
-Complete Assay (Techlab). Samples from these patients were stored in
-Cary-Blair medium at -80C. All patients consented to participation in
-the study. Donor stool for this study was selected from one patient that
-tested positive for \emph{C. difficile} infection and five patients that
-presented with diarrhea and tested negative for \emph{C. difficile}.
-Healthy donor stool was collected from adult volunteers living in the
-Ann Arbor, MI area. 100ul of stool slurry was gavaged into each mouse.
+University of Michigan Health System. Sample collection methods and the
+study were approved by the Institutional Review Board at the University
+of Michigan. Stool samples from University of Michigan Health System
+patients were submitted to the microbiology lab for \emph{C. difficile}
+testing. \emph{C. difficile} presence was measured using C.Diff Quik
+Chek Complete Assay (Techlab). Fecal samples from these patients were
+stored in Cary-Blair medium at -80C. All patients consented to
+participation in the study. Donor stool was selected from one patient
+that tested positive for \emph{C. difficile} infection and five patients
+that presented with diarrhea and tested negative for \emph{C.
+difficile}. Healthy donor stool was collected from adult volunteers
+living in the Ann Arbor, MI area. 100ul of stool slurry was gavaged into
+each mouse.
 
@@
 strain 431 is an 027 ribotype clinical isolate isolated from a patient
-with severe CDI {[}21{]}. After gavage, the remaining inoculum was
+with severe CDI {[}23{]}. After gavage, the remaining inoculum was
 diluted and plated to confirm infection dose. Bacteria were cultured
-anaerobically on TCCFA plates and incubated at 37C {[}22{]}. During
+anaerobically on TCCFA plates and incubated at 37C {[}24{]}. During
 infection, two fresh stool pellets were collected from each mouse daily.
@@
 amplification of the V4 region of the 16S rRNA gene and fragments were
-sequenced on an Illumina MiSeq as previously described {[}23{]}.
+sequenced on an Illumina MiSeq as previously described {[}25{]}.
 
@@
 
-13. Kim HB, Zhang Q, Sun X, Beamer G, Wang Y, Tzipori S. Beneficial
+13. Pawlowski SW, Calabrese G, Kolling GL, Freire R, AlcantaraWarren C,
+Liu B, et al. Murine model of clostridium difficile infection with aged
+gnotobiotic c57BL/6 mice and a BI/NAP1 strain. The Journal of Infectious
+Diseases. Oxford University Press (OUP); 2010;202: 1708--1712.
+doi:\href{http://dx.doi.org/10.1086/657086}{10.1086/657086}
+
+14. Kim HB, Zhang Q, Sun X, Beamer G, Wang Y, Tzipori S. Beneficial
 effect of oral tigecycline treatment on clostridium difficile infection
@@
 
-14. Diao H, Yan HL, Xiao Y, Yu B, Yu J, He J, et al. Intestinal
+15. Diao H, Yan HL, Xiao Y, Yu B, Yu J, He J, et al. Intestinal
 microbiota could transfer host gut characteristics from pigs to mice.
@@
 
-15. Steele J, Feng H, Parry N, Tzipori S. Piglet models of acute or
+16. Steele J, Feng H, Parry N, Tzipori S. Piglet models of acute or
 chronic clostridium difficile illness. The Journal of Infectious
@@
 
-16. Collins J, Auchtung JM, Schaefer L, Eaton KA, Britton RA. Humanized
+17. Collins J, Auchtung JM, Schaefer L, Eaton KA, Britton RA. Humanized
 microbiota mice as a model of recurrent clostridium difficile disease.
@@
 
-17. Calle ML, Urrea V, Boulesteix A-L, Malats N. AUC-RF: A new strategy
+18. Calle ML, Urrea V, Boulesteix A-L, Malats N. AUC-RF: A new strategy
 for genomic profiling with random forest. Human Heredity. S. Karger AG;
@@
 
-18. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
+19. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
 improves the sensitivity of fecal immunochemical test for detecting
@@
 
-19. Furuya-Kanamori L, Marquess J, Yakob L, Riley TV, Paterson DL,
+20. Heinlen L, Ballard JD. Clostridium difficile infection. The American
+Journal of the Medical Sciences. Elsevier BV; 2010;340: 247--252.
+doi:\href{http://dx.doi.org/10.1097/maj.0b013e3181e939d8}{10.1097/maj.0b013e3181e939d8}
+
+21. Furuya-Kanamori L, Marquess J, Yakob L, Riley TV, Paterson DL,
 Foster NF, et al. Asymptomatic clostridium difficile colonization:
@@
 
-20. Gerding DN, Meyer T, Lee C, Cohen SH, Murthy UK, Poirier A, et al.
+22. Gerding DN, Meyer T, Lee C, Cohen SH, Murthy UK, Poirier A, et al.
 Administration of spores of nontoxigenic clostridium difficile strain m3
@@
 
-21. Carlson PE, Walk ST, Bourgis AE, Liu MW, Kopliku F, Lo E, et al. The
+23. Carlson PE, Walk ST, Bourgis AE, Liu MW, Kopliku F, Lo E, et al. The
 relationship between phenotype, ribotype, and clinical disease in human
@@
 
-22. Bliss DZ, Johnson S, Clabots CR, Savik K, Gerding DN. Comparison of
+24. Bliss DZ, Johnson S, Clabots CR, Savik K, Gerding DN. Comparison of
 cycloserine-cefoxitin-fructose agar (CCFA) and taurocholate-CCFA for
@@
 
-23. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
+25. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD.
 Development of a dual-index sequencing strategy and curation pipeline

### 2017-01-06 code/HumanCdGF_cdiff_colonization.Rmd
nmds on day 0 comm for colonization level

@@
 `r summary(base_wo_hilo_rsq)`
+
+***  
+***
+
+### Regression cannot be used to predict colonization levels in this dataset
+
+Since the % Variation explained by the RF model decreases to ~0% when removing the extreme values, the suggests that we do not have the data to make accurate predictions of the colonization level due to the low number of samples at the extremes.
+However we do seem to have a separation between the high and low but we cannot model the middle levels with linear regression.
+So another way we could look at this would to classify the groups.
+
+##### NMDS of colonization level
+
+```{r NMDS, eval = F}
+#export shared file to generate theta-yc distances in mothur
+#read in full shared file
+full_shared_file <- 'data/process/full_shared.shared'
+full_shared_file <- read.table(full_shared_file, sep = '\t', header = T)
+
+nmds_df <- colonized_cfu %>% 
+    select(mouse_id, human_source, log_cfu_med) %>% 
+    right_join(shared_day_0) %>% 
+    filter(!is.na(log_cfu_med))
+nmds_df <- data.frame(colonization_level= ifelse(nmds_df$log_cfu_med >= 8.25, 'hi',
+                                     ifelse(nmds_df$log_cfu_med <= 6.5, 'lo', 'mid')),
+                       select(nmds_df, -log_cfu_med))
+nmds_shared <- full_shared_file[full_shared_file$Group %in% 
+                                  paste0(gsub('_','-',nmds_df$mouse_id),'-D0'),]
+write.table(nmds_shared, file= '~/Desktop/day_0_communities.shared', sep = '\t', 
+            row.names = F, quote = F)
+nmds_CL_design <- data.frame(paste0(gsub('_','-',nmds_df$mouse_id),'-D0'), nmds_df$colonization_level)
+write.table(nmds_CL_design, file= '~/Desktop/day_0_communities.CL.design', sep = '\t', 
+      col.names = F, row.names = F, quote = F)
+nmds_src_design <- data.frame(paste0(gsub('_','-',nmds_df$mouse_id),'-D0'), nmds_df$human_source)
+write.table(nmds_src_design, file= '~/Desktop/day_0_communities.SRC.design', sep = '\t', 
+      col.names = F, row.names = F, quote = F)
+#min(apply(nmds_shared[,-c(1:3)],1,sum)) # 2423
+#run dist command in mothur
+#dist.shared(shared=~/Desktop/day_0_communities.shared, calc=thetayc, subsample=2400)
+#nmds(phylip=~/Desktop/day_0_communities.thetayc.0.03.lt.ave.dist)
+Number of dimensions:	2
+Lowest stress :	0.297773
+R-squared for configuration:	0.55791
+#nmds(phylip=~/Desktop/day_0_communities.thetayc.0.03.lt.ave.dist, mindim=3, maxdim=3)
+Number of dimensions:	3
+Lowest stress :	0.191484
+R-squared for configuration:	0.71623
+#amova(phylip=~/Desktop/day_0_communities.thetayc.0.03.lt.ave.dist, design=~/Desktop/day_0_communities.CL.design)
+hi-lo-mid	Among	Within	Total
+SS	4.48543	8.08716	12.5726
+df	2	30	32
+MS	2.24272	0.269572
+
+Fs:	8.31955
+p-value: <0.001*
+
+hi-lo	Among	Within	Total
+SS	2.4372	0.676826	3.11403
+df	1	10	11
+MS	2.4372	0.0676826
+
+Fs:	36.0093
+p-value: <0.001*
+
+hi-mid	Among	Within	Total
+SS	2.08215	7.41077	9.49293
+df	1	23	24
+MS	2.08215	0.322207
+
+Fs:	6.46215
+p-value: <0.001*
+
+lo-mid	Among	Within	Total
+SS	2.30066	8.08673	10.3874
+df	1	27	28
+MS	2.30066	0.299508
+
+Fs:	7.68145
+p-value: <0.001*
+#significantly different centroids
+
+#homova(phylip=~/Desktop/day_0_communities.thetayc.0.03.lt.ave.dist, design=~/Desktop/day_0_communities.CL.design)
+HOMOVA	BValue	P-value	SSwithin/(Ni-1)_values
+hi-lo-mid	21.6225	<0.001*	0.000145266	0.0966271	0.370517
+hi-lo	14.1638	<0.001*	0.000145266	0.0966271
+hi-mid	18.2515	<0.001*	0.000145266	0.370517
+lo-mid	3.48295	<0.001*	0.0966271	0.370517
+# all groups have significantly different variation (hi = 0.0001, mid = 0.37, lo = 0.097)
+
+dist_cl <- vegdist(nmds_df[,-c(1,2)])
+nmds <- metaMDS(nmds_df[,-c(1:2)])
+stressplot(nmds, dist_cl)
+plot(nmds, type = 'n')
+points(nmds, display = 'sites', pch = 21, bg = as.numeric(as.factor(nmds_df$human_source)))
+#text(nmds, display = 'species', cex = 0.5)
+envfit(nmds ~ )
+```
+
+#### Classification of Colonization Level
+
+```{r classify colonization level}
+
+```
+
+#### Features driving classification
+
+```{r features of col level}
+
+```
+
+While these features seem to drive the grouping, it still needs to be further validated/investigated to see if these results repeat in similar microbiomes since the high and low are largely homogenous communities and the middle group is made of heterogenous commnunities.

### 2017-01-09 code/HumanCdGF_cdiff_colonization.Rmd
pull dist generated dist in mothur into R

@@
 # all groups have significantly different variation (hi = 0.0001, mid = 0.37, lo = 0.097)
-
-dist_cl <- vegdist(nmds_df[,-c(1,2)])
-nmds <- metaMDS(nmds_df[,-c(1:2)])
-stressplot(nmds, dist_cl)
+```
+
+```{r NMDS of Colonization Level}
+
+full_shared_file <- 'data/process/full_shared.shared'
+full_shared_file <- read.table(full_shared_file, sep = '\t', header = T)
+
+dist_file <- 'data/process/nmds_colonization_level/day_0_communities.thetayc.0.03.lt.ave.dist'
+dist_day_0_comm <- scan(dist_file, what = 'character') # load dist file as vector
+n_samples_dist <- as.numeric(dist_day_0_comm[1]) # save sample number and remove from dist vector
+dist_day_0_comm <- dist_day_0_comm[-1]
+dist_matrix <- data.frame(matrix(NA, nrow = n_samples_dist, ncol = n_samples_dist)) # matrix size of samples
+diag_value <- 0
+diag(dist_matrix) <- diag_value # fills in diagonal
+c <- 1 # c keep track of position in dist vector
+for(i in 1:n_samples_dist){
+  sample <- dist_day_0_comm[c]
+  colnames(dist_matrix)[i] <- sample
+  rownames(dist_matrix)[i] <- sample
+  if(i > 1){
+    dist_matrix[i,1:(i-1)] <- dist_day_0_comm[(c+1):(c+i-1)] # fill in matrix
+  }
+  c <- c+i # maintain position in dist vector
+}
+dist_matrix[upper.tri(dist_matrix)] <- t(dist_matrix)[upper.tri(dist_matrix)] #fill in upper triangle
+
+nmds <- metaMDS(dist_matrix)
+stressplot(nmds)
 plot(nmds, type = 'n')
-points(nmds, display = 'sites', pch = 21, bg = as.numeric(as.factor(nmds_df$human_source)))
+
+#points(nmds, display = 'sites', pch = 21, bg = as.numeric(as.factor(nmds_df$human_source)))
 #text(nmds, display = 'species', cex = 0.5)
-envfit(nmds ~ )
+#envfit(nmds ~ )
 ```

### 2017-01-09 code/HumanCdGF_cdiff_colonization.Rmd

@@
   set.seed(seed)
-  randomized_cfu_src_rf <- randomForest(cfu_all_src_shared_df[ , -1], random_cfu,
+  randomized_cfu_src_rf <- randomForest(cfu_all_src_shared_df[ , -c(1,2)], random_cfu,
                            ntree = 501,
@@
   set.seed(seed)
-  randomized_wo581_rf <- randomForest(cfu_wo581_df[ , -1], random_cfu,
+  randomized_wo581_rf <- randomForest(cfu_wo581_df[ , -c(1,2)], random_cfu,
                            ntree = 501,
@@
   set.seed(seed)
-  randomized_wo_hilo_rf <- randomForest(cfu_wo_hilo_df[ , -1], random_cfu,
+  randomized_wo_hilo_rf <- randomForest(cfu_wo_hilo_df[ , -c(1,2)], random_cfu,
                            ntree = 501,

### 2017-01-09 nicks model report.Rmd
added histo code and nicks rmd

--- 2016-11-23 code/CdGF_AUCRF.Rmd
@@
+---
+title: "nick report"
+output:
+  pdf_document: default
+  html_document: default
+---
 ---
@@
 
-meta_file   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
-shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
-tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/gf_new.an.taxonomy'
-tax_function <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R'
+meta_file   <- 'data/process/human_CdGF_metadata.txt'
+shared_file <- 'data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
+tax_file <- 'data/process/gf_new.an.taxonomy'
+tax_function <- 'code/tax_level.R'
 

### 2017-01-09 histo_scores.R

@@ -6,12 +6,17 @@
 #
 #
 # Output: 
-
+library('tidyr')
+library('ggplot2')
 # load files and trim
 meta_file   <- 'data/process/human_CdGF_metadata.txt'
 meta_file   <- read.table(meta_file, sep = '\t', header = T, row.names = 'sample_id')
 histo_file <- 'data/process/histology_scores_txt.txt'
 histo_file <- read.table(histo_file, sep = '\t', header = T)
+
+#make mouse_id column to make easier for merging later
+histo_file <- unite_(histo_file, "mouse_id", c("cage_ID", "mouse_ID"), sep="_", remove=FALSE)
+histo_full <- merge(meta_file, histo_file, by ="mouse_id")
 
 #plot of summary score by donor
 ggplot(histo_file, aes(human_source, summary_score)) + geom_point(aes(fill=outcome, color=outcome), size=2) +ggtitle("histology summary score")

### 2017-01-11 code/CdGF_AUCRF.Rmd
intialize calibration plot code

--- 2016-11-23 code/CdGF_AUCRF.Rmd
@@
 author: "Nicholas Lesniak"
-date: "September 15, 2016"
+date: "January 10, 2016"
 output: html_document
@@
 otu_euth_probs <- predict(rf_early_aucrf$RFopt, type = 'prob')
+
+xyplot(calibration(Predict_early_euth_df$Euth_Early~otu_euth_probs[,1]))
+
+cal_plots <- function(observed_col, prediction_df){
+  histo_plot <- ggplot(data.frame(prediction_df), aes(x=X1)) + 
+    stat_bin(bins = 20, aes(y = ..count../sum(..count..))) +
+    scale_x_continuous(breaks=c(seq(0,1,0.2))) +
+    labs(x = 'Predicted Value', y = 'Fraction of Samples') + 
+    theme_bw()
+  cal_plot <- data.frame(observed = observed_col, 
+                         predicted = prediction_df[,2],
+                         bin = cut(prediction_df[,2], 
+                                   breaks =c(-Inf,seq(0.1,0.9,0.1), Inf),
+                                   labels = seq(0.05,0.95,0.1))) %>% 
+    group_by(bin) %>% 
+    summarise(positive_portion = sum(observed==1)/length(observed)) %>% 
+    mutate(bin = as.numeric(as.character(bin))) %>% 
+    ggplot(aes(x=bin, y=positive_portion)) + geom_point() +
+      labs(x='Predicted Value', y='Fraction of Positives') +
+      geom_abline(color='gray', intercept = 0, slope = 1) + 
+      scale_x_continuous(breaks=c(seq(0,1,0.2)), limits=c(0,1)) +
+      scale_y_continuous(breaks=c(seq(0,1,0.2))) +
+      theme_bw()
+  return(list(histo_plot, cal_plot))
+}
+test <- cal_plots(observed_col = Predict_early_euth_df$Euth_Early,
+          prediction_df = otu_euth_probs)
+test[[2]]
+
 otu_euth_roc <- roc(Predict_early_euth_df$Euth_Early ~ otu_euth_probs[ , 2])

### 2017-01-11 code/CdGF_AUCRF.Rmd
fix object labels for leave one source out

--- 2016-11-23 code/CdGF_AUCRF.Rmd
@@
 otu_euth_probs <- predict(rf_early_aucrf$RFopt, type = 'prob')
+
+all_mice_probs <- data.frame(obs = Predict_early_euth_df$Euth_Early,
+                             pred = otu_euth_probs[,2])
+
+calibration_plots <- function(prediction_df){
+  # prediction_df must have obs in col 1 and predict in col 2
+  histo_plot <- ggplot(data.frame(prediction_df), aes(x=pred)) + 
+    stat_bin(bins = 20, aes(y = ..count../sum(..count..))) +
+    scale_x_continuous(breaks=c(seq(0,1,0.2))) +
+    labs(x = 'Predicted Value', y = 'Fraction of Samples') + 
+    theme_bw()
+  cal_plot <- data.frame(prediction_df,
+                         bin = cut(prediction_df[,2], 
+                                   breaks =c(-Inf,seq(0.1,0.9,0.1), Inf),
+                                   labels = seq(0.05,0.95,0.1))) %>% 
+    group_by(bin) %>% 
+    summarise(positive_portion = sum(obs==1)/length(obs)) %>% 
+    mutate(bin = as.numeric(as.character(bin))) %>% 
+    ggplot(aes(x=bin, y=positive_portion)) + geom_point() +
+      labs(x='Predicted Value', y='Fraction of Positives') +
+      geom_abline(color='gray', intercept = 0, slope = 1) + 
+      scale_x_continuous(breaks=c(seq(0,1,0.2)), limits=c(0,1)) +
+      scale_y_continuous(breaks=c(seq(0,1,0.2)), limits=c(0,1)) +
+      theme_bw()
+  return(list(histo_plot, cal_plot))
+}
+all_mice_cal <- calibration_plots(all_mice_probs)
+
 otu_euth_roc <- roc(Predict_early_euth_df$Euth_Early ~ otu_euth_probs[ , 2])
@@
 }
+om_cal <- calibration_plots(om_probs)
 om_roc <- roc(om_all$obs~om_all$pred)
@@
 #### leave-one-source-out
-LOcO_df <- Predict_df %>% 
+LOsO_df <- Predict_df %>% 
                 select(Euth_Early, human_source, one_of(om_import_otus))
-source_list <- as.character(unique(LOcO_df$human_source))
+source_list <- as.character(unique(LOsO_df$human_source))
 
-LOcO_probs <- c()
-LOcO_rf <- c()
+LOsO_probs <- c()
+LOsO_rf <- c()
 for(i in 1:length(source_list)){
-  source <- LOcO_df$human_source == source_list[i]
-  LOcO_predict <- select(LOcO_df, -human_source)
-  train <- LOcO_predict[!source,]
-  test <- LOcO_predict[source,]
+  source <- LOsO_df$human_source == source_list[i]
+  LOsO_predict <- select(LOsO_df, -human_source)
+  train <- LOsO_predict[!source,]
+  test <- LOsO_predict[source,]
   set.seed(seed)
   temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
-  LOcO_rf[[i]] <- temp_model$RFopt
-  LOcO_probs[LOcO_df$human_source == source_list[i]] <- predict(temp_model$RFopt, test, type='prob')[,2]
+  LOsO_rf[[i]] <- temp_model$RFopt
+  LOsO_probs[LOsO_df$human_source == source_list[i]] <- predict(temp_model$RFopt, test, type='prob')[,2]
 }
-LOcO_roc <- roc(LOcO_df$Euth_Early~LOcO_probs)
+LOsO_roc <- roc(LOsO_df$Euth_Early~LOsO_probs)
 
 #### leave-2-out
-LOdO_df <- Predict_df %>% 
+LTsO_df <- Predict_df %>% 
                 select(Euth_Early, human_source, one_of(om_import_otus))
-donor_list <- as.character(unique(LOdO_df$human_source))
+donor_list <- as.character(unique(LTsO_df$human_source))
 
-LOdO_probs <- c()
-LOdO_rf <- c()
+LTsO_probs <- c()
+LTsO_rf <- c()
 for(i in seq(1,length(donor_list),2)){
@@
   donors <- sample(donor_list)
-  by_donor <- LOdO_df$human_source %in% donors[i:(i+1)]
-  LOdO_predict <- select(LOdO_df, -human_source)
-  train <- LOdO_predict[!by_donor,]
-  test <- LOdO_predict[by_donor,]
+  by_donor <- LTsO_df$human_source %in% donors[i:(i+1)]
+  LTsO_predict <- select(LTsO_df, -human_source)
+  train <- LTsO_predict[!by_donor,]
+  test <- LTsO_predict[by_donor,]
   set.seed(seed)
   temp_model <- AUCRF(Euth_Early~., data=train, pdel=0.99, ntree=500)
-  LOdO_rf[[i]] <- temp_model$RFopt
-  LOdO_probs[LOdO_df$human_source %in% donors[i:(i+1)]] <- predict(temp_model$RFopt, test, type='prob')[,2]
+  LTsO_rf[[i]] <- temp_model$RFopt
+  LTsO_probs[LTsO_df$human_source %in% donors[i:(i+1)]] <- predict(temp_model$RFopt, test, type='prob')[,2]
 }
-LOdO_roc <- roc(LOdO_df$Euth_Early~LOdO_probs)
+LTsO_roc <- roc(LTsO_df$Euth_Early~LTsO_probs)
 
@@
 plot(om_roc, col='black', lwd=2, add=T, lty=1)
-plot(LOcO_roc, col='purple', lwd=2, add=T, lty=3)
-plot(LOdO_roc, col='red', lwd=2, add=T, lty=3)
+plot(LOsO_roc, col='purple', lwd=2, add=T, lty=3)
+plot(LTsO_roc, col='red', lwd=2, add=T, lty=3)
 plot(om_loo_roc, col='blue', lwd=2, add=T, lty=2)
@@
 legend('bottomright', legend=c(sprintf('One-mouse (AUC: %.3g)',om_roc$auc),
-                           sprintf('Leave-1 source-out (AUC: %.3g)',LOcO_roc$auc),
-                           sprintf('Leave-2 sources-out (AUC: %.3g)',LOdO_roc$auc),
+                           sprintf('Leave-1 source-out (AUC: %.3g)',LOsO_roc$auc),
+                           sprintf('Leave-2 sources-out (AUC: %.3g)',LTsO_roc$auc),
                            sprintf('One-mouse-Leave-1 source-out (AUC: %.3g)',om_loo_roc$auc)),

### 2017-01-11 code/CdGF_AUCRF.Rmd
add cal plot for each model

@@
 LOsO_roc <- roc(LOsO_df$Euth_Early~LOsO_probs)
-
+LOsO_cal <- calibration_plots(data.frame(obs=LOsO_df$Euth_Early,
+                                         pred=LOsO_probs))
 #### leave-2-out
@@
 LTsO_roc <- roc(LTsO_df$Euth_Early~LTsO_probs)
+LTsO_cal <- calibration_plots(data.frame(obs=LTsO_df$Euth_Early,
+                                         pred=LTsO_probs))
 
@@
 om_loo_roc <- roc(om_loo_all$obs~om_loo_all$pred) 
+om_loo_cal <- calibration_plots(om_loo_all)
 

### 2017-01-11 code/HumanCdGF_cdiff_colonization.Rmd
add IQR to mid-range col level samples

@@
 ```{r setup environment}
-pack_used <- c('randomForest','tidyr','ggplot2','plyr', 'dplyr', 'caret')
+pack_used <- c('randomForest','tidyr','ggplot2','plyr', 'dplyr', 'caret', 'vegan')
 for (dep in pack_used){
@@
   summarise(log_cfu_med = median(log_cfu)) %>% 
+  ungroup()
+
+colonized_cfu_range <- meta_data %>% 
+  filter(day %in% c(4:10)) %>% 
+  group_by(mouse_id, cage_id, human_source) %>% 
+  summarise(log_cfu_med = median(log_cfu),
+            low_iqr = quantile(log_cfu)['25%'],
+            high_iqr = quantile(log_cfu)['75%']) %>% 
   ungroup()
@@
   set.seed(seed)
-  randomized_wo_hilo_rf <- randomForest(cfu_wo_hilo_df[ , -c(1,2)], random_cfu,
-                           ntree = 501,
-                           mtry = 15)
+  randomized_wo_hilo_rf <- randomForest(select(cfu_wo_hilo_df, contains('Otu')), 
+                                        random_cfu, ntree = 501, mtry = 15)
   randomized_wo_hilo <-   rbind(randomized_wo_hilo, 
@@
 set.seed(seed)
-cfu_wo_hilo_rf <- randomForest(cfu_wo_hilo_df[ , -2], cfu_wo_hilo_df$log_cfu_med,
-                           ntree = 1001,
-                           mtry = 15,
-                           importance = T)
+cfu_wo_hilo_rf <- randomForest(select(cfu_wo_hilo_df, contains('Otu')), 
+                               cfu_wo_hilo_df$log_cfu_med, ntree = 1001,
+                               mtry = 15, importance = T)
 top_features_wo_hilo <- head(sort(importance(cfu_wo_hilo_rf)[,1],decreasing = T), n_features)
@@
 So another way we could look at this would to classify the groups.
+
+Furthermore, the variation in the steady state CFU seems to have a high degree of overlap in this middle grouping of colonization level, thus it does not appear possible to precisely define a CFU level to predict relative to the others.
+
+```{r IQR of CFU in middle range, fig.height=4, fig.width=5}
+meta_data %>% 
+  filter(day %in% c(4:10)) %>% 
+  group_by(mouse_id, cage_id, human_source) %>% 
+  merge(data.frame(colonized_cfu %>% 
+    select(mouse_id, human_source, log_cfu_med) %>% 
+    right_join(community_all_d0) %>% 
+    filter(!human_source %in% c('DA00581', 'DA01134','DA01146')) %>% 
+    filter(!is.na(log_cfu_med)) %>% 
+      select(mouse_id), 
+                   predicted = cfu_wo_hilo_rf$predicted)) %>% 
+  ggplot(aes(x=predicted, y=log_cfu, group=mouse_id))+
+     stat_summary(fun.data = 'median_hilow', fun.args=(conf.int=0.5)) + 
+     xlim(7,8) + ylim(5,9) +
+     labs(x= "Predicted CFU (log10)", y='Observed CFU (log10)\nMedian with IQR of CFU from day 4-10') +
+     theme(panel.background = element_rect(linetype = 1, size = 1, fill = 'white', color = 'black'),
+          panel.grid.major.x = element_blank(),
+          panel.grid.minor.x = element_blank(),
+          panel.grid.major.y = element_blank(),
+          panel.grid.minor.y = element_blank())
+```
 
@@
 
-```{r NMDS of Colonization Level}
-
+```{r NMDS of Colonization Level, eval=F}
+setwd('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/')
 full_shared_file <- 'data/process/full_shared.shared'

### 2017-01-23 code/HumanCdGF_cdiff_colonization.Rmd

@@
 
-```{r NMDS of Colonization Level, eval=F}
+```{r NMDS of Colonization Level, fig.height=6, fig.width=6}
 setwd('~/Documents/Github/Schubert_humanCdGF_XXXX_2016/')
@@
 
+nmds_group <- colonized_cfu %>% 
+    select(mouse_id, human_source, log_cfu_med) %>% 
+    right_join(community_all_d0) %>% 
+    filter(!is.na(log_cfu_med)) %>% 
+    mutate(colonization_level= ifelse(log_cfu_med >= 8.25, 'hi',
+                                     ifelse(log_cfu_med <= 6.5, 'lo', 'mid'))) %>% 
+    select(colonization_level, mouse_id)
+
+rownames(dist_matrix) <- gsub('_D0','',rownames(dist_matrix))
+colnames(dist_matrix) <- gsub('_D0','',colnames(dist_matrix))
+
 nmds <- metaMDS(dist_matrix)
-stressplot(nmds)
+#stressplot(nmds)
+colors <- c('red','green','orange')
+
 plot(nmds, type = 'n')
-
-#points(nmds, display = 'sites', pch = 21, bg = as.numeric(as.factor(nmds_df$human_source)))
-#text(nmds, display = 'species', cex = 0.5)
-#envfit(nmds ~ )
-```
+points(nmds, display = 'sites', pch = 21, bg = colors[as.numeric(factor(nmds_group$colonization_level))])
+legend('topright', c('High (> 8.25 Log10 CFU)','Middle (8.25-6.5 Log10 CFU)','Low (< 6.5 Log10 CFU)'), 
+       pch =21, pt.bg=c('red','orange','green'), inset=0.09, box.lty = 0)
+#adonis(dist_matrix~colonization_level, data = nmds_group)
+```
+
+Stress = `r nmds$stress`
+
+##### AMOVA
+hi-lo-mid	Among	  Within	Total
+SS	      4.48543	8.08716	12.5726
+df	      2	      30	    32
+MS	      2.24272	0.269572
+
+Fs:	8.31955
+p-value: <0.001*
+
+hi-lo	Among	  Within	  Total
+SS	  2.4372	0.676826	3.11403
+df	  1	      10	      11
+MS	  2.4372	0.0676826
+
+Fs:	36.0093
+p-value: <0.001*
+
+hi-mid	Among	  Within	Total
+SS	    2.08215 7.41077	9.49293
+df	    1	      23	    24
+MS	    2.08215	0.322207
+
+Fs:	6.46215
+p-value: <0.001*
+
+lo-mid	Among	    Within	Total
+SS	    2.30066	  8.08673	10.3874
+df	    1	        27	    28
+MS	    2.30066	  0.299508
+
+Fs:	7.68145
+p-value: <0.001*
+###### significantly different centroids
+
+#### HOMOVA
+HOMOVA	  BValue  	P-value	  SSwithin/(Ni-1)_values
+hi-lo-mid	21.6225	  <0.001*	  0.000145266	0.0966271	0.370517
+hi-lo	    14.1638	  <0.001*	  0.000145266	0.0966271
+hi-mid	  18.2515 	<0.001*	  0.000145266	0.370517
+lo-mid	  3.48295 	<0.001*	  0.0966271	0.370517
 

### 2017-02-01 code/CdGF_disease_severity.Rmd

@@
 author: "Nicholas Lesniak"
-date: "October 27, 2016"
+date: "January 11, 2017"
 output: html_document
@@
 
-```{r setup, include=FALSE}
+```{r process data, include=FALSE}
 meta_file   <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF_metadata.txt'
 shared_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/human_CdGF.an.unique_list.0.03.subsample.shared'
-tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/mothur/gf_all.an.cons.taxonomy'
+tax_file <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/data/process/gf_new.an.taxonomy'
 tax_function <- '~/Documents/Github/Schubert_humanCdGF_XXXX_2016/code/tax_level.R'
@@
   ggplot(aes(x = factor(day), y = percent_weightLoss, color = factor(Early_Euth))) + 
-        geom_jitter(width = 0.5, alpha = 0.1) + 
+        geom_jitter(width = 0.3, alpha = 0.1) + 
         scale_color_manual(values = c('black', 'red'), breaks=c("FALSE", "TRUE"),
@@
         labs(x = 'Day', y = ' Percent Relative to Day 0', title = 'Weight Loss') +
+          stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
+        theme_bw() +
         theme(legend.justification=c(1,0), legend.position=c(1,0.1), 
-                  legend.title=element_blank()) +
-          stat_summary(fun.data = 'median_hilow')
+              legend.title=element_blank())
 
@@
   ggplot(aes(x = factor(day), y = log_cfu, color = factor(Early_Euth))) + 
-        geom_jitter(width = 0.5, alpha = 0.2) + 
+        geom_jitter(width = 0.3, alpha = 0.2) + 
         scale_color_manual(values = c('black', 'red'), breaks=c("FALSE", "TRUE"),
@@
                   legend.title=element_blank()) +
-        stat_summary(fun.data = 'median_hilow')
+        stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) + 
+        theme_bw() +
+        theme(legend.justification=c(1,0), legend.position=c(1,0.2), 
+              legend.title=element_blank())
+
 ```
+
+Points are median of Disease Severity Group and lines are IQR

### 2020-03-03 exploratory/scratch/previous_analysis/submission/Schubert_humanCdGF_2015.Rmd
move old analysis files from 2016 to exploratory/scratch

---
title: 'Human-derived microbiota challenged with C.diff'
author: 'Alyxandria Schubert'
date: '10 June 2015'
---

# Colonization and Infection Resistance by Human-dervied Microbiotas against _Clostridium difficile_

**Running title:** Colonization and Infection Resistance against _Clostridium difficile_

Alyxandria M. Schubert, Jhansi Leslie, and Patrick D. Schloss\*

Correspondence:

> Department of Microbiology and Immunology
> University of Michigan
> 1520A Medical Science Research Bldg. I
> 1500 W. Medical Center Dr.
> Ann Arbor, MI 48109
> pschloss@umich.edu
> 734.647.5801

```{r, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
opts_chunk$set("dev" = c("png", "pdf"))
opts_chunk$set(results = "hold")
opts_chunk$set(fig.show = "hold")
opts_chunk$set(fig.align = "center")
opts_chunk$set(cache = FALSE)
opts_chunk$set(echo = FALSE)
opts_chunk$set(warning = FALSE)
```

### Abstract

### (Importance)

### Introduction

### Results

```{r summary_stats}

metadata <- read.table(file="~/Documents/Schloss/Data/GF/GFmetadata.txt", header=T, na.strings="NA", stringsAsFactors = FALSE)
num_sources <- length(unique(metadata$human_source)) 
## add source group column

num_cages <- length(unique(metadata$cage_id))
num_cages <- length(unique(metadata$cage_id))
num_mice <- length(unique(metadata$mouse_id))
plot( metadata$day, metadata$cdiff_cfu, log="y", ylim=c(.01, 10e11))
```

Stool from `r num_sources` human subjects were selected for gavage into germ-free mice. Of these subjects, 
X were inpatient and X were healthy subjects.

### Discussion

### Materials and Methods

### Acknowledgements

### Figure Legends

### 2020-03-03 code/make_files_file.R .R
fix path issues

@@ -58,5 +58,5 @@ mock_fastqs <- fastq_df %>%
 
 files_file <- bind_rows(experiment_fastqs, inoculum_fastqs, mock_fastqs)	
 
-write.table(files_file, 'data/mothur/humanGF_cdiff.files', sep = '\t',
+write.table(files_file, 'data/raw/humanGF_cdiff.files', sep = '\t',
 	quote = FALSE, row.names = FALSE, col.names = FALSE)

### 2020-03-24 code/get_cdiff_16S_variation.R
add v4 gene dist plot

@@ -55,6 +55,15 @@ sequence_identity_df <- map_dfr(unique_seqs, function(i){
 
 write_lines(concensus_seq, 'data/process/cdiff_16S/consensus_cdifficile_16S_V4.txt')
 
+#cdiff_v4_seq_dist_plot <- sequence_identity_df %>% 
+#	left_join(sequence_distribution, by = c('sequence')) %>% 
+#	ggplot(aes(x = percent_identity, y = n)) + 
+#		geom_bar(stat='identity', width = .125) + 
+#		labs(x = 'Percent Identity to C. difficile Consensus Sequence', y = 'Sequence Counts in C. difficile genomes', 
+#			title = 'Distribution of C. difficile 16S V4 gene sequences',
+#			subtitle = '85 Genomes with 819 Sequences') +
+#		coord_cartesian(xlim = c(96,100.2)) + theme_bw() + coord_flip()
+
 # select sequences below 97% similar to consensus sequence
 outlier_sequence_16S <- sequence_identity_df %>% 
 	filter(percent_identity < 97) %>% 

### 2020-03-24 exploratory/scratch/previous_analysis/submission/study.Rmd

--- 2016-11-10 submission/study.Rmd
@@
 ### Introduction
-Clostridium difficile is an opportunistic pathogen of the human lower gastrointestinal tract.
-Disruption of the native microbial community of the gut by antibiotics is the most common risk factor for development of C.
-difficile infection (CDI) (1).
-C.
-difficile is a spore-forming bacteria and can persist on abiotic surfaces and is not readily killed by ethanol-based hand-sanitizers, putting hospital patients particularly at risk.
-Indeed, ~12% of hospital acquired infections in the United States are due to C.
-difficile and result in up to 15,000 deaths annually (2).
-
-**need to add a paragraph about colonization resistence**
+_Clostridium difficile_ is an opportunistic pathogen of the human lower gastrointestinal tract.
+_C.
+difficile_ forms spores that can persist on abiotic surfaces and are not readily killed by ethanol-based hand-sanitizers, putting hospital patients particularly at risk.
+Indeed, ~12% of hospital acquired infections in the United States are due to _C.
+difficile_ and result in up to 15,000 deaths annually (2).
+Disruption of the native microbial community is the most common risk factor for development of _C.
+difficile_ infection (1).
+Antibiotic use and inflammatory bowel diseases are associated with loss of colonization resistance to _C.
+difficile_ infection through the loss of potentially protective bacterial families such as _Barnesiella_ and _Lachnospiraceae_ in both mouse models and human association experiments (cite alyx 2014 and vincent 2013).
+The composition of the community is clearly important for the acquisition, resistance and treatment of _C.
+difficile_ infection, as giving patients a healthy fecal microbiome transplant is the most effective treatment for this disease (cite Anna 2014).
+The precise mechanisms of colonization resistance and restoration of a healthy community are yet to be discovered.
 
 Murine models to study CDI typically rely on treating conventionally-raised mice with antibiotics either in drinking water or by injection to induce susceptibility (3, 4).
-This model provides a convenient way to study C.
-difficile pathogenesis and virulence factors.
-Numerous microbiome studies have been performed using this model to determine the antibiotic classes (5), starting microbial community (6) and metabolites (7) that impact development and severity of CDI.
+This model provides a convenient way to study _C.
+difficile_ pathogenesis and virulence factors.
+Numerous microbiome studies have been performed using this model to determine the antibiotic classes (5), starting microbial community (6) and metabolites (7) that impact development and severity of _C.
+difficile_ infection.
 While informative, these studies are somewhat removed from human disease because they only examine mouse-associated microbial communities.
@@
 Gnotobiotic or germ-free mouse models have been used for a range of studies of CDI, including assessment of species-specific interactions between C.
-difficile and competing microbial community members (8), analysis of nutrient restriction (9), _in vivo_ transcriptomics of C.
+difficile and competing microbial community members (8), analysis of nutrient restriction (9), in vivo transcriptomics of C.
 difficile and examination of host immune response to CDI (10).
@@
 
+The goal of this study was to design a laboratory model that allowed for studying and modeling microbial coumminity interactions with _C.
+difficile_ infection in mice with human derived microbiomes.
 To test the impact of individual human microbiomes on CDI, we colonized germ-free mice with 16 different human stool donors.

### 2020-03-24 exploratory/scratch/previous_analysis/submission/study.tex

--- 2016-10-27 submission/study.tex
@@
-\documentclass[11pt,]{article}
+\documentclass[]{article}
 \usepackage{lmodern}
@@
 }{}
-\usepackage[margin=1.0in]{geometry}
 \ifxetex
@@
             pdfauthor={},
-            pdftitle={Microbiota predict Clostridium difficile severity in germ-free mice colonized with human feces},
+            pdftitle={},
             colorlinks=true,
@@
 \setlength{\droptitle}{-2em}
-  \title{\textbf{Microbiota predict \emph{Clostridium difficile} severity in
-germ-free mice colonized with human feces}}
-  \pretitle{\vspace{\droptitle}\centering\huge}
-  \posttitle{\par}
+  \title{}
+  \pretitle{\vspace{\droptitle}}
+  \posttitle{}
   \author{}
@@
 in the United States are due to C. difficile and result in up to 15,000
-deaths annually (2).
+deaths annually (2). This is a test (@Britton12).
 

### 2021-04-13 code/convert_OTU_labels.R
add read_dist.R from clearance repo

@@ -1,56 +1,0 @@
-###################
-#
-# convert_OTU_labels.R
-#
-# creates table of tax labels for OTUs
-# created by Nicholas Lesniak in 2016
-# updated by Nicholas Lesniak in 2019
-#
-# convert mothur output to a dataframe 
-# 
-# taxa_file - taxonomy file location
-#
-###################
-
-library(dplyr)
-library(tidyr)
-library(readr)
-
-input_file <- as.character(commandArgs(trailingOnly=TRUE))
-
-taxonomy_file <- paste0('data/process/', input_file)
-print(paste('Converting', taxonomy_file))
-taxonomy_df <- read.table(taxonomy_file, sep = '\t', header = T)
-
-levels <- c('Kingdom','Phylum','Class','Order','Family','Genus')
-
-# convert mothur taxonomy string to columns
-taxonomy_df_edit <- taxonomy_df %>% 
-	separate(Taxonomy, levels, sep = '\\(\\d{2,3}\\);', extra = 'drop') %>% 
-	select(-Size) 
-# in older version of mothur unclassified are listed as unclassified
-# without information from higher level classification
-# for those cases, append with lowest identified classification
-if(any(grepl('nclassified', taxonomy_df_edit$Genus))){ 
-	taxonomy_df_edit <- taxonomy_df_edit %>% 
-		gather(Level, Classification, -OTU) %>% # convert to long form to group classification by OTU 
-		mutate(Level = factor(Level, levels))  %>% # order classification level
-		left_join(group_by(., OTU) %>% # create dataframe with OTU and lowest level classification
-				filter(Classification != 'unclassified') %>% # remove unclassifieds
-				filter(Level == levels[max(as.numeric(Level))]) %>% # select lowest level classification
-				select(OTU, Lowest_classified = Classification), 
-			by = 'OTU') %>% 
-		mutate(Classification = ifelse(Classification == 'unclassified', 
-				paste(Lowest_classified, Classification, sep= '_'), # append unclassified with lowest classification
-				Classification)) %>% 
-		select(-Lowest_classified) %>% 
-		spread(Level, Classification)
-}
-
-# edit output to have cleaned up OTU
-taxonomy_df_edit <- taxonomy_df_edit %>%  
-    mutate(tax_otu_label = paste0(Genus, ' (',  
-        gsub('tu0*', 'TU ', OTU),')'), # create labels 
[…]

### 2021-04-13 code/tidy_taxonomy.R
convert convert_taxonomy to tidy_taxonomy.R

@@ -1,0 +1,55 @@
+###################
+#
+# tidy_taxonomy.R
+#
+# creates table of tax labels for OTUs
+# created by Nicholas Lesniak in 2016
+# updated by Nicholas Lesniak in 2019, 2021
+#
+# convert mothur output to a dataframe 
+# 
+# input - mothur taxonomy file 
+#	data/mothur/final.taxonomy
+#
+# output - tsv with columns for OTU and each taxonomic group
+#	data/process/final.taxonomy.tidy.tsv
+#
+###################
+
+library(tidyverse)
+
+taxonomy_df <- read_tsv('data/mothur/final.taxonomy',
+	col_types = 'cdc')
+
+levels <- c('Kingdom','Phylum','Class','Order','Family','Genus')
+
+# convert mothur taxonomy string to columns
+taxonomy_df_edit <- taxonomy_df %>% 
+	separate(Taxonomy, levels, sep = '\\(\\d{2,3}\\);', extra = 'drop') %>% 
+	select(-Size) 
+# in older version of mothur unclassified are listed as unclassified
+# without information from higher level classification
+# for those cases, append with lowest identified classification
+if(any(grepl('nclassified', taxonomy_df_edit$Genus))){ 
+	taxonomy_df_edit <- taxonomy_df_edit %>% 
+		gather(Level, Classification, -OTU) %>% # convert to long form to group classification by OTU 
+		mutate(Level = factor(Level, levels))  %>% # order classification level
+		left_join(group_by(., OTU) %>% # create dataframe with OTU and lowest level classification
+				filter(Classification != 'unclassified') %>% # remove unclassifieds
+				filter(Level == levels[max(as.numeric(Level))]) %>% # select lowest level classification
+				select(OTU, Lowest_classified = Classification), 
+			by = 'OTU') %>% 
+		mutate(Classification = ifelse(Classification == 'unclassified', 
+				paste(Lowest_classified, Classification, sep= '_'), # append unclassified with lowest classification
+				Classification)) %>% 
+		select(-Lowest_classified) %>% 
+		spread(Level, Classification)
+}
+
+# edit output to have cleaned up OTU
+taxonomy_df_edit <- taxonomy_df_edit %>%  
+    mutate(tax_otu_label = paste0(Genus, ' (',  
+        gsub('tu0*', 'TU ', OTU),')'), # create labels 
[…]

### 2021-04-13 exploratory/notebook/initial_analysis/initial_analysis_notebook.Rmd
add initial anlysis notebook with skecth of figs1-3

---
title: "initial_analysis_notebook"
author: "Nick Lesniak"
date: "4/13/2021"
output: rmarkdown::github_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
#knitr::opts_knit$set(root.dir = '../../..')
```

### Process sequencing data

- removed unused fastq files

### run fastqs through mothur

```{r read in data}
library(tidyverse)

# calculate error rate
error_df <- read.table('../../../data/mothur/sample.error.count', header = T)
total_seqs <- sum(error_df$Sequences)
total_errors <- sum(error_df$Errors * error_df$Sequences)
error_rate <- total_errors / (total_seqs * 253) * 100

read_tsv('../../../data/mothur/sample.final.shared',
         col_types = cols(.default = col_double(),
                          Group = col_character())) %>% 
  select(-numOtus, -label) %>% 
  pivot_longer(cols = -Group, values_to = 'count', names_to = 'otu') %>% 
  group_by(Group) %>% 
  summarise(counts = sum(count)) %>% 
  filter(counts < 2020) %>% 
  arrange(counts) %>% 
  mutate(list = paste0(Group, ' (', counts, ' seqs)')) %>% 
  pull(list) %>% 
  paste(., collapse = ', ')
```

subsampled to 2017, eliminating the 25 samples listed above with their counts
error rate for the data was `r round(error_rate, 2)`%

```{r}
# 
source('../../../code/Sum_OTU_by_Tax.R')

# load in data
metadata <- read_tsv('../../../data/process/metadata_tidy.tsv',
                     col_type = 'cccccDddddDcdddddcdl') 
toxin <- read_tsv('../../../data/process/toxin_tidy.tsv',
                  col_type = 'cccccdc')
histology <- read_tsv('../../../data/process/histology_tidy.tsv',
                      col_type = 'ccccdddcdcc')
human_source <- read_tsv('../../../data/process/human_source_tidy.tsv',
                         col_type = cols(.default = col_character(), age = col_double()))
shared <- read_tsv('../../../data/mothur/sample.final.0.03.subsample.shared',
         col_types = cols(.default = col_double(),
                          Group = col_character())) %>% 
  mutate(Group = ifelse(grepl('_D|DA', Group), Group,
      gsub('D', '_D', Group))) %>% 
  select(-label, -numOtus) %>% 
  pivot_longer(-Group, names_to = 'otu', values_to = 'counts') %>% 
  filter(counts > 0) %>% 
  mutate(relative_abundance = counts/2107 * 100)
taxonomy <- read_tsv('../../../data/process/final.taxonomy.tidy.tsv',
                     col_types = cols(.default = col_character()))
nmds_df <- read_tsv('../../../data/mothur/sample.final.thetayc.0.03.lt.ave.nmds.axes',
         col_types ='cdd') %>% 
    mutate(group = ifelse(grepl('_D|DA', group), group,
                        gsub('D', '_D', group)))
source('../../../code/read_dist.R')
beta_df <- read_dist('../../../data/mothur/sample.final.thetayc.0.03.lt.ave.dist') %>% 
      mutate(rows = ifelse(grepl('_D|DA', rows), rows,
                        gsub('D', '_D', rows)),
             columns = ifelse(grepl('_D|DA', columns), columns,
                        gsub('D', '_D', columns)))
donor_colors <- c("#41bbc5", "#7e2640", "#9ad859", "#682dbd", "#728e24", "#e26df8", "#23980d", "#fd048f", "#46e33e", "#fa1bfc", "#155126", "#f9b2d1", "#154975", "#e2c627", "#6294ce")
```

# Figure 1 - Human communities reporducibly colonize mice

Sampled spread of diverse donors for inoculating germ-free mice

```{r fig.width=7, fig.height=5}

human_source_list <- shared %>% 
  filter(grepl('DA', Group)) %>% 
  pull(Group) %>% unique

# plot distribution of donors by their stool status - case/diarrheal/nondiarrheal control
#
#human_source %>% 
#  filter(sample_id %in% human_source_list) %>% 
#  select(sample_id, disease_stat) %>% 
#  left_join(nmds_df, by = c('sample_id' = 'group')) %>% 
#  ggplot(aes(x = axis1, y = axis2, color = disease_stat)) + 
#    geom_point(size = 5) + 
#    labs(x = 'NMDS axis 1', y = 'NMDS axis 2', color = NULL) + 
#    theme_bw()
#
# status doesnt matter to this story as much as the outcome of the challenge

donor_by_outcome <- histology %>% 
  select(human_source, epithelial_damage) %>% 
  group_by(human_source) %>% 
  summarise(epithelial_damage = mean(epithelial_damage)) %>% 
  mutate(epithelial_damage = epithelial_damage > 1) %>% 
  left_join(unique(select(metadata, human_source, early_euth, cdiff_strain),
            by = c('human_source'))) %>% 
  filter(cdiff_strain == 431) %>% 
  mutate(outcome = case_when(early_euth == T ~ 'moribund',
                             epithelial_damage == F ~ 'mild',
                             epithelial_damage == T ~ 'severe',
                             T ~ 'NA'),
         outcome = factor(outcome, levels = c('mild', 'severe', 'moribund'))) %>% 
  select(human_source, outcome)

donor_by_outcome %>% 
  left_join(nmds_df, by = c('human_source' = 'group')) %>% 
  ggplot(aes(x = axis1, y = axis2, color = outcome)) + 
    geom_point() + 
    labs(x = 'NMDS axis 1', y = 'NMDS axis 2', color = NULL) + 
    theme_bw()

```

> recalc nmds with just these samples?

Unique communities without conserved structure
```{r}
shared %>% 
  filter(grepl('DA', Group)) %>% 
  left_join(donor_by_outcome, 
            by = c('Group' = 'human_source')) %>% 
#  mutate(label = case_when(disease_stat == 'Case' ~ 'CDI_431',
#                           disease_stat == 'NonDiarrhealControl' ~ paste0('NDC_', gsub('DA0*', '', Group)),
#                           disease_stat == 'DiarrhealControl' ~ paste0('DC_', gsub('DA0*', '', Group)))) %>% 
  left_join(taxonomy, by = c('otu' = 'OTU')) %>% 
  rename(taxa_level = Class) %>% 
  group_by(Group, taxa_level, outcome) %>% 
  summarise(relative_abundance = log10(sum(relative_abundance) + 0.0001)) %>% 
  group_by(taxa_level) %>% 
  mutate(median_ra = median(relative_abundance)) %>% 
  ggplot(aes(x = Group, 
             y = reorder(taxa_level, median_ra), 
             fill = relative_abundance)) + 
    geom_tile() + 
    scale_fill_gradient2(low="white", mid='#0B775E', high = 'black',
		  	limits = c(-2,2), na.value = NA, midpoint = .3,
			  breaks = c(-2.5, -1, 0, 1, 2), labels = c('', '0.1', '1', '10', '100')) + 
	  theme_bw() + 
    labs(x = NULL, y = NULL, fill = NULL) + 
    scale_x_discrete(guide = guide_axis(angle = 45)) + 
    facet_grid(.~outcome, scales = 'free_x')
```

> does outcome matter at this point? in this figure?

Donor communities colonize mice and shift community, ~~but consistently~~

```{r fig.width = 10}
shared %>% 
  filter(grepl('D0', Group)) %>% 
  left_join(select(metadata, group, human_source, mouse_id), by = c('Group' = 'group')) %>% 
  left_join(donor_by_outcome, by = c('human_source')) %>% 
#  mutate(label = case_when(disease_stat == 'Case' ~ paste('CDI', gsub('DA0*', '', human_source), 
#                                                          gsub('^.*_', '', mouse_id), sep = '_'), 
#                           disease_stat == 'NonDiarrhealControl' ~ paste('NDC', gsub('DA0*', '', human_source), 
#                                                                         gsub('^.*_', '', mouse_id), sep = '_'),
#                           disease_stat == 'DiarrhealControl' ~ paste('DC', gsub('DA0*', '', human_source), 
#                                                                      gsub('^.*_', '', mouse_id), sep = '_'))) %>% 
  left_join(taxonomy, by = c('otu' = 'OTU')) %>% 
  rename(taxa_level = Class) %>% 
  group_by(Group, taxa_level, human_source) %>% 
  summarise(relative_abundance = log10(sum(relative_abundance) + 0.0001)) %>% 
  group_by(taxa_level) %>% 
  mutate(median_ra = median(relative_abundance)) %>% 
  ggplot(aes(x = Group, 
             y = reorder(taxa_level, median_ra), 
             fill = relative_abundance)) + 
    geom_tile() + 
    scale_fill_gradient2(low="white", mid='#0B775E', high = 'black',
		  	limits = c(-2,2), na.value = NA, midpoint = .3,
			  breaks = c(-2.5, -1, 0, 1, 2), labels = c('', '0.1', '1', '10', '100')) + 
	  theme_bw() + 
    labs(x = NULL, y = NULL, fill = NULL) + 
    scale_x_discrete(guide = guide_axis(angle = 45)) + 
    facet_grid(.~human_source, scales = 'free_x')

```

```{r}
nmds_df %>% 
  left_join(filter(select(metadata, group, human_source, day), day %in% c(0, 10)), by = c('group')) %>% 
  mutate(human_source = case_when(!is.na(human_source) ~ human_source,
                                  grepl('DA', group) ~ group,
                                  grepl('CON3', group) ~ 'DA00430',
                                  T ~ 'Unknown'),
         sample_type = ifelse(grepl('DA', group), 'human', 'murine')) %>% 
  filter(sample_type == 'human' | day == 0) %>% 
  ggplot(aes(x = axis1, y = axis2, color = human_source)) +
    geom_point() + 
    scale_color_manual(values = donor_colors) + 
    facet_grid(~sample_type) + 
    theme_bw() + 
    guides(color = 'none') + 
    labs(x = 'NMDS axis 1', y = 'NMDS axis 2')
```

```{r}
day_0_samples <- metadata %>% 
  filter(day == 0) %>% 
  select(group, human_source)
# compare distance between:
#   donor and recipient
#   co-recipients
#   other cages
#   other donors
donor_v_recipent <- bind_rows(
    day_0_samples %>% 
      inner_join(beta_df, by = c('group' = 'rows', 'human_source' = 'columns')),
    day_0_samples %>% 
      inner_join(beta_df, by = c('group' = 'columns', 'human_source' = 'rows'))) %>% 
  mutate(comparison = 'Donor\nvs\nRecipient')

corecipient <- day_0_samples %>% 
  left_join(day_0_samples, by = c('human_source')) %>% 
  filter(group.x != group.y) %>% 
  inner_join(beta_df, by = c('group.x' = 'rows', 'group.y' = 'columns')) %>% 
  mutate(comparison = 'Recipients\ncompared to\nothers w/same donor')

other_donors <- beta_df %>% 
  filter((grepl('DA', rows) & grepl('D0', columns)) | 
         (grepl('D0', rows) & grepl('DA', columns))) %>% 
  anti_join(donor_v_recipent, by = c('rows' = 'group', 'columns' = 'human_source')) %>% 
  anti_join(donor_v_recipent, by = c('rows' = 'human_source', 'columns' = 'group')) %>% 
  mutate(comparison = 'Donor\nvs\nNon-recipients')

other_recipients <- beta_df %>% 
  filter(grepl('D0', rows) & grepl('D0', columns)) %>% 
  anti_join(corecipient, by = c('rows' = 'group.x', 'columns' = 'group.y')) %>% 
  mutate(comparison = 'Recipients\ncompared to\nothers w/different donor')

bind_rows(donor_v_recipent, corecipient, other_donors, other_recipients) %>% 
  mutate(comparison = factor(comparison, levels = 
                              c('Donor\nvs\nRecipient', 'Recipients\ncompared to\nothers w/same donor', 
                                'Donor\nvs\nNon-recipients', 'Recipients\ncompared to\nothers w/different donor'))) %>% 
  ggplot(aes(x = comparison, y = distances)) + 
    geom_boxplot() + 
    labs(x = NULL, y = 'Theta YC') +
    theme_bw()

corecipient %>% 
  separate(group.x, c('cageA', 'mouseA', 'dayA')) %>% 
  separate(group.y, c('cageB', 'mouseB', 'dayB')) %>% 
  mutate(same_cage = cageA == cageB,) %>% 
  ggplot(aes(x = human_source, y = distances, color = human_source, shape = same_cage)) +
    geom_jitter(aes(color = human_source), width = 0.2) + 
    scale_color_manual(values = donor_colors) + 
    scale_shape_manual(values = c(18,16)) + 
    scale_x_discrete(guide = guide_axis(angle = 45)) + 
    theme_bw() +
    theme(legend.position = 'bottom') + 
    labs(y = 'Theta YC', x = 'Donor Source', color = NULL, shape = 'Same Cage') + 
    guides(color = 'none')
```

```{r}
#genus <- sum_otu_by_taxa(taxonomy_df = taxonomy,
#                         otu_df = select(shared, Group = sample_id, contains('Otu')),
#                         taxa_level = 'Genus', 
#                         top_n = 12)
#donor_genus <- sum_otu_by_taxa(taxonomy_df = donor_taxonomy,
#                         otu_df = donor,
#                         taxa_level = 'Genus', 
#                         top_n = 12)
#
#donor_order <- donor_genus %>% 
#    filter(taxa == 'Other') %>% 
#    arrange(desc(abundance)) %>% 
#    pull(Group)
#
#
#donor_genus %>% 
#  mutate(rel_abun = abundance/2779 * 100,
#         taxa = factor(taxa,
#                       levels = c("Alistipes", "Bacteroides", "Blautia", 
#                                  "Clostridium_sensu_stricto", "Cronobacter",
#                                  "Enterococcus", "Faecalibacterium", 
#                                  "Lachnospiraceae_unclassified", "Lactobacillus", 
#                                  "Prevotella", "Ruminococcaceae_unclassified", "Other")),
#          Group = factor(Group, levels = donor_order)) %>% 
#  ggplot(aes(x = Group, y = taxa, fill = rel_abun)) + 
#    labs(x = NULL, y = NULL, fill = 'Relative Abundance') + 
#    geom_tile() + 
#    scale_fill_gradient(low = 'white', high = 'black')
#    coord_flip()
# 
##    scale_fill_gradient2(low="white", mid='blue3', high = 'black',
##  			limits = c(-2,2), na.value = NA, midpoint = .3,
##  			breaks = c(-2.5, -1, 0, 1, 2), labels = c('', '0.1', '1', '10', '100'))

```

# Figure 2 - Mice were colonized without any perturbation

```{r fig.height = 14}
metadata %>% 
  filter(cdiff_strain == 431) %>% 
  mutate(human_source = factor(human_source, levels = 
                                c('DA01324', 'DA00953', 'DA00581', 'DA01134', 'DA10148', 
                                  'DA00369', 'DA00430', 'DA10093', 'DA10027', 'DA10034',
                                  'DA10082', 'DA00884', 'DA00578', 'DA00431', 'DA01245'))) %>% 
  ggplot(aes(x = day, y = log10(cdiff_cfu + 60), color = human_source)) +
    stat_summary(fun = 'median', geom = 'line') + 
    geom_point(alpha = 0.3) +
    facet_grid(human_source~.) + 
    scale_color_manual(values = donor_colors) + 
    guides(color = 'none') + 
    theme_bw() + 
    labs(x = 'Day', y = 'C. difficile CFU (log10)')

metadata %>% 
  filter(cdiff_strain == 431) %>% 
  mutate(human_source = factor(human_source, levels = 
                                c('DA01324', 'DA00953', 'DA00581', 'DA01134', 'DA10148', 
                                  'DA00369', 'DA00430', 'DA10093', 'DA10027', 'DA10034',
                                  'DA10082', 'DA00884', 'DA00578', 'DA00431', 'DA01245'))) %>% 
  ggplot(aes(x = day, y = percent_weightLoss, color = human_source)) +
    stat_summary(fun = 'median', geom = 'line') + 
    geom_point(alpha = 0.3) +
    facet_grid(human_source~.) + 
    scale_color_manual(values = donor_colors) + 
    guides(color = 'none') + 
    theme_bw() + 
    labs(x = 'Day', y = 'Weight (Relative to initial)')

```

```{r}
metadata %>% 
  filter(cdiff_strain == 431,
         day %in% c(1,10)) %>% 
  mutate(human_source = factor(human_source, levels = 
                                c('DA01324', 'DA00953', 'DA00581', 'DA01134', 'DA10148', 
                                  'DA00369', 'DA00430', 'DA10093', 'DA10027', 'DA10034',
                                  'DA10082', 'DA00884', 'DA00578', 'DA00431', 'DA01245'))) %>% 
  ggplot(aes(x = human_source, y = log10(cdiff_cfu + 60), color = human_source)) +
    geom_jitter(width = 0.2) + 
    facet_grid(day~.)+ 
    scale_x_discrete(guide = guide_axis(angle = 45)) + 
    scale_color_manual(values = donor_colors) + 
    theme_bw() + 
    labs(x = 'Donor Source', y = 'C. difficile CFU (log10)') + 
    guides(color = 'none')

metadata %>% 
  filter(cdiff_strain == 431,
         day %in% c(2,10)) %>% 
  mutate(human_source = factor(human_source, levels = 
                                c('DA01324', 'DA00953', 'DA00581', 'DA01134', 'DA10148', 
                                  'DA00369', 'DA00430', 'DA10093', 'DA10027', 'DA10034',
                                  'DA10082', 'DA00884', 'DA00578', 'DA00431', 'DA01245'))) %>% 
  ggplot(aes(x = human_source, y = percent_weightLoss, color = human_source)) +
    geom_jitter(width = 0.2) + 
    facet_grid(day~.)+ 
    scale_x_discrete(guide = guide_axis(angle = 45)) + 
    scale_color_manual(values = donor_colors) + 
    theme_bw() + 
    labs(x = 'Donor Source', y = 'Weight (Relative to initial)') + 
    guides(color = 'none')
```

```{r}
metadata %>% 
  filter(day == 0 | day == mouse_endpoint) %>% 
  mutate(time_point = ifelse(day == 0, 'Initial', 'Endpoint'),
         time_point = factor(time_point, levels = c('Initial', 'Endpoint'))) %>% 
  left_join(nmds_df, by = c('group')) %>% 
  left_join(donor_by_outcome, by = c('human_source')) %>% 
  ggplot(aes(x = axis1, y = axis2, color = early_euth)) +
    geom_point() + 
    facet_grid(.~time_point) +
    #scale_color_manual(values  = donor_colors) + 
    theme_bw() + 
    labs(x = 'NMDS axis 1', y = 'NMDS axis 2', color = 'Moribund') + 
    theme(legend.position = 'bottom')

```

> NAs on Day 0 for DA01134, DA10034, DA00884  
> Last samples for moribund mice is from cecum  
> Switch to cage color?  
> Recalc nmds?

# Figure 3 - Difference in severity

```{r histology}

histology %>% 
  left_join(distinct(select(metadata, mouse_id, early_euth)), by = 'mouse_id') %>% 
  filter(!is.na(early_euth)) %>% 
  mutate(early_euth = ifelse(early_euth, 'Moribund', 'Colonized')) %>% 
  left_join(donor_by_outcome, by = c('human_source')) %>% 
  #filter(outcome == 'mild', epithelial_damage > 0)
  pivot_longer(cols = c(edema_tissue, inflammation_tissue, epithelial_damage, summary_score),
               names_to = 'histology', values_to = 'score') %>% 
  ggplot(aes(x = histology, y = score, fill = outcome)) +
    geom_boxplot() + 
    #geom_point(position = position_jitterdodge(), alpha = 0.5) + 
    theme_bw() + 
    labs(x = NULL, y = 'Histology Score', fill = 'Outcome') +
    scale_x_discrete(guide = guide_axis(angle = 45)) + 
    facet_grid(.~early_euth)
```

Mice were challenged with a strain isolated that matched Clostridioides difficile ribotype 027

> what is published about histology of 027 in C57/B6 mice?

```{r toxin}

toxin %>% 
  filter(!is.na(ear_tag),
         !is.na(toxin_sample_day)) %>% 
  left_join(select(metadata, group, human_source), by = c('group')) %>% 
  left_join(donor_by_outcome, by = c('human_source')) %>% 
  filter(!is.na(outcome)) %>% 
  mutate(day = as.numeric(toxin_sample_day)) %>% 
  ggplot(aes(x = day, y = Log_repiricoal_dilution, fill = outcome)) + 
#    geom_line(aes(color = outcome, group = mouse_id)) + 
    geom_boxplot(aes(group = interaction(day, outcome))) +
    labs(x = 'Day', y = 'Toxin activity') + 
#    facet_grid(outcome~.) + 
    theme_bw()
# check for correlation between toxin and cfu
#toxin %>% 
#  filter(!is.na(ear_tag),
#         !is.na(toxin_sample_day)) %>% 
#  left_join(select(metadata, group, cdiff_cfu), by = 'group') %>% 
#  ggplot(aes(x = log10(cdiff_cfu + 0.01), y = Log_repiricoal_dilution)) + 
#    geom_point(alpha = 0.2) +
#    labs(x = 'CFU', y = 'Toxin activity') + 
#    theme_bw()
# does this matter?

# how many samples tested are positive for toxin per mouse?
#toxin %>% 
#  group_by(mouse_id) %>% 
#  summarise(n_samples = length(toxin_sample_day),
#            n_positive = sum(Log_repiricoal_dilution > 1))
# is this relevant?
```

```{r, individual cfu plots}
#metadata %>% 
#  left_join(mutate(histology, mouse_id = paste0(cage_ID, '_', mouse_ID)) %>% select(-day),
#            by = c('mouse_id')) %>% 
#  mutate(outcome = ifelse(summary_score < 4, 'Low', outcome)) %>% 
#ggplot(aes(x = day, 
#           y = log_cfu, 
#           color = outcome, 
#           group = mouse_id)) + 
#  geom_line() + 
#  facet_wrap(outcome~., ncol = 1) +
#  theme_bw()
```

### 2021-04-23 code/install_mothur.sh
fix mothur to install on osx

@@ -7,8 +7,8 @@
 # The zip archive contains a director called "mothur" so we can extract it directly
 # to code/
 
-wget -P code/mothur/ -nc https://github.com/mothur/mothur/releases/download/v1.44.3/Mothur.linux.zip
-unzip -n -d code/ code/mothur/Mothur.linux.zip
+wget -P code/mothur/ -nc https://github.com/mothur/mothur/releases/download/v1.45.2/Mothur.OSX-10.14.zip
+unzip -n -d code/ code/mothur/Mothur.OSX-10.14.zip
 
 if [[ $? -eq 0 ]]
 then

### 2021-04-23 code/run_lefse_analysis.R
create script to subset dataframe and run lefse

@@ -1,0 +1,57 @@
+
+
+library(tidyverse)
+
+metadata <- read_tsv('data/process/metadata_tidy.tsv',
+		col_type = 'cccccDddddDcdddddcdl') 
+shared <- read_tsv('data/mothur/sample.final.0.03.subsample.shared',
+		col_types = cols(.default = col_double(),
+			Group = col_character())) %>% 
+	mutate(Group = ifelse(grepl('_D|DA', Group), Group,
+		gsub('D', '_D', Group)))
+taxonomy <- read_tsv('data/process/final.taxonomy.tidy.tsv',
+                     col_types = cols(.default = col_character()))
+
+run_lefse <- function(sample_df_name, tax_level){
+	i <- sample_df_name
+	current_df <- get(i)
+
+	current_shared <- shared %>% 
+		filter(Group %in% current_df$Group) %>% 
+		pivot_longer(cols = -c('label', 'Group', 'numOtus'),
+			names_to = "OTU", values_to = 'value') %>% 
+		left_join(taxonomy, by = c('OTU')) %>% 
+		group_by(label, Group, numOtus, .data[[tax_level]]) %>% 
+		summarise(value = sum(value)) %>% 
+		pivot_wider(names_from = .data[[tax_level]], values_from = value) %>% 
+		ungroup
+	# remove otus that either have 0 or only present in 5 or fewer samples
+	present_otus <- current_shared %>% 
+		select(-label, -Group, -numOtus) %>% 
+		map_dbl(~ sum(. > 0)) %>% 
+		which(x = (. > 5)) %>% 
+		names
+	current_shared <- current_shared %>% 
+		select(label, Group, numOtus, one_of(present_otus)) %>% 
+		mutate(numOtus = length(present_otus))
+
+	# write files to be used in mothur for lefse analysis
+	write_tsv(path = paste0('data/process/', i, '_', tax_level, '.shared'), 
+		x = filter(current_shared, Group %in% current_df$Group))
+	write_tsv(path = paste0('data/process/', i, '_', tax_level, '.design'), 
+		x = filter(current_df, Group %in% current_shared$Group))
+	print(tax_level)
+	# run lefse
+	system(paste0('code/mothur/mothur "#set.dir(input=data/process, output=data/process);
+		lefse(shared=', i, '_', tax_level, '.shared, design=', i, '_', tax_level, '.design)"'))
+}
+
+# lefse - severe disease by OTU
[…]

### 2021-04-23 exploratory/notebook/initial_analysis/initial_analysis_notebook.Rmd
add lefse and correlation of toxin/histology

@@
 
-# plot distribution of donors by their stool status - case/diarrheal/nondiarrheal control
-#
-#human_source %>% 
-#  filter(sample_id %in% human_source_list) %>% 
-#  select(sample_id, disease_stat) %>% 
-#  left_join(nmds_df, by = c('sample_id' = 'group')) %>% 
-#  ggplot(aes(x = axis1, y = axis2, color = disease_stat)) + 
-#    geom_point(size = 5) + 
+#nmds_df %>% 
+#  filter(group %in% human_source_list) %>% 
+#  ggplot(aes(x = axis1, y = axis2, color = outcome)) + 
+#    geom_point() + 
 #    labs(x = 'NMDS axis 1', y = 'NMDS axis 2', color = NULL) + 
 #    theme_bw()
-#
-# status doesnt matter to this story as much as the outcome of the challenge
-
-donor_by_outcome <- histology %>% 
-  select(human_source, epithelial_damage) %>% 
-  group_by(human_source) %>% 
-  summarise(epithelial_damage = mean(epithelial_damage)) %>% 
-  mutate(epithelial_damage = epithelial_damage > 1) %>% 
-  left_join(unique(select(metadata, human_source, early_euth, cdiff_strain),
-            by = c('human_source'))) %>% 
-  filter(cdiff_strain == 431) %>% 
-  mutate(outcome = case_when(early_euth == T ~ 'moribund',
-                             epithelial_damage == F ~ 'mild',
-                             epithelial_damage == T ~ 'severe',
-                             T ~ 'NA'),
-         outcome = factor(outcome, levels = c('mild', 'severe', 'moribund'))) %>% 
-  select(human_source, outcome)
-
-donor_by_outcome %>% 
-  left_join(nmds_df, by = c('human_source' = 'group')) %>% 
-  ggplot(aes(x = axis1, y = axis2, color = outcome)) + 
-    geom_point() + 
-    labs(x = 'NMDS axis 1', y = 'NMDS axis 2', color = NULL) + 
-    theme_bw()
-
+# theta yc shows comparison better
 ```
-
-> recalc nmds with just these samples?
 
@@
   filter(grepl('DA', Group)) %>% 
-  left_join(donor_by_outcome, 
-            by = c('Group' = 'human_source')) %>% 
-#  mutate(label = case_when(disease_stat == 'Case' ~ 'CDI_431',
-#                           disease_stat == 'NonDiarrhealControl' ~ paste0('NDC_', gsub('DA0*', '', Group)),
-#                           disease_stat == 'DiarrhealControl' ~ paste0('DC_', gsub('DA0*', '', Group)))) %>% 
   left_join(taxonomy, by = c('otu' = 'OTU')) %>% 
   rename(taxa_level = Class) %>% 
-  group_by(Group, taxa_level, outcome) %>% 
+  group_by(Group, taxa_level) %>% 
   summarise(relative_abundance = log10(sum(relative_abundance) + 0.0001)) %>% 
@@
     scale_x_discrete(guide = guide_axis(angle = 45)) + 
-    facet_grid(.~outcome, scales = 'free_x')
+    facet_grid(.~Group, scales = 'free_x')
 ```
-
-> does outcome matter at this point? in this figure?
 
@@
   left_join(select(metadata, group, human_source, mouse_id), by = c('Group' = 'group')) %>% 
-  left_join(donor_by_outcome, by = c('human_source')) %>% 
-#  mutate(label = case_when(disease_stat == 'Case' ~ paste('CDI', gsub('DA0*', '', human_source), 
-#                                                          gsub('^.*_', '', mouse_id), sep = '_'), 
-#                           disease_stat == 'NonDiarrhealControl' ~ paste('NDC', gsub('DA0*', '', human_source), 
-#                                                                         gsub('^.*_', '', mouse_id), sep = '_'),
-#                           disease_stat == 'DiarrhealControl' ~ paste('DC', gsub('DA0*', '', human_source), 
-#                                                                      gsub('^.*_', '', mouse_id), sep = '_'))) %>% 
   left_join(taxonomy, by = c('otu' = 'OTU')) %>% 
@@
     scale_x_discrete(guide = guide_axis(angle = 45)) + 
-    facet_grid(.~human_source, scales = 'free_x')
+    facet_grid(.~human_source, scales = 'free_x') + 
+    theme(axis.ticks.x = element_blank(),
+          axis.text.x = element_blank())
 
@@
   select(group, human_source)
+day_10_samples <- metadata %>% 
+  filter(day == 10) %>% 
+  select(group, human_source)
+
 # compare distance between:
@@
 #   other donors
+
+donor_v_donor <- beta_df %>% 
+  filter(grepl('DA', rows),
+         grepl('DA', columns)) %>% 
+  mutate(comparison = 'Donor\nvs\nDonor')
+
 donor_v_recipent <- bind_rows(
@@
   mutate(comparison = 'Donor\nvs\nRecipient')
+
+donor_v_recipent_end <- bind_rows(
+    day_10_samples %>% 
+      inner_join(beta_df, by = c('group' = 'rows', 'human_source' = 'columns')),
+    day_10_samples %>% 
+      inner_join(beta_df, by = c('group' = 'columns', 'human_source' = 'rows'))) %>% 
+  mutate(comparison = 'Donor\nvs\nRecipient End')
 
@@
 
-bind_rows(donor_v_recipent, corecipient, other_donors, other_recipients) %>% 
+day0_v_day10 <- beta_df %>% 
+  separate(rows, c('cageA', 'mouseA', 'dayA')) %>% 
+  separate(columns, c('cageB', 'mouseB', 'dayB')) %>% 
+  filter(mouseA == mouseB & cageA == cageB) %>% 
+  filter((dayA == 'D0' & dayB == 'D10') |
+         (dayA == 'D10' & dayB == 'D0')) %>% 
+  mutate(comparison = 'Day 0\nvs\nDay 10')
+
+bind_rows(donor_v_donor, donor_v_recipent, corecipient, other_donors, 
+          other_recipients, day0_v_day10, donor_v_recipent_end) %>% 
   mutate(comparison = factor(comparison, levels = 
-                              c('Donor\nvs\nRecipient', 'Recipients\ncompared to\nothers w/same donor', 
-                                'Donor\nvs\nNon-recipients', 'Recipients\ncompared to\nothers w/different donor'))) %>% 
+                              c('Donor\nvs\nDonor',
+                                'Donor\nvs\nRecipient', 
+                                'Recipients\ncompared to\nothers w/same donor', 
+                                'Day 0\nvs\nDay 10',
+                                'Donor\nvs\nRecipient End',
+                                'Donor\nvs\nNon-recipients', 
+                                'Recipients\ncompared to\nothers w/different donor'))) %>% 
   ggplot(aes(x = comparison, y = distances)) + 
-    geom_boxplot() + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
+    geom_jitter(alpha = 0.1, width = 0.3) + 
     labs(x = NULL, y = 'Theta YC') +
     theme_bw()
+
+# how do donors distances look by donor?
+#donor_v_donor %>% 
+#  pivot_longer(cols = c(rows, columns)) %>% 
+#  ggplot(aes(x = value, y = distances, color = value)) + geom_point()
+#    geom_jitter(width = 0.25) + 
+#    scale_color_manual(values = donor_colors) %>% 
+#    theme_bw()
 
@@
   separate(group.y, c('cageB', 'mouseB', 'dayB')) %>% 
-  mutate(same_cage = cageA == cageB,) %>% 
+  mutate(same_cage = cageA == cageB) %>% 
   ggplot(aes(x = human_source, y = distances, color = human_source, shape = same_cage)) +
-    geom_jitter(aes(color = human_source), width = 0.2) + 
+    geom_point(position = position_jitterdodge(), width = 0.2) + 
     scale_color_manual(values = donor_colors) + 
@@
     guides(color = 'none')
-```
-
-```{r}
-#genus <- sum_otu_by_taxa(taxonomy_df = taxonomy,
-#                         otu_df = select(shared, Group = sample_id, contains('Otu')),
-#                         taxa_level = 'Genus', 
-#                         top_n = 12)
-#donor_genus <- sum_otu_by_taxa(taxonomy_df = donor_taxonomy,
-#                         otu_df = donor,
-#                         taxa_level = 'Genus', 
-#                         top_n = 12)
-#
-#donor_order <- donor_genus %>% 
-#    filter(taxa == 'Other') %>% 
-#    arrange(desc(abundance)) %>% 
-#    pull(Group)
-#
-#
-#donor_genus %>% 
-#  mutate(rel_abun = abundance/2779 * 100,
-#         taxa = factor(taxa,
-#                       levels = c("Alistipes", "Bacteroides", "Blautia", 
-#                                  "Clostridium_sensu_stricto", "Cronobacter",
-#                                  "Enterococcus", "Faecalibacterium", 
-#                                  "Lachnospiraceae_unclassified", "Lactobacillus", 
-#                                  "Prevotella", "Ruminococcaceae_unclassified", "Other")),
-#          Group = factor(Group, levels = donor_order)) %>% 
-#  ggplot(aes(x = Group, y = taxa, fill = rel_abun)) + 
-#    labs(x = NULL, y = NULL, fill = 'Relative Abundance') + 
-#    geom_tile() + 
-#    scale_fill_gradient(low = 'white', high = 'black')
-#    coord_flip()
-# 
-##    scale_fill_gradient2(low="white", mid='blue3', high = 'black',
-##  			limits = c(-2,2), na.value = NA, midpoint = .3,
-##  			breaks = c(-2.5, -1, 0, 1, 2), labels = c('', '0.1', '1', '10', '100'))
-
 ```
@@
     labs(x = 'Day', y = 'C. difficile CFU (log10)')
-
-metadata %>% 
-  filter(cdiff_strain == 431) %>% 
-  mutate(human_source = factor(human_source, levels = 
-                                c('DA01324', 'DA00953', 'DA00581', 'DA01134', 'DA10148', 
-                                  'DA00369', 'DA00430', 'DA10093', 'DA10027', 'DA10034',
-                                  'DA10082', 'DA00884', 'DA00578', 'DA00431', 'DA01245'))) %>% 
-  ggplot(aes(x = day, y = percent_weightLoss, color = human_source)) +
-    stat_summary(fun = 'median', geom = 'line') + 
-    geom_point(alpha = 0.3) +
-    facet_grid(human_source~.) + 
-    scale_color_manual(values = donor_colors) + 
-    guides(color = 'none') + 
-    theme_bw() + 
-    labs(x = 'Day', y = 'Weight (Relative to initial)')
-
 ```
-
 ```{r}
@@
     guides(color = 'none')
+```
 
+DA00581 is made up of 4 cages, and at day 1 we see 1 cage at ~10^7.5, 2 cages just above 10^5 and 1 cage at 0.
+For the cage at 0 there are three mice, one remains at 0, the other two have transient low appearance of cdiff  as seen below
+
+```{r}
+metadata %>% filter(human_source == 'DA00581') %>% 
+  ggplot(aes(x = day, y = log10(cdiff_cfu + 60), shape = mouse_id, color = cage_id)) +
+    geom_line() + 
+    geom_point() + 
+    scale_shape_manual(values = 1:11) + 
+    theme_bw() + 
+    guides(shape = 'none') + 
+    labs(x = 'Day', y = 'C. difficile CFU (log10)')
+
+```
+
+Points are added to allow identification of individuals. Here we see one mouse has low cfu on days 6 and 7, another mouse on day 10.
+
+# Figure supplemental - weight loss
+
+```{r fig.height = 14}
+metadata %>% 
+  filter(cdiff_strain == 431) %>% 
+  mutate(human_source = factor(human_source, levels = 
+                                c('DA01324', 'DA00953', 'DA00581', 'DA01134', 'DA10148', 
+                                  'DA00369', 'DA00430', 'DA10093', 'DA10027', 'DA10034',
+                                  'DA10082', 'DA00884', 'DA00578', 'DA00431', 'DA01245'))) %>% 
+  ggplot(aes(x = day, y = percent_weightLoss, color = human_source)) +
+    stat_summary(fun = 'median', geom = 'line') + 
+    geom_point(alpha = 0.3) +
+    facet_grid(human_source~.) + 
+    scale_color_manual(values = donor_colors) + 
+    guides(color = 'none') + 
+    theme_bw() + 
+    labs(x = 'Day', y = 'Weight (Relative to initial)')
+```
+
+```{r}
 metadata %>% 
@@
 ```{r}
-metadata %>% 
-  filter(day == 0 | day == mouse_endpoint) %>% 
-  mutate(time_point = ifelse(day == 0, 'Initial', 'Endpoint'),
-         time_point = factor(time_point, levels = c('Initial', 'Endpoint'))) %>% 
-  left_join(nmds_df, by = c('group')) %>% 
-  left_join(donor_by_outcome, by = c('human_source')) %>% 
-  ggplot(aes(x = axis1, y = axis2, color = early_euth)) +
-    geom_point() + 
-    facet_grid(.~time_point) +
-    #scale_color_manual(values  = donor_colors) + 
-    theme_bw() + 
-    labs(x = 'NMDS axis 1', y = 'NMDS axis 2', color = 'Moribund') + 
-    theme(legend.position = 'bottom')
+#metadata %>% 
+#  filter(day == 0 | day == mouse_endpoint) %>% 
+#  mutate(time_point = ifelse(day == 0, 'Initial', 'Endpoint'),
+#         time_point = factor(time_point, levels = c('Initial', 'Endpoint'))) %>% 
+#  left_join(nmds_df, by = c('group')) %>% 
+#  ggplot(aes(x = axis1, y = axis2, color = early_euth)) +
+#    geom_point() + 
+#    facet_grid(.~time_point) +
+#    #scale_color_manual(values  = donor_colors) + 
+#    theme_bw() + 
+#    labs(x = 'NMDS axis 1', y = 'NMDS axis 2', color = 'Moribund') + 
+#    theme(legend.position = 'bottom')
+#
+# not sure if this adds to this figure
 
@@
 > Last samples for moribund mice is from cecum
-> Switch to cage color?  
-> Recalc nmds?
 
@@
 ```{r histology}
+mouse_by_outcome <- histology %>% 
+  mutate(mouse_id = paste0(cage_id, '_', ear_tag)) %>% 
+  select(mouse_id, human_source, epithelial_damage, summary_score) %>% 
+  left_join(unique(select(metadata, mouse_id, early_euth, cdiff_strain),
+            by = c('mouse_id'))) %>% 
+  filter(cdiff_strain == 431) %>% 
+  group_by(human_source) %>% 
+  mutate(severity = mean(epithelial_damage) < 1) %>% 
+# using epithelial damage
+  mutate(outcome = case_when(early_euth == T ~ 'moribund',
+                             severity == T ~ 'mild',
+                             severity == F ~ 'severe',
+                             T ~ 'NA')) %>% 
+# using summary score
+#    mutate(outcome_ss = case_when(early_euth == T ~ 'moribund',
+#                             summary_score < 4.4 ~ 'mild', # mean = 4.4, median = 5
+#                             summary_score > 4.4 ~ 'severe',
+#                             T ~ 'NA')) %>% 
+  mutate(outcome = factor(outcome, levels = c('mild', 'severe', 'moribund'))) %>% 
+  select(mouse_id, outcome)
 
@@
   filter(!is.na(early_euth)) %>% 
-  mutate(early_euth = ifelse(early_euth, 'Moribund', 'Colonized')) %>% 
-  left_join(donor_by_outcome, by = c('human_source')) %>% 
-  #filter(outcome == 'mild', epithelial_damage > 0)
+  mutate(early_euth = ifelse(early_euth, 'Moribund', 'Colonized'),
+         mouse_id = paste0(cage_id, '_', ear_tag)) %>% 
+  inner_join(mouse_by_outcome, by = c('mouse_id')) %>%
   pivot_longer(cols = c(edema_tissue, inflammation_tissue, epithelial_damage, summary_score),
                names_to = 'histology', values_to = 'score') %>% 
-  ggplot(aes(x = histology, y = score, fill = outcome)) +
-    geom_boxplot() + 
+  ggplot(aes(x = histology, y = score, color = outcome)) +
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5), position = position_dodge(width = 0.4)) +
+    #geom_boxplot() + 
     #geom_point(position = position_jitterdodge(), alpha = 0.5) + 
@@
 
+> **Splitting by mean epithelial damage for donor < 1 = 12 mild, 15 severe, 17 moribund**  
+> Splitting of median summary score of 5 =  10 mild, 11 severe, 17 moribund  
+>        (6 have a summary score of 5)  
+> Splitting of mean summary score of 4.4 =  10 mild, 17 severe, 17 moribund
+
+ What is toxin distribution by outcome?
+
+```{r toxin}
+toxin %>% 
+  inner_join(select(metadata, group, human_source), by = c('group')) %>% 
+  left_join(mouse_by_outcome, by = c('human_source')) %>% 
+  filter(!is.na(outcome)) %>% 
+  mutate(day = as.numeric(toxin_sample_day)) %>% 
+  ggplot(aes(x = day, y = Log_repiricoal_dilution, color = outcome)) + 
+#    geom_line(aes(color = outcome, group = mouse_id)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5), position = position_dodge(width = 0.4)) +
+    labs(x = 'Day', y = 'Toxin activity') + 
+#    facet_grid(outcome~.) + 
+    theme_bw()
+```
+
+## Can we use toxin to split severity?
+do mice split into +/- toxin based on outcome of moribund/mild/severe?
+
+```{r}
+toxin %>% 
+  mutate(mouse_id = ifelse(toxin_sample_type == 'cecalcon', 
+                           gsub('C$', '', mouse_id), mouse_id)) %>% 
+  group_by(mouse_id) %>% 
+  summarise(max_toxin = max(Log_repiricoal_dilution),
+            mean_toxin = mean(Log_repiricoal_dilution)) %>% 
+  left_join(mouse_by_outcome, by = c('mouse_id')) %>% 
+  filter(!is.na(human_source)) %>% 
+  pivot_longer(c(max_toxin, mean_toxin)) %>% 
+  ggplot(aes(x = outcome, y = value)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
+    geom_jitter(alpha = 0.3, width = 0.3) + 
+    facet_wrap(.~name, scales = 'free_y') + 
+    labs(x = NULL, y = 'Log reciprocal toxin') + 
+    theme_bw()
+```
+
+## Does toxin explain histological findings?
+
+```{r}
+toxin %>% 
+  mutate(mouse_id = ifelse(toxin_sample_type == 'cecalcon', 
+                           gsub('C$', '', mouse_id), mouse_id)) %>% 
+  group_by(mouse_id) %>% 
+  mutate(last_sample = max(toxin_sample_day)) %>% 
+  filter(toxin_sample_day == last_sample) %>% 
+  left_join(select(histology, mouse_id, summary_score), by = c('mouse_id')) %>% 
+  left_join(mouse_by_outcome, by = c('mouse_id')) %>% 
+  filter(!is.na(summary_score)) %>% 
+  ggplot(aes(y = Log_repiricoal_dilution, x = summary_score)) + 
+    geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 0.5,
+                 aes(fill = as.factor(summary_score))) + 
+    #stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5),
+    #             aes(color = as.factor(summary_score))) +
+    #geom_jitter(width = 0.25, alpha = 0.5) + 
+    scale_color_manual(values = donor_colors) + 
+    scale_fill_manual(values = donor_colors) + 
+    theme_bw() + 
+    scale_x_continuous(breaks = 0:10) + 
+    guides(color = 'none', fill = 'none')
+
+toxin %>% 
+  mutate(mouse_id = ifelse(toxin_sample_type == 'cecalcon', 
+                           gsub('C$', '', mouse_id), mouse_id)) %>% 
+  group_by(mouse_id) %>% 
+  mutate(last_sample = max(toxin_sample_day)) %>% 
+  filter(toxin_sample_day == last_sample) %>% 
+  left_join(select(histology, mouse_id, epithelial_damage), by = c('mouse_id')) %>% 
+  left_join(mouse_by_outcome, by = c('mouse_id')) %>% 
+  filter(!is.na(epithelial_damage)) %>% 
+  ggplot(aes(y = Log_repiricoal_dilution, x = epithelial_damage, 
+             color = as.factor(epithelial_damage))) + 
+    geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 0.5,
+                 aes(fill = as.factor(epithelial_damage))) + 
+    #stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
+    #geom_jitter(width = 0.25, alpha = 0.5) + 
+    scale_color_manual(values = donor_colors) + 
+    scale_fill_manual(values = donor_colors) + 
+    theme_bw() + 
+    scale_x_continuous(breaks = 0:10) + 
+    guides(color = 'none', fill = 'none')
+
+# is this relevant?
+```
+
 Mice were challenged with a strain isolated that matched Clostridioides difficile ribotype 027
@@
 > what is published about histology of 027 in C57/B6 mice?   
-
-```{r toxin}
-
-toxin %>% 
-  filter(!is.na(ear_tag),
-         !is.na(toxin_sample_day)) %>% 
-  left_join(select(metadata, group, human_source), by = c('group')) %>% 
-  left_join(donor_by_outcome, by = c('human_source')) %>% 
-  filter(!is.na(outcome)) %>% 
-  mutate(day = as.numeric(toxin_sample_day)) %>% 
-  ggplot(aes(x = day, y = Log_repiricoal_dilution, fill = outcome)) + 
-#    geom_line(aes(color = outcome, group = mouse_id)) + 
-    geom_boxplot(aes(group = interaction(day, outcome))) +
-    labs(x = 'Day', y = 'Toxin activity') + 
-#    facet_grid(outcome~.) + 
-    theme_bw()
-# check for correlation between toxin and cfu
-#toxin %>% 
-#  filter(!is.na(ear_tag),
-#         !is.na(toxin_sample_day)) %>% 
-#  left_join(select(metadata, group, cdiff_cfu), by = 'group') %>% 
-#  ggplot(aes(x = log10(cdiff_cfu + 0.01), y = Log_repiricoal_dilution)) + 
-#    geom_point(alpha = 0.2) +
-#    labs(x = 'CFU', y = 'Toxin activity') + 
-#    theme_bw()
-# does this matter?
-
-# how many samples tested are positive for toxin per mouse?
-#toxin %>% 
-#  group_by(mouse_id) %>% 
-#  summarise(n_samples = length(toxin_sample_day),
-#            n_positive = sum(Log_repiricoal_dilution > 1))
-# is this relevant?
-```
+> how much variation is there in individual isolates of RT027?
+> RT027/BI/NAP1, toxinotype III
+> VPI 10463 is ribotype 087, toxinotype 0
 
@@
 #  theme_bw()
+
+shared <- read_tsv('../../../data/mothur/sample.final.0.03.subsample.shared',
+         col_types = cols(.default = col_double(),
+                          Group = col_character())) %>% 
+  mutate(Group = ifelse(grepl('_D|DA', Group), Group,
+      gsub('D', '_D', Group))) %>% 
+  select(-label, -numOtus) %>% 
+  pivot_longer(-Group, names_to = 'otu', values_to = 'counts') %>% 
+  mutate(relative_abundance = counts/2107 * 100)
+
+# correlations
+# toxin to otu
+toxin_otu_corr <- metadata %>% 
+  inner_join(toxin, by = 'group') %>% 
+  filter(cdiff_strain == 431) %>% 
+  select(Group = group, Log_repiricoal_dilution) %>% 
+  left_join(shared, by = c('Group')) %>% 
+  left_join(taxonomy, by = c('otu' = 'OTU')) %>% 
+  filter(!is.na(otu)) %>% 
+  group_by(Group, Genus, Log_repiricoal_dilution) %>%
+  summarise(counts = sum(counts),
+            relative_abundance = sum(relative_abundance)) %>% 
+  group_by(Genus) %>% 
+  mutate(otu = Genus,
+        sd = sd(counts)) %>% 
+  filter(sd != 0) %>% 
+  nest() %>% 
+	mutate(spearman_p = map(.x = data, .f = ~ cor.test(.x$counts, .x$Log_repiricoal_dilution, 
+	                                                   method = 'spearman', exact = F)$p.value),
+	       spearman_rho = map(.x = data, .f = ~ cor.test(.x$counts, .x$Log_repiricoal_dilution, 
+	                                                     method = 'spearman', exact = F)$estimate)) %>% # compare cleared vs colonized
+	unnest(c(spearman_p, spearman_rho)) %>% 
+	mutate(pvalue = p.adjust(spearman_p, method = 'BH', n = nrow(.))) %>% # correct p values
+	filter(pvalue < 0.05) %>% # select only those above 0.05 after pvalue correction
+	unnest(data)
+
+toxin_otu_corr %>% 
+  filter(spearman_rho < -0.35 | spearman_rho > 0.35) %>% 
+  mutate(relative_abundance = ifelse(relative_abundance == 0, 0.045, relative_abundance)) %>% 
+  ggplot(aes(x = relative_abundance, y = Log_repiricoal_dilution, color = otu)) + 
+    geom_point(alpha = 0.3) + 
+    theme_bw() + 
+    facet_wrap(otu~.) + 
+    scale_x_log10() + 
+    guides(color = 'none') + 
+    geom_smooth(method = 'lm', se=F) + 
+    labs(x = 'Relative Abundance', y = 'Toxin (log reciprocal dilution)')
+
 ```
+
+```{r}
+# correlation of histology to otu
+plot_hist_corr <- function(i){
+  metadata %>% 
+    filter(cdiff_strain == 431,
+           day == mouse_endpoint) %>% 
+    inner_join(histology, by = 'mouse_id') %>% 
+    filter(early_euth == F) %>% 
+    select(Group = group, test = .data[[i]]) %>% 
+    left_join(shared, by = c('Group')) %>% 
+    group_by(otu) %>%
+    mutate(sd = sd(counts)) %>% 
+    filter(!is.na(otu),
+           sd != 0) %>% 
+    nest() %>% 
+  	mutate(spearman_p = map(.x = data, .f = ~ cor.test(.x$counts, .x$test, 
+  	                                                   method = 'spearman', exact = F)$p.value),
+  	       spearman_rho = map(.x = data, .f = ~ cor.test(.x$counts, .x$test, 
+  	                                                     method = 'spearman', exact = F)$estimate)) %>% # compare cleared vs colonized
+  	unnest(c(spearman_p, spearman_rho)) %>% 
+  	mutate(pvalue = p.adjust(spearman_p, method = 'BH', n = nrow(.))) %>% # correct p values
+  	filter(pvalue < 0.05) %>% # select only those above 0.05 after pvalue correction
+  	unnest(data) %>% 
+    left_join(taxonomy, by = c('otu' = 'OTU')) %>% 
+    mutate(tax_otu_label = gsub(' \\(', '\n\\(', tax_otu_label)) %>% 
+    ggplot(aes(x = relative_abundance, y = test, color = tax_otu_label)) + 
+      geom_point(alpha = 0.3) +
+      geom_smooth(method = 'lm', se = F) +
+      facet_wrap(tax_otu_label ~ .) + 
+      theme_bw() + 
+      labs(x = 'Relative Abundance', y = i, title = i) + 
+      guides(color = 'none')
+}
+
+plot_hist_corr('summary_score')
+plot_hist_corr('edema_tissue')
+plot_hist_corr('inflammation_tissue')
+plot_hist_corr('epithelial_damage')
+```
+```{r}
+# correlation of histology to otu
+plot_hist_corr <- function(i){
+  metadata %>% 
+    filter(cdiff_strain == 431,
+           day == mouse_endpoint) %>% 
+    inner_join(histology, by = 'mouse_id') %>% 
+    filter(early_euth == F) %>% 
+    select(Group = group, test = .data[[i]]) %>% 
+    left_join(shared, by = c('Group')) %>% 
+    group_by(otu) %>%
+    mutate(sd = sd(counts)) %>% 
+    filter(!is.na(otu),
+           sd != 0) %>% 
+    nest() %>% 
+  	mutate(spearman_p = map(.x = data, .f = ~ cor.test(.x$counts, .x$test, 
+  	                                                   method = 'spearman', exact = F)$p.value),
+  	       spearman_rho = map(.x = data, .f = ~ cor.test(.x$counts, .x$test, 
+  	                                                     method = 'spearman', exact = F)$estimate)) %>% # compare cleared vs colonized
+  	unnest(c(spearman_p, spearman_rho)) %>% 
+  	mutate(pvalue = p.adjust(spearman_p, method = 'BH', n = nrow(.))) %>% # correct p values
+  	filter(pvalue < 0.05) %>% # select only those above 0.05 after pvalue correction
+  	unnest(data) %>% 
+    left_join(taxonomy, by = c('otu' = 'OTU')) %>% 
+    mutate(tax_otu_label = gsub(' \\(', '\n\\(', tax_otu_label)) %>% 
+    ggplot(aes(x = relative_abundance, y = test, color = tax_otu_label)) + 
+      geom_point(alpha = 0.3) +
+      geom_smooth(method = 'lm', se = F) +
+      facet_wrap(tax_otu_label ~ .) + 
+      theme_bw() + 
+      labs(x = 'Relative Abundance', y = i, title = i) + 
+      guides(color = 'none')
+}
+
+plot_hist_corr('summary_score')
+plot_hist_corr('edema_tissue')
+plot_hist_corr('inflammation_tissue')
+plot_hist_corr('epithelial_damage')
+```
+
+```{r}
+metadata <- read_tsv('../../../data/process/metadata_tidy.tsv',
+		col_type = 'cccccDddddDcdddddcdl') 
+shared <- read_tsv('../../../data/mothur/sample.final.0.03.subsample.shared',
+		col_types = cols(.default = col_double(),
+			Group = col_character())) %>% 
+	mutate(Group = ifelse(grepl('_D|DA', Group), Group,
+		gsub('D', '_D', Group)))
+taxonomy <- read_tsv('../../../data/process/final.taxonomy.tidy.tsv',
+                     col_types = cols(.default = col_character()))
+
+lefse_df <- data.table::fread(paste0('../../../data/process/severe_disease_Genus.0.03.lefse_summary'),
+		fill = T) %>% 
+	filter(!is.na(LDA)) %>% 
+	top_n(20, LDA) %>% 
+	mutate(order = LDA + 10 * as.logical(Class),
+		tax_otu_label = OTU)
+
+lefse_plot <- lefse_df %>% 
+	ggplot(aes(y = LDA, fill = Class, 
+			x = reorder(tax_otu_label, order))) +
+		geom_col() + 
+		theme_bw() + 
+		coord_flip() + 
+		guides(fill = 'none') + 
+		scale_fill_manual(values = c(1,2),
+			limits = c(TRUE, FALSE),
+			labels = c('Severe', 'Non-severe')) + 
+		labs(x = NULL, y = 'LDA Score (log 10)')
+
+abundance_plot <- metadata %>% 
+	filter(cdiff_strain == 431,
+		day == 0) %>% 
+	left_join(shared, by = c('group' = 'Group')) %>% 
+	gather(OTU, abundance, matches('Otu\\d*')) %>% 
+	left_join(taxonomy, by = 'OTU') %>% 
+	right_join(lefse_df, by = c('Genus' = 'OTU')) %>% 
+	group_by(group, Genus, order, early_euth) %>% 
+	summarise(abundance = sum(abundance)) %>% 
+	mutate(tax_otu_label = Genus) %>% 
+	mutate(abundance = abundance / 21.07,
+		abundance = ifelse(abundance == 0, 0.045, abundance)) %>% 
+	ggplot(aes(x = reorder(tax_otu_label, order), 
+			y = abundance, color = as.factor(early_euth))) + 
+		geom_jitter(position = position_jitterdodge(dodge.width = 0.5), alpha = 0.25) + 
+		stat_summary(fun.data = 'median_hilow', aes(group = early_euth),
+			fun.args = (conf.int=0.5), position = position_dodge(width = 0.5)) +
+		theme_bw() + labs(x = NULL, y = 'Relative Abundance', color = NULL) + 
+		scale_y_log10(breaks=c(0.01, 0.1, 1, 10, 100),
+			labels=c("0","0.1","1","10","100")) +
+		scale_color_manual(values = c(1,2),
+						   limits = c(TRUE, FALSE),
+						   labels = c('Severe', 'Non-severe')) + 
+		geom_hline(yintercept = 0.047, linetype = 'dashed', lwd = 0.3) + 
+		theme(panel.grid.major.y = element_blank(), 
+			panel.grid.minor = element_blank(),
+			axis.ticks.y = element_blank(),
+			axis.text.y = element_blank(),
+			legend.position = 'right') + 
+		coord_flip()
+
+cowplot::plot_grid(lefse_plot, abundance_plot, nrow = 1)
+
+```

### 2021-04-27 exploratory/notebook/initial_analysis/initial_analysis_notebook.Rmd
add lefse and interactive feature plot

@@
 date: "4/13/2021"
-output: rmarkdown::github_document
+always_allow_html: true
+output: 
+  html_document: default
 ---
@@
 ```{r}
+library(plotly)
 metadata <- read_tsv('../../../data/process/metadata_tidy.tsv',
@@
 		guides(fill = 'none') + 
-		scale_fill_manual(values = c(1,2),
+		scale_fill_manual(values = c('#8de4d3','#f75b5f'),
 			limits = c(TRUE, FALSE),
@@
 	right_join(lefse_df, by = c('Genus' = 'OTU')) %>% 
-	group_by(group, Genus, order, early_euth) %>% 
+	group_by(mouse_id, group, Genus, order, early_euth) %>% 
 	summarise(abundance = sum(abundance)) %>% 
@@
 		abundance = ifelse(abundance == 0, 0.045, abundance)) %>% 
+  highlight_key(~mouse_id) %>% 
 	ggplot(aes(x = reorder(tax_otu_label, order), 
-			y = abundance, color = as.factor(early_euth))) + 
-		geom_jitter(position = position_jitterdodge(dodge.width = 0.5), alpha = 0.25) + 
+			y = abundance, color = as.factor(early_euth), shape = mouse_id)) + 
 		stat_summary(fun.data = 'median_hilow', aes(group = early_euth),
@@
 			labels=c("0","0.1","1","10","100")) +
-		scale_color_manual(values = c(1,2),
+		scale_color_manual(values = c('#8de4d3','#f75b5f'),
 						   limits = c(TRUE, FALSE),
 						   labels = c('Severe', 'Non-severe')) + 
+    scale_shape_manual(values = rep(16, 54)) + 
 		geom_hline(yintercept = 0.047, linetype = 'dashed', lwd = 0.3) + 
@@
 			panel.grid.minor = element_blank(),
-			axis.ticks.y = element_blank(),
-			axis.text.y = element_blank(),
 			legend.position = 'right') + 
-		coord_flip()
+		coord_flip() + 
+    guides(shape = 'none')
 
-cowplot::plot_grid(lefse_plot, abundance_plot, nrow = 1)
+cowplot::plot_grid(lefse_plot, 
+                   abundance_plot + 
+                     geom_jitter(position = position_jitterdodge(dodge.width = 0.5), alpha = 0.25) + 
+                     theme(axis.ticks.y = element_blank(),
+                           axis.text.y = element_blank()), 
+                   nrow = 1)
 
+gg <- ggplotly(abundance_plot + geom_jitter(position = position_jitterdodge(dodge.width = 0.5)), tooltip = c('mouse_id'))
+highlight(hide_legend(gg), on = "plotly_click", off = "plotly_doubleclick", 
+           color = "black", selectize = T, opacityDim = .2,
+           selected = attrs_selected(showlegend = F, marker = list(symbol = 'X')))
+str(gg)
 ```

### 2021-04-27 exploratory/notebook/initial_analysis/initial_analysis_notebook.Rmd
change back to github doc
combine initial communities with cowplot

@@
 date: "4/13/2021"
-always_allow_html: true
-output: 
-  html_document: default
+output: rmarkdown::github_document
 ---
@@
 Unique communities without conserved structure
-```{r}
-shared %>% 
+```{r, fig.height= 6, fig.width=6}
+donor_plot <- shared %>% 
   filter(grepl('DA', Group)) %>% 
@@
     scale_x_discrete(guide = guide_axis(angle = 45)) + 
-    facet_grid(.~Group, scales = 'free_x')
-```
+    facet_grid(.~Group, scales = 'free_x') + 
+    theme(axis.ticks.x = element_blank(),
+          axis.text.x = element_blank(),
+          legend.position = 'none')
 
-Donor communities colonize mice and shift community, ~~but consistently~~
-
-```{r fig.width = 10}
-shared %>% 
-  filter(grepl('D0', Group)) %>% 
-  left_join(select(metadata, group, human_source, mouse_id), by = c('Group' = 'group')) %>% 
+day_0_plot <- metadata %>% 
+  filter(day == 0,
+         cdiff_strain == 431) %>% 
+  select(group, human_source, mouse_id) %>% 
+  left_join(shared, by = c('group' = 'Group')) %>% 
   left_join(taxonomy, by = c('otu' = 'OTU')) %>% 
   rename(taxa_level = Class) %>% 
-  group_by(Group, taxa_level, human_source) %>% 
+  group_by(group, taxa_level, human_source) %>% 
   summarise(relative_abundance = log10(sum(relative_abundance) + 0.0001)) %>% 
@@
   mutate(median_ra = median(relative_abundance)) %>% 
-  ggplot(aes(x = Group, 
+  filter(!is.na(taxa_level)) %>% 
+  ggplot(aes(x = group, 
              y = reorder(taxa_level, median_ra), 
@@
     theme(axis.ticks.x = element_blank(),
-          axis.text.x = element_blank())
+          axis.text.x = element_blank(),
+          legend.position = 'bottom')
 
+cowplot::plot_grid(
+  cowplot::plot_grid(NULL, donor_plot, nrow = 1, rel_widths = c(.95,13)),
+  day_0_plot, ncol = 1, rel_heights = c(8,9))
 ```

### 2021-04-27 exploratory/notebook/initial_analysis/initial_analysis_notebook.Rmd
convert temporal cfu/weight loss plots to single line plots

@@
 
-```{r fig.height = 14}
+```{r}
 metadata %>% 
@@
     geom_point(alpha = 0.3) +
-    facet_grid(human_source~.) + 
     scale_color_manual(values = donor_colors) + 

### 2021-04-27 exploratory/notebook/initial_analysis/initial_analysis_notebook.Rmd
plot histology/toxin by moribund not mild/moderate/severe

@@
                names_to = 'histology', values_to = 'score') %>% 
-  ggplot(aes(x = histology, y = score, color = outcome)) +
-    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5), position = position_dodge(width = 0.4)) +
-    #geom_boxplot() + 
-    #geom_point(position = position_jitterdodge(), alpha = 0.5) + 
+  ggplot(aes(x = early_euth, y = score, , color = early_euth, fill = early_euth)) +
+    #stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5), position = position_dodge(width = 0.4)) +
+    geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 0.5) + 
+    #geom_point(position = position_jitterdodge(jitter.width = 0.3), alpha = 0.5) + 
     theme_bw() + 
@@
     scale_x_discrete(guide = guide_axis(angle = 45)) + 
-    facet_grid(.~early_euth)
+    facet_grid(.~histology) + 
+    guides(color = 'none') + 
+    theme(legend.position = 'bottom')
 ```
@@
   filter(!is.na(outcome)) %>% 
-  mutate(day = as.numeric(toxin_sample_day)) %>% 
-  ggplot(aes(x = day, y = Log_repiricoal_dilution, color = outcome)) + 
-#    geom_line(aes(color = outcome, group = mouse_id)) + 
+  mutate(day = as.numeric(toxin_sample_day),
+         outcome = ifelse(outcome == 'moribund', as.character(outcome), 'colonized')) %>% 
+  ggplot(aes(x = outcome, y = Log_repiricoal_dilution, fill = outcome, color = outcome)) + 
     stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5), position = position_dodge(width = 0.4)) +
+    geom_jitter(width= 0.3, height = 0, alpha = 0.3) + 
     labs(x = 'Day', y = 'Toxin activity') + 
-#    facet_grid(outcome~.) + 
-    theme_bw()
+    facet_grid(.~day) + 
+    theme_bw() +
+    theme(legend.position = 'bottom')
 ```

### 2021-04-27 exploratory/notebook/initial_analysis/initial_analysis_notebook.Rmd
remove hist v toxin

@@
 
+```{r}
 ## Does toxin explain histological findings?
-
-```{r}
-toxin %>% 
-  mutate(mouse_id = ifelse(toxin_sample_type == 'cecalcon', 
-                           gsub('C$', '', mouse_id), mouse_id)) %>% 
-  group_by(mouse_id) %>% 
-  mutate(last_sample = max(toxin_sample_day)) %>% 
-  filter(toxin_sample_day == last_sample) %>% 
-  left_join(select(histology, mouse_id, summary_score), by = c('mouse_id')) %>% 
-  left_join(mouse_by_outcome, by = c('mouse_id')) %>% 
-  filter(!is.na(summary_score)) %>% 
-  ggplot(aes(y = Log_repiricoal_dilution, x = summary_score)) + 
-    geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 0.5,
-                 aes(fill = as.factor(summary_score))) + 
-    #stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5),
-    #             aes(color = as.factor(summary_score))) +
-    #geom_jitter(width = 0.25, alpha = 0.5) + 
-    scale_color_manual(values = donor_colors) + 
-    scale_fill_manual(values = donor_colors) + 
-    theme_bw() + 
-    scale_x_continuous(breaks = 0:10) + 
-    guides(color = 'none', fill = 'none')
-
-toxin %>% 
-  mutate(mouse_id = ifelse(toxin_sample_type == 'cecalcon', 
-                           gsub('C$', '', mouse_id), mouse_id)) %>% 
-  group_by(mouse_id) %>% 
-  mutate(last_sample = max(toxin_sample_day)) %>% 
-  filter(toxin_sample_day == last_sample) %>% 
-  left_join(select(histology, mouse_id, epithelial_damage), by = c('mouse_id')) %>% 
-  left_join(mouse_by_outcome, by = c('mouse_id')) %>% 
-  filter(!is.na(epithelial_damage)) %>% 
-  ggplot(aes(y = Log_repiricoal_dilution, x = epithelial_damage, 
-             color = as.factor(epithelial_damage))) + 
-    geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 0.5,
-                 aes(fill = as.factor(epithelial_damage))) + 
-    #stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
-    #geom_jitter(width = 0.25, alpha = 0.5) + 
-    scale_color_manual(values = donor_colors) + 
-    scale_fill_manual(values = donor_colors) + 
-    theme_bw() + 
-    scale_x_continuous(breaks = 0:10) + 
-    guides(color = 'none', fill = 'none')
-
-# is this relevant?
+#toxin %>% 
+#  mutate(mouse_id = ifelse(toxin_sample_type == 'cecalcon', 
+#                           gsub('C$', '', mouse_id), mouse_id)) %>% 
+#  group_by(mouse_id) %>% 
+#  mutate(last_sample = max(toxin_sample_day)) %>% 
+#  filter(toxin_sample_day == last_sample) %>% 
+#  left_join(select(histology, mouse_id, summary_score), by = c('mouse_id')) %>% 
+#  left_join(mouse_by_outcome, by = c('mouse_id')) %>% 
+#  filter(!is.na(summary_score)) %>% 
+#  ggplot(aes(y = Log_repiricoal_dilution, x = summary_score)) + 
+#    geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 0.5,
+#                 aes(fill = as.factor(summary_score))) + 
+#    #stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5),
+#    #             aes(color = as.factor(summary_score))) +
+#    #geom_jitter(width = 0.25, alpha = 0.5) + 
+#    scale_color_manual(values = donor_colors) + 
+#    scale_fill_manual(values = donor_colors) + 
+#    theme_bw() + 
+#    scale_x_continuous(breaks = 0:10) + 
+#    guides(color = 'none', fill = 'none')
+#
+#toxin %>% 
+#  mutate(mouse_id = ifelse(toxin_sample_type == 'cecalcon', 
+#                           gsub('C$', '', mouse_id), mouse_id)) %>% 
+#  group_by(mouse_id) %>% 
+#  mutate(last_sample = max(toxin_sample_day)) %>% 
+#  filter(toxin_sample_day == last_sample) %>% 
+#  left_join(select(histology, mouse_id, epithelial_damage), by = c('mouse_id')) %>% 
+#  left_join(mouse_by_outcome, by = c('mouse_id')) %>% 
+#  filter(!is.na(epithelial_damage)) %>% 
+#  ggplot(aes(y = Log_repiricoal_dilution, x = epithelial_damage, 
+#             color = as.factor(epithelial_damage))) + 
+#    geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 0.5,
+#                 aes(fill = as.factor(epithelial_damage))) + 
+#    #stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
+#    #geom_jitter(width = 0.25, alpha = 0.5) + 
+#    scale_color_manual(values = donor_colors) + 
+#    scale_fill_manual(values = donor_colors) + 
+#    theme_bw() + 
+#    scale_x_continuous(breaks = 0:10) + 
+#    guides(color = 'none', fill = 'none')
+#
+# not applicable to our story
 ```

### 2021-04-27 exploratory/notebook/initial_analysis/initial_analysis_notebook.Rmd
remove interactive plot

@@
 ```{r}
-library(plotly)
+
 metadata <- read_tsv('../../../data/process/metadata_tidy.tsv',
@@
                    nrow = 1)
+#library(plotly)
+#gg <- ggplotly(abundance_plot + geom_jitter(position = position_jitterdodge(dodge.width = 0.5)), tooltip = c('mouse_id'))
+#highlight(hide_legend(gg), on = "plotly_click", off = "plotly_doubleclick", 
+#           color = "black", selectize = T, opacityDim = .2,
+#           selected = attrs_selected(showlegend = F, marker = list(symbol = 'X')))
 
-gg <- ggplotly(abundance_plot + geom_jitter(position = position_jitterdodge(dodge.width = 0.5)), tooltip = c('mouse_id'))
-highlight(hide_legend(gg), on = "plotly_click", off = "plotly_doubleclick", 
-           color = "black", selectize = T, opacityDim = .2,
-           selected = attrs_selected(showlegend = F, marker = list(symbol = 'X')))
-str(gg)
 ```
+## Severe disease
+```{r}
+plot_lefse_data('severe_disease')
+```
+```

### 2021-04-27 exploratory/notebook/initial_analysis/initial_analysis_notebook.Rmd
setup lefse plot for other comparisons

@@
 
-lefse_df <- data.table::fread(paste0('../../../data/process/severe_disease_Genus.0.03.lefse_summary'),
+plot_lefse_data <- function(file_name){
+    lefse_design <- data.table::fread(paste0('../../../data/process/', file_name, '_Genus.design'))
+    colnames(lefse_design) <- c('Group', 'class')
+
+    lefse_df <- data.table::fread(paste0('../../../data/process/', file_name, '_Genus.0.03.lefse_summary'),
   		fill = T) %>% 
@@
   		guides(fill = 'none') + 
-		scale_fill_manual(values = c('#8de4d3','#f75b5f'),
-			limits = c(TRUE, FALSE),
-			labels = c('Severe', 'Non-severe')) + 
+  		scale_fill_manual(values = c('#8de4d3','#f75b5f')) + 
   		labs(x = NULL, y = 'LDA Score (log 10)')
 
-abundance_plot <- metadata %>% 
-	filter(cdiff_strain == 431,
-		day == 0) %>% 
-	left_join(shared, by = c('group' = 'Group')) %>% 
+  abundance_plot <- shared %>% 
   	gather(OTU, abundance, matches('Otu\\d*')) %>% 
@@
   	right_join(lefse_df, by = c('Genus' = 'OTU')) %>% 
-	group_by(mouse_id, group, Genus, order, early_euth) %>% 
+    left_join(lefse_design, by = c('Group')) %>% 
+  	group_by(Group, Genus, order, class) %>% # removed mouse_id from interactive plot
   	summarise(abundance = sum(abundance)) %>% 
@@
   		abundance = ifelse(abundance == 0, 0.045, abundance)) %>% 
-  highlight_key(~mouse_id) %>% 
+    #highlight_key(~mouse_id) %>% 
   	ggplot(aes(x = reorder(tax_otu_label, order), 
-			y = abundance, color = as.factor(early_euth), shape = mouse_id)) + 
-		stat_summary(fun.data = 'median_hilow', aes(group = early_euth),
+  			y = abundance, color = class)) + # , shape = mouse_id remove mouse_id from interactive plot
+  		stat_summary(fun.data = 'median_hilow', aes(group = class),
   			fun.args = (conf.int=0.5), position = position_dodge(width = 0.5)) +
@@
   			labels=c("0","0.1","1","10","100")) +
-		scale_color_manual(values = c('#8de4d3','#f75b5f'),
-						   limits = c(TRUE, FALSE),
-						   labels = c('Severe', 'Non-severe')) + 
+  		scale_color_manual(values = c('#8de4d3','#f75b5f')) + 
       scale_shape_manual(values = rep(16, 54)) + 
@@
                      nrow = 1)
+}
+
 #library(plotly)
@@
 ```
+
+## Toxin presence
+```{r}
+plot_lefse_data('toxin_presence')
 ```
+
+## Summary score
+```{r}
+plot_lefse_data('summary_score_hilo')
+```
+
+## Epithelial damage
+```{r}
+plot_lefse_data('epithelial_damage_presence')
+```

### 2021-05-11 exploratory/notebook/initial_analysis/initial_analysis_notebook.Rmd
add alpha temporal plot

@@
                      col_types = cols(.default = col_character()))
+alpha_df <- read_tsv('../../../data/mothur/sample.final.groups.ave-std.summary',
+         col_types ='dccdddddd') %>% 
+  filter(method == 'ave') %>% 
+  mutate(group = ifelse(grepl('_D|DA', group), group,
+                        gsub('D', '_D', group)))
 nmds_df <- read_tsv('../../../data/mothur/sample.final.thetayc.0.03.lt.ave.nmds.axes',
@@
 ```{r}
+alpha_df <- alpha_df %>% 
+  select(group, sobs, invsimpson) %>% 
+  left_join(bind_rows(metadata, 
+                      tibble(group = unique(metadata$human_source),
+                             human_source = unique(metadata$human_source),
+                             sample_type = 'donor')),
+            by = c('group')) %>% 
+  mutate(sample = case_when(sample_type == 'donor' ~ 'Donor',
+                            day %in% c(0:10) ~ paste('Day', day),
+                            day == mouse_endpoint ~ 'Endpoint',
+                            T ~ 'NA'),
+         sample = factor(sample, levels = c('Donor', 'Day 0', 'Day 1', 'Day 2', 'Day 3', 
+                                            'Day 4', 'Day 5', 'Day 6', 'Day 7', 'Day 8', 'Day 9', 'Day 10'))) %>% 
+  filter(sample != 'NA')
+
+alpha_df %>% 
+  ggplot(aes(x = sample, y = invsimpson, color = human_source)) + 
+    geom_point(alpha = 0.3) + 
+    theme_bw() + 
+    facet_wrap(human_source~.) + 
+    guides(color = 'none')
+
+alpha_df %>% 
+  ggplot(aes(x = sample, y = sobs, color = human_source)) + 
+    geom_point(alpha = 0.3) + 
+    theme_bw() + 
+    facet_wrap(human_source~.) + 
+    guides(color = 'none')
+
 nmds_df %>% 

### 2021-05-11 exploratory/notebook/initial_analysis/initial_analysis_notebook.Rmd
remove mild/severe categorization

@@
     theme(legend.position = 'bottom')
+
+#if trying to split colonized
+#> **Splitting by mean epithelial damage for donor < 1 = 12 mild, 15 severe, 17 moribund**  
+#> Splitting of median summary score of 5 =  10 mild, 11 severe, 17 moribund  
+#>        (6 have a summary score of 5)  
+#> Splitting of mean summary score of 4.4 =  10 mild, 17 severe, 17 moribund
+
 ```
-
-> **Splitting by mean epithelial damage for donor < 1 = 12 mild, 15 severe, 17 moribund**  
-> Splitting of median summary score of 5 =  10 mild, 11 severe, 17 moribund  
->        (6 have a summary score of 5)  
-> Splitting of mean summary score of 4.4 =  10 mild, 17 severe, 17 moribund
 
@@
 
+```{r}
 ## Can we use toxin to split severity?
-do mice split into +/- toxin based on outcome of moribund/mild/severe?
+#do mice split into +/- toxin based on outcome of moribund/mild/severe?
 
-```{r}
-toxin %>% 
-  mutate(mouse_id = ifelse(toxin_sample_type == 'cecalcon', 
-                           gsub('C$', '', mouse_id), mouse_id)) %>% 
-  group_by(mouse_id) %>% 
-  summarise(max_toxin = max(Log_repiricoal_dilution),
-            mean_toxin = mean(Log_repiricoal_dilution)) %>% 
-  left_join(mouse_by_outcome, by = c('mouse_id')) %>% 
-  filter(!is.na(human_source)) %>% 
-  pivot_longer(c(max_toxin, mean_toxin)) %>% 
-  ggplot(aes(x = outcome, y = value)) + 
-    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
-    geom_jitter(alpha = 0.3, width = 0.3) + 
-    facet_wrap(.~name, scales = 'free_y') + 
-    labs(x = NULL, y = 'Log reciprocal toxin') + 
-    theme_bw()
+#toxin %>% 
+#  mutate(mouse_id = ifelse(toxin_sample_type == 'cecalcon', 
+#                           gsub('C$', '', mouse_id), mouse_id)) %>% 
+#  group_by(mouse_id) %>% 
+#  summarise(max_toxin = max(Log_repiricoal_dilution),
+#            mean_toxin = mean(Log_repiricoal_dilution)) %>% 
+#  left_join(mouse_by_outcome, by = c('mouse_id')) %>% 
+#  filter(!is.na(human_source)) %>% 
+#  pivot_longer(c(max_toxin, mean_toxin)) %>% 
+#  ggplot(aes(x = outcome, y = value)) + 
+#    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
+#    geom_jitter(alpha = 0.3, width = 0.3) + 
+#    facet_wrap(.~name, scales = 'free_y') + 
+#    labs(x = NULL, y = 'Log reciprocal toxin') + 
+#    theme_bw()
 ```

### 2021-05-11 exploratory/notebook/initial_analysis/initial_analysis_notebook.Rmd
setup correlations

@@
 
-```{r, individual cfu plots}
+```{r}
+#does cdiff correlate with toxin or histology?
+
+metadata %>% 
+  inner_join(toxin, by = 'group') %>% 
+  filter(cdiff_strain == 431,
+         cdiff_cfu != 0) %>% 
+  mutate(cfu = log10(cdiff_cfu)) %>% 
+	mutate(spearman_p = cor.test(cfu, Log_repiricoal_dilution, 
+	                             method = 'spearman', exact = F)$p.value,
+	       spearman_rho = cor.test(cfu, Log_repiricoal_dilution, 
+	                               method = 'spearman', exact = F)$estimate)  %>% 
+  ggplot(aes(x = cfu, y = Log_repiricoal_dilution)) + 
+    geom_point(alpha = 0.3) + 
+    theme_bw() + 
+    guides(color = 'none') + 
+    geom_smooth(method = 'lm', se=F) + 
+    labs(x = 'C. difficile CFU (log 10)', y = 'Toxin (log reciprocal dilution)')
+
+metadata %>% 
+  filter(day == mouse_endpoint,
+         cdiff_strain == 431) %>% 
+  left_join(histology, by = c('mouse_id')) %>% 
+  mutate(cfu = log10(cdiff_cfu)) %>% 
+  pivot_longer(cols = c(edema_tissue, inflammation_tissue, epithelial_damage, summary_score),
+               names_to = 'hist_score', values_to = 'score') %>% 
+  group_by(hist_score) %>% 
+  nest() %>% 
+	mutate(spearman_p = map(.x = data, .f = ~ cor.test(.x$cfu, .x$score, 
+	                                                   method = 'spearman', exact = F)$p.value),
+	       spearman_rho = map(.x = data, .f = ~ cor.test(.x$cfu, .x$score, 
+	                                                     method = 'spearman', exact = F)$estimate)) %>%
+	unnest(c(spearman_p, spearman_rho)) %>% 
+	mutate(pvalue = p.adjust(spearman_p, method = 'BH', n = 5)) %>% 
+	unnest(data) %>% 
+  ggplot(aes(x = cfu, y = score)) + 
+    geom_point(alpha = 0.3) + 
+    theme_bw() + 
+    guides(color = 'none') + 
+    geom_smooth(method = 'lm', se=F) + 
+    labs(x = 'C. difficile CFU (log 10)', y = 'Histology Score') + 
+    facet_wrap(hist_score~., scale = 'free_y')
+
+```
+
+Correlations between CFU and histology score not significant after BH correction.
+
+```{r}
 #metadata %>% 
@@
 	       spearman_rho = map(.x = data, .f = ~ cor.test(.x$counts, .x$Log_repiricoal_dilution, 
-	                                                     method = 'spearman', exact = F)$estimate)) %>% # compare cleared vs colonized
+	                                                     method = 'spearman', exact = F)$estimate)) %>%
 	unnest(c(spearman_p, spearman_rho)) %>% 
-	mutate(pvalue = p.adjust(spearman_p, method = 'BH', n = nrow(.))) %>% # correct p values
+	mutate(ncomparisons = nrow(.),
+  	  pvalue = p.adjust(spearman_p, method = 'BH', n = ncomparisons)) %>% # correct p values
 	filter(pvalue < 0.05) %>% # select only those above 0.05 after pvalue correction
@@
     theme_bw() + 
-    facet_wrap(otu~.) + 
+    facet_wrap(otu~., scales = 'free_x') + 
     scale_x_log10() + 

### 2021-05-11 exploratory/notebook/initial_analysis/initial_analysis_notebook.Rmd
edit lefse comparisons

@@
 
-  lefse_plot <- lefse_df %>% 
-  	ggplot(aes(y = LDA, fill = Class, 
-  			x = reorder(tax_otu_label, order))) +
-  		geom_col() + 
-  		theme_bw() + 
-  		coord_flip() + 
-  		guides(fill = 'none') + 
-  		scale_fill_manual(values = c('#8de4d3','#f75b5f')) + 
-  		labs(x = NULL, y = 'LDA Score (log 10)')
-
-  abundance_plot <- shared %>% 
+  shared %>% 
   	gather(OTU, abundance, matches('Otu\\d*')) %>% 
@@
   	right_join(lefse_df, by = c('Genus' = 'OTU')) %>% 
-    left_join(lefse_design, by = c('Group')) %>% 
+    right_join(lefse_design, by = c('Group')) %>% 
   	group_by(Group, Genus, order, class) %>% # removed mouse_id from interactive plot
@@
   			y = abundance, color = class)) + # , shape = mouse_id remove mouse_id from interactive plot
+  		geom_hline(yintercept = 0.047, linetype = 'dashed', lwd = 0.3) + 
       stat_summary(fun.data = 'median_hilow', aes(group = class),
   			fun.args = (conf.int=0.5), position = position_dodge(width = 0.5)) +
-  		theme_bw() + labs(x = NULL, y = 'Relative Abundance', color = NULL) + 
+  		theme_bw() + 
+      labs(x = NULL, y = 'Relative Abundance', color = NULL) + 
   		scale_y_log10(breaks=c(0.01, 0.1, 1, 10, 100),
   			labels=c("0","0.1","1","10","100")) +
-  		scale_color_manual(values = c('#8de4d3','#f75b5f')) + 
-      scale_shape_manual(values = rep(16, 54)) + 
-  		geom_hline(yintercept = 0.047, linetype = 'dashed', lwd = 0.3) + 
+  		#scale_color_manual(values = c('#8de4d3','#f75b5f')) + 
       theme(panel.grid.major.y = element_blank(), 
   			panel.grid.minor = element_blank(),
-  			legend.position = 'right') + 
+  			legend.position = 'bottom') + 
   		coord_flip() + 
-      guides(shape = 'none')
-
-  cowplot::plot_grid(lefse_plot, 
-                     abundance_plot + 
-                       geom_jitter(position = position_jitterdodge(dodge.width = 0.5), alpha = 0.25) + 
-                       theme(axis.ticks.y = element_blank(),
-                             axis.text.y = element_blank()), 
-                     nrow = 1)
+      geom_jitter(position = position_jitterdodge(dodge.width = 0.5), alpha = 0.2)
 }
@@
 
+## Toxin presence for mild outcome
+```{r}
+plot_lefse_data('toxin_mild')
+```
+
+## Toxin presence for severe outcome
+```{r fig.height = 4}
+plot_lefse_data('toxin_moribund')
+```
+
+## Toxin presence and outcome severity
+```{r, fig.height = 12}
+plot_lefse_data('toxin_severity')
+```
+
 ## Summary score

### 2021-05-26 code/run_ml.R
setup final ml run
using rf w/ 0.6 training fraction

@@ -22,7 +22,7 @@ library('mikropml')
 # get seed
 current_seed <- as.numeric(commandArgs(TRUE))
 set.seed <- current_seed
-fraction <- 0.8
+fraction <- 0.6
 
 # read in data
 print('Reading in data')
@@ -56,47 +56,47 @@ day_10_histology <- preprocess_data(day_10_histology,
 
 
 # run logistic regression
-new_hp <- list(alpha = 0,
-			   lambda = c(1e-2, 1e-1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1e0, 1e1, 1e2))
-print('Running Logistic Regression on same day toxin presence')
-same_day_toxin_lr <- run_ml(same_day_toxin,
-	   'glmnet',
-       outcome_colname = 'toxin',
-	   training_frac = fraction,
-       hyperparameters = new_hp,
-       #find_feature_importance = TRUE,
-       seed = current_seed)
-new_hp <- list(alpha = 0,
-			   lambda = c(1e-0, 1e1, 1e2, 2e2, 3e2, 4e2, 5e2, 6e2, 7e2, 8e2, 9e2, 1e3, 1e4))
-print('Running Logistic Regression on day 0 to predict toxin production')
-day_0_predict_future_toxin_lr <- run_ml(day_0_predict_future_toxin,
-	   'glmnet',
-       outcome_colname = 'toxin',
-       training_frac = fraction,
-	   hyperparameters = new_hp,
-	   #find_feature_importance = TRUE,
-       seed = current_seed)
-
-new_hp <- list(alpha = 0,
-			   lambda = c(1e-0, 1e1, 1e2, 2e2, 3e2, 4e2, 5e2, 6e2, 7e2, 8e2, 9e2, 1e3, 1e4))
-print('Running Logistic Regression on day 0 to predict severity')
-day_0_moribund_lr <- run_ml(day_0_moribund,
-	   'glmnet',
-       outcome_colname = 'early_euth',
-       training_frac = fraction,
-	   hyperparameters = new_hp,
-	   #find_feature_importance = TRUE,
-       seed = current_seed)
-new_hp <- list(alpha = 0,
-			   lambda = c(1e-1, 1e-0, 1e1, 1e2, 2e2, 3e2, 4e2, 5e2, 6e2, 7e2, 8e2, 9e2, 1e3, 1e4))
[…]

### 2021-05-26 exploratory/notebook/ml_analysis/ml_analysis_notebook.Rmd
start notebook analyzing ml performance

---
title: "ml_output_notebook"
author: "Nick Lesniak"
date: "5/21/2021"
output: rmarkdown::github_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
#knitr::opts_knit$set(root.dir = '../../..')
```

```{r read in data}
library(tidyverse)

# load in data
performance_df <- read_tsv('../../../data/process/ml/combined_ml_performance.tsv',
                     col_type = 'dddddddddddddddcdc') 
features_df <- read_tsv('../../../data/process/ml/combined_ml_feature_imp.tsv',
                     col_type = 'cccccDddddDcdddddcdl') 
hyperparameter_df <- read_tsv('../../../data/process/ml/combined_ml_hp_performance.tsv',
                     col_type = 'ddcdcc')

training_frac_df <- map_dfr(c(50, 60, 70, 80, 90), function(x){
  read_tsv(
      paste0('../../../data/process/ml/combined_ml_performance_', x, '.tsv'),
      col_type = 'dddddddddddddddcdc') %>% 
    mutate(training_frac = x/100)
})
```

## Model Performance

```{r}
performance_df %>% 
  select(method, seed, dataset, cv_auc = cv_metric_AUC, test_auc = AUC) %>% 
  pivot_longer(cols = c(cv_auc, test_auc), 
               names_to = 'metric', 
               values_to = 'Performance') %>% 
  ggplot(aes(x = method, y = Performance, color = metric)) + 
    geom_boxplot() + 
    geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
    facet_wrap(dataset~.) + 
    theme_bw() + 
    coord_flip()

```

```{r}
training_frac_df %>% 
  select(method, seed, dataset, cv_auc = cv_metric_AUC, test_auc = AUC, training_frac) %>% 
  pivot_longer(cols = c(cv_auc, test_auc), 
               names_to = 'metric', 
               values_to = 'Performance') %>% 
  ggplot(aes(x = method, y = Performance, color = metric)) + 
    geom_boxplot() + 
    geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
    facet_grid(dataset~training_frac) + 
    theme_bw() + 
    coord_flip()

```

## Hyperparameter Performance

```{r}
hyperparameter_df %>% 
  #filter(model == 'glmnet', dataset == 'same_day_toxin') %>% 
  #filter(model == 'glmnet', dataset == 'day_0_predict_future_toxin') %>% 
  #filter(model == 'glmnet', dataset == 'day_0_moribund') %>% 
  #filter(model == 'glmnet', dataset == 'day_10_histology') %>% 
  filter(model == 'glmnet') %>% 
    ggplot(aes(x = value, y = AUC, color = params)) + 
    geom_point(alpha = 0.25)+ 
    stat_summary(fun = median, geom = 'line') + 
    geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
    facet_grid(model~dataset, scales = 'free_x') + 
    theme_bw() + 
    scale_x_log10()

hyperparameter_df %>% 
  #filter(model == 'rf', dataset == 'same_day_toxin') %>% 
  #filter(model == 'rf', dataset == 'day_0_predict_future_toxin') %>% 
  #filter(model == 'rf', dataset == 'day_0_moribund') %>% 
  #filter(model == 'rf', dataset == 'day_10_histology') %>% 
  filter(model == 'rf') %>% 
  ggplot(aes(x = value, y = AUC, color = params)) + 
    geom_point(alpha = 0.25)+ 
    stat_summary(fun = median, geom = 'line') + 
    ylim(c(0.5,1)) + 
    facet_grid(dataset~model, scales = 'free') + 
    theme_bw()

hyperparameter_df %>% 
  filter(model == 'glmnet') %>% 
  pull(value) %>% 
  unique()
```

### 2021-06-09 exploratory/notebook/ml_analysis/ml_analysis_notebook.Rmd
update with otu only models

@@
 
-# load in data
-performance_df <- read_tsv('../../../data/process/ml/combined_ml_performance.tsv',
-                     col_type = 'dddddddddddddddcdc') 
-features_df <- read_tsv('../../../data/process/ml/combined_ml_feature_imp.tsv',
-                     col_type = 'cccccDddddDcdddddcdl') 
-hyperparameter_df <- read_tsv('../../../data/process/ml/combined_ml_hp_performance.tsv',
-                     col_type = 'ddcdcc')
-
+# load in hyperparameter/training data
 training_frac_df <- map_dfr(c(50, 60, 70, 80, 90), function(x){
   read_tsv(
-      paste0('../../../data/process/ml/combined_ml_performance_', x, '.tsv'),
+      paste0('../../../data/process/ml/training_frac/combined_ml_performance_', x, '.tsv'),
       col_type = 'dddddddddddddddcdc') %>% 
+    mutate(training_frac = x/100)
+})
+
+training_frac_hp_df <- map_dfr(c(50, 60, 70, 80, 90), function(x){
+  read_tsv(
+      paste0('../../../data/process/ml/training_frac/combined_ml_hp_performance_', x, '.tsv'),
+      col_type = 'ddcdcc') %>% 
     mutate(training_frac = x/100)
@@
 
-## Model Performance
-
-```{r}
-performance_df %>% 
-  select(method, seed, dataset, cv_auc = cv_metric_AUC, test_auc = AUC) %>% 
-  pivot_longer(cols = c(cv_auc, test_auc), 
-               names_to = 'metric', 
-               values_to = 'Performance') %>% 
-  ggplot(aes(x = method, y = Performance, color = metric)) + 
-    geom_boxplot() + 
-    geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
-    facet_wrap(dataset~.) + 
-    theme_bw() + 
-    coord_flip()
-
-```
+## Hyperparameter Performance
 
@@
                values_to = 'Performance') %>% 
+  mutate(dataset = factor(dataset, c('same_day_toxin', 'day_0_predict_future_toxin',
+                                     'day_0_moribund', 'day_10_histology'))) %>% 
   ggplot(aes(x = method, y = Performance, color = metric)) + 
@@
 
-## Hyperparameter Performance
-
 ```{r}
-hyperparameter_df %>% 
-  #filter(model == 'glmnet', dataset == 'same_day_toxin') %>% 
-  #filter(model == 'glmnet', dataset == 'day_0_predict_future_toxin') %>% 
-  #filter(model == 'glmnet', dataset == 'day_0_moribund') %>% 
-  #filter(model == 'glmnet', dataset == 'day_10_histology') %>% 
+training_frac_hp_df %>% 
   filter(model == 'glmnet') %>% 
+  mutate(dataset = factor(dataset, c('same_day_toxin', 'day_0_predict_future_toxin',
+                                     'day_0_moribund', 'day_10_histology'))) %>% 
     ggplot(aes(x = value, y = AUC, color = params)) + 
@@
     geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
-    facet_grid(model~dataset, scales = 'free_x') + 
+    facet_grid(dataset~training_frac, scales = 'free_x') + 
     theme_bw() + 
@@
 
-hyperparameter_df %>% 
-  #filter(model == 'rf', dataset == 'same_day_toxin') %>% 
-  #filter(model == 'rf', dataset == 'day_0_predict_future_toxin') %>% 
-  #filter(model == 'rf', dataset == 'day_0_moribund') %>% 
-  #filter(model == 'rf', dataset == 'day_10_histology') %>% 
+training_frac_hp_df %>% 
   filter(model == 'rf') %>% 
+  mutate(dataset = factor(dataset, c('same_day_toxin', 'day_0_predict_future_toxin',
+                                     'day_0_moribund', 'day_10_histology'))) %>% 
   ggplot(aes(x = value, y = AUC, color = params)) + 
@@
     ylim(c(0.5,1)) + 
-    facet_grid(dataset~model, scales = 'free') + 
+    facet_grid(dataset~training_frac, scales = 'free') + 
+    theme_bw()
+```
+
+## Model Performance
+
+Features  
+-  same day toxin presence - CFU, taxonomic abundances  
+-  day 0 future toxin presence - taxonomic abundances  
+-  day 0 future moribund - max cfu, median cfu, max toxin, median toxin, toxin presence, taxonomic abundances  
+-  day 10 histology (high/low) - cfu, max toxin, median toxin, toxin level, toxin presence, taxonomic abundances
+
+```{r}
+
+performance_df <- read_tsv('../../../data/process/ml/all_taxa_levels_all_features_ml_performance.tsv',
+                     col_type = 'dddddddddddddddcdcc') 
+features_df <- read_tsv('../../../data/process/ml/all_taxa_levels_all_features_ml_feature_imp.tsv',
+                     col_type = 'ddcccdcc') 
+hyperparameter_df <- read_tsv('../../../data/process/ml/all_taxa_levels_all_features_ml_hp_performance.tsv',
+                     col_type = 'ddcdccc') 
+taxonomy_df <- read_tsv('../../../data/process/final.taxonomy.tidy.tsv',
+                        col_type = cols(.default = col_character()))
+
+performance_df %>% 
+  select(method, seed, dataset, cv_auc = cv_metric_AUC, test_auc = AUC, taxonomic_level) %>% 
+  pivot_longer(cols = c(cv_auc, test_auc), 
+               names_to = 'metric', 
+               values_to = 'Performance') %>% 
+  mutate(dataset = factor(dataset, c('same_day_toxin', 'day_0_predict_future_toxin',
+                                     'day_0_moribund', 'day_10_histology')),
+         taxonomic_level = factor(taxonomic_level, 
+                                  levels = c('OTU', 'Genus', 'Family', 'Order', 'Class', 'Phylum'))) %>% 
+  ggplot(aes(x = taxonomic_level, y = Performance, color = metric)) + 
+    geom_boxplot() + 
+    geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
+    facet_grid(dataset~method, scales = 'free_y') + 
+    theme_bw() 
+```
+
+*day 10 histology is only predicting the high/low histology scores
+
+```{r}
+hyperparameter_df %>% 
+  filter(model == 'glmnet') %>% 
+  mutate(dataset = factor(dataset, c('same_day_toxin', 'day_0_predict_future_toxin',
+                                     'day_0_moribund', 'day_10_histology')),
+         taxonomic_level = factor(taxonomic_level, 
+                                  levels = c('OTU', 'Genus', 'Family', 'Order', 'Class', 'Phylum'))) %>% 
+  ggplot(aes(x = value, y = AUC, color = taxonomic_level)) + 
+    geom_point(alpha = 0.25) + 
+    stat_summary(fun = median, geom = 'line') + 
+    ylim(c(0.5,1)) + 
+    facet_grid(dataset~taxonomic_level, scales = 'free') + 
+    theme_bw() + 
+    scale_x_log10()
+
+```
+
+```{r}
+hyperparameter_df %>% 
+  filter(model == 'rf') %>% 
+  mutate(dataset = factor(dataset, c('same_day_toxin', 'day_0_predict_future_toxin',
+                                     'day_0_moribund', 'day_10_histology')),
+         taxonomic_level = factor(taxonomic_level, 
+                                  levels = c('OTU', 'Genus', 'Family', 'Order', 'Class', 'Phylum'))) %>% 
+  ggplot(aes(x = value, y = AUC, color = taxonomic_level)) + 
+    geom_point(alpha = 0.25) + 
+    stat_summary(fun = median, geom = 'line') + 
+    ylim(c(0.5,1)) + 
+    facet_grid(dataset~taxonomic_level, scales = 'free') + 
+    theme_bw() + 
+    scale_x_log10()
+
+```
+
+## Models and features
+
+```{r}
+performance_df %>% 
+  filter(dataset == 'same_day_toxin') %>% 
+  select(method, seed, dataset, cv_auc = cv_metric_AUC, test_auc = AUC, taxonomic_level) %>% 
+  pivot_longer(cols = c(cv_auc, test_auc), 
+               names_to = 'metric', 
+               values_to = 'Performance') %>% 
+  mutate(taxonomic_level = factor(taxonomic_level, 
+                                  levels = c('OTU', 'Genus', 'Family', 'Order', 'Class', 'Phylum'))) %>% 
+  ggplot(aes(x = taxonomic_level, y = Performance, color = metric)) + 
+    geom_boxplot() + 
+    geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
+    facet_grid(dataset~method, scales = 'free_y') + 
+    theme_bw() +
+    geom_rect(data = data.frame(ymin = 0.7, ymax = 0.95, method = 'glmnet', taxonomic_level = 6, Performance = 1),
+              aes(xmin = taxonomic_level - 0.5, xmax = taxonomic_level + 0.5, ymin = ymin, ymax = ymax, color = NA), 
+              fill = NA)
+```
+
+Using Logistic regression at the Phylum level the features with median differences greater than 0 are:
+
+```{r}
+
+features_df %>% 
+  filter(dataset == 'same_day_toxin',
+         method == 'glmnet',
+         taxonomic_level == 'Family') %>% 
+  group_by(names) %>% 
+  mutate(median_diff = median(perf_metric_diff)) %>% 
+  ungroup() %>% 
+  filter(median_diff > 0) %>%   
+  ggplot(aes(x = reorder(names, median_diff), y = perf_metric_diff)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
+    coord_flip() + 
+    facet_grid(method~dataset) + 
+    labs(x = NULL, y = 'AUC Diff with feature permuted') +
     theme_bw()
 
-hyperparameter_df %>% 
-  filter(model == 'glmnet') %>% 
-  pull(value) %>% 
-  unique()
 ```
+
+```{r}
+performance_df %>% 
+  filter(dataset == 'day_0_predict_future_toxin') %>% 
+  select(method, seed, dataset, cv_auc = cv_metric_AUC, test_auc = AUC, taxonomic_level) %>% 
+  pivot_longer(cols = c(cv_auc, test_auc), 
+               names_to = 'metric', 
+               values_to = 'Performance') %>% 
+  mutate(taxonomic_level = factor(taxonomic_level, 
+                                  levels = c('OTU', 'Genus', 'Family', 'Order', 'Class', 'Phylum'))) %>% 
+  ggplot(aes(x = taxonomic_level, y = Performance, color = metric)) + 
+    geom_boxplot() + 
+    geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
+    facet_grid(dataset~method, scales = 'free_y') + 
+    theme_bw() +
+    geom_rect(data = data.frame(ymin = 0.35, ymax = 0.98, method = 'glmnet', taxonomic_level = 2, Performance = 1),
+              aes(xmin = taxonomic_level - 0.5, xmax = taxonomic_level + 0.5, ymin = ymin, ymax = ymax, color = NA), 
+              fill = NA) +
+    geom_rect(data = data.frame(ymin = 0.48, ymax = 0.99, method = 'rf', taxonomic_level = 6, Performance = 1),
+              aes(xmin = taxonomic_level - 0.5, xmax = taxonomic_level + 0.5, ymin = ymin, ymax = ymax, color = NA), 
+              fill = NA)
+```
+
+Using Logistic regression at the Genus level or RF at the Phylum level the features with median differences greater than 0 are:
+
+```{r}
+features_df %>% 
+  filter(dataset == 'day_0_predict_future_toxin',
+         (method == 'rf' & taxonomic_level == 'Phylum') | (method == 'glmnet' & taxonomic_level == 'Genus')) %>% 
+  group_by(names) %>% 
+  mutate(median_diff = median(perf_metric_diff)) %>% 
+  ungroup() %>% 
+  filter(median_diff > 0) %>% 
+  ggplot(aes(x = reorder(names,median_diff), y = perf_metric_diff)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
+    #geom_jitter(alpha = 0.1, width = 0.1) + 
+    coord_flip() + 
+    facet_grid(method~dataset, scales= 'free_y', space = 'free_y') + 
+    labs(x = NULL, y = 'Decrease AUC') +
+    theme_bw()
+```
+
+```{r}
+performance_df %>% 
+  filter(dataset == 'day_0_moribund') %>% 
+  select(method, seed, dataset, cv_auc = cv_metric_AUC, test_auc = AUC, taxonomic_level) %>% 
+  pivot_longer(cols = c(cv_auc, test_auc), 
+               names_to = 'metric', 
+               values_to = 'Performance') %>% 
+  mutate(taxonomic_level = factor(taxonomic_level, 
+                                  levels = c('OTU', 'Genus', 'Family', 'Order', 'Class', 'Phylum'))) %>% 
+  ggplot(aes(x = taxonomic_level, y = Performance, color = metric)) + 
+    geom_boxplot() + 
+    geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
+    facet_grid(dataset~method, scales = 'free_y') + 
+    theme_bw() +
+    geom_rect(data = data.frame(ymin = 0.65, ymax = 1.01, method = 'glmnet', taxonomic_level = 5, Performance = 1),
+              aes(xmin = taxonomic_level - 0.5, xmax = taxonomic_level + 0.5, ymin = ymin, ymax = ymax, color = NA), 
+              fill = NA)
+```
+
+Using Logistic regression at the Class level the features with median differences greater than 0 are:
+
+```{r}
+features_df %>% 
+  filter(dataset == 'day_0_moribund',
+         method == 'glmnet',
+         taxonomic_level == 'Class') %>% 
+  group_by(names) %>% 
+  mutate(median_diff = median(perf_metric_diff)) %>% 
+  ungroup() %>% 
+  filter(median_diff > 0) %>% 
+  ggplot(aes(x = reorder(names,median_diff), y = perf_metric_diff)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
+    #geom_jitter(alpha = 0.1, width = 0.1) + 
+    coord_flip() + 
+    facet_grid(method~dataset) + 
+    labs(x = NULL, y = 'AUC Diff with feature permuted') +
+    theme_bw()
+```
+
+## Using only OTUs
+
+```{r}
+
+performance_df <- read_tsv('../../../data/process/ml/otu_only_all_taxa_levels_ml_performance.tsv',
+                     col_type = 'dddddddddddddddcdcc') 
+features_df <- read_tsv('../../../data/process/ml/otu_only_all_taxa_levels_ml_feature_imp.tsv',
+                     col_type = 'ddcccdcc') 
+hyperparameter_df <- read_tsv('../../../data/process/ml/otu_only_all_taxa_levels_ml_hp_performance.tsv',
+                     col_type = 'ddcdccc')
+
+performance_df %>% 
+  select(method, seed, dataset, cv_auc = cv_metric_AUC, test_auc = AUC, taxonomic_level) %>% 
+  pivot_longer(cols = c(cv_auc, test_auc), 
+               names_to = 'metric', 
+               values_to = 'Performance') %>% 
+  mutate(dataset = factor(dataset, c('same_day_toxin', 'day_0_predict_future_toxin',
+                                     'day_0_moribund', 'day_10_histology')),
+         taxonomic_level = factor(taxonomic_level, 
+                                  levels = c('OTU', 'Genus', 'Family', 'Order', 'Class', 'Phylum'))) %>% 
+  ggplot(aes(x = taxonomic_level, y = Performance, color = metric)) + 
+    geom_boxplot() + 
+    geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
+    facet_grid(dataset~method, scales = 'free_y') + 
+    theme_bw() 
+```
+
+```{r}
+performance_df %>% 
+  filter(dataset == 'same_day_toxin') %>% 
+  select(method, seed, dataset, cv_auc = cv_metric_AUC, test_auc = AUC, taxonomic_level) %>% 
+  pivot_longer(cols = c(cv_auc, test_auc), 
+               names_to = 'metric', 
+               values_to = 'Performance') %>% 
+  mutate(taxonomic_level = factor(taxonomic_level, 
+                                  levels = c('OTU', 'Genus', 'Family', 'Order', 'Class', 'Phylum'))) %>% 
+  ggplot(aes(x = taxonomic_level, y = Performance, color = metric)) + 
+    geom_boxplot() + 
+    geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
+    facet_grid(dataset~method, scales = 'free_y') + 
+    theme_bw()  +
+    geom_rect(data = data.frame(ymin = 0.71, ymax = 0.94, taxonomic_level = 3, Performance = 1),
+              aes(xmin = taxonomic_level - 0.5, xmax = taxonomic_level + 0.5, ymin = ymin, ymax = ymax, color = NA), 
+              fill = NA)
+
+```
+
+Using either Logistic regression or RF at the Family level the features with median differences greater than 0 are:
+
+```{r}
+
+features_df %>% 
+  filter(dataset == 'same_day_toxin',
+         method == 'glmnet',
+         taxonomic_level == 'Family') %>% 
+  group_by(names) %>% 
+  mutate(median_diff = median(perf_metric_diff)) %>% 
+  ungroup() %>% 
+  filter(median_diff > 0) %>%   
+  ggplot(aes(x = reorder(names, median_diff), y = perf_metric_diff)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
+    coord_flip() + 
+    facet_grid(method~dataset) + 
+    labs(x = NULL, y = 'AUC Diff with feature permuted') +
+    theme_bw()
+
+features_df %>% 
+  filter(dataset == 'same_day_toxin',
+         method == 'rf',
+         taxonomic_level == 'Family') %>% 
+  group_by(names) %>% 
+  mutate(median_diff = median(perf_metric_diff)) %>% 
+  ungroup() %>% 
+  filter(median_diff > 0) %>%   
+  ggplot(aes(x = reorder(names, median_diff), y = perf_metric_diff)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
+    coord_flip() + 
+    facet_grid(method~dataset) + 
+    labs(x = NULL, y = 'AUC Diff with feature permuted') +
+    theme_bw()
+
+```
+
+```{r}
+performance_df %>% 
+  filter(dataset == 'day_0_predict_future_toxin') %>% 
+  select(method, seed, dataset, cv_auc = cv_metric_AUC, test_auc = AUC, taxonomic_level) %>% 
+  pivot_longer(cols = c(cv_auc, test_auc), 
+               names_to = 'metric', 
+               values_to = 'Performance') %>% 
+  mutate(taxonomic_level = factor(taxonomic_level, 
+                                  levels = c('OTU', 'Genus', 'Family', 'Order', 'Class', 'Phylum'))) %>% 
+  ggplot(aes(x = taxonomic_level, y = Performance, color = metric)) + 
+    geom_boxplot() + 
+    geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
+    facet_grid(dataset~method, scales = 'free_y') + 
+    theme_bw()  +
+    geom_rect(data = data.frame(ymin = 0.38, ymax = 0.97, method = 'glmnet', taxonomic_level = 2, Performance = 1),
+              aes(xmin = taxonomic_level - 0.5, xmax = taxonomic_level + 0.5, ymin = ymin, ymax = ymax, color = NA), 
+              fill = NA) +
+    geom_rect(data = data.frame(ymin = 0.48, ymax = 0.99, method = 'rf', taxonomic_level = 6, Performance = 1),
+              aes(xmin = taxonomic_level - 0.5, xmax = taxonomic_level + 0.5, ymin = ymin, ymax = ymax, color = NA), 
+              fill = NA)
+
+```
+
+Using Logistic regression at the Genus level or RF at the Phylum level the features with median differences greater than 0 are:
+
+```{r}
+features_df %>% 
+  filter(dataset == 'day_0_predict_future_toxin',
+         (method == 'rf' & taxonomic_level == 'Phylum') | (method == 'glmnet' & taxonomic_level == 'Genus')) %>% 
+  group_by(names) %>% 
+  mutate(median_diff = median(perf_metric_diff)) %>% 
+  ungroup() %>% 
+  filter(median_diff > 0) %>% 
+  ggplot(aes(x = reorder(names, median_diff), y = perf_metric_diff)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
+    #geom_jitter(alpha = 0.1, width = 0.1) + 
+    coord_flip() + 
+    facet_grid(method~dataset, scales= 'free_y', space = 'free_y') + 
+    labs(x = NULL, y = 'Decrease AUC') +
+    theme_bw()
+```
+
+```{r}
+performance_df %>% 
+  filter(dataset == 'day_0_moribund') %>% 
+  select(method, seed, dataset, cv_auc = cv_metric_AUC, test_auc = AUC, taxonomic_level) %>% 
+  pivot_longer(cols = c(cv_auc, test_auc), 
+               names_to = 'metric', 
+               values_to = 'Performance') %>% 
+  mutate(taxonomic_level = factor(taxonomic_level, 
+                                  levels = c('OTU', 'Genus', 'Family', 'Order', 'Class', 'Phylum'))) %>% 
+  ggplot(aes(x = taxonomic_level, y = Performance, color = metric)) + 
+    geom_boxplot() + 
+    geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
+    facet_grid(dataset~method, scales = 'free_y') + 
+    theme_bw()  +
+    geom_rect(data = data.frame(ymin = 0.63, ymax = 1.01, taxonomic_level = 5, Performance = 1),
+              aes(xmin = taxonomic_level - 0.5, xmax = taxonomic_level + 0.5, ymin = ymin, ymax = ymax, color = NA), 
+              fill = NA)
+```
+
+Using Logistic regression or RF at the Class level features with median differences greater than 0 are:
+
+```{r}
+features_df %>% 
+  filter(dataset == 'day_0_moribund',
+         method == 'glmnet',
+         taxonomic_level == 'Class') %>% 
+  group_by(names) %>% 
+  mutate(median_diff = median(perf_metric_diff)) %>% 
+  ungroup() %>% 
+  filter(median_diff > 0) %>% 
+  ggplot(aes(x = reorder(names,median_diff), y = perf_metric_diff)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
+    #geom_jitter(alpha = 0.1, width = 0.1) + 
+    coord_flip() + 
+    facet_grid(method~dataset) + 
+    labs(x = NULL, y = 'AUC Diff with feature permuted') +
+    theme_bw()
+
+features_df %>% 
+  filter(dataset == 'day_0_moribund',
+         method == 'rf',
+         taxonomic_level == 'Class') %>% 
+  group_by(names) %>% 
+  mutate(median_diff = median(perf_metric_diff)) %>% 
+  ungroup() %>% 
+  filter(median_diff > 0) %>% 
+  ggplot(aes(x = reorder(names,median_diff), y = perf_metric_diff)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
+    #geom_jitter(alpha = 0.1, width = 0.1) + 
+    coord_flip() + 
+    facet_grid(method~dataset) + 
+    labs(x = NULL, y = 'AUC Diff with feature permuted') +
+    theme_bw()
+```
+
+## Use multiclass for endpoint histology scores
+
+```{r}
+hist_performance_df <- read_tsv('../../../data/process/ml/otu_only_all_taxa_levels_hist_ml_performance.tsv')  %>% 
+  mutate(taxonomic_level = factor(taxonomic_level, 
+                                  levels = c('OTU', 'Genus', 'Family', 'Order', 'Class', 'Phylum')))
+hist_features_df <- read_tsv('../../../data/process/ml/otu_only_all_taxa_levels_hist_ml_feature_imp.tsv') 
+hist_hyperparameter_df <- read_tsv('../../../data/process/ml/otu_only_all_taxa_levels_hist_ml_hp_performance.tsv')
+
+hist_performance_df %>% 
+  select(method, seed, dataset, training =cv_metric_logLoss, test = logLoss, taxonomic_level, null_model) %>% 
+  pivot_longer(cols = c(training, test), 
+               names_to = 'metric', 
+               values_to = 'Performance') %>% 
+  mutate(metric = ifelse(grepl('NULL', dataset), paste0(metric, '_NULL'), metric)) %>% 
+  ggplot(aes(x = taxonomic_level, y = Performance, color = metric)) + 
+    geom_boxplot() + 
+    #geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
+    facet_grid(.~method, scales = 'free_y') + 
+    theme_bw() + 
+    coord_cartesian(ylim = c(0,2)) +
+    geom_rect(data = data.frame(ymin = 0.1, ymax = 0.93, method = 'rf', taxonomic_level = 6, Performance = 1),
+              aes(xmin = taxonomic_level - 0.5, xmax = taxonomic_level + 0.5, ymin = ymin, ymax = ymax, color = NA), 
+              fill = NA)
+```
+
+Using multiclass RF at the Phylum level the features with median differences greater than 0 are:
+
+```{r}
+hist_features_df %>% 
+  filter(dataset == 'day_10_histology',
+         method == 'rf',
+         taxonomic_level == 'Phylum') %>% 
+  group_by(names) %>% 
+  mutate(median_diff = median(perf_metric_diff)) %>% 
+  ungroup() %>% 
+  filter(median_diff > 0) %>% 
+  ggplot(aes(x = reorder(names,median_diff), y = perf_metric_diff)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
+    geom_jitter(alpha = 0.1, width = 0.1) + 
+    coord_flip() + 
+    facet_grid(method~dataset) + 
+    labs(x = NULL, y = 'AUC Diff with feature permuted') + 
+    theme_bw()
+
+```

### 2021-06-09 exploratory/notebook/ml_analysis/ml_analysis_notebook.Rmd
summary md of modeling otus/toxin/cfu

--- 2021-05-26 exploratory/notebook/ml_analysis/ml_analysis_notebook.Rmd
@@
     ylim(c(0.5,1)) + 
-    facet_grid(dataset~model, scales = 'free') + 
+    facet_grid(dataset~training_frac, scales = 'free') + 
+    theme_bw()
+```
+
+## Model Performance
+
+Features  
+-  same day toxin presence - CFU, taxonomic abundances  
+-  day 0 future toxin presence - taxonomic abundances  
+-  day 0 future moribund - median cfu, day w/toxin presence (prior to day 3), taxonomic abundances  
+      to avoid duplicating similar data, i sumarized cfu/toxin to be a single feature to represent the amount or presence over the first two days of challenge since all mice were present for at least 2 days
+-  day 10 histology (high/mid/low) - cfu, toxin level, taxonomic abundances
+
+```{r}
+
+performance_df <- read_tsv('../../../data/process/ml/ml_performance.tsv',
+                     col_type = 'dddddddddddddddcdcc') %>% 
+  mutate(dataset = factor(dataset, c('same_day_toxin', 'day_0_predict_future_toxin',
+                                     'day_0_moribund', 'day_10_histology')),
+         taxonomic_level = factor(taxonomic_level, 
+                                  levels = c('OTU', 'Genus', 'Family', 'Order', 'Class', 'Phylum')))  
+features_df <- read_tsv('../../../data/process/ml/ml_feature_imp.tsv',
+                     col_type = 'ddcccdcc') 
+hyperparameter_df <- read_tsv('../../../data/process/ml/ml_hp_performance.tsv',
+                     col_type = 'ddcdccc') 
+taxonomy_df <- read_tsv('../../../data/process/final.taxonomy.tidy.tsv',
+                        col_type = cols(.default = col_character()))
+
+performance_df %>% 
+  select(method, seed, dataset, cv_auc = cv_metric_AUC, test_auc = AUC, taxonomic_level) %>% 
+  pivot_longer(cols = c(cv_auc, test_auc), 
+               names_to = 'metric', 
+               values_to = 'Performance') %>% 
+  ggplot(aes(x = taxonomic_level, y = Performance, color = metric)) + 
+    geom_boxplot() + 
+    geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
+    facet_grid(dataset~method, scales = 'free_y') + 
+    theme_bw() 
+```
+
+```{r}
+hyperparameter_df %>% 
+  filter(model == 'glmnet') %>% 
+  ggplot(aes(x = value, y = AUC, color = taxonomic_level)) + 
+    geom_point(alpha = 0.25) + 
+    stat_summary(fun = median, geom = 'line') + 
+    ylim(c(0.5,1)) + 
+    facet_grid(dataset~taxonomic_level, scales = 'free') + 
+    theme_bw() + 
+    scale_x_log10()
+
+```
+
+```{r}
+hyperparameter_df %>% 
+  filter(model == 'rf') %>% 
+  ggplot(aes(x = value, y = AUC, color = taxonomic_level)) + 
+    geom_point(alpha = 0.25) + 
+    stat_summary(fun = median, geom = 'line') + 
+    ylim(c(0.5,1)) + 
+    facet_grid(dataset~taxonomic_level, scales = 'free') + 
+    theme_bw() + 
+    scale_x_log10()
+
+```
+
+## Models and features
+
+#### Modeling future production of toxin based on initial community
+
+```{r}
+performance_df %>% 
+  filter(dataset == 'day_0_predict_future_toxin') %>% 
+  select(method, seed, dataset, cv_auc = cv_metric_AUC, test_auc = AUC, taxonomic_level) %>% 
+  pivot_longer(cols = c(cv_auc, test_auc), 
+               names_to = 'metric', 
+               values_to = 'Performance') %>% 
+  mutate(taxonomic_level = factor(taxonomic_level, 
+                                  levels = c('OTU', 'Genus', 'Family', 'Order', 'Class', 'Phylum'))) %>% 
+  ggplot(aes(x = taxonomic_level, y = Performance, color = metric)) + 
+    geom_boxplot() + 
+    geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
+    facet_grid(dataset~method, scales = 'free_y') + 
+    theme_bw() +
+    geom_rect(data = data.frame(ymin = 0.35, ymax = 0.98, method = 'glmnet', taxonomic_level = 2, Performance = 1),
+              aes(xmin = taxonomic_level - 0.5, xmax = taxonomic_level + 0.5, ymin = ymin, ymax = ymax, color = NA), 
+              fill = NA) +
+    geom_rect(data = data.frame(ymin = 0.48, ymax = 0.99, method = 'rf', taxonomic_level = 6, Performance = 1),
+              aes(xmin = taxonomic_level - 0.5, xmax = taxonomic_level + 0.5, ymin = ymin, ymax = ymax, color = NA), 
+              fill = NA)
+```
+
+Using Logistic regression at the Genus level or RF at the Phylum level the features with median differences greater than 0 are:
+
+```{r}
+features_df %>% 
+  filter(dataset == 'day_0_predict_future_toxin',
+         (method == 'rf' & taxonomic_level == 'Phylum') | (method == 'glmnet' & taxonomic_level == 'Genus')) %>% 
+  group_by(names) %>% 
+  mutate(median_diff = median(perf_metric_diff)) %>% 
+  ungroup() %>% 
+  filter(median_diff > 0) %>% 
+  ggplot(aes(x = reorder(names,median_diff), y = perf_metric_diff)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
+    #geom_jitter(alpha = 0.1, width = 0.1) + 
+    coord_flip() + 
+    facet_grid(method~dataset, scales= 'free_y', space = 'free_y') + 
+    labs(x = NULL, y = 'Decrease AUC') +
+    theme_bw()
+```
+
+#### Modeling production of toxin with the community and CFU data from that same day
+
+```{r}
+performance_df %>% 
+  filter(dataset == 'same_day_toxin') %>% 
+  select(method, seed, dataset, cv_auc = cv_metric_AUC, test_auc = AUC, taxonomic_level) %>% 
+  pivot_longer(cols = c(cv_auc, test_auc), 
+               names_to = 'metric', 
+               values_to = 'Performance') %>% 
+  ggplot(aes(x = taxonomic_level, y = Performance, color = metric)) + 
+    geom_boxplot() + 
+    geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
+    facet_grid(dataset~method, scales = 'free_y') + 
+    theme_bw() +
+    geom_rect(data = data.frame(ymin = 0.7, ymax = 0.95, method = 'glmnet', taxonomic_level = 6, Performance = 1),
+              aes(xmin = taxonomic_level - 0.5, xmax = taxonomic_level + 0.5, ymin = ymin, ymax = ymax, color = NA), 
+              fill = NA)
+```
+
+Using Logistic regression at the Phylum level the features with median differences greater than 0 are:
+
+```{r}
+
+features_df %>% 
+  filter(dataset == 'same_day_toxin',
+         method == 'glmnet',
+         taxonomic_level == 'Phylum') %>% 
+  group_by(names) %>% 
+  mutate(median_diff = median(perf_metric_diff)) %>% 
+  ungroup() %>% 
+  filter(median_diff > 0) %>%   
+  ggplot(aes(x = reorder(names, median_diff), y = perf_metric_diff)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
+    coord_flip() + 
+    facet_grid(method~dataset) + 
+    labs(x = NULL, y = 'AUC Diff with feature permuted') +
     theme_bw()
 
-hyperparameter_df %>% 
-  filter(model == 'glmnet') %>% 
-  pull(value) %>% 
-  unique()
 ```
+
+#### Modeling severe disease (moribund) from the initial community
+
+```{r}
+performance_df %>% 
+  filter(dataset == 'day_0_moribund') %>% 
+  select(method, seed, dataset, cv_auc = cv_metric_AUC, test_auc = AUC, taxonomic_level) %>% 
+  pivot_longer(cols = c(cv_auc, test_auc), 
+               names_to = 'metric', 
+               values_to = 'Performance') %>% 
+  mutate(taxonomic_level = factor(taxonomic_level, 
+                                  levels = c('OTU', 'Genus', 'Family', 'Order', 'Class', 'Phylum'))) %>% 
+  ggplot(aes(x = taxonomic_level, y = Performance, color = metric)) + 
+    geom_boxplot() + 
+    geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
+    facet_grid(dataset~method, scales = 'free_y') + 
+    theme_bw() +
+    geom_rect(data = data.frame(ymin = 0.65, ymax = 1.01, method = 'glmnet', taxonomic_level = 5, Performance = 1),
+              aes(xmin = taxonomic_level - 0.5, xmax = taxonomic_level + 0.5, ymin = ymin, ymax = ymax, color = NA), 
+              fill = NA)
+```
+
+Using Logistic regression at the Class level the features with median differences greater than 0 are:
+
+```{r}
+features_df %>% 
+  filter(dataset == 'day_0_moribund',
+         method == 'glmnet',
+         taxonomic_level == 'Class') %>% 
+  group_by(names) %>% 
+  mutate(median_diff = median(perf_metric_diff)) %>% 
+  ungroup() %>% 
+  filter(median_diff > 0) %>% 
+  ggplot(aes(x = reorder(names,median_diff), y = perf_metric_diff)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
+    #geom_jitter(alpha = 0.1, width = 0.1) + 
+    coord_flip() + 
+    facet_grid(method~dataset) + 
+    labs(x = NULL, y = 'AUC Diff with feature permuted') +
+    theme_bw()
+```
+
+#### Modeling endpoint histology scores
+
+```{r}
+hist_performance_df <- read_tsv('../../../data/process/ml/hist_ml_performance.tsv')
+hist_features_df <- read_tsv('../../../data/process/ml/hist_ml_feature_imp.tsv') 
+hist_hyperparameter_df <- read_tsv('../../../data/process/ml/hist_ml_hp_performance.tsv')
+
+hist_performance_df %>% 
+  select(method, seed, dataset, training =cv_metric_logLoss, test = logLoss, taxonomic_level, null_model) %>% 
+  pivot_longer(cols = c(training, test), 
+               names_to = 'metric', 
+               values_to = 'Performance') %>% 
+  mutate(metric = ifelse(grepl('NULL', dataset), paste0(metric, '_NULL'), metric)) %>% 
+  ggplot(aes(x = taxonomic_level, y = Performance, color = metric)) + 
+    geom_boxplot() + 
+    #geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
+    facet_grid(.~method, scales = 'free_y') + 
+    theme_bw() + 
+    coord_cartesian(ylim = c(0,2)) +
+    labs(x = NULL, y = 'Performance - logLoss') + 
+    geom_rect(data = data.frame(ymin = 0.1, ymax = 0.93, method = 'rf', taxonomic_level = 6, Performance = 1),
+              aes(xmin = taxonomic_level - 0.5, xmax = taxonomic_level + 0.5, ymin = ymin, ymax = ymax, color = NA), 
+              fill = NA)
+```
+
+Using multiclass RF at the Phylum level the features with median differences greater than 0 are:
+
+```{r}
+hist_features_df %>% 
+  filter(dataset == 'day_10_histology',
+         method == 'rf',
+         taxonomic_level == 'Phylum') %>% 
+  group_by(names) %>% 
+  mutate(median_diff = median(perf_metric_diff)) %>% 
+  ungroup() %>% 
+  filter(median_diff > 0) %>% 
+  ggplot(aes(x = reorder(names,median_diff), y = perf_metric_diff)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
+    geom_jitter(alpha = 0.1, width = 0.1) + 
+    coord_flip() + 
+    facet_grid(method~dataset) + 
+    labs(x = NULL, y = 'AUC Diff with feature permuted') + 
+    theme_bw()
+
+```

### 2021-06-10 exploratory/notebook/ml_analysis/ml_analysis_notebook.Rmd
add abundance information for features

@@
 
+metadata <- read_tsv('../../../data/process/metadata_tidy.tsv',
+					 col_type = 'cccccDddddDcdddddcdl') %>% 
+	filter(cdiff_strain == 431) %>% 
+	mutate(cdiff_cfu = ifelse(cdiff_cfu == 0, log10(60), log10(cdiff_cfu)))
+
+toxin <- read_tsv('../../../data/process/toxin_tidy.tsv',
+				  col_type = 'cccccdc') %>% 
+	mutate(toxin = ifelse(Log_repiricoal_dilution > 1, 'present', 'absent'))
+
+histology <- read_tsv('../../../data/process/histology_tidy.tsv',
+					  col_type = 'ccccdddcdcc') %>% 
+	mutate(hist_score = case_when(summary_score < 5 ~ 'low',
+								  summary_score > 5 ~ 'high',
+								  summary_score == 5 ~ 'mid',
+								  T ~ 'NA'))
+
+shared <- read_tsv('../../../data/mothur/sample.final.0.03.subsample.shared',
+         col_types = cols(.default = col_double(),
+                          Group = col_character())) %>% 
+  mutate(Group = ifelse(grepl('_D|DA', Group), Group,
+      gsub('D', '_D', Group))) %>% 
+  select(-label, -numOtus)
+
+same_day_toxin_df <- toxin %>% 
+	select(group, toxin) %>% 
+	inner_join(shared, by = c('group' = 'Group')) %>% 
+	left_join(select(metadata, group, C_difficile_CFU = cdiff_cfu), by = c('group'))
+
+#	from day 0?
+day_0_predict_future_toxin_df <- toxin %>% 
+	group_by(mouse_id) %>% 
+	summarise(toxin = ifelse(max(Log_repiricoal_dilution) > 1, 'present', 'absent')) %>% 
+	left_join(select(metadata, mouse_id, day, group), 
+			  by = c('mouse_id')) %>% 
+	filter(day == 0) %>% 
+	inner_join(shared, by = c('group' = 'Group'))
+
+# setup data to predict moribunidty (microbiome, cdiff, toxin)
+toxin_summary <- toxin %>% 
+	filter(toxin_sample_day  %in% c(1,2)) %>% 
+	group_by(mouse_id) %>% 
+	summarise(toxin_positive_days = sum(Log_repiricoal_dilution > 1))
+cfu_summary <- metadata %>% 
+	filter(day %in% c(1,2)) %>% 
+	group_by(mouse_id, early_euth) %>% 
+	summarise(median_CFU = median(cdiff_cfu))
+day_0_moribund <- metadata %>% 
+	select(group, mouse_id, day) %>% 
+	filter(day == 0) %>% 
+	left_join(cfu_summary, by = c('mouse_id')) %>% 
+	left_join(toxin_summary, by = c('mouse_id')) %>% 
+	inner_join(shared, by = c('group' = 'Group')) %>% 
+	mutate(early_euth = ifelse(early_euth, 'severe', 'mild'))
+
+# setup data to predict histology severity of colonized mice (microbiome, cdiff, toxin)
+day_10_histology_df <- metadata %>% 
+	filter(day == 10,
+		early_euth == F) %>% 
+	select(group, C_difficile_CFU = cdiff_cfu, mouse_id) %>% 
+	left_join(select(toxin, group, toxin_level = Log_repiricoal_dilution),
+		by = c('group')) %>% 
+	inner_join(shared, by = c('group' = 'Group')) %>% 
+	inner_join(select(histology, mouse_id, hist_score), by = c('mouse_id')) 
+```
+
+```{r}
 performance_df %>% 
@@
 ```{r}
-features_df %>% 
+day_0_toxin_production <- features_df %>% 
   filter(dataset == 'day_0_predict_future_toxin',
-         (method == 'rf' & taxonomic_level == 'Phylum') | (method == 'glmnet' & taxonomic_level == 'Genus')) %>% 
+         (method == 'rf' & taxonomic_level == 'Phylum')# | (method == 'glmnet' & taxonomic_level == 'Genus')
+         ) %>% 
   group_by(names) %>% 
@@
   ungroup() %>% 
-  filter(median_diff > 0) %>% 
+  filter(median_diff > 0) 
+day_0_toxin_production %>%
   ggplot(aes(x = reorder(names,median_diff), y = perf_metric_diff)) + 
@@
     theme_bw()
+```
+
+```{r}
+day_0_toxin_features <- day_0_toxin_production %>% 
+  select(names, median_diff) %>% 
+  unique %>% 
+  arrange(median_diff) %>% 
+  pull(names)
+
+day_0_predict_future_toxin_df %>% 
+  pivot_longer(cols = starts_with('Otu'), names_to = 'OTU', values_to = 'counts') %>% 
+  left_join(taxonomy_df, by = "OTU") %>% 
+  filter(Phylum %in% day_0_toxin_features) %>% 
+  group_by(group, toxin, Phylum) %>% 
+  summarise(counts = sum(counts)/21.07) %>% 
+  mutate(Phylum = factor(Phylum, day_0_toxin_features)) %>% 
+  ggplot(aes(x = Phylum, y = counts, color = toxin)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5),
+                 position = position_dodge(width = 0.7)) +
+    geom_jitter(position = position_jitterdodge(dodge.width = 0.7, jitter.width = 0.3), 
+                alpha = 0.2) + 
+    scale_y_log10() + 
+    theme_bw() + 
+    coord_flip() + 
+    scale_color_manual(values = c('red', 'black'),
+                       breaks = c('absent', 'present'), 
+                       labels = c('False', 'True')) + 
+    labs(x = NULL, y = 'Relative Abundance (%)', color  = 'Toxin Detected')
+
 ```
@@
 
-features_df %>% 
+same_day_toxin <- features_df %>% 
   filter(dataset == 'same_day_toxin',
@@
   ungroup() %>% 
-  filter(median_diff > 0) %>%   
+  filter(median_diff > 0) 
+same_day_toxin %>%   
   ggplot(aes(x = reorder(names, median_diff), y = perf_metric_diff)) + 
@@
     theme_bw()
+
+```
+
+```{r}
+same_day_toxin_features <- same_day_toxin %>% 
+  select(names, median_diff) %>% 
+  unique %>% 
+  arrange(median_diff) %>% 
+  pull(names)
+
+same_day_toxin_df %>% 
+  ggplot(aes(x = toxin, y = C_difficile_CFU, color = toxin)) + 
+    geom_jitter(alpha = 0.2, width  = 0.3) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5),
+                 position = position_dodge(width = 0.7)) +
+    coord_flip() + 
+    theme_bw() + 
+    labs(x = NULL, y = 'C. difficile CFU', color  = 'Toxin Detected')
+```
+
+```{r}
+same_day_toxin_df %>% 
+  pivot_longer(cols = starts_with('Otu'), names_to = 'OTU', values_to = 'counts') %>% 
+  left_join(taxonomy_df, by = "OTU") %>% 
+  filter(Phylum %in% same_day_toxin_features) %>% 
+  group_by(group, toxin, Phylum) %>% 
+  summarise(counts = sum(counts)/21.07) %>% 
+  mutate(Phylum = factor(Phylum, same_day_toxin_features)) %>% 
+  ggplot(aes(x = Phylum, y = counts, color = toxin)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5),
+                 position = position_dodge(width = 0.7)) +
+    geom_jitter(position = position_jitterdodge(dodge.width = 0.7, jitter.width = 0.3), 
+                alpha = 0.05) + 
+    scale_y_log10() + 
+    theme_bw() + 
+    coord_flip() + 
+    scale_color_manual(values = c('red', 'black'),
+                       breaks = c('absent', 'present'), 
+                       labels = c('False', 'True')) + 
+    labs(x = NULL, y = 'Relative Abundance (%)', color  = 'Toxin Detected')
 
@@
 ```{r}
-features_df %>% 
+day_0_moribund_plot_df <- features_df %>% 
   filter(dataset == 'day_0_moribund',
@@
   ungroup() %>% 
-  filter(median_diff > 0) %>% 
+  filter(median_diff > 0) 
+day_0_moribund_plot_df %>% 
   ggplot(aes(x = reorder(names,median_diff), y = perf_metric_diff)) + 
@@
     theme_bw()
+```
+
+```{r}
+day_0_moribund_features <- day_0_moribund_plot_df %>% 
+  select(names, median_diff) %>% 
+  unique %>% 
+  arrange(median_diff) %>% 
+  pull(names)
+
+day_0_moribund %>% 
+  pivot_longer(cols = starts_with('Otu'), names_to = 'OTU', values_to = 'counts') %>% 
+  left_join(taxonomy_df, by = "OTU") %>% 
+  filter(Class %in% day_0_moribund_features) %>% 
+  group_by(group, early_euth, Class) %>% 
+  summarise(counts = sum(counts)/21.07) %>% 
+  mutate(Phylum = factor(Class, day_0_moribund_features)) %>% 
+  ggplot(aes(x = Phylum, y = counts, color = early_euth)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5),
+                 position = position_dodge(width = 0.7)) +
+    geom_jitter(position = position_jitterdodge(dodge.width = 0.7, jitter.width = 0.3), 
+                alpha = 0.2) + 
+    scale_y_log10() + 
+    theme_bw() + 
+    coord_flip() + 
+#    scale_color_manual(values = c('red', 'black'),
+ #                      breaks = c('absent', 'present'), 
+  #                     labels = c('False', 'True')) + 
+    labs(x = NULL, y = 'Relative Abundance (%)', color  = 'Disease Severity')
+
 ```
@@
 ```{r}
-hist_features_df %>% 
+hist_features_plot_df <- hist_features_df %>% 
   filter(dataset == 'day_10_histology',
@@
   ungroup() %>% 
-  filter(median_diff > 0) %>% 
+  filter(median_diff > 0)
+hist_features_plot_df %>% 
   ggplot(aes(x = reorder(names,median_diff), y = perf_metric_diff)) + 
@@
 ```
+
+```{r}
+hist_features <- hist_features_plot_df %>% 
+  select(names, median_diff) %>% 
+  unique %>% 
+  arrange(median_diff) %>% 
+  pull(names)
+
+day_10_histology_df %>% 
+  pivot_longer(cols = starts_with('Otu'), names_to = 'OTU', values_to = 'counts') %>% 
+  left_join(taxonomy_df, by = "OTU") %>% 
+  filter(Phylum %in% hist_features) %>% 
+  group_by(group, hist_score, Phylum) %>% 
+  summarise(counts = sum(counts)/21.07) %>% 
+  mutate(Phylum = factor(Phylum, hist_features),
+         hist_score = factor(hist_score, c('low', 'mid', 'high'))) %>% 
+  ggplot(aes(x = Phylum, y = counts, color = hist_score)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5),
+                 position = position_dodge(width = 0.7)) +
+    geom_jitter(position = position_jitterdodge(dodge.width = 0.9, jitter.width = 0.3), 
+                alpha = 0.2) + 
+    scale_y_log10() + 
+    theme_bw() + 
+    coord_flip() + 
+#    scale_color_manual(values = c('red', 'black'),
+ #                      breaks = c('absent', 'present'), 
+  #                     labels = c('False', 'True')) + 
+    labs(x = NULL, y = 'Relative Abundance (%)', color  = 'Histology Score')
+
+```

### 2021-09-17 code/ml/concat_ml_results.sh
setup separate tuning run of model and include day 0 hist

@@ -1,0 +1,46 @@
+#!/bin/bash
+
+# Author: Nick Lesniak
+# Date: 2021-05-21
+#
+#######################################################################################
+# This script will:
+#	Combine them together to have the results for 100 datasplits in one file
+#   
+########################################################################################
+
+SEARCH_DIR=data/process/ml/temp
+FINAL_DIR=data/process/ml/
+
+# Keep the first line of File1 and remove the first line of all the others and combine
+# combine binary models
+for taxa_level in "OTU" "Genus" "Family" "Order" "Class" "Phylum"
+	do
+	head -1 $SEARCH_DIR/ml_performance_"$taxa_level"_1.tsv  > $SEARCH_DIR/combined_ml_performance_"$taxa_level".tsv; tail -n +2 -q $SEARCH_DIR/ml_performance_"$taxa_level"_+([0-9]).tsv >> $SEARCH_DIR/combined_ml_performance_"$taxa_level".tsv
+	head -1 $SEARCH_DIR/ml_feature_imp_"$taxa_level"_1.tsv  > $SEARCH_DIR/combined_ml_feature_imp_"$taxa_level".tsv; tail -n +2 -q $SEARCH_DIR/ml_feature_imp_"$taxa_level"_+([0-9]).tsv >> $SEARCH_DIR/combined_ml_feature_imp_"$taxa_level".tsv
+	head -1 $SEARCH_DIR/ml_hp_performance_"$taxa_level"_1.tsv  > $SEARCH_DIR/combined_ml_hp_performance_"$taxa_level".tsv; tail -n +2 -q $SEARCH_DIR/ml_hp_performance_"$taxa_level"_+([0-9]).tsv >> $SEARCH_DIR/combined_ml_hp_performance_"$taxa_level".tsv
+
+	mv $SEARCH_DIR/combined_ml_performance_"$taxa_level".tsv $FINAL_DIR/combined_ml_performance_"$taxa_level".tsv
+	mv $SEARCH_DIR/combined_ml_feature_imp_"$taxa_level".tsv $FINAL_DIR/combined_ml_feature_imp_"$taxa_level".tsv
[…]

### 2021-09-17 code/ml/setup_ml_data.R

@@ -89,14 +89,19 @@ day_0_moribund <- metadata %>%
 write_tsv(day_0_moribund, 'data/process/ml/day_0_moribund.tsv')	
 
 # setup data to predict histology severity of colonized mice (microbiome, cdiff, toxin)
-day_10_histology_df <- metadata %>% 
-	filter(day == 10,
-		early_euth == F) %>% 
-	select(group, C_difficile_CFU = cdiff_cfu, mouse_id) %>% 
-	left_join(select(toxin, group, toxin_level = Log_repiricoal_dilution),
-		by = c('group')) %>% 
+toxin_summary <- toxin %>% 
+	group_by(mouse_id) %>% 
+	summarise(toxin_positive_days = sum(Log_repiricoal_dilution > 1))
+day_0_histology_df <- metadata %>% 
+	select(group, mouse_id, day) %>% 
+	filter(day == 0) %>% 
+	left_join(cfu_summary, by = c('mouse_id')) %>% 
+	left_join(toxin_summary, by = c('mouse_id')) %>% 
 	inner_join(shared, by = c('group' = 'Group')) %>% 
+	ungroup %>% 
+	filter(early_euth == F) %>% 
 	inner_join(select(histology, mouse_id, hist_score), by = c('mouse_id')) %>% 
-	select(-mouse_id, -group) %>% 
+	select(-mouse_id, -group, -early_euth, -day) %>% 
+	filter(hist_score != 'mid') %>% 
 	relocate(hist_score)
-write_tsv(day_10_histology_df, 'data/process/ml/day_10_histology.tsv')	
+write_tsv(day_0_histology_df, 'data/process/ml/day_0_histology.tsv')	

### 2021-09-17 code/ml/tune_ml.R

@@ -1,0 +1,187 @@
+################################################################################
+#
+# run_ml.R
+#
+# This script runs mikropml
+# 	outputs
+#
+# Dependencies...
+# * data/process/ml/same_day_toxin.tsv
+# * data/process/ml/day_0_predict_future_toxin.tsv
+# * data/process/ml/day_0_moribund.tsv
+# * data/process/ml/day_0_histology.tsv
+#
+# Output...
+#
+################################################################################
+
+# load packages
+library('tidyverse')
+library('mikropml')
+
+# get seed
+current_seed <- as.numeric(commandArgs(TRUE)[1])
+taxonomic_level <- as.character(commandArgs(TRUE)[2])
+set.seed(current_seed)
+fraction <- as.numeric(commandArgs(TRUE)[3])
+
+# read in data
+print('Reading in data')
+same_day_toxin <- read_tsv('data/process/ml/same_day_toxin.tsv',
+						   col_type = cols(.default = col_double(), 
+						   				   toxin = col_character()))
+day_0_predict_future_toxin <- read_tsv('data/process/ml/day_0_predict_future_toxin.tsv',
+						   col_type = cols(.default = col_double(), 
+						   				   toxin = col_character()))
+day_0_moribund <- read_tsv('data/process/ml/day_0_moribund.tsv',
+						   col_type = cols(.default = col_double(), 
+						   				   early_euth = col_character())) 
+day_0_histology <- read_tsv('data/process/ml/day_0_histology.tsv',
+						   col_type = cols(.default = col_double(), 
+						   				   hist_score = col_character())) 
+taxonomy_df <- read_tsv('data/process/final.taxonomy.tidy.tsv',
+                        col_type = cols(.default = col_character()))
+
+# preprocess data
+print('Preprocessing data')
+setup_ml_df <- function(input_df, outcome_column){
+	tax_df <- taxonomy_df %>% 
+		mutate(tax_level = get(taxonomic_level)) %>% 
+		select(OTU, tax_level)
+	input_df %>% 
+	# select taxonomic level to be used 
+		mutate(row = rownames(.)) %>% 
+		pivot_longer(cols = contains('Otu'), names_to = 'OTU', values_to = 'value') %>% 
[…]

### 2021-09-17 code/tidy_raw_data.R
fix ear tag typo

@@ -52,8 +52,10 @@ toxin_data <- read_xlsx(input_toxin, sheet = 'Sheet1',
 		col_types = c('text', 'text', 'text', 'numeric')) %>% 
 		rename(group = Cage_Mouse,
 			   toxin_sample_day = Experiment_day,
-			   toxin_sample_type = Sample_source) %>% 
+			   toxin_sample_type = Sample_source) %>%
 		separate(group, c('cage_id', 'ear_tag', 'day'), remove = F, fill = 'left') %>% 
+		mutate(ear_tag = ifelse(ear_tag == '986', '896', ear_tag), # fix ear tag typo
+			   ear_tag = gsub('C', '', ear_tag)) %>%  # remove C for cecal label from ear tag
 		mutate(mouse_id = paste0(cage_id, '_', ear_tag), # create unique mouse id
 			   group = gsub('-', '_', group), # adjust sample labels to be mothur friendly
 			   mouse_id = ifelse(mouse_id == 'NA_NA', group, mouse_id)) %>% # set mouse_id for inoculum

### 2021-09-17 exploratory/notebook/ml_analysis/ml_analysis_notebook.Rmd
add data type for added column

@@
 features_df <- read_tsv('../../../data/process/ml/ml_feature_imp.tsv',
-                     col_type = 'ddcccdcc') 
+                     col_type = 'ddcccddcc') 
 hyperparameter_df <- read_tsv('../../../data/process/ml/ml_hp_performance.tsv',

### 2021-09-17 exploratory/notebook/ml_analysis/ml_analysis_notebook.Rmd
add link for log loss description

@@
               fill = NA)
+# helpful logloss explanation https://stats.stackexchange.com/questions/276067/whats-considered-a-good-log-loss
 ```

### 2021-10-21 code/plot_figure_6.R
create script for figure 6

@@ -1,0 +1,94 @@
+library(tidyverse)
+library(here)
+
+
+source(here('code/Sum_OTU_by_Tax.R'))
+
+# load in data
+metadata <- read_tsv(here('data/process/metadata_tidy.tsv'),
+                     col_type = 'cccccDddddDcdddddcdl') %>% 
+  mutate(isolate_431 = ifelse(cdiff_strain == 431, '431', 'Other'),
+         isolate = factor(cdiff_strain, levels = c('431', '299', '395','458')))
+  
+toxin <- read_tsv(here('data/process/toxin_tidy.tsv'),
+                  col_type = 'cccccdc')
+histology <- read_tsv(here('data/process/histology_tidy.tsv'),
+                      col_type = 'ccccdddcdcc') %>% 
+  select(summary_score, mouse_id)
+
+donor_colors <- c("#41bbc5", "#7e2640", "#9ad859", "#682dbd", "#728e24", "#e26df8", "#23980d", "#fd048f", "#46e33e", "#fa1bfc", "#155126", "#f9b2d1", "#154975", "#e2c627", "#6294ce")
+
+
+non_431_donors <- metadata %>% 
+  filter(cdiff_strain != 431) %>% 
+  pull(human_source) %>% 
+  unique
+
+cfu_plot <- metadata %>% 
+  filter(human_source %in% non_431_donors,
+         day == 10) %>% 
+  add_row(human_source = 'DA00578', cdiff_cfu = NA, isolate_431 = 'Other') %>% 
+  left_join(toxin, by = 'group') %>% 
+  mutate(cdiff_cfu = ifelse(cdiff_cfu == 0, log10(60), log10(cdiff_cfu)),
+         donor_source = case_when(human_source == 'DA00369' ~ 'F',
+                                  human_source == 'DA00430' ~ 'G',
+                                  human_source == 'DA00578' ~ 'M'),
+         donor_source = factor(donor_source, levels = c('G' ,'F', 'M'))) %>% 
+  ggplot(aes(x = donor_source, y = cdiff_cfu, color = isolate_431)) + 
+    geom_point(position = position_jitterdodge(jitter.width = 0.1)) + 
+    geom_hline(yintercept = 2, linetype = 'dashed', color = 'lightgray') + 
+    geom_label(x = 3, y = 2, label = "LOD", fill = 'white', color = 'white') + 
+    geom_text(x = 3, y = 2, label = "LOD", color = 'lightgray', size = 12/.pt) + 
[…]

### 2021-10-21 code/run_lefse_analysis.R
refine lefse analysis

@@ -1,20 +1,22 @@
 
 
 library(tidyverse)
+library(here)
 
-metadata <- read_tsv('data/process/metadata_tidy.tsv',
+metadata <- read_tsv(here('data/process/metadata_tidy.tsv'),
 		col_type = 'cccccDddddDcdddddcdl') 
-shared <- read_tsv('data/mothur/sample.final.0.03.subsample.shared',
+shared <- read_tsv(here('data/mothur/sample.final.0.03.subsample.shared'),
 		col_types = cols(.default = col_double(),
 			Group = col_character())) %>% 
 	mutate(Group = ifelse(grepl('_D|DA', Group), Group,
 		gsub('D', '_D', Group)))
-toxin <- read_tsv('data/process/toxin_tidy.tsv',
-                  col_type = 'cccccdc')
-histology <- read_tsv('data/process/histology_tidy.tsv',
+toxin <- read_tsv(here('data/process/toxin_tidy.tsv'),
+                  col_type = 'cccccdc') %>% 
+  filter(mouse_id %in% metadata$mouse_id)
+histology <- read_tsv(here('data/process/histology_tidy.tsv'),
                       col_type = 'ccccdddcdcc') %>% 
 	             mutate(mouse_id = paste0(cage_id, '_', ear_tag))
-taxonomy <- read_tsv('data/process/final.taxonomy.tidy.tsv',
+taxonomy <- read_tsv(here('data/process/final.taxonomy.tidy.tsv'),
                      col_types = cols(.default = col_character()))
 
 run_lefse <- function(sample_df_name, tax_level){
@@ -51,66 +53,90 @@ run_lefse <- function(sample_df_name, tax_level){
 		mutate(numOtus = length(present_otus))
 
 	# write files to be used in mothur for lefse analysis
-	write_tsv(file = paste0('data/process/', i, '_', tax_level, '.shared'), 
+	write_tsv(file = paste0('data/process/lefse/', i, '_', tax_level, '.shared'), 
 		x = filter(current_shared, Group %in% current_df$Group))
-	write_tsv(file = paste0('data/process/', i, '_', tax_level, '.design'), 
+	write_tsv(file = paste0('data/process/lefse/', i, '_', tax_level, '.design'), 
 		x = filter(current_df, Group %in% current_shared$Group))
 	print(tax_level)
 	# run lefse
[…]

### 2021-10-21 exploratory/notebook/ml_analysis/ml_analysis_notebook.Rmd
add manuscript figures to rmd notebooks

@@
 knitr::opts_chunk$set(warning = FALSE)
-#knitr::opts_knit$set(root.dir = '../../..')
+
 ```
@@
 library(tidyverse)
+library(here)
 
 # load in hyperparameter/training data
-training_frac_df <- map_dfr(c(50, 60, 70, 80, 90), function(x){
-  read_tsv(
-      paste0('../../../data/process/ml/training_frac/combined_ml_performance_', x, '.tsv'),
-      col_type = 'dddddddddddddddcdc') %>% 
-    mutate(training_frac = x/100)
-})
+training_frac_df <- read_tsv(
+      here(paste0('data/process/ml/ml_tune_performance.tsv')),
+      col_type = 'dddddddddddddddcdccd')
 
-training_frac_hp_df <- map_dfr(c(50, 60, 70, 80, 90), function(x){
-  read_tsv(
-      paste0('../../../data/process/ml/training_frac/combined_ml_hp_performance_', x, '.tsv'),
-      col_type = 'ddcdcc') %>% 
-    mutate(training_frac = x/100)
-})
+training_frac_hp_df <- read_tsv(
+      here(paste0('data/process/ml/ml_tune_hp_performance.tsv')),
+      col_type = 'ddcdcdcc')
+
 ```
@@
 ```{r}
+tax_levels <- c('Phylum', 'Class', 'Order', 'Family', 'Genus', 'OTU')
+plot_model_performance <- function(plot_dataset){
   training_frac_df %>% 
-  select(method, seed, dataset, cv_auc = cv_metric_AUC, test_auc = AUC, training_frac) %>% 
+    filter(dataset == plot_dataset,
+           method == 'rf') %>% 
+    select(method, seed, cv_auc = cv_metric_AUC, test_auc = AUC, training_fraction, taxonomic_level) %>% 
     pivot_longer(cols = c(cv_auc, test_auc), 
@@
                  values_to = 'Performance') %>% 
-  mutate(dataset = factor(dataset, c('same_day_toxin', 'day_0_predict_future_toxin',
-                                     'day_0_moribund', 'day_10_histology'))) %>% 
-  ggplot(aes(x = method, y = Performance, color = metric)) + 
+    mutate(taxonomic_level = factor(taxonomic_level, tax_levels)) %>% 
+    ggplot(aes(x = taxonomic_level, y = Performance, color = metric)) + 
       geom_boxplot() + 
       geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
-    facet_grid(dataset~training_frac) + 
+      facet_grid(method~training_fraction) + 
       theme_bw() + 
       coord_flip()
+}
 
+plot_lr_hp <- function(plot_dataset){
+  training_frac_hp_df %>% 
+    filter(model == 'glmnet',
+           dataset == plot_dataset) %>% 
+    mutate(taxonomic_level = factor(taxonomic_level, tax_levels)) %>% 
+      ggplot(aes(x = value, y = AUC, color = params)) + 
+      geom_point(alpha = 0.1)+ 
+      stat_summary(fun = median, geom = 'line', color = 'black') + 
+      #geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
+      facet_grid(taxonomic_level~training_fraction, scales = 'free') + 
+      theme_bw() + 
+      scale_x_log10()
+}
+plot_rf_hp <- function(plot_dataset){
+  training_frac_hp_df %>% 
+    filter(model == 'rf',
+           dataset == plot_dataset) %>% 
+    mutate(taxonomic_level = factor(taxonomic_level, tax_levels)) %>% 
+    ggplot(aes(x = value, y = AUC, color = params)) + 
+      geom_point(alpha = 0.1)+ 
+      stat_summary(fun = median, geom = 'line', color = 'black') + 
+      xlim(c(0,10)) + 
+      facet_grid(taxonomic_level~training_fraction, scales = 'free') + 
+      theme_bw()
+}
 ```
 
+#### Same day toxin
 ```{r}
-training_frac_hp_df %>% 
-  filter(model == 'glmnet') %>% 
-  mutate(dataset = factor(dataset, c('same_day_toxin', 'day_0_predict_future_toxin',
-                                     'day_0_moribund', 'day_10_histology'))) %>% 
-    ggplot(aes(x = value, y = AUC, color = params)) + 
-    geom_point(alpha = 0.25)+ 
-    stat_summary(fun = median, geom = 'line') + 
-    geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
-    facet_grid(dataset~training_frac, scales = 'free_x') + 
-    theme_bw() + 
-    scale_x_log10()
+plot_model_performance('same_day_toxin')
+plot_lr_hp('same_day_toxin')
+plot_rf_hp('same_day_toxin')
+```
 
-training_frac_hp_df %>% 
-  filter(model == 'rf') %>% 
-  mutate(dataset = factor(dataset, c('same_day_toxin', 'day_0_predict_future_toxin',
-                                     'day_0_moribund', 'day_10_histology'))) %>% 
-  ggplot(aes(x = value, y = AUC, color = params)) + 
-    geom_point(alpha = 0.25)+ 
-    stat_summary(fun = median, geom = 'line') + 
-    ylim(c(0.5,1)) + 
-    facet_grid(dataset~training_frac, scales = 'free') + 
-    theme_bw()
+#### Day 0 predict future toxin
+```{r}
+plot_model_performance('day_0_predict_future_toxin')
+plot_lr_hp('day_0_predict_future_toxin')
+plot_rf_hp('day_0_predict_future_toxin')
+```
+
+#### Day 0 predict moribund outcome
+```{r}
+plot_model_performance('day_0_moribund')
+plot_lr_hp('day_0_moribund')
+plot_rf_hp('day_0_moribund')
+```
+
+#### Day 0 predict high/low clinical score on day 10 
+```{r}
+plot_model_performance('day_0_histology')
+plot_lr_hp('day_0_histology')
+plot_rf_hp('day_0_histology')
 ```
@@
 -  day 0 future toxin presence - taxonomic abundances
+
 -  day 0 future moribund - median cfu, day w/toxin presence (prior to day 3), taxonomic abundances   
-      to avoid duplicating similar data, i sumarized cfu/toxin to be a single feature to represent the amount or presence over the first two days of challenge since all mice were present for at least 2 days
--  day 10 histology (high/mid/low) - cfu, toxin level, taxonomic abundances
+  -    to avoid duplicating similar data, i sumarized cfu/toxin to be a single feature to represent the amount or presence over the first two days of challenge since all mice were present for at least 2 days
+-  day 0 histology (high vs low) - median cfu, days w/toxin, taxonomic abundances
 
@@
 
-performance_df <- read_tsv('../../../data/process/ml/ml_performance.tsv',
+performance_df <- read_tsv(here('data/process/ml/ml_performance.tsv'),
                      col_type = 'dddddddddddddddcdcc') %>% 
-  mutate(dataset = factor(dataset, c('same_day_toxin', 'day_0_predict_future_toxin',
-                                     'day_0_moribund', 'day_10_histology')),
+  filter(dataset != 'day_0_histology_50') %>% 
+  mutate(dataset = gsub('_80', '', dataset),
+         dataset = factor(dataset, c('same_day_toxin', 'day_0_predict_future_toxin',
+                                     'day_0_moribund', 'day_0_histology')),
          taxonomic_level = factor(taxonomic_level, 
                                   levels = c('OTU', 'Genus', 'Family', 'Order', 'Class', 'Phylum')))  
-features_df <- read_tsv('../../../data/process/ml/ml_feature_imp.tsv',
-                     col_type = 'ddcccddcc') 
-hyperparameter_df <- read_tsv('../../../data/process/ml/ml_hp_performance.tsv',
-                     col_type = 'ddcdccc') 
-taxonomy_df <- read_tsv('../../../data/process/final.taxonomy.tidy.tsv',
+features_df <- read_tsv(here('data/process/ml/ml_feature_imp.tsv'),
+                     col_type = 'ddcccddcc') %>% 
+  filter(dataset != 'day_0_histology_50') %>% 
+  mutate(dataset = gsub('_80', '', dataset))
+hyperparameter_df <- read_tsv(here('data/process/ml/ml_hp_performance.tsv'),
+                     col_type = 'ddcdccc') %>% 
+  filter(dataset != 'day_0_histology_50') %>% 
+  mutate(dataset = gsub('_80', '', dataset),
+         dataset = factor(dataset, c('same_day_toxin', 'day_0_predict_future_toxin',
+                                     'day_0_moribund', 'day_0_histology')),
+         taxonomic_level = factor(taxonomic_level, 
+                                  levels = c('OTU', 'Genus', 'Family', 'Order', 'Class', 'Phylum')))   
+taxonomy_df <- read_tsv(here('data/process/final.taxonomy.tidy.tsv'),
                         col_type = cols(.default = col_character()))
 
-metadata <- read_tsv('../../../data/process/metadata_tidy.tsv',
+metadata <- read_tsv(here('data/process/metadata_tidy.tsv'),
 					 col_type = 'cccccDddddDcdddddcdl') %>% 
@@
 
-toxin <- read_tsv('../../../data/process/toxin_tidy.tsv',
+toxin <- read_tsv(here('data/process/toxin_tidy.tsv'),
 				  col_type = 'cccccdc') %>% 
@@
 
-histology <- read_tsv('../../../data/process/histology_tidy.tsv',
+histology <- read_tsv(here('data/process/histology_tidy.tsv'),
 					  col_type = 'ccccdddcdcc') %>% 
@@
 
-shared <- read_tsv('../../../data/mothur/sample.final.0.03.subsample.shared',
+shared <- read_tsv(here('data/mothur/sample.final.0.03.subsample.shared'),
          col_types = cols(.default = col_double(),
@@
 # setup data to predict histology severity of colonized mice (microbiome, cdiff, toxin)
-day_10_histology_df <- metadata %>% 
-	filter(day == 10,
-		early_euth == F) %>% 
-	select(group, C_difficile_CFU = cdiff_cfu, mouse_id) %>% 
-	left_join(select(toxin, group, toxin_level = Log_repiricoal_dilution),
-		by = c('group')) %>% 
+toxin_summary <- toxin %>% 
+	group_by(mouse_id) %>% 
+	summarise(toxin_positive_days = sum(Log_repiricoal_dilution > 1))
+day_0_histology <- metadata %>% 
+	select(group, mouse_id, day) %>% 
+	filter(day == 0) %>% 
+	left_join(cfu_summary, by = c('mouse_id')) %>% 
+	left_join(toxin_summary, by = c('mouse_id')) %>% 
 	inner_join(shared, by = c('group' = 'Group')) %>% 
+	ungroup %>% 
+	filter(early_euth == F) %>% 
 	inner_join(select(histology, mouse_id, hist_score), by = c('mouse_id'))
@@
 ```{r}
-hyperparameter_df %>% 
-  filter(model == 'glmnet') %>% 
-  ggplot(aes(x = value, y = AUC, color = taxonomic_level)) + 
-    geom_point(alpha = 0.25) + 
-    stat_summary(fun = median, geom = 'line') + 
-    ylim(c(0.5,1)) + 
-    facet_grid(dataset~taxonomic_level, scales = 'free') + 
-    theme_bw() + 
-    scale_x_log10()
+# did not run logistic regression in final run as using rf for final model
+#hyperparameter_df %>% 
+#  filter(model == 'glmnet') %>% 
+#  ggplot(aes(x = value, y = AUC, color = taxonomic_level)) + 
+#    geom_point(alpha = 0.25) + 
+#    stat_summary(fun = median, geom = 'line') + 
+#    ylim(c(0.5,1)) + 
+#    facet_grid(dataset~taxonomic_level, scales = 'free') + 
+#    theme_bw() + 
+#    scale_x_log10() +
+#    theme(legend.position = 'none')
 
@@
     theme_bw() + 
-    scale_x_log10()
+    scale_x_log10() + 
+    theme(legend.position = 'none')
 
@@
 ```{r}
-performance_df %>% 
+day_0_toxin_perf_plot <- performance_df %>% 
   filter(dataset == 'day_0_predict_future_toxin') %>% 
@@
     theme_bw() +
-    geom_rect(data = data.frame(ymin = 0.35, ymax = 0.98, method = 'glmnet', taxonomic_level = 2, Performance = 1),
-              aes(xmin = taxonomic_level - 0.5, xmax = taxonomic_level + 0.5, ymin = ymin, ymax = ymax, color = NA), 
-              fill = NA) +
+#    geom_rect(data = data.frame(ymin = 0.35, ymax = 0.98, method = 'glmnet', taxonomic_level = 2, Performance = 1),
+#              aes(xmin = taxonomic_level - 0.5, xmax = taxonomic_level + 0.5, ymin = ymin, ymax = ymax, color = NA), 
+#              fill = NA) +
     geom_rect(data = data.frame(ymin = 0.48, ymax = 0.99, method = 'rf', taxonomic_level = 6, Performance = 1),
@@
               fill = NA)
+day_0_toxin_perf_plot
 ```
@@
   filter(dataset == 'day_0_predict_future_toxin',
-         (method == 'rf' & taxonomic_level == 'Phylum')# | (method == 'glmnet' & taxonomic_level == 'Genus')
-         ) %>% 
+         (method == 'rf' & taxonomic_level == 'Phylum')) %>% 
   group_by(names) %>% 
@@
   filter(median_diff > 0) 
-day_0_toxin_production %>%
+day_0_toxin_feature_plot <- day_0_toxin_production %>%
   ggplot(aes(x = reorder(names,median_diff), y = perf_metric_diff)) + 
@@
 
-day_0_predict_future_toxin_df %>% 
+day_0_toxin_feature_abundance_plot <- day_0_predict_future_toxin_df %>% 
   pivot_longer(cols = starts_with('Otu'), names_to = 'OTU', values_to = 'counts') %>% 
@@
   summarise(counts = sum(counts)/21.07) %>% 
-  mutate(Phylum = factor(Phylum, day_0_toxin_features)) %>% 
-  ggplot(aes(x = Phylum, y = counts, color = toxin)) + 
+  mutate(taxon = factor(Phylum, day_0_toxin_features)) %>% 
+  ggplot(aes(x = taxon, y = counts, color = toxin)) + 
     stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5),
@@
 ```{r}
-performance_df %>% 
+same_day_toxin_perf_plot <- performance_df %>% 
   filter(dataset == 'same_day_toxin') %>% 
@@
     theme_bw() +
-    geom_rect(data = data.frame(ymin = 0.7, ymax = 0.95, method = 'glmnet', taxonomic_level = 6, Performance = 1),
+    geom_rect(data = data.frame(ymin = 0.7, ymax = 0.95, method = 'rf', taxonomic_level = 6, Performance = 1),
               aes(xmin = taxonomic_level - 0.5, xmax = taxonomic_level + 0.5, ymin = ymin, ymax = ymax, color = NA), 
               fill = NA)
+same_day_toxin_perf_plot
 ```
 
-Using Logistic regression at the Phylum level the features with median differences greater than 0 are:
+Using Random Forest at the Phylum level the features with median differences greater than 0 are:
 
@@
   filter(dataset == 'same_day_toxin',
-         method == 'glmnet',
+         method == 'rf',
          taxonomic_level == 'Phylum') %>% 
@@
   filter(median_diff > 0) 
-same_day_toxin %>%   
+same_day_toxin_feature_plot <- same_day_toxin %>%   
   ggplot(aes(x = reorder(names, median_diff), y = perf_metric_diff)) + 
@@
     theme_bw()
-
+same_day_toxin_feature_plot
 ```
@@
 ```{r}
-same_day_toxin_df %>% 
+same_day_toxin_feature_abundance_plot <- same_day_toxin_df %>% 
   pivot_longer(cols = starts_with('Otu'), names_to = 'OTU', values_to = 'counts') %>% 
@@
   summarise(counts = sum(counts)/21.07) %>% 
-  mutate(Phylum = factor(Phylum, same_day_toxin_features)) %>% 
-  ggplot(aes(x = Phylum, y = counts, color = toxin)) + 
+  mutate(taxon = factor(Phylum, same_day_toxin_features)) %>% 
+  ggplot(aes(x = taxon, y = counts, color = toxin)) + 
     stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5),
@@
 ```{r}
-performance_df %>% 
+day_0_moribund_perf_plot <- performance_df %>% 
   filter(dataset == 'day_0_moribund') %>% 
@@
     theme_bw() +
-    geom_rect(data = data.frame(ymin = 0.65, ymax = 1.01, method = 'glmnet', taxonomic_level = 5, Performance = 1),
+    geom_rect(data = data.frame(ymin = 0.65, ymax = 1.01, method = 'rf', taxonomic_level = 5, Performance = 1),
               aes(xmin = taxonomic_level - 0.5, xmax = taxonomic_level + 0.5, ymin = ymin, ymax = ymax, color = NA), 
               fill = NA)
+day_0_moribund_perf_plot
 ```
 
-Using Logistic regression at the Class level the features with median differences greater than 0 are:
+Using random forest at the Class level the features with median differences greater than 0 are:
 
@@
   filter(dataset == 'day_0_moribund',
-         method == 'glmnet',
+         method == 'rf',
          taxonomic_level == 'Class') %>% 
@@
   filter(median_diff > 0) 
-day_0_moribund_plot_df %>% 
+day_0_moribund_feature_plot <- day_0_moribund_plot_df %>% 
   ggplot(aes(x = reorder(names,median_diff), y = perf_metric_diff)) + 
@@
     theme_bw()
+day_0_moribund_feature_plot
 ```
@@
 
-day_0_moribund %>% 
+day_0_moribund_feature_abundance_plot <- day_0_moribund %>% 
   pivot_longer(cols = starts_with('Otu'), names_to = 'OTU', values_to = 'counts') %>% 
@@
   summarise(counts = sum(counts)/21.07) %>% 
-  mutate(Phylum = factor(Class, day_0_moribund_features)) %>% 
-  ggplot(aes(x = Phylum, y = counts, color = early_euth)) + 
+  mutate(taxon = factor(Class, day_0_moribund_features)) %>% 
+  ggplot(aes(x = taxon, y = counts, color = early_euth)) + 
     stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5),
@@
     labs(x = NULL, y = 'Relative Abundance (%)', color  = 'Disease Severity')
-
+day_0_moribund_feature_abundance_plot
 ```
@@
 
-```{r}
+```{r, include = F, eval = F}
 hist_performance_df <- read_tsv('../../../data/process/ml/hist_ml_performance.tsv')
@@
 
-Using multiclass RF at the Phylum level the features with median differences greater than 0 are:
+```{r, include = F, eval = F}
+#Using multiclass RF at the Phylum level the features with median differences greater than 0 are:
 
-```{r}
 hist_features_plot_df <- hist_features_df %>% 
@@
 
-```{r}
+```{r, include = F, eval = F}
 hist_features <- hist_features_plot_df %>% 
@@
 ```
+
+```{r}
+day_0_hist_perf_plot <- performance_df %>% 
+  filter(dataset == 'day_0_histology') %>% 
+  select(method, seed, dataset, cv_auc = cv_metric_AUC, test_auc = AUC, taxonomic_level) %>% 
+  pivot_longer(cols = c(cv_auc, test_auc), 
+               names_to = 'metric', 
+               values_to = 'Performance') %>% 
+  mutate(taxonomic_level = factor(taxonomic_level, 
+                                  levels = c('OTU', 'Genus', 'Family', 'Order', 'Class', 'Phylum'))) %>% 
+  ggplot(aes(x = taxonomic_level, y = Performance, color = metric)) + 
+    geom_boxplot() + 
+    geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
+    #facet_grid(dataset~method, scales = 'free_y') + 
+    theme_bw() +
+    geom_rect(data = data.frame(ymin = 0.48, ymax = 1.01, method = 'rf', taxonomic_level = 2, Performance = 1),
+              aes(xmin = taxonomic_level - 0.5, xmax = taxonomic_level + 0.5, ymin = ymin, ymax = ymax, color = NA), 
+              fill = NA)
+day_0_hist_perf_plot
+```
+
+Using Logistic regression at the Class level the features with median differences greater than 0 are:
+
+```{r}
+day_0_histology_plot_df <- features_df %>% 
+  filter(dataset == 'day_0_histology',
+         method == 'rf',
+         taxonomic_level == 'Genus') %>% 
+  group_by(names) %>% 
+  mutate(median_diff = mean(perf_metric_diff)) %>% 
+  ungroup() %>% 
+  filter(median_diff > 0) 
+day_0_hist_feature_plot <- day_0_histology_plot_df %>% 
+  ggplot(aes(x = reorder(names,median_diff), y = perf_metric_diff)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
+    geom_jitter(alpha = 0.1, width = 0.1) + 
+    coord_flip() + 
+    facet_grid(method~dataset) + 
+    labs(x = NULL, y = 'AUC Diff with feature permuted') +
+    theme_bw()
+day_0_hist_feature_plot
+```
+
+```{r}
+day_0_histology_features <- day_0_histology_plot_df %>% 
+  select(names, median_diff) %>% 
+  separate(names, into = c('A', 'B')) %>% 
+  unique %>% 
+  arrange(median_diff) %>% 
+  pivot_longer(cols = c('A', 'B'), values_to = 'names') %>% 
+  pull(names) %>% 
+  unique
+
+day_0_hist_feature_abundance_plot <- day_0_histology %>% 
+  filter(hist_score %in% c('high', 'low')) %>% 
+  pivot_longer(cols = starts_with('Otu'), names_to = 'OTU', values_to = 'counts') %>% 
+  left_join(taxonomy_df, by = "OTU") %>% 
+  filter(Genus %in% day_0_histology_features) %>% 
+  group_by(group, hist_score, Genus) %>% 
+  summarise(counts = sum(counts)/21.07) %>% 
+  mutate(taxon = factor(Genus, day_0_histology_features)) %>% 
+  ggplot(aes(x = taxon, y = counts, color = hist_score)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5),
+                 position = position_dodge(width = 0.7)) +
+    geom_jitter(position = position_jitterdodge(dodge.width = 0.7, jitter.width = 0.3), 
+                alpha = 0.2) + 
+    scale_y_log10() + 
+    theme_bw() + 
+    coord_flip() + 
+#    scale_color_manual(values = c('red', 'black'),
+ #                      breaks = c('absent', 'present'), 
+  #                     labels = c('False', 'True')) + 
+    labs(x = NULL, y = 'Relative Abundance (%)', color  = 'Disease Severity')
+day_0_hist_feature_abundance_plot  
+```
+
+```{r}
+day_0_toxin_perf_plot
+day_0_hist_perf_plot
+day_0_moribund_perf_plot
+same_day_toxin_perf_plot
+
+ggsave(here('results/figures/Figure_S2.jpg'),
+  cowplot::plot_grid(
+      cowplot::plot_grid(day_0_toxin_perf_plot + coord_flip(), day_0_toxin_feature_plot, nrow = 1),
+      #cowplot::plot_grid(same_day_toxin_perf_plot + coord_flip(), same_day_toxin_feature_plot, nrow = 1),
+      cowplot::plot_grid(day_0_moribund_perf_plot + coord_flip(), day_0_moribund_feature_plot, nrow = 1),
+      cowplot::plot_grid(day_0_hist_perf_plot + coord_flip(), day_0_hist_feature_plot, nrow = 1),
+      ncol = 1),  height = 9, width = 6.875, unit = 'in')
+
+ggsave(here('results/figures/Figure_5.jpg'),
+  cowplot::plot_grid(
+#    cowplot::plot_grid(
+     cowplot::plot_grid(NULL,
+      day_0_toxin_feature_abundance_plot + 
+        theme(legend.position = 'right') + 
+        labs(color = 'Toxin\nDetected'),
+      nrow = 1, rel_widths = c(1, 25)),
+#      same_day_toxin_feature_abundance_plot + 
+#        theme(legend.position = 'none'),
+#      ncol = 1),
+#    cowplot::plot_grid(  
+      day_0_moribund_feature_abundance_plot + 
+        theme(legend.position = 'right') + 
+        labs(color = 'Disease\nSeverity'),
+     cowplot::plot_grid(NULL,     
+      day_0_hist_feature_abundance_plot + 
+        theme(legend.position = 'right') + 
+        scale_color_manual(values = c('dark green', 'light green')) + 
+        labs(color = 'Clinical\nScore'),
+      nrow = 1, rel_widths = c(1, 20)),
+      rel_heights = c(9, 15, 14),
+      labels = c('A', 'B', 'C'),
+      ncol = 1), 
+#    nrow = 1),
+  height = 9, width = 6, unit = 'in')
+
+rf_performance <- performance_df %>% 
+  filter((dataset == 'day_0_predict_future_toxin' & taxonomic_level == 'Phylum') | 
+         (dataset == 'day_0_moribund' & taxonomic_level == 'Class') |
+         (dataset == 'day_0_histology' & taxonomic_level == 'Genus')) %>% 
+  group_by(dataset) %>% 
+  summarise(AUC = round(mean(AUC), 2)) %>% 
+  pivot_wider(names_from = dataset, values_from = AUC)
+
+```
+
+## What are the temporal dynamics of the taxa identified in the Lefse/RF models?
+
+```{r}
+taxonomy_df %>% 
+  filter(Genus == 'Helicobacter') %>% 
+  left_join(pivot_longer(shared, 
+                         cols = -Group, names_to = 'OTU', values_to = 'counts')) %>% 
+  left_join(metadata, by = c('Group' = 'group')) %>% 
+  #filter(early_euth == F) %>% 
+  group_by(mouse_id, day, Genus, early_euth) %>% 
+  summarise(counts = sum(counts)) %>% 
+  inner_join(select(histology, mouse_id, hist_score), by = 'mouse_id') %>% 
+  mutate(outcome = ifelse(early_euth, 'severe', hist_score),
+         outcome = factor(outcome, levels = c('low', 'high', 'severe'))) %>% 
+  filter(outcome != 'mid') %>% 
+  ggplot(aes(x = day, y = counts, group = interaction(day, outcome), color = outcome)) + 
+    geom_boxplot() + 
+    facet_wrap(Genus~.) + 
+    theme_bw()
+
+taxonomy_df %>% 
+  filter(Family %in% c('Enterobacteriaceae', 'Pseudomonadaceae', 'Neisseriaceae', 'Sutterellaceae')) %>% 
+  left_join(pivot_longer(shared, 
+                         cols = -Group, names_to = 'OTU', values_to = 'counts')) %>% 
+  left_join(metadata, by = c('Group' = 'group')) %>% 
+  #filter(early_euth == F) %>% 
+  group_by(mouse_id, day, Family, early_euth) %>% 
+  summarise(counts = sum(counts)) %>% 
+  inner_join(select(histology, mouse_id, hist_score), by = 'mouse_id') %>% 
+  mutate(outcome = ifelse(early_euth, 'severe', hist_score),
+         outcome = factor(outcome, levels = c('low', 'high', 'severe'))) %>% 
+  filter(outcome != 'mid') %>% 
+  group_by(Family) %>% 
+  mutate(present = sum(counts > 0)) %>% 
+  filter(present > 30) %>% 
+  ggplot(aes(x = day, y = counts, group = interaction(day, outcome), color = outcome)) + 
+    geom_boxplot() + 
+    facet_wrap(Family~., scales = 'free_y') + 
+    theme_bw()
+
+plot_temporal_genera <- function(taxa_list){
+  taxonomy_df %>% 
+    filter(Genus %in% taxa_list) %>% 
+    left_join(pivot_longer(shared, 
+                           cols = -Group, names_to = 'OTU', values_to = 'counts')) %>% 
+    left_join(metadata, by = c('Group' = 'group')) %>%
+    group_by(mouse_id, day, Genus, early_euth) %>% 
+    summarise(counts = sum(counts)) %>% 
+    inner_join(select(histology, mouse_id, hist_score), by = 'mouse_id') %>% 
+    mutate(outcome = ifelse(early_euth, 'severe', hist_score),
+           outcome = factor(outcome, levels = c('low', 'high', 'severe'))) %>% 
+    filter(outcome != 'mid') %>% 
+    group_by(Genus) %>% 
+    mutate(present = sum(counts > 0)) %>% 
+    filter(present > 30) %>% 
+    ggplot(aes(x = day, y = counts, group = interaction(day, outcome), color = outcome)) + 
+      geom_boxplot() + 
+      facet_wrap(Genus~., scales = 'free_y') + 
+      theme_bw()
+}
+
+# manual selection/exploration
+plot_temporal_genera(c('Limosilactobacillus', 'Ihubacter', 'Anaeroplasma', 
+                      'Murimonas', 'Sellimonas', 'Lactonifactor', 
+                      'Enterocloster', 'Akkermansia', 'Clostridium', 
+                      'Anaerostipes', 'Monoglobus', 'Coprobacillus'))
+
+plot_temporal_genera(c('Bacteroides', 'Prevotellaceae_unclassified', 'Parabacteroides',
+                  'Muribaculaceae_unclassified', 'Duncaniella', 'Phocaeicola'))
+
+plot_temporal_genera(c('Bifidobacterium', 'Eggerthella', 'Deinococcus',
+                'Ligilactobacillus', 'Clostridium_IV', 'Turicibacter',
+                'Clostridium_sensu_stricto'))
+```
+
+#### identify temporal patterns differentiating high/low clinical score
+
+potential last figure or supplemental to show significant temporal trends
+
+```{r}
+
+#lefse_ temporal trend hi/low day 0-10
+lefse_hilo <- c("Anaeromassilibacillus", "Anaeroplasma", "Bacteroidales_unclassified", "Bifidobacterium", "Blautia", "Clostridioides", "Clostridium_IV", "Clostridium_XlVa", "Duncaniella", "Escherichia/Shigella", "Faecalibacterium", "Fournierella", "Helicobacter", "Hungatella", "Ihubacter", "Intestinimonas", "Klebsiella", "Kosakonia", "Lactobacillales_unclassified", "Legionella", "Muribaculaceae_unclassified", "Neglecta", "Odoribacter", "Oscillibacter", "Paraclostridium", "Paramuribaculum", "Prevotellaceae_unclassified", "Ruthenibacterium")
+
+temporal_lefse_plot <- taxonomy_df %>% 
+  filter(Genus %in% lefse_hilo) %>% 
+  left_join(pivot_longer(shared, 
+                         cols = -Group, names_to = 'OTU', values_to = 'counts')) %>% 
+  left_join(metadata, by = c('Group' = 'group')) %>%
+  group_by(mouse_id, day, Genus, early_euth) %>% 
+  summarise(counts = sum(counts)) %>% 
+  inner_join(select(histology, mouse_id, hist_score), by = 'mouse_id') %>% 
+  mutate(outcome = ifelse(early_euth, 'severe', hist_score),
+         outcome = factor(outcome, levels = c('low', 'high', 'severe')),
+         rel_abun = counts/2107 * 100) %>% 
+  filter(outcome != 'mid') %>% 
+  group_by(Genus) %>% 
+  mutate(present = sum(counts > 0)) %>% 
+  filter(present > 60) %>% 
+  ggplot(aes(x = day, y = rel_abun, group = interaction(day, outcome), color = outcome)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5),
+                 position = position_dodge(width = 0.7)) +
+    geom_hline(yintercept = 1/21.07, linetype = 'dashed', size = 0.25) + 
+    facet_wrap(Genus~., scales = 'free_y') + 
+    scale_y_log10() + 
+    theme_bw()
+
+ggsave(here('results/figures/Figure_XX_temporal_lefse.jpg'),
+       temporal_lefse_plot,
+       height = 9, width = 6)
+```
+
+```{r}
+# presentation plot
+#
+#strain_exp_donor_selection_plot <- taxonomy_df %>% 
+#  filter(Genus %in% c('Anaerotignum',
+#                      'Hungatella', 'Eisenbergiella', 'Clostridium_sensu_stricto', 
+#                      'Enterocloster')) %>% 
+#  left_join(pivot_longer(shared, 
+#                         cols = -Group, names_to = 'OTU', values_to = 'counts')) %>% 
+#  left_join(metadata, by = c('Group' = 'group')) %>%
+#  filter(day == 0) %>% 
+#  group_by(human_source, mouse_id, Genus, early_euth) %>% 
+#  summarise(counts = sum(counts)) %>% 
+#  inner_join(select(histology, mouse_id, hist_score), by = 'mouse_id') %>% 
+#  mutate(Genus = gsub('_', ' ', Genus),
+#  	     Genus = paste0('*', Genus, '*'),
+#         outcome = case_when(early_euth == T ~ 'Severe', 
+#                             hist_score == 'low' ~ 'Low Clinical Score',
+#                             hist_score == 'high' ~ 'High Clinical Score'),
+#         outcome = factor(outcome, levels = c('Low Clinical Score', 'High Clinical Score', 'Severe')),
+#         relative_abundance = counts/2107 * 100,
+#         relative_abundance = ifelse(relative_abundance == 0, 0.035, relative_abundance)) %>% 
+#  filter(human_source %in% c('DA00369', 'DA00430', 'DA00578')) %>% 
+#  mutate(donor_source = case_when(human_source == 'DA00369' ~ 'F',
+#                                    human_source == 'DA00430' ~ 'G',
+#                                    human_source == 'DA00578' ~ 'M'),
+#         donor_source = factor(donor_source, levels = c('G' ,'F', 'M'))) %>% 
+#  ggplot(aes(x = donor_source, y = relative_abundance, color = outcome)) + 
+#    geom_point() + 
+#    geom_hline(yintercept = 100/2107, linetype = 'dashed', size = 0.25) +
+#    facet_grid(.~Genus) + 
+#    scale_color_manual(breaks = c('Low Clinical Score', 'High Clinical Score', 'Severe'),
+#                       values = c("#23980d", "#e26df8", "#154975")) + 
+#    scale_y_log10(breaks = c(0.1, 1, 10), labels = c('0.1', '1', '10')) + 
+#    theme_bw() +
+#    labs(x = 'Donor Source', y = 'Relative Abundance', color = NULL) +
+#    theme(legend.position = 'bottom',
+#          strip.text = ggtext::element_markdown(),
+#          strip.background = element_blank())
+#
+#
+#
+#ggsave('~/Desktop/donor_selection.jpg',
+#       strain_exp_donor_selection_plot,
+#       height = 4, width = 8)
+```

### 2021-10-21 submission/manuscript.Rmd
initial draft of manuscript

---
csl: mbio.csl
fontsize: 12pt
output:
  word_document: default
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
bibliography: references.bib
---

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}
library(knitr)
library(readr)

opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("cache" = FALSE)
opts_chunk$set("message" = FALSE)

```

\pagenumbering{arabic}
\linenumbers
\doublespacing

# Gut microbiota drives differences in Clostridiodes difficile infection severity.

\vspace{30mm}

**Running title:** Microbiota drives *Clostridioides difficile* infection severity

\vspace{20mm}

Nicholas A. Lesniak$^1$, Alyxandria M. Schubert$^1$, Hamide Sinani$^1$, Patrick D. Schloss$^{1,\dagger}$

\vspace{30mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu  
1\. Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI

\newpage

## Abstract

Clostridioides difficile infection has become more common and severe in the last few decades. Patient age, white blood cell count, creatinine levels as well as C. difficile ribotype and the presence of toxin genes have been associated with disease severity. However, models based on those data to identify patients with CDI that will develop severe disease have not been robust enough to be broadly applied. These models are input from our understanding of C. difficile interactions with the gut environment. The gut bacteria are a key determinant in C. difficile infection. CDI is dependent on perturbations in the bacterial community in order to colonize the gut. The gut microbiota also impairs C. difficile colonization through bile acid metabolism,  nutrient consumption and bacteriocin production. However, it is unclear if the gut bacteria affect the disease severity resulting from that colonization. Here we have shown gut bacterial communities contribute to disease severity variation. We created diverse gut communities by colonizing germ-free mice with human fecal communities. They were then infected with C. difficile which resulted in differed disease severities. The severity grouped by the human fecal community they received. Generally, facultative anaerobes with pathogenic potential, such as Escherichia, Helicobacter  and Klebsiella, were associated with more severe disease. Bacterial groups associated with dietary fiber degradation, such as Coprobacillus, were associated with reduced disease severity. Lastly, we showed communities that resulted in either low or high histopathologic scores or severe disease had the same outcome when infected with a different C. difficile isolate.

## Importance

Clostridioides difficile infection (CDI) causes a range of disease from asymptomatic to mild diarrhea to severe outcomes of recurrent infections and even death. Models that predict severity and treatment decisions are based on clinical factors and C. difficile capabilities. But currently the gut microbiome, the primary protector against CDI, has not been accounted for its effect on CDI disease severity. We demonstrated variation in the microbiome of mice colonized with human feces was sufficient to result in a range of disease severities. Our results show groups of bacteria contribute to the development of disease in C. difficile infections. Gut bacterial community data from patients with CDI could improve our ability to identify patients more at risk for severe disease and improve interventions which treat the gut bacteria to reduce host damage.

\newpage

## Introduction

Clostridioides difficile infections (CDI) have been increasing in incidence and severity for the past two decades. *C. difficile*, first identified in 1931, wasn't categorized as a pathogen until the late 1970s after it was determined to be the driver of antibiotic-associated psuedomembranous colitis following the introduction of broad-spectrum antibiotics. In the early 2000s CDI became more problematic with increasing incidences, severity and death. People colonized with *C. difficile* can experience a range of disease severities from asymptomatic to mild diarhea to toxic megacolon and death. CDI is primarily treated with a course of antibiotics, either metronidazole or vancomycin depending on CDI severity, but can also be treated with cessation of other antibiotics for milder cases or with a fecal microbiota transplantion for CDI cases recalitrant to antibiotics.

The degree of CDI disease severity has been connected to aspects of C. difficile strain and host factors. Since the first identification of C. difficile, more virulent strains have emerged, such as ribotype 027. Virulence is increased by the presence of the genes for Toxin A/B or fluoroquinolon resistence. The binary toxin, while not sufficient on its own to allow virulence, has been associated with increased virulence in the presence of Toxin A/B. Mutations in tcd C, the negative regulator of of toxin transcription, also increased the virulence of the C. difficile strain. In addition to the virulent capability of the colonizing strain, the immune response of the person infected also affects the severity of the disease. The delayed or reduced production of antibodies IgG and IgA can result in more severe disease. Increased neutrophil inflitration can lead to worse outcomes. Beyond the immune response, increased density of the toxin receptor on the host cells can increase the severity of the infection.

There has been great interest in the interactions of *C. difficile* and the intestinal bacterial community in CDI. Indigenous bacteria of a healthy intestinal community provide a protective barrier preventing *C. difficile* from colonizing and causing infection. The disruption of this community from a range of pertubations, such as many classes of antibiotics, medications, or even dietary changes, allows C. difficile to colonize. We have shown that different pertubations lead to differences in community structure and the ability for C. difficile to maintain colonization. C. difficile has also been shown to modify its metabolism and expression of virulence genes in response to the nutritional environment. However, it is unclear if the bacterial community affect the disease severity of CDI. Here, we tested if variablity in the intestinal bacterial community can produce and explain variation in the disease severity caused by a single C. difficile strain, ribotype 027.

## Results

**Germ-free mice colonized with human fecal commmunities have diverse and unique structures.** Based on our previous observation mice with variation in their microbiota showed differences in colonization of *C. difficile*, we sought to explore the affect communties with greater variation had on *C. difficile* challenge. To produce communities with greater variation than mice with their indigenous communties, we used human fecal communities to establish the gut commuities in germ-free mice. We inoculated a cage of germ-free C57BL/6 mice (2 -4 mice per cage) with homogenized feces from one of 15 different human donors via oral gavage. The fecal community was allowed to colonize and stabilize for two weeks post inoculation. We then assessed the gut bacterial communities by sequencing the V4 region of the 16S rRNA gene extracted from fecal pellets. The communities established in the mice group most closely to other mice recieving the same donor than to either their respective donor or any mice who received stool from a different donor (Figure 1A). Most of the communities  highly abundant classes were *Clostridia*, *Bacteroidia*, *Erysipelotrichia*, *Bacilli*, and *Gammaproteobacteria*. However, each group of mice harbored unique combination of class abunadances in their gut bacterial communities.
**Any alpha? compare to murine variation? Number of unique otus?**

**C. difficile is able to colonize mice without perturbation.** Our goal is to test the effect of the variation in the microbiota on the effect of a *C. difficile* challenge. A typical mouse model of CDI requires pretreatment with antibiotics such as clindamycin to become susceptible to *C. difficile* colonization [CITATION]. However, we wanted to avoid modifying the communties with an antibiotic to maintain the community structures and differences. Since these communties came from human donor which may have become susceptible through their own exposures, we decided to test if *C. difficile* was able to colonize these communities without further treatment. We hypothesized that *C. difficile* would most likely be able to colonize the mice with communities that came from donors which had a history of antibiotic use, one of the primary risk factors for developing CDI (TABLE, CITATION). After two weeks post incoluation with human fecal communities, the mice were then challenged with 10^**XXXXX** *C. difficile* ribotype 027 spores (**Add schematic?**). The mice were followed for 10 days post challenge and their stool was collected and plated for *C. difficile* CFU to detect colonization level. Suprisingly, communities from all donors were colonized (Figure 2). The only two mice that were able to resist *C. difficile* colonization both recieved their community from the same donor. Therefore, the transplant communties were susceptible to colonization without further modification of the communities.

**Infection seveity varies by initial community structure.** After challenging the mice with *C. difficile*, we looked at the effects of the different communities on the outcome from the *C. difficile* challenge. We followed the mice for signs of disease as well as *C. difficile* colonization and toxin production for 10 days post challenge. Some mice were not initally colonized on the day after challenge and took a few days to become fully colonized (**SUPP FIG?**). There were a group of mice, all from a set of 6 donors, highly colonized one day post challenge that became moribund within 3 days post challenge. The remaining mice, except for one cage, maintained *C. difficile* colonization through the end of the experiment (Figure 2). After the  mice were humanely euthanised on the experiment endpoint, day 10 post challenge or earlier if they succumbed to the infection, their cecal tissue was collected and scored for histological signs of disease and fecal samples were assayed for toxin (Figure 3). Overall, there was greater toxin activity detected in the stool of the moribund mice (*P* = 0.003). However, when looking at each group of mice, we saw there was a range in toxin activity for both the mice with mild disease and moribund mice (Figure 3A). Some donor groups that maintained colonization without severe disease had similarlly high activity levels as was detected in the moribund mice. Additionally, some moribund mice did not have toxin activity detected in their stool. These oberserved variations in toxin activity and relation to disease severity have been previously characterized [@Akerlund2006, @Vitucci2020]. Next we examined the histological damage caused by the *C. difficile* infection. For mice that succumb to the infection prior to day 10 post challenge, we observed high levels of epithelial damage, tissue edema and inflammation, similar to previously reported histological findings for *C. difficile* ribotype 027 [CITATION, @Cowardin2016]. As observed with toxin activity, the moribund mice had higher disease clinical scores than the mice with mild disease (*P* = 3.0e-9). However, unlike the toxin activity, all moribund mice had consistently high clinical scores. The mice which maintained persistent level of colonization, we saw a range in tissue damage from no signs of disease to nearly the same level as the moribund mice, which grouped by community donor. Together, the toxin activity, clinical score, and moribundity showed variation across the donors but were largely consistent within each group of mice recieving the same human fecal community.

**Populations of the microbial communities explain variation in disease severity.** Since the disease charateristics grouped by human fecal community donor, we next looked if the composition of the community could explain the variation we observed. First, we tested for associations between taxonomic groups and the outcomes of the *C. difficile* challenge. We used the linear discriminant analysis (LDA) effect size (LEfSe) method to identify the Genus level differences in the initial bacterial communities between the severity level and toxin production. We split the mice into groups by severity level based on their if they became moribund (severe), or if their clinical score was above 5 (moderate) or below 5 (mild). This analysis revealed 20 genera that were significantly different by the disease severity (Figure 4A). The presence of a group of populations of *Turicibacter*, *Streptococcus*, *Staphylococcus*, *Pseudomonas*, *Phocaeicola*, *Parabacteroides*, *Bacteroides*, and *Escherichia/Shigella* were detected at higher levels in the moribund group. Populations of *Anaerotignum*, *Coprobacillus*, *Enterocloster*, and *Murimonas* were increased in the group of mice that only had mild disease. There were many genera that did not have a correlation of relative abundance and disease severity, such as the populations of *Monoglobus* and *Hungatella*, which were decreased or increased, respectively, in only the group of mice with high clinical scores but did not succumb to the infection. To understand these cases better we used LEfSe to identify the genera differentiating the communities that we detected toxin in from those that did not (Figure 4B). There were many genera the were associated with toxin production as we would expect from their association with severe disease, such as populations of *Escherichia/Shigella* and *Bacteroides*. Likewise, there were genera such as *Anaerotignum*, *Enterocloster*, and *Murimonas* that were higher in communities that had mild disease and no toxin was produced. However, communities without a linear trend between relative abundance and disease severity can be better understood by their association with toxin production. Populations of *Hungatella* were increased in the group with moderate disease but not with the production of toxin. Lastly, we looked for associations between the clinical score and the genera populations at that same time (Figure 4C). The populatons of *Bacteroides* at the end matched the trend observed in the initial community, increased population correlated with increased clinical score. Populations of *Klebsiella* and *Prevotellaceae* positively correlated with clinic score and were increased in the group that produced toxin. These tests have shown which individual populations associated with the different outcomes.

We next wanted to understand how the populations in the context of each other lead to the different outcomes from CDI. We trained random forest models with the intital bacterial community relative abundance data at each taxonomic rank to predict toxin production, moribundity or final clinical score. Overall for predicting toxin production, microbial populations aggregated by their phylum level classificiation performed as well as models using lower taxonomic ranks (AUROC = 0.83, Figure S2). The initial community prior to challenge, populations of *Verrucomicrobia*, *Campilobacterota* and *Proteobacteria* contributed the most to the correct prediction of toxin production (Figure 5A). *C. difficile* was more likely to produce toxin when the initial community had smaller populations of *Verrucomicrobia* and *Campilobacterota* and had larger popluations of *Proteobacteria*. Next, we looked at the ability of the community to predict moribundity. The community grouped at the class rank was sufficient to predict if the mouse would succumb to the infection before the end of the experiment (AUROC = 0.91, Figure S2). Groups of bacteria belonging to *Bacilli*, *Firmicutes*, and *Erysipelotrichia* contributed the most to the correct prediction. Larger populations of *Bacilli* and *Firmicutes* and reduced populations of *Erysipelotrichia* were more likely to result in moribundity (Figure 5B). Aside from *Erysipelotrichia*, the only other population decreased in moribund mice was *Clostridia*, all other features had increased populations in the moribund mice. Lastly, the initial community was able to predict if the mice would have a clinical score above or below 5 with the relative abundances from the genus rank (AUROC = 0.99, Figure S2). There were no genera with a significantly greater effect on the model performance than any others, indicating the model is reliant on many genera for the correct prediction. The model used some of the genera identified in the LEfSe analysis, such as populations of *Coprobacillus*, *Anaerostipes*, and *Hungatella*. Communities with greater populations of *Hungatella*, *Eggerthella*, *Bifidobacterium*, *Duncaniella* and *Neisseria* were more likely to have higher clinical scores.

**Disease severity consistent by donor community across strains/isolates.** We used a single *C. difficile* isolate to characterize the range of disease severity and the features of the bacterial community associated with severity. Based on these results, we hypothesized the community was responsible for the disease severity not the specific *C. difficile* isolate we used in our experiments. To test this hypothesis, we selected three communities based on the result from our analysis thus far to select one community we would expect to result in a low clinical score (< 5), a high clinical score (> 5), and one with a high clinical score which became moribund. Using genus level data to select the communities, we selected communities based on the relative abundance patterns of *Sellimonas* and *Anaerotignum* (higher abundance in mice with a low clinical score), *Lachnospiracea* (lower abundance in moribund mice), *Hungatella* and *Eisenbergiella* (higher abundance in moderate mice and lower in moribund mice), Clostridium sensu stricto (higher in moribund mice by both LEfSe analysis and RF model) and *Enterocloster* (lower in moribund mice and a negative linear trend with disease severity by LEfSe analysis). With this set of human fecal communities selected, we inoculated mice via oral gavage of the human fecal slurry and allowed two weeks for the a community to establish. We then challenged the mice with a different *C. difficile* [@Carlson2013]. The disease severity for this set of experiments closely matched that in our original set which were challenged with *C. difficile* isolate 431 (Figure 6).

## Discussion

Challenging mice colonized with different human fecal communities with a single *C. difficile* isolate allowed us to identify the effect of microbiome variation on the disease severity caused by *C. difficile* infection. Our LEfSe analyses and random forest models revealed the relationships between the bacterial community members and the CDI disease severity. We used the relative abundance patterns of community members with strong associations with disease outcome to select communities to test if the predicted severity were *C. difficile* isolate specific. The disease outcome from the *C. difficile* chalenge with the new isolate matched the expected disease severities. Overall, these results have demonstrated that the bacterial community is capable of modulating the disease severity of a *C. difficile* infection.

CDI severity can range from no symptoms, mild symptoms such as diarrhea, or critical symptoms such as pseudomembranous colitis, toxic megacolon or death. At the clinical level, physicians use prediction tools that identify patients most at risk of developing a severe infection using white blood cell counts, serum albumin level, or serum creatine level [@Lungulescu2011, @McDonald2018, @Zar2007]. Those levels are being driven by the activity in the intestine [@diMasi2018]. Research into the drivers of this variation have revealed factors that make *C. difficile* more virulent. Strains are categorized for their virulence by the presence and production toxin A and toxin B and the prevelance in outbreaks, such as 027 and 078 [@AbernathyClose2020, @Theriot2011, @Goorhuis2008, @OConnor2009, @Chen2008]. However, other studies have shown that virulence is not necessarily linked with toxin production [@Roa2015] or the strain [@Walk2012]. Furthermore, there is variation in the genome, growth rate, sporulation, germination, and toxin production in different isolates of a strain [@Carlson2013, @He2010]. These variations may help explain why the severe CDI prediction tools often miss identifying many patients with CDI that will develop severe disease [@Dieterle2020, @Butt2013, @Perry2021, @vanBeurden2017]. Therefore, it is necessary to gain a full understanding of all factors contributing to disease variation to improve our ability to predict severity.

The state of the intestinal bacterial community determines the ability of *C. difficile* to colonize, persist and even cause recurrent infections. *C. difficile* is unable to colonize an unperturbed healthy gut community and is only able to entry after a perturbation [@Schubert2015]. Once colonized, the different communities lead to different metabolic responses and dynamics of the *C. difficile* population [@Jenior2017, Jenior2018, @Lesniak2021]. The gut bacteria metabolize primary bile acids into secondary bile acids [@Staley2016, @Long2017]. The concentration of these bile acids affect the germination, growth, toxin production and biofilm formation [@Sorg2008, @Sorg2010, @Thanissery2017, @Dubois2019]. Members of the bacterial community also affect other metabolites *C. difficile* utilizes. *Bacteroides thetaiotaomicron* produce sialidases which release sialic acid from the mucosa for *C. difficile* to utilize [@Ng2013, @Ferreyra2014]. The nutrient environemt affects toxin production [@MartinVerstraete2016]. Thus, many of the actions of the gut bacteria modulate *C. difficile* in ways that would affect the disease severity driven by CDI.

The gut community has not been demostrated to modulate CDI disease severity. Myriad studies have explored the relationship between the microbiome and CDI but none have ventured into the variation of disease severity due to the microbiome. Most CDI studies employ either an *in vitro* or *in vivo* model using a homogenous bacterial community. Collins et al. used multiple human communities to colonize mice, however the communities were pooled prior to gavaging into germ-free mice [@Collins2015], resulting in a single community. Studies examining difference in disease often use different strains in mice with similar microbiota as a proxy for variation in disease, such as 630 for non-severe and 027 for severe [@AbernathyClose2020, @Theriot2011, @Lawley2012, @Chen2008]. There has been studies demonstrating vairation in severity, but by tapering antibiotic dosage [@Lesniak2021, @Schubert2015, @Reeves2011] or by reducing the amount of cells or spores used for the challenge [@Chen2008, @Reeves2011]. Our group has recently been uncovering the variation the microbiome affects CDI but was limited to *C. difficile* colonization [@Lesniak2021, @Tomkovich2020].

With our recent observations that the initial community affects the ability of *C. difficile* to persist in an environment and the existing research describing the myriad ways the microbiome affects the lifestyle of *C. difficile*, we hypothesized that the microbiome directly and indirectly modulates the disease severity of CDI. Our data demostrate bacterial community member abundances associate with variation in toxin production, histological scoring of the intestine and severe disease. This anaylsis has revealed populations of *Akkermansia*, *Anaerostipes*, *Coprobacillus*, *Enterocloster*, *Lactonifactor*, and *Monoglobus* were more abundant in the microbiome of mice which little to no disease was detected. These genera providing a protective role are supported by previous studies. *Coprobacillus*, *Lactonifactor*, and *Monoglobus* have been shown to be involved in dietary fiber fermentation and associated with healthy communities [@Mabrok2011, @Kim2019, @doPrado2021, @Muthuramalingam2019]. *Anaerostipes* and *Coprobacillus* produce short chain fatty acids which also have been associated with healthy communities [@Han2019, @Duncan2004, @Ye2018]. Furthermore, *Coprobacillus*, which is abundant in mice with low clinical disease scores but rare in all other mice, has been shwon to contain putative type I lantibiotic gene cluster and inhibit *C. difficile* colonization [@Walsh2017, @Sandiford2018, @Stein2013]. *Akkermansia* and *Enterocloster* were also idenitified as more abundant in mice which had a low clinical disease but have contridictory supporting evidence in the current literature. In our data, *Akkermansia* is most abundant in the mice with least disease but there were some mice with severe disease which had increased populations of *Akkermansia*. This could be attributed to either a more protective mucus layer had developed in the mice with little disease and *Akkermansia* is inhibiting colonization [@Stein2013, @Nakashima2021] or *Akkermansia* could be crossfeeding or exposing a niche via its mucus consumption [@Geerlings2018, @Deng2018, @Engevik2020]. Similarly with *Enterocloster*, in our data this genus was more abundant and associated with low clinical disease scores. This genus has been associated with healthy populations and has been used to mono-colonize germ-free mice to reduce the ability of *C. difficile* to colonize [@Reeves2012, @Ma2021]. However, *Enterocloster* has mostly been indentified in infections that result in disease [@Haas2020, @Finegold2005]. These data demostrate a population of bacteria has the potential to be either protective or harmful. The disease outcome is not based on the abundance of a single population of bacteria rather the result of the interactions between the host, microbiome and infecting bacteria.

The groups of bacteria that were associated with either a higher clinical disease score or severe disease are members of an indigenous gut community that also have been associated with disease. These are often referred to as opportunitic pathogens, however based on the the results reported in this study it is more appropriate to apply the damage-response framework for microbial pathogenesis [@Pirofski2008, @Casadevall2014]. The disease is not driven by a single entity, rather it is an emergent property of the responses of the host immune system, infecting microbe and the microbes of the indigenous community. Many of the populations with pathgenic potential that associated with worse outcomes are also facultative anaerobes.  *Enterococcus*, *Klebsiella*, *Shigella/Escherichia*, *Staphylococcus*, and *Streptococcus* have been shown to expand their population with antibiotic use [@VanInsberghe2020, @GarzaGonzalez2019, @Schubert2014] and are commonly detected in CDI cases [@Shafiq2020, @Keith2020, @Zackular2016, @Berkell2021]. In addition to these populations, another set of bacteria, *Eggerthella*, *Prevotellaceae* and *Helicobacter*, associated with worse outcomes have also been associated with intestinal inflammation [@Gardiner2014, @Iljazovic2020, @Nagalingam2013]. *Helicobacter hepaticus* has been demostrated to be sufficient to cause susceptibility to CDI in IL-10^(-/-) B57BL/6 mice [@AbernathyClose2021]. In our mice, if *Helicobacter* was detected the mouse resulted in either severe disease or a high clinical score. While we did not use IL-10^(-/-) mice, it is possible the bacterial communtity or host response are similarly modified by *Helicobacter* allowing C. difficile infection and resulting in host damage.

The damage-response framework for pathogenesis best explains how bacterial populations identified in this study potentiate CDI disease severity through the varied interactions with the host and *C. difficile*. In this set of experiments, we use the same host, C57BL/6 mice, the same infecting microbe, *C. difficile* RT027 isolate 431, with different gut bacterial communities. The members of those communities were often present in multiple levels of disease severity. Thus, it is not merely, the presence of the microbe but their activity in response to the other microbes and host which determine their effect on host damage. Additionally, while we are using the same host and infecting microbe, they too are reacting to the specific members of the bacterial community. Disease severity is driven by the cummulative effect of the host immune response and the activity of the gut bacteria and *C. difficile*. *C. difficile* can be driving host damage through the production of toxin. The gut microbiota could be contributing to the host damage through the balance of metabolic and competitive activities, such as bacteriocin production or mucin degradation. Low levels of mucin degradation can provide nutrients to other community members producing a diverse non-damaging community []. However, if mucin degradation becomes too great that it reducing the its protective function and exposes the epithelial cells []. This over-harvesting contributes to the host damage due to other members producing toxin. Thus disease is a balance of many activities of the community. Host damage is the emergent property of numerous damage-response curves, one for host immune response, one for *C. difficile* activity and another for microbiome community activity, each of which are a composite curve of the individual activities from each group, such as antibody production, neutrophil infiltration, toxin production, sporulation, fiber and mucin degradation, etc.

Many studies have investigated individual activities contributing to CDI. Here we have shown myraid bacterial groups and their associated activities result in variation in CDI disease severity. We must continue to explore how the host, microbiota, and *C. difficile* interactions result in severe disease. We can reduce the risk of severity with a better understanding of which interactions, whether be driven by specific community functions or groups of bacteria, reduce the host damage. Approaching this problem from a damage-response framework allows us to experiment with modifying subsets of activities to reduce overall host damage. Our current treatment approach treats an individual group in this tripartite system. Most commonly, CDI is treated with an antibiotic to eliminate *C. difficile*. Alternatively, another treatment uses antibodies to neutralize *C. difficile* toxins. These treatments are only addressing a single entity of the system, leaving the potential for the other two maintain responses that result in continued host damage. However we can continue this research to elucidate treatments that drive all three entities towards reduced host damage. When a patient is diagnosed with CDI, the gut community composition, in addition to information on the immune response and *C. difficile* ribotype, can inform the predicted severity and appropriate treatment. Treating the microbiome may contribute to reduced severity. From our results, promoting fiber metabolizing bacteria and reducing facultative anerobes could bolster the efficacy of current CDI treatments, reducing disease severities.

## Materials and Methods

**Animal care.** 5.
64 mice, 33F and 31M, ages 6-13 weeks old, 3-4 mice/group (except 2 for C/DA00581 and MOUT/DA10093) each run once except for DA00581 which was run 4 times because we observed colonization resistance in one of the experiments 
15 human fecal samples, selected to create a subset list with diverse responses to health questionaire

***C. difficile* challenge.** M.

**Sample collection.** F.

**DNA sequencing.** T.

**Sequence curation.** S.

**Statistical analysis and modeling.** D.

**Code availability.** Scripts necessary to reproduce our analysis and this paper are available in an online repository (https://github.com/SchlossLab/Lesniak_XXXX).

**Sequence data accession number.** All 16S rRNA gene sequence data and associated metadata are available through the Sequence Read Archive via accession XXXX.

## Acknowledgements

This work was supported by several grants from the National Institutes for Health R01GM099514, U19AI090871, U01AI12455, and P30DK034933. Additionally, NAL was supported by the Molecular Mechanisms of Microbial Pathogenesis training grant (NIH T32 AI007528).
The funding agencies had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.

\newpage

## References

<div id="refs"></div>

\newpage

![](../results/figures/figure_1.jpg)

**Figure 1. R.** (A-C).

\hfill\break

![](../results/figures/figure_2.jpg)

**Figure 2. M.** F.

\hfill\break

![](../results/figures/figure_3.jpg)

**Figure 3. O.** F.

\hfill\break

![](../results/figures/figure_4.jpg)

**Figure 4. E.** A.

\hfill\break

![](../results/figures/figure_5.jpg)

**Figure 5. D.** (A) L

\hfill\break

![](../results/figures/figure_6.jpg)

**Figure 6. D.** (A) L

\hfill\break

![](../results/figures/figure_S1.jpg)

**Figure S1. D.** (A) L

\hfill\break

![](../results/figures/figure_S2.jpg)

**Figure S2. D.** (A) L

\hfill\break

![](../results/figures/figure_S3.jpg)

**Figure S3. D.** (A) L

\hfill\break

**Table 1. D.** (A) L

samples_used <- readxl::read_xlsx('data/raw/humanGF_ids.xlsx') %>% 
	pull(human_source) %>% 
	unique
donor_data <- readxl::read_xlsx('data/raw/MIMARKS_cdclinical.xlsx') %>% 
	filter(sample_id %in% samples_used) %>% 
	select(sample_id, biome, age, gender, "antibiotics >3mo", protonpump, 
		antacid, Healthworker, historyCdiff, Surgery6mos, Vegetarian, weight, disease_stat)

**Table 2. D.** (A) L

read.csv(here('data/raw/Hanna_in_vitro_data_plus_germination.txt')) %>% 
  rownames_to_column('isolate') %>% 
  filter(isolate %in% c('DA00299', 'DA00395', 'DA00458', 'DA00431'))

### 2021-10-21 exploratory/notebook/ml_analysis/feature_correlation.Rmd
render rmd notebooks

---
title: "feature_correlation_notebook"
author: "Nick Lesniak"
date: "6/16/2021"
output:
  rmarkdown::github_document: default
  pdf_document: default
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
#knitr::opts_knit$set(root.dir = '../../..')
```

```{r setup data}
library(tidyverse)

performance_df <- read_tsv('../../../data/process/ml/ml_performance.tsv',
                     col_type = 'dddddddddddddddcdcc') %>% 
  mutate(dataset = factor(dataset, c('same_day_toxin', 'day_0_predict_future_toxin',
                                     'day_0_moribund', 'day_10_histology')),
         taxonomic_level = factor(taxonomic_level, 
                                  levels = c('OTU', 'Genus', 'Family', 'Order', 'Class', 'Phylum')))  
features_df <- read_tsv('../../../data/process/ml/ml_feature_imp.tsv',
                     col_type = 'ddcccdcc') 
hyperparameter_df <- read_tsv('../../../data/process/ml/ml_hp_performance.tsv',
                     col_type = 'ddcdccc') 
taxonomy_df <- read_tsv('../../../data/process/final.taxonomy.tidy.tsv',
                        col_type = cols(.default = col_character()))

metadata <- read_tsv('../../../data/process/metadata_tidy.tsv',
					 col_type = 'cccccDddddDcdddddcdl') %>% 
	filter(cdiff_strain == 431) %>% 
	mutate(cdiff_cfu = ifelse(cdiff_cfu == 0, log10(60), log10(cdiff_cfu)))

toxin <- read_tsv('../../../data/process/toxin_tidy.tsv',
				  col_type = 'cccccdc') %>% 
	mutate(toxin = ifelse(Log_repiricoal_dilution > 1, 'present', 'absent'))

histology <- read_tsv('../../../data/process/histology_tidy.tsv',
					  col_type = 'ccccdddcdcc') %>% 
	mutate(hist_score = case_when(summary_score < 5 ~ 'low',
								  summary_score > 5 ~ 'high',
								  summary_score == 5 ~ 'mid',
								  T ~ 'NA'))

shared <- read_tsv('../../../data/mothur/sample.final.0.03.subsample.shared',
         col_types = cols(.default = col_double(),
                          Group = col_character())) %>% 
  mutate(Group = ifelse(grepl('_D|DA', Group), Group,
      gsub('D', '_D', Group))) %>% 
  select(-label, -numOtus)

same_day_toxin_df <- toxin %>% 
	select(group, toxin) %>% 
	inner_join(shared, by = c('group' = 'Group')) %>% 
	left_join(select(metadata, group, C_difficile_CFU = cdiff_cfu), by = c('group'))

#	from day 0?
day_0_predict_future_toxin_df <- toxin %>% 
	group_by(mouse_id) %>% 
	summarise(toxin = ifelse(max(Log_repiricoal_dilution) > 1, 'present', 'absent')) %>% 
	left_join(select(metadata, mouse_id, day, group), 
			  by = c('mouse_id')) %>% 
	filter(day == 0) %>% 
	inner_join(shared, by = c('group' = 'Group'))

# setup data to predict moribunidty (microbiome, cdiff, toxin)
toxin_summary <- toxin %>% 
	filter(toxin_sample_day  %in% c(1,2)) %>% 
	group_by(mouse_id) %>% 
	summarise(toxin_positive_days = sum(Log_repiricoal_dilution > 1))
cfu_summary <- metadata %>% 
	filter(day %in% c(1,2)) %>% 
	group_by(mouse_id, early_euth) %>% 
	summarise(median_CFU = median(cdiff_cfu))
day_0_moribund <- metadata %>% 
	select(group, mouse_id, day) %>% 
	filter(day == 0) %>% 
	left_join(cfu_summary, by = c('mouse_id')) %>% 
	left_join(toxin_summary, by = c('mouse_id')) %>% 
	inner_join(shared, by = c('group' = 'Group')) %>% 
	mutate(early_euth = ifelse(early_euth, 'severe', 'mild'))

# setup data to predict histology severity of colonized mice (microbiome, cdiff, toxin)
day_10_histology_df <- metadata %>% 
	filter(day == 10,
		early_euth == F) %>% 
	select(group, C_difficile_CFU = cdiff_cfu, mouse_id) %>% 
	left_join(select(toxin, group, toxin_level = Log_repiricoal_dilution),
		by = c('group')) %>% 
	inner_join(shared, by = c('group' = 'Group')) %>% 
	inner_join(select(histology, mouse_id, hist_score), by = c('mouse_id')) 
```

```{r}
performance_df %>% 
  filter(dataset == 'day_0_predict_future_toxin',
         method == 'rf') %>% 
  select(method, seed, dataset, cv_auc = cv_metric_AUC, test_auc = AUC, taxonomic_level) %>% 
  pivot_longer(cols = c(cv_auc, test_auc), 
               names_to = 'metric', 
               values_to = 'Performance') %>% 
  mutate(taxonomic_level = factor(taxonomic_level, 
                                  levels = c('OTU', 'Genus', 'Family', 'Order', 'Class', 'Phylum'))) %>% 
  ggplot(aes(x = taxonomic_level, y = Performance, color = metric)) + 
    geom_boxplot() + 
    geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
    facet_grid(dataset~method, scales = 'free_y') + 
    theme_bw()
```

Using RF at the Phylum level the features with median differences greater than 0 are:

```{r}
day_0_toxin_production <- features_df %>% 
  filter(dataset == 'day_0_predict_future_toxin',
         (method == 'rf' & taxonomic_level == 'Phylum')) %>% 
  group_by(names) %>% 
  mutate(median_diff = median(perf_metric_diff)) %>% 
  ungroup() %>% 
  filter(median_diff > 0) 
day_0_toxin_production %>%
  ggplot(aes(x = reorder(names,median_diff), y = perf_metric_diff)) + 
    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
    #geom_jitter(alpha = 0.1, width = 0.1) + 
    coord_flip() + 
    facet_grid(method~dataset, scales= 'free_y', space = 'free_y') + 
    labs(x = NULL, y = 'Decrease AUC') +
    theme_bw()
```

#### Verrucomicrobia (Akkermansia) has a median decrease in AUC of 0.12

Using RF at the Genus level the features with median differences greater than 0 are:

```{r}
day_0_toxin_production <- features_df %>% 
  filter(dataset == 'day_0_predict_future_toxin',
         (method == 'rf' & taxonomic_level == 'Genus')
         ) %>% 
  group_by(names) %>% 
  mutate(median_diff = median(perf_metric_diff),
         n = length(perf_metric_diff)) %>% 
  ungroup() %>% 
  filter(median_diff > 0,
         n > 1) %>% 
  left_join(select(taxonomy_df, Genus, Phylum), by = c('names' = 'Genus')) %>% 
  mutate(names = paste(names, '(', Phylum, ')'))
day_0_toxin_production %>%
  ggplot(aes(x = reorder(names,median_diff), y = perf_metric_diff)) + 
    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
    #geom_jitter(alpha = 0.1, width = 0.1) + 
    coord_flip() + 
    facet_grid(method~dataset, scales= 'free_y', space = 'free_y') + 
    labs(x = NULL, y = 'Decrease AUC') +
    theme_bw()
```

#### Akkermansia now has a median decrease in AUC of less than 0.005

runnning my model at all taxa levels, Verrucomicrobia (akkermansia) is top feature with median decrease in AUC of 0.12 for phylum but at the genus level it drops to 0.005 but both models overall perform the same. only one genus correlates close to akkermansia (0.8) but has low effect on auc. so i was trying to think if its possible another set of features could be replace the information lost by akkermansia, but unsure how to show or evaluate that possibility  
-  run ml to predict akkermansia  
-  add multiple genera together and check for correlation between akkermansia and combined genera sets  
-  compare distibution of abundance relative to outcome

Are other genera or combinations of genera able to replace signal from akkermansia?

```{r}
corr_df <- day_0_predict_future_toxin_df %>% 
  pivot_longer(cols = starts_with('Otu'), names_to = 'OTU', values_to = 'counts') %>% 
  left_join(taxonomy_df, by = "OTU") %>% 
  group_by(group, toxin, Genus) %>% 
  summarise(counts = sum(counts)/21.07) %>% 
  group_by(Genus) %>% 
  mutate(max = max(counts)) %>% 
  filter(max > 0) %>% 
  ungroup
akk_df <- corr_df %>% 
  filter(Genus == 'Akkermansia') %>% 
  pull(counts)

#corr_df %>% 
#  group_by(Genus) %>% 
#  nest() %>% 
#  mutate(pvalue = map(data, function(x) cor.test(akk_df, x$counts, method = 'pearson')$p.value),         
#         corr = map(data, function(x) cor.test(akk_df, x$counts, method = 'pearson')$estimate)
#         ) %>% 
#  unnest(pvalue, corr) %>% 
#  filter(!is.na(pvalue)) %>% 
#  arrange(desc(abs(corr))) %>% 
#	mutate(pvalue_BH = p.adjust(pvalue, method = 'BH', n = nrow(.))) %>% 
#  filter(pvalue_BH< 0.05)
```
  Correlation with Akkermansia

|Feature|P value| Rho|  
|-----------------|-----|-----|  
|Akkermansia | 0.00e00	| 1.00 |  
|Rubneribacter | 1.50e-12	| 0.812 |

Try combining features and then testing correlations w/ Akkermansia

```{r}
#output <- c()
#for(i in corr_df$Genus){
#  for(j in corr_df$Genus){
#    comparison <- paste(i, j, sep = '_')
#    comparison_values <- pull(filter(corr_df, Genus == i), counts) + pull(filter(corr_df, Genus == j), counts)
#    correlation_output <- cor.test(akk_df, comparison_values, method = 'pearson')
#    output <- bind_rows(output, tibble(feature = comparison, 
#                                       pvalue = correlation_output$p.value, 
#                                       correlation = correlation_output$estimate))
#  }
#}
#
#tail(output)
#output %>% 
#  filter(correlation < -0.5 | correlation > 0.5) %>% 
#  distinct %>% 
#  filter(!grepl('Akkermansia', feature),
#         !grepl('Rubneribacter', feature))
```

Correlation to Akkermansia (only running up to the combination of  Ileibacterium_Rikenellaceae_unclassified)

|Feature|P value| Rho|  
|-----------------|-----|-----|  
|Alloiococcus | 1.412915e-03	| 0.443538316 |  
|Enterobacterales_unclassified | 1.027612e-01	| 0.235874618 |  
|Alloiococcus_Enterobacterales_unclassified | 0.0002539063	| 0.5000263 |

Could the effect in decrease AUC be due to the number of features being much larger than the number of samples?

This model has 49 samples  
-  the phylum level model has 11 features  
-  the genus level model has 111 features

The literature suggests as few as 5 samples/feature but Ploeg et al says 20-50 samples/feature for LR and >200 was insufficient in some cases for RF, but LR can use penalization to reduce the number of features
https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/1471-2288-14-137

```{r, eval=F}

day_0_toxin_production <- features_df %>% 
  filter(dataset == 'day_0_predict_future_toxin',
         method == 'glmnet' & taxonomic_level == 'Genus') %>% 
  group_by(names) %>% 
  mutate(median_diff = median(perf_metric_diff)) %>% 
  ungroup() %>% 
  filter(median_diff > 0) 
day_0_toxin_production %>%
  ggplot(aes(x = reorder(names,median_diff), y = perf_metric_diff)) + 
    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
    #geom_jitter(alpha = 0.1, width = 0.1) + 
    coord_flip() + 
    facet_grid(method~dataset, scales= 'free_y', space = 'free_y') + 
    labs(x = NULL, y = 'Decrease AUC') +
    theme_bw()
```

```{r, eval =F}
day_0_toxin_production <- features_df %>% 
  filter(dataset == 'day_0_predict_future_toxin',
         method == 'glmnet' & taxonomic_level == 'Family') %>% 
  group_by(names) %>% 
  mutate(median_diff = median(perf_metric_diff)) %>% 
  ungroup() %>% 
  filter(median_diff > 0) 
day_0_toxin_production %>%
  ggplot(aes(x = reorder(names,median_diff), y = perf_metric_diff)) + 
    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
    #geom_jitter(alpha = 0.1, width = 0.1) + 
    coord_flip() + 
    facet_grid(method~dataset, scales= 'free_y', space = 'free_y') + 
    labs(x = NULL, y = 'Decrease AUC') +
    theme_bw()

```

### 2021-10-21 exploratory/notebook/plot_lefse_analysis.R

@@ -1,0 +1,119 @@
+library(tidyverse)
+
+metadata <- read_tsv('data/process/metadata_tidy.tsv',
+		col_type = 'cccccDddddDcdddddcdl') 
+shared <- read_tsv('data/mothur/sample.final.0.03.subsample.shared',
+		col_types = cols(.default = col_double(),
+			Group = col_character())) %>% 
+	mutate(Group = ifelse(grepl('_D|DA', Group), Group,
+		gsub('D', '_D', Group)))
+taxonomy <- read_tsv('data/process/final.taxonomy.tidy.tsv',
+                     col_types = cols(.default = col_character()))
+
+ # plot lefse results
+lefse_df <- data.table::fread('data/process/severe_disease_OTU.0.03.lefse_summary',
+		fill = T) %>% 
+	filter(!is.na(LDA)) %>% 
+	top_n(20, LDA) %>% 
+	left_join(select(taxonomy, OTU, tax_otu_label), by = 'OTU') %>% 
+	mutate(order = LDA + 10 * as.logical(Class),
+		tax_otu_label = gsub('unclassified', 'UC', tax_otu_label))
+
+lefse_plot <- lefse_df %>% 
+	ggplot(aes(y = LDA, fill = Class, 
+			x = reorder(tax_otu_label, order))) +
+		geom_col() + 
+		theme_bw() + 
+		coord_flip() + 
+		guides(fill = 'none') + 
+		scale_fill_manual(values = c(1,2),
+			limits = c(TRUE, FALSE),
+			labels = c('Severe', 'Non-severe')) + 
+		labs(x = NULL, y = 'LDA Score (log 10)')
+
+abundance_plot <- metadata %>% 
+	filter(cdiff_strain == 431,
+		day == 0) %>%   
+	left_join(shared, by = c('group' = 'Group')) %>% 
+	select(group, early_euth, one_of(lefse_df$OTU)) %>%  
+	gather(OTU, abundance, matches('Otu\\d')) %>% 
+	left_join(lefse_df, by = 'OTU') %>% 
+	mutate(abundance = abundance / 21.07,
+		abundance = ifelse(abundance == 0, 0.045, abundance)) %>% 
+	ggplot(aes(x = reorder(tax_otu_label, order), 
+			y = abundance, color = as.factor(early_euth))) + 
+		geom_jitter(position = position_jitterdodge(dodge.width = 0.5), alpha = 0.25) + 
[…]

### 2021-10-27 code/plot_figure_1.R
create plot rscripts from rmd notebooks

@@ -1,0 +1,119 @@
+###############################################################################
+#
+# Figure 1
+#	main idea - germ0free mice colonized with human feces produce diverse communities
+# 	plot - relative abundance of communities for each sample and 
+#		beta diversity to compare distance from mice recieving same donor to others
+#
+# Nick Lesniak 2021-10-26
+###############################################################################
+# setup environment
+###############################################################################
+source(here::here('code/utilities.R'))
+
+###############################################################################
+# laod data
+###############################################################################
+metadata <- read_metadata()
+
+taxonomy <- read_taxonomy()
+
+shared <- read_shared() %>% 
+  select(-label, -numOtus) %>% 
+  pivot_longer(-Group, names_to = 'otu', values_to = 'counts') %>% 
+  mutate(relative_abundance = counts/2107 * 100)
+beta_df <- read_beta()
+
+###############################################################################
+#  setup data
+###############################################################################
+day_0_samples <- metadata %>% 
+  filter(day == 0) %>% 
+  select(group, human_source)
+
+corecipient <- day_0_samples %>% 
+  left_join(day_0_samples, by = c('human_source')) %>% 
+  filter(group.x != group.y) %>% 
+  inner_join(beta_df, by = c('group.x' = 'rows', 'group.y' = 'columns')) %>% 
+  mutate(comparison = 'Recipients compared to\nothers w/same donor')
+
+other_recipients <- beta_df %>% 
+  filter(grepl('D0', rows) & grepl('D0', columns)) %>% 
+  anti_join(corecipient, by = c('rows' = 'group.x', 'columns' = 'group.y')) %>% 
+  mutate(comparison = 'Recipients compared to\nothers w/different donor')
+
+###############################################################################
+#  analyze data
+###############################################################################
+wilcox.test(corecipient$distances, other_recipients$distances)
+
+###############################################################################
+#  plot data
+###############################################################################
+day_0_plot <- metadata %>% 
+  filter(day == 0,
+         cdiff_strain == 431) %>% 
+  select(group, human_source, mouse_id) %>% 
+  left_join(filter(shared, counts > 0), by = c('group' = 'Group')) %>% 
[…]

### 2021-10-27 code/plot_figure_2.R

@@ -1,0 +1,81 @@
+###############################################################################
+#
+# Figure 2
+#	main idea - Mice were colonized without any perturbation
+# 	plot - CFU of each mouse at day 0 and day 10
+#
+# Nick Lesniak 2021-10-26
+###############################################################################
+# setup environment
+###############################################################################
+source(here::here('code/utilities.R'))
+
+###############################################################################
+# load data
+###############################################################################
+metadata <- read_metadata
+
+donor_aes <- donor_df
+
+###############################################################################
+#  setup data
+###############################################################################
+
+###############################################################################
+#  plot data
+###############################################################################
+cfu_plot <- metadata %>% 
+  filter(cdiff_strain == 431,
+         day %in% c(1,10)) %>% 
+  mutate(human_source = as.numeric(gsub('DA', '', human_source)),
+         human_source = factor(human_source, levels = 
+                                c('1324', '953', '581', '1134', '10148', 
+                                  '369', '430', '10093', '10027', '10034',
+                                  '10082', '884', '578', '431', '1245')),
+         day = factor(paste('Day', day), levels = c('Day 1', 'Day 10'))) %>% 
+  ggplot(aes(x = human_source, y = log10(cdiff_cfu + 60), color = human_source)) +
+    geom_jitter(width = 0.2) + 
+    facet_grid(day~.)+ 
+    scale_x_discrete(guide = guide_axis(angle = 45)) + 
+    scale_color_manual(values = donor_colors) + 
+    theme_bw() + 
+    labs(x = 'Donor Source', y = 'C. difficile CFU (log10)') + 
+    guides(color = 'none')
+
+# presentation plot
+#presentation_cfu_plot <- metadata %>% 
+#  filter(cdiff_strain == 431,
+#         day %in% c(1,10)) %>% 
+#  mutate(human_source = as.numeric(gsub('DA', '', human_source)),
+#         human_source = factor(human_source, levels = 
+#                                c('1324', '953', '581', '1134', '10148', 
+#                                  '369', '430', '10093', '10027', '10034',
+#                                  '10082', '884', '578', '431', '1245'),
+#                               labels = LETTERS[1:15]),
+#         day = factor(paste('Day', day), levels = c('Day 1', 'Day 10'))) %>% 
[…]

### 2021-10-27 code/plot_figure_3.R

@@ -1,0 +1,121 @@
+###############################################################################
+#
+# Figure 3
+#	main idea - 
+# 	plot -
+#
+# Nick Lesniak 2021-10-26
+###############################################################################
+# setup environment
+###############################################################################
+source(here::here('code/utilities.R'))
+
+###############################################################################
+# load data
+###############################################################################
+metadata <- read_metadata
+
+toxin <- read_toxin
+
+histology <- read_histology
+
+donor_aes <- donor_df
+
+###############################################################################
+#  plot data
+###############################################################################
+histology_plot <- histology %>% 
+  left_join(distinct(select(metadata, mouse_id, early_euth)), by = 'mouse_id') %>% 
+  filter(!is.na(early_euth)) %>% 
+  mutate(early_euth = ifelse(early_euth, 'Severe', 'Mild'),
+         mouse_id = paste0(cage_id, '_', ear_tag)) %>% 
+  #left_join(mouse_by_outcome, by = c('mouse_id')) %>%
+  pivot_longer(cols = c(edema_tissue, inflammation_tissue, epithelial_damage, summary_score),
+               names_to = 'histology', values_to = 'score') %>% 
+  filter(histology == 'summary_score') %>% 
+  mutate(histology = ifelse(histology == 'summary_score', 'End point', NA),
+         human_source = gsub('DA0{0,2}', '', human_source)) %>% 
+  ggplot(aes(x = human_source, y = score, , color = early_euth, fill = early_euth)) +
+    #stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5), position = position_dodge(width = 0.4)) +
+    geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 2) + 
+    #geom_point(position = position_jitterdodge(jitter.width = 0.3), alpha = 0.5) + 
+    theme_bw() + 
+    labs(x = NULL, y = 'Clinical\nScore', fill = 'Outcome') +
+    scale_x_discrete(guide = guide_axis(angle = 45)) + 
+    scale_y_continuous(breaks = c(1,3,5,7,9), 
+                       labels = c(1,3,5,7,9)) + 
+    facet_grid(histology~early_euth, scales = 'free', space = 'free_x') + 
+    guides(color = 'none') + 
[…]

### 2021-10-27 code/plot_figure_4.R

@@ -1,0 +1,122 @@
+###############################################################################
+#
+# Figure 4
+#	main idea - 
+# 	plot -
+#
+# Nick Lesniak 2021-10-26
+###############################################################################
+# setup environment
+###############################################################################
+source(here::here('code/utilities.R'))
+
+###############################################################################
+# load data
+###############################################################################
+metadata <- read_metadata()
+
+taxonomy <- read_taxonomy()
+
+shared <- read_shared()
+
+###############################################################################
+#  create plot lefse data function
+###############################################################################
+plot_lefse <- function(file_name){
+  lefse_design <- data.table::fread(here(paste0('data/process/lefse/', file_name, '.design')))
+      colnames(lefse_design) <- c('Group', 'class')
+    
+  lefse_df <- data.table::fread(here(paste0('data/process/lefse/', file_name, '.0.03.lefse_summary')),
+    	fill = T) %>% 
+    filter(!is.na(LDA)) %>% 
+    top_n(20, LDA) %>% 
+    mutate(fctr_class = factor(Class),
+      order = LDA + 10 * as.numeric(fctr_class),
+    	tax_otu_label = OTU)
+    
+   shared %>% 
+    gather(OTU, abundance, matches('Otu\\d*')) %>% 
+  	left_join(taxonomy, by = 'OTU') %>% 
+  	right_join(lefse_df, by = c('Genus' = 'OTU')) %>% 
+    right_join(lefse_design, by = c('Group')) %>% 
+  	group_by(Group, Genus, order, class) %>% # removed mouse_id from interactive plot
+  	summarise(abundance = sum(abundance)) %>% 
+  	mutate(tax_otu_label = Genus) %>% 
+  	mutate(abundance = abundance / 21.07,
+  		abundance = ifelse(abundance == 0, 0.045, abundance)) %>% 
+    #highlight_key(~mouse_id) %>% 
+  	ggplot(aes(x = tax_otu_label, 
+  			y = abundance, color = class)) + # , shape = mouse_id remove mouse_id from interactive plot
+  		geom_hline(yintercept = 0.047, linetype = 'dashed', lwd = 0.3) + 
+      stat_summary(fun.data = 'median_hilow', aes(group = class),
+  			fun.args = (conf.int=0.5), position = position_dodge(width = .8)) +
+  		theme_bw() + 
+      labs(x = NULL, y = 'Relative Abundance', color = NULL, title = file_name) + 
[…]

### 2021-10-27 code/plot_figure_5.R

@@ -1,0 +1,179 @@
+###############################################################################
+#
+# Figure 5
+#	main idea - 
+# 	plot -
+#
+# need files:
+# 	data/process/
+#	data/mothur/
+#	data/process/
+#
+# Nick Lesniak 2021-10-26
+###############################################################################
+# setup environment
+###############################################################################
+source(here::here('code/utilities.R'))
+
+###############################################################################
+# load data
+###############################################################################
+metadata <- read_metadata()
+toxin <- read_toxin()
+histology <- read_histology()
+taxonomy_df <- read_taxonomy()
+shared <- read_shared()
+
+
+###############################################################################
+#  setup data
+###############################################################################
+# setup data from model prediction toxin production from day 0
+day_0_predict_future_toxin_df <- toxin %>% 
+	group_by(mouse_id) %>% 
+	summarise(toxin = ifelse(max(Log_repiricoal_dilution) > 1, 'present', 'absent')) %>% 
+	left_join(select(metadata, mouse_id, day, group), 
+			  by = c('mouse_id')) %>% 
+	filter(day == 0) %>% 
+	inner_join(shared, by = c('group' = 'Group'))
+day_0_toxin_features <- day_0_toxin_production %>% 
+  select(names, median_diff) %>% 
+  unique %>% 
+  arrange(median_diff) %>% 
+  pull(names)
+# setup data for model predicting moribund from day 0
+toxin_summary <- toxin %>% 
+	filter(toxin_sample_day  %in% c(1,2)) %>% 
+	group_by(mouse_id) %>% 
+	summarise(toxin_positive_days = sum(Log_repiricoal_dilution > 1))
+cfu_summary <- metadata %>% 
+	filter(day %in% c(1,2)) %>% 
+	group_by(mouse_id, early_euth) %>% 
+	summarise(median_CFU = median(cdiff_cfu))
+day_0_moribund <- metadata %>% 
+	select(group, mouse_id, day) %>% 
+	filter(day == 0) %>% 
+	left_join(cfu_summary, by = c('mouse_id')) %>% 
+	left_join(toxin_summary, by = c('mouse_id')) %>% 
+	inner_join(shared, by = c('group' = 'Group')) %>% 
[…]

### 2021-10-27 code/plot_figure_6.R

@@ -1,29 +1,40 @@
-library(tidyverse)
-library(here)
+###############################################################################
+#
+# Figure 6
+#	main idea - 
+#   plot -
+#
+# Nick Lesniak 2021-10-26
+###############################################################################
+# setup environment
+###############################################################################
+source(here::here('code/utilities.R'))
 
-
-source(here('code/Sum_OTU_by_Tax.R'))
-
-# load in data
-metadata <- read_tsv(here('data/process/metadata_tidy.tsv'),
-                     col_type = 'cccccDddddDcdddddcdl') %>% 
+###############################################################################
+# load data
+###############################################################################
+metadata <- read_metadata %>% 
   mutate(isolate_431 = ifelse(cdiff_strain == 431, '431', 'Other'),
          isolate = factor(cdiff_strain, levels = c('431', '299', '395','458')))
   
-toxin <- read_tsv(here('data/process/toxin_tidy.tsv'),
-                  col_type = 'cccccdc')
-histology <- read_tsv(here('data/process/histology_tidy.tsv'),
-                      col_type = 'ccccdddcdcc') %>% 
+toxin <- read_toxin()
+
+histology <- read_histology() %>% 
   select(summary_score, mouse_id)
 
-donor_colors <- c("#41bbc5", "#7e2640", "#9ad859", "#682dbd", "#728e24", "#e26df8", "#23980d", "#fd048f", "#46e33e", "#fa1bfc", "#155126", "#f9b2d1", "#154975", "#e2c627", "#6294ce")
+donor_aes <- donor_df
 
-
+###############################################################################
+#  setup data
+###############################################################################
 non_431_donors <- metadata %>% 
   filter(cdiff_strain != 431) %>% 
   pull(human_source) %>% 
   unique
 
+###############################################################################
+#  plot data
+###############################################################################
 cfu_plot <- metadata %>% 
   filter(human_source %in% non_431_donors,
          day == 10) %>% 
@@ -39,8 +50,10 @@ cfu_plot <- metadata %>%
     geom_hline(yintercept = 2, linetype = 'dashed', color = 'lightgray') + 
     geom_label(x = 3, y = 2, label = "LOD", fill = 'white', color = 'white') + 
[…]

### 2021-10-27 code/plot_figure_S1.R

@@ -1,0 +1,50 @@
+###############################################################################
+#
+# Figure S1
+#	main idea - 
+# 	plot -
+#
+# Nick Lesniak 2021-10-26
+###############################################################################
+# setup environment
+###############################################################################
+source(here::here('code/utilities.R'))
+
+###############################################################################
+# load data
+###############################################################################
+metadata <- read_metadata()
+
+###############################################################################
+#  setup data
+###############################################################################
+wt_loss_df <- metadata %>% 
+  filter(cdiff_strain == 431,
+         day %in% c(2,10)) %>% 
+  mutate(human_source = as.numeric(gsub('DA', '', human_source)),
+         human_source = factor(human_source, levels = 
+                                c('1324', '953', '581', '1134', '10148', 
+                                  '369', '430', '10093', '10027', '10034',
+                                  '10082', '884', '578', '431', '1245')),
+         day = factor(paste('Day', day), levels = c('Day 2', 'Day 10'))) %>% 
+
+###############################################################################
+#  plot data
+###############################################################################
+wt_loss_plot <- wt_loss_df %>% 
+  ggplot(aes(x = human_source, y = percent_weightLoss, color = human_source)) +
+    geom_jitter(width = 0.2) + 
+    facet_grid(day~.)+ 
+    scale_x_discrete(guide = guide_axis(angle = 45)) + 
+    scale_color_manual(values = donor_colors) + 
+    theme_bw() + 
+    labs(x = 'Donor Source', y = 'Weight (% relative to initial)') + 
+    guides(color = 'none')
+
+###############################################################################
+#  save plot
+###############################################################################
+ggsave(here('results/figures/Figure_S1.jpg'),
+       wt_loss_plot,
+       width = 6, height = 4)
+###############################################################################

### 2021-10-27 code/plot_figure_S2.R

@@ -1,0 +1,161 @@
+###############################################################################
+#
+# Figure 
+# 	plot -
+#	main idea - 
+#
+# need files:
+# 	data/process/
+#	data/mothur/
+#	data/process/
+#
+# Nick Lesniak 2021-10-26
+###############################################################################
+# setup environment
+###############################################################################
+library(tidyverse)
+library(cowplot)
+
+###############################################################################
+# load data
+###############################################################################
+performance_df <- read_tsv(here('data/process/ml/ml_performance.tsv'),
+                     col_type = 'dddddddddddddddcdcc') %>% 
+  filter(dataset != 'day_0_histology_50') %>% 
+  mutate(dataset = gsub('_80', '', dataset),
+         dataset = factor(dataset, c('same_day_toxin', 'day_0_predict_future_toxin',
+                                     'day_0_moribund', 'day_0_histology')),
+         taxonomic_level = factor(taxonomic_level, 
+                                  levels = c('OTU', 'Genus', 'Family', 'Order', 'Class', 'Phylum')))  
+
+features_df <- read_tsv(here('data/process/ml/ml_feature_imp.tsv'),
+                     col_type = 'ddcccddcc') %>% 
+  filter(dataset != 'day_0_histology_50') %>% 
+  mutate(dataset = gsub('_80', '', dataset))
+
+###############################################################################
+#  setup data
+###############################################################################
+day_0_toxin_production <- features_df %>% 
+  filter(dataset == 'day_0_predict_future_toxin',
+         (method == 'rf' & taxonomic_level == 'Phylum')) %>% 
+  group_by(names) %>% 
+  mutate(median_diff = median(perf_metric_diff)) %>% 
+  ungroup() %>% 
+  filter(median_diff > 0) 
+
+day_0_moribund_plot_df <- features_df %>% 
+  filter(dataset == 'day_0_moribund',
+         method == 'rf',
+         taxonomic_level == 'Class') %>% 
+  group_by(names) %>% 
+  mutate(median_diff = median(perf_metric_diff)) %>% 
+  ungroup() %>% 
+  filter(median_diff > 0) 
+
[…]

### 2021-10-27 code/plot_figure_S3.R

@@ -1,0 +1,148 @@
+###############################################################################
+#
+# Figure 
+#   plot -
+# main idea - 
+#
+# need files:
+#   data/process/
+# data/mothur/
+# data/process/
+#
+# Nick Lesniak 2021-10-26
+###############################################################################
+# setup environment
+###############################################################################
+library(tidyverse)
+library(here)
+source(here('code/Sum_OTU_by_Tax.R'))
+
+###############################################################################
+# load and steup data
+###############################################################################
+metadata <- read_tsv(here('data/process/metadata_tidy.tsv'),
+                     col_type = 'cccccDddddDcdddddcdl') %>% 
+  filter(human_source %in% c('DA00430', 'DA00578', 'DA00369'),
+         day == 0) %>% 
+  mutate(human_source = as.character(as.numeric(gsub('DA', '', human_source))),
+         human_source = factor(human_source, levels = c('430','369', '578')))
+         
+shared <- read_tsv(here('data/mothur/sample.final.0.03.subsample.shared'),
+                   col_types = cols(.default = col_double(),
+                                    Group = col_character())) %>% 
+  mutate(Group = ifelse(grepl('_D|DA', Group), Group,
+                        gsub('D', '_D', Group))) %>% 
+  select(-label, -numOtus) %>% 
+  pivot_longer(-Group, names_to = 'otu', values_to = 'counts') %>% 
+  filter(counts > 0) %>% 
+  mutate(relative_abundance = counts/2107 * 100)
+
+taxonomy <- read_tsv(here('data/process/final.taxonomy.tidy.tsv'),
+                     col_types = cols(.default = col_character()))
+
+features_df <- read_tsv(here('data/process/ml/ml_feature_imp.tsv'),
+                        col_type = 'ddcccddcc') %>% 
+  filter(dataset != 'day_0_histology_50') %>% 
+  mutate(dataset = gsub('_80', '', dataset))
+
+day_0_moribund <- metadata %>% 
+  select(group, mouse_id, day, early_euth) %>% 
+  filter(day == 0) %>% 
+  inner_join(shared, by = c('group' = 'Group')) %>% 
+  mutate(early_euth = ifelse(early_euth, 'severe', 'mild'))
+
+###############################################################################
[…]

### 2021-10-27 code/plot_figure_S4.R

@@ -1,0 +1,71 @@
+###############################################################################
+#
+# Figure 
+# 	plot -
+#	main idea - 
+#
+# need files:
+# 	data/process/
+#	data/mothur/
+#	data/process/
+#
+# Nick Lesniak 2021-10-26
+###############################################################################
+# setup environment
+###############################################################################
+source(here::here('code/utilities.R'))
+
+###############################################################################
+# load data
+###############################################################################
+taxonomy_df <- read_taxonomy()
+metadata <- read_metadata()
+histology <- read_histology()
+# outcome from run_lefse_temporal_trend.R
+lefse_hilo <- c("Anaeromassilibacillus", "Anaeroplasma", 
+	"Bacteroidales_unclassified", "Bifidobacterium", "Blautia", "Clostridioides", 
+	"Clostridium_IV", "Clostridium_XlVa", "Duncaniella", "Escherichia/Shigella", 
+	"Faecalibacterium", "Fournierella", "Helicobacter", "Hungatella", "Ihubacter", 
+	"Intestinimonas", "Klebsiella", "Kosakonia", "Lactobacillales_unclassified", 
+	"Legionella", "Muribaculaceae_unclassified", "Neglecta", "Odoribacter", 
+	"Oscillibacter", "Paraclostridium", "Paramuribaculum", 
+	"Prevotellaceae_unclassified", "Ruthenibacterium")
+
+###############################################################################
+#  setup data
+###############################################################################
+temporal_lefse_df <- taxonomy_df %>% 
+  filter(Genus %in% lefse_hilo) %>% 
+  left_join(pivot_longer(shared, 
+                         cols = -Group, names_to = 'OTU', values_to = 'counts')) %>% 
+  left_join(metadata, by = c('Group' = 'group')) %>%
+  group_by(mouse_id, day, Genus, early_euth) %>% 
+  summarise(counts = sum(counts)) %>% 
+  inner_join(select(histology, mouse_id, hist_score), by = 'mouse_id') %>% 
+  mutate(outcome = ifelse(early_euth, 'severe', hist_score),
+         outcome = factor(outcome, levels = c('low', 'high', 'severe')),
+         rel_abun = counts/2107 * 100) %>% 
+  filter(outcome != 'mid') %>% 
+  group_by(Genus) %>% 
+  mutate(present = sum(counts > 0)) %>% 
+  filter(present > 60)
+
+###############################################################################
+#  plot data
+###############################################################################
[…]

### 2021-10-27 code/utilities.R

@@ -1,0 +1,66 @@
+###############################################################################
+#
+#	Common functions used across scripts
+#
+# need files:
+# 	data/process/metadata_tidy.tsv
+
+
+#
+# Nick Lesniak 2021-10-26
+###############################################################################
+# setup environment
+###############################################################################
+library(tidyverse)
+library(cowplot)
+library(here)
+
+###############################################################################
+# functions to load data
+###############################################################################
+read_taxonomy <- function(){
+	read_tsv(here('data/process/final.taxonomy.tidy.tsv'),
+		col_types = cols(.default = col_character()))
+}
+
+read_metadata <- function(){
+	read_tsv(here('data/process/metadata_tidy.tsv'),
+		col_type = 'cccccDddddDcdddddcdl') 
+}
+
+read_shared <- function(){
+	read_tsv(here('data/mothur/sample.final.0.03.subsample.shared'),
+		col_types = cols(.default = col_double(),
+			Group = col_character())) %>% 
+	mutate(Group = ifelse(grepl('_D|DA', Group), Group,
+		gsub('D', '_D', Group)))
+}
+
+read_beta <- function(){
+	source(here('code/read_dist.R'))
+	read_dist(here('data/mothur/sample.final.thetayc.0.03.lt.ave.dist')) %>% 
+		mutate(rows = ifelse(grepl('_D|DA', rows), rows,
+				gsub('D', '_D', rows)),
+			columns = ifelse(grepl('_D|DA', columns), columns,
+				gsub('D', '_D', columns)))
+}
+
+read_toxin <- function(){
+	read_tsv(here('data/process/toxin_tidy.tsv'),
+		col_type = 'cccccdc')
+}
+
+read_histology <- function(){
+	read_tsv(here('data/process/histology_tidy.tsv'),
+		col_type = 'ccccdddcdcc')
+}
+
+
+donor_df <- tibble(	
+	donor_colors = c("#41bbc5", "#7e2640", "#9ad859", "#682dbd", "#728e24",
+		"#e26df8", "#23980d", "#fd048f", "#46e33e", "#fa1bfc",
+		"#155126", "#f9b2d1", "#154975", "#e2c627", "#6294ce"),
+	human_source = c('DA01324', 'DA00953', 'DA00581', 'DA01134', 'DA10148',
+		'DA00369', 'DA00430', 'DA10093', 'DA10027', 'DA10034', 
[…]

### 2021-10-27 submission/manuscript.Rmd
revise draft with initial proof read

@@
 output:
-  word_document: default
   pdf_document:
@@
     keep_tex: yes
+  word_document: default
 bibliography: references.bib
@@
 
-# Gut microbiota drives differences in Clostridiodes difficile infection severity.
+# The gut microbiota potentiates *Clostridiodes difficile* infection severity.
 
@@
 
-**Running title:** Microbiota drives *Clostridioides difficile* infection severity
+**Running title:** Microbiota potentiates *Clostridioides difficile* infection severity
 
@@
 
-Nicholas A. Lesniak$^1$, Alyxandria M. Schubert$^1$, Hamide Sinani$^1$, Patrick D. Schloss$^{1,\dagger}$
+Nicholas A. Lesniak$^1$, Alyxandria M. Schubert$^1$, Kaitlyn Flynn$^1$, Hamide Sinani$^1$, Kathyrn A. Eaton$^1$, Ingrid L. Bergin$^2$, Patrick D. Schloss$^{1,\dagger}$
 
@@
 1\. Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI  
+2\. Unit for Laboratory Animal Medicine, University of Michigan, Ann Arbor, MI
 
@@
 
-Clostridioides difficile infection has become more common and severe in the last few decades.
-Patient age, white blood cell count, creatinine levels as well as C.
-difficile ribotype and the presence of toxin genes have been associated with disease severity.
+*Clostridioides difficile* infection has become more common and severe in the last few decades.
+Patient age, white blood cell count, creatinine levels as well as *C.
+difficile* ribotype and the presence of toxin genes have been associated with disease severity.
 However, models based on those data to identify patients with CDI that will develop severe disease have not been robust enough to be broadly applied.
-These models are input from our understanding of C.
-difficile interactions with the gut environment.
-The gut bacteria are a key determinant in C.
-difficile infection.
+These models are input from our understanding of *C.
+difficile* interactions with the gut environment.
+The gut bacteria are a key determinant in *C.
+difficile* infection.
 CDI is dependent on perturbations in the bacterial community in order to colonize the gut.
-The gut microbiota also impairs C.
-difficile colonization through bile acid metabolism,  nutrient consumption and bacteriocin production.
+The gut microbiota also impairs *C.
+difficile* colonization through bile acid metabolism, nutrient consumption and bacteriocin production.
 However, it is unclear if the gut bacteria affect the disease severity resulting from that colonization.
@@
 We created diverse gut communities by colonizing germ-free mice with human fecal communities.
-They were then infected with C.
-difficile which resulted in differed disease severities.
+They were then infected with *C.
+difficile* which resulted in differed disease severities.
 The severity grouped by the human fecal community they received.
@@
 Bacterial groups associated with dietary fiber degradation, such as Coprobacillus, were associated with reduced disease severity.
-Lastly, we showed communities that resulted in either low or high histopathologic scores or severe disease had the same outcome when infected with a different C.
-difficile isolate.
+Lastly, we showed communities that resulted in either low or high histopathologic scores or severe disease had the same outcome when infected with a different *C.
+difficile* isolate.
 
@@
 Clostridioides difficile infection (CDI) causes a range of disease from asymptomatic to mild diarrhea to severe outcomes of recurrent infections and even death.
-Models that predict severity and treatment decisions are based on clinical factors and C.
-difficile capabilities.
+Models that predict severity and treatment decisions are based on clinical factors and *C.
+difficile* capabilities.
 But currently the gut microbiome, the primary protector against CDI, has not been accounted for its effect on CDI disease severity.
 We demonstrated variation in the microbiome of mice colonized with human feces was sufficient to result in a range of disease severities.
-Our results show groups of bacteria contribute to the development of disease in C.
-difficile infections.
+Our results show groups of bacteria contribute to the development of disease in *C.
+difficile* infections.
 Gut bacterial community data from patients with CDI could improve our ability to identify patients more at risk for severe disease and improve interventions which treat the gut bacteria to reduce host damage.
@@
 
-Clostridioides difficile infections (CDI) have been increasing in incidence and severity for the past two decades.
-*C.
-difficile*, first identified in 1931, wasn't categorized as a pathogen until the late 1970s after it was determined to be the driver of antibiotic-associated psuedomembranous colitis following the introduction of broad-spectrum antibiotics.
-In the early 2000s CDI became more problematic with increasing incidences, severity and death.
-People colonized with *C.
-difficile* can experience a range of disease severities from asymptomatic to mild diarhea to toxic megacolon and death.
-CDI is primarily treated with a course of antibiotics, either metronidazole or vancomycin depending on CDI severity, but can also be treated with cessation of other antibiotics for milder cases or with a fecal microbiota transplantion for CDI cases recalitrant to antibiotics.
+*Clostridioides difficile* infections (CDI) have been increasing in incidence and severity since *C.
+difficile* was first identified as a pathogen.
+It was first identified in 1931 in a healthy fecal community [CITATION].
+But then in the late 1970s,  following the introduction of broad-spectrum antibiotics, *C.
+difficile* was identified as the cause of antibiotic-associated psuedomembranous colitis [CITATION].
+Beginning in the early 2000s, CDI became more problematic with increasing incidences and severity [CITATION].
+Disease severities can range from asymptomatic to mild diarrhea to toxic megacolon and death.
+This severity has been associated with characteristics of *C.
+difficile* ribotypes and host factors.
+Over time more pathogenic ribotypes have emerged, such as ribotype 027 in the early 2000s.
+The presence of the genes for Toxin A/B or fluoroquinolon resistence has been associated with more severe disease [CITATION].
+Binary toxin, while alone insufficient to cause damage on it own, when in the presence of Toxin A/B has been associated with increased virulence [CITATION].
+Mutations in tcd C, the negative regulator of of toxin transcription, also increased the virulence of *C.
+difficile* [CITATION].
+In addition to the pathogenic potential of the infecting strain, the response by the host also affects the disease severity.
+Delayed or reduced production of antibodies IgG and IgA can result in more severe disease [CITATION].
+Increased neutrophil inflitration can lead to worse outcomes [CITATION].
+In addition to the immune response, increased density of the toxin receptor on the host cells can increase the severity of the infection [CITATION].
+While there is a thorough understanding of many of the mechanisms driving in CDI incidence and disease severity, there still is not effective tools to reduce the risk of severe disease.
 
-The degree of CDI disease severity has been connected to aspects of C.
-difficile strain and host factors.
-Since the first identification of C.
-difficile, more virulent strains have emerged, such as ribotype 027.
-Virulence is increased by the presence of the genes for Toxin A/B or fluoroquinolon resistence.
-The binary toxin, while not sufficient on its own to allow virulence, has been associated with increased virulence in the presence of Toxin A/B.
-Mutations in tcd C, the negative regulator of of toxin transcription, also increased the virulence of the C.
-difficile strain.
-In addition to the virulent capability of the colonizing strain, the immune response of the person infected also affects the severity of the disease.
-The delayed or reduced production of antibodies IgG and IgA can result in more severe disease.
-Increased neutrophil inflitration can lead to worse outcomes.
-Beyond the immune response, increased density of the toxin receptor on the host cells can increase the severity of the infection.
+Models have been developed to identify patients at risk for severe CDI but have not been robust for broad application [@Dieterle2020, @Butt2013, @Perry2021, @vanBeurden2017].
+These models utilize patient related data, such as patient age, white blood cell count, serum albumin levels, and creatinine levels, to predict disease severity.
+In their initial publication, these models perform well enough for publication.
+However when Perry et al applied many of the most current models to a multi-center external validation, their performance suffered and often had more false positives than true positives [@Perry2021].
+Thus, there seems to be a missing factor contributing to disease severity.
 
-There has been great interest in the interactions of *C.
-difficile* and the intestinal bacterial community in CDI.
-Indigenous bacteria of a healthy intestinal community provide a protective barrier preventing *C.
-difficile* from colonizing and causing infection.
-The disruption of this community from a range of pertubations, such as many classes of antibiotics, medications, or even dietary changes, allows C.
-difficile to colonize.
-We have shown that different pertubations lead to differences in community structure and the ability for C.
-difficile to maintain colonization.
-C.
-difficile has also been shown to modify its metabolism and expression of virulence genes in response to the nutritional environment.
-However, it is unclear if the bacterial community affect the disease severity of CDI.
-Here, we tested if variablity in the intestinal bacterial community can produce and explain variation in the disease severity caused by a single C.
-difficile strain, ribotype 027.
+Missing from these models are the gut bacteria which also interact with *C.
+difficile* during the course of the infection.
+The indigenous bacteria of a healthy intestinal community provide a protective barrier preventing *C.
+difficile* from infecting the gut.
+Only after this community is disrupted from a range of pertubations, such as many classes of antibiotics, medications, or even dietary changes, can *C.
+difficile* infect the intestine.
+Once established, the gut bacteria can either promote or inhibit *C.
+difficile* through producing molecules or modifying the environment [Abbas2020].
+Bile acids metabolized by the gut bacteria can inhibit *C.
+difficile* growth and affect toxin production [@Sorg2008, @Thanissery2017].
+Bacteria in the gut also can compete more directly with *C.
+difficile* through antibiotic production or nutrient consumption [@Aguirre2021, @Kang2019].
+To eliminate CDI, the primary treatment is a course of antibiotics.
+Patients who become reinfected and recalitrant to antibiotics, a fecal microbiota transplantion often is sufficient to restore the gut community to eliminate *C.
+difficile*.
+While there has been many studies to investigate the interaction between the gut bacteria and *C.
+difficile* during the infection, it has not been demostrated the bacterial community can modulate the disease severity of CDI.
+Recently, Tomkovich et al showed the variation in the bacterial communtiies between mice from different colonies and vendors was sufficient to cause differences in the temporal dynamics of the *C.
+difficile* infection [@Tomkovich2020].
+Since the gut bacteria interact with *C.
+difficile* throughout the infection and community difference resulted in different colonization dynamics, we hypothesized that the gut bacteria contribute to the variation in CDI disease severity.
+Here, we colonized germ-free C57BL/6 mice with human fecal communities to test if the variablity in the gut bacteria can produce and explain variation in the disease severity caused by a single *C.
+difficile* isolate.
 
@@
 
-**Germ-free mice colonized with human fecal commmunities have diverse and unique structures.** Based on our previous observation mice with variation in their microbiota showed differences in colonization of *C.
-difficile*, we sought to explore the affect communties with greater variation had on *C.
+**Germ-free mice colonized with human fecal commmunities have diverse and unique community structures.** Based on our previous observation mice with variation in their microbiota showed differences in colonization of *C.
+difficile* [@Tomkovich2020], we sought to explore the effect communties with greater variation had on *C.
 difficile* challenge.
-To produce communities with greater variation than mice with their indigenous communties, we used human fecal communities to establish the gut commuities in germ-free mice.
-We inoculated a cage of germ-free C57BL/6 mice (2 -4 mice per cage) with homogenized feces from one of 15 different human donors via oral gavage.
-The fecal community was allowed to colonize and stabilize for two weeks post inoculation.
-We then assessed the gut bacterial communities by sequencing the V4 region of the 16S rRNA gene extracted from fecal pellets.
-The communities established in the mice group most closely to other mice recieving the same donor than to either their respective donor or any mice who received stool from a different donor (Figure 1A).
-Most of the communities  highly abundant classes were *Clostridia*, *Bacteroidia*, *Erysipelotrichia*, *Bacilli*, and *Gammaproteobacteria*.
+To produce communities with greater variation than mouse colonies with their indigenous communties, we used human fecal communities to establish a community in the intestine of germ-free mice.
+We inoculated a cage of germ-free C57BL/6 mice (2-4 mice per cage) with homogenized feces from one of 15 different human donors via oral gavage.
+The fecal community was allowed to colonize and stabilize for two weeks post inoculation [CITATION?].
+We then assessed the gut bacterial communities by sequencing the V4 region of the 16S rRNA gene extracted from fecal pellets (Figure 1A).
+The communities established in the mice group most closely to other mice recieving the same human fecal donor community than to either their respective donor or any mice who received stool from a different donor (Figure 1B).
+Most of the communities were primairily composed of populations of *Clostridia*, *Bacteroidia*, *Erysipelotrichia*, *Bacilli*, and *Gammaproteobacteria*.
 However, each group of mice harbored unique combination of class abunadances in their gut bacterial communities.
-**Any alpha? compare to murine variation? Number of unique otus?**
+*Any alpha? compare to murine variation? Number of unique otus?*
 
-**C.
-difficile is able to colonize mice without perturbation.** Our goal is to test the effect of the variation in the microbiota on the effect of a *C.
-difficile* challenge.
+***C.
+difficile* is able to colonize mice without perturbation.** Our goal was to test the effect of the variation in the microbiota on *C.
+difficile* infection.
 A typical mouse model of CDI requires pretreatment with antibiotics such as clindamycin to become susceptible to *C.
 difficile* colonization [CITATION].
-However, we wanted to avoid modifying the communties with an antibiotic to maintain the community structures and differences.
-Since these communties came from human donor which may have become susceptible through their own exposures, we decided to test if *C.
-difficile* was able to colonize these communities without further treatment.
+However, we wanted to avoid modifying the communties with an antibiotic to maintain their community structures and differences.
+Since these communties came from human donors which may have become susceptible through their own exposures, we decided to test if *C.
+difficile* was able to colonize these communities without any perturbation.
+After two weeks post incoluation with human fecal communities, the mice were then challenged with 10^**4?** *C.
+difficile* ribotype 027 isolate 431 spores (**Add schematic?**) [@Carlson2013].
+The mice were followed for 10 days post challenge and their stool was collected and plated for *C.
+difficile* CFU to detect colonization level.
 We hypothesized that *C.
 difficile* would most likely be able to colonize the mice with communities that came from donors which had a history of antibiotic use, one of the primary risk factors for developing CDI (TABLE, CITATION).
-After two weeks post incoluation with human fecal communities, the mice were then challenged with 10^**XXXXX** *C.
-difficile* ribotype 027 spores (**Add schematic?**).
-The mice were followed for 10 days post challenge and their stool was collected and plated for *C.
-difficile* CFU to detect colonization level.
 Suprisingly, communities from all donors were colonized (Figure 2).
@@
 difficile* colonization both recieved their community from the same donor.
-Therefore, the transplant communties were susceptible to colonization without further modification of the communities.
+Therefore, the transplanted human fecal communties were susceptible to *C.
+difficile* colonization without the need for any perturbation.
 
 **Infection seveity varies by initial community structure.** After challenging the mice with *C.
-difficile*, we looked at the effects of the different communities on the outcome from the *C.
-difficile* challenge.
+difficile*, we looked at the effect of the community variation on the outcome from the *C.
+difficile* infection.
 We followed the mice for signs of disease as well as *C.
 difficile* colonization and toxin production for 10 days post challenge.
-Some mice were not initally colonized on the day after challenge and took a few days to become fully colonized (**SUPP FIG?**).
-There were a group of mice, all from a set of 6 donors, highly colonized one day post challenge that became moribund within 3 days post challenge.
+Some mice were not initally colonized on the day after *C.
+difficile* challenge but became colonized with a few days post challenge (**SUPP FIG?**).
+A subset of mice, all that recieved their comunity from one of 6 donors, suffered from severe disease, which became highly colonized one day post challenge and moribund within 3 days post challenge.
 The remaining mice, except for one cage, maintained *C.
 difficile* colonization through the end of the experiment (Figure 2).
-After the  mice were humanely euthanised on the experiment endpoint, day 10 post challenge or earlier if they succumbed to the infection, their cecal tissue was collected and scored for histological signs of disease and fecal samples were assayed for toxin (Figure 3).
+After the  mice were humanely euthanised at the endpoint, day 10 post challenge or earlier if they became moribund, their cecal tissue was collected and scored for histological signs of disease and fecal samples were assayed for toxin (Figure 3).
 Overall, there was greater toxin activity detected in the stool of the moribund mice (*P* = 0.003).
-However, when looking at each group of mice, we saw there was a range in toxin activity for both the mice with mild disease and moribund mice (Figure 3A).
-Some donor groups that maintained colonization without severe disease had similarlly high activity levels as was detected in the moribund mice.
-Additionally, some moribund mice did not have toxin activity detected in their stool.
+However, when looking at each group of mice, we saw there was a range in toxin activity for both the mice with non-severe disease and severe disease (Figure 3A).
+Some donor groups that maintained colonization without severe disease had similarlly high activity levels as was detected in the mice with severe disease.
+Additionally, some mice with severe disease did not have toxin activity detected in their stool.
 These oberserved variations in toxin activity and relation to disease severity have been previously characterized [@Akerlund2006, @Vitucci2020].
-Next we examined the histological damage caused by the *C.
+Next we examined the cecal tissue for histopathological damage.
+For mice with severe disease, we observed high levels of epithelial damage, tissue edema and inflammation (Figure SXXXX), similar to previously reported histopathologic findings for *C.
+difficile* ribotype 027 [CITATION, @Cowardin2016].
+As observed with toxin activity, the mice with severe disease had higher histopathologic scores than the mice with non-severe disease (*P* = 3.0e-9).
+However, unlike the toxin activity, all moribund mice had consistently high histopathologic scores.
+The mice which maintained persistent population size, we saw a range in tissue damage from no signs of disease to nearly the same level as the mice with severe disease, which grouped by community donor.
+Together, the toxin activity, histopathologic score, and moribundity showed variation across the donors but were largely consistent within each group of mice recieving the same human donor fecal community.
+
+**Populations of the microbial communities explain variation in CDI disease severity.** Since the disease charateristics grouped by human donor fecal community, we next investigated the community composition for its ability to explain the variation in disease severity.
+First, we tested for associations between taxonomic groups and the outcomes of the *C.
 difficile* infection.
-For mice that succumb to the infection prior to day 10 post challenge, we observed high levels of epithelial damage, tissue edema and inflammation, similar to previously reported histological findings for *C.
-difficile* ribotype 027 [CITATION, @Cowardin2016].
-As observed with toxin activity, the moribund mice had higher disease clinical scores than the mice with mild disease (*P* = 3.0e-9).
-However, unlike the toxin activity, all moribund mice had consistently high clinical scores.
-The mice which maintained persistent level of colonization, we saw a range in tissue damage from no signs of disease to nearly the same level as the moribund mice, which grouped by community donor.
-Together, the toxin activity, clinical score, and moribundity showed variation across the donors but were largely consistent within each group of mice recieving the same human fecal community.
-
-**Populations of the microbial communities explain variation in disease severity.** Since the disease charateristics grouped by human fecal community donor, we next looked if the composition of the community could explain the variation we observed.
-First, we tested for associations between taxonomic groups and the outcomes of the *C.
-difficile* challenge.
-We used the linear discriminant analysis (LDA) effect size (LEfSe) method to identify the Genus level differences in the initial bacterial communities between the severity level and toxin production.
+We used the linear discriminant analysis (LDA) effect size (LEfSe) method to identify differences in the genera of the initial bacterial communities between the severity level and toxin production.
 We split the mice into groups by severity level based on their if they became moribund (severe), or if their clinical score was above 5 (moderate) or below 5 (mild).
 This analysis revealed 20 genera that were significantly different by the disease severity (Figure 4A).
-The presence of a group of populations of *Turicibacter*, *Streptococcus*, *Staphylococcus*, *Pseudomonas*, *Phocaeicola*, *Parabacteroides*, *Bacteroides*, and *Escherichia/Shigella* were detected at higher levels in the moribund group.
-Populations of *Anaerotignum*, *Coprobacillus*, *Enterocloster*, and *Murimonas* were increased in the group of mice that only had mild disease.
-There were many genera that did not have a correlation of relative abundance and disease severity, such as the populations of *Monoglobus* and *Hungatella*, which were decreased or increased, respectively, in only the group of mice with high clinical scores but did not succumb to the infection.
-To understand these cases better we used LEfSe to identify the genera differentiating the communities that we detected toxin in from those that did not (Figure 4B).
+The presence of populations of *Turicibacter*, *Streptococcus*, *Staphylococcus*, *Pseudomonas*, *Phocaeicola*, *Parabacteroides*, *Bacteroides*, and *Escherichia/Shigella* were detected at higher levels in the mice that developed severe disease.
+Populations of *Anaerotignum*, *Coprobacillus*, *Enterocloster*, and *Murimonas* were increased in the mice that would experience mild disease.
+There were many genera that did not have a correlation of relative abundance and disease severity, such as the populations of *Monoglobus* and *Hungatella*, which were decreased or increased, respectively, in only the group of mice that would develop moderate disease.
+To understand these cases better we used LEfSe to identify the genera differentiating the communities that we detected toxin in from those that we did not (Figure 4B).
 There were many genera the were associated with toxin production as we would expect from their association with severe disease, such as populations of *Escherichia/Shigella* and *Bacteroides*.
@@
 Populations of *Hungatella* were increased in the group with moderate disease but not with the production of toxin.
-Lastly, we looked for associations between the clinical score and the genera populations at that same time (Figure 4C).
-The populatons of *Bacteroides* at the end matched the trend observed in the initial community, increased population correlated with increased clinical score.
+Lastly, we looked for associations between the histopathologic score and the populations of genera at that same time (Figure 4C).
+The populatons of *Bacteroides* at the end matched the trend observed in the initial community, increased population correlated with increased histopathologic score.
 Populations of *Klebsiella* and *Prevotellaceae* positively correlated with clinic score and were increased in the group that produced toxin.
-These tests have shown which individual populations associated with the different outcomes.
+These tests have identified which individual populations of bacteria that associated with each disease severity.
 
-We next wanted to understand how the populations in the context of each other lead to the different outcomes from CDI.
-We trained random forest models with the intital bacterial community relative abundance data at each taxonomic rank to predict toxin production, moribundity or final clinical score.
+We next wanted to understand how the populations in the context of each other lead to the different disease severities.
+We trained random forest models with the intital bacterial community relative abundance data at each taxonomic rank to predict toxin production, severe disease or final histopathologic score.
 Overall for predicting toxin production, microbial populations aggregated by their phylum level classificiation performed as well as models using lower taxonomic ranks (AUROC = 0.83, Figure S2).
-The initial community prior to challenge, populations of *Verrucomicrobia*, *Campilobacterota* and *Proteobacteria* contributed the most to the correct prediction of toxin production (Figure 5A).
+In the initial community populations of *Verrucomicrobia*, *Campilobacterota* and *Proteobacteria* contributed the most to the correct prediction of toxin production (Figure 5A).
 *C.
 difficile* was more likely to produce toxin when the initial community had smaller populations of *Verrucomicrobia* and *Campilobacterota* and had larger popluations of *Proteobacteria*.
-Next, we looked at the ability of the community to predict moribundity.
-The community grouped at the class rank was sufficient to predict if the mouse would succumb to the infection before the end of the experiment (AUROC = 0.91, Figure S2).
+Next, we assessed the ability of the community to predict severe disease.
+Class rank classification was sufficient to predict if the mouse would succumb to the infection before the end of the experiment (AUROC = 0.91, Figure S2).
 Groups of bacteria belonging to *Bacilli*, *Firmicutes*, and *Erysipelotrichia* contributed the most to the correct prediction.
@@
 Aside from *Erysipelotrichia*, the only other population decreased in moribund mice was *Clostridia*, all other features had increased populations in the moribund mice.
-Lastly, the initial community was able to predict if the mice would have a clinical score above or below 5 with the relative abundances from the genus rank (AUROC = 0.99, Figure S2).
+Lastly, the initial community was able to predict if the mice would have a histopathologic score above or below the median score (median histopathologic score = 5) with the genera relative abundances (AUROC = 0.99, Figure S2).
 There were no genera with a significantly greater effect on the model performance than any others, indicating the model is reliant on many genera for the correct prediction.
 The model used some of the genera identified in the LEfSe analysis, such as populations of *Coprobacillus*, *Anaerostipes*, and *Hungatella*.
-Communities with greater populations of *Hungatella*, *Eggerthella*, *Bifidobacterium*, *Duncaniella* and *Neisseria* were more likely to have higher clinical scores.
+Communities with greater populations of *Hungatella*, *Eggerthella*, *Bifidobacterium*, *Duncaniella* and *Neisseria* were more likely to have greater histopathologic scores.
 
 **Disease severity consistent by donor community across strains/isolates.** We used a single *C.
-difficile* isolate to characterize the range of disease severity and the features of the bacterial community associated with severity.
-Based on these results, we hypothesized the community was responsible for the disease severity not the specific *C.
-difficile* isolate we used in our experiments.
-To test this hypothesis, we selected three communities based on the result from our analysis thus far to select one community we would expect to result in a low clinical score (< 5), a high clinical score (> 5), and one with a high clinical score which became moribund.
-Using genus level data to select the communities, we selected communities based on the relative abundance patterns of *Sellimonas* and *Anaerotignum* (higher abundance in mice with a low clinical score), *Lachnospiracea* (lower abundance in moribund mice), *Hungatella* and *Eisenbergiella* (higher abundance in moderate mice and lower in moribund mice), Clostridium sensu stricto (higher in moribund mice by both LEfSe analysis and RF model) and *Enterocloster* (lower in moribund mice and a negative linear trend with disease severity by LEfSe analysis).
+difficile* isolate, ribotype 027 isolate 431 to characterize the range of disease severity and the features of the bacterial community associated with severity [@Carlson2013].
+Since we had used the same mice and *C.
+difficile* isolate in our first set of experiemnts, we next wanted to test if the effect of the community would apply to other *C.
+difficile* isolates.
+We selected three communities based on the result from our analysis thus far to select one community we would expect to result in a low clinical score (< 5), a high clinical score (> 5), and one with a high clinical score which becomes moribund.
+Using genus level data to select the communities, we selected communities based on the relative abundance patterns of *Sellimonas* and *Anaerotignum* (higher abundance in mice with a low clinical score), *Lachnospiracea* (lower abundance in moribund mice), *Hungatella* and *Eisenbergiella* (higher abundance in moderate mice and lower in moribund mice), Clostridium sensu stricto (higher in moribund mice by both LEfSe analysis and RF model) and *Enterocloster* (lower in moribund mice and a negative linear trend with disease severity by LEfSe analysis) (Figure S3).
 With this set of human fecal communities selected, we inoculated mice via oral gavage of the human fecal slurry and allowed two weeks for the a community to establish.
@@
 difficile* [@Carlson2013].
-The disease severity for this set of experiments closely matched that in our original set which were challenged with *C.
+The disease severity for this set of experiments closely matched that in our first set of experiments with *C.
 difficile* isolate 431 (Figure 6).
@@
 Challenging mice colonized with different human fecal communities with a single *C.
-difficile* isolate allowed us to identify the effect of microbiome variation on the disease severity caused by *C.
-difficile* infection.
-Our LEfSe analyses and random forest models revealed the relationships between the bacterial community members and the CDI disease severity.
-We used the relative abundance patterns of community members with strong associations with disease outcome to select communities to test if the predicted severity were *C.
-difficile* isolate specific.
+difficile* isolate allowed us to identify the effect of microbiome variation on *C.
+difficile* infection disease severity.
+Our LEfSe analyses and random forest models revealed the relationships between the bacterial community members and the degree of disease severity.
+We used the relative abundance patterns of community members with strong associations with disease severity to select communities to test if the observed disease severity is conserved when other *C.
+difficile* isolates are used.
 The disease outcome from the *C.
 difficile* chalenge with the new isolate matched the expected disease severities.
-Overall, these results have demonstrated that the bacterial community is capable of modulating the disease severity of a *C.
+Overall, these results have demonstrated that the bacterial community is capable of modulating the disease severity of *C.
 difficile* infection.
 
-CDI severity can range from no symptoms, mild symptoms such as diarrhea, or critical symptoms such as pseudomembranous colitis, toxic megacolon or death.
-At the clinical level, physicians use prediction tools that identify patients most at risk of developing a severe infection using white blood cell counts, serum albumin level, or serum creatine level [@Lungulescu2011, @McDonald2018, @Zar2007].
-Those levels are being driven by the activity in the intestine [@diMasi2018].
+CDI severity can range from no symptoms, mild symptoms such as diarrhea, or severe symptoms such as toxic megacolon or death.
+Physicians use prediction tools to identify patients most at risk of developing a severe infection using white blood cell counts, serum albumin level, or serum creatine level [@Lungulescu2011, @McDonald2018, @Zar2007].
+Those levels are being driven by the activities in the intestine [@diMasi2018].
 Research into the drivers of this variation have revealed factors that make *C.
 difficile* more virulent.
-Strains are categorized for their virulence by the presence and production toxin A and toxin B and the prevelance in outbreaks, such as 027 and 078 [@AbernathyClose2020, @Theriot2011, @Goorhuis2008, @OConnor2009, @Chen2008].
-However, other studies have shown that virulence is not necessarily linked with toxin production [@Roa2015] or the strain [@Walk2012].
+Strains are categorized for their virulence by the presence and production toxin A and toxin B and the prevelance in outbreaks, such as ribotype 027 and 078 [@AbernathyClose2020, @Theriot2011, @Goorhuis2008, @OConnor2009, @Chen2008].
+However, other studies have shown that virulence is not necessarily linked with toxin production **@Roa2015** or the strain **@Walk2012**.
 Furthermore, there is variation in the genome, growth rate, sporulation, germination, and toxin production in different isolates of a strain [@Carlson2013, @He2010].
-These variations may help explain why the severe CDI prediction tools often miss identifying many patients with CDI that will develop severe disease [@Dieterle2020, @Butt2013, @Perry2021, @vanBeurden2017].
+This variation may help explain why severe CDI prediction tools often miss identifying many patients with CDI that will develop severe disease [@Dieterle2020, @Butt2013, @Perry2021, @vanBeurden2017].
 Therefore, it is necessary to gain a full understanding of all factors contributing to disease variation to improve our ability to predict severity.
@@
 Most CDI studies employ either an *in vitro* or *in vivo* model using a homogenous bacterial community.
-Collins et al.
-used multiple human communities to colonize mice, however the communities were pooled prior to gavaging into germ-free mice [@Collins2015], resulting in a single community.
-Studies examining difference in disease often use different strains in mice with similar microbiota as a proxy for variation in disease, such as 630 for non-severe and 027 for severe [@AbernathyClose2020, @Theriot2011, @Lawley2012, @Chen2008].
-There has been studies demonstrating vairation in severity, but by tapering antibiotic dosage [@Lesniak2021, @Schubert2015, @Reeves2011] or by reducing the amount of cells or spores used for the challenge [@Chen2008, @Reeves2011].
-Our group has recently been uncovering the variation the microbiome affects CDI but was limited to *C.
+Collins et al used multiple human communities to colonize mice, however the communities were pooled prior to gavaging into germ-free mice [@Collins2015], resulting in a single community.
+Studies examining difference in disease often use different *C.
+difficile* strains or ribotypes in mice with similar microbiota as a proxy for variation in disease, such as strain 630 for non-severe and ribotype 027 for severe [@AbernathyClose2020, @Theriot2011, @Lawley2012, @Chen2008].
+There has been studies demonstrating variation in severity, but through tapering antibiotic dosage [@Lesniak2021, @Schubert2015, @Reeves2011] or by reducing the amount of cells or spores used for the challenge [@Chen2008, @Reeves2011].
+Our group has recently been uncovering how variation in the microbiome affects CDI but have been limited to *C.
 difficile* colonization [@Lesniak2021, @Tomkovich2020].
@@
 With our recent observations that the initial community affects the ability of *C.
-difficile* to persist in an environment and the existing research describing the myriad ways the microbiome affects the lifestyle of *C.
-difficile*, we hypothesized that the microbiome directly and indirectly modulates the disease severity of CDI.
-Our data demostrate bacterial community member abundances associate with variation in toxin production, histological scoring of the intestine and severe disease.
+difficile* to persist in an environment and the existing research describing the myriad ways the microbiome interacts with *C.
+difficile*, we hypothesized that the microbiome directly and indirectly modulates CDI disease severity.
+Our data demostrate gut bacterial relative abundances associate with variation in toxin production, histopathologic scoring of the cecum and severe disease.
 This anaylsis has revealed populations of *Akkermansia*, *Anaerostipes*, *Coprobacillus*, *Enterocloster*, *Lactonifactor*, and *Monoglobus* were more abundant in the microbiome of mice which little to no disease was detected.
@@
 *Anaerostipes* and *Coprobacillus* produce short chain fatty acids which also have been associated with healthy communities [@Han2019, @Duncan2004, @Ye2018].
-Furthermore, *Coprobacillus*, which is abundant in mice with low clinical disease scores but rare in all other mice, has been shwon to contain putative type I lantibiotic gene cluster and inhibit *C.
+Furthermore, *Coprobacillus*, which is abundant in mice with low histopathologic scores but rare in all other mice, has been shwon to contain putative type I lantibiotic gene cluster and inhibit *C.
 difficile* colonization [@Walsh2017, @Sandiford2018, @Stein2013].
@@
 In our data, *Akkermansia* is most abundant in the mice with least disease but there were some mice with severe disease which had increased populations of *Akkermansia*.
-This could be attributed to either a more protective mucus layer had developed in the mice with little disease and *Akkermansia* is inhibiting colonization [@Stein2013, @Nakashima2021] or *Akkermansia* could be crossfeeding or exposing a niche via its mucus consumption [@Geerlings2018, @Deng2018, @Engevik2020].
-Similarly with *Enterocloster*, in our data this genus was more abundant and associated with low clinical disease scores.
-This genus has been associated with healthy populations and has been used to mono-colonize germ-free mice to reduce the ability of *C.
+This could be attributed to either a more protective mucus layer had developed in the mice with little disease and *Akkermansia* is inhibiting colonization [@Stein2013, @Nakashima2021] or *Akkermansia* could be crossfeeding *C.
+difficile* or exposing a niche via its mucus consumption [@Geerlings2018, @Deng2018, @Engevik2020].
+Similarly with *Enterocloster*, in our data this genus was more abundant and associated with low histopathologic scores.
+It has been associated with healthy populations and has been used to mono-colonize germ-free mice to reduce the ability of *C.
 difficile* to colonize [@Reeves2012, @Ma2021].
@@
 These data demostrate a population of bacteria has the potential to be either protective or harmful.
-The disease outcome is not based on the abundance of a single population of bacteria rather the result of the interactions between the host, microbiome and infecting bacteria.
+The disease outcome is not likely based on the abundance of a single population of bacteria rather the result of the interactions of many populations.
 
-The groups of bacteria that were associated with either a higher clinical disease score or severe disease are members of an indigenous gut community that also have been associated with disease.
+The groups of bacteria that were associated with either a higher histopathologic score or severe disease are members of the indigenous gut community that also have been associated with disease.
 These are often referred to as opportunitic pathogens, however based on the the results reported in this study it is more appropriate to apply the damage-response framework for microbial pathogenesis [@Pirofski2008, @Casadevall2014].
-The disease is not driven by a single entity, rather it is an emergent property of the responses of the host immune system, infecting microbe and the microbes of the indigenous community.
+The disease is not driven by a single entity, rather it is an emergent property of the responses of the host immune system, infecting microbe, and the microbes of the indigenous community.
 Many of the populations with pathgenic potential that associated with worse outcomes are also facultative anaerobes.
 *Enterococcus*, *Klebsiella*, *Shigella/Escherichia*, *Staphylococcus*, and *Streptococcus* have been shown to expand their population with antibiotic use [@VanInsberghe2020, @GarzaGonzalez2019, @Schubert2014] and are commonly detected in CDI cases [@Shafiq2020, @Keith2020, @Zackular2016, @Berkell2021].
-In addition to these populations, another set of bacteria, *Eggerthella*, *Prevotellaceae* and *Helicobacter*, associated with worse outcomes have also been associated with intestinal inflammation [@Gardiner2014, @Iljazovic2020, @Nagalingam2013].
-*Helicobacter hepaticus* has been demostrated to be sufficient to cause susceptibility to CDI in IL-10^(-/-) B57BL/6 mice [@AbernathyClose2021].
-In our mice, if *Helicobacter* was detected the mouse resulted in either severe disease or a high clinical score.
-While we did not use IL-10^(-/-) mice, it is possible the bacterial communtity or host response are similarly modified by *Helicobacter* allowing C.
-difficile infection and resulting in host damage.
+In addition to these populations, another set of bacteria, populations of *Eggerthella*, *Prevotellaceae* and *Helicobacter*, associated with worse outcomes have also been associated with intestinal inflammation [@Gardiner2014, @Iljazovic2020, @Nagalingam2013].
+Recently, *Helicobacter hepaticus* has been demostrated to be sufficient to cause susceptibility to CDI in IL-10^(-/-) B57BL/6 mice [@AbernathyClose2021].
+In our experiments, when *Helicobacter* was present, the mouse resulted in either severe disease or a high histopathologic score.
+While we did not use IL-10^(-/-) mice, it is possible the bacterial communtity or host response are similarly modified by *Helicobacter* allowing *C.
+difficile* infection and resulting in host damage.
+However, the populations identified other than *Helicobacter* are not as clear with their relationship between their relative abundance and the resulting disease severity.
 
-The damage-response framework for pathogenesis best explains how bacterial populations identified in this study potentiate CDI disease severity through the varied interactions with the host and *C.
+The damage-response framework for pathogenesis helps to understand how bacterial populations identified in this study potentiate CDI disease severity through the varied interactions with the host and *C.
 difficile*.
 In this set of experiments, we use the same host, C57BL/6 mice, the same infecting microbe, *C.
-difficile* RT027 isolate 431, with different gut bacterial communities.
+difficile* ribotype 027 isolate 431, with different gut bacterial communities.
 The members of those communities were often present in multiple levels of disease severity.
-Thus, it is not merely, the presence of the microbe but their activity in response to the other microbes and host which determine their effect on host damage.
+Thus, it is not merely, the presence of the bacteria but their activity in response to the other microbes and host which determine their effect on host damage.
 Additionally, while we are using the same host and infecting microbe, they too are reacting to the specific members of the bacterial community.
@@
 Low levels of mucin degradation can provide nutrients to other community members producing a diverse non-damaging community [].
-However, if mucin degradation becomes too great that it reducing the its protective function and exposes the epithelial cells [].
+However, if mucin degradation becomes too great it reduces the protective function of the mucin layer and exposes the epithelial cells [].
 This over-harvesting contributes to the host damage due to other members producing toxin.
@@
 Host damage is the emergent property of numerous damage-response curves, one for host immune response, one for *C.
-difficile* activity and another for microbiome community activity, each of which are a composite curve of the individual activities from each group, such as antibody production, neutrophil infiltration, toxin production, sporulation, fiber and mucin degradation, etc.
+difficile* activity and another for microbiome community activity, each of which are a composite curve of the individual activities from each group, such as antibody production, neutrophil infiltration, toxin production, sporulation, fiber and mucin degradation.
+Therefore, while we have identified populations of interest, it is likely targeting one specifically will reduce disease severity.
 
 Many studies have investigated individual activities contributing to CDI.
-Here we have shown myraid bacterial groups and their associated activities result in variation in CDI disease severity.
+Here we have shown myraid bacterial groups and their relative abundances associate with variation in CDI disease severity.
 We must continue to explore how the host, microbiota, and *C.
@@
 Approaching this problem from a damage-response framework allows us to experiment with modifying subsets of activities to reduce overall host damage.
-Our current treatment approach treats an individual group in this tripartite system.
+Our current clinical treatment targets an individual group in this tripartite system.
 Most commonly, CDI is treated with an antibiotic to eliminate *C.
@@
 difficile* toxins.
-These treatments are only addressing a single entity of the system, leaving the potential for the other two maintain responses that result in continued host damage.
-However we can continue this research to elucidate treatments that drive all three entities towards reduced host damage.
-When a patient is diagnosed with CDI, the gut community composition, in addition to information on the immune response and *C.
-difficile* ribotype, can inform the predicted severity and appropriate treatment.
+These treatments are only addressing a single entity of the system, leaving the potential for the others to maintain responses that maintain host damage.
+However, we can continue this research to elucidate treatments that drive all three entities towards reduced host damage.
+When a patient is diagnosed with CDI, the gut community composition, in addition to the traditionally obtained information, may improve our severity prediction and appropriate treatment.
 Treating the microbiome may contribute to reduced severity.

### 2021-10-27 submission/manuscript.tex
revise draft with initial proof read

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}
\usepackage[labelfont=bf]{caption}

\usepackage[none]{hyphenat}

\usepackage{setspace}
\doublespacing
\setlength{\parskip}{1em}

\usepackage{lineno}

\usepackage{pdfpages}
\floatplacement{figure}{H} % Keep the figure up top of the page
\DeclareUnicodeCharacter{0301}{*************************************}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {}%
  {\par}

\author{}
\date{\vspace{-2.5em}}

\begin{document}

\pagenumbering{arabic}
\linenumbers
\doublespacing

\hypertarget{the-gut-microbiota-potentiates-clostridiodes-difficile-infection-severity.}{%
\section{\texorpdfstring{The gut microbiota potentiates
\emph{Clostridiodes difficile} infection
severity.}{The gut microbiota potentiates Clostridiodes difficile infection severity.}}\label{the-gut-microbiota-potentiates-clostridiodes-difficile-infection-severity.}}

\vspace{30mm}

\textbf{Running title:} Microbiota potentiates \emph{Clostridioides
difficile} infection severity

\vspace{20mm}

Nicholas A. Lesniak\(^1\), Alyxandria M. Schubert\(^1\), Kaitlyn
Flynn\(^1\), Hamide Sinani\(^1\), Kathyrn A. Eaton\(^1\), Ingrid L.
Bergin\(^2\), Patrick D. Schloss\(^{1,\dagger}\)

\vspace{30mm}

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}\\
1. Department of Microbiology and Immunology, University of Michigan,
Ann Arbor, MI\\
2. Unit for Laboratory Animal Medicine, University of Michigan, Ann
Arbor, MI

\newpage

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

\emph{Clostridioides difficile} infection has become more common and
severe in the last few decades. Patient age, white blood cell count,
creatinine levels as well as \emph{C. difficile} ribotype and the
presence of toxin genes have been associated with disease severity.
However, models based on those data to identify patients with CDI that
will develop severe disease have not been robust enough to be broadly
applied. These models are input from our understanding of \emph{C.
difficile} interactions with the gut environment. The gut bacteria are a
key determinant in \emph{C. difficile} infection. CDI is dependent on
perturbations in the bacterial community in order to colonize the gut.
The gut microbiota also impairs \emph{C. difficile} colonization through
bile acid metabolism, nutrient consumption and bacteriocin production.
However, it is unclear if the gut bacteria affect the disease severity
resulting from that colonization. Here we have shown gut bacterial
communities contribute to disease severity variation. We created diverse
gut communities by colonizing germ-free mice with human fecal
communities. They were then infected with \emph{C. difficile} which
resulted in differed disease severities. The severity grouped by the
human fecal community they received. Generally, facultative anaerobes
with pathogenic potential, such as Escherichia, Helicobacter and
Klebsiella, were associated with more severe disease. Bacterial groups
associated with dietary fiber degradation, such as Coprobacillus, were
associated with reduced disease severity. Lastly, we showed communities
that resulted in either low or high histopathologic scores or severe
disease had the same outcome when infected with a different \emph{C.
difficile} isolate.

\hypertarget{importance}{%
\subsection{Importance}\label{importance}}

Clostridioides difficile infection (CDI) causes a range of disease from
asymptomatic to mild diarrhea to severe outcomes of recurrent infections
and even death. Models that predict severity and treatment decisions are
based on clinical factors and \emph{C. difficile} capabilities. But
currently the gut microbiome, the primary protector against CDI, has not
been accounted for its effect on CDI disease severity. We demonstrated
variation in the microbiome of mice colonized with human feces was
sufficient to result in a range of disease severities. Our results show
groups of bacteria contribute to the development of disease in \emph{C.
difficile} infections. Gut bacterial community data from patients with
CDI could improve our ability to identify patients more at risk for
severe disease and improve interventions which treat the gut bacteria to
reduce host damage.

\newpage

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

\emph{Clostridioides difficile} infections (CDI) have been increasing in
incidence and severity since \emph{C. difficile} was first identified as
a pathogen. It was first identified in 1931 in a healthy fecal community
{[}CITATION{]}. But then in the late 1970s, following the introduction
of broad-spectrum antibiotics, \emph{C. difficile} was identified as the
cause of antibiotic-associated psuedomembranous colitis {[}CITATION{]}.
Beginning in the early 2000s, CDI became more problematic with
increasing incidences and severity {[}CITATION{]}. Disease severities
can range from asymptomatic to mild diarrhea to toxic megacolon and
death. This severity has been associated with characteristics of
\emph{C. difficile} ribotypes and host factors. Over time more
pathogenic ribotypes have emerged, such as ribotype 027 in the early
2000s. The presence of the genes for Toxin A/B or fluoroquinolon
resistence has been associated with more severe disease {[}CITATION{]}.
Binary toxin, while alone insufficient to cause damage on it own, when
in the presence of Toxin A/B has been associated with increased
virulence {[}CITATION{]}. Mutations in tcd C, the negative regulator of
of toxin transcription, also increased the virulence of \emph{C.
difficile} {[}CITATION{]}. In addition to the pathogenic potential of
the infecting strain, the response by the host also affects the disease
severity. Delayed or reduced production of antibodies IgG and IgA can
result in more severe disease {[}CITATION{]}. Increased neutrophil
inflitration can lead to worse outcomes {[}CITATION{]}. In addition to
the immune response, increased density of the toxin receptor on the host
cells can increase the severity of the infection {[}CITATION{]}. While
there is a thorough understanding of many of the mechanisms driving in
CDI incidence and disease severity, there still is not effective tools
to reduce the risk of severe disease.

Models have been developed to identify patients at risk for severe CDI
but have not been robust for broad application (1). These models utilize
patient related data, such as patient age, white blood cell count, serum
albumin levels, and creatinine levels, to predict disease severity. In
their initial publication, these models perform well enough for
publication. However when Perry et al applied many of the most current
models to a multi-center external validation, their performance suffered
and often had more false positives than true positives (3). Thus, there
seems to be a missing factor contributing to disease severity.

Missing from these models are the gut bacteria which also interact with
\emph{C. difficile} during the course of the infection. The indigenous
bacteria of a healthy intestinal community provide a protective barrier
preventing \emph{C. difficile} from infecting the gut. Only after this
community is disrupted from a range of pertubations, such as many
classes of antibiotics, medications, or even dietary changes, can
\emph{C. difficile} infect the intestine. Once established, the gut
bacteria can either promote or inhibit \emph{C. difficile} through
producing molecules or modifying the environment {[}Abbas2020{]}. Bile
acids metabolized by the gut bacteria can inhibit \emph{C. difficile}
growth and affect toxin production (5). Bacteria in the gut also can
compete more directly with \emph{C. difficile} through antibiotic
production or nutrient consumption (7). To eliminate CDI, the primary
treatment is a course of antibiotics. Patients who become reinfected and
recalitrant to antibiotics, a fecal microbiota transplantion often is
sufficient to restore the gut community to eliminate \emph{C.
difficile}. While there has been many studies to investigate the
interaction between the gut bacteria and \emph{C. difficile} during the
infection, it has not been demostrated the bacterial community can
modulate the disease severity of CDI. Recently, Tomkovich et al showed
the variation in the bacterial communtiies between mice from different
colonies and vendors was sufficient to cause differences in the temporal
dynamics of the \emph{C. difficile} infection (9). Since the gut
bacteria interact with \emph{C. difficile} throughout the infection and
community difference resulted in different colonization dynamics, we
hypothesized that the gut bacteria contribute to the variation in CDI
disease severity. Here, we colonized germ-free C57BL/6 mice with human
fecal communities to test if the variablity in the gut bacteria can
produce and explain variation in the disease severity caused by a single
\emph{C. difficile} isolate.

\hypertarget{results}{%
\subsection{Results}\label{results}}

\textbf{Germ-free mice colonized with human fecal commmunities have
diverse and unique community structures.} Based on our previous
observation mice with variation in their microbiota showed differences
in colonization of \emph{C. difficile} (9), we sought to explore the
effect communties with greater variation had on \emph{C. difficile}
challenge. To produce communities with greater variation than mouse
colonies with their indigenous communties, we used human fecal
communities to establish a community in the intestine of germ-free mice.
We inoculated a cage of germ-free C57BL/6 mice (2-4 mice per cage) with
homogenized feces from one of 15 different human donors via oral gavage.
The fecal community was allowed to colonize and stabilize for two weeks
post inoculation {[}CITATION?{]}. We then assessed the gut bacterial
communities by sequencing the V4 region of the 16S rRNA gene extracted
from fecal pellets (Figure 1A). The communities established in the mice
group most closely to other mice recieving the same human fecal donor
community than to either their respective donor or any mice who received
stool from a different donor (Figure 1B). Most of the communities were
primairily composed of populations of \emph{Clostridia},
\emph{Bacteroidia}, \emph{Erysipelotrichia}, \emph{Bacilli}, and
\emph{Gammaproteobacteria}. However, each group of mice harbored unique
combination of class abunadances in their gut bacterial communities.
\emph{Any alpha? compare to murine variation? Number of unique otus?}

\textbf{\emph{C. difficile} is able to colonize mice without
perturbation.} Our goal was to test the effect of the variation in the
microbiota on \emph{C. difficile} infection. A typical mouse model of
CDI requires pretreatment with antibiotics such as clindamycin to become
susceptible to \emph{C. difficile} colonization {[}CITATION{]}. However,
we wanted to avoid modifying the communties with an antibiotic to
maintain their community structures and differences. Since these
communties came from human donors which may have become susceptible
through their own exposures, we decided to test if \emph{C. difficile}
was able to colonize these communities without any perturbation. After
two weeks post incoluation with human fecal communities, the mice were
then challenged with 10\^{}\textbf{4?} \emph{C. difficile} ribotype 027
isolate 431 spores (\textbf{Add schematic?}) (10). The mice were
followed for 10 days post challenge and their stool was collected and
plated for \emph{C. difficile} CFU to detect colonization level. We
hypothesized that \emph{C. difficile} would most likely be able to
colonize the mice with communities that came from donors which had a
history of antibiotic use, one of the primary risk factors for
developing CDI (TABLE, CITATION). Suprisingly, communities from all
donors were colonized (Figure 2). The only two mice that were able to
resist \emph{C. difficile} colonization both recieved their community
from the same donor. Therefore, the transplanted human fecal communties
were susceptible to \emph{C. difficile} colonization without the need
for any perturbation.

\textbf{Infection seveity varies by initial community structure.} After
challenging the mice with \emph{C. difficile}, we looked at the effect
of the community variation on the outcome from the \emph{C. difficile}
infection. We followed the mice for signs of disease as well as \emph{C.
difficile} colonization and toxin production for 10 days post challenge.
Some mice were not initally colonized on the day after \emph{C.
difficile} challenge but became colonized with a few days post challenge
(\textbf{SUPP FIG?}). A subset of mice, all that recieved their comunity
from one of 6 donors, suffered from severe disease, which became highly
colonized one day post challenge and moribund within 3 days post
challenge. The remaining mice, except for one cage, maintained \emph{C.
difficile} colonization through the end of the experiment (Figure 2).
After the mice were humanely euthanised at the endpoint, day 10 post
challenge or earlier if they became moribund, their cecal tissue was
collected and scored for histological signs of disease and fecal samples
were assayed for toxin (Figure 3). Overall, there was greater toxin
activity detected in the stool of the moribund mice (\emph{P} = 0.003).
However, when looking at each group of mice, we saw there was a range in
toxin activity for both the mice with non-severe disease and severe
disease (Figure 3A). Some donor groups that maintained colonization
without severe disease had similarlly high activity levels as was
detected in the mice with severe disease. Additionally, some mice with
severe disease did not have toxin activity detected in their stool.
These oberserved variations in toxin activity and relation to disease
severity have been previously characterized (11). Next we examined the
cecal tissue for histopathological damage. For mice with severe disease,
we observed high levels of epithelial damage, tissue edema and
inflammation (Figure SXXXX), similar to previously reported
histopathologic findings for \emph{C. difficile} ribotype 027 (13). As
observed with toxin activity, the mice with severe disease had higher
histopathologic scores than the mice with non-severe disease (\emph{P} =
3.0e-9). However, unlike the toxin activity, all moribund mice had
consistently high histopathologic scores. The mice which maintained
persistent population size, we saw a range in tissue damage from no
signs of disease to nearly the same level as the mice with severe
disease, which grouped by community donor. Together, the toxin activity,
histopathologic score, and moribundity showed variation across the
donors but were largely consistent within each group of mice recieving
the same human donor fecal community.

\textbf{Populations of the microbial communities explain variation in
CDI disease severity.} Since the disease charateristics grouped by human
donor fecal community, we next investigated the community composition
for its ability to explain the variation in disease severity. First, we
tested for associations between taxonomic groups and the outcomes of the
\emph{C. difficile} infection. We used the linear discriminant analysis
(LDA) effect size (LEfSe) method to identify differences in the genera
of the initial bacterial communities between the severity level and
toxin production. We split the mice into groups by severity level based
on their if they became moribund (severe), or if their clinical score
was above 5 (moderate) or below 5 (mild). This analysis revealed 20
genera that were significantly different by the disease severity (Figure
4A). The presence of populations of \emph{Turicibacter},
\emph{Streptococcus}, \emph{Staphylococcus}, \emph{Pseudomonas},
\emph{Phocaeicola}, \emph{Parabacteroides}, \emph{Bacteroides}, and
\emph{Escherichia/Shigella} were detected at higher levels in the mice
that developed severe disease. Populations of \emph{Anaerotignum},
\emph{Coprobacillus}, \emph{Enterocloster}, and \emph{Murimonas} were
increased in the mice that would experience mild disease. There were
many genera that did not have a correlation of relative abundance and
disease severity, such as the populations of \emph{Monoglobus} and
\emph{Hungatella}, which were decreased or increased, respectively, in
only the group of mice that would develop moderate disease. To
understand these cases better we used LEfSe to identify the genera
differentiating the communities that we detected toxin in from those
that we did not (Figure 4B). There were many genera the were associated
with toxin production as we would expect from their association with
severe disease, such as populations of \emph{Escherichia/Shigella} and
\emph{Bacteroides}. Likewise, there were genera such as
\emph{Anaerotignum}, \emph{Enterocloster}, and \emph{Murimonas} that
were higher in communities that had mild disease and no toxin was
produced. However, communities without a linear trend between relative
abundance and disease severity can be better understood by their
association with toxin production. Populations of \emph{Hungatella} were
increased in the group with moderate disease but not with the production
of toxin. Lastly, we looked for associations between the histopathologic
score and the populations of genera at that same time (Figure 4C). The
populatons of \emph{Bacteroides} at the end matched the trend observed
in the initial community, increased population correlated with increased
histopathologic score. Populations of \emph{Klebsiella} and
\emph{Prevotellaceae} positively correlated with clinic score and were
increased in the group that produced toxin. These tests have identified
which individual populations of bacteria that associated with each
disease severity.

We next wanted to understand how the populations in the context of each
other lead to the different disease severities. We trained random forest
models with the intital bacterial community relative abundance data at
each taxonomic rank to predict toxin production, severe disease or final
histopathologic score. Overall for predicting toxin production,
microbial populations aggregated by their phylum level classificiation
performed as well as models using lower taxonomic ranks (AUROC = 0.83,
Figure S2). In the initial community populations of
\emph{Verrucomicrobia}, \emph{Campilobacterota} and
\emph{Proteobacteria} contributed the most to the correct prediction of
toxin production (Figure 5A). \emph{C. difficile} was more likely to
produce toxin when the initial community had smaller populations of
\emph{Verrucomicrobia} and \emph{Campilobacterota} and had larger
popluations of \emph{Proteobacteria}. Next, we assessed the ability of
the community to predict severe disease. Class rank classification was
sufficient to predict if the mouse would succumb to the infection before
the end of the experiment (AUROC = 0.91, Figure S2). Groups of bacteria
belonging to \emph{Bacilli}, \emph{Firmicutes}, and
\emph{Erysipelotrichia} contributed the most to the correct prediction.
Larger populations of \emph{Bacilli} and \emph{Firmicutes} and reduced
populations of \emph{Erysipelotrichia} were more likely to result in
moribundity (Figure 5B). Aside from \emph{Erysipelotrichia}, the only
other population decreased in moribund mice was \emph{Clostridia}, all
other features had increased populations in the moribund mice. Lastly,
the initial community was able to predict if the mice would have a
histopathologic score above or below the median score (median
histopathologic score = 5) with the genera relative abundances (AUROC =
0.99, Figure S2). There were no genera with a significantly greater
effect on the model performance than any others, indicating the model is
reliant on many genera for the correct prediction. The model used some
of the genera identified in the LEfSe analysis, such as populations of
\emph{Coprobacillus}, \emph{Anaerostipes}, and \emph{Hungatella}.
Communities with greater populations of \emph{Hungatella},
\emph{Eggerthella}, \emph{Bifidobacterium}, \emph{Duncaniella} and
\emph{Neisseria} were more likely to have greater histopathologic
scores.

\textbf{Disease severity consistent by donor community across
strains/isolates.} We used a single \emph{C. difficile} isolate,
ribotype 027 isolate 431 to characterize the range of disease severity
and the features of the bacterial community associated with severity
(10). Since we had used the same mice and \emph{C. difficile} isolate in
our first set of experiemnts, we next wanted to test if the effect of
the community would apply to other \emph{C. difficile} isolates. We
selected three communities based on the result from our analysis thus
far to select one community we would expect to result in a low clinical
score (\textless{} 5), a high clinical score (\textgreater{} 5), and one
with a high clinical score which becomes moribund. Using genus level
data to select the communities, we selected communities based on the
relative abundance patterns of \emph{Sellimonas} and \emph{Anaerotignum}
(higher abundance in mice with a low clinical score),
\emph{Lachnospiracea} (lower abundance in moribund mice),
\emph{Hungatella} and \emph{Eisenbergiella} (higher abundance in
moderate mice and lower in moribund mice), Clostridium sensu stricto
(higher in moribund mice by both LEfSe analysis and RF model) and
\emph{Enterocloster} (lower in moribund mice and a negative linear trend
with disease severity by LEfSe analysis) (Figure S3). With this set of
human fecal communities selected, we inoculated mice via oral gavage of
the human fecal slurry and allowed two weeks for the a community to
establish. We then challenged the mice with a different \emph{C.
difficile} (10). The disease severity for this set of experiments
closely matched that in our first set of experiments with \emph{C.
difficile} isolate 431 (Figure 6).

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

Challenging mice colonized with different human fecal communities with a
single \emph{C. difficile} isolate allowed us to identify the effect of
microbiome variation on \emph{C. difficile} infection disease severity.
Our LEfSe analyses and random forest models revealed the relationships
between the bacterial community members and the degree of disease
severity. We used the relative abundance patterns of community members
with strong associations with disease severity to select communities to
test if the observed disease severity is conserved when other \emph{C.
difficile} isolates are used. The disease outcome from the \emph{C.
difficile} chalenge with the new isolate matched the expected disease
severities. Overall, these results have demonstrated that the bacterial
community is capable of modulating the disease severity of \emph{C.
difficile} infection.

CDI severity can range from no symptoms, mild symptoms such as diarrhea,
or severe symptoms such as toxic megacolon or death. Physicians use
prediction tools to identify patients most at risk of developing a
severe infection using white blood cell counts, serum albumin level, or
serum creatine level (14). Those levels are being driven by the
activities in the intestine (17). Research into the drivers of this
variation have revealed factors that make \emph{C. difficile} more
virulent. Strains are categorized for their virulence by the presence
and production toxin A and toxin B and the prevelance in outbreaks, such
as ribotype 027 and 078 (18). However, other studies have shown that
virulence is not necessarily linked with toxin production
\textbf{\textbf{???}} or the strain \textbf{\textbf{???}}. Furthermore,
there is variation in the genome, growth rate, sporulation, germination,
and toxin production in different isolates of a strain (10). This
variation may help explain why severe CDI prediction tools often miss
identifying many patients with CDI that will develop severe disease (1).
Therefore, it is necessary to gain a full understanding of all factors
contributing to disease variation to improve our ability to predict
severity.

The state of the intestinal bacterial community determines the ability
of \emph{C. difficile} to colonize, persist and even cause recurrent
infections. \emph{C. difficile} is unable to colonize an unperturbed
healthy gut community and is only able to entry after a perturbation
(24). Once colonized, the different communities lead to different
metabolic responses and dynamics of the \emph{C. difficile} population
(25). The gut bacteria metabolize primary bile acids into secondary bile
acids (27). The concentration of these bile acids affect the
germination, growth, toxin production and biofilm formation (5). Members
of the bacterial community also affect other metabolites \emph{C.
difficile} utilizes. \emph{Bacteroides thetaiotaomicron} produce
sialidases which release sialic acid from the mucosa for \emph{C.
difficile} to utilize (31). The nutrient environemt affects toxin
production (33). Thus, many of the actions of the gut bacteria modulate
\emph{C. difficile} in ways that would affect the disease severity
driven by CDI.

The gut community has not been demostrated to modulate CDI disease
severity. Myriad studies have explored the relationship between the
microbiome and CDI but none have ventured into the variation of disease
severity due to the microbiome. Most CDI studies employ either an
\emph{in vitro} or \emph{in vivo} model using a homogenous bacterial
community. Collins et al used multiple human communities to colonize
mice, however the communities were pooled prior to gavaging into
germ-free mice (34), resulting in a single community. Studies examining
difference in disease often use different \emph{C. difficile} strains or
ribotypes in mice with similar microbiota as a proxy for variation in
disease, such as strain 630 for non-severe and ribotype 027 for severe
(18). There has been studies demonstrating variation in severity, but
through tapering antibiotic dosage (26) or by reducing the amount of
cells or spores used for the challenge (22). Our group has recently been
uncovering how variation in the microbiome affects CDI but have been
limited to \emph{C. difficile} colonization (26).

With our recent observations that the initial community affects the
ability of \emph{C. difficile} to persist in an environment and the
existing research describing the myriad ways the microbiome interacts
with \emph{C. difficile}, we hypothesized that the microbiome directly
and indirectly modulates CDI disease severity. Our data demostrate gut
bacterial relative abundances associate with variation in toxin
production, histopathologic scoring of the cecum and severe disease.
This anaylsis has revealed populations of \emph{Akkermansia},
\emph{Anaerostipes}, \emph{Coprobacillus}, \emph{Enterocloster},
\emph{Lactonifactor}, and \emph{Monoglobus} were more abundant in the
microbiome of mice which little to no disease was detected. These genera
providing a protective role are supported by previous studies.
\emph{Coprobacillus}, \emph{Lactonifactor}, and \emph{Monoglobus} have
been shown to be involved in dietary fiber fermentation and associated
with healthy communities (37). \emph{Anaerostipes} and
\emph{Coprobacillus} produce short chain fatty acids which also have
been associated with healthy communities (41). Furthermore,
\emph{Coprobacillus}, which is abundant in mice with low histopathologic
scores but rare in all other mice, has been shwon to contain putative
type I lantibiotic gene cluster and inhibit \emph{C. difficile}
colonization (44). \emph{Akkermansia} and \emph{Enterocloster} were also
idenitified as more abundant in mice which had a low clinical disease
but have contridictory supporting evidence in the current literature. In
our data, \emph{Akkermansia} is most abundant in the mice with least
disease but there were some mice with severe disease which had increased
populations of \emph{Akkermansia}. This could be attributed to either a
more protective mucus layer had developed in the mice with little
disease and \emph{Akkermansia} is inhibiting colonization (46) or
\emph{Akkermansia} could be crossfeeding \emph{C. difficile} or exposing
a niche via its mucus consumption (48). Similarly with
\emph{Enterocloster}, in our data this genus was more abundant and
associated with low histopathologic scores. It has been associated with
healthy populations and has been used to mono-colonize germ-free mice to
reduce the ability of \emph{C. difficile} to colonize (51). However,
\emph{Enterocloster} has mostly been indentified in infections that
result in disease (53). These data demostrate a population of bacteria
has the potential to be either protective or harmful. The disease
outcome is not likely based on the abundance of a single population of
bacteria rather the result of the interactions of many populations.

The groups of bacteria that were associated with either a higher
histopathologic score or severe disease are members of the indigenous
gut community that also have been associated with disease. These are
often referred to as opportunitic pathogens, however based on the the
results reported in this study it is more appropriate to apply the
damage-response framework for microbial pathogenesis (55). The disease
is not driven by a single entity, rather it is an emergent property of
the responses of the host immune system, infecting microbe, and the
microbes of the indigenous community. Many of the populations with
pathgenic potential that associated with worse outcomes are also
facultative anaerobes. \emph{Enterococcus}, \emph{Klebsiella},
\emph{Shigella/Escherichia}, \emph{Staphylococcus}, and
\emph{Streptococcus} have been shown to expand their population with
antibiotic use (57) and are commonly detected in CDI cases (60). In
addition to these populations, another set of bacteria, populations of
\emph{Eggerthella}, \emph{Prevotellaceae} and \emph{Helicobacter},
associated with worse outcomes have also been associated with intestinal
inflammation (64). Recently, \emph{Helicobacter hepaticus} has been
demostrated to be sufficient to cause susceptibility to CDI in
IL-10\^{}(-/-) B57BL/6 mice (67). In our experiments, when
\emph{Helicobacter} was present, the mouse resulted in either severe
disease or a high histopathologic score. While we did not use
IL-10\^{}(-/-) mice, it is possible the bacterial communtity or host
response are similarly modified by \emph{Helicobacter} allowing \emph{C.
difficile} infection and resulting in host damage. However, the
populations identified other than \emph{Helicobacter} are not as clear
with their relationship between their relative abundance and the
resulting disease severity.

The damage-response framework for pathogenesis helps to understand how
bacterial populations identified in this study potentiate CDI disease
severity through the varied interactions with the host and \emph{C.
difficile}. In this set of experiments, we use the same host, C57BL/6
mice, the same infecting microbe, \emph{C. difficile} ribotype 027
isolate 431, with different gut bacterial communities. The members of
those communities were often present in multiple levels of disease
severity. Thus, it is not merely, the presence of the bacteria but their
activity in response to the other microbes and host which determine
their effect on host damage. Additionally, while we are using the same
host and infecting microbe, they too are reacting to the specific
members of the bacterial community. Disease severity is driven by the
cummulative effect of the host immune response and the activity of the
gut bacteria and \emph{C. difficile}. \emph{C. difficile} can be driving
host damage through the production of toxin. The gut microbiota could be
contributing to the host damage through the balance of metabolic and
competitive activities, such as bacteriocin production or mucin
degradation. Low levels of mucin degradation can provide nutrients to
other community members producing a diverse non-damaging community
{[}{]}. However, if mucin degradation becomes too great it reduces the
protective function of the mucin layer and exposes the epithelial cells
{[}{]}. This over-harvesting contributes to the host damage due to other
members producing toxin. Thus disease is a balance of many activities of
the community. Host damage is the emergent property of numerous
damage-response curves, one for host immune response, one for \emph{C.
difficile} activity and another for microbiome community activity, each
of which are a composite curve of the individual activities from each
group, such as antibody production, neutrophil infiltration, toxin
production, sporulation, fiber and mucin degradation. Therefore, while
we have identified populations of interest, it is likely targeting one
specifically will reduce disease severity.

Many studies have investigated individual activities contributing to
CDI. Here we have shown myraid bacterial groups and their relative
abundances associate with variation in CDI disease severity. We must
continue to explore how the host, microbiota, and \emph{C. difficile}
interactions result in severe disease. We can reduce the risk of
severity with a better understanding of which interactions, whether be
driven by specific community functions or groups of bacteria, reduce the
host damage. Approaching this problem from a damage-response framework
allows us to experiment with modifying subsets of activities to reduce
overall host damage. Our current clinical treatment targets an
individual group in this tripartite system. Most commonly, CDI is
treated with an antibiotic to eliminate \emph{C. difficile}.
Alternatively, another treatment uses antibodies to neutralize \emph{C.
difficile} toxins. These treatments are only addressing a single entity
of the system, leaving the potential for the others to maintain
responses that maintain host damage. However, we can continue this
research to elucidate treatments that drive all three entities towards
reduced host damage. When a patient is diagnosed with CDI, the gut
community composition, in addition to the traditionally obtained
information, may improve our severity prediction and appropriate
treatment. Treating the microbiome may contribute to reduced severity.
From our results, promoting fiber metabolizing bacteria and reducing
facultative anerobes could bolster the efficacy of current CDI
treatments, reducing disease severities.

\hypertarget{materials-and-methods}{%
\subsection{Materials and Methods}\label{materials-and-methods}}

\textbf{Animal care.} 5. 64 mice, 33F and 31M, ages 6-13 weeks old, 3-4
mice/group (except 2 for C/DA00581 and MOUT/DA10093) each run once
except for DA00581 which was run 4 times because we observed
colonization resistance in one of the experiments 15 human fecal
samples, selected to create a subset list with diverse responses to
health questionaire

\textbf{\emph{C. difficile} challenge.} M.

\textbf{Sample collection.} F.

\textbf{DNA sequencing.} T.

\textbf{Sequence curation.} S.

\textbf{Statistical analysis and modeling.} D.

\textbf{Code availability.} Scripts necessary to reproduce our analysis
and this paper are available in an online repository
(\url{https://github.com/SchlossLab/Lesniak_XXXX}).

\textbf{Sequence data accession number.} All 16S rRNA gene sequence data
and associated metadata are available through the Sequence Read Archive
via accession XXXX.

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

This work was supported by several grants from the National Institutes
for Health R01GM099514, U19AI090871, U01AI12455, and P30DK034933.
Additionally, NAL was supported by the Molecular Mechanisms of Microbial
Pathogenesis training grant (NIH T32 AI007528). The funding agencies had
no role in study design, data collection and analysis, decision to
publish, or preparation of the manuscript.

\newpage

\hypertarget{references}{%
\subsection{References}\label{references}}

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-Dieterle2020}{}%
1. \textbf{Dieterle MG}, \textbf{Putler R}, \textbf{Perry DA},
\textbf{Menon A}, \textbf{Abernathy-Close L}, \textbf{Perlman NS},
\textbf{Penkevich A}, \textbf{Standke A}, \textbf{Keidan M},
\textbf{Vendrov KC}, \textbf{Bergin IL}, \textbf{Young VB}, \textbf{Rao
K}. 2020. Systemic inflammatory mediators are effective biomarkers for
predicting adverse outcomes in \emph{Clostridioides difficile}
infection. mBio \textbf{11}.
doi:\href{https://doi.org/10.1128/mbio.00180-20}{10.1128/mbio.00180-20}.

\leavevmode\hypertarget{ref-Butt2013}{}%
2. \textbf{Butt E}, \textbf{Foster JA}, \textbf{Keedwell E},
\textbf{Bell JE}, \textbf{Titball RW}, \textbf{Bhangu A},
\textbf{Michell SL}, \textbf{Sheridan R}. 2013. Derivation and
validation of a simple, accurate and robust prediction rule for risk of
mortality in patients with \emph{Clostridium difficile} infection. BMC
Infectious Diseases \textbf{13}.
doi:\href{https://doi.org/10.1186/1471-2334-13-316}{10.1186/1471-2334-13-316}.

\leavevmode\hypertarget{ref-Perry2021}{}%
3. \textbf{Perry DA}, \textbf{Shirley D}, \textbf{Micic D},
\textbf{Patel CP}, \textbf{Putler R}, \textbf{Menon A}, \textbf{Young
VB}, \textbf{Rao K}. 2021. External validation and comparison of
\emph{Clostridioides difficile} severity scoring systems. Clinical
Infectious Diseases.
doi:\href{https://doi.org/10.1093/cid/ciab737}{10.1093/cid/ciab737}.

\leavevmode\hypertarget{ref-vanBeurden2017}{}%
4. \textbf{Beurden YH van}, \textbf{Hensgens MPM}, \textbf{Dekkers OM},
\textbf{Cessie SL}, \textbf{Mulder CJJ}, \textbf{Vandenbroucke-Grauls
CMJE}. 2017. External validation of three prediction tools for patients
at risk of a complicated course of \emph{Clostridium difficile}
infection: Disappointing in an outbreak setting. Infection Control \&
Hospital Epidemiology \textbf{38}:897--905.
doi:\href{https://doi.org/10.1017/ice.2017.89}{10.1017/ice.2017.89}.

\leavevmode\hypertarget{ref-Sorg2008}{}%
5. \textbf{Sorg JA}, \textbf{Sonenshein AL}. 2008. Bile salts and
glycine as cogerminants for \emph{Clostridium difficile} spores. Journal
of Bacteriology \textbf{190}:2505--2512.
doi:\href{https://doi.org/10.1128/jb.01765-07}{10.1128/jb.01765-07}.

\leavevmode\hypertarget{ref-Thanissery2017}{}%
6. \textbf{Thanissery R}, \textbf{Winston JA}, \textbf{Theriot CM}.
2017. Inhibition of spore germination, growth, and toxin activity of
clinically relevant \emph{C. difficile} strains by gut microbiota
derived secondary bile acids. Anaerobe \textbf{45}:86--100.
doi:\href{https://doi.org/10.1016/j.anaerobe.2017.03.004}{10.1016/j.anaerobe.2017.03.004}.

\leavevmode\hypertarget{ref-Aguirre2021}{}%
7. \textbf{Aguirre AM}, \textbf{Yalcinkaya N}, \textbf{Wu Q},
\textbf{Swennes A}, \textbf{Tessier ME}, \textbf{Roberts P},
\textbf{Miyajima F}, \textbf{Savidge T}, \textbf{Sorg JA}. 2021. Bile
acid-independent protection against \emph{Clostridioides difficile}
infection \textbf{17}:e1010015.
doi:\href{https://doi.org/10.1371/journal.ppat.1010015}{10.1371/journal.ppat.1010015}.

\leavevmode\hypertarget{ref-Kang2019}{}%
8. \textbf{Kang JD}, \textbf{Myers CJ}, \textbf{Harris SC},
\textbf{Kakiyama G}, \textbf{Lee I-K}, \textbf{Yun B-S},
\textbf{Matsuzaki K}, \textbf{Furukawa M}, \textbf{Min H-K},
\textbf{Bajaj JS}, \textbf{Zhou H}, \textbf{Hylemon PB}. 2019. Bile acid
7\(\alpha\)-dehydroxylating gut bacteria secrete antibiotics that
inhibit \emph{Clostridium difficile}: Role of secondary bile acids
\textbf{26}:27--34.e4.
doi:\href{https://doi.org/10.1016/j.chembiol.2018.10.003}{10.1016/j.chembiol.2018.10.003}.

\leavevmode\hypertarget{ref-Tomkovich2020}{}%
9. \textbf{Tomkovich S}, \textbf{Stough JMA}, \textbf{Bishop L},
\textbf{Schloss PD}. 2020. The initial gut microbiota and response to
antibiotic perturbation influence \emph{Clostridioides difficile}
clearance in mice. mSphere \textbf{5}.
doi:\href{https://doi.org/10.1128/msphere.00869-20}{10.1128/msphere.00869-20}.

\leavevmode\hypertarget{ref-Carlson2013}{}%
10. \textbf{Carlson PE}, \textbf{Walk ST}, \textbf{Bourgis AET},
\textbf{Liu MW}, \textbf{Kopliku F}, \textbf{Lo E}, \textbf{Young VB},
\textbf{Aronoff DM}, \textbf{Hanna PC}. 2013. The relationship between
phenotype, ribotype, and clinical disease in human \emph{Clostridium
difficile} isolates. Anaerobe \textbf{24}:109--116.
doi:\href{https://doi.org/10.1016/j.anaerobe.2013.04.003}{10.1016/j.anaerobe.2013.04.003}.

\leavevmode\hypertarget{ref-Akerlund2006}{}%
11. 2006. Correlation of disease severity with fecal toxin levels in
patients with \emph{Clostridium difficile}-associated diarrhea and
distribution of PCR ribotypes and toxin yields in vitro of corresponding
isolates. Journal of Clinical Microbiology \textbf{44}:353--358.
doi:\href{https://doi.org/10.1128/jcm.44.2.353-358.2006}{10.1128/jcm.44.2.353-358.2006}.

\leavevmode\hypertarget{ref-Vitucci2020}{}%
12. \textbf{Vitucci JC}, \textbf{Pulse M}, \textbf{Tabor-Simecka L},
\textbf{Simecka J}. 2020. Epidemic ribotypes of \emph{clostridium} (now
\emph{clostridioides}) \emph{difficile} are likely to be more virulent
than non-epidemic ribotypes in animal models. BMC Microbiology
\textbf{20}.
doi:\href{https://doi.org/10.1186/s12866-020-1710-5}{10.1186/s12866-020-1710-5}.

\leavevmode\hypertarget{ref-Cowardin2016}{}%
13. \textbf{Cowardin CA}, \textbf{Buonomo EL}, \textbf{Saleh MM},
\textbf{Wilson MG}, \textbf{Burgess SL}, \textbf{Kuehne SA},
\textbf{Schwan C}, \textbf{Eichhoff AM}, \textbf{Koch-Nolte F},
\textbf{Lyras D}, \textbf{Aktories K}, \textbf{Minton NP}, \textbf{Petri
WA}. 2016. The binary toxin CDT enhances \emph{Clostridium difficile}
virulence by suppressing protective colonic eosinophilia. Nature
Microbiology \textbf{1}.
doi:\href{https://doi.org/10.1038/nmicrobiol.2016.108}{10.1038/nmicrobiol.2016.108}.

\leavevmode\hypertarget{ref-Lungulescu2011}{}%
14. \textbf{Lungulescu OA}, \textbf{Cao W}, \textbf{Gatskevich E},
\textbf{Tlhabano L}, \textbf{Stratidis JG}. 2011. CSI: A severity index
for \emph{Clostridium difficile} infection at the time of admission.
Journal of Hospital Infection \textbf{79}:151--154.
doi:\href{https://doi.org/10.1016/j.jhin.2011.04.017}{10.1016/j.jhin.2011.04.017}.

\leavevmode\hypertarget{ref-McDonald2018}{}%
15. \textbf{McDonald LC}, \textbf{Gerding DN}, \textbf{Johnson S},
\textbf{Bakken JS}, \textbf{Carroll KC}, \textbf{Coffin SE},
\textbf{Dubberke ER}, \textbf{Garey KW}, \textbf{Gould CV},
\textbf{Kelly C}, \textbf{Loo V}, \textbf{Sammons JS}, \textbf{Sandora
TJ}, \textbf{Wilcox MH}. 2018. Clinical practice guidelines for
\emph{Clostridium difficile} infection in adults and children: 2017
update by the infectious diseases society of america (IDSA) and society
for healthcare epidemiology of america (SHEA). Clinical Infectious
Diseases \textbf{66}:e1--e48.
doi:\href{https://doi.org/10.1093/cid/cix1085}{10.1093/cid/cix1085}.

\leavevmode\hypertarget{ref-Zar2007}{}%
16. \textbf{Zar FA}, \textbf{Bakkanagari SR}, \textbf{Moorthi KMLST},
\textbf{Davis MB}. 2007. A comparison of vancomycin and metronidazole
for the treatment of \emph{Clostridium difficile}-associated diarrhea,
stratified by disease severity. Clinical Infectious Diseases
\textbf{45}:302--307.
doi:\href{https://doi.org/10.1086/519265}{10.1086/519265}.

\leavevmode\hypertarget{ref-diMasi2018}{}%
17. \textbf{Masi A di}, \textbf{Leboffe L}, \textbf{Polticelli F},
\textbf{Tonon F}, \textbf{Zennaro C}, \textbf{Caterino M}, \textbf{Stano
P}, \textbf{Fischer S}, \textbf{Hägele M}, \textbf{Müller M},
\textbf{Kleger A}, \textbf{Papatheodorou P}, \textbf{Nocca G},
\textbf{Arcovito A}, \textbf{Gori A}, \textbf{Ruoppolo M}, \textbf{Barth
H}, \textbf{Petrosillo N}, \textbf{Ascenzi P}, \textbf{Bella SD}. 2018.
Human serum albumin is an essential component of the host defense
mechanism against \emph{Clostridium difficile} intoxication. The Journal
of Infectious Diseases \textbf{218}:1424--1435.
doi:\href{https://doi.org/10.1093/infdis/jiy338}{10.1093/infdis/jiy338}.

\leavevmode\hypertarget{ref-AbernathyClose2020}{}%
18. \textbf{Abernathy-Close L}, \textbf{Dieterle MG}, \textbf{Vendrov
KC}, \textbf{Bergin IL}, \textbf{Rao K}, \textbf{Young VB}. 2020. Aging
dampens the intestinal innate immune response during severe
\emph{Clostridioides difficile} infection and is associated with altered
cytokine levels and granulocyte mobilization. Infection and Immunity
\textbf{88}.
doi:\href{https://doi.org/10.1128/iai.00960-19}{10.1128/iai.00960-19}.

\leavevmode\hypertarget{ref-Theriot2011}{}%
19. \textbf{Theriot CM}, \textbf{Koumpouras CC}, \textbf{Carlson PE},
\textbf{Bergin II}, \textbf{Aronoff DM}, \textbf{Young VB}. 2011.
Cefoperazone-treated mice as an experimental platform to assess
differential virulence of \emph{Clostridium difficile} strains. Gut
Microbes \textbf{2}:326--334.
doi:\href{https://doi.org/10.4161/gmic.19142}{10.4161/gmic.19142}.

\leavevmode\hypertarget{ref-Goorhuis2008}{}%
20. \textbf{Goorhuis A}, \textbf{Bakker D}, \textbf{Corver J},
\textbf{Debast SB}, \textbf{Harmanus C}, \textbf{Notermans DW},
\textbf{Bergwerff AA}, \textbf{Dekker FW}, \textbf{Kuijper EJ}. 2008.
Emergence of \emph{Clostridium difficile} infection due to a new
hypervirulent strain, polymerase chain reaction ribotype 078. Clinical
Infectious Diseases \textbf{47}:1162--1170.
doi:\href{https://doi.org/10.1086/592257}{10.1086/592257}.

\leavevmode\hypertarget{ref-OConnor2009}{}%
21. \textbf{O'Connor JR}, \textbf{Johnson S}, \textbf{Gerding DN}. 2009.
\emph{Clostridium difficile} infection caused by the epidemic
BI/NAP1/027 strain. Gastroenterology \textbf{136}:1913--1924.
doi:\href{https://doi.org/10.1053/j.gastro.2009.02.073}{10.1053/j.gastro.2009.02.073}.

\leavevmode\hypertarget{ref-Chen2008}{}%
22. \textbf{Chen X}, \textbf{Katchar K}, \textbf{Goldsmith JD},
\textbf{Nanthakumar N}, \textbf{Cheknis A}, \textbf{Gerding DN},
\textbf{Kelly CP}. 2008. A mouse model of \emph{Clostridium
difficile}-associated disease. Gastroenterology \textbf{135}:1984--1992.
doi:\href{https://doi.org/10.1053/j.gastro.2008.09.002}{10.1053/j.gastro.2008.09.002}.

\leavevmode\hypertarget{ref-He2010}{}%
23. \textbf{He M}, \textbf{Sebaihia M}, \textbf{Lawley TD},
\textbf{Stabler RA}, \textbf{Dawson LF}, \textbf{Martin MJ},
\textbf{Holt KE}, \textbf{Seth-Smith HMB}, \textbf{Quail MA},
\textbf{Rance R}, \textbf{Brooks K}, \textbf{Churcher C}, \textbf{Harris
D}, \textbf{Bentley SD}, \textbf{Burrows C}, \textbf{Clark L},
\textbf{Corton C}, \textbf{Murray V}, \textbf{Rose G}, \textbf{Thurston
S}, \textbf{Tonder A van}, \textbf{Walker D}, \textbf{Wren BW},
\textbf{Dougan G}, \textbf{Parkhill J}. 2010. Evolutionary dynamics of
\emph{Clostridium difficile} over short and long time scales.
Proceedings of the National Academy of Sciences \textbf{107}:7527--7532.
doi:\href{https://doi.org/10.1073/pnas.0914322107}{10.1073/pnas.0914322107}.

\leavevmode\hypertarget{ref-Schubert2015}{}%
24. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
Antibiotic-induced alterations of the murine gut microbiota and
subsequent effects on colonization resistance against \emph{Clostridium
difficile}. mBio \textbf{6}.
doi:\href{https://doi.org/10.1128/mbio.00974-15}{10.1128/mbio.00974-15}.

\leavevmode\hypertarget{ref-Jenior2017}{}%
25. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
\textbf{Schloss PD}. 2017. \emph{Clostridium difficile} colonizes
alternative nutrient niches during infection across distinct murine gut
microbiomes. mSystems \textbf{2}.
doi:\href{https://doi.org/10.1128/msystems.00063-17}{10.1128/msystems.00063-17}.

\leavevmode\hypertarget{ref-Lesniak2021}{}%
26. \textbf{Lesniak NA}, \textbf{Schubert AM}, \textbf{Sinani H},
\textbf{Schloss PD}. 2021. Clearance of \emph{Clostridioides difficile}
colonization is associated with antibiotic-specific bacterial changes.
mSphere \textbf{6}.
doi:\href{https://doi.org/10.1128/msphere.01238-20}{10.1128/msphere.01238-20}.

\leavevmode\hypertarget{ref-Staley2016}{}%
27. \textbf{Staley C}, \textbf{Weingarden AR}, \textbf{Khoruts A},
\textbf{Sadowsky MJ}. 2016. Interaction of gut microbiota with bile acid
metabolism and its influence on disease states. Applied Microbiology and
Biotechnology \textbf{101}:47--64.
doi:\href{https://doi.org/10.1007/s00253-016-8006-6}{10.1007/s00253-016-8006-6}.

\leavevmode\hypertarget{ref-Long2017}{}%
28. \textbf{Long SL}, \textbf{Gahan CGM}, \textbf{Joyce SA}. 2017.
Interactions between gut bacteria and bile in health and disease.
Molecular Aspects of Medicine \textbf{56}:54--65.
doi:\href{https://doi.org/10.1016/j.mam.2017.06.002}{10.1016/j.mam.2017.06.002}.

\leavevmode\hypertarget{ref-Sorg2010}{}%
29. \textbf{Sorg JA}, \textbf{Sonenshein AL}. 2010. Inhibiting the
initiation of \emph{Clostridium difficile} spore germination using
analogs of chenodeoxycholic acid, a bile acid. Journal of Bacteriology
\textbf{192}:4983--4990.
doi:\href{https://doi.org/10.1128/jb.00610-10}{10.1128/jb.00610-10}.

\leavevmode\hypertarget{ref-Dubois2019}{}%
30. \textbf{Dubois T}, \textbf{Tremblay YDN}, \textbf{Hamiot A},
\textbf{Martin-Verstraete I}, \textbf{Deschamps J}, \textbf{Monot M},
\textbf{Briandet R}, \textbf{Dupuy B}. 2019. A microbiota-generated bile
salt induces biofilm formation in \emph{Clostridium difficile}. npj
Biofilms and Microbiomes \textbf{5}.
doi:\href{https://doi.org/10.1038/s41522-019-0087-4}{10.1038/s41522-019-0087-4}.

\leavevmode\hypertarget{ref-Ng2013}{}%
31. \textbf{Ng KM}, \textbf{Ferreyra JA}, \textbf{Higginbottom SK},
\textbf{Lynch JB}, \textbf{Kashyap PC}, \textbf{Gopinath S},
\textbf{Naidu N}, \textbf{Choudhury B}, \textbf{Weimer BC},
\textbf{Monack DM}, \textbf{Sonnenburg JL}. 2013. Microbiota-liberated
host sugars facilitate post-antibiotic expansion of enteric pathogens.
Nature \textbf{502}:96--99.
doi:\href{https://doi.org/10.1038/nature12503}{10.1038/nature12503}.

\leavevmode\hypertarget{ref-Ferreyra2014}{}%
32. \textbf{Ferreyra JA}, \textbf{Wu KJ}, \textbf{Hryckowian AJ},
\textbf{Bouley DM}, \textbf{Weimer BC}, \textbf{Sonnenburg JL}. 2014.
Gut microbiota-produced succinate promotes \emph{C. difficile} infection
after antibiotic treatment or motility disturbance. Cell Host \& Microbe
\textbf{16}:770--777.
doi:\href{https://doi.org/10.1016/j.chom.2014.11.003}{10.1016/j.chom.2014.11.003}.

\leavevmode\hypertarget{ref-MartinVerstraete2016}{}%
33. \textbf{Martin-Verstraete I}, \textbf{Peltier J}, \textbf{Dupuy B}.
2016. The regulatory networks that control \emph{Clostridium difficile}
toxin synthesis. Toxins \textbf{8}:153.
doi:\href{https://doi.org/10.3390/toxins8050153}{10.3390/toxins8050153}.

\leavevmode\hypertarget{ref-Collins2015}{}%
34. \textbf{Collins J}, \textbf{Auchtung JM}, \textbf{Schaefer L},
\textbf{Eaton KA}, \textbf{Britton RA}. 2015. Humanized microbiota mice
as a model of recurrent \emph{Clostridium difficile} disease. Microbiome
\textbf{3}.
doi:\href{https://doi.org/10.1186/s40168-015-0097-2}{10.1186/s40168-015-0097-2}.

\leavevmode\hypertarget{ref-Lawley2012}{}%
35. \textbf{Lawley TD}, \textbf{Clare S}, \textbf{Walker AW},
\textbf{Stares MD}, \textbf{Connor TR}, \textbf{Raisen C},
\textbf{Goulding D}, \textbf{Rad R}, \textbf{Schreiber F},
\textbf{Brandt C}, \textbf{Deakin LJ}, \textbf{Pickard DJ},
\textbf{Duncan SH}, \textbf{Flint HJ}, \textbf{Clark TG},
\textbf{Parkhill J}, \textbf{Dougan G}. 2012. Targeted restoration of
the intestinal microbiota with a simple, defined bacteriotherapy
resolves relapsing \emph{Clostridium difficile} disease in mice. PLoS
Pathogens \textbf{8}:e1002995.
doi:\href{https://doi.org/10.1371/journal.ppat.1002995}{10.1371/journal.ppat.1002995}.

\leavevmode\hypertarget{ref-Reeves2011}{}%
36. \textbf{Reeves AE}, \textbf{Theriot CM}, \textbf{Bergin IL},
\textbf{Huffnagle GB}, \textbf{Schloss PD}, \textbf{Young VB}. 2011. The
interplay between microbiome dynamics and pathogen dynamics in a murine
model of \emph{Clostridium difficile} infection. Gut Microbes
\textbf{2}:145--158.
doi:\href{https://doi.org/10.4161/gmic.2.3.16333}{10.4161/gmic.2.3.16333}.

\leavevmode\hypertarget{ref-Mabrok2011}{}%
37. \textbf{Mabrok HB}, \textbf{Klopfleisch R}, \textbf{Ghanem KZ},
\textbf{Clavel T}, \textbf{Blaut M}, \textbf{Loh G}. 2011. Lignan
transformation by gut bacteria lowers tumor burden in a gnotobiotic rat
model of breast cancer. Carcinogenesis \textbf{33}:203--208.
doi:\href{https://doi.org/10.1093/carcin/bgr256}{10.1093/carcin/bgr256}.

\leavevmode\hypertarget{ref-Kim2019}{}%
38. \textbf{Kim CC}, \textbf{Healey GR}, \textbf{Kelly WJ},
\textbf{Patchett ML}, \textbf{Jordens Z}, \textbf{Tannock GW},
\textbf{Sims IM}, \textbf{Bell TJ}, \textbf{Hedderley D},
\textbf{Henrissat B}, \textbf{Rosendale DI}. 2019. Genomic insights from
\emph{Monoglobus pectinilyticus}: A pectin-degrading specialist
bacterium in the human colon. The ISME Journal \textbf{13}:1437--1456.
doi:\href{https://doi.org/10.1038/s41396-019-0363-6}{10.1038/s41396-019-0363-6}.

\leavevmode\hypertarget{ref-doPrado2021}{}%
39. \textbf{Prado SBR do}, \textbf{Minguzzi BT}, \textbf{Hoffmann C},
\textbf{Fabi JP}. 2021. Modulation of human gut microbiota by dietary
fibers from unripe and ripe papayas: Distinct polysaccharide degradation
using a colonic in vitro fermentation model. Food Chemistry
\textbf{348}:129071.
doi:\href{https://doi.org/10.1016/j.foodchem.2021.129071}{10.1016/j.foodchem.2021.129071}.

\leavevmode\hypertarget{ref-Muthuramalingam2019}{}%
40. \textbf{Muthuramalingam K}, \textbf{Singh V}, \textbf{Choi C},
\textbf{Choi SI}, \textbf{Kim YM}, \textbf{Unno T}, \textbf{Cho M}.
2019. Dietary intervention using (1, 3)/(1, 6)-\(\beta\)-glucan, a
fungus-derived soluble prebiotic ameliorates high-fat diet-induced
metabolic distress and alters beneficially the gut microbiota in mice
model. European Journal of Nutrition \textbf{59}:2617--2629.
doi:\href{https://doi.org/10.1007/s00394-019-02110-5}{10.1007/s00394-019-02110-5}.

\leavevmode\hypertarget{ref-Han2019}{}%
41. \textbf{Han S-H}, \textbf{Yi J}, \textbf{Kim J-H}, \textbf{Lee S},
\textbf{Moon H-W}. 2019. Composition of gut microbiota in patients with
toxigenic \emph{clostridioides} (\emph{clostridium}) \emph{difficile}:
Comparison between subgroups according to clinical criteria and toxin
gene load. PLOS ONE \textbf{14}:e0212626.
doi:\href{https://doi.org/10.1371/journal.pone.0212626}{10.1371/journal.pone.0212626}.

\leavevmode\hypertarget{ref-Duncan2004}{}%
42. \textbf{Duncan SH}, \textbf{Louis P}, \textbf{Flint HJ}. 2004.
Lactate-utilizing bacteria, isolated from human feces, that produce
butyrate as a major fermentation product. Applied and Environmental
Microbiology \textbf{70}:5810--5817.
doi:\href{https://doi.org/10.1128/aem.70.10.5810-5817.2004}{10.1128/aem.70.10.5810-5817.2004}.

\leavevmode\hypertarget{ref-Ye2018}{}%
43. \textbf{Ye J}, \textbf{Lv L}, \textbf{Wu W}, \textbf{Li Y},
\textbf{Shi D}, \textbf{Fang D}, \textbf{Guo F}, \textbf{Jiang H},
\textbf{Yan R}, \textbf{Ye W}, \textbf{Li L}. 2018. Butyrate protects
mice against methionineCholine-deficient diet-induced non-alcoholic
steatohepatitis by improving gut barrier function, attenuating
inflammation and reducing endotoxin levels. Frontiers in Microbiology
\textbf{9}.
doi:\href{https://doi.org/10.3389/fmicb.2018.01967}{10.3389/fmicb.2018.01967}.

\leavevmode\hypertarget{ref-Walsh2017}{}%
44. \textbf{Walsh CJ}, \textbf{Guinane CM}, \textbf{O'Toole PW},
\textbf{Cotter PD}. 2017. A profile hidden markov model to investigate
the distribution and frequency of LanB-encoding lantibiotic modification
genes in the human oral and gut microbiome. PeerJ \textbf{5}:e3254.
doi:\href{https://doi.org/10.7717/peerj.3254}{10.7717/peerj.3254}.

\leavevmode\hypertarget{ref-Sandiford2018}{}%
45. \textbf{Sandiford SK}. 2018. Current developments in lantibiotic
discovery for treating \emph{Clostridium difficile} infection. Expert
Opinion on Drug Discovery \textbf{14}:71--79.
doi:\href{https://doi.org/10.1080/17460441.2019.1549032}{10.1080/17460441.2019.1549032}.

\leavevmode\hypertarget{ref-Stein2013}{}%
46. \textbf{Stein RR}, \textbf{Bucci V}, \textbf{Toussaint NC},
\textbf{Buffie CG}, \textbf{Rätsch G}, \textbf{Pamer EG}, \textbf{Sander
C}, \textbf{Xavier JB}. 2013. Ecological modeling from time-series
inference: Insight into dynamics and stability of intestinal microbiota.
PLoS Computational Biology \textbf{9}:e1003388.
doi:\href{https://doi.org/10.1371/journal.pcbi.1003388}{10.1371/journal.pcbi.1003388}.

\leavevmode\hypertarget{ref-Nakashima2021}{}%
47. \textbf{Nakashima T}, \textbf{Fujii K}, \textbf{Seki T},
\textbf{Aoyama M}, \textbf{Azuma A}, \textbf{Kawasome H}. 2021. Novel
gut microbiota modulator, which markedly increases \emph{Akkermansia
muciniphila} occupancy, ameliorates experimental colitis in rats.
Digestive Diseases and Sciences.
doi:\href{https://doi.org/10.1007/s10620-021-07131-x}{10.1007/s10620-021-07131-x}.

\leavevmode\hypertarget{ref-Geerlings2018}{}%
48. \textbf{Geerlings S}, \textbf{Kostopoulos I}, \textbf{Vos W de},
\textbf{Belzer C}. 2018. \emph{Akkermansia muciniphila} in the human
gastrointestinal tract: When, where, and how? Microorganisms
\textbf{6}:75.
doi:\href{https://doi.org/10.3390/microorganisms6030075}{10.3390/microorganisms6030075}.

\leavevmode\hypertarget{ref-Deng2018}{}%
49. \textbf{Deng H}, \textbf{Yang S}, \textbf{Zhang Y}, \textbf{Qian K},
\textbf{Zhang Z}, \textbf{Liu Y}, \textbf{Wang Y}, \textbf{Bai Y},
\textbf{Fan H}, \textbf{Zhao X}, \textbf{Zhi F}. 2018. \emph{Bacteroides
fragilis} prevents \emph{Clostridium difficile} infection in a mouse
model by restoring gut barrier and microbiome regulation. Frontiers in
Microbiology \textbf{9}.
doi:\href{https://doi.org/10.3389/fmicb.2018.02976}{10.3389/fmicb.2018.02976}.

\leavevmode\hypertarget{ref-Engevik2020}{}%
50. \textbf{Engevik MA}, \textbf{Engevik AC}, \textbf{Engevik KA},
\textbf{Auchtung JM}, \textbf{Chang-Graham AL}, \textbf{Ruan W},
\textbf{Luna RA}, \textbf{Hyser JM}, \textbf{Spinler JK},
\textbf{Versalovic J}. 2020. Mucin-degrading microbes release
monosaccharides that chemoattract \emph{Clostridioides difficile} and
facilitate colonization of the human intestinal mucus layer. ACS
Infectious Diseases \textbf{7}:1126--1142.
doi:\href{https://doi.org/10.1021/acsinfecdis.0c00634}{10.1021/acsinfecdis.0c00634}.

\leavevmode\hypertarget{ref-Reeves2012}{}%
51. \textbf{Reeves AE}, \textbf{Koenigsknecht MJ}, \textbf{Bergin IL},
\textbf{Young VB}. 2012. Suppression of \emph{Clostridium difficile} in
the gastrointestinal tracts of germfree mice inoculated with a murine
isolate from the family \emph{lachnospiraceae}. Infection and Immunity
\textbf{80}:3786--3794.
doi:\href{https://doi.org/10.1128/iai.00647-12}{10.1128/iai.00647-12}.

\leavevmode\hypertarget{ref-Ma2021}{}%
52. \textbf{Ma L}, \textbf{Keng J}, \textbf{Cheng M}, \textbf{Pan H},
\textbf{Feng B}, \textbf{Hu Y}, \textbf{Feng T}, \textbf{Yang F}. 2021.
Gut microbiome and serum metabolome alterations associated with isolated
dystonia. mSphere \textbf{6}.
doi:\href{https://doi.org/10.1128/msphere.00283-21}{10.1128/msphere.00283-21}.

\leavevmode\hypertarget{ref-Haas2020}{}%
53. \textbf{Haas KN}, \textbf{Blanchard JL}. 2020. Reclassification of
the \emph{Clostridium clostridioforme} and \emph{Clostridium sphenoides}
clades as \emph{enterocloster} gen. Nov. And \emph{lacrimispora} gen.
Nov., including reclassification of 15 taxa. International Journal of
Systematic and Evolutionary Microbiology \textbf{70}:23--34.
doi:\href{https://doi.org/10.1099/ijsem.0.003698}{10.1099/ijsem.0.003698}.

\leavevmode\hypertarget{ref-Finegold2005}{}%
54. \textbf{Finegold SM}, \textbf{Song Y}, \textbf{Liu C}, \textbf{Hecht
DW}, \textbf{Summanen P}, \textbf{Könönen E}, \textbf{Allen SD}. 2005.
\emph{Clostridium clostridioforme}: A mixture of three clinically
important species. European Journal of Clinical Microbiology \&
Infectious Diseases \textbf{24}:319--324.
doi:\href{https://doi.org/10.1007/s10096-005-1334-6}{10.1007/s10096-005-1334-6}.

\leavevmode\hypertarget{ref-Pirofski2008}{}%
55. \textbf{Pirofski L-a}, \textbf{Casadevall A}. 2008. The
damage-response framework of microbial pathogenesis and infectious
diseases, pp. 135--146. \emph{In}. Springer New York.

\leavevmode\hypertarget{ref-Casadevall2014}{}%
56. \textbf{Casadevall A}, \textbf{Pirofski L-a}. 2014. What is a host?
Incorporating the microbiota into the damage-response framework
\textbf{83}:2--7.
doi:\href{https://doi.org/10.1128/iai.02627-14}{10.1128/iai.02627-14}.

\leavevmode\hypertarget{ref-VanInsberghe2020}{}%
57. \textbf{VanInsberghe D}, \textbf{Elsherbini JA}, \textbf{Varian B},
\textbf{Poutahidis T}, \textbf{Erdman S}, \textbf{Polz MF}. 2020.
Diarrhoeal events can trigger long-term \emph{Clostridium difficile}
colonization with recurrent blooms \textbf{5}:642--650.
doi:\href{https://doi.org/10.1038/s41564-020-0668-2}{10.1038/s41564-020-0668-2}.

\leavevmode\hypertarget{ref-GarzaGonzalez2019}{}%
58. \textbf{Garza-González E}, \textbf{Mendoza-Olazarán S},
\textbf{Morfin-Otero R}, \textbf{Ramírez-Fontes A},
\textbf{Rodríguez-Zulueta P}, \textbf{Flores-Treviño S},
\textbf{Bocanegra-Ibarias P}, \textbf{Maldonado-Garza H},
\textbf{Camacho-Ortiz A}. 2019. Intestinal microbiome changes in fecal
microbiota transplant (FMT) vs. FMT enriched with \emph{lactobacillus}
in the treatment of recurrent \emph{Clostridioides difficile} infection
\textbf{2019}:1--7.
doi:\href{https://doi.org/10.1155/2019/4549298}{10.1155/2019/4549298}.

\leavevmode\hypertarget{ref-Schubert2014}{}%
59. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
\textbf{Mogle J}, \textbf{Petrosino JP}, \textbf{Young VB},
\textbf{Aronoff DM}, \textbf{Schloss PD}. 2014. Microbiome data
distinguish patients with \emph{Clostridium difficile} infection and
non-\emph{C. difficile}-associated diarrhea from healthy controls
\textbf{5}.
doi:\href{https://doi.org/10.1128/mbio.01021-14}{10.1128/mbio.01021-14}.

\leavevmode\hypertarget{ref-Shafiq2020}{}%
60. \textbf{Shafiq M}, \textbf{Alturkmani H}, \textbf{Zafar Y},
\textbf{Mittal V}, \textbf{Lodhi H}, \textbf{Ullah W}, \textbf{Brewer
J}. 2020. Effects of co-infection on the clinical outcomes of
\emph{Clostridium difficile} infection \textbf{12}.
doi:\href{https://doi.org/10.1186/s13099-020-00348-7}{10.1186/s13099-020-00348-7}.

\leavevmode\hypertarget{ref-Keith2020}{}%
61. \textbf{Keith JW}, \textbf{Dong Q}, \textbf{Sorbara MT},
\textbf{Becattini S}, \textbf{Sia JK}, \textbf{Gjonbalaj M},
\textbf{Seok R}, \textbf{Leiner IM}, \textbf{Littmann ER}, \textbf{Pamer
EG}. 2020. Impact of antibiotic-resistant bacteria on immune activation
and \emph{Clostridioides difficile} infection in the mouse intestine
\textbf{88}.
doi:\href{https://doi.org/10.1128/iai.00362-19}{10.1128/iai.00362-19}.

\leavevmode\hypertarget{ref-Zackular2016}{}%
62. \textbf{Zackular JP}, \textbf{Moore JL}, \textbf{Jordan AT},
\textbf{Juttukonda LJ}, \textbf{Noto MJ}, \textbf{Nicholson MR},
\textbf{Crews JD}, \textbf{Semler MW}, \textbf{Zhang Y}, \textbf{Ware
LB}, \textbf{Washington MK}, \textbf{Chazin WJ}, \textbf{Caprioli RM},
\textbf{Skaar EP}. 2016. Dietary zinc alters the microbiota and
decreases resistance to \emph{Clostridium difficile} infection
\textbf{22}:1330--1334.
doi:\href{https://doi.org/10.1038/nm.4174}{10.1038/nm.4174}.

\leavevmode\hypertarget{ref-Berkell2021}{}%
63. \textbf{Berkell M}, \textbf{Mysara M}, \textbf{Xavier BB},
\textbf{Werkhoven CH van}, \textbf{Monsieurs P}, \textbf{Lammens C},
\textbf{Ducher A}, \textbf{Vehreschild MJGT}, \textbf{Goossens H},
\textbf{Gunzburg J de}, \textbf{Bonten MJM}, \textbf{Malhotra-Kumar S}.
2021. Microbiota-based markers predictive of development of
\emph{Clostridioides difficile} infection \textbf{12}.
doi:\href{https://doi.org/10.1038/s41467-021-22302-0}{10.1038/s41467-021-22302-0}.

\leavevmode\hypertarget{ref-Gardiner2014}{}%
64. \textbf{Gardiner BJ}, \textbf{Tai AY}, \textbf{Kotsanas D},
\textbf{Francis MJ}, \textbf{Roberts SA}, \textbf{Ballard SA},
\textbf{Junckerstorff RK}, \textbf{Korman TM}. 2014. Clinical and
microbiological characteristics of \emph{Eggerthella lenta} bacteremia
\textbf{53}:626--635.
doi:\href{https://doi.org/10.1128/jcm.02926-14}{10.1128/jcm.02926-14}.

\leavevmode\hypertarget{ref-Iljazovic2020}{}%
65. \textbf{Iljazovic A}, \textbf{Roy U}, \textbf{Gálvez EJC},
\textbf{Lesker TR}, \textbf{Zhao B}, \textbf{Gronow A}, \textbf{Amend
L}, \textbf{Will SE}, \textbf{Hofmann JD}, \textbf{Pils MC},
\textbf{Schmidt-Hohagen K}, \textbf{Neumann-Schaal M}, \textbf{Strowig
T}. 2020. Perturbation of the gut microbiome by \emph{prevotella} spp.
Enhances host susceptibility to mucosal inflammation
\textbf{14}:113--124.
doi:\href{https://doi.org/10.1038/s41385-020-0296-4}{10.1038/s41385-020-0296-4}.

\leavevmode\hypertarget{ref-Nagalingam2013}{}%
66. \textbf{Nagalingam NA}, \textbf{Robinson CJ}, \textbf{Bergin IL},
\textbf{Eaton KA}, \textbf{Huffnagle GB}, \textbf{Young VB}. 2013. The
effects of intestinal microbial community structure on disease
manifestation in IL-\(10^{-/-}\) mice infected with helicobacter
hepaticus \textbf{1}.
doi:\href{https://doi.org/10.1186/2049-2618-1-15}{10.1186/2049-2618-1-15}.

\leavevmode\hypertarget{ref-AbernathyClose2021}{}%
67. \textbf{Abernathy-Close L}, \textbf{Barron MR}, \textbf{George JM},
\textbf{Dieterle MG}, \textbf{Vendrov KC}, \textbf{Bergin IL},
\textbf{Young VB}. 2021. Intestinal inflammation and altered gut
microbiota associated with inflammatory bowel disease render mice
susceptible to \emph{Clostridioides difficile} colonization and
infection.
doi:\href{https://doi.org/10.1128/mbio.02733-20}{10.1128/mbio.02733-20}.
\end{cslreferences}

\newpage

\includegraphics{../results/figures/figure_1.jpg}

\textbf{Figure 1. R.} (A-C).

\hfill\break

\includegraphics{../results/figures/figure_2.jpg}

\textbf{Figure 2. M.} F.

\hfill\break

\includegraphics{../results/figures/figure_3.jpg}

\textbf{Figure 3. O.} F.

\hfill\break

\includegraphics{../results/figures/figure_4.jpg}

\textbf{Figure 4. E.} A.

\hfill\break

\includegraphics{../results/figures/figure_5.jpg}

\textbf{Figure 5. D.} (A) L

\hfill\break

\includegraphics{../results/figures/figure_6.jpg}

\textbf{Figure 6. D.} (A) L

\hfill\break

\includegraphics{../results/figures/figure_S1.jpg}

\textbf{Figure S1. D.} (A) L

\hfill\break

\includegraphics{../results/figures/figure_S2.jpg}

\textbf{Figure S2. D.} (A) L

\hfill\break

\includegraphics{../results/figures/figure_S3.jpg}

\textbf{Figure S3. D.} (A) L

\hfill\break

\textbf{Table 1. D.} (A) L

samples\_used \textless-
readxl::read\_xlsx(`data/raw/humanGF\_ids.xlsx') \%\textgreater\%
pull(human\_source) \%\textgreater\% unique donor\_data \textless-
readxl::read\_xlsx(`data/raw/MIMARKS\_cdclinical.xlsx') \%\textgreater\%
filter(sample\_id \%in\% samples\_used) \%\textgreater\%
select(sample\_id, biome, age, gender, ``antibiotics \textgreater3mo'',
protonpump, antacid, Healthworker, historyCdiff, Surgery6mos,
Vegetarian, weight, disease\_stat)

\textbf{Table 2. D.} (A) L

read.csv(here(`data/raw/Hanna\_in\_vitro\_data\_plus\_germination.txt'))
\%\textgreater\% rownames\_to\_column(`isolate') \%\textgreater\%
filter(isolate \%in\% c(`DA00299', `DA00395', `DA00458', `DA00431'))

\end{document}

### 2021-11-30 submission/manuscript.Rmd
revisions based on ST and SL feedback

@@
 
-# The gut microbiota potentiates *Clostridiodes difficile* infection severity.
+# The gut bacterial community potentiates *Clostridiodes difficile* infection severity.
 
@@
 
-Nicholas A. Lesniak$^1$, Alyxandria M. Schubert$^1$, Kaitlyn Flynn$^1$, Hamide Sinani$^1$, Kathyrn A. Eaton$^1$, Ingrid L. Bergin$^2$, Patrick D. Schloss$^{1,\dagger}$
+Nicholas A. Lesniak$^1$, Alyxandria M. Schubert$^1$, Kaitlyn Flynn$^1$, Jhansi L Leslie$^1$, Hamide Sinani$^1$, Ingrid L. Bergin$^3$, Vincent B Young$^2$, Patrick D. Schloss$^{1,\dagger}$
 
@@
 1\. Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI  
-2\. Unit for Laboratory Animal Medicine, University of Michigan, Ann Arbor, MI
+2\. Division of Infectious Diseases, Department of Internal Medicine, University of Michigan Medical School, Ann Arbor, MI
+3\. Unit for Laboratory Animal Medicine, University of Michigan, Ann Arbor, MI
 
@@
 
-*Clostridioides difficile* infection has become more common and severe in the last few decades.
+*Clostridioides difficile* infection (CDI) has become more common and severe in the last few decades.
 Patient age, white blood cell count, creatinine levels as well as *C.
 difficile* ribotype and the presence of toxin genes have been associated with disease severity.
-However, models based on those data to identify patients with CDI that will develop severe disease have not been robust enough to be broadly applied.
-These models are input from our understanding of *C.
-difficile* interactions with the gut environment.
-The gut bacteria are a key determinant in *C.
-difficile* infection.
-CDI is dependent on perturbations in the bacterial community in order to colonize the gut.
+However, current models of CDI disease severity that incorporate these factors are not robust enough to be broadly applied.
+These models are missing input from the gut environment.
+CDI is dependent on perturbations of the gut bacterial community in order to colonize the gut.
 The gut microbiota also impairs *C.
 difficile* colonization through bile acid metabolism, nutrient consumption and bacteriocin production.
-However, it is unclear if the gut bacteria affect the disease severity resulting from that colonization.
-Here we have shown gut bacterial communities contribute to disease severity variation.
-We created diverse gut communities by colonizing germ-free mice with human fecal communities.
-They were then infected with *C.
-difficile* which resulted in differed disease severities.
+However, it is unclear if the gut bacteria affect CDI disease severity.
+Here we demonstrate that gut bacterial communities contribute to disease severity variation.
+We derived diverse gut communities by colonizing germ-free mice with human fecal communities.
+The mice were then infected with a single *C.
+difficile* RT027 isolate which resulted in mortality and histopathologic differences.
 The severity grouped by the human fecal community they received.
-Generally, facultative anaerobes with pathogenic potential, such as Escherichia, Helicobacter  and Klebsiella, were associated with more severe disease.
-Bacterial groups associated with dietary fiber degradation, such as Coprobacillus, were associated with reduced disease severity.
+Generally, facultative anaerobes with pathogenic potential, such as *Escherichia*, *Helicobacter*,  and *Klebsiella*, were associated with more severe disease.
+Bacterial groups associated with dietary fiber degradation, such as *Coprobacillus*, were associated with reduced disease severity.
 Lastly, we showed communities that resulted in either low or high histopathologic scores or severe disease had the same outcome when infected with a different *C.
@@
 
-Clostridioides difficile infection (CDI) causes a range of disease from asymptomatic to mild diarrhea to severe outcomes of recurrent infections and even death.
-Models that predict severity and treatment decisions are based on clinical factors and *C.
-difficile* capabilities.
-But currently the gut microbiome, the primary protector against CDI, has not been accounted for its effect on CDI disease severity.
-We demonstrated variation in the microbiome of mice colonized with human feces was sufficient to result in a range of disease severities.
-Our results show groups of bacteria contribute to the development of disease in *C.
+*Clostridioides difficile* infection (CDI) can be asymptomatic, or range in severity from mild diarrhea to severe outcomes of surgical interventions and even death.
+Models that predict severity and guide treatment decisions are based on clinical factors and *C.
+difficile* characteristics.
+The gut microbiome plays a role in potecting against CDI, though has not been investigated for its effect on CDI disease severity.
+We demonstrate that variation in the microbiome of mice colonized with human feces was sufficient to result in a range of disease severities.
+The results reveal groups of bacteria associated with both severe and more mild *C.
 difficile* infections.
-Gut bacterial community data from patients with CDI could improve our ability to identify patients more at risk for severe disease and improve interventions which treat the gut bacteria to reduce host damage.
+Gut bacterial community data from patients with CDI could improve our ability to identify patients at risk for severe disease and improve interventions which treat the gut bacteria to reduce host damage.
 
@@
 *Clostridioides difficile* infections (CDI) have been increasing in incidence and severity since *C.
-difficile* was first identified as a pathogen.
-It was first identified in 1931 in a healthy fecal community [CITATION].
-But then in the late 1970s,  following the introduction of broad-spectrum antibiotics, *C.
-difficile* was identified as the cause of antibiotic-associated psuedomembranous colitis [CITATION].
-Beginning in the early 2000s, CDI became more problematic with increasing incidences and severity [CITATION].
-Disease severities can range from asymptomatic to mild diarrhea to toxic megacolon and death.
-This severity has been associated with characteristics of *C.
-difficile* ribotypes and host factors.
-Over time more pathogenic ribotypes have emerged, such as ribotype 027 in the early 2000s.
-The presence of the genes for Toxin A/B or fluoroquinolon resistence has been associated with more severe disease [CITATION].
-Binary toxin, while alone insufficient to cause damage on it own, when in the presence of Toxin A/B has been associated with increased virulence [CITATION].
-Mutations in tcd C, the negative regulator of of toxin transcription, also increased the virulence of *C.
-difficile* [CITATION].
-In addition to the pathogenic potential of the infecting strain, the response by the host also affects the disease severity.
-Delayed or reduced production of antibodies IgG and IgA can result in more severe disease [CITATION].
-Increased neutrophil inflitration can lead to worse outcomes [CITATION].
-In addition to the immune response, increased density of the toxin receptor on the host cells can increase the severity of the infection [CITATION].
-While there is a thorough understanding of many of the mechanisms driving in CDI incidence and disease severity, there still is not effective tools to reduce the risk of severe disease.
+difficile* was first identified as the cause of antibiotic-associated psuedomembranous colitis [@Kelly2008].
+Disease severity can range from asymptomatic to mild diarrhea to toxic megacolon and death.
+Treatment for CDI is driven by the disease severity.
+The Infectious Diseases Society of America (IDSA) and Society for Healthcare Epidemiology of America (SHEA) guidelines categorize severe disease based on white blood cell count >15,000 cells/mm^3 and/or a serum creatinine >1.5 mg/dL, and complicated disease based on systolic blood pressure <90, ileus or toxic megacolon [@McDonald2018].
+Since these measures are only present once the disease has advanced, the IDSA classification does not allow for identification of patients at risk of severe CDI.
+Thus, we have limited ability to prevent patients from developing severe CDI.
+Ideally, we could identify patients that are high-risk of severe disease use that to adjust their treatments to halt the disease before it worsens.
 
-Models have been developed to identify patients at risk for severe CDI but have not been robust for broad application [@Dieterle2020, @Butt2013, @Perry2021, @vanBeurden2017].
-These models utilize patient related data, such as patient age, white blood cell count, serum albumin levels, and creatinine levels, to predict disease severity.
-In their initial publication, these models perform well enough for publication.
-However when Perry et al applied many of the most current models to a multi-center external validation, their performance suffered and often had more false positives than true positives [@Perry2021].
-Thus, there seems to be a missing factor contributing to disease severity.
+Prediction models have been developed to score a patients risk for severe CDI outcomes but have not been robust for broad application [@Dieterle2020; @Butt2013; @Perry2021; @vanBeurden2017].
+Firstly, there is no widely accepted definition of severe disease [@Ressler2021].
+While the IDSA guideline is frequently used, many studies develop their own severity definitions.
+Aside from the features utilized by the IDSA guidelines, other severity measures use patient demographics, vital information, laboratory test results, disease complications, or infection outcomes such as ICU admission, surgery and/or death [@Dieterle2020; @Zar2007; @Gujja2009; @Jardin2013].
+However, a recent multi-center external validation of these scoring systems using a consistent severity definition - transfer or admission to the intensive care unit, colectomy, and/or death within 30 days of diagnosis - demonstrated a decline in performance and high false positivity rate [@Perry2021].
+Thus, it may be necessary to incorporate information in addition to patient and clinical factors to accurately idenitfy patients at risk of severe CDI.
 
-Missing from these models are the gut bacteria which also interact with *C.
-difficile* during the course of the infection.
+CDI disease severity is the result of damage to the host.
+Severity has been associated with characteristics of *C.
+difficile* ribotypes, such as 027, and host factors [@Bauer2011].
+Ribotypes with genes for toxin TcdA/TcdB have the potential to disrupt the tight junctions between the intestinal epithelial cells [@Abt2016].
+The presence of binary toxin CDT or mutations in *tcdC*, the negative regulator of toxin transcription, have been associated with increased virulence of TcdA/TcdB [@Abt2016].
+In addition to the pathogenic potential of *C.
+difficile*, the response by the host also affects the disease severity.
+IgG antibody activity towards the TcdA/TcdB are decreased in both patients with symptomatic and recurrent infections [@Kelly2011].
+In addition to controlling the infection, the immune response can also cause intestinal injury.
+Increased neutrophil inflitration into the mucosa after the cellular tight junctions have been compromised can exacerbate the inflammatory response [@Farooq2015].
+While we have have elucidated these major drivers of host damage in CDI, these do not explain the observed variation in disease severity.
+
+Missing from our CDI disease model are the effects of gut bacteria.
+*C.
+difficile* interacts with the intestinal community in many ways.
 The indigenous bacteria of a healthy intestinal community provide a protective barrier preventing *C.
 difficile* from infecting the gut.
-Only after this community is disrupted from a range of pertubations, such as many classes of antibiotics, medications, or even dietary changes, can *C.
-difficile* infect the intestine.
-Once established, the gut bacteria can either promote or inhibit *C.
-difficile* through producing molecules or modifying the environment [Abbas2020].
+A range of mechanisms can disrupt this barrier, including antibiotics, medications, or dietary changes, and lead to increased susceptiblity to CDI [@Britton2014; @Hryckowian2018; @VichVila2020].
+Once in the gut, the indigenous bacteria can either promote or inhibit *C.
+difficile* through producing molecules or modifying the environment [@Abbas2020].
 Bile acids metabolized by the gut bacteria can inhibit *C.
-difficile* growth and affect toxin production [@Sorg2008, @Thanissery2017].
+difficile* growth and affect toxin production [@Sorg2008; @Thanissery2017].
 Bacteria in the gut also can compete more directly with *C.
-difficile* through antibiotic production or nutrient consumption [@Aguirre2021, @Kang2019].
-To eliminate CDI, the primary treatment is a course of antibiotics.
-Patients who become reinfected and recalitrant to antibiotics, a fecal microbiota transplantion often is sufficient to restore the gut community to eliminate *C.
-difficile*.
-While there has been many studies to investigate the interaction between the gut bacteria and *C.
-difficile* during the infection, it has not been demostrated the bacterial community can modulate the disease severity of CDI.
-Recently, Tomkovich et al showed the variation in the bacterial communtiies between mice from different colonies and vendors was sufficient to cause differences in the temporal dynamics of the *C.
-difficile* infection [@Tomkovich2020].
-Since the gut bacteria interact with *C.
-difficile* throughout the infection and community difference resulted in different colonization dynamics, we hypothesized that the gut bacteria contribute to the variation in CDI disease severity.
-Here, we colonized germ-free C57BL/6 mice with human fecal communities to test if the variablity in the gut bacteria can produce and explain variation in the disease severity caused by a single *C.
-difficile* isolate.
+difficile* through antibiotic production or nutrient consumption [@Aguirre2021; @Kang2019].
+While the interaction between the gut bacteria and *C.
+difficile* is established, the effect the gut bacteria on CDI disease severity is unclear.
+Recent studies that have begun to uncover this relationship but have either used clinical data which does not permit control of the host or *C.
+difficile* variation [@Seekatz2016] or murine experiments which lack variation in the bacterial communities [@Dieterle2020].
+Recently, Tomkovich et al.
+showed the variation in the bacterial communities between mice from different mouse colonies and vendors resulted in different clearance rates of *C.
+difficile* [@Tomkovich2020].
+Therefore it is clear the gut bacteria modulate the activity and infection dynmaics of *C.
+difficile*, we hypothesized that gut bacteria also contribute to the variation in CDI disease severity.
+Here, we colonized germ-free C57BL/6 mice with human fecal communities to test if the variablity in the gut bacteria can produce and explain variation in the CDI disease severity.
 
@@
 
-**Germ-free mice colonized with human fecal commmunities have diverse and unique community structures.** Based on our previous observation mice with variation in their microbiota showed differences in colonization of *C.
-difficile* [@Tomkovich2020], we sought to explore the effect communties with greater variation had on *C.
-difficile* challenge.
-To produce communities with greater variation than mouse colonies with their indigenous communties, we used human fecal communities to establish a community in the intestine of germ-free mice.
-We inoculated a cage of germ-free C57BL/6 mice (2-4 mice per cage) with homogenized feces from one of 15 different human donors via oral gavage.
-The fecal community was allowed to colonize and stabilize for two weeks post inoculation [CITATION?].
-We then assessed the gut bacterial communities by sequencing the V4 region of the 16S rRNA gene extracted from fecal pellets (Figure 1A).
-The communities established in the mice group most closely to other mice recieving the same human fecal donor community than to either their respective donor or any mice who received stool from a different donor (Figure 1B).
+***C.
+difficile* is able to infect germ-free mice colonized with human fecal microbial communities without perturbation.**  Based on our previous observation that variation in the murine microbiota associated with different *C.
+difficile* colonization dynamics [@Tomkovich2020], we sought to explore how variation in murine microbial communities affects the outcome of *C.
+difficile* infection.
+To produce communities with greater variation than mouse colonies with their indigenous communties, we used human fecal communities to establish a microbial community in the intestine of germ-free mice [@Nagpal2018].
+We inoculated germ-free C57BL/6 mice with homogenized feces from one of 15 human fecal samples via oral gavage.
+The human fecal samples selected were chosen to provide diverse communtiy structures and based on community clusters [@Schubert2014].
+The gut community was allowed to establish itself for two weeks post inoculation [@Gillilland2012].
+We then surveyed the bacterial members of the gut communities by 16S rRNA gene sequencing of fecal pellets (Figure 1A).
+The bacterial communities from each mouse grouped more closely to those communities from mice that received the same human fecal donor community than to either their respective donor community or any mice who received a fecal homogenate from a different donor (Figure 1B).
 Most of the communities were primairily composed of populations of *Clostridia*, *Bacteroidia*, *Erysipelotrichia*, *Bacilli*, and *Gammaproteobacteria*.
-However, each group of mice harbored unique combination of class abunadances in their gut bacterial communities.
-*Any alpha? compare to murine variation? Number of unique otus?*
+However, each group of mice harbored unique combination of class abundances in their gut bacterial communities.
 
-***C.
-difficile* is able to colonize mice without perturbation.** Our goal was to test the effect of the variation in the microbiota on *C.
+Next, we tested this set of mice with their human-derived gut microbial communities for susceptibility to *C.
 difficile* infection.
 A typical mouse model of CDI requires pretreatment with antibiotics such as clindamycin to become susceptible to *C.
-difficile* colonization [CITATION].
-However, we wanted to avoid modifying the communties with an antibiotic to maintain their community structures and differences.
-Since these communties came from human donors which may have become susceptible through their own exposures, we decided to test if *C.
-difficile* was able to colonize these communities without any perturbation.
-After two weeks post incoluation with human fecal communities, the mice were then challenged with 10^**4?** *C.
-difficile* ribotype 027 isolate 431 spores (**Add schematic?**) [@Carlson2013].
-The mice were followed for 10 days post challenge and their stool was collected and plated for *C.
-difficile* CFU to detect colonization level.
+difficile* colonization [@Chen2008; @Schubert2015].
+However, we wanted to avoid modifying the communties with an antibiotic to maintain their unique microbial compositions and ecological relationships.
+Since these communties came from people who have increased risk of CDI, such as hospitalization or recent antibiotic use [@Schubert2014], or the communities became susceptible through the murine gut colonization process, we decided to test if *C.
+difficile* was able to infect these mice without any antibiotic perturbation.
 We hypothesized that *C.
-difficile* would most likely be able to colonize the mice with communities that came from donors which had a history of antibiotic use, one of the primary risk factors for developing CDI (TABLE, CITATION).
+difficile* would be able to colonize a subset of mice whose communities were susceptible from perturbations either the donor experienced or from the ecological selection colonizing the murine gut.
+After two weeks post incoluation with human fecal communities, the mice were then challenged with $10^{3}$ *C.
+difficile* ribotype 027 human isolate 431 spores.
+The mice were followed for 10 days post-challenge and their stool was collected and plated for *C.
+difficile* colony forming units (CFU) to detect infection level.
 Suprisingly, communities from all donors were colonized (Figure 2).
-The only two mice that were able to resist *C.
-difficile* colonization both recieved their community from the same donor.
-Therefore, the transplanted human fecal communties were susceptible to *C.
-difficile* colonization without the need for any perturbation.
+Only two mice were able to resist *C.
+difficile* infection and both received their community from the same donor, Donor C.
+The remaining mice demonstrated susceptiblity to *C.
+difficile* infection without the need for any perturbation.
+Thus, *C.
+difficile* infected the mice without needing to perturb the established human-derived gut microbial communities, allowing them to maintain their unique community structures.
 
-**Infection seveity varies by initial community structure.** After challenging the mice with *C.
-difficile*, we looked at the effect of the community variation on the outcome from the *C.
-difficile* infection.
-We followed the mice for signs of disease as well as *C.
-difficile* colonization and toxin production for 10 days post challenge.
-Some mice were not initally colonized on the day after *C.
-difficile* challenge but became colonized with a few days post challenge (**SUPP FIG?**).
-A subset of mice, all that recieved their comunity from one of 6 donors, suffered from severe disease, which became highly colonized one day post challenge and moribund within 3 days post challenge.
-The remaining mice, except for one cage, maintained *C.
-difficile* colonization through the end of the experiment (Figure 2).
-After the  mice were humanely euthanised at the endpoint, day 10 post challenge or earlier if they became moribund, their cecal tissue was collected and scored for histological signs of disease and fecal samples were assayed for toxin (Figure 3).
+**Infection severity varies by initial community.** After challenging the mice with *C.
+difficile*, we investigated the outcome from the *C.
+difficile* infection and its relationship to the initial community.
+We followed the mice for 10 days post-challenge for *C.
+difficile* CFU, toxin production, and mortality.
+Seven mice, from Donors A, B and E, were not initally colonized on the day after *C.
+difficile* challenge but were infected ($>10^{6}$) by the end of the experiment.
+All mice that received their community from Donor J through O suffered from severe disease, which became highly colonized one day post-challenge and moribund within 3 days post-challenge.
+The remaining mice, except the Donor C mice which were not infected, maintained *C.
+difficile* infection through the end of the experiment (Figure 2).
+At 10 days post-challenge, or earlier for the moribund mice, mice were euthanised and cecal tissue was collected and scored for histopathologic signs of disease and fecal pellets were assayed for toxin activity (Figure 3).
 Overall, there was greater toxin activity detected in the stool of the moribund mice (*P* = 0.003).
-However, when looking at each group of mice, we saw there was a range in toxin activity for both the mice with non-severe disease and severe disease (Figure 3A).
-Some donor groups that maintained colonization without severe disease had similarlly high activity levels as was detected in the mice with severe disease.
-Additionally, some mice with severe disease did not have toxin activity detected in their stool.
-These oberserved variations in toxin activity and relation to disease severity have been previously characterized [@Akerlund2006, @Vitucci2020].
-Next we examined the cecal tissue for histopathological damage.
-For mice with severe disease, we observed high levels of epithelial damage, tissue edema and inflammation (Figure SXXXX), similar to previously reported histopathologic findings for *C.
-difficile* ribotype 027 [CITATION, @Cowardin2016].
-As observed with toxin activity, the mice with severe disease had higher histopathologic scores than the mice with non-severe disease (*P* = 3.0e-9).
-However, unlike the toxin activity, all moribund mice had consistently high histopathologic scores.
-The mice which maintained persistent population size, we saw a range in tissue damage from no signs of disease to nearly the same level as the mice with severe disease, which grouped by community donor.
+However, when looking at each group of mice, we saw there was a range in toxin activity for both the moribund and non-moribund mice (Figure 3A).
+Non-moribund mice from Donors D through I had comparable toxin activity as the moribund mice.
+Additionally, not all moribund mice had toxin activity detected in their stool.
+This discordance between toxin activity and disease severity has been previously described [@Akerlund2006; @Vitucci2020].
+Next we examined the cecal tissue for histopathologic damage.
+For moribund mice, we observed high levels of epithelial damage, tissue edema and inflammation (Figure S1), similar to previously reported histopathologic findings for *C.
+difficile* ribotype 027 [@Cowardin2016].
+As observed with toxin activity, the moribund mice had higher histopathologic scores than the mice with non-severe disease (*P* = 3.0e-9).
+However, unlike the toxin activity, all moribund mice had consistently high histopathologic summary scores (Figure 3B).
+For Donor groups A through I, where all mice were non-moribund, we saw a range in tissue damage from no signs of disease to nearly the same level as the mice with severe disease, which grouped by community donor.
 Together, the toxin activity, histopathologic score, and moribundity showed variation across the donors but were largely consistent within each group of mice recieving the same human donor fecal community.
 
-**Populations of the microbial communities explain variation in CDI disease severity.** Since the disease charateristics grouped by human donor fecal community, we next investigated the community composition for its ability to explain the variation in disease severity.
-First, we tested for associations between taxonomic groups and the outcomes of the *C.
-difficile* infection.
-We used the linear discriminant analysis (LDA) effect size (LEfSe) method to identify differences in the genera of the initial bacterial communities between the severity level and toxin production.
-We split the mice into groups by severity level based on their if they became moribund (severe), or if their clinical score was above 5 (moderate) or below 5 (mild).
+**Microbial community members explain variation in CDI severity.** Since the disease charateristics grouped by human donor fecal community, we next investigated specific bacterial populations that could explain the variation in disease severity.
+To this aim, we interogated the day 0 bacterial community for their relationship to our CDI severity measurements using linear discriminant analysis (LDA) effect size (LEfSe) analysis.
+We split the mice into groups by severity level based on their moribundity and histopathologic score.
+We dichotomized the histopathologic scores into high and low bins by splitting on the median score.
 This analysis revealed 20 genera that were significantly different by the disease severity (Figure 4A).
-The presence of populations of *Turicibacter*, *Streptococcus*, *Staphylococcus*, *Pseudomonas*, *Phocaeicola*, *Parabacteroides*, *Bacteroides*, and *Escherichia/Shigella* were detected at higher levels in the mice that developed severe disease.
-Populations of *Anaerotignum*, *Coprobacillus*, *Enterocloster*, and *Murimonas* were increased in the mice that would experience mild disease.
-There were many genera that did not have a correlation of relative abundance and disease severity, such as the populations of *Monoglobus* and *Hungatella*, which were decreased or increased, respectively, in only the group of mice that would develop moderate disease.
-To understand these cases better we used LEfSe to identify the genera differentiating the communities that we detected toxin in from those that we did not (Figure 4B).
-There were many genera the were associated with toxin production as we would expect from their association with severe disease, such as populations of *Escherichia/Shigella* and *Bacteroides*.
-Likewise, there were genera such as *Anaerotignum*, *Enterocloster*, and *Murimonas* that were higher in communities that had mild disease and no toxin was produced.
-However, communities without a linear trend between relative abundance and disease severity can be better understood by their association with toxin production.
-Populations of *Hungatella* were increased in the group with moderate disease but not with the production of toxin.
-Lastly, we looked for associations between the histopathologic score and the populations of genera at that same time (Figure 4C).
-The populatons of *Bacteroides* at the end matched the trend observed in the initial community, increased population correlated with increased histopathologic score.
-Populations of *Klebsiella* and *Prevotellaceae* positively correlated with clinic score and were increased in the group that produced toxin.
-These tests have identified which individual populations of bacteria that associated with each disease severity.
+Bacterial genera *Turicibacter*, *Streptococcus*, *Staphylococcus*, *Pseudomonas*, *Phocaeicola*, *Parabacteroides*, *Bacteroides*, and *Escherichia/Shigella* were detected at higher relative abundances in the mice that became moribund.
+Populations of *Anaerotignum*, *Coprobacillus*, *Enterocloster*, and *Murimonas* were more abundant in the mice that would develop only low intestinal injury.
+To understand the role of toxin activity in disease severity, we applied LEfSe to identify the genera most likely to explain the observed differences in toxin activity (Figure 4B).
+Many genera the were associated with toxin production as we would expect from their association with moribundity, such as populations of *Escherichia/Shigella* and *Bacteroides*.
+Likewise, there were genera such as *Anaerotignum*, *Enterocloster*, and *Murimonas* that exhibited greater relative abundance in communities from non-moribund mice with a low histopathologic score and no detected toxin.
+Overall, communities without genera that demonstrate a linear trend between relative abundance and disease severity can be better understood by their association with toxin activity.
+Lastly, we tested for correlations between the histopathologic score and the endpoint relative abundances of bacterial operational taxonomic units (OTUs) (Figure 4C).
+The endpoint relative abundance of *Bacteroides* was positively correlated with histopathologic score, as its day 0 relative abundance did with disease severity (Figure 4A).
+Populations of *Klebsiella* and *Prevotellaceae* were positively correlated with the histopathologic score and were increased in the group of mice with detectable toxin.
+Our analysis have revealed bacterial taxa that are associated with three different measures of CDI severity in our model: moribundity, histopathologic score, and toxin activity.
 
-We next wanted to understand how the populations in the context of each other lead to the different disease severities.
-We trained random forest models with the intital bacterial community relative abundance data at each taxonomic rank to predict toxin production, severe disease or final histopathologic score.
+We next wanted to determine whether collectively, bacterial community membership and relative abundance could be predictive of CDI severity.
+We trained random forest models with the day 0 bacterial community relative abundance data at each taxonomic rank to predict toxin production, severe disease or final histopathologic score.
 Overall for predicting toxin production, microbial populations aggregated by their phylum level classificiation performed as well as models using lower taxonomic ranks (AUROC = 0.83, Figure S2).
-In the initial community populations of *Verrucomicrobia*, *Campilobacterota* and *Proteobacteria* contributed the most to the correct prediction of toxin production (Figure 5A).
+Collectively, members of *Verrucomicrobia*, *Campilobacterota* and *Proteobacteria* contributed the most to the correct prediction of toxin production (Figure 5A).
 *C.
-difficile* was more likely to produce toxin when the initial community had smaller populations of *Verrucomicrobia* and *Campilobacterota* and had larger popluations of *Proteobacteria*.
-Next, we assessed the ability of the community to predict severe disease.
+difficile* was more likely to produce toxin when the community it infected had less abundant populations of *Verrucomicrobia* and *Campilobacterota* and had more abundant populations of *Proteobacteria*.
+Next, we assessed the ability of the community to predict moribundity.
 Class rank classification was sufficient to predict if the mouse would succumb to the infection before the end of the experiment (AUROC = 0.91, Figure S2).
-Groups of bacteria belonging to *Bacilli*, *Firmicutes*, and *Erysipelotrichia* contributed the most to the correct prediction.
+Bacterial taxa belonging to *Bacilli*, *Firmicutes*, and *Erysipelotrichia* hasd the greatest effect on model accuracy.
 Larger populations of *Bacilli* and *Firmicutes* and reduced populations of *Erysipelotrichia* were more likely to result in moribundity (Figure 5B).
-Aside from *Erysipelotrichia*, the only other population decreased in moribund mice was *Clostridia*, all other features had increased populations in the moribund mice.
-Lastly, the initial community was able to predict if the mice would have a histopathologic score above or below the median score (median histopathologic score = 5) with the genera relative abundances (AUROC = 0.99, Figure S2).
-There were no genera with a significantly greater effect on the model performance than any others, indicating the model is reliant on many genera for the correct prediction.
-The model used some of the genera identified in the LEfSe analysis, such as populations of *Coprobacillus*, *Anaerostipes*, and *Hungatella*.
-Communities with greater populations of *Hungatella*, *Eggerthella*, *Bifidobacterium*, *Duncaniella* and *Neisseria* were more likely to have greater histopathologic scores.
+The only other class that was decreased in moribund mice was *Clostridia*.
+Lastly, the relative abundances of genera were able to predict a high or low histopathologic score (histopathologic score dichotomized as in previous analysis, AUROC = 0.99, Figure S2).
+No genera had a significantly greater effect on the model performance than any others, indicating the model is reliant on many genera for the correct prediction.
+The model used some of the genera identified in the LEfSe analysis, such as *Coprobacillus*, *Anaerostipes*, and *Hungatella*.
+Communities with greater abundances of *Hungatella*, *Eggerthella*, *Bifidobacterium*, *Duncaniella* and *Neisseria* were more likely to have high histopathologic scores.
 
-**Disease severity consistent by donor community across strains/isolates.** We used a single *C.
-difficile* isolate, ribotype 027 isolate 431 to characterize the range of disease severity and the features of the bacterial community associated with severity [@Carlson2013].
-Since we had used the same mice and *C.
-difficile* isolate in our first set of experiemnts, we next wanted to test if the effect of the community would apply to other *C.
+**Disease severity consistent by donor community across isolates.** Since Carlson et al.
+showed clinical disease severity varied by *C.
+difficile* isolate characteristics [@Carlson2013; @Carlson2015], we next wanted to test if the effect of the community on CDI severity is conserved across *C.
 difficile* isolates.
-We selected three communities based on the result from our analysis thus far to select one community we would expect to result in a low clinical score (< 5), a high clinical score (> 5), and one with a high clinical score which becomes moribund.
-Using genus level data to select the communities, we selected communities based on the relative abundance patterns of *Sellimonas* and *Anaerotignum* (higher abundance in mice with a low clinical score), *Lachnospiracea* (lower abundance in moribund mice), *Hungatella* and *Eisenbergiella* (higher abundance in moderate mice and lower in moribund mice), Clostridium sensu stricto (higher in moribund mice by both LEfSe analysis and RF model) and *Enterocloster* (lower in moribund mice and a negative linear trend with disease severity by LEfSe analysis) (Figure S3).
-With this set of human fecal communities selected, we inoculated mice via oral gavage of the human fecal slurry and allowed two weeks for the a community to establish.
+To facilitate this analysis, we selected a subset of donor fecal samples based on their associated mouse transplant communities that fit the following criteria: one community we would expect to result in a low histopathologic score (< 5), a high histopathologic score (> 5), and a high histopathologic score which becomes moribund.
+Additionally, we selected the donors based their relative abundances of genera identified through the our analysis (Figure S3).
+We inoculated mice via oral gavage of the human fecal slurry of either Donor F, G, or M and allowed two weeks for the a community to establish.
 We then challenged the mice with a different *C.
-difficile* [@Carlson2013].
-The disease severity for this set of experiments closely matched that in our first set of experiments with *C.
-difficile* isolate 431 (Figure 6).
+difficile* isolate and all mice were infected.
+The moribundity and histopathologic score outcomes from challenge with the second *C.
+difficile* human isolate were comparable to the outcomes of challenges with *C.
+difficile* ribotype 027 human isolate 431 by their donor group (Figure 6).
 
@@
 Our LEfSe analyses and random forest models revealed the relationships between the bacterial community members and the degree of disease severity.
-We used the relative abundance patterns of community members with strong associations with disease severity to select communities to test if the observed disease severity is conserved when other *C.
-difficile* isolates are used.
+We used the relative abundance patterns of community members with strong associations with disease severity to select communities to test if the observed disease severity is conserved when a second *C.
+difficile* isolate is used.
 The disease outcome from the *C.
 difficile* chalenge with the new isolate matched the expected disease severities.
-Overall, these results have demonstrated that the bacterial community is capable of modulating the disease severity of *C.
+Therefore, the effect of the community variation on disease severity *in vivo* is greater than the effect of *C.
+difficile* isolate phenotype variation [@Carlson2013; @Carlson2015].
+Overall, these results have demonstrated that the bacterial community contributes to the disease severity of *C.
 difficile* infection.
@@
 CDI severity can range from no symptoms, mild symptoms such as diarrhea, or severe symptoms such as toxic megacolon or death.
-Physicians use prediction tools to identify patients most at risk of developing a severe infection using white blood cell counts, serum albumin level, or serum creatine level [@Lungulescu2011, @McDonald2018, @Zar2007].
+Physicians use prediction tools to identify patients most at risk of developing a severe infection using white blood cell counts, serum albumin level, or serum creatine level [@Lungulescu2011; @McDonald2018; @Zar2007].
 Those levels are being driven by the activities in the intestine [@diMasi2018].
@@
 difficile* more virulent.
-Strains are categorized for their virulence by the presence and production toxin A and toxin B and the prevelance in outbreaks, such as ribotype 027 and 078 [@AbernathyClose2020, @Theriot2011, @Goorhuis2008, @OConnor2009, @Chen2008].
-However, other studies have shown that virulence is not necessarily linked with toxin production **@Roa2015** or the strain **@Walk2012**.
-Furthermore, there is variation in the genome, growth rate, sporulation, germination, and toxin production in different isolates of a strain [@Carlson2013, @He2010].
-This variation may help explain why severe CDI prediction tools often miss identifying many patients with CDI that will develop severe disease [@Dieterle2020, @Butt2013, @Perry2021, @vanBeurden2017].
+Strains are categorized for their virulence by the presence and production toxin A, toxin B, and binary toxin and the prevelance in outbreaks, such as ribotype 027 and 078 [@AbernathyClose2020; @Theriot2011; @Goorhuis2008; @OConnor2009; @Chen2008].
+However, other studies have shown that virulence is not necessarily linked with toxin production [@Rao2015] or the strain [@Walk2012].
+Furthermore, there is variation in the genome, growth rate, sporulation, germination, and toxin production in different isolates of a strain [@Carlson2013; @He2010].
+This variation may help explain why severe CDI prediction tools often miss identifying many patients with CDI that will develop severe disease [@Dieterle2020; @Butt2013; @Perry2021; @vanBeurden2017].
 Therefore, it is necessary to gain a full understanding of all factors contributing to disease variation to improve our ability to predict severity.
@@
 *C.
-difficile* is unable to colonize an unperturbed healthy gut community and is only able to entry after a perturbation [@Schubert2015].
+difficile* is unable to colonize an unperturbed healthy gut community and is only able to become established after a perturbation [@Schubert2015].
 Once colonized, the different communities lead to different metabolic responses and dynamics of the *C.
-difficile* population [@Jenior2017, Jenior2018, @Lesniak2021].
-The gut bacteria metabolize primary bile acids into secondary bile acids [@Staley2016, @Long2017].
-The concentration of these bile acids affect the germination, growth, toxin production and biofilm formation [@Sorg2008, @Sorg2010, @Thanissery2017, @Dubois2019].
+difficile* population [@Jenior2017, Jenior2018; @Lesniak2021].
+The gut bacteria metabolize primary bile acids into secondary bile acids [@Staley2016; @Long2017].
+The concentration of these bile acids affect the germination, growth, toxin production and biofilm formation [@Sorg2008; @Sorg2010; @Thanissery2017; @Dubois2019].
 Members of the bacterial community also affect other metabolites *C.
@@
 *Bacteroides thetaiotaomicron* produce sialidases which release sialic acid from the mucosa for *C.
-difficile* to utilize [@Ng2013, @Ferreyra2014].
+difficile* to utilize [@Ng2013; @Ferreyra2014].
 The nutrient environemt affects toxin production [@MartinVerstraete2016].
@@
 
-The gut community has not been demostrated to modulate CDI disease severity.
-Myriad studies have explored the relationship between the microbiome and CDI but none have ventured into the variation of disease severity due to the microbiome.
+Myriad studies have explored the relationship between the microbiome and CDI but none have established a relationship between the variation of disease severity and the microbiome.
 Most CDI studies employ either an *in vitro* or *in vivo* model using a homogenous bacterial community.
-Collins et al used multiple human communities to colonize mice, however the communities were pooled prior to gavaging into germ-free mice [@Collins2015], resulting in a single community.
+Collins et al.
+used multiple human communities to colonize mice, however the communities were pooled prior to gavaging into germ-free mice [@Collins2015], resulting in a single community.
 Studies examining difference in disease often use different *C.
-difficile* strains or ribotypes in mice with similar microbiota as a proxy for variation in disease, such as strain 630 for non-severe and ribotype 027 for severe [@AbernathyClose2020, @Theriot2011, @Lawley2012, @Chen2008].
-There has been studies demonstrating variation in severity, but through tapering antibiotic dosage [@Lesniak2021, @Schubert2015, @Reeves2011] or by reducing the amount of cells or spores used for the challenge [@Chen2008, @Reeves2011].
+difficile* strains or ribotypes in mice with similar microbiota as a proxy for variation in disease, such as strain 630 for non-severe and ribotype 027 for severe [@AbernathyClose2020; @Theriot2011; @Lawley2012; @Chen2008].
+There have also been studies demonstrating variation in severity through tapering antibiotic dosage [@Lesniak2021; @Schubert2015; @Reeves2011] or by reducing the amount of cells or spores used for the challenge [@Chen2008; @Reeves2011].
 Our group has recently been uncovering how variation in the microbiome affects CDI but have been limited to *C.
-difficile* colonization [@Lesniak2021, @Tomkovich2020].
+difficile* colonization [@Lesniak2021; @Tomkovich2020].
 
@@
 difficile*, we hypothesized that the microbiome directly and indirectly modulates CDI disease severity.
-Our data demostrate gut bacterial relative abundances associate with variation in toxin production, histopathologic scoring of the cecum and severe disease.
-This anaylsis has revealed populations of *Akkermansia*, *Anaerostipes*, *Coprobacillus*, *Enterocloster*, *Lactonifactor*, and *Monoglobus* were more abundant in the microbiome of mice which little to no disease was detected.
-These genera providing a protective role are supported by previous studies.
-*Coprobacillus*, *Lactonifactor*, and *Monoglobus* have been shown to be involved in dietary fiber fermentation and associated with healthy communities [@Mabrok2011, @Kim2019, @doPrado2021, @Muthuramalingam2019].
-*Anaerostipes* and *Coprobacillus* produce short chain fatty acids which also have been associated with healthy communities [@Han2019, @Duncan2004, @Ye2018].
-Furthermore, *Coprobacillus*, which is abundant in mice with low histopathologic scores but rare in all other mice, has been shwon to contain putative type I lantibiotic gene cluster and inhibit *C.
-difficile* colonization [@Walsh2017, @Sandiford2018, @Stein2013].
-*Akkermansia* and *Enterocloster* were also idenitified as more abundant in mice which had a low clinical disease but have contridictory supporting evidence in the current literature.
-In our data, *Akkermansia* is most abundant in the mice with least disease but there were some mice with severe disease which had increased populations of *Akkermansia*.
-This could be attributed to either a more protective mucus layer had developed in the mice with little disease and *Akkermansia* is inhibiting colonization [@Stein2013, @Nakashima2021] or *Akkermansia* could be crossfeeding *C.
-difficile* or exposing a niche via its mucus consumption [@Geerlings2018, @Deng2018, @Engevik2020].
+Our data demostrate gut bacterial relative abundances associate with variation in toxin production, histopathologic scoring of the cecum and mortality.
+This anaylsis revealed populations of *Akkermansia*, *Anaerostipes*, *Coprobacillus*, *Enterocloster*, *Lactonifactor*, and *Monoglobus* were more abundant in the microbiome of mice which had low histopathologic scores and no mortality or detected toxin.
+The protective role of these genera are supported by previous studies.
+*Coprobacillus*, *Lactonifactor*, and *Monoglobus* have been shown to be involved in dietary fiber fermentation and associated with healthy communities [@Mabrok2011; @Kim2019; @doPrado2021; @Muthuramalingam2019].
+*Anaerostipes* and *Coprobacillus* produce short chain fatty acids which also have been associated with healthy communities [@Han2019; @Duncan2004; @Ye2018].
+Furthermore, *Coprobacillus*, which is abundant in mice with low histopathologic scores but rare in all other mice, has been shown to contain putative type I lantibiotic gene cluster and inhibit *C.
+difficile* colonization [@Walsh2017; @Sandiford2018; @Stein2013].
+*Akkermansia* and *Enterocloster* were also idenitified as more abundant in mice which had a low histopathologic scores but have contradictory supporting evidence in the current literature.
+In our data, *Akkermansia* is most abundant in the non-moribund mice with low histopathologic scores but there were some moribund mice which had increased populations of *Akkermansia*.
+This could be attributed to either a more protective mucus layer was present inhibiting colonization [@Stein2013; @Nakashima2021] or mucus consumption by *Akkermansia* could have been crossfeeding *C.
+difficile* or exposing a niche for *C.
+difficile* [@Geerlings2018; @Deng2018; @Engevik2020].
 Similarly with *Enterocloster*, in our data this genus was more abundant and associated with low histopathologic scores.
 It has been associated with healthy populations and has been used to mono-colonize germ-free mice to reduce the ability of *C.
-difficile* to colonize [@Reeves2012, @Ma2021].
-However, *Enterocloster* has mostly been indentified in infections that result in disease [@Haas2020, @Finegold2005].
-These data demostrate a population of bacteria has the potential to be either protective or harmful.
+difficile* to colonize [@Reeves2012; @Ma2021].
+However, *Enterocloster*, a member of indigenous gut communities, has been involved in infections, such as bacteremia [@Haas2020; @Finegold2005].
+These data demonstrate a population of bacteria has the potential to be either protective or harmful.
 The disease outcome is not likely based on the abundance of a single population of bacteria rather the result of the interactions of many populations.
 
-The groups of bacteria that were associated with either a higher histopathologic score or severe disease are members of the indigenous gut community that also have been associated with disease.
-These are often referred to as opportunitic pathogens, however based on the the results reported in this study it is more appropriate to apply the damage-response framework for microbial pathogenesis [@Pirofski2008, @Casadevall2014].
-The disease is not driven by a single entity, rather it is an emergent property of the responses of the host immune system, infecting microbe, and the microbes of the indigenous community.
-Many of the populations with pathgenic potential that associated with worse outcomes are also facultative anaerobes.
-*Enterococcus*, *Klebsiella*, *Shigella/Escherichia*, *Staphylococcus*, and *Streptococcus* have been shown to expand their population with antibiotic use [@VanInsberghe2020, @GarzaGonzalez2019, @Schubert2014] and are commonly detected in CDI cases [@Shafiq2020, @Keith2020, @Zackular2016, @Berkell2021].
-In addition to these populations, another set of bacteria, populations of *Eggerthella*, *Prevotellaceae* and *Helicobacter*, associated with worse outcomes have also been associated with intestinal inflammation [@Gardiner2014, @Iljazovic2020, @Nagalingam2013].
-Recently, *Helicobacter hepaticus* has been demostrated to be sufficient to cause susceptibility to CDI in IL-10^(-/-) B57BL/6 mice [@AbernathyClose2021].
-In our experiments, when *Helicobacter* was present, the mouse resulted in either severe disease or a high histopathologic score.
-While we did not use IL-10^(-/-) mice, it is possible the bacterial communtity or host response are similarly modified by *Helicobacter* allowing *C.
+The groups of bacteria that were associated with either a higher histopathologic score or moribundity are members of the indigenous gut community that also have been associated with disease, often referred to as opportunitic pathogens.
+Many of the populations with pathogenic potential that associated with worse outcomes are also facultative anaerobes.
+*Enterococcus*, *Klebsiella*, *Shigella/Escherichia*, *Staphylococcus*, and *Streptococcus* have been shown to expand after antibiotic use [@VanInsberghe2020; @GarzaGonzalez2019; @Schubert2014] and are commonly detected in CDI cases [@Shafiq2020; @Keith2020; @Zackular2016; @Berkell2021].
+In addition to these populations, *Eggerthella*, *Prevotellaceae* and *Helicobacter*, associated with worse outcomes have also been associated with intestinal inflammation [@Gardiner2014; @Iljazovic2020; @Nagalingam2013].
+Recently, *Helicobacter hepaticus* has been demostrated to be sufficient to cause susceptibility to CDI in IL-$10^{-/-}$ C57BL/6 mice [@AbernathyClose2021].
+In our experiments, when *Helicobacter* was present, the mouse resulted in a high histopathologic score.
+While we did not use IL-$10^{-/-}$ mice, it is possible the bacterial communtity or host response are similarly modified by *Helicobacter* allowing *C.
 difficile* infection and resulting in host damage.
-However, the populations identified other than *Helicobacter* are not as clear with their relationship between their relative abundance and the resulting disease severity.
+Aside from *Helicobacter*, these groups of bacteria associated with more severe outcomes did not have a conserved association between their relative abundance and the disease severity across all mice.
 
-The damage-response framework for pathogenesis helps to understand how bacterial populations identified in this study potentiate CDI disease severity through the varied interactions with the host and *C.
-difficile*.
-In this set of experiments, we use the same host, C57BL/6 mice, the same infecting microbe, *C.
-difficile* ribotype 027 isolate 431, with different gut bacterial communities.
-The members of those communities were often present in multiple levels of disease severity.
-Thus, it is not merely, the presence of the bacteria but their activity in response to the other microbes and host which determine their effect on host damage.
-Additionally, while we are using the same host and infecting microbe, they too are reacting to the specific members of the bacterial community.
+The relationship between a group of bacteria and severity may seem to vary, until we evaluate in the context of the environment inwhich there interact.
+Based on our results, it is more appropriate to apply the damage-response framework for microbial pathogenesis to CDI [@Pirofski2008; @Casadevall2014].
+Disease is not driven by a single entity, rather it is an emergent property of the responses of the host immune system, infecting microbe, *C.
+difficile*, and the microbes of the indigenous community.
+In the first set of experiments, we use the same host, C57BL/6 mice, the same infecting microbe, *C.
+difficile* ribotype 027 human isolate 431, with different gut bacterial communities.
+The bacterial groups in those communities were often present in both moribund and non-moribund and across the range of histopathologic scores.
+Thus, it is not merely, the presence of the bacteria but their activity in response to the other microbes and host which affect the extent of the host damage.
+Additionally, while each mouse and *C.
+difficile* has the same genetic background, they too are reacting to the specific microbial community.
 Disease severity is driven by the cummulative effect of the host immune response and the activity of the gut bacteria and *C.
@@
 However, if mucin degradation becomes too great it reduces the protective function of the mucin layer and exposes the epithelial cells [].
-This over-harvesting contributes to the host damage due to other members producing toxin.
-Thus disease is a balance of many activities of the community.
+This over-harvesting contributes to the host damage due to other members producing toxin [].
+Thus, the resultant intestinal damage is the balance of all activities in the gut environment.
 Host damage is the emergent property of numerous damage-response curves, one for host immune response, one for *C.
 difficile* activity and another for microbiome community activity, each of which are a composite curve of the individual activities from each group, such as antibody production, neutrophil infiltration, toxin production, sporulation, fiber and mucin degradation.
-Therefore, while we have identified populations of interest, it is likely targeting one specifically will reduce disease severity.
+Therefore, while we have identified populations of interest, it is likely necessary to target multiple to reduce those with community interactions contributing to host damage.
 
@@
 difficile* interactions result in severe disease.
-We can reduce the risk of severity with a better understanding of which interactions, whether be driven by specific community functions or groups of bacteria, reduce the host damage.
+We can reduce the risk of severity with a better understanding of which interactions, whether driven by specific community functions or groups of bacteria, reduce host damage.
 Approaching this problem from a damage-response framework allows us to experiment with modifying subsets of activities to reduce overall host damage.
-Our current clinical treatment targets an individual group in this tripartite system.
+Our current clinical treatment for primary CDI targets an individual group of this tripartite system.
 Most commonly, CDI is treated with an antibiotic to eliminate *C.
@@
 When a patient is diagnosed with CDI, the gut community composition, in addition to the traditionally obtained information, may improve our severity prediction and appropriate treatment.
-Treating the microbiome may contribute to reduced severity.
-From our results, promoting fiber metabolizing bacteria and reducing facultative anerobes could bolster the efficacy of current CDI treatments, reducing disease severities.
+Treating the microbiome at the time of diagnosis may prevent the infection from becoming more severity.
+Based on the general trends of our results, promoting fiber metabolizing bacteria and reducing facultative anerobes could bolster the efficacy of current CDI treatments, reducing disease severities.
 
@@
 
-**Animal care.** 5.
-64 mice, 33F and 31M, ages 6-13 weeks old, 3-4 mice/group (except 2 for C/DA00581 and MOUT/DA10093) each run once except for DA00581 which was run 4 times because we observed colonization resistance in one of the experiments
-15 human fecal samples, selected to create a subset list with diverse responses to health questionaire
+**Animal care.** 6- to 13-week old male and female germ-free C57BL/6 were obtained from a single breeding colony in the University of Michigan Germ-free Mouse Core.
+Mice were housed in cages of 2-4 mice per cage and maintained in germ-free isolators at the University of Michigan  germ-free facility.
+All mouse experiments were approved by the University Committee on Use and Care of Animals at the University of Michigan.
+**Were the mice maintained in gnotobiotic isolators throughout the experiment, one per donor?
+Or were they transferred to BSL2 conditions prior to administering the human FMTs?**
 
-***C. difficile* challenge.** M.
+***C.
+difficile* experiments.** Human fecal samples were obtained from Schubert et al.
+and selected based on community clusters [@Schubert2014] to result in diverse community structures.
+Feces were homogenized by mixing 200 mg of sample with 5 ml of PBS.
+Mice were inoculated with 100 $\mu$l of the fecal homogenate via oral gavage.
+Two weeks after the fecal community inoculation, mice were challenged with *C.
+difficile*.
+*C.
+difficile* isolates came from Carlson et al.
+which had previously been isolated and characterized [@Carlson2013; @Carlson2015].
+Spores concentration were determined both before and after challenge [@Sorg2009].
+$10^{3}$ *C.
+difficile* spores were given to each mouse via oral gavage.
+**How were human FMTs prepared (anaerobically vs aerobically, etc)?**
 
-**Sample collection.** F.
+**Sample collection.** Fecal samples were collected on the day of *C.
+difficile* challenge and the following 10 days.
+Each day, a fecal sample was also collected and weighed and the remaining sample was frozen at -20$^\circ$C.
+Anaerobically, the weighed fecal samples were serially diluted in PBS, plated on TCCFA plates, and incubated at 37$^\circ$C for 24 hours.
+The plates were then counted for the number of colony forming units (CFU) [@Winston2016].
 
-**DNA sequencing.** T.
+**DNA sequencing.** From the frozen fecal samples, total bacterial DNA was extracted using MOBIO PowerSoil-htp 96-well soil DNA isolation kit.
+We amplified the 16S rRNA gene V4 region and sequenced the resulting amplicons using an Illumina MiSeq as described previously [@Kozich2013].
 
-**Sequence curation.** S.
+**Sequence curation.** Sequences were processed with mothur(v.1.44.3) as previously described [@Kozich2013; @Schloss2009].
+In short, we used a 3% dissimilarity cutoff to group sequences into operational taxonomic units (OTUs).
+We used a naive Bayesian classifier with the Ribosomal Database Project training set (version 18) to assign taxonomic classifications to each OTU [@Wang2007].
+We sequenced a mock community of a known community composition and 16s rRNA gene sequences.
+We processed this mock community with our samples to calculate the error rate for our sequence curation, which was an error rate of 0.19%.
 
-**Statistical analysis and modeling.** D.
+**Toxin cytotoxicity assay.** Fecal sample remaining after plating for CFU and DNA extraction were collected in a preweighted sterile tube and stored at −80°C.
+To prepare the sample for the activity assay, samples were diluted 1:10 weight per volume using sterile PBS and then filter sterilized through a 0.22-$\mu$m filter.
+Toxin activity was assessed using a Vero cell rounding-based cytotoxicity assay as described previously [@Theriot2011].
+The cytotoxicity titer was determined for each sample as the last dilution, which resulted in at least 80% cell rounding.
+Toxin titers are reported as the log10 of the reciprocal of the cytotoxicity titer.
 
-**Code availability.** Scripts necessary to reproduce our analysis and this paper are available in an online repository (https://github.com/SchlossLab/Lesniak_XXXX).
+**Histopathology evaluation.** Mouse ceca tissue was placed in histopathology cassettes and fixed in 10% formalin, then stored in 70% ethanol.
+McClinchey Histology Labs, Inc.
+(Stockbridge, MI) embedded the samples in paraffin, sectioned, and created the hematoxylin and eosin-stained slides.
+The slides were scored using previously described criteria by a board-certified veterinary pathologist scored who was blinded to the experimental groups [@Theriot2011].
+
+**Statistical analysis and modeling.** To compare community structures, we calculated Yue and Clayton dissimilarity matrices in mothur [@Yue2005].
+We rarified samples to 2,107 sequences per sample to limit uneven sampling biases.
+We tested for differences in individual taxonomic groups that would explain the outcome differences with LEfSe [@Segata2011] in mothur.
+Remaining statistical analysis and data visualization was performed in R (v4.0.5) with the tidyverse package (v1.3.1).
+We tested for significant differences in $\beta$-diversity ($\theta$~YC~)using the Wilcoxon rank sum test.
+We used Spearman's corrlation to identified the OTU which had a correlation between their relative abundance and the histopathologic summary score.
+*P* values were then corrected for multiple comparisons with a Benjamini and Hochberg adjustment for a type I error rate of 0.05 [@Benjamini1995].
+We built random forest models using the mikropml package [@Topcuoglu2021] with OTUs from day 0 samples using mtry values of 1 through 10, 15, 20, 25, 40, 50, 100.
+The split for training and testing varied by model to avoid overfitting the data.
+The toxin and moribundity models were trained with 60% of the data.
+The histopathologic score model was trained with 80% of the data.
+Lastly, we did not analyze murine communities to donor community or clinical data because germ-free mice colonized with human fecal communities become more like indigenous murine communities once stabilized.
+Futhermore, we are not trying to make any claims about human bacteria and their association with human CDI.
+
+**Code availability.** Scripts necessary to reproduce our analysis and this paper are available in an online repository (https://github.com/SchlossLab/Lesniak_Severity_XXXX_2021).
 
@@
 
+Thank you to Sarah Lucas and Sarah Tomkovich for critical discussion in the development and execution of this project.
+We also thank the University of Michigan Germ-free Mouse Core for assitance with our germfree mice, funded in part by U2CDK110768.
 This work was supported by several grants from the National Institutes for Health R01GM099514, U19AI090871, U01AI12455, and P30DK034933.
@@
 
-**Figure 1. R.** (A-C).
+**Figure 1.
+Human fecal microbial communtities establish diverse gut bacterial communties in germ-free mice.** (A) Relative abundace of 10 most abundant bacterial classes of the gut bacterial communities on timepoint day 0, 14 days post-inoculation of germ-free C57Bl/6 mice with human fecal samples.
+Each column of abundances represent an individual mouse.
+Each group of mice that recieved the same donor are group together and labeled with a letter (A through O).
+(B) Median (points) and interquartile range (lines) of $\beta$-diversity ($\theta$~YC~) between an individual mouse and either all others which were inoculated with feces from the same donor or from a different donor.
+The $\beta$-diversity amongst in the same donor comparison group was significantly less than the $\beta$-diversity of the different donor group (*P* < 0.05, calculated by Wilcoxon rank sum test).
 
@@
 
-**Figure 2. M.** F.
+**Figure 2.
+All donor groups resulted in *C.
+difficile* infection but with different outcomes.** *C.
+difficile* CFU per gram of stool measured the day after challenge with $10^{3}$ *C.
+difficile* ribotype 027 human isolate 431 spores and at the end of the experiment, 10 days post-challenge.
+Each point is an individual mouse.
+Mice are grouped by donor and labeled by the donor letter.
+Points are colored by donor group.
+Mice from donor groups J through O succumbed to the infection prior to day 10 and were not plated.
+LOD = Limit of detection.
 
@@
 
-**Figure 3. O.** F.
+**Figure 3.
+Histopathologic score and toxin activity varied across donor groups.** (A) Fecal toxin activity detected in some mice post *C.
+difficile challenge in both moribund and non-moribund mice.
+(B) Histopathologic summary score for mice at the end of the experiment.
+Samples were collected for histopathologic scoring on day 10 post-challenge for non-moribund mice or the day the mouse succumbed to the infection for the moribund group (day 2-3 post-challenge).
+Each point is an individual mouse.
+Mice are grouped by donor and labeled by the donor letter.
+Points are colored by donor group.
+Missing points were from mice with insufficient sample collected for assaying fecal toxin.
 
@@
 
-**Figure 4. E.** A.
+**Figure 4.
+Individual fecal bacterial community members of the murine gut associate with *C .difficile* infection outcomes.** (A and B) Relative abundance of genera at the time of *C.
+difficile* challenge (Day 0) that vary significantly by the moribundity and histopathologic summary score or toxin activity by LEfSe analysis.
+Median (points) and interquartile range (lines) are plotted.
+Genera are ordered alphabetically to ease comparisons across analyses.
+(A) Comparison across infection outcome of moribund (colored black) or non-moribund with either a high histopathologic score (score greater than the median score of 5, colored green) or a low histopathologic summary score (score less than the median score of 5, colored light green).
+(B) Comparison between mice which toxin activity was detected (Toxigenic, colored dark purple) and mice which no toxin activity was detected (Non-toxigenic, colored light purple).
+(C) Endpoint bacterial OTUs correlate with histopathologic summary score.
+Each individual mouse is plotted (transparent gray point).
+Spearman correlatations are statistically significant after Benjamini-Hochberg correction for multiple comparisons.
+* indicates bacterial group was unclassified at lower classification ranks.
 
@@
 
-**Figure 5. D.** (A) L
+**Figure 5.
+Fecal bacterial community members of the murine gut at the time of *C.
+difficile* (Day 0) predict outcomes of the infection.** Day 0 bacterial community members grouped by different classification rank were modeled with random forest to predict the infection outcome.
+The models used the highest taxonomic classification rank that performed as well as the lower ranks.
+Median (solid points) and interquartile range (lines) of the group relative abundance and each individual (transparent points) relative abundances are plotted.
+Bacterial groups are ordered by their importance to the model; top taxonomic group has the greatest decrease in performance which its relative abundances are permutted.
+* indicates bacterial group was unclassified at lower classification ranks.
+(A) Bacterial members grouped by phyla predicted which mice would have toxin activity detected at any point thoughout the infection (Toxigenic, dark purple).
+(B) Bacterial members grouped by class predicted which mice would become moribund (dark blue).
+(C) Bacterial members grouped by genera predicted if the mice would have a high (score greater than the median score of 5, colored dark green) or low (score less than the median score of 5, colored light green) histopathologic summary score.
 
@@
 
-**Figure 6. D.** (A) L
+**Figure 6.
+*C.
+difficile* infection outcome were similar within human fecal donor groups across different *C.
+difficile* isolates.** Mice colonized with human fecal donor communities G, F, or M were challenged with the isolate from the previous experiments in this study, *C.
+difficile* ribotype 027 human isolate 431 (blue points), or a different *C.
+difficile* isolate (purple points).
+*C.
+difficile* CFUs were enumerated at the end of the experiment (day 10 post-challenge) and plotted grouped by donor source (top plot).
+Histopathologic scores from the tissue collected at the endpoint (either day 10 post-challenge or day mice succumbed to infection) were plotted grouped by donor source (bottom plot).
+Mice that received fecal communities from human donor M succumbed to the infection prior to day 10 post-challenge so were not plated for CFU (DECD = deceased).
+RT = ribotype, Isolate = human isolate id, LOD = limit of detection.
 
@@
 
-**Figure S1. D.** (A) L
+**Figure S1.
+Histopathologic score of tissue damage at the endpoint of the infection.** Tissue collected at the endpoint, either day 10 post-challenge (Non-moribund) or day mice succumbed to infection (Moribund), were scored from histopathologic damage.
+Mice (points) are grouped and colored by their human fecal communtity donor.
+Missing points were from mice with insufficient sample for histopathologic scoring.
 
@@
 
-**Figure S2. D.** (A) L
+**Figure S2.
+Random forest models predict outcomes of the *C.
+difficile* challenge.** (A-C) Taxonomic classification rank model performance.
+Relative abundance at the time of *C.
+difficile* challenge (Day 0) of the bacterial community members grouped by different classification rank were modeled with random forest to predict the infection outcome.
+The models used the highest taxonomic classification rank that performed as well as the lower ranks.
+Black rectagle highlights classification rank used to model each outome.
+(D-F) Model feature importance.
+Bacterial groups are ordered by their deacrease in area under receiver-operator curve (AUC) when its relative abundances are permutted.
+Individual relative abundances were added to F since differences are in AUC outside the interquartile range.
+* indicates bacterial group was unclassified at lower classification ranks.
+For all plots, median (solid points) and interquartile range (lines) are plotted.
+(A) Toxin production modeled which mice would have toxin detected during the experiment.
+(B) Moribundity modeled which mice would succumb to the infection prior to day 10 post-challenge.
+(C) Histopathologic score modeled which mice would have a high (score greater than the median score of 5) or low (score less than the median score of 5) histopathologic summary score.
+(D) Bacterial phyla which affect the performance of predicting toxin activity when permutted.
+(E) Bacterial classes which affect the performance of predicting moribundity when permutted.
+(D) Bacterial genera which affect the performance of predicting histopathologic score when permutted.
 
@@
 
-**Figure S3. D.** (A) L
-
-\hfill\break
-
-**Table 1. D.** (A) L
-
-samples_used <- readxl::read_xlsx('data/raw/humanGF_ids.xlsx') %>% 
-	pull(human_source) %>% 
-	unique
-donor_data <- readxl::read_xlsx('data/raw/MIMARKS_cdclinical.xlsx') %>% 
-	filter(sample_id %in% samples_used) %>% 
-	select(sample_id, biome, age, gender, "antibiotics >3mo", protonpump, 
-		antacid, Healthworker, historyCdiff, Surgery6mos, Vegetarian, weight, disease_stat)
-
-**Table 2. D.** (A) L
-
-read.csv(here('data/raw/Hanna_in_vitro_data_plus_germination.txt')) %>% 
-  rownames_to_column('isolate') %>% 
-  filter(isolate %in% c('DA00299', 'DA00395', 'DA00458', 'DA00431'))
+**Figure S3.
+Selection of communities to test for conserved *C.
+difficile* infection outcome with different *C.
+difficile* isolate.** (A) Relative abundance of genera in mice from previous experiments which received human fecal donor communities F, G, or M.
+Each column is an individual mouse.
+(B) Features from the random forest model that predicted moribundity using relative abundance of bacteria grouped by their genus classification rank and ordered by their decrease in AUC when each feature is permutted.
+Median (solid points) and interquartile range (lines) change in AUC are plotted.
+(C) Relative abundance of the most important genera predicting moribundity.
+Median (solid points) and interquartile range (lines) of the group relative abundance and each individual (transparent points) relative abundances are plotted.
+* indicates bacterial group was unclassified at lower classification ranks.

### 2021-11-30 code/plot_figure_1.R

@@ -90,7 +90,7 @@ day_0_plot <- relative_abundance_data %>% # only plot top 10
 	  theme_bw() + 
     labs(x = NULL, y = NULL, fill = NULL) + 
     scale_x_discrete(guide = guide_axis(angle = 45)) + 
-    facet_grid(.~donor_labels, scales = 'free_x') + 
+    facet_grid(.~donor_labels, scales = 'free_x', space = 'free_x') + 
     theme(axis.ticks.x = element_blank(),
           axis.text.x = element_blank(),
           legend.position = 'bottom',
@@ -111,11 +111,13 @@ beta_plot <- beta_data %>%
 #  save plot
 ###############################################################################
 ggsave(here('results/figures/Figure_1.jpg'),
-  cowplot::plot_grid(day_0_plot + theme(text=element_text(size = 9)), 
+  cowplot::plot_grid(cowplot::plot_grid(
+                      day_0_plot + theme(text=element_text(size = 9)),
+                      labels = c('       Donor'), label_size = 9, label_fontface = 'plain'), 
                      NULL,
                      beta_plot + theme(text=element_text(size = 9)), 
                      ncol = 1, 
                      rel_heights = c(6, 0.3, 1.5),
                      labels = c('A', 'B', NULL)),
-  height = 9.0625, width = 6.875, unit = 'in')
+  height = 5.5, width = 4.5, unit = 'in')
 ###############################################################################

### 2021-11-30 code/plot_figure_2.R

@@ -21,13 +21,14 @@ donor_aes <- donor_df
 #  setup data
 ###############################################################################
 LOD_df <- data.frame(x = 14.5, y = 2)
-DECD_df <- data.frame(x = 12, y = 3, day = as.factor('Day 10'))
+DECD_df <- data.frame(x = 12, y = 3, day = as.factor('10 days post-challenge'))
 
 cfu_data <- metadata %>% 
 	filter(cdiff_strain == 431,
 		day %in% c(1,10),
 		!is.na(cdiff_cfu)) %>% 
-	mutate(day = factor(paste('Day', day), levels = c('Day 1', 'Day 10')),
+	mutate(day = ifelse(day == 1, '1 day post-challenge', '10 days post-challenge'),
+		day = factor(day, levels = c('1 day post-challenge', '10 days post-challenge')),
 		cdiff_cfu = ifelse(cdiff_cfu > 0, log10(cdiff_cfu), log10(60))) %>% 
 	left_join(donor_aes, by = c('human_source'))
 
@@ -36,31 +37,31 @@ cfu_data <- metadata %>%
 ###############################################################################
 cfu_plot <- cfu_data %>% 
 	ggplot(aes(x = donor_labels, y = cdiff_cfu, color = donor_colors)) +
+		geom_hline(yintercept = 2, linetype = 'dashed', 
+			size = .5, color = 'white') + 
 		geom_jitter(width = 0.2) + 
 		scale_color_identity() + 
-		facet_grid(day~.) + 
+		facet_wrap(.~day, ncol = 1) + 
 		scale_y_continuous(limits = c(1.5,8.5),
 			breaks = c(2,4,6,8),
 			labels = c('10^2', '10^4', '10^6', '10^8')) + 
 		theme_bw() + 
-		labs(x = 'Donor Source', y = '*C. difficile* CFU') + 
+		labs(x = 'Donor', y = '*C. difficile* CFU') + 
 		guides(color = 'none') + 
 		theme(axis.text.y = ggtext::element_markdown(),
 			axis.title.y = ggtext::element_markdown(),
 			strip.text.y = element_text(angle = 0, size = 12),
 			strip.background = element_blank(),
-			panel.spacing = unit(2, "lines")) + 
-		geom_hline(yintercept = 2, linetype = 'dashed', 
-			size = .5, color = 'white') + 
+			panel.spacing = unit(1, "lines")) + 
[…]

### 2021-11-30 code/plot_figure_3.R

@@ -13,7 +13,8 @@ source(here::here('code/utilities.R'))
 ###############################################################################
 # load data
 ###############################################################################
-metadata <- read_metadata()
+metadata <- read_metadata() %>% 
+  filter(cdiff_strain == '431')
 
 toxin <- read_toxin()
 
@@ -26,17 +27,19 @@ donor_aes <- donor_df
 ###############################################################################
 toxin <- toxin %>% 
   inner_join(select(metadata, group, human_source, early_euth), by = c('group')) %>% 
-  mutate(outcome = ifelse(early_euth == T, 'Severe', 'Non-severe')) %>% 
+  mutate(outcome = ifelse(early_euth == T, 'Moribund', 'Non-moribund'),
+         outcome = factor(outcome, levels = c('Non-moribund', 'Moribund'))) %>% 
   filter(!is.na(outcome)) %>%
   mutate(day = as.numeric(toxin_sample_day)) %>% 
   left_join(donor_aes, by = 'human_source') %>% 
-  filter(!day %in% c(4,7),
+  filter(!day %in% c(3,4,7),
          toxin_sample_type == 'stool') 
 
 histology <- histology %>% 
   left_join(distinct(select(metadata, mouse_id, early_euth)), by = 'mouse_id') %>% 
   filter(!is.na(early_euth)) %>% 
-  mutate(early_euth = ifelse(early_euth, 'Severe', 'Non-severe'),
+  mutate(early_euth = ifelse(early_euth, 'Moribund', 'Non-moribund'),
+         early_euth = factor(early_euth, levels = c('Non-moribund', 'Moribund')),
          mouse_id = paste0(cage_id, '_', ear_tag),
          score = summary_score) %>% 
   left_join(donor_aes, by = 'human_source')
@@ -47,15 +50,20 @@ histology <- histology %>%
 toxin_plot <- toxin %>% 
   ggplot(aes(x = donor_labels, y = Log_repiricoal_dilution, 
              fill = donor_colors, color = donor_colors)) + 
-    geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 2) + 
+    geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 1.5) + 
     scale_fill_identity() + 
     scale_color_identity() + 
-    labs(x = NULL, y = 'Toxin activity') + 
+    labs(x = NULL, y = expression(Toxin ~ Titer ~ Log[10])) + 
[…]

### 2021-11-30 code/plot_figure_4.R

@@ -55,19 +55,19 @@ plot_lefse <- function(file_name){
 		mutate(taxa_label = Genus,
 			relative_abundance = ifelse(relative_abundance == 0, 0.045, 
 				relative_abundance),
-			class = case_when(class == 'severe' ~ 'Severe',
-				class == 'moderate' ~ 'Non-severe, High histopathologic score',
-				class == 'mild' ~ 'Non-severe, Low histopathologic score',
+			class = case_when(class == 'severe' ~ 'Moribund',
+				class == 'moderate' ~ 'Non-moribund, High histopathologic score',
+				class == 'mild' ~ 'Non-moribund, Low histopathologic score',
 				class == FALSE ~ 'Non-toxigenic',
 				class == TRUE ~ 'Toxigenic', 
 				T ~ 'NA'),
-			class = factor(class, levels = c('Non-severe, Low histopathologic score',
-				'Non-severe, High histopathologic score', 'Severe',
+			class = factor(class, levels = c('Non-moribund, Low histopathologic score',
+				'Non-moribund, High histopathologic score', 'Moribund',
 				'Non-toxigenic', 'Toxigenic')),
 			taxa_label = gsub('_', ' ', taxa_label),
 			taxa_label = paste0('*', taxa_label, '*'),
 			taxa_label = ifelse(grepl('unclassified', taxa_label), 
-				paste('Unclassified', gsub(' unclassified\\*', '*', taxa_label)),
+				gsub(' unclassified\\*', '* *', taxa_label),
 				taxa_label)) %>% 
 		ggplot(aes(x = taxa_label, y = relative_abundance, color = class)) +
 			stat_summary(fun.data = 'median_hilow', aes(group = class),
@@ -82,7 +82,9 @@ plot_lefse <- function(file_name){
 				panel.grid.major.y = element_line(color = 'gray95'),
 				panel.grid.major.x = element_line(color = 'gray85'),
 				panel.grid.minor.x = element_blank(),
-				legend.position = 'top') + 
+				legend.position = 'top',
+				legend.margin=margin(2,0,0,0),
+    	    	legend.box.margin=margin(2,-10,-10,-10)) + 
 			coord_flip()
 }
 
@@ -121,20 +123,19 @@ day_10_hist_genus_corr_plot <- day_10_hist_genus_corr %>%
 \\(', tax_otu_label),
[…]

### 2021-11-30 code/plot_figure_5.R

@@ -80,7 +80,7 @@ day_0_moribund <- metadata %>%
 	select(group, mouse_id, day, early_euth) %>% 
 	filter(day == 0) %>% 
 	inner_join(shared, by = c('group' = 'Group')) %>% 
-	mutate(early_euth = ifelse(early_euth, 'Severe', 'Non-severe')) %>% 
+	mutate(early_euth = ifelse(early_euth, 'Moribund', 'Non-moribund')) %>% 
 	inner_join(day_0_moribund_features, by = c('Class' = 'names')) %>% 
 	group_by(group, early_euth, taxa_label) %>% 
 	summarise(counts = sum(counts)) %>% 
@@ -109,10 +109,11 @@ day_0_histology <- metadata %>%
 	filter(early_euth == F) %>% 
 	inner_join(select(histology, mouse_id, summary_score), 
 		by = c('mouse_id')) %>% 
-	mutate(hist_score = case_when(summary_score > 5 ~ 'High',
-		summary_score < 5 ~ 'Low',
+	mutate(hist_score = case_when(summary_score > 5 ~ 'High histopathologic score',
+		summary_score < 5 ~ 'Low histopathologic score',
 		T ~ 'NA')) %>% 
-	filter(hist_score %in% c('High', 'Low')) %>% 
+	filter(hist_score %in% c('High histopathologic score', 
+		'Low histopathologic score')) %>% 
 	inner_join(day_0_histology_features, by = c('Genus' = 'names')) %>% 
 	group_by(group, hist_score, taxa_label) %>% 
 	summarise(counts = sum(counts)) %>% 
@@ -125,7 +126,7 @@ day_0_histology <- metadata %>%
 day_0_toxin_feature_abundance_plot <- day_0_predict_future_toxin_df %>% 
 	ggplot(aes(x = taxon, y = counts, color = toxin)) + 
 		geom_jitter(position = position_jitterdodge(dodge.width = 0.7, jitter.width = 0.3), 
-			alpha = 0.2) + 
+			alpha = 0.1) + 
 		stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5),
 			position = position_dodge(width = 0.7)) +
 		scale_y_log10() + 
@@ -141,7 +142,7 @@ day_0_toxin_feature_abundance_plot <- day_0_predict_future_toxin_df %>%
 day_0_moribund_feature_abundance_plot <- day_0_moribund %>% 
[…]

### 2021-11-30 code/plot_figure_6.R

@@ -2,7 +2,7 @@
 #
 # Figure 6
 #	main idea - Is our observation conserved across isolates?
-#   plot - CFU and histopathology score with different isolates
+#	plot - CFU and histopathology score with different isolates
 #
 # Nick Lesniak 2021-10-26
 ###############################################################################
@@ -13,97 +13,130 @@ source(here::here('code/utilities.R'))
 ###############################################################################
 # load data
 ###############################################################################
+isolate_levels <- c('     RT         \nIsolate         ', 
+	'027\n431', 'UM11\n299', '027\n395','027\n458')
+
 metadata <- read_metadata() %>% 
-  mutate(isolate_431 = ifelse(cdiff_strain == 431, '431', 'Other'),
-         isolate = factor(cdiff_strain, levels = c('431', '299', '395','458')))
-  
+	mutate(isolate_431 = ifelse(cdiff_strain == 431, '431', 'Other'),
+		isolate = ifelse(cdiff_strain == 299, paste0('UM11\n', cdiff_strain),
+			paste0('027\n', cdiff_strain)),
+		isolate = factor(isolate, levels = isolate_levels))
+	
 toxin <- read_toxin()
 
 histology <- read_histology() %>% 
-  select(summary_score, mouse_id)
+	select(summary_score, mouse_id)
 
 donor_aes <- donor_df
 
+donor_levels <- c(' ', 'G' ,'F', 'M')
+
 ###############################################################################
-#  setup data
+#	setup data
 ###############################################################################
 non_431_donors <- metadata %>% 
-  filter(cdiff_strain != 431) %>% 
-  pull(human_source) %>% 
-  unique
+	filter(cdiff_strain != 431) %>% 
+	pull(human_source) %>% 
+	unique
 
 cfu_data <- metadata %>% 
-  filter(human_source %in% non_431_donors,
-         day == 10) %>% 
-  add_row(human_source = 'DA00578', cdiff_cfu = NA, isolate_431 = 'Other') %>% 
-  left_join(toxin, by = 'group') %>% 
-  left_join(donor_aes, by = c('human_source')) %>% 
-  mutate(cdiff_cfu = ifelse(cdiff_cfu == 0, log10(60), log10(cdiff_cfu)),
[…]

### 2021-11-30 code/plot_figure_S1.R

@@ -1,8 +1,9 @@
 ###############################################################################
 #
 # Figure S1
-#	main idea - mice lost weight with challenge
-# 	plot - % weight loss through experiment
+#	main idea - moribund mice had high levels of 
+#     epithelial damage, tissue edema and inflammation
+# 	plot - epithelial damage, tissue edema and inflammation
 #
 # Nick Lesniak 2021-10-26
 ###############################################################################
@@ -15,43 +16,50 @@ source(here::here('code/utilities.R'))
 ###############################################################################
 metadata <- read_metadata()
 
+histology <- read_histology()
+
 donor_aes <- donor_df
 
 ###############################################################################
 #  setup data
 ###############################################################################
-wt_loss_df <- metadata %>% 
-  filter(cdiff_strain == 431,
-         day %in% c(2,10)) %>% 
-  mutate(day = factor(paste('Day', day), levels = c('Day 2', 'Day 10'))) %>% 
-  left_join(donor_aes, by = 'human_source')
+histology <- histology %>% 
+  left_join(distinct(select(metadata, mouse_id, early_euth)), by = 'mouse_id') %>% 
+  filter(!is.na(early_euth)) %>% 
+  mutate(early_euth = ifelse(early_euth, 'Moribund', 'Non-moribund'),
+         early_euth = factor(early_euth, levels = c('Non-moribund', 'Moribund')),
+         mouse_id = paste0(cage_id, '_', ear_tag)) %>% 
+  left_join(donor_aes, by = 'human_source') %>% 
+  rename('Tissue edema' = edema_tissue,
+    'Tissue inflammation' = inflammation_tissue,
+    'Epithelial damage' = epithelial_damage) %>% 
+  pivot_longer(cols = c('Tissue edema', 'Tissue inflammation', 'Epithelial damage'),
+    names_to = 'type', values_to = 'score')
 
-DECD_df <- data.frame(x = 12, y = 75, day = as.factor('Day 10'))
 ###############################################################################
 #  plot data
 ###############################################################################
-wt_loss_plot <- wt_loss_df %>% 
-  ggplot(aes(x = donor_labels, y = percent_weightLoss, color = donor_colors)) +
-    geom_jitter(width = 0.2) + 
-	scale_color_identity() + 
-    facet_grid(day~.)+ 
+histology_plot <- histology %>%  
[…]

### 2021-11-30 code/plot_figure_S2.R

@@ -82,10 +82,11 @@ day_0_histology_plot_df <- features_df %>%
 day_0_toxin_perf_plot <- performance_df %>% 
   filter(dataset == 'day_0_predict_future_toxin') %>% 
   ggplot(aes(x = taxonomic_level, y = Performance, color = metric)) + 
-    geom_boxplot() + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5), 
+      position = position_dodge(width = .5)) +
     geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
     theme_bw() +
-    geom_rect(data = data.frame(ymin = 0.48, ymax = 0.99, method = 'rf', taxonomic_level = 6, Performance = 1),
+    geom_rect(data = data.frame(ymin = 0.75, ymax = 0.9, method = 'rf', taxonomic_level = 6, Performance = 1),
               aes(xmin = taxonomic_level - 0.5, xmax = taxonomic_level + 0.5, ymin = ymin, ymax = ymax, color = NA), 
               fill = NA) + 
     labs(x = NULL, y = 'Performance (AUC)', color = NULL)
@@ -102,10 +103,11 @@ day_0_toxin_feature_plot <- day_0_toxin_production %>%
 day_0_moribund_perf_plot <- performance_df %>% 
   filter(dataset == 'day_0_moribund') %>% 
   ggplot(aes(x = taxonomic_level, y = Performance, color = metric)) + 
-    geom_boxplot() + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5), 
+      position = position_dodge(width = .5)) +
     geom_hline(yintercept = 0.5, linetype = 'dashed', size = 0.25) +
     theme_bw() +
-    geom_rect(data = data.frame(ymin = 0.65, ymax = 1.01, method = 'rf', taxonomic_level = 5, Performance = 1),
+    geom_rect(data = data.frame(ymin = 0.85, ymax = .98, method = 'rf', taxonomic_level = 5, Performance = 1),
               aes(xmin = taxonomic_level - 0.5, xmax = taxonomic_level + 0.5, ymin = ymin, ymax = ymax, color = NA), 
               fill = NA) + 
     labs(x = NULL, y = 'Performance (AUC)', color = NULL)
[…]

### 2021-11-30 code/plot_figure_S3.R

@@ -39,7 +39,7 @@ day_0_moribund <- metadata %>%
 	select(group, mouse_id, day, early_euth) %>% 
 	filter(day == 0) %>% 
 	inner_join(shared, by = c('group' = 'Group')) %>% 
-	mutate(early_euth = ifelse(early_euth, 'Severe', 'Non-severe'))
+	mutate(early_euth = ifelse(early_euth, 'Moribund', 'Non-moribund'))
 
 ###############################################################################
 # create relative abundance plot
@@ -70,7 +70,7 @@ day_0_plot <- metadata %>%
 		theme_bw() + 
 		labs(x = NULL, y = NULL, fill = NULL) + 
 		scale_x_discrete(guide = guide_axis(angle = 45)) + 
-		facet_grid(.~donor_labels, scales = 'free_x') + 
+		facet_grid(.~donor_labels, scales = 'free_x', space = 'free_x') + 
 		theme(axis.ticks.x = element_blank(),
 			axis.text.x = element_blank(),
 			legend.position = 'bottom',
@@ -97,7 +97,7 @@ day_0_moribund_feature_plot <- day_0_moribund_plot_df %>%
 	ggplot(aes(x = reorder(taxa_level, median_diff), y = perf_metric_diff)) + 
 	stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
 	coord_flip() + 
-	labs(x = NULL, y = expression(Delta~AUC), title = 'Features of severe disease model') +
+	labs(x = NULL, y = 'Decrease AUC', title = 'Features of moribund model') +
 	theme_bw() +
 	theme(axis.text.y = ggtext::element_markdown())
 
@@ -138,7 +138,7 @@ ggsave(here('results/figures/Figure_S3.jpg'),
 				 cowplot::plot_grid(
 					 day_0_moribund_feature_plot + theme(text=element_text(size = 7)),
 					 day_0_moribund_feature_abundance_plot + theme(text=element_text(size = 7)),
-					 ncol = 1),
-				 nrow = 1),
+					 ncol = 1, labels = c('B', 'C')),
+				 nrow = 1, labels = c('A', NULL)),
 			 height = 7, width = 7, unit = 'in')
 ###############################################################################

### 2021-12-09 code/plot_figure_1.R
refine figures
remove transparent points and switch to N/M instead of letter labels

@@ -119,5 +119,5 @@ ggsave(here('results/figures/Figure_1.jpg'),
                      ncol = 1, 
                      rel_heights = c(6, 0.3, 1.5),
                      labels = c('A', 'B', NULL)),
-  height = 5.5, width = 4.5, unit = 'in')
+  height = 5.5, width = 5.1, unit = 'in')
 ###############################################################################

### 2021-12-09 code/plot_figure_5.R

@@ -125,8 +125,8 @@ day_0_histology <- metadata %>%
 ###############################################################################
 day_0_toxin_feature_abundance_plot <- day_0_predict_future_toxin_df %>% 
 	ggplot(aes(x = taxon, y = counts, color = toxin)) + 
-		geom_jitter(position = position_jitterdodge(dodge.width = 0.7, jitter.width = 0.3), 
-			alpha = 0.1) + 
+		#geom_jitter(position = position_jitterdodge(dodge.width = 0.7, jitter.width = 0.3), 
+		#	alpha = 0.1) + 
 		stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5),
 			position = position_dodge(width = 0.7)) +
 		scale_y_log10() + 
@@ -141,8 +141,8 @@ day_0_toxin_feature_abundance_plot <- day_0_predict_future_toxin_df %>%
 
 day_0_moribund_feature_abundance_plot <- day_0_moribund %>% 
 	ggplot(aes(x = taxon, y = counts, color = early_euth)) + 
-		geom_jitter(position = position_jitterdodge(dodge.width = 0.7, jitter.width = 0.3), 
-			alpha = 0.1) + 
+		#geom_jitter(position = position_jitterdodge(dodge.width = 0.7, jitter.width = 0.3), 
+		#	alpha = 0.1) + 
 		stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5),
 			position = position_dodge(width = 0.7)) +
 		scale_y_log10() + 
@@ -157,8 +157,8 @@ day_0_moribund_feature_abundance_plot <- day_0_moribund %>%
 
 day_0_hist_feature_abundance_plot <- day_0_histology %>% 
 	ggplot(aes(x = taxon, y = counts, color = hist_score)) + 
-		geom_jitter(position = position_jitterdodge(dodge.width = 0.7, jitter.width = 0.3), 
-			alpha = 0.1) + 
+		#geom_jitter(position = position_jitterdodge(dodge.width = 0.7, jitter.width = 0.3), 
+		#	alpha = 0.1) + 
 		stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5),
 			position = position_dodge(width = 0.7)) +
 		scale_y_log10() + 
@@ -185,9 +185,9 @@ ggsave(here('results/figures/Figure_5.jpg'),
 		 cowplot::plot_grid(NULL,		 
[…]

### 2021-12-09 code/plot_figure_6.R

@@ -29,7 +29,7 @@ histology <- read_histology() %>%
 
 donor_aes <- donor_df
 
-donor_levels <- c(' ', 'G' ,'F', 'M')
+donor_levels <- c(' ', 'N7' ,'N6', 'M4')
 
 ###############################################################################
 #	setup data
@@ -73,16 +73,16 @@ cfu_plot <- cfu_data %>%
 		geom_point(position = position_jitterdodge(jitter.width = 0.1)) + 
 		geom_hline(yintercept = 2, linetype = 'dashed', color = 'lightgray') + 
 		geom_label(data = data.frame(x = 1, y = 2, 
-			donor_labels = factor('G', levels = donor_levels)), aes(x = x, y = y), 
+			donor_labels = factor('N7', levels = donor_levels)), aes(x = x, y = y), 
 			label = "LOD", fill = 'white', color = 'white', size = 8/.pt, inherit.aes = F) + 
 		geom_text(data = data.frame(x = 1, y = 2, 
-			donor_labels = factor('G', levels = donor_levels)), aes(x = x, y = y), 
+			donor_labels = factor('N7', levels = donor_levels)), aes(x = x, y = y), 
 			label = "LOD", color = 'lightgray', size = 8/.pt, inherit.aes = F) + 
 		geom_text(data = data.frame(x = 1, y = 1.65, 
-			donor_labels = factor('M', levels = donor_levels)), aes(x = x, y = y), 
+			donor_labels = factor('M4', levels = donor_levels)), aes(x = x, y = y), 
 			label = "DECD", color = "#3588d1", size = 8/.pt, inherit.aes = F) + 
 		geom_text(data = data.frame(x = 2, y = 1.65, 
-			donor_labels = factor('M', levels = donor_levels)), aes(x = x, y = y), 
+			donor_labels = factor('M4', levels = donor_levels)), aes(x = x, y = y), 
 			label = "DECD", color = "#76406b", size = 8/.pt, inherit.aes = F) + 
 		scale_color_manual(values = c("#3588d1", "#76406b")) + 
 		scale_y_continuous(limits = c(1.5, 8.5),

### 2021-12-09 code/plot_figure_S3.R

@@ -120,8 +120,8 @@ day_0_moribund_feature_abundance_plot <- day_0_moribund %>%
 	ggplot(aes(x = taxa_label, y = relative_abundance, color = early_euth)) + 
 		stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5),
 								 position = position_dodge(width = 0.7)) +
-		geom_jitter(position = position_jitterdodge(dodge.width = 0.7, jitter.width = 0.3), 
-								alpha = 0.2) + 
+		#geom_jitter(position = position_jitterdodge(dodge.width = 0.7, jitter.width = 0.3), 
+		#						alpha = 0.2) + 
 		scale_y_log10() + 
 		theme_bw() + 
 		coord_flip() + 

### 2021-12-09 submission/manuscript.Rmd
update manuscript with PDS feedback

@@
 output:
+  word_document: default
   pdf_document:
@@
     keep_tex: yes
-  word_document: default
 bibliography: references.bib
@@
 
-*Clostridioides difficile* infection (CDI) has become more common and severe in the last few decades.
+*Clostridioides difficile* infection (CDI) disease severity has increased over the last few decades.
 Patient age, white blood cell count, creatinine levels as well as *C.
-difficile* ribotype and the presence of toxin genes have been associated with disease severity.
+difficile* ribotype and toxin genes have been associated with disease severity.
 However, current models of CDI disease severity that incorporate these factors are not robust enough to be broadly applied.
-These models are missing input from the gut environment.
-CDI is dependent on perturbations of the gut bacterial community in order to colonize the gut.
-The gut microbiota also impairs *C.
+These models lack data describing the gut microbiota.
+Perturbations to the gut microbiota are necessary for *C.
+difficile* to colonize the gut.
+The gut microbiota is thought to impair *C.
 difficile* colonization through bile acid metabolism, nutrient consumption and bacteriocin production.
-However, it is unclear if the gut bacteria affect CDI disease severity.
-Here we demonstrate that gut bacterial communities contribute to disease severity variation.
-We derived diverse gut communities by colonizing germ-free mice with human fecal communities.
+It has been unclear whether the gut microbiota affect CDI disease severity.
+However, here we demonstrate that gut bacterial communities can contribute to disease severity.
+We derived diverse gut communities by colonizing germ-free mice with different human fecal communities.
 The mice were then infected with a single *C.
-difficile* RT027 isolate which resulted in mortality and histopathologic differences.
-The severity grouped by the human fecal community they received.
-Generally, facultative anaerobes with pathogenic potential, such as *Escherichia*, *Helicobacter*,  and *Klebsiella*, were associated with more severe disease.
-Bacterial groups associated with dietary fiber degradation, such as *Coprobacillus*, were associated with reduced disease severity.
-Lastly, we showed communities that resulted in either low or high histopathologic scores or severe disease had the same outcome when infected with a different *C.
-difficile* isolate.
+difficile* RT027 clinical isolate which resulted in moribundity and histopathologic differences.
+The variation in severity was associated with the human fecal community that the mice received.
+Generally, facultative anaerobes with pathogenic potential, such as *Escherichia*, *Helicobacter*,  and *Klebsiella*, were associated with more severe outcomes.
+Bacterial groups associated with fiber degradation, such as *Coprobacillus*, were associated with less severe outcomes.
+Finally, we showed that, when infected with a different *C.
+difficile* isolate, the outcome of the infection was similar between the mice that received the same fecal donor community.
+These data have demonstrated that the gut bacteria can influence CDI disease severity and may improve clinical models.
 
@@
 
-*Clostridioides difficile* infection (CDI) can be asymptomatic, or range in severity from mild diarrhea to severe outcomes of surgical interventions and even death.
+*Clostridioides difficile* infections (CDI) can be asymptomatic or range in severity from mild diarrhea to severe outcomes including surgical interventions and death.
 Models that predict severity and guide treatment decisions are based on clinical factors and *C.
 difficile* characteristics.
-The gut microbiome plays a role in potecting against CDI, though has not been investigated for its effect on CDI disease severity.
-We demonstrate that variation in the microbiome of mice colonized with human feces was sufficient to result in a range of disease severities.
-The results reveal groups of bacteria associated with both severe and more mild *C.
-difficile* infections.
-Gut bacterial community data from patients with CDI could improve our ability to identify patients at risk for severe disease and improve interventions which treat the gut bacteria to reduce host damage.
+Although the gut microbiome plays a role in protecting against CDI, it has not been investigated for its effect on CDI disease severity or incorporated into attempts to predict disease severity.
+We demonstrated that variation in the microbiome of mice colonized with human feces was sufficient to result in a range of disease outcomes.
+These results revealed groups of bacteria associated with both severe and mild *C.
+difficile* infection outcomes.
+Gut bacterial community data from patients with CDI could improve our ability to identify patients at risk of developing more severe disease and improve interventions which treat the gut bacteria to reduce host damage.
 
@@
 
-*Clostridioides difficile* infections (CDI) have been increasing in incidence and severity since *C.
-difficile* was first identified as the cause of antibiotic-associated psuedomembranous colitis [@Kelly2008].
-Disease severity can range from asymptomatic to mild diarrhea to toxic megacolon and death.
+*Clostridioides difficile* infections (CDI) have increased in incidence and severity since *C.
+difficile* was first identified as the cause of antibiotic-associated pseudomembranous colitis [@Kelly2008].
+CDI disease severity can range from asymptomatic to mild diarrhea to toxic megacolon and death.
 Treatment for CDI is driven by the disease severity.
-The Infectious Diseases Society of America (IDSA) and Society for Healthcare Epidemiology of America (SHEA) guidelines categorize severe disease based on white blood cell count >15,000 cells/mm^3 and/or a serum creatinine >1.5 mg/dL, and complicated disease based on systolic blood pressure <90, ileus or toxic megacolon [@McDonald2018].
+The Infectious Diseases Society of America (IDSA) and Society for Healthcare Epidemiology of America (SHEA) guidelines categorize severe disease based on white blood cell count greater than 15,000 cells/mm$^3$ and/or a serum creatinine greater than 1.5 mg/dL, and complicated disease based on systolic blood pressure below 90, ileus, or toxic megacolon [@McDonald2018].
 Since these measures are only present once the disease has advanced, the IDSA classification does not allow for identification of patients at risk of severe CDI.
 Thus, we have limited ability to prevent patients from developing severe CDI.
-Ideally, we could identify patients that are high-risk of severe disease use that to adjust their treatments to halt the disease before it worsens.
+Ideally, we could identify patients that are high-risk of severe disease and adjust their treatments to halt the disease before it worsens.
 
 Prediction models have been developed to score a patients risk for severe CDI outcomes but have not been robust for broad application [@Dieterle2020; @Butt2013; @Perry2021; @vanBeurden2017].
-Firstly, there is no widely accepted definition of severe disease [@Ressler2021].
+There is no widely accepted definition of severe disease [@Ressler2021].
 While the IDSA guideline is frequently used, many studies develop their own severity definitions.
@@
 However, a recent multi-center external validation of these scoring systems using a consistent severity definition - transfer or admission to the intensive care unit, colectomy, and/or death within 30 days of diagnosis - demonstrated a decline in performance and high false positivity rate [@Perry2021].
-Thus, it may be necessary to incorporate information in addition to patient and clinical factors to accurately idenitfy patients at risk of severe CDI.
+Thus, patient and clinical factors have not been sufficient to accurately identify patients at risk of severe CDI and may benefit from additional information involved in the development of CDI disease.
 
-CDI disease severity is the result of damage to the host.
 Severity has been associated with characteristics of *C.
@@
 Increased neutrophil inflitration into the mucosa after the cellular tight junctions have been compromised can exacerbate the inflammatory response [@Farooq2015].
-While we have have elucidated these major drivers of host damage in CDI, these do not explain the observed variation in disease severity.
+While we have have elucidated these major drivers of host damage in CDI, these are also the drivers of features of the poorly performing prediction models.
 
-Missing from our CDI disease model are the effects of gut bacteria.
+Missing from CDI severity prediction models are the effects of the indigenous gut bacteria.
 *C.
-difficile* interacts with the intestinal community in many ways.
+difficile* interacts with the gut community in many ways.
 The indigenous bacteria of a healthy intestinal community provide a protective barrier preventing *C.
@@
 difficile* through antibiotic production or nutrient consumption [@Aguirre2021; @Kang2019].
-While the interaction between the gut bacteria and *C.
-difficile* is established, the effect the gut bacteria on CDI disease severity is unclear.
-Recent studies that have begun to uncover this relationship but have either used clinical data which does not permit control of the host or *C.
-difficile* variation [@Seekatz2016] or murine experiments which lack variation in the bacterial communities [@Dieterle2020].
+While the relationship between the gut bacteria and *C.
+difficile* has been established, the effect the gut bacteria can have on CDI disease severity is unclear.
+Recent studies have begun to uncover this relationship but have either used clinical data, which does not permit control of the host or *C.
+difficile* variation [@Seekatz2016] or murine experiments, which lack variation in the bacterial communities [@Dieterle2020].
 Recently, Tomkovich et al.
-showed the variation in the bacterial communities between mice from different mouse colonies and vendors resulted in different clearance rates of *C.
+showed the variation in the bacterial communities between mice from different mouse colonies resulted in different clearance rates of *C.
 difficile* [@Tomkovich2020].
-Therefore it is clear the gut bacteria modulate the activity and infection dynmaics of *C.
-difficile*, we hypothesized that gut bacteria also contribute to the variation in CDI disease severity.
-Here, we colonized germ-free C57BL/6 mice with human fecal communities to test if the variablity in the gut bacteria can produce and explain variation in the CDI disease severity.
+Therefore, it is clear the gut bacteria modulate the activity and infection dynmaics of *C.
+difficile*.
+We hypothesized that gut bacteria also contribute to the variation in CDI disease severity.
+Here, we colonized germ-free C57BL/6 mice with human fecal communities to test whether the variability in the gut bacteria can produce and explain variation in disease severity.
 
@@
 ***C.
-difficile* is able to infect germ-free mice colonized with human fecal microbial communities without perturbation.**  Based on our previous observation that variation in the murine microbiota associated with different *C.
-difficile* colonization dynamics [@Tomkovich2020], we sought to explore how variation in murine microbial communities affects the outcome of *C.
-difficile* infection.
-To produce communities with greater variation than mouse colonies with their indigenous communties, we used human fecal communities to establish a microbial community in the intestine of germ-free mice [@Nagpal2018].
+difficile* is able to infect germ-free mice colonized with human fecal microbial communities without perturbation.**  To produce gut microbiomes with greater variation than those found in conventional mouse colonies, we colonized germ-free mice with bacteria from human feces [@Nagpal2018].
 We inoculated germ-free C57BL/6 mice with homogenized feces from one of 15 human fecal samples via oral gavage.
-The human fecal samples selected were chosen to provide diverse communtiy structures and based on community clusters [@Schubert2014].
-The gut community was allowed to establish itself for two weeks post inoculation [@Gillilland2012].
+The human fecal samples were selected because they represented diverse community structures based on community clustering [@Schubert2014].
+The gut communities were allowed to establish themselves for two weeks post inoculation [@Gillilland2012].
 We then surveyed the bacterial members of the gut communities by 16S rRNA gene sequencing of fecal pellets (Figure 1A).
-The bacterial communities from each mouse grouped more closely to those communities from mice that received the same human fecal donor community than to either their respective donor community or any mice who received a fecal homogenate from a different donor (Figure 1B).
-Most of the communities were primairily composed of populations of *Clostridia*, *Bacteroidia*, *Erysipelotrichia*, *Bacilli*, and *Gammaproteobacteria*.
-However, each group of mice harbored unique combination of class abundances in their gut bacterial communities.
+The bacterial communities from each mouse grouped more closely to those communities from mice that received the same human fecal donor community than to the mice who received a different human fecal donor community (Figure 1B).
+The communities were primarily composed of populations of *Clostridia*, *Bacteroidia*, *Erysipelotrichia*, *Bacilli*, and *Gammaproteobacteria*.
+However, each group of mice harbored unique combination of class relative abundances in their gut bacterial communities.
 
@@
 difficile* infection.
-A typical mouse model of CDI requires pretreatment with antibiotics such as clindamycin to become susceptible to *C.
+A typical mouse model of CDI requires pre-treatment with antibiotics, such as clindamycin, to become susceptible to *C.
 difficile* colonization [@Chen2008; @Schubert2015].
-However, we wanted to avoid modifying the communties with an antibiotic to maintain their unique microbial compositions and ecological relationships.
-Since these communties came from people who have increased risk of CDI, such as hospitalization or recent antibiotic use [@Schubert2014], or the communities became susceptible through the murine gut colonization process, we decided to test if *C.
+However, we wanted to avoid modifying the communities with an antibiotic to maintain their unique microbial compositions and ecological relationships.
+Since these communities came from people who have increased risk of CDI, such as hospitalization or recent antibiotic use [@Schubert2014], we decided to test if *C.
 difficile* was able to infect these mice without any antibiotic perturbation.
 We hypothesized that *C.
-difficile* would be able to colonize a subset of mice whose communities were susceptible from perturbations either the donor experienced or from the ecological selection colonizing the murine gut.
-After two weeks post incoluation with human fecal communities, the mice were then challenged with $10^{3}$ *C.
+difficile* would be able to colonize a subset of mice whose communities were susceptible from perturbations either the donor experienced or from the ecological selection when colonizing the murine gut.
+Mice were challenged with 10$^{3}$ *C.
 difficile* ribotype 027 human isolate 431 spores.
 The mice were followed for 10 days post-challenge and their stool was collected and plated for *C.
-difficile* colony forming units (CFU) to detect infection level.
-Suprisingly, communities from all donors were colonized (Figure 2).
+difficile* colony forming units (CFU) to determine the extent of the infection.
+Suprisingly, mice from all donors were colonized (Figure 2).
 Only two mice were able to resist *C.
-difficile* infection and both received their community from the same donor, Donor C.
-The remaining mice demonstrated susceptiblity to *C.
-difficile* infection without the need for any perturbation.
-Thus, *C.
-difficile* infected the mice without needing to perturb the established human-derived gut microbial communities, allowing them to maintain their unique community structures.
+difficile* infection and both received their community from the same donor, N3.
+By colonizing germ-free mice with different human fecal communities, we were able to generate diverse gut communities in mice, which were susceptible to *C.
+difficile* infection without further modification of the gut community.
 
-**Infection severity varies by initial community.** After challenging the mice with *C.
-difficile*, we investigated the outcome from the *C.
-difficile* infection and its relationship to the initial community.
-We followed the mice for 10 days post-challenge for *C.
-difficile* CFU, toxin production, and mortality.
-Seven mice, from Donors A, B and E, were not initally colonized on the day after *C.
-difficile* challenge but were infected ($>10^{6}$) by the end of the experiment.
-All mice that received their community from Donor J through O suffered from severe disease, which became highly colonized one day post-challenge and moribund within 3 days post-challenge.
-The remaining mice, except the Donor C mice which were not infected, maintained *C.
+**Infection severity varies by initial community.** After we challenged the mice with *C.
+difficile*, we investigated the outcome from the infection and its relationship to the initial community.
+We followed the mice for 10 days post-challenge for colonization density, toxin production, and mortality.
+Seven mice, from Donors N1, N2 and N5, were not colonized on the day after *C.
+difficile* challenge but were infected (>10$^{6}$) by the end of the experiment.
+All mice that received their community from Donor M1 through M6 succumbed to the infection and became moribund within 3 days post-challenge.
+The remaining mice, except the uninfected Donor N3 mice, maintained *C.
 difficile* infection through the end of the experiment (Figure 2).
-At 10 days post-challenge, or earlier for the moribund mice, mice were euthanised and cecal tissue was collected and scored for histopathologic signs of disease and fecal pellets were assayed for toxin activity (Figure 3).
+At 10 days post-challenge, or earlier for the moribund mice, mice were euthanised and fecal pellets were assayed for toxin activity and cecal tissue was collected and scored for histopathologic signs of disease (Figure 3).
 Overall, there was greater toxin activity detected in the stool of the moribund mice (*P* = 0.003).
 However, when looking at each group of mice, we saw there was a range in toxin activity for both the moribund and non-moribund mice (Figure 3A).
-Non-moribund mice from Donors D through I had comparable toxin activity as the moribund mice.
+Non-moribund mice from Donors N4 through N9 had comparable toxin activity as the moribund mice.
 Additionally, not all moribund mice had toxin activity detected in their stool.
@@
 Next we examined the cecal tissue for histopathologic damage.
-For moribund mice, we observed high levels of epithelial damage, tissue edema and inflammation (Figure S1), similar to previously reported histopathologic findings for *C.
+Moribund mice had high levels of epithelial damage, tissue edema and inflammation (Figure S1), similar to previously reported histopathologic findings for *C.
 difficile* ribotype 027 [@Cowardin2016].
-As observed with toxin activity, the moribund mice had higher histopathologic scores than the mice with non-severe disease (*P* = 3.0e-9).
+As observed with toxin activity, the moribund mice had higher histopathologic scores than the non-moribund mice (*P* = 3.0e-9).
 However, unlike the toxin activity, all moribund mice had consistently high histopathologic summary scores (Figure 3B).
-For Donor groups A through I, where all mice were non-moribund, we saw a range in tissue damage from no signs of disease to nearly the same level as the mice with severe disease, which grouped by community donor.
-Together, the toxin activity, histopathologic score, and moribundity showed variation across the donors but were largely consistent within each group of mice recieving the same human donor fecal community.
+The non-moribund mice, Donor groups N1 through N9, had a range in tissue damage from none detected to similar levels as the moribund mice, which grouped by community donor.
+Together, the toxin activity, histopathologic score, and moribundity showed variation across the donor groups but were largely consistent within each donor group.
 
-**Microbial community members explain variation in CDI severity.** Since the disease charateristics grouped by human donor fecal community, we next investigated specific bacterial populations that could explain the variation in disease severity.
-To this aim, we interogated the day 0 bacterial community for their relationship to our CDI severity measurements using linear discriminant analysis (LDA) effect size (LEfSe) analysis.
+**Microbial community members explain variation in CDI severity.** We next interrogated the bacterial communities at the time of *C.
+difficile* challenge (day 0) for their relationship to infection outcomes using linear discriminant analysis (LDA) effect size (LEfSe) analysis to identify individual bacterial populations that could explain the variation in disease severity.
 We split the mice into groups by severity level based on their moribundity and histopathologic score.
-We dichotomized the histopathologic scores into high and low bins by splitting on the median score.
+We dichotomized the histopathologic scores into high and low groups by splitting on the median score.
 This analysis revealed 20 genera that were significantly different by the disease severity (Figure 4A).
 Bacterial genera *Turicibacter*, *Streptococcus*, *Staphylococcus*, *Pseudomonas*, *Phocaeicola*, *Parabacteroides*, *Bacteroides*, and *Escherichia/Shigella* were detected at higher relative abundances in the mice that became moribund.
-Populations of *Anaerotignum*, *Coprobacillus*, *Enterocloster*, and *Murimonas* were more abundant in the mice that would develop only low intestinal injury.
+Populations of *Anaerotignum*, *Coprobacillus*, *Enterocloster*, and *Murimonas* were more abundant in the non-moribund mice that would develop only low intestinal injury.
 To understand the role of toxin activity in disease severity, we applied LEfSe to identify the genera most likely to explain the observed differences in toxin activity (Figure 4B).
-Many genera the were associated with toxin production as we would expect from their association with moribundity, such as populations of *Escherichia/Shigella* and *Bacteroides*.
-Likewise, there were genera such as *Anaerotignum*, *Enterocloster*, and *Murimonas* that exhibited greater relative abundance in communities from non-moribund mice with a low histopathologic score and no detected toxin.
-Overall, communities without genera that demonstrate a linear trend between relative abundance and disease severity can be better understood by their association with toxin activity.
-Lastly, we tested for correlations between the histopathologic score and the endpoint relative abundances of bacterial operational taxonomic units (OTUs) (Figure 4C).
+Many genera that associated with toxin production were also associated with moribundity, such as populations of *Escherichia/Shigella* and *Bacteroides*.
+Likewise, there were genera such as *Anaerotignum*, *Enterocloster*, and *Murimonas* that associated with no toxin that also exhibited greater relative abundance in communities from non-moribund mice with a low histopathologic score.
+Lastly, we tested for correlations between the endpoint relative abundances of bacterial operational taxonomic units (OTUs) and the histopathologic summary score (Figure 4C).
 The endpoint relative abundance of *Bacteroides* was positively correlated with histopathologic score, as its day 0 relative abundance did with disease severity (Figure 4A).
 Populations of *Klebsiella* and *Prevotellaceae* were positively correlated with the histopathologic score and were increased in the group of mice with detectable toxin.
-Our analysis have revealed bacterial taxa that are associated with three different measures of CDI severity in our model: moribundity, histopathologic score, and toxin activity.
+This analysis has revealed bacterial genera that associated with the variation in moribundity, histopathologic score, and toxin activity of CDI.
 
-We next wanted to determine whether collectively, bacterial community membership and relative abundance could be predictive of CDI severity.
-We trained random forest models with the day 0 bacterial community relative abundance data at each taxonomic rank to predict toxin production, severe disease or final histopathologic score.
-Overall for predicting toxin production, microbial populations aggregated by their phylum level classificiation performed as well as models using lower taxonomic ranks (AUROC = 0.83, Figure S2).
-Collectively, members of *Verrucomicrobia*, *Campilobacterota* and *Proteobacteria* contributed the most to the correct prediction of toxin production (Figure 5A).
+We next determined whether collectively, bacterial community membership and relative abundance could be predictive of the CDI disease outcome.
+We trained random forest models with bacterial community relative abundance data from the day of colonization at each taxonomic rank to predict toxin production, moribundity, and day 10 post-challenge histopathologic summary score.
+For predicting if toxin would be produced, microbial populations aggregated by phylum rank classification performed similarly as models using lower taxonomic ranks (AUROC = 0.83, Figure S2).
 *C.
-difficile* was more likely to produce toxin when the community it infected had less abundant populations of *Verrucomicrobia* and *Campilobacterota* and had more abundant populations of *Proteobacteria*.
+difficile* was more likely to produce toxin when the community it infected had less abundant populations of *Verrucomicrobia* and *Campilobacterota* and had more abundant populations of *Proteobacteria* (Figure 5A).
 Next, we assessed the ability of the community to predict moribundity.
-Class rank classification was sufficient to predict if the mouse would succumb to the infection before the end of the experiment (AUROC = 0.91, Figure S2).
-Bacterial taxa belonging to *Bacilli*, *Firmicutes*, and *Erysipelotrichia* hasd the greatest effect on model accuracy.
-Larger populations of *Bacilli* and *Firmicutes* and reduced populations of *Erysipelotrichia* were more likely to result in moribundity (Figure 5B).
-The only other class that was decreased in moribund mice was *Clostridia*.
-Lastly, the relative abundances of genera were able to predict a high or low histopathologic score (histopathologic score dichotomized as in previous analysis, AUROC = 0.99, Figure S2).
+Bacteria grouped by class rank classification was sufficient to predict which mice would succumb to the infection before the end of the experiment (AUROC = 0.91, Figure S2).
+THe features with the greatest effect showed that communities with greater populations of bacteria belonging to *Bacilli* and *Firmicutes* and reduced populations of *Erysipelotrichia* were more likely to result in moribundity (Figure 5B).
+Only one other class of bacteria was decreased in moribund mice, a group of unclassified *Clostridia*.
+Lastly, the relative abundances of genera were able to predict a high or low histopathologic score (histopathologic scores were dichotomized as in previous analysis, AUROC = 0.99, Figure S2).
 No genera had a significantly greater effect on the model performance than any others, indicating the model is reliant on many genera for the correct prediction.
@@
 Communities with greater abundances of *Hungatella*, *Eggerthella*, *Bifidobacterium*, *Duncaniella* and *Neisseria* were more likely to have high histopathologic scores.
+These models have shown that the relative abundance of bacterial populations and their relationship to each other could be used to predict the variation in moribundity, histopathologic score, and toxin activity of CDI.
 
@@
 showed clinical disease severity varied by *C.
-difficile* isolate characteristics [@Carlson2013; @Carlson2015], we next wanted to test if the effect of the community on CDI severity is conserved across *C.
+difficile* isolate characteristics, such as germination rate [@Carlson2013; @Carlson2015], we next tested whether the effect of the community on CDI outcomes was conserved across *C.
 difficile* isolates.
-To facilitate this analysis, we selected a subset of donor fecal samples based on their associated mouse transplant communities that fit the following criteria: one community we would expect to result in a low histopathologic score (< 5), a high histopathologic score (> 5), and a high histopathologic score which becomes moribund.
-Additionally, we selected the donors based their relative abundances of genera identified through the our analysis (Figure S3).
-We inoculated mice via oral gavage of the human fecal slurry of either Donor F, G, or M and allowed two weeks for the a community to establish.
+We colonized mice with human fecal slurry of either Donor N7, N6, or M4 and allowed two weeks for the community to establish.
 We then challenged the mice with a different *C.
 difficile* isolate and all mice were infected.
-The moribundity and histopathologic score outcomes from challenge with the second *C.
+The moribundity and histopathologic score outcomes when challenged with a different *C.
 difficile* human isolate were comparable to the outcomes of challenges with *C.
 difficile* ribotype 027 human isolate 431 by their donor group (Figure 6).
+For the isolates and communities we tested, the effect of the gut bacterial community on disease severity was greater than the effect of the variation in these isolates.
 
@@
 Challenging mice colonized with different human fecal communities with a single *C.
-difficile* isolate allowed us to identify the effect of microbiome variation on *C.
+difficile* isolate, ribotype 027 isolate 431, demonstrated the variation in the gut bacteria affects the *C.
 difficile* infection disease severity.
-Our LEfSe analyses and random forest models revealed the relationships between the bacterial community members and the degree of disease severity.
-We used the relative abundance patterns of community members with strong associations with disease severity to select communities to test if the observed disease severity is conserved when a second *C.
-difficile* isolate is used.
-The disease outcome from the *C.
-difficile* chalenge with the new isolate matched the expected disease severities.
-Therefore, the effect of the community variation on disease severity *in vivo* is greater than the effect of *C.
-difficile* isolate phenotype variation [@Carlson2013; @Carlson2015].
-Overall, these results have demonstrated that the bacterial community contributes to the disease severity of *C.
+Our analysis revealed an association between the relative abundance of bacterial community members and disease severity.
+We used the relative abundance patterns of community members associated with disease outcomes to select donor communities to test whether the disease severity was conserved when challenged with a different *C.
+difficile* isolate.
+The disease outcomes from the challenge with the second *C.
+difficile* isolate matched the outcomes of the challenge with *C.
+difficile* ribotype 027 isolate 431.
+Therefore, the effect of the community variation on disease severity *in vivo* was greater than the effect of *C.
+difficile* isolate variation [@Carlson2013; @Carlson2015].
+Overall, these results have demonstrated that the bacterial community contributed to the disease severity of *C.
 difficile* infection.
 
-CDI severity can range from no symptoms, mild symptoms such as diarrhea, or severe symptoms such as toxic megacolon or death.
-Physicians use prediction tools to identify patients most at risk of developing a severe infection using white blood cell counts, serum albumin level, or serum creatine level [@Lungulescu2011; @McDonald2018; @Zar2007].
-Those levels are being driven by the activities in the intestine [@diMasi2018].
+CDI severity can range from no symptoms to mild symptoms, such as diarrhea, to severe symptoms, such as toxic megacolon or death.
+Physicians use prediction tools to identify patients most at risk of developing a severe infection using white blood cell counts, serum albumin level, or serum creatinine level [@Lungulescu2011; @McDonald2018; @Zar2007].
+Those levels are driven by the activities in the intestine [@diMasi2018].
 Research into the drivers of this variation have revealed factors that make *C.
 difficile* more virulent.
-Strains are categorized for their virulence by the presence and production toxin A, toxin B, and binary toxin and the prevelance in outbreaks, such as ribotype 027 and 078 [@AbernathyClose2020; @Theriot2011; @Goorhuis2008; @OConnor2009; @Chen2008].
+Strains are categorized for their virulence by the presence and production toxin TcdA, toxin TcdB, and binary toxin and the prevelance in outbreaks, such as ribotypes 027 and 078 [@AbernathyClose2020; @Theriot2011; @Goorhuis2008; @OConnor2009; @Chen2008].
 However, other studies have shown that virulence is not necessarily linked with toxin production [@Rao2015] or the strain [@Walk2012].
-Furthermore, there is variation in the genome, growth rate, sporulation, germination, and toxin production in different isolates of a strain [@Carlson2013; @He2010].
+Furthermore, there is variation in the genome, growth rate, sporulation, germination, and toxin production in different isolates of a strain [@Carlson2013; @Carlson2015; @He2010].
 This variation may help explain why severe CDI prediction tools often miss identifying many patients with CDI that will develop severe disease [@Dieterle2020; @Butt2013; @Perry2021; @vanBeurden2017].
@@
 
-The state of the intestinal bacterial community determines the ability of *C.
-difficile* to colonize, persist and even cause recurrent infections.
+The state of the gut bacterial community determines the ability of *C.
+difficile* to colonize, persist in the intestine.
 *C.
@@
 Once colonized, the different communities lead to different metabolic responses and dynamics of the *C.
-difficile* population [@Jenior2017, Jenior2018; @Lesniak2021].
+difficile* population [@Jenior2017; @Jenior2018; @Lesniak2021].
 The gut bacteria metabolize primary bile acids into secondary bile acids [@Staley2016; @Long2017].
@@
 difficile* to utilize [@Ng2013; @Ferreyra2014].
-The nutrient environemt affects toxin production [@MartinVerstraete2016].
+The nutrient environment affects toxin production [@MartinVerstraete2016].
 Thus, many of the actions of the gut bacteria modulate *C.
-difficile* in ways that would affect the disease severity driven by CDI.
+difficile* in ways that could affect the infection and resultant disease.
 
 Myriad studies have explored the relationship between the microbiome and CDI but none have established a relationship between the variation of disease severity and the microbiome.
-Most CDI studies employ either an *in vitro* or *in vivo* model using a homogenous bacterial community.
+CDI studies often use an infection model with a single homogeneous bacterial community.
 Collins et al.
@@
 difficile* strains or ribotypes in mice with similar microbiota as a proxy for variation in disease, such as strain 630 for non-severe and ribotype 027 for severe [@AbernathyClose2020; @Theriot2011; @Lawley2012; @Chen2008].
-There have also been studies demonstrating variation in severity through tapering antibiotic dosage [@Lesniak2021; @Schubert2015; @Reeves2011] or by reducing the amount of cells or spores used for the challenge [@Chen2008; @Reeves2011].
-Our group has recently been uncovering how variation in the microbiome affects CDI but have been limited to *C.
-difficile* colonization [@Lesniak2021; @Tomkovich2020].
+Studies have also demonstrated variation in infection through tapering antibiotic dosage [@Lesniak2021; @Schubert2015; @Reeves2011] or by reducing the amount of *C.
+difficile* cells or spores used for the challenge [@Chen2008; @Reeves2011].
+These studies either lack variation in the microbiome or have variation in the *C.
+difficile* infection itself, therefore we cannot make any association between variation in severity and the microbiome.
 
-With our recent observations that the initial community affects the ability of *C.
-difficile* to persist in an environment and the existing research describing the myriad ways the microbiome interacts with *C.
-difficile*, we hypothesized that the microbiome directly and indirectly modulates CDI disease severity.
-Our data demostrate gut bacterial relative abundances associate with variation in toxin production, histopathologic scoring of the cecum and mortality.
-This anaylsis revealed populations of *Akkermansia*, *Anaerostipes*, *Coprobacillus*, *Enterocloster*, *Lactonifactor*, and *Monoglobus* were more abundant in the microbiome of mice which had low histopathologic scores and no mortality or detected toxin.
+With our recent observations that the initial community affected the ability of *C.
+difficile* to persist in the gut [@Tomkovich2020; @Lesniak2021] and the existing research describing the myriad interactions between the microbiome and *C.
+difficile*, we hypothesized that the microbiome can modulate CDI disease severity.
+Our data have demonstrated gut bacterial relative abundances associate with variation in toxin production, histopathologic scoring of the cecal tissue and mortality.
+This analysis revealed populations of *Akkermansia*, *Anaerostipes*, *Coprobacillus*, *Enterocloster*, *Lactonifactor*, and *Monoglobus* were more abundant in the microbiome of non-moribund mice which had low histopathologic scores and no detected toxin.
 The protective role of these genera are supported by previous studies.
 *Coprobacillus*, *Lactonifactor*, and *Monoglobus* have been shown to be involved in dietary fiber fermentation and associated with healthy communities [@Mabrok2011; @Kim2019; @doPrado2021; @Muthuramalingam2019].
-*Anaerostipes* and *Coprobacillus* produce short chain fatty acids which also have been associated with healthy communities [@Han2019; @Duncan2004; @Ye2018].
-Furthermore, *Coprobacillus*, which is abundant in mice with low histopathologic scores but rare in all other mice, has been shown to contain putative type I lantibiotic gene cluster and inhibit *C.
+*Anaerostipes* and *Coprobacillus*, which produce short chain fatty acids, have been associated with healthy communities [@Han2019; @Duncan2004; @Ye2018].
+Furthermore, *Coprobacillus*, which was abundant in mice with low histopathologic scores but rare in all other mice, has been shown to contain a putative type I lantibiotic gene cluster and inhibit *C.
 difficile* colonization [@Walsh2017; @Sandiford2018; @Stein2013].
-*Akkermansia* and *Enterocloster* were also idenitified as more abundant in mice which had a low histopathologic scores but have contradictory supporting evidence in the current literature.
-In our data, *Akkermansia* is most abundant in the non-moribund mice with low histopathologic scores but there were some moribund mice which had increased populations of *Akkermansia*.
+*Akkermansia* and *Enterocloster* were also identified as more abundant in mice which had a low histopathologic scores but have contradictory supporting evidence in the current literature.
+In our data, *Akkermansia* was most abundant in the non-moribund mice with low histopathologic scores but there were some moribund mice which had increased populations of *Akkermansia*.
 This could be attributed to either a more protective mucus layer was present inhibiting colonization [@Stein2013; @Nakashima2021] or mucus consumption by *Akkermansia* could have been crossfeeding *C.
@@
 However, *Enterocloster*, a member of indigenous gut communities, has been involved in infections, such as bacteremia [@Haas2020; @Finegold2005].
-These data demonstrate a population of bacteria has the potential to be either protective or harmful.
-The disease outcome is not likely based on the abundance of a single population of bacteria rather the result of the interactions of many populations.
+These data have exemplified populations of bacteria that have the potential to be either protective or harmful.
+Thus, the disease outcome is not likely based on the abundance of individual populations of bacteria, rather it is the result of the interactions of the community.
 
-The groups of bacteria that were associated with either a higher histopathologic score or moribundity are members of the indigenous gut community that also have been associated with disease, often referred to as opportunitic pathogens.
+The groups of bacteria that were associated with either a higher histopathologic score or moribundity are members of the indigenous gut community that also have been associated with disease, often referred to as opportunistic pathogens.
 Many of the populations with pathogenic potential that associated with worse outcomes are also facultative anaerobes.
 *Enterococcus*, *Klebsiella*, *Shigella/Escherichia*, *Staphylococcus*, and *Streptococcus* have been shown to expand after antibiotic use [@VanInsberghe2020; @GarzaGonzalez2019; @Schubert2014] and are commonly detected in CDI cases [@Shafiq2020; @Keith2020; @Zackular2016; @Berkell2021].
-In addition to these populations, *Eggerthella*, *Prevotellaceae* and *Helicobacter*, associated with worse outcomes have also been associated with intestinal inflammation [@Gardiner2014; @Iljazovic2020; @Nagalingam2013].
-Recently, *Helicobacter hepaticus* has been demostrated to be sufficient to cause susceptibility to CDI in IL-$10^{-/-}$ C57BL/6 mice [@AbernathyClose2021].
-In our experiments, when *Helicobacter* was present, the mouse resulted in a high histopathologic score.
-While we did not use IL-$10^{-/-}$ mice, it is possible the bacterial communtity or host response are similarly modified by *Helicobacter* allowing *C.
-difficile* infection and resulting in host damage.
-Aside from *Helicobacter*, these groups of bacteria associated with more severe outcomes did not have a conserved association between their relative abundance and the disease severity across all mice.
+In addition to these populations, *Eggerthella*, *Prevotellaceae* and *Helicobacter*, which associated with worse outcomes, have also been associated with intestinal inflammation [@Gardiner2014; @Iljazovic2020; @Nagalingam2013].
+Recently, *Helicobacter hepaticus* was shown to be sufficient to cause susceptibility to CDI in IL-10 deficient C57BL/6 mice [@AbernathyClose2021].
+In our experiments, when *Helicobacter* was present, the infection resulted in a high histopathologic score.
+While we did not use IL-10 deficient mice, it is possible the bacterial community or host response are similarly modified by *Helicobacter*, allowing *C.
+difficile* infection and host damage.
+Aside from *Helicobacter*, these groups of bacteria that associated with more severe outcomes did not have a conserved association between their relative abundance and the disease severity across all mice.
 
-The relationship between a group of bacteria and severity may seem to vary, until we evaluate in the context of the environment inwhich there interact.
-Based on our results, it is more appropriate to apply the damage-response framework for microbial pathogenesis to CDI [@Pirofski2008; @Casadevall2014].
+Since we observed groups of bacteria that were associated with less severe disease in some case and either no effect or more severe disease in others, it may be more appropriate to apply the damage-response framework for microbial pathogenesis to CDI [@Pirofski2008; @Casadevall2014].
 Disease is not driven by a single entity, rather it is an emergent property of the responses of the host immune system, infecting microbe, *C.
-difficile*, and the microbes of the indigenous community.
-In the first set of experiments, we use the same host, C57BL/6 mice, the same infecting microbe, *C.
+difficile*, and the indigenous microbes at the site of infection.
+In the first set of experiments, we used the same host, C57BL/6 mice, the same infecting microbe, *C.
 difficile* ribotype 027 human isolate 431, with different gut bacterial communities.
 The bacterial groups in those communities were often present in both moribund and non-moribund and across the range of histopathologic scores.
-Thus, it is not merely, the presence of the bacteria but their activity in response to the other microbes and host which affect the extent of the host damage.
+Thus, it was not merely the presence of the bacteria but their activity in response to the other microbes and host which affect the extent of the host damage.
 Additionally, while each mouse and *C.
-difficile* has the same genetic background, they too are reacting to the specific microbial community.
-Disease severity is driven by the cummulative effect of the host immune response and the activity of the gut bacteria and *C.
-difficile*.
+difficile* population had the same genetic background, they too were reacting to the specific microbial community.
+Disease severity is driven by the cumulative effect of the host immune response and the activity of  *C.
+difficile* and the gut bacteria.
 *C.
-difficile* can be driving host damage through the production of toxin.
-The gut microbiota could be contributing to the host damage through the balance of metabolic and competitive activities, such as bacteriocin production or mucin degradation.
-Low levels of mucin degradation can provide nutrients to other community members producing a diverse non-damaging community [].
-However, if mucin degradation becomes too great it reduces the protective function of the mucin layer and exposes the epithelial cells [].
-This over-harvesting contributes to the host damage due to other members producing toxin [].
+difficile* can drive host damage through the production of toxin.
+The gut microbiota can contribute to the host damage through the balance of metabolic and competitive activities, such as bacteriocin production or mucin degradation.
+For example, low levels of mucin degradation can provide nutrients to other community members producing a diverse non-damaging community [@Tailford2015].
+However, if mucin degradation becomes too great it reduces the protective function of the mucin layer and exposes the epithelial cells.
+This over-harvesting can contribute to the host damage due to other members producing toxin.
 Thus, the resultant intestinal damage is the balance of all activities in the gut environment.
-Host damage is the emergent property of numerous damage-response curves, one for host immune response, one for *C.
+Host damage is the emergent property of numerous damage-response curves, such as one for host immune response, one for *C.
 difficile* activity and another for microbiome community activity, each of which are a composite curve of the individual activities from each group, such as antibody production, neutrophil infiltration, toxin production, sporulation, fiber and mucin degradation.
-Therefore, while we have identified populations of interest, it is likely necessary to target multiple to reduce those with community interactions contributing to host damage.
+Therefore, while we have identified populations of interest, it may be necessary to target multiple types of bacteria to reduce the community interactions contributing to host damage.
 
-Many studies have investigated individual activities contributing to CDI.
-Here we have shown myraid bacterial groups and their relative abundances associate with variation in CDI disease severity.
-We must continue to explore how the host, microbiota, and *C.
-difficile* interactions result in severe disease.
-We can reduce the risk of severity with a better understanding of which interactions, whether driven by specific community functions or groups of bacteria, reduce host damage.
-Approaching this problem from a damage-response framework allows us to experiment with modifying subsets of activities to reduce overall host damage.
+Many studies have investigated individual activities that contribute to CDI.
+Here we have shown myriad bacterial groups and their relative abundances associated with variation in severity of CDI disease.
+We must continue to explore how host, microbiota, and *C.
+difficile* interactions result in severe outcomes.
+We may be able to reduce the risk of severity with a better understanding of which interactions, whether driven by specific community functions or groups of bacteria, reduce host damage.
+Approaching this problem from a damage-response framework would expand the treatment to modifying multiple groups of bacteria and activities to reduce overall host damage.
 Our current clinical treatment for primary CDI targets an individual group of this tripartite system.
@@
 difficile* toxins.
-These treatments are only addressing a single entity of the system, leaving the potential for the others to maintain responses that maintain host damage.
-However, we can continue this research to elucidate treatments that drive all three entities towards reduced host damage.
-When a patient is diagnosed with CDI, the gut community composition, in addition to the traditionally obtained information, may improve our severity prediction and appropriate treatment.
-Treating the microbiome at the time of diagnosis may prevent the infection from becoming more severity.
-Based on the general trends of our results, promoting fiber metabolizing bacteria and reducing facultative anerobes could bolster the efficacy of current CDI treatments, reducing disease severities.
+These treatments are only addressing a single entity of the system, leaving the potential for the others to maintain responses that contribute to host damage.
+However, we may be able to improve CDI treatment through elucidating interventions that drive all three entities towards reduced host damage.
+When a patient is diagnosed with CDI, the gut community composition, in addition to the traditionally obtained clinical information, may improve our severity prediction and guide profilactic treatment.
+Treating the microbiome at the time of diagnosis may prevent the infection from becoming more severe.
+Based on the general trends of our results, promoting fiber metabolizing bacteria and reducing facultative anaerobes may bolster the efficacy of current CDI treatments, reducing disease severity.
 
@@
 All mouse experiments were approved by the University Committee on Use and Care of Animals at the University of Michigan.
-**Were the mice maintained in gnotobiotic isolators throughout the experiment, one per donor?
+**Were the mice maintained in gnotobiotic isolators throughout the experiment?
 Or were they transferred to BSL2 conditions prior to administering the human FMTs?**
@@
 Mice were inoculated with 100 $\mu$l of the fecal homogenate via oral gavage.
+For the set of experiments testing multiple *C.
+difficile* clinical isolates, we selected a subset of donor fecal samples based on their associated mouse transplant communities and the outcomes in the experiments with *C.
+difficile* ribotype 027 isolate 431 that fit the following criteria: one community we would expect to result in a low histopathologic score (< 5), a high histopathologic score (> 5), and a high histopathologic score which becomes moribund.
+Additionally, we selected the donors based their relative abundances of genera identified through the our analysis (Figure S3).
 Two weeks after the fecal community inoculation, mice were challenged with *C.
@@
 
-**Toxin cytotoxicity assay.** Fecal sample remaining after plating for CFU and DNA extraction were collected in a preweighted sterile tube and stored at −80°C.
+**Toxin cytotoxicity assay.** Fecal sample remaining after plating for CFU and DNA extraction were collected in a preweighted sterile tube and stored at -80°C.
 To prepare the sample for the activity assay, samples were diluted 1:10 weight per volume using sterile PBS and then filter sterilized through a 0.22-$\mu$m filter.
@@
 
-**Code availability.** Scripts necessary to reproduce our analysis and this paper are available in an online repository (https://github.com/SchlossLab/Lesniak_Severity_XXXX_2021).
+**Code availability.** Scripts necessary to reproduce our analysis and this paper are available in an online repository (https://github.com/SchlossLab/Lesniak_Severity_XXXX_2022).
 
@@
 Thank you to Sarah Lucas and Sarah Tomkovich for critical discussion in the development and execution of this project.
-We also thank the University of Michigan Germ-free Mouse Core for assitance with our germfree mice, funded in part by U2CDK110768.
+We also thank the University of Michigan Germ-free Mouse Core for assistance with our germfree mice, funded in part by U2CDK110768.
 This work was supported by several grants from the National Institutes for Health R01GM099514, U19AI090871, U01AI12455, and P30DK034933.
@@
 **Figure 1.
-Human fecal microbial communtities establish diverse gut bacterial communties in germ-free mice.** (A) Relative abundace of 10 most abundant bacterial classes of the gut bacterial communities on timepoint day 0, 14 days post-inoculation of germ-free C57Bl/6 mice with human fecal samples.
-Each column of abundances represent an individual mouse.
-Each group of mice that recieved the same donor are group together and labeled with a letter (A through O).
+Human fecal microbial communities established diverse gut bacterial communties in germ-free mice.** (A) Relative abundance of 10 most abundant bacterial classes of the gut bacterial communities on timepoint day 0, 14 days post-colonization of germ-free C57Bl/6 mice with human fecal samples.
+Each column of abundances represents an individual mouse.
+Each group of mice that received the same donor are grouped together and labeled with a letter (N for non-moribund mice and M for moribund mice) and number.
 (B) Median (points) and interquartile range (lines) of $\beta$-diversity ($\theta$~YC~) between an individual mouse and either all others which were inoculated with feces from the same donor or from a different donor.
@@
 difficile* infection but with different outcomes.** *C.
-difficile* CFU per gram of stool measured the day after challenge with $10^{3}$ *C.
+difficile* CFU per gram of stool was measured the day after challenge with $10^{3}$ *C.
 difficile* ribotype 027 human isolate 431 spores and at the end of the experiment, 10 days post-challenge.
 Each point is an individual mouse.
-Mice are grouped by donor and labeled by the donor letter.
+Mice are grouped by donor and labeled by the donor letter (N for non-moribund mice and M for moribund mice) and number.
 Points are colored by donor group.
-Mice from donor groups J through O succumbed to the infection prior to day 10 and were not plated.
+Mice from donor groups N1 through N6 succumbed to the infection prior to day 10 and were not plated on day 10 post-challenge.
 LOD = Limit of detection.
@@
 **Figure 3.
-Histopathologic score and toxin activity varied across donor groups.** (A) Fecal toxin activity detected in some mice post *C.
+Histopathologic score and toxin activity varied across donor groups.** (A) Fecal toxin activity was detected in some mice post *C.
 difficile challenge in both moribund and non-moribund mice.
-(B) Histopathologic summary score for mice at the end of the experiment.
+(B) Cecum scored for histopathologic damage from mice at the end of the experiment.
 Samples were collected for histopathologic scoring on day 10 post-challenge for non-moribund mice or the day the mouse succumbed to the infection for the moribund group (day 2-3 post-challenge).
 Each point is an individual mouse.
-Mice are grouped by donor and labeled by the donor letter.
+Mice are grouped by donor and labeled by the donor letter (N for non-moribund mice and M for moribund mice) and number.
 Points are colored by donor group.
-Missing points were from mice with insufficient sample collected for assaying fecal toxin.
+Missing points are from mice that had insufficient sample collected for assaying fecal toxin.
 
@@
 **Figure 4.
-Individual fecal bacterial community members of the murine gut associate with *C .difficile* infection outcomes.** (A and B) Relative abundance of genera at the time of *C.
-difficile* challenge (Day 0) that vary significantly by the moribundity and histopathologic summary score or toxin activity by LEfSe analysis.
+Individual fecal bacterial community members of the murine gut associated with *C .difficile* infection outcomes.** (A and B) Relative abundance of genera at the time of *C.
+difficile* challenge (Day 0) that varied significantly by the moribundity and histopathologic summary score or toxin activity by LEfSe analysis.
 Median (points) and interquartile range (lines) are plotted.
 Genera are ordered alphabetically to ease comparisons across analyses.
-(A) Comparison across infection outcome of moribund (colored black) or non-moribund with either a high histopathologic score (score greater than the median score of 5, colored green) or a low histopathologic summary score (score less than the median score of 5, colored light green).
-(B) Comparison between mice which toxin activity was detected (Toxigenic, colored dark purple) and mice which no toxin activity was detected (Non-toxigenic, colored light purple).
-(C) Endpoint bacterial OTUs correlate with histopathologic summary score.
+(A) Relative abundances were compared across infection outcome of moribund (colored black) or non-moribund with either a high histopathologic score (score greater than the median score of 5, colored green) or a low histopathologic summary score (score less than the median score of 5, colored light green).
+(B) Relative abundances were compared between mice which toxin activity was detected (Toxigenic, colored dark purple) and which no toxin activity was detected (Non-toxigenic, colored light purple).
+(C) Endpoint bacterial OTUs correlated with histopathologic summary score.
 Each individual mouse is plotted (transparent gray point).
-Spearman correlatations are statistically significant after Benjamini-Hochberg correction for multiple comparisons.
+Spearman correlations were statistically significant after Benjamini-Hochberg correction for multiple comparisons.
 * indicates bacterial group was unclassified at lower classification ranks.
@@
 Fecal bacterial community members of the murine gut at the time of *C.
-difficile* (Day 0) predict outcomes of the infection.** Day 0 bacterial community members grouped by different classification rank were modeled with random forest to predict the infection outcome.
+difficile* (Day 0) predicted outcomes of the infection.** Day 0 bacterial community members grouped by different classification rank were modeled with random forest to predict the infection outcome.
 The models used the highest taxonomic classification rank that performed as well as the lower ranks.
-Median (solid points) and interquartile range (lines) of the group relative abundance and each individual (transparent points) relative abundances are plotted.
-Bacterial groups are ordered by their importance to the model; top taxonomic group has the greatest decrease in performance which its relative abundances are permutted.
+Median (solid points) and interquartile range (lines) of the group relative abundance are plotted.
+Bacterial groups are ordered by their importance to the model; taxonomic group at the top of the plot had the greatest decrease in performance when its relative abundances were permuted.
 * indicates bacterial group was unclassified at lower classification ranks.
-(A) Bacterial members grouped by phyla predicted which mice would have toxin activity detected at any point thoughout the infection (Toxigenic, dark purple).
+(A) Bacterial members grouped by phyla predicted which mice would have toxin activity detected at any point throughout the infection (Toxigenic, dark purple).
 (B) Bacterial members grouped by class predicted which mice would become moribund (dark blue).
@@
 difficile* infection outcome were similar within human fecal donor groups across different *C.
-difficile* isolates.** Mice colonized with human fecal donor communities G, F, or M were challenged with the isolate from the previous experiments in this study, *C.
+difficile* isolates.** Mice colonized with human fecal donor communities N7, N6, or M4 were challenged with the isolate from the previous experiments in this study, *C.
 difficile* ribotype 027 human isolate 431 (blue points), or a different *C.
@@
 *C.
-difficile* CFUs were enumerated at the end of the experiment (day 10 post-challenge) and plotted grouped by donor source (top plot).
-Histopathologic scores from the tissue collected at the endpoint (either day 10 post-challenge or day mice succumbed to infection) were plotted grouped by donor source (bottom plot).
-Mice that received fecal communities from human donor M succumbed to the infection prior to day 10 post-challenge so were not plated for CFU (DECD = deceased).
-RT = ribotype, Isolate = human isolate id, LOD = limit of detection.
+difficile* CFUs were enumerated at the end of the experiment (day 10 post-challenge) and are plotted grouped by donor source (top plot).
+Histopathologic scores from the tissue collected at the endpoint (either day 10 post-challenge or day mice succumbed to infection) are plotted grouped by donor source (bottom plot).
+Mice that had received fecal communities from human donor M4 succumbed to the infection prior to day 10 post-challenge so were not plated for CFU on day 10 post-challenge (DECD = deceased).
+RT = ribotype, Isolate = *C.
+difficile* human isolate id, LOD = limit of detection.
 
@@
 Histopathologic score of tissue damage at the endpoint of the infection.** Tissue collected at the endpoint, either day 10 post-challenge (Non-moribund) or day mice succumbed to infection (Moribund), were scored from histopathologic damage.
-Mice (points) are grouped and colored by their human fecal communtity donor.
-Missing points were from mice with insufficient sample for histopathologic scoring.
+Mice (points) are grouped and colored by their human fecal community donor.
+Missing points are from mice that had insufficient sample for histopathologic scoring.
 
@@
 **Figure S2.
-Random forest models predict outcomes of the *C.
+Random forest models predicted outcomes of the *C.
 difficile* challenge.** (A-C) Taxonomic classification rank model performance.
@@
 The models used the highest taxonomic classification rank that performed as well as the lower ranks.
-Black rectagle highlights classification rank used to model each outome.
+Black rectangle highlights classification rank used to model each outcome.
 (D-F) Model feature importance.
-Bacterial groups are ordered by their deacrease in area under receiver-operator curve (AUC) when its relative abundances are permutted.
-Individual relative abundances were added to F since differences are in AUC outside the interquartile range.
+Bacterial groups are ordered by their decrease in area under receiver-operator curve (AUC) when its relative abundances was permuted.
+Individual relative abundances were added to F since differences in AUC were outside the interquartile range.
 * indicates bacterial group was unclassified at lower classification ranks.
@@
 (C) Histopathologic score modeled which mice would have a high (score greater than the median score of 5) or low (score less than the median score of 5) histopathologic summary score.
-(D) Bacterial phyla which affect the performance of predicting toxin activity when permutted.
-(E) Bacterial classes which affect the performance of predicting moribundity when permutted.
-(D) Bacterial genera which affect the performance of predicting histopathologic score when permutted.
+(D) Bacterial phyla which affected the performance of predicting toxin activity when permuted.
+(E) Bacterial classes which affected the performance of predicting moribundity when permuted.
+(D) Bacterial genera which affected the performance of predicting histopathologic score when permuted.
 
@@
 **Figure S3.
-Selection of communities to test for conserved *C.
-difficile* infection outcome with different *C.
-difficile* isolate.** (A) Relative abundance of genera in mice from previous experiments which received human fecal donor communities F, G, or M.
+Subset of donor communities were selected to test for conserved *C.
+difficile* infection outcome with a different *C.
+difficile* isolate.** (A) Relative abundance of genera in mice from previous experiments which received human fecal donor communities N7, N6, or M4.
 Each column is an individual mouse.
-(B) Features from the random forest model that predicted moribundity using relative abundance of bacteria grouped by their genus classification rank and ordered by their decrease in AUC when each feature is permutted.
+(B) Features from the random forest model that predicted moribundity using relative abundance of bacteria were grouped by their genus classification rank and are ordered by their decrease in AUC when each feature was permuted.
 Median (solid points) and interquartile range (lines) change in AUC are plotted.
-(C) Relative abundance of the most important genera predicting moribundity.
-Median (solid points) and interquartile range (lines) of the group relative abundance and each individual (transparent points) relative abundances are plotted.
+(C) Relative abundance of the most important genera used by the model to predict moribundity.
+Median (solid points) and interquartile range (lines) of the group relative abundances are plotted.
 * indicates bacterial group was unclassified at lower classification ranks.

### 2021-12-09 code/utilities.R

@@ -69,4 +69,5 @@ donor_df <- tibble(
 	human_source = c('DA01324', 'DA00953', 'DA00581', 'DA01134', 'DA10148',
 		'DA00369', 'DA00430', 'DA10093', 'DA10027', 'DA10034', 
 		'DA10082', 'DA00884', 'DA00578', 'DA00431', 'DA01245'),
-	donor_labels = LETTERS[1:15])
+	donor_labels = factor(c(paste0('N', 1:9), paste0('M', 1:6)), 
+			levels = c(paste0('N', 1:9), paste0('M', 1:6))))

### 2021-12-14 code/fix_con2_label.sh
fix sequence label

@@ -1,0 +1,10 @@
+mv data/raw/CON3-2328-D10_S139_L001_R1_001.fastq data/raw/CON2-2328-D10_S139_L001_R1_001.fastq
+mv data/raw/CON3-2328-D10_S139_L001_R2_001.fastq data/raw/CON2-2328-D10_S139_L001_R2_001.fastq
+mv data/raw/CON3-2328-D7_S106_L001_R1_001.fastq data/raw/CON2-2328-D7_S106_L001_R1_001.fastq
+mv data/raw/CON3-2328-D7_S106_L001_R2_001.fastq data/raw/CON2-2328-D7_S106_L001_R2_001.fastq
+mv data/raw/CON3-2328-D9_S127_L001_R1_001.fastq data/raw/CON2-2328-D9_S127_L001_R1_001.fastq
+mv data/raw/CON3-2328-D9_S127_L001_R2_001.fastq data/raw/CON2-2328-D9_S127_L001_R2_001.fastq
+mv data/raw/CON3-2376-D10_S140_L001_R1_001.fastq data/raw/CON2-2376-D10_S140_L001_R1_001.fastq
+mv data/raw/CON3-2376-D10_S140_L001_R2_001.fastq data/raw/CON2-2376-D10_S140_L001_R2_001.fastq
+mv data/raw/CON3-2376-D9_S128_L001_R1_001.fastq data/raw/CON2-2376-D9_S128_L001_R1_001.fastq
+mv data/raw/CON3-2376-D9_S128_L001_R2_001.fastq data/raw/CON2-2376-D9_S128_L001_R2_001.fastq

### 2021-12-14 code/make_files_file.R
remove unused - inocula

@@ -20,12 +20,6 @@ library(tidyverse)
 # read in the names of all fastq available from project
 fastq_list <- list.files(path = 'data/raw/', pattern = 'fastq')
 
-# create dataframe with human source names and numbers
-sources <- read_tsv('data/process/human_source_tidy.tsv',
-		col_type = cols(.default = col_character())) %>% 
-	mutate(source_number = gsub('DA0*','',sample_id)) %>% 
-	select(source = sample_id, source_number)
-
 # separate fastqs by sample type - mock, inoculum, experiment since each has a different name structure
 fastq_df <- tibble(fastq = fastq_list) %>% 
 	mutate(sample_names = gsub('_S.*', '', fastq), # remove sequencing file naming suffixes
@@ -39,23 +33,16 @@ experiment_fastqs <- fastq_df %>%
 	select(sample_names, fastq) %>% 
 	mutate(read = str_extract(fastq, 'R[12]')) %>%
 	spread(read, fastq, fill = 'NA')
-# format inoculum samples into files format and add DA0* prefix
-inoculum_fastqs <- fastq_df %>% 
-	filter(sample_type == 'inoculum') %>% 
-	left_join(sources, by = c('sample_names' = 'source_number')) %>% 
-	select(sample_names = source, fastq) %>% 
-	mutate(read = str_extract(fastq, 'R[12]')) %>%
-	spread(read, fastq, fill = 'NA')
 # format inoculum samples into files format and add S## suffix to differentiate mocks
 mock_fastqs <- fastq_df %>% 
-	filter(sample_type == 'mock') %>% 
+	filter(sample_names %in% c('mock', 'mock1', 'mock2', 'Mock1')) %>% 
 	mutate(sample_names = gsub('.*(S\\d{2,3}).*', 'mock_\\1', fastq)) %>% 
 	select(sample_names, fastq) %>% 
 	mutate(read = str_extract(fastq, 'R[12]')) %>%
 	spread(read, fastq, fill = 'NA')
 
 # replace all - with _  since mothur uses - to split sames
-files_file <- bind_rows(experiment_fastqs, inoculum_fastqs, mock_fastqs) %>% 
+files_file <- bind_rows(experiment_fastqs, mock_fastqs) %>% 
 	mutate(sample_names = gsub('-', '_', sample_names))
 
 # output files file

### 2021-12-14 code/plot_figure_1.R
make taxa labels horizontal

@@ -89,12 +89,11 @@ day_0_plot <- relative_abundance_data %>% # only plot top 10
 			  breaks = c(-2.5, -1, 0, 1, 2), labels = c('', '0.1', '1', '10', '100')) + 
 	  theme_bw() + 
     labs(x = NULL, y = NULL, fill = NULL) + 
-    scale_x_discrete(guide = guide_axis(angle = 45)) + 
     facet_grid(.~donor_labels, scales = 'free_x', space = 'free_x') + 
     theme(axis.ticks.x = element_blank(),
           axis.text.x = element_blank(),
           legend.position = 'bottom',
-          axis.text.y = ggtext::element_markdown(angle = 45),
+          axis.text.y = ggtext::element_markdown(),
           strip.background = element_blank(),
           legend.margin=margin(t=-0.1, r=0, b=-0.1, l=0, unit="cm"),
           legend.key.height = unit(0.3, 'cm'))

### 2021-12-14 code/plot_figure_4.R

@@ -78,7 +78,7 @@ plot_lefse <- function(file_name){
 			scale_y_log10(breaks=c(0.01, 0.1, 1, 10, 100),
 				labels=c("0","0.1","1","10","100")) + 
 			theme(
-				axis.text.y = ggtext::element_markdown(angle = 45),
+				axis.text.y = ggtext::element_markdown(),
 				panel.grid.major.y = element_line(color = 'gray95'),
 				panel.grid.major.x = element_line(color = 'gray85'),
 				panel.grid.minor.x = element_blank(),

### 2021-12-14 code/plot_figure_2.R
label facets with labels to the right as in fig 3

@@ -21,15 +21,13 @@ donor_aes <- donor_df
 #  setup data
 ###############################################################################
 LOD_df <- data.frame(x = 14.5, y = 2)
-DECD_df <- data.frame(x = 12, y = 3, day = as.factor('10 days post-challenge'))
+DECD_df <- data.frame(x = 12, y = 3, day = 10)
 
 cfu_data <- metadata %>% 
 	filter(cdiff_strain == 431,
 		day %in% c(1,10),
 		!is.na(cdiff_cfu)) %>% 
-	mutate(day = ifelse(day == 1, '1 day post-challenge', '10 days post-challenge'),
-		day = factor(day, levels = c('1 day post-challenge', '10 days post-challenge')),
-		cdiff_cfu = ifelse(cdiff_cfu > 0, log10(cdiff_cfu), log10(60))) %>% 
+	mutate(cdiff_cfu = ifelse(cdiff_cfu > 0, log10(cdiff_cfu), log10(60))) %>% 
 	left_join(donor_aes, by = c('human_source'))
 
 ###############################################################################
@@ -41,10 +39,12 @@ cfu_plot <- cfu_data %>%
 			size = .5, color = 'white') + 
 		geom_jitter(width = 0.2) + 
 		scale_color_identity() + 
-		facet_wrap(.~day, ncol = 1) + 
+		facet_grid(day~.) + 
 		scale_y_continuous(limits = c(1.5,8.5),
 			breaks = c(2,4,6,8),
-			labels = c('10^2', '10^4', '10^6', '10^8')) + 
+			labels = c('10^2', '10^4', '10^6', '10^8'),
+			sec.axis = dup_axis(labels = NULL, breaks = c(2,4,6,8),
+                                           name = "Days post-challenge")) + 
 		theme_bw() + 
 		labs(x = 'Donor', y = '*C. difficile* CFU') + 
 		guides(color = 'none') + 
@@ -52,7 +52,7 @@ cfu_plot <- cfu_data %>%
 			axis.title.y = ggtext::element_markdown(),
 			strip.text.y = element_text(angle = 0, size = 12),
 			strip.background = element_blank(),
-			panel.spacing = unit(1, "lines")) + 
+			axis.ticks = element_blank()) + 
 		geom_label(data = LOD_df, aes(x = x, y = y), label = "LOD", 
 			fill = 'white', color = 'white', inherit.aes = F) + 
[…]

### 2021-12-14 code/plot_figure_6.R

@@ -71,7 +71,7 @@ histology_data <- metadata %>%
 cfu_plot <- cfu_data %>% 
 	ggplot(aes(x = isolate, y = cdiff_cfu, color = isolate_431)) + 
 		geom_point(position = position_jitterdodge(jitter.width = 0.1)) + 
-		geom_hline(yintercept = 2, linetype = 'dashed', color = 'lightgray') + 
+		geom_hline(yintercept = 2, linetype = 'dashed', color = 'white') + 
 		geom_label(data = data.frame(x = 1, y = 2, 
 			donor_labels = factor('N7', levels = donor_levels)), aes(x = x, y = y), 
 			label = "LOD", fill = 'white', color = 'white', size = 8/.pt, inherit.aes = F) + 
@@ -87,7 +87,9 @@ cfu_plot <- cfu_data %>%
 		scale_color_manual(values = c("#3588d1", "#76406b")) + 
 		scale_y_continuous(limits = c(1.5, 8.5),
 			breaks = c(2,4,6,8), 
-			labels = c('10^2', '10^4', '10^6', '10^8')) + 
+			labels = c('10^2', '10^4', '10^6', '10^8'),
+			sec.axis = dup_axis(labels = NULL, breaks = c(2,4,6,8),
+                                name = "Day 10 post-challenge")) + 
 		facet_grid(.~donor_labels, scale = 'free_x', margins = c(1, 2, 2)) + 
 		theme_bw() + 
 		theme(legend.position = 'none',
@@ -95,15 +97,14 @@ cfu_plot <- cfu_data %>%
 			strip.background = element_rect(fill = 'white'),
 			axis.text.y = ggtext::element_markdown(),
 			axis.title.y = ggtext::element_markdown(),
-			axis.ticks.x = element_blank(),
-			axis.text.x = element_text(#hjust = c(1, .5),
-				size = 8),
+			axis.ticks = element_blank(),
+			axis.text.x = element_blank(),
 			legend.title = ggtext::element_markdown(),
 			plot.title = ggtext::element_markdown(hjust = 0.5),
-			panel.grid.major = element_blank(),
+			#panel.grid.major = element_blank(),
 			panel.grid.minor = element_blank()) + 
 		labs(x = NULL, y = '*C. difficile* CFU', 
-			title = 'Day 10 post-challenge', subtitle = 'Donor source',
+			subtitle = 'Donor source',
 			color = '*C. difficile*	
[…]

### 2021-12-14 code/tidy_raw_data.R
update script to use utilities.R

@@ -17,18 +17,19 @@
 ################################################################################
 
 # load packages
-library(tidyverse)
 library(readxl)
+source('code/utilities.R')
 
 # file inputs/outputs
-input_metadata <- 'data/raw/humanGF_ids.xlsx'
-input_toxin <- 'data/raw/Alyx_Humice_toxinassay_results.xlsx'
-input_histology <- 'data/raw/histopathology_scores_raw.xlsx'
-input_human_source <- 'data/raw/MIMARKS_cdclinical.xlsx'
-output_metadata <- 'data/process/metadata_tidy.tsv'
-output_toxin <- 'data/process/toxin_tidy.tsv'
-output_histology <- 'data/process/histology_tidy.tsv'
-output_human_source <- 'data/process/human_source_tidy.tsv'
+input_metadata <- here('data/raw/humanGF_ids.xlsx')
+input_toxin <- here('data/raw/Alyx_Humice_toxinassay_results.xlsx')
+input_histology <- here('data/raw/histopathology_scores_raw.xlsx')
+input_human_source <- here('data/raw/MIMARKS_cdclinical.xlsx')
+output_metadata <- here('data/process/metadata_tidy.tsv')
+output_toxin <- here('data/process/toxin_tidy.tsv')
+output_histology <- here('data/process/histology_tidy.tsv')
+output_human_source <- here('data/process/human_source_tidy.tsv')
+output_supp_table1 <- here('submission/Table_S1.csv')
 
 # Read in metadata
 metadata <- read_xlsx(input_metadata, 
@@ -75,13 +76,16 @@ human_source_data <- read_xlsx(input_human_source,
 		sheet = 'Sheet1', range = 'A1:AW353') %>% 
 	filter(sample_id %in% c(unique(metadata$human_source), 
 							'DA00299', 'DA00395', 'DA00458')) %>% 
-	select(sample_id, biome, samp_mat_process, age, race, gender, 
+	select(sample_id, biome, samp_mat_process, age, gender, 
 		   recent_antibiotic_use = "antibiotics >3mo", protonpump, h2receptor, 
 		   antacid, Healthworker, historyCdiff, Surgery6mos, Vegetarian, weight, 
-		   disease_stat)
+		   disease_stat) %>% 
+	left_join(select(donor_df, human_source, plot_labels = donor_labels),
+		by = c('sample_id' = 'human_source'))
 
 #output tidied data
-write_tsv(metadata, path = output_metadata)
-write_tsv(toxin_data, path = output_toxin)
[…]

### 2021-12-14 submission/manuscript.Rmd
update manuscript with pds edits round 2

@@
 
-# The gut bacterial community potentiates *Clostridiodes difficile* infection severity.
+# The gut bacterial community potentiates *Clostridioides difficile* infection severity.
 
@@
 
-*Clostridioides difficile* infection (CDI) disease severity has increased over the last few decades.
+The severity of *Clostridioides difficile* infections (CDI) has increased over the last few decades.
 Patient age, white blood cell count, creatinine levels as well as *C.
 difficile* ribotype and toxin genes have been associated with disease severity.
-However, current models of CDI disease severity that incorporate these factors are not robust enough to be broadly applied.
-These models lack data describing the gut microbiota.
+However, it is unclear whether there is an associate between the gut microbiota disease severity.
+The gut microbiota is known to interact with *C.
+difficile* during infection.
 Perturbations to the gut microbiota are necessary for *C.
@@
 difficile* colonization through bile acid metabolism, nutrient consumption and bacteriocin production.
-It has been unclear whether the gut microbiota affect CDI disease severity.
-However, here we demonstrate that gut bacterial communities can contribute to disease severity.
+It has not been shown whether the gut microbiota can affect CDI disease severity.
+Here we demonstrate that gut bacterial communities can contribute to disease severity.
 We derived diverse gut communities by colonizing germ-free mice with different human fecal communities.
 The mice were then infected with a single *C.
-difficile* RT027 clinical isolate which resulted in moribundity and histopathologic differences.
+difficile* ribotype 027 clinical isolate which resulted in moribundity and histopathologic differences.
 The variation in severity was associated with the human fecal community that the mice received.
@@
 difficile* isolate, the outcome of the infection was similar between the mice that received the same fecal donor community.
-These data have demonstrated that the gut bacteria can influence CDI disease severity and may improve clinical models.
+These data indicate that gut bacteria can influence CDI disease severity.
 
@@
 Although the gut microbiome plays a role in protecting against CDI, it has not been investigated for its effect on CDI disease severity or incorporated into attempts to predict disease severity.
-We demonstrated that variation in the microbiome of mice colonized with human feces was sufficient to result in a range of disease outcomes.
+We demonstrated that variation in the microbiome of mice colonized with human feces yielded a range of disease outcomes.
 These results revealed groups of bacteria associated with both severe and mild *C.
@@
 CDI disease severity can range from asymptomatic to mild diarrhea to toxic megacolon and death.
-Treatment for CDI is driven by the disease severity.
 The Infectious Diseases Society of America (IDSA) and Society for Healthcare Epidemiology of America (SHEA) guidelines categorize severe disease based on white blood cell count greater than 15,000 cells/mm$^3$ and/or a serum creatinine greater than 1.5 mg/dL, and complicated disease based on systolic blood pressure below 90, ileus, or toxic megacolon [@McDonald2018].
 Since these measures are only present once the disease has advanced, the IDSA classification does not allow for identification of patients at risk of severe CDI.
+Prediction models have been developed to score a patient's risk for severe CDI outcomes based on clinical factors but have not been robust for broad application [@Dieterle2020; @Butt2013; @Perry2021; @vanBeurden2017].
 Thus, we have limited ability to prevent patients from developing severe CDI.
-Ideally, we could identify patients that are high-risk of severe disease and adjust their treatments to halt the disease before it worsens.
-
-Prediction models have been developed to score a patients risk for severe CDI outcomes but have not been robust for broad application [@Dieterle2020; @Butt2013; @Perry2021; @vanBeurden2017].
-There is no widely accepted definition of severe disease [@Ressler2021].
-While the IDSA guideline is frequently used, many studies develop their own severity definitions.
-Aside from the features utilized by the IDSA guidelines, other severity measures use patient demographics, vital information, laboratory test results, disease complications, or infection outcomes such as ICU admission, surgery and/or death [@Dieterle2020; @Zar2007; @Gujja2009; @Jardin2013].
-However, a recent multi-center external validation of these scoring systems using a consistent severity definition - transfer or admission to the intensive care unit, colectomy, and/or death within 30 days of diagnosis - demonstrated a decline in performance and high false positivity rate [@Perry2021].
-Thus, patient and clinical factors have not been sufficient to accurately identify patients at risk of severe CDI and may benefit from additional information involved in the development of CDI disease.
-
-Severity has been associated with characteristics of *C.
-difficile* ribotypes, such as 027, and host factors [@Bauer2011].
-Ribotypes with genes for toxin TcdA/TcdB have the potential to disrupt the tight junctions between the intestinal epithelial cells [@Abt2016].
-The presence of binary toxin CDT or mutations in *tcdC*, the negative regulator of toxin transcription, have been associated with increased virulence of TcdA/TcdB [@Abt2016].
-In addition to the pathogenic potential of *C.
-difficile*, the response by the host also affects the disease severity.
-IgG antibody activity towards the TcdA/TcdB are decreased in both patients with symptomatic and recurrent infections [@Kelly2011].
-In addition to controlling the infection, the immune response can also cause intestinal injury.
-Increased neutrophil inflitration into the mucosa after the cellular tight junctions have been compromised can exacerbate the inflammatory response [@Farooq2015].
-While we have have elucidated these major drivers of host damage in CDI, these are also the drivers of features of the poorly performing prediction models.
+Ideally, we could identify patients that are high-risk of severe disease so their treatments could be adjusted to halt the infection before the disease worsens.
 
@@
 difficile* from infecting the gut.
-A range of mechanisms can disrupt this barrier, including antibiotics, medications, or dietary changes, and lead to increased susceptiblity to CDI [@Britton2014; @Hryckowian2018; @VichVila2020].
+A range of mechanisms can disrupt this barrier, including antibiotics, medications, or dietary changes, and lead to increased susceptibility to CDI [@Britton2014; @Hryckowian2018; @VichVila2020].
 Once in the gut, the indigenous bacteria can either promote or inhibit *C.
-difficile* through producing molecules or modifying the environment [@Abbas2020].
+difficile* through producing molecules or modifying the environment [@Abbas2020; @Jenior2017].
 Bile acids metabolized by the gut bacteria can inhibit *C.
@@
 difficile* has been established, the effect the gut bacteria can have on CDI disease severity is unclear.
-Recent studies have begun to uncover this relationship but have either used clinical data, which does not permit control of the host or *C.
-difficile* variation [@Seekatz2016] or murine experiments, which lack variation in the bacterial communities [@Dieterle2020].
-Recently, Tomkovich et al.
-showed the variation in the bacterial communities between mice from different mouse colonies resulted in different clearance rates of *C.
+
+Studies investigating the severity of CDI disease involving the microbiome have not shown clear evidence of a relationship between the microbiome and disease severity.
+These studies were not setup to identify a relationship between the microbiome and disease severity.
+Studies that have used clinical data cannot control for host, microbiome or *C.
+difficile* ribotype variation [@Seekatz2016].
+While disease severity varies, it cannot be attributed to specifically the microbiome since all factors vary.
+Murine experiments typically use a single mouse colony and different *C.
+difficile* ribotypes to create severity differences [@Dieterle2020].
+Recently, our group has begun uncovering the effect microbiome variation has on *C.
+difficile* infection.
+We showed the variation in the bacterial communities between mice from different mouse colonies resulted in different clearance rates of *C.
 difficile* [@Tomkovich2020].
-Therefore, it is clear the gut bacteria modulate the activity and infection dynmaics of *C.
-difficile*.
-We hypothesized that gut bacteria also contribute to the variation in CDI disease severity.
-Here, we colonized germ-free C57BL/6 mice with human fecal communities to test whether the variability in the gut bacteria can produce and explain variation in disease severity.
+We also showed varied ability of mice to spontaneously eliminate *C.
+difficile* infection when they were treated with different antibiotics prior to *C.
+difficile* challenge [@Lesniak2021].
+However, there have not been any studies investigating whether variation in the microbiome affects the resultant disease severity of *C.
+difficile* infections.
+
+Here, we sought to determine whether the gut bacteria contribute to disease severity.
+We hypothesized that since the gut bacteria affect the metabolism of *C.
+difficile* and its infection dynamics, it can also affect the disease severity of the infection.
+To test this hypothesis, we colonized germ-free C57BL/6 mice with human fecal communities to create varied gut communities.
+We then challenged the mice with a single *C.
+difficile* isolate, ribotype 027 (RT027) human isolate 431, and followed the mice for the development of severe outcomes of moribundity and histopathologic cecal tissue damage.
+Since the murine host and *C.
+difficile* isolate were the same and only the gut community varied, the variation in disease severity we observed was attributable to the microbiome.
 
@@
 The communities were primarily composed of populations of *Clostridia*, *Bacteroidia*, *Erysipelotrichia*, *Bacilli*, and *Gammaproteobacteria*.
-However, each group of mice harbored unique combination of class relative abundances in their gut bacterial communities.
+However, the gut bacterial communities of each donor group of mice harbored unique relative abundance distributions of the shared bacterial classes.
 
@@
 However, we wanted to avoid modifying the communities with an antibiotic to maintain their unique microbial compositions and ecological relationships.
-Since these communities came from people who have increased risk of CDI, such as hospitalization or recent antibiotic use [@Schubert2014], we decided to test if *C.
+Since these communities came from people who had increased risk of CDI, such as hospitalization or recent antibiotic use [@Schubert2014], we decided to test if *C.
 difficile* was able to infect these mice without any antibiotic perturbation.
@@
 Mice were challenged with 10$^{3}$ *C.
-difficile* ribotype 027 human isolate 431 spores.
+difficile* RT027 human isolate 431 spores.
 The mice were followed for 10 days post-challenge and their stool was collected and plated for *C.
 difficile* colony forming units (CFU) to determine the extent of the infection.
-Suprisingly, mice from all donors were colonized (Figure 2).
+Surprisingly, mice from all donors were colonized (Figure 2).
 Only two mice were able to resist *C.
@@
 Overall, there was greater toxin activity detected in the stool of the moribund mice (*P* = 0.003).
-However, when looking at each group of mice, we saw there was a range in toxin activity for both the moribund and non-moribund mice (Figure 3A).
+However, when looking at each group of mice, we observed a range in toxin activity for both the moribund and non-moribund mice (Figure 3A).
 Non-moribund mice from Donors N4 through N9 had comparable toxin activity as the moribund mice.
@@
 Next we examined the cecal tissue for histopathologic damage.
-Moribund mice had high levels of epithelial damage, tissue edema and inflammation (Figure S1), similar to previously reported histopathologic findings for *C.
-difficile* ribotype 027 [@Cowardin2016].
-As observed with toxin activity, the moribund mice had higher histopathologic scores than the non-moribund mice (*P* = 3.0e-9).
+Moribund mice had high levels of epithelial damage, tissue edema, and inflammation (Figure S1) similar to previously reported histopathologic findings for *C.
+difficile* RT027 [@Cowardin2016].
+As observed with toxin activity, the moribund mice had higher histopathologic scores than the non-moribund mice (*P* < 0.001).
 However, unlike the toxin activity, all moribund mice had consistently high histopathologic summary scores (Figure 3B).
@@
 Many genera that associated with toxin production were also associated with moribundity, such as populations of *Escherichia/Shigella* and *Bacteroides*.
-Likewise, there were genera such as *Anaerotignum*, *Enterocloster*, and *Murimonas* that associated with no toxin that also exhibited greater relative abundance in communities from non-moribund mice with a low histopathologic score.
+Likewise, there were genera such as *Anaerotignum*, *Enterocloster*, and *Murimonas* that were associated with no toxin production that also exhibited greater relative abundance in communities from non-moribund mice with a low histopathologic score.
 Lastly, we tested for correlations between the endpoint relative abundances of bacterial operational taxonomic units (OTUs) and the histopathologic summary score (Figure 4C).
@@
 Populations of *Klebsiella* and *Prevotellaceae* were positively correlated with the histopathologic score and were increased in the group of mice with detectable toxin.
-This analysis has revealed bacterial genera that associated with the variation in moribundity, histopathologic score, and toxin activity of CDI.
+This analysis identified bacterial genera that were associated with the variation in moribundity, histopathologic score, and toxin activity of CDI.
 
@@
 Bacteria grouped by class rank classification was sufficient to predict which mice would succumb to the infection before the end of the experiment (AUROC = 0.91, Figure S2).
-THe features with the greatest effect showed that communities with greater populations of bacteria belonging to *Bacilli* and *Firmicutes* and reduced populations of *Erysipelotrichia* were more likely to result in moribundity (Figure 5B).
+The features with the greatest effect showed that communities with greater populations of bacteria belonging to *Bacilli* and *Firmicutes* and reduced populations of *Erysipelotrichia* were more likely to result in moribundity (Figure 5B).
 Only one other class of bacteria was decreased in moribund mice, a group of unclassified *Clostridia*.
 Lastly, the relative abundances of genera were able to predict a high or low histopathologic score (histopathologic scores were dichotomized as in previous analysis, AUROC = 0.99, Figure S2).
-No genera had a significantly greater effect on the model performance than any others, indicating the model is reliant on many genera for the correct prediction.
+No genera had a significantly greater effect on the model performance than any others, indicating the model was reliant on many genera for the correct prediction.
 The model used some of the genera identified in the LEfSe analysis, such as *Coprobacillus*, *Anaerostipes*, and *Hungatella*.
@@
 **Disease severity consistent by donor community across isolates.** Since Carlson et al.
-showed clinical disease severity varied by *C.
-difficile* isolate characteristics, such as germination rate [@Carlson2013; @Carlson2015], we next tested whether the effect of the community on CDI outcomes was conserved across *C.
-difficile* isolates.
+showed *C.
+difficile* isolates with lower germination rates were more likely to result in severe disease [@Carlson2015], we next tested whether the effect of the community on CDI outcomes was conserved across *C.
+difficile* isolates with higher rates of germination.
 We colonized mice with human fecal slurry of either Donor N7, N6, or M4 and allowed two weeks for the community to establish.
@@
 difficile* human isolate were comparable to the outcomes of challenges with *C.
-difficile* ribotype 027 human isolate 431 by their donor group (Figure 6).
+difficile* RT027 human isolate 431 by their donor group (Figure 6).
 For the isolates and communities we tested, the effect of the gut bacterial community on disease severity was greater than the effect of the variation in these isolates.
@@
 Challenging mice colonized with different human fecal communities with a single *C.
-difficile* isolate, ribotype 027 isolate 431, demonstrated the variation in the gut bacteria affects the *C.
+difficile* isolate, RT027 isolate 431, demonstrated that variation in the gut bacteria affects the *C.
 difficile* infection disease severity.
@@
 difficile* isolate matched the outcomes of the challenge with *C.
-difficile* ribotype 027 isolate 431.
+difficile* RT027 isolate 431.
 Therefore, the effect of the community variation on disease severity *in vivo* was greater than the effect of *C.
@@
 difficile* more virulent.
-Strains are categorized for their virulence by the presence and production toxin TcdA, toxin TcdB, and binary toxin and the prevelance in outbreaks, such as ribotypes 027 and 078 [@AbernathyClose2020; @Theriot2011; @Goorhuis2008; @OConnor2009; @Chen2008].
+Strains are categorized for their virulence by the presence and production toxin TcdA, toxin TcdB, and binary toxin and the prevalence in outbreaks, such as ribotypes 027 and 078 [@AbernathyClose2020; @Theriot2011; @Goorhuis2008; @OConnor2009; @Chen2008].
 However, other studies have shown that virulence is not necessarily linked with toxin production [@Rao2015] or the strain [@Walk2012].
-Furthermore, there is variation in the genome, growth rate, sporulation, germination, and toxin production in different isolates of a strain [@Carlson2013; @Carlson2015; @He2010].
+Furthermore, there is variation in the genome, growth rate, sporulation, germination, and toxin production in different isolates of a strain [@Carlson2013; @Carlson2015; @Saund2021; @He2010].
 This variation may help explain why severe CDI prediction tools often miss identifying many patients with CDI that will develop severe disease [@Dieterle2020; @Butt2013; @Perry2021; @vanBeurden2017].
@@
 The state of the gut bacterial community determines the ability of *C.
-difficile* to colonize, persist in the intestine.
+difficile* to colonize and persist in the intestine.
 *C.
@@
 Studies examining difference in disease often use different *C.
-difficile* strains or ribotypes in mice with similar microbiota as a proxy for variation in disease, such as strain 630 for non-severe and ribotype 027 for severe [@AbernathyClose2020; @Theriot2011; @Lawley2012; @Chen2008].
+difficile* strains or ribotypes in mice with similar microbiota as a proxy for variation in disease, such as strain 630 for non-severe and RT027 for severe [@AbernathyClose2020; @Theriot2011; @Lawley2012; @Chen2008].
 Studies have also demonstrated variation in infection through tapering antibiotic dosage [@Lesniak2021; @Schubert2015; @Reeves2011] or by reducing the amount of *C.
 difficile* cells or spores used for the challenge [@Chen2008; @Reeves2011].
-These studies either lack variation in the microbiome or have variation in the *C.
+These studies either lack variation in the initial microbiome or have variation in the *C.
 difficile* infection itself, therefore we cannot make any association between variation in severity and the microbiome.
@@
 difficile* [@Geerlings2018; @Deng2018; @Engevik2020].
-Similarly with *Enterocloster*, in our data this genus was more abundant and associated with low histopathologic scores.
+Similarly, *Enterocloster* was more abundant and associated with low histopathologic scores.
 It has been associated with healthy populations and has been used to mono-colonize germ-free mice to reduce the ability of *C.
 difficile* to colonize [@Reeves2012; @Ma2021].
-However, *Enterocloster*, a member of indigenous gut communities, has been involved in infections, such as bacteremia [@Haas2020; @Finegold2005].
+However, *Enterocloster* has also been involved in infections, such as bacteremia [@Haas2020; @Finegold2005].
 These data have exemplified populations of bacteria that have the potential to be either protective or harmful.
@@
 
-Since we observed groups of bacteria that were associated with less severe disease in some case and either no effect or more severe disease in others, it may be more appropriate to apply the damage-response framework for microbial pathogenesis to CDI [@Pirofski2008; @Casadevall2014].
-Disease is not driven by a single entity, rather it is an emergent property of the responses of the host immune system, infecting microbe, *C.
+Since we observed groups of bacteria that were associated with less severe disease it may be appropriate to apply the damage-response framework for microbial pathogenesis to CDI [@Pirofski2008; @Casadevall2014].
+This framework posits that disease is not driven by a single entity, rather it is an emergent property of the responses of the host immune system, infecting microbe, *C.
 difficile*, and the indigenous microbes at the site of infection.
 In the first set of experiments, we used the same host, C57BL/6 mice, the same infecting microbe, *C.
-difficile* ribotype 027 human isolate 431, with different gut bacterial communities.
+difficile* RT027 human isolate 431, with different gut bacterial communities.
 The bacterial groups in those communities were often present in both moribund and non-moribund and across the range of histopathologic scores.
@@
 
-Many studies have investigated individual activities that contribute to CDI.
 Here we have shown myriad bacterial groups and their relative abundances associated with variation in severity of CDI disease.
-We must continue to explore how host, microbiota, and *C.
-difficile* interactions result in severe outcomes.
-We may be able to reduce the risk of severity with a better understanding of which interactions, whether driven by specific community functions or groups of bacteria, reduce host damage.
-Approaching this problem from a damage-response framework would expand the treatment to modifying multiple groups of bacteria and activities to reduce overall host damage.
-Our current clinical treatment for primary CDI targets an individual group of this tripartite system.
-Most commonly, CDI is treated with an antibiotic to eliminate *C.
-difficile*.
-Alternatively, another treatment uses antibodies to neutralize *C.
-difficile* toxins.
-These treatments are only addressing a single entity of the system, leaving the potential for the others to maintain responses that contribute to host damage.
-However, we may be able to improve CDI treatment through elucidating interventions that drive all three entities towards reduced host damage.
-When a patient is diagnosed with CDI, the gut community composition, in addition to the traditionally obtained clinical information, may improve our severity prediction and guide profilactic treatment.
+Further understanding of how the microbiome affects severity in clinical patients could reduce the amount of adverse outcomes of CDI.
+When a patient is diagnosed with CDI, the gut community composition, in addition to the traditionally obtained clinical information, may improve our severity prediction and guide prophylactic treatment.
 Treating the microbiome at the time of diagnosis may prevent the infection from becoming more severe.
@@
 ***C.
-difficile* experiments.** Human fecal samples were obtained from Schubert et al.
+difficile* experiments.** Human fecal samples were obtained as part of Schubert et al.
 and selected based on community clusters [@Schubert2014] to result in diverse community structures.
@@
 difficile* clinical isolates, we selected a subset of donor fecal samples based on their associated mouse transplant communities and the outcomes in the experiments with *C.
-difficile* ribotype 027 isolate 431 that fit the following criteria: one community we would expect to result in a low histopathologic score (< 5), a high histopathologic score (> 5), and a high histopathologic score which becomes moribund.
+difficile* RT027 isolate 431 that fit the following criteria: one community we would expect to result in a low histopathologic score (< 5), a high histopathologic score (> 5), and a high histopathologic score which becomes moribund.
 Additionally, we selected the donors based their relative abundances of genera identified through the our analysis (Figure S3).
@@
 difficile* isolates came from Carlson et al.
-which had previously been isolated and characterized [@Carlson2013; @Carlson2015].
+which had previously been isolated and characterized [@Carlson2013; @Carlson2015] and have recently been further characterized [@Saund2021].
 Spores concentration were determined both before and after challenge [@Sorg2009].
@@
 
-**Toxin cytotoxicity assay.** Fecal sample remaining after plating for CFU and DNA extraction were collected in a preweighted sterile tube and stored at -80°C.
-To prepare the sample for the activity assay, samples were diluted 1:10 weight per volume using sterile PBS and then filter sterilized through a 0.22-$\mu$m filter.
+**Toxin cytotoxicity assay.** To prepare the sample for the activity assay, fecal material was diluted 1:10 weight per volume using sterile PBS and then filter sterilized through a 0.22-$\mu$m filter.
 Toxin activity was assessed using a Vero cell rounding-based cytotoxicity assay as described previously [@Theriot2011].
@@
 
-**Histopathology evaluation.** Mouse ceca tissue was placed in histopathology cassettes and fixed in 10% formalin, then stored in 70% ethanol.
+**Histopathology evaluation.** Mouse cecal tissue was placed in histopathology cassettes and fixed in 10% formalin, then stored in 70% ethanol.
 McClinchey Histology Labs, Inc.
@@
 
-**Statistical analysis and modeling.** To compare community structures, we calculated Yue and Clayton dissimilarity matrices in mothur [@Yue2005].
-We rarified samples to 2,107 sequences per sample to limit uneven sampling biases.
+**Statistical analysis and modeling.** To compare community structures, we calculated Yue and Clayton dissimilarity matrices ($\theta$~YC~) in mothur [@Yue2005].
+We rarefied samples to 2,107 sequences per sample to limit uneven sampling biases.
 We tested for differences in individual taxonomic groups that would explain the outcome differences with LEfSe [@Segata2011] in mothur.
 Remaining statistical analysis and data visualization was performed in R (v4.0.5) with the tidyverse package (v1.3.1).
-We tested for significant differences in $\beta$-diversity ($\theta$~YC~)using the Wilcoxon rank sum test.
-We used Spearman's corrlation to identified the OTU which had a correlation between their relative abundance and the histopathologic summary score.
+We tested for significant differences in $\beta$-diversity ($\theta$~YC~) using the Wilcoxon rank sum test.
+We used Spearman's correlation to identified which OTUs that had a correlation between their relative abundance and the histopathologic summary score.
 *P* values were then corrected for multiple comparisons with a Benjamini and Hochberg adjustment for a type I error rate of 0.05 [@Benjamini1995].
@@
 Lastly, we did not analyze murine communities to donor community or clinical data because germ-free mice colonized with human fecal communities become more like indigenous murine communities once stabilized.
-Futhermore, we are not trying to make any claims about human bacteria and their association with human CDI.
+Furthermore, we are not trying to make any claims about human bacteria and their association with human CDI.
 
@@
 
-**Sequence data accession number.** All 16S rRNA gene sequence data and associated metadata are available through the Sequence Read Archive via accession XXXX.
+**Sequence data accession number.** All 16S rRNA gene sequence data and associated metadata are available through the Sequence Read Archive via accession PRJNA787941.
 
@@
 **Figure 1.
-Human fecal microbial communities established diverse gut bacterial communties in germ-free mice.** (A) Relative abundance of 10 most abundant bacterial classes of the gut bacterial communities on timepoint day 0, 14 days post-colonization of germ-free C57Bl/6 mice with human fecal samples.
+Human fecal microbial communities established diverse gut bacterial communities in germ-free mice.** (A) Relative abundance of 10 most abundant bacterial classes of the gut bacterial communities on timepoint day 0, 14 days post-colonization of germ-free C57Bl/6 mice with human fecal samples.
 Each column of abundances represents an individual mouse.
@@
 (B) Median (points) and interquartile range (lines) of $\beta$-diversity ($\theta$~YC~) between an individual mouse and either all others which were inoculated with feces from the same donor or from a different donor.
-The $\beta$-diversity amongst in the same donor comparison group was significantly less than the $\beta$-diversity of the different donor group (*P* < 0.05, calculated by Wilcoxon rank sum test).
+The $\beta$-diversity among the same donor comparison group was significantly less than the $\beta$-diversity of the different donor group (*P* < 0.05, calculated by Wilcoxon rank sum test).
 
@@
 difficile* infection but with different outcomes.** *C.
-difficile* CFU per gram of stool was measured the day after challenge with $10^{3}$ *C.
-difficile* ribotype 027 human isolate 431 spores and at the end of the experiment, 10 days post-challenge.
-Each point is an individual mouse.
+difficile* CFU per gram of stool was measured the day after challenge with 10$^{3}$ *C.
+difficile* RT027 human isolate 431 spores and at the end of the experiment, 10 days post-challenge.
+Each point represents an individual mouse.
 Mice are grouped by donor and labeled by the donor letter (N for non-moribund mice and M for moribund mice) and number.
@@
 Histopathologic score and toxin activity varied across donor groups.** (A) Fecal toxin activity was detected in some mice post *C.
-difficile challenge in both moribund and non-moribund mice.
+difficile* challenge in both moribund and non-moribund mice.
 (B) Cecum scored for histopathologic damage from mice at the end of the experiment.
 Samples were collected for histopathologic scoring on day 10 post-challenge for non-moribund mice or the day the mouse succumbed to the infection for the moribund group (day 2-3 post-challenge).
-Each point is an individual mouse.
+Each point represents an individual mouse.
 Mice are grouped by donor and labeled by the donor letter (N for non-moribund mice and M for moribund mice) and number.
@@
 Each individual mouse is plotted (transparent gray point).
-Spearman correlations were statistically significant after Benjamini-Hochberg correction for multiple comparisons.
-* indicates bacterial group was unclassified at lower classification ranks.
+Spearman's correlations were statistically significant after Benjamini-Hochberg correction for multiple comparisons.
+* next to a name indicates that the bacterial group was unclassified at lower classification ranks.
 
@@
 Bacterial groups are ordered by their importance to the model; taxonomic group at the top of the plot had the greatest decrease in performance when its relative abundances were permuted.
-* indicates bacterial group was unclassified at lower classification ranks.
+* next to the name indicates that the bacterial group was unclassified at lower classification ranks.
 (A) Bacterial members grouped by phyla predicted which mice would have toxin activity detected at any point throughout the infection (Toxigenic, dark purple).
@@
 difficile* isolates.** Mice colonized with human fecal donor communities N7, N6, or M4 were challenged with the isolate from the previous experiments in this study, *C.
-difficile* ribotype 027 human isolate 431 (blue points), or a different *C.
+difficile* RT027 human isolate 431 (blue points), or a different *C.
 difficile* isolate (purple points).
@@
 Histopathologic scores from the tissue collected at the endpoint (either day 10 post-challenge or day mice succumbed to infection) are plotted grouped by donor source (bottom plot).
+Each point represents an individual mouse.
 Mice that had received fecal communities from human donor M4 succumbed to the infection prior to day 10 post-challenge so were not plated for CFU on day 10 post-challenge (DECD = deceased).
@@
 Histopathologic score of tissue damage at the endpoint of the infection.** Tissue collected at the endpoint, either day 10 post-challenge (Non-moribund) or day mice succumbed to infection (Moribund), were scored from histopathologic damage.
+Each point represents an individual mouse.
 Mice (points) are grouped and colored by their human fecal community donor.
@@
 difficile* isolate.** (A) Relative abundance of genera in mice from previous experiments which received human fecal donor communities N7, N6, or M4.
-Each column is an individual mouse.
+Each column represents an individual mouse.
 (B) Features from the random forest model that predicted moribundity using relative abundance of bacteria were grouped by their genus classification rank and are ordered by their decrease in AUC when each feature was permuted.
@@
 Median (solid points) and interquartile range (lines) of the group relative abundances are plotted.
-* indicates bacterial group was unclassified at lower classification ranks.
+* next to a name indicates that the bacterial group was unclassified at lower classification ranks.

### 2021-12-14 submission/manuscript.tex
update manuscript with pds edits round 2

@@
 
-\hypertarget{the-gut-microbiota-potentiates-clostridiodes-difficile-infection-severity.}{%
-\section{\texorpdfstring{The gut microbiota potentiates
-\emph{Clostridiodes difficile} infection
-severity.}{The gut microbiota potentiates Clostridiodes difficile infection severity.}}\label{the-gut-microbiota-potentiates-clostridiodes-difficile-infection-severity.}}
+\hypertarget{the-gut-bacterial-community-potentiates-clostridioides-difficile-infection-severity.}{%
+\section{\texorpdfstring{The gut bacterial community potentiates
+\emph{Clostridioides difficile} infection
+severity.}{The gut bacterial community potentiates Clostridioides difficile infection severity.}}\label{the-gut-bacterial-community-potentiates-clostridioides-difficile-infection-severity.}}
 
@@
 Nicholas A. Lesniak\(^1\), Alyxandria M. Schubert\(^1\), Kaitlyn
-Flynn\(^1\), Hamide Sinani\(^1\), Kathyrn A. Eaton\(^1\), Ingrid L.
-Bergin\(^2\), Patrick D. Schloss\(^{1,\dagger}\)
+Flynn\(^1\), Jhansi L Leslie\(^1\), Hamide Sinani\(^1\), Ingrid L.
+Bergin\(^3\), Vincent B Young\(^2\), Patrick D. Schloss\(^{1,\dagger}\)
 
@@
 Ann Arbor, MI\\
-2. Unit for Laboratory Animal Medicine, University of Michigan, Ann
+2. Division of Infectious Diseases, Department of Internal Medicine,
+University of Michigan Medical School, Ann Arbor, MI\\
+3. Unit for Laboratory Animal Medicine, University of Michigan, Ann
 Arbor, MI
@@
 
-\emph{Clostridioides difficile} infection has become more common and
-severe in the last few decades. Patient age, white blood cell count,
-creatinine levels as well as \emph{C. difficile} ribotype and the
-presence of toxin genes have been associated with disease severity.
-However, models based on those data to identify patients with CDI that
-will develop severe disease have not been robust enough to be broadly
-applied. These models are input from our understanding of \emph{C.
-difficile} interactions with the gut environment. The gut bacteria are a
-key determinant in \emph{C. difficile} infection. CDI is dependent on
-perturbations in the bacterial community in order to colonize the gut.
-The gut microbiota also impairs \emph{C. difficile} colonization through
+The severity of \emph{Clostridioides difficile} infections (CDI) has
+increased over the last few decades. Patient age, white blood cell
+count, creatinine levels as well as \emph{C. difficile} ribotype and
+toxin genes have been associated with disease severity. However, it is
+unclear whether there is an associate between the gut microbiota disease
+severity. The gut microbiota is known to interact with \emph{C.
+difficile} during infection. Perturbations to the gut microbiota are
+necessary for \emph{C. difficile} to colonize the gut. The gut
+microbiota is thought to impair \emph{C. difficile} colonization through
 bile acid metabolism, nutrient consumption and bacteriocin production.
-However, it is unclear if the gut bacteria affect the disease severity
-resulting from that colonization. Here we have shown gut bacterial
-communities contribute to disease severity variation. We created diverse
-gut communities by colonizing germ-free mice with human fecal
-communities. They were then infected with \emph{C. difficile} which
-resulted in differed disease severities. The severity grouped by the
-human fecal community they received. Generally, facultative anaerobes
-with pathogenic potential, such as Escherichia, Helicobacter and
-Klebsiella, were associated with more severe disease. Bacterial groups
-associated with dietary fiber degradation, such as Coprobacillus, were
-associated with reduced disease severity. Lastly, we showed communities
-that resulted in either low or high histopathologic scores or severe
-disease had the same outcome when infected with a different \emph{C.
-difficile} isolate.
+It has not been shown whether the gut microbiota can affect CDI disease
+severity. Here we demonstrate that gut bacterial communities can
+contribute to disease severity. We derived diverse gut communities by
+colonizing germ-free mice with different human fecal communities. The
+mice were then infected with a single \emph{C. difficile} ribotype 027
+clinical isolate which resulted in moribundity and histopathologic
+differences. The variation in severity was associated with the human
+fecal community that the mice received. Generally, facultative anaerobes
+with pathogenic potential, such as \emph{Escherichia},
+\emph{Helicobacter}, and \emph{Klebsiella}, were associated with more
+severe outcomes. Bacterial groups associated with fiber degradation,
+such as \emph{Coprobacillus}, were associated with less severe outcomes.
+Finally, we showed that, when infected with a different \emph{C.
+difficile} isolate, the outcome of the infection was similar between the
+mice that received the same fecal donor community. These data indicate
+that gut bacteria can influence CDI disease severity.
 
@@
 
-Clostridioides difficile infection (CDI) causes a range of disease from
-asymptomatic to mild diarrhea to severe outcomes of recurrent infections
-and even death. Models that predict severity and treatment decisions are
-based on clinical factors and \emph{C. difficile} capabilities. But
-currently the gut microbiome, the primary protector against CDI, has not
-been accounted for its effect on CDI disease severity. We demonstrated
-variation in the microbiome of mice colonized with human feces was
-sufficient to result in a range of disease severities. Our results show
-groups of bacteria contribute to the development of disease in \emph{C.
-difficile} infections. Gut bacterial community data from patients with
-CDI could improve our ability to identify patients more at risk for
-severe disease and improve interventions which treat the gut bacteria to
-reduce host damage.
+\emph{Clostridioides difficile} infections (CDI) can be asymptomatic or
+range in severity from mild diarrhea to severe outcomes including
+surgical interventions and death. Models that predict severity and guide
+treatment decisions are based on clinical factors and \emph{C.
+difficile} characteristics. Although the gut microbiome plays a role in
+protecting against CDI, it has not been investigated for its effect on
+CDI disease severity or incorporated into attempts to predict disease
+severity. We demonstrated that variation in the microbiome of mice
+colonized with human feces yielded a range of disease outcomes. These
+results revealed groups of bacteria associated with both severe and mild
+\emph{C. difficile} infection outcomes. Gut bacterial community data
+from patients with CDI could improve our ability to identify patients at
+risk of developing more severe disease and improve interventions which
+treat the gut bacteria to reduce host damage.
 
@@
 
-\emph{Clostridioides difficile} infections (CDI) have been increasing in
+\emph{Clostridioides difficile} infections (CDI) have increased in
 incidence and severity since \emph{C. difficile} was first identified as
-a pathogen. It was first identified in 1931 in a healthy fecal community
-{[}CITATION{]}. But then in the late 1970s, following the introduction
-of broad-spectrum antibiotics, \emph{C. difficile} was identified as the
-cause of antibiotic-associated psuedomembranous colitis {[}CITATION{]}.
-Beginning in the early 2000s, CDI became more problematic with
-increasing incidences and severity {[}CITATION{]}. Disease severities
-can range from asymptomatic to mild diarrhea to toxic megacolon and
-death. This severity has been associated with characteristics of
-\emph{C. difficile} ribotypes and host factors. Over time more
-pathogenic ribotypes have emerged, such as ribotype 027 in the early
-2000s. The presence of the genes for Toxin A/B or fluoroquinolon
-resistence has been associated with more severe disease {[}CITATION{]}.
-Binary toxin, while alone insufficient to cause damage on it own, when
-in the presence of Toxin A/B has been associated with increased
-virulence {[}CITATION{]}. Mutations in tcd C, the negative regulator of
-of toxin transcription, also increased the virulence of \emph{C.
-difficile} {[}CITATION{]}. In addition to the pathogenic potential of
-the infecting strain, the response by the host also affects the disease
-severity. Delayed or reduced production of antibodies IgG and IgA can
-result in more severe disease {[}CITATION{]}. Increased neutrophil
-inflitration can lead to worse outcomes {[}CITATION{]}. In addition to
-the immune response, increased density of the toxin receptor on the host
-cells can increase the severity of the infection {[}CITATION{]}. While
-there is a thorough understanding of many of the mechanisms driving in
-CDI incidence and disease severity, there still is not effective tools
-to reduce the risk of severe disease.
-
-Models have been developed to identify patients at risk for severe CDI
-but have not been robust for broad application (1). These models utilize
-patient related data, such as patient age, white blood cell count, serum
-albumin levels, and creatinine levels, to predict disease severity. In
-their initial publication, these models perform well enough for
-publication. However when Perry et al applied many of the most current
-models to a multi-center external validation, their performance suffered
-and often had more false positives than true positives (3). Thus, there
-seems to be a missing factor contributing to disease severity.
-
-Missing from these models are the gut bacteria which also interact with
-\emph{C. difficile} during the course of the infection. The indigenous
-bacteria of a healthy intestinal community provide a protective barrier
-preventing \emph{C. difficile} from infecting the gut. Only after this
-community is disrupted from a range of pertubations, such as many
-classes of antibiotics, medications, or even dietary changes, can
-\emph{C. difficile} infect the intestine. Once established, the gut
+the cause of antibiotic-associated pseudomembranous colitis (1). CDI
+disease severity can range from asymptomatic to mild diarrhea to toxic
+megacolon and death. The Infectious Diseases Society of America (IDSA)
+and Society for Healthcare Epidemiology of America (SHEA) guidelines
+categorize severe disease based on white blood cell count greater than
+15,000 cells/mm\(^3\) and/or a serum creatinine greater than 1.5 mg/dL,
+and complicated disease based on systolic blood pressure below 90,
+ileus, or toxic megacolon (2). Since these measures are only present
+once the disease has advanced, the IDSA classification does not allow
+for identification of patients at risk of severe CDI. Prediction models
+have been developed to score a patient's risk for severe CDI outcomes
+based on clinical factors but have not been robust for broad application
+(3--6). Thus, we have limited ability to prevent patients from
+developing severe CDI. Ideally, we could identify patients that are
+high-risk of severe disease so their treatments could be adjusted to
+halt the infection before the disease worsens.
+
+Missing from CDI severity prediction models are the effects of the
+indigenous gut bacteria. \emph{C. difficile} interacts with the gut
+community in many ways. The indigenous bacteria of a healthy intestinal
+community provide a protective barrier preventing \emph{C. difficile}
+from infecting the gut. A range of mechanisms can disrupt this barrier,
+including antibiotics, medications, or dietary changes, and lead to
+increased susceptibility to CDI (7--9). Once in the gut, the indigenous
 bacteria can either promote or inhibit \emph{C. difficile} through
-producing molecules or modifying the environment {[}Abbas2020{]}. Bile
-acids metabolized by the gut bacteria can inhibit \emph{C. difficile}
-growth and affect toxin production (5). Bacteria in the gut also can
+producing molecules or modifying the environment (10, 11). Bile acids
+metabolized by the gut bacteria can inhibit \emph{C. difficile} growth
+and affect toxin production (12, 13). Bacteria in the gut also can
 compete more directly with \emph{C. difficile} through antibiotic
-production or nutrient consumption (7). To eliminate CDI, the primary
-treatment is a course of antibiotics. Patients who become reinfected and
-recalitrant to antibiotics, a fecal microbiota transplantion often is
-sufficient to restore the gut community to eliminate \emph{C.
-difficile}. While there has been many studies to investigate the
-interaction between the gut bacteria and \emph{C. difficile} during the
-infection, it has not been demostrated the bacterial community can
-modulate the disease severity of CDI. Recently, Tomkovich et al showed
-the variation in the bacterial communtiies between mice from different
-colonies and vendors was sufficient to cause differences in the temporal
-dynamics of the \emph{C. difficile} infection (9). Since the gut
-bacteria interact with \emph{C. difficile} throughout the infection and
-community difference resulted in different colonization dynamics, we
-hypothesized that the gut bacteria contribute to the variation in CDI
-disease severity. Here, we colonized germ-free C57BL/6 mice with human
-fecal communities to test if the variablity in the gut bacteria can
-produce and explain variation in the disease severity caused by a single
-\emph{C. difficile} isolate.
+production or nutrient consumption (14, 15). While the relationship
+between the gut bacteria and \emph{C. difficile} has been established,
+the effect the gut bacteria can have on CDI disease severity is unclear.
+
+Studies investigating the severity of CDI disease involving the
+microbiome have not shown clear evidence of a relationship between the
+microbiome and disease severity. These studies were not setup to
+identify a relationship between the microbiome and disease severity.
+Studies that have used clinical data cannot control for host, microbiome
+or \emph{C. difficile} ribotype variation (16). While disease severity
+varies, it cannot be attributed to specifically the microbiome since all
+factors vary. Murine experiments typically use a single mouse colony and
+different \emph{C. difficile} ribotypes to create severity differences
+(3). Recently, our group has begun uncovering the effect microbiome
+variation has on \emph{C. difficile} infection. We showed the variation
+in the bacterial communities between mice from different mouse colonies
+resulted in different clearance rates of \emph{C. difficile} (17). We
+also showed varied ability of mice to spontaneously eliminate \emph{C.
+difficile} infection when they were treated with different antibiotics
+prior to \emph{C. difficile} challenge (18). However, there have not
+been any studies investigating whether variation in the microbiome
+affects the resultant disease severity of \emph{C. difficile}
+infections.
+
+Here, we sought to determine whether the gut bacteria contribute to
+disease severity. We hypothesized that since the gut bacteria affect the
+metabolism of \emph{C. difficile} and its infection dynamics, it can
+also affect the disease severity of the infection. To test this
+hypothesis, we colonized germ-free C57BL/6 mice with human fecal
+communities to create varied gut communities. We then challenged the
+mice with a single \emph{C. difficile} isolate, ribotype 027 (RT027)
+human isolate 431, and followed the mice for the development of severe
+outcomes of moribundity and histopathologic cecal tissue damage. Since
+the murine host and \emph{C. difficile} isolate were the same and only
+the gut community varied, the variation in disease severity we observed
+was attributable to the microbiome.
 
@@
 
-\textbf{Germ-free mice colonized with human fecal commmunities have
-diverse and unique community structures.} Based on our previous
-observation mice with variation in their microbiota showed differences
-in colonization of \emph{C. difficile} (9), we sought to explore the
-effect communties with greater variation had on \emph{C. difficile}
-challenge. To produce communities with greater variation than mouse
-colonies with their indigenous communties, we used human fecal
-communities to establish a community in the intestine of germ-free mice.
-We inoculated a cage of germ-free C57BL/6 mice (2-4 mice per cage) with
-homogenized feces from one of 15 different human donors via oral gavage.
-The fecal community was allowed to colonize and stabilize for two weeks
-post inoculation {[}CITATION?{]}. We then assessed the gut bacterial
-communities by sequencing the V4 region of the 16S rRNA gene extracted
-from fecal pellets (Figure 1A). The communities established in the mice
-group most closely to other mice recieving the same human fecal donor
-community than to either their respective donor or any mice who received
-stool from a different donor (Figure 1B). Most of the communities were
-primairily composed of populations of \emph{Clostridia},
+\textbf{\emph{C. difficile} is able to infect germ-free mice colonized
+with human fecal microbial communities without perturbation.} To produce
+gut microbiomes with greater variation than those found in conventional
+mouse colonies, we colonized germ-free mice with bacteria from human
+feces (19). We inoculated germ-free C57BL/6 mice with homogenized feces
+from one of 15 human fecal samples via oral gavage. The human fecal
+samples were selected because they represented diverse community
+structures based on community clustering (20). The gut communities were
+allowed to establish themselves for two weeks post inoculation (21). We
+then surveyed the bacterial members of the gut communities by 16S rRNA
+gene sequencing of fecal pellets (Figure 1A). The bacterial communities
+from each mouse grouped more closely to those communities from mice that
+received the same human fecal donor community than to the mice who
+received a different human fecal donor community (Figure 1B). The
+communities were primarily composed of populations of \emph{Clostridia},
 \emph{Bacteroidia}, \emph{Erysipelotrichia}, \emph{Bacilli}, and
-\emph{Gammaproteobacteria}. However, each group of mice harbored unique
-combination of class abunadances in their gut bacterial communities.
-\emph{Any alpha? compare to murine variation? Number of unique otus?}
-
-\textbf{\emph{C. difficile} is able to colonize mice without
-perturbation.} Our goal was to test the effect of the variation in the
-microbiota on \emph{C. difficile} infection. A typical mouse model of
-CDI requires pretreatment with antibiotics such as clindamycin to become
-susceptible to \emph{C. difficile} colonization {[}CITATION{]}. However,
-we wanted to avoid modifying the communties with an antibiotic to
-maintain their community structures and differences. Since these
-communties came from human donors which may have become susceptible
-through their own exposures, we decided to test if \emph{C. difficile}
-was able to colonize these communities without any perturbation. After
-two weeks post incoluation with human fecal communities, the mice were
-then challenged with 10\^{}\textbf{4?} \emph{C. difficile} ribotype 027
-isolate 431 spores (\textbf{Add schematic?}) (10). The mice were
-followed for 10 days post challenge and their stool was collected and
-plated for \emph{C. difficile} CFU to detect colonization level. We
-hypothesized that \emph{C. difficile} would most likely be able to
-colonize the mice with communities that came from donors which had a
-history of antibiotic use, one of the primary risk factors for
-developing CDI (TABLE, CITATION). Suprisingly, communities from all
-donors were colonized (Figure 2). The only two mice that were able to
-resist \emph{C. difficile} colonization both recieved their community
-from the same donor. Therefore, the transplanted human fecal communties
-were susceptible to \emph{C. difficile} colonization without the need
-for any perturbation.
-
-\textbf{Infection seveity varies by initial community structure.} After
-challenging the mice with \emph{C. difficile}, we looked at the effect
-of the community variation on the outcome from the \emph{C. difficile}
-infection. We followed the mice for signs of disease as well as \emph{C.
-difficile} colonization and toxin production for 10 days post challenge.
-Some mice were not initally colonized on the day after \emph{C.
-difficile} challenge but became colonized with a few days post challenge
-(\textbf{SUPP FIG?}). A subset of mice, all that recieved their comunity
-from one of 6 donors, suffered from severe disease, which became highly
-colonized one day post challenge and moribund within 3 days post
-challenge. The remaining mice, except for one cage, maintained \emph{C.
-difficile} colonization through the end of the experiment (Figure 2).
-After the mice were humanely euthanised at the endpoint, day 10 post
-challenge or earlier if they became moribund, their cecal tissue was
-collected and scored for histological signs of disease and fecal samples
-were assayed for toxin (Figure 3). Overall, there was greater toxin
-activity detected in the stool of the moribund mice (\emph{P} = 0.003).
-However, when looking at each group of mice, we saw there was a range in
-toxin activity for both the mice with non-severe disease and severe
-disease (Figure 3A). Some donor groups that maintained colonization
-without severe disease had similarlly high activity levels as was
-detected in the mice with severe disease. Additionally, some mice with
-severe disease did not have toxin activity detected in their stool.
-These oberserved variations in toxin activity and relation to disease
-severity have been previously characterized (11). Next we examined the
-cecal tissue for histopathological damage. For mice with severe disease,
-we observed high levels of epithelial damage, tissue edema and
-inflammation (Figure SXXXX), similar to previously reported
-histopathologic findings for \emph{C. difficile} ribotype 027 (13). As
-observed with toxin activity, the mice with severe disease had higher
-histopathologic scores than the mice with non-severe disease (\emph{P} =
-3.0e-9). However, unlike the toxin activity, all moribund mice had
-consistently high histopathologic scores. The mice which maintained
-persistent population size, we saw a range in tissue damage from no
-signs of disease to nearly the same level as the mice with severe
-disease, which grouped by community donor. Together, the toxin activity,
-histopathologic score, and moribundity showed variation across the
-donors but were largely consistent within each group of mice recieving
-the same human donor fecal community.
-
-\textbf{Populations of the microbial communities explain variation in
-CDI disease severity.} Since the disease charateristics grouped by human
-donor fecal community, we next investigated the community composition
-for its ability to explain the variation in disease severity. First, we
-tested for associations between taxonomic groups and the outcomes of the
-\emph{C. difficile} infection. We used the linear discriminant analysis
-(LDA) effect size (LEfSe) method to identify differences in the genera
-of the initial bacterial communities between the severity level and
-toxin production. We split the mice into groups by severity level based
-on their if they became moribund (severe), or if their clinical score
-was above 5 (moderate) or below 5 (mild). This analysis revealed 20
-genera that were significantly different by the disease severity (Figure
-4A). The presence of populations of \emph{Turicibacter},
-\emph{Streptococcus}, \emph{Staphylococcus}, \emph{Pseudomonas},
-\emph{Phocaeicola}, \emph{Parabacteroides}, \emph{Bacteroides}, and
-\emph{Escherichia/Shigella} were detected at higher levels in the mice
-that developed severe disease. Populations of \emph{Anaerotignum},
+\emph{Gammaproteobacteria}. However, the gut bacterial communities of
+each donor group of mice harbored unique relative abundance
+distributions of the shared bacterial classes.
+
+Next, we tested this set of mice with their human-derived gut microbial
+communities for susceptibility to \emph{C. difficile} infection. A
+typical mouse model of CDI requires pre-treatment with antibiotics, such
+as clindamycin, to become susceptible to \emph{C. difficile}
+colonization (22, 23). However, we wanted to avoid modifying the
+communities with an antibiotic to maintain their unique microbial
+compositions and ecological relationships. Since these communities came
+from people who had increased risk of CDI, such as hospitalization or
+recent antibiotic use (20), we decided to test if \emph{C. difficile}
+was able to infect these mice without any antibiotic perturbation. We
+hypothesized that \emph{C. difficile} would be able to colonize a subset
+of mice whose communities were susceptible from perturbations either the
+donor experienced or from the ecological selection when colonizing the
+murine gut. Mice were challenged with 10\(^{3}\) \emph{C. difficile}
+RT027 human isolate 431 spores. The mice were followed for 10 days
+post-challenge and their stool was collected and plated for \emph{C.
+difficile} colony forming units (CFU) to determine the extent of the
+infection. Surprisingly, mice from all donors were colonized (Figure 2).
+Only two mice were able to resist \emph{C. difficile} infection and both
+received their community from the same donor, N3. By colonizing
+germ-free mice with different human fecal communities, we were able to
+generate diverse gut communities in mice, which were susceptible to
+\emph{C. difficile} infection without further modification of the gut
+community.
+
+\textbf{Infection severity varies by initial community.} After we
+challenged the mice with \emph{C. difficile}, we investigated the
+outcome from the infection and its relationship to the initial
+community. We followed the mice for 10 days post-challenge for
+colonization density, toxin production, and mortality. Seven mice, from
+Donors N1, N2 and N5, were not colonized on the day after \emph{C.
+difficile} challenge but were infected (\textgreater10\(^{6}\)) by the
+end of the experiment. All mice that received their community from Donor
+M1 through M6 succumbed to the infection and became moribund within 3
+days post-challenge. The remaining mice, except the uninfected Donor N3
+mice, maintained \emph{C. difficile} infection through the end of the
+experiment (Figure 2). At 10 days post-challenge, or earlier for the
+moribund mice, mice were euthanised and fecal pellets were assayed for
+toxin activity and cecal tissue was collected and scored for
+histopathologic signs of disease (Figure 3). Overall, there was greater
+toxin activity detected in the stool of the moribund mice (\emph{P} =
+0.003). However, when looking at each group of mice, we observed a range
+in toxin activity for both the moribund and non-moribund mice (Figure
+3A). Non-moribund mice from Donors N4 through N9 had comparable toxin
+activity as the moribund mice. Additionally, not all moribund mice had
+toxin activity detected in their stool. This discordance between toxin
+activity and disease severity has been previously described (24, 25).
+Next we examined the cecal tissue for histopathologic damage. Moribund
+mice had high levels of epithelial damage, tissue edema, and
+inflammation (Figure S1) similar to previously reported histopathologic
+findings for \emph{C. difficile} RT027 (26). As observed with toxin
+activity, the moribund mice had higher histopathologic scores than the
+non-moribund mice (\emph{P} \textless{} 0.001). However, unlike the
+toxin activity, all moribund mice had consistently high histopathologic
+summary scores (Figure 3B). The non-moribund mice, Donor groups N1
+through N9, had a range in tissue damage from none detected to similar
+levels as the moribund mice, which grouped by community donor. Together,
+the toxin activity, histopathologic score, and moribundity showed
+variation across the donor groups but were largely consistent within
+each donor group.
+
+\textbf{Microbial community members explain variation in CDI severity.}
+We next interrogated the bacterial communities at the time of \emph{C.
+difficile} challenge (day 0) for their relationship to infection
+outcomes using linear discriminant analysis (LDA) effect size (LEfSe)
+analysis to identify individual bacterial populations that could explain
+the variation in disease severity. We split the mice into groups by
+severity level based on their moribundity and histopathologic score. We
+dichotomized the histopathologic scores into high and low groups by
+splitting on the median score. This analysis revealed 20 genera that
+were significantly different by the disease severity (Figure 4A).
+Bacterial genera \emph{Turicibacter}, \emph{Streptococcus},
+\emph{Staphylococcus}, \emph{Pseudomonas}, \emph{Phocaeicola},
+\emph{Parabacteroides}, \emph{Bacteroides}, and
+\emph{Escherichia/Shigella} were detected at higher relative abundances
+in the mice that became moribund. Populations of \emph{Anaerotignum},
 \emph{Coprobacillus}, \emph{Enterocloster}, and \emph{Murimonas} were
-increased in the mice that would experience mild disease. There were
-many genera that did not have a correlation of relative abundance and
-disease severity, such as the populations of \emph{Monoglobus} and
-\emph{Hungatella}, which were decreased or increased, respectively, in
-only the group of mice that would develop moderate disease. To
-understand these cases better we used LEfSe to identify the genera
-differentiating the communities that we detected toxin in from those
-that we did not (Figure 4B). There were many genera the were associated
-with toxin production as we would expect from their association with
-severe disease, such as populations of \emph{Escherichia/Shigella} and
+more abundant in the non-moribund mice that would develop only low
+intestinal injury. To understand the role of toxin activity in disease
+severity, we applied LEfSe to identify the genera most likely to explain
+the observed differences in toxin activity (Figure 4B). Many genera that
+associated with toxin production were also associated with moribundity,
+such as populations of \emph{Escherichia/Shigella} and
 \emph{Bacteroides}. Likewise, there were genera such as
 \emph{Anaerotignum}, \emph{Enterocloster}, and \emph{Murimonas} that
-were higher in communities that had mild disease and no toxin was
-produced. However, communities without a linear trend between relative
-abundance and disease severity can be better understood by their
-association with toxin production. Populations of \emph{Hungatella} were
-increased in the group with moderate disease but not with the production
-of toxin. Lastly, we looked for associations between the histopathologic
-score and the populations of genera at that same time (Figure 4C). The
-populatons of \emph{Bacteroides} at the end matched the trend observed
-in the initial community, increased population correlated with increased
-histopathologic score. Populations of \emph{Klebsiella} and
-\emph{Prevotellaceae} positively correlated with clinic score and were
-increased in the group that produced toxin. These tests have identified
-which individual populations of bacteria that associated with each
-disease severity.
-
-We next wanted to understand how the populations in the context of each
-other lead to the different disease severities. We trained random forest
-models with the intital bacterial community relative abundance data at
-each taxonomic rank to predict toxin production, severe disease or final
-histopathologic score. Overall for predicting toxin production,
-microbial populations aggregated by their phylum level classificiation
-performed as well as models using lower taxonomic ranks (AUROC = 0.83,
-Figure S2). In the initial community populations of
-\emph{Verrucomicrobia}, \emph{Campilobacterota} and
-\emph{Proteobacteria} contributed the most to the correct prediction of
-toxin production (Figure 5A). \emph{C. difficile} was more likely to
-produce toxin when the initial community had smaller populations of
-\emph{Verrucomicrobia} and \emph{Campilobacterota} and had larger
-popluations of \emph{Proteobacteria}. Next, we assessed the ability of
-the community to predict severe disease. Class rank classification was
-sufficient to predict if the mouse would succumb to the infection before
-the end of the experiment (AUROC = 0.91, Figure S2). Groups of bacteria
-belonging to \emph{Bacilli}, \emph{Firmicutes}, and
-\emph{Erysipelotrichia} contributed the most to the correct prediction.
-Larger populations of \emph{Bacilli} and \emph{Firmicutes} and reduced
-populations of \emph{Erysipelotrichia} were more likely to result in
-moribundity (Figure 5B). Aside from \emph{Erysipelotrichia}, the only
-other population decreased in moribund mice was \emph{Clostridia}, all
-other features had increased populations in the moribund mice. Lastly,
-the initial community was able to predict if the mice would have a
-histopathologic score above or below the median score (median
-histopathologic score = 5) with the genera relative abundances (AUROC =
-0.99, Figure S2). There were no genera with a significantly greater
-effect on the model performance than any others, indicating the model is
+were associated with no toxin production that also exhibited greater
+relative abundance in communities from non-moribund mice with a low
+histopathologic score. Lastly, we tested for correlations between the
+endpoint relative abundances of bacterial operational taxonomic units
+(OTUs) and the histopathologic summary score (Figure 4C). The endpoint
+relative abundance of \emph{Bacteroides} was positively correlated with
+histopathologic score, as its day 0 relative abundance did with disease
+severity (Figure 4A). Populations of \emph{Klebsiella} and
+\emph{Prevotellaceae} were positively correlated with the
+histopathologic score and were increased in the group of mice with
+detectable toxin. This analysis identified bacterial genera that were
+associated with the variation in moribundity, histopathologic score, and
+toxin activity of CDI.
+
+We next determined whether collectively, bacterial community membership
+and relative abundance could be predictive of the CDI disease outcome.
+We trained random forest models with bacterial community relative
+abundance data from the day of colonization at each taxonomic rank to
+predict toxin production, moribundity, and day 10 post-challenge
+histopathologic summary score. For predicting if toxin would be
+produced, microbial populations aggregated by phylum rank classification
+performed similarly as models using lower taxonomic ranks (AUROC = 0.83,
+Figure S2). \emph{C. difficile} was more likely to produce toxin when
+the community it infected had less abundant populations of
+\emph{Verrucomicrobia} and \emph{Campilobacterota} and had more abundant
+populations of \emph{Proteobacteria} (Figure 5A). Next, we assessed the
+ability of the community to predict moribundity. Bacteria grouped by
+class rank classification was sufficient to predict which mice would
+succumb to the infection before the end of the experiment (AUROC = 0.91,
+Figure S2). The features with the greatest effect showed that
+communities with greater populations of bacteria belonging to
+\emph{Bacilli} and \emph{Firmicutes} and reduced populations of
+\emph{Erysipelotrichia} were more likely to result in moribundity
+(Figure 5B). Only one other class of bacteria was decreased in moribund
+mice, a group of unclassified \emph{Clostridia}. Lastly, the relative
+abundances of genera were able to predict a high or low histopathologic
+score (histopathologic scores were dichotomized as in previous analysis,
+AUROC = 0.99, Figure S2). No genera had a significantly greater effect
+on the model performance than any others, indicating the model was
 reliant on many genera for the correct prediction. The model used some
-of the genera identified in the LEfSe analysis, such as populations of
+of the genera identified in the LEfSe analysis, such as
 \emph{Coprobacillus}, \emph{Anaerostipes}, and \emph{Hungatella}.
-Communities with greater populations of \emph{Hungatella},
+Communities with greater abundances of \emph{Hungatella},
 \emph{Eggerthella}, \emph{Bifidobacterium}, \emph{Duncaniella} and
-\emph{Neisseria} were more likely to have greater histopathologic
-scores.
-
-\textbf{Disease severity consistent by donor community across
-strains/isolates.} We used a single \emph{C. difficile} isolate,
-ribotype 027 isolate 431 to characterize the range of disease severity
-and the features of the bacterial community associated with severity
-(10). Since we had used the same mice and \emph{C. difficile} isolate in
-our first set of experiemnts, we next wanted to test if the effect of
-the community would apply to other \emph{C. difficile} isolates. We
-selected three communities based on the result from our analysis thus
-far to select one community we would expect to result in a low clinical
-score (\textless{} 5), a high clinical score (\textgreater{} 5), and one
-with a high clinical score which becomes moribund. Using genus level
-data to select the communities, we selected communities based on the
-relative abundance patterns of \emph{Sellimonas} and \emph{Anaerotignum}
-(higher abundance in mice with a low clinical score),
-\emph{Lachnospiracea} (lower abundance in moribund mice),
-\emph{Hungatella} and \emph{Eisenbergiella} (higher abundance in
-moderate mice and lower in moribund mice), Clostridium sensu stricto
-(higher in moribund mice by both LEfSe analysis and RF model) and
-\emph{Enterocloster} (lower in moribund mice and a negative linear trend
-with disease severity by LEfSe analysis) (Figure S3). With this set of
-human fecal communities selected, we inoculated mice via oral gavage of
-the human fecal slurry and allowed two weeks for the a community to
-establish. We then challenged the mice with a different \emph{C.
-difficile} (10). The disease severity for this set of experiments
-closely matched that in our first set of experiments with \emph{C.
-difficile} isolate 431 (Figure 6).
+\emph{Neisseria} were more likely to have high histopathologic scores.
+These models have shown that the relative abundance of bacterial
+populations and their relationship to each other could be used to
+predict the variation in moribundity, histopathologic score, and toxin
+activity of CDI.
+
+\textbf{Disease severity consistent by donor community across isolates.}
+Since Carlson et al.~showed \emph{C. difficile} isolates with lower
+germination rates were more likely to result in severe disease (27), we
+next tested whether the effect of the community on CDI outcomes was
+conserved across \emph{C. difficile} isolates with higher rates of
+germination. We colonized mice with human fecal slurry of either Donor
+N7, N6, or M4 and allowed two weeks for the community to establish. We
+then challenged the mice with a different \emph{C. difficile} isolate
+and all mice were infected. The moribundity and histopathologic score
+outcomes when challenged with a different \emph{C. difficile} human
+isolate were comparable to the outcomes of challenges with \emph{C.
+difficile} RT027 human isolate 431 by their donor group (Figure 6). For
+the isolates and communities we tested, the effect of the gut bacterial
+community on disease severity was greater than the effect of the
+variation in these isolates.
 
@@
 Challenging mice colonized with different human fecal communities with a
-single \emph{C. difficile} isolate allowed us to identify the effect of
-microbiome variation on \emph{C. difficile} infection disease severity.
-Our LEfSe analyses and random forest models revealed the relationships
-between the bacterial community members and the degree of disease
-severity. We used the relative abundance patterns of community members
-with strong associations with disease severity to select communities to
-test if the observed disease severity is conserved when other \emph{C.
-difficile} isolates are used. The disease outcome from the \emph{C.
-difficile} chalenge with the new isolate matched the expected disease
-severities. Overall, these results have demonstrated that the bacterial
-community is capable of modulating the disease severity of \emph{C.
-difficile} infection.
-
-CDI severity can range from no symptoms, mild symptoms such as diarrhea,
-or severe symptoms such as toxic megacolon or death. Physicians use
-prediction tools to identify patients most at risk of developing a
-severe infection using white blood cell counts, serum albumin level, or
-serum creatine level (14). Those levels are being driven by the
-activities in the intestine (17). Research into the drivers of this
-variation have revealed factors that make \emph{C. difficile} more
-virulent. Strains are categorized for their virulence by the presence
-and production toxin A and toxin B and the prevelance in outbreaks, such
-as ribotype 027 and 078 (18). However, other studies have shown that
-virulence is not necessarily linked with toxin production
-\textbf{\textbf{???}} or the strain \textbf{\textbf{???}}. Furthermore,
-there is variation in the genome, growth rate, sporulation, germination,
-and toxin production in different isolates of a strain (10). This
-variation may help explain why severe CDI prediction tools often miss
-identifying many patients with CDI that will develop severe disease (1).
-Therefore, it is necessary to gain a full understanding of all factors
-contributing to disease variation to improve our ability to predict
-severity.
-
-The state of the intestinal bacterial community determines the ability
-of \emph{C. difficile} to colonize, persist and even cause recurrent
-infections. \emph{C. difficile} is unable to colonize an unperturbed
-healthy gut community and is only able to entry after a perturbation
-(24). Once colonized, the different communities lead to different
-metabolic responses and dynamics of the \emph{C. difficile} population
-(25). The gut bacteria metabolize primary bile acids into secondary bile
-acids (27). The concentration of these bile acids affect the
-germination, growth, toxin production and biofilm formation (5). Members
-of the bacterial community also affect other metabolites \emph{C.
-difficile} utilizes. \emph{Bacteroides thetaiotaomicron} produce
-sialidases which release sialic acid from the mucosa for \emph{C.
-difficile} to utilize (31). The nutrient environemt affects toxin
-production (33). Thus, many of the actions of the gut bacteria modulate
-\emph{C. difficile} in ways that would affect the disease severity
-driven by CDI.
-
-The gut community has not been demostrated to modulate CDI disease
-severity. Myriad studies have explored the relationship between the
-microbiome and CDI but none have ventured into the variation of disease
-severity due to the microbiome. Most CDI studies employ either an
-\emph{in vitro} or \emph{in vivo} model using a homogenous bacterial
-community. Collins et al used multiple human communities to colonize
-mice, however the communities were pooled prior to gavaging into
-germ-free mice (34), resulting in a single community. Studies examining
-difference in disease often use different \emph{C. difficile} strains or
-ribotypes in mice with similar microbiota as a proxy for variation in
-disease, such as strain 630 for non-severe and ribotype 027 for severe
-(18). There has been studies demonstrating variation in severity, but
-through tapering antibiotic dosage (26) or by reducing the amount of
-cells or spores used for the challenge (22). Our group has recently been
-uncovering how variation in the microbiome affects CDI but have been
-limited to \emph{C. difficile} colonization (26).
-
-With our recent observations that the initial community affects the
-ability of \emph{C. difficile} to persist in an environment and the
-existing research describing the myriad ways the microbiome interacts
-with \emph{C. difficile}, we hypothesized that the microbiome directly
-and indirectly modulates CDI disease severity. Our data demostrate gut
+single \emph{C. difficile} isolate, RT027 isolate 431, demonstrated that
+variation in the gut bacteria affects the \emph{C. difficile} infection
+disease severity. Our analysis revealed an association between the
+relative abundance of bacterial community members and disease severity.
+We used the relative abundance patterns of community members associated
+with disease outcomes to select donor communities to test whether the
+disease severity was conserved when challenged with a different \emph{C.
+difficile} isolate. The disease outcomes from the challenge with the
+second \emph{C. difficile} isolate matched the outcomes of the challenge
+with \emph{C. difficile} RT027 isolate 431. Therefore, the effect of the
+community variation on disease severity \emph{in vivo} was greater than
+the effect of \emph{C. difficile} isolate variation (27, 28). Overall,
+these results have demonstrated that the bacterial community contributed
+to the disease severity of \emph{C. difficile} infection.
+
+CDI severity can range from no symptoms to mild symptoms, such as
+diarrhea, to severe symptoms, such as toxic megacolon or death.
+Physicians use prediction tools to identify patients most at risk of
+developing a severe infection using white blood cell counts, serum
+albumin level, or serum creatinine level (2, 29, 30). Those levels are
+driven by the activities in the intestine (31). Research into the
+drivers of this variation have revealed factors that make \emph{C.
+difficile} more virulent. Strains are categorized for their virulence by
+the presence and production toxin TcdA, toxin TcdB, and binary toxin and
+the prevalence in outbreaks, such as ribotypes 027 and 078 (22, 32--35).
+However, other studies have shown that virulence is not necessarily
+linked with toxin production (36) or the strain (37). Furthermore, there
+is variation in the genome, growth rate, sporulation, germination, and
+toxin production in different isolates of a strain (27, 28, 38, 39).
+This variation may help explain why severe CDI prediction tools often
+miss identifying many patients with CDI that will develop severe disease
+(3--6). Therefore, it is necessary to gain a full understanding of all
+factors contributing to disease variation to improve our ability to
+predict severity.
+
+The state of the gut bacterial community determines the ability of
+\emph{C. difficile} to colonize and persist in the intestine. \emph{C.
+difficile} is unable to colonize an unperturbed healthy gut community
+and is only able to become established after a perturbation (23). Once
+colonized, the different communities lead to different metabolic
+responses and dynamics of the \emph{C. difficile} population (11, 18,
+40). The gut bacteria metabolize primary bile acids into secondary bile
+acids (41, 42). The concentration of these bile acids affect the
+germination, growth, toxin production and biofilm formation (12, 13, 43,
+44). Members of the bacterial community also affect other metabolites
+\emph{C. difficile} utilizes. \emph{Bacteroides thetaiotaomicron}
+produce sialidases which release sialic acid from the mucosa for
+\emph{C. difficile} to utilize (45, 46). The nutrient environment
+affects toxin production (47). Thus, many of the actions of the gut
+bacteria modulate \emph{C. difficile} in ways that could affect the
+infection and resultant disease.
+
+Myriad studies have explored the relationship between the microbiome and
+CDI but none have established a relationship between the variation of
+disease severity and the microbiome. CDI studies often use an infection
+model with a single homogeneous bacterial community. Collins et al.~used
+multiple human communities to colonize mice, however the communities
+were pooled prior to gavaging into germ-free mice (48), resulting in a
+single community. Studies examining difference in disease often use
+different \emph{C. difficile} strains or ribotypes in mice with similar
+microbiota as a proxy for variation in disease, such as strain 630 for
+non-severe and RT027 for severe (22, 32, 33, 49). Studies have also
+demonstrated variation in infection through tapering antibiotic dosage
+(18, 23, 50) or by reducing the amount of \emph{C. difficile} cells or
+spores used for the challenge (22, 50). These studies either lack
+variation in the initial microbiome or have variation in the \emph{C.
+difficile} infection itself, therefore we cannot make any association
+between variation in severity and the microbiome.
+
+With our recent observations that the initial community affected the
+ability of \emph{C. difficile} to persist in the gut (17, 18) and the
+existing research describing the myriad interactions between the
+microbiome and \emph{C. difficile}, we hypothesized that the microbiome
+can modulate CDI disease severity. Our data have demonstrated gut
 bacterial relative abundances associate with variation in toxin
-production, histopathologic scoring of the cecum and severe disease.
-This anaylsis has revealed populations of \emph{Akkermansia},
+production, histopathologic scoring of the cecal tissue and mortality.
+This analysis revealed populations of \emph{Akkermansia},
 \emph{Anaerostipes}, \emph{Coprobacillus}, \emph{Enterocloster},
 \emph{Lactonifactor}, and \emph{Monoglobus} were more abundant in the
-microbiome of mice which little to no disease was detected. These genera
-providing a protective role are supported by previous studies.
-\emph{Coprobacillus}, \emph{Lactonifactor}, and \emph{Monoglobus} have
-been shown to be involved in dietary fiber fermentation and associated
-with healthy communities (37). \emph{Anaerostipes} and
-\emph{Coprobacillus} produce short chain fatty acids which also have
-been associated with healthy communities (41). Furthermore,
-\emph{Coprobacillus}, which is abundant in mice with low histopathologic
-scores but rare in all other mice, has been shwon to contain putative
-type I lantibiotic gene cluster and inhibit \emph{C. difficile}
-colonization (44). \emph{Akkermansia} and \emph{Enterocloster} were also
-idenitified as more abundant in mice which had a low clinical disease
-but have contridictory supporting evidence in the current literature. In
-our data, \emph{Akkermansia} is most abundant in the mice with least
-disease but there were some mice with severe disease which had increased
-populations of \emph{Akkermansia}. This could be attributed to either a
-more protective mucus layer had developed in the mice with little
-disease and \emph{Akkermansia} is inhibiting colonization (46) or
-\emph{Akkermansia} could be crossfeeding \emph{C. difficile} or exposing
-a niche via its mucus consumption (48). Similarly with
-\emph{Enterocloster}, in our data this genus was more abundant and
-associated with low histopathologic scores. It has been associated with
-healthy populations and has been used to mono-colonize germ-free mice to
-reduce the ability of \emph{C. difficile} to colonize (51). However,
-\emph{Enterocloster} has mostly been indentified in infections that
-result in disease (53). These data demostrate a population of bacteria
-has the potential to be either protective or harmful. The disease
-outcome is not likely based on the abundance of a single population of
-bacteria rather the result of the interactions of many populations.
+microbiome of non-moribund mice which had low histopathologic scores and
+no detected toxin. The protective role of these genera are supported by
+previous studies. \emph{Coprobacillus}, \emph{Lactonifactor}, and
+\emph{Monoglobus} have been shown to be involved in dietary fiber
+fermentation and associated with healthy communities (51--54).
+\emph{Anaerostipes} and \emph{Coprobacillus}, which produce short chain
+fatty acids, have been associated with healthy communities (55--57).
+Furthermore, \emph{Coprobacillus}, which was abundant in mice with low
+histopathologic scores but rare in all other mice, has been shown to
+contain a putative type I lantibiotic gene cluster and inhibit \emph{C.
+difficile} colonization (58--60). \emph{Akkermansia} and
+\emph{Enterocloster} were also identified as more abundant in mice which
+had a low histopathologic scores but have contradictory supporting
+evidence in the current literature. In our data, \emph{Akkermansia} was
+most abundant in the non-moribund mice with low histopathologic scores
+but there were some moribund mice which had increased populations of
+\emph{Akkermansia}. This could be attributed to either a more protective
+mucus layer was present inhibiting colonization (60, 61) or mucus
+consumption by \emph{Akkermansia} could have been crossfeeding \emph{C.
+difficile} or exposing a niche for \emph{C. difficile} (62--64).
+Similarly, \emph{Enterocloster} was more abundant and associated with
+low histopathologic scores. It has been associated with healthy
+populations and has been used to mono-colonize germ-free mice to reduce
+the ability of \emph{C. difficile} to colonize (65, 66). However,
+\emph{Enterocloster} has also been involved in infections, such as
+bacteremia (67, 68). These data have exemplified populations of bacteria
+that have the potential to be either protective or harmful. Thus, the
+disease outcome is not likely based on the abundance of individual
+populations of bacteria, rather it is the result of the interactions of
+the community.
 
 The groups of bacteria that were associated with either a higher
-histopathologic score or severe disease are members of the indigenous
-gut community that also have been associated with disease. These are
-often referred to as opportunitic pathogens, however based on the the
-results reported in this study it is more appropriate to apply the
-damage-response framework for microbial pathogenesis (55). The disease
-is not driven by a single entity, rather it is an emergent property of
-the responses of the host immune system, infecting microbe, and the
-microbes of the indigenous community. Many of the populations with
-pathgenic potential that associated with worse outcomes are also
-facultative anaerobes. \emph{Enterococcus}, \emph{Klebsiella},
+histopathologic score or moribundity are members of the indigenous gut
+community that also have been associated with disease, often referred to
+as opportunistic pathogens. Many of the populations with pathogenic
+potential that associated with worse outcomes are also facultative
+anaerobes. \emph{Enterococcus}, \emph{Klebsiella},
 \emph{Shigella/Escherichia}, \emph{Staphylococcus}, and
-\emph{Streptococcus} have been shown to expand their population with
-antibiotic use (57) and are commonly detected in CDI cases (60). In
-addition to these populations, another set of bacteria, populations of
-\emph{Eggerthella}, \emph{Prevotellaceae} and \emph{Helicobacter},
-associated with worse outcomes have also been associated with intestinal
-inflammation (64). Recently, \emph{Helicobacter hepaticus} has been
-demostrated to be sufficient to cause susceptibility to CDI in
-IL-10\^{}(-/-) B57BL/6 mice (67). In our experiments, when
-\emph{Helicobacter} was present, the mouse resulted in either severe
-disease or a high histopathologic score. While we did not use
-IL-10\^{}(-/-) mice, it is possible the bacterial communtity or host
-response are similarly modified by \emph{Helicobacter} allowing \emph{C.
-difficile} infection and resulting in host damage. However, the
-populations identified other than \emph{Helicobacter} are not as clear
-with their relationship between their relative abundance and the
-resulting disease severity.
-
-The damage-response framework for pathogenesis helps to understand how
-bacterial populations identified in this study potentiate CDI disease
-severity through the varied interactions with the host and \emph{C.
-difficile}. In this set of experiments, we use the same host, C57BL/6
-mice, the same infecting microbe, \emph{C. difficile} ribotype 027
-isolate 431, with different gut bacterial communities. The members of
-those communities were often present in multiple levels of disease
-severity. Thus, it is not merely, the presence of the bacteria but their
-activity in response to the other microbes and host which determine
-their effect on host damage. Additionally, while we are using the same
-host and infecting microbe, they too are reacting to the specific
-members of the bacterial community. Disease severity is driven by the
-cummulative effect of the host immune response and the activity of the
-gut bacteria and \emph{C. difficile}. \emph{C. difficile} can be driving
-host damage through the production of toxin. The gut microbiota could be
-contributing to the host damage through the balance of metabolic and
+\emph{Streptococcus} have been shown to expand after antibiotic use (20,
+69, 70) and are commonly detected in CDI cases (71--74). In addition to
+these populations, \emph{Eggerthella}, \emph{Prevotellaceae} and
+\emph{Helicobacter}, which associated with worse outcomes, have also
+been associated with intestinal inflammation (75--77). Recently,
+\emph{Helicobacter hepaticus} was shown to be sufficient to cause
+susceptibility to CDI in IL-10 deficient C57BL/6 mice (78). In our
+experiments, when \emph{Helicobacter} was present, the infection
+resulted in a high histopathologic score. While we did not use IL-10
+deficient mice, it is possible the bacterial community or host response
+are similarly modified by \emph{Helicobacter}, allowing \emph{C.
+difficile} infection and host damage. Aside from \emph{Helicobacter},
+these groups of bacteria that associated with more severe outcomes did
+not have a conserved association between their relative abundance and
+the disease severity across all mice.
+
+Since we observed groups of bacteria that were associated with less
+severe disease it may be appropriate to apply the damage-response
+framework for microbial pathogenesis to CDI (79, 80). This framework
+posits that disease is not driven by a single entity, rather it is an
+emergent property of the responses of the host immune system, infecting
+microbe, \emph{C. difficile}, and the indigenous microbes at the site of
+infection. In the first set of experiments, we used the same host,
+C57BL/6 mice, the same infecting microbe, \emph{C. difficile} RT027
+human isolate 431, with different gut bacterial communities. The
+bacterial groups in those communities were often present in both
+moribund and non-moribund and across the range of histopathologic
+scores. Thus, it was not merely the presence of the bacteria but their
+activity in response to the other microbes and host which affect the
+extent of the host damage. Additionally, while each mouse and \emph{C.
+difficile} population had the same genetic background, they too were
+reacting to the specific microbial community. Disease severity is driven
+by the cumulative effect of the host immune response and the activity of
+\emph{C. difficile} and the gut bacteria. \emph{C. difficile} can drive
+host damage through the production of toxin. The gut microbiota can
+contribute to the host damage through the balance of metabolic and
 competitive activities, such as bacteriocin production or mucin
-degradation. Low levels of mucin degradation can provide nutrients to
-other community members producing a diverse non-damaging community
-{[}{]}. However, if mucin degradation becomes too great it reduces the
-protective function of the mucin layer and exposes the epithelial cells
-{[}{]}. This over-harvesting contributes to the host damage due to other
-members producing toxin. Thus disease is a balance of many activities of
-the community. Host damage is the emergent property of numerous
-damage-response curves, one for host immune response, one for \emph{C.
-difficile} activity and another for microbiome community activity, each
-of which are a composite curve of the individual activities from each
-group, such as antibody production, neutrophil infiltration, toxin
-production, sporulation, fiber and mucin degradation. Therefore, while
-we have identified populations of interest, it is likely targeting one
-specifically will reduce disease severity.
-
-Many studies have investigated individual activities contributing to
-CDI. Here we have shown myraid bacterial groups and their relative
-abundances associate with variation in CDI disease severity. We must
-continue to explore how the host, microbiota, and \emph{C. difficile}
-interactions result in severe disease. We can reduce the risk of
-severity with a better understanding of which interactions, whether be
-driven by specific community functions or groups of bacteria, reduce the
-host damage. Approaching this problem from a damage-response framework
-allows us to experiment with modifying subsets of activities to reduce
-overall host damage. Our current clinical treatment targets an
-individual group in this tripartite system. Most commonly, CDI is
-treated with an antibiotic to eliminate \emph{C. difficile}.
-Alternatively, another treatment uses antibodies to neutralize \emph{C.
-difficile} toxins. These treatments are only addressing a single entity
-of the system, leaving the potential for the others to maintain
-responses that maintain host damage. However, we can continue this
-research to elucidate treatments that drive all three entities towards
-reduced host damage. When a patient is diagnosed with CDI, the gut
-community composition, in addition to the traditionally obtained
-information, may improve our severity prediction and appropriate
-treatment. Treating the microbiome may contribute to reduced severity.
-From our results, promoting fiber metabolizing bacteria and reducing
-facultative anerobes could bolster the efficacy of current CDI
-treatments, reducing disease severities.
+degradation. For example, low levels of mucin degradation can provide
+nutrients to other community members producing a diverse non-damaging
+community (81). However, if mucin degradation becomes too great it
+reduces the protective function of the mucin layer and exposes the
+epithelial cells. This over-harvesting can contribute to the host damage
+due to other members producing toxin. Thus, the resultant intestinal
+damage is the balance of all activities in the gut environment. Host
+damage is the emergent property of numerous damage-response curves, such
+as one for host immune response, one for \emph{C. difficile} activity
+and another for microbiome community activity, each of which are a
+composite curve of the individual activities from each group, such as
+antibody production, neutrophil infiltration, toxin production,
+sporulation, fiber and mucin degradation. Therefore, while we have
+identified populations of interest, it may be necessary to target
+multiple types of bacteria to reduce the community interactions
+contributing to host damage.
+
+Here we have shown myriad bacterial groups and their relative abundances
+associated with variation in severity of CDI disease. Further
+understanding of how the microbiome affects severity in clinical
+patients could reduce the amount of adverse outcomes of CDI. When a
+patient is diagnosed with CDI, the gut community composition, in
+addition to the traditionally obtained clinical information, may improve
+our severity prediction and guide prophylactic treatment. Treating the
+microbiome at the time of diagnosis may prevent the infection from
+becoming more severe. Based on the general trends of our results,
+promoting fiber metabolizing bacteria and reducing facultative anaerobes
+may bolster the efficacy of current CDI treatments, reducing disease
+severity.
 
@@
 
-\textbf{Animal care.} 5. 64 mice, 33F and 31M, ages 6-13 weeks old, 3-4
-mice/group (except 2 for C/DA00581 and MOUT/DA10093) each run once
-except for DA00581 which was run 4 times because we observed
-colonization resistance in one of the experiments 15 human fecal
-samples, selected to create a subset list with diverse responses to
-health questionaire
-
-\textbf{\emph{C. difficile} challenge.} M.
-
-\textbf{Sample collection.} F.
-
-\textbf{DNA sequencing.} T.
-
-\textbf{Sequence curation.} S.
-
-\textbf{Statistical analysis and modeling.} D.
+\textbf{Animal care.} 6- to 13-week old male and female germ-free
+C57BL/6 were obtained from a single breeding colony in the University of
+Michigan Germ-free Mouse Core. Mice were housed in cages of 2-4 mice per
+cage and maintained in germ-free isolators at the University of Michigan
+germ-free facility. All mouse experiments were approved by the
+University Committee on Use and Care of Animals at the University of
+Michigan. \textbf{Were the mice maintained in gnotobiotic isolators
+throughout the experiment? Or were they transferred to BSL2 conditions
+prior to administering the human FMTs?}
+
+\textbf{\emph{C. difficile} experiments.} Human fecal samples were
+obtained as part of Schubert et al.~and selected based on community
+clusters (20) to result in diverse community structures. Feces were
+homogenized by mixing 200 mg of sample with 5 ml of PBS. Mice were
+inoculated with 100 \(\mu\)l of the fecal homogenate via oral gavage.
+For the set of experiments testing multiple \emph{C. difficile} clinical
+isolates, we selected a subset of donor fecal samples based on their
+associated mouse transplant communities and the outcomes in the
+experiments with \emph{C. difficile} RT027 isolate 431 that fit the
+following criteria: one community we would expect to result in a low
+histopathologic score (\textless{} 5), a high histopathologic score
+(\textgreater{} 5), and a high histopathologic score which becomes
+moribund. Additionally, we selected the donors based their relative
+abundances of genera identified through the our analysis (Figure S3).
+Two weeks after the fecal community inoculation, mice were challenged
+with \emph{C. difficile}. \emph{C. difficile} isolates came from Carlson
+et al.~which had previously been isolated and characterized (27, 28) and
+have recently been further characterized (38). Spores concentration were
+determined both before and after challenge (82). \(10^{3}\) \emph{C.
+difficile} spores were given to each mouse via oral gavage. \textbf{How
+were human FMTs prepared (anaerobically vs aerobically, etc)?}
+
+\textbf{Sample collection.} Fecal samples were collected on the day of
+\emph{C. difficile} challenge and the following 10 days. Each day, a
+fecal sample was also collected and weighed and the remaining sample was
+frozen at -20\(^\circ\)C. Anaerobically, the weighed fecal samples were
+serially diluted in PBS, plated on TCCFA plates, and incubated at
+37\(^\circ\)C for 24 hours. The plates were then counted for the number
+of colony forming units (CFU) (83).
+
+\textbf{DNA sequencing.} From the frozen fecal samples, total bacterial
+DNA was extracted using MOBIO PowerSoil-htp 96-well soil DNA isolation
+kit. We amplified the 16S rRNA gene V4 region and sequenced the
+resulting amplicons using an Illumina MiSeq as described previously
+(84).
+
+\textbf{Sequence curation.} Sequences were processed with
+mothur(v.1.44.3) as previously described (84, 85). In short, we used a
+3\% dissimilarity cutoff to group sequences into operational taxonomic
+units (OTUs). We used a naive Bayesian classifier with the Ribosomal
+Database Project training set (version 18) to assign taxonomic
+classifications to each OTU (86). We sequenced a mock community of a
+known community composition and 16s rRNA gene sequences. We processed
+this mock community with our samples to calculate the error rate for our
+sequence curation, which was an error rate of 0.19\%.
+
+\textbf{Toxin cytotoxicity assay.} To prepare the sample for the
+activity assay, fecal material was diluted 1:10 weight per volume using
+sterile PBS and then filter sterilized through a 0.22-\(\mu\)m filter.
+Toxin activity was assessed using a Vero cell rounding-based
+cytotoxicity assay as described previously (33). The cytotoxicity titer
+was determined for each sample as the last dilution, which resulted in
+at least 80\% cell rounding. Toxin titers are reported as the log10 of
+the reciprocal of the cytotoxicity titer.
+
+\textbf{Histopathology evaluation.} Mouse cecal tissue was placed in
+histopathology cassettes and fixed in 10\% formalin, then stored in 70\%
+ethanol. McClinchey Histology Labs, Inc.~(Stockbridge, MI) embedded the
+samples in paraffin, sectioned, and created the hematoxylin and
+eosin-stained slides. The slides were scored using previously described
+criteria by a board-certified veterinary pathologist scored who was
+blinded to the experimental groups (33).
+
+\textbf{Statistical analysis and modeling.} To compare community
+structures, we calculated Yue and Clayton dissimilarity matrices
+(\(\theta\)\textsubscript{YC}) in mothur (87). We rarefied samples to
+2,107 sequences per sample to limit uneven sampling biases. We tested
+for differences in individual taxonomic groups that would explain the
+outcome differences with LEfSe (88) in mothur. Remaining statistical
+analysis and data visualization was performed in R (v4.0.5) with the
+tidyverse package (v1.3.1). We tested for significant differences in
+\(\beta\)-diversity (\(\theta\)\textsubscript{YC}) using the Wilcoxon
+rank sum test. We used Spearman's correlation to identified which OTUs
+that had a correlation between their relative abundance and the
+histopathologic summary score. \emph{P} values were then corrected for
+multiple comparisons with a Benjamini and Hochberg adjustment for a type
+I error rate of 0.05 (89). We built random forest models using the
+mikropml package (90) with OTUs from day 0 samples using mtry values of
+1 through 10, 15, 20, 25, 40, 50, 100. The split for training and
+testing varied by model to avoid overfitting the data. The toxin and
+moribundity models were trained with 60\% of the data. The
+histopathologic score model was trained with 80\% of the data. Lastly,
+we did not analyze murine communities to donor community or clinical
+data because germ-free mice colonized with human fecal communities
+become more like indigenous murine communities once stabilized.
+Furthermore, we are not trying to make any claims about human bacteria
+and their association with human CDI.
 
@@
 and this paper are available in an online repository
-(\url{https://github.com/SchlossLab/Lesniak_XXXX}).
+(\url{https://github.com/SchlossLab/Lesniak_Severity_XXXX_2022}).
 
@@
 and associated metadata are available through the Sequence Read Archive
-via accession XXXX.
+via accession PRJNA787941.
 
@@
 
-This work was supported by several grants from the National Institutes
-for Health R01GM099514, U19AI090871, U01AI12455, and P30DK034933.
-Additionally, NAL was supported by the Molecular Mechanisms of Microbial
-Pathogenesis training grant (NIH T32 AI007528). The funding agencies had
-no role in study design, data collection and analysis, decision to
-publish, or preparation of the manuscript.
+Thank you to Sarah Lucas and Sarah Tomkovich for critical discussion in
+the development and execution of this project. We also thank the
+University of Michigan Germ-free Mouse Core for assistance with our
+germfree mice, funded in part by U2CDK110768. This work was supported by
+several grants from the National Institutes for Health R01GM099514,
+U19AI090871, U01AI12455, and P30DK034933. Additionally, NAL was
+supported by the Molecular Mechanisms of Microbial Pathogenesis training
+grant (NIH T32 AI007528). The funding agencies had no role in study
+design, data collection and analysis, decision to publish, or
+preparation of the manuscript.
 
@@
 \begin{cslreferences}
-\leavevmode\hypertarget{ref-Dieterle2020}{}%
-1. \textbf{Dieterle MG}, \textbf{Putler R}, \textbf{Perry DA},
-\textbf{Menon A}, \textbf{Abernathy-Close L}, \textbf{Perlman NS},
-\textbf{Penkevich A}, \textbf{Standke A}, \textbf{Keidan M},
-\textbf{Vendrov KC}, \textbf{Bergin IL}, \textbf{Young VB}, \textbf{Rao
-K}. 2020. Systemic inflammatory mediators are effective biomarkers for
-predicting adverse outcomes in \emph{Clostridioides difficile}
-infection. mBio \textbf{11}.
-doi:\href{https://doi.org/10.1128/mbio.00180-20}{10.1128/mbio.00180-20}.
-
-\leavevmode\hypertarget{ref-Butt2013}{}%
-2. \textbf{Butt E}, \textbf{Foster JA}, \textbf{Keedwell E},
-\textbf{Bell JE}, \textbf{Titball RW}, \textbf{Bhangu A},
-\textbf{Michell SL}, \textbf{Sheridan R}. 2013. Derivation and
-validation of a simple, accurate and robust prediction rule for risk of
-mortality in patients with \emph{Clostridium difficile} infection. BMC
-Infectious Diseases \textbf{13}.
-doi:\href{https://doi.org/10.1186/1471-2334-13-316}{10.1186/1471-2334-13-316}.
-
-\leavevmode\hypertarget{ref-Perry2021}{}%
-3. \textbf{Perry DA}, \textbf{Shirley D}, \textbf{Micic D},
-\textbf{Patel CP}, \textbf{Putler R}, \textbf{Menon A}, \textbf{Young
-VB}, \textbf{Rao K}. 2021. External validation and comparison of
-\emph{Clostridioides difficile} severity scoring systems. Clinical
-Infectious Diseases.
-doi:\href{https://doi.org/10.1093/cid/ciab737}{10.1093/cid/ciab737}.
-
-\leavevmode\hypertarget{ref-vanBeurden2017}{}%
-4. \textbf{Beurden YH van}, \textbf{Hensgens MPM}, \textbf{Dekkers OM},
-\textbf{Cessie SL}, \textbf{Mulder CJJ}, \textbf{Vandenbroucke-Grauls
-CMJE}. 2017. External validation of three prediction tools for patients
-at risk of a complicated course of \emph{Clostridium difficile}
-infection: Disappointing in an outbreak setting. Infection Control \&
-Hospital Epidemiology \textbf{38}:897--905.
-doi:\href{https://doi.org/10.1017/ice.2017.89}{10.1017/ice.2017.89}.
-
-\leavevmode\hypertarget{ref-Sorg2008}{}%
-5. \textbf{Sorg JA}, \textbf{Sonenshein AL}. 2008. Bile salts and
-glycine as cogerminants for \emph{Clostridium difficile} spores. Journal
-of Bacteriology \textbf{190}:2505--2512.
-doi:\href{https://doi.org/10.1128/jb.01765-07}{10.1128/jb.01765-07}.
-
-\leavevmode\hypertarget{ref-Thanissery2017}{}%
-6. \textbf{Thanissery R}, \textbf{Winston JA}, \textbf{Theriot CM}.
-2017. Inhibition of spore germination, growth, and toxin activity of
-clinically relevant \emph{C. difficile} strains by gut microbiota
-derived secondary bile acids. Anaerobe \textbf{45}:86--100.
-doi:\href{https://doi.org/10.1016/j.anaerobe.2017.03.004}{10.1016/j.anaerobe.2017.03.004}.
-
-\leavevmode\hypertarget{ref-Aguirre2021}{}%
-7. \textbf{Aguirre AM}, \textbf{Yalcinkaya N}, \textbf{Wu Q},
-\textbf{Swennes A}, \textbf{Tessier ME}, \textbf{Roberts P},
-\textbf{Miyajima F}, \textbf{Savidge T}, \textbf{Sorg JA}. 2021. Bile
-acid-independent protection against \emph{Clostridioides difficile}
-infection \textbf{17}:e1010015.
-doi:\href{https://doi.org/10.1371/journal.ppat.1010015}{10.1371/journal.ppat.1010015}.
-
-\leavevmode\hypertarget{ref-Kang2019}{}%
-8. \textbf{Kang JD}, \textbf{Myers CJ}, \textbf{Harris SC},
-\textbf{Kakiyama G}, \textbf{Lee I-K}, \textbf{Yun B-S},
-\textbf{Matsuzaki K}, \textbf{Furukawa M}, \textbf{Min H-K},
-\textbf{Bajaj JS}, \textbf{Zhou H}, \textbf{Hylemon PB}. 2019. Bile acid
-7\(\alpha\)-dehydroxylating gut bacteria secrete antibiotics that
-inhibit \emph{Clostridium difficile}: Role of secondary bile acids
-\textbf{26}:27--34.e4.
-doi:\href{https://doi.org/10.1016/j.chembiol.2018.10.003}{10.1016/j.chembiol.2018.10.003}.
-
-\leavevmode\hypertarget{ref-Tomkovich2020}{}%
-9. \textbf{Tomkovich S}, \textbf{Stough JMA}, \textbf{Bishop L},
-\textbf{Schloss PD}. 2020. The initial gut microbiota and response to
-antibiotic perturbation influence \emph{Clostridioides difficile}
-clearance in mice. mSphere \textbf{5}.
-doi:\href{https://doi.org/10.1128/msphere.00869-20}{10.1128/msphere.00869-20}.
-
-\leavevmode\hypertarget{ref-Carlson2013}{}%
-10. \textbf{Carlson PE}, \textbf{Walk ST}, \textbf{Bourgis AET},
-\textbf{Liu MW}, \textbf{Kopliku F}, \textbf{Lo E}, \textbf{Young VB},
-\textbf{Aronoff DM}, \textbf{Hanna PC}. 2013. The relationship between
-phenotype, ribotype, and clinical disease in human \emph{Clostridium
-difficile} isolates. Anaerobe \textbf{24}:109--116.
-doi:\href{https://doi.org/10.1016/j.anaerobe.2013.04.003}{10.1016/j.anaerobe.2013.04.003}.
-
-\leavevmode\hypertarget{ref-Akerlund2006}{}%
-11. 2006. Correlation of disease severity with fecal toxin levels in
-patients with \emph{Clostridium difficile}-associated diarrhea and
-distribution of PCR ribotypes and toxin yields in vitro of corresponding
-isolates. Journal of Clinical Microbiology \textbf{44}:353--358.
-doi:\href{https://doi.org/10.1128/jcm.44.2.353-358.2006}{10.1128/jcm.44.2.353-358.2006}.
-
-\leavevmode\hypertarget{ref-Vitucci2020}{}%
-12. \textbf{Vitucci JC}, \textbf{Pulse M}, \textbf{Tabor-Simecka L},
-\textbf{Simecka J}. 2020. Epidemic ribotypes of \emph{clostridium} (now
-\emph{clostridioides}) \emph{difficile} are likely to be more virulent
-than non-epidemic ribotypes in animal models. BMC Microbiology
-\textbf{20}.
-doi:\href{https://doi.org/10.1186/s12866-020-1710-5}{10.1186/s12866-020-1710-5}.
-
-\leavevmode\hypertarget{ref-Cowardin2016}{}%
-13. \textbf{Cowardin CA}, \textbf{Buonomo EL}, \textbf{Saleh MM},
-\textbf{Wilson MG}, \textbf{Burgess SL}, \textbf{Kuehne SA},
-\textbf{Schwan C}, \textbf{Eichhoff AM}, \textbf{Koch-Nolte F},
-\textbf{Lyras D}, \textbf{Aktories K}, \textbf{Minton NP}, \textbf{Petri
-WA}. 2016. The binary toxin CDT enhances \emph{Clostridium difficile}
-virulence by suppressing protective colonic eosinophilia. Nature
-Microbiology \textbf{1}.
-doi:\href{https://doi.org/10.1038/nmicrobiol.2016.108}{10.1038/nmicrobiol.2016.108}.
-
-\leavevmode\hypertarget{ref-Lungulescu2011}{}%
-14. \textbf{Lungulescu OA}, \textbf{Cao W}, \textbf{Gatskevich E},
-\textbf{Tlhabano L}, \textbf{Stratidis JG}. 2011. CSI: A severity index
-for \emph{Clostridium difficile} infection at the time of admission.
-Journal of Hospital Infection \textbf{79}:151--154.
-doi:\href{https://doi.org/10.1016/j.jhin.2011.04.017}{10.1016/j.jhin.2011.04.017}.
+\leavevmode\hypertarget{ref-Kelly2008}{}%
+1. \textbf{Kelly CP}, \textbf{LaMont JT}. 2008. \emph{Clostridium
+difficile} more difficult than ever. New England Journal of Medicine
+\textbf{359}:1932--1940.
+doi:\href{https://doi.org/10.1056/nejmra0707500}{10.1056/nejmra0707500}.
 
 \leavevmode\hypertarget{ref-McDonald2018}{}%
-15. \textbf{McDonald LC}, \textbf{Gerding DN}, \textbf{Johnson S},
+2. \textbf{McDonald LC}, \textbf{Gerding DN}, \textbf{Johnson S},
 \textbf{Bakken JS}, \textbf{Carroll KC}, \textbf{Coffin SE},
@@
 
+\leavevmode\hypertarget{ref-Dieterle2020}{}%
+3. \textbf{Dieterle MG}, \textbf{Putler R}, \textbf{Perry DA},
+\textbf{Menon A}, \textbf{Abernathy-Close L}, \textbf{Perlman NS},
+\textbf{Penkevich A}, \textbf{Standke A}, \textbf{Keidan M},
+\textbf{Vendrov KC}, \textbf{Bergin IL}, \textbf{Young VB}, \textbf{Rao
+K}. 2020. Systemic inflammatory mediators are effective biomarkers for
+predicting adverse outcomes in \emph{Clostridioides difficile}
+infection. mBio \textbf{11}.
+doi:\href{https://doi.org/10.1128/mbio.00180-20}{10.1128/mbio.00180-20}.
+
+\leavevmode\hypertarget{ref-Butt2013}{}%
+4. \textbf{Butt E}, \textbf{Foster JA}, \textbf{Keedwell E},
+\textbf{Bell JE}, \textbf{Titball RW}, \textbf{Bhangu A},
+\textbf{Michell SL}, \textbf{Sheridan R}. 2013. Derivation and
+validation of a simple, accurate and robust prediction rule for risk of
+mortality in patients with \emph{Clostridium difficile} infection. BMC
+Infectious Diseases \textbf{13}.
+doi:\href{https://doi.org/10.1186/1471-2334-13-316}{10.1186/1471-2334-13-316}.
+
+\leavevmode\hypertarget{ref-Perry2021}{}%
+5. \textbf{Perry DA}, \textbf{Shirley D}, \textbf{Micic D},
+\textbf{Patel CP}, \textbf{Putler R}, \textbf{Menon A}, \textbf{Young
+VB}, \textbf{Rao K}. 2021. External validation and comparison of
+\emph{Clostridioides difficile} severity scoring systems. Clinical
+Infectious Diseases.
+doi:\href{https://doi.org/10.1093/cid/ciab737}{10.1093/cid/ciab737}.
+
+\leavevmode\hypertarget{ref-vanBeurden2017}{}%
+6. \textbf{Beurden YH van}, \textbf{Hensgens MPM}, \textbf{Dekkers OM},
+\textbf{Cessie SL}, \textbf{Mulder CJJ}, \textbf{Vandenbroucke-Grauls
+CMJE}. 2017. External validation of three prediction tools for patients
+at risk of a complicated course of \emph{Clostridium difficile}
+infection: Disappointing in an outbreak setting. Infection Control \&
+Hospital Epidemiology \textbf{38}:897--905.
+doi:\href{https://doi.org/10.1017/ice.2017.89}{10.1017/ice.2017.89}.
+
+\leavevmode\hypertarget{ref-Britton2014}{}%
+7. \textbf{Britton RA}, \textbf{Young VB}. 2014. Role of the intestinal
+microbiota in resistance to colonization by \emph{Clostridium
+difficile}. Gastroenterology \textbf{146}:1547--1553.
+doi:\href{https://doi.org/10.1053/j.gastro.2014.01.059}{10.1053/j.gastro.2014.01.059}.
+
+\leavevmode\hypertarget{ref-Hryckowian2018}{}%
+8. \textbf{Hryckowian AJ}, \textbf{Treuren WV}, \textbf{Smits SA},
+\textbf{Davis NM}, \textbf{Gardner JO}, \textbf{Bouley DM},
+\textbf{Sonnenburg JL}. 2018. Microbiota-accessible carbohydrates
+suppress \emph{Clostridium difficile} infection in a murine model.
+Nature Microbiology \textbf{3}:662--669.
+doi:\href{https://doi.org/10.1038/s41564-018-0150-6}{10.1038/s41564-018-0150-6}.
+
+\leavevmode\hypertarget{ref-VichVila2020}{}%
+9. \textbf{Vila AV}, \textbf{Collij V}, \textbf{Sanna S}, \textbf{Sinha
+T}, \textbf{Imhann F}, \textbf{Bourgonje AR}, \textbf{Mujagic Z},
+\textbf{Jonkers DMAE}, \textbf{Masclee AAM}, \textbf{Fu J},
+\textbf{Kurilshikov A}, \textbf{Wijmenga C}, \textbf{Zhernakova A},
+\textbf{Weersma RK}. 2020. Impact of commonly used drugs on the
+composition and metabolic function of the gut microbiota. Nature
+Communications \textbf{11}.
+doi:\href{https://doi.org/10.1038/s41467-019-14177-z}{10.1038/s41467-019-14177-z}.
+
+\leavevmode\hypertarget{ref-Abbas2020}{}%
+10. \textbf{Abbas A}, \textbf{Zackular JP}. 2020. Microbe-microbe
+interactions during \emph{Clostridioides difficile} infection. Current
+Opinion in Microbiology \textbf{53}:19--25.
+doi:\href{https://doi.org/10.1016/j.mib.2020.01.016}{10.1016/j.mib.2020.01.016}.
+
+\leavevmode\hypertarget{ref-Jenior2017}{}%
+11. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
+\textbf{Schloss PD}. 2017. \emph{Clostridium difficile} colonizes
+alternative nutrient niches during infection across distinct murine gut
+microbiomes. mSystems \textbf{2}.
+doi:\href{https://doi.org/10.1128/msystems.00063-17}{10.1128/msystems.00063-17}.
+
+\leavevmode\hypertarget{ref-Sorg2008}{}%
+12. \textbf{Sorg JA}, \textbf{Sonenshein AL}. 2008. Bile salts and
+glycine as cogerminants for \emph{Clostridium difficile} spores. Journal
+of Bacteriology \textbf{190}:2505--2512.
+doi:\href{https://doi.org/10.1128/jb.01765-07}{10.1128/jb.01765-07}.
+
+\leavevmode\hypertarget{ref-Thanissery2017}{}%
+13. \textbf{Thanissery R}, \textbf{Winston JA}, \textbf{Theriot CM}.
+2017. Inhibition of spore germination, growth, and toxin activity of
+clinically relevant \emph{C. difficile} strains by gut microbiota
+derived secondary bile acids. Anaerobe \textbf{45}:86--100.
+doi:\href{https://doi.org/10.1016/j.anaerobe.2017.03.004}{10.1016/j.anaerobe.2017.03.004}.
+
+\leavevmode\hypertarget{ref-Aguirre2021}{}%
+14. \textbf{Aguirre AM}, \textbf{Yalcinkaya N}, \textbf{Wu Q},
+\textbf{Swennes A}, \textbf{Tessier ME}, \textbf{Roberts P},
+\textbf{Miyajima F}, \textbf{Savidge T}, \textbf{Sorg JA}. 2021. Bile
+acid-independent protection against \emph{Clostridioides difficile}
+infection. PLOS Pathogens \textbf{17}:e1010015.
+doi:\href{https://doi.org/10.1371/journal.ppat.1010015}{10.1371/journal.ppat.1010015}.
+
+\leavevmode\hypertarget{ref-Kang2019}{}%
+15. \textbf{Kang JD}, \textbf{Myers CJ}, \textbf{Harris SC},
+\textbf{Kakiyama G}, \textbf{Lee I-K}, \textbf{Yun B-S},
+\textbf{Matsuzaki K}, \textbf{Furukawa M}, \textbf{Min H-K},
+\textbf{Bajaj JS}, \textbf{Zhou H}, \textbf{Hylemon PB}. 2019. Bile acid
+7\(\alpha\)-dehydroxylating gut bacteria secrete antibiotics that
+inhibit \emph{Clostridium difficile}: Role of secondary bile acids. Cell
+Chemical Biology \textbf{26}:27--34.e4.
+doi:\href{https://doi.org/10.1016/j.chembiol.2018.10.003}{10.1016/j.chembiol.2018.10.003}.
+
+\leavevmode\hypertarget{ref-Seekatz2016}{}%
+16. \textbf{Seekatz AM}, \textbf{Rao K}, \textbf{Santhosh K},
+\textbf{Young VB}. 2016. Dynamics of the fecal microbiome in patients
+with recurrent and nonrecurrent \emph{Clostridium difficile} infection.
+Genome Medicine \textbf{8}.
+doi:\href{https://doi.org/10.1186/s13073-016-0298-8}{10.1186/s13073-016-0298-8}.
+
+\leavevmode\hypertarget{ref-Tomkovich2020}{}%
+17. \textbf{Tomkovich S}, \textbf{Stough JMA}, \textbf{Bishop L},
+\textbf{Schloss PD}. 2020. The initial gut microbiota and response to
+antibiotic perturbation influence \emph{Clostridioides difficile}
+clearance in mice. mSphere \textbf{5}.
+doi:\href{https://doi.org/10.1128/msphere.00869-20}{10.1128/msphere.00869-20}.
+
+\leavevmode\hypertarget{ref-Lesniak2021}{}%
+18. \textbf{Lesniak NA}, \textbf{Schubert AM}, \textbf{Sinani H},
+\textbf{Schloss PD}. 2021. Clearance of \emph{Clostridioides difficile}
+colonization is associated with antibiotic-specific bacterial changes.
+mSphere \textbf{6}.
+doi:\href{https://doi.org/10.1128/msphere.01238-20}{10.1128/msphere.01238-20}.
+
+\leavevmode\hypertarget{ref-Nagpal2018}{}%
+19. \textbf{Nagpal R}, \textbf{Wang S}, \textbf{Woods LCS},
+\textbf{Seshie O}, \textbf{Chung ST}, \textbf{Shively CA},
+\textbf{Register TC}, \textbf{Craft S}, \textbf{McClain DA},
+\textbf{Yadav H}. 2018. Comparative microbiome signatures and
+short-chain fatty acids in mouse, rat, non-human primate, and human
+feces. Frontiers in Microbiology \textbf{9}.
+doi:\href{https://doi.org/10.3389/fmicb.2018.02897}{10.3389/fmicb.2018.02897}.
+
+\leavevmode\hypertarget{ref-Schubert2014}{}%
+20. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
+\textbf{Mogle J}, \textbf{Petrosino JP}, \textbf{Young VB},
+\textbf{Aronoff DM}, \textbf{Schloss PD}. 2014. Microbiome data
+distinguish patients with \emph{Clostridium difficile} infection and
+non-\emph{C. difficile}-associated diarrhea from healthy controls. mBio
+\textbf{5}.
+doi:\href{https://doi.org/10.1128/mbio.01021-14}{10.1128/mbio.01021-14}.
+
+\leavevmode\hypertarget{ref-Gillilland2012}{}%
+21. \textbf{Gillilland MG}, \textbf{Erb-Downward JR}, \textbf{Bassis
+CM}, \textbf{Shen MC}, \textbf{Toews GB}, \textbf{Young VB},
+\textbf{Huffnagle GB}. 2012. Ecological succession of bacterial
+communities during conventionalization of germ-free mice. Applied and
+Environmental Microbiology \textbf{78}:2359--2366.
+doi:\href{https://doi.org/10.1128/aem.05239-11}{10.1128/aem.05239-11}.
+
+\leavevmode\hypertarget{ref-Chen2008}{}%
+22. \textbf{Chen X}, \textbf{Katchar K}, \textbf{Goldsmith JD},
+\textbf{Nanthakumar N}, \textbf{Cheknis A}, \textbf{Gerding DN},
+\textbf{Kelly CP}. 2008. A mouse model of \emph{Clostridium
+difficile}-associated disease. Gastroenterology \textbf{135}:1984--1992.
+doi:\href{https://doi.org/10.1053/j.gastro.2008.09.002}{10.1053/j.gastro.2008.09.002}.
+
+\leavevmode\hypertarget{ref-Schubert2015}{}%
+23. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
+Antibiotic-induced alterations of the murine gut microbiota and
+subsequent effects on colonization resistance against \emph{Clostridium
+difficile}. mBio \textbf{6}.
+doi:\href{https://doi.org/10.1128/mbio.00974-15}{10.1128/mbio.00974-15}.
+
+\leavevmode\hypertarget{ref-Akerlund2006}{}%
+24. \textbf{Åkerlund T}, \textbf{Svenungsson B}, \textbf{Lagergren},
+\textbf{Burman LG}. 2006. Correlation of disease severity with fecal
+toxin levels in patients with \emph{Clostridium difficile}-associated
+diarrhea and distribution of PCR ribotypes and toxin yields in vitro of
+corresponding isolates. Journal of Clinical Microbiology
+\textbf{44}:353--358.
+doi:\href{https://doi.org/10.1128/jcm.44.2.353-358.2006}{10.1128/jcm.44.2.353-358.2006}.
+
+\leavevmode\hypertarget{ref-Vitucci2020}{}%
+25. \textbf{Vitucci JC}, \textbf{Pulse M}, \textbf{Tabor-Simecka L},
+\textbf{Simecka J}. 2020. Epidemic ribotypes of \emph{Clostridium} (now
+\emph{Clostridioides}) \emph{difficile} are likely to be more virulent
+than non-epidemic ribotypes in animal models. BMC Microbiology
+\textbf{20}.
+doi:\href{https://doi.org/10.1186/s12866-020-1710-5}{10.1186/s12866-020-1710-5}.
+
+\leavevmode\hypertarget{ref-Cowardin2016}{}%
+26. \textbf{Cowardin CA}, \textbf{Buonomo EL}, \textbf{Saleh MM},
+\textbf{Wilson MG}, \textbf{Burgess SL}, \textbf{Kuehne SA},
+\textbf{Schwan C}, \textbf{Eichhoff AM}, \textbf{Koch-Nolte F},
+\textbf{Lyras D}, \textbf{Aktories K}, \textbf{Minton NP}, \textbf{Petri
+WA}. 2016. The binary toxin CDT enhances \emph{Clostridium difficile}
+virulence by suppressing protective colonic eosinophilia. Nature
+Microbiology \textbf{1}.
+doi:\href{https://doi.org/10.1038/nmicrobiol.2016.108}{10.1038/nmicrobiol.2016.108}.
+
+\leavevmode\hypertarget{ref-Carlson2015}{}%
+27. \textbf{Carlson PE}, \textbf{Kaiser AM}, \textbf{McColm SA},
+\textbf{Bauer JM}, \textbf{Young VB}, \textbf{Aronoff DM}, \textbf{Hanna
+PC}. 2015. Variation in germination of \emph{Clostridium difficile}
+clinical isolates correlates to disease severity. Anaerobe
+\textbf{33}:64--70.
+doi:\href{https://doi.org/10.1016/j.anaerobe.2015.02.003}{10.1016/j.anaerobe.2015.02.003}.
+
+\leavevmode\hypertarget{ref-Carlson2013}{}%
+28. \textbf{Carlson PE}, \textbf{Walk ST}, \textbf{Bourgis AET},
+\textbf{Liu MW}, \textbf{Kopliku F}, \textbf{Lo E}, \textbf{Young VB},
+\textbf{Aronoff DM}, \textbf{Hanna PC}. 2013. The relationship between
+phenotype, ribotype, and clinical disease in human \emph{Clostridium
+difficile} isolates. Anaerobe \textbf{24}:109--116.
+doi:\href{https://doi.org/10.1016/j.anaerobe.2013.04.003}{10.1016/j.anaerobe.2013.04.003}.
+
+\leavevmode\hypertarget{ref-Lungulescu2011}{}%
+29. \textbf{Lungulescu OA}, \textbf{Cao W}, \textbf{Gatskevich E},
+\textbf{Tlhabano L}, \textbf{Stratidis JG}. 2011. CSI: A severity index
+for \emph{Clostridium difficile} infection at the time of admission.
+Journal of Hospital Infection \textbf{79}:151--154.
+doi:\href{https://doi.org/10.1016/j.jhin.2011.04.017}{10.1016/j.jhin.2011.04.017}.
+
 \leavevmode\hypertarget{ref-Zar2007}{}%
-16. \textbf{Zar FA}, \textbf{Bakkanagari SR}, \textbf{Moorthi KMLST},
+30. \textbf{Zar FA}, \textbf{Bakkanagari SR}, \textbf{Moorthi KMLST},
 \textbf{Davis MB}. 2007. A comparison of vancomycin and metronidazole
@@
 \leavevmode\hypertarget{ref-diMasi2018}{}%
-17. \textbf{Masi A di}, \textbf{Leboffe L}, \textbf{Polticelli F},
+31. \textbf{Masi A di}, \textbf{Leboffe L}, \textbf{Polticelli F},
 \textbf{Tonon F}, \textbf{Zennaro C}, \textbf{Caterino M}, \textbf{Stano
@@
 \leavevmode\hypertarget{ref-AbernathyClose2020}{}%
-18. \textbf{Abernathy-Close L}, \textbf{Dieterle MG}, \textbf{Vendrov
+32. \textbf{Abernathy-Close L}, \textbf{Dieterle MG}, \textbf{Vendrov
 KC}, \textbf{Bergin IL}, \textbf{Rao K}, \textbf{Young VB}. 2020. Aging
@@
 \leavevmode\hypertarget{ref-Theriot2011}{}%
-19. \textbf{Theriot CM}, \textbf{Koumpouras CC}, \textbf{Carlson PE},
+33. \textbf{Theriot CM}, \textbf{Koumpouras CC}, \textbf{Carlson PE},
 \textbf{Bergin II}, \textbf{Aronoff DM}, \textbf{Young VB}. 2011.
@@
 \leavevmode\hypertarget{ref-Goorhuis2008}{}%
-20. \textbf{Goorhuis A}, \textbf{Bakker D}, \textbf{Corver J},
+34. \textbf{Goorhuis A}, \textbf{Bakker D}, \textbf{Corver J},
 \textbf{Debast SB}, \textbf{Harmanus C}, \textbf{Notermans DW},
@@
 \leavevmode\hypertarget{ref-OConnor2009}{}%
-21. \textbf{O'Connor JR}, \textbf{Johnson S}, \textbf{Gerding DN}. 2009.
+35. \textbf{O'Connor JR}, \textbf{Johnson S}, \textbf{Gerding DN}. 2009.
 \emph{Clostridium difficile} infection caused by the epidemic
@@
 
-\leavevmode\hypertarget{ref-Chen2008}{}%
-22. \textbf{Chen X}, \textbf{Katchar K}, \textbf{Goldsmith JD},
-\textbf{Nanthakumar N}, \textbf{Cheknis A}, \textbf{Gerding DN},
-\textbf{Kelly CP}. 2008. A mouse model of \emph{Clostridium
-difficile}-associated disease. Gastroenterology \textbf{135}:1984--1992.
-doi:\href{https://doi.org/10.1053/j.gastro.2008.09.002}{10.1053/j.gastro.2008.09.002}.
+\leavevmode\hypertarget{ref-Rao2015}{}%
+36. \textbf{Rao K}, \textbf{Micic D}, \textbf{Natarajan M},
+\textbf{Winters S}, \textbf{Kiel MJ}, \textbf{Walk ST}, \textbf{Santhosh
+K}, \textbf{Mogle JA}, \textbf{Galecki AT}, \textbf{LeBar W},
+\textbf{Higgins PDR}, \textbf{Young VB}, \textbf{Aronoff DM}. 2015.
+\emph{Clostridium difficile} ribotype 027: Relationship to age,
+detectability of toxins A or B in stool with rapid testing, severe
+infection, and mortality. Clinical Infectious Diseases
+\textbf{61}:233--241.
+doi:\href{https://doi.org/10.1093/cid/civ254}{10.1093/cid/civ254}.
+
+\leavevmode\hypertarget{ref-Walk2012}{}%
+37. \textbf{Walk ST}, \textbf{Micic D}, \textbf{Jain R}, \textbf{Lo ES},
+\textbf{Trivedi I}, \textbf{Liu EW}, \textbf{Almassalha LM},
+\textbf{Ewing SA}, \textbf{Ring C}, \textbf{Galecki AT}, \textbf{Rogers
+MAM}, \textbf{Washer L}, \textbf{Newton DW}, \textbf{Malani PN},
+\textbf{Young VB}, \textbf{Aronoff DM}. 2012. \emph{Clostridium
+difficile} ribotype does not predict severe infection. Clinical
+Infectious Diseases \textbf{55}:1661--1668.
+doi:\href{https://doi.org/10.1093/cid/cis786}{10.1093/cid/cis786}.
+
+\leavevmode\hypertarget{ref-Saund2021}{}%
+38. \textbf{Saund K}, \textbf{Pirani A}, \textbf{Lacy B}, \textbf{Hanna
+PC}, \textbf{Snitkin ES}. 2021. Strain variation in \emph{Clostridioides
+difficile} toxin activity associated with genomic variation at both
+PaLoc and non-PaLoc loci.
+doi:\href{https://doi.org/10.1101/2021.12.08.471880}{10.1101/2021.12.08.471880}.
 
 \leavevmode\hypertarget{ref-He2010}{}%
-23. \textbf{He M}, \textbf{Sebaihia M}, \textbf{Lawley TD},
+39. \textbf{He M}, \textbf{Sebaihia M}, \textbf{Lawley TD},
 \textbf{Stabler RA}, \textbf{Dawson LF}, \textbf{Martin MJ},
@@
 
-\leavevmode\hypertarget{ref-Schubert2015}{}%
-24. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
-Antibiotic-induced alterations of the murine gut microbiota and
-subsequent effects on colonization resistance against \emph{Clostridium
-difficile}. mBio \textbf{6}.
-doi:\href{https://doi.org/10.1128/mbio.00974-15}{10.1128/mbio.00974-15}.
-
-\leavevmode\hypertarget{ref-Jenior2017}{}%
-25. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
-\textbf{Schloss PD}. 2017. \emph{Clostridium difficile} colonizes
-alternative nutrient niches during infection across distinct murine gut
-microbiomes. mSystems \textbf{2}.
-doi:\href{https://doi.org/10.1128/msystems.00063-17}{10.1128/msystems.00063-17}.
-
-\leavevmode\hypertarget{ref-Lesniak2021}{}%
-26. \textbf{Lesniak NA}, \textbf{Schubert AM}, \textbf{Sinani H},
-\textbf{Schloss PD}. 2021. Clearance of \emph{Clostridioides difficile}
-colonization is associated with antibiotic-specific bacterial changes.
-mSphere \textbf{6}.
-doi:\href{https://doi.org/10.1128/msphere.01238-20}{10.1128/msphere.01238-20}.
+\leavevmode\hypertarget{ref-Jenior2018}{}%
+40. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
+\textbf{Schloss PD}. 2018. \emph{Clostridium difficile} alters the
+structure and metabolism of distinct cecal microbiomes during initial
+infection to promote sustained colonization. mSphere \textbf{3}.
+doi:\href{https://doi.org/10.1128/msphere.00261-18}{10.1128/msphere.00261-18}.
 
 \leavevmode\hypertarget{ref-Staley2016}{}%
-27. \textbf{Staley C}, \textbf{Weingarden AR}, \textbf{Khoruts A},
+41. \textbf{Staley C}, \textbf{Weingarden AR}, \textbf{Khoruts A},
 \textbf{Sadowsky MJ}. 2016. Interaction of gut microbiota with bile acid
@@
 \leavevmode\hypertarget{ref-Long2017}{}%
-28. \textbf{Long SL}, \textbf{Gahan CGM}, \textbf{Joyce SA}. 2017.
+42. \textbf{Long SL}, \textbf{Gahan CGM}, \textbf{Joyce SA}. 2017.
 Interactions between gut bacteria and bile in health and disease.
@@
 \leavevmode\hypertarget{ref-Sorg2010}{}%
-29. \textbf{Sorg JA}, \textbf{Sonenshein AL}. 2010. Inhibiting the
+43. \textbf{Sorg JA}, \textbf{Sonenshein AL}. 2010. Inhibiting the
 initiation of \emph{Clostridium difficile} spore germination using
@@
 \leavevmode\hypertarget{ref-Dubois2019}{}%
-30. \textbf{Dubois T}, \textbf{Tremblay YDN}, \textbf{Hamiot A},
+44. \textbf{Dubois T}, \textbf{Tremblay YDN}, \textbf{Hamiot A},
 \textbf{Martin-Verstraete I}, \textbf{Deschamps J}, \textbf{Monot M},
@@
 \leavevmode\hypertarget{ref-Ng2013}{}%
-31. \textbf{Ng KM}, \textbf{Ferreyra JA}, \textbf{Higginbottom SK},
+45. \textbf{Ng KM}, \textbf{Ferreyra JA}, \textbf{Higginbottom SK},
 \textbf{Lynch JB}, \textbf{Kashyap PC}, \textbf{Gopinath S},
@@
 \leavevmode\hypertarget{ref-Ferreyra2014}{}%
-32. \textbf{Ferreyra JA}, \textbf{Wu KJ}, \textbf{Hryckowian AJ},
+46. \textbf{Ferreyra JA}, \textbf{Wu KJ}, \textbf{Hryckowian AJ},
 \textbf{Bouley DM}, \textbf{Weimer BC}, \textbf{Sonnenburg JL}. 2014.
@@
 \leavevmode\hypertarget{ref-MartinVerstraete2016}{}%
-33. \textbf{Martin-Verstraete I}, \textbf{Peltier J}, \textbf{Dupuy B}.
+47. \textbf{Martin-Verstraete I}, \textbf{Peltier J}, \textbf{Dupuy B}.
 2016. The regulatory networks that control \emph{Clostridium difficile}
@@
 \leavevmode\hypertarget{ref-Collins2015}{}%
-34. \textbf{Collins J}, \textbf{Auchtung JM}, \textbf{Schaefer L},
+48. \textbf{Collins J}, \textbf{Auchtung JM}, \textbf{Schaefer L},
 \textbf{Eaton KA}, \textbf{Britton RA}. 2015. Humanized microbiota mice
@@
 \leavevmode\hypertarget{ref-Lawley2012}{}%
-35. \textbf{Lawley TD}, \textbf{Clare S}, \textbf{Walker AW},
+49. \textbf{Lawley TD}, \textbf{Clare S}, \textbf{Walker AW},
 \textbf{Stares MD}, \textbf{Connor TR}, \textbf{Raisen C},
@@
 \leavevmode\hypertarget{ref-Reeves2011}{}%
-36. \textbf{Reeves AE}, \textbf{Theriot CM}, \textbf{Bergin IL},
+50. \textbf{Reeves AE}, \textbf{Theriot CM}, \textbf{Bergin IL},
 \textbf{Huffnagle GB}, \textbf{Schloss PD}, \textbf{Young VB}. 2011. The
@@
 \leavevmode\hypertarget{ref-Mabrok2011}{}%
-37. \textbf{Mabrok HB}, \textbf{Klopfleisch R}, \textbf{Ghanem KZ},
+51. \textbf{Mabrok HB}, \textbf{Klopfleisch R}, \textbf{Ghanem KZ},
 \textbf{Clavel T}, \textbf{Blaut M}, \textbf{Loh G}. 2011. Lignan
@@
 \leavevmode\hypertarget{ref-Kim2019}{}%
-38. \textbf{Kim CC}, \textbf{Healey GR}, \textbf{Kelly WJ},
+52. \textbf{Kim CC}, \textbf{Healey GR}, \textbf{Kelly WJ},
 \textbf{Patchett ML}, \textbf{Jordens Z}, \textbf{Tannock GW},
@@
 \leavevmode\hypertarget{ref-doPrado2021}{}%
-39. \textbf{Prado SBR do}, \textbf{Minguzzi BT}, \textbf{Hoffmann C},
+53. \textbf{Prado SBR do}, \textbf{Minguzzi BT}, \textbf{Hoffmann C},
 \textbf{Fabi JP}. 2021. Modulation of human gut microbiota by dietary
@@
 \leavevmode\hypertarget{ref-Muthuramalingam2019}{}%
-40. \textbf{Muthuramalingam K}, \textbf{Singh V}, \textbf{Choi C},
+54. \textbf{Muthuramalingam K}, \textbf{Singh V}, \textbf{Choi C},
 \textbf{Choi SI}, \textbf{Kim YM}, \textbf{Unno T}, \textbf{Cho M}.
@@
 \leavevmode\hypertarget{ref-Han2019}{}%
-41. \textbf{Han S-H}, \textbf{Yi J}, \textbf{Kim J-H}, \textbf{Lee S},
+55. \textbf{Han S-H}, \textbf{Yi J}, \textbf{Kim J-H}, \textbf{Lee S},
 \textbf{Moon H-W}. 2019. Composition of gut microbiota in patients with
-toxigenic \emph{clostridioides} (\emph{clostridium}) \emph{difficile}:
+toxigenic \emph{Clostridioides} (\emph{Clostridium}) \emph{difficile}:
 Comparison between subgroups according to clinical criteria and toxin
@@
 \leavevmode\hypertarget{ref-Duncan2004}{}%
-42. \textbf{Duncan SH}, \textbf{Louis P}, \textbf{Flint HJ}. 2004.
+56. \textbf{Duncan SH}, \textbf{Louis P}, \textbf{Flint HJ}. 2004.
 Lactate-utilizing bacteria, isolated from human feces, that produce
@@
 \leavevmode\hypertarget{ref-Ye2018}{}%
-43. \textbf{Ye J}, \textbf{Lv L}, \textbf{Wu W}, \textbf{Li Y},
+57. \textbf{Ye J}, \textbf{Lv L}, \textbf{Wu W}, \textbf{Li Y},
 \textbf{Shi D}, \textbf{Fang D}, \textbf{Guo F}, \textbf{Jiang H},
 \textbf{Yan R}, \textbf{Ye W}, \textbf{Li L}. 2018. Butyrate protects
-mice against methionineCholine-deficient diet-induced non-alcoholic
+mice against methionine--choline-deficient diet-induced non-alcoholic
 steatohepatitis by improving gut barrier function, attenuating
@@
 \leavevmode\hypertarget{ref-Walsh2017}{}%
-44. \textbf{Walsh CJ}, \textbf{Guinane CM}, \textbf{O'Toole PW},
+58. \textbf{Walsh CJ}, \textbf{Guinane CM}, \textbf{O'Toole PW},
 \textbf{Cotter PD}. 2017. A profile hidden markov model to investigate
@@
 \leavevmode\hypertarget{ref-Sandiford2018}{}%
-45. \textbf{Sandiford SK}. 2018. Current developments in lantibiotic
+59. \textbf{Sandiford SK}. 2018. Current developments in lantibiotic
 discovery for treating \emph{Clostridium difficile} infection. Expert
@@
 \leavevmode\hypertarget{ref-Stein2013}{}%
-46. \textbf{Stein RR}, \textbf{Bucci V}, \textbf{Toussaint NC},
+60. \textbf{Stein RR}, \textbf{Bucci V}, \textbf{Toussaint NC},
 \textbf{Buffie CG}, \textbf{Rätsch G}, \textbf{Pamer EG}, \textbf{Sander
@@
 \leavevmode\hypertarget{ref-Nakashima2021}{}%
-47. \textbf{Nakashima T}, \textbf{Fujii K}, \textbf{Seki T},
+61. \textbf{Nakashima T}, \textbf{Fujii K}, \textbf{Seki T},
 \textbf{Aoyama M}, \textbf{Azuma A}, \textbf{Kawasome H}. 2021. Novel
@@
 \leavevmode\hypertarget{ref-Geerlings2018}{}%
-48. \textbf{Geerlings S}, \textbf{Kostopoulos I}, \textbf{Vos W de},
+62. \textbf{Geerlings S}, \textbf{Kostopoulos I}, \textbf{Vos W de},
 \textbf{Belzer C}. 2018. \emph{Akkermansia muciniphila} in the human
@@
 \leavevmode\hypertarget{ref-Deng2018}{}%
-49. \textbf{Deng H}, \textbf{Yang S}, \textbf{Zhang Y}, \textbf{Qian K},
+63. \textbf{Deng H}, \textbf{Yang S}, \textbf{Zhang Y}, \textbf{Qian K},
 \textbf{Zhang Z}, \textbf{Liu Y}, \textbf{Wang Y}, \textbf{Bai Y},
@@
 \leavevmode\hypertarget{ref-Engevik2020}{}%
-50. \textbf{Engevik MA}, \textbf{Engevik AC}, \textbf{Engevik KA},
+64. \textbf{Engevik MA}, \textbf{Engevik AC}, \textbf{Engevik KA},
 \textbf{Auchtung JM}, \textbf{Chang-Graham AL}, \textbf{Ruan W},
@@
 \leavevmode\hypertarget{ref-Reeves2012}{}%
-51. \textbf{Reeves AE}, \textbf{Koenigsknecht MJ}, \textbf{Bergin IL},
+65. \textbf{Reeves AE}, \textbf{Koenigsknecht MJ}, \textbf{Bergin IL},
 \textbf{Young VB}. 2012. Suppression of \emph{Clostridium difficile} in
 the gastrointestinal tracts of germfree mice inoculated with a murine
-isolate from the family \emph{lachnospiraceae}. Infection and Immunity
+isolate from the family \emph{Lachnospiraceae}. Infection and Immunity
 \textbf{80}:3786--3794.
@@
 \leavevmode\hypertarget{ref-Ma2021}{}%
-52. \textbf{Ma L}, \textbf{Keng J}, \textbf{Cheng M}, \textbf{Pan H},
+66. \textbf{Ma L}, \textbf{Keng J}, \textbf{Cheng M}, \textbf{Pan H},
 \textbf{Feng B}, \textbf{Hu Y}, \textbf{Feng T}, \textbf{Yang F}. 2021.
@@
 \leavevmode\hypertarget{ref-Haas2020}{}%
-53. \textbf{Haas KN}, \textbf{Blanchard JL}. 2020. Reclassification of
+67. \textbf{Haas KN}, \textbf{Blanchard JL}. 2020. Reclassification of
 the \emph{Clostridium clostridioforme} and \emph{Clostridium sphenoides}
-clades as \emph{enterocloster} gen. Nov. And \emph{lacrimispora} gen.
-Nov., including reclassification of 15 taxa. International Journal of
+clades as \emph{Enterocloster} gen. nov. And \emph{Lacrimispora} gen.
+nov., Including reclassification of 15 taxa. International Journal of
 Systematic and Evolutionary Microbiology \textbf{70}:23--34.
@@
 \leavevmode\hypertarget{ref-Finegold2005}{}%
-54. \textbf{Finegold SM}, \textbf{Song Y}, \textbf{Liu C}, \textbf{Hecht
+68. \textbf{Finegold SM}, \textbf{Song Y}, \textbf{Liu C}, \textbf{Hecht
 DW}, \textbf{Summanen P}, \textbf{Könönen E}, \textbf{Allen SD}. 2005.
@@
 
-\leavevmode\hypertarget{ref-Pirofski2008}{}%
-55. \textbf{Pirofski L-a}, \textbf{Casadevall A}. 2008. The
-damage-response framework of microbial pathogenesis and infectious
-diseases, pp. 135--146. \emph{In}. Springer New York.
-
-\leavevmode\hypertarget{ref-Casadevall2014}{}%
-56. \textbf{Casadevall A}, \textbf{Pirofski L-a}. 2014. What is a host?
-Incorporating the microbiota into the damage-response framework
-\textbf{83}:2--7.
-doi:\href{https://doi.org/10.1128/iai.02627-14}{10.1128/iai.02627-14}.
-
 \leavevmode\hypertarget{ref-VanInsberghe2020}{}%
-57. \textbf{VanInsberghe D}, \textbf{Elsherbini JA}, \textbf{Varian B},
+69. \textbf{VanInsberghe D}, \textbf{Elsherbini JA}, \textbf{Varian B},
 \textbf{Poutahidis T}, \textbf{Erdman S}, \textbf{Polz MF}. 2020.
 Diarrhoeal events can trigger long-term \emph{Clostridium difficile}
-colonization with recurrent blooms \textbf{5}:642--650.
+colonization with recurrent blooms. Nature Microbiology
+\textbf{5}:642--650.
 doi:\href{https://doi.org/10.1038/s41564-020-0668-2}{10.1038/s41564-020-0668-2}.
@@
 \leavevmode\hypertarget{ref-GarzaGonzalez2019}{}%
-58. \textbf{Garza-González E}, \textbf{Mendoza-Olazarán S},
+70. \textbf{Garza-González E}, \textbf{Mendoza-Olazarán S},
 \textbf{Morfin-Otero R}, \textbf{Ramírez-Fontes A},
@@
 \textbf{Camacho-Ortiz A}. 2019. Intestinal microbiome changes in fecal
-microbiota transplant (FMT) vs. FMT enriched with \emph{lactobacillus}
-in the treatment of recurrent \emph{Clostridioides difficile} infection
-\textbf{2019}:1--7.
+microbiota transplant (FMT) vs. FMT enriched with \emph{Lactobacillus}
+in the treatment of recurrent \emph{Clostridioides difficile} infection.
+Canadian Journal of Gastroenterology and Hepatology \textbf{2019}:1--7.
 doi:\href{https://doi.org/10.1155/2019/4549298}{10.1155/2019/4549298}.
 
-\leavevmode\hypertarget{ref-Schubert2014}{}%
-59. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
-\textbf{Mogle J}, \textbf{Petrosino JP}, \textbf{Young VB},
-\textbf{Aronoff DM}, \textbf{Schloss PD}. 2014. Microbiome data
-distinguish patients with \emph{Clostridium difficile} infection and
-non-\emph{C. difficile}-associated diarrhea from healthy controls
-\textbf{5}.
-doi:\href{https://doi.org/10.1128/mbio.01021-14}{10.1128/mbio.01021-14}.
-
 \leavevmode\hypertarget{ref-Shafiq2020}{}%
-60. \textbf{Shafiq M}, \textbf{Alturkmani H}, \textbf{Zafar Y},
+71. \textbf{Shafiq M}, \textbf{Alturkmani H}, \textbf{Zafar Y},
 \textbf{Mittal V}, \textbf{Lodhi H}, \textbf{Ullah W}, \textbf{Brewer
 J}. 2020. Effects of co-infection on the clinical outcomes of
-\emph{Clostridium difficile} infection \textbf{12}.
+\emph{Clostridium difficile} infection. Gut Pathogens \textbf{12}.
 doi:\href{https://doi.org/10.1186/s13099-020-00348-7}{10.1186/s13099-020-00348-7}.
@@
 \leavevmode\hypertarget{ref-Keith2020}{}%
-61. \textbf{Keith JW}, \textbf{Dong Q}, \textbf{Sorbara MT},
+72. \textbf{Keith JW}, \textbf{Dong Q}, \textbf{Sorbara MT},
 \textbf{Becattini S}, \textbf{Sia JK}, \textbf{Gjonbalaj M},
@@
 EG}. 2020. Impact of antibiotic-resistant bacteria on immune activation
-and \emph{Clostridioides difficile} infection in the mouse intestine
-\textbf{88}.
+and \emph{Clostridioides difficile} infection in the mouse intestine.
+Infection and Immunity \textbf{88}.
 doi:\href{https://doi.org/10.1128/iai.00362-19}{10.1128/iai.00362-19}.
@@
 \leavevmode\hypertarget{ref-Zackular2016}{}%
-62. \textbf{Zackular JP}, \textbf{Moore JL}, \textbf{Jordan AT},
+73. \textbf{Zackular JP}, \textbf{Moore JL}, \textbf{Jordan AT},
 \textbf{Juttukonda LJ}, \textbf{Noto MJ}, \textbf{Nicholson MR},
@@
 \textbf{Skaar EP}. 2016. Dietary zinc alters the microbiota and
-decreases resistance to \emph{Clostridium difficile} infection
-\textbf{22}:1330--1334.
+decreases resistance to \emph{Clostridium difficile} infection. Nature
+Medicine \textbf{22}:1330--1334.
 doi:\href{https://doi.org/10.1038/nm.4174}{10.1038/nm.4174}.
@@
 \leavevmode\hypertarget{ref-Berkell2021}{}%
-63. \textbf{Berkell M}, \textbf{Mysara M}, \textbf{Xavier BB},
+74. \textbf{Berkell M}, \textbf{Mysara M}, \textbf{Xavier BB},
 \textbf{Werkhoven CH van}, \textbf{Monsieurs P}, \textbf{Lammens C},
@@
 2021. Microbiota-based markers predictive of development of
-\emph{Clostridioides difficile} infection \textbf{12}.
+\emph{Clostridioides difficile} infection. Nature Communications
+\textbf{12}.
 doi:\href{https://doi.org/10.1038/s41467-021-22302-0}{10.1038/s41467-021-22302-0}.
@@
 \leavevmode\hypertarget{ref-Gardiner2014}{}%
-64. \textbf{Gardiner BJ}, \textbf{Tai AY}, \textbf{Kotsanas D},
+75. \textbf{Gardiner BJ}, \textbf{Tai AY}, \textbf{Kotsanas D},
 \textbf{Francis MJ}, \textbf{Roberts SA}, \textbf{Ballard SA},
 \textbf{Junckerstorff RK}, \textbf{Korman TM}. 2014. Clinical and
-microbiological characteristics of \emph{Eggerthella lenta} bacteremia
-\textbf{53}:626--635.
+microbiological characteristics of \emph{Eggerthella lenta} bacteremia.
+Journal of Clinical Microbiology \textbf{53}:626--635.
 doi:\href{https://doi.org/10.1128/jcm.02926-14}{10.1128/jcm.02926-14}.
@@
 \leavevmode\hypertarget{ref-Iljazovic2020}{}%
-65. \textbf{Iljazovic A}, \textbf{Roy U}, \textbf{Gálvez EJC},
+76. \textbf{Iljazovic A}, \textbf{Roy U}, \textbf{Gálvez EJC},
 \textbf{Lesker TR}, \textbf{Zhao B}, \textbf{Gronow A}, \textbf{Amend
@@
 \textbf{Schmidt-Hohagen K}, \textbf{Neumann-Schaal M}, \textbf{Strowig
-T}. 2020. Perturbation of the gut microbiome by \emph{prevotella} spp.
-Enhances host susceptibility to mucosal inflammation
+T}. 2020. Perturbation of the gut microbiome by \emph{Prevotella} spp.
+enhances host susceptibility to mucosal inflammation. Mucosal Immunology
 \textbf{14}:113--124.
@@
 \leavevmode\hypertarget{ref-Nagalingam2013}{}%
-66. \textbf{Nagalingam NA}, \textbf{Robinson CJ}, \textbf{Bergin IL},
+77. \textbf{Nagalingam NA}, \textbf{Robinson CJ}, \textbf{Bergin IL},
 \textbf{Eaton KA}, \textbf{Huffnagle GB}, \textbf{Young VB}. 2013. The
 effects of intestinal microbial community structure on disease
-manifestation in IL-\(10^{-/-}\) mice infected with helicobacter
-hepaticus \textbf{1}.
+manifestation in IL-\(10^{-/-}\) mice infected with \emph{Helicobacter
+hepaticus}. Microbiome \textbf{1}.
 doi:\href{https://doi.org/10.1186/2049-2618-1-15}{10.1186/2049-2618-1-15}.
@@
 \leavevmode\hypertarget{ref-AbernathyClose2021}{}%
-67. \textbf{Abernathy-Close L}, \textbf{Barron MR}, \textbf{George JM},
+78. \textbf{Abernathy-Close L}, \textbf{Barron MR}, \textbf{George JM},
 \textbf{Dieterle MG}, \textbf{Vendrov KC}, \textbf{Bergin IL},
@@
 susceptible to \emph{Clostridioides difficile} colonization and
-infection.
+infection. mBio.
 doi:\href{https://doi.org/10.1128/mbio.02733-20}{10.1128/mbio.02733-20}.
+
+\leavevmode\hypertarget{ref-Pirofski2008}{}%
+79. \textbf{Pirofski L-a}, \textbf{Casadevall A}. 2008. The
+damage-response framework of microbial pathogenesis and infectious
+diseases, pp. 135--146. \emph{In} Advances in experimental medicine and
+biology. Springer New York.
+
+\leavevmode\hypertarget{ref-Casadevall2014}{}%
+80. \textbf{Casadevall A}, \textbf{Pirofski L-a}. 2014. What is a host?
+Incorporating the microbiota into the damage-response framework.
+Infection and Immunity \textbf{83}:2--7.
+doi:\href{https://doi.org/10.1128/iai.02627-14}{10.1128/iai.02627-14}.
+
+\leavevmode\hypertarget{ref-Tailford2015}{}%
+81. \textbf{Tailford LE}, \textbf{Crost EH}, \textbf{Kavanaugh D},
+\textbf{Juge N}. 2015. Mucin glycan foraging in the human gut
+microbiome. Frontiers in Genetics \textbf{6}.
+doi:\href{https://doi.org/10.3389/fgene.2015.00081}{10.3389/fgene.2015.00081}.
+
+\leavevmode\hypertarget{ref-Sorg2009}{}%
+82. \textbf{Sorg JA}, \textbf{Dineen SS}. 2009. Laboratory maintenance
+of \emph{Clostridium difficile}. Current Protocols in Microbiology
+\textbf{12}.
+doi:\href{https://doi.org/10.1002/9780471729259.mc09a01s12}{10.1002/9780471729259.mc09a01s12}.
+
+\leavevmode\hypertarget{ref-Winston2016}{}%
+83. \textbf{Winston JA}, \textbf{Thanissery R}, \textbf{Montgomery SA},
+\textbf{Theriot CM}. 2016. Cefoperazone-treated mouse model of
+clinically-relevant \emph{Clostridium difficile} strain R20291. Journal
+of Visualized Experiments.
+doi:\href{https://doi.org/10.3791/54850}{10.3791/54850}.
+
+\leavevmode\hypertarget{ref-Kozich2013}{}%
+84. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
+dual-index sequencing strategy and curation pipeline for analyzing
+amplicon sequence data on the MiSeq illumina sequencing platform.
+Applied and Environmental Microbiology \textbf{79}:5112--5120.
+doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
+
+\leavevmode\hypertarget{ref-Schloss2009}{}%
+85. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
+\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
+\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
+\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
+Introducing mothur: Open-source, platform-independent,
+community-supported software for describing and comparing microbial
+communities. Applied and Environmental Microbiology
+\textbf{75}:7537--7541.
+doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.
+
+\leavevmode\hypertarget{ref-Wang2007}{}%
+86. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
+\textbf{Cole JR}. 2007. Naïve bayesian classifier for rapid assignment
+of rRNA sequences into the new bacterial taxonomy. Applied and
+Environmental Microbiology \textbf{73}:5261--5267.
+doi:\href{https://doi.org/10.1128/aem.00062-07}{10.1128/aem.00062-07}.
+
+\leavevmode\hypertarget{ref-Yue2005}{}%
+87. \textbf{Yue JC}, \textbf{Clayton MK}. 2005. A similarity measure
+based on species proportions. Communications in Statistics - Theory and
+Methods \textbf{34}:2123--2131.
+doi:\href{https://doi.org/10.1080/sta-200066418}{10.1080/sta-200066418}.
+
+\leavevmode\hypertarget{ref-Segata2011}{}%
+88. \textbf{Segata N}, \textbf{Izard J}, \textbf{Waldron L},
+\textbf{Gevers D}, \textbf{Miropolsky L}, \textbf{Garrett WS},
+\textbf{Huttenhower C}. 2011. Metagenomic biomarker discovery and
+explanation. Genome Biology \textbf{12}:R60.
+doi:\href{https://doi.org/10.1186/gb-2011-12-6-r60}{10.1186/gb-2011-12-6-r60}.
+
+\leavevmode\hypertarget{ref-Benjamini1995}{}%
+89. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
+false discovery rate: A practical and powerful approach to multiple
+testing. Journal of the Royal Statistical Society: Series B
+(Methodological) \textbf{57}:289--300.
+doi:\href{https://doi.org/10.1111/j.2517-6161.1995.tb02031.x}{10.1111/j.2517-6161.1995.tb02031.x}.
+
+\leavevmode\hypertarget{ref-Topcuoglu2021}{}%
+90. \textbf{Topçuoğlu B}, \textbf{Lapp Z}, \textbf{Sovacool K},
+\textbf{Snitkin E}, \textbf{Wiens J}, \textbf{Schloss P}. 2021.
+Mikropml: User-friendly R package for supervised machine learning
+pipelines. Journal of Open Source Software \textbf{6}:3073.
+doi:\href{https://doi.org/10.21105/joss.03073}{10.21105/joss.03073}.
 \end{cslreferences}
@@
 
-\textbf{Figure 1. R.} (A-C).
+\textbf{Figure 1. Human fecal microbial communities established diverse
+gut bacterial communities in germ-free mice.} (A) Relative abundance of
+10 most abundant bacterial classes of the gut bacterial communities on
+timepoint day 0, 14 days post-colonization of germ-free C57Bl/6 mice
+with human fecal samples. Each column of abundances represents an
+individual mouse. Each group of mice that received the same donor are
+grouped together and labeled with a letter (N for non-moribund mice and
+M for moribund mice) and number. (B) Median (points) and interquartile
+range (lines) of \(\beta\)-diversity (\(\theta\)\textsubscript{YC})
+between an individual mouse and either all others which were inoculated
+with feces from the same donor or from a different donor. The
+\(\beta\)-diversity among the same donor comparison group was
+significantly less than the \(\beta\)-diversity of the different donor
+group (\emph{P} \textless{} 0.05, calculated by Wilcoxon rank sum test).
 
@@
 
-\textbf{Figure 2. M.} F.
+\textbf{Figure 2. All donor groups resulted in \emph{C. difficile}
+infection but with different outcomes.} \emph{C. difficile} CFU per gram
+of stool was measured the day after challenge with 10\(^{3}\) \emph{C.
+difficile} RT027 human isolate 431 spores and at the end of the
+experiment, 10 days post-challenge. Each point represents an individual
+mouse. Mice are grouped by donor and labeled by the donor letter (N for
+non-moribund mice and M for moribund mice) and number. Points are
+colored by donor group. Mice from donor groups N1 through N6 succumbed
+to the infection prior to day 10 and were not plated on day 10
+post-challenge. LOD = Limit of detection.
 
@@
 
-\textbf{Figure 3. O.} F.
+\textbf{Figure 3. Histopathologic score and toxin activity varied across
+donor groups.} (A) Fecal toxin activity was detected in some mice post
+\emph{C. difficile} challenge in both moribund and non-moribund mice.
+(B) Cecum scored for histopathologic damage from mice at the end of the
+experiment. Samples were collected for histopathologic scoring on day 10
+post-challenge for non-moribund mice or the day the mouse succumbed to
+the infection for the moribund group (day 2-3 post-challenge). Each
+point represents an individual mouse. Mice are grouped by donor and
+labeled by the donor letter (N for non-moribund mice and M for moribund
+mice) and number. Points are colored by donor group. Missing points are
+from mice that had insufficient sample collected for assaying fecal
+toxin.
 
@@
 
-\textbf{Figure 4. E.} A.
+\textbf{Figure 4. Individual fecal bacterial community members of the
+murine gut associated with \emph{C .difficile} infection outcomes.} (A
+and B) Relative abundance of genera at the time of \emph{C. difficile}
+challenge (Day 0) that varied significantly by the moribundity and
+histopathologic summary score or toxin activity by LEfSe analysis.
+Median (points) and interquartile range (lines) are plotted. Genera are
+ordered alphabetically to ease comparisons across analyses. (A) Relative
+abundances were compared across infection outcome of moribund (colored
+black) or non-moribund with either a high histopathologic score (score
+greater than the median score of 5, colored green) or a low
+histopathologic summary score (score less than the median score of 5,
+colored light green). (B) Relative abundances were compared between mice
+which toxin activity was detected (Toxigenic, colored dark purple) and
+which no toxin activity was detected (Non-toxigenic, colored light
+purple). (C) Endpoint bacterial OTUs correlated with histopathologic
+summary score. Each individual mouse is plotted (transparent gray
+point). Spearman's correlations were statistically significant after
+Benjamini-Hochberg correction for multiple comparisons. * next to a name
+indicates that the bacterial group was unclassified at lower
+classification ranks.
 
@@
 
-\textbf{Figure 5. D.} (A) L
+\textbf{Figure 5. Fecal bacterial community members of the murine gut at
+the time of \emph{C. difficile} (Day 0) predicted outcomes of the
+infection.} Day 0 bacterial community members grouped by different
+classification rank were modeled with random forest to predict the
+infection outcome. The models used the highest taxonomic classification
+rank that performed as well as the lower ranks. Median (solid points)
+and interquartile range (lines) of the group relative abundance are
+plotted. Bacterial groups are ordered by their importance to the model;
+taxonomic group at the top of the plot had the greatest decrease in
+performance when its relative abundances were permuted. * next to the
+name indicates that the bacterial group was unclassified at lower
+classification ranks. (A) Bacterial members grouped by phyla predicted
+which mice would have toxin activity detected at any point throughout
+the infection (Toxigenic, dark purple). (B) Bacterial members grouped by
+class predicted which mice would become moribund (dark blue). (C)
+Bacterial members grouped by genera predicted if the mice would have a
+high (score greater than the median score of 5, colored dark green) or
+low (score less than the median score of 5, colored light green)
+histopathologic summary score.
 
@@
 
-\textbf{Figure 6. D.} (A) L
+\textbf{Figure 6. \emph{C. difficile} infection outcome were similar
+within human fecal donor groups across different \emph{C. difficile}
+isolates.} Mice colonized with human fecal donor communities N7, N6, or
+M4 were challenged with the isolate from the previous experiments in
+this study, \emph{C. difficile} RT027 human isolate 431 (blue points),
+or a different \emph{C. difficile} isolate (purple points). \emph{C.
+difficile} CFUs were enumerated at the end of the experiment (day 10
+post-challenge) and are plotted grouped by donor source (top plot).
+Histopathologic scores from the tissue collected at the endpoint (either
+day 10 post-challenge or day mice succumbed to infection) are plotted
+grouped by donor source (bottom plot). Each point represents an
+individual mouse. Mice that had received fecal communities from human
+donor M4 succumbed to the infection prior to day 10 post-challenge so
+were not plated for CFU on day 10 post-challenge (DECD = deceased). RT =
+ribotype, Isolate = \emph{C. difficile} human isolate id, LOD = limit of
+detection.
 
@@
 
-\textbf{Figure S1. D.} (A) L
+\textbf{Figure S1. Histopathologic score of tissue damage at the
+endpoint of the infection.} Tissue collected at the endpoint, either day
+10 post-challenge (Non-moribund) or day mice succumbed to infection
+(Moribund), were scored from histopathologic damage. Each point
+represents an individual mouse. Mice (points) are grouped and colored by
+their human fecal community donor. Missing points are from mice that had
+insufficient sample for histopathologic scoring.
 
@@
 
-\textbf{Figure S2. D.} (A) L
+\textbf{Figure S2. Random forest models predicted outcomes of the
+\emph{C. difficile} challenge.} (A-C) Taxonomic classification rank
+model performance. Relative abundance at the time of \emph{C. difficile}
+challenge (Day 0) of the bacterial community members grouped by
+different classification rank were modeled with random forest to predict
+the infection outcome. The models used the highest taxonomic
+classification rank that performed as well as the lower ranks. Black
+rectangle highlights classification rank used to model each outcome.
+(D-F) Model feature importance. Bacterial groups are ordered by their
+decrease in area under receiver-operator curve (AUC) when its relative
+abundances was permuted. Individual relative abundances were added to F
+since differences in AUC were outside the interquartile range. *
+indicates bacterial group was unclassified at lower classification
+ranks. For all plots, median (solid points) and interquartile range
+(lines) are plotted. (A) Toxin production modeled which mice would have
+toxin detected during the experiment. (B) Moribundity modeled which mice
+would succumb to the infection prior to day 10 post-challenge. (C)
+Histopathologic score modeled which mice would have a high (score
+greater than the median score of 5) or low (score less than the median
+score of 5) histopathologic summary score. (D) Bacterial phyla which
+affected the performance of predicting toxin activity when permuted. (E)
+Bacterial classes which affected the performance of predicting
+moribundity when permuted. (D) Bacterial genera which affected the
+performance of predicting histopathologic score when permuted.
 
@@
 
-\textbf{Figure S3. D.} (A) L
-
-\hfill\break
-
-\textbf{Table 1. D.} (A) L
-
-samples\_used \textless-
-readxl::read\_xlsx(`data/raw/humanGF\_ids.xlsx') \%\textgreater\%
-pull(human\_source) \%\textgreater\% unique donor\_data \textless-
-readxl::read\_xlsx(`data/raw/MIMARKS\_cdclinical.xlsx') \%\textgreater\%
-filter(sample\_id \%in\% samples\_used) \%\textgreater\%
-select(sample\_id, biome, age, gender, ``antibiotics \textgreater3mo'',
-protonpump, antacid, Healthworker, historyCdiff, Surgery6mos,
-Vegetarian, weight, disease\_stat)
-
-\textbf{Table 2. D.} (A) L
-
-read.csv(here(`data/raw/Hanna\_in\_vitro\_data\_plus\_germination.txt'))
-\%\textgreater\% rownames\_to\_column(`isolate') \%\textgreater\%
-filter(isolate \%in\% c(`DA00299', `DA00395', `DA00458', `DA00431'))
+\textbf{Figure S3. Subset of donor communities were selected to test for
+conserved \emph{C. difficile} infection outcome with a different
+\emph{C. difficile} isolate.} (A) Relative abundance of genera in mice
+from previous experiments which received human fecal donor communities
+N7, N6, or M4. Each column represents an individual mouse. (B) Features
+from the random forest model that predicted moribundity using relative
+abundance of bacteria were grouped by their genus classification rank
+and are ordered by their decrease in AUC when each feature was permuted.
+Median (solid points) and interquartile range (lines) change in AUC are
+plotted. (C) Relative abundance of the most important genera used by the
+model to predict moribundity. Median (solid points) and interquartile
+range (lines) of the group relative abundances are plotted. * next to a
+name indicates that the bacterial group was unclassified at lower
+classification ranks.
 

### 2022-01-22 submission/manuscript.Rmd
incorporate coauthor feedback

@@
 
-Nicholas A. Lesniak$^1$, Alyxandria M. Schubert$^1$, Kaitlyn Flynn$^1$, Jhansi L Leslie$^1$, Hamide Sinani$^1$, Ingrid L. Bergin$^3$, Vincent B Young$^2$, Patrick D. Schloss$^{1,\dagger}$
+Nicholas A. Lesniak$^1$, Alyxandria M. Schubert$^1$, Kaitlyn J. Flynn$^1$, Jhansi L. Leslie$^{1,4}$, Hamide Sinani$^1$, Ingrid L. Bergin$^3$, Vincent B. Young$^{1,2}$, Patrick D. Schloss$^{1,\dagger}$
 
@@
 3\. Unit for Laboratory Animal Medicine, University of Michigan, Ann Arbor, MI  
+4\. Current affiliation: Department of Medicine, Division of International Health and Infectious Diseases, University of Virginia School of Medicine, Charlottesville, Virginia, USA
 
@@
 difficile* ribotype and toxin genes have been associated with disease severity.
-However, it is unclear whether there is an associate between the gut microbiota disease severity.
+However, it is unclear whether there is an association between the gut microbiota and disease severity.
 The gut microbiota is known to interact with *C.
@@
 difficile* colonization through bile acid metabolism, nutrient consumption and bacteriocin production.
-It has not been shown whether the gut microbiota can affect CDI disease severity.
-Here we demonstrate that gut bacterial communities can contribute to disease severity.
+Here we sought to demonstrate that members of the gut bacterial communities can contribute to disease severity.
 We derived diverse gut communities by colonizing germ-free mice with different human fecal communities.
@@
 The variation in severity was associated with the human fecal community that the mice received.
-Generally, facultative anaerobes with pathogenic potential, such as *Escherichia*, *Helicobacter*,  and *Klebsiella*, were associated with more severe outcomes.
-Bacterial groups associated with fiber degradation, such as *Coprobacillus*, were associated with less severe outcomes.
-Finally, we showed that, when infected with a different *C.
-difficile* isolate, the outcome of the infection was similar between the mice that received the same fecal donor community.
+Generally, bacterial populations with pathogenic potential, such as *Escherichia*, *Helicobacter*,  and *Klebsiella*, were associated with more severe outcomes.
+Bacterial groups associated with fiber degradation, bile acid metabolism and lantibiotic production, such as *Anaerostipes* and *Coprobacillus*, were associated with less severe outcomes.
 These data indicate that gut bacteria can influence CDI disease severity.
@@
 
-*Clostridioides difficile* infections (CDI) can be asymptomatic or range in severity from mild diarrhea to severe outcomes including surgical interventions and death.
+*Clostridioides difficile* colonization can result in no symptoms or develop into an infection, ranging in severity from mild diarrhea to toxic megacolon, sepsis and death.
 Models that predict severity and guide treatment decisions are based on clinical factors and *C.
@@
 difficile* infection outcomes.
-Gut bacterial community data from patients with CDI could improve our ability to identify patients at risk of developing more severe disease and improve interventions which treat the gut bacteria to reduce host damage.
+Gut bacterial community data from patients with CDI could improve our ability to identify patients at risk of developing more severe disease and improve interventions which target *C.
+difficile* and the gut bacteria to reduce host damage.
 
@@
 CDI disease severity can range from asymptomatic to mild diarrhea to toxic megacolon and death.
-The Infectious Diseases Society of America (IDSA) and Society for Healthcare Epidemiology of America (SHEA) guidelines categorize severe disease based on white blood cell count greater than 15,000 cells/mm$^3$ and/or a serum creatinine greater than 1.5 mg/dL, and complicated disease based on systolic blood pressure below 90, ileus, or toxic megacolon [@McDonald2018].
-Since these measures are only present once the disease has advanced, the IDSA classification does not allow for identification of patients at risk of severe CDI.
-Prediction models have been developed to score a patient's risk for severe CDI outcomes based on clinical factors but have not been robust for broad application [@Dieterle2020; @Butt2013; @Perry2021; @vanBeurden2017].
+The Infectious Diseases Society of America (IDSA) and Society for Healthcare Epidemiology of America (SHEA) guidelines defines severe CDI in terms of a white blood cell count greater than 15,000 cells/mm$^3$ and/or a serum creatinine greater than 1.5 mg/dL.
+Patients who develop shock or hypotension, ileus, or toxic megacolon are considered to have fulminant CDI [@McDonald2018].
+Since these measures are outcomes of CDI, these measures have limited ablity to predict risk of severe CDI when CDI is fiest detected.
+Schemes have been developed to score a patient's risk for severe CDI outcomes based on clinical factors but have not been robust for broad application [@Perry2021].
 Thus, we have limited ability to prevent patients from developing severe CDI.
-Ideally, we could identify patients that are high-risk of severe disease so their treatments could be adjusted to halt the infection before the disease worsens.
+Ideally, we could identify patients that are high-risk of severe disease so their treatments could be adjusted to alter the clinical trajectory.
 
@@
 A range of mechanisms can disrupt this barrier, including antibiotics, medications, or dietary changes, and lead to increased susceptibility to CDI [@Britton2014; @Hryckowian2018; @VichVila2020].
-Once in the gut, the indigenous bacteria can either promote or inhibit *C.
+Once *C.
+difficile* overcomes the protective barrier and colonizes the intestine, the indigenous bacteria can either promote or inhibit *C.
 difficile* through producing molecules or modifying the environment [@Abbas2020; @Jenior2017].
@@
 Bacteria in the gut also can compete more directly with *C.
-difficile* through antibiotic production or nutrient consumption [@Aguirre2021; @Kang2019].
+difficile* through antibiotic production or nutrient consumption [@Aguirre2021; @Kang2019; @Leslie2021].
 While the relationship between the gut bacteria and *C.
 difficile* has been established, the effect the gut bacteria can have on CDI disease severity is unclear.
-
-Studies investigating the severity of CDI disease involving the microbiome have not shown clear evidence of a relationship between the microbiome and disease severity.
-These studies were not setup to identify a relationship between the microbiome and disease severity.
-Studies that have used clinical data cannot control for host, microbiome or *C.
-difficile* ribotype variation [@Seekatz2016].
-While disease severity varies, it cannot be attributed to specifically the microbiome since all factors vary.
-Murine experiments typically use a single mouse colony and different *C.
-difficile* ribotypes to create severity differences [@Dieterle2020].
-Recently, our group has begun uncovering the effect microbiome variation has on *C.
-difficile* infection.
-We showed the variation in the bacterial communities between mice from different mouse colonies resulted in different clearance rates of *C.
-difficile* [@Tomkovich2020].
-We also showed varied ability of mice to spontaneously eliminate *C.
-difficile* infection when they were treated with different antibiotics prior to *C.
-difficile* challenge [@Lesniak2021].
-However, there have not been any studies investigating whether variation in the microbiome affects the resultant disease severity of *C.
-difficile* infections.
 
@@
 difficile* and its infection dynamics, it can also affect the disease severity of the infection.
-To test this hypothesis, we colonized germ-free C57BL/6 mice with human fecal communities to create varied gut communities.
+To test this hypothesis, we colonized germ-free C57BL/6 mice with human fecal samples to create varied gut communities.
 We then challenged the mice with a single *C.
-difficile* isolate, ribotype 027 (RT027) human isolate 431, and followed the mice for the development of severe outcomes of moribundity and histopathologic cecal tissue damage.
+difficile* isolate, ribotype 027 (RT027) clinical isolate 431, and followed the mice for the development of severe outcomes of moribundity and histopathologic cecal tissue damage.
 Since the murine host and *C.
@@
 ***C.
-difficile* is able to infect germ-free mice colonized with human fecal microbial communities without perturbation.**  To produce gut microbiomes with greater variation than those found in conventional mouse colonies, we colonized germ-free mice with bacteria from human feces [@Nagpal2018].
+difficile* is able to infect germ-free mice colonized with human fecal microbial communities without antibiotics**  To produce gut microbiomes with greater variation than those found in conventional mouse colonies, we colonized germ-free mice with bacteria from human feces [@Nagpal2018].
 We inoculated germ-free C57BL/6 mice with homogenized feces from one of 15 human fecal samples via oral gavage.
-The human fecal samples were selected because they represented diverse community structures based on community clustering [@Schubert2014].
-The gut communities were allowed to establish themselves for two weeks post inoculation [@Gillilland2012].
-We then surveyed the bacterial members of the gut communities by 16S rRNA gene sequencing of fecal pellets (Figure 1A).
+These human fecal samples were selected because they represented diverse community structures based on community clustering [@Schubert2014].
+The gut communities were allowed to equlibriate for two weeks post inoculation [@Gillilland2012].
+We then surveyed the bacterial members of the gut communities by 16S rRNA gene sequencing of murine fecal pellets (Figure 1A).
 The bacterial communities from each mouse grouped more closely to those communities from mice that received the same human fecal donor community than to the mice who received a different human fecal donor community (Figure 1B).
@@
 difficile* infection.
-A typical mouse model of CDI requires pre-treatment with antibiotics, such as clindamycin, to become susceptible to *C.
+A typical mouse model of CDI requires pre-treatment of conventional mice with antibiotics, such as clindamycin, to become susceptible to *C.
 difficile* colonization [@Chen2008; @Schubert2015].
-However, we wanted to avoid modifying the communities with an antibiotic to maintain their unique microbial compositions and ecological relationships.
-Since these communities came from people who had increased risk of CDI, such as hospitalization or recent antibiotic use [@Schubert2014], we decided to test if *C.
+However, we wanted to avoid modifying the gut communities with an antibiotic to maintain their unique microbial compositions and ecological relationships.
+Since some of these communities came from people at increased risk of CDI, for example due to hospitalization or recent antibiotic use [@Schubert2014], we decided to test if *C.
 difficile* was able to infect these mice without any antibiotic perturbation.
 We hypothesized that *C.
-difficile* would be able to colonize a subset of mice whose communities were susceptible from perturbations either the donor experienced or from the ecological selection when colonizing the murine gut.
+difficile* would be able to colonize the mice who received their gut communities from a donor with perturbed community.
 Mice were challenged with 10$^{3}$ *C.
-difficile* RT027 human isolate 431 spores.
-The mice were followed for 10 days post-challenge and their stool was collected and plated for *C.
+difficile* RT027 clinical isolate 431 spores.
+The mice were followed for 10 days post-challenge, and their stool was collected and plated for *C.
 difficile* colony forming units (CFU) to determine the extent of the infection.
-Surprisingly, mice from all donors were colonized (Figure 2).
-Only two mice were able to resist *C.
-difficile* infection and both received their community from the same donor, N3.
+Surprisingly, communities from all donors were able to be colonized (Figure 2).
+Two mice were able to resist *C.
+difficile* colonization, both received their community donor N3, which may be attributed to experimental variation since this group also had more mice.
 By colonizing germ-free mice with different human fecal communities, we were able to generate diverse gut communities in mice, which were susceptible to *C.
@@
 We followed the mice for 10 days post-challenge for colonization density, toxin production, and mortality.
-Seven mice, from Donors N1, N2 and N5, were not colonized on the day after *C.
+Seven mice, from Donors N1, N2 and N5, were not colonized at detectable levels on the day after *C.
 difficile* challenge but were infected (>10$^{6}$) by the end of the experiment.
@@
 difficile* infection through the end of the experiment (Figure 2).
-At 10 days post-challenge, or earlier for the moribund mice, mice were euthanised and fecal pellets were assayed for toxin activity and cecal tissue was collected and scored for histopathologic signs of disease (Figure 3).
+At 10 days post-challenge, or earlier for the moribund mice, mice were euthanised and fecal material were assayed for toxin activity and cecal tissue was collected and scored for histopathologic signs of disease (Figure 3).
 Overall, there was greater toxin activity detected in the stool of the moribund mice (*P* = 0.003).
@@
 Additionally, not all moribund mice had toxin activity detected in their stool.
-This discordance between toxin activity and disease severity has been previously described [@Akerlund2006; @Vitucci2020].
 Next we examined the cecal tissue for histopathologic damage.
@@
 We split the mice into groups by severity level based on their moribundity and histopathologic score.
-We dichotomized the histopathologic scores into high and low groups by splitting on the median score.
+We dichotomized the histopathologic scores into high and low groups by splitting on the median score of 5.
 This analysis revealed 20 genera that were significantly different by the disease severity (Figure 4A).
@@
 Populations of *Anaerotignum*, *Coprobacillus*, *Enterocloster*, and *Murimonas* were more abundant in the non-moribund mice that would develop only low intestinal injury.
-To understand the role of toxin activity in disease severity, we applied LEfSe to identify the genera most likely to explain the observed differences in toxin activity (Figure 4B).
+To understand the role of toxin activity in disease severity, we applied LEfSe to identify the genera most likely to explain the differences between the presence and absence of detected toxin activity (Figure 4B).
 Many genera that associated with toxin production were also associated with moribundity, such as populations of *Escherichia/Shigella* and *Bacteroides*.
-Likewise, there were genera such as *Anaerotignum*, *Enterocloster*, and *Murimonas* that were associated with no toxin production that also exhibited greater relative abundance in communities from non-moribund mice with a low histopathologic score.
+Likewise, there were genera such as *Anaerotignum*, *Enterocloster*, and *Murimonas* that were associated with no detected toxin production that also exhibited greater relative abundance in communities from non-moribund mice with a low histopathologic score.
 Lastly, we tested for correlations between the endpoint relative abundances of bacterial operational taxonomic units (OTUs) and the histopathologic summary score (Figure 4C).
@@
 
-**Disease severity consistent by donor community across isolates.** Since Carlson et al.
-showed *C.
-difficile* isolates with lower germination rates were more likely to result in severe disease [@Carlson2015], we next tested whether the effect of the community on CDI outcomes was conserved across *C.
-difficile* isolates with higher rates of germination.
-We colonized mice with human fecal slurry of either Donor N7, N6, or M4 and allowed two weeks for the community to establish.
-We then challenged the mice with a different *C.
-difficile* isolate and all mice were infected.
-The moribundity and histopathologic score outcomes when challenged with a different *C.
-difficile* human isolate were comparable to the outcomes of challenges with *C.
-difficile* RT027 human isolate 431 by their donor group (Figure 6).
-For the isolates and communities we tested, the effect of the gut bacterial community on disease severity was greater than the effect of the variation in these isolates.
-
 ## Discussion
@@
 Our analysis revealed an association between the relative abundance of bacterial community members and disease severity.
-We used the relative abundance patterns of community members associated with disease outcomes to select donor communities to test whether the disease severity was conserved when challenged with a different *C.
-difficile* isolate.
-The disease outcomes from the challenge with the second *C.
-difficile* isolate matched the outcomes of the challenge with *C.
-difficile* RT027 isolate 431.
-Therefore, the effect of the community variation on disease severity *in vivo* was greater than the effect of *C.
-difficile* isolate variation [@Carlson2013; @Carlson2015].
-Overall, these results have demonstrated that the bacterial community contributed to the disease severity of *C.
+Previous studies investigating the severity of CDI disease involving the microbiome have had limited ability to interrogate this relationship between the microbiome and disease severity.
+Studies that have used clinical data have limited ability to control variation in the host, microbiome or *C.
+difficile* ribotype [@Seekatz2016].
+Murine experiments typically use a single mouse colony and different *C.
+difficile* ribotypes to create severity differences [@Dieterle2020].
+Recently, our group has begun uncovering the effect microbiome variation has on *C.
+difficile* infection.
+We showed the variation in the bacterial communities between mice from different mouse colonies resulted in different clearance rates of *C.
+difficile* [@Tomkovich2020].
+We also showed varied ability of mice to spontaneously eliminate *C.
+difficile* infection when they were treated with different antibiotics prior to *C.
+difficile* challenge [@Lesniak2021].
+Overall, the results presented here have demonstrated that the gut bacterial community contributed to the severity of *C.
 difficile* infection.
 
-CDI severity can range from no symptoms to mild symptoms, such as diarrhea, to severe symptoms, such as toxic megacolon or death.
-Physicians use prediction tools to identify patients most at risk of developing a severe infection using white blood cell counts, serum albumin level, or serum creatinine level [@Lungulescu2011; @McDonald2018; @Zar2007].
+*C.
+difficile* can lead to asymptomatic colonization or infections with severity ranging from mild diarrhea to death.
+Physicians use classification tools to identify patients most at risk of developing a severe infection using white blood cell counts, serum albumin level, or serum creatinine level [@Lungulescu2011; @McDonald2018; @Zar2007].
 Those levels are driven by the activities in the intestine [@diMasi2018].
@@
 difficile* more virulent.
-Strains are categorized for their virulence by the presence and production toxin TcdA, toxin TcdB, and binary toxin and the prevalence in outbreaks, such as ribotypes 027 and 078 [@AbernathyClose2020; @Theriot2011; @Goorhuis2008; @OConnor2009; @Chen2008].
-However, other studies have shown that virulence is not necessarily linked with toxin production [@Rao2015] or the strain [@Walk2012].
+Strains are categorized for their virulence by the presence and production of the toxins TcdA, TcdB, and binary toxin and the prevalence in outbreaks, such as ribotypes 027 and 078 [@AbernathyClose2020; @Theriot2011; @Goorhuis2008; @OConnor2009; @Chen2008].
+However, other studies have shown that disease is not necessarily linked with toxin production [@Rao2015] or the strain [@Walk2012].
 Furthermore, there is variation in the genome, growth rate, sporulation, germination, and toxin production in different isolates of a strain [@Carlson2013; @Carlson2015; @Saund2021; @He2010].
@@
 difficile* population [@Jenior2017; @Jenior2018; @Lesniak2021].
-The gut bacteria metabolize primary bile acids into secondary bile acids [@Staley2016; @Long2017].
-The concentration of these bile acids affect the germination, growth, toxin production and biofilm formation [@Sorg2008; @Sorg2010; @Thanissery2017; @Dubois2019].
+Gut bacteria metabolize primary bile acids into secondary bile acids [@Staley2016; @Long2017].
+The concentration of these bile acids affects germination, growth, toxin production and biofilm formation [@Sorg2008; @Sorg2010; @Thanissery2017; @Dubois2019].
 Members of the bacterial community also affect other metabolites *C.
@@
 
-Myriad studies have explored the relationship between the microbiome and CDI but none have established a relationship between the variation of disease severity and the microbiome.
+A myriad of studies have explored the relationship between the microbiome and CDI disease.
 CDI studies often use an infection model with a single homogeneous bacterial community.
-Collins et al.
-used multiple human communities to colonize mice, however the communities were pooled prior to gavaging into germ-free mice [@Collins2015], resulting in a single community.
+Collins *et al.* used multiple human communities to colonize mice, however the communities were pooled prior to gavaging into germ-free mice [@Collins2015], resulting in a single human-derived community.
 Studies examining difference in disease often use different *C.
@@
 These studies either lack variation in the initial microbiome or have variation in the *C.
-difficile* infection itself, therefore we cannot make any association between variation in severity and the microbiome.
-
+difficile* infection itself, confounding any association between variation in severity and the microbiome.
+**Add Hikiro et al**
 With our recent observations that the initial community affected the ability of *C.
-difficile* to persist in the gut [@Tomkovich2020; @Lesniak2021] and the existing research describing the myriad interactions between the microbiome and *C.
+difficile* to persist in the gut [@Tomkovich2020; @Lesniak2021] and the existing research describing the numerous interactions between the microbiome and *C.
 difficile*, we hypothesized that the microbiome can modulate CDI disease severity.
@@
 Recently, *Helicobacter hepaticus* was shown to be sufficient to cause susceptibility to CDI in IL-10 deficient C57BL/6 mice [@AbernathyClose2021].
-In our experiments, when *Helicobacter* was present, the infection resulted in a high histopathologic score.
+In our experiments, when *Helicobacter* was present, the infection resulted in a high histopathologic score (Figure 4C).
 While we did not use IL-10 deficient mice, it is possible the bacterial community or host response are similarly modified by *Helicobacter*, allowing *C.
@@
 difficile*, and the indigenous microbes at the site of infection.
-In the first set of experiments, we used the same host, C57BL/6 mice, the same infecting microbe, *C.
-difficile* RT027 human isolate 431, with different gut bacterial communities.
+In the first set of experiments, we used the same host background, C57BL/6 mice, the same infecting microbe, *C.
+difficile* RT027 clinical isolate 431, with different gut bacterial communities.
 The bacterial groups in those communities were often present in both moribund and non-moribund and across the range of histopathologic scores.
@@
 *C.
-difficile* can drive host damage through the production of toxin.
-The gut microbiota can contribute to the host damage through the balance of metabolic and competitive activities, such as bacteriocin production or mucin degradation.
+difficile* drives host damage through the production of toxin.
+The gut microbiota can modulate host damage through the balance of metabolic and competitive interactions with *C.
+difficile*, such as bacteriocin production or mucin degradation, and interactions with the host, such as host mucus glycosylation or intestinal IL-33 expression [@Frisbee2019; @NagaoKitamoto2020].
 For example, low levels of mucin degradation can provide nutrients to other community members producing a diverse non-damaging community [@Tailford2015].
@@
 
-Here we have shown myriad bacterial groups and their relative abundances associated with variation in severity of CDI disease.
-Further understanding of how the microbiome affects severity in clinical patients could reduce the amount of adverse outcomes of CDI.
+Here we have shown several bacterial groups and their relative abundances associated with variation in CDI disease severity.
+Further understanding how the microbiome affects severity in patients could reduce the amount of adverse CDI outcomes.
 When a patient is diagnosed with CDI, the gut community composition, in addition to the traditionally obtained clinical information, may improve our severity prediction and guide prophylactic treatment.
-Treating the microbiome at the time of diagnosis may prevent the infection from becoming more severe.
-Based on the general trends of our results, promoting fiber metabolizing bacteria and reducing facultative anaerobes may bolster the efficacy of current CDI treatments, reducing disease severity.
+Treating the microbiome at the time of diagnosis, in addition to *C.
+difficile*, may prevent the infection from becoming more severe.
 
@@
 **Animal care.** 6- to 13-week old male and female germ-free C57BL/6 were obtained from a single breeding colony in the University of Michigan Germ-free Mouse Core.
-Mice were housed in cages of 2-4 mice per cage and maintained in germ-free isolators at the University of Michigan  germ-free facility.
+Mice (N1 n=11, N2 n=7, N3 n=3, N4 n=3, N5 n=3, N6 n=3, N7 n=7, N8 n=3, N9 n=2, M1 n=3, M2 n=3, M3 n=3, M4 n=3, M5 n=7, M6 n=3) were housed in cages of 2-4 mice per cage and maintained in germ-free isolators at the University of Michigan  germ-free facility.
 All mouse experiments were approved by the University Committee on Use and Care of Animals at the University of Michigan.
-**Were the mice maintained in gnotobiotic isolators throughout the experiment?
-Or were they transferred to BSL2 conditions prior to administering the human FMTs?**
 
 ***C.
-difficile* experiments.** Human fecal samples were obtained as part of Schubert et al.
-and selected based on community clusters [@Schubert2014] to result in diverse community structures.
+difficile* experiments.** Human fecal samples were obtained as part of Schubert *et al.* and selected based on community clusters [@Schubert2014] to result in diverse community structures.
 Feces were homogenized by mixing 200 mg of sample with 5 ml of PBS.
 Mice were inoculated with 100 $\mu$l of the fecal homogenate via oral gavage.
-For the set of experiments testing multiple *C.
-difficile* clinical isolates, we selected a subset of donor fecal samples based on their associated mouse transplant communities and the outcomes in the experiments with *C.
-difficile* RT027 isolate 431 that fit the following criteria: one community we would expect to result in a low histopathologic score (< 5), a high histopathologic score (> 5), and a high histopathologic score which becomes moribund.
-Additionally, we selected the donors based their relative abundances of genera identified through the our analysis (Figure S3).
 Two weeks after the fecal community inoculation, mice were challenged with *C.
@@
 *C.
-difficile* isolates came from Carlson et al.
-which had previously been isolated and characterized [@Carlson2013; @Carlson2015] and have recently been further characterized [@Saund2021].
+difficile* isolate came from Carlson *et al.* which had previously been isolated and characterized [@Carlson2013; @Carlson2015] and has recently been further characterized [@Saund2021].
 Spores concentration were determined both before and after challenge [@Sorg2009].
@@
 difficile* spores were given to each mouse via oral gavage.
-**How were human FMTs prepared (anaerobically vs aerobically, etc)?**
 
@@
 difficile* challenge and the following 10 days.
-Each day, a fecal sample was also collected and weighed and the remaining sample was frozen at -20$^\circ$C.
+Each day, a fecal sample was collected and a portion was weighed for plating (approximately 30 mg) and the remaining sample was frozen at -20$^\circ$C.
 Anaerobically, the weighed fecal samples were serially diluted in PBS, plated on TCCFA plates, and incubated at 37$^\circ$C for 24 hours.
@@
 We tested for significant differences in $\beta$-diversity ($\theta$~YC~) using the Wilcoxon rank sum test.
-We used Spearman's correlation to identified which OTUs that had a correlation between their relative abundance and the histopathologic summary score.
+We used Spearman's correlation to identify which OTUs that had a correlation between their relative abundance and the histopathologic summary score.
 *P* values were then corrected for multiple comparisons with a Benjamini and Hochberg adjustment for a type I error rate of 0.05 [@Benjamini1995].
@@
 The split for training and testing varied by model to avoid overfitting the data.
+To determine the optimal split, we tested splits (50%, 60%, 70%, 80%, 90% data used for training) to find the greatest portion of data that could be used to train the model while still maintaining the same performance for the training model as the model with the held-out test data.
 The toxin and moribundity models were trained with 60% of the data.
 The histopathologic score model was trained with 80% of the data.
-Lastly, we did not analyze murine communities to donor community or clinical data because germ-free mice colonized with human fecal communities become more like indigenous murine communities once stabilized.
-Furthermore, we are not trying to make any claims about human bacteria and their association with human CDI.
+Lastly, we did not compare murine communities to donor community or clinical data because germ-free mice colonized with non-murine fecal communities have been shown to more closely resemble the recipient species than the donor species [@Rawls2006].
+Furthermore, it is not our intention to make any inferences regarding human associated bacteria and their relationship with human CDI outcome.
 
@@
 **Figure 1.
-Human fecal microbial communities established diverse gut bacterial communities in germ-free mice.** (A) Relative abundance of 10 most abundant bacterial classes of the gut bacterial communities on timepoint day 0, 14 days post-colonization of germ-free C57Bl/6 mice with human fecal samples.
+Human fecal microbial communities established diverse gut bacterial communities in germ-free mice.** (A) Relative abundances of the 10 most abundant bacterial classes observed in the feces of previously germ-free C57Bl/6 mice 14 days post-colonization with human fecal samples (i.e., day 0 relative to *C.
+difficile* challenge).
 Each column of abundances represents an individual mouse.
-Each group of mice that received the same donor are grouped together and labeled with a letter (N for non-moribund mice and M for moribund mice) and number.
+Mice that received the same donor feces are grouped together and labeled above with a letter (N for non-moribund mice and M for moribund mice) and number (ordered by mean histopathologic score).
++ indicates the mice which did not have detectable *C.
+difficile* CFU in Figure 2.
 (B) Median (points) and interquartile range (lines) of $\beta$-diversity ($\theta$~YC~) between an individual mouse and either all others which were inoculated with feces from the same donor or from a different donor.
@@
 difficile* CFU per gram of stool was measured the day after challenge with 10$^{3}$ *C.
-difficile* RT027 human isolate 431 spores and at the end of the experiment, 10 days post-challenge.
+difficile* RT027 clinical isolate 431 spores and at the end of the experiment, 10 days post-challenge.
 Each point represents an individual mouse.
-Mice are grouped by donor and labeled by the donor letter (N for non-moribund mice and M for moribund mice) and number.
+Mice are grouped by donor and labeled by the donor letter (N for non-moribund mice and M for moribund mice) and number (ordered by mean histopathologic score).
 Points are colored by donor group.
@@
 (B) Cecum scored for histopathologic damage from mice at the end of the experiment.
-Samples were collected for histopathologic scoring on day 10 post-challenge for non-moribund mice or the day the mouse succumbed to the infection for the moribund group (day 2-3 post-challenge).
+Samples were collected for histopathologic scoring on day 10 post-challenge for non-moribund mice or the day the mouse succumbed to the infection for the moribund group (day 2 or 3 post-challenge).
 Each point represents an individual mouse.
-Mice are grouped by donor and labeled by the donor letter (N for non-moribund mice and M for moribund mice) and number.
+Mice are grouped by donor and labeled by the donor letter (N for non-moribund mice and M for moribund mice) and number (ordered by mean histopathologic score).
 Points are colored by donor group.
-Missing points are from mice that had insufficient sample collected for assaying fecal toxin.
+Mice in group N1 that have a summary score of 0 are the mice which did not have detectable *C.
+difficile* CFU in Figure 2.
+Missing points are from mice that had insufficient fecal sample collected for assaying toxin or cecum for histopathologic scoring.
 
@@
 (A) Relative abundances were compared across infection outcome of moribund (colored black) or non-moribund with either a high histopathologic score (score greater than the median score of 5, colored green) or a low histopathologic summary score (score less than the median score of 5, colored light green).
-(B) Relative abundances were compared between mice which toxin activity was detected (Toxigenic, colored dark purple) and which no toxin activity was detected (Non-toxigenic, colored light purple).
+(B) Relative abundances were compared between mice which toxin activity was detected (Toxin +, colored dark purple) and which no toxin activity was detected (Toxin -, colored light purple).
 (C) Endpoint bacterial OTUs correlated with histopathologic summary score.
@@
 Spearman's correlations were statistically significant after Benjamini-Hochberg correction for multiple comparisons.
-* next to a name indicates that the bacterial group was unclassified at lower classification ranks.
+All bacterial groups are ordered alphabetically.
+* indicates that the bacterial group was unclassified at lower taxonomic classification ranks.
 
@@
 Bacterial groups are ordered by their importance to the model; taxonomic group at the top of the plot had the greatest decrease in performance when its relative abundances were permuted.
-* next to the name indicates that the bacterial group was unclassified at lower classification ranks.
-(A) Bacterial members grouped by phyla predicted which mice would have toxin activity detected at any point throughout the infection (Toxigenic, dark purple).
+* indicates that the bacterial group was unclassified at lower taxonomic classification ranks.
+(A) Bacterial members grouped by phyla predicted which mice would have toxin activity detected at any point throughout the infection (Toxin +, dark purple).
 (B) Bacterial members grouped by class predicted which mice would become moribund (dark blue).
 (C) Bacterial members grouped by genera predicted if the mice would have a high (score greater than the median score of 5, colored dark green) or low (score less than the median score of 5, colored light green) histopathologic summary score.
-
-\hfill\break
-
-![](../results/figures/figure_6.jpg)
-
-**Figure 6.
-*C.
-difficile* infection outcome were similar within human fecal donor groups across different *C.
-difficile* isolates.** Mice colonized with human fecal donor communities N7, N6, or M4 were challenged with the isolate from the previous experiments in this study, *C.
-difficile* RT027 human isolate 431 (blue points), or a different *C.
-difficile* isolate (purple points).
-*C.
-difficile* CFUs were enumerated at the end of the experiment (day 10 post-challenge) and are plotted grouped by donor source (top plot).
-Histopathologic scores from the tissue collected at the endpoint (either day 10 post-challenge or day mice succumbed to infection) are plotted grouped by donor source (bottom plot).
-Each point represents an individual mouse.
-Mice that had received fecal communities from human donor M4 succumbed to the infection prior to day 10 post-challenge so were not plated for CFU on day 10 post-challenge (DECD = deceased).
-RT = ribotype, Isolate = *C.
-difficile* human isolate id, LOD = limit of detection.
 
@@
 difficile* challenge (Day 0) of the bacterial community members grouped by different classification rank were modeled with random forest to predict the infection outcome.
-The models used the highest taxonomic classification rank that performed as well as the lower ranks.
+The models used the highest taxonomic classification rank performed as well as the lower ranks.
 Black rectangle highlights classification rank used to model each outcome.
@@
 Individual relative abundances were added to F since differences in AUC were outside the interquartile range.
-* indicates bacterial group was unclassified at lower classification ranks.
+* indicates bacterial group was unclassified at lower taxonomic classification ranks.
 For all plots, median (solid points) and interquartile range (lines) are plotted.
@@
 (D) Bacterial genera which affected the performance of predicting histopathologic score when permuted.
-
-\hfill\break
-
-![](../results/figures/figure_S3.jpg)
-
-**Figure S3.
-Subset of donor communities were selected to test for conserved *C.
-difficile* infection outcome with a different *C.
-difficile* isolate.** (A) Relative abundance of genera in mice from previous experiments which received human fecal donor communities N7, N6, or M4.
-Each column represents an individual mouse.
-(B) Features from the random forest model that predicted moribundity using relative abundance of bacteria were grouped by their genus classification rank and are ordered by their decrease in AUC when each feature was permuted.
-Median (solid points) and interquartile range (lines) change in AUC are plotted.
-(C) Relative abundance of the most important genera used by the model to predict moribundity.
-Median (solid points) and interquartile range (lines) of the group relative abundances are plotted.
-* next to a name indicates that the bacterial group was unclassified at lower classification ranks.

### 2022-01-22 submission/manuscript.tex
incorporate coauthor feedback

@@
 
-Nicholas A. Lesniak\(^1\), Alyxandria M. Schubert\(^1\), Kaitlyn
-Flynn\(^1\), Jhansi L Leslie\(^1\), Hamide Sinani\(^1\), Ingrid L.
-Bergin\(^3\), Vincent B Young\(^2\), Patrick D. Schloss\(^{1,\dagger}\)
+Nicholas A. Lesniak\(^1\), Alyxandria M. Schubert\(^1\), Kaitlyn J.
+Flynn\(^1\), Jhansi L. Leslie\(^{1,4}\), Hamide Sinani\(^1\), Ingrid L.
+Bergin\(^3\), Vincent B. Young\(^{1,2}\), Patrick D.
+Schloss\(^{1,\dagger}\)
 
@@
 3. Unit for Laboratory Animal Medicine, University of Michigan, Ann
-Arbor, MI
+Arbor, MI\\
+4. Current affiliation: Department of Medicine, Division of
+International Health and Infectious Diseases, University of Virginia
+School of Medicine, Charlottesville, Virginia, USA
 
@@
 toxin genes have been associated with disease severity. However, it is
-unclear whether there is an associate between the gut microbiota disease
-severity. The gut microbiota is known to interact with \emph{C.
+unclear whether there is an association between the gut microbiota and
+disease severity. The gut microbiota is known to interact with \emph{C.
 difficile} during infection. Perturbations to the gut microbiota are
@@
 bile acid metabolism, nutrient consumption and bacteriocin production.
-It has not been shown whether the gut microbiota can affect CDI disease
-severity. Here we demonstrate that gut bacterial communities can
-contribute to disease severity. We derived diverse gut communities by
-colonizing germ-free mice with different human fecal communities. The
-mice were then infected with a single \emph{C. difficile} ribotype 027
-clinical isolate which resulted in moribundity and histopathologic
-differences. The variation in severity was associated with the human
-fecal community that the mice received. Generally, facultative anaerobes
-with pathogenic potential, such as \emph{Escherichia},
-\emph{Helicobacter}, and \emph{Klebsiella}, were associated with more
-severe outcomes. Bacterial groups associated with fiber degradation,
-such as \emph{Coprobacillus}, were associated with less severe outcomes.
-Finally, we showed that, when infected with a different \emph{C.
-difficile} isolate, the outcome of the infection was similar between the
-mice that received the same fecal donor community. These data indicate
-that gut bacteria can influence CDI disease severity.
+Here we sought to demonstrate that members of the gut bacterial
+communities can contribute to disease severity. We derived diverse gut
+communities by colonizing germ-free mice with different human fecal
+communities. The mice were then infected with a single \emph{C.
+difficile} ribotype 027 clinical isolate which resulted in moribundity
+and histopathologic differences. The variation in severity was
+associated with the human fecal community that the mice received.
+Generally, bacterial populations with pathogenic potential, such as
+\emph{Escherichia}, \emph{Helicobacter}, and \emph{Klebsiella}, were
+associated with more severe outcomes. Bacterial groups associated with
+fiber degradation, bile acid metabolism and lantibiotic production, such
+as \emph{Anaerostipes} and \emph{Coprobacillus}, were associated with
+less severe outcomes. These data indicate that gut bacteria can
+influence CDI disease severity.
 
@@
 
-\emph{Clostridioides difficile} infections (CDI) can be asymptomatic or
-range in severity from mild diarrhea to severe outcomes including
-surgical interventions and death. Models that predict severity and guide
-treatment decisions are based on clinical factors and \emph{C.
+\emph{Clostridioides difficile} colonization can result in no symptoms
+or develop into an infection, ranging in severity from mild diarrhea to
+toxic megacolon, sepsis and death. Models that predict severity and
+guide treatment decisions are based on clinical factors and \emph{C.
 difficile} characteristics. Although the gut microbiome plays a role in
@@
 risk of developing more severe disease and improve interventions which
-treat the gut bacteria to reduce host damage.
+target \emph{C. difficile} and the gut bacteria to reduce host damage.
 
@@
 and Society for Healthcare Epidemiology of America (SHEA) guidelines
-categorize severe disease based on white blood cell count greater than
-15,000 cells/mm\(^3\) and/or a serum creatinine greater than 1.5 mg/dL,
-and complicated disease based on systolic blood pressure below 90,
-ileus, or toxic megacolon (2). Since these measures are only present
-once the disease has advanced, the IDSA classification does not allow
-for identification of patients at risk of severe CDI. Prediction models
-have been developed to score a patient's risk for severe CDI outcomes
-based on clinical factors but have not been robust for broad application
-(3--6). Thus, we have limited ability to prevent patients from
-developing severe CDI. Ideally, we could identify patients that are
-high-risk of severe disease so their treatments could be adjusted to
-halt the infection before the disease worsens.
+defines severe CDI in terms of a white blood cell count greater than
+15,000 cells/mm\(^3\) and/or a serum creatinine greater than 1.5 mg/dL.
+Patients who develop shock or hypotension, ileus, or toxic megacolon are
+considered to have fulminant CDI (2). Since these measures are outcomes
+of CDI, these measures have limited ablity to predict risk of severe CDI
+when CDI is fiest detected. Schemes have been developed to score a
+patient's risk for severe CDI outcomes based on clinical factors but
+have not been robust for broad application (3). Thus, we have limited
+ability to prevent patients from developing severe CDI. Ideally, we
+could identify patients that are high-risk of severe disease so their
+treatments could be adjusted to alter the clinical trajectory.
 
@@
 including antibiotics, medications, or dietary changes, and lead to
-increased susceptibility to CDI (7--9). Once in the gut, the indigenous
-bacteria can either promote or inhibit \emph{C. difficile} through
-producing molecules or modifying the environment (10, 11). Bile acids
-metabolized by the gut bacteria can inhibit \emph{C. difficile} growth
-and affect toxin production (12, 13). Bacteria in the gut also can
+increased susceptibility to CDI (4--6). Once \emph{C. difficile}
+overcomes the protective barrier and colonizes the intestine, the
+indigenous bacteria can either promote or inhibit \emph{C. difficile}
+through producing molecules or modifying the environment (7, 8). Bile
+acids metabolized by the gut bacteria can inhibit \emph{C. difficile}
+growth and affect toxin production (9, 10). Bacteria in the gut also can
 compete more directly with \emph{C. difficile} through antibiotic
-production or nutrient consumption (14, 15). While the relationship
+production or nutrient consumption (11--13). While the relationship
 between the gut bacteria and \emph{C. difficile} has been established,
 the effect the gut bacteria can have on CDI disease severity is unclear.
-
-Studies investigating the severity of CDI disease involving the
-microbiome have not shown clear evidence of a relationship between the
-microbiome and disease severity. These studies were not setup to
-identify a relationship between the microbiome and disease severity.
-Studies that have used clinical data cannot control for host, microbiome
-or \emph{C. difficile} ribotype variation (16). While disease severity
-varies, it cannot be attributed to specifically the microbiome since all
-factors vary. Murine experiments typically use a single mouse colony and
-different \emph{C. difficile} ribotypes to create severity differences
-(3). Recently, our group has begun uncovering the effect microbiome
-variation has on \emph{C. difficile} infection. We showed the variation
-in the bacterial communities between mice from different mouse colonies
-resulted in different clearance rates of \emph{C. difficile} (17). We
-also showed varied ability of mice to spontaneously eliminate \emph{C.
-difficile} infection when they were treated with different antibiotics
-prior to \emph{C. difficile} challenge (18). However, there have not
-been any studies investigating whether variation in the microbiome
-affects the resultant disease severity of \emph{C. difficile}
-infections.
 
@@
 also affect the disease severity of the infection. To test this
-hypothesis, we colonized germ-free C57BL/6 mice with human fecal
-communities to create varied gut communities. We then challenged the
-mice with a single \emph{C. difficile} isolate, ribotype 027 (RT027)
-human isolate 431, and followed the mice for the development of severe
+hypothesis, we colonized germ-free C57BL/6 mice with human fecal samples
+to create varied gut communities. We then challenged the mice with a
+single \emph{C. difficile} isolate, ribotype 027 (RT027) clinical
+isolate 431, and followed the mice for the development of severe
 outcomes of moribundity and histopathologic cecal tissue damage. Since
@@
 \textbf{\emph{C. difficile} is able to infect germ-free mice colonized
-with human fecal microbial communities without perturbation.} To produce
+with human fecal microbial communities without antibiotics} To produce
 gut microbiomes with greater variation than those found in conventional
 mouse colonies, we colonized germ-free mice with bacteria from human
-feces (19). We inoculated germ-free C57BL/6 mice with homogenized feces
-from one of 15 human fecal samples via oral gavage. The human fecal
+feces (14). We inoculated germ-free C57BL/6 mice with homogenized feces
+from one of 15 human fecal samples via oral gavage. These human fecal
 samples were selected because they represented diverse community
-structures based on community clustering (20). The gut communities were
-allowed to establish themselves for two weeks post inoculation (21). We
-then surveyed the bacterial members of the gut communities by 16S rRNA
-gene sequencing of fecal pellets (Figure 1A). The bacterial communities
-from each mouse grouped more closely to those communities from mice that
-received the same human fecal donor community than to the mice who
-received a different human fecal donor community (Figure 1B). The
-communities were primarily composed of populations of \emph{Clostridia},
-\emph{Bacteroidia}, \emph{Erysipelotrichia}, \emph{Bacilli}, and
-\emph{Gammaproteobacteria}. However, the gut bacterial communities of
-each donor group of mice harbored unique relative abundance
-distributions of the shared bacterial classes.
+structures based on community clustering (15). The gut communities were
+allowed to equlibriate for two weeks post inoculation (16). We then
+surveyed the bacterial members of the gut communities by 16S rRNA gene
+sequencing of murine fecal pellets (Figure 1A). The bacterial
+communities from each mouse grouped more closely to those communities
+from mice that received the same human fecal donor community than to the
+mice who received a different human fecal donor community (Figure 1B).
+The communities were primarily composed of populations of
+\emph{Clostridia}, \emph{Bacteroidia}, \emph{Erysipelotrichia},
+\emph{Bacilli}, and \emph{Gammaproteobacteria}. However, the gut
+bacterial communities of each donor group of mice harbored unique
+relative abundance distributions of the shared bacterial classes.
 
@@
 communities for susceptibility to \emph{C. difficile} infection. A
-typical mouse model of CDI requires pre-treatment with antibiotics, such
-as clindamycin, to become susceptible to \emph{C. difficile}
-colonization (22, 23). However, we wanted to avoid modifying the
-communities with an antibiotic to maintain their unique microbial
-compositions and ecological relationships. Since these communities came
-from people who had increased risk of CDI, such as hospitalization or
-recent antibiotic use (20), we decided to test if \emph{C. difficile}
-was able to infect these mice without any antibiotic perturbation. We
-hypothesized that \emph{C. difficile} would be able to colonize a subset
-of mice whose communities were susceptible from perturbations either the
-donor experienced or from the ecological selection when colonizing the
-murine gut. Mice were challenged with 10\(^{3}\) \emph{C. difficile}
-RT027 human isolate 431 spores. The mice were followed for 10 days
-post-challenge and their stool was collected and plated for \emph{C.
-difficile} colony forming units (CFU) to determine the extent of the
-infection. Surprisingly, mice from all donors were colonized (Figure 2).
-Only two mice were able to resist \emph{C. difficile} infection and both
-received their community from the same donor, N3. By colonizing
-germ-free mice with different human fecal communities, we were able to
-generate diverse gut communities in mice, which were susceptible to
-\emph{C. difficile} infection without further modification of the gut
-community.
+typical mouse model of CDI requires pre-treatment of conventional mice
+with antibiotics, such as clindamycin, to become susceptible to \emph{C.
+difficile} colonization (17, 18). However, we wanted to avoid modifying
+the gut communities with an antibiotic to maintain their unique
+microbial compositions and ecological relationships. Since some of these
+communities came from people at increased risk of CDI, for example due
+to hospitalization or recent antibiotic use (15), we decided to test if
+\emph{C. difficile} was able to infect these mice without any antibiotic
+perturbation. We hypothesized that \emph{C. difficile} would be able to
+colonize the mice who received their gut communities from a donor with
+perturbed community. Mice were challenged with 10\(^{3}\) \emph{C.
+difficile} RT027 clinical isolate 431 spores. The mice were followed for
+10 days post-challenge, and their stool was collected and plated for
+\emph{C. difficile} colony forming units (CFU) to determine the extent
+of the infection. Surprisingly, communities from all donors were able to
+be colonized (Figure 2). Two mice were able to resist \emph{C.
+difficile} colonization, both received their community donor N3, which
+may be attributed to experimental variation since this group also had
+more mice. By colonizing germ-free mice with different human fecal
+communities, we were able to generate diverse gut communities in mice,
+which were susceptible to \emph{C. difficile} infection without further
+modification of the gut community.
 
@@
 colonization density, toxin production, and mortality. Seven mice, from
-Donors N1, N2 and N5, were not colonized on the day after \emph{C.
-difficile} challenge but were infected (\textgreater10\(^{6}\)) by the
-end of the experiment. All mice that received their community from Donor
-M1 through M6 succumbed to the infection and became moribund within 3
-days post-challenge. The remaining mice, except the uninfected Donor N3
-mice, maintained \emph{C. difficile} infection through the end of the
-experiment (Figure 2). At 10 days post-challenge, or earlier for the
-moribund mice, mice were euthanised and fecal pellets were assayed for
-toxin activity and cecal tissue was collected and scored for
-histopathologic signs of disease (Figure 3). Overall, there was greater
-toxin activity detected in the stool of the moribund mice (\emph{P} =
-0.003). However, when looking at each group of mice, we observed a range
-in toxin activity for both the moribund and non-moribund mice (Figure
-3A). Non-moribund mice from Donors N4 through N9 had comparable toxin
-activity as the moribund mice. Additionally, not all moribund mice had
-toxin activity detected in their stool. This discordance between toxin
-activity and disease severity has been previously described (24, 25).
-Next we examined the cecal tissue for histopathologic damage. Moribund
-mice had high levels of epithelial damage, tissue edema, and
+Donors N1, N2 and N5, were not colonized at detectable levels on the day
+after \emph{C. difficile} challenge but were infected
+(\textgreater10\(^{6}\)) by the end of the experiment. All mice that
+received their community from Donor M1 through M6 succumbed to the
+infection and became moribund within 3 days post-challenge. The
+remaining mice, except the uninfected Donor N3 mice, maintained \emph{C.
+difficile} infection through the end of the experiment (Figure 2). At 10
+days post-challenge, or earlier for the moribund mice, mice were
+euthanised and fecal material were assayed for toxin activity and cecal
+tissue was collected and scored for histopathologic signs of disease
+(Figure 3). Overall, there was greater toxin activity detected in the
+stool of the moribund mice (\emph{P} = 0.003). However, when looking at
+each group of mice, we observed a range in toxin activity for both the
+moribund and non-moribund mice (Figure 3A). Non-moribund mice from
+Donors N4 through N9 had comparable toxin activity as the moribund mice.
+Additionally, not all moribund mice had toxin activity detected in their
+stool. Next we examined the cecal tissue for histopathologic damage.
+Moribund mice had high levels of epithelial damage, tissue edema, and
 inflammation (Figure S1) similar to previously reported histopathologic
-findings for \emph{C. difficile} RT027 (26). As observed with toxin
+findings for \emph{C. difficile} RT027 (19). As observed with toxin
 activity, the moribund mice had higher histopathologic scores than the
@@
 dichotomized the histopathologic scores into high and low groups by
-splitting on the median score. This analysis revealed 20 genera that
-were significantly different by the disease severity (Figure 4A).
+splitting on the median score of 5. This analysis revealed 20 genera
+that were significantly different by the disease severity (Figure 4A).
 Bacterial genera \emph{Turicibacter}, \emph{Streptococcus},
@@
 severity, we applied LEfSe to identify the genera most likely to explain
-the observed differences in toxin activity (Figure 4B). Many genera that
-associated with toxin production were also associated with moribundity,
-such as populations of \emph{Escherichia/Shigella} and
-\emph{Bacteroides}. Likewise, there were genera such as
-\emph{Anaerotignum}, \emph{Enterocloster}, and \emph{Murimonas} that
-were associated with no toxin production that also exhibited greater
-relative abundance in communities from non-moribund mice with a low
-histopathologic score. Lastly, we tested for correlations between the
-endpoint relative abundances of bacterial operational taxonomic units
-(OTUs) and the histopathologic summary score (Figure 4C). The endpoint
-relative abundance of \emph{Bacteroides} was positively correlated with
-histopathologic score, as its day 0 relative abundance did with disease
-severity (Figure 4A). Populations of \emph{Klebsiella} and
-\emph{Prevotellaceae} were positively correlated with the
-histopathologic score and were increased in the group of mice with
-detectable toxin. This analysis identified bacterial genera that were
-associated with the variation in moribundity, histopathologic score, and
-toxin activity of CDI.
+the differences between the presence and absence of detected toxin
+activity (Figure 4B). Many genera that associated with toxin production
+were also associated with moribundity, such as populations of
+\emph{Escherichia/Shigella} and \emph{Bacteroides}. Likewise, there were
+genera such as \emph{Anaerotignum}, \emph{Enterocloster}, and
+\emph{Murimonas} that were associated with no detected toxin production
+that also exhibited greater relative abundance in communities from
+non-moribund mice with a low histopathologic score. Lastly, we tested
+for correlations between the endpoint relative abundances of bacterial
+operational taxonomic units (OTUs) and the histopathologic summary score
+(Figure 4C). The endpoint relative abundance of \emph{Bacteroides} was
+positively correlated with histopathologic score, as its day 0 relative
+abundance did with disease severity (Figure 4A). Populations of
+\emph{Klebsiella} and \emph{Prevotellaceae} were positively correlated
+with the histopathologic score and were increased in the group of mice
+with detectable toxin. This analysis identified bacterial genera that
+were associated with the variation in moribundity, histopathologic
+score, and toxin activity of CDI.
 
@@
 
-\textbf{Disease severity consistent by donor community across isolates.}
-Since Carlson et al.~showed \emph{C. difficile} isolates with lower
-germination rates were more likely to result in severe disease (27), we
-next tested whether the effect of the community on CDI outcomes was
-conserved across \emph{C. difficile} isolates with higher rates of
-germination. We colonized mice with human fecal slurry of either Donor
-N7, N6, or M4 and allowed two weeks for the community to establish. We
-then challenged the mice with a different \emph{C. difficile} isolate
-and all mice were infected. The moribundity and histopathologic score
-outcomes when challenged with a different \emph{C. difficile} human
-isolate were comparable to the outcomes of challenges with \emph{C.
-difficile} RT027 human isolate 431 by their donor group (Figure 6). For
-the isolates and communities we tested, the effect of the gut bacterial
-community on disease severity was greater than the effect of the
-variation in these isolates.
-
 \hypertarget{discussion}{%
@@
 relative abundance of bacterial community members and disease severity.
-We used the relative abundance patterns of community members associated
-with disease outcomes to select donor communities to test whether the
-disease severity was conserved when challenged with a different \emph{C.
-difficile} isolate. The disease outcomes from the challenge with the
-second \emph{C. difficile} isolate matched the outcomes of the challenge
-with \emph{C. difficile} RT027 isolate 431. Therefore, the effect of the
-community variation on disease severity \emph{in vivo} was greater than
-the effect of \emph{C. difficile} isolate variation (27, 28). Overall,
-these results have demonstrated that the bacterial community contributed
-to the disease severity of \emph{C. difficile} infection.
-
-CDI severity can range from no symptoms to mild symptoms, such as
-diarrhea, to severe symptoms, such as toxic megacolon or death.
-Physicians use prediction tools to identify patients most at risk of
-developing a severe infection using white blood cell counts, serum
-albumin level, or serum creatinine level (2, 29, 30). Those levels are
-driven by the activities in the intestine (31). Research into the
-drivers of this variation have revealed factors that make \emph{C.
-difficile} more virulent. Strains are categorized for their virulence by
-the presence and production toxin TcdA, toxin TcdB, and binary toxin and
-the prevalence in outbreaks, such as ribotypes 027 and 078 (22, 32--35).
-However, other studies have shown that virulence is not necessarily
-linked with toxin production (36) or the strain (37). Furthermore, there
-is variation in the genome, growth rate, sporulation, germination, and
-toxin production in different isolates of a strain (27, 28, 38, 39).
-This variation may help explain why severe CDI prediction tools often
-miss identifying many patients with CDI that will develop severe disease
-(3--6). Therefore, it is necessary to gain a full understanding of all
-factors contributing to disease variation to improve our ability to
+Previous studies investigating the severity of CDI disease involving the
+microbiome have had limited ability to interrogate this relationship
+between the microbiome and disease severity. Studies that have used
+clinical data have limited ability to control variation in the host,
+microbiome or \emph{C. difficile} ribotype (20). Murine experiments
+typically use a single mouse colony and different \emph{C. difficile}
+ribotypes to create severity differences (21). Recently, our group has
+begun uncovering the effect microbiome variation has on \emph{C.
+difficile} infection. We showed the variation in the bacterial
+communities between mice from different mouse colonies resulted in
+different clearance rates of \emph{C. difficile} (22). We also showed
+varied ability of mice to spontaneously eliminate \emph{C. difficile}
+infection when they were treated with different antibiotics prior to
+\emph{C. difficile} challenge (23). Overall, the results presented here
+have demonstrated that the gut bacterial community contributed to the
+severity of \emph{C. difficile} infection.
+
+\emph{C. difficile} can lead to asymptomatic colonization or infections
+with severity ranging from mild diarrhea to death. Physicians use
+classification tools to identify patients most at risk of developing a
+severe infection using white blood cell counts, serum albumin level, or
+serum creatinine level (2, 24, 25). Those levels are driven by the
+activities in the intestine (26). Research into the drivers of this
+variation have revealed factors that make \emph{C. difficile} more
+virulent. Strains are categorized for their virulence by the presence
+and production of the toxins TcdA, TcdB, and binary toxin and the
+prevalence in outbreaks, such as ribotypes 027 and 078 (17, 27--30).
+However, other studies have shown that disease is not necessarily linked
+with toxin production (31) or the strain (32). Furthermore, there is
+variation in the genome, growth rate, sporulation, germination, and
+toxin production in different isolates of a strain (33--36). This
+variation may help explain why severe CDI prediction tools often miss
+identifying many patients with CDI that will develop severe disease (3,
+21, 37, 38). Therefore, it is necessary to gain a full understanding of
+all factors contributing to disease variation to improve our ability to
 predict severity.
@@
 difficile} is unable to colonize an unperturbed healthy gut community
-and is only able to become established after a perturbation (23). Once
+and is only able to become established after a perturbation (18). Once
 colonized, the different communities lead to different metabolic
-responses and dynamics of the \emph{C. difficile} population (11, 18,
-40). The gut bacteria metabolize primary bile acids into secondary bile
-acids (41, 42). The concentration of these bile acids affect the
-germination, growth, toxin production and biofilm formation (12, 13, 43,
-44). Members of the bacterial community also affect other metabolites
+responses and dynamics of the \emph{C. difficile} population (8, 23,
+39). Gut bacteria metabolize primary bile acids into secondary bile
+acids (40, 41). The concentration of these bile acids affects
+germination, growth, toxin production and biofilm formation (9, 10, 42,
+43). Members of the bacterial community also affect other metabolites
 \emph{C. difficile} utilizes. \emph{Bacteroides thetaiotaomicron}
 produce sialidases which release sialic acid from the mucosa for
-\emph{C. difficile} to utilize (45, 46). The nutrient environment
-affects toxin production (47). Thus, many of the actions of the gut
+\emph{C. difficile} to utilize (44, 45). The nutrient environment
+affects toxin production (46). Thus, many of the actions of the gut
 bacteria modulate \emph{C. difficile} in ways that could affect the
@@
 
-Myriad studies have explored the relationship between the microbiome and
-CDI but none have established a relationship between the variation of
-disease severity and the microbiome. CDI studies often use an infection
-model with a single homogeneous bacterial community. Collins et al.~used
-multiple human communities to colonize mice, however the communities
-were pooled prior to gavaging into germ-free mice (48), resulting in a
-single community. Studies examining difference in disease often use
-different \emph{C. difficile} strains or ribotypes in mice with similar
-microbiota as a proxy for variation in disease, such as strain 630 for
-non-severe and RT027 for severe (22, 32, 33, 49). Studies have also
-demonstrated variation in infection through tapering antibiotic dosage
-(18, 23, 50) or by reducing the amount of \emph{C. difficile} cells or
-spores used for the challenge (22, 50). These studies either lack
-variation in the initial microbiome or have variation in the \emph{C.
-difficile} infection itself, therefore we cannot make any association
-between variation in severity and the microbiome.
-
-With our recent observations that the initial community affected the
-ability of \emph{C. difficile} to persist in the gut (17, 18) and the
-existing research describing the myriad interactions between the
-microbiome and \emph{C. difficile}, we hypothesized that the microbiome
-can modulate CDI disease severity. Our data have demonstrated gut
-bacterial relative abundances associate with variation in toxin
-production, histopathologic scoring of the cecal tissue and mortality.
-This analysis revealed populations of \emph{Akkermansia},
-\emph{Anaerostipes}, \emph{Coprobacillus}, \emph{Enterocloster},
-\emph{Lactonifactor}, and \emph{Monoglobus} were more abundant in the
-microbiome of non-moribund mice which had low histopathologic scores and
-no detected toxin. The protective role of these genera are supported by
-previous studies. \emph{Coprobacillus}, \emph{Lactonifactor}, and
-\emph{Monoglobus} have been shown to be involved in dietary fiber
-fermentation and associated with healthy communities (51--54).
-\emph{Anaerostipes} and \emph{Coprobacillus}, which produce short chain
-fatty acids, have been associated with healthy communities (55--57).
-Furthermore, \emph{Coprobacillus}, which was abundant in mice with low
-histopathologic scores but rare in all other mice, has been shown to
-contain a putative type I lantibiotic gene cluster and inhibit \emph{C.
-difficile} colonization (58--60). \emph{Akkermansia} and
-\emph{Enterocloster} were also identified as more abundant in mice which
-had a low histopathologic scores but have contradictory supporting
-evidence in the current literature. In our data, \emph{Akkermansia} was
-most abundant in the non-moribund mice with low histopathologic scores
-but there were some moribund mice which had increased populations of
-\emph{Akkermansia}. This could be attributed to either a more protective
-mucus layer was present inhibiting colonization (60, 61) or mucus
-consumption by \emph{Akkermansia} could have been crossfeeding \emph{C.
-difficile} or exposing a niche for \emph{C. difficile} (62--64).
-Similarly, \emph{Enterocloster} was more abundant and associated with
-low histopathologic scores. It has been associated with healthy
-populations and has been used to mono-colonize germ-free mice to reduce
-the ability of \emph{C. difficile} to colonize (65, 66). However,
-\emph{Enterocloster} has also been involved in infections, such as
-bacteremia (67, 68). These data have exemplified populations of bacteria
-that have the potential to be either protective or harmful. Thus, the
-disease outcome is not likely based on the abundance of individual
-populations of bacteria, rather it is the result of the interactions of
-the community.
+A myriad of studies have explored the relationship between the
+microbiome and CDI disease. CDI studies often use an infection model
+with a single homogeneous bacterial community. Collins \emph{et al.}
+used multiple human communities to colonize mice, however the
+communities were pooled prior to gavaging into germ-free mice (47),
+resulting in a single human-derived community. Studies examining
+difference in disease often use different \emph{C. difficile} strains or
+ribotypes in mice with similar microbiota as a proxy for variation in
+disease, such as strain 630 for non-severe and RT027 for severe (17, 27,
+28, 48). Studies have also demonstrated variation in infection through
+tapering antibiotic dosage (18, 23, 49) or by reducing the amount of
+\emph{C. difficile} cells or spores used for the challenge (17, 49).
+These studies either lack variation in the initial microbiome or have
+variation in the \emph{C. difficile} infection itself, confounding any
+association between variation in severity and the microbiome.
+\textbf{Add Hikiro et al} With our recent observations that the initial
+community affected the ability of \emph{C. difficile} to persist in the
+gut (22, 23) and the existing research describing the numerous
+interactions between the microbiome and \emph{C. difficile}, we
+hypothesized that the microbiome can modulate CDI disease severity. Our
+data have demonstrated gut bacterial relative abundances associate with
+variation in toxin production, histopathologic scoring of the cecal
+tissue and mortality. This analysis revealed populations of
+\emph{Akkermansia}, \emph{Anaerostipes}, \emph{Coprobacillus},
+\emph{Enterocloster}, \emph{Lactonifactor}, and \emph{Monoglobus} were
+more abundant in the microbiome of non-moribund mice which had low
+histopathologic scores and no detected toxin. The protective role of
+these genera are supported by previous studies. \emph{Coprobacillus},
+\emph{Lactonifactor}, and \emph{Monoglobus} have been shown to be
+involved in dietary fiber fermentation and associated with healthy
+communities (50--53). \emph{Anaerostipes} and \emph{Coprobacillus},
+which produce short chain fatty acids, have been associated with healthy
+communities (54--56). Furthermore, \emph{Coprobacillus}, which was
+abundant in mice with low histopathologic scores but rare in all other
+mice, has been shown to contain a putative type I lantibiotic gene
+cluster and inhibit \emph{C. difficile} colonization (57--59).
+\emph{Akkermansia} and \emph{Enterocloster} were also identified as more
+abundant in mice which had a low histopathologic scores but have
+contradictory supporting evidence in the current literature. In our
+data, \emph{Akkermansia} was most abundant in the non-moribund mice with
+low histopathologic scores but there were some moribund mice which had
+increased populations of \emph{Akkermansia}. This could be attributed to
+either a more protective mucus layer was present inhibiting colonization
+(59, 60) or mucus consumption by \emph{Akkermansia} could have been
+crossfeeding \emph{C. difficile} or exposing a niche for \emph{C.
+difficile} (61--63). Similarly, \emph{Enterocloster} was more abundant
+and associated with low histopathologic scores. It has been associated
+with healthy populations and has been used to mono-colonize germ-free
+mice to reduce the ability of \emph{C. difficile} to colonize (64, 65).
+However, \emph{Enterocloster} has also been involved in infections, such
+as bacteremia (66, 67). These data have exemplified populations of
+bacteria that have the potential to be either protective or harmful.
+Thus, the disease outcome is not likely based on the abundance of
+individual populations of bacteria, rather it is the result of the
+interactions of the community.
 
@@
 \emph{Shigella/Escherichia}, \emph{Staphylococcus}, and
-\emph{Streptococcus} have been shown to expand after antibiotic use (20,
-69, 70) and are commonly detected in CDI cases (71--74). In addition to
+\emph{Streptococcus} have been shown to expand after antibiotic use (15,
+68, 69) and are commonly detected in CDI cases (70--73). In addition to
 these populations, \emph{Eggerthella}, \emph{Prevotellaceae} and
 \emph{Helicobacter}, which associated with worse outcomes, have also
-been associated with intestinal inflammation (75--77). Recently,
+been associated with intestinal inflammation (74--76). Recently,
 \emph{Helicobacter hepaticus} was shown to be sufficient to cause
-susceptibility to CDI in IL-10 deficient C57BL/6 mice (78). In our
+susceptibility to CDI in IL-10 deficient C57BL/6 mice (77). In our
 experiments, when \emph{Helicobacter} was present, the infection
-resulted in a high histopathologic score. While we did not use IL-10
-deficient mice, it is possible the bacterial community or host response
-are similarly modified by \emph{Helicobacter}, allowing \emph{C.
-difficile} infection and host damage. Aside from \emph{Helicobacter},
-these groups of bacteria that associated with more severe outcomes did
-not have a conserved association between their relative abundance and
-the disease severity across all mice.
+resulted in a high histopathologic score (Figure 4C). While we did not
+use IL-10 deficient mice, it is possible the bacterial community or host
+response are similarly modified by \emph{Helicobacter}, allowing
+\emph{C. difficile} infection and host damage. Aside from
+\emph{Helicobacter}, these groups of bacteria that associated with more
+severe outcomes did not have a conserved association between their
+relative abundance and the disease severity across all mice.
 
@@
 severe disease it may be appropriate to apply the damage-response
-framework for microbial pathogenesis to CDI (79, 80). This framework
+framework for microbial pathogenesis to CDI (78, 79). This framework
 posits that disease is not driven by a single entity, rather it is an
@@
 microbe, \emph{C. difficile}, and the indigenous microbes at the site of
-infection. In the first set of experiments, we used the same host,
-C57BL/6 mice, the same infecting microbe, \emph{C. difficile} RT027
-human isolate 431, with different gut bacterial communities. The
-bacterial groups in those communities were often present in both
-moribund and non-moribund and across the range of histopathologic
-scores. Thus, it was not merely the presence of the bacteria but their
-activity in response to the other microbes and host which affect the
-extent of the host damage. Additionally, while each mouse and \emph{C.
-difficile} population had the same genetic background, they too were
-reacting to the specific microbial community. Disease severity is driven
-by the cumulative effect of the host immune response and the activity of
-\emph{C. difficile} and the gut bacteria. \emph{C. difficile} can drive
-host damage through the production of toxin. The gut microbiota can
-contribute to the host damage through the balance of metabolic and
-competitive activities, such as bacteriocin production or mucin
-degradation. For example, low levels of mucin degradation can provide
+infection. In the first set of experiments, we used the same host
+background, C57BL/6 mice, the same infecting microbe, \emph{C.
+difficile} RT027 clinical isolate 431, with different gut bacterial
+communities. The bacterial groups in those communities were often
+present in both moribund and non-moribund and across the range of
+histopathologic scores. Thus, it was not merely the presence of the
+bacteria but their activity in response to the other microbes and host
+which affect the extent of the host damage. Additionally, while each
+mouse and \emph{C. difficile} population had the same genetic
+background, they too were reacting to the specific microbial community.
+Disease severity is driven by the cumulative effect of the host immune
+response and the activity of \emph{C. difficile} and the gut bacteria.
+\emph{C. difficile} drives host damage through the production of toxin.
+The gut microbiota can modulate host damage through the balance of
+metabolic and competitive interactions with \emph{C. difficile}, such as
+bacteriocin production or mucin degradation, and interactions with the
+host, such as host mucus glycosylation or intestinal IL-33 expression
+(80, 81). For example, low levels of mucin degradation can provide
 nutrients to other community members producing a diverse non-damaging
-community (81). However, if mucin degradation becomes too great it
+community (82). However, if mucin degradation becomes too great it
 reduces the protective function of the mucin layer and exposes the
@@
 
-Here we have shown myriad bacterial groups and their relative abundances
-associated with variation in severity of CDI disease. Further
-understanding of how the microbiome affects severity in clinical
-patients could reduce the amount of adverse outcomes of CDI. When a
-patient is diagnosed with CDI, the gut community composition, in
-addition to the traditionally obtained clinical information, may improve
-our severity prediction and guide prophylactic treatment. Treating the
-microbiome at the time of diagnosis may prevent the infection from
-becoming more severe. Based on the general trends of our results,
-promoting fiber metabolizing bacteria and reducing facultative anaerobes
-may bolster the efficacy of current CDI treatments, reducing disease
-severity.
+Here we have shown several bacterial groups and their relative
+abundances associated with variation in CDI disease severity. Further
+understanding how the microbiome affects severity in patients could
+reduce the amount of adverse CDI outcomes. When a patient is diagnosed
+with CDI, the gut community composition, in addition to the
+traditionally obtained clinical information, may improve our severity
+prediction and guide prophylactic treatment. Treating the microbiome at
+the time of diagnosis, in addition to \emph{C. difficile}, may prevent
+the infection from becoming more severe.
 
@@
 C57BL/6 were obtained from a single breeding colony in the University of
-Michigan Germ-free Mouse Core. Mice were housed in cages of 2-4 mice per
-cage and maintained in germ-free isolators at the University of Michigan
-germ-free facility. All mouse experiments were approved by the
-University Committee on Use and Care of Animals at the University of
-Michigan. \textbf{Were the mice maintained in gnotobiotic isolators
-throughout the experiment? Or were they transferred to BSL2 conditions
-prior to administering the human FMTs?}
+Michigan Germ-free Mouse Core. Mice (N1 n=11, N2 n=7, N3 n=3, N4 n=3, N5
+n=3, N6 n=3, N7 n=7, N8 n=3, N9 n=2, M1 n=3, M2 n=3, M3 n=3, M4 n=3, M5
+n=7, M6 n=3) were housed in cages of 2-4 mice per cage and maintained in
+germ-free isolators at the University of Michigan germ-free facility.
+All mouse experiments were approved by the University Committee on Use
+and Care of Animals at the University of Michigan.
 
 \textbf{\emph{C. difficile} experiments.} Human fecal samples were
-obtained as part of Schubert et al.~and selected based on community
-clusters (20) to result in diverse community structures. Feces were
-homogenized by mixing 200 mg of sample with 5 ml of PBS. Mice were
+obtained as part of Schubert \emph{et al.} and selected based on
+community clusters (15) to result in diverse community structures. Feces
+were homogenized by mixing 200 mg of sample with 5 ml of PBS. Mice were
 inoculated with 100 \(\mu\)l of the fecal homogenate via oral gavage.
-For the set of experiments testing multiple \emph{C. difficile} clinical
-isolates, we selected a subset of donor fecal samples based on their
-associated mouse transplant communities and the outcomes in the
-experiments with \emph{C. difficile} RT027 isolate 431 that fit the
-following criteria: one community we would expect to result in a low
-histopathologic score (\textless{} 5), a high histopathologic score
-(\textgreater{} 5), and a high histopathologic score which becomes
-moribund. Additionally, we selected the donors based their relative
-abundances of genera identified through the our analysis (Figure S3).
 Two weeks after the fecal community inoculation, mice were challenged
-with \emph{C. difficile}. \emph{C. difficile} isolates came from Carlson
-et al.~which had previously been isolated and characterized (27, 28) and
-have recently been further characterized (38). Spores concentration were
-determined both before and after challenge (82). \(10^{3}\) \emph{C.
-difficile} spores were given to each mouse via oral gavage. \textbf{How
-were human FMTs prepared (anaerobically vs aerobically, etc)?}
+with \emph{C. difficile}. \emph{C. difficile} isolate came from Carlson
+\emph{et al.} which had previously been isolated and characterized (33,
+34) and has recently been further characterized (35). Spores
+concentration were determined both before and after challenge (83).
+\(10^{3}\) \emph{C. difficile} spores were given to each mouse via oral
+gavage.
 
@@
 \emph{C. difficile} challenge and the following 10 days. Each day, a
-fecal sample was also collected and weighed and the remaining sample was
-frozen at -20\(^\circ\)C. Anaerobically, the weighed fecal samples were
-serially diluted in PBS, plated on TCCFA plates, and incubated at
-37\(^\circ\)C for 24 hours. The plates were then counted for the number
-of colony forming units (CFU) (83).
+fecal sample was collected and a portion was weighed for plating
+(approximately 30 mg) and the remaining sample was frozen at
+-20\(^\circ\)C. Anaerobically, the weighed fecal samples were serially
+diluted in PBS, plated on TCCFA plates, and incubated at 37\(^\circ\)C
+for 24 hours. The plates were then counted for the number of colony
+forming units (CFU) (84).
 
@@
 resulting amplicons using an Illumina MiSeq as described previously
-(84).
+(85).
 
 \textbf{Sequence curation.} Sequences were processed with
-mothur(v.1.44.3) as previously described (84, 85). In short, we used a
+mothur(v.1.44.3) as previously described (85, 86). In short, we used a
 3\% dissimilarity cutoff to group sequences into operational taxonomic
@@
 Database Project training set (version 18) to assign taxonomic
-classifications to each OTU (86). We sequenced a mock community of a
+classifications to each OTU (87). We sequenced a mock community of a
 known community composition and 16s rRNA gene sequences. We processed
@@
 Toxin activity was assessed using a Vero cell rounding-based
-cytotoxicity assay as described previously (33). The cytotoxicity titer
+cytotoxicity assay as described previously (28). The cytotoxicity titer
 was determined for each sample as the last dilution, which resulted in
@@
 criteria by a board-certified veterinary pathologist scored who was
-blinded to the experimental groups (33).
+blinded to the experimental groups (28).
 
@@
 structures, we calculated Yue and Clayton dissimilarity matrices
-(\(\theta\)\textsubscript{YC}) in mothur (87). We rarefied samples to
+(\(\theta\)\textsubscript{YC}) in mothur (88). We rarefied samples to
 2,107 sequences per sample to limit uneven sampling biases. We tested
 for differences in individual taxonomic groups that would explain the
-outcome differences with LEfSe (88) in mothur. Remaining statistical
+outcome differences with LEfSe (89) in mothur. Remaining statistical
 analysis and data visualization was performed in R (v4.0.5) with the
@@
 \(\beta\)-diversity (\(\theta\)\textsubscript{YC}) using the Wilcoxon
-rank sum test. We used Spearman's correlation to identified which OTUs
+rank sum test. We used Spearman's correlation to identify which OTUs
 that had a correlation between their relative abundance and the
@@
 multiple comparisons with a Benjamini and Hochberg adjustment for a type
-I error rate of 0.05 (89). We built random forest models using the
-mikropml package (90) with OTUs from day 0 samples using mtry values of
+I error rate of 0.05 (90). We built random forest models using the
+mikropml package (91) with OTUs from day 0 samples using mtry values of
 1 through 10, 15, 20, 25, 40, 50, 100. The split for training and
-testing varied by model to avoid overfitting the data. The toxin and
+testing varied by model to avoid overfitting the data. To determine the
+optimal split, we tested splits (50\%, 60\%, 70\%, 80\%, 90\% data used
+for training) to find the greatest portion of data that could be used to
+train the model while still maintaining the same performance for the
+training model as the model with the held-out test data. The toxin and
 moribundity models were trained with 60\% of the data. The
 histopathologic score model was trained with 80\% of the data. Lastly,
-we did not analyze murine communities to donor community or clinical
-data because germ-free mice colonized with human fecal communities
-become more like indigenous murine communities once stabilized.
-Furthermore, we are not trying to make any claims about human bacteria
-and their association with human CDI.
+we did not compare murine communities to donor community or clinical
+data because germ-free mice colonized with non-murine fecal communities
+have been shown to more closely resemble the recipient species than the
+donor species (92). Furthermore, it is not our intention to make any
+inferences regarding human associated bacteria and their relationship
+with human CDI outcome.
 
@@
 
+\leavevmode\hypertarget{ref-Perry2021}{}%
+3. \textbf{Perry DA}, \textbf{Shirley D}, \textbf{Micic D},
+\textbf{Patel CP}, \textbf{Putler R}, \textbf{Menon A}, \textbf{Young
+VB}, \textbf{Rao K}. 2021. External validation and comparison of
+\emph{Clostridioides difficile} severity scoring systems. Clinical
+Infectious Diseases.
+doi:\href{https://doi.org/10.1093/cid/ciab737}{10.1093/cid/ciab737}.
+
+\leavevmode\hypertarget{ref-Britton2014}{}%
+4. \textbf{Britton RA}, \textbf{Young VB}. 2014. Role of the intestinal
+microbiota in resistance to colonization by \emph{Clostridium
+difficile}. Gastroenterology \textbf{146}:1547--1553.
+doi:\href{https://doi.org/10.1053/j.gastro.2014.01.059}{10.1053/j.gastro.2014.01.059}.
+
+\leavevmode\hypertarget{ref-Hryckowian2018}{}%
+5. \textbf{Hryckowian AJ}, \textbf{Treuren WV}, \textbf{Smits SA},
+\textbf{Davis NM}, \textbf{Gardner JO}, \textbf{Bouley DM},
+\textbf{Sonnenburg JL}. 2018. Microbiota-accessible carbohydrates
+suppress \emph{Clostridium difficile} infection in a murine model.
+Nature Microbiology \textbf{3}:662--669.
+doi:\href{https://doi.org/10.1038/s41564-018-0150-6}{10.1038/s41564-018-0150-6}.
+
+\leavevmode\hypertarget{ref-VichVila2020}{}%
+6. \textbf{Vila AV}, \textbf{Collij V}, \textbf{Sanna S}, \textbf{Sinha
+T}, \textbf{Imhann F}, \textbf{Bourgonje AR}, \textbf{Mujagic Z},
+\textbf{Jonkers DMAE}, \textbf{Masclee AAM}, \textbf{Fu J},
+\textbf{Kurilshikov A}, \textbf{Wijmenga C}, \textbf{Zhernakova A},
+\textbf{Weersma RK}. 2020. Impact of commonly used drugs on the
+composition and metabolic function of the gut microbiota. Nature
+Communications \textbf{11}.
+doi:\href{https://doi.org/10.1038/s41467-019-14177-z}{10.1038/s41467-019-14177-z}.
+
+\leavevmode\hypertarget{ref-Abbas2020}{}%
+7. \textbf{Abbas A}, \textbf{Zackular JP}. 2020. Microbe-microbe
+interactions during \emph{Clostridioides difficile} infection. Current
+Opinion in Microbiology \textbf{53}:19--25.
+doi:\href{https://doi.org/10.1016/j.mib.2020.01.016}{10.1016/j.mib.2020.01.016}.
+
+\leavevmode\hypertarget{ref-Jenior2017}{}%
+8. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
+\textbf{Schloss PD}. 2017. \emph{Clostridium difficile} colonizes
+alternative nutrient niches during infection across distinct murine gut
+microbiomes. mSystems \textbf{2}.
+doi:\href{https://doi.org/10.1128/msystems.00063-17}{10.1128/msystems.00063-17}.
+
+\leavevmode\hypertarget{ref-Sorg2008}{}%
+9. \textbf{Sorg JA}, \textbf{Sonenshein AL}. 2008. Bile salts and
+glycine as cogerminants for \emph{Clostridium difficile} spores. Journal
+of Bacteriology \textbf{190}:2505--2512.
+doi:\href{https://doi.org/10.1128/jb.01765-07}{10.1128/jb.01765-07}.
+
+\leavevmode\hypertarget{ref-Thanissery2017}{}%
+10. \textbf{Thanissery R}, \textbf{Winston JA}, \textbf{Theriot CM}.
+2017. Inhibition of spore germination, growth, and toxin activity of
+clinically relevant \emph{C. difficile} strains by gut microbiota
+derived secondary bile acids. Anaerobe \textbf{45}:86--100.
+doi:\href{https://doi.org/10.1016/j.anaerobe.2017.03.004}{10.1016/j.anaerobe.2017.03.004}.
+
+\leavevmode\hypertarget{ref-Aguirre2021}{}%
+11. \textbf{Aguirre AM}, \textbf{Yalcinkaya N}, \textbf{Wu Q},
+\textbf{Swennes A}, \textbf{Tessier ME}, \textbf{Roberts P},
+\textbf{Miyajima F}, \textbf{Savidge T}, \textbf{Sorg JA}. 2021. Bile
+acid-independent protection against \emph{Clostridioides difficile}
+infection. PLOS Pathogens \textbf{17}:e1010015.
+doi:\href{https://doi.org/10.1371/journal.ppat.1010015}{10.1371/journal.ppat.1010015}.
+
+\leavevmode\hypertarget{ref-Kang2019}{}%
+12. \textbf{Kang JD}, \textbf{Myers CJ}, \textbf{Harris SC},
+\textbf{Kakiyama G}, \textbf{Lee I-K}, \textbf{Yun B-S},
+\textbf{Matsuzaki K}, \textbf{Furukawa M}, \textbf{Min H-K},
+\textbf{Bajaj JS}, \textbf{Zhou H}, \textbf{Hylemon PB}. 2019. Bile acid
+7\(\alpha\)-dehydroxylating gut bacteria secrete antibiotics that
+inhibit \emph{Clostridium difficile}: Role of secondary bile acids. Cell
+Chemical Biology \textbf{26}:27--34.e4.
+doi:\href{https://doi.org/10.1016/j.chembiol.2018.10.003}{10.1016/j.chembiol.2018.10.003}.
+
+\leavevmode\hypertarget{ref-Leslie2021}{}%
+13. \textbf{Leslie JL}, \textbf{Jenior ML}, \textbf{Vendrov KC},
+\textbf{Standke AK}, \textbf{Barron MR}, \textbf{O'Brien TJ},
+\textbf{Unverdorben L}, \textbf{Thaprawat P}, \textbf{Bergin IL},
+\textbf{Schloss PD}, \textbf{Young VB}. 2021. Protection from lethal
+clostridioides difficile infection via intraspecies competition for
+cogerminant. mBio \textbf{12}.
+doi:\href{https://doi.org/10.1128/mbio.00522-21}{10.1128/mbio.00522-21}.
+
+\leavevmode\hypertarget{ref-Nagpal2018}{}%
+14. \textbf{Nagpal R}, \textbf{Wang S}, \textbf{Woods LCS},
+\textbf{Seshie O}, \textbf{Chung ST}, \textbf{Shively CA},
+\textbf{Register TC}, \textbf{Craft S}, \textbf{McClain DA},
+\textbf{Yadav H}. 2018. Comparative microbiome signatures and
+short-chain fatty acids in mouse, rat, non-human primate, and human
+feces. Frontiers in Microbiology \textbf{9}.
+doi:\href{https://doi.org/10.3389/fmicb.2018.02897}{10.3389/fmicb.2018.02897}.
+
+\leavevmode\hypertarget{ref-Schubert2014}{}%
+15. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
+\textbf{Mogle J}, \textbf{Petrosino JP}, \textbf{Young VB},
+\textbf{Aronoff DM}, \textbf{Schloss PD}. 2014. Microbiome data
+distinguish patients with \emph{Clostridium difficile} infection and
+non-\emph{C. difficile}-associated diarrhea from healthy controls. mBio
+\textbf{5}.
+doi:\href{https://doi.org/10.1128/mbio.01021-14}{10.1128/mbio.01021-14}.
+
+\leavevmode\hypertarget{ref-Gillilland2012}{}%
+16. \textbf{Gillilland MG}, \textbf{Erb-Downward JR}, \textbf{Bassis
+CM}, \textbf{Shen MC}, \textbf{Toews GB}, \textbf{Young VB},
+\textbf{Huffnagle GB}. 2012. Ecological succession of bacterial
+communities during conventionalization of germ-free mice. Applied and
+Environmental Microbiology \textbf{78}:2359--2366.
+doi:\href{https://doi.org/10.1128/aem.05239-11}{10.1128/aem.05239-11}.
+
+\leavevmode\hypertarget{ref-Chen2008}{}%
+17. \textbf{Chen X}, \textbf{Katchar K}, \textbf{Goldsmith JD},
+\textbf{Nanthakumar N}, \textbf{Cheknis A}, \textbf{Gerding DN},
+\textbf{Kelly CP}. 2008. A mouse model of \emph{Clostridium
+difficile}-associated disease. Gastroenterology \textbf{135}:1984--1992.
+doi:\href{https://doi.org/10.1053/j.gastro.2008.09.002}{10.1053/j.gastro.2008.09.002}.
+
+\leavevmode\hypertarget{ref-Schubert2015}{}%
+18. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
+Antibiotic-induced alterations of the murine gut microbiota and
+subsequent effects on colonization resistance against \emph{Clostridium
+difficile}. mBio \textbf{6}.
+doi:\href{https://doi.org/10.1128/mbio.00974-15}{10.1128/mbio.00974-15}.
+
+\leavevmode\hypertarget{ref-Cowardin2016}{}%
+19. \textbf{Cowardin CA}, \textbf{Buonomo EL}, \textbf{Saleh MM},
+\textbf{Wilson MG}, \textbf{Burgess SL}, \textbf{Kuehne SA},
+\textbf{Schwan C}, \textbf{Eichhoff AM}, \textbf{Koch-Nolte F},
+\textbf{Lyras D}, \textbf{Aktories K}, \textbf{Minton NP}, \textbf{Petri
+WA}. 2016. The binary toxin CDT enhances \emph{Clostridium difficile}
+virulence by suppressing protective colonic eosinophilia. Nature
+Microbiology \textbf{1}.
+doi:\href{https://doi.org/10.1038/nmicrobiol.2016.108}{10.1038/nmicrobiol.2016.108}.
+
+\leavevmode\hypertarget{ref-Seekatz2016}{}%
+20. \textbf{Seekatz AM}, \textbf{Rao K}, \textbf{Santhosh K},
+\textbf{Young VB}. 2016. Dynamics of the fecal microbiome in patients
+with recurrent and nonrecurrent \emph{Clostridium difficile} infection.
+Genome Medicine \textbf{8}.
+doi:\href{https://doi.org/10.1186/s13073-016-0298-8}{10.1186/s13073-016-0298-8}.
+
 \leavevmode\hypertarget{ref-Dieterle2020}{}%
-3. \textbf{Dieterle MG}, \textbf{Putler R}, \textbf{Perry DA},
+21. \textbf{Dieterle MG}, \textbf{Putler R}, \textbf{Perry DA},
 \textbf{Menon A}, \textbf{Abernathy-Close L}, \textbf{Perlman NS},
@@
 
-\leavevmode\hypertarget{ref-Butt2013}{}%
-4. \textbf{Butt E}, \textbf{Foster JA}, \textbf{Keedwell E},
-\textbf{Bell JE}, \textbf{Titball RW}, \textbf{Bhangu A},
-\textbf{Michell SL}, \textbf{Sheridan R}. 2013. Derivation and
-validation of a simple, accurate and robust prediction rule for risk of
-mortality in patients with \emph{Clostridium difficile} infection. BMC
-Infectious Diseases \textbf{13}.
-doi:\href{https://doi.org/10.1186/1471-2334-13-316}{10.1186/1471-2334-13-316}.
-
-\leavevmode\hypertarget{ref-Perry2021}{}%
-5. \textbf{Perry DA}, \textbf{Shirley D}, \textbf{Micic D},
-\textbf{Patel CP}, \textbf{Putler R}, \textbf{Menon A}, \textbf{Young
-VB}, \textbf{Rao K}. 2021. External validation and comparison of
-\emph{Clostridioides difficile} severity scoring systems. Clinical
-Infectious Diseases.
-doi:\href{https://doi.org/10.1093/cid/ciab737}{10.1093/cid/ciab737}.
-
-\leavevmode\hypertarget{ref-vanBeurden2017}{}%
-6. \textbf{Beurden YH van}, \textbf{Hensgens MPM}, \textbf{Dekkers OM},
-\textbf{Cessie SL}, \textbf{Mulder CJJ}, \textbf{Vandenbroucke-Grauls
-CMJE}. 2017. External validation of three prediction tools for patients
-at risk of a complicated course of \emph{Clostridium difficile}
-infection: Disappointing in an outbreak setting. Infection Control \&
-Hospital Epidemiology \textbf{38}:897--905.
-doi:\href{https://doi.org/10.1017/ice.2017.89}{10.1017/ice.2017.89}.
-
-\leavevmode\hypertarget{ref-Britton2014}{}%
-7. \textbf{Britton RA}, \textbf{Young VB}. 2014. Role of the intestinal
-microbiota in resistance to colonization by \emph{Clostridium
-difficile}. Gastroenterology \textbf{146}:1547--1553.
-doi:\href{https://doi.org/10.1053/j.gastro.2014.01.059}{10.1053/j.gastro.2014.01.059}.
-
-\leavevmode\hypertarget{ref-Hryckowian2018}{}%
-8. \textbf{Hryckowian AJ}, \textbf{Treuren WV}, \textbf{Smits SA},
-\textbf{Davis NM}, \textbf{Gardner JO}, \textbf{Bouley DM},
-\textbf{Sonnenburg JL}. 2018. Microbiota-accessible carbohydrates
-suppress \emph{Clostridium difficile} infection in a murine model.
-Nature Microbiology \textbf{3}:662--669.
-doi:\href{https://doi.org/10.1038/s41564-018-0150-6}{10.1038/s41564-018-0150-6}.
-
-\leavevmode\hypertarget{ref-VichVila2020}{}%
-9. \textbf{Vila AV}, \textbf{Collij V}, \textbf{Sanna S}, \textbf{Sinha
-T}, \textbf{Imhann F}, \textbf{Bourgonje AR}, \textbf{Mujagic Z},
-\textbf{Jonkers DMAE}, \textbf{Masclee AAM}, \textbf{Fu J},
-\textbf{Kurilshikov A}, \textbf{Wijmenga C}, \textbf{Zhernakova A},
-\textbf{Weersma RK}. 2020. Impact of commonly used drugs on the
-composition and metabolic function of the gut microbiota. Nature
-Communications \textbf{11}.
-doi:\href{https://doi.org/10.1038/s41467-019-14177-z}{10.1038/s41467-019-14177-z}.
-
-\leavevmode\hypertarget{ref-Abbas2020}{}%
-10. \textbf{Abbas A}, \textbf{Zackular JP}. 2020. Microbe-microbe
-interactions during \emph{Clostridioides difficile} infection. Current
-Opinion in Microbiology \textbf{53}:19--25.
-doi:\href{https://doi.org/10.1016/j.mib.2020.01.016}{10.1016/j.mib.2020.01.016}.
-
-\leavevmode\hypertarget{ref-Jenior2017}{}%
-11. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
-\textbf{Schloss PD}. 2017. \emph{Clostridium difficile} colonizes
-alternative nutrient niches during infection across distinct murine gut
-microbiomes. mSystems \textbf{2}.
-doi:\href{https://doi.org/10.1128/msystems.00063-17}{10.1128/msystems.00063-17}.
-
-\leavevmode\hypertarget{ref-Sorg2008}{}%
-12. \textbf{Sorg JA}, \textbf{Sonenshein AL}. 2008. Bile salts and
-glycine as cogerminants for \emph{Clostridium difficile} spores. Journal
-of Bacteriology \textbf{190}:2505--2512.
-doi:\href{https://doi.org/10.1128/jb.01765-07}{10.1128/jb.01765-07}.
-
-\leavevmode\hypertarget{ref-Thanissery2017}{}%
-13. \textbf{Thanissery R}, \textbf{Winston JA}, \textbf{Theriot CM}.
-2017. Inhibition of spore germination, growth, and toxin activity of
-clinically relevant \emph{C. difficile} strains by gut microbiota
-derived secondary bile acids. Anaerobe \textbf{45}:86--100.
-doi:\href{https://doi.org/10.1016/j.anaerobe.2017.03.004}{10.1016/j.anaerobe.2017.03.004}.
-
-\leavevmode\hypertarget{ref-Aguirre2021}{}%
-14. \textbf{Aguirre AM}, \textbf{Yalcinkaya N}, \textbf{Wu Q},
-\textbf{Swennes A}, \textbf{Tessier ME}, \textbf{Roberts P},
-\textbf{Miyajima F}, \textbf{Savidge T}, \textbf{Sorg JA}. 2021. Bile
-acid-independent protection against \emph{Clostridioides difficile}
-infection. PLOS Pathogens \textbf{17}:e1010015.
-doi:\href{https://doi.org/10.1371/journal.ppat.1010015}{10.1371/journal.ppat.1010015}.
-
-\leavevmode\hypertarget{ref-Kang2019}{}%
-15. \textbf{Kang JD}, \textbf{Myers CJ}, \textbf{Harris SC},
-\textbf{Kakiyama G}, \textbf{Lee I-K}, \textbf{Yun B-S},
-\textbf{Matsuzaki K}, \textbf{Furukawa M}, \textbf{Min H-K},
-\textbf{Bajaj JS}, \textbf{Zhou H}, \textbf{Hylemon PB}. 2019. Bile acid
-7\(\alpha\)-dehydroxylating gut bacteria secrete antibiotics that
-inhibit \emph{Clostridium difficile}: Role of secondary bile acids. Cell
-Chemical Biology \textbf{26}:27--34.e4.
-doi:\href{https://doi.org/10.1016/j.chembiol.2018.10.003}{10.1016/j.chembiol.2018.10.003}.
-
-\leavevmode\hypertarget{ref-Seekatz2016}{}%
-16. \textbf{Seekatz AM}, \textbf{Rao K}, \textbf{Santhosh K},
-\textbf{Young VB}. 2016. Dynamics of the fecal microbiome in patients
-with recurrent and nonrecurrent \emph{Clostridium difficile} infection.
-Genome Medicine \textbf{8}.
-doi:\href{https://doi.org/10.1186/s13073-016-0298-8}{10.1186/s13073-016-0298-8}.
-
 \leavevmode\hypertarget{ref-Tomkovich2020}{}%
-17. \textbf{Tomkovich S}, \textbf{Stough JMA}, \textbf{Bishop L},
+22. \textbf{Tomkovich S}, \textbf{Stough JMA}, \textbf{Bishop L},
 \textbf{Schloss PD}. 2020. The initial gut microbiota and response to
@@
 \leavevmode\hypertarget{ref-Lesniak2021}{}%
-18. \textbf{Lesniak NA}, \textbf{Schubert AM}, \textbf{Sinani H},
+23. \textbf{Lesniak NA}, \textbf{Schubert AM}, \textbf{Sinani H},
 \textbf{Schloss PD}. 2021. Clearance of \emph{Clostridioides difficile}
@@
 
-\leavevmode\hypertarget{ref-Nagpal2018}{}%
-19. \textbf{Nagpal R}, \textbf{Wang S}, \textbf{Woods LCS},
-\textbf{Seshie O}, \textbf{Chung ST}, \textbf{Shively CA},
-\textbf{Register TC}, \textbf{Craft S}, \textbf{McClain DA},
-\textbf{Yadav H}. 2018. Comparative microbiome signatures and
-short-chain fatty acids in mouse, rat, non-human primate, and human
-feces. Frontiers in Microbiology \textbf{9}.
-doi:\href{https://doi.org/10.3389/fmicb.2018.02897}{10.3389/fmicb.2018.02897}.
-
-\leavevmode\hypertarget{ref-Schubert2014}{}%
-20. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
-\textbf{Mogle J}, \textbf{Petrosino JP}, \textbf{Young VB},
-\textbf{Aronoff DM}, \textbf{Schloss PD}. 2014. Microbiome data
-distinguish patients with \emph{Clostridium difficile} infection and
-non-\emph{C. difficile}-associated diarrhea from healthy controls. mBio
-\textbf{5}.
-doi:\href{https://doi.org/10.1128/mbio.01021-14}{10.1128/mbio.01021-14}.
-
-\leavevmode\hypertarget{ref-Gillilland2012}{}%
-21. \textbf{Gillilland MG}, \textbf{Erb-Downward JR}, \textbf{Bassis
-CM}, \textbf{Shen MC}, \textbf{Toews GB}, \textbf{Young VB},
-\textbf{Huffnagle GB}. 2012. Ecological succession of bacterial
-communities during conventionalization of germ-free mice. Applied and
-Environmental Microbiology \textbf{78}:2359--2366.
-doi:\href{https://doi.org/10.1128/aem.05239-11}{10.1128/aem.05239-11}.
-
-\leavevmode\hypertarget{ref-Chen2008}{}%
-22. \textbf{Chen X}, \textbf{Katchar K}, \textbf{Goldsmith JD},
-\textbf{Nanthakumar N}, \textbf{Cheknis A}, \textbf{Gerding DN},
-\textbf{Kelly CP}. 2008. A mouse model of \emph{Clostridium
-difficile}-associated disease. Gastroenterology \textbf{135}:1984--1992.
-doi:\href{https://doi.org/10.1053/j.gastro.2008.09.002}{10.1053/j.gastro.2008.09.002}.
-
-\leavevmode\hypertarget{ref-Schubert2015}{}%
-23. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
-Antibiotic-induced alterations of the murine gut microbiota and
-subsequent effects on colonization resistance against \emph{Clostridium
-difficile}. mBio \textbf{6}.
-doi:\href{https://doi.org/10.1128/mbio.00974-15}{10.1128/mbio.00974-15}.
-
-\leavevmode\hypertarget{ref-Akerlund2006}{}%
-24. \textbf{Åkerlund T}, \textbf{Svenungsson B}, \textbf{Lagergren},
-\textbf{Burman LG}. 2006. Correlation of disease severity with fecal
-toxin levels in patients with \emph{Clostridium difficile}-associated
-diarrhea and distribution of PCR ribotypes and toxin yields in vitro of
-corresponding isolates. Journal of Clinical Microbiology
-\textbf{44}:353--358.
-doi:\href{https://doi.org/10.1128/jcm.44.2.353-358.2006}{10.1128/jcm.44.2.353-358.2006}.
-
-\leavevmode\hypertarget{ref-Vitucci2020}{}%
-25. \textbf{Vitucci JC}, \textbf{Pulse M}, \textbf{Tabor-Simecka L},
-\textbf{Simecka J}. 2020. Epidemic ribotypes of \emph{Clostridium} (now
-\emph{Clostridioides}) \emph{difficile} are likely to be more virulent
-than non-epidemic ribotypes in animal models. BMC Microbiology
-\textbf{20}.
-doi:\href{https://doi.org/10.1186/s12866-020-1710-5}{10.1186/s12866-020-1710-5}.
-
-\leavevmode\hypertarget{ref-Cowardin2016}{}%
-26. \textbf{Cowardin CA}, \textbf{Buonomo EL}, \textbf{Saleh MM},
-\textbf{Wilson MG}, \textbf{Burgess SL}, \textbf{Kuehne SA},
-\textbf{Schwan C}, \textbf{Eichhoff AM}, \textbf{Koch-Nolte F},
-\textbf{Lyras D}, \textbf{Aktories K}, \textbf{Minton NP}, \textbf{Petri
-WA}. 2016. The binary toxin CDT enhances \emph{Clostridium difficile}
-virulence by suppressing protective colonic eosinophilia. Nature
-Microbiology \textbf{1}.
-doi:\href{https://doi.org/10.1038/nmicrobiol.2016.108}{10.1038/nmicrobiol.2016.108}.
-
-\leavevmode\hypertarget{ref-Carlson2015}{}%
-27. \textbf{Carlson PE}, \textbf{Kaiser AM}, \textbf{McColm SA},
-\textbf{Bauer JM}, \textbf{Young VB}, \textbf{Aronoff DM}, \textbf{Hanna
-PC}. 2015. Variation in germination of \emph{Clostridium difficile}
-clinical isolates correlates to disease severity. Anaerobe
-\textbf{33}:64--70.
-doi:\href{https://doi.org/10.1016/j.anaerobe.2015.02.003}{10.1016/j.anaerobe.2015.02.003}.
-
-\leavevmode\hypertarget{ref-Carlson2013}{}%
-28. \textbf{Carlson PE}, \textbf{Walk ST}, \textbf{Bourgis AET},
-\textbf{Liu MW}, \textbf{Kopliku F}, \textbf{Lo E}, \textbf{Young VB},
-\textbf{Aronoff DM}, \textbf{Hanna PC}. 2013. The relationship between
-phenotype, ribotype, and clinical disease in human \emph{Clostridium
-difficile} isolates. Anaerobe \textbf{24}:109--116.
-doi:\href{https://doi.org/10.1016/j.anaerobe.2013.04.003}{10.1016/j.anaerobe.2013.04.003}.
-
 \leavevmode\hypertarget{ref-Lungulescu2011}{}%
-29. \textbf{Lungulescu OA}, \textbf{Cao W}, \textbf{Gatskevich E},
+24. \textbf{Lungulescu OA}, \textbf{Cao W}, \textbf{Gatskevich E},
 \textbf{Tlhabano L}, \textbf{Stratidis JG}. 2011. CSI: A severity index
@@
 \leavevmode\hypertarget{ref-Zar2007}{}%
-30. \textbf{Zar FA}, \textbf{Bakkanagari SR}, \textbf{Moorthi KMLST},
+25. \textbf{Zar FA}, \textbf{Bakkanagari SR}, \textbf{Moorthi KMLST},
 \textbf{Davis MB}. 2007. A comparison of vancomycin and metronidazole
@@
 \leavevmode\hypertarget{ref-diMasi2018}{}%
-31. \textbf{Masi A di}, \textbf{Leboffe L}, \textbf{Polticelli F},
+26. \textbf{Masi A di}, \textbf{Leboffe L}, \textbf{Polticelli F},
 \textbf{Tonon F}, \textbf{Zennaro C}, \textbf{Caterino M}, \textbf{Stano
@@
 \leavevmode\hypertarget{ref-AbernathyClose2020}{}%
-32. \textbf{Abernathy-Close L}, \textbf{Dieterle MG}, \textbf{Vendrov
+27. \textbf{Abernathy-Close L}, \textbf{Dieterle MG}, \textbf{Vendrov
 KC}, \textbf{Bergin IL}, \textbf{Rao K}, \textbf{Young VB}. 2020. Aging
@@
 \leavevmode\hypertarget{ref-Theriot2011}{}%
-33. \textbf{Theriot CM}, \textbf{Koumpouras CC}, \textbf{Carlson PE},
+28. \textbf{Theriot CM}, \textbf{Koumpouras CC}, \textbf{Carlson PE},
 \textbf{Bergin II}, \textbf{Aronoff DM}, \textbf{Young VB}. 2011.
@@
 \leavevmode\hypertarget{ref-Goorhuis2008}{}%
-34. \textbf{Goorhuis A}, \textbf{Bakker D}, \textbf{Corver J},
+29. \textbf{Goorhuis A}, \textbf{Bakker D}, \textbf{Corver J},
 \textbf{Debast SB}, \textbf{Harmanus C}, \textbf{Notermans DW},
@@
 \leavevmode\hypertarget{ref-OConnor2009}{}%
-35. \textbf{O'Connor JR}, \textbf{Johnson S}, \textbf{Gerding DN}. 2009.
+30. \textbf{O'Connor JR}, \textbf{Johnson S}, \textbf{Gerding DN}. 2009.
 \emph{Clostridium difficile} infection caused by the epidemic
@@
 \leavevmode\hypertarget{ref-Rao2015}{}%
-36. \textbf{Rao K}, \textbf{Micic D}, \textbf{Natarajan M},
+31. \textbf{Rao K}, \textbf{Micic D}, \textbf{Natarajan M},
 \textbf{Winters S}, \textbf{Kiel MJ}, \textbf{Walk ST}, \textbf{Santhosh
@@
 \leavevmode\hypertarget{ref-Walk2012}{}%
-37. \textbf{Walk ST}, \textbf{Micic D}, \textbf{Jain R}, \textbf{Lo ES},
+32. \textbf{Walk ST}, \textbf{Micic D}, \textbf{Jain R}, \textbf{Lo ES},
 \textbf{Trivedi I}, \textbf{Liu EW}, \textbf{Almassalha LM},
@@
 
+\leavevmode\hypertarget{ref-Carlson2013}{}%
+33. \textbf{Carlson PE}, \textbf{Walk ST}, \textbf{Bourgis AET},
+\textbf{Liu MW}, \textbf{Kopliku F}, \textbf{Lo E}, \textbf{Young VB},
+\textbf{Aronoff DM}, \textbf{Hanna PC}. 2013. The relationship between
+phenotype, ribotype, and clinical disease in human \emph{Clostridium
+difficile} isolates. Anaerobe \textbf{24}:109--116.
+doi:\href{https://doi.org/10.1016/j.anaerobe.2013.04.003}{10.1016/j.anaerobe.2013.04.003}.
+
+\leavevmode\hypertarget{ref-Carlson2015}{}%
+34. \textbf{Carlson PE}, \textbf{Kaiser AM}, \textbf{McColm SA},
+\textbf{Bauer JM}, \textbf{Young VB}, \textbf{Aronoff DM}, \textbf{Hanna
+PC}. 2015. Variation in germination of \emph{Clostridium difficile}
+clinical isolates correlates to disease severity. Anaerobe
+\textbf{33}:64--70.
+doi:\href{https://doi.org/10.1016/j.anaerobe.2015.02.003}{10.1016/j.anaerobe.2015.02.003}.
+
 \leavevmode\hypertarget{ref-Saund2021}{}%
-38. \textbf{Saund K}, \textbf{Pirani A}, \textbf{Lacy B}, \textbf{Hanna
+35. \textbf{Saund K}, \textbf{Pirani A}, \textbf{Lacy B}, \textbf{Hanna
 PC}, \textbf{Snitkin ES}. 2021. Strain variation in \emph{Clostridioides
@@
 \leavevmode\hypertarget{ref-He2010}{}%
-39. \textbf{He M}, \textbf{Sebaihia M}, \textbf{Lawley TD},
+36. \textbf{He M}, \textbf{Sebaihia M}, \textbf{Lawley TD},
 \textbf{Stabler RA}, \textbf{Dawson LF}, \textbf{Martin MJ},
@@
 
+\leavevmode\hypertarget{ref-Butt2013}{}%
+37. \textbf{Butt E}, \textbf{Foster JA}, \textbf{Keedwell E},
+\textbf{Bell JE}, \textbf{Titball RW}, \textbf{Bhangu A},
+\textbf{Michell SL}, \textbf{Sheridan R}. 2013. Derivation and
+validation of a simple, accurate and robust prediction rule for risk of
+mortality in patients with \emph{Clostridium difficile} infection. BMC
+Infectious Diseases \textbf{13}.
+doi:\href{https://doi.org/10.1186/1471-2334-13-316}{10.1186/1471-2334-13-316}.
+
+\leavevmode\hypertarget{ref-vanBeurden2017}{}%
+38. \textbf{Beurden YH van}, \textbf{Hensgens MPM}, \textbf{Dekkers OM},
+\textbf{Cessie SL}, \textbf{Mulder CJJ}, \textbf{Vandenbroucke-Grauls
+CMJE}. 2017. External validation of three prediction tools for patients
+at risk of a complicated course of \emph{Clostridium difficile}
+infection: Disappointing in an outbreak setting. Infection Control \&
+Hospital Epidemiology \textbf{38}:897--905.
+doi:\href{https://doi.org/10.1017/ice.2017.89}{10.1017/ice.2017.89}.
+
 \leavevmode\hypertarget{ref-Jenior2018}{}%
-40. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
+39. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
 \textbf{Schloss PD}. 2018. \emph{Clostridium difficile} alters the
@@
 \leavevmode\hypertarget{ref-Staley2016}{}%
-41. \textbf{Staley C}, \textbf{Weingarden AR}, \textbf{Khoruts A},
+40. \textbf{Staley C}, \textbf{Weingarden AR}, \textbf{Khoruts A},
 \textbf{Sadowsky MJ}. 2016. Interaction of gut microbiota with bile acid
@@
 \leavevmode\hypertarget{ref-Long2017}{}%
-42. \textbf{Long SL}, \textbf{Gahan CGM}, \textbf{Joyce SA}. 2017.
+41. \textbf{Long SL}, \textbf{Gahan CGM}, \textbf{Joyce SA}. 2017.
 Interactions between gut bacteria and bile in health and disease.
@@
 \leavevmode\hypertarget{ref-Sorg2010}{}%
-43. \textbf{Sorg JA}, \textbf{Sonenshein AL}. 2010. Inhibiting the
+42. \textbf{Sorg JA}, \textbf{Sonenshein AL}. 2010. Inhibiting the
 initiation of \emph{Clostridium difficile} spore germination using
@@
 \leavevmode\hypertarget{ref-Dubois2019}{}%
-44. \textbf{Dubois T}, \textbf{Tremblay YDN}, \textbf{Hamiot A},
+43. \textbf{Dubois T}, \textbf{Tremblay YDN}, \textbf{Hamiot A},
 \textbf{Martin-Verstraete I}, \textbf{Deschamps J}, \textbf{Monot M},
@@
 \leavevmode\hypertarget{ref-Ng2013}{}%
-45. \textbf{Ng KM}, \textbf{Ferreyra JA}, \textbf{Higginbottom SK},
+44. \textbf{Ng KM}, \textbf{Ferreyra JA}, \textbf{Higginbottom SK},
 \textbf{Lynch JB}, \textbf{Kashyap PC}, \textbf{Gopinath S},
@@
 \leavevmode\hypertarget{ref-Ferreyra2014}{}%
-46. \textbf{Ferreyra JA}, \textbf{Wu KJ}, \textbf{Hryckowian AJ},
+45. \textbf{Ferreyra JA}, \textbf{Wu KJ}, \textbf{Hryckowian AJ},
 \textbf{Bouley DM}, \textbf{Weimer BC}, \textbf{Sonnenburg JL}. 2014.
@@
 \leavevmode\hypertarget{ref-MartinVerstraete2016}{}%
-47. \textbf{Martin-Verstraete I}, \textbf{Peltier J}, \textbf{Dupuy B}.
+46. \textbf{Martin-Verstraete I}, \textbf{Peltier J}, \textbf{Dupuy B}.
 2016. The regulatory networks that control \emph{Clostridium difficile}
@@
 \leavevmode\hypertarget{ref-Collins2015}{}%
-48. \textbf{Collins J}, \textbf{Auchtung JM}, \textbf{Schaefer L},
+47. \textbf{Collins J}, \textbf{Auchtung JM}, \textbf{Schaefer L},
 \textbf{Eaton KA}, \textbf{Britton RA}. 2015. Humanized microbiota mice
@@
 \leavevmode\hypertarget{ref-Lawley2012}{}%
-49. \textbf{Lawley TD}, \textbf{Clare S}, \textbf{Walker AW},
+48. \textbf{Lawley TD}, \textbf{Clare S}, \textbf{Walker AW},
 \textbf{Stares MD}, \textbf{Connor TR}, \textbf{Raisen C},
@@
 \leavevmode\hypertarget{ref-Reeves2011}{}%
-50. \textbf{Reeves AE}, \textbf{Theriot CM}, \textbf{Bergin IL},
+49. \textbf{Reeves AE}, \textbf{Theriot CM}, \textbf{Bergin IL},
 \textbf{Huffnagle GB}, \textbf{Schloss PD}, \textbf{Young VB}. 2011. The
@@
 \leavevmode\hypertarget{ref-Mabrok2011}{}%
-51. \textbf{Mabrok HB}, \textbf{Klopfleisch R}, \textbf{Ghanem KZ},
+50. \textbf{Mabrok HB}, \textbf{Klopfleisch R}, \textbf{Ghanem KZ},
 \textbf{Clavel T}, \textbf{Blaut M}, \textbf{Loh G}. 2011. Lignan
@@
 \leavevmode\hypertarget{ref-Kim2019}{}%
-52. \textbf{Kim CC}, \textbf{Healey GR}, \textbf{Kelly WJ},
+51. \textbf{Kim CC}, \textbf{Healey GR}, \textbf{Kelly WJ},
 \textbf{Patchett ML}, \textbf{Jordens Z}, \textbf{Tannock GW},
@@
 \leavevmode\hypertarget{ref-doPrado2021}{}%
-53. \textbf{Prado SBR do}, \textbf{Minguzzi BT}, \textbf{Hoffmann C},
+52. \textbf{Prado SBR do}, \textbf{Minguzzi BT}, \textbf{Hoffmann C},
 \textbf{Fabi JP}. 2021. Modulation of human gut microbiota by dietary
@@
 \leavevmode\hypertarget{ref-Muthuramalingam2019}{}%
-54. \textbf{Muthuramalingam K}, \textbf{Singh V}, \textbf{Choi C},
+53. \textbf{Muthuramalingam K}, \textbf{Singh V}, \textbf{Choi C},
 \textbf{Choi SI}, \textbf{Kim YM}, \textbf{Unno T}, \textbf{Cho M}.
@@
 \leavevmode\hypertarget{ref-Han2019}{}%
-55. \textbf{Han S-H}, \textbf{Yi J}, \textbf{Kim J-H}, \textbf{Lee S},
+54. \textbf{Han S-H}, \textbf{Yi J}, \textbf{Kim J-H}, \textbf{Lee S},
 \textbf{Moon H-W}. 2019. Composition of gut microbiota in patients with
@@
 \leavevmode\hypertarget{ref-Duncan2004}{}%
-56. \textbf{Duncan SH}, \textbf{Louis P}, \textbf{Flint HJ}. 2004.
+55. \textbf{Duncan SH}, \textbf{Louis P}, \textbf{Flint HJ}. 2004.
 Lactate-utilizing bacteria, isolated from human feces, that produce
@@
 \leavevmode\hypertarget{ref-Ye2018}{}%
-57. \textbf{Ye J}, \textbf{Lv L}, \textbf{Wu W}, \textbf{Li Y},
+56. \textbf{Ye J}, \textbf{Lv L}, \textbf{Wu W}, \textbf{Li Y},
 \textbf{Shi D}, \textbf{Fang D}, \textbf{Guo F}, \textbf{Jiang H},
@@
 \leavevmode\hypertarget{ref-Walsh2017}{}%
-58. \textbf{Walsh CJ}, \textbf{Guinane CM}, \textbf{O'Toole PW},
+57. \textbf{Walsh CJ}, \textbf{Guinane CM}, \textbf{O'Toole PW},
 \textbf{Cotter PD}. 2017. A profile hidden markov model to investigate
@@
 \leavevmode\hypertarget{ref-Sandiford2018}{}%
-59. \textbf{Sandiford SK}. 2018. Current developments in lantibiotic
+58. \textbf{Sandiford SK}. 2018. Current developments in lantibiotic
 discovery for treating \emph{Clostridium difficile} infection. Expert
@@
 \leavevmode\hypertarget{ref-Stein2013}{}%
-60. \textbf{Stein RR}, \textbf{Bucci V}, \textbf{Toussaint NC},
+59. \textbf{Stein RR}, \textbf{Bucci V}, \textbf{Toussaint NC},
 \textbf{Buffie CG}, \textbf{Rätsch G}, \textbf{Pamer EG}, \textbf{Sander
@@
 \leavevmode\hypertarget{ref-Nakashima2021}{}%
-61. \textbf{Nakashima T}, \textbf{Fujii K}, \textbf{Seki T},
+60. \textbf{Nakashima T}, \textbf{Fujii K}, \textbf{Seki T},
 \textbf{Aoyama M}, \textbf{Azuma A}, \textbf{Kawasome H}. 2021. Novel
@@
 \leavevmode\hypertarget{ref-Geerlings2018}{}%
-62. \textbf{Geerlings S}, \textbf{Kostopoulos I}, \textbf{Vos W de},
+61. \textbf{Geerlings S}, \textbf{Kostopoulos I}, \textbf{Vos W de},
 \textbf{Belzer C}. 2018. \emph{Akkermansia muciniphila} in the human
@@
 \leavevmode\hypertarget{ref-Deng2018}{}%
-63. \textbf{Deng H}, \textbf{Yang S}, \textbf{Zhang Y}, \textbf{Qian K},
+62. \textbf{Deng H}, \textbf{Yang S}, \textbf{Zhang Y}, \textbf{Qian K},
 \textbf{Zhang Z}, \textbf{Liu Y}, \textbf{Wang Y}, \textbf{Bai Y},
@@
 \leavevmode\hypertarget{ref-Engevik2020}{}%
-64. \textbf{Engevik MA}, \textbf{Engevik AC}, \textbf{Engevik KA},
+63. \textbf{Engevik MA}, \textbf{Engevik AC}, \textbf{Engevik KA},
 \textbf{Auchtung JM}, \textbf{Chang-Graham AL}, \textbf{Ruan W},
@@
 \leavevmode\hypertarget{ref-Reeves2012}{}%
-65. \textbf{Reeves AE}, \textbf{Koenigsknecht MJ}, \textbf{Bergin IL},
+64. \textbf{Reeves AE}, \textbf{Koenigsknecht MJ}, \textbf{Bergin IL},
 \textbf{Young VB}. 2012. Suppression of \emph{Clostridium difficile} in
@@
 \leavevmode\hypertarget{ref-Ma2021}{}%
-66. \textbf{Ma L}, \textbf{Keng J}, \textbf{Cheng M}, \textbf{Pan H},
+65. \textbf{Ma L}, \textbf{Keng J}, \textbf{Cheng M}, \textbf{Pan H},
 \textbf{Feng B}, \textbf{Hu Y}, \textbf{Feng T}, \textbf{Yang F}. 2021.
@@
 \leavevmode\hypertarget{ref-Haas2020}{}%
-67. \textbf{Haas KN}, \textbf{Blanchard JL}. 2020. Reclassification of
+66. \textbf{Haas KN}, \textbf{Blanchard JL}. 2020. Reclassification of
 the \emph{Clostridium clostridioforme} and \emph{Clostridium sphenoides}
@@
 \leavevmode\hypertarget{ref-Finegold2005}{}%
-68. \textbf{Finegold SM}, \textbf{Song Y}, \textbf{Liu C}, \textbf{Hecht
+67. \textbf{Finegold SM}, \textbf{Song Y}, \textbf{Liu C}, \textbf{Hecht
 DW}, \textbf{Summanen P}, \textbf{Könönen E}, \textbf{Allen SD}. 2005.
@@
 \leavevmode\hypertarget{ref-VanInsberghe2020}{}%
-69. \textbf{VanInsberghe D}, \textbf{Elsherbini JA}, \textbf{Varian B},
+68. \textbf{VanInsberghe D}, \textbf{Elsherbini JA}, \textbf{Varian B},
 \textbf{Poutahidis T}, \textbf{Erdman S}, \textbf{Polz MF}. 2020.
@@
 \leavevmode\hypertarget{ref-GarzaGonzalez2019}{}%
-70. \textbf{Garza-González E}, \textbf{Mendoza-Olazarán S},
+69. \textbf{Garza-González E}, \textbf{Mendoza-Olazarán S},
 \textbf{Morfin-Otero R}, \textbf{Ramírez-Fontes A},
@@
 \leavevmode\hypertarget{ref-Shafiq2020}{}%
-71. \textbf{Shafiq M}, \textbf{Alturkmani H}, \textbf{Zafar Y},
+70. \textbf{Shafiq M}, \textbf{Alturkmani H}, \textbf{Zafar Y},
 \textbf{Mittal V}, \textbf{Lodhi H}, \textbf{Ullah W}, \textbf{Brewer
@@
 \leavevmode\hypertarget{ref-Keith2020}{}%
-72. \textbf{Keith JW}, \textbf{Dong Q}, \textbf{Sorbara MT},
+71. \textbf{Keith JW}, \textbf{Dong Q}, \textbf{Sorbara MT},
 \textbf{Becattini S}, \textbf{Sia JK}, \textbf{Gjonbalaj M},
@@
 \leavevmode\hypertarget{ref-Zackular2016}{}%
-73. \textbf{Zackular JP}, \textbf{Moore JL}, \textbf{Jordan AT},
+72. \textbf{Zackular JP}, \textbf{Moore JL}, \textbf{Jordan AT},
 \textbf{Juttukonda LJ}, \textbf{Noto MJ}, \textbf{Nicholson MR},
@@
 \leavevmode\hypertarget{ref-Berkell2021}{}%
-74. \textbf{Berkell M}, \textbf{Mysara M}, \textbf{Xavier BB},
+73. \textbf{Berkell M}, \textbf{Mysara M}, \textbf{Xavier BB},
 \textbf{Werkhoven CH van}, \textbf{Monsieurs P}, \textbf{Lammens C},
@@
 \leavevmode\hypertarget{ref-Gardiner2014}{}%
-75. \textbf{Gardiner BJ}, \textbf{Tai AY}, \textbf{Kotsanas D},
+74. \textbf{Gardiner BJ}, \textbf{Tai AY}, \textbf{Kotsanas D},
 \textbf{Francis MJ}, \textbf{Roberts SA}, \textbf{Ballard SA},
@@
 \leavevmode\hypertarget{ref-Iljazovic2020}{}%
-76. \textbf{Iljazovic A}, \textbf{Roy U}, \textbf{Gálvez EJC},
+75. \textbf{Iljazovic A}, \textbf{Roy U}, \textbf{Gálvez EJC},
 \textbf{Lesker TR}, \textbf{Zhao B}, \textbf{Gronow A}, \textbf{Amend
@@
 \leavevmode\hypertarget{ref-Nagalingam2013}{}%
-77. \textbf{Nagalingam NA}, \textbf{Robinson CJ}, \textbf{Bergin IL},
+76. \textbf{Nagalingam NA}, \textbf{Robinson CJ}, \textbf{Bergin IL},
 \textbf{Eaton KA}, \textbf{Huffnagle GB}, \textbf{Young VB}. 2013. The
@@
 \leavevmode\hypertarget{ref-AbernathyClose2021}{}%
-78. \textbf{Abernathy-Close L}, \textbf{Barron MR}, \textbf{George JM},
+77. \textbf{Abernathy-Close L}, \textbf{Barron MR}, \textbf{George JM},
 \textbf{Dieterle MG}, \textbf{Vendrov KC}, \textbf{Bergin IL},
@@
 \leavevmode\hypertarget{ref-Pirofski2008}{}%
-79. \textbf{Pirofski L-a}, \textbf{Casadevall A}. 2008. The
+78. \textbf{Pirofski L-a}, \textbf{Casadevall A}. 2008. The
 damage-response framework of microbial pathogenesis and infectious
@@
 \leavevmode\hypertarget{ref-Casadevall2014}{}%
-80. \textbf{Casadevall A}, \textbf{Pirofski L-a}. 2014. What is a host?
+79. \textbf{Casadevall A}, \textbf{Pirofski L-a}. 2014. What is a host?
 Incorporating the microbiota into the damage-response framework.
@@
 
+\leavevmode\hypertarget{ref-Frisbee2019}{}%
+80. \textbf{Frisbee AL}, \textbf{Saleh MM}, \textbf{Young MK},
+\textbf{Leslie JL}, \textbf{Simpson ME}, \textbf{Abhyankar MM},
+\textbf{Cowardin CA}, \textbf{Ma JZ}, \textbf{Pramoonjago P},
+\textbf{Turner SD}, \textbf{Liou AP}, \textbf{Buonomo EL}, \textbf{Petri
+WA}. 2019. IL-33 drives group 2 innate lymphoid cell-mediated protection
+during \emph{Clostridium difficile} infection. Nature Communications
+\textbf{10}.
+doi:\href{https://doi.org/10.1038/s41467-019-10733-9}{10.1038/s41467-019-10733-9}.
+
+\leavevmode\hypertarget{ref-NagaoKitamoto2020}{}%
+81. \textbf{Nagao-Kitamoto H}, \textbf{Leslie JL}, \textbf{Kitamoto S},
+\textbf{Jin C}, \textbf{Thomsson KA}, \textbf{Gillilland MG},
+\textbf{Kuffa P}, \textbf{Goto Y}, \textbf{Jenq RR}, \textbf{Ishii C},
+\textbf{Hirayama A}, \textbf{Seekatz AM}, \textbf{Martens EC},
+\textbf{Eaton KA}, \textbf{Kao JY}, \textbf{Fukuda S}, \textbf{Higgins
+PDR}, \textbf{Karlsson NG}, \textbf{Young VB}, \textbf{Kamada N}. 2020.
+Interleukin-22-mediated host glycosylation prevents \emph{Clostridioides
+difficile} infection by modulating the metabolic activity of the gut
+microbiota. Nature Medicine \textbf{26}:608--617.
+doi:\href{https://doi.org/10.1038/s41591-020-0764-0}{10.1038/s41591-020-0764-0}.
+
 \leavevmode\hypertarget{ref-Tailford2015}{}%
-81. \textbf{Tailford LE}, \textbf{Crost EH}, \textbf{Kavanaugh D},
+82. \textbf{Tailford LE}, \textbf{Crost EH}, \textbf{Kavanaugh D},
 \textbf{Juge N}. 2015. Mucin glycan foraging in the human gut
@@
 \leavevmode\hypertarget{ref-Sorg2009}{}%
-82. \textbf{Sorg JA}, \textbf{Dineen SS}. 2009. Laboratory maintenance
+83. \textbf{Sorg JA}, \textbf{Dineen SS}. 2009. Laboratory maintenance
 of \emph{Clostridium difficile}. Current Protocols in Microbiology
@@
 \leavevmode\hypertarget{ref-Winston2016}{}%
-83. \textbf{Winston JA}, \textbf{Thanissery R}, \textbf{Montgomery SA},
+84. \textbf{Winston JA}, \textbf{Thanissery R}, \textbf{Montgomery SA},
 \textbf{Theriot CM}. 2016. Cefoperazone-treated mouse model of
@@
 \leavevmode\hypertarget{ref-Kozich2013}{}%
-84. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+85. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 \leavevmode\hypertarget{ref-Schloss2009}{}%
-85. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+86. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \leavevmode\hypertarget{ref-Wang2007}{}%
-86. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
+87. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
 \textbf{Cole JR}. 2007. Naïve bayesian classifier for rapid assignment
@@
 \leavevmode\hypertarget{ref-Yue2005}{}%
-87. \textbf{Yue JC}, \textbf{Clayton MK}. 2005. A similarity measure
+88. \textbf{Yue JC}, \textbf{Clayton MK}. 2005. A similarity measure
 based on species proportions. Communications in Statistics - Theory and
@@
 \leavevmode\hypertarget{ref-Segata2011}{}%
-88. \textbf{Segata N}, \textbf{Izard J}, \textbf{Waldron L},
+89. \textbf{Segata N}, \textbf{Izard J}, \textbf{Waldron L},
 \textbf{Gevers D}, \textbf{Miropolsky L}, \textbf{Garrett WS},
@@
 \leavevmode\hypertarget{ref-Benjamini1995}{}%
-89. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
+90. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
 false discovery rate: A practical and powerful approach to multiple
@@
 \leavevmode\hypertarget{ref-Topcuoglu2021}{}%
-90. \textbf{Topçuoğlu B}, \textbf{Lapp Z}, \textbf{Sovacool K},
+91. \textbf{Topçuoğlu B}, \textbf{Lapp Z}, \textbf{Sovacool K},
 \textbf{Snitkin E}, \textbf{Wiens J}, \textbf{Schloss P}. 2021.
@@
 doi:\href{https://doi.org/10.21105/joss.03073}{10.21105/joss.03073}.
+
+\leavevmode\hypertarget{ref-Rawls2006}{}%
+92. \textbf{Rawls JF}, \textbf{Mahowald MA}, \textbf{Ley RE},
+\textbf{Gordon JI}. 2006. Reciprocal gut microbiota transplants from
+zebrafish and mice to germ-free recipients reveal host habitat
+selection. Cell \textbf{127}:423--433.
+doi:\href{https://doi.org/10.1016/j.cell.2006.08.043}{10.1016/j.cell.2006.08.043}.
 \end{cslreferences}
@@
 \textbf{Figure 1. Human fecal microbial communities established diverse
-gut bacterial communities in germ-free mice.} (A) Relative abundance of
-10 most abundant bacterial classes of the gut bacterial communities on
-timepoint day 0, 14 days post-colonization of germ-free C57Bl/6 mice
-with human fecal samples. Each column of abundances represents an
-individual mouse. Each group of mice that received the same donor are
-grouped together and labeled with a letter (N for non-moribund mice and
-M for moribund mice) and number. (B) Median (points) and interquartile
-range (lines) of \(\beta\)-diversity (\(\theta\)\textsubscript{YC})
-between an individual mouse and either all others which were inoculated
-with feces from the same donor or from a different donor. The
-\(\beta\)-diversity among the same donor comparison group was
-significantly less than the \(\beta\)-diversity of the different donor
-group (\emph{P} \textless{} 0.05, calculated by Wilcoxon rank sum test).
+gut bacterial communities in germ-free mice.} (A) Relative abundances of
+the 10 most abundant bacterial classes observed in the feces of
+previously germ-free C57Bl/6 mice 14 days post-colonization with human
+fecal samples (i.e., day 0 relative to \emph{C. difficile} challenge).
+Each column of abundances represents an individual mouse. Mice that
+received the same donor feces are grouped together and labeled above
+with a letter (N for non-moribund mice and M for moribund mice) and
+number (ordered by mean histopathologic score). + indicates the mice
+which did not have detectable \emph{C. difficile} CFU in Figure 2. (B)
+Median (points) and interquartile range (lines) of \(\beta\)-diversity
+(\(\theta\)\textsubscript{YC}) between an individual mouse and either
+all others which were inoculated with feces from the same donor or from
+a different donor. The \(\beta\)-diversity among the same donor
+comparison group was significantly less than the \(\beta\)-diversity of
+the different donor group (\emph{P} \textless{} 0.05, calculated by
+Wilcoxon rank sum test).
 
@@
 of stool was measured the day after challenge with 10\(^{3}\) \emph{C.
-difficile} RT027 human isolate 431 spores and at the end of the
+difficile} RT027 clinical isolate 431 spores and at the end of the
 experiment, 10 days post-challenge. Each point represents an individual
 mouse. Mice are grouped by donor and labeled by the donor letter (N for
-non-moribund mice and M for moribund mice) and number. Points are
-colored by donor group. Mice from donor groups N1 through N6 succumbed
-to the infection prior to day 10 and were not plated on day 10
-post-challenge. LOD = Limit of detection.
+non-moribund mice and M for moribund mice) and number (ordered by mean
+histopathologic score). Points are colored by donor group. Mice from
+donor groups N1 through N6 succumbed to the infection prior to day 10
+and were not plated on day 10 post-challenge. LOD = Limit of detection.
 
@@
 post-challenge for non-moribund mice or the day the mouse succumbed to
-the infection for the moribund group (day 2-3 post-challenge). Each
+the infection for the moribund group (day 2 or 3 post-challenge). Each
 point represents an individual mouse. Mice are grouped by donor and
 labeled by the donor letter (N for non-moribund mice and M for moribund
-mice) and number. Points are colored by donor group. Missing points are
-from mice that had insufficient sample collected for assaying fecal
-toxin.
+mice) and number (ordered by mean histopathologic score). Points are
+colored by donor group. Mice in group N1 that have a summary score of 0
+are the mice which did not have detectable \emph{C. difficile} CFU in
+Figure 2. Missing points are from mice that had insufficient fecal
+sample collected for assaying toxin or cecum for histopathologic
+scoring.
 
@@
 colored light green). (B) Relative abundances were compared between mice
-which toxin activity was detected (Toxigenic, colored dark purple) and
-which no toxin activity was detected (Non-toxigenic, colored light
-purple). (C) Endpoint bacterial OTUs correlated with histopathologic
-summary score. Each individual mouse is plotted (transparent gray
-point). Spearman's correlations were statistically significant after
-Benjamini-Hochberg correction for multiple comparisons. * next to a name
-indicates that the bacterial group was unclassified at lower
-classification ranks.
+which toxin activity was detected (Toxin +, colored dark purple) and
+which no toxin activity was detected (Toxin -, colored light purple).
+(C) Endpoint bacterial OTUs correlated with histopathologic summary
+score. Each individual mouse is plotted (transparent gray point).
+Spearman's correlations were statistically significant after
+Benjamini-Hochberg correction for multiple comparisons. All bacterial
+groups are ordered alphabetically. * indicates that the bacterial group
+was unclassified at lower taxonomic classification ranks.
 
@@
 taxonomic group at the top of the plot had the greatest decrease in
-performance when its relative abundances were permuted. * next to the
-name indicates that the bacterial group was unclassified at lower
-classification ranks. (A) Bacterial members grouped by phyla predicted
-which mice would have toxin activity detected at any point throughout
-the infection (Toxigenic, dark purple). (B) Bacterial members grouped by
-class predicted which mice would become moribund (dark blue). (C)
-Bacterial members grouped by genera predicted if the mice would have a
-high (score greater than the median score of 5, colored dark green) or
-low (score less than the median score of 5, colored light green)
-histopathologic summary score.
-
-\hfill\break
-
-\includegraphics{../results/figures/figure_6.jpg}
-
-\textbf{Figure 6. \emph{C. difficile} infection outcome were similar
-within human fecal donor groups across different \emph{C. difficile}
-isolates.} Mice colonized with human fecal donor communities N7, N6, or
-M4 were challenged with the isolate from the previous experiments in
-this study, \emph{C. difficile} RT027 human isolate 431 (blue points),
-or a different \emph{C. difficile} isolate (purple points). \emph{C.
-difficile} CFUs were enumerated at the end of the experiment (day 10
-post-challenge) and are plotted grouped by donor source (top plot).
-Histopathologic scores from the tissue collected at the endpoint (either
-day 10 post-challenge or day mice succumbed to infection) are plotted
-grouped by donor source (bottom plot). Each point represents an
-individual mouse. Mice that had received fecal communities from human
-donor M4 succumbed to the infection prior to day 10 post-challenge so
-were not plated for CFU on day 10 post-challenge (DECD = deceased). RT =
-ribotype, Isolate = \emph{C. difficile} human isolate id, LOD = limit of
-detection.
+performance when its relative abundances were permuted. * indicates that
+the bacterial group was unclassified at lower taxonomic classification
+ranks. (A) Bacterial members grouped by phyla predicted which mice would
+have toxin activity detected at any point throughout the infection
+(Toxin +, dark purple). (B) Bacterial members grouped by class predicted
+which mice would become moribund (dark blue). (C) Bacterial members
+grouped by genera predicted if the mice would have a high (score greater
+than the median score of 5, colored dark green) or low (score less than
+the median score of 5, colored light green) histopathologic summary
+score.
 
@@
 the infection outcome. The models used the highest taxonomic
-classification rank that performed as well as the lower ranks. Black
+classification rank performed as well as the lower ranks. Black
 rectangle highlights classification rank used to model each outcome.
@@
 since differences in AUC were outside the interquartile range. *
-indicates bacterial group was unclassified at lower classification
-ranks. For all plots, median (solid points) and interquartile range
-(lines) are plotted. (A) Toxin production modeled which mice would have
-toxin detected during the experiment. (B) Moribundity modeled which mice
-would succumb to the infection prior to day 10 post-challenge. (C)
-Histopathologic score modeled which mice would have a high (score
-greater than the median score of 5) or low (score less than the median
-score of 5) histopathologic summary score. (D) Bacterial phyla which
-affected the performance of predicting toxin activity when permuted. (E)
-Bacterial classes which affected the performance of predicting
-moribundity when permuted. (D) Bacterial genera which affected the
-performance of predicting histopathologic score when permuted.
-
-\hfill\break
-
-\includegraphics{../results/figures/figure_S3.jpg}
-
-\textbf{Figure S3. Subset of donor communities were selected to test for
-conserved \emph{C. difficile} infection outcome with a different
-\emph{C. difficile} isolate.} (A) Relative abundance of genera in mice
-from previous experiments which received human fecal donor communities
-N7, N6, or M4. Each column represents an individual mouse. (B) Features
-from the random forest model that predicted moribundity using relative
-abundance of bacteria were grouped by their genus classification rank
-and are ordered by their decrease in AUC when each feature was permuted.
-Median (solid points) and interquartile range (lines) change in AUC are
-plotted. (C) Relative abundance of the most important genera used by the
-model to predict moribundity. Median (solid points) and interquartile
-range (lines) of the group relative abundances are plotted. * next to a
-name indicates that the bacterial group was unclassified at lower
-classification ranks.
+indicates bacterial group was unclassified at lower taxonomic
+classification ranks. For all plots, median (solid points) and
+interquartile range (lines) are plotted. (A) Toxin production modeled
+which mice would have toxin detected during the experiment. (B)
+Moribundity modeled which mice would succumb to the infection prior to
+day 10 post-challenge. (C) Histopathologic score modeled which mice
+would have a high (score greater than the median score of 5) or low
+(score less than the median score of 5) histopathologic summary score.
+(D) Bacterial phyla which affected the performance of predicting toxin
+activity when permuted. (E) Bacterial classes which affected the
+performance of predicting moribundity when permuted. (D) Bacterial
+genera which affected the performance of predicting histopathologic
+score when permuted.
 

### 2022-01-22 code/plot_figure_1.R

@@ -57,7 +57,7 @@ relative_abundance_data <- metadata %>%
   left_join(filter(shared, counts > 0), by = c('group' = 'Group')) %>% 
   left_join(taxonomy, by = c('otu' = 'OTU')) %>% 
   rename(taxa_level = Class) %>% 
-  group_by(group, taxa_level, human_source) %>% 
+  group_by(group, taxa_level, human_source, mouse_id) %>% 
   summarise(relative_abundance = log10(sum(relative_abundance) + 0.0001)) %>% 
   group_by(taxa_level) %>% 
   filter(!is.na(taxa_level)) %>% 
@@ -67,7 +67,8 @@ relative_abundance_data <- metadata %>%
          taxa_level = paste0('*', taxa_level, '*'),
          taxa_level = ifelse(grepl('unclassified', taxa_level), 
          		paste('Unclassified', gsub(' unclassified\\*', '*', taxa_level)),
-         		taxa_level)) %>% 
+         		taxa_level),
+         mouse_id = ifelse(mouse_id %in% c('NP1_2597', 'NP1_2725'), '+', '')) %>% 
   filter(median_ra > -0.677)
 
 
@@ -87,12 +88,13 @@ day_0_plot <- relative_abundance_data %>% # only plot top 10
     scale_fill_gradient2(low="white", mid='#0B775E', high = 'black',
 		  	limits = c(-2,2), na.value = NA, midpoint = .3,
 			  breaks = c(-2.5, -1, 0, 1, 2), labels = c('', '0.1', '1', '10', '100')) + 
+    scale_x_discrete(labels = c(rep('', 8), rep('+', 2), rep('', 41))) + 
 	  theme_bw() + 
     labs(x = NULL, y = NULL, fill = NULL) + 
     facet_grid(.~donor_labels, scales = 'free_x', space = 'free_x') + 
     theme(axis.ticks.x = element_blank(),
-          axis.text.x = element_blank(),
           legend.position = 'bottom',
+          axis.text.x = element_text(margin = margin(t = -3)),
           axis.text.y = ggtext::element_markdown(),
           strip.background = element_blank(),
           legend.margin=margin(t=-0.1, r=0, b=-0.1, l=0, unit="cm"),

### 2022-01-22 code/plot_figure_2.R

@@ -46,7 +46,7 @@ cfu_plot <- cfu_data %>%
 			sec.axis = dup_axis(labels = NULL, breaks = c(2,4,6,8),
                                            name = "Days post-challenge")) + 
 		theme_bw() + 
-		labs(x = 'Donor', y = '*C. difficile* CFU') + 
+		labs(x = 'Donor', y = '*C. difficile* CFU/g stool') + 
 		guides(color = 'none') + 
 		theme(axis.text.y = ggtext::element_markdown(),
 			axis.title.y = ggtext::element_markdown(),

### 2022-01-22 code/plot_figure_3.R

@@ -22,6 +22,10 @@ histology <- read_histology()
 
 donor_aes <- donor_df
 
+LOD_df <- data.frame(x = 1.15, y = 2)
+DECD_df <- data.frame(x = 2, y = 2.5, day = 10, 
+  outcome = factor('Moribund', levels = c('Non-moribund', 'Moribund')))
+
 ###############################################################################
 #  setup data
 ###############################################################################
@@ -50,6 +54,8 @@ histology <- histology %>%
 toxin_plot <- toxin %>% 
   ggplot(aes(x = donor_labels, y = Log_repiricoal_dilution, 
              fill = donor_colors, color = donor_colors)) + 
+    geom_hline(yintercept = 2, linetype = 'dashed', 
+      size = 1, color = 'white') + 
     geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 1.5) + 
     scale_fill_identity() + 
     scale_color_identity() + 
@@ -64,8 +70,17 @@ toxin_plot <- toxin %>%
           strip.background = element_blank(),
           axis.text.x = element_blank(),
           axis.ticks = element_blank(),
-          panel.grid.minor.y = element_blank())
-
+          panel.grid.minor.y = element_blank()) + 
+    geom_label(data = LOD_df, aes(x = x-.25, y = y), label = "LOD", 
+      fill = 'white', size = 10/.pt, color = 'white', inherit.aes = F) + 
+    geom_text(data = LOD_df, aes(x = x, y = y), label = "LOD", 
+      color = 'lightgray', inherit.aes = F) + 
+    geom_segment(data = DECD_df, aes(x = 1, xend = 6, y = 2.5, yend = 2.5), 
+      size = .25, color = 'black', inherit.aes = F) + 
+    geom_label(data = DECD_df, aes(x = 3.5, y = 2.5), label = "Deceased", 
+      fill = 'white', color = 'white', inherit.aes = F) + 
+    geom_text(data = DECD_df, aes(x = 3.5, y = 2.5), label = "Deceased", 
+      color = 'black', inherit.aes = F)
 
 histology_plot <- histology %>%  
   ggplot(aes(x = donor_labels, y = score, 

### 2022-01-22 code/plot_figure_4.R

@@ -58,18 +58,21 @@ plot_lefse <- function(file_name){
 			class = case_when(class == 'severe' ~ 'Moribund',
 				class == 'moderate' ~ 'Non-moribund, High histopathologic score',
 				class == 'mild' ~ 'Non-moribund, Low histopathologic score',
-				class == FALSE ~ 'Non-toxigenic',
-				class == TRUE ~ 'Toxigenic', 
+				class == FALSE ~ 'Toxin -',
+				class == TRUE ~ 'Toxin +', 
 				T ~ 'NA'),
-			class = factor(class, levels = c('Non-moribund, Low histopathologic score',
-				'Non-moribund, High histopathologic score', 'Moribund',
-				'Non-toxigenic', 'Toxigenic')),
+			class = factor(class, levels = c('Moribund',
+				'Non-moribund, High histopathologic score', 
+				'Non-moribund, Low histopathologic score',
+				'Toxin +', 'Toxin -')),
 			taxa_label = gsub('_', ' ', taxa_label),
+			taxa_label = gsub('2', '', taxa_label),
 			taxa_label = paste0('*', taxa_label, '*'),
 			taxa_label = ifelse(grepl('unclassified', taxa_label), 
 				gsub(' unclassified\\*', '* *', taxa_label),
 				taxa_label)) %>% 
-		ggplot(aes(x = taxa_label, y = relative_abundance, color = class)) +
+		ggplot(aes(x = reorder(taxa_label, desc(taxa_label)), y = relative_abundance, 
+				color = class)) +
 			stat_summary(fun.data = 'median_hilow', aes(group = class),
 				fun.args = (conf.int=0.5), position = position_dodge(width = .5)) +
 			geom_hline(yintercept = 0.047, linetype = 'dashed', lwd = 0.3) + 
@@ -85,6 +88,7 @@ plot_lefse <- function(file_name){
 				legend.position = 'top',
 				legend.margin=margin(2,0,0,0),
     	    	legend.box.margin=margin(2,-10,-10,-10)) + 
+			guides(color = guide_legend(reverse = TRUE)) + 
 			coord_flip()
 }
 
@@ -141,10 +145,10 @@ day_10_hist_genus_corr_plot <- day_10_hist_genus_corr %>%
 			strip.text = ggtext::element_markdown())
 
 day_0_toxin_production_plot <- plot_lefse('day_0_toxin_production_Genus') +
-	scale_color_manual(values = c("#e4b5ff", "#6a45c2"))
[…]

### 2022-01-22 code/plot_figure_5.R

@@ -50,7 +50,8 @@ day_0_toxin_features <- features_df %>%
 day_0_predict_future_toxin_df <- toxin %>% 
 	group_by(mouse_id) %>% 
 	summarise(toxin = ifelse(max(Log_repiricoal_dilution) > 1, 
-		'Toxigenic', 'Non-toxigenic')) %>% 
+		'Toxin +', 'Toxin -'),
+		toxin = factor(toxin, levels = c('Toxin +', 'Toxin -'))) %>% 
 	left_join(select(metadata, mouse_id, day, group), 
 		by = c('mouse_id')) %>% 
 	filter(day == 0) %>% 
@@ -132,12 +133,14 @@ day_0_toxin_feature_abundance_plot <- day_0_predict_future_toxin_df %>%
 		scale_y_log10() + 
 		theme_bw() + 
 		coord_flip() + 
-		scale_color_manual(values = c("#e4b5ff", "#6a45c2")) + 
+		scale_color_manual(breaks = c('Toxin -', 'Toxin +'),
+			values = c("#e4b5ff", "#6a45c2")) + 
 		labs(x = NULL, y = 'Relative Abundance (%)', color	= NULL) + 
 		theme(legend.position = 'top',
 			axis.text.y = ggtext::element_markdown(),
 			legend.margin=margin(2,0,0,0),
         	legend.box.margin=margin(2,-10,-10,-10))
+
 
 day_0_moribund_feature_abundance_plot <- day_0_moribund %>% 
 	ggplot(aes(x = taxon, y = counts, color = early_euth)) + 
@@ -148,7 +151,8 @@ day_0_moribund_feature_abundance_plot <- day_0_moribund %>%
 		scale_y_log10() + 
 		theme_bw() + 
 		coord_flip() + 
-		scale_color_manual(values = c("4EBFA6", "#0c1b37")) + 
+		scale_color_manual(breaks = c('Non-moribund', 'Moribund'),
+			values = c("4EBFA6", "#0c1b37")) + 
 		labs(x = NULL, y = 'Relative Abundance (%)', color	= NULL) + 
 		theme(legend.position = 'top',
 			axis.text.y = ggtext::element_markdown(),

### 2022-01-22 code/utilities.R

@@ -66,8 +66,8 @@ donor_df <- tibble(
 	donor_colors = c("#41bbc5", "#7e2640", "#9ad859", "#682dbd", "#728e24",
 		"#e26df8", "#23980d", "#fd048f", "#46e33e", "#fa1bfc",
 		"#155126", "#f9b2d1", "#154975", "#e2c627", "#6294ce"),
-	human_source = c('DA01324', 'DA00953', 'DA00581', 'DA01134', 'DA10148',
-		'DA00369', 'DA00430', 'DA10093', 'DA10027', 'DA10034', 
-		'DA10082', 'DA00884', 'DA00578', 'DA00431', 'DA01245'),
+	human_source = c('DA00581', 'DA00430', 'DA01324', 'DA00953', 'DA10148',
+		'DA01134', 'DA00369', 'DA10027', 'DA10093', 
+		'DA10082', 'DA00884', 'DA10034', 'DA01245', 'DA00578', 'DA00431'),
 	donor_labels = factor(c(paste0('N', 1:9), paste0('M', 1:6)), 
 			levels = c(paste0('N', 1:9), paste0('M', 1:6))))

### 2022-01-23 submission/manuscript.Rmd
proofread coathur edits

@@
 difficile* ribotype and toxin genes have been associated with disease severity.
-However, it is unclear whether there is an association between the gut microbiota and disease severity.
+However, it is unclear whether there is an association between members of the gut microbiota and disease severity.
 The gut microbiota is known to interact with *C.
@@
 difficile* to colonize the gut.
-The gut microbiota is thought to impair *C.
+The gut microbiota can inhibit *C.
 difficile* colonization through bile acid metabolism, nutrient consumption and bacteriocin production.
-Here we sought to demonstrate that members of the gut bacterial communities can contribute to disease severity.
+Here we sought to demonstrate that members of the gut bacterial communities can also contribute to disease severity.
 We derived diverse gut communities by colonizing germ-free mice with different human fecal communities.
@@
 Bacterial groups associated with fiber degradation, bile acid metabolism and lantibiotic production, such as *Anaerostipes* and *Coprobacillus*, were associated with less severe outcomes.
-These data indicate that gut bacteria can influence CDI disease severity.
+These data indicate that, in addition to the host and *C.
+difficile*, populations of gut bacteria can influence CDI disease severity.
 
@@
 
-*Clostridioides difficile* colonization can result in no symptoms or develop into an infection, ranging in severity from mild diarrhea to toxic megacolon, sepsis and death.
+*Clostridioides difficile* colonization can result in no symptoms or develop into an infection, ranging in severity from mild diarrhea to toxic megacolon, sepsis, and death.
 Models that predict severity and guide treatment decisions are based on clinical factors and *C.
 difficile* characteristics.
-Although the gut microbiome plays a role in protecting against CDI, it has not been investigated for its effect on CDI disease severity or incorporated into attempts to predict disease severity.
+Although the gut microbiome plays a role in protecting against CDI, it has its effect on CDI disease severity is unclear and has not been incorporated into disease severity models.
 We demonstrated that variation in the microbiome of mice colonized with human feces yielded a range of disease outcomes.
@@
 difficile* was first identified as the cause of antibiotic-associated pseudomembranous colitis [@Kelly2008].
-CDI disease severity can range from asymptomatic to mild diarrhea to toxic megacolon and death.
-The Infectious Diseases Society of America (IDSA) and Society for Healthcare Epidemiology of America (SHEA) guidelines defines severe CDI in terms of a white blood cell count greater than 15,000 cells/mm$^3$ and/or a serum creatinine greater than 1.5 mg/dL.
+CDI disease severity can range from mild diarrhea to toxic megacolon and death.
+The Infectious Diseases Society of America (IDSA) and Society for Healthcare Epidemiology of America (SHEA) guidelines define severe CDI in terms of a white blood cell count greater than 15,000 cells/mm$^3$ and/or a serum creatinine greater than 1.5 mg/dL.
 Patients who develop shock or hypotension, ileus, or toxic megacolon are considered to have fulminant CDI [@McDonald2018].
-Since these measures are outcomes of CDI, these measures have limited ablity to predict risk of severe CDI when CDI is fiest detected.
+Since these measures are CDI outcomes, they have limited ability to predict risk of severe CDI when the infection is first detected.
 Schemes have been developed to score a patient's risk for severe CDI outcomes based on clinical factors but have not been robust for broad application [@Perry2021].
 Thus, we have limited ability to prevent patients from developing severe CDI.
-Ideally, we could identify patients that are high-risk of severe disease so their treatments could be adjusted to alter the clinical trajectory.
 
@@
 
-Here, we sought to determine whether the gut bacteria contribute to disease severity.
-We hypothesized that since the gut bacteria affect the metabolism of *C.
-difficile* and its infection dynamics, it can also affect the disease severity of the infection.
+Recently, Nagao-Kitamoto *et al.* showed germ-free mice which received a microbial community from different human donors resulted in variation in a subsequent *C.
+difficile* infection [@NagaoKitamoto2020].
+Here, we sought to further elucidate the relationship between members of the gut bacterial community and CDI disease severity.
+We hypothesized that since specific groups of gut bacteria affect the metabolism of *C.
+difficile* and its infection dynamics, we can also identify groups of bacteria that affect the disease severity of the infection.
 To test this hypothesis, we colonized germ-free C57BL/6 mice with human fecal samples to create varied gut communities.
-We then challenged the mice with a single *C.
-difficile* isolate, ribotype 027 (RT027) clinical isolate 431, and followed the mice for the development of severe outcomes of moribundity and histopathologic cecal tissue damage.
+We then challenged the mice with *C.
+difficile* ribotype 027 (RT027) and followed the mice for the development of severe outcomes of moribundity and histopathologic cecal tissue damage.
 Since the murine host and *C.
-difficile* isolate were the same and only the gut community varied, the variation in disease severity we observed was attributable to the microbiome.
+difficile* isolate were the same and only the gut community varied, the variation in disease severity we observed was attributable to the gut microbiome.
 
@@
 These human fecal samples were selected because they represented diverse community structures based on community clustering [@Schubert2014].
-The gut communities were allowed to equlibriate for two weeks post inoculation [@Gillilland2012].
+The gut communities were allowed to equilibrate for two weeks post-inoculation [@Gillilland2012].
 We then surveyed the bacterial members of the gut communities by 16S rRNA gene sequencing of murine fecal pellets (Figure 1A).
@@
 However, we wanted to avoid modifying the gut communities with an antibiotic to maintain their unique microbial compositions and ecological relationships.
-Since some of these communities came from people at increased risk of CDI, for example due to hospitalization or recent antibiotic use [@Schubert2014], we decided to test if *C.
+Since some of these communities came from people at increased risk of CDI, such as recent hospitalization or antibiotic use [@Schubert2014], we decided to test if *C.
 difficile* was able to infect these mice without any antibiotic perturbation.
 We hypothesized that *C.
-difficile* would be able to colonize the mice who received their gut communities from a donor with perturbed community.
+difficile* would be able to colonize the mice who received their gut communities from a donor with a perturbed community.
 Mice were challenged with 10$^{3}$ *C.
-difficile* RT027 clinical isolate 431 spores.
+difficile* RT027 clinical isolate spores.
 The mice were followed for 10 days post-challenge, and their stool was collected and plated for *C.
@@
 Two mice were able to resist *C.
-difficile* colonization, both received their community donor N3, which may be attributed to experimental variation since this group also had more mice.
+difficile* colonization, both received their community donor N1, which may be attributed to experimental variation since this group also had more mice.
 By colonizing germ-free mice with different human fecal communities, we were able to generate diverse gut communities in mice, which were susceptible to *C.
@@
 We followed the mice for 10 days post-challenge for colonization density, toxin production, and mortality.
-Seven mice, from Donors N1, N2 and N5, were not colonized at detectable levels on the day after *C.
+Seven mice, from Donors N1, N3, N4, and N5, were not colonized at detectable levels on the day after *C.
 difficile* challenge but were infected (>10$^{6}$) by the end of the experiment.
 All mice that received their community from Donor M1 through M6 succumbed to the infection and became moribund within 3 days post-challenge.
-The remaining mice, except the uninfected Donor N3 mice, maintained *C.
+The remaining mice, except the uninfected Donor N1 mice, maintained *C.
 difficile* infection through the end of the experiment (Figure 2).
@@
 However, when looking at each group of mice, we observed a range in toxin activity for both the moribund and non-moribund mice (Figure 3A).
-Non-moribund mice from Donors N4 through N9 had comparable toxin activity as the moribund mice.
+Non-moribund mice from Donors N2 and N5 through N9 had comparable toxin activity as the moribund mice.
 Additionally, not all moribund mice had toxin activity detected in their stool.
-Next we examined the cecal tissue for histopathologic damage.
+Next, we examined the cecal tissue for histopathologic damage.
 Moribund mice had high levels of epithelial damage, tissue edema, and inflammation (Figure S1) similar to previously reported histopathologic findings for *C.
@@
 To understand the role of toxin activity in disease severity, we applied LEfSe to identify the genera most likely to explain the differences between the presence and absence of detected toxin activity (Figure 4B).
-Many genera that associated with toxin production were also associated with moribundity, such as populations of *Escherichia/Shigella* and *Bacteroides*.
-Likewise, there were genera such as *Anaerotignum*, *Enterocloster*, and *Murimonas* that were associated with no detected toxin production that also exhibited greater relative abundance in communities from non-moribund mice with a low histopathologic score.
+Many genera that associated with the presence of toxin were also associated with moribundity, such as populations of *Escherichia/Shigella* and *Bacteroides*.
+Likewise, there were genera such as *Anaerotignum*, *Enterocloster*, and *Murimonas* that were associated with no detected toxin that also exhibited greater relative abundance in communities from non-moribund mice with a low histopathologic score.
 Lastly, we tested for correlations between the endpoint relative abundances of bacterial operational taxonomic units (OTUs) and the histopathologic summary score (Figure 4C).
@@
 Populations of *Klebsiella* and *Prevotellaceae* were positively correlated with the histopathologic score and were increased in the group of mice with detectable toxin.
-This analysis identified bacterial genera that were associated with the variation in moribundity, histopathologic score, and toxin activity of CDI.
+This analysis identified bacterial genera that were associated with the variation in moribundity, histopathologic score, and toxin.
 
-We next determined whether collectively, bacterial community membership and relative abundance could be predictive of the CDI disease outcome.
-We trained random forest models with bacterial community relative abundance data from the day of colonization at each taxonomic rank to predict toxin production, moribundity, and day 10 post-challenge histopathologic summary score.
-For predicting if toxin would be produced, microbial populations aggregated by phylum rank classification performed similarly as models using lower taxonomic ranks (AUROC = 0.83, Figure S2).
+We next determined whether, collectively, bacterial community membership and relative abundance could be predictive of the CDI disease outcome.
+We trained random forest models with bacterial community relative abundance data from the day of colonization at each taxonomic rank to predict toxin, moribundity, and day 10 post-challenge histopathologic summary score.
+For predicting if detectable toxin would be produced, microbial populations aggregated by phylum rank classification performed similarly as models using lower taxonomic ranks (AUROC = 0.83, Figure S2).
 *C.
-difficile* was more likely to produce toxin when the community it infected had less abundant populations of *Verrucomicrobia* and *Campilobacterota* and had more abundant populations of *Proteobacteria* (Figure 5A).
+difficile* was more likely to produce detectable toxin when the community infected had less abundant populations of *Verrucomicrobia* and *Campilobacterota* and had more abundant populations of *Proteobacteria* (Figure 5A).
 Next, we assessed the ability of the community to predict moribundity.
@@
 Communities with greater abundances of *Hungatella*, *Eggerthella*, *Bifidobacterium*, *Duncaniella* and *Neisseria* were more likely to have high histopathologic scores.
-These models have shown that the relative abundance of bacterial populations and their relationship to each other could be used to predict the variation in moribundity, histopathologic score, and toxin activity of CDI.
+These models have shown that the relative abundance of bacterial populations and their relationship to each other could be used to predict the variation in moribundity, histopathologic score, and detectable toxin of CDI.
 
@@
 
-Challenging mice colonized with different human fecal communities with a single *C.
-difficile* isolate, RT027 isolate 431, demonstrated that variation in the gut bacteria affects the *C.
+Challenging mice colonized with different human fecal communities with *C.
+difficile* RT027 demonstrated that variation in members of the gut microbiome affects *C.
 difficile* infection disease severity.
@@
 difficile* infection itself, confounding any association between variation in severity and the microbiome.
-**Add Hikiro et al**
-With our recent observations that the initial community affected the ability of *C.
-difficile* to persist in the gut [@Tomkovich2020; @Lesniak2021] and the existing research describing the numerous interactions between the microbiome and *C.
-difficile*, we hypothesized that the microbiome can modulate CDI disease severity.
+
+Recently, Nagao-Kitamoto *et al.* demonstrated when only allowing the microbiome to vary in a murine *C.
+difficile* infection experiemnt that the resultant disease variation associated with the microbiome communities [@NagaoKitamoto2020].
 Our data have demonstrated gut bacterial relative abundances associate with variation in toxin production, histopathologic scoring of the cecal tissue and mortality.
@@
 *C.
-difficile* isolate came from Carlson *et al.* which had previously been isolated and characterized [@Carlson2013; @Carlson2015] and has recently been further characterized [@Saund2021].
+difficile* clinical isolate 431 came from Carlson *et al.* which had previously been isolated and characterized [@Carlson2013; @Carlson2015] and has recently been further characterized [@Saund2021].
 Spores concentration were determined both before and after challenge [@Sorg2009].
@@
 *P* values were then corrected for multiple comparisons with a Benjamini and Hochberg adjustment for a type I error rate of 0.05 [@Benjamini1995].
-We built random forest models using the mikropml package [@Topcuoglu2021] with OTUs from day 0 samples using mtry values of 1 through 10, 15, 20, 25, 40, 50, 100.
+We built random forest models using the mikropml package [@Topcuoglu2021] with relative abundance summed by taxonomic ranks from day 0 samples using mtry values of 1 through 10, 15, 20, 25, 40, 50, 100.
 The split for training and testing varied by model to avoid overfitting the data.
@@
 The histopathologic score model was trained with 80% of the data.
-Lastly, we did not compare murine communities to donor community or clinical data because germ-free mice colonized with non-murine fecal communities have been shown to more closely resemble the recipient species than the donor species [@Rawls2006].
+Lastly, we did not compare murine communities to donor community or clinical data because germ-free mice colonized with non-murine fecal communities have been shown to more closely resemble the murine communities than the donor species community [@Rawls2006].
 Furthermore, it is not our intention to make any inferences regarding human associated bacteria and their relationship with human CDI outcome.
@@
 Each column of abundances represents an individual mouse.
-Mice that received the same donor feces are grouped together and labeled above with a letter (N for non-moribund mice and M for moribund mice) and number (ordered by mean histopathologic score).
+Mice that received the same donor feces are grouped together and labeled above with a letter (N for non-moribund mice and M for moribund mice) and number (ordered by mean histopathologic score of the donor group).
 + indicates the mice which did not have detectable *C.
-difficile* CFU in Figure 2.
+difficile* CFU (Figure 2).
 (B) Median (points) and interquartile range (lines) of $\beta$-diversity ($\theta$~YC~) between an individual mouse and either all others which were inoculated with feces from the same donor or from a different donor.
@@
 Each point represents an individual mouse.
-Mice are grouped by donor and labeled by the donor letter (N for non-moribund mice and M for moribund mice) and number (ordered by mean histopathologic score).
+Mice are grouped by donor and labeled by the donor letter (N for non-moribund mice and M for moribund mice) and number (ordered by mean histopathologic score of the donor group).
 Points are colored by donor group.
@@
 Each point represents an individual mouse.
-Mice are grouped by donor and labeled by the donor letter (N for non-moribund mice and M for moribund mice) and number (ordered by mean histopathologic score).
+Mice are grouped by donor and labeled by the donor letter (N for non-moribund mice and M for moribund mice) and number (ordered by mean histopathologic score of the donor group).
 Points are colored by donor group.
 Mice in group N1 that have a summary score of 0 are the mice which did not have detectable *C.
-difficile* CFU in Figure 2.
+difficile* CFU (Figure 2).
 Missing points are from mice that had insufficient fecal sample collected for assaying toxin or cecum for histopathologic scoring.
+LOD = Limit of detection.
 
@@
 Individual fecal bacterial community members of the murine gut associated with *C .difficile* infection outcomes.** (A and B) Relative abundance of genera at the time of *C.
-difficile* challenge (Day 0) that varied significantly by the moribundity and histopathologic summary score or toxin activity by LEfSe analysis.
+difficile* challenge (Day 0) that varied significantly by the moribundity and histopathologic summary score or detected toxin by LEfSe analysis.
 Median (points) and interquartile range (lines) are plotted.
@@
 Fecal bacterial community members of the murine gut at the time of *C.
-difficile* (Day 0) predicted outcomes of the infection.** Day 0 bacterial community members grouped by different classification rank were modeled with random forest to predict the infection outcome.
+difficile* infection predicted outcomes of the infection.** On the day of infection (Day 0), bacterial community members grouped by different classification rank were modeled with random forest to predict the infection outcome.
 The models used the highest taxonomic classification rank that performed as well as the lower ranks.
@@
 (C) Histopathologic score modeled which mice would have a high (score greater than the median score of 5) or low (score less than the median score of 5) histopathologic summary score.
-(D) Bacterial phyla which affected the performance of predicting toxin activity when permuted.
+(D) Bacterial phyla which affected the performance of predicting detectable toxin activity when permuted.
 (E) Bacterial classes which affected the performance of predicting moribundity when permuted.

### 2022-01-23 submission/manuscript.tex
proofread coathur edits

@@
 toxin genes have been associated with disease severity. However, it is
-unclear whether there is an association between the gut microbiota and
-disease severity. The gut microbiota is known to interact with \emph{C.
-difficile} during infection. Perturbations to the gut microbiota are
-necessary for \emph{C. difficile} to colonize the gut. The gut
-microbiota is thought to impair \emph{C. difficile} colonization through
+unclear whether there is an association between members of the gut
+microbiota and disease severity. The gut microbiota is known to interact
+with \emph{C. difficile} during infection. Perturbations to the gut
+microbiota are necessary for \emph{C. difficile} to colonize the gut.
+The gut microbiota can inhibit \emph{C. difficile} colonization through
 bile acid metabolism, nutrient consumption and bacteriocin production.
 Here we sought to demonstrate that members of the gut bacterial
-communities can contribute to disease severity. We derived diverse gut
-communities by colonizing germ-free mice with different human fecal
+communities can also contribute to disease severity. We derived diverse
+gut communities by colonizing germ-free mice with different human fecal
 communities. The mice were then infected with a single \emph{C.
@@
 as \emph{Anaerostipes} and \emph{Coprobacillus}, were associated with
-less severe outcomes. These data indicate that gut bacteria can
-influence CDI disease severity.
+less severe outcomes. These data indicate that, in addition to the host
+and \emph{C. difficile}, populations of gut bacteria can influence CDI
+disease severity.
 
@@
 or develop into an infection, ranging in severity from mild diarrhea to
-toxic megacolon, sepsis and death. Models that predict severity and
+toxic megacolon, sepsis, and death. Models that predict severity and
 guide treatment decisions are based on clinical factors and \emph{C.
 difficile} characteristics. Although the gut microbiome plays a role in
-protecting against CDI, it has not been investigated for its effect on
-CDI disease severity or incorporated into attempts to predict disease
-severity. We demonstrated that variation in the microbiome of mice
-colonized with human feces yielded a range of disease outcomes. These
-results revealed groups of bacteria associated with both severe and mild
-\emph{C. difficile} infection outcomes. Gut bacterial community data
-from patients with CDI could improve our ability to identify patients at
-risk of developing more severe disease and improve interventions which
-target \emph{C. difficile} and the gut bacteria to reduce host damage.
+protecting against CDI, it has its effect on CDI disease severity is
+unclear and has not been incorporated into disease severity models. We
+demonstrated that variation in the microbiome of mice colonized with
+human feces yielded a range of disease outcomes. These results revealed
+groups of bacteria associated with both severe and mild \emph{C.
+difficile} infection outcomes. Gut bacterial community data from
+patients with CDI could improve our ability to identify patients at risk
+of developing more severe disease and improve interventions which target
+\emph{C. difficile} and the gut bacteria to reduce host damage.
 
@@
 the cause of antibiotic-associated pseudomembranous colitis (1). CDI
-disease severity can range from asymptomatic to mild diarrhea to toxic
-megacolon and death. The Infectious Diseases Society of America (IDSA)
-and Society for Healthcare Epidemiology of America (SHEA) guidelines
-defines severe CDI in terms of a white blood cell count greater than
-15,000 cells/mm\(^3\) and/or a serum creatinine greater than 1.5 mg/dL.
-Patients who develop shock or hypotension, ileus, or toxic megacolon are
-considered to have fulminant CDI (2). Since these measures are outcomes
-of CDI, these measures have limited ablity to predict risk of severe CDI
-when CDI is fiest detected. Schemes have been developed to score a
-patient's risk for severe CDI outcomes based on clinical factors but
-have not been robust for broad application (3). Thus, we have limited
-ability to prevent patients from developing severe CDI. Ideally, we
-could identify patients that are high-risk of severe disease so their
-treatments could be adjusted to alter the clinical trajectory.
+disease severity can range from mild diarrhea to toxic megacolon and
+death. The Infectious Diseases Society of America (IDSA) and Society for
+Healthcare Epidemiology of America (SHEA) guidelines define severe CDI
+in terms of a white blood cell count greater than 15,000 cells/mm\(^3\)
+and/or a serum creatinine greater than 1.5 mg/dL. Patients who develop
+shock or hypotension, ileus, or toxic megacolon are considered to have
+fulminant CDI (2). Since these measures are CDI outcomes, they have
+limited ability to predict risk of severe CDI when the infection is
+first detected. Schemes have been developed to score a patient's risk
+for severe CDI outcomes based on clinical factors but have not been
+robust for broad application (3). Thus, we have limited ability to
+prevent patients from developing severe CDI.
 
@@
 
-Here, we sought to determine whether the gut bacteria contribute to
-disease severity. We hypothesized that since the gut bacteria affect the
-metabolism of \emph{C. difficile} and its infection dynamics, it can
-also affect the disease severity of the infection. To test this
+Recently, Nagao-Kitamoto \emph{et al.} showed germ-free mice which
+received a microbial community from different human donors resulted in
+variation in a subsequent \emph{C. difficile} infection (14). Here, we
+sought to further elucidate the relationship between members of the gut
+bacterial community and CDI disease severity. We hypothesized that since
+specific groups of gut bacteria affect the metabolism of \emph{C.
+difficile} and its infection dynamics, we can also identify groups of
+bacteria that affect the disease severity of the infection. To test this
 hypothesis, we colonized germ-free C57BL/6 mice with human fecal samples
-to create varied gut communities. We then challenged the mice with a
-single \emph{C. difficile} isolate, ribotype 027 (RT027) clinical
-isolate 431, and followed the mice for the development of severe
-outcomes of moribundity and histopathologic cecal tissue damage. Since
-the murine host and \emph{C. difficile} isolate were the same and only
-the gut community varied, the variation in disease severity we observed
-was attributable to the microbiome.
+to create varied gut communities. We then challenged the mice with
+\emph{C. difficile} ribotype 027 (RT027) and followed the mice for the
+development of severe outcomes of moribundity and histopathologic cecal
+tissue damage. Since the murine host and \emph{C. difficile} isolate
+were the same and only the gut community varied, the variation in
+disease severity we observed was attributable to the gut microbiome.
 
@@
 mouse colonies, we colonized germ-free mice with bacteria from human
-feces (14). We inoculated germ-free C57BL/6 mice with homogenized feces
+feces (15). We inoculated germ-free C57BL/6 mice with homogenized feces
 from one of 15 human fecal samples via oral gavage. These human fecal
 samples were selected because they represented diverse community
-structures based on community clustering (15). The gut communities were
-allowed to equlibriate for two weeks post inoculation (16). We then
+structures based on community clustering (16). The gut communities were
+allowed to equilibrate for two weeks post-inoculation (17). We then
 surveyed the bacterial members of the gut communities by 16S rRNA gene
@@
 with antibiotics, such as clindamycin, to become susceptible to \emph{C.
-difficile} colonization (17, 18). However, we wanted to avoid modifying
+difficile} colonization (18, 19). However, we wanted to avoid modifying
 the gut communities with an antibiotic to maintain their unique
 microbial compositions and ecological relationships. Since some of these
-communities came from people at increased risk of CDI, for example due
-to hospitalization or recent antibiotic use (15), we decided to test if
-\emph{C. difficile} was able to infect these mice without any antibiotic
+communities came from people at increased risk of CDI, such as recent
+hospitalization or antibiotic use (16), we decided to test if \emph{C.
+difficile} was able to infect these mice without any antibiotic
 perturbation. We hypothesized that \emph{C. difficile} would be able to
-colonize the mice who received their gut communities from a donor with
+colonize the mice who received their gut communities from a donor with a
 perturbed community. Mice were challenged with 10\(^{3}\) \emph{C.
-difficile} RT027 clinical isolate 431 spores. The mice were followed for
-10 days post-challenge, and their stool was collected and plated for
+difficile} RT027 clinical isolate spores. The mice were followed for 10
+days post-challenge, and their stool was collected and plated for
 \emph{C. difficile} colony forming units (CFU) to determine the extent
@@
 be colonized (Figure 2). Two mice were able to resist \emph{C.
-difficile} colonization, both received their community donor N3, which
+difficile} colonization, both received their community donor N1, which
 may be attributed to experimental variation since this group also had
@@
 colonization density, toxin production, and mortality. Seven mice, from
-Donors N1, N2 and N5, were not colonized at detectable levels on the day
-after \emph{C. difficile} challenge but were infected
+Donors N1, N3, N4, and N5, were not colonized at detectable levels on
+the day after \emph{C. difficile} challenge but were infected
 (\textgreater10\(^{6}\)) by the end of the experiment. All mice that
@@
 infection and became moribund within 3 days post-challenge. The
-remaining mice, except the uninfected Donor N3 mice, maintained \emph{C.
+remaining mice, except the uninfected Donor N1 mice, maintained \emph{C.
 difficile} infection through the end of the experiment (Figure 2). At 10
@@
 moribund and non-moribund mice (Figure 3A). Non-moribund mice from
-Donors N4 through N9 had comparable toxin activity as the moribund mice.
-Additionally, not all moribund mice had toxin activity detected in their
-stool. Next we examined the cecal tissue for histopathologic damage.
-Moribund mice had high levels of epithelial damage, tissue edema, and
-inflammation (Figure S1) similar to previously reported histopathologic
-findings for \emph{C. difficile} RT027 (19). As observed with toxin
-activity, the moribund mice had higher histopathologic scores than the
-non-moribund mice (\emph{P} \textless{} 0.001). However, unlike the
-toxin activity, all moribund mice had consistently high histopathologic
-summary scores (Figure 3B). The non-moribund mice, Donor groups N1
-through N9, had a range in tissue damage from none detected to similar
-levels as the moribund mice, which grouped by community donor. Together,
-the toxin activity, histopathologic score, and moribundity showed
-variation across the donor groups but were largely consistent within
-each donor group.
+Donors N2 and N5 through N9 had comparable toxin activity as the
+moribund mice. Additionally, not all moribund mice had toxin activity
+detected in their stool. Next, we examined the cecal tissue for
+histopathologic damage. Moribund mice had high levels of epithelial
+damage, tissue edema, and inflammation (Figure S1) similar to previously
+reported histopathologic findings for \emph{C. difficile} RT027 (20). As
+observed with toxin activity, the moribund mice had higher
+histopathologic scores than the non-moribund mice (\emph{P} \textless{}
+0.001). However, unlike the toxin activity, all moribund mice had
+consistently high histopathologic summary scores (Figure 3B). The
+non-moribund mice, Donor groups N1 through N9, had a range in tissue
+damage from none detected to similar levels as the moribund mice, which
+grouped by community donor. Together, the toxin activity,
+histopathologic score, and moribundity showed variation across the donor
+groups but were largely consistent within each donor group.
 
@@
 the differences between the presence and absence of detected toxin
-activity (Figure 4B). Many genera that associated with toxin production
-were also associated with moribundity, such as populations of
+activity (Figure 4B). Many genera that associated with the presence of
+toxin were also associated with moribundity, such as populations of
 \emph{Escherichia/Shigella} and \emph{Bacteroides}. Likewise, there were
 genera such as \emph{Anaerotignum}, \emph{Enterocloster}, and
-\emph{Murimonas} that were associated with no detected toxin production
-that also exhibited greater relative abundance in communities from
-non-moribund mice with a low histopathologic score. Lastly, we tested
-for correlations between the endpoint relative abundances of bacterial
+\emph{Murimonas} that were associated with no detected toxin that also
+exhibited greater relative abundance in communities from non-moribund
+mice with a low histopathologic score. Lastly, we tested for
+correlations between the endpoint relative abundances of bacterial
 operational taxonomic units (OTUs) and the histopathologic summary score
@@
 were associated with the variation in moribundity, histopathologic
-score, and toxin activity of CDI.
-
-We next determined whether collectively, bacterial community membership
+score, and toxin.
+
+We next determined whether, collectively, bacterial community membership
 and relative abundance could be predictive of the CDI disease outcome.
@@
 abundance data from the day of colonization at each taxonomic rank to
-predict toxin production, moribundity, and day 10 post-challenge
-histopathologic summary score. For predicting if toxin would be
-produced, microbial populations aggregated by phylum rank classification
-performed similarly as models using lower taxonomic ranks (AUROC = 0.83,
-Figure S2). \emph{C. difficile} was more likely to produce toxin when
-the community it infected had less abundant populations of
+predict toxin, moribundity, and day 10 post-challenge histopathologic
+summary score. For predicting if detectable toxin would be produced,
+microbial populations aggregated by phylum rank classification performed
+similarly as models using lower taxonomic ranks (AUROC = 0.83, Figure
+S2). \emph{C. difficile} was more likely to produce detectable toxin
+when the community infected had less abundant populations of
 \emph{Verrucomicrobia} and \emph{Campilobacterota} and had more abundant
@@
 populations and their relationship to each other could be used to
-predict the variation in moribundity, histopathologic score, and toxin
-activity of CDI.
+predict the variation in moribundity, histopathologic score, and
+detectable toxin of CDI.
 
@@
 
-Challenging mice colonized with different human fecal communities with a
-single \emph{C. difficile} isolate, RT027 isolate 431, demonstrated that
-variation in the gut bacteria affects the \emph{C. difficile} infection
-disease severity. Our analysis revealed an association between the
-relative abundance of bacterial community members and disease severity.
-Previous studies investigating the severity of CDI disease involving the
-microbiome have had limited ability to interrogate this relationship
-between the microbiome and disease severity. Studies that have used
-clinical data have limited ability to control variation in the host,
-microbiome or \emph{C. difficile} ribotype (20). Murine experiments
-typically use a single mouse colony and different \emph{C. difficile}
-ribotypes to create severity differences (21). Recently, our group has
-begun uncovering the effect microbiome variation has on \emph{C.
-difficile} infection. We showed the variation in the bacterial
-communities between mice from different mouse colonies resulted in
-different clearance rates of \emph{C. difficile} (22). We also showed
-varied ability of mice to spontaneously eliminate \emph{C. difficile}
-infection when they were treated with different antibiotics prior to
-\emph{C. difficile} challenge (23). Overall, the results presented here
-have demonstrated that the gut bacterial community contributed to the
-severity of \emph{C. difficile} infection.
+Challenging mice colonized with different human fecal communities with
+\emph{C. difficile} RT027 demonstrated that variation in members of the
+gut microbiome affects \emph{C. difficile} infection disease severity.
+Our analysis revealed an association between the relative abundance of
+bacterial community members and disease severity. Previous studies
+investigating the severity of CDI disease involving the microbiome have
+had limited ability to interrogate this relationship between the
+microbiome and disease severity. Studies that have used clinical data
+have limited ability to control variation in the host, microbiome or
+\emph{C. difficile} ribotype (21). Murine experiments typically use a
+single mouse colony and different \emph{C. difficile} ribotypes to
+create severity differences (22). Recently, our group has begun
+uncovering the effect microbiome variation has on \emph{C. difficile}
+infection. We showed the variation in the bacterial communities between
+mice from different mouse colonies resulted in different clearance rates
+of \emph{C. difficile} (23). We also showed varied ability of mice to
+spontaneously eliminate \emph{C. difficile} infection when they were
+treated with different antibiotics prior to \emph{C. difficile}
+challenge (24). Overall, the results presented here have demonstrated
+that the gut bacterial community contributed to the severity of \emph{C.
+difficile} infection.
 
@@
 severe infection using white blood cell counts, serum albumin level, or
-serum creatinine level (2, 24, 25). Those levels are driven by the
-activities in the intestine (26). Research into the drivers of this
+serum creatinine level (2, 25, 26). Those levels are driven by the
+activities in the intestine (27). Research into the drivers of this
 variation have revealed factors that make \emph{C. difficile} more
@@
 and production of the toxins TcdA, TcdB, and binary toxin and the
-prevalence in outbreaks, such as ribotypes 027 and 078 (17, 27--30).
+prevalence in outbreaks, such as ribotypes 027 and 078 (18, 28--31).
 However, other studies have shown that disease is not necessarily linked
-with toxin production (31) or the strain (32). Furthermore, there is
+with toxin production (32) or the strain (33). Furthermore, there is
 variation in the genome, growth rate, sporulation, germination, and
-toxin production in different isolates of a strain (33--36). This
+toxin production in different isolates of a strain (34--37). This
 variation may help explain why severe CDI prediction tools often miss
 identifying many patients with CDI that will develop severe disease (3,
-21, 37, 38). Therefore, it is necessary to gain a full understanding of
+22, 38, 39). Therefore, it is necessary to gain a full understanding of
 all factors contributing to disease variation to improve our ability to
@@
 difficile} is unable to colonize an unperturbed healthy gut community
-and is only able to become established after a perturbation (18). Once
+and is only able to become established after a perturbation (19). Once
 colonized, the different communities lead to different metabolic
-responses and dynamics of the \emph{C. difficile} population (8, 23,
-39). Gut bacteria metabolize primary bile acids into secondary bile
-acids (40, 41). The concentration of these bile acids affects
-germination, growth, toxin production and biofilm formation (9, 10, 42,
-43). Members of the bacterial community also affect other metabolites
+responses and dynamics of the \emph{C. difficile} population (8, 24,
+40). Gut bacteria metabolize primary bile acids into secondary bile
+acids (41, 42). The concentration of these bile acids affects
+germination, growth, toxin production and biofilm formation (9, 10, 43,
+44). Members of the bacterial community also affect other metabolites
 \emph{C. difficile} utilizes. \emph{Bacteroides thetaiotaomicron}
 produce sialidases which release sialic acid from the mucosa for
-\emph{C. difficile} to utilize (44, 45). The nutrient environment
-affects toxin production (46). Thus, many of the actions of the gut
+\emph{C. difficile} to utilize (45, 46). The nutrient environment
+affects toxin production (47). Thus, many of the actions of the gut
 bacteria modulate \emph{C. difficile} in ways that could affect the
@@
 used multiple human communities to colonize mice, however the
-communities were pooled prior to gavaging into germ-free mice (47),
+communities were pooled prior to gavaging into germ-free mice (48),
 resulting in a single human-derived community. Studies examining
@@
 ribotypes in mice with similar microbiota as a proxy for variation in
-disease, such as strain 630 for non-severe and RT027 for severe (17, 27,
-28, 48). Studies have also demonstrated variation in infection through
-tapering antibiotic dosage (18, 23, 49) or by reducing the amount of
-\emph{C. difficile} cells or spores used for the challenge (17, 49).
+disease, such as strain 630 for non-severe and RT027 for severe (18, 28,
+29, 49). Studies have also demonstrated variation in infection through
+tapering antibiotic dosage (19, 24, 50) or by reducing the amount of
+\emph{C. difficile} cells or spores used for the challenge (18, 50).
 These studies either lack variation in the initial microbiome or have
@@
 association between variation in severity and the microbiome.
-\textbf{Add Hikiro et al} With our recent observations that the initial
-community affected the ability of \emph{C. difficile} to persist in the
-gut (22, 23) and the existing research describing the numerous
-interactions between the microbiome and \emph{C. difficile}, we
-hypothesized that the microbiome can modulate CDI disease severity. Our
-data have demonstrated gut bacterial relative abundances associate with
-variation in toxin production, histopathologic scoring of the cecal
-tissue and mortality. This analysis revealed populations of
-\emph{Akkermansia}, \emph{Anaerostipes}, \emph{Coprobacillus},
-\emph{Enterocloster}, \emph{Lactonifactor}, and \emph{Monoglobus} were
-more abundant in the microbiome of non-moribund mice which had low
-histopathologic scores and no detected toxin. The protective role of
-these genera are supported by previous studies. \emph{Coprobacillus},
-\emph{Lactonifactor}, and \emph{Monoglobus} have been shown to be
-involved in dietary fiber fermentation and associated with healthy
-communities (50--53). \emph{Anaerostipes} and \emph{Coprobacillus},
-which produce short chain fatty acids, have been associated with healthy
-communities (54--56). Furthermore, \emph{Coprobacillus}, which was
-abundant in mice with low histopathologic scores but rare in all other
-mice, has been shown to contain a putative type I lantibiotic gene
-cluster and inhibit \emph{C. difficile} colonization (57--59).
-\emph{Akkermansia} and \emph{Enterocloster} were also identified as more
-abundant in mice which had a low histopathologic scores but have
-contradictory supporting evidence in the current literature. In our
-data, \emph{Akkermansia} was most abundant in the non-moribund mice with
-low histopathologic scores but there were some moribund mice which had
-increased populations of \emph{Akkermansia}. This could be attributed to
-either a more protective mucus layer was present inhibiting colonization
-(59, 60) or mucus consumption by \emph{Akkermansia} could have been
-crossfeeding \emph{C. difficile} or exposing a niche for \emph{C.
-difficile} (61--63). Similarly, \emph{Enterocloster} was more abundant
-and associated with low histopathologic scores. It has been associated
-with healthy populations and has been used to mono-colonize germ-free
-mice to reduce the ability of \emph{C. difficile} to colonize (64, 65).
-However, \emph{Enterocloster} has also been involved in infections, such
-as bacteremia (66, 67). These data have exemplified populations of
-bacteria that have the potential to be either protective or harmful.
-Thus, the disease outcome is not likely based on the abundance of
-individual populations of bacteria, rather it is the result of the
-interactions of the community.
+
+Recently, Nagao-Kitamoto \emph{et al.} demonstrated when only allowing
+the microbiome to vary in a murine \emph{C. difficile} infection
+experiemnt that the resultant disease variation associated with the
+microbiome communities (14). Our data have demonstrated gut bacterial
+relative abundances associate with variation in toxin production,
+histopathologic scoring of the cecal tissue and mortality. This analysis
+revealed populations of \emph{Akkermansia}, \emph{Anaerostipes},
+\emph{Coprobacillus}, \emph{Enterocloster}, \emph{Lactonifactor}, and
+\emph{Monoglobus} were more abundant in the microbiome of non-moribund
+mice which had low histopathologic scores and no detected toxin. The
+protective role of these genera are supported by previous studies.
+\emph{Coprobacillus}, \emph{Lactonifactor}, and \emph{Monoglobus} have
+been shown to be involved in dietary fiber fermentation and associated
+with healthy communities (51--54). \emph{Anaerostipes} and
+\emph{Coprobacillus}, which produce short chain fatty acids, have been
+associated with healthy communities (55--57). Furthermore,
+\emph{Coprobacillus}, which was abundant in mice with low
+histopathologic scores but rare in all other mice, has been shown to
+contain a putative type I lantibiotic gene cluster and inhibit \emph{C.
+difficile} colonization (58--60). \emph{Akkermansia} and
+\emph{Enterocloster} were also identified as more abundant in mice which
+had a low histopathologic scores but have contradictory supporting
+evidence in the current literature. In our data, \emph{Akkermansia} was
+most abundant in the non-moribund mice with low histopathologic scores
+but there were some moribund mice which had increased populations of
+\emph{Akkermansia}. This could be attributed to either a more protective
+mucus layer was present inhibiting colonization (60, 61) or mucus
+consumption by \emph{Akkermansia} could have been crossfeeding \emph{C.
+difficile} or exposing a niche for \emph{C. difficile} (62--64).
+Similarly, \emph{Enterocloster} was more abundant and associated with
+low histopathologic scores. It has been associated with healthy
+populations and has been used to mono-colonize germ-free mice to reduce
+the ability of \emph{C. difficile} to colonize (65, 66). However,
+\emph{Enterocloster} has also been involved in infections, such as
+bacteremia (67, 68). These data have exemplified populations of bacteria
+that have the potential to be either protective or harmful. Thus, the
+disease outcome is not likely based on the abundance of individual
+populations of bacteria, rather it is the result of the interactions of
+the community.
 
@@
 \emph{Shigella/Escherichia}, \emph{Staphylococcus}, and
-\emph{Streptococcus} have been shown to expand after antibiotic use (15,
-68, 69) and are commonly detected in CDI cases (70--73). In addition to
+\emph{Streptococcus} have been shown to expand after antibiotic use (16,
+69, 70) and are commonly detected in CDI cases (71--74). In addition to
 these populations, \emph{Eggerthella}, \emph{Prevotellaceae} and
 \emph{Helicobacter}, which associated with worse outcomes, have also
-been associated with intestinal inflammation (74--76). Recently,
+been associated with intestinal inflammation (75--77). Recently,
 \emph{Helicobacter hepaticus} was shown to be sufficient to cause
-susceptibility to CDI in IL-10 deficient C57BL/6 mice (77). In our
+susceptibility to CDI in IL-10 deficient C57BL/6 mice (78). In our
 experiments, when \emph{Helicobacter} was present, the infection
@@
 severe disease it may be appropriate to apply the damage-response
-framework for microbial pathogenesis to CDI (78, 79). This framework
+framework for microbial pathogenesis to CDI (79, 80). This framework
 posits that disease is not driven by a single entity, rather it is an
@@
 host, such as host mucus glycosylation or intestinal IL-33 expression
-(80, 81). For example, low levels of mucin degradation can provide
+(14, 81). For example, low levels of mucin degradation can provide
 nutrients to other community members producing a diverse non-damaging
@@
 obtained as part of Schubert \emph{et al.} and selected based on
-community clusters (15) to result in diverse community structures. Feces
+community clusters (16) to result in diverse community structures. Feces
 were homogenized by mixing 200 mg of sample with 5 ml of PBS. Mice were
@@
 Two weeks after the fecal community inoculation, mice were challenged
-with \emph{C. difficile}. \emph{C. difficile} isolate came from Carlson
-\emph{et al.} which had previously been isolated and characterized (33,
-34) and has recently been further characterized (35). Spores
-concentration were determined both before and after challenge (83).
-\(10^{3}\) \emph{C. difficile} spores were given to each mouse via oral
-gavage.
+with \emph{C. difficile}. \emph{C. difficile} clinical isolate 431 came
+from Carlson \emph{et al.} which had previously been isolated and
+characterized (34, 35) and has recently been further characterized (36).
+Spores concentration were determined both before and after challenge
+(83). \(10^{3}\) \emph{C. difficile} spores were given to each mouse via
+oral gavage.
 
@@
 Toxin activity was assessed using a Vero cell rounding-based
-cytotoxicity assay as described previously (28). The cytotoxicity titer
+cytotoxicity assay as described previously (29). The cytotoxicity titer
 was determined for each sample as the last dilution, which resulted in
@@
 criteria by a board-certified veterinary pathologist scored who was
-blinded to the experimental groups (28).
+blinded to the experimental groups (29).
 
@@
 I error rate of 0.05 (90). We built random forest models using the
-mikropml package (91) with OTUs from day 0 samples using mtry values of
-1 through 10, 15, 20, 25, 40, 50, 100. The split for training and
-testing varied by model to avoid overfitting the data. To determine the
-optimal split, we tested splits (50\%, 60\%, 70\%, 80\%, 90\% data used
-for training) to find the greatest portion of data that could be used to
-train the model while still maintaining the same performance for the
-training model as the model with the held-out test data. The toxin and
-moribundity models were trained with 60\% of the data. The
-histopathologic score model was trained with 80\% of the data. Lastly,
-we did not compare murine communities to donor community or clinical
-data because germ-free mice colonized with non-murine fecal communities
-have been shown to more closely resemble the recipient species than the
-donor species (92). Furthermore, it is not our intention to make any
-inferences regarding human associated bacteria and their relationship
-with human CDI outcome.
+mikropml package (91) with relative abundance summed by taxonomic ranks
+from day 0 samples using mtry values of 1 through 10, 15, 20, 25, 40,
+50, 100. The split for training and testing varied by model to avoid
+overfitting the data. To determine the optimal split, we tested splits
+(50\%, 60\%, 70\%, 80\%, 90\% data used for training) to find the
+greatest portion of data that could be used to train the model while
+still maintaining the same performance for the training model as the
+model with the held-out test data. The toxin and moribundity models were
+trained with 60\% of the data. The histopathologic score model was
+trained with 80\% of the data. Lastly, we did not compare murine
+communities to donor community or clinical data because germ-free mice
+colonized with non-murine fecal communities have been shown to more
+closely resemble the murine communities than the donor species community
+(92). Furthermore, it is not our intention to make any inferences
+regarding human associated bacteria and their relationship with human
+CDI outcome.
 
@@
 \textbf{Schloss PD}, \textbf{Young VB}. 2021. Protection from lethal
-clostridioides difficile infection via intraspecies competition for
-cogerminant. mBio \textbf{12}.
+\emph{Clostridioides difficile} infection via intraspecies competition
+for cogerminant. mBio \textbf{12}.
 doi:\href{https://doi.org/10.1128/mbio.00522-21}{10.1128/mbio.00522-21}.
 
+\leavevmode\hypertarget{ref-NagaoKitamoto2020}{}%
+14. \textbf{Nagao-Kitamoto H}, \textbf{Leslie JL}, \textbf{Kitamoto S},
+\textbf{Jin C}, \textbf{Thomsson KA}, \textbf{Gillilland MG},
+\textbf{Kuffa P}, \textbf{Goto Y}, \textbf{Jenq RR}, \textbf{Ishii C},
+\textbf{Hirayama A}, \textbf{Seekatz AM}, \textbf{Martens EC},
+\textbf{Eaton KA}, \textbf{Kao JY}, \textbf{Fukuda S}, \textbf{Higgins
+PDR}, \textbf{Karlsson NG}, \textbf{Young VB}, \textbf{Kamada N}. 2020.
+Interleukin-22-mediated host glycosylation prevents \emph{Clostridioides
+difficile} infection by modulating the metabolic activity of the gut
+microbiota. Nature Medicine \textbf{26}:608--617.
+doi:\href{https://doi.org/10.1038/s41591-020-0764-0}{10.1038/s41591-020-0764-0}.
+
 \leavevmode\hypertarget{ref-Nagpal2018}{}%
-14. \textbf{Nagpal R}, \textbf{Wang S}, \textbf{Woods LCS},
+15. \textbf{Nagpal R}, \textbf{Wang S}, \textbf{Woods LCS},
 \textbf{Seshie O}, \textbf{Chung ST}, \textbf{Shively CA},
@@
 \leavevmode\hypertarget{ref-Schubert2014}{}%
-15. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
+16. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
 \textbf{Mogle J}, \textbf{Petrosino JP}, \textbf{Young VB},
@@
 \leavevmode\hypertarget{ref-Gillilland2012}{}%
-16. \textbf{Gillilland MG}, \textbf{Erb-Downward JR}, \textbf{Bassis
+17. \textbf{Gillilland MG}, \textbf{Erb-Downward JR}, \textbf{Bassis
 CM}, \textbf{Shen MC}, \textbf{Toews GB}, \textbf{Young VB},
@@
 \leavevmode\hypertarget{ref-Chen2008}{}%
-17. \textbf{Chen X}, \textbf{Katchar K}, \textbf{Goldsmith JD},
+18. \textbf{Chen X}, \textbf{Katchar K}, \textbf{Goldsmith JD},
 \textbf{Nanthakumar N}, \textbf{Cheknis A}, \textbf{Gerding DN},
@@
 \leavevmode\hypertarget{ref-Schubert2015}{}%
-18. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
+19. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
 Antibiotic-induced alterations of the murine gut microbiota and
@@
 \leavevmode\hypertarget{ref-Cowardin2016}{}%
-19. \textbf{Cowardin CA}, \textbf{Buonomo EL}, \textbf{Saleh MM},
+20. \textbf{Cowardin CA}, \textbf{Buonomo EL}, \textbf{Saleh MM},
 \textbf{Wilson MG}, \textbf{Burgess SL}, \textbf{Kuehne SA},
@@
 \leavevmode\hypertarget{ref-Seekatz2016}{}%
-20. \textbf{Seekatz AM}, \textbf{Rao K}, \textbf{Santhosh K},
+21. \textbf{Seekatz AM}, \textbf{Rao K}, \textbf{Santhosh K},
 \textbf{Young VB}. 2016. Dynamics of the fecal microbiome in patients
@@
 \leavevmode\hypertarget{ref-Dieterle2020}{}%
-21. \textbf{Dieterle MG}, \textbf{Putler R}, \textbf{Perry DA},
+22. \textbf{Dieterle MG}, \textbf{Putler R}, \textbf{Perry DA},
 \textbf{Menon A}, \textbf{Abernathy-Close L}, \textbf{Perlman NS},
@@
 \leavevmode\hypertarget{ref-Tomkovich2020}{}%
-22. \textbf{Tomkovich S}, \textbf{Stough JMA}, \textbf{Bishop L},
+23. \textbf{Tomkovich S}, \textbf{Stough JMA}, \textbf{Bishop L},
 \textbf{Schloss PD}. 2020. The initial gut microbiota and response to
@@
 \leavevmode\hypertarget{ref-Lesniak2021}{}%
-23. \textbf{Lesniak NA}, \textbf{Schubert AM}, \textbf{Sinani H},
+24. \textbf{Lesniak NA}, \textbf{Schubert AM}, \textbf{Sinani H},
 \textbf{Schloss PD}. 2021. Clearance of \emph{Clostridioides difficile}
@@
 \leavevmode\hypertarget{ref-Lungulescu2011}{}%
-24. \textbf{Lungulescu OA}, \textbf{Cao W}, \textbf{Gatskevich E},
+25. \textbf{Lungulescu OA}, \textbf{Cao W}, \textbf{Gatskevich E},
 \textbf{Tlhabano L}, \textbf{Stratidis JG}. 2011. CSI: A severity index
@@
 \leavevmode\hypertarget{ref-Zar2007}{}%
-25. \textbf{Zar FA}, \textbf{Bakkanagari SR}, \textbf{Moorthi KMLST},
+26. \textbf{Zar FA}, \textbf{Bakkanagari SR}, \textbf{Moorthi KMLST},
 \textbf{Davis MB}. 2007. A comparison of vancomycin and metronidazole
@@
 \leavevmode\hypertarget{ref-diMasi2018}{}%
-26. \textbf{Masi A di}, \textbf{Leboffe L}, \textbf{Polticelli F},
+27. \textbf{Masi A di}, \textbf{Leboffe L}, \textbf{Polticelli F},
 \textbf{Tonon F}, \textbf{Zennaro C}, \textbf{Caterino M}, \textbf{Stano
@@
 \leavevmode\hypertarget{ref-AbernathyClose2020}{}%
-27. \textbf{Abernathy-Close L}, \textbf{Dieterle MG}, \textbf{Vendrov
+28. \textbf{Abernathy-Close L}, \textbf{Dieterle MG}, \textbf{Vendrov
 KC}, \textbf{Bergin IL}, \textbf{Rao K}, \textbf{Young VB}. 2020. Aging
@@
 \leavevmode\hypertarget{ref-Theriot2011}{}%
-28. \textbf{Theriot CM}, \textbf{Koumpouras CC}, \textbf{Carlson PE},
+29. \textbf{Theriot CM}, \textbf{Koumpouras CC}, \textbf{Carlson PE},
 \textbf{Bergin II}, \textbf{Aronoff DM}, \textbf{Young VB}. 2011.
@@
 \leavevmode\hypertarget{ref-Goorhuis2008}{}%
-29. \textbf{Goorhuis A}, \textbf{Bakker D}, \textbf{Corver J},
+30. \textbf{Goorhuis A}, \textbf{Bakker D}, \textbf{Corver J},
 \textbf{Debast SB}, \textbf{Harmanus C}, \textbf{Notermans DW},
@@
 \leavevmode\hypertarget{ref-OConnor2009}{}%
-30. \textbf{O'Connor JR}, \textbf{Johnson S}, \textbf{Gerding DN}. 2009.
+31. \textbf{O'Connor JR}, \textbf{Johnson S}, \textbf{Gerding DN}. 2009.
 \emph{Clostridium difficile} infection caused by the epidemic
@@
 \leavevmode\hypertarget{ref-Rao2015}{}%
-31. \textbf{Rao K}, \textbf{Micic D}, \textbf{Natarajan M},
+32. \textbf{Rao K}, \textbf{Micic D}, \textbf{Natarajan M},
 \textbf{Winters S}, \textbf{Kiel MJ}, \textbf{Walk ST}, \textbf{Santhosh
@@
 \leavevmode\hypertarget{ref-Walk2012}{}%
-32. \textbf{Walk ST}, \textbf{Micic D}, \textbf{Jain R}, \textbf{Lo ES},
+33. \textbf{Walk ST}, \textbf{Micic D}, \textbf{Jain R}, \textbf{Lo ES},
 \textbf{Trivedi I}, \textbf{Liu EW}, \textbf{Almassalha LM},
@@
 \leavevmode\hypertarget{ref-Carlson2013}{}%
-33. \textbf{Carlson PE}, \textbf{Walk ST}, \textbf{Bourgis AET},
+34. \textbf{Carlson PE}, \textbf{Walk ST}, \textbf{Bourgis AET},
 \textbf{Liu MW}, \textbf{Kopliku F}, \textbf{Lo E}, \textbf{Young VB},
@@
 \leavevmode\hypertarget{ref-Carlson2015}{}%
-34. \textbf{Carlson PE}, \textbf{Kaiser AM}, \textbf{McColm SA},
+35. \textbf{Carlson PE}, \textbf{Kaiser AM}, \textbf{McColm SA},
 \textbf{Bauer JM}, \textbf{Young VB}, \textbf{Aronoff DM}, \textbf{Hanna
@@
 \leavevmode\hypertarget{ref-Saund2021}{}%
-35. \textbf{Saund K}, \textbf{Pirani A}, \textbf{Lacy B}, \textbf{Hanna
+36. \textbf{Saund K}, \textbf{Pirani A}, \textbf{Lacy B}, \textbf{Hanna
 PC}, \textbf{Snitkin ES}. 2021. Strain variation in \emph{Clostridioides
@@
 \leavevmode\hypertarget{ref-He2010}{}%
-36. \textbf{He M}, \textbf{Sebaihia M}, \textbf{Lawley TD},
+37. \textbf{He M}, \textbf{Sebaihia M}, \textbf{Lawley TD},
 \textbf{Stabler RA}, \textbf{Dawson LF}, \textbf{Martin MJ},
@@
 \leavevmode\hypertarget{ref-Butt2013}{}%
-37. \textbf{Butt E}, \textbf{Foster JA}, \textbf{Keedwell E},
+38. \textbf{Butt E}, \textbf{Foster JA}, \textbf{Keedwell E},
 \textbf{Bell JE}, \textbf{Titball RW}, \textbf{Bhangu A},
@@
 \leavevmode\hypertarget{ref-vanBeurden2017}{}%
-38. \textbf{Beurden YH van}, \textbf{Hensgens MPM}, \textbf{Dekkers OM},
+39. \textbf{Beurden YH van}, \textbf{Hensgens MPM}, \textbf{Dekkers OM},
 \textbf{Cessie SL}, \textbf{Mulder CJJ}, \textbf{Vandenbroucke-Grauls
@@
 \leavevmode\hypertarget{ref-Jenior2018}{}%
-39. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
+40. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
 \textbf{Schloss PD}. 2018. \emph{Clostridium difficile} alters the
@@
 \leavevmode\hypertarget{ref-Staley2016}{}%
-40. \textbf{Staley C}, \textbf{Weingarden AR}, \textbf{Khoruts A},
+41. \textbf{Staley C}, \textbf{Weingarden AR}, \textbf{Khoruts A},
 \textbf{Sadowsky MJ}. 2016. Interaction of gut microbiota with bile acid
@@
 \leavevmode\hypertarget{ref-Long2017}{}%
-41. \textbf{Long SL}, \textbf{Gahan CGM}, \textbf{Joyce SA}. 2017.
+42. \textbf{Long SL}, \textbf{Gahan CGM}, \textbf{Joyce SA}. 2017.
 Interactions between gut bacteria and bile in health and disease.
@@
 \leavevmode\hypertarget{ref-Sorg2010}{}%
-42. \textbf{Sorg JA}, \textbf{Sonenshein AL}. 2010. Inhibiting the
+43. \textbf{Sorg JA}, \textbf{Sonenshein AL}. 2010. Inhibiting the
 initiation of \emph{Clostridium difficile} spore germination using
@@
 \leavevmode\hypertarget{ref-Dubois2019}{}%
-43. \textbf{Dubois T}, \textbf{Tremblay YDN}, \textbf{Hamiot A},
+44. \textbf{Dubois T}, \textbf{Tremblay YDN}, \textbf{Hamiot A},
 \textbf{Martin-Verstraete I}, \textbf{Deschamps J}, \textbf{Monot M},
@@
 \leavevmode\hypertarget{ref-Ng2013}{}%
-44. \textbf{Ng KM}, \textbf{Ferreyra JA}, \textbf{Higginbottom SK},
+45. \textbf{Ng KM}, \textbf{Ferreyra JA}, \textbf{Higginbottom SK},
 \textbf{Lynch JB}, \textbf{Kashyap PC}, \textbf{Gopinath S},
@@
 \leavevmode\hypertarget{ref-Ferreyra2014}{}%
-45. \textbf{Ferreyra JA}, \textbf{Wu KJ}, \textbf{Hryckowian AJ},
+46. \textbf{Ferreyra JA}, \textbf{Wu KJ}, \textbf{Hryckowian AJ},
 \textbf{Bouley DM}, \textbf{Weimer BC}, \textbf{Sonnenburg JL}. 2014.
@@
 \leavevmode\hypertarget{ref-MartinVerstraete2016}{}%
-46. \textbf{Martin-Verstraete I}, \textbf{Peltier J}, \textbf{Dupuy B}.
+47. \textbf{Martin-Verstraete I}, \textbf{Peltier J}, \textbf{Dupuy B}.
 2016. The regulatory networks that control \emph{Clostridium difficile}
@@
 \leavevmode\hypertarget{ref-Collins2015}{}%
-47. \textbf{Collins J}, \textbf{Auchtung JM}, \textbf{Schaefer L},
+48. \textbf{Collins J}, \textbf{Auchtung JM}, \textbf{Schaefer L},
 \textbf{Eaton KA}, \textbf{Britton RA}. 2015. Humanized microbiota mice
@@
 \leavevmode\hypertarget{ref-Lawley2012}{}%
-48. \textbf{Lawley TD}, \textbf{Clare S}, \textbf{Walker AW},
+49. \textbf{Lawley TD}, \textbf{Clare S}, \textbf{Walker AW},
 \textbf{Stares MD}, \textbf{Connor TR}, \textbf{Raisen C},
@@
 \leavevmode\hypertarget{ref-Reeves2011}{}%
-49. \textbf{Reeves AE}, \textbf{Theriot CM}, \textbf{Bergin IL},
+50. \textbf{Reeves AE}, \textbf{Theriot CM}, \textbf{Bergin IL},
 \textbf{Huffnagle GB}, \textbf{Schloss PD}, \textbf{Young VB}. 2011. The
@@
 \leavevmode\hypertarget{ref-Mabrok2011}{}%
-50. \textbf{Mabrok HB}, \textbf{Klopfleisch R}, \textbf{Ghanem KZ},
+51. \textbf{Mabrok HB}, \textbf{Klopfleisch R}, \textbf{Ghanem KZ},
 \textbf{Clavel T}, \textbf{Blaut M}, \textbf{Loh G}. 2011. Lignan
@@
 \leavevmode\hypertarget{ref-Kim2019}{}%
-51. \textbf{Kim CC}, \textbf{Healey GR}, \textbf{Kelly WJ},
+52. \textbf{Kim CC}, \textbf{Healey GR}, \textbf{Kelly WJ},
 \textbf{Patchett ML}, \textbf{Jordens Z}, \textbf{Tannock GW},
@@
 \leavevmode\hypertarget{ref-doPrado2021}{}%
-52. \textbf{Prado SBR do}, \textbf{Minguzzi BT}, \textbf{Hoffmann C},
+53. \textbf{Prado SBR do}, \textbf{Minguzzi BT}, \textbf{Hoffmann C},
 \textbf{Fabi JP}. 2021. Modulation of human gut microbiota by dietary
@@
 \leavevmode\hypertarget{ref-Muthuramalingam2019}{}%
-53. \textbf{Muthuramalingam K}, \textbf{Singh V}, \textbf{Choi C},
+54. \textbf{Muthuramalingam K}, \textbf{Singh V}, \textbf{Choi C},
 \textbf{Choi SI}, \textbf{Kim YM}, \textbf{Unno T}, \textbf{Cho M}.
@@
 \leavevmode\hypertarget{ref-Han2019}{}%
-54. \textbf{Han S-H}, \textbf{Yi J}, \textbf{Kim J-H}, \textbf{Lee S},
+55. \textbf{Han S-H}, \textbf{Yi J}, \textbf{Kim J-H}, \textbf{Lee S},
 \textbf{Moon H-W}. 2019. Composition of gut microbiota in patients with
@@
 \leavevmode\hypertarget{ref-Duncan2004}{}%
-55. \textbf{Duncan SH}, \textbf{Louis P}, \textbf{Flint HJ}. 2004.
+56. \textbf{Duncan SH}, \textbf{Louis P}, \textbf{Flint HJ}. 2004.
 Lactate-utilizing bacteria, isolated from human feces, that produce
@@
 \leavevmode\hypertarget{ref-Ye2018}{}%
-56. \textbf{Ye J}, \textbf{Lv L}, \textbf{Wu W}, \textbf{Li Y},
+57. \textbf{Ye J}, \textbf{Lv L}, \textbf{Wu W}, \textbf{Li Y},
 \textbf{Shi D}, \textbf{Fang D}, \textbf{Guo F}, \textbf{Jiang H},
@@
 \leavevmode\hypertarget{ref-Walsh2017}{}%
-57. \textbf{Walsh CJ}, \textbf{Guinane CM}, \textbf{O'Toole PW},
+58. \textbf{Walsh CJ}, \textbf{Guinane CM}, \textbf{O'Toole PW},
 \textbf{Cotter PD}. 2017. A profile hidden markov model to investigate
@@
 \leavevmode\hypertarget{ref-Sandiford2018}{}%
-58. \textbf{Sandiford SK}. 2018. Current developments in lantibiotic
+59. \textbf{Sandiford SK}. 2018. Current developments in lantibiotic
 discovery for treating \emph{Clostridium difficile} infection. Expert
@@
 \leavevmode\hypertarget{ref-Stein2013}{}%
-59. \textbf{Stein RR}, \textbf{Bucci V}, \textbf{Toussaint NC},
+60. \textbf{Stein RR}, \textbf{Bucci V}, \textbf{Toussaint NC},
 \textbf{Buffie CG}, \textbf{Rätsch G}, \textbf{Pamer EG}, \textbf{Sander
@@
 \leavevmode\hypertarget{ref-Nakashima2021}{}%
-60. \textbf{Nakashima T}, \textbf{Fujii K}, \textbf{Seki T},
+61. \textbf{Nakashima T}, \textbf{Fujii K}, \textbf{Seki T},
 \textbf{Aoyama M}, \textbf{Azuma A}, \textbf{Kawasome H}. 2021. Novel
@@
 \leavevmode\hypertarget{ref-Geerlings2018}{}%
-61. \textbf{Geerlings S}, \textbf{Kostopoulos I}, \textbf{Vos W de},
+62. \textbf{Geerlings S}, \textbf{Kostopoulos I}, \textbf{Vos W de},
 \textbf{Belzer C}. 2018. \emph{Akkermansia muciniphila} in the human
@@
 \leavevmode\hypertarget{ref-Deng2018}{}%
-62. \textbf{Deng H}, \textbf{Yang S}, \textbf{Zhang Y}, \textbf{Qian K},
+63. \textbf{Deng H}, \textbf{Yang S}, \textbf{Zhang Y}, \textbf{Qian K},
 \textbf{Zhang Z}, \textbf{Liu Y}, \textbf{Wang Y}, \textbf{Bai Y},
@@
 \leavevmode\hypertarget{ref-Engevik2020}{}%
-63. \textbf{Engevik MA}, \textbf{Engevik AC}, \textbf{Engevik KA},
+64. \textbf{Engevik MA}, \textbf{Engevik AC}, \textbf{Engevik KA},
 \textbf{Auchtung JM}, \textbf{Chang-Graham AL}, \textbf{Ruan W},
@@
 \leavevmode\hypertarget{ref-Reeves2012}{}%
-64. \textbf{Reeves AE}, \textbf{Koenigsknecht MJ}, \textbf{Bergin IL},
+65. \textbf{Reeves AE}, \textbf{Koenigsknecht MJ}, \textbf{Bergin IL},
 \textbf{Young VB}. 2012. Suppression of \emph{Clostridium difficile} in
@@
 \leavevmode\hypertarget{ref-Ma2021}{}%
-65. \textbf{Ma L}, \textbf{Keng J}, \textbf{Cheng M}, \textbf{Pan H},
+66. \textbf{Ma L}, \textbf{Keng J}, \textbf{Cheng M}, \textbf{Pan H},
 \textbf{Feng B}, \textbf{Hu Y}, \textbf{Feng T}, \textbf{Yang F}. 2021.
@@
 \leavevmode\hypertarget{ref-Haas2020}{}%
-66. \textbf{Haas KN}, \textbf{Blanchard JL}. 2020. Reclassification of
+67. \textbf{Haas KN}, \textbf{Blanchard JL}. 2020. Reclassification of
 the \emph{Clostridium clostridioforme} and \emph{Clostridium sphenoides}
@@
 \leavevmode\hypertarget{ref-Finegold2005}{}%
-67. \textbf{Finegold SM}, \textbf{Song Y}, \textbf{Liu C}, \textbf{Hecht
+68. \textbf{Finegold SM}, \textbf{Song Y}, \textbf{Liu C}, \textbf{Hecht
 DW}, \textbf{Summanen P}, \textbf{Könönen E}, \textbf{Allen SD}. 2005.
@@
 \leavevmode\hypertarget{ref-VanInsberghe2020}{}%
-68. \textbf{VanInsberghe D}, \textbf{Elsherbini JA}, \textbf{Varian B},
+69. \textbf{VanInsberghe D}, \textbf{Elsherbini JA}, \textbf{Varian B},
 \textbf{Poutahidis T}, \textbf{Erdman S}, \textbf{Polz MF}. 2020.
@@
 \leavevmode\hypertarget{ref-GarzaGonzalez2019}{}%
-69. \textbf{Garza-González E}, \textbf{Mendoza-Olazarán S},
+70. \textbf{Garza-González E}, \textbf{Mendoza-Olazarán S},
 \textbf{Morfin-Otero R}, \textbf{Ramírez-Fontes A},
@@
 \leavevmode\hypertarget{ref-Shafiq2020}{}%
-70. \textbf{Shafiq M}, \textbf{Alturkmani H}, \textbf{Zafar Y},
+71. \textbf{Shafiq M}, \textbf{Alturkmani H}, \textbf{Zafar Y},
 \textbf{Mittal V}, \textbf{Lodhi H}, \textbf{Ullah W}, \textbf{Brewer
@@
 \leavevmode\hypertarget{ref-Keith2020}{}%
-71. \textbf{Keith JW}, \textbf{Dong Q}, \textbf{Sorbara MT},
+72. \textbf{Keith JW}, \textbf{Dong Q}, \textbf{Sorbara MT},
 \textbf{Becattini S}, \textbf{Sia JK}, \textbf{Gjonbalaj M},
@@
 \leavevmode\hypertarget{ref-Zackular2016}{}%
-72. \textbf{Zackular JP}, \textbf{Moore JL}, \textbf{Jordan AT},
+73. \textbf{Zackular JP}, \textbf{Moore JL}, \textbf{Jordan AT},
 \textbf{Juttukonda LJ}, \textbf{Noto MJ}, \textbf{Nicholson MR},
@@
 \leavevmode\hypertarget{ref-Berkell2021}{}%
-73. \textbf{Berkell M}, \textbf{Mysara M}, \textbf{Xavier BB},
+74. \textbf{Berkell M}, \textbf{Mysara M}, \textbf{Xavier BB},
 \textbf{Werkhoven CH van}, \textbf{Monsieurs P}, \textbf{Lammens C},
@@
 \leavevmode\hypertarget{ref-Gardiner2014}{}%
-74. \textbf{Gardiner BJ}, \textbf{Tai AY}, \textbf{Kotsanas D},
+75. \textbf{Gardiner BJ}, \textbf{Tai AY}, \textbf{Kotsanas D},
 \textbf{Francis MJ}, \textbf{Roberts SA}, \textbf{Ballard SA},
@@
 \leavevmode\hypertarget{ref-Iljazovic2020}{}%
-75. \textbf{Iljazovic A}, \textbf{Roy U}, \textbf{Gálvez EJC},
+76. \textbf{Iljazovic A}, \textbf{Roy U}, \textbf{Gálvez EJC},
 \textbf{Lesker TR}, \textbf{Zhao B}, \textbf{Gronow A}, \textbf{Amend
@@
 \leavevmode\hypertarget{ref-Nagalingam2013}{}%
-76. \textbf{Nagalingam NA}, \textbf{Robinson CJ}, \textbf{Bergin IL},
+77. \textbf{Nagalingam NA}, \textbf{Robinson CJ}, \textbf{Bergin IL},
 \textbf{Eaton KA}, \textbf{Huffnagle GB}, \textbf{Young VB}. 2013. The
@@
 \leavevmode\hypertarget{ref-AbernathyClose2021}{}%
-77. \textbf{Abernathy-Close L}, \textbf{Barron MR}, \textbf{George JM},
+78. \textbf{Abernathy-Close L}, \textbf{Barron MR}, \textbf{George JM},
 \textbf{Dieterle MG}, \textbf{Vendrov KC}, \textbf{Bergin IL},
@@
 \leavevmode\hypertarget{ref-Pirofski2008}{}%
-78. \textbf{Pirofski L-a}, \textbf{Casadevall A}. 2008. The
+79. \textbf{Pirofski L-a}, \textbf{Casadevall A}. 2008. The
 damage-response framework of microbial pathogenesis and infectious
@@
 \leavevmode\hypertarget{ref-Casadevall2014}{}%
-79. \textbf{Casadevall A}, \textbf{Pirofski L-a}. 2014. What is a host?
+80. \textbf{Casadevall A}, \textbf{Pirofski L-a}. 2014. What is a host?
 Incorporating the microbiota into the damage-response framework.
@@
 \leavevmode\hypertarget{ref-Frisbee2019}{}%
-80. \textbf{Frisbee AL}, \textbf{Saleh MM}, \textbf{Young MK},
+81. \textbf{Frisbee AL}, \textbf{Saleh MM}, \textbf{Young MK},
 \textbf{Leslie JL}, \textbf{Simpson ME}, \textbf{Abhyankar MM},
@@
 doi:\href{https://doi.org/10.1038/s41467-019-10733-9}{10.1038/s41467-019-10733-9}.
-
-\leavevmode\hypertarget{ref-NagaoKitamoto2020}{}%
-81. \textbf{Nagao-Kitamoto H}, \textbf{Leslie JL}, \textbf{Kitamoto S},
-\textbf{Jin C}, \textbf{Thomsson KA}, \textbf{Gillilland MG},
-\textbf{Kuffa P}, \textbf{Goto Y}, \textbf{Jenq RR}, \textbf{Ishii C},
-\textbf{Hirayama A}, \textbf{Seekatz AM}, \textbf{Martens EC},
-\textbf{Eaton KA}, \textbf{Kao JY}, \textbf{Fukuda S}, \textbf{Higgins
-PDR}, \textbf{Karlsson NG}, \textbf{Young VB}, \textbf{Kamada N}. 2020.
-Interleukin-22-mediated host glycosylation prevents \emph{Clostridioides
-difficile} infection by modulating the metabolic activity of the gut
-microbiota. Nature Medicine \textbf{26}:608--617.
-doi:\href{https://doi.org/10.1038/s41591-020-0764-0}{10.1038/s41591-020-0764-0}.
 
@@
 with a letter (N for non-moribund mice and M for moribund mice) and
-number (ordered by mean histopathologic score). + indicates the mice
-which did not have detectable \emph{C. difficile} CFU in Figure 2. (B)
-Median (points) and interquartile range (lines) of \(\beta\)-diversity
-(\(\theta\)\textsubscript{YC}) between an individual mouse and either
-all others which were inoculated with feces from the same donor or from
-a different donor. The \(\beta\)-diversity among the same donor
-comparison group was significantly less than the \(\beta\)-diversity of
-the different donor group (\emph{P} \textless{} 0.05, calculated by
-Wilcoxon rank sum test).
+number (ordered by mean histopathologic score of the donor group). +
+indicates the mice which did not have detectable \emph{C. difficile} CFU
+(Figure 2). (B) Median (points) and interquartile range (lines) of
+\(\beta\)-diversity (\(\theta\)\textsubscript{YC}) between an individual
+mouse and either all others which were inoculated with feces from the
+same donor or from a different donor. The \(\beta\)-diversity among the
+same donor comparison group was significantly less than the
+\(\beta\)-diversity of the different donor group (\emph{P} \textless{}
+0.05, calculated by Wilcoxon rank sum test).
 
@@
 non-moribund mice and M for moribund mice) and number (ordered by mean
-histopathologic score). Points are colored by donor group. Mice from
-donor groups N1 through N6 succumbed to the infection prior to day 10
-and were not plated on day 10 post-challenge. LOD = Limit of detection.
+histopathologic score of the donor group). Points are colored by donor
+group. Mice from donor groups N1 through N6 succumbed to the infection
+prior to day 10 and were not plated on day 10 post-challenge. LOD =
+Limit of detection.
 
@@
 labeled by the donor letter (N for non-moribund mice and M for moribund
-mice) and number (ordered by mean histopathologic score). Points are
-colored by donor group. Mice in group N1 that have a summary score of 0
-are the mice which did not have detectable \emph{C. difficile} CFU in
-Figure 2. Missing points are from mice that had insufficient fecal
-sample collected for assaying toxin or cecum for histopathologic
-scoring.
+mice) and number (ordered by mean histopathologic score of the donor
+group). Points are colored by donor group. Mice in group N1 that have a
+summary score of 0 are the mice which did not have detectable \emph{C.
+difficile} CFU (Figure 2). Missing points are from mice that had
+insufficient fecal sample collected for assaying toxin or cecum for
+histopathologic scoring. LOD = Limit of detection.
 
@@
 challenge (Day 0) that varied significantly by the moribundity and
-histopathologic summary score or toxin activity by LEfSe analysis.
+histopathologic summary score or detected toxin by LEfSe analysis.
 Median (points) and interquartile range (lines) are plotted. Genera are
@@
 \textbf{Figure 5. Fecal bacterial community members of the murine gut at
-the time of \emph{C. difficile} (Day 0) predicted outcomes of the
-infection.} Day 0 bacterial community members grouped by different
-classification rank were modeled with random forest to predict the
-infection outcome. The models used the highest taxonomic classification
-rank that performed as well as the lower ranks. Median (solid points)
-and interquartile range (lines) of the group relative abundance are
-plotted. Bacterial groups are ordered by their importance to the model;
-taxonomic group at the top of the plot had the greatest decrease in
-performance when its relative abundances were permuted. * indicates that
-the bacterial group was unclassified at lower taxonomic classification
-ranks. (A) Bacterial members grouped by phyla predicted which mice would
-have toxin activity detected at any point throughout the infection
-(Toxin +, dark purple). (B) Bacterial members grouped by class predicted
-which mice would become moribund (dark blue). (C) Bacterial members
-grouped by genera predicted if the mice would have a high (score greater
-than the median score of 5, colored dark green) or low (score less than
-the median score of 5, colored light green) histopathologic summary
-score.
+the time of \emph{C. difficile} infection predicted outcomes of the
+infection.} On the day of infection (Day 0), bacterial community members
+grouped by different classification rank were modeled with random forest
+to predict the infection outcome. The models used the highest taxonomic
+classification rank that performed as well as the lower ranks. Median
+(solid points) and interquartile range (lines) of the group relative
+abundance are plotted. Bacterial groups are ordered by their importance
+to the model; taxonomic group at the top of the plot had the greatest
+decrease in performance when its relative abundances were permuted. *
+indicates that the bacterial group was unclassified at lower taxonomic
+classification ranks. (A) Bacterial members grouped by phyla predicted
+which mice would have toxin activity detected at any point throughout
+the infection (Toxin +, dark purple). (B) Bacterial members grouped by
+class predicted which mice would become moribund (dark blue). (C)
+Bacterial members grouped by genera predicted if the mice would have a
+high (score greater than the median score of 5, colored dark green) or
+low (score less than the median score of 5, colored light green)
+histopathologic summary score.
 
@@
 (score less than the median score of 5) histopathologic summary score.
-(D) Bacterial phyla which affected the performance of predicting toxin
-activity when permuted. (E) Bacterial classes which affected the
-performance of predicting moribundity when permuted. (D) Bacterial
-genera which affected the performance of predicting histopathologic
-score when permuted.
+(D) Bacterial phyla which affected the performance of predicting
+detectable toxin activity when permuted. (E) Bacterial classes which
+affected the performance of predicting moribundity when permuted. (D)
+Bacterial genera which affected the performance of predicting
+histopathologic score when permuted.
 

### 2022-01-24 submission/manuscript.Rmd
add ILB feedback

@@
 (Stockbridge, MI) embedded the samples in paraffin, sectioned, and created the hematoxylin and eosin-stained slides.
-The slides were scored using previously described criteria by a board-certified veterinary pathologist scored who was blinded to the experimental groups [@Theriot2011].
+The slides were scored using previously described criteria by a board-certified veterinary pathologist who was blinded to the experimental groups [@Theriot2011].
+Slides were scored as 0-4 for parameters of epithelial damage, tissue edema, and inflammation and a summary score of 0-12 was generated by summing the three individual parameter scores.
 

### 2022-01-24 submission/manuscript.tex
add ILB feedback

@@
 eosin-stained slides. The slides were scored using previously described
-criteria by a board-certified veterinary pathologist scored who was
-blinded to the experimental groups (29).
+criteria by a board-certified veterinary pathologist who was blinded to
+the experimental groups (29). Slides were scored as 0-4 for parameters
+of epithelial damage, tissue edema, and inflammation and a summary score
+of 0-12 was generated by summing the three individual parameter scores.
 

### 2022-01-26 exploratory/notebook/initial_analysis/initial_analysis_notebook.Rmd

@@
 knitr::opts_chunk$set(warning = FALSE)
-#knitr::opts_knit$set(root.dir = '../../..')
 ```
@@
 library(tidyverse)
+library(here)
 
 # calculate error rate
-error_df <- read.table('../../../data/mothur/sample.error.count', header = T)
+error_df <- read.table(here('data/mothur/sample.error.count'), header = T)
 total_seqs <- sum(error_df$Sequences)
@@
 
-read_tsv('../../../data/mothur/sample.final.shared',
+read_tsv(here('data/mothur/sample.final.shared'),
          col_types = cols(.default = col_double(),
@@
 # 
-source('../../../code/Sum_OTU_by_Tax.R')
+source(here('code/Sum_OTU_by_Tax.R'))
 
 # load in data
-metadata <- read_tsv('../../../data/process/metadata_tidy.tsv',
+metadata <- read_tsv(here('data/process/metadata_tidy.tsv'),
                      col_type = 'cccccDddddDcdddddcdl') 
-toxin <- read_tsv('../../../data/process/toxin_tidy.tsv',
+toxin <- read_tsv(here('data/process/toxin_tidy.tsv'),
                   col_type = 'cccccdc')
-histology <- read_tsv('../../../data/process/histology_tidy.tsv',
+histology <- read_tsv(here('data/process/histology_tidy.tsv'),
                       col_type = 'ccccdddcdcc')
-human_source <- read_tsv('../../../data/process/human_source_tidy.tsv',
+human_source <- read_tsv(here('data/process/human_source_tidy.tsv'),
                          col_type = cols(.default = col_character(), age = col_double()))
-shared <- read_tsv('../../../data/mothur/sample.final.0.03.subsample.shared',
+shared <- read_tsv(here('data/mothur/sample.final.0.03.subsample.shared'),
          col_types = cols(.default = col_double(),
@@
   pivot_longer(-Group, names_to = 'otu', values_to = 'counts') %>% 
-  filter(counts > 0) %>% 
   mutate(relative_abundance = counts/2107 * 100)
-taxonomy <- read_tsv('../../../data/process/final.taxonomy.tidy.tsv',
+taxonomy <- read_tsv(here('data/process/final.taxonomy.tidy.tsv'),
                      col_types = cols(.default = col_character()))
-alpha_df <- read_tsv('../../../data/mothur/sample.final.groups.ave-std.summary',
+alpha_df <- read_tsv(here('data/mothur/sample.final.groups.ave-std.summary'),
          col_types ='dccdddddd') %>% 
@@
                         gsub('D', '_D', group)))
-nmds_df <- read_tsv('../../../data/mothur/sample.final.thetayc.0.03.lt.ave.nmds.axes',
+nmds_df <- read_tsv(here('data/mothur/sample.final.thetayc.0.03.lt.ave.nmds.axes'),
          col_types ='cdd') %>% 
@@
                         gsub('D', '_D', group)))
-source('../../../code/read_dist.R')
-beta_df <- read_dist('../../../data/mothur/sample.final.thetayc.0.03.lt.ave.dist') %>% 
+source(here('code/read_dist.R'))
+beta_df <- read_dist(here('data/mothur/sample.final.thetayc.0.03.lt.ave.dist')) %>% 
       mutate(rows = ifelse(grepl('_D|DA', rows), rows,
@@
                         gsub('D', '_D', columns)))
-donor_colors <- c("#41bbc5", "#7e2640", "#9ad859", "#682dbd", "#728e24", "#e26df8", "#23980d", "#fd048f", "#46e33e", "#fa1bfc", "#155126", "#f9b2d1", "#154975", "#e2c627", "#6294ce")
+donor_colors <- c("#41bbc5", "#7e2640", "#9ad859", "#682dbd", "#728e24", 
+                  "#e26df8", "#23980d", "#fd048f", "#46e33e", "#fa1bfc", 
+                  "#155126", "#f9b2d1", "#154975", "#e2c627", "#6294ce")
 ```
@@
 donor_plot <- shared %>% 
-  filter(grepl('DA', Group)) %>% 
+  filter(counts > 0,
+         grepl('DA', Group)) %>% 
   left_join(taxonomy, by = c('otu' = 'OTU')) %>% 
@@
   select(group, human_source, mouse_id) %>% 
-  left_join(shared, by = c('group' = 'Group')) %>% 
+  left_join(filter(shared, counts > 0), by = c('group' = 'Group')) %>% 
   left_join(taxonomy, by = c('otu' = 'OTU')) %>% 
@@
   group_by(taxa_level) %>% 
-  mutate(median_ra = median(relative_abundance)) %>% 
   filter(!is.na(taxa_level)) %>% 
+  mutate(median_ra = mean(relative_abundance),
+         human_source = as.character(as.numeric(gsub('DA', '', human_source))),
+         human_source = factor(human_source, levels =
+                                 c('10093', '10148', '431', '884', '1245',
+                                   '581', '1324', '953', '1134',
+                                   '369', '430', '10027', '10034',
+                                   '10082', '578')),
+         taxa_level = gsub('_unclassified', '', taxa_level),
+         taxa_level = gsub('_', ' ', taxa_level),
+         taxa_level = paste0('*', taxa_level, '*')) %>% 
   ggplot(aes(x = group, 
@@
           axis.text.x = element_blank(),
-          legend.position = 'bottom')
-
-cowplot::plot_grid(
-  cowplot::plot_grid(NULL, donor_plot, nrow = 1, rel_widths = c(.95,13)),
-  day_0_plot, ncol = 1, rel_heights = c(8,9))
-```
+          legend.position = 'bottom',
+          axis.text.y = ggtext::element_markdown(angle = 45),
+          strip.text.x = element_text(angle = 45),
+          strip.background = element_blank(),
+          legend.margin=margin(t=-0.1, r=0, b=-0.1, l=0, unit="cm"),
+          legend.key.height = unit(0.3, 'cm'))
+
+#cowplot::plot_grid(
+#  cowplot::plot_grid(NULL, donor_plot, nrow = 1, rel_widths = c(.95,13)),
+#  day_0_plot, ncol = 1, rel_heights = c(8,9))
+day_0_plot
+
+```
+
+Unsure about alpha plots
 
@@
 alpha_df %>% 
-  ggplot(aes(x = sample, y = invsimpson, color = human_source)) + 
+  filter(day == 0) %>% 
+  #kruskal.test(x = .$invsimpson, g = as.factor(.$human_source)) # p = 0.002
+  ggplot(aes(x = human_source, y = invsimpson, color = human_source)) + 
     geom_point(alpha = 0.3) + 
     theme_bw() + 
-    facet_wrap(human_source~.) + 
-    guides(color = 'none')
+    #facet_wrap(human_source~.) + 
+    guides(color = 'none') +  
+    ylim(0,NA)
 
 alpha_df %>% 
-  ggplot(aes(x = sample, y = sobs, color = human_source)) + 
+  filter(day == 0) %>% 
+  #kruskal.test(x = .$sobs, g = as.factor(.$human_source)) # p = 9.9e-5
+  ggplot(aes(x = human_source, y = sobs, color = human_source)) + 
     geom_point(alpha = 0.3) + 
     theme_bw() + 
-    facet_wrap(human_source~.) + 
-    guides(color = 'none')
-
-nmds_df %>% 
+    #facet_wrap(human_source~.) + 
+    guides(color = 'none') +
+    ylim(0,NA)
+
+# how many otus are unique to each donor and what portion of community?
+otu_dist <- alpha_df %>% 
+  filter(day == 0) %>% 
+  select(group, human_source) %>% 
+  left_join(shared, by = c('group' = 'Group')) %>% 
+  filter(relative_abundance > 0.1) %>% 
+  select(human_source, otu) 
+# 353 otus
+#otu_dist %>% 
+#  pull(otu) %>% 
+#  unique %>% length 
+otu_dist %>% 
+  distinct %>% 
+  group_by(otu) %>% 
+  mutate(otu_present_n_samples = length(human_source)) %>% 
+  left_join(select(taxonomy, tax_otu_label, OTU), by = c('otu' = 'OTU')) %>% 
+  left_join(shared, by = c('otu')) %>% 
+  ggplot(aes(x = as.factor(otu_present_n_samples), y = relative_abundance)) + 
+    geom_violin() +
+    scale_y_log10() + 
+    facet_wrap(human_source~.)
+
+```
+
+```{r nmds, eval = F, include = F}
+initial_nmds_plot <- nmds_df %>% 
   left_join(filter(select(metadata, group, human_source, day), day %in% c(0, 10)), by = c('group')) %>% 
@@
          sample_type = ifelse(grepl('DA', group), 'human', 'murine')) %>% 
-  filter(sample_type == 'human' | day == 0) %>% 
+  filter(sample_type != 'human', day == 0) %>% 
   ggplot(aes(x = axis1, y = axis2, color = human_source)) +
@@
     scale_color_manual(values = donor_colors) + 
-    facet_grid(~sample_type) + 
+    #facet_grid(~sample_type) + 
     theme_bw() + 
@@
     labs(x = 'NMDS axis 1', y = 'NMDS axis 2')
+
+# plot for presentation
+#ggsave('~/Desktop/intial_nmds.jpg',
+#       initial_nmds_plot,
+#       height = 4, width = 4)
+
+initial_nmds_plot
+
 ```
@@
          grepl('DA', columns)) %>% 
-  mutate(comparison = 'Donor\nvs\nDonor')
+  mutate(comparison = 'Donor vs Donor')
 
@@
       inner_join(beta_df, by = c('group' = 'columns', 'human_source' = 'rows'))) %>% 
-  mutate(comparison = 'Donor\nvs\nRecipient')
+  mutate(comparison = 'Donor vs Recipient Initial')
 
@@
       inner_join(beta_df, by = c('group' = 'columns', 'human_source' = 'rows'))) %>% 
-  mutate(comparison = 'Donor\nvs\nRecipient End')
+  mutate(comparison = 'Donor vs Recipient End')
 
@@
   inner_join(beta_df, by = c('group.x' = 'rows', 'group.y' = 'columns')) %>% 
-  mutate(comparison = 'Recipients\ncompared to\nothers w/same donor')
+  mutate(comparison = 'Recipients compared to\nothers w/same donor')
 
@@
   anti_join(donor_v_recipent, by = c('rows' = 'human_source', 'columns' = 'group')) %>% 
-  mutate(comparison = 'Donor\nvs\nNon-recipients')
+  mutate(comparison = 'Donor vs Non-recipients')
 
@@
   anti_join(corecipient, by = c('rows' = 'group.x', 'columns' = 'group.y')) %>% 
-  mutate(comparison = 'Recipients\ncompared to\nothers w/different donor')
+  mutate(comparison = 'Recipients compared to\nothers w/different donor')
 
@@
          (dayA == 'D10' & dayB == 'D0')) %>% 
-  mutate(comparison = 'Day 0\nvs\nDay 10')
-
-bind_rows(donor_v_donor, donor_v_recipent, corecipient, other_donors, 
+  mutate(comparison = 'Initial vs End')
+
+beta_plot <- bind_rows(donor_v_donor, donor_v_recipent, corecipient, other_donors, 
           other_recipients, day0_v_day10, donor_v_recipent_end) %>% 
   mutate(comparison = factor(comparison, levels = 
-                              c('Donor\nvs\nDonor',
-                                'Donor\nvs\nRecipient', 
-                                'Recipients\ncompared to\nothers w/same donor', 
-                                'Day 0\nvs\nDay 10',
-                                'Donor\nvs\nRecipient End',
-                                'Donor\nvs\nNon-recipients', 
-                                'Recipients\ncompared to\nothers w/different donor'))) %>% 
+                              c('Donor vs Donor',
+                                'Donor vs Non-recipients', 
+                                'Recipients compared to\nothers w/different donor',
+                                'Recipients compared to\nothers w/same donor', 
+                                'Donor vs Recipient End',
+                                'Initial vs End',
+                                'Donor vs Recipient Initial'))) %>% 
   ggplot(aes(x = comparison, y = distances)) + 
@@
     labs(x = NULL, y = 'Theta YC') +
+    theme_bw() + 
+    coord_flip()
+
+refined_beta_plot <- bind_rows(corecipient, other_recipients) %>% 
+  mutate(comparison = ifelse(comparison == 'Recipients compared to\nothers w/different donor',
+                             'Different donor', 'Same donor'),
+    comparison = factor(comparison, levels = 
+                              c('Same donor',
+                                'Different donor'))) %>% 
+  ggplot(aes(x = comparison, y = distances)) + 
+    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5)) +
+    #geom_jitter(alpha = 0.1, width = 0.3) + 
+    labs(x = NULL, y = 'Similarity') +
+    scale_y_continuous(breaks = c(0, 1), limits = c(0,1),
+                       labels = c('Complete\nsimilarity', 'No similarity')) + 
     theme_bw()
+#wilcox.test(corecipient$distances, other_recipients$distances)
+# presentation plot
+#ggsave('~/Desktop/beta_plot.jpg',
+#       refined_beta_plot,
+#       width = 3, height = 5)
 
@@
     guides(color = 'none')
+
+ggsave(here('results/figures/Figure_1.jpg'),
+  cowplot::plot_grid(day_0_plot + theme(text=element_text(size = 9)), 
+                     NULL,
+                     beta_plot + theme(text=element_text(size = 9)), 
+                     ncol = 1, 
+                     rel_heights = c(6, 0.3, 3),
+                     labels = c('A', 'B', NULL)),
+  height = 9.0625, width = 6.875, unit = 'in')
 ```
@@
 
-```{r}
+```{r, include = T, eval = F}
 metadata %>% 
@@
 ```{r}
-metadata %>% 
+cfu_plot <- metadata %>% 
   filter(cdiff_strain == 431,
          day %in% c(1,10)) %>% 
-  mutate(human_source = factor(human_source, levels = 
-                                c('DA01324', 'DA00953', 'DA00581', 'DA01134', 'DA10148', 
-                                  'DA00369', 'DA00430', 'DA10093', 'DA10027', 'DA10034',
-                                  'DA10082', 'DA00884', 'DA00578', 'DA00431', 'DA01245'))) %>% 
+  mutate(human_source = as.numeric(gsub('DA', '', human_source)),
+         human_source = factor(human_source, levels = 
+                                c('1324', '953', '581', '1134', '10148', 
+                                  '369', '430', '10093', '10027', '10034',
+                                  '10082', '884', '578', '431', '1245')),
+         day = factor(paste('Day', day), levels = c('Day 1', 'Day 10'))) %>% 
   ggplot(aes(x = human_source, y = log10(cdiff_cfu + 60), color = human_source)) +
@@
     guides(color = 'none')
+
+ggsave(here('results/figures/Figure_2.jpg'),
+       cfu_plot,
+       width = 6, height = 4)
+
+# presentation plot
+#presentation_cfu_plot <- metadata %>% 
+#  filter(cdiff_strain == 431,
+#         day %in% c(1,10)) %>% 
+#  mutate(human_source = as.numeric(gsub('DA', '', human_source)),
+#         human_source = factor(human_source, levels = 
+#                                c('1324', '953', '581', '1134', '10148', 
+#                                  '369', '430', '10093', '10027', '10034',
+#                                  '10082', '884', '578', '431', '1245'),
+#                               labels = LETTERS[1:15]),
+#         day = factor(paste('Day', day), levels = c('Day 1', 'Day 10'))) %>% 
+#  ggplot(aes(x = human_source, y = cdiff_cfu, color = human_source)) +
+#    geom_jitter(width = 0.2) + 
+#    facet_grid(day~.) + 
+#    scale_color_manual(values = donor_colors) + 
+#    scale_y_continuous(limits = c(1.5,8.5),
+#                       breaks = c(2,4,6,8),
+#                       labels = c('10^2', '10^4', '10^6', '10^8')) + 
+#    theme_bw() + 
+#    labs(x = 'Donor Source', y = '*C. difficile* CFU') + 
+#    guides(color = 'none') + 
+#    theme(axis.text.y = ggtext::element_markdown(),
+#          axis.title.y = ggtext::element_markdown(),
+#          strip.text.y = element_text(angle = 0, size = 12),
+#          strip.background = element_blank(),
+#          panel.spacing = unit(2, "lines")) + 
+#			geom_hline(yintercept = 2, linetype = 'dashed', size = .5, color = 'white') + 
+#			geom_label(x = 14, y = 2, label = "LOD", fill = 'white', color = 'white') + 
+#			geom_text(x = 14, y = 2, label = "LOD", color = 'lightgray', size = 12/.pt)
+#ggsave('~/Desktop/cfu_plot.jpg',
+#       presentation_cfu_plot,
+#       width = 8, height = 5)
+#
 ```
@@
 
-```{r}
+```{r, include = F, eval = F}
 metadata %>% 
@@
 ```{r}
-metadata %>% 
+wt_loss_plot <- metadata %>% 
   filter(cdiff_strain == 431,
          day %in% c(2,10)) %>% 
-  mutate(human_source = factor(human_source, levels = 
-                                c('DA01324', 'DA00953', 'DA00581', 'DA01134', 'DA10148', 
-                                  'DA00369', 'DA00430', 'DA10093', 'DA10027', 'DA10034',
-                                  'DA10082', 'DA00884', 'DA00578', 'DA00431', 'DA01245'))) %>% 
+  mutate(human_source = as.numeric(gsub('DA', '', human_source)),
+         human_source = factor(human_source, levels = 
+                                c('1324', '953', '581', '1134', '10148', 
+                                  '369', '430', '10093', '10027', '10034',
+                                  '10082', '884', '578', '431', '1245')),
+         day = factor(paste('Day', day), levels = c('Day 2', 'Day 10'))) %>% 
   ggplot(aes(x = human_source, y = percent_weightLoss, color = human_source)) +
@@
     theme_bw() + 
-    labs(x = 'Donor Source', y = 'Weight (Relative to initial)') + 
+    labs(x = 'Donor Source', y = 'Weight (% relative to initial)') + 
     guides(color = 'none')
+
+wt_loss_plot
+
+ggsave(here('results/figures/Figure_S1.jpg'),
+       wt_loss_plot,
+       width = 6, height = 4)
 ```
@@
 
+ What is toxin distribution by outcome?
+
 ```{r histology}
-mouse_by_outcome <- histology %>% 
-  mutate(mouse_id = paste0(cage_id, '_', ear_tag)) %>% 
-  select(mouse_id, human_source, epithelial_damage, summary_score) %>% 
-  left_join(unique(select(metadata, mouse_id, early_euth, cdiff_strain),
-            by = c('mouse_id'))) %>% 
-  filter(cdiff_strain == 431) %>% 
-  group_by(human_source) %>% 
-  mutate(severity = mean(epithelial_damage) < 1) %>% 
-# using epithelial damage
-  mutate(outcome = case_when(early_euth == T ~ 'moribund',
-                             severity == T ~ 'mild',
-                             severity == F ~ 'severe',
-                             T ~ 'NA')) %>% 
-# using summary score
-#    mutate(outcome_ss = case_when(early_euth == T ~ 'moribund',
-#                             summary_score < 4.4 ~ 'mild', # mean = 4.4, median = 5
-#                             summary_score > 4.4 ~ 'severe',
+#mouse_by_outcome <- histology %>% 
+#  mutate(mouse_id = paste0(cage_id, '_', ear_tag)) %>% 
+#  select(mouse_id, human_source, epithelial_damage, summary_score) %>% 
+#  left_join(unique(select(metadata, mouse_id, early_euth, cdiff_strain),
+#            by = c('mouse_id'))) %>% 
+#  filter(cdiff_strain == 431) %>% 
+#  group_by(human_source) %>% 
+#  mutate(severity = mean(epithelial_damage) < 1) %>% 
+## using epithelial damage
+#  mutate(outcome = case_when(early_euth == T ~ 'moribund',
+#                             severity == T ~ 'mild',
+#                             severity == F ~ 'severe',
 #                             T ~ 'NA')) %>% 
-  mutate(outcome = factor(outcome, levels = c('mild', 'severe', 'moribund'))) %>% 
-  select(mouse_id, outcome)
-
-histology %>% 
+## using summary score
+##    mutate(outcome_ss = case_when(early_euth == T ~ 'moribund',
+##                             summary_score < 4.4 ~ 'mild', # mean = 4.4, median = 5
+##                             summary_score > 4.4 ~ 'severe',
+##                             T ~ 'NA')) %>% 
+#  mutate(outcome = factor(outcome, levels = c('mild', 'severe', 'moribund'))) %>% 
+#  ungroup %>% 
+#  select(mouse_id, outcome)
+# if not using histology to split non-severe into mild/moderate, no need for mouse_by_outcome
+
+histology_plot <- histology %>% 
   left_join(distinct(select(metadata, mouse_id, early_euth)), by = 'mouse_id') %>% 
   filter(!is.na(early_euth)) %>% 
-  mutate(early_euth = ifelse(early_euth, 'Moribund', 'Colonized'),
+  mutate(early_euth = ifelse(early_euth, 'Severe', 'Mild'),
          mouse_id = paste0(cage_id, '_', ear_tag)) %>% 
-  inner_join(mouse_by_outcome, by = c('mouse_id')) %>%
+  #left_join(mouse_by_outcome, by = c('mouse_id')) %>%
   pivot_longer(cols = c(edema_tissue, inflammation_tissue, epithelial_damage, summary_score),
                names_to = 'histology', values_to = 'score') %>% 
-  ggplot(aes(x = early_euth, y = score, , color = early_euth, fill = early_euth)) +
+  filter(histology == 'summary_score') %>% 
+  mutate(histology = ifelse(histology == 'summary_score', 'End point', NA),
+         human_source = gsub('DA0{0,2}', '', human_source)) %>% 
+  ggplot(aes(x = human_source, y = score, , color = early_euth, fill = early_euth)) +
     #stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5), position = position_dodge(width = 0.4)) +
-    geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 0.5) + 
+    geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 2) + 
     #geom_point(position = position_jitterdodge(jitter.width = 0.3), alpha = 0.5) + 
     theme_bw() + 
-    labs(x = NULL, y = 'Histology Score', fill = 'Outcome') +
+    labs(x = NULL, y = 'Clinical\nScore', fill = 'Outcome') +
     scale_x_discrete(guide = guide_axis(angle = 45)) + 
-    facet_grid(.~histology) + 
+    scale_y_continuous(breaks = c(1,3,5,7,9), 
+                       labels = c(1,3,5,7,9)) + 
+    facet_grid(histology~early_euth, scales = 'free', space = 'free_x') + 
     guides(color = 'none') + 
-    theme(legend.position = 'bottom')
+    theme(legend.position = 'bottom',
+          strip.background.x = element_blank(),
+          strip.text.x = element_blank(),
+          legend.margin=margin(t=-0.3, r=0, b=-0.2, l=0, unit="cm"),
+          legend.key.height = unit(0, 'cm'))
 
@@
 
-```
-
- What is toxin distribution by outcome?
-
-```{r toxin}
-toxin %>% 
-  inner_join(select(metadata, group, human_source), by = c('group')) %>% 
-  left_join(mouse_by_outcome, by = c('human_source')) %>% 
+toxin_plot <- toxin %>% 
+  inner_join(select(metadata, group, human_source, early_euth), by = c('group')) %>% 
+#  left_join(mouse_by_outcome, by = c('mouse_id')) %>% 
+  mutate(outcome = ifelse(early_euth == T, 'Severe', 'Mild')) %>% 
   filter(!is.na(outcome)) %>%
   mutate(day = as.numeric(toxin_sample_day),
-         outcome = ifelse(outcome == 'moribund', as.character(outcome), 'colonized')) %>% 
-  ggplot(aes(x = outcome, y = Log_repiricoal_dilution, fill = outcome, color = outcome)) + 
-    stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5), position = position_dodge(width = 0.4)) +
-    geom_jitter(width= 0.3, height = 0, alpha = 0.3) + 
-    labs(x = 'Day', y = 'Toxin activity') + 
-    facet_grid(.~day) + 
-    theme_bw() +
-    theme(legend.position = 'bottom')
+         human_source = gsub('DA0{0,2}', '', human_source)) %>% 
+  filter(!day %in% c(4,7),
+         toxin_sample_type == 'stool') %>% 
+  ggplot(aes(x = human_source, y = Log_repiricoal_dilution, 
+             fill = outcome, color = outcome)) + 
+    #stat_summary(fun.data = 'median_hilow', fun.args = (conf.int=0.5), position = position_dodge(width = 0.4)) +
+    #geom_jitter(width= 0.3, height = 0, alpha = 0.3) + 
+    geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 2) + 
+    labs(x = NULL, y = 'Toxin activity\n ') + 
+    facet_grid(day~outcome, scales = 'free_x', space = 'free_x') + 
+    theme_bw() +
+    theme(legend.position = 'none',
+          axis.text.x = element_blank(),
+          axis.ticks.x = element_blank(),
+          panel.grid.minor.y = element_blank())
+
+toxin_plot
+histology_plot
+
+ggsave(here('results/figures/Figure_3.jpg'),
+  cowplot::plot_grid(toxin_plot, NULL, histology_plot, ncol = 1, 
+                   rel_heights = c(8, 1, 4),
+                   labels = c('A', 'B', NULL)),
+  width = 6, height = 4)
+
+# presentation plot
+#presentation_histology_plot <- histology %>% 
+#  left_join(distinct(select(metadata, mouse_id, early_euth)), by = 'mouse_id') %>% 
+#  filter(!is.na(early_euth)) %>% 
+#  mutate(early_euth = ifelse(early_euth, 'Severe', 'Non-severe'),
+#         mouse_id = paste0(cage_id, '_', ear_tag)) %>% 
+#  #left_join(mouse_by_outcome, by = c('mouse_id')) %>%
+#  pivot_longer(cols = c(edema_tissue, inflammation_tissue, epithelial_damage, summary_score),
+#               names_to = 'histology', values_to = 'score') %>% 
+#  filter(histology == 'summary_score') %>% 
+#  mutate(histology = ifelse(histology == 'summary_score', 'End point', NA),
+#         human_source = gsub('DA0{0,2}', '', human_source),
+#         human_source = factor(human_source, levels = 
+#                                 c('1324', '953', '581', '1134', '10148', 
+#                                   '369', '430', '10093', '10027', '10034',
+#                                   '10082', '884', '578', '431', '1245'),
+#                               labels = LETTERS[1:15])) %>% 
+#  ggplot(aes(x = human_source, y = score, fill = human_source, color = human_source)) +
+#    geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 1) + 
+#    theme_bw() + 
+#    labs(x = NULL, y = 'Clinical Score') +
+#    scale_y_continuous(breaks = c(1,3,5,7,9), 
+#                       labels = c(1,3,5,7,9)) + 
+#    scale_fill_manual(values = donor_colors) + 
+#    scale_color_manual(values = donor_colors) + 
+#    facet_grid(.~early_euth, scales = 'free', space = 'free_x') + 
+#    theme(legend.position = 'none',
+#          strip.background.x = element_blank(),
+#          strip.text.x = element_text(size = 12),
+#          legend.margin=margin(t=-0.3, r=0, b=-0.2, l=0, unit="cm"),
+#          legend.key.height = unit(0, 'cm'))
+#
+#ggsave('~/Desktop/histology_plot.jpg',
+#       presentation_histology_plot,
+#       width = 8, height = 3.5)
+#
+```
+
+```{r, toxin/hist pvalues}
+
+toxin %>% 
+  filter(toxin_sample_day < 3,
+         toxin_sample_type == 'stool') %>% 
+  left_join(
+    unique(select(metadata, mouse_id, early_euth)),
+    by = c('mouse_id')) %>% 
+  select(early_euth, Log_repiricoal_dilution) %>% 
+  filter(!is.na(early_euth)) %>% 
+  summarise(pvalue = 
+    wilcox.test(
+      pull(filter(., early_euth ==T), Log_repiricoal_dilution),
+      pull(filter(., early_euth ==F), Log_repiricoal_dilution),
+      alternative = 'greater')$p.value)
+
+histology %>% 
+  left_join(distinct(select(metadata, mouse_id, early_euth)), by = 'mouse_id') %>% 
+  filter(!is.na(early_euth)) %>% 
+  summarise(pvalue = 
+    wilcox.test(
+      pull(filter(., early_euth ==T), summary_score),
+      pull(filter(., early_euth ==F), summary_score),
+      alternative = 'greater')$p.value)
+
 ```
@@
 
-metadata %>% 
+cfu_toxin_corr <- metadata %>% 
   inner_join(toxin, by = 'group') %>% 
@@
 	       spearman_rho = cor.test(cfu, Log_repiricoal_dilution, 
-	                               method = 'spearman', exact = F)$estimate)  %>% 
+	                               method = 'spearman', exact = F)$estimate) 
+cfu_toxin_corr %>% 
   ggplot(aes(x = cfu, y = Log_repiricoal_dilution)) + 
@@
     geom_smooth(method = 'lm', se=F) + 
-    labs(x = 'C. difficile CFU (log 10)', y = 'Toxin (log reciprocal dilution)')
-
-metadata %>% 
-  filter(day == mouse_endpoint,
-         cdiff_strain == 431) %>% 
-  left_join(histology, by = c('mouse_id')) %>% 
-  mutate(cfu = log10(cdiff_cfu)) %>% 
-  pivot_longer(cols = c(edema_tissue, inflammation_tissue, epithelial_damage, summary_score),
-               names_to = 'hist_score', values_to = 'score') %>% 
-  group_by(hist_score) %>% 
-  nest() %>% 
-	mutate(spearman_p = map(.x = data, .f = ~ cor.test(.x$cfu, .x$score, 
-	                                                   method = 'spearman', exact = F)$p.value),
-	       spearman_rho = map(.x = data, .f = ~ cor.test(.x$cfu, .x$score, 
-	                                                     method = 'spearman', exact = F)$estimate)) %>%
-	unnest(c(spearman_p, spearman_rho)) %>% 
-	mutate(pvalue = p.adjust(spearman_p, method = 'BH', n = 5)) %>% 
-	unnest(data) %>% 
-  ggplot(aes(x = cfu, y = score)) + 
-    geom_point(alpha = 0.3) + 
-    theme_bw() + 
-    guides(color = 'none') + 
-    geom_smooth(method = 'lm', se=F) + 
-    labs(x = 'C. difficile CFU (log 10)', y = 'Histology Score') + 
-    facet_wrap(hist_score~., scale = 'free_y')
-
-```
-
-Correlations between CFU and histology score not significant after BH correction.
+    labs(x = 'C. difficile CFU (log 10)', y = 'Toxin (log reciprocal dilution)',
+         subtitle = paste0('rho = ', unique(cfu_toxin_corr$spearman_rho), 
+                           ', pvalue = ', unique(cfu_toxin_corr$spearman_p)))
+
+#hist_cfu_corr <- metadata %>% 
+#  filter(day == mouse_endpoint,
+#         cdiff_strain == 431) %>% 
+#  left_join(histology, by = c('mouse_id')) %>% 
+#  mutate(cfu = log10(cdiff_cfu),
+#         score = summary_score) %>% 
+#	mutate(spearman_p = cor.test(cfu, score, method = 'spearman', exact = F)$p.value,
+#	       spearman_rho = cor.test(cfu, score, method = 'spearman', exact = F)$estimate)
+#hist_cfu_corr %>% 
+#  ggplot(aes(x = cfu, y = score)) + 
+#    geom_point(alpha = 0.3) + 
+#    theme_bw() + 
+#    guides(color = 'none') + 
+#    geom_smooth(method = 'lm', se=F) + 
+#    labs(x = 'C. difficile CFU (log 10)', y = 'Histology Score',
+#         subtitle = paste0('rho = ', unique(hist_cfu_corr$spearman_rho), 
+#                           ', pvalue = ', unique(hist_cfu_corr$spearman_p)))
+# correlation of cfu and summary score, rho = 0.15, p value = 0.29
+
+```
+Expected correlation, since no cfu == no toxin, and cfu when greater popultion will produce more toxin
+
+Did not calculation correlation for toxin/clinical score since toxin only detected in 5 of 53 (with high clinica score) however within range of others with similar clinical score and no toxin and bias in number of samples w/o toxin
+Correlations between CFU and histology score not significant.
+
+Same day toxin vs Genus
 
@@
     labs(x = 'Relative Abundance', y = 'Toxin (log reciprocal dilution)')
-
-```
+```
+
+```{r significant_same_day_toxin, fig,height = 14}
+# use lefse instead
+#metadata %>% 
+#  inner_join(toxin, by = 'group') %>% 
+#  filter(cdiff_strain == 431) %>% 
+#  mutate(toxin = Log_repiricoal_dilution > 1) %>% 
+#  select(Group = group, toxin) %>% 
+#  left_join(shared, by = c('Group')) %>% 
+#  left_join(taxonomy, by = c('otu' = 'OTU')) %>% 
+#  filter(!is.na(otu)) %>% 
+#  group_by(Group, Genus, toxin) %>%
+#  summarise(counts = sum(counts),
+#            relative_abundance = sum(relative_abundance)) %>% 
+#  group_by(Genus) %>% 
+#  mutate(otu = Genus,
+#        sd = sd(counts)) %>% 
+#  filter(sd != 0) %>% 
+#  nest() %>% 
+#	mutate(wilcox_p = map(.x = data, .f = ~ wilcox.test(pull(filter(.x, toxin ==T), counts),
+#    pull(filter(.x, toxin ==F), counts))$p.value)) %>%
+#	unnest(wilcox_p) %>% 
+#	mutate(pvalue = p.adjust(wilcox_p, method = 'BH', n = nrow(.))) %>% # correct p values
+#	filter(pvalue < 0.05) %>% # select only those above 0.05 after pvalue correction
+#	unnest(data) %>% 
+#  group_by(Genus) %>% 
+#  mutate(Genus = gsub('_unclassified', '', Genus),
+#         median_ra = median(relative_abundance),
+#         relative_abundance = ifelse(relative_abundance == 0, 0.045, relative_abundance)) %>% 
+#  ggplot(aes(x = reorder(Genus, median_ra), y = relative_abundance, color = toxin)) + 
+#    geom_boxplot() + 
+#    coord_flip() + 
+#    scale_y_log10() + 
+#    theme_bw() + 
+#    labs(x = NULL, y = 'Relative Abundance')
+#
+```
+
+```{r significant_day_0_toxin, fig,height = 14}
+#Significant Day 0 OTUs for communities that do/dont produce toxin
+# use lefse instead
+#toxin %>% 
+#  group_by(mouse_id) %>% 
+#  summarise(toxin = max(Log_repiricoal_dilution) > 1) %>% 
+#  inner_join(filter(metadata, cdiff_strain == 431, day == 0), 
+#             by = 'mouse_id') %>% 
+#  select(Group = group, toxin) %>% 
+#  left_join(shared, by = c('Group')) %>% 
+#  left_join(taxonomy, by = c('otu' = 'OTU')) %>% 
+#  filter(!is.na(otu)) %>% 
+#  group_by(Group, Genus, toxin) %>%
+#  summarise(counts = sum(counts),
+#            relative_abundance = sum(relative_abundance)) %>% 
+#  group_by(Genus) %>% 
+#  mutate(otu = Genus,
+#        sd = sd(counts)) %>% 
+#  filter(sd != 0) %>% 
+#  nest() %>% 
+#	mutate(wilcox_p = map(.x = data, .f = ~ wilcox.test(pull(filter(.x, toxin ==T), counts),
+#    pull(filter(.x, toxin ==F), counts))$p.value)) %>%
+#	unnest(wilcox_p) %>% 
+#	mutate(pvalue = p.adjust(wilcox_p, method = 'BH', n = nrow(.))) %>% # correct p values
+#	filter(pvalue < 0.05) %>% # select only those above 0.05 after pvalue correction
+#	unnest(data) %>% 
+#  group_by(Genus) %>% 
+#  mutate(Genus = gsub('_unclassified', '', Genus),
+#         median_ra = median(relative_abundance),
+#         relative_abundance = ifelse(relative_abundance == 0, 0.045, relative_abundance)) %>% 
+#  ggplot(aes(x = reorder(Genus, median_ra), y = relative_abundance, color = toxin)) + 
+#    geom_boxplot() + 
+#    coord_flip() + 
+#    scale_y_log10() + 
+#    theme_bw() + 
+#    labs(x = NULL, y = 'Relative Abundance')
+
+```
+
+Correlation of day 0 otus with day 10 histology scores
 
@@
 # correlation of histology to otu
-plot_hist_corr <- function(i){
 metadata %>% 
+  filter(cdiff_strain == 431,
+         day == 0) %>% 
+  inner_join(histology, by = 'mouse_id') %>% 
+  filter(early_euth == F) %>% 
+  select(Group = group, test = summary_score) %>% 
+  left_join(shared, by = c('Group')) %>% 
+  group_by(otu) %>%
+  mutate(sd = sd(counts)) %>% 
+  filter(!is.na(otu),
+         sd != 0) %>% 
+  nest() %>% 
+  mutate(spearman_p = map(.x = data, .f = ~ cor.test(.x$counts, .x$test, 
+                                                     method = 'spearman', exact = F)$p.value),
+         spearman_rho = map(.x = data, .f = ~ cor.test(.x$counts, .x$test, 
+                                                       method = 'spearman', exact = F)$estimate)) %>% # compare cleared vs colonized
+  unnest(c(spearman_p, spearman_rho)) %>% 
+  mutate(pvalue = p.adjust(spearman_p, method = 'BH', n = nrow(.))) %>% # correct p values
+  filter(pvalue < 1) %>% # select only those above 0.05 after pvalue correction
+  unnest(data) %>% 
+  left_join(taxonomy, by = c('otu' = 'OTU')) %>% 
+  mutate(tax_otu_label = gsub('_unclassified', '', tax_otu_label),
+         tax_otu_label = paste0(tax_otu_label, '\nrho = ', round(spearman_rho, 3), 
+                                ', *P* = ', round(pvalue, 3))) %>%  
+  ggplot(aes(x = relative_abundance + 0.001, y = test, color = tax_otu_label)) + 
+    geom_point(alpha = 0.3) +
+    geom_smooth(method = 'lm', se = F) +
+    scale_x_log10() + 
+    facet_wrap(tax_otu_label ~ ., scales = 'free_x') + 
+    theme_bw() + 
+    labs(x = 'Relative Abundance', y = 'Clinical Score') + 
+      guides(color = 'none')
+
+```
+
+Correlation of day 10 otus with histology scores
+
+```{r}
+# correlation of histology to otu
+day_10_hist_genus_corr_plot <- metadata %>% 
   filter(cdiff_strain == 431,
@@
   filter(early_euth == F) %>% 
-    select(Group = group, test = .data[[i]]) %>% 
+  select(Group = group, test = summary_score) %>% 
     left_join(shared, by = c('Group')) %>% 
@@
     mutate(tax_otu_label = gsub(' \\(', '\n\\(', tax_otu_label)) %>% 
-    ggplot(aes(x = relative_abundance, y = test, color = tax_otu_label)) + 
+    ggplot(aes(x = relative_abundance + 0.001, y = test, color = tax_otu_label)) + 
       geom_point(alpha = 0.3) +
       geom_smooth(method = 'lm', se = F) +
-      facet_wrap(tax_otu_label ~ .) + 
-      theme_bw() + 
-      labs(x = 'Relative Abundance', y = i, title = i) + 
+      scale_x_log10() + 
+      facet_wrap(tax_otu_label ~ ., scales = 'free_x') + 
+      theme_bw() + 
+      labs(x = 'Relative Abundance', y = 'Clinical Score') + 
       guides(color = 'none')
-}
-
-plot_hist_corr('summary_score')
-plot_hist_corr('edema_tissue')
-plot_hist_corr('inflammation_tissue')
-plot_hist_corr('epithelial_damage')
-```
-```{r}
-# correlation of histology to otu
-plot_hist_corr <- function(i){
-  metadata %>% 
-    filter(cdiff_strain == 431,
-           day == mouse_endpoint) %>% 
-    inner_join(histology, by = 'mouse_id') %>% 
-    filter(early_euth == F) %>% 
-    select(Group = group, test = .data[[i]]) %>% 
-    left_join(shared, by = c('Group')) %>% 
-    group_by(otu) %>%
-    mutate(sd = sd(counts)) %>% 
-    filter(!is.na(otu),
-           sd != 0) %>% 
-    nest() %>% 
-  	mutate(spearman_p = map(.x = data, .f = ~ cor.test(.x$counts, .x$test, 
-  	                                                   method = 'spearman', exact = F)$p.value),
-  	       spearman_rho = map(.x = data, .f = ~ cor.test(.x$counts, .x$test, 
-  	                                                     method = 'spearman', exact = F)$estimate)) %>% # compare cleared vs colonized
-  	unnest(c(spearman_p, spearman_rho)) %>% 
-  	mutate(pvalue = p.adjust(spearman_p, method = 'BH', n = nrow(.))) %>% # correct p values
-  	filter(pvalue < 0.05) %>% # select only those above 0.05 after pvalue correction
-  	unnest(data) %>% 
-    left_join(taxonomy, by = c('otu' = 'OTU')) %>% 
-    mutate(tax_otu_label = gsub(' \\(', '\n\\(', tax_otu_label)) %>% 
-    ggplot(aes(x = relative_abundance, y = test, color = tax_otu_label)) + 
-      geom_point(alpha = 0.3) +
-      geom_smooth(method = 'lm', se = F) +
-      facet_wrap(tax_otu_label ~ .) + 
-      theme_bw() + 
-      labs(x = 'Relative Abundance', y = i, title = i) + 
-      guides(color = 'none')
-}
-
-plot_hist_corr('summary_score')
-plot_hist_corr('edema_tissue')
-plot_hist_corr('inflammation_tissue')
-plot_hist_corr('epithelial_damage')
-```
-
-```{r}
-
-metadata <- read_tsv('../../../data/process/metadata_tidy.tsv',
+
+day_10_hist_genus_corr_plot
+```
+```
+
+### LEfSe
+
+```{r, fig.height = 9, fig.width = 6}
+# Figures should be no more than 6.87 inches wide (approximately 17.46 cm or 41.25 picas wide) and 9.06 inches high (approximately 23.01 cm or 54.37 picas high)
+
+metadata <- read_tsv(here('data/process/metadata_tidy.tsv'),
 		col_type = 'cccccDddddDcdddddcdl') 
-shared <- read_tsv('../../../data/mothur/sample.final.0.03.subsample.shared',
+shared <- read_tsv(here('data/mothur/sample.final.0.03.subsample.shared'),
 		col_types = cols(.default = col_double(),
@@
 		gsub('D', '_D', Group)))
-taxonomy <- read_tsv('../../../data/process/final.taxonomy.tidy.tsv',
+taxonomy <- read_tsv(here('data/process/final.taxonomy.tidy.tsv'),
                      col_types = cols(.default = col_character()))
 
-plot_lefse_data <- function(file_name){
-    lefse_design <- data.table::fread(paste0('../../../data/process/', file_name, '_Genus.design'))
+lefse_design <- data.table::fread(here('data/process/lefse/toxin_severity_Genus.design'))
     colnames(lefse_design) <- c('Group', 'class')
 
-    lefse_df <- data.table::fread(paste0('../../../data/process/', file_name, '_Genus.0.03.lefse_summary'),
+lefse_df <- data.table::fread(here('data/process/lefse/toxin_severity_Genus.0.03.lefse_summary'),
   	fill = T) %>% 
@@
   top_n(20, LDA) %>% 
-  	mutate(order = LDA + 10 * as.logical(Class),
+  mutate(fctr_class = factor(Class),
+    order = LDA + 10 * as.numeric(fctr_class),
+  	tax_otu_label = OTU)
+
+#shared_lefse <- shared %>% 
+#  gather(OTU, abundance, matches('Otu\\d*')) %>% 
+#	left_join(taxonomy, by = 'OTU') %>% 
+#	right_join(lefse_df, by = c('Genus' = 'OTU')) %>% 
+#  right_join(lefse_design, by = c('Group')) %>% 
+#	group_by(Group, Genus, order, class) %>% # removed mouse_id from interactive plot
+#	summarise(abundance = sum(abundance)) %>% 
+#	mutate(tax_otu_label = Genus) %>% 
+#	mutate(abundance = abundance / 21.07,
+#		abundance = ifelse(abundance == 0, 0.045, abundance)) %>% 
+#  #highlight_key(~mouse_id) %>% 
+#	ggplot(aes(x = reorder(tax_otu_label, order), 
+#			y = abundance, color = class)) + # , shape = mouse_id remove mouse_id from interactive plot
+#		geom_hline(yintercept = 0.047, linetype = 'dashed', lwd = 0.3) + 
+#    stat_summary(fun.data = 'median_hilow', aes(group = class),
+#			fun.args = (conf.int=0.5), position = position_dodge(width = .8)) +
+#		theme_bw() + 
+#    labs(x = NULL, y = 'Relative Abundance', color = NULL) + 
+#		scale_y_log10(breaks=c(0.01, 0.1, 1, 10, 100),
+#			labels=c("0","0.1","1","10","100")) +
+#		#scale_color_manual(values = c('#8de4d3','#f75b5f')) + 
+#    theme(panel.grid.major.y = element_blank(),
+#          panel.grid.minor.y = element_line(color = 'black'),
+#			legend.position = 'top') + 
+#		coord_flip() + 
+#  geom_jitter(position = position_jitterdodge(dodge.width = .8, jitter.width = 0.1),
+#              alpha = 0.1)
+#
+#cfu_lefse <- metadata %>% 
+#  left_join(mutate(lefse_design, mouse_id = gsub('D0', '', Group)), 
+#      by = c('mouse_id')) %>% 
+#  mutate(cdiff_cfu = ifelse(cdiff_cfu == 0, log10(60), log10(cdiff_cfu)),
+#         xaxis = 'C. difficile') %>% 
+#	ggplot(aes(x = xaxis, y = cdiff_cfu, color = class)) + 
+#		geom_hline(yintercept = log10(100), linetype = 'dashed', lwd = 0.3) + 
+#    stat_summary(fun.data = 'median_hilow', aes(group = class),
+#			fun.args = (conf.int=0.5), position = position_dodge(width = 1)) +
+#		theme_bw() + 
+#    labs(x = NULL, y = 'CFU', color = NULL) + 
+#		#scale_color_manual(values = c('#8de4d3','#f75b5f')) + 
+#    theme(panel.grid.major.y = element_blank(), 
+#			panel.grid.minor = element_blank(),
+#			legend.position = 'none') + 
+#		coord_flip() + 
+#  geom_jitter(position = position_jitterdodge(dodge.width = 1, jitter.width = 0.2), 
+#              alpha = 0.1)
+#
+#cowplot::plot_grid(shared_lefse, 
+#                   cowplot::plot_grid(NULL, cfu_lefse, nrow = 1, rel_widths = c(1.2, 8)),
+#                   ncol = 1, rel_heights = c(9,1))
+#
+
+#library(plotly)
+#gg <- ggplotly(abundance_plot + geom_jitter(position = position_jitterdodge(dodge.width = 0.5)), tooltip = c('mouse_id'))
+#highlight(hide_legend(gg), on = "plotly_click", off = "plotly_doubleclick", 
+#           color = "black", selectize = T, opacityDim = .2,
+#           selected = attrs_selected(showlegend = F, marker = list(symbol = 'X')))
+
+```
+
+```{r, fig.height = 9}
+
+plot_lefse <- function(file_name){
+  lefse_design <- data.table::fread(here(paste0('data/process/lefse/', file_name, '.design')))
+      colnames(lefse_design) <- c('Group', 'class')
+
+  lefse_df <- data.table::fread(here(paste0('data/process/lefse/', file_name, '.0.03.lefse_summary')),
+    	fill = T) %>% 
+    filter(!is.na(LDA)) %>% 
+    top_n(20, LDA) %>% 
+    mutate(fctr_class = factor(Class),
+      order = LDA + 10 * as.numeric(fctr_class),
     	tax_otu_label = OTU)
@@
     #highlight_key(~mouse_id) %>% 
-  	ggplot(aes(x = reorder(tax_otu_label, order), 
+  	ggplot(aes(x = tax_otu_label, 
   			y = abundance, color = class)) + # , shape = mouse_id remove mouse_id from interactive plot
@@
       stat_summary(fun.data = 'median_hilow', aes(group = class),
-  			fun.args = (conf.int=0.5), position = position_dodge(width = 0.5)) +
-  		theme_bw() + 
-      labs(x = NULL, y = 'Relative Abundance', color = NULL) + 
+  			fun.args = (conf.int=0.5), position = position_dodge(width = .8)) +
+  		theme_bw() + 
+      labs(x = NULL, y = 'Relative Abundance', color = NULL, title = file_name) + 
   		scale_y_log10(breaks=c(0.01, 0.1, 1, 10, 100),
@@
       theme(panel.grid.major.y = element_blank(),
-  			panel.grid.minor = element_blank(),
-  			legend.position = 'bottom') + 
+            panel.grid.minor.y = element_line(color = 'black'),
+  			legend.position = 'top') + 
   		coord_flip() + 
-      geom_jitter(position = position_jitterdodge(dodge.width = 0.5), alpha = 0.2)
+    geom_jitter(position = position_jitterdodge(dodge.width = .8, jitter.width = 0.1),
+                alpha = 0.1)
 }
-
-#library(plotly)
-#gg <- ggplotly(abundance_plot + geom_jitter(position = position_jitterdodge(dodge.width = 0.5)), tooltip = c('mouse_id'))
-#highlight(hide_legend(gg), on = "plotly_click", off = "plotly_doubleclick", 
-#           color = "black", selectize = T, opacityDim = .2,
-#           selected = attrs_selected(showlegend = F, marker = list(symbol = 'X')))
-
-```
-## Severe disease
-```{r}
-plot_lefse_data('severe_disease')
-```
-
-## Toxin presence
-```{r}
-plot_lefse_data('toxin_presence')
-```
-
-## Toxin presence for mild outcome
-```{r}
-plot_lefse_data('toxin_mild')
-```
-
-## Toxin presence for severe outcome
-```{r fig.height = 4}
-plot_lefse_data('toxin_moribund')
-```
-
-## Toxin presence and outcome severity
-```{r, fig.height = 12}
-plot_lefse_data('toxin_severity')
-```
-
-## Summary score
-```{r}
-plot_lefse_data('summary_score_hilo')
-```
-
-## Epithelial damage
-```{r}
-plot_lefse_data('epithelial_damage_presence')
-```
+```
+
+#### day 0 prediction of future
+
+```{r, fig.height = 9}
+plot_lefse('day_0_toxin_production_Genus')
+plot_lefse('severe_disease_Genus')
+#plot_lefse('toxin_severity_Genus') # decieving because communities/cages get split like (ie the one mouse that was moribund w/o toxin)
+
+plot_lefse('summary_score_hilo_day0_Genus')
+plot_lefse('day_0_severity_Genus')
+```
+
+#### same day
+
+```{r, fig.height = 9}
+plot_lefse('toxin_presence_Genus')
+plot_lefse('summary_score_hilo_day10_Genus')
+
+day_0_toxin_production_plot <- plot_lefse('day_0_toxin_production_Genus')
+day_0_severity_Genus_plot <- plot_lefse('day_0_severity_Genus')
+
+ggsave(here('results/figures/Figure_4.jpg'),
+  cowplot::plot_grid(day_0_severity_Genus_plot + theme(text=element_text(size = 9)),
+                     cowplot::plot_grid(day_0_toxin_production_plot + theme(text=element_text(size = 9)),
+                                        day_10_hist_genus_corr_plot + theme(text=element_text(size = 9)),
+                                        ncol = 1, rel_heights = c(3,2),
+                                        labels = c('B', 'C')),
+                     nrow = 1, 
+                     labels = c('A', NULL)),
+  height = 7, width = 6.875, unit = 'in')
+
+```
+
+```{r presentation lefse plot}
+#  lefse_design <- data.table::fread(here(paste0('data/process/lefse/day_0_severity_Genus.design')))
+#      colnames(lefse_design) <- c('Group', 'class')
+#    
+#  lefse_df <- data.table::fread(here(paste0('data/process/lefse/day_0_severity_Genus.0.03.lefse_summary')),
+#    	fill = T) %>% 
+#    filter(!is.na(LDA)) %>% 
+#    top_n(10, LDA) %>% 
+#    mutate(fctr_class = factor(Class),
+#      order = LDA + 10 * as.numeric(fctr_class),
+#    	tax_otu_label = OTU)
+#    
+#lefse_presentation_plot <- shared %>% 
+#    gather(OTU, abundance, matches('Otu\\d*')) %>% 
+#  	left_join(taxonomy, by = 'OTU') %>% 
+#  	right_join(lefse_df, by = c('Genus' = 'OTU')) %>% 
+#    right_join(lefse_design, by = c('Group')) %>% 
+#  	group_by(Group, Genus, order, class) %>% # removed mouse_id from interactive plot
+#  	summarise(abundance = sum(abundance)) %>% 
+#  	mutate(tax_otu_label = Genus) %>% 
+#  	mutate(tax_otu_label = gsub('_', ' ', tax_otu_label),
+#  	       tax_otu_label = paste0('*', tax_otu_label, '*'),
+#  	       abundance = abundance / 21.07,
+#  		     abundance = ifelse(abundance == 0, 0.035, abundance),
+#  		     class = case_when(class == 'mild' ~ 'Low Clinical Score',
+#  		                         class == 'moderate' ~ 'High Clinical Score',
+#  		                         class == 'severe' ~ 'Severe'),
+#  		     class = factor(class, levels = c('Low Clinical Score', 'High Clinical Score', 'Severe'))) %>% 
+#  	ggplot(aes(x = tax_otu_label, y = abundance, color = class)) +
+#      stat_summary(fun.data = 'median_hilow', aes(group = class),
+#  			fun.args = (conf.int=0.5), position = position_dodge(width = .4)) +
+#  		geom_hline(yintercept = 0.047, linetype = 'dashed', lwd = 0.3) + 
+#  		theme_bw() + 
+#      labs(x = NULL, y = 'Relative Abundance', color = NULL) + 
+#  		scale_y_log10(breaks=c(0.01, 0.1, 1, 10, 100),
+#  			labels=c("0","0.1","1","10","100")) +
+#	    scale_x_discrete(guide = guide_axis(angle = 45)) + 
+#  		scale_color_manual(values = c("#62ebc9", "#399283", "#0c1b37")) + 
+#      theme(panel.grid.minor.y = element_blank(),
+#  			    legend.position = 'top',
+#            axis.text.x = ggtext::element_markdown())
+#
+#ggsave('~/Desktop/lefse_presentation_plot.jpg',
+#       lefse_presentation_plot,
+#       width = 9, height = 5)
+```
+
+What is the genus Clostridioides made up of?
+Only stool?
+toxin + disease mild = limited access to epithelial
+
+Microbe-microbe Interactions during Clostridioides difficile Infection
+  Arwa Abbas and Joseph P. Zackular. Curr Opin Microbiol. 2020 Feb; 53: 19–25. doi: 10.1016/j.mib.2020.01.016
+
+    C.
+difficile infection (CDI) can cause a spectrum of disease from mild diarrhea to severe complications such as pseudomembranous colitis, toxic megacolon and death (reviewed in [3]).
+This broad spectrum of C.
+difficile-associated disease may be explained in part by bacterial genetic factors such as variation in the pathogenicity locus [4] and increased accessory gene content [5,6].
+There is also likely a strong role for exogenous factors such as host genetics, comorbidities, treatment modalities and previous drug exposures.
+Here we focus on how resident microbiota can manipulate pathogen behavior and virulence.
+        competition for essential nutrients, limiting access to mucosal surfaces, direct production of antimicrobial molecules, modulating the intestinal metabolome, and activating the host immune system against the pathogen of interest
+    Exposure to low concentrations of deoxycholate, one of the most abundant cecal bile acids [39,40], reduced toxin production by most strains, without a concomitant reduction in general vegetative cell growth
+    B. thetatiotaomicron cross-feeds sialic acid to C. difficile
+    succinate appears necessary for C. difficile expansion in the gut
+    inhibit via production of bile acids, antibiotics, acidification
+    lacto-conditioned media reduce toxin and ability to adhere to host cells
+    Enterobacteriaceae and Enterococcus are known to thrive during intestinal inflammation and during CDI 
+    Cdiff induces indole production via E. coli which inhibits anaerobes
+    Cdiff ferments tyrosine to p-cresol
+    C. difficile can also secrete proline-based cyclic dipeptides that can inhibit gut bacteria, including commensal Clostridium species
+
+microbiome/Cdifficile/host interactions
+  lactobacillus inhibits cdiff
+  biofilm can protect cdiff
+  microbiome can secrete inhibitory compounds - turbomycin, lantibiotics
+  microbiome can cross feed microbiome via sialic acid/succinate or even proline or AAs
+  cdiff can inhibit microbiome w/ indole or para-cresol

### 2022-01-26 submission/manuscript.Rmd

@@
 
-*Clostridioides difficile* colonization can result in no symptoms or develop into an infection, ranging in severity from mild diarrhea to toxic megacolon, sepsis, and death.
+*Clostridioides difficile* colonization can be asymptomatic or develop into an infection, ranging in severity from mild diarrhea to toxic megacolon, sepsis, and death.
 Models that predict severity and guide treatment decisions are based on clinical factors and *C.
 difficile* characteristics.
-Although the gut microbiome plays a role in protecting against CDI, it has its effect on CDI disease severity is unclear and has not been incorporated into disease severity models.
+Although the gut microbiome plays a role in protecting against CDI, its effect on CDI disease severity is unclear and has not been incorporated into disease severity models.
 We demonstrated that variation in the microbiome of mice colonized with human feces yielded a range of disease outcomes.
@@
 
-Recently, Nagao-Kitamoto *et al.* showed germ-free mice which received a microbial community from different human donors resulted in variation in a subsequent *C.
-difficile* infection [@NagaoKitamoto2020].
-Here, we sought to further elucidate the relationship between members of the gut bacterial community and CDI disease severity.
+Recent studies have demonstrated that when mice with diverse microbial communities were challenged with a high-toxigenic strain resulted in varied disease severity [@NagaoKitamoto2020] and when challenged with a low-toxigenic strain members of the gut microbial community associated with variation in colonization [@Tomkovich2020].
+Here, we sought to further elucidate the relationship between members of the gut bacterial community and CDI disease severity when challenged with a high-toxigenic strain, *C.
+difficile* ribotype 027 (RT027).
 We hypothesized that since specific groups of gut bacteria affect the metabolism of *C.
@@
 We then challenged the mice with *C.
-difficile* ribotype 027 (RT027) and followed the mice for the development of severe outcomes of moribundity and histopathologic cecal tissue damage.
+difficile* RT027 and followed the mice for the development of severe outcomes of moribundity and histopathologic cecal tissue damage.
 Since the murine host and *C.
@@
 ***C.
-difficile* is able to infect germ-free mice colonized with human fecal microbial communities without antibiotics**  To produce gut microbiomes with greater variation than those found in conventional mouse colonies, we colonized germ-free mice with bacteria from human feces [@Nagpal2018].
-We inoculated germ-free C57BL/6 mice with homogenized feces from one of 15 human fecal samples via oral gavage.
+difficile* is able to infect germ-free mice colonized with human fecal microbial communities without antibiotics.**  To produce gut microbiomes with greater variation than those found in conventional mouse colonies, we colonized germ-free mice with bacteria from human feces [@Nagpal2018].
+We inoculated germ-free C57BL/6 mice with homogenized feces from each of 15 human fecal samples via oral gavage.
 These human fecal samples were selected because they represented diverse community structures based on community clustering [@Schubert2014].
@@
 However, we wanted to avoid modifying the gut communities with an antibiotic to maintain their unique microbial compositions and ecological relationships.
-Since some of these communities came from people at increased risk of CDI, such as recent hospitalization or antibiotic use [@Schubert2014], we decided to test if *C.
-difficile* was able to infect these mice without any antibiotic perturbation.
+Since some of these communities came from people at increased risk of CDI, such as recent hospitalization or antibiotic use [@Schubert2014], we tested whether *C.
+difficile* was able to infect these mice without an antibiotic perturbation.
 We hypothesized that *C.
@@
 *C.
-difficile* is unable to colonize an unperturbed healthy gut community and is only able to become established after a perturbation [@Schubert2015].
+difficile* is unable to colonize an unperturbed healthy murine gut community and is only able to become established after a perturbation [@Schubert2015].
 Once colonized, the different communities lead to different metabolic responses and dynamics of the *C.
@@
 A myriad of studies have explored the relationship between the microbiome and CDI disease.
-CDI studies often use an infection model with a single homogeneous bacterial community.
-Collins *et al.* used multiple human communities to colonize mice, however the communities were pooled prior to gavaging into germ-free mice [@Collins2015], resulting in a single human-derived community.
 Studies examining difference in disease often use different *C.
@@
 difficile* cells or spores used for the challenge [@Chen2008; @Reeves2011].
-These studies either lack variation in the initial microbiome or have variation in the *C.
+These studies often either lack variation in the initial microbiome or have variation in the *C.
 difficile* infection itself, confounding any association between variation in severity and the microbiome.
+Recent studies have shown variation in the initial microbiome, via different murine colonies or colonizing germ-free mice with human feces, that were challenged with *C.
+difficile* resulted in varied outcomes of the infection [@Tomkovich2020; @NagaoKitamoto2020].
 
-Recently, Nagao-Kitamoto *et al.* demonstrated when only allowing the microbiome to vary in a murine *C.
-difficile* infection experiemnt that the resultant disease variation associated with the microbiome communities [@NagaoKitamoto2020].
 Our data have demonstrated gut bacterial relative abundances associate with variation in toxin production, histopathologic scoring of the cecal tissue and mortality.

### 2022-01-26 submission/manuscript.tex

@@
 
-\emph{Clostridioides difficile} colonization can result in no symptoms
-or develop into an infection, ranging in severity from mild diarrhea to
+\emph{Clostridioides difficile} colonization can be asymptomatic or
+develop into an infection, ranging in severity from mild diarrhea to
 toxic megacolon, sepsis, and death. Models that predict severity and
@@
 difficile} characteristics. Although the gut microbiome plays a role in
-protecting against CDI, it has its effect on CDI disease severity is
-unclear and has not been incorporated into disease severity models. We
+protecting against CDI, its effect on CDI disease severity is unclear
+and has not been incorporated into disease severity models. We
 demonstrated that variation in the microbiome of mice colonized with
@@
 
-Recently, Nagao-Kitamoto \emph{et al.} showed germ-free mice which
-received a microbial community from different human donors resulted in
-variation in a subsequent \emph{C. difficile} infection (14). Here, we
-sought to further elucidate the relationship between members of the gut
-bacterial community and CDI disease severity. We hypothesized that since
-specific groups of gut bacteria affect the metabolism of \emph{C.
-difficile} and its infection dynamics, we can also identify groups of
-bacteria that affect the disease severity of the infection. To test this
-hypothesis, we colonized germ-free C57BL/6 mice with human fecal samples
-to create varied gut communities. We then challenged the mice with
-\emph{C. difficile} ribotype 027 (RT027) and followed the mice for the
-development of severe outcomes of moribundity and histopathologic cecal
-tissue damage. Since the murine host and \emph{C. difficile} isolate
-were the same and only the gut community varied, the variation in
-disease severity we observed was attributable to the gut microbiome.
+Recent studies have demonstrated that when mice with diverse microbial
+communities were challenged with a high-toxigenic strain resulted in
+varied disease severity (14) and when challenged with a low-toxigenic
+strain members of the gut microbial community associated with variation
+in colonization (15). Here, we sought to further elucidate the
+relationship between members of the gut bacterial community and CDI
+disease severity when challenged with a high-toxigenic strain, \emph{C.
+difficile} ribotype 027 (RT027). We hypothesized that since specific
+groups of gut bacteria affect the metabolism of \emph{C. difficile} and
+its infection dynamics, we can also identify groups of bacteria that
+affect the disease severity of the infection. To test this hypothesis,
+we colonized germ-free C57BL/6 mice with human fecal samples to create
+varied gut communities. We then challenged the mice with \emph{C.
+difficile} RT027 and followed the mice for the development of severe
+outcomes of moribundity and histopathologic cecal tissue damage. Since
+the murine host and \emph{C. difficile} isolate were the same and only
+the gut community varied, the variation in disease severity we observed
+was attributable to the gut microbiome.
 
@@
 \textbf{\emph{C. difficile} is able to infect germ-free mice colonized
-with human fecal microbial communities without antibiotics} To produce
+with human fecal microbial communities without antibiotics.} To produce
 gut microbiomes with greater variation than those found in conventional
 mouse colonies, we colonized germ-free mice with bacteria from human
-feces (15). We inoculated germ-free C57BL/6 mice with homogenized feces
-from one of 15 human fecal samples via oral gavage. These human fecal
+feces (16). We inoculated germ-free C57BL/6 mice with homogenized feces
+from each of 15 human fecal samples via oral gavage. These human fecal
 samples were selected because they represented diverse community
-structures based on community clustering (16). The gut communities were
-allowed to equilibrate for two weeks post-inoculation (17). We then
+structures based on community clustering (17). The gut communities were
+allowed to equilibrate for two weeks post-inoculation (18). We then
 surveyed the bacterial members of the gut communities by 16S rRNA gene
@@
 with antibiotics, such as clindamycin, to become susceptible to \emph{C.
-difficile} colonization (18, 19). However, we wanted to avoid modifying
+difficile} colonization (19, 20). However, we wanted to avoid modifying
 the gut communities with an antibiotic to maintain their unique
@@
 communities came from people at increased risk of CDI, such as recent
-hospitalization or antibiotic use (16), we decided to test if \emph{C.
-difficile} was able to infect these mice without any antibiotic
+hospitalization or antibiotic use (17), we tested whether \emph{C.
+difficile} was able to infect these mice without an antibiotic
 perturbation. We hypothesized that \emph{C. difficile} would be able to
@@
 damage, tissue edema, and inflammation (Figure S1) similar to previously
-reported histopathologic findings for \emph{C. difficile} RT027 (20). As
+reported histopathologic findings for \emph{C. difficile} RT027 (21). As
 observed with toxin activity, the moribund mice had higher
@@
 have limited ability to control variation in the host, microbiome or
-\emph{C. difficile} ribotype (21). Murine experiments typically use a
+\emph{C. difficile} ribotype (22). Murine experiments typically use a
 single mouse colony and different \emph{C. difficile} ribotypes to
-create severity differences (22). Recently, our group has begun
+create severity differences (23). Recently, our group has begun
 uncovering the effect microbiome variation has on \emph{C. difficile}
@@
 mice from different mouse colonies resulted in different clearance rates
-of \emph{C. difficile} (23). We also showed varied ability of mice to
+of \emph{C. difficile} (15). We also showed varied ability of mice to
 spontaneously eliminate \emph{C. difficile} infection when they were
@@
 and production of the toxins TcdA, TcdB, and binary toxin and the
-prevalence in outbreaks, such as ribotypes 027 and 078 (18, 28--31).
+prevalence in outbreaks, such as ribotypes 027 and 078 (19, 28--31).
 However, other studies have shown that disease is not necessarily linked
@@
 identifying many patients with CDI that will develop severe disease (3,
-22, 38, 39). Therefore, it is necessary to gain a full understanding of
+23, 38, 39). Therefore, it is necessary to gain a full understanding of
 all factors contributing to disease variation to improve our ability to
@@
 \emph{C. difficile} to colonize and persist in the intestine. \emph{C.
-difficile} is unable to colonize an unperturbed healthy gut community
-and is only able to become established after a perturbation (19). Once
-colonized, the different communities lead to different metabolic
-responses and dynamics of the \emph{C. difficile} population (8, 24,
-40). Gut bacteria metabolize primary bile acids into secondary bile
-acids (41, 42). The concentration of these bile acids affects
+difficile} is unable to colonize an unperturbed healthy murine gut
+community and is only able to become established after a perturbation
+(20). Once colonized, the different communities lead to different
+metabolic responses and dynamics of the \emph{C. difficile} population
+(8, 24, 40). Gut bacteria metabolize primary bile acids into secondary
+bile acids (41, 42). The concentration of these bile acids affects
 germination, growth, toxin production and biofilm formation (9, 10, 43,
@@
 A myriad of studies have explored the relationship between the
-microbiome and CDI disease. CDI studies often use an infection model
-with a single homogeneous bacterial community. Collins \emph{et al.}
-used multiple human communities to colonize mice, however the
-communities were pooled prior to gavaging into germ-free mice (48),
-resulting in a single human-derived community. Studies examining
-difference in disease often use different \emph{C. difficile} strains or
-ribotypes in mice with similar microbiota as a proxy for variation in
-disease, such as strain 630 for non-severe and RT027 for severe (18, 28,
-29, 49). Studies have also demonstrated variation in infection through
-tapering antibiotic dosage (19, 24, 50) or by reducing the amount of
-\emph{C. difficile} cells or spores used for the challenge (18, 50).
-These studies either lack variation in the initial microbiome or have
+microbiome and CDI disease. Studies examining difference in disease
+often use different \emph{C. difficile} strains or ribotypes in mice
+with similar microbiota as a proxy for variation in disease, such as
+strain 630 for non-severe and RT027 for severe (19, 28, 29, 48). Studies
+have also demonstrated variation in infection through tapering
+antibiotic dosage (20, 24, 49) or by reducing the amount of \emph{C.
+difficile} cells or spores used for the challenge (19, 49). These
+studies often either lack variation in the initial microbiome or have
 variation in the \emph{C. difficile} infection itself, confounding any
-association between variation in severity and the microbiome.
-
-Recently, Nagao-Kitamoto \emph{et al.} demonstrated when only allowing
-the microbiome to vary in a murine \emph{C. difficile} infection
-experiemnt that the resultant disease variation associated with the
-microbiome communities (14). Our data have demonstrated gut bacterial
-relative abundances associate with variation in toxin production,
-histopathologic scoring of the cecal tissue and mortality. This analysis
-revealed populations of \emph{Akkermansia}, \emph{Anaerostipes},
-\emph{Coprobacillus}, \emph{Enterocloster}, \emph{Lactonifactor}, and
-\emph{Monoglobus} were more abundant in the microbiome of non-moribund
-mice which had low histopathologic scores and no detected toxin. The
-protective role of these genera are supported by previous studies.
-\emph{Coprobacillus}, \emph{Lactonifactor}, and \emph{Monoglobus} have
-been shown to be involved in dietary fiber fermentation and associated
-with healthy communities (51--54). \emph{Anaerostipes} and
-\emph{Coprobacillus}, which produce short chain fatty acids, have been
-associated with healthy communities (55--57). Furthermore,
-\emph{Coprobacillus}, which was abundant in mice with low
-histopathologic scores but rare in all other mice, has been shown to
-contain a putative type I lantibiotic gene cluster and inhibit \emph{C.
-difficile} colonization (58--60). \emph{Akkermansia} and
-\emph{Enterocloster} were also identified as more abundant in mice which
-had a low histopathologic scores but have contradictory supporting
-evidence in the current literature. In our data, \emph{Akkermansia} was
-most abundant in the non-moribund mice with low histopathologic scores
-but there were some moribund mice which had increased populations of
-\emph{Akkermansia}. This could be attributed to either a more protective
-mucus layer was present inhibiting colonization (60, 61) or mucus
-consumption by \emph{Akkermansia} could have been crossfeeding \emph{C.
-difficile} or exposing a niche for \emph{C. difficile} (62--64).
-Similarly, \emph{Enterocloster} was more abundant and associated with
-low histopathologic scores. It has been associated with healthy
-populations and has been used to mono-colonize germ-free mice to reduce
-the ability of \emph{C. difficile} to colonize (65, 66). However,
-\emph{Enterocloster} has also been involved in infections, such as
-bacteremia (67, 68). These data have exemplified populations of bacteria
-that have the potential to be either protective or harmful. Thus, the
-disease outcome is not likely based on the abundance of individual
-populations of bacteria, rather it is the result of the interactions of
-the community.
+association between variation in severity and the microbiome. Recent
+studies have shown variation in the initial microbiome, via different
+murine colonies or colonizing germ-free mice with human feces, that were
+challenged with \emph{C. difficile} resulted in varied outcomes of the
+infection (14, 15).
+
+Our data have demonstrated gut bacterial relative abundances associate
+with variation in toxin production, histopathologic scoring of the cecal
+tissue and mortality. This analysis revealed populations of
+\emph{Akkermansia}, \emph{Anaerostipes}, \emph{Coprobacillus},
+\emph{Enterocloster}, \emph{Lactonifactor}, and \emph{Monoglobus} were
+more abundant in the microbiome of non-moribund mice which had low
+histopathologic scores and no detected toxin. The protective role of
+these genera are supported by previous studies. \emph{Coprobacillus},
+\emph{Lactonifactor}, and \emph{Monoglobus} have been shown to be
+involved in dietary fiber fermentation and associated with healthy
+communities (50--53). \emph{Anaerostipes} and \emph{Coprobacillus},
+which produce short chain fatty acids, have been associated with healthy
+communities (54--56). Furthermore, \emph{Coprobacillus}, which was
+abundant in mice with low histopathologic scores but rare in all other
+mice, has been shown to contain a putative type I lantibiotic gene
+cluster and inhibit \emph{C. difficile} colonization (57--59).
+\emph{Akkermansia} and \emph{Enterocloster} were also identified as more
+abundant in mice which had a low histopathologic scores but have
+contradictory supporting evidence in the current literature. In our
+data, \emph{Akkermansia} was most abundant in the non-moribund mice with
+low histopathologic scores but there were some moribund mice which had
+increased populations of \emph{Akkermansia}. This could be attributed to
+either a more protective mucus layer was present inhibiting colonization
+(59, 60) or mucus consumption by \emph{Akkermansia} could have been
+crossfeeding \emph{C. difficile} or exposing a niche for \emph{C.
+difficile} (61--63). Similarly, \emph{Enterocloster} was more abundant
+and associated with low histopathologic scores. It has been associated
+with healthy populations and has been used to mono-colonize germ-free
+mice to reduce the ability of \emph{C. difficile} to colonize (64, 65).
+However, \emph{Enterocloster} has also been involved in infections, such
+as bacteremia (66, 67). These data have exemplified populations of
+bacteria that have the potential to be either protective or harmful.
+Thus, the disease outcome is not likely based on the abundance of
+individual populations of bacteria, rather it is the result of the
+interactions of the community.
 
@@
 \emph{Shigella/Escherichia}, \emph{Staphylococcus}, and
-\emph{Streptococcus} have been shown to expand after antibiotic use (16,
-69, 70) and are commonly detected in CDI cases (71--74). In addition to
+\emph{Streptococcus} have been shown to expand after antibiotic use (17,
+68, 69) and are commonly detected in CDI cases (70--73). In addition to
 these populations, \emph{Eggerthella}, \emph{Prevotellaceae} and
 \emph{Helicobacter}, which associated with worse outcomes, have also
-been associated with intestinal inflammation (75--77). Recently,
+been associated with intestinal inflammation (74--76). Recently,
 \emph{Helicobacter hepaticus} was shown to be sufficient to cause
-susceptibility to CDI in IL-10 deficient C57BL/6 mice (78). In our
+susceptibility to CDI in IL-10 deficient C57BL/6 mice (77). In our
 experiments, when \emph{Helicobacter} was present, the infection
@@
 host, such as host mucus glycosylation or intestinal IL-33 expression
-(14, 81). For example, low levels of mucin degradation can provide
+(14, 80). For example, low levels of mucin degradation can provide
 nutrients to other community members producing a diverse non-damaging
-community (82). However, if mucin degradation becomes too great it
+community (81). However, if mucin degradation becomes too great it
 reduces the protective function of the mucin layer and exposes the
@@
 obtained as part of Schubert \emph{et al.} and selected based on
-community clusters (16) to result in diverse community structures. Feces
+community clusters (17) to result in diverse community structures. Feces
 were homogenized by mixing 200 mg of sample with 5 ml of PBS. Mice were
@@
 Spores concentration were determined both before and after challenge
-(83). \(10^{3}\) \emph{C. difficile} spores were given to each mouse via
+(82). \(10^{3}\) \emph{C. difficile} spores were given to each mouse via
 oral gavage.
@@
 for 24 hours. The plates were then counted for the number of colony
-forming units (CFU) (84).
+forming units (CFU) (83).
 
@@
 resulting amplicons using an Illumina MiSeq as described previously
-(85).
+(84).
 
 \textbf{Sequence curation.} Sequences were processed with
-mothur(v.1.44.3) as previously described (85, 86). In short, we used a
+mothur(v.1.44.3) as previously described (84, 85). In short, we used a
 3\% dissimilarity cutoff to group sequences into operational taxonomic
@@
 Database Project training set (version 18) to assign taxonomic
-classifications to each OTU (87). We sequenced a mock community of a
+classifications to each OTU (86). We sequenced a mock community of a
 known community composition and 16s rRNA gene sequences. We processed
@@
 structures, we calculated Yue and Clayton dissimilarity matrices
-(\(\theta\)\textsubscript{YC}) in mothur (88). We rarefied samples to
+(\(\theta\)\textsubscript{YC}) in mothur (87). We rarefied samples to
 2,107 sequences per sample to limit uneven sampling biases. We tested
 for differences in individual taxonomic groups that would explain the
-outcome differences with LEfSe (89) in mothur. Remaining statistical
+outcome differences with LEfSe (88) in mothur. Remaining statistical
 analysis and data visualization was performed in R (v4.0.5) with the
@@
 multiple comparisons with a Benjamini and Hochberg adjustment for a type
-I error rate of 0.05 (90). We built random forest models using the
-mikropml package (91) with relative abundance summed by taxonomic ranks
+I error rate of 0.05 (89). We built random forest models using the
+mikropml package (90) with relative abundance summed by taxonomic ranks
 from day 0 samples using mtry values of 1 through 10, 15, 20, 25, 40,
@@
 closely resemble the murine communities than the donor species community
-(92). Furthermore, it is not our intention to make any inferences
+(91). Furthermore, it is not our intention to make any inferences
 regarding human associated bacteria and their relationship with human
@@
 
+\leavevmode\hypertarget{ref-Tomkovich2020}{}%
+15. \textbf{Tomkovich S}, \textbf{Stough JMA}, \textbf{Bishop L},
+\textbf{Schloss PD}. 2020. The initial gut microbiota and response to
+antibiotic perturbation influence \emph{Clostridioides difficile}
+clearance in mice. mSphere \textbf{5}.
+doi:\href{https://doi.org/10.1128/msphere.00869-20}{10.1128/msphere.00869-20}.
+
 \leavevmode\hypertarget{ref-Nagpal2018}{}%
-15. \textbf{Nagpal R}, \textbf{Wang S}, \textbf{Woods LCS},
+16. \textbf{Nagpal R}, \textbf{Wang S}, \textbf{Woods LCS},
 \textbf{Seshie O}, \textbf{Chung ST}, \textbf{Shively CA},
@@
 \leavevmode\hypertarget{ref-Schubert2014}{}%
-16. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
+17. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
 \textbf{Mogle J}, \textbf{Petrosino JP}, \textbf{Young VB},
@@
 \leavevmode\hypertarget{ref-Gillilland2012}{}%
-17. \textbf{Gillilland MG}, \textbf{Erb-Downward JR}, \textbf{Bassis
+18. \textbf{Gillilland MG}, \textbf{Erb-Downward JR}, \textbf{Bassis
 CM}, \textbf{Shen MC}, \textbf{Toews GB}, \textbf{Young VB},
@@
 \leavevmode\hypertarget{ref-Chen2008}{}%
-18. \textbf{Chen X}, \textbf{Katchar K}, \textbf{Goldsmith JD},
+19. \textbf{Chen X}, \textbf{Katchar K}, \textbf{Goldsmith JD},
 \textbf{Nanthakumar N}, \textbf{Cheknis A}, \textbf{Gerding DN},
@@
 \leavevmode\hypertarget{ref-Schubert2015}{}%
-19. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
+20. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
 Antibiotic-induced alterations of the murine gut microbiota and
@@
 \leavevmode\hypertarget{ref-Cowardin2016}{}%
-20. \textbf{Cowardin CA}, \textbf{Buonomo EL}, \textbf{Saleh MM},
+21. \textbf{Cowardin CA}, \textbf{Buonomo EL}, \textbf{Saleh MM},
 \textbf{Wilson MG}, \textbf{Burgess SL}, \textbf{Kuehne SA},
@@
 \leavevmode\hypertarget{ref-Seekatz2016}{}%
-21. \textbf{Seekatz AM}, \textbf{Rao K}, \textbf{Santhosh K},
+22. \textbf{Seekatz AM}, \textbf{Rao K}, \textbf{Santhosh K},
 \textbf{Young VB}. 2016. Dynamics of the fecal microbiome in patients
@@
 \leavevmode\hypertarget{ref-Dieterle2020}{}%
-22. \textbf{Dieterle MG}, \textbf{Putler R}, \textbf{Perry DA},
+23. \textbf{Dieterle MG}, \textbf{Putler R}, \textbf{Perry DA},
 \textbf{Menon A}, \textbf{Abernathy-Close L}, \textbf{Perlman NS},
@@
 doi:\href{https://doi.org/10.1128/mbio.00180-20}{10.1128/mbio.00180-20}.
-
-\leavevmode\hypertarget{ref-Tomkovich2020}{}%
-23. \textbf{Tomkovich S}, \textbf{Stough JMA}, \textbf{Bishop L},
-\textbf{Schloss PD}. 2020. The initial gut microbiota and response to
-antibiotic perturbation influence \emph{Clostridioides difficile}
-clearance in mice. mSphere \textbf{5}.
-doi:\href{https://doi.org/10.1128/msphere.00869-20}{10.1128/msphere.00869-20}.
 
@@
 
-\leavevmode\hypertarget{ref-Collins2015}{}%
-48. \textbf{Collins J}, \textbf{Auchtung JM}, \textbf{Schaefer L},
-\textbf{Eaton KA}, \textbf{Britton RA}. 2015. Humanized microbiota mice
-as a model of recurrent \emph{Clostridium difficile} disease. Microbiome
-\textbf{3}.
-doi:\href{https://doi.org/10.1186/s40168-015-0097-2}{10.1186/s40168-015-0097-2}.
-
 \leavevmode\hypertarget{ref-Lawley2012}{}%
-49. \textbf{Lawley TD}, \textbf{Clare S}, \textbf{Walker AW},
+48. \textbf{Lawley TD}, \textbf{Clare S}, \textbf{Walker AW},
 \textbf{Stares MD}, \textbf{Connor TR}, \textbf{Raisen C},
@@
 \leavevmode\hypertarget{ref-Frisbee2019}{}%
-81. \textbf{Frisbee AL}, \textbf{Saleh MM}, \textbf{Young MK},
+80. \textbf{Frisbee AL}, \textbf{Saleh MM}, \textbf{Young MK},
 \textbf{Leslie JL}, \textbf{Simpson ME}, \textbf{Abhyankar MM},
@@
 \leavevmode\hypertarget{ref-Tailford2015}{}%
-82. \textbf{Tailford LE}, \textbf{Crost EH}, \textbf{Kavanaugh D},
+81. \textbf{Tailford LE}, \textbf{Crost EH}, \textbf{Kavanaugh D},
 \textbf{Juge N}. 2015. Mucin glycan foraging in the human gut
@@
 \leavevmode\hypertarget{ref-Sorg2009}{}%
-83. \textbf{Sorg JA}, \textbf{Dineen SS}. 2009. Laboratory maintenance
+82. \textbf{Sorg JA}, \textbf{Dineen SS}. 2009. Laboratory maintenance
 of \emph{Clostridium difficile}. Current Protocols in Microbiology
@@
 \leavevmode\hypertarget{ref-Winston2016}{}%
-84. \textbf{Winston JA}, \textbf{Thanissery R}, \textbf{Montgomery SA},
+83. \textbf{Winston JA}, \textbf{Thanissery R}, \textbf{Montgomery SA},
 \textbf{Theriot CM}. 2016. Cefoperazone-treated mouse model of
@@
 \leavevmode\hypertarget{ref-Kozich2013}{}%
-85. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+84. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 \leavevmode\hypertarget{ref-Schloss2009}{}%
-86. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+85. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \leavevmode\hypertarget{ref-Wang2007}{}%
-87. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
+86. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
 \textbf{Cole JR}. 2007. Naïve bayesian classifier for rapid assignment
@@
 \leavevmode\hypertarget{ref-Yue2005}{}%
-88. \textbf{Yue JC}, \textbf{Clayton MK}. 2005. A similarity measure
+87. \textbf{Yue JC}, \textbf{Clayton MK}. 2005. A similarity measure
 based on species proportions. Communications in Statistics - Theory and
@@
 \leavevmode\hypertarget{ref-Segata2011}{}%
-89. \textbf{Segata N}, \textbf{Izard J}, \textbf{Waldron L},
+88. \textbf{Segata N}, \textbf{Izard J}, \textbf{Waldron L},
 \textbf{Gevers D}, \textbf{Miropolsky L}, \textbf{Garrett WS},
@@
 \leavevmode\hypertarget{ref-Benjamini1995}{}%
-90. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
+89. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
 false discovery rate: A practical and powerful approach to multiple
@@
 \leavevmode\hypertarget{ref-Topcuoglu2021}{}%
-91. \textbf{Topçuoğlu B}, \textbf{Lapp Z}, \textbf{Sovacool K},
+90. \textbf{Topçuoğlu B}, \textbf{Lapp Z}, \textbf{Sovacool K},
 \textbf{Snitkin E}, \textbf{Wiens J}, \textbf{Schloss P}. 2021.
@@
 \leavevmode\hypertarget{ref-Rawls2006}{}%
-92. \textbf{Rawls JF}, \textbf{Mahowald MA}, \textbf{Ley RE},
+91. \textbf{Rawls JF}, \textbf{Mahowald MA}, \textbf{Ley RE},
 \textbf{Gordon JI}. 2006. Reciprocal gut microbiota transplants from

### 2022-04-22 results/tables/header.tex
update manuscript w/revisions

--- 2022-01-26 submission/manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  12pt,
-]{article}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage[margin=1in]{geometry}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 \usepackage{booktabs}
@@
 \usepackage{makecell}
-\usepackage{caption}
-\usepackage{hyperref}
+\pagenumbering{gobble}
 \usepackage{helvet} % Helvetica font
@@
 \usepackage[T1]{fontenc}
-\usepackage[labelfont=bf]{caption}
-
-\usepackage[none]{hyphenat}
-
+\usepackage[labelformat=empty]{caption} % Table caption label NULL
+\usepackage[textfont=bf]{caption} % Table caption text bold
 \usepackage{setspace}
@@
 \setlength{\parskip}{1em}
-
-\usepackage{lineno}
-
-\usepackage{pdfpages}
-\floatplacement{figure}{H} % Keep the figure up top of the page
-\DeclareUnicodeCharacter{0301}{*************************************}
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newenvironment{cslreferences}%
-  {}%
-  {\par}
-
-\author{}
-\date{\vspace{-2.5em}}
-
-\begin{document}
-
-\pagenumbering{arabic}
-\linenumbers
-\doublespacing
-
-\hypertarget{the-gut-bacterial-community-potentiates-clostridioides-difficile-infection-severity.}{%
-\section{\texorpdfstring{The gut bacterial community potentiates
-\emph{Clostridioides difficile} infection
-severity.}{The gut bacterial community potentiates Clostridioides difficile infection severity.}}\label{the-gut-bacterial-community-potentiates-clostridioides-difficile-infection-severity.}}
-
-\vspace{30mm}
-
-\textbf{Running title:} Microbiota potentiates \emph{Clostridioides
-difficile} infection severity
-
-\vspace{20mm}
-
-Nicholas A. Lesniak\(^1\), Alyxandria M. Schubert\(^1\), Kaitlyn J.
-Flynn\(^1\), Jhansi L. Leslie\(^{1,4}\), Hamide Sinani\(^1\), Ingrid L.
-Bergin\(^3\), Vincent B. Young\(^{1,2}\), Patrick D.
-Schloss\(^{1,\dagger}\)
-
-\vspace{30mm}
-
-\(\dagger\) To whom correspondence should be addressed:
-\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}\\
-1. Department of Microbiology and Immunology, University of Michigan,
-Ann Arbor, MI\\
-2. Division of Infectious Diseases, Department of Internal Medicine,
-University of Michigan Medical School, Ann Arbor, MI\\
-3. Unit for Laboratory Animal Medicine, University of Michigan, Ann
-Arbor, MI\\
-4. Current affiliation: Department of Medicine, Division of
-International Health and Infectious Diseases, University of Virginia
-School of Medicine, Charlottesville, Virginia, USA
-
-\newpage
-
-\hypertarget{abstract}{%
-\subsection{Abstract}\label{abstract}}
-
-The severity of \emph{Clostridioides difficile} infections (CDI) has
-increased over the last few decades. Patient age, white blood cell
-count, creatinine levels as well as \emph{C. difficile} ribotype and
-toxin genes have been associated with disease severity. However, it is
-unclear whether there is an association between members of the gut
-microbiota and disease severity. The gut microbiota is known to interact
-with \emph{C. difficile} during infection. Perturbations to the gut
-microbiota are necessary for \emph{C. difficile} to colonize the gut.
-The gut microbiota can inhibit \emph{C. difficile} colonization through
-bile acid metabolism, nutrient consumption and bacteriocin production.
-Here we sought to demonstrate that members of the gut bacterial
-communities can also contribute to disease severity. We derived diverse
-gut communities by colonizing germ-free mice with different human fecal
-communities. The mice were then infected with a single \emph{C.
-difficile} ribotype 027 clinical isolate which resulted in moribundity
-and histopathologic differences. The variation in severity was
-associated with the human fecal community that the mice received.
-Generally, bacterial populations with pathogenic potential, such as
-\emph{Escherichia}, \emph{Helicobacter}, and \emph{Klebsiella}, were
-associated with more severe outcomes. Bacterial groups associated with
-fiber degradation, bile acid metabolism and lantibiotic production, such
-as \emph{Anaerostipes} and \emph{Coprobacillus}, were associated with
-less severe outcomes. These data indicate that, in addition to the host
-and \emph{C. difficile}, populations of gut bacteria can influence CDI
-disease severity.
-
-\hypertarget{importance}{%
-\subsection{Importance}\label{importance}}
-
-\emph{Clostridioides difficile} colonization can be asymptomatic or
-develop into an infection, ranging in severity from mild diarrhea to
-toxic megacolon, sepsis, and death. Models that predict severity and
-guide treatment decisions are based on clinical factors and \emph{C.
-difficile} characteristics. Although the gut microbiome plays a role in
-protecting against CDI, its effect on CDI disease severity is unclear
-and has not been incorporated into disease severity models. We
-demonstrated that variation in the microbiome of mice colonized with
-human feces yielded a range of disease outcomes. These results revealed
-groups of bacteria associated with both severe and mild \emph{C.
-difficile} infection outcomes. Gut bacterial community data from
-patients with CDI could improve our ability to identify patients at risk
-of developing more severe disease and improve interventions which target
-\emph{C. difficile} and the gut bacteria to reduce host damage.
-
-\newpage
-
-\hypertarget{introduction}{%
-\subsection{Introduction}\label{introduction}}
-
-\emph{Clostridioides difficile} infections (CDI) have increased in
-incidence and severity since \emph{C. difficile} was first identified as
-the cause of antibiotic-associated pseudomembranous colitis (1). CDI
-disease severity can range from mild diarrhea to toxic megacolon and
-death. The Infectious Diseases Society of America (IDSA) and Society for
-Healthcare Epidemiology of America (SHEA) guidelines define severe CDI
-in terms of a white blood cell count greater than 15,000 cells/mm\(^3\)
-and/or a serum creatinine greater than 1.5 mg/dL. Patients who develop
-shock or hypotension, ileus, or toxic megacolon are considered to have
-fulminant CDI (2). Since these measures are CDI outcomes, they have
-limited ability to predict risk of severe CDI when the infection is
-first detected. Schemes have been developed to score a patient's risk
-for severe CDI outcomes based on clinical factors but have not been
-robust for broad application (3). Thus, we have limited ability to
-prevent patients from developing severe CDI.
-
-Missing from CDI severity prediction models are the effects of the
-indigenous gut bacteria. \emph{C. difficile} interacts with the gut
-community in many ways. The indigenous bacteria of a healthy intestinal
-community provide a protective barrier preventing \emph{C. difficile}
-from infecting the gut. A range of mechanisms can disrupt this barrier,
-including antibiotics, medications, or dietary changes, and lead to
-increased susceptibility to CDI (4--6). Once \emph{C. difficile}
-overcomes the protective barrier and colonizes the intestine, the
-indigenous bacteria can either promote or inhibit \emph{C. difficile}
-through producing molecules or modifying the environment (7, 8). Bile
-acids metabolized by the gut bacteria can inhibit \emph{C. difficile}
-growth and affect toxin production (9, 10). Bacteria in the gut also can
-compete more directly with \emph{C. difficile} through antibiotic
-production or nutrient consumption (11--13). While the relationship
-between the gut bacteria and \emph{C. difficile} has been established,
-the effect the gut bacteria can have on CDI disease severity is unclear.
-
-Recent studies have demonstrated that when mice with diverse microbial
-communities were challenged with a high-toxigenic strain resulted in
-varied disease severity (14) and when challenged with a low-toxigenic
-strain members of the gut microbial community associated with variation
-in colonization (15). Here, we sought to further elucidate the
-relationship between members of the gut bacterial community and CDI
-disease severity when challenged with a high-toxigenic strain, \emph{C.
-difficile} ribotype 027 (RT027). We hypothesized that since specific
-groups of gut bacteria affect the metabolism of \emph{C. difficile} and
-its infection dynamics, we can also identify groups of bacteria that
-affect the disease severity of the infection. To test this hypothesis,
-we colonized germ-free C57BL/6 mice with human fecal samples to create
-varied gut communities. We then challenged the mice with \emph{C.
-difficile} RT027 and followed the mice for the development of severe
-outcomes of moribundity and histopathologic cecal tissue damage. Since
-the murine host and \emph{C. difficile} isolate were the same and only
-the gut community varied, the variation in disease severity we observed
-was attributable to the gut microbiome.
-
-\hypertarget{results}{%
-\subsection{Results}\label{results}}
-
-\textbf{\emph{C. difficile} is able to infect germ-free mice colonized
-with human fecal microbial communities without antibiotics.} To produce
-gut microbiomes with greater variation than those found in conventional
-mouse colonies, we colonized germ-free mice with bacteria from human
-feces (16). We inoculated germ-free C57BL/6 mice with homogenized feces
-from each of 15 human fecal samples via oral gavage. These human fecal
-samples were selected because they represented diverse community
-structures based on community clustering (17). The gut communities were
-allowed to equilibrate for two weeks post-inoculation (18). We then
-surveyed the bacterial members of the gut communities by 16S rRNA gene
-sequencing of murine fecal pellets (Figure 1A). The bacterial
-communities from each mouse grouped more closely to those communities
-from mice that received the same human fecal donor community than to the
-mice who received a different human fecal donor community (Figure 1B).
-The communities were primarily composed of populations of
-\emph{Clostridia}, \emph{Bacteroidia}, \emph{Erysipelotrichia},
-\emph{Bacilli}, and \emph{Gammaproteobacteria}. However, the gut
-bacterial communities of each donor group of mice harbored unique
-relative abundance distributions of the shared bacterial classes.
-
-Next, we tested this set of mice with their human-derived gut microbial
-communities for susceptibility to \emph{C. difficile} infection. A
-typical mouse model of CDI requires pre-treatment of conventional mice
-with antibiotics, such as clindamycin, to become susceptible to \emph{C.
-difficile} colonization (19, 20). However, we wanted to avoid modifying
-the gut communities with an antibiotic to maintain their unique
-microbial compositions and ecological relationships. Since some of these
-communities came from people at increased risk of CDI, such as recent
-hospitalization or antibiotic use (17), we tested whether \emph{C.
-difficile} was able to infect these mice without an antibiotic
-perturbation. We hypothesized that \emph{C. difficile} would be able to
-colonize the mice who received their gut communities from a donor with a
-perturbed community. Mice were challenged with 10\(^{3}\) \emph{C.
-difficile} RT027 clinical isolate spores. The mice were followed for 10
-days post-challenge, and their stool was collected and plated for
-\emph{C. difficile} colony forming units (CFU) to determine the extent
-of the infection. Surprisingly, communities from all donors were able to
-be colonized (Figure 2). Two mice were able to resist \emph{C.
-difficile} colonization, both received their community donor N1, which
-may be attributed to experimental variation since this group also had
-more mice. By colonizing germ-free mice with different human fecal
-communities, we were able to generate diverse gut communities in mice,
-which were susceptible to \emph{C. difficile} infection without further
-modification of the gut community.
-
-\textbf{Infection severity varies by initial community.} After we
-challenged the mice with \emph{C. difficile}, we investigated the
-outcome from the infection and its relationship to the initial
-community. We followed the mice for 10 days post-challenge for
-colonization density, toxin production, and mortality. Seven mice, from
-Donors N1, N3, N4, and N5, were not colonized at detectable levels on
-the day after \emph{C. difficile} challenge but were infected
-(\textgreater10\(^{6}\)) by the end of the experiment. All mice that
-received their community from Donor M1 through M6 succumbed to the
-infection and became moribund within 3 days post-challenge. The
-remaining mice, except the uninfected Donor N1 mice, maintained \emph{C.
-difficile} infection through the end of the experiment (Figure 2). At 10
-days post-challenge, or earlier for the moribund mice, mice were
-euthanised and fecal material were assayed for toxin activity and cecal
-tissue was collected and scored for histopathologic signs of disease
-(Figure 3). Overall, there was greater toxin activity detected in the
-stool of the moribund mice (\emph{P} = 0.003). However, when looking at
-each group of mice, we observed a range in toxin activity for both the
-moribund and non-moribund mice (Figure 3A). Non-moribund mice from
-Donors N2 and N5 through N9 had comparable toxin activity as the
-moribund mice. Additionally, not all moribund mice had toxin activity
-detected in their stool. Next, we examined the cecal tissue for
-histopathologic damage. Moribund mice had high levels of epithelial
-damage, tissue edema, and inflammation (Figure S1) similar to previously
-reported histopathologic findings for \emph{C. difficile} RT027 (21). As
-observed with toxin activity, the moribund mice had higher
-histopathologic scores than the non-moribund mice (\emph{P} \textless{}
-0.001). However, unlike the toxin activity, all moribund mice had
-consistently high histopathologic summary scores (Figure 3B). The
-non-moribund mice, Donor groups N1 through N9, had a range in tissue
-damage from none detected to similar levels as the moribund mice, which
-grouped by community donor. Together, the toxin activity,
-histopathologic score, and moribundity showed variation across the donor
-groups but were largely consistent within each donor group.
-
-\textbf{Microbial community members explain variation in CDI severity.}
-We next interrogated the bacterial communities at the time of \emph{C.
-difficile} challenge (day 0) for their relationship to infection
-outcomes using linear discriminant analysis (LDA) effect size (LEfSe)
-analysis to identify individual bacterial populations that could explain
-the variation in disease severity. We split the mice into groups by
-severity level based on their moribundity and histopathologic score. We
-dichotomized the histopathologic scores into high and low groups by
-splitting on the median score of 5. This analysis revealed 20 genera
-that were significantly different by the disease severity (Figure 4A).
-Bacterial genera \emph{Turicibacter}, \emph{Streptococcus},
-\emph{Staphylococcus}, \emph{Pseudomonas}, \emph{Phocaeicola},
-\emph{Parabacteroides}, \emph{Bacteroides}, and
-\emph{Escherichia/Shigella} were detected at higher relative abundances
-in the mice that became moribund. Populations of \emph{Anaerotignum},
-\emph{Coprobacillus}, \emph{Enterocloster}, and \emph{Murimonas} were
-more abundant in the non-moribund mice that would develop only low
-intestinal injury. To understand the role of toxin activity in disease
-severity, we applied LEfSe to identify the genera most likely to explain
-the differences between the presence and absence of detected toxin
-activity (Figure 4B). Many genera that associated with the presence of
-toxin were also associated with moribundity, such as populations of
-\emph{Escherichia/Shigella} and \emph{Bacteroides}. Likewise, there were
-genera such as \emph{Anaerotignum}, \emph{Enterocloster}, and
-\emph{Murimonas} that were associated with no detected toxin that also
-exhibited greater relative abundance in communities from non-moribund
-mice with a low histopathologic score. Lastly, we tested for
-correlations between the endpoint relative abundances of bacterial
-operational taxonomic units (OTUs) and the histopathologic summary score
-(Figure 4C). The endpoint relative abundance of \emph{Bacteroides} was
-positively correlated with histopathologic score, as its day 0 relative
-abundance did with disease severity (Figure 4A). Populations of
-\emph{Klebsiella} and \emph{Prevotellaceae} were positively correlated
-with the histopathologic score and were increased in the group of mice
-with detectable toxin. This analysis identified bacterial genera that
-were associated with the variation in moribundity, histopathologic
-score, and toxin.
-
-We next determined whether, collectively, bacterial community membership
-and relative abundance could be predictive of the CDI disease outcome.
-We trained random forest models with bacterial community relative
-abundance data from the day of colonization at each taxonomic rank to
-predict toxin, moribundity, and day 10 post-challenge histopathologic
-summary score. For predicting if detectable toxin would be produced,
-microbial populations aggregated by phylum rank classification performed
-similarly as models using lower taxonomic ranks (AUROC = 0.83, Figure
-S2). \emph{C. difficile} was more likely to produce detectable toxin
-when the community infected had less abundant populations of
-\emph{Verrucomicrobia} and \emph{Campilobacterota} and had more abundant
-populations of \emph{Proteobacteria} (Figure 5A). Next, we assessed the
-ability of the community to predict moribundity. Bacteria grouped by
-class rank classification was sufficient to predict which mice would
-succumb to the infection before the end of the experiment (AUROC = 0.91,
-Figure S2). The features with the greatest effect showed that
-communities with greater populations of bacteria belonging to
-\emph{Bacilli} and \emph{Firmicutes} and reduced populations of
-\emph{Erysipelotrichia} were more likely to result in moribundity
-(Figure 5B). Only one other class of bacteria was decreased in moribund
-mice, a group of unclassified \emph{Clostridia}. Lastly, the relative
-abundances of genera were able to predict a high or low histopathologic
-score (histopathologic scores were dichotomized as in previous analysis,
-AUROC = 0.99, Figure S2). No genera had a significantly greater effect
-on the model performance than any others, indicating the model was
-reliant on many genera for the correct prediction. The model used some
-of the genera identified in the LEfSe analysis, such as
-\emph{Coprobacillus}, \emph{Anaerostipes}, and \emph{Hungatella}.
-Communities with greater abundances of \emph{Hungatella},
-\emph{Eggerthella}, \emph{Bifidobacterium}, \emph{Duncaniella} and
-\emph{Neisseria} were more likely to have high histopathologic scores.
-These models have shown that the relative abundance of bacterial
-populations and their relationship to each other could be used to
-predict the variation in moribundity, histopathologic score, and
-detectable toxin of CDI.
-
-\hypertarget{discussion}{%
-\subsection{Discussion}\label{discussion}}
-
-Challenging mice colonized with different human fecal communities with
-\emph{C. difficile} RT027 demonstrated that variation in members of the
-gut microbiome affects \emph{C. difficile} infection disease severity.
-Our analysis revealed an association between the relative abundance of
-bacterial community members and disease severity. Previous studies
-investigating the severity of CDI disease involving the microbiome have
-had limited ability to interrogate this relationship between the
-microbiome and disease severity. Studies that have used clinical data
-have limited ability to control variation in the host, microbiome or
-\emph{C. difficile} ribotype (22). Murine experiments typically use a
-single mouse colony and different \emph{C. difficile} ribotypes to
-create severity differences (23). Recently, our group has begun
-uncovering the effect microbiome variation has on \emph{C. difficile}
-infection. We showed the variation in the bacterial communities between
-mice from different mouse colonies resulted in different clearance rates
-of \emph{C. difficile} (15). We also showed varied ability of mice to
-spontaneously eliminate \emph{C. difficile} infection when they were
-treated with different antibiotics prior to \emph{C. difficile}
-challenge (24). Overall, the results presented here have demonstrated
-that the gut bacterial community contributed to the severity of \emph{C.
-difficile} infection.
-
-\emph{C. difficile} can lead to asymptomatic colonization or infections
-with severity ranging from mild diarrhea to death. Physicians use
-classification tools to identify patients most at risk of developing a
-severe infection using white blood cell counts, serum albumin level, or
-serum creatinine level (2, 25, 26). Those levels are driven by the
-activities in the intestine (27). Research into the drivers of this
-variation have revealed factors that make \emph{C. difficile} more
-virulent. Strains are categorized for their virulence by the presence
-and production of the toxins TcdA, TcdB, and binary toxin and the
-prevalence in outbreaks, such as ribotypes 027 and 078 (19, 28--31).
-However, other studies have shown that disease is not necessarily linked
-with toxin production (32) or the strain (33). Furthermore, there is
-variation in the genome, growth rate, sporulation, germination, and
-toxin production in different isolates of a strain (34--37). This
-variation may help explain why severe CDI prediction tools often miss
-identifying many patients with CDI that will develop severe disease (3,
-23, 38, 39). Therefore, it is necessary to gain a full understanding of
-all factors contributing to disease variation to improve our ability to
-predict severity.
-
-The state of the gut bacterial community determines the ability of
-\emph{C. difficile} to colonize and persist in the intestine. \emph{C.
-difficile} is unable to colonize an unperturbed healthy murine gut
-community and is only able to become established after a perturbation
-(20). Once colonized, the different communities lead to different
-metabolic responses and dynamics of the \emph{C. difficile} population
-(8, 24, 40). Gut bacteria metabolize primary bile acids into secondary
-bile acids (41, 42). The concentration of these bile acids affects
-germination, growth, toxin production and biofilm formation (9, 10, 43,
-44). Members of the bacterial community also affect other metabolites
-\emph{C. difficile} utilizes. \emph{Bacteroides thetaiotaomicron}
-produce sialidases which release sialic acid from the mucosa for
-\emph{C. difficile} to utilize (45, 46). The nutrient environment
-affects toxin production (47). Thus, many of the actions of the gut
-bacteria modulate \emph{C. difficile} in ways that could affect the
-infection and resultant disease.
-
-A myriad of studies have explored the relationship between the
-microbiome and CDI disease. Studies examining difference in disease
-often use different \emph{C. difficile} strains or ribotypes in mice
-with similar microbiota as a proxy for variation in disease, such as
-strain 630 for non-severe and RT027 for severe (19, 28, 29, 48). Studies
-have also demonstrated variation in infection through tapering
-antibiotic dosage (20, 24, 49) or by reducing the amount of \emph{C.
-difficile} cells or spores used for the challenge (19, 49). These
-studies often either lack variation in the initial microbiome or have
-variation in the \emph{C. difficile} infection itself, confounding any
-association between variation in severity and the microbiome. Recent
-studies have shown variation in the initial microbiome, via different
-murine colonies or colonizing germ-free mice with human feces, that were
-challenged with \emph{C. difficile} resulted in varied outcomes of the
-infection (14, 15).
-
-Our data have demonstrated gut bacterial relative abundances associate
-with variation in toxin production, histopathologic scoring of the cecal
-tissue and mortality. This analysis revealed populations of
-\emph{Akkermansia}, \emph{Anaerostipes}, \emph{Coprobacillus},
-\emph{Enterocloster}, \emph{Lactonifactor}, and \emph{Monoglobus} were
-more abundant in the microbiome of non-moribund mice which had low
-histopathologic scores and no detected toxin. The protective role of
-these genera are supported by previous studies. \emph{Coprobacillus},
-\emph{Lactonifactor}, and \emph{Monoglobus} have been shown to be
-involved in dietary fiber fermentation and associated with healthy
-communities (50--53). \emph{Anaerostipes} and \emph{Coprobacillus},
-which produce short chain fatty acids, have been associated with healthy
-communities (54--56). Furthermore, \emph{Coprobacillus}, which was
-abundant in mice with low histopathologic scores but rare in all other
-mice, has been shown to contain a putative type I lantibiotic gene
-cluster and inhibit \emph{C. difficile} colonization (57--59).
-\emph{Akkermansia} and \emph{Enterocloster} were also identified as more
-abundant in mice which had a low histopathologic scores but have
-contradictory supporting evidence in the current literature. In our
-data, \emph{Akkermansia} was most abundant in the non-moribund mice with
-low histopathologic scores but there were some moribund mice which had
-increased populations of \emph{Akkermansia}. This could be attributed to
-either a more protective mucus layer was present inhibiting colonization
-(59, 60) or mucus consumption by \emph{Akkermansia} could have been
-crossfeeding \emph{C. difficile} or exposing a niche for \emph{C.
-difficile} (61--63). Similarly, \emph{Enterocloster} was more abundant
-and associated with low histopathologic scores. It has been associated
-with healthy populations and has been used to mono-colonize germ-free
-mice to reduce the ability of \emph{C. difficile} to colonize (64, 65).
-However, \emph{Enterocloster} has also been involved in infections, such
-as bacteremia (66, 67). These data have exemplified populations of
-bacteria that have the potential to be either protective or harmful.
-Thus, the disease outcome is not likely based on the abundance of
-individual populations of bacteria, rather it is the result of the
-interactions of the community.
-
-The groups of bacteria that were associated with either a higher
-histopathologic score or moribundity are members of the indigenous gut
-community that also have been associated with disease, often referred to
-as opportunistic pathogens. Many of the populations with pathogenic
-potential that associated with worse outcomes are also facultative
-anaerobes. \emph{Enterococcus}, \emph{Klebsiella},
-\emph{Shigella/Escherichia}, \emph{Staphylococcus}, and
-\emph{Streptococcus} have been shown to expand after antibiotic use (17,
-68, 69) and are commonly detected in CDI cases (70--73). In addition to
-these populations, \emph{Eggerthella}, \emph{Prevotellaceae} and
-\emph{Helicobacter}, which associated with worse outcomes, have also
-been associated with intestinal inflammation (74--76). Recently,
-\emph{Helicobacter hepaticus} was shown to be sufficient to cause
-susceptibility to CDI in IL-10 deficient C57BL/6 mice (77). In our
-experiments, when \emph{Helicobacter} was present, the infection
-resulted in a high histopathologic score (Figure 4C). While we did not
-use IL-10 deficient mice, it is possible the bacterial community or host
-response are similarly modified by \emph{Helicobacter}, allowing
-\emph{C. difficile} infection and host damage. Aside from
-\emph{Helicobacter}, these groups of bacteria that associated with more
-severe outcomes did not have a conserved association between their
-relative abundance and the disease severity across all mice.
-
-Since we observed groups of bacteria that were associated with less
-severe disease it may be appropriate to apply the damage-response
-framework for microbial pathogenesis to CDI (78, 79). This framework
-posits that disease is not driven by a single entity, rather it is an
-emergent property of the responses of the host immune system, infecting
-microbe, \emph{C. difficile}, and the indigenous microbes at the site of
-infection. In the first set of experiments, we used the same host
-background, C57BL/6 mice, the same infecting microbe, \emph{C.
-difficile} RT027 clinical isolate 431, with different gut bacterial
-communities. The bacterial groups in those communities were often
-present in both moribund and non-moribund and across the range of
-histopathologic scores. Thus, it was not merely the presence of the
-bacteria but their activity in response to the other microbes and host
-which affect the extent of the host damage. Additionally, while each
-mouse and \emph{C. difficile} population had the same genetic
-background, they too were reacting to the specific microbial community.
-Disease severity is driven by the cumulative effect of the host immune
-response and the activity of \emph{C. difficile} and the gut bacteria.
-\emph{C. difficile} drives host damage through the production of toxin.
-The gut microbiota can modulate host damage through the balance of
-metabolic and competitive interactions with \emph{C. difficile}, such as
-bacteriocin production or mucin degradation, and interactions with the
-host, such as host mucus glycosylation or intestinal IL-33 expression
-(14, 80). For example, low levels of mucin degradation can provide
-nutrients to other community members producing a diverse non-damaging
-community (81). However, if mucin degradation becomes too great it
-reduces the protective function of the mucin layer and exposes the
-epithelial cells. This over-harvesting can contribute to the host damage
-due to other members producing toxin. Thus, the resultant intestinal
-damage is the balance of all activities in the gut environment. Host
-damage is the emergent property of numerous damage-response curves, such
-as one for host immune response, one for \emph{C. difficile} activity
-and another for microbiome community activity, each of which are a
-composite curve of the individual activities from each group, such as
-antibody production, neutrophil infiltration, toxin production,
-sporulation, fiber and mucin degradation. Therefore, while we have
-identified populations of interest, it may be necessary to target
-multiple types of bacteria to reduce the community interactions
-contributing to host damage.
-
-Here we have shown several bacterial groups and their relative
-abundances associated with variation in CDI disease severity. Further
-understanding how the microbiome affects severity in patients could
-reduce the amount of adverse CDI outcomes. When a patient is diagnosed
-with CDI, the gut community composition, in addition to the
-traditionally obtained clinical information, may improve our severity
-prediction and guide prophylactic treatment. Treating the microbiome at
-the time of diagnosis, in addition to \emph{C. difficile}, may prevent
-the infection from becoming more severe.
-
-\hypertarget{materials-and-methods}{%
-\subsection{Materials and Methods}\label{materials-and-methods}}
-
-\textbf{Animal care.} 6- to 13-week old male and female germ-free
-C57BL/6 were obtained from a single breeding colony in the University of
-Michigan Germ-free Mouse Core. Mice (N1 n=11, N2 n=7, N3 n=3, N4 n=3, N5
-n=3, N6 n=3, N7 n=7, N8 n=3, N9 n=2, M1 n=3, M2 n=3, M3 n=3, M4 n=3, M5
-n=7, M6 n=3) were housed in cages of 2-4 mice per cage and maintained in
-germ-free isolators at the University of Michigan germ-free facility.
-All mouse experiments were approved by the University Committee on Use
-and Care of Animals at the University of Michigan.
-
-\textbf{\emph{C. difficile} experiments.} Human fecal samples were
-obtained as part of Schubert \emph{et al.} and selected based on
-community clusters (17) to result in diverse community structures. Feces
-were homogenized by mixing 200 mg of sample with 5 ml of PBS. Mice were
-inoculated with 100 \(\mu\)l of the fecal homogenate via oral gavage.
-Two weeks after the fecal community inoculation, mice were challenged
-with \emph{C. difficile}. \emph{C. difficile} clinical isolate 431 came
-from Carlson \emph{et al.} which had previously been isolated and
-characterized (34, 35) and has recently been further characterized (36).
-Spores concentration were determined both before and after challenge
-(82). \(10^{3}\) \emph{C. difficile} spores were given to each mouse via
-oral gavage.
-
-\textbf{Sample collection.} Fecal samples were collected on the day of
-\emph{C. difficile} challenge and the following 10 days. Each day, a
-fecal sample was collected and a portion was weighed for plating
-(approximately 30 mg) and the remaining sample was frozen at
--20\(^\circ\)C. Anaerobically, the weighed fecal samples were serially
-diluted in PBS, plated on TCCFA plates, and incubated at 37\(^\circ\)C
-for 24 hours. The plates were then counted for the number of colony
-forming units (CFU) (83).
-
-\textbf{DNA sequencing.} From the frozen fecal samples, total bacterial
-DNA was extracted using MOBIO PowerSoil-htp 96-well soil DNA isolation
-kit. We amplified the 16S rRNA gene V4 region and sequenced the
-resulting amplicons using an Illumina MiSeq as described previously
-(84).
-
-\textbf{Sequence curation.} Sequences were processed with
-mothur(v.1.44.3) as previously described (84, 85). In short, we used a
-3\% dissimilarity cutoff to group sequences into operational taxonomic
-units (OTUs). We used a naive Bayesian classifier with the Ribosomal
-Database Project training set (version 18) to assign taxonomic
-classifications to each OTU (86). We sequenced a mock community of a
-known community composition and 16s rRNA gene sequences. We processed
-this mock community with our samples to calculate the error rate for our
-sequence curation, which was an error rate of 0.19\%.
-
-\textbf{Toxin cytotoxicity assay.} To prepare the sample for the
-activity assay, fecal material was diluted 1:10 weight per volume using
-sterile PBS and then filter sterilized through a 0.22-\(\mu\)m filter.
-Toxin activity was assessed using a Vero cell rounding-based
-cytotoxicity assay as described previously (29). The cytotoxicity titer
-was determined for each sample as the last dilution, which resulted in
-at least 80\% cell rounding. Toxin titers are reported as the log10 of
-the reciprocal of the cytotoxicity titer.
-
-\textbf{Histopathology evaluation.} Mouse cecal tissue was placed in
-histopathology cassettes and fixed in 10\% formalin, then stored in 70\%
-ethanol. McClinchey Histology Labs, Inc.~(Stockbridge, MI) embedded the
-samples in paraffin, sectioned, and created the hematoxylin and
-eosin-stained slides. The slides were scored using previously described
-criteria by a board-certified veterinary pathologist who was blinded to
-the experimental groups (29). Slides were scored as 0-4 for parameters
-of epithelial damage, tissue edema, and inflammation and a summary score
-of 0-12 was generated by summing the three individual parameter scores.
-
-\textbf{Statistical analysis and modeling.} To compare community
-structures, we calculated Yue and Clayton dissimilarity matrices
-(\(\theta\)\textsubscript{YC}) in mothur (87). We rarefied samples to
-2,107 sequences per sample to limit uneven sampling biases. We tested
-for differences in individual taxonomic groups that would explain the
-outcome differences with LEfSe (88) in mothur. Remaining statistical
-analysis and data visualization was performed in R (v4.0.5) with the
-tidyverse package (v1.3.1). We tested for significant differences in
-\(\beta\)-diversity (\(\theta\)\textsubscript{YC}) using the Wilcoxon
-rank sum test. We used Spearman's correlation to identify which OTUs
-that had a correlation between their relative abundance and the
-histopathologic summary score. \emph{P} values were then corrected for
-multiple comparisons with a Benjamini and Hochberg adjustment for a type
-I error rate of 0.05 (89). We built random forest models using the
-mikropml package (90) with relative abundance summed by taxonomic ranks
-from day 0 samples using mtry values of 1 through 10, 15, 20, 25, 40,
-50, 100. The split for training and testing varied by model to avoid
-overfitting the data. To determine the optimal split, we tested splits
-(50\%, 60\%, 70\%, 80\%, 90\% data used for training) to find the
-greatest portion of data that could be used to train the model while
-still maintaining the same performance for the training model as the
-model with the held-out test data. The toxin and moribundity models were
-trained with 60\% of the data. The histopathologic score model was
-trained with 80\% of the data. Lastly, we did not compare murine
-communities to donor community or clinical data because germ-free mice
-colonized with non-murine fecal communities have been shown to more
-closely resemble the murine communities than the donor species community
-(91). Furthermore, it is not our intention to make any inferences
-regarding human associated bacteria and their relationship with human
-CDI outcome.
-
-\textbf{Code availability.} Scripts necessary to reproduce our analysis
-and this paper are available in an online repository
-(\url{https://github.com/SchlossLab/Lesniak_Severity_XXXX_2022}).
-
-\textbf{Sequence data accession number.} All 16S rRNA gene sequence data
-and associated metadata are available through the Sequence Read Archive
-via accession PRJNA787941.
-
-\hypertarget{acknowledgements}{%
-\subsection{Acknowledgements}\label{acknowledgements}}
-
-Thank you to Sarah Lucas and Sarah Tomkovich for critical discussion in
-the development and execution of this project. We also thank the
-University of Michigan Germ-free Mouse Core for assistance with our
-germfree mice, funded in part by U2CDK110768. This work was supported by
-several grants from the National Institutes for Health R01GM099514,
-U19AI090871, U01AI12455, and P30DK034933. Additionally, NAL was
-supported by the Molecular Mechanisms of Microbial Pathogenesis training
-grant (NIH T32 AI007528). The funding agencies had no role in study
-design, data collection and analysis, decision to publish, or
-preparation of the manuscript.
-
-\newpage
-
-\hypertarget{references}{%
-\subsection{References}\label{references}}
-
-\hypertarget{refs}{}
-\begin{cslreferences}
-\leavevmode\hypertarget{ref-Kelly2008}{}%
-1. \textbf{Kelly CP}, \textbf{LaMont JT}. 2008. \emph{Clostridium
-difficile} more difficult than ever. New England Journal of Medicine
-\textbf{359}:1932--1940.
-doi:\href{https://doi.org/10.1056/nejmra0707500}{10.1056/nejmra0707500}.
-
-\leavevmode\hypertarget{ref-McDonald2018}{}%
-2. \textbf{McDonald LC}, \textbf{Gerding DN}, \textbf{Johnson S},
-\textbf{Bakken JS}, \textbf{Carroll KC}, \textbf{Coffin SE},
-\textbf{Dubberke ER}, \textbf{Garey KW}, \textbf{Gould CV},
-\textbf{Kelly C}, \textbf{Loo V}, \textbf{Sammons JS}, \textbf{Sandora
-TJ}, \textbf{Wilcox MH}. 2018. Clinical practice guidelines for
-\emph{Clostridium difficile} infection in adults and children: 2017
-update by the infectious diseases society of america (IDSA) and society
-for healthcare epidemiology of america (SHEA). Clinical Infectious
-Diseases \textbf{66}:e1--e48.
-doi:\href{https://doi.org/10.1093/cid/cix1085}{10.1093/cid/cix1085}.
-
-\leavevmode\hypertarget{ref-Perry2021}{}%
-3. \textbf{Perry DA}, \textbf{Shirley D}, \textbf{Micic D},
-\textbf{Patel CP}, \textbf{Putler R}, \textbf{Menon A}, \textbf{Young
-VB}, \textbf{Rao K}. 2021. External validation and comparison of
-\emph{Clostridioides difficile} severity scoring systems. Clinical
-Infectious Diseases.
-doi:\href{https://doi.org/10.1093/cid/ciab737}{10.1093/cid/ciab737}.
-
-\leavevmode\hypertarget{ref-Britton2014}{}%
-4. \textbf{Britton RA}, \textbf{Young VB}. 2014. Role of the intestinal
-microbiota in resistance to colonization by \emph{Clostridium
-difficile}. Gastroenterology \textbf{146}:1547--1553.
-doi:\href{https://doi.org/10.1053/j.gastro.2014.01.059}{10.1053/j.gastro.2014.01.059}.
-
-\leavevmode\hypertarget{ref-Hryckowian2018}{}%
-5. \textbf{Hryckowian AJ}, \textbf{Treuren WV}, \textbf{Smits SA},
-\textbf{Davis NM}, \textbf{Gardner JO}, \textbf{Bouley DM},
-\textbf{Sonnenburg JL}. 2018. Microbiota-accessible carbohydrates
-suppress \emph{Clostridium difficile} infection in a murine model.
-Nature Microbiology \textbf{3}:662--669.
-doi:\href{https://doi.org/10.1038/s41564-018-0150-6}{10.1038/s41564-018-0150-6}.
-
-\leavevmode\hypertarget{ref-VichVila2020}{}%
-6. \textbf{Vila AV}, \textbf{Collij V}, \textbf{Sanna S}, \textbf{Sinha
-T}, \textbf{Imhann F}, \textbf{Bourgonje AR}, \textbf{Mujagic Z},
-\textbf{Jonkers DMAE}, \textbf{Masclee AAM}, \textbf{Fu J},
-\textbf{Kurilshikov A}, \textbf{Wijmenga C}, \textbf{Zhernakova A},
-\textbf{Weersma RK}. 2020. Impact of commonly used drugs on the
-composition and metabolic function of the gut microbiota. Nature
-Communications \textbf{11}.
-doi:\href{https://doi.org/10.1038/s41467-019-14177-z}{10.1038/s41467-019-14177-z}.
-
-\leavevmode\hypertarget{ref-Abbas2020}{}%
-7. \textbf{Abbas A}, \textbf{Zackular JP}. 2020. Microbe-microbe
-interactions during \emph{Clostridioides difficile} infection. Current
-Opinion in Microbiology \textbf{53}:19--25.
-doi:\href{https://doi.org/10.1016/j.mib.2020.01.016}{10.1016/j.mib.2020.01.016}.
-
-\leavevmode\hypertarget{ref-Jenior2017}{}%
-8. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
-\textbf{Schloss PD}. 2017. \emph{Clostridium difficile} colonizes
-alternative nutrient niches during infection across distinct murine gut
-microbiomes. mSystems \textbf{2}.
-doi:\href{https://doi.org/10.1128/msystems.00063-17}{10.1128/msystems.00063-17}.
-
-\leavevmode\hypertarget{ref-Sorg2008}{}%
-9. \textbf{Sorg JA}, \textbf{Sonenshein AL}. 2008. Bile salts and
-glycine as cogerminants for \emph{Clostridium difficile} spores. Journal
-of Bacteriology \textbf{190}:2505--2512.
-doi:\href{https://doi.org/10.1128/jb.01765-07}{10.1128/jb.01765-07}.
-
-\leavevmode\hypertarget{ref-Thanissery2017}{}%
-10. \textbf{Thanissery R}, \textbf{Winston JA}, \textbf{Theriot CM}.
-2017. Inhibition of spore germination, growth, and toxin activity of
-clinically relevant \emph{C. difficile} strains by gut microbiota
-derived secondary bile acids. Anaerobe \textbf{45}:86--100.
-doi:\href{https://doi.org/10.1016/j.anaerobe.2017.03.004}{10.1016/j.anaerobe.2017.03.004}.
-
-\leavevmode\hypertarget{ref-Aguirre2021}{}%
-11. \textbf{Aguirre AM}, \textbf{Yalcinkaya N}, \textbf{Wu Q},
-\textbf{Swennes A}, \textbf{Tessier ME}, \textbf{Roberts P},
-\textbf{Miyajima F}, \textbf{Savidge T}, \textbf{Sorg JA}. 2021. Bile
-acid-independent protection against \emph{Clostridioides difficile}
-infection. PLOS Pathogens \textbf{17}:e1010015.
-doi:\href{https://doi.org/10.1371/journal.ppat.1010015}{10.1371/journal.ppat.1010015}.
-
-\leavevmode\hypertarget{ref-Kang2019}{}%
-12. \textbf{Kang JD}, \textbf{Myers CJ}, \textbf{Harris SC},
-\textbf{Kakiyama G}, \textbf{Lee I-K}, \textbf{Yun B-S},
-\textbf{Matsuzaki K}, \textbf{Furukawa M}, \textbf{Min H-K},
-\textbf{Bajaj JS}, \textbf{Zhou H}, \textbf{Hylemon PB}. 2019. Bile acid
-7\(\alpha\)-dehydroxylating gut bacteria secrete antibiotics that
-inhibit \emph{Clostridium difficile}: Role of secondary bile acids. Cell
-Chemical Biology \textbf{26}:27--34.e4.
-doi:\href{https://doi.org/10.1016/j.chembiol.2018.10.003}{10.1016/j.chembiol.2018.10.003}.
-
-\leavevmode\hypertarget{ref-Leslie2021}{}%
-13. \textbf{Leslie JL}, \textbf{Jenior ML}, \textbf{Vendrov KC},
-\textbf{Standke AK}, \textbf{Barron MR}, \textbf{O'Brien TJ},
-\textbf{Unverdorben L}, \textbf{Thaprawat P}, \textbf{Bergin IL},
-\textbf{Schloss PD}, \textbf{Young VB}. 2021. Protection from lethal
-\emph{Clostridioides difficile} infection via intraspecies competition
-for cogerminant. mBio \textbf{12}.
-doi:\href{https://doi.org/10.1128/mbio.00522-21}{10.1128/mbio.00522-21}.
-
-\leavevmode\hypertarget{ref-NagaoKitamoto2020}{}%
-14. \textbf{Nagao-Kitamoto H}, \textbf{Leslie JL}, \textbf{Kitamoto S},
-\textbf{Jin C}, \textbf{Thomsson KA}, \textbf{Gillilland MG},
-\textbf{Kuffa P}, \textbf{Goto Y}, \textbf{Jenq RR}, \textbf{Ishii C},
-\textbf{Hirayama A}, \textbf{Seekatz AM}, \textbf{Martens EC},
-\textbf{Eaton KA}, \textbf{Kao JY}, \textbf{Fukuda S}, \textbf{Higgins
-PDR}, \textbf{Karlsson NG}, \textbf{Young VB}, \textbf{Kamada N}. 2020.
-Interleukin-22-mediated host glycosylation prevents \emph{Clostridioides
-difficile} infection by modulating the metabolic activity of the gut
-microbiota. Nature Medicine \textbf{26}:608--617.
-doi:\href{https://doi.org/10.1038/s41591-020-0764-0}{10.1038/s41591-020-0764-0}.
-
-\leavevmode\hypertarget{ref-Tomkovich2020}{}%
-15. \textbf{Tomkovich S}, \textbf{Stough JMA}, \textbf{Bishop L},
-\textbf{Schloss PD}. 2020. The initial gut microbiota and response to
-antibiotic perturbation influence \emph{Clostridioides difficile}
-clearance in mice. mSphere \textbf{5}.
-doi:\href{https://doi.org/10.1128/msphere.00869-20}{10.1128/msphere.00869-20}.
-
-\leavevmode\hypertarget{ref-Nagpal2018}{}%
-16. \textbf{Nagpal R}, \textbf{Wang S}, \textbf{Woods LCS},
-\textbf{Seshie O}, \textbf{Chung ST}, \textbf{Shively CA},
-\textbf{Register TC}, \textbf{Craft S}, \textbf{McClain DA},
-\textbf{Yadav H}. 2018. Comparative microbiome signatures and
-short-chain fatty acids in mouse, rat, non-human primate, and human
-feces. Frontiers in Microbiology \textbf{9}.
-doi:\href{https://doi.org/10.3389/fmicb.2018.02897}{10.3389/fmicb.2018.02897}.
-
-\leavevmode\hypertarget{ref-Schubert2014}{}%
-17. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
-\textbf{Mogle J}, \textbf{Petrosino JP}, \textbf{Young VB},
-\textbf{Aronoff DM}, \textbf{Schloss PD}. 2014. Microbiome data
-distinguish patients with \emph{Clostridium difficile} infection and
-non-\emph{C. difficile}-associated diarrhea from healthy controls. mBio
-\textbf{5}.
-doi:\href{https://doi.org/10.1128/mbio.01021-14}{10.1128/mbio.01021-14}.
-
-\leavevmode\hypertarget{ref-Gillilland2012}{}%
-18. \textbf{Gillilland MG}, \textbf{Erb-Downward JR}, \textbf{Bassis
-CM}, \textbf{Shen MC}, \textbf{Toews GB}, \textbf{Young VB},
-\textbf{Huffnagle GB}. 2012. Ecological succession of bacterial
-communities during conventionalization of germ-free mice. Applied and
-Environmental Microbiology \textbf{78}:2359--2366.
-doi:\href{https://doi.org/10.1128/aem.05239-11}{10.1128/aem.05239-11}.
-
-\leavevmode\hypertarget{ref-Chen2008}{}%
-19. \textbf{Chen X}, \textbf{Katchar K}, \textbf{Goldsmith JD},
-\textbf{Nanthakumar N}, \textbf{Cheknis A}, \textbf{Gerding DN},
-\textbf{Kelly CP}. 2008. A mouse model of \emph{Clostridium
-difficile}-associated disease. Gastroenterology \textbf{135}:1984--1992.
-doi:\href{https://doi.org/10.1053/j.gastro.2008.09.002}{10.1053/j.gastro.2008.09.002}.
-
-\leavevmode\hypertarget{ref-Schubert2015}{}%
-20. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
-Antibiotic-induced alterations of the murine gut microbiota and
-subsequent effects on colonization resistance against \emph{Clostridium
-difficile}. mBio \textbf{6}.
-doi:\href{https://doi.org/10.1128/mbio.00974-15}{10.1128/mbio.00974-15}.
-
-\leavevmode\hypertarget{ref-Cowardin2016}{}%
-21. \textbf{Cowardin CA}, \textbf{Buonomo EL}, \textbf{Saleh MM},
-\textbf{Wilson MG}, \textbf{Burgess SL}, \textbf{Kuehne SA},
-\textbf{Schwan C}, \textbf{Eichhoff AM}, \textbf{Koch-Nolte F},
-\textbf{Lyras D}, \textbf{Aktories K}, \textbf{Minton NP}, \textbf{Petri
-WA}. 2016. The binary toxin CDT enhances \emph{Clostridium difficile}
-virulence by suppressing protective colonic eosinophilia. Nature
-Microbiology \textbf{1}.
-doi:\href{https://doi.org/10.1038/nmicrobiol.2016.108}{10.1038/nmicrobiol.2016.108}.
-
-\leavevmode\hypertarget{ref-Seekatz2016}{}%
-22. \textbf{Seekatz AM}, \textbf{Rao K}, \textbf{Santhosh K},
-\textbf{Young VB}. 2016. Dynamics of the fecal microbiome in patients
-with recurrent and nonrecurrent \emph{Clostridium difficile} infection.
-Genome Medicine \textbf{8}.
-doi:\href{https://doi.org/10.1186/s13073-016-0298-8}{10.1186/s13073-016-0298-8}.
-
-\leavevmode\hypertarget{ref-Dieterle2020}{}%
-23. \textbf{Dieterle MG}, \textbf{Putler R}, \textbf{Perry DA},
-\textbf{Menon A}, \textbf{Abernathy-Close L}, \textbf{Perlman NS},
-\textbf{Penkevich A}, \textbf{Standke A}, \textbf{Keidan M},
-\textbf{Vendrov KC}, \textbf{Bergin IL}, \textbf{Young VB}, \textbf{Rao
-K}. 2020. Systemic inflammatory mediators are effective biomarkers for
-predicting adverse outcomes in \emph{Clostridioides difficile}
-infection. mBio \textbf{11}.
-doi:\href{https://doi.org/10.1128/mbio.00180-20}{10.1128/mbio.00180-20}.
-
-\leavevmode\hypertarget{ref-Lesniak2021}{}%
-24. \textbf{Lesniak NA}, \textbf{Schubert AM}, \textbf{Sinani H},
-\textbf{Schloss PD}. 2021. Clearance of \emph{Clostridioides difficile}
-colonization is associated with antibiotic-specific bacterial changes.
-mSphere \textbf{6}.
-doi:\href{https://doi.org/10.1128/msphere.01238-20}{10.1128/msphere.01238-20}.
-
-\leavevmode\hypertarget{ref-Lungulescu2011}{}%
-25. \textbf{Lungulescu OA}, \textbf{Cao W}, \textbf{Gatskevich E},
-\textbf{Tlhabano L}, \textbf{Stratidis JG}. 2011. CSI: A severity index
-for \emph{Clostridium difficile} infection at the time of admission.
-Journal of Hospital Infection \textbf{79}:151--154.
-doi:\href{https://doi.org/10.1016/j.jhin.2011.04.017}{10.1016/j.jhin.2011.04.017}.
-
-\leavevmode\hypertarget{ref-Zar2007}{}%
-26. \textbf{Zar FA}, \textbf{Bakkanagari SR}, \textbf{Moorthi KMLST},
-\textbf{Davis MB}. 2007. A comparison of vancomycin and metronidazole
-for the treatment of \emph{Clostridium difficile}-associated diarrhea,
-stratified by disease severity. Clinical Infectious Diseases
-\textbf{45}:302--307.
-doi:\href{https://doi.org/10.1086/519265}{10.1086/519265}.
-
-\leavevmode\hypertarget{ref-diMasi2018}{}%
-27. \textbf{Masi A di}, \textbf{Leboffe L}, \textbf{Polticelli F},
-\textbf{Tonon F}, \textbf{Zennaro C}, \textbf{Caterino M}, \textbf{Stano
-P}, \textbf{Fischer S}, \textbf{Hägele M}, \textbf{Müller M},
-\textbf{Kleger A}, \textbf{Papatheodorou P}, \textbf{Nocca G},
-\textbf{Arcovito A}, \textbf{Gori A}, \textbf{Ruoppolo M}, \textbf{Barth
-H}, \textbf{Petrosillo N}, \textbf{Ascenzi P}, \textbf{Bella SD}. 2018.
-Human serum albumin is an essential component of the host defense
-mechanism against \emph{Clostridium difficile} intoxication. The Journal
-of Infectious Diseases \textbf{218}:1424--1435.
-doi:\href{https://doi.org/10.1093/infdis/jiy338}{10.1093/infdis/jiy338}.
-
-\leavevmode\hypertarget{ref-AbernathyClose2020}{}%
-28. \textbf{Abernathy-Close L}, \textbf{Dieterle MG}, \textbf{Vendrov
-KC}, \textbf{Bergin IL}, \textbf{Rao K}, \textbf{Young VB}. 2020. Aging
-dampens the intestinal innate immune response during severe
-\emph{Clostridioides difficile} infection and is associated with altered
-cytokine levels and granulocyte mobilization. Infection and Immunity
-\textbf{88}.
-doi:\href{https://doi.org/10.1128/iai.00960-19}{10.1128/iai.00960-19}.
-
-\leavevmode\hypertarget{ref-Theriot2011}{}%
-29. \textbf{Theriot CM}, \textbf{Koumpouras CC}, \textbf{Carlson PE},
-\textbf{Bergin II}, \textbf{Aronoff DM}, \textbf{Young VB}. 2011.
-Cefoperazone-treated mice as an experimental platform to assess
-differential virulence of \emph{Clostridium difficile} strains. Gut
-Microbes \textbf{2}:326--334.
-doi:\href{https://doi.org/10.4161/gmic.19142}{10.4161/gmic.19142}.
-
-\leavevmode\hypertarget{ref-Goorhuis2008}{}%
-30. \textbf{Goorhuis A}, \textbf{Bakker D}, \textbf{Corver J},
-\textbf{Debast SB}, \textbf{Harmanus C}, \textbf{Notermans DW},
-\textbf{Bergwerff AA}, \textbf{Dekker FW}, \textbf{Kuijper EJ}. 2008.
-Emergence of \emph{Clostridium difficile} infection due to a new
-hypervirulent strain, polymerase chain reaction ribotype 078. Clinical
-Infectious Diseases \textbf{47}:1162--1170.
-doi:\href{https://doi.org/10.1086/592257}{10.1086/592257}.
-
-\leavevmode\hypertarget{ref-OConnor2009}{}%
-31. \textbf{O'Connor JR}, \textbf{Johnson S}, \textbf{Gerding DN}. 2009.
-\emph{Clostridium difficile} infection caused by the epidemic
-BI/NAP1/027 strain. Gastroenterology \textbf{136}:1913--1924.
-doi:\href{https://doi.org/10.1053/j.gastro.2009.02.073}{10.1053/j.gastro.2009.02.073}.
-
-\leavevmode\hypertarget{ref-Rao2015}{}%
-32. \textbf{Rao K}, \textbf{Micic D}, \textbf{Natarajan M},
-\textbf{Winters S}, \textbf{Kiel MJ}, \textbf{Walk ST}, \textbf{Santhosh
-K}, \textbf{Mogle JA}, \textbf{Galecki AT}, \textbf{LeBar W},
-\textbf{Higgins PDR}, \textbf{Young VB}, \textbf{Aronoff DM}. 2015.
-\emph{Clostridium difficile} ribotype 027: Relationship to age,
-detectability of toxins A or B in stool with rapid testing, severe
-infection, and mortality. Clinical Infectious Diseases
-\textbf{61}:233--241.
-doi:\href{https://doi.org/10.1093/cid/civ254}{10.1093/cid/civ254}.
-
-\leavevmode\hypertarget{ref-Walk2012}{}%
-33. \textbf{Walk ST}, \textbf{Micic D}, \textbf{Jain R}, \textbf{Lo ES},
-\textbf{Trivedi I}, \textbf{Liu EW}, \textbf{Almassalha LM},
-\textbf{Ewing SA}, \textbf{Ring C}, \textbf{Galecki AT}, \textbf{Rogers
-MAM}, \textbf{Washer L}, \textbf{Newton DW}, \textbf{Malani PN},
-\textbf{Young VB}, \textbf{Aronoff DM}. 2012. \emph{Clostridium
-difficile} ribotype does not predict severe infection. Clinical
-Infectious Diseases \textbf{55}:1661--1668.
-doi:\href{https://doi.org/10.1093/cid/cis786}{10.1093/cid/cis786}.
-
-\leavevmode\hypertarget{ref-Carlson2013}{}%
-34. \textbf{Carlson PE}, \textbf{Walk ST}, \textbf{Bourgis AET},
-\textbf{Liu MW}, \textbf{Kopliku F}, \textbf{Lo E}, \textbf{Young VB},
-\textbf{Aronoff DM}, \textbf{Hanna PC}. 2013. The relationship between
-phenotype, ribotype, and clinical disease in human \emph{Clostridium
-difficile} isolates. Anaerobe \textbf{24}:109--116.
-doi:\href{https://doi.org/10.1016/j.anaerobe.2013.04.003}{10.1016/j.anaerobe.2013.04.003}.
-
-\leavevmode\hypertarget{ref-Carlson2015}{}%
-35. \textbf{Carlson PE}, \textbf{Kaiser AM}, \textbf{McColm SA},
-\textbf{Bauer JM}, \textbf{Young VB}, \textbf{Aronoff DM}, \textbf{Hanna
-PC}. 2015. Variation in germination of \emph{Clostridium difficile}
-clinical isolates correlates to disease severity. Anaerobe
-\textbf{33}:64--70.
-doi:\href{https://doi.org/10.1016/j.anaerobe.2015.02.003}{10.1016/j.anaerobe.2015.02.003}.
-
-\leavevmode\hypertarget{ref-Saund2021}{}%
-36. \textbf{Saund K}, \textbf{Pirani A}, \textbf{Lacy B}, \textbf{Hanna
-PC}, \textbf{Snitkin ES}. 2021. Strain variation in \emph{Clostridioides
-difficile} toxin activity associated with genomic variation at both
-PaLoc and non-PaLoc loci.
-doi:\href{https://doi.org/10.1101/2021.12.08.471880}{10.1101/2021.12.08.471880}.
-
-\leavevmode\hypertarget{ref-He2010}{}%
-37. \textbf{He M}, \textbf{Sebaihia M}, \textbf{Lawley TD},
-\textbf{Stabler RA}, \textbf{Dawson LF}, \textbf{Martin MJ},
-\textbf{Holt KE}, \textbf{Seth-Smith HMB}, \textbf{Quail MA},
-\textbf{Rance R}, \textbf{Brooks K}, \textbf{Churcher C}, \textbf{Harris
-D}, \textbf{Bentley SD}, \textbf{Burrows C}, \textbf{Clark L},
-\textbf{Corton C}, \textbf{Murray V}, \textbf{Rose G}, \textbf{Thurston
-S}, \textbf{Tonder A van}, \textbf{Walker D}, \textbf{Wren BW},
-\textbf{Dougan G}, \textbf{Parkhill J}. 2010. Evolutionary dynamics of
-\emph{Clostridium difficile} over short and long time scales.
-Proceedings of the National Academy of Sciences \textbf{107}:7527--7532.
-doi:\href{https://doi.org/10.1073/pnas.0914322107}{10.1073/pnas.0914322107}.
-
-\leavevmode\hypertarget{ref-Butt2013}{}%
-38. \textbf{Butt E}, \textbf{Foster JA}, \textbf{Keedwell E},
-\textbf{Bell JE}, \textbf{Titball RW}, \textbf{Bhangu A},
-\textbf{Michell SL}, \textbf{Sheridan R}. 2013. Derivation and
-validation of a simple, accurate and robust prediction rule for risk of
-mortality in patients with \emph{Clostridium difficile} infection. BMC
-Infectious Diseases \textbf{13}.
-doi:\href{https://doi.org/10.1186/1471-2334-13-316}{10.1186/1471-2334-13-316}.
-
-\leavevmode\hypertarget{ref-vanBeurden2017}{}%
-39. \textbf{Beurden YH van}, \textbf{Hensgens MPM}, \textbf{Dekkers OM},
-\textbf{Cessie SL}, \textbf{Mulder CJJ}, \textbf{Vandenbroucke-Grauls
-CMJE}. 2017. External validation of three prediction tools for patients
-at risk of a complicated course of \emph{Clostridium difficile}
-infection: Disappointing in an outbreak setting. Infection Control \&
-Hospital Epidemiology \textbf{38}:897--905.
-doi:\href{https://doi.org/10.1017/ice.2017.89}{10.1017/ice.2017.89}.
-
-\leavevmode\hypertarget{ref-Jenior2018}{}%
-40. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
-\textbf{Schloss PD}. 2018. \emph{Clostridium difficile} alters the
-structure and metabolism of distinct cecal microbiomes during initial
-infection to promote sustained colonization. mSphere \textbf{3}.
-doi:\href{https://doi.org/10.1128/msphere.00261-18}{10.1128/msphere.00261-18}.
-
-\leavevmode\hypertarget{ref-Staley2016}{}%
-41. \textbf{Staley C}, \textbf{Weingarden AR}, \textbf{Khoruts A},
-\textbf{Sadowsky MJ}. 2016. Interaction of gut microbiota with bile acid
-metabolism and its influence on disease states. Applied Microbiology and
-Biotechnology \textbf{101}:47--64.
-doi:\href{https://doi.org/10.1007/s00253-016-8006-6}{10.1007/s00253-016-8006-6}.
-
-\leavevmode\hypertarget{ref-Long2017}{}%
-42. \textbf{Long SL}, \textbf{Gahan CGM}, \textbf{Joyce SA}. 2017.
-Interactions between gut bacteria and bile in health and disease.
-Molecular Aspects of Medicine \textbf{56}:54--65.
-doi:\href{https://doi.org/10.1016/j.mam.2017.06.002}{10.1016/j.mam.2017.06.002}.
-
-\leavevmode\hypertarget{ref-Sorg2010}{}%
-43. \textbf{Sorg JA}, \textbf{Sonenshein AL}. 2010. Inhibiting the
-initiation of \emph{Clostridium difficile} spore germination using
-analogs of chenodeoxycholic acid, a bile acid. Journal of Bacteriology
-\textbf{192}:4983--4990.
-doi:\href{https://doi.org/10.1128/jb.00610-10}{10.1128/jb.00610-10}.
-
-\leavevmode\hypertarget{ref-Dubois2019}{}%
-44. \textbf{Dubois T}, \textbf{Tremblay YDN}, \textbf{Hamiot A},
-\textbf{Martin-Verstraete I}, \textbf{Deschamps J}, \textbf{Monot M},
-\textbf{Briandet R}, \textbf{Dupuy B}. 2019. A microbiota-generated bile
-salt induces biofilm formation in \emph{Clostridium difficile}. npj
-Biofilms and Microbiomes \textbf{5}.
-doi:\href{https://doi.org/10.1038/s41522-019-0087-4}{10.1038/s41522-019-0087-4}.
-
-\leavevmode\hypertarget{ref-Ng2013}{}%
-45. \textbf{Ng KM}, \textbf{Ferreyra JA}, \textbf{Higginbottom SK},
-\textbf{Lynch JB}, \textbf{Kashyap PC}, \textbf{Gopinath S},
-\textbf{Naidu N}, \textbf{Choudhury B}, \textbf{Weimer BC},
-\textbf{Monack DM}, \textbf{Sonnenburg JL}. 2013. Microbiota-liberated
-host sugars facilitate post-antibiotic expansion of enteric pathogens.
-Nature \textbf{502}:96--99.
-doi:\href{https://doi.org/10.1038/nature12503}{10.1038/nature12503}.
-
-\leavevmode\hypertarget{ref-Ferreyra2014}{}%
-46. \textbf{Ferreyra JA}, \textbf{Wu KJ}, \textbf{Hryckowian AJ},
-\textbf{Bouley DM}, \textbf{Weimer BC}, \textbf{Sonnenburg JL}. 2014.
-Gut microbiota-produced succinate promotes \emph{C. difficile} infection
-after antibiotic treatment or motility disturbance. Cell Host \& Microbe
-\textbf{16}:770--777.
-doi:\href{https://doi.org/10.1016/j.chom.2014.11.003}{10.1016/j.chom.2014.11.003}.
-
-\leavevmode\hypertarget{ref-MartinVerstraete2016}{}%
-47. \textbf{Martin-Verstraete I}, \textbf{Peltier J}, \textbf{Dupuy B}.
-2016. The regulatory networks that control \emph{Clostridium difficile}
-toxin synthesis. Toxins \textbf{8}:153.
-doi:\href{https://doi.org/10.3390/toxins8050153}{10.3390/toxins8050153}.
-
-\leavevmode\hypertarget{ref-Lawley2012}{}%
-48. \textbf{Lawley TD}, \textbf{Clare S}, \textbf{Walker AW},
-\textbf{Stares MD}, \textbf{Connor TR}, \textbf{Raisen C},
-\textbf{Goulding D}, \textbf{Rad R}, \textbf{Schreiber F},
-\textbf{Brandt C}, \textbf{Deakin LJ}, \textbf{Pickard DJ},
-\textbf{Duncan SH}, \textbf{Flint HJ}, \textbf{Clark TG},
-\textbf{Parkhill J}, \textbf{Dougan G}. 2012. Targeted restoration of
-the intestinal microbiota with a simple, defined bacteriotherapy
-resolves relapsing \emph{Clostridium difficile} disease in mice. PLoS
-Pathogens \textbf{8}:e1002995.
-doi:\href{https://doi.org/10.1371/journal.ppat.1002995}{10.1371/journal.ppat.1002995}.
-
-\leavevmode\hypertarget{ref-Reeves2011}{}%
-49. \textbf{Reeves AE}, \textbf{Theriot CM}, \textbf{Bergin IL},
-\textbf{Huffnagle GB}, \textbf{Schloss PD}, \textbf{Young VB}. 2011. The
-interplay between microbiome dynamics and pathogen dynamics in a murine
-model of \emph{Clostridium difficile} infection. Gut Microbes
-\textbf{2}:145--158.
-doi:\href{https://doi.org/10.4161/gmic.2.3.16333}{10.4161/gmic.2.3.16333}.
-
-\leavevmode\hypertarget{ref-Mabrok2011}{}%
-50. \textbf{Mabrok HB}, \textbf{Klopfleisch R}, \textbf{Ghanem KZ},
-\textbf{Clavel T}, \textbf{Blaut M}, \textbf{Loh G}. 2011. Lignan
-transformation by gut bacteria lowers tumor burden in a gnotobiotic rat
-model of breast cancer. Carcinogenesis \textbf{33}:203--208.
-doi:\href{https://doi.org/10.1093/carcin/bgr256}{10.1093/carcin/bgr256}.
-
-\leavevmode\hypertarget{ref-Kim2019}{}%
-51. \textbf{Kim CC}, \textbf{Healey GR}, \textbf{Kelly WJ},
-\textbf{Patchett ML}, \textbf{Jordens Z}, \textbf{Tannock GW},
-\textbf{Sims IM}, \textbf{Bell TJ}, \textbf{Hedderley D},
-\textbf{Henrissat B}, \textbf{Rosendale DI}. 2019. Genomic insights from
-\emph{Monoglobus pectinilyticus}: A pectin-degrading specialist
-bacterium in the human colon. The ISME Journal \textbf{13}:1437--1456.
-doi:\href{https://doi.org/10.1038/s41396-019-0363-6}{10.1038/s41396-019-0363-6}.
-
-\leavevmode\hypertarget{ref-doPrado2021}{}%
-52. \textbf{Prado SBR do}, \textbf{Minguzzi BT}, \textbf{Hoffmann C},
-\textbf{Fabi JP}. 2021. Modulation of human gut microbiota by dietary
-fibers from unripe and ripe papayas: Distinct polysaccharide degradation
-using a colonic in vitro fermentation model. Food Chemistry
-\textbf{348}:129071.
-doi:\href{https://doi.org/10.1016/j.foodchem.2021.129071}{10.1016/j.foodchem.2021.129071}.
-
-\leavevmode\hypertarget{ref-Muthuramalingam2019}{}%
-53. \textbf{Muthuramalingam K}, \textbf{Singh V}, \textbf{Choi C},
-\textbf{Choi SI}, \textbf{Kim YM}, \textbf{Unno T}, \textbf{Cho M}.
-2019. Dietary intervention using (1, 3)/(1, 6)-\(\beta\)-glucan, a
-fungus-derived soluble prebiotic ameliorates high-fat diet-induced
-metabolic distress and alters beneficially the gut microbiota in mice
-model. European Journal of Nutrition \textbf{59}:2617--2629.
-doi:\href{https://doi.org/10.1007/s00394-019-02110-5}{10.1007/s00394-019-02110-5}.
-
-\leavevmode\hypertarget{ref-Han2019}{}%
-54. \textbf{Han S-H}, \textbf{Yi J}, \textbf{Kim J-H}, \textbf{Lee S},
-\textbf{Moon H-W}. 2019. Composition of gut microbiota in patients with
-toxigenic \emph{Clostridioides} (\emph{Clostridium}) \emph{difficile}:
-Comparison between subgroups according to clinical criteria and toxin
-gene load. PLOS ONE \textbf{14}:e0212626.
-doi:\href{https://doi.org/10.1371/journal.pone.0212626}{10.1371/journal.pone.0212626}.
-
-\leavevmode\hypertarget{ref-Duncan2004}{}%
-55. \textbf{Duncan SH}, \textbf{Louis P}, \textbf{Flint HJ}. 2004.
-Lactate-utilizing bacteria, isolated from human feces, that produce
-butyrate as a major fermentation product. Applied and Environmental
-Microbiology \textbf{70}:5810--5817.
-doi:\href{https://doi.org/10.1128/aem.70.10.5810-5817.2004}{10.1128/aem.70.10.5810-5817.2004}.
-
-\leavevmode\hypertarget{ref-Ye2018}{}%
-56. \textbf{Ye J}, \textbf{Lv L}, \textbf{Wu W}, \textbf{Li Y},
-\textbf{Shi D}, \textbf{Fang D}, \textbf{Guo F}, \textbf{Jiang H},
-\textbf{Yan R}, \textbf{Ye W}, \textbf{Li L}. 2018. Butyrate protects
-mice against methionine--choline-deficient diet-induced non-alcoholic
-steatohepatitis by improving gut barrier function, attenuating
-inflammation and reducing endotoxin levels. Frontiers in Microbiology
-\textbf{9}.
-doi:\href{https://doi.org/10.3389/fmicb.2018.01967}{10.3389/fmicb.2018.01967}.
-
-\leavevmode\hypertarget{ref-Walsh2017}{}%
-57. \textbf{Walsh CJ}, \textbf{Guinane CM}, \textbf{O'Toole PW},
-\textbf{Cotter PD}. 2017. A profile hidden markov model to investigate
-the distribution and frequency of LanB-encoding lantibiotic modification
-genes in the human oral and gut microbiome. PeerJ \textbf{5}:e3254.
-doi:\href{https://doi.org/10.7717/peerj.3254}{10.7717/peerj.3254}.
-
-\leavevmode\hypertarget{ref-Sandiford2018}{}%
-58. \textbf{Sandiford SK}. 2018. Current developments in lantibiotic
-discovery for treating \emph{Clostridium difficile} infection. Expert
-Opinion on Drug Discovery \textbf{14}:71--79.
-doi:\href{https://doi.org/10.1080/17460441.2019.1549032}{10.1080/17460441.2019.1549032}.
-
-\leavevmode\hypertarget{ref-Stein2013}{}%
-59. \textbf{Stein RR}, \textbf{Bucci V}, \textbf{Toussaint NC},
-\textbf{Buffie CG}, \textbf{Rätsch G}, \textbf{Pamer EG}, \textbf{Sander
-C}, \textbf{Xavier JB}. 2013. Ecological modeling from time-series
-inference: Insight into dynamics and stability of intestinal microbiota.
-PLoS Computational Biology \textbf{9}:e1003388.
-doi:\href{https://doi.org/10.1371/journal.pcbi.1003388}{10.1371/journal.pcbi.1003388}.
-
-\leavevmode\hypertarget{ref-Nakashima2021}{}%
-60. \textbf{Nakashima T}, \textbf{Fujii K}, \textbf{Seki T},
-\textbf{Aoyama M}, \textbf{Azuma A}, \textbf{Kawasome H}. 2021. Novel
-gut microbiota modulator, which markedly increases \emph{Akkermansia
-muciniphila} occupancy, ameliorates experimental colitis in rats.
-Digestive Diseases and Sciences.
-doi:\href{https://doi.org/10.1007/s10620-021-07131-x}{10.1007/s10620-021-07131-x}.
-
-\leavevmode\hypertarget{ref-Geerlings2018}{}%
-61. \textbf{Geerlings S}, \textbf{Kostopoulos I}, \textbf{Vos W de},
-\textbf{Belzer C}. 2018. \emph{Akkermansia muciniphila} in the human
-gastrointestinal tract: When, where, and how? Microorganisms
-\textbf{6}:75.
-doi:\href{https://doi.org/10.3390/microorganisms6030075}{10.3390/microorganisms6030075}.
-
-\leavevmode\hypertarget{ref-Deng2018}{}%
-62. \textbf{Deng H}, \textbf{Yang S}, \textbf{Zhang Y}, \textbf{Qian K},
-\textbf{Zhang Z}, \textbf{Liu Y}, \textbf{Wang Y}, \textbf{Bai Y},
-\textbf{Fan H}, \textbf{Zhao X}, \textbf{Zhi F}. 2018. \emph{Bacteroides
-fragilis} prevents \emph{Clostridium difficile} infection in a mouse
-model by restoring gut barrier and microbiome regulation. Frontiers in
-Microbiology \textbf{9}.
-doi:\href{https://doi.org/10.3389/fmicb.2018.02976}{10.3389/fmicb.2018.02976}.
-
-\leavevmode\hypertarget{ref-Engevik2020}{}%
-63. \textbf{Engevik MA}, \textbf{Engevik AC}, \textbf{Engevik KA},
-\textbf{Auchtung JM}, \textbf{Chang-Graham AL}, \textbf{Ruan W},
-\textbf{Luna RA}, \textbf{Hyser JM}, \textbf{Spinler JK},
-\textbf{Versalovic J}. 2020. Mucin-degrading microbes release
-monosaccharides that chemoattract \emph{Clostridioides difficile} and
-facilitate colonization of the human intestinal mucus layer. ACS
-Infectious Diseases \textbf{7}:1126--1142.
-doi:\href{https://doi.org/10.1021/acsinfecdis.0c00634}{10.1021/acsinfecdis.0c00634}.
-
-\leavevmode\hypertarget{ref-Reeves2012}{}%
-64. \textbf{Reeves AE}, \textbf{Koenigsknecht MJ}, \textbf{Bergin IL},
-\textbf{Young VB}. 2012. Suppression of \emph{Clostridium difficile} in
-the gastrointestinal tracts of germfree mice inoculated with a murine
-isolate from the family \emph{Lachnospiraceae}. Infection and Immunity
-\textbf{80}:3786--3794.
-doi:\href{https://doi.org/10.1128/iai.00647-12}{10.1128/iai.00647-12}.
-
-\leavevmode\hypertarget{ref-Ma2021}{}%
-65. \textbf{Ma L}, \textbf{Keng J}, \textbf{Cheng M}, \textbf{Pan H},
-\textbf{Feng B}, \textbf{Hu Y}, \textbf{Feng T}, \textbf{Yang F}. 2021.
-Gut microbiome and serum metabolome alterations associated with isolated
-dystonia. mSphere \textbf{6}.
-doi:\href{https://doi.org/10.1128/msphere.00283-21}{10.1128/msphere.00283-21}.
-
-\leavevmode\hypertarget{ref-Haas2020}{}%
-66. \textbf{Haas KN}, \textbf{Blanchard JL}. 2020. Reclassification of
-the \emph{Clostridium clostridioforme} and \emph{Clostridium sphenoides}
-clades as \emph{Enterocloster} gen. nov. And \emph{Lacrimispora} gen.
-nov., Including reclassification of 15 taxa. International Journal of
-Systematic and Evolutionary Microbiology \textbf{70}:23--34.
-doi:\href{https://doi.org/10.1099/ijsem.0.003698}{10.1099/ijsem.0.003698}.
-
-\leavevmode\hypertarget{ref-Finegold2005}{}%
-67. \textbf{Finegold SM}, \textbf{Song Y}, \textbf{Liu C}, \textbf{Hecht
-DW}, \textbf{Summanen P}, \textbf{Könönen E}, \textbf{Allen SD}. 2005.
-\emph{Clostridium clostridioforme}: A mixture of three clinically
-important species. European Journal of Clinical Microbiology \&
-Infectious Diseases \textbf{24}:319--324.
-doi:\href{https://doi.org/10.1007/s10096-005-1334-6}{10.1007/s10096-005-1334-6}.
-
-\leavevmode\hypertarget{ref-VanInsberghe2020}{}%
-68. \textbf{VanInsberghe D}, \textbf{Elsherbini JA}, \textbf{Varian B},
-\textbf{Poutahidis T}, \textbf{Erdman S}, \textbf{Polz MF}. 2020.
-Diarrhoeal events can trigger long-term \emph{Clostridium difficile}
-colonization with recurrent blooms. Nature Microbiology
-\textbf{5}:642--650.
-doi:\href{https://doi.org/10.1038/s41564-020-0668-2}{10.1038/s41564-020-0668-2}.
-
-\leavevmode\hypertarget{ref-GarzaGonzalez2019}{}%
-69. \textbf{Garza-González E}, \textbf{Mendoza-Olazarán S},
-\textbf{Morfin-Otero R}, \textbf{Ramírez-Fontes A},
-\textbf{Rodríguez-Zulueta P}, \textbf{Flores-Treviño S},
-\textbf{Bocanegra-Ibarias P}, \textbf{Maldonado-Garza H},
-\textbf{Camacho-Ortiz A}. 2019. Intestinal microbiome changes in fecal
-microbiota transplant (FMT) vs. FMT enriched with \emph{Lactobacillus}
-in the treatment of recurrent \emph{Clostridioides difficile} infection.
-Canadian Journal of Gastroenterology and Hepatology \textbf{2019}:1--7.
-doi:\href{https://doi.org/10.1155/2019/4549298}{10.1155/2019/4549298}.
-
-\leavevmode\hypertarget{ref-Shafiq2020}{}%
-70. \textbf{Shafiq M}, \textbf{Alturkmani H}, \textbf{Zafar Y},
-\textbf{Mittal V}, \textbf{Lodhi H}, \textbf{Ullah W}, \textbf{Brewer
-J}. 2020. Effects of co-infection on the clinical outcomes of
-\emph{Clostridium difficile} infection. Gut Pathogens \textbf{12}.
-doi:\href{https://doi.org/10.1186/s13099-020-00348-7}{10.1186/s13099-020-00348-7}.
-
-\leavevmode\hypertarget{ref-Keith2020}{}%
-71. \textbf{Keith JW}, \textbf{Dong Q}, \textbf{Sorbara MT},
-\textbf{Becattini S}, \textbf{Sia JK}, \textbf{Gjonbalaj M},
-\textbf{Seok R}, \textbf{Leiner IM}, \textbf{Littmann ER}, \textbf{Pamer
-EG}. 2020. Impact of antibiotic-resistant bacteria on immune activation
-and \emph{Clostridioides difficile} infection in the mouse intestine.
-Infection and Immunity \textbf{88}.
-doi:\href{https://doi.org/10.1128/iai.00362-19}{10.1128/iai.00362-19}.
-
-\leavevmode\hypertarget{ref-Zackular2016}{}%
-72. \textbf{Zackular JP}, \textbf{Moore JL}, \textbf{Jordan AT},
-\textbf{Juttukonda LJ}, \textbf{Noto MJ}, \textbf{Nicholson MR},
-\textbf{Crews JD}, \textbf{Semler MW}, \textbf{Zhang Y}, \textbf{Ware
-LB}, \textbf{Washington MK}, \textbf{Chazin WJ}, \textbf{Caprioli RM},
-\textbf{Skaar EP}. 2016. Dietary zinc alters the microbiota and
-decreases resistance to \emph{Clostridium difficile} infection. Nature
-Medicine \textbf{22}:1330--1334.
-doi:\href{https://doi.org/10.1038/nm.4174}{10.1038/nm.4174}.
-
-\leavevmode\hypertarget{ref-Berkell2021}{}%
-73. \textbf{Berkell M}, \textbf{Mysara M}, \textbf{Xavier BB},
-\textbf{Werkhoven CH van}, \textbf{Monsieurs P}, \textbf{Lammens C},
-\textbf{Ducher A}, \textbf{Vehreschild MJGT}, \textbf{Goossens H},
-\textbf{Gunzburg J de}, \textbf{Bonten MJM}, \textbf{Malhotra-Kumar S}.
-2021. Microbiota-based markers predictive of development of
-\emph{Clostridioides difficile} infection. Nature Communications
-\textbf{12}.
-doi:\href{https://doi.org/10.1038/s41467-021-22302-0}{10.1038/s41467-021-22302-0}.
-
-\leavevmode\hypertarget{ref-Gardiner2014}{}%
-74. \textbf{Gardiner BJ}, \textbf{Tai AY}, \textbf{Kotsanas D},
-\textbf{Francis MJ}, \textbf{Roberts SA}, \textbf{Ballard SA},
-\textbf{Junckerstorff RK}, \textbf{Korman TM}. 2014. Clinical and
-microbiological characteristics of \emph{Eggerthella lenta} bacteremia.
-Journal of Clinical Microbiology \textbf{53}:626--635.
-doi:\href{https://doi.org/10.1128/jcm.02926-14}{10.1128/jcm.02926-14}.
-
-\leavevmode\hypertarget{ref-Iljazovic2020}{}%
-75. \textbf{Iljazovic A}, \textbf{Roy U}, \textbf{Gálvez EJC},
-\textbf{Lesker TR}, \textbf{Zhao B}, \textbf{Gronow A}, \textbf{Amend
-L}, \textbf{Will SE}, \textbf{Hofmann JD}, \textbf{Pils MC},
-\textbf{Schmidt-Hohagen K}, \textbf{Neumann-Schaal M}, \textbf{Strowig
-T}. 2020. Perturbation of the gut microbiome by \emph{Prevotella} spp.
-enhances host susceptibility to mucosal inflammation. Mucosal Immunology
-\textbf{14}:113--124.
-doi:\href{https://doi.org/10.1038/s41385-020-0296-4}{10.1038/s41385-020-0296-4}.
-
-\leavevmode\hypertarget{ref-Nagalingam2013}{}%
-76. \textbf{Nagalingam NA}, \textbf{Robinson CJ}, \textbf{Bergin IL},
-\textbf{Eaton KA}, \textbf{Huffnagle GB}, \textbf{Young VB}. 2013. The
-effects of intestinal microbial community structure on disease
-manifestation in IL-\(10^{-/-}\) mice infected with \emph{Helicobacter
-hepaticus}. Microbiome \textbf{1}.
-doi:\href{https://doi.org/10.1186/2049-2618-1-15}{10.1186/2049-2618-1-15}.
-
-\leavevmode\hypertarget{ref-AbernathyClose2021}{}%
-77. \textbf{Abernathy-Close L}, \textbf{Barron MR}, \textbf{George JM},
-\textbf{Dieterle MG}, \textbf{Vendrov KC}, \textbf{Bergin IL},
-\textbf{Young VB}. 2021. Intestinal inflammation and altered gut
-microbiota associated with inflammatory bowel disease render mice
-susceptible to \emph{Clostridioides difficile} colonization and
-infection. mBio.
-doi:\href{https://doi.org/10.1128/mbio.02733-20}{10.1128/mbio.02733-20}.
-
-\leavevmode\hypertarget{ref-Pirofski2008}{}%
-78. \textbf{Pirofski L-a}, \textbf{Casadevall A}. 2008. The
-damage-response framework of microbial pathogenesis and infectious
-diseases, pp. 135--146. \emph{In} Advances in experimental medicine and
-biology. Springer New York.
-
-\leavevmode\hypertarget{ref-Casadevall2014}{}%
-79. \textbf{Casadevall A}, \textbf{Pirofski L-a}. 2014. What is a host?
-Incorporating the microbiota into the damage-response framework.
-Infection and Immunity \textbf{83}:2--7.
-doi:\href{https://doi.org/10.1128/iai.02627-14}{10.1128/iai.02627-14}.
-
-\leavevmode\hypertarget{ref-Frisbee2019}{}%
-80. \textbf{Frisbee AL}, \textbf{Saleh MM}, \textbf{Young MK},
-\textbf{Leslie JL}, \textbf{Simpson ME}, \textbf{Abhyankar MM},
-\textbf{Cowardin CA}, \textbf{Ma JZ}, \textbf{Pramoonjago P},
-\textbf{Turner SD}, \textbf{Liou AP}, \textbf{Buonomo EL}, \textbf{Petri
-WA}. 2019. IL-33 drives group 2 innate lymphoid cell-mediated protection
-during \emph{Clostridium difficile} infection. Nature Communications
-\textbf{10}.
-doi:\href{https://doi.org/10.1038/s41467-019-10733-9}{10.1038/s41467-019-10733-9}.
-
-\leavevmode\hypertarget{ref-Tailford2015}{}%
-81. \textbf{Tailford LE}, \textbf{Crost EH}, \textbf{Kavanaugh D},
-\textbf{Juge N}. 2015. Mucin glycan foraging in the human gut
-microbiome. Frontiers in Genetics \textbf{6}.
-doi:\href{https://doi.org/10.3389/fgene.2015.00081}{10.3389/fgene.2015.00081}.
-
-\leavevmode\hypertarget{ref-Sorg2009}{}%
-82. \textbf{Sorg JA}, \textbf{Dineen SS}. 2009. Laboratory maintenance
-of \emph{Clostridium difficile}. Current Protocols in Microbiology
-\textbf{12}.
-doi:\href{https://doi.org/10.1002/9780471729259.mc09a01s12}{10.1002/9780471729259.mc09a01s12}.
-
-\leavevmode\hypertarget{ref-Winston2016}{}%
-83. \textbf{Winston JA}, \textbf{Thanissery R}, \textbf{Montgomery SA},
-\textbf{Theriot CM}. 2016. Cefoperazone-treated mouse model of
-clinically-relevant \emph{Clostridium difficile} strain R20291. Journal
-of Visualized Experiments.
-doi:\href{https://doi.org/10.3791/54850}{10.3791/54850}.
-
-\leavevmode\hypertarget{ref-Kozich2013}{}%
-84. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
-\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
-dual-index sequencing strategy and curation pipeline for analyzing
-amplicon sequence data on the MiSeq illumina sequencing platform.
-Applied and Environmental Microbiology \textbf{79}:5112--5120.
-doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
-
-\leavevmode\hypertarget{ref-Schloss2009}{}%
-85. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
-\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
-\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
-\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
-\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
-Introducing mothur: Open-source, platform-independent,
-community-supported software for describing and comparing microbial
-communities. Applied and Environmental Microbiology
-\textbf{75}:7537--7541.
-doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.
-
-\leavevmode\hypertarget{ref-Wang2007}{}%
-86. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
-\textbf{Cole JR}. 2007. Naïve bayesian classifier for rapid assignment
-of rRNA sequences into the new bacterial taxonomy. Applied and
-Environmental Microbiology \textbf{73}:5261--5267.
-doi:\href{https://doi.org/10.1128/aem.00062-07}{10.1128/aem.00062-07}.
-
-\leavevmode\hypertarget{ref-Yue2005}{}%
-87. \textbf{Yue JC}, \textbf{Clayton MK}. 2005. A similarity measure
-based on species proportions. Communications in Statistics - Theory and
-Methods \textbf{34}:2123--2131.
-doi:\href{https://doi.org/10.1080/sta-200066418}{10.1080/sta-200066418}.
-
-\leavevmode\hypertarget{ref-Segata2011}{}%
-88. \textbf{Segata N}, \textbf{Izard J}, \textbf{Waldron L},
-\textbf{Gevers D}, \textbf{Miropolsky L}, \textbf{Garrett WS},
-\textbf{Huttenhower C}. 2011. Metagenomic biomarker discovery and
-explanation. Genome Biology \textbf{12}:R60.
-doi:\href{https://doi.org/10.1186/gb-2011-12-6-r60}{10.1186/gb-2011-12-6-r60}.
-
-\leavevmode\hypertarget{ref-Benjamini1995}{}%
-89. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
-false discovery rate: A practical and powerful approach to multiple
-testing. Journal of the Royal Statistical Society: Series B
-(Methodological) \textbf{57}:289--300.
-doi:\href{https://doi.org/10.1111/j.2517-6161.1995.tb02031.x}{10.1111/j.2517-6161.1995.tb02031.x}.
-
-\leavevmode\hypertarget{ref-Topcuoglu2021}{}%
-90. \textbf{Topçuoğlu B}, \textbf{Lapp Z}, \textbf{Sovacool K},
-\textbf{Snitkin E}, \textbf{Wiens J}, \textbf{Schloss P}. 2021.
-Mikropml: User-friendly R package for supervised machine learning
-pipelines. Journal of Open Source Software \textbf{6}:3073.
-doi:\href{https://doi.org/10.21105/joss.03073}{10.21105/joss.03073}.
-
-\leavevmode\hypertarget{ref-Rawls2006}{}%
-91. \textbf{Rawls JF}, \textbf{Mahowald MA}, \textbf{Ley RE},
-\textbf{Gordon JI}. 2006. Reciprocal gut microbiota transplants from
-zebrafish and mice to germ-free recipients reveal host habitat
-selection. Cell \textbf{127}:423--433.
-doi:\href{https://doi.org/10.1016/j.cell.2006.08.043}{10.1016/j.cell.2006.08.043}.
-\end{cslreferences}
-
-\newpage
-
-\includegraphics{../results/figures/figure_1.jpg}
-
-\textbf{Figure 1. Human fecal microbial communities established diverse
-gut bacterial communities in germ-free mice.} (A) Relative abundances of
-the 10 most abundant bacterial classes observed in the feces of
-previously germ-free C57Bl/6 mice 14 days post-colonization with human
-fecal samples (i.e., day 0 relative to \emph{C. difficile} challenge).
-Each column of abundances represents an individual mouse. Mice that
-received the same donor feces are grouped together and labeled above
-with a letter (N for non-moribund mice and M for moribund mice) and
-number (ordered by mean histopathologic score of the donor group). +
-indicates the mice which did not have detectable \emph{C. difficile} CFU
-(Figure 2). (B) Median (points) and interquartile range (lines) of
-\(\beta\)-diversity (\(\theta\)\textsubscript{YC}) between an individual
-mouse and either all others which were inoculated with feces from the
-same donor or from a different donor. The \(\beta\)-diversity among the
-same donor comparison group was significantly less than the
-\(\beta\)-diversity of the different donor group (\emph{P} \textless{}
-0.05, calculated by Wilcoxon rank sum test).
-
-\hfill\break
-
-\includegraphics{../results/figures/figure_2.jpg}
-
-\textbf{Figure 2. All donor groups resulted in \emph{C. difficile}
-infection but with different outcomes.} \emph{C. difficile} CFU per gram
-of stool was measured the day after challenge with 10\(^{3}\) \emph{C.
-difficile} RT027 clinical isolate 431 spores and at the end of the
-experiment, 10 days post-challenge. Each point represents an individual
-mouse. Mice are grouped by donor and labeled by the donor letter (N for
-non-moribund mice and M for moribund mice) and number (ordered by mean
-histopathologic score of the donor group). Points are colored by donor
-group. Mice from donor groups N1 through N6 succumbed to the infection
-prior to day 10 and were not plated on day 10 post-challenge. LOD =
-Limit of detection.
-
-\hfill\break
-
-\includegraphics{../results/figures/figure_3.jpg}
-
-\textbf{Figure 3. Histopathologic score and toxin activity varied across
-donor groups.} (A) Fecal toxin activity was detected in some mice post
-\emph{C. difficile} challenge in both moribund and non-moribund mice.
-(B) Cecum scored for histopathologic damage from mice at the end of the
-experiment. Samples were collected for histopathologic scoring on day 10
-post-challenge for non-moribund mice or the day the mouse succumbed to
-the infection for the moribund group (day 2 or 3 post-challenge). Each
-point represents an individual mouse. Mice are grouped by donor and
-labeled by the donor letter (N for non-moribund mice and M for moribund
-mice) and number (ordered by mean histopathologic score of the donor
-group). Points are colored by donor group. Mice in group N1 that have a
-summary score of 0 are the mice which did not have detectable \emph{C.
-difficile} CFU (Figure 2). Missing points are from mice that had
-insufficient fecal sample collected for assaying toxin or cecum for
-histopathologic scoring. LOD = Limit of detection.
-
-\hfill\break
-
-\includegraphics{../results/figures/figure_4.jpg}
-
-\textbf{Figure 4. Individual fecal bacterial community members of the
-murine gut associated with \emph{C .difficile} infection outcomes.} (A
-and B) Relative abundance of genera at the time of \emph{C. difficile}
-challenge (Day 0) that varied significantly by the moribundity and
-histopathologic summary score or detected toxin by LEfSe analysis.
-Median (points) and interquartile range (lines) are plotted. Genera are
-ordered alphabetically to ease comparisons across analyses. (A) Relative
-abundances were compared across infection outcome of moribund (colored
-black) or non-moribund with either a high histopathologic score (score
-greater than the median score of 5, colored green) or a low
-histopathologic summary score (score less than the median score of 5,
-colored light green). (B) Relative abundances were compared between mice
-which toxin activity was detected (Toxin +, colored dark purple) and
-which no toxin activity was detected (Toxin -, colored light purple).
-(C) Endpoint bacterial OTUs correlated with histopathologic summary
-score. Each individual mouse is plotted (transparent gray point).
-Spearman's correlations were statistically significant after
-Benjamini-Hochberg correction for multiple comparisons. All bacterial
-groups are ordered alphabetically. * indicates that the bacterial group
-was unclassified at lower taxonomic classification ranks.
-
-\hfill\break
-
-\includegraphics{../results/figures/figure_5.jpg}
-
-\textbf{Figure 5. Fecal bacterial community members of the murine gut at
-the time of \emph{C. difficile} infection predicted outcomes of the
-infection.} On the day of infection (Day 0), bacterial community members
-grouped by different classification rank were modeled with random forest
-to predict the infection outcome. The models used the highest taxonomic
-classification rank that performed as well as the lower ranks. Median
-(solid points) and interquartile range (lines) of the group relative
-abundance are plotted. Bacterial groups are ordered by their importance
-to the model; taxonomic group at the top of the plot had the greatest
-decrease in performance when its relative abundances were permuted. *
-indicates that the bacterial group was unclassified at lower taxonomic
-classification ranks. (A) Bacterial members grouped by phyla predicted
-which mice would have toxin activity detected at any point throughout
-the infection (Toxin +, dark purple). (B) Bacterial members grouped by
-class predicted which mice would become moribund (dark blue). (C)
-Bacterial members grouped by genera predicted if the mice would have a
-high (score greater than the median score of 5, colored dark green) or
-low (score less than the median score of 5, colored light green)
-histopathologic summary score.
-
-\hfill\break
-
-\includegraphics{../results/figures/figure_S1.jpg}
-
-\textbf{Figure S1. Histopathologic score of tissue damage at the
-endpoint of the infection.} Tissue collected at the endpoint, either day
-10 post-challenge (Non-moribund) or day mice succumbed to infection
-(Moribund), were scored from histopathologic damage. Each point
-represents an individual mouse. Mice (points) are grouped and colored by
-their human fecal community donor. Missing points are from mice that had
-insufficient sample for histopathologic scoring.
-
-\hfill\break
-
-\includegraphics{../results/figures/figure_S2.jpg}
-
-\textbf{Figure S2. Random forest models predicted outcomes of the
-\emph{C. difficile} challenge.} (A-C) Taxonomic classification rank
-model performance. Relative abundance at the time of \emph{C. difficile}
-challenge (Day 0) of the bacterial community members grouped by
-different classification rank were modeled with random forest to predict
-the infection outcome. The models used the highest taxonomic
-classification rank performed as well as the lower ranks. Black
-rectangle highlights classification rank used to model each outcome.
-(D-F) Model feature importance. Bacterial groups are ordered by their
-decrease in area under receiver-operator curve (AUC) when its relative
-abundances was permuted. Individual relative abundances were added to F
-since differences in AUC were outside the interquartile range. *
-indicates bacterial group was unclassified at lower taxonomic
-classification ranks. For all plots, median (solid points) and
-interquartile range (lines) are plotted. (A) Toxin production modeled
-which mice would have toxin detected during the experiment. (B)
-Moribundity modeled which mice would succumb to the infection prior to
-day 10 post-challenge. (C) Histopathologic score modeled which mice
-would have a high (score greater than the median score of 5) or low
-(score less than the median score of 5) histopathologic summary score.
-(D) Bacterial phyla which affected the performance of predicting
-detectable toxin activity when permuted. (E) Bacterial classes which
-affected the performance of predicting moribundity when permuted. (D)
-Bacterial genera which affected the performance of predicting
-histopathologic score when permuted.
-
-\end{document}

### 2022-04-22 submission/manuscript.tex
update manuscript w/revisions

@@
 \usepackage{lineno}
+\pagenumbering{arabic}
+\linenumbers
 
@@
 \floatplacement{figure}{H} % Keep the figure up top of the page
-\DeclareUnicodeCharacter{0301}{*************************************}
+\usepackage{setspace}
+\doublespacing
+\usepackage{lineno}
+\linenumbers
 \newlength{\cslhangindent}
@@
 \begin{document}
-
-\pagenumbering{arabic}
-\linenumbers
-\doublespacing
 
@@
 toxin genes have been associated with disease severity. However, it is
-unclear whether there is an association between members of the gut
-microbiota and disease severity. The gut microbiota is known to interact
+unclear whether specific members of the gut microbiota associate with
+variation in disease severity. The gut microbiota is known to interact
 with \emph{C. difficile} during infection. Perturbations to the gut
@@
 Generally, bacterial populations with pathogenic potential, such as
-\emph{Escherichia}, \emph{Helicobacter}, and \emph{Klebsiella}, were
+\emph{Enterococcus}, \emph{Helicobacter}, and \emph{Klebsiella}, were
 associated with more severe outcomes. Bacterial groups associated with
-fiber degradation, bile acid metabolism and lantibiotic production, such
-as \emph{Anaerostipes} and \emph{Coprobacillus}, were associated with
-less severe outcomes. These data indicate that, in addition to the host
-and \emph{C. difficile}, populations of gut bacteria can influence CDI
-disease severity.
+fiber degradation and bile acid metabolism, such as \emph{Anaerotignum},
+\emph{Blautia}, \emph{Lactonifactor}, and \emph{Monoglobus}, were
+associated with less severe outcomes. These data indicate that, in
+addition to the host and \emph{C. difficile} subtype, populations of gut
+bacteria can influence CDI disease severity.
 
@@
 community in many ways. The indigenous bacteria of a healthy intestinal
-community provide a protective barrier preventing \emph{C. difficile}
-from infecting the gut. A range of mechanisms can disrupt this barrier,
-including antibiotics, medications, or dietary changes, and lead to
-increased susceptibility to CDI (4--6). Once \emph{C. difficile}
-overcomes the protective barrier and colonizes the intestine, the
-indigenous bacteria can either promote or inhibit \emph{C. difficile}
-through producing molecules or modifying the environment (7, 8). Bile
-acids metabolized by the gut bacteria can inhibit \emph{C. difficile}
-growth and affect toxin production (9, 10). Bacteria in the gut also can
-compete more directly with \emph{C. difficile} through antibiotic
-production or nutrient consumption (11--13). While the relationship
-between the gut bacteria and \emph{C. difficile} has been established,
-the effect the gut bacteria can have on CDI disease severity is unclear.
+community prevent \emph{C. difficile} from infecting the gut (4). A
+range of mechanisms can disrupt this inhibition, including antibiotics,
+medications, or dietary changes, and lead to increased susceptibility to
+CDI (5--7). Once \emph{C. difficile} overcomes the inhibition and
+colonizes the intestine, the indigenous bacteria can either promote or
+inhibit \emph{C. difficile} through producing molecules or modifying the
+environment (8, 9). Bile acids metabolized by the gut bacteria can
+inhibit \emph{C. difficile} growth and affect toxin production (4, 10,
+11). Bacteria in the gut also can compete more directly with \emph{C.
+difficile} through antibiotic production or nutrient consumption
+(12--14). While the relationship between the gut bacteria and \emph{C.
+difficile} has been established, the effect the gut bacteria can have on
+CDI disease severity is unclear.
 
@@
 communities were challenged with a high-toxigenic strain resulted in
-varied disease severity (14) and when challenged with a low-toxigenic
+varied disease severity (15) and when challenged with a low-toxigenic
 strain members of the gut microbial community associated with variation
-in colonization (15). Here, we sought to further elucidate the
+in colonization (16). Here, we sought to further elucidate the
 relationship between members of the gut bacterial community and CDI
@@
 groups of gut bacteria affect the metabolism of \emph{C. difficile} and
-its infection dynamics, we can also identify groups of bacteria that
-affect the disease severity of the infection. To test this hypothesis,
-we colonized germ-free C57BL/6 mice with human fecal samples to create
-varied gut communities. We then challenged the mice with \emph{C.
-difficile} RT027 and followed the mice for the development of severe
-outcomes of moribundity and histopathologic cecal tissue damage. Since
-the murine host and \emph{C. difficile} isolate were the same and only
-the gut community varied, the variation in disease severity we observed
-was attributable to the gut microbiome.
+its clearance rate, specific groups of bacteria associate with variation
+in CDI disease severity. To test this hypothesis, we colonized germ-free
+C57BL/6 mice with human fecal samples to create varied gut communities.
+We then challenged the mice with \emph{C. difficile} RT027 and followed
+the mice for the development of severe outcomes of moribundity and
+histopathologic cecal tissue damage. Since the murine host and \emph{C.
+difficile} isolate were the same and only the gut community varied, the
+variation in disease severity we observed was attributable to the gut
+microbiome.
 
@@
 mouse colonies, we colonized germ-free mice with bacteria from human
-feces (16). We inoculated germ-free C57BL/6 mice with homogenized feces
+feces (17). We inoculated germ-free C57BL/6 mice with homogenized feces
 from each of 15 human fecal samples via oral gavage. These human fecal
 samples were selected because they represented diverse community
-structures based on community clustering (17). The gut communities were
-allowed to equilibrate for two weeks post-inoculation (18). We then
+structures based on community clustering (18). The gut communities were
+allowed to colonize for two weeks post-inoculation (19). We then
 surveyed the bacterial members of the gut communities by 16S rRNA gene
@@
 with antibiotics, such as clindamycin, to become susceptible to \emph{C.
-difficile} colonization (19, 20). However, we wanted to avoid modifying
+difficile} colonization (20, 21). However, we wanted to avoid modifying
 the gut communities with an antibiotic to maintain their unique
@@
 communities came from people at increased risk of CDI, such as recent
-hospitalization or antibiotic use (17), we tested whether \emph{C.
+hospitalization or antibiotic use (18), we tested whether \emph{C.
 difficile} was able to infect these mice without an antibiotic
@@
 (Figure 3). Overall, there was greater toxin activity detected in the
-stool of the moribund mice (\emph{P} = 0.003). However, when looking at
-each group of mice, we observed a range in toxin activity for both the
+stool of the moribund mice (Figure S1). However, when looking at each
+group of mice, we observed a range in toxin activity for both the
 moribund and non-moribund mice (Figure 3A). Non-moribund mice from
 Donors N2 and N5 through N9 had comparable toxin activity as the
-moribund mice. Additionally, not all moribund mice had toxin activity
-detected in their stool. Next, we examined the cecal tissue for
-histopathologic damage. Moribund mice had high levels of epithelial
-damage, tissue edema, and inflammation (Figure S1) similar to previously
-reported histopathologic findings for \emph{C. difficile} RT027 (21). As
-observed with toxin activity, the moribund mice had higher
-histopathologic scores than the non-moribund mice (\emph{P} \textless{}
-0.001). However, unlike the toxin activity, all moribund mice had
-consistently high histopathologic summary scores (Figure 3B). The
-non-moribund mice, Donor groups N1 through N9, had a range in tissue
+moribund mice at 2 days post-challenge. Additionally, not all moribund
+mice had toxin activity detected in their stool. Next, we examined the
+cecal tissue for histopathologic damage. Moribund mice had high levels
+of epithelial damage, tissue edema, and inflammation (Figure S2) similar
+to previously reported histopathologic findings for \emph{C. difficile}
+RT027 (22). As observed with toxin activity, the moribund mice had
+higher histopathologic scores than the non-moribund mice (\emph{P}
+\textless{} 0.001). However, unlike the toxin activity, all moribund
+mice had consistently high histopathologic summary scores (Figure 3B).
+The non-moribund mice, Donor groups N1 through N9, had a range in tissue
 damage from none detected to similar levels as the moribund mice, which
@@
 the variation in disease severity. We split the mice into groups by
-severity level based on their moribundity and histopathologic score. We
-dichotomized the histopathologic scores into high and low groups by
-splitting on the median score of 5. This analysis revealed 20 genera
-that were significantly different by the disease severity (Figure 4A).
-Bacterial genera \emph{Turicibacter}, \emph{Streptococcus},
-\emph{Staphylococcus}, \emph{Pseudomonas}, \emph{Phocaeicola},
-\emph{Parabacteroides}, \emph{Bacteroides}, and
-\emph{Escherichia/Shigella} were detected at higher relative abundances
-in the mice that became moribund. Populations of \emph{Anaerotignum},
-\emph{Coprobacillus}, \emph{Enterocloster}, and \emph{Murimonas} were
-more abundant in the non-moribund mice that would develop only low
-intestinal injury. To understand the role of toxin activity in disease
-severity, we applied LEfSe to identify the genera most likely to explain
-the differences between the presence and absence of detected toxin
-activity (Figure 4B). Many genera that associated with the presence of
-toxin were also associated with moribundity, such as populations of
-\emph{Escherichia/Shigella} and \emph{Bacteroides}. Likewise, there were
-genera such as \emph{Anaerotignum}, \emph{Enterocloster}, and
-\emph{Murimonas} that were associated with no detected toxin that also
+severity level based on moribundity or 10 days post infection (dpi)
+histopathologic score for non-moribund. This analysis revealed bacterial
+operational taxonomic units (OTUs) that were significantly different at
+the time of challenge by the disease severity (Figure 4A). OTUs
+associated with \emph{Akkermansia}, \emph{Bacteroides},
+\emph{Clostridium sensu stricto}, and \emph{Turicibacter} were detected
+at higher relative abundances in the mice that became moribund. OTUs
+associated with \emph{Anaerotignum}, \emph{Enterocloster}, and
+\emph{Murimonas} were more abundant in the non-moribund mice that would
+develop only low intestinal injury. To understand the role of toxin
+activity in disease severity, we applied LEfSe to identify the OTUs at
+the time of challenge that most likely explain the differences between
+communities that had toxin activity detected at anytime point to those
+that did not (Figure 4B). An OTU associated with \emph{Bacteroides}, OTU
+7, associated with the presence of toxin also associated with
+moribundity. Likewise, OTUs associated with \emph{Enterocloster} and
+\emph{Murimonas} that were associated with no detected toxin also
 exhibited greater relative abundance in communities from non-moribund
 mice with a low histopathologic score. Lastly, we tested for
-correlations between the endpoint relative abundances of bacterial
-operational taxonomic units (OTUs) and the histopathologic summary score
-(Figure 4C). The endpoint relative abundance of \emph{Bacteroides} was
-positively correlated with histopathologic score, as its day 0 relative
-abundance did with disease severity (Figure 4A). Populations of
-\emph{Klebsiella} and \emph{Prevotellaceae} were positively correlated
-with the histopathologic score and were increased in the group of mice
-with detectable toxin. This analysis identified bacterial genera that
-were associated with the variation in moribundity, histopathologic
-score, and toxin.
+correlations between the endpoint (10 dpi) relative abundances of OTUs
+and the histopathologic summary score (Figure 4C). The endpoint relative
+abundance of \emph{Bacteroides}, OTU 17, was positively correlated with
+histopathologic score, as its day 0 relative abundance did with disease
+severity (Figure 4A). A population of \emph{Bacteroides}, OTU 17, was
+positively correlated with the histopathologic score and were increased
+in the group of mice with detectable toxin. We also tested for
+correlations between the endpoint relative abundances of OTUs and toxin
+activity but none were significant. This analysis identified bacterial
+populations that were associated with the variation in moribundity,
+histopathologic score, and toxin.
 
@@
 and relative abundance could be predictive of the CDI disease outcome.
-We trained random forest models with bacterial community relative
+We trained logistic regression models with bacterial community relative
 abundance data from the day of colonization at each taxonomic rank to
-predict toxin, moribundity, and day 10 post-challenge histopathologic
-summary score. For predicting if detectable toxin would be produced,
-microbial populations aggregated by phylum rank classification performed
-similarly as models using lower taxonomic ranks (AUROC = 0.83, Figure
-S2). \emph{C. difficile} was more likely to produce detectable toxin
-when the community infected had less abundant populations of
-\emph{Verrucomicrobia} and \emph{Campilobacterota} and had more abundant
-populations of \emph{Proteobacteria} (Figure 5A). Next, we assessed the
-ability of the community to predict moribundity. Bacteria grouped by
-class rank classification was sufficient to predict which mice would
-succumb to the infection before the end of the experiment (AUROC = 0.91,
-Figure S2). The features with the greatest effect showed that
-communities with greater populations of bacteria belonging to
-\emph{Bacilli} and \emph{Firmicutes} and reduced populations of
-\emph{Erysipelotrichia} were more likely to result in moribundity
-(Figure 5B). Only one other class of bacteria was decreased in moribund
-mice, a group of unclassified \emph{Clostridia}. Lastly, the relative
-abundances of genera were able to predict a high or low histopathologic
-score (histopathologic scores were dichotomized as in previous analysis,
-AUROC = 0.99, Figure S2). No genera had a significantly greater effect
-on the model performance than any others, indicating the model was
-reliant on many genera for the correct prediction. The model used some
-of the genera identified in the LEfSe analysis, such as
-\emph{Coprobacillus}, \emph{Anaerostipes}, and \emph{Hungatella}.
-Communities with greater abundances of \emph{Hungatella},
-\emph{Eggerthella}, \emph{Bifidobacterium}, \emph{Duncaniella} and
-\emph{Neisseria} were more likely to have high histopathologic scores.
-These models have shown that the relative abundance of bacterial
-populations and their relationship to each other could be used to
-predict the variation in moribundity, histopathologic score, and
-detectable toxin of CDI.
+predict toxin, moribundity, and histopathologic summary score. For
+predicting if detectable toxin would be produced, microbial populations
+aggregated by genus rank classification performed similarly as models
+using lower taxonomic ranks (mean AUROC = 0.787, Figure S3). \emph{C.
+difficile} increased odds of producing detectable toxin when the
+community infected had less abundant populations of \emph{Monoglobus},
+\emph{Akkermansia}, \emph{Extibacter}, \emph{Intestinimonas} and
+\emph{Holdemania} and had more abundant populations of
+\emph{Lachnospiraceae} (Figure 5A). Next, we assessed the ability of the
+community to predict moribundity. Bacteria grouped by order rank
+classification was sufficient to predict which mice would succumb to the
+infection before the end of the experiment (mean AUROC = 0.9205, Figure
+S3). Many populations contributed to an increase odds of moribundity
+(Figure 5B). Populations related to \emph{Bifidobacteriales} and
+\emph{Clostridia} decreased the odds of a moribund outcome. Lastly, the
+relative abundances of OTUs were able to predict a high or low
+histopathologic score 10 dpi (histopathologic scores were dichotomized
+as in previous analysis, mean AUROC = 0.99, Figure S3). The model
+identified some similar OTUs as the LEfSe analysis, such as
+\emph{Murimonas} (OTU 48), \emph{Bacteroides} (OTU 7), and
+\emph{Hungatella} (OTU 24). These models have shown that the relative
+abundance of bacterial populations and their relationship to each other
+could be used to predict the variation in moribundity, histopathologic
+score, and detectable toxin of CDI.
 
@@
 have limited ability to control variation in the host, microbiome or
-\emph{C. difficile} ribotype (22). Murine experiments typically use a
+\emph{C. difficile} ribotype (23). Murine experiments typically use a
 single mouse colony and different \emph{C. difficile} ribotypes to
-create severity differences (23). Recently, our group has begun
+create severity differences (24). Recently, our group has begun
 uncovering the effect microbiome variation has on \emph{C. difficile}
@@
 mice from different mouse colonies resulted in different clearance rates
-of \emph{C. difficile} (15). We also showed varied ability of mice to
+of \emph{C. difficile} (16). We also showed varied ability of mice to
 spontaneously eliminate \emph{C. difficile} infection when they were
 treated with different antibiotics prior to \emph{C. difficile}
-challenge (24). Overall, the results presented here have demonstrated
+challenge (25). Overall, the results presented here have demonstrated
 that the gut bacterial community contributed to the severity of \emph{C.
@@
 severe infection using white blood cell counts, serum albumin level, or
-serum creatinine level (2, 25, 26). Those levels are driven by the
-activities in the intestine (27). Research into the drivers of this
+serum creatinine level (2, 26, 27). Those levels are driven by the
+activities in the intestine (28). Research into the drivers of this
 variation have revealed factors that make \emph{C. difficile} more
@@
 and production of the toxins TcdA, TcdB, and binary toxin and the
-prevalence in outbreaks, such as ribotypes 027 and 078 (19, 28--31).
+prevalence in outbreaks, such as ribotypes 027 and 078 (20, 29--32).
 However, other studies have shown that disease is not necessarily linked
-with toxin production (32) or the strain (33). Furthermore, there is
+with toxin production (33) or the strain (34). Furthermore, there is
 variation in the genome, growth rate, sporulation, germination, and
-toxin production in different isolates of a strain (34--37). This
+toxin production in different isolates of a strain (35--38). This
 variation may help explain why severe CDI prediction tools often miss
 identifying many patients with CDI that will develop severe disease (3,
-23, 38, 39). Therefore, it is necessary to gain a full understanding of
+24, 39, 40). Therefore, it is necessary to gain a full understanding of
 all factors contributing to disease variation to improve our ability to
@@
 community and is only able to become established after a perturbation
-(20). Once colonized, the different communities lead to different
+(21). Once colonized, the different communities lead to different
 metabolic responses and dynamics of the \emph{C. difficile} population
-(8, 24, 40). Gut bacteria metabolize primary bile acids into secondary
-bile acids (41, 42). The concentration of these bile acids affects
-germination, growth, toxin production and biofilm formation (9, 10, 43,
-44). Members of the bacterial community also affect other metabolites
+(9, 25, 41). Gut bacteria metabolize primary bile acids into secondary
+bile acids (4, 42, 43). The concentration of these bile acids affects
+germination, growth, toxin production and biofilm formation (10, 11, 44,
+45). Members of the bacterial community also affect other metabolites
 \emph{C. difficile} utilizes. \emph{Bacteroides thetaiotaomicron}
 produce sialidases which release sialic acid from the mucosa for
-\emph{C. difficile} to utilize (45, 46). The nutrient environment
-affects toxin production (47). Thus, many of the actions of the gut
+\emph{C. difficile} to utilize (46, 47). The nutrient environment
+affects toxin production (48). Thus, many of the actions of the gut
 bacteria modulate \emph{C. difficile} in ways that could affect the
@@
 with similar microbiota as a proxy for variation in disease, such as
-strain 630 for non-severe and RT027 for severe (19, 28, 29, 48). Studies
+strain 630 for non-severe and RT027 for severe (20, 29, 30, 49). Studies
 have also demonstrated variation in infection through tapering
-antibiotic dosage (20, 24, 49) or by reducing the amount of \emph{C.
-difficile} cells or spores used for the challenge (19, 49). These
+antibiotic dosage (21, 25, 50) or by reducing the amount of \emph{C.
+difficile} cells or spores used for the challenge (20, 50). These
 studies often either lack variation in the initial microbiome or have
@@
 challenged with \emph{C. difficile} resulted in varied outcomes of the
-infection (14, 15).
+infection (15, 16, 51).
 
@@
 tissue and mortality. This analysis revealed populations of
-\emph{Akkermansia}, \emph{Anaerostipes}, \emph{Coprobacillus},
+\emph{Akkermansia}, \emph{Anaerotignum}, \emph{Blautia},
 \emph{Enterocloster}, \emph{Lactonifactor}, and \emph{Monoglobus} were
@@
 histopathologic scores and no detected toxin. The protective role of
-these genera are supported by previous studies. \emph{Coprobacillus},
+these bacteria are supported by previous studies. \emph{Blautia},
 \emph{Lactonifactor}, and \emph{Monoglobus} have been shown to be
 involved in dietary fiber fermentation and associated with healthy
-communities (50--53). \emph{Anaerostipes} and \emph{Coprobacillus},
-which produce short chain fatty acids, have been associated with healthy
-communities (54--56). Furthermore, \emph{Coprobacillus}, which was
-abundant in mice with low histopathologic scores but rare in all other
-mice, has been shown to contain a putative type I lantibiotic gene
-cluster and inhibit \emph{C. difficile} colonization (57--59).
+communities (52--54). \emph{Anaerotignum}, which produce short chain
+fatty acids, has been associated with healthy communities (55, 56).
 \emph{Akkermansia} and \emph{Enterocloster} were also identified as more
@@
 contradictory supporting evidence in the current literature. In our
-data, \emph{Akkermansia} was most abundant in the non-moribund mice with
-low histopathologic scores but there were some moribund mice which had
-increased populations of \emph{Akkermansia}. This could be attributed to
-either a more protective mucus layer was present inhibiting colonization
-(59, 60) or mucus consumption by \emph{Akkermansia} could have been
-crossfeeding \emph{C. difficile} or exposing a niche for \emph{C.
-difficile} (61--63). Similarly, \emph{Enterocloster} was more abundant
-and associated with low histopathologic scores. It has been associated
-with healthy populations and has been used to mono-colonize germ-free
-mice to reduce the ability of \emph{C. difficile} to colonize (64, 65).
-However, \emph{Enterocloster} has also been involved in infections, such
-as bacteremia (66, 67). These data have exemplified populations of
-bacteria that have the potential to be either protective or harmful.
-Thus, the disease outcome is not likely based on the abundance of
-individual populations of bacteria, rather it is the result of the
-interactions of the community.
+data, a population of \emph{Akkermansia}, OTU 5, was most abundant in
+the non-moribund mice with low histopathologic scores but moribund mice
+had increased population of \emph{Akkermansia}, OTU 8. This difference
+could indicate either a more protective mucus layer was present
+inhibiting colonization (57, 58) or mucus consumption by
+\emph{Akkermansia} could have been crossfeeding \emph{C. difficile} or
+exposing a niche for \emph{C. difficile} (59--61). Similarly,
+\emph{Enterocloster} was more abundant and associated with low
+histopathologic scores. It has been associated with healthy populations
+and has been used to mono-colonize germ-free mice to reduce the ability
+of \emph{C. difficile} to colonize (62, 63). However,
+\emph{Enterocloster} has also been involved in infections, such as
+bacteremia (64, 65). These data have exemplified populations of bacteria
+that have the potential to be either protective or harmful. Thus, the
+disease outcome is not likely based on the abundance of individual
+populations of bacteria, rather it is the result of the interactions of
+the community.
 
@@
 community that also have been associated with disease, often referred to
-as opportunistic pathogens. Many of the populations with pathogenic
-potential that associated with worse outcomes are also facultative
-anaerobes. \emph{Enterococcus}, \emph{Klebsiella},
-\emph{Shigella/Escherichia}, \emph{Staphylococcus}, and
-\emph{Streptococcus} have been shown to expand after antibiotic use (17,
-68, 69) and are commonly detected in CDI cases (70--73). In addition to
-these populations, \emph{Eggerthella}, \emph{Prevotellaceae} and
-\emph{Helicobacter}, which associated with worse outcomes, have also
-been associated with intestinal inflammation (74--76). Recently,
-\emph{Helicobacter hepaticus} was shown to be sufficient to cause
-susceptibility to CDI in IL-10 deficient C57BL/6 mice (77). In our
-experiments, when \emph{Helicobacter} was present, the infection
-resulted in a high histopathologic score (Figure 4C). While we did not
-use IL-10 deficient mice, it is possible the bacterial community or host
-response are similarly modified by \emph{Helicobacter}, allowing
-\emph{C. difficile} infection and host damage. Aside from
+as opportunistic pathogens. Some of the populations of
+\emph{Bacteroides}, \emph{Enterococcus}, and \emph{Klebsiella} that
+associated with worse outcomes, have been shown to have pathogenic
+potential, expand after antibiotic use, and are commonly detected in CDI
+cases (66--69). In addition to these populations, \emph{Eggerthella},
+\emph{Prevotellaceae} and \emph{Helicobacter}, which associated with
+worse outcomes, have also been associated with intestinal inflammation
+(70--72). Recently, \emph{Helicobacter hepaticus} was shown to be
+sufficient to cause susceptibility to CDI in IL-10 deficient C57BL/6
+mice (73). In our experiments, when \emph{Helicobacter} was present, the
+infection resulted in a high histopathologic score (Figure 4C). While we
+did not use IL-10 deficient mice, it is possible the bacterial community
+or host response are similarly modified by \emph{Helicobacter}, allowing
+\emph{C. difficile} infection and host damage. These bacteria groups
+increased in severe outcomes maintained their differences throughout the
+length of the experiment (Figure S4). These results agreed Aside from
 \emph{Helicobacter}, these groups of bacteria that associated with more
@@
 severe disease it may be appropriate to apply the damage-response
-framework for microbial pathogenesis to CDI (78, 79). This framework
+framework for microbial pathogenesis to CDI (74, 75). This framework
 posits that disease is not driven by a single entity, rather it is an
@@
 background, they too were reacting to the specific microbial community.
-Disease severity is driven by the cumulative effect of the host immune
-response and the activity of \emph{C. difficile} and the gut bacteria.
-\emph{C. difficile} drives host damage through the production of toxin.
-The gut microbiota can modulate host damage through the balance of
-metabolic and competitive interactions with \emph{C. difficile}, such as
-bacteriocin production or mucin degradation, and interactions with the
-host, such as host mucus glycosylation or intestinal IL-33 expression
-(14, 80). For example, low levels of mucin degradation can provide
-nutrients to other community members producing a diverse non-damaging
-community (81). However, if mucin degradation becomes too great it
-reduces the protective function of the mucin layer and exposes the
-epithelial cells. This over-harvesting can contribute to the host damage
-due to other members producing toxin. Thus, the resultant intestinal
-damage is the balance of all activities in the gut environment. Host
-damage is the emergent property of numerous damage-response curves, such
-as one for host immune response, one for \emph{C. difficile} activity
-and another for microbiome community activity, each of which are a
-composite curve of the individual activities from each group, such as
-antibody production, neutrophil infiltration, toxin production,
-sporulation, fiber and mucin degradation. Therefore, while we have
-identified populations of interest, it may be necessary to target
-multiple types of bacteria to reduce the community interactions
-contributing to host damage.
+Different gut microbial communities can also have different effects on
+the host immune responses (76). Disease severity is driven by the
+cumulative effect of the host immune response and the activity of
+\emph{C. difficile} and the gut bacteria. \emph{C. difficile} drives
+host damage through the production of toxin. The gut microbiota can
+modulate host damage through the balance of metabolic and competitive
+interactions with \emph{C. difficile}, such as bacteriocin production or
+mucin degradation, and interactions with the host, such as host mucus
+glycosylation or intestinal IL-33 expression (15, 77). For example, low
+levels of mucin degradation can provide nutrients to other community
+members producing a diverse non-damaging community (78). However, if
+mucin degradation becomes too great it reduces the protective function
+of the mucin layer and exposes the epithelial cells. This
+over-harvesting can contribute to the host damage due to other members
+producing toxin. Thus, the resultant intestinal damage is the balance of
+all activities in the gut environment. Host damage is the emergent
+property of numerous damage-response curves, such as one for host immune
+response, one for \emph{C. difficile} activity and another for
+microbiome community activity, each of which are a composite curve of
+the individual activities from each group, such as antibody production,
+neutrophil infiltration, toxin production, sporulation, fiber and mucin
+degradation. Therefore, while we have identified populations of
+interest, it may be necessary to target multiple types of bacteria to
+reduce the community interactions contributing to host damage.
 
@@
 C57BL/6 were obtained from a single breeding colony in the University of
-Michigan Germ-free Mouse Core. Mice (N1 n=11, N2 n=7, N3 n=3, N4 n=3, N5
-n=3, N6 n=3, N7 n=7, N8 n=3, N9 n=2, M1 n=3, M2 n=3, M3 n=3, M4 n=3, M5
-n=7, M6 n=3) were housed in cages of 2-4 mice per cage and maintained in
+Michigan Germ-free Mouse Core. Mice (M1 n=3, M2 n=3, M3 n=3, M4 n=3, M5
+n=7, M6 n=3, N1 n=11, N2 n=7, N3 n=3, N4 n=3, N5 n=3, N6 n=3, N7 n=7, N8
+n=3, N9 n=2) were housed in cages of 2-4 mice per cage and maintained in
 germ-free isolators at the University of Michigan germ-free facility.
@@
 obtained as part of Schubert \emph{et al.} and selected based on
-community clusters (17) to result in diverse community structures. Feces
-were homogenized by mixing 200 mg of sample with 5 ml of PBS. Mice were
-inoculated with 100 \(\mu\)l of the fecal homogenate via oral gavage.
-Two weeks after the fecal community inoculation, mice were challenged
-with \emph{C. difficile}. \emph{C. difficile} clinical isolate 431 came
+community clusters (18) to result in diverse community structures (Table
+S1). Feces were homogenized by mixing 200 mg of sample with 5 ml of PBS.
+Mice were inoculated with 100 \(\mu\)l of the fecal homogenate via oral
+gavage. Two weeks after the fecal community inoculation, mice were
+challenged with \emph{C. difficile}. Stool samples from each mouse were
+collected one day prior to \emph{C. difficile} and plated for \emph{C.
+difficile} enumeration to confirm no \emph{C. difficile} was detected in
+stool prior to challenge. \emph{C. difficile} clinical isolate 431 came
 from Carlson \emph{et al.} which had previously been isolated and
-characterized (34, 35) and has recently been further characterized (36).
+characterized (35, 36) and has recently been further characterized (37).
 Spores concentration were determined both before and after challenge
-(82). \(10^{3}\) \emph{C. difficile} spores were given to each mouse via
+(79). \(10^{3}\) \emph{C. difficile} spores were given to each mouse via
 oral gavage.
@@
 for 24 hours. The plates were then counted for the number of colony
-forming units (CFU) (83).
+forming units (CFU) (80).
 
@@
 resulting amplicons using an Illumina MiSeq as described previously
-(84).
+(81).
 
 \textbf{Sequence curation.} Sequences were processed with
-mothur(v.1.44.3) as previously described (84, 85). In short, we used a
+mothur(v.1.44.3) as previously described (81, 82). In short, we used a
 3\% dissimilarity cutoff to group sequences into operational taxonomic
@@
 Database Project training set (version 18) to assign taxonomic
-classifications to each OTU (86). We sequenced a mock community of a
+classifications to each OTU (83). We sequenced a mock community of a
 known community composition and 16s rRNA gene sequences. We processed
@@
 Toxin activity was assessed using a Vero cell rounding-based
-cytotoxicity assay as described previously (29). The cytotoxicity titer
+cytotoxicity assay as described previously (30). The cytotoxicity titer
 was determined for each sample as the last dilution, which resulted in
@@
 criteria by a board-certified veterinary pathologist who was blinded to
-the experimental groups (29). Slides were scored as 0-4 for parameters
+the experimental groups (30). Slides were scored as 0-4 for parameters
 of epithelial damage, tissue edema, and inflammation and a summary score
 of 0-12 was generated by summing the three individual parameter scores.
+For non-moribund mice, histopathological summary scores used for LEfSe
+and logistic regression were split into high and low groups based on
+greater or less than the median summary score of 5 because the had a
+bimodal distribution (\emph{P} \textless{} 0.05).
 
@@
 structures, we calculated Yue and Clayton dissimilarity matrices
-(\(\theta\)\textsubscript{YC}) in mothur (87). We rarefied samples to
-2,107 sequences per sample to limit uneven sampling biases. We tested
-for differences in individual taxonomic groups that would explain the
-outcome differences with LEfSe (88) in mothur. Remaining statistical
-analysis and data visualization was performed in R (v4.0.5) with the
-tidyverse package (v1.3.1). We tested for significant differences in
-\(\beta\)-diversity (\(\theta\)\textsubscript{YC}) using the Wilcoxon
-rank sum test. We used Spearman's correlation to identify which OTUs
+(\(\theta\)\textsubscript{YC}) in mothur (84). For this calculation, we
+averaged of 1000 sub-samples, or rarified, samples to 2,107 sequence
+reads per sample to limit uneven sampling biases. We tested for
+differences in individual taxonomic groups that would explain the
+outcome differences with LEfSe (85) in mothur (default parameters, LDA
+\textgreater{} 4). We tested for differences in temporal trends through
+fitting a linear model to each OTU and testing for differences between
+histopathological summary scores with LEfSe (85) in mothur (default
+parameters, LDA \textgreater{} 3). Remaining statistical analysis and
+data visualization was performed in R (v4.0.5) with the tidyverse
+package (v1.3.1). We tested for significant differences in
+\(\beta\)-diversity (\(\theta\)\textsubscript{YC}), histopathological
+scores, and toxin activity using the Wilcoxon rank sum test,
+non-unimodality to non-moribund histopathological summary score using
+Hartigans' dip test, and toxin detection in mice using the Pearson's
+Chi-square test. We used Spearman's correlation to identify which OTUs
 that had a correlation between their relative abundance and the
@@
 multiple comparisons with a Benjamini and Hochberg adjustment for a type
-I error rate of 0.05 (89). We built random forest models using the
-mikropml package (90) with relative abundance summed by taxonomic ranks
-from day 0 samples using mtry values of 1 through 10, 15, 20, 25, 40,
-50, 100. The split for training and testing varied by model to avoid
-overfitting the data. To determine the optimal split, we tested splits
-(50\%, 60\%, 70\%, 80\%, 90\% data used for training) to find the
-greatest portion of data that could be used to train the model while
-still maintaining the same performance for the training model as the
-model with the held-out test data. The toxin and moribundity models were
-trained with 60\% of the data. The histopathologic score model was
-trained with 80\% of the data. Lastly, we did not compare murine
-communities to donor community or clinical data because germ-free mice
-colonized with non-murine fecal communities have been shown to more
-closely resemble the murine communities than the donor species community
-(91). Furthermore, it is not our intention to make any inferences
-regarding human associated bacteria and their relationship with human
-CDI outcome.
+I error rate of 0.05 (86). We built L2 logistic regression models using
+the mikropml package (87). Sequence counts were summed by taxonomic
+ranks from day 0 samples, normalized by centering to the feature mean
+and scaling by the standard deviation, and features positively or
+negatively correlated were collapsed into a single feature. We ran our
+models using alpha = 0 and lambda values of 1e-0, 1e1, 1e2, 2e2, 3e2,
+4e2, 5e2, 6e2, 7e2, 8e2, 9e2, 1e3, 1e4 with a split of 80\% of the data
+for training and 20\% of the data for testing. Lastly, we did not
+compare murine communities to donor community or clinical data because
+germ-free mice colonized with non-murine fecal communities have been
+shown to more closely resemble the murine communities than the donor
+species community (88). Furthermore, it is not our intention to make any
+inferences regarding human associated bacteria and their relationship
+with human CDI outcome.
 
@@
 
+Conceptualization: N.A.L., A.M.S., K.J.F., P.D.S.; Data curation:
+N.A.L., K.J.F.; Formal analysis: N.A.L., K.J.F., J.L.L., I.L.B.;
+Investigation: N.A.L., A.M.S., H.S., I.L.B., V.B.Y., P.D.S.;
+Methodology: N.A.L., A.M.S., K.J.F., J.L.L., H.S., I.L.B., V.B.Y.,
+P.D.S.; Resources: N.A.L., A.M.S., P.D.S.; Software: NAL; Visualization:
+N.A.L., K.J.F., P.D.S.; Writing - original draft: N.A.L.; Writing -
+review \& editing: N.A.L., A.M.S., K.J.F., J.L.L., H.S., I.L.B., V.B.Y.,
+P.D.S.; Funding acquisition: V.B.Y.; Project administration: P.D.S.;
+Supervision: P.D.S.
+
 \newpage
@@
 1. \textbf{Kelly CP}, \textbf{LaMont JT}. 2008. \emph{Clostridium
-difficile} more difficult than ever. New England Journal of Medicine
+difficile} --- more difficult than ever. New England Journal of Medicine
 \textbf{359}:1932--1940.
@@
 
+\leavevmode\hypertarget{ref-Buffie2014}{}%
+4. \textbf{Buffie CG}, \textbf{Bucci V}, \textbf{Stein RR},
+\textbf{McKenney PT}, \textbf{Ling L}, \textbf{Gobourne A}, \textbf{No
+D}, \textbf{Liu H}, \textbf{Kinnebrew M}, \textbf{Viale A},
+\textbf{Littmann E}, \textbf{Brink MRM van den}, \textbf{Jenq RR},
+\textbf{Taur Y}, \textbf{Sander C}, \textbf{Cross JR}, \textbf{Toussaint
+NC}, \textbf{Xavier JB}, \textbf{Pamer EG}. 2014. Precision microbiome
+reconstitution restores bile acid mediated resistance to
+\emph{Clostridium difficile}. Nature \textbf{517}:205--208.
+doi:\href{https://doi.org/10.1038/nature13828}{10.1038/nature13828}.
+
 \leavevmode\hypertarget{ref-Britton2014}{}%
-4. \textbf{Britton RA}, \textbf{Young VB}. 2014. Role of the intestinal
+5. \textbf{Britton RA}, \textbf{Young VB}. 2014. Role of the intestinal
 microbiota in resistance to colonization by \emph{Clostridium
@@
 \leavevmode\hypertarget{ref-Hryckowian2018}{}%
-5. \textbf{Hryckowian AJ}, \textbf{Treuren WV}, \textbf{Smits SA},
+6. \textbf{Hryckowian AJ}, \textbf{Treuren WV}, \textbf{Smits SA},
 \textbf{Davis NM}, \textbf{Gardner JO}, \textbf{Bouley DM},
@@
 \leavevmode\hypertarget{ref-VichVila2020}{}%
-6. \textbf{Vila AV}, \textbf{Collij V}, \textbf{Sanna S}, \textbf{Sinha
+7. \textbf{Vila AV}, \textbf{Collij V}, \textbf{Sanna S}, \textbf{Sinha
 T}, \textbf{Imhann F}, \textbf{Bourgonje AR}, \textbf{Mujagic Z},
@@
 \leavevmode\hypertarget{ref-Abbas2020}{}%
-7. \textbf{Abbas A}, \textbf{Zackular JP}. 2020. Microbe-microbe
+8. \textbf{Abbas A}, \textbf{Zackular JP}. 2020. Microbe-microbe
 interactions during \emph{Clostridioides difficile} infection. Current
@@
 \leavevmode\hypertarget{ref-Jenior2017}{}%
-8. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
+9. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
 \textbf{Schloss PD}. 2017. \emph{Clostridium difficile} colonizes
@@
 \leavevmode\hypertarget{ref-Sorg2008}{}%
-9. \textbf{Sorg JA}, \textbf{Sonenshein AL}. 2008. Bile salts and
+10. \textbf{Sorg JA}, \textbf{Sonenshein AL}. 2008. Bile salts and
 glycine as cogerminants for \emph{Clostridium difficile} spores. Journal
@@
 \leavevmode\hypertarget{ref-Thanissery2017}{}%
-10. \textbf{Thanissery R}, \textbf{Winston JA}, \textbf{Theriot CM}.
+11. \textbf{Thanissery R}, \textbf{Winston JA}, \textbf{Theriot CM}.
 2017. Inhibition of spore germination, growth, and toxin activity of
@@
 \leavevmode\hypertarget{ref-Aguirre2021}{}%
-11. \textbf{Aguirre AM}, \textbf{Yalcinkaya N}, \textbf{Wu Q},
+12. \textbf{Aguirre AM}, \textbf{Yalcinkaya N}, \textbf{Wu Q},
 \textbf{Swennes A}, \textbf{Tessier ME}, \textbf{Roberts P},
@@
 \leavevmode\hypertarget{ref-Kang2019}{}%
-12. \textbf{Kang JD}, \textbf{Myers CJ}, \textbf{Harris SC},
+13. \textbf{Kang JD}, \textbf{Myers CJ}, \textbf{Harris SC},
 \textbf{Kakiyama G}, \textbf{Lee I-K}, \textbf{Yun B-S},
@@
 \leavevmode\hypertarget{ref-Leslie2021}{}%
-13. \textbf{Leslie JL}, \textbf{Jenior ML}, \textbf{Vendrov KC},
+14. \textbf{Leslie JL}, \textbf{Jenior ML}, \textbf{Vendrov KC},
 \textbf{Standke AK}, \textbf{Barron MR}, \textbf{O'Brien TJ},
@@
 \leavevmode\hypertarget{ref-NagaoKitamoto2020}{}%
-14. \textbf{Nagao-Kitamoto H}, \textbf{Leslie JL}, \textbf{Kitamoto S},
+15. \textbf{Nagao-Kitamoto H}, \textbf{Leslie JL}, \textbf{Kitamoto S},
 \textbf{Jin C}, \textbf{Thomsson KA}, \textbf{Gillilland MG},
@@
 \leavevmode\hypertarget{ref-Tomkovich2020}{}%
-15. \textbf{Tomkovich S}, \textbf{Stough JMA}, \textbf{Bishop L},
+16. \textbf{Tomkovich S}, \textbf{Stough JMA}, \textbf{Bishop L},
 \textbf{Schloss PD}. 2020. The initial gut microbiota and response to
@@
 \leavevmode\hypertarget{ref-Nagpal2018}{}%
-16. \textbf{Nagpal R}, \textbf{Wang S}, \textbf{Woods LCS},
+17. \textbf{Nagpal R}, \textbf{Wang S}, \textbf{Woods LCS},
 \textbf{Seshie O}, \textbf{Chung ST}, \textbf{Shively CA},
@@
 \leavevmode\hypertarget{ref-Schubert2014}{}%
-17. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
+18. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
 \textbf{Mogle J}, \textbf{Petrosino JP}, \textbf{Young VB},
@@
 \leavevmode\hypertarget{ref-Gillilland2012}{}%
-18. \textbf{Gillilland MG}, \textbf{Erb-Downward JR}, \textbf{Bassis
+19. \textbf{Gillilland MG}, \textbf{Erb-Downward JR}, \textbf{Bassis
 CM}, \textbf{Shen MC}, \textbf{Toews GB}, \textbf{Young VB},
@@
 \leavevmode\hypertarget{ref-Chen2008}{}%
-19. \textbf{Chen X}, \textbf{Katchar K}, \textbf{Goldsmith JD},
+20. \textbf{Chen X}, \textbf{Katchar K}, \textbf{Goldsmith JD},
 \textbf{Nanthakumar N}, \textbf{Cheknis A}, \textbf{Gerding DN},
@@
 \leavevmode\hypertarget{ref-Schubert2015}{}%
-20. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
+21. \textbf{Schubert AM}, \textbf{Sinani H}, \textbf{Schloss PD}. 2015.
 Antibiotic-induced alterations of the murine gut microbiota and
@@
 \leavevmode\hypertarget{ref-Cowardin2016}{}%
-21. \textbf{Cowardin CA}, \textbf{Buonomo EL}, \textbf{Saleh MM},
+22. \textbf{Cowardin CA}, \textbf{Buonomo EL}, \textbf{Saleh MM},
 \textbf{Wilson MG}, \textbf{Burgess SL}, \textbf{Kuehne SA},
@@
 \leavevmode\hypertarget{ref-Seekatz2016}{}%
-22. \textbf{Seekatz AM}, \textbf{Rao K}, \textbf{Santhosh K},
+23. \textbf{Seekatz AM}, \textbf{Rao K}, \textbf{Santhosh K},
 \textbf{Young VB}. 2016. Dynamics of the fecal microbiome in patients
@@
 \leavevmode\hypertarget{ref-Dieterle2020}{}%
-23. \textbf{Dieterle MG}, \textbf{Putler R}, \textbf{Perry DA},
+24. \textbf{Dieterle MG}, \textbf{Putler R}, \textbf{Perry DA},
 \textbf{Menon A}, \textbf{Abernathy-Close L}, \textbf{Perlman NS},
@@
 \leavevmode\hypertarget{ref-Lesniak2021}{}%
-24. \textbf{Lesniak NA}, \textbf{Schubert AM}, \textbf{Sinani H},
+25. \textbf{Lesniak NA}, \textbf{Schubert AM}, \textbf{Sinani H},
 \textbf{Schloss PD}. 2021. Clearance of \emph{Clostridioides difficile}
@@
 \leavevmode\hypertarget{ref-Lungulescu2011}{}%
-25. \textbf{Lungulescu OA}, \textbf{Cao W}, \textbf{Gatskevich E},
+26. \textbf{Lungulescu OA}, \textbf{Cao W}, \textbf{Gatskevich E},
 \textbf{Tlhabano L}, \textbf{Stratidis JG}. 2011. CSI: A severity index
@@
 \leavevmode\hypertarget{ref-Zar2007}{}%
-26. \textbf{Zar FA}, \textbf{Bakkanagari SR}, \textbf{Moorthi KMLST},
+27. \textbf{Zar FA}, \textbf{Bakkanagari SR}, \textbf{Moorthi KMLST},
 \textbf{Davis MB}. 2007. A comparison of vancomycin and metronidazole
@@
 \leavevmode\hypertarget{ref-diMasi2018}{}%
-27. \textbf{Masi A di}, \textbf{Leboffe L}, \textbf{Polticelli F},
+28. \textbf{Masi A di}, \textbf{Leboffe L}, \textbf{Polticelli F},
 \textbf{Tonon F}, \textbf{Zennaro C}, \textbf{Caterino M}, \textbf{Stano
@@
 \leavevmode\hypertarget{ref-AbernathyClose2020}{}%
-28. \textbf{Abernathy-Close L}, \textbf{Dieterle MG}, \textbf{Vendrov
+29. \textbf{Abernathy-Close L}, \textbf{Dieterle MG}, \textbf{Vendrov
 KC}, \textbf{Bergin IL}, \textbf{Rao K}, \textbf{Young VB}. 2020. Aging
@@
 \leavevmode\hypertarget{ref-Theriot2011}{}%
-29. \textbf{Theriot CM}, \textbf{Koumpouras CC}, \textbf{Carlson PE},
+30. \textbf{Theriot CM}, \textbf{Koumpouras CC}, \textbf{Carlson PE},
 \textbf{Bergin II}, \textbf{Aronoff DM}, \textbf{Young VB}. 2011.
@@
 \leavevmode\hypertarget{ref-Goorhuis2008}{}%
-30. \textbf{Goorhuis A}, \textbf{Bakker D}, \textbf{Corver J},
+31. \textbf{Goorhuis A}, \textbf{Bakker D}, \textbf{Corver J},
 \textbf{Debast SB}, \textbf{Harmanus C}, \textbf{Notermans DW},
@@
 \leavevmode\hypertarget{ref-OConnor2009}{}%
-31. \textbf{O'Connor JR}, \textbf{Johnson S}, \textbf{Gerding DN}. 2009.
+32. \textbf{O'Connor JR}, \textbf{Johnson S}, \textbf{Gerding DN}. 2009.
 \emph{Clostridium difficile} infection caused by the epidemic
@@
 \leavevmode\hypertarget{ref-Rao2015}{}%
-32. \textbf{Rao K}, \textbf{Micic D}, \textbf{Natarajan M},
+33. \textbf{Rao K}, \textbf{Micic D}, \textbf{Natarajan M},
 \textbf{Winters S}, \textbf{Kiel MJ}, \textbf{Walk ST}, \textbf{Santhosh
@@
 \leavevmode\hypertarget{ref-Walk2012}{}%
-33. \textbf{Walk ST}, \textbf{Micic D}, \textbf{Jain R}, \textbf{Lo ES},
+34. \textbf{Walk ST}, \textbf{Micic D}, \textbf{Jain R}, \textbf{Lo ES},
 \textbf{Trivedi I}, \textbf{Liu EW}, \textbf{Almassalha LM},
@@
 \leavevmode\hypertarget{ref-Carlson2013}{}%
-34. \textbf{Carlson PE}, \textbf{Walk ST}, \textbf{Bourgis AET},
+35. \textbf{Carlson PE}, \textbf{Walk ST}, \textbf{Bourgis AET},
 \textbf{Liu MW}, \textbf{Kopliku F}, \textbf{Lo E}, \textbf{Young VB},
@@
 \leavevmode\hypertarget{ref-Carlson2015}{}%
-35. \textbf{Carlson PE}, \textbf{Kaiser AM}, \textbf{McColm SA},
+36. \textbf{Carlson PE}, \textbf{Kaiser AM}, \textbf{McColm SA},
 \textbf{Bauer JM}, \textbf{Young VB}, \textbf{Aronoff DM}, \textbf{Hanna
@@
 \leavevmode\hypertarget{ref-Saund2021}{}%
-36. \textbf{Saund K}, \textbf{Pirani A}, \textbf{Lacy B}, \textbf{Hanna
+37. \textbf{Saund K}, \textbf{Pirani A}, \textbf{Lacy B}, \textbf{Hanna
 PC}, \textbf{Snitkin ES}. 2021. Strain variation in \emph{Clostridioides
@@
 \leavevmode\hypertarget{ref-He2010}{}%
-37. \textbf{He M}, \textbf{Sebaihia M}, \textbf{Lawley TD},
+38. \textbf{He M}, \textbf{Sebaihia M}, \textbf{Lawley TD},
 \textbf{Stabler RA}, \textbf{Dawson LF}, \textbf{Martin MJ},
@@
 \leavevmode\hypertarget{ref-Butt2013}{}%
-38. \textbf{Butt E}, \textbf{Foster JA}, \textbf{Keedwell E},
+39. \textbf{Butt E}, \textbf{Foster JA}, \textbf{Keedwell E},
 \textbf{Bell JE}, \textbf{Titball RW}, \textbf{Bhangu A},
@@
 \leavevmode\hypertarget{ref-vanBeurden2017}{}%
-39. \textbf{Beurden YH van}, \textbf{Hensgens MPM}, \textbf{Dekkers OM},
+40. \textbf{Beurden YH van}, \textbf{Hensgens MPM}, \textbf{Dekkers OM},
 \textbf{Cessie SL}, \textbf{Mulder CJJ}, \textbf{Vandenbroucke-Grauls
@@
 \leavevmode\hypertarget{ref-Jenior2018}{}%
-40. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
+41. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
 \textbf{Schloss PD}. 2018. \emph{Clostridium difficile} alters the
@@
 \leavevmode\hypertarget{ref-Staley2016}{}%
-41. \textbf{Staley C}, \textbf{Weingarden AR}, \textbf{Khoruts A},
+42. \textbf{Staley C}, \textbf{Weingarden AR}, \textbf{Khoruts A},
 \textbf{Sadowsky MJ}. 2016. Interaction of gut microbiota with bile acid
@@
 \leavevmode\hypertarget{ref-Long2017}{}%
-42. \textbf{Long SL}, \textbf{Gahan CGM}, \textbf{Joyce SA}. 2017.
+43. \textbf{Long SL}, \textbf{Gahan CGM}, \textbf{Joyce SA}. 2017.
 Interactions between gut bacteria and bile in health and disease.
@@
 \leavevmode\hypertarget{ref-Sorg2010}{}%
-43. \textbf{Sorg JA}, \textbf{Sonenshein AL}. 2010. Inhibiting the
+44. \textbf{Sorg JA}, \textbf{Sonenshein AL}. 2010. Inhibiting the
 initiation of \emph{Clostridium difficile} spore germination using
@@
 \leavevmode\hypertarget{ref-Dubois2019}{}%
-44. \textbf{Dubois T}, \textbf{Tremblay YDN}, \textbf{Hamiot A},
+45. \textbf{Dubois T}, \textbf{Tremblay YDN}, \textbf{Hamiot A},
 \textbf{Martin-Verstraete I}, \textbf{Deschamps J}, \textbf{Monot M},
@@
 \leavevmode\hypertarget{ref-Ng2013}{}%
-45. \textbf{Ng KM}, \textbf{Ferreyra JA}, \textbf{Higginbottom SK},
+46. \textbf{Ng KM}, \textbf{Ferreyra JA}, \textbf{Higginbottom SK},
 \textbf{Lynch JB}, \textbf{Kashyap PC}, \textbf{Gopinath S},
@@
 \leavevmode\hypertarget{ref-Ferreyra2014}{}%
-46. \textbf{Ferreyra JA}, \textbf{Wu KJ}, \textbf{Hryckowian AJ},
+47. \textbf{Ferreyra JA}, \textbf{Wu KJ}, \textbf{Hryckowian AJ},
 \textbf{Bouley DM}, \textbf{Weimer BC}, \textbf{Sonnenburg JL}. 2014.
@@
 \leavevmode\hypertarget{ref-MartinVerstraete2016}{}%
-47. \textbf{Martin-Verstraete I}, \textbf{Peltier J}, \textbf{Dupuy B}.
+48. \textbf{Martin-Verstraete I}, \textbf{Peltier J}, \textbf{Dupuy B}.
 2016. The regulatory networks that control \emph{Clostridium difficile}
@@
 
+\leavevmode\hypertarget{ref-Battaglioli2018}{}%
+51. \textbf{Battaglioli EJ}, \textbf{Hale VL}, \textbf{Chen J},
+\textbf{Jeraldo P}, \textbf{Ruiz-Mojica C}, \textbf{Schmidt BA},
+\textbf{Rekdal VM}, \textbf{Till LM}, \textbf{Huq L}, \textbf{Smits SA},
+\textbf{Moor WJ}, \textbf{Jones-Hall Y}, \textbf{Smyrk T},
+\textbf{Khanna S}, \textbf{Pardi DS}, \textbf{Grover M}, \textbf{Patel
+R}, \textbf{Chia N}, \textbf{Nelson H}, \textbf{Sonnenburg JL},
+\textbf{Farrugia G}, \textbf{Kashyap PC}. 2018. \emph{Clostridioides
+difficile} uses amino acids associated with gut microbial dysbiosis in a
+subset of patients with diarrhea. Science Translational Medicine
+\textbf{10}.
+doi:\href{https://doi.org/10.1126/scitranslmed.aam7019}{10.1126/scitranslmed.aam7019}.
+
+\leavevmode\hypertarget{ref-Liu2021}{}%
+52. \textbf{Liu X}, \textbf{Mao B}, \textbf{Gu J}, \textbf{Wu J},
+\textbf{Cui S}, \textbf{Wang G}, \textbf{Zhao J}, \textbf{Zhang H},
+\textbf{Chen W}. 2021. \emph{Blautia} --- a new functional genus with
+potential probiotic properties? Gut Microbes \textbf{13}.
+doi:\href{https://doi.org/10.1080/19490976.2021.1875796}{10.1080/19490976.2021.1875796}.
+
 \leavevmode\hypertarget{ref-Mabrok2011}{}%
-50. \textbf{Mabrok HB}, \textbf{Klopfleisch R}, \textbf{Ghanem KZ},
+53. \textbf{Mabrok HB}, \textbf{Klopfleisch R}, \textbf{Ghanem KZ},
 \textbf{Clavel T}, \textbf{Blaut M}, \textbf{Loh G}. 2011. Lignan
@@
 \leavevmode\hypertarget{ref-Kim2019}{}%
-51. \textbf{Kim CC}, \textbf{Healey GR}, \textbf{Kelly WJ},
+54. \textbf{Kim CC}, \textbf{Healey GR}, \textbf{Kelly WJ},
 \textbf{Patchett ML}, \textbf{Jordens Z}, \textbf{Tannock GW},
@@
 
-\leavevmode\hypertarget{ref-doPrado2021}{}%
-52. \textbf{Prado SBR do}, \textbf{Minguzzi BT}, \textbf{Hoffmann C},
-\textbf{Fabi JP}. 2021. Modulation of human gut microbiota by dietary
-fibers from unripe and ripe papayas: Distinct polysaccharide degradation
-using a colonic in vitro fermentation model. Food Chemistry
-\textbf{348}:129071.
-doi:\href{https://doi.org/10.1016/j.foodchem.2021.129071}{10.1016/j.foodchem.2021.129071}.
-
-\leavevmode\hypertarget{ref-Muthuramalingam2019}{}%
-53. \textbf{Muthuramalingam K}, \textbf{Singh V}, \textbf{Choi C},
-\textbf{Choi SI}, \textbf{Kim YM}, \textbf{Unno T}, \textbf{Cho M}.
-2019. Dietary intervention using (1, 3)/(1, 6)-\(\beta\)-glucan, a
-fungus-derived soluble prebiotic ameliorates high-fat diet-induced
-metabolic distress and alters beneficially the gut microbiota in mice
-model. European Journal of Nutrition \textbf{59}:2617--2629.
-doi:\href{https://doi.org/10.1007/s00394-019-02110-5}{10.1007/s00394-019-02110-5}.
-
-\leavevmode\hypertarget{ref-Han2019}{}%
-54. \textbf{Han S-H}, \textbf{Yi J}, \textbf{Kim J-H}, \textbf{Lee S},
-\textbf{Moon H-W}. 2019. Composition of gut microbiota in patients with
-toxigenic \emph{Clostridioides} (\emph{Clostridium}) \emph{difficile}:
-Comparison between subgroups according to clinical criteria and toxin
-gene load. PLOS ONE \textbf{14}:e0212626.
-doi:\href{https://doi.org/10.1371/journal.pone.0212626}{10.1371/journal.pone.0212626}.
-
-\leavevmode\hypertarget{ref-Duncan2004}{}%
-55. \textbf{Duncan SH}, \textbf{Louis P}, \textbf{Flint HJ}. 2004.
-Lactate-utilizing bacteria, isolated from human feces, that produce
-butyrate as a major fermentation product. Applied and Environmental
-Microbiology \textbf{70}:5810--5817.
-doi:\href{https://doi.org/10.1128/aem.70.10.5810-5817.2004}{10.1128/aem.70.10.5810-5817.2004}.
-
-\leavevmode\hypertarget{ref-Ye2018}{}%
-56. \textbf{Ye J}, \textbf{Lv L}, \textbf{Wu W}, \textbf{Li Y},
-\textbf{Shi D}, \textbf{Fang D}, \textbf{Guo F}, \textbf{Jiang H},
-\textbf{Yan R}, \textbf{Ye W}, \textbf{Li L}. 2018. Butyrate protects
-mice against methionine--choline-deficient diet-induced non-alcoholic
-steatohepatitis by improving gut barrier function, attenuating
-inflammation and reducing endotoxin levels. Frontiers in Microbiology
-\textbf{9}.
-doi:\href{https://doi.org/10.3389/fmicb.2018.01967}{10.3389/fmicb.2018.01967}.
-
-\leavevmode\hypertarget{ref-Walsh2017}{}%
-57. \textbf{Walsh CJ}, \textbf{Guinane CM}, \textbf{O'Toole PW},
-\textbf{Cotter PD}. 2017. A profile hidden markov model to investigate
-the distribution and frequency of LanB-encoding lantibiotic modification
-genes in the human oral and gut microbiome. PeerJ \textbf{5}:e3254.
-doi:\href{https://doi.org/10.7717/peerj.3254}{10.7717/peerj.3254}.
-
-\leavevmode\hypertarget{ref-Sandiford2018}{}%
-58. \textbf{Sandiford SK}. 2018. Current developments in lantibiotic
-discovery for treating \emph{Clostridium difficile} infection. Expert
-Opinion on Drug Discovery \textbf{14}:71--79.
-doi:\href{https://doi.org/10.1080/17460441.2019.1549032}{10.1080/17460441.2019.1549032}.
+\leavevmode\hypertarget{ref-Choi2019}{}%
+55. \textbf{Choi S-H}, \textbf{Kim J-S}, \textbf{Park J-E}, \textbf{Lee
+KC}, \textbf{Eom MK}, \textbf{Oh BS}, \textbf{Yu SY}, \textbf{Kang SW},
+\textbf{Han K-I}, \textbf{Suh MK}, \textbf{Lee DH}, \textbf{Yoon H},
+\textbf{Kim B-Y}, \textbf{Lee JH}, \textbf{Lee JH}, \textbf{Lee J-S},
+\textbf{Park S-H}. 2019. \emph{Anaerotignum faecicola} sp. Nov.,
+isolated from human faeces. Journal of Microbiology
+\textbf{57}:1073--1078.
+doi:\href{https://doi.org/10.1007/s12275-019-9268-3}{10.1007/s12275-019-9268-3}.
+
+\leavevmode\hypertarget{ref-Ueki2017}{}%
+56. \textbf{Ueki A}, \textbf{Goto K}, \textbf{Ohtaki Y}, \textbf{Kaku
+N}, \textbf{Ueki K}. 2017. Description of \emph{Anaerotignum
+aminivorans} gen. Nov., sp. Nov., a strictly anaerobic,
+amino-acid-decomposing bacterium isolated from a methanogenic reactor,
+and reclassification of \emph{Clostridium propionicum},
+\emph{Clostridium neopropionicum} and \emph{Clostridium
+lactatifermentans} as species of the genus \emph{anaerotignum}.
+International Journal of Systematic and Evolutionary Microbiology
+\textbf{67}:4146--4153.
+doi:\href{https://doi.org/10.1099/ijsem.0.002268}{10.1099/ijsem.0.002268}.
 
 \leavevmode\hypertarget{ref-Stein2013}{}%
-59. \textbf{Stein RR}, \textbf{Bucci V}, \textbf{Toussaint NC},
+57. \textbf{Stein RR}, \textbf{Bucci V}, \textbf{Toussaint NC},
 \textbf{Buffie CG}, \textbf{Rätsch G}, \textbf{Pamer EG}, \textbf{Sander
@@
 \leavevmode\hypertarget{ref-Nakashima2021}{}%
-60. \textbf{Nakashima T}, \textbf{Fujii K}, \textbf{Seki T},
+58. \textbf{Nakashima T}, \textbf{Fujii K}, \textbf{Seki T},
 \textbf{Aoyama M}, \textbf{Azuma A}, \textbf{Kawasome H}. 2021. Novel
@@
 \leavevmode\hypertarget{ref-Geerlings2018}{}%
-61. \textbf{Geerlings S}, \textbf{Kostopoulos I}, \textbf{Vos W de},
+59. \textbf{Geerlings S}, \textbf{Kostopoulos I}, \textbf{Vos W de},
 \textbf{Belzer C}. 2018. \emph{Akkermansia muciniphila} in the human
@@
 \leavevmode\hypertarget{ref-Deng2018}{}%
-62. \textbf{Deng H}, \textbf{Yang S}, \textbf{Zhang Y}, \textbf{Qian K},
+60. \textbf{Deng H}, \textbf{Yang S}, \textbf{Zhang Y}, \textbf{Qian K},
 \textbf{Zhang Z}, \textbf{Liu Y}, \textbf{Wang Y}, \textbf{Bai Y},
@@
 \leavevmode\hypertarget{ref-Engevik2020}{}%
-63. \textbf{Engevik MA}, \textbf{Engevik AC}, \textbf{Engevik KA},
+61. \textbf{Engevik MA}, \textbf{Engevik AC}, \textbf{Engevik KA},
 \textbf{Auchtung JM}, \textbf{Chang-Graham AL}, \textbf{Ruan W},
@@
 \leavevmode\hypertarget{ref-Reeves2012}{}%
-64. \textbf{Reeves AE}, \textbf{Koenigsknecht MJ}, \textbf{Bergin IL},
+62. \textbf{Reeves AE}, \textbf{Koenigsknecht MJ}, \textbf{Bergin IL},
 \textbf{Young VB}. 2012. Suppression of \emph{Clostridium difficile} in
@@
 \leavevmode\hypertarget{ref-Ma2021}{}%
-65. \textbf{Ma L}, \textbf{Keng J}, \textbf{Cheng M}, \textbf{Pan H},
+63. \textbf{Ma L}, \textbf{Keng J}, \textbf{Cheng M}, \textbf{Pan H},
 \textbf{Feng B}, \textbf{Hu Y}, \textbf{Feng T}, \textbf{Yang F}. 2021.
@@
 \leavevmode\hypertarget{ref-Haas2020}{}%
-66. \textbf{Haas KN}, \textbf{Blanchard JL}. 2020. Reclassification of
+64. \textbf{Haas KN}, \textbf{Blanchard JL}. 2020. Reclassification of
 the \emph{Clostridium clostridioforme} and \emph{Clostridium sphenoides}
@@
 \leavevmode\hypertarget{ref-Finegold2005}{}%
-67. \textbf{Finegold SM}, \textbf{Song Y}, \textbf{Liu C}, \textbf{Hecht
+65. \textbf{Finegold SM}, \textbf{Song Y}, \textbf{Liu C}, \textbf{Hecht
 DW}, \textbf{Summanen P}, \textbf{Könönen E}, \textbf{Allen SD}. 2005.
@@
 
-\leavevmode\hypertarget{ref-VanInsberghe2020}{}%
-68. \textbf{VanInsberghe D}, \textbf{Elsherbini JA}, \textbf{Varian B},
-\textbf{Poutahidis T}, \textbf{Erdman S}, \textbf{Polz MF}. 2020.
-Diarrhoeal events can trigger long-term \emph{Clostridium difficile}
-colonization with recurrent blooms. Nature Microbiology
-\textbf{5}:642--650.
-doi:\href{https://doi.org/10.1038/s41564-020-0668-2}{10.1038/s41564-020-0668-2}.
-
-\leavevmode\hypertarget{ref-GarzaGonzalez2019}{}%
-69. \textbf{Garza-González E}, \textbf{Mendoza-Olazarán S},
-\textbf{Morfin-Otero R}, \textbf{Ramírez-Fontes A},
-\textbf{Rodríguez-Zulueta P}, \textbf{Flores-Treviño S},
-\textbf{Bocanegra-Ibarias P}, \textbf{Maldonado-Garza H},
-\textbf{Camacho-Ortiz A}. 2019. Intestinal microbiome changes in fecal
-microbiota transplant (FMT) vs. FMT enriched with \emph{Lactobacillus}
-in the treatment of recurrent \emph{Clostridioides difficile} infection.
-Canadian Journal of Gastroenterology and Hepatology \textbf{2019}:1--7.
-doi:\href{https://doi.org/10.1155/2019/4549298}{10.1155/2019/4549298}.
-
-\leavevmode\hypertarget{ref-Shafiq2020}{}%
-70. \textbf{Shafiq M}, \textbf{Alturkmani H}, \textbf{Zafar Y},
-\textbf{Mittal V}, \textbf{Lodhi H}, \textbf{Ullah W}, \textbf{Brewer
-J}. 2020. Effects of co-infection on the clinical outcomes of
-\emph{Clostridium difficile} infection. Gut Pathogens \textbf{12}.
-doi:\href{https://doi.org/10.1186/s13099-020-00348-7}{10.1186/s13099-020-00348-7}.
+\leavevmode\hypertarget{ref-Tomkovich2021}{}%
+66. \textbf{Tomkovich S}, \textbf{Taylor A}, \textbf{King J},
+\textbf{Colovas J}, \textbf{Bishop L}, \textbf{McBride K},
+\textbf{Royzenblat S}, \textbf{Lesniak NA}, \textbf{Bergin IL},
+\textbf{Schloss PD}. 2021. An osmotic laxative renders mice susceptible
+to prolonged \emph{Clostridioides difficile} colonization and hinders
+clearance. mSphere \textbf{6}.
+doi:\href{https://doi.org/10.1128/msphere.00629-21}{10.1128/msphere.00629-21}.
 
 \leavevmode\hypertarget{ref-Keith2020}{}%
-71. \textbf{Keith JW}, \textbf{Dong Q}, \textbf{Sorbara MT},
+67. \textbf{Keith JW}, \textbf{Dong Q}, \textbf{Sorbara MT},
 \textbf{Becattini S}, \textbf{Sia JK}, \textbf{Gjonbalaj M},
@@
 \leavevmode\hypertarget{ref-Zackular2016}{}%
-72. \textbf{Zackular JP}, \textbf{Moore JL}, \textbf{Jordan AT},
+68. \textbf{Zackular JP}, \textbf{Moore JL}, \textbf{Jordan AT},
 \textbf{Juttukonda LJ}, \textbf{Noto MJ}, \textbf{Nicholson MR},
@@
 \leavevmode\hypertarget{ref-Berkell2021}{}%
-73. \textbf{Berkell M}, \textbf{Mysara M}, \textbf{Xavier BB},
+69. \textbf{Berkell M}, \textbf{Mysara M}, \textbf{Xavier BB},
 \textbf{Werkhoven CH van}, \textbf{Monsieurs P}, \textbf{Lammens C},
@@
 \leavevmode\hypertarget{ref-Gardiner2014}{}%
-74. \textbf{Gardiner BJ}, \textbf{Tai AY}, \textbf{Kotsanas D},
+70. \textbf{Gardiner BJ}, \textbf{Tai AY}, \textbf{Kotsanas D},
 \textbf{Francis MJ}, \textbf{Roberts SA}, \textbf{Ballard SA},
@@
 \leavevmode\hypertarget{ref-Iljazovic2020}{}%
-75. \textbf{Iljazovic A}, \textbf{Roy U}, \textbf{Gálvez EJC},
+71. \textbf{Iljazovic A}, \textbf{Roy U}, \textbf{Gálvez EJC},
 \textbf{Lesker TR}, \textbf{Zhao B}, \textbf{Gronow A}, \textbf{Amend
@@
 \leavevmode\hypertarget{ref-Nagalingam2013}{}%
-76. \textbf{Nagalingam NA}, \textbf{Robinson CJ}, \textbf{Bergin IL},
+72. \textbf{Nagalingam NA}, \textbf{Robinson CJ}, \textbf{Bergin IL},
 \textbf{Eaton KA}, \textbf{Huffnagle GB}, \textbf{Young VB}. 2013. The
@@
 \leavevmode\hypertarget{ref-AbernathyClose2021}{}%
-77. \textbf{Abernathy-Close L}, \textbf{Barron MR}, \textbf{George JM},
+73. \textbf{Abernathy-Close L}, \textbf{Barron MR}, \textbf{George JM},
 \textbf{Dieterle MG}, \textbf{Vendrov KC}, \textbf{Bergin IL},
@@
 \leavevmode\hypertarget{ref-Pirofski2008}{}%
-78. \textbf{Pirofski L-a}, \textbf{Casadevall A}. 2008. The
+74. \textbf{Pirofski L-a}, \textbf{Casadevall A}. 2008. The
 damage-response framework of microbial pathogenesis and infectious
@@
 \leavevmode\hypertarget{ref-Casadevall2014}{}%
-79. \textbf{Casadevall A}, \textbf{Pirofski L-a}. 2014. What is a host?
+75. \textbf{Casadevall A}, \textbf{Pirofski L-a}. 2014. What is a host?
 Incorporating the microbiota into the damage-response framework.
@@
 
+\leavevmode\hypertarget{ref-Lundberg2020}{}%
+76. \textbf{Lundberg R}, \textbf{Toft MF}, \textbf{Metzdorff SB},
+\textbf{Hansen CHF}, \textbf{Licht TR}, \textbf{Bahl MI}, \textbf{Hansen
+AK}. 2020. Human microbiota-transplanted C57BL/6 mice and offspring
+display reduced establishment of key bacteria and reduced immune
+stimulation compared to mouse microbiota-transplantation. Scientific
+Reports \textbf{10}.
+doi:\href{https://doi.org/10.1038/s41598-020-64703-z}{10.1038/s41598-020-64703-z}.
+
 \leavevmode\hypertarget{ref-Frisbee2019}{}%
-80. \textbf{Frisbee AL}, \textbf{Saleh MM}, \textbf{Young MK},
+77. \textbf{Frisbee AL}, \textbf{Saleh MM}, \textbf{Young MK},
 \textbf{Leslie JL}, \textbf{Simpson ME}, \textbf{Abhyankar MM},
@@
 \leavevmode\hypertarget{ref-Tailford2015}{}%
-81. \textbf{Tailford LE}, \textbf{Crost EH}, \textbf{Kavanaugh D},
+78. \textbf{Tailford LE}, \textbf{Crost EH}, \textbf{Kavanaugh D},
 \textbf{Juge N}. 2015. Mucin glycan foraging in the human gut
@@
 \leavevmode\hypertarget{ref-Sorg2009}{}%
-82. \textbf{Sorg JA}, \textbf{Dineen SS}. 2009. Laboratory maintenance
+79. \textbf{Sorg JA}, \textbf{Dineen SS}. 2009. Laboratory maintenance
 of \emph{Clostridium difficile}. Current Protocols in Microbiology
@@
 \leavevmode\hypertarget{ref-Winston2016}{}%
-83. \textbf{Winston JA}, \textbf{Thanissery R}, \textbf{Montgomery SA},
+80. \textbf{Winston JA}, \textbf{Thanissery R}, \textbf{Montgomery SA},
 \textbf{Theriot CM}. 2016. Cefoperazone-treated mouse model of
@@
 \leavevmode\hypertarget{ref-Kozich2013}{}%
-84. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
+81. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
 \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
@@
 \leavevmode\hypertarget{ref-Schloss2009}{}%
-85. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+82. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
 \textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
@@
 \leavevmode\hypertarget{ref-Wang2007}{}%
-86. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
+83. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
 \textbf{Cole JR}. 2007. Naïve bayesian classifier for rapid assignment
@@
 \leavevmode\hypertarget{ref-Yue2005}{}%
-87. \textbf{Yue JC}, \textbf{Clayton MK}. 2005. A similarity measure
+84. \textbf{Yue JC}, \textbf{Clayton MK}. 2005. A similarity measure
 based on species proportions. Communications in Statistics - Theory and
@@
 \leavevmode\hypertarget{ref-Segata2011}{}%
-88. \textbf{Segata N}, \textbf{Izard J}, \textbf{Waldron L},
+85. \textbf{Segata N}, \textbf{Izard J}, \textbf{Waldron L},
 \textbf{Gevers D}, \textbf{Miropolsky L}, \textbf{Garrett WS},
@@
 \leavevmode\hypertarget{ref-Benjamini1995}{}%
-89. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
+86. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
 false discovery rate: A practical and powerful approach to multiple
@@
 \leavevmode\hypertarget{ref-Topcuoglu2021}{}%
-90. \textbf{Topçuoğlu B}, \textbf{Lapp Z}, \textbf{Sovacool K},
+87. \textbf{Topçuoğlu B}, \textbf{Lapp Z}, \textbf{Sovacool K},
 \textbf{Snitkin E}, \textbf{Wiens J}, \textbf{Schloss P}. 2021.
@@
 \leavevmode\hypertarget{ref-Rawls2006}{}%
-91. \textbf{Rawls JF}, \textbf{Mahowald MA}, \textbf{Ley RE},
+88. \textbf{Rawls JF}, \textbf{Mahowald MA}, \textbf{Ley RE},
 \textbf{Gordon JI}. 2006. Reciprocal gut microbiota transplants from
@@
 \newpage
-
-\includegraphics{../results/figures/figure_1.jpg}
 
@@
 same donor comparison group was significantly less than the
-\(\beta\)-diversity of the different donor group (\emph{P} \textless{}
-0.05, calculated by Wilcoxon rank sum test).
+\(\beta\)-diversity of either the different donor group or the donor
+community (\emph{P} \textless{} 0.05, calculated by Wilcoxon rank sum
+test).
 
 \hfill\break
-
-\includegraphics{../results/figures/figure_2.jpg}
 
@@
 prior to day 10 and were not plated on day 10 post-challenge. LOD =
-Limit of detection.
+Limit of detection. -Deceased- indicates mice were deceased at that time
+point so no sample was available.
 
 \hfill\break
-
-\includegraphics{../results/figures/figure_3.jpg}
 
@@
 insufficient fecal sample collected for assaying toxin or cecum for
-histopathologic scoring. LOD = Limit of detection.
+histopathologic scoring. * indicates significant difference between
+non-moribund and moribund groups of mice by Wilcoxon test (\emph{P}
+\textless{} 0.002). LOD = Limit of detection. -Deceased- indicates mice
+were deceased at that time point so no sample was available.
 
 \hfill\break
-
-\includegraphics{../results/figures/figure_4.jpg}
 
@@
 murine gut associated with \emph{C .difficile} infection outcomes.} (A
-and B) Relative abundance of genera at the time of \emph{C. difficile}
+and B) Relative abundance of OTUs at the time of \emph{C. difficile}
 challenge (Day 0) that varied significantly by the moribundity and
 histopathologic summary score or detected toxin by LEfSe analysis.
-Median (points) and interquartile range (lines) are plotted. Genera are
-ordered alphabetically to ease comparisons across analyses. (A) Relative
-abundances were compared across infection outcome of moribund (colored
-black) or non-moribund with either a high histopathologic score (score
-greater than the median score of 5, colored green) or a low
+Median (points) and interquartile range (lines) are plotted. (A) Day 0
+relative abundances were compared across infection outcome of moribund
+(colored black) or non-moribund with either a high histopathologic score
+(score greater than the median score of 5, colored green) or a low
 histopathologic summary score (score less than the median score of 5,
-colored light green). (B) Relative abundances were compared between mice
-which toxin activity was detected (Toxin +, colored dark purple) and
-which no toxin activity was detected (Toxin -, colored light purple).
-(C) Endpoint bacterial OTUs correlated with histopathologic summary
-score. Each individual mouse is plotted (transparent gray point).
-Spearman's correlations were statistically significant after
-Benjamini-Hochberg correction for multiple comparisons. All bacterial
-groups are ordered alphabetically. * indicates that the bacterial group
-was unclassified at lower taxonomic classification ranks.
+colored light green). (B) Day 0 relative abundances were compared
+between mice which toxin activity was detected (Toxin +, colored dark
+purple) and which no toxin activity was detected (Toxin -, colored light
+purple). (C) Day 10 bacterial OTU relative abundances correlated with
+histopathologic summary score. Each individual mouse is plotted and
+colored according to their categorization in panel A. Points at the
+median score of 5 (gray points) were not included in panel A. Spearman's
+correlations were statistically significant after Benjamini-Hochberg
+correction for multiple comparisons. All bacterial groups are ordered by
+the LDA score. * indicates that the bacterial group was unclassified at
+lower taxonomic classification ranks.
 
 \hfill\break
-
-\includegraphics{../results/figures/figure_5.jpg}
 
@@
 infection.} On the day of infection (Day 0), bacterial community members
-grouped by different classification rank were modeled with random forest
-to predict the infection outcome. The models used the highest taxonomic
-classification rank that performed as well as the lower ranks. Median
-(solid points) and interquartile range (lines) of the group relative
-abundance are plotted. Bacterial groups are ordered by their importance
-to the model; taxonomic group at the top of the plot had the greatest
-decrease in performance when its relative abundances were permuted. *
-indicates that the bacterial group was unclassified at lower taxonomic
-classification ranks. (A) Bacterial members grouped by phyla predicted
-which mice would have toxin activity detected at any point throughout
-the infection (Toxin +, dark purple). (B) Bacterial members grouped by
-class predicted which mice would become moribund (dark blue). (C)
-Bacterial members grouped by genera predicted if the mice would have a
-high (score greater than the median score of 5, colored dark green) or
-low (score less than the median score of 5, colored light green)
-histopathologic summary score.
+grouped by different classification rank were modeled with logistic
+regression to predict the infection outcome. The models used the highest
+taxonomic classification rank without a decrease in performance. Models
+used all community members but plotted are those members with a mean
+odds ratio not equal to 1. Median (solid points) and interquartile range
+(lines) of the group relative abundance are plotted. Bacterial groups
+are ordered by their odds ratio. * indicates that the bacterial group
+was unclassified at lower taxonomic classification ranks. (A) Bacterial
+members grouped by genus predicted which mice would have toxin activity
+detected at any point throughout the infection (Toxin +, dark purple).
+(B) Bacterial members grouped by order predicted which mice would become
+moribund (dark blue). (C) Bacterial members grouped by OTU predicted if
+the mice would have a high (score greater than the median score of 5,
+colored dark green) or low (score less than the median score of 5,
+colored light green) histopathologic summary score.
 
@@
 
-\includegraphics{../results/figures/figure_S1.jpg}
-
-\textbf{Figure S1. Histopathologic score of tissue damage at the
+\textbf{Figure S1. Toxin detect in mice based on outcome of the
+infection.} Comparison of the distribution of number of either
+non-moribund or moribund mice which toxin was detected in the first
+three days post infection. Bars are colored by whether toxin was
+detected in stool from the mouse (dark purple) or not (light purple).
+Moribund mice had significantly more mice with toxin detected (\emph{P}
+\textless{} 0.008) by Pearson's Chi-square test.
+
+\hfill\break
+
+\textbf{Figure S2. Histopathologic score of tissue damage at the
 endpoint of the infection.} Tissue collected at the endpoint, either day
@@
 their human fecal community donor. Missing points are from mice that had
-insufficient sample for histopathologic scoring.
+insufficient sample for histopathologic scoring. * indicates significant
+difference between non-moribund and moribund groups of mice by Wilcoxon
+test (\emph{P} \textless{} 0.002).
 
@@
 
-\includegraphics{../results/figures/figure_S2.jpg}
-
-\textbf{Figure S2. Random forest models predicted outcomes of the
+\textbf{Figure S3. Logistic regression models predicted outcomes of the
 \emph{C. difficile} challenge.} (A-C) Taxonomic classification rank
@@
 classification rank performed as well as the lower ranks. Black
-rectangle highlights classification rank used to model each outcome.
-(D-F) Model feature importance. Bacterial groups are ordered by their
-decrease in area under receiver-operator curve (AUC) when its relative
-abundances was permuted. Individual relative abundances were added to F
-since differences in AUC were outside the interquartile range. *
-indicates bacterial group was unclassified at lower taxonomic
-classification ranks. For all plots, median (solid points) and
-interquartile range (lines) are plotted. (A) Toxin production modeled
-which mice would have toxin detected during the experiment. (B)
-Moribundity modeled which mice would succumb to the infection prior to
-day 10 post-challenge. (C) Histopathologic score modeled which mice
-would have a high (score greater than the median score of 5) or low
-(score less than the median score of 5) histopathologic summary score.
-(D) Bacterial phyla which affected the performance of predicting
-detectable toxin activity when permuted. (E) Bacterial classes which
-affected the performance of predicting moribundity when permuted. (D)
-Bacterial genera which affected the performance of predicting
-histopathologic score when permuted.
+rectangle highlights classification rank used to model each outcome. For
+all plots, median (large solid points), interquartile range (lines), and
+individual models (small transparent points) are plotted. (A) Toxin
+production modeled which mice would have toxin detected during the
+experiment. (B) Moribundity modeled which mice would succumb to the
+infection prior to day 10 post-challenge. (C) Histopathologic score
+modeled which mice would have a high (score greater than the median
+score of 5) or low (score less than the median score of 5)
+histopathologic summary score.
+
+\hfill\break
+
+\textbf{Figure S4. Temporal dynamics of OTUs that differed between
+histopathologic summary score.} Relative abundance of OTUs on each day
+relative to the time of \emph{C. difficile} challenge (Day 0) that have
+a significantly different temporal trend by the histopathologic summary
+score by LEfSe analysis. Median (points) and interquartile range (lines)
+are plotted. Points and lines are colored by infection outcome of
+moribund (colored black) or non-moribund with either a high
+histopathologic score (score greater than the median score of 5, colored
+green) or a low histopathologic summary score (score less than the
+median score of 5, colored light green).
 

### 2022-04-22 code/fix_con2_label.sh

@@ -1,10 +1,0 @@
-mv data/raw/CON3-2328-D10_S139_L001_R1_001.fastq data/raw/CON2-2328-D10_S139_L001_R1_001.fastq
-mv data/raw/CON3-2328-D10_S139_L001_R2_001.fastq data/raw/CON2-2328-D10_S139_L001_R2_001.fastq
-mv data/raw/CON3-2328-D7_S106_L001_R1_001.fastq data/raw/CON2-2328-D7_S106_L001_R1_001.fastq
-mv data/raw/CON3-2328-D7_S106_L001_R2_001.fastq data/raw/CON2-2328-D7_S106_L001_R2_001.fastq
-mv data/raw/CON3-2328-D9_S127_L001_R1_001.fastq data/raw/CON2-2328-D9_S127_L001_R1_001.fastq
-mv data/raw/CON3-2328-D9_S127_L001_R2_001.fastq data/raw/CON2-2328-D9_S127_L001_R2_001.fastq
-mv data/raw/CON3-2376-D10_S140_L001_R1_001.fastq data/raw/CON2-2376-D10_S140_L001_R1_001.fastq
-mv data/raw/CON3-2376-D10_S140_L001_R2_001.fastq data/raw/CON2-2376-D10_S140_L001_R2_001.fastq
-mv data/raw/CON3-2376-D9_S128_L001_R1_001.fastq data/raw/CON2-2376-D9_S128_L001_R1_001.fastq
-mv data/raw/CON3-2376-D9_S128_L001_R2_001.fastq data/raw/CON2-2376-D9_S128_L001_R2_001.fastq

### 2022-04-22 code/ml/run_ml.R

@@ -67,44 +67,42 @@ day_0_moribund <- setup_ml_df(day_0_moribund, 'early_euth')
 day_0_histology <- setup_ml_df(day_0_histology, 'hist_score')
 
 
-# run random forest
-new_hp <- list(mtry = c(1:10, 15, 20, 25, 40, 50, 100))
-print('Running Random Forest on day 0 to predict toxin production')
-day_0_predict_future_toxin_rf <- run_ml(day_0_predict_future_toxin,
-	   'rf',
+## run logistic regression
+fraction <- 0.8
+new_hp <- list(alpha = 0,
+			   lambda = c(1e-0, 1e1, 1e2, 2e2, 3e2, 4e2, 5e2, 6e2, 7e2, 8e2, 9e2, 1e3, 1e4))
+print('Running Logistic Regression on day 0 to predict toxin production')
+day_0_predict_future_toxin_lr <- run_ml(day_0_predict_future_toxin,
+	   'glmnet',
        outcome_colname = 'toxin',
-       training_frac = 0.6,
+       training_frac = fraction,
 	   hyperparameters = new_hp,
 	   find_feature_importance = TRUE,
        seed = current_seed)
-print('Running Random Forest on day 0 to predict severity')
-day_0_moribund_rf <- run_ml(day_0_moribund,
-	   'rf',
+new_hp <- list(alpha = 0,
+			   lambda = c(1e-0, 1e1, 1e2, 2e2, 3e2, 4e2, 5e2, 6e2, 7e2, 8e2, 9e2, 1e3, 1e4))
+print('Running Logistic Regression on day 0 to predict severity')
+day_0_moribund_lr <- run_ml(day_0_moribund,
+	   'glmnet',
        outcome_colname = 'early_euth',
-       training_frac = 0.6,
+       training_frac = fraction,
 	   hyperparameters = new_hp,
 	   find_feature_importance = TRUE,
        seed = current_seed)
-print('Running Random Forest on day 0 to predict day 10 severity')
-day_0_histology_rf_80 <- run_ml(day_0_histology,
-	   'rf',
+new_hp <- list(alpha = 0,
+			   lambda = c(1e-0, 1e1, 1e2, 2e2, 3e2, 4e2, 5e2, 6e2, 7e2, 8e2, 9e2, 1e3, 1e4))
+print('Running Logistic Regression on day 0 to predict day 10 severity')
+day_0_histology_lr <- run_ml(day_0_histology,
+	   'glmnet',
        outcome_colname = 'hist_score',
[…]

### 2022-04-22 code/plot_figure_2.R

@@ -24,8 +24,7 @@ LOD_df <- data.frame(x = 14.5, y = 2)
 DECD_df <- data.frame(x = 12, y = 3, day = 10)
 
 cfu_data <- metadata %>% 
-	filter(cdiff_strain == 431,
-		day %in% c(1,10),
+	filter(day %in% c(1,10),
 		!is.na(cdiff_cfu)) %>% 
 	mutate(cdiff_cfu = ifelse(cdiff_cfu > 0, log10(cdiff_cfu), log10(60))) %>% 
 	left_join(donor_aes, by = c('human_source'))
@@ -57,17 +56,20 @@ cfu_plot <- cfu_data %>%
 			fill = 'white', color = 'white', inherit.aes = F) + 
 		geom_text(data = LOD_df, aes(x = x, y = y), label = "LOD", 
 			color = 'lightgray', size = 12/.pt, inherit.aes = F) + 
-		geom_segment(data = DECD_df, aes(x = 10, xend = 15, y = 5, yend = 5), 
+		geom_segment(data = DECD_df, aes(x = 10, xend = 10.8, y = 1.8, yend = 1.8), 
 			size = .25, color = 'black') + 
-		geom_label(data = DECD_df, aes(x = 12.5, y = 5), label = "Deceased", 
-			fill = 'white', color = 'white', inherit.aes = F) + 
-		geom_text(data = DECD_df, aes(x = 12.5, y = 5), label = "Deceased", 
+		geom_segment(data = DECD_df, aes(x = 14.2, xend = 15, y = 1.8, yend = 1.8), 
+			size = .25, color = 'black') + 
+		#geom_label(data = DECD_df, aes(x = 12.5, y = 1.8), label = "Deceased", 
+		#	fill = 'white', color = 'white', inherit.aes = F) + 
+		geom_text(data = DECD_df, aes(x = 12.5, y = 1.8), label = "Deceased", 
 			color = 'black', size = 12/.pt, inherit.aes = F)
 
 ###############################################################################
 #  save plot
 ###############################################################################
-ggsave(here('results/figures/Figure_2.jpg'),
+ggsave(here('submission/Figure_2.tiff'),
        cfu_plot,
-       width = 4.5, height = 3)
+       width = 4.5, height = 3, unit = 'in',
+  compression = 'lzw')
 ###############################################################################

### 2022-04-22 code/plot_figure_3.R

@@ -13,8 +13,7 @@ source(here::here('code/utilities.R'))
 ###############################################################################
 # load data
 ###############################################################################
-metadata <- read_metadata() %>% 
-  filter(cdiff_strain == '431')
+metadata <- read_metadata()
 
 toxin <- read_toxin()
 
@@ -23,7 +22,7 @@ histology <- read_histology()
 donor_aes <- donor_df
 
 LOD_df <- data.frame(x = 1.15, y = 2)
-DECD_df <- data.frame(x = 2, y = 2.5, day = 10, 
+DECD_df <- data.frame(x = 2, y = 2.5, day = factor('10'), 
   outcome = factor('Moribund', levels = c('Non-moribund', 'Moribund')))
 
 ###############################################################################
@@ -37,7 +36,9 @@ toxin <- toxin %>%
   mutate(day = as.numeric(toxin_sample_day)) %>% 
   left_join(donor_aes, by = 'human_source') %>% 
   filter(!day %in% c(3,4,7),
-         toxin_sample_type == 'stool') 
+         toxin_sample_type == 'stool') %>% 
+  mutate(day = ifelse(day == 1, '1*', as.character(day)),
+    day = factor(day, levels = c('1*', '2', '10'))) 
 
 histology <- histology %>% 
   left_join(distinct(select(metadata, mouse_id, early_euth)), by = 'mouse_id') %>% 
@@ -47,6 +48,69 @@ histology <- histology %>%
          mouse_id = paste0(cage_id, '_', ear_tag),
          score = summary_score) %>% 
   left_join(donor_aes, by = 'human_source')
+
+toxin_present <- read_toxin() %>% 
+  inner_join(select(metadata, group, human_source, early_euth), by = c('group')) %>% 
+  mutate(outcome = ifelse(early_euth == T, 'Moribund', 'Non-moribund'),
+         outcome = factor(outcome, levels = c('Non-moribund', 'Moribund'))) %>% 
+  filter(!is.na(outcome)) %>%
+  mutate(day = as.numeric(toxin_sample_day)) %>% 
+  left_join(donor_aes, by = 'human_source') %>% 
+  filter(day < 4) %>% 
+  group_by(mouse_id, outcome) %>% 
+  summarize(toxin = max(Log_repiricoal_dilution) > 1) %>% 
+  ungroup %>% 
+  mutate(toxin = ifelse(toxin, 'Toxin +', 'Toxin -'))
+
[…]

### 2022-04-22 code/plot_figure_5.R

@@ -11,187 +11,114 @@
 ###############################################################################
 # setup environment
 ###############################################################################
+library(cowplot)
+
 source(here::here('code/utilities.R'))
 
 ###############################################################################
 # load data
 ###############################################################################
-metadata <- read_metadata()
-toxin <- read_toxin()
-histology <- read_histology()
+#metadata <- read_metadata()
+#toxin <- read_toxin()
+#histology <- read_histology()
 taxonomy_df <- read_taxonomy()
-shared <- read_shared() %>% 
-	pivot_longer(cols = starts_with('Otu'), names_to = 'OTU', values_to = 'counts') %>% 
-	left_join(taxonomy_df, by = "OTU") %>% 
-	mutate(counts = counts/2107 * 100)
+#shared <- read_shared() %>% 
+#	pivot_longer(cols = starts_with('Otu'), names_to = 'OTU', values_to = 'counts') %>% 
+#	left_join(taxonomy_df, by = "OTU") %>% 
+#	mutate(counts = counts/2107 * 100)
 	
 
-features_df <- read_tsv(here('data/process/ml/ml_feature_imp.tsv'),
-		col_type = 'ddcccddcc') %>% 
-	filter(dataset != 'day_0_histology_50') %>% 
-	mutate(dataset = gsub('_80', '', dataset))
+features_df <- read_tsv(here('data/process/ml/lr_feature_imp.tsv'),
+		col_type = 'ddcccddcc')  %>% 
+	mutate(sign = ifelse(coefficient < 0,'negative', 'positive'),
+	    p_outcome =  1 - (exp(coefficient)/(1+exp(coefficient))), 
+	    OR_outcome = p_outcome/(1-p_outcome),
+	    logit_outcome = log(OR_outcome)) %>% 
+	group_by(names, dataset, method, taxonomic_level) %>% 
+	mutate(median_coef = median(coefficient),
+		mean_diff = mean(perf_metric_diff),
+		median_diff = median(perf_metric_diff)) %>% 
+	ungroup
 
 ###############################################################################
 #	setup data
 ###############################################################################
[…]

### 2022-04-22 code/plot_figure_6.R

@@ -1,145 +1,0 @@
-###############################################################################
-#
-# Figure 6
-#	main idea - Is our observation conserved across isolates?
-#	plot - CFU and histopathology score with different isolates
-#
-# Nick Lesniak 2021-10-26
-###############################################################################
-# setup environment
-###############################################################################
-source(here::here('code/utilities.R'))
-
-###############################################################################
-# load data
-###############################################################################
-isolate_levels <- c('     RT         \nIsolate         ', 
-	'027\n431', 'UM11\n299', '027\n395','027\n458')
-
-metadata <- read_metadata() %>% 
-	mutate(isolate_431 = ifelse(cdiff_strain == 431, '431', 'Other'),
-		isolate = ifelse(cdiff_strain == 299, paste0('UM11\n', cdiff_strain),
-			paste0('027\n', cdiff_strain)),
-		isolate = factor(isolate, levels = isolate_levels))
-	
-toxin <- read_toxin()
-
-histology <- read_histology() %>% 
-	select(summary_score, mouse_id)
-
-donor_aes <- donor_df
-
-donor_levels <- c(' ', 'N7' ,'N6', 'M4')
-
-###############################################################################
-#	setup data
-###############################################################################
-non_431_donors <- metadata %>% 
-	filter(cdiff_strain != 431) %>% 
-	pull(human_source) %>% 
-	unique
-
-cfu_data <- metadata %>% 
-	filter(human_source %in% non_431_donors,
-				 day == 10) %>% 
-	add_row(human_source = 'DA00578', cdiff_cfu = NA, isolate_431 = 'Other',
-		isolate = factor('027\n431', levels = isolate_levels)) %>% 
-	add_row(human_source = 'DA00578', cdiff_cfu = NA, isolate_431 = 'Other',
-		isolate = factor('027\n395', levels = isolate_levels)) %>% 
-	left_join(toxin, by = 'group') %>% 
-	left_join(donor_aes, by = c('human_source')) %>% 
-	add_row(donor_labels = factor(' ', levels = donor_levels),
-		isolate = factor('     RT         \nIsolate         ', levels = isolate_levels)) %>% 
[…]

### 2022-04-22 code/plot_figure_S4.R

@@ -34,18 +34,23 @@ lefse_design <- data.table::fread(here('data/process/lefse/temporal_trend_hilo.d
     
 lefse_df <- data.table::fread(here('data/process/lefse/temporal_trend_hilo.0.03.lefse_summary'),
 		fill = T) %>% 
-	filter(!is.na(LDA)) %>% 
-	rename(Genus = OTU) %>% 
+	filter(!is.na(LDA),
+    LDA > 3) %>% 
+	rename(tax_otu_label = OTU) %>% 
 	mutate(fctr_class = factor(Class),
 		order = LDA + 10 * as.numeric(fctr_class),
-		taxa_label = gsub('_unclassified', '*', Genus),
+    tax_otu_label = gsub('_', ' ', tax_otu_label),
+    tax_otu_label = gsub(' cla', '_cla', tax_otu_label),
+		taxa_label = gsub('_unclassified', '*', tax_otu_label),
 		taxa_label = gsub('_', ' ', taxa_label),
-		taxa_label = paste0('*', taxa_label, '*'))
+		taxa_label = paste0('*', taxa_label),
+    taxa_label = gsub(' \\(', '*  
+      \\(', taxa_label))
 ###############################################################################
 #  setup data
 ###############################################################################
 temporal_lefse_df <- taxonomy_df %>% 
-  inner_join(lefse_df, by = 'Genus') %>% 
+  inner_join(lefse_df, by = 'tax_otu_label') %>% 
   left_join(shared, by = 'OTU') %>% 
   left_join(metadata, by = c('Group' = 'group')) %>%
   group_by(mouse_id, day, taxa_label, early_euth) %>% 
@@ -55,10 +60,11 @@ temporal_lefse_df <- taxonomy_df %>%
          outcome = factor(outcome, levels = c('Low', 'High', 'Severe'))) %>% 
   filter(outcome != 'NA') %>% 
   group_by(taxa_label) %>% 
-  mutate(present = sum(relative_abundance > 0),
+  mutate(median_ra = median(relative_abundance),
+    present = sum(relative_abundance > 0),
   	relative_abundance = ifelse(relative_abundance == 0, 0.045, 
   		relative_abundance)) %>% 
-  filter(present > 60)
+  filter(present > 86)
 
 ###############################################################################
 #  plot data
@@ -70,7 +76,7 @@ temporal_lefse_plot <- temporal_lefse_df %>%
[…]

### 2022-04-22 code/render_table_S1.sh

@@ -1,0 +1,10 @@
+#!/usr/bin/env bash
+
+# author: Nick Lesniak
+# input: 
+#	results/tables/Table_S1.Rmd
+#	results/tables/header.tex
+# outputs: submission/Table_S1.pdf
+
+Rscript -e "rmarkdown::render('results/tables/Table_S1.Rmd')"
+mv results/tables/Table_S1.pdf submission/Table_S1.pdf

### 2022-04-22 exploratory/notebook/previous_cdiff_analysis_notebooks_analysis_in_separate_branch/cdiff_sequence_investigation.R

@@ -1,205 +1,0 @@
-Otu000033	3061	Bacteria(100);Firmicutes(100);Clostridia(100);Clostridiales(100);Peptostreptococcaceae(100);Clostridium_XI(100);
-Otu000152	8701	Bacteria(100);Firmicutes(100);Clostridia(100);Clostridiales(100);Peptostreptococcaceae(100);Clostridium_XI(98);
-Otu000153	29231	Bacteria(100);Firmicutes(100);Clostridia(100);Clostridiales(100);Peptostreptococcaceae(100);Clostridium_XI(100);
-Otu000628	2805	Bacteria(100);Firmicutes(100);Clostridia(100);Clostridiales(100);Peptostreptococcaceae(100);Clostridium_XI(100);
-Otu001100	228	Bacteria(100);Firmicutes(100);Clostridia(100);Clostridiales(100);Peptostreptococcaceae(100);Clostridium_XI(100);
-Otu001148	13	Bacteria(100);Firmicutes(100);Clostridia(100);Clostridiales(100);Peptostreptococcaceae(93);
-Otu001149	108	Bacteria(100);Firmicutes(100);Clostridia(100);Clostridiales(100);Peptostreptococcaceae(100);Clostridium_XI(100);
-Otu001521	9	Bacteria(100);Firmicutes(100);Clostridia(100);Clostridiales(100);Peptostreptococcaceae(100);Clostridium_XI(100);
-Otu001824	209	Bacteria(100);Firmicutes(100);Clostridia(100);Clostridiales(100);Peptostreptococcaceae(100);Peptostreptococcus(100);
-Otu001845	7	Bacteria(100);Firmicutes(100);Clostridia(100);Clostridiales(100);Peptostreptococcaceae(100);Clostridium_XI(100);
-Otu002284	3	Bacteria(100);Firmicutes(100);Clostridia(100);Clostridiales(100);Peptostreptococcaceae(100);Clostridium_XI(100);
-Otu002316	4	Bacteria(100);Firmicutes(100);Clostridia(100);Clostridiales(100);Peptostreptococcaceae(100);Clostridium_XI(100);
-Otu003821	9	Bacteria(100);Firmicutes(100);Clostridia(100);Clostridiales(100);Peptostreptococcaceae(100);
-Otu003946	4	Bacteria(100);Firmicutes(100);Clostridia(100);Clostridiales(100);Peptostreptococcaceae(100);Clostridium_XI(100);
[…]

### 2022-04-22 results/tables/Table_S1.Rmd

---
title: ""
output:
  pdf_document:
    keep_tex: false
    includes:
      in_header: header.tex
geometry: margin=1.0in
---

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE,  warning=FALSE, message=FALSE}
######################################################################
#----------------- Read in necessary libraries -------------------#
######################################################################
deps = c("knitr","rmarkdown","tidyverse");
for (dep in deps){
  if (dep %in% installed.packages()[,"Package"] == FALSE){
    install.packages(as.character(dep), quiet=TRUE);
  }
  library(dep, verbose=FALSE, character.only=TRUE)
}
######################################################################
#-------------- Define knitr options to allow kable ----------------#
######################################################################
library(kableExtra)
options(kableExtra.latex.load_packages = FALSE)
options(kableExtra.auto_format = FALSE)
options(knitr.table.format = "latex")
```

```{r table_2, echo=FALSE, results='asis'}
######################################################################
# We load a human_source_tidy.tsv file to make Table S1
# This file is generated from subsetting MIMARKS_cdclinical.xlsx 
# from Schubert et al. mBio 2014 in code/tidy_raw_data.R
######################################################################

human_src_meta <- read_tsv(here::here('data/process/human_source_tidy.tsv'),
                           col_types = cols(.default = col_character())) %>% 
    select(plot_labels, everything()) %>%
    rename_if(is.character,
           ~ c("Sample Label", "Subject ID", "Subject Location", 
                     "Sample Medium", "Subject Age", "Subject Sex", 
                     "Recent Antibiotic Use", "Proton Pump Inhibitor Use",
                     "H2 Receptor Blocker Use", "Antacid Use", "Healthcare Worker",
                     "History of CDI", "Surgery within 6 Months", "Vegetarian",
                     "Weight", "Disease Status")) %>% 
    arrange(`Sample Label`)

#------------------------Table S1 generation------------------------#
TableS1 <- kable(human_src_meta,
                "latex",
                caption = "Table S1. Demographic information of subjects whose stool samples used to colonize germ-free mice.",
                booktabs = T, 
                linesep = "") %>%
  # latex options make table to be condensed
  # table itseld is left aligned
  kable_styling(latex_options = c("striped", "scale_down", "hold_position"),
                position = "left") %>%
  # Adjust header widths
  column_spec(1, width = '4em') %>%
  column_spec(3, width = '5em') %>%
  column_spec(4, width = '7em') %>%
  column_spec(5, width = '3em') %>%
  column_spec(6, width = '5em') %>%
  column_spec(7, width = '5em') %>%
  column_spec(8, width = '5em') %>%
  column_spec(9, width = '5em') %>%
  column_spec(10, width = '5em') %>%
  column_spec(11, width = '5em') %>%
  column_spec(12, width = '5em') %>%
  column_spec(13, width = '5em') %>%
  # Add bold header
  row_spec(0,bold=TRUE)

TableS1
```

### 2022-04-22 submission/manuscript.Rmd

@@
     keep_tex: yes
+header-includes:
+    - \usepackage{setspace}
+    - \doublespacing
+    - \usepackage{lineno}
+    - \linenumbers    
 bibliography: references.bib
@@
 ```
-
-\pagenumbering{arabic}
-\linenumbers
-\doublespacing
 
@@
 difficile* ribotype and toxin genes have been associated with disease severity.
-However, it is unclear whether there is an association between members of the gut microbiota and disease severity.
+However, it is unclear whether specific members of the gut microbiota associate with variation in disease severity.
 The gut microbiota is known to interact with *C.
@@
 The variation in severity was associated with the human fecal community that the mice received.
-Generally, bacterial populations with pathogenic potential, such as *Escherichia*, *Helicobacter*,  and *Klebsiella*, were associated with more severe outcomes.
-Bacterial groups associated with fiber degradation, bile acid metabolism and lantibiotic production, such as *Anaerostipes* and *Coprobacillus*, were associated with less severe outcomes.
+Generally, bacterial populations with pathogenic potential, such as *Enterococcus*, *Helicobacter*, and *Klebsiella*, were associated with more severe outcomes.
+Bacterial groups associated with fiber degradation and bile acid metabolism, such as *Anaerotignum*, *Blautia*, *Lactonifactor*, and *Monoglobus*, were associated with less severe outcomes.
 These data indicate that, in addition to the host and *C.
-difficile*, populations of gut bacteria can influence CDI disease severity.
+difficile* subtype, populations of gut bacteria can influence CDI disease severity.
 
@@
 difficile* interacts with the gut community in many ways.
-The indigenous bacteria of a healthy intestinal community provide a protective barrier preventing *C.
-difficile* from infecting the gut.
-A range of mechanisms can disrupt this barrier, including antibiotics, medications, or dietary changes, and lead to increased susceptibility to CDI [@Britton2014; @Hryckowian2018; @VichVila2020].
+The indigenous bacteria of a healthy intestinal community prevent *C.
+difficile* from infecting the gut [@Buffie2014].
+A range of mechanisms can disrupt this inhibition, including antibiotics, medications, or dietary changes, and lead to increased susceptibility to CDI [@Britton2014; @Hryckowian2018; @VichVila2020].
 Once *C.
-difficile* overcomes the protective barrier and colonizes the intestine, the indigenous bacteria can either promote or inhibit *C.
+difficile* overcomes the inhibition and colonizes the intestine, the indigenous bacteria can either promote or inhibit *C.
 difficile* through producing molecules or modifying the environment [@Abbas2020; @Jenior2017].
 Bile acids metabolized by the gut bacteria can inhibit *C.
-difficile* growth and affect toxin production [@Sorg2008; @Thanissery2017].
+difficile* growth and affect toxin production [@Buffie2014; @Sorg2008; @Thanissery2017].
 Bacteria in the gut also can compete more directly with *C.
@@
 We hypothesized that since specific groups of gut bacteria affect the metabolism of *C.
-difficile* and its infection dynamics, we can also identify groups of bacteria that affect the disease severity of the infection.
+difficile* and its clearance rate, specific groups of bacteria associate with variation in CDI disease severity.
 To test this hypothesis, we colonized germ-free C57BL/6 mice with human fecal samples to create varied gut communities.
@@
 These human fecal samples were selected because they represented diverse community structures based on community clustering [@Schubert2014].
-The gut communities were allowed to equilibrate for two weeks post-inoculation [@Gillilland2012].
+The gut communities were allowed to colonize for two weeks post-inoculation [@Gillilland2012].
 We then surveyed the bacterial members of the gut communities by 16S rRNA gene sequencing of murine fecal pellets (Figure 1A).
@@
 At 10 days post-challenge, or earlier for the moribund mice, mice were euthanised and fecal material were assayed for toxin activity and cecal tissue was collected and scored for histopathologic signs of disease (Figure 3).
-Overall, there was greater toxin activity detected in the stool of the moribund mice (*P* = 0.003).
+Overall, there was greater toxin activity detected in the stool of the moribund mice (Figure S1).
 However, when looking at each group of mice, we observed a range in toxin activity for both the moribund and non-moribund mice (Figure 3A).
-Non-moribund mice from Donors N2 and N5 through N9 had comparable toxin activity as the moribund mice.
+Non-moribund mice from Donors N2 and N5 through N9 had comparable toxin activity as the moribund mice at 2 days post-challenge.
 Additionally, not all moribund mice had toxin activity detected in their stool.
 Next, we examined the cecal tissue for histopathologic damage.
-Moribund mice had high levels of epithelial damage, tissue edema, and inflammation (Figure S1) similar to previously reported histopathologic findings for *C.
+Moribund mice had high levels of epithelial damage, tissue edema, and inflammation (Figure S2) similar to previously reported histopathologic findings for *C.
 difficile* RT027 [@Cowardin2016].
@@
 difficile* challenge (day 0) for their relationship to infection outcomes using linear discriminant analysis (LDA) effect size (LEfSe) analysis to identify individual bacterial populations that could explain the variation in disease severity.
-We split the mice into groups by severity level based on their moribundity and histopathologic score.
-We dichotomized the histopathologic scores into high and low groups by splitting on the median score of 5.
-This analysis revealed 20 genera that were significantly different by the disease severity (Figure 4A).
-Bacterial genera *Turicibacter*, *Streptococcus*, *Staphylococcus*, *Pseudomonas*, *Phocaeicola*, *Parabacteroides*, *Bacteroides*, and *Escherichia/Shigella* were detected at higher relative abundances in the mice that became moribund.
-Populations of *Anaerotignum*, *Coprobacillus*, *Enterocloster*, and *Murimonas* were more abundant in the non-moribund mice that would develop only low intestinal injury.
-To understand the role of toxin activity in disease severity, we applied LEfSe to identify the genera most likely to explain the differences between the presence and absence of detected toxin activity (Figure 4B).
-Many genera that associated with the presence of toxin were also associated with moribundity, such as populations of *Escherichia/Shigella* and *Bacteroides*.
-Likewise, there were genera such as *Anaerotignum*, *Enterocloster*, and *Murimonas* that were associated with no detected toxin that also exhibited greater relative abundance in communities from non-moribund mice with a low histopathologic score.
-Lastly, we tested for correlations between the endpoint relative abundances of bacterial operational taxonomic units (OTUs) and the histopathologic summary score (Figure 4C).
-The endpoint relative abundance of *Bacteroides* was positively correlated with histopathologic score, as its day 0 relative abundance did with disease severity (Figure 4A).
-Populations of *Klebsiella* and *Prevotellaceae* were positively correlated with the histopathologic score and were increased in the group of mice with detectable toxin.
-This analysis identified bacterial genera that were associated with the variation in moribundity, histopathologic score, and toxin.
+We split the mice into groups by severity level based on moribundity or 10 days post infection (dpi) histopathologic score for non-moribund.
+This analysis revealed bacterial operational taxonomic units (OTUs) that were significantly different at the time of challenge by the disease severity (Figure 4A).
+OTUs associated with *Akkermansia*, *Bacteroides*, *Clostridium sensu stricto*, and *Turicibacter* were detected at higher relative abundances in the mice that became moribund.
+OTUs associated with *Anaerotignum*, *Enterocloster*, and *Murimonas* were more abundant in the non-moribund mice that would develop only low intestinal injury.
+To understand the role of toxin activity in disease severity, we applied LEfSe to identify the OTUs at the time of challenge that most likely  explain the differences between communities that had toxin activity detected at anytime point to those that did not (Figure 4B).
+An OTU associated with *Bacteroides*, OTU 7, associated with the presence of toxin also associated with moribundity.
+Likewise, OTUs associated with *Enterocloster* and *Murimonas* that were associated with no detected toxin also exhibited greater relative abundance in communities from non-moribund mice with a low histopathologic score.
+Lastly, we tested for correlations between the endpoint (10 dpi) relative abundances of OTUs and the histopathologic summary score (Figure 4C).
+The endpoint relative abundance of *Bacteroides*, OTU 17, was positively correlated with histopathologic score, as its day 0 relative abundance did with disease severity (Figure 4A).
+A population of *Bacteroides*, OTU 17, was positively correlated with the histopathologic score and were increased in the group of mice with detectable toxin.
+We also tested for correlations between the endpoint relative abundances of OTUs and toxin activity but none were significant.
+This analysis identified bacterial populations that were associated with the variation in moribundity, histopathologic score, and toxin.
 
 We next determined whether, collectively, bacterial community membership and relative abundance could be predictive of the CDI disease outcome.
-We trained random forest models with bacterial community relative abundance data from the day of colonization at each taxonomic rank to predict toxin, moribundity, and day 10 post-challenge histopathologic summary score.
-For predicting if detectable toxin would be produced, microbial populations aggregated by phylum rank classification performed similarly as models using lower taxonomic ranks (AUROC = 0.83, Figure S2).
+We trained logistic regression models with bacterial community relative abundance data from the day of colonization at each taxonomic rank to predict toxin, moribundity, and histopathologic summary score.
+For predicting if detectable toxin would be produced, microbial populations aggregated by genus rank classification performed similarly as models using lower taxonomic ranks (mean AUROC = 0.787, Figure S3).
 *C.
-difficile* was more likely to produce detectable toxin when the community infected had less abundant populations of *Verrucomicrobia* and *Campilobacterota* and had more abundant populations of *Proteobacteria* (Figure 5A).
+difficile* increased odds of producing detectable toxin when the community infected had less abundant populations of *Monoglobus*, *Akkermansia*, *Extibacter*, *Intestinimonas* and *Holdemania* and had more abundant populations of *Lachnospiraceae* (Figure 5A).
 Next, we assessed the ability of the community to predict moribundity.
-Bacteria grouped by class rank classification was sufficient to predict which mice would succumb to the infection before the end of the experiment (AUROC = 0.91, Figure S2).
-The features with the greatest effect showed that communities with greater populations of bacteria belonging to *Bacilli* and *Firmicutes* and reduced populations of *Erysipelotrichia* were more likely to result in moribundity (Figure 5B).
-Only one other class of bacteria was decreased in moribund mice, a group of unclassified *Clostridia*.
-Lastly, the relative abundances of genera were able to predict a high or low histopathologic score (histopathologic scores were dichotomized as in previous analysis, AUROC = 0.99, Figure S2).
-No genera had a significantly greater effect on the model performance than any others, indicating the model was reliant on many genera for the correct prediction.
-The model used some of the genera identified in the LEfSe analysis, such as *Coprobacillus*, *Anaerostipes*, and *Hungatella*.
-Communities with greater abundances of *Hungatella*, *Eggerthella*, *Bifidobacterium*, *Duncaniella* and *Neisseria* were more likely to have high histopathologic scores.
+Bacteria grouped by order rank classification was sufficient to predict which mice would succumb to the infection before the end of the experiment (mean AUROC = 0.9205, Figure S3).
+Many populations contributed to an increase odds of moribundity (Figure 5B).
+Populations related to *Bifidobacteriales* and *Clostridia* decreased the odds of a moribund outcome.
+Lastly, the relative abundances of OTUs were able to predict a high or low histopathologic score 10 dpi (histopathologic scores were dichotomized as in previous analysis, mean AUROC = 0.99, Figure S3).
+The model identified some similar OTUs as the LEfSe analysis, such as *Murimonas* (OTU 48), *Bacteroides* (OTU 7), and *Hungatella* (OTU 24).
 These models have shown that the relative abundance of bacterial populations and their relationship to each other could be used to predict the variation in moribundity, histopathologic score, and detectable toxin of CDI.
@@
 difficile* population [@Jenior2017; @Jenior2018; @Lesniak2021].
-Gut bacteria metabolize primary bile acids into secondary bile acids [@Staley2016; @Long2017].
+Gut bacteria metabolize primary bile acids into secondary bile acids [@Buffie2014; @Staley2016; @Long2017].
 The concentration of these bile acids affects germination, growth, toxin production and biofilm formation [@Sorg2008; @Sorg2010; @Thanissery2017; @Dubois2019].
@@
 Recent studies have shown variation in the initial microbiome, via different murine colonies or colonizing germ-free mice with human feces, that were challenged with *C.
-difficile* resulted in varied outcomes of the infection [@Tomkovich2020; @NagaoKitamoto2020].
+difficile* resulted in varied outcomes of the infection [@Tomkovich2020; @NagaoKitamoto2020; @Battaglioli2018].
 
 Our data have demonstrated gut bacterial relative abundances associate with variation in toxin production, histopathologic scoring of the cecal tissue and mortality.
-This analysis revealed populations of *Akkermansia*, *Anaerostipes*, *Coprobacillus*, *Enterocloster*, *Lactonifactor*, and *Monoglobus* were more abundant in the microbiome of non-moribund mice which had low histopathologic scores and no detected toxin.
-The protective role of these genera are supported by previous studies.
-*Coprobacillus*, *Lactonifactor*, and *Monoglobus* have been shown to be involved in dietary fiber fermentation and associated with healthy communities [@Mabrok2011; @Kim2019; @doPrado2021; @Muthuramalingam2019].
-*Anaerostipes* and *Coprobacillus*, which produce short chain fatty acids, have been associated with healthy communities [@Han2019; @Duncan2004; @Ye2018].
-Furthermore, *Coprobacillus*, which was abundant in mice with low histopathologic scores but rare in all other mice, has been shown to contain a putative type I lantibiotic gene cluster and inhibit *C.
-difficile* colonization [@Walsh2017; @Sandiford2018; @Stein2013].
+This analysis revealed populations of *Akkermansia*, *Anaerotignum*, *Blautia*, *Enterocloster*, *Lactonifactor*, and *Monoglobus* were more abundant in the microbiome of non-moribund mice which had low histopathologic scores and no detected toxin.
+The protective role of these bacteria are supported by previous studies.
+*Blautia*, *Lactonifactor*, and *Monoglobus* have been shown to be involved in dietary fiber fermentation and associated with healthy communities [@Liu2021; @Mabrok2011; @Kim2019].
+*Anaerotignum*, which produce short chain fatty acids, has been associated with healthy communities [@Choi2019; @Ueki2017].
 *Akkermansia* and *Enterocloster* were also identified as more abundant in mice which had a low histopathologic scores but have contradictory supporting evidence in the current literature.
-In our data, *Akkermansia* was most abundant in the non-moribund mice with low histopathologic scores but there were some moribund mice which had increased populations of *Akkermansia*.
-This could be attributed to either a more protective mucus layer was present inhibiting colonization [@Stein2013; @Nakashima2021] or mucus consumption by *Akkermansia* could have been crossfeeding *C.
+In our data, a population of *Akkermansia*, OTU 5, was most abundant in the non-moribund mice with low histopathologic scores but moribund mice had increased population of *Akkermansia*, OTU 8.
+This difference could indicate either a more protective mucus layer was present inhibiting colonization [@Stein2013; @Nakashima2021] or mucus consumption by *Akkermansia* could have been crossfeeding *C.
 difficile* or exposing a niche for *C.
@@
 The groups of bacteria that were associated with either a higher histopathologic score or moribundity are members of the indigenous gut community that also have been associated with disease, often referred to as opportunistic pathogens.
-Many of the populations with pathogenic potential that associated with worse outcomes are also facultative anaerobes.
-*Enterococcus*, *Klebsiella*, *Shigella/Escherichia*, *Staphylococcus*, and *Streptococcus* have been shown to expand after antibiotic use [@VanInsberghe2020; @GarzaGonzalez2019; @Schubert2014] and are commonly detected in CDI cases [@Shafiq2020; @Keith2020; @Zackular2016; @Berkell2021].
+Some of the populations of *Bacteroides*, *Enterococcus*, and *Klebsiella* that associated with worse outcomes, have been shown to have pathogenic potential, expand after antibiotic use, and are commonly detected in CDI cases [@Tomkovich2021; @Keith2020; @Zackular2016; @Berkell2021].
 In addition to these populations, *Eggerthella*, *Prevotellaceae* and *Helicobacter*, which associated with worse outcomes, have also been associated with intestinal inflammation [@Gardiner2014; @Iljazovic2020; @Nagalingam2013].
@@
 difficile* infection and host damage.
-Aside from *Helicobacter*, these groups of bacteria that associated with more severe outcomes did not have a conserved association between their relative abundance and the disease severity across all mice.
+These bacteria groups increased in severe outcomes maintained their differences throughout the length of the experiment (Figure S4).
+These results agreed Aside from *Helicobacter*, these groups of bacteria that associated with more severe outcomes did not have a conserved association between their relative abundance and the disease severity across all mice.
 
@@
 difficile* population had the same genetic background, they too were reacting to the specific microbial community.
+Different gut microbial communities can also have different effects on the host immune responses [@Lundberg2020].
 Disease severity is driven by the cumulative effect of the host immune response and the activity of  *C.
@@
 **Animal care.** 6- to 13-week old male and female germ-free C57BL/6 were obtained from a single breeding colony in the University of Michigan Germ-free Mouse Core.
-Mice (N1 n=11, N2 n=7, N3 n=3, N4 n=3, N5 n=3, N6 n=3, N7 n=7, N8 n=3, N9 n=2, M1 n=3, M2 n=3, M3 n=3, M4 n=3, M5 n=7, M6 n=3) were housed in cages of 2-4 mice per cage and maintained in germ-free isolators at the University of Michigan germ-free facility.
+Mice (M1 n=3, M2 n=3, M3 n=3, M4 n=3, M5 n=7, M6 n=3, N1 n=11, N2 n=7, N3 n=3, N4 n=3, N5 n=3, N6 n=3, N7 n=7, N8 n=3, N9 n=2) were housed in cages of 2-4 mice per cage and maintained in germ-free isolators at the University of Michigan germ-free facility.
 All mouse experiments were approved by the University Committee on Use and Care of Animals at the University of Michigan.
@@
 ***C.
-difficile* experiments.** Human fecal samples were obtained as part of Schubert *et al.* and selected based on community clusters [@Schubert2014] to result in diverse community structures.
+difficile* experiments.** Human fecal samples were obtained as part of Schubert *et al.* and selected based on community clusters [@Schubert2014] to result in diverse community structures (Table S1).
 Feces were homogenized by mixing 200 mg of sample with 5 ml of PBS.
@@
 difficile*.
+Stool samples from each mouse were collected one day prior to *C.
+difficile* and plated for *C.
+difficile* enumeration to confirm no *C.
+difficile* was detected in stool prior to challenge.
 *C.
@@
 Slides were scored as 0-4 for parameters of epithelial damage, tissue edema, and inflammation and a summary score of 0-12 was generated by summing the three individual parameter scores.
+For non-moribund mice, histopathological summary scores used for LEfSe and logistic regression were split into high and low groups based on greater or less than the median summary score of 5 because the had a bimodal distribution (*P* < 0.05).
 
 **Statistical analysis and modeling.** To compare community structures, we calculated Yue and Clayton dissimilarity matrices ($\theta$~YC~) in mothur [@Yue2005].
-We rarefied samples to 2,107 sequences per sample to limit uneven sampling biases.
-We tested for differences in individual taxonomic groups that would explain the outcome differences with LEfSe [@Segata2011] in mothur.
+For this calculation, we averaged of 1000 sub-samples, or rarified, samples to 2,107 sequence reads per sample to limit uneven sampling biases.
+We tested for differences in individual taxonomic groups that would explain the outcome differences with LEfSe [@Segata2011] in mothur (default parameters, LDA > 4).
+We tested for differences in temporal trends through fitting a linear model to each OTU and testing for differences between histopathological summary scores with LEfSe [@Segata2011] in mothur (default parameters, LDA > 3).
 Remaining statistical analysis and data visualization was performed in R (v4.0.5) with the tidyverse package (v1.3.1).
-We tested for significant differences in $\beta$-diversity ($\theta$~YC~) using the Wilcoxon rank sum test.
+We tested for significant differences in $\beta$-diversity ($\theta$~YC~), histopathological scores, and toxin activity using the Wilcoxon rank sum test, non-unimodality to non-moribund histopathological summary score using Hartigans' dip test, and  toxin detection in mice using the Pearson's Chi-square test.
 We used Spearman's correlation to identify which OTUs that had a correlation between their relative abundance and the histopathologic summary score.
 *P* values were then corrected for multiple comparisons with a Benjamini and Hochberg adjustment for a type I error rate of 0.05 [@Benjamini1995].
-We built random forest models using the mikropml package [@Topcuoglu2021] with relative abundance summed by taxonomic ranks from day 0 samples using mtry values of 1 through 10, 15, 20, 25, 40, 50, 100.
-The split for training and testing varied by model to avoid overfitting the data.
-To determine the optimal split, we tested splits (50%, 60%, 70%, 80%, 90% data used for training) to find the greatest portion of data that could be used to train the model while still maintaining the same performance for the training model as the model with the held-out test data.
-The toxin and moribundity models were trained with 60% of the data.
-The histopathologic score model was trained with 80% of the data.
+We built L2 logistic regression models using the mikropml package [@Topcuoglu2021].
+Sequence counts were summed by taxonomic ranks from day 0 samples, normalized by centering to the feature mean and scaling by the standard deviation, and features positively or negatively correlated were collapsed into a single feature.
+We ran our models using alpha = 0 and lambda values of 1e-0, 1e1, 1e2, 2e2, 3e2, 4e2, 5e2, 6e2, 7e2, 8e2, 9e2, 1e3, 1e4 with a split of 80% of the data for training and 20% of the data for testing.
 Lastly, we did not compare murine communities to donor community or clinical data because germ-free mice colonized with non-murine fecal communities have been shown to more closely resemble the murine communities than the donor species community [@Rawls2006].
@@
 
+Conceptualization: N.A.L., A.M.S., K.J.F., P.D.S.; Data curation: N.A.L., K.J.F.; Formal analysis: N.A.L., K.J.F., J.L.L., I.L.B.; Investigation: N.A.L., A.M.S., H.S., I.L.B., V.B.Y., P.D.S.; Methodology: N.A.L., A.M.S., K.J.F., J.L.L., H.S., I.L.B., V.B.Y., P.D.S.; Resources: N.A.L., A.M.S., P.D.S.; Software: NAL; Visualization: N.A.L., K.J.F., P.D.S.; Writing - original draft: N.A.L.; Writing - review & editing: N.A.L., A.M.S., K.J.F., J.L.L., H.S., I.L.B., V.B.Y., P.D.S.; Funding acquisition: V.B.Y.; Project administration: P.D.S.; Supervision: P.D.S.
+
 \newpage
@@
 \newpage
-
-![](../results/figures/figure_1.jpg)
 
@@
 (B) Median (points) and interquartile range (lines) of $\beta$-diversity ($\theta$~YC~) between an individual mouse and either all others which were inoculated with feces from the same donor or from a different donor.
-The $\beta$-diversity among the same donor comparison group was significantly less than the $\beta$-diversity of the different donor group (*P* < 0.05, calculated by Wilcoxon rank sum test).
+The $\beta$-diversity among the same donor comparison group was significantly less than the $\beta$-diversity of either the different donor group or the donor community (*P* < 0.05, calculated by Wilcoxon rank sum test).
 
 \hfill\break
-
-![](../results/figures/figure_2.jpg)
 
@@
 LOD = Limit of detection.
+-Deceased- indicates mice were deceased at that time point so no sample was available.
 
 \hfill\break
-
-![](../results/figures/figure_3.jpg)
 
@@
 Missing points are from mice that had insufficient fecal sample collected for assaying toxin or cecum for histopathologic scoring.
+* indicates significant difference between non-moribund and moribund groups of mice by Wilcoxon test (*P* < 0.002).
 LOD = Limit of detection.
+-Deceased- indicates mice were deceased at that time point so no sample was available.
 
@@
 
-![](../results/figures/figure_4.jpg)
-
 **Figure 4.
-Individual fecal bacterial community members of the murine gut associated with *C .difficile* infection outcomes.** (A and B) Relative abundance of genera at the time of *C.
+Individual fecal bacterial community members of the murine gut associated with *C .difficile* infection outcomes.** (A and B) Relative abundance of OTUs at the time of *C.
 difficile* challenge (Day 0) that varied significantly by the moribundity and histopathologic summary score or detected toxin by LEfSe analysis.
 Median (points) and interquartile range (lines) are plotted.
-Genera are ordered alphabetically to ease comparisons across analyses.
-(A) Relative abundances were compared across infection outcome of moribund (colored black) or non-moribund with either a high histopathologic score (score greater than the median score of 5, colored green) or a low histopathologic summary score (score less than the median score of 5, colored light green).
-(B) Relative abundances were compared between mice which toxin activity was detected (Toxin +, colored dark purple) and which no toxin activity was detected (Toxin -, colored light purple).
-(C) Endpoint bacterial OTUs correlated with histopathologic summary score.
-Each individual mouse is plotted (transparent gray point).
+(A) Day 0 relative abundances were compared across infection outcome of moribund (colored black) or non-moribund with either a high histopathologic score (score greater than the median score of 5, colored green) or a low histopathologic summary score (score less than the median score of 5, colored light green).
+(B) Day 0 relative abundances were compared between mice which toxin activity was detected (Toxin +, colored dark purple) and which no toxin activity was detected (Toxin -, colored light purple).
+(C) Day 10 bacterial OTU relative abundances correlated with histopathologic summary score.
+Each individual mouse is plotted and colored according to their categorization in panel A.
+Points at the median score of 5 (gray points) were not included in panel A.
 Spearman's correlations were statistically significant after Benjamini-Hochberg correction for multiple comparisons.
-All bacterial groups are ordered alphabetically.
+All bacterial groups are ordered by the LDA score.
 * indicates that the bacterial group was unclassified at lower taxonomic classification ranks.
@@
 
-![](../results/figures/figure_5.jpg)
-
 **Figure 5.
 Fecal bacterial community members of the murine gut at the time of *C.
-difficile* infection predicted outcomes of the infection.** On the day of infection (Day 0), bacterial community members grouped by different classification rank were modeled with random forest to predict the infection outcome.
-The models used the highest taxonomic classification rank that performed as well as the lower ranks.
+difficile* infection predicted outcomes of the infection.** On the day of infection (Day 0), bacterial community members grouped by different classification rank were modeled with logistic regression to predict the infection outcome.
+The models used the highest taxonomic classification rank without a decrease in performance.
+Models used all community members but plotted are those members with a mean odds ratio not equal to 1.
 Median (solid points) and interquartile range (lines) of the group relative abundance are plotted.
-Bacterial groups are ordered by their importance to the model; taxonomic group at the top of the plot had the greatest decrease in performance when its relative abundances were permuted.
+Bacterial groups are ordered by their odds ratio.
 * indicates that the bacterial group was unclassified at lower taxonomic classification ranks.
-(A) Bacterial members grouped by phyla predicted which mice would have toxin activity detected at any point throughout the infection (Toxin +, dark purple).
-(B) Bacterial members grouped by class predicted which mice would become moribund (dark blue).
-(C) Bacterial members grouped by genera predicted if the mice would have a high (score greater than the median score of 5, colored dark green) or low (score less than the median score of 5, colored light green) histopathologic summary score.
+(A) Bacterial members grouped by genus predicted which mice would have toxin activity detected at any point throughout the infection (Toxin +, dark purple).
+(B) Bacterial members grouped by order predicted which mice would become moribund (dark blue).
+(C) Bacterial members grouped by OTU predicted if the mice would have a high (score greater than the median score of 5, colored dark green) or low (score less than the median score of 5, colored light green) histopathologic summary score.
 
@@
 
-![](../results/figures/figure_S1.jpg)
+**Figure S1.
+Toxin detect in mice based on outcome of the infection.** Comparison of the distribution of number of either non-moribund or moribund mice which toxin was detected in the first three days post infection.
+Bars are colored by whether toxin was detected in stool from the mouse (dark purple) or not (light purple).
+Moribund mice had significantly more mice with toxin detected (*P* < 0.008) by Pearson's Chi-square test.
 
-**Figure S1.
+\hfill\break
+
+**Figure S2.
 Histopathologic score of tissue damage at the endpoint of the infection.** Tissue collected at the endpoint, either day 10 post-challenge (Non-moribund) or day mice succumbed to infection (Moribund), were scored from histopathologic damage.
@@
 Missing points are from mice that had insufficient sample for histopathologic scoring.
+* indicates significant difference between non-moribund and moribund groups of mice by Wilcoxon test (*P* < 0.002).
 
@@
 
-![](../results/figures/figure_S2.jpg)
-
-**Figure S2.
-Random forest models predicted outcomes of the *C.
+**Figure S3.
+Logistic regression models predicted outcomes of the *C.
 difficile* challenge.** (A-C) Taxonomic classification rank model performance.
@@
 Black rectangle highlights classification rank used to model each outcome.
-(D-F) Model feature importance.
-Bacterial groups are ordered by their decrease in area under receiver-operator curve (AUC) when its relative abundances was permuted.
-Individual relative abundances were added to F since differences in AUC were outside the interquartile range.
-* indicates bacterial group was unclassified at lower taxonomic classification ranks.
-For all plots, median (solid points) and interquartile range (lines) are plotted.
+For all plots, median (large solid points), interquartile range (lines), and individual models (small transparent points) are plotted.
 (A) Toxin production modeled which mice would have toxin detected during the experiment.
@@
 (C) Histopathologic score modeled which mice would have a high (score greater than the median score of 5) or low (score less than the median score of 5) histopathologic summary score.
-(D) Bacterial phyla which affected the performance of predicting detectable toxin activity when permuted.
-(E) Bacterial classes which affected the performance of predicting moribundity when permuted.
-(D) Bacterial genera which affected the performance of predicting histopathologic score when permuted.
+
+\hfill\break
+
+**Figure S4.
+Temporal dynamics of OTUs that differed between histopathologic summary score.** Relative abundance of OTUs on each day relative to the time of *C.
+difficile* challenge (Day 0) that have a significantly different temporal trend by the histopathologic summary score by LEfSe analysis.
+Median (points) and interquartile range (lines) are plotted.
+Points and lines are colored by infection outcome of moribund (colored black) or non-moribund with either a high histopathologic score (score greater than the median score of 5, colored green) or a low histopathologic summary score (score less than the median score of 5, colored light green).

### 2022-04-25 submission/manuscript.tex
incorporate PDS feedback in revision

@@
 samples were selected because they represented diverse community
-structures based on community clustering (18). The gut communities were
-allowed to colonize for two weeks post-inoculation (19). We then
-surveyed the bacterial members of the gut communities by 16S rRNA gene
-sequencing of murine fecal pellets (Figure 1A). The bacterial
-communities from each mouse grouped more closely to those communities
-from mice that received the same human fecal donor community than to the
-mice who received a different human fecal donor community (Figure 1B).
-The communities were primarily composed of populations of
-\emph{Clostridia}, \emph{Bacteroidia}, \emph{Erysipelotrichia},
-\emph{Bacilli}, and \emph{Gammaproteobacteria}. However, the gut
-bacterial communities of each donor group of mice harbored unique
-relative abundance distributions of the shared bacterial classes.
+structures based on community clustering (18). After the gut communities
+had colonized for two weeks, we confirmed them to be \emph{C. difficile}
+negative by culture (19). We then surveyed the bacterial members of the
+gut communities by 16S rRNA gene sequencing of murine fecal pellets
+(Figure 1A). The bacterial communities from each mouse grouped more
+closely to those communities from mice that received the same human
+fecal donor community than to the mice who received a different human
+fecal donor community (Figure 1B). The communities were primarily
+composed of populations of \emph{Clostridia}, \emph{Bacteroidia},
+\emph{Erysipelotrichia}, \emph{Bacilli}, and \emph{Gammaproteobacteria}.
+However, the gut bacterial communities of each donor group of mice
+harbored unique relative abundance distributions of the shared bacterial
+classes.
 
@@
 and scaling by the standard deviation, and features positively or
-negatively correlated were collapsed into a single feature. We ran our
-models using alpha = 0 and lambda values of 1e-0, 1e1, 1e2, 2e2, 3e2,
-4e2, 5e2, 6e2, 7e2, 8e2, 9e2, 1e3, 1e4 with a split of 80\% of the data
-for training and 20\% of the data for testing. Lastly, we did not
-compare murine communities to donor community or clinical data because
-germ-free mice colonized with non-murine fecal communities have been
-shown to more closely resemble the murine communities than the donor
-species community (88). Furthermore, it is not our intention to make any
-inferences regarding human associated bacteria and their relationship
-with human CDI outcome.
+negatively correlated were collapsed into a single feature. For each L2
+logistic regression model, we ran 100 random iterations using values of
+1e-0, 1e1, 1e2, 2e2, 3e2, 4e2, 5e2, 6e2, 7e2, 8e2, 9e2, 1e3, 1e4 for the
+L2 regularization penalty with a split of 80\% of the data for training
+and 20\% of the data for testing. Lastly, we did not compare murine
+communities to donor community or clinical data because germ-free mice
+colonized with non-murine fecal communities have been shown to more
+closely resemble the murine communities than the donor species community
+(88). Furthermore, it is not our intention to make any inferences
+regarding human associated bacteria and their relationship with human
+CDI outcome.
 
@@
 P.D.S.; Funding acquisition: V.B.Y.; Project administration: P.D.S.;
-Supervision: P.D.S.
-
-\newpage
+Supervision: P.D.S. \newpage
 

### 2022-04-25 submission/manuscript.Rmd

@@
 These human fecal samples were selected because they represented diverse community structures based on community clustering [@Schubert2014].
-The gut communities were allowed to colonize for two weeks post-inoculation [@Gillilland2012].
+After the gut communities had colonized for two weeks, we confirmed them to be *C.
+difficile* negative by culture [@Gillilland2012].
 We then surveyed the bacterial members of the gut communities by 16S rRNA gene sequencing of murine fecal pellets (Figure 1A).
@@
 Sequence counts were summed by taxonomic ranks from day 0 samples, normalized by centering to the feature mean and scaling by the standard deviation, and features positively or negatively correlated were collapsed into a single feature.
-We ran our models using alpha = 0 and lambda values of 1e-0, 1e1, 1e2, 2e2, 3e2, 4e2, 5e2, 6e2, 7e2, 8e2, 9e2, 1e3, 1e4 with a split of 80% of the data for training and 20% of the data for testing.
+For each L2 logistic regression model, we ran 100 random iterations using values of 1e-0, 1e1, 1e2, 2e2, 3e2, 4e2, 5e2, 6e2, 7e2, 8e2, 9e2, 1e3, 1e4 for the L2 regularization penalty with a split of 80% of the data for training and 20% of the data for testing.
 Lastly, we did not compare murine communities to donor community or clinical data because germ-free mice colonized with non-murine fecal communities have been shown to more closely resemble the murine communities than the donor species community [@Rawls2006].
@@
 Conceptualization: N.A.L., A.M.S., K.J.F., P.D.S.; Data curation: N.A.L., K.J.F.; Formal analysis: N.A.L., K.J.F., J.L.L., I.L.B.; Investigation: N.A.L., A.M.S., H.S., I.L.B., V.B.Y., P.D.S.; Methodology: N.A.L., A.M.S., K.J.F., J.L.L., H.S., I.L.B., V.B.Y., P.D.S.; Resources: N.A.L., A.M.S., P.D.S.; Software: NAL; Visualization: N.A.L., K.J.F., P.D.S.; Writing - original draft: N.A.L.; Writing - review & editing: N.A.L., A.M.S., K.J.F., J.L.L., H.S., I.L.B., V.B.Y., P.D.S.; Funding acquisition: V.B.Y.; Project administration: P.D.S.; Supervision: P.D.S.
-
 \newpage

### 2022-06-15 submission/manuscript.Rmd
add Table S1 legend to main text
update manuscript with reviewer feedback

@@
 Two mice were able to resist *C.
-difficile* colonization, both received their community donor N1, which may be attributed to experimental variation since this group also had more mice.
+difficile* colonization, both received their community from Donor N1, which may be attributed to experimental variation since this group also had more mice.
 By colonizing germ-free mice with different human fecal communities, we were able to generate diverse gut communities in mice, which were susceptible to *C.
@@
 difficile* challenge (day 0) for their relationship to infection outcomes using linear discriminant analysis (LDA) effect size (LEfSe) analysis to identify individual bacterial populations that could explain the variation in disease severity.
-We split the mice into groups by severity level based on moribundity or 10 days post infection (dpi) histopathologic score for non-moribund.
+We split the mice into groups by severity level based on moribundity or 10 days post infection (dpi) histopathologic score for non-moribund mice.
 This analysis revealed bacterial operational taxonomic units (OTUs) that were significantly different at the time of challenge by the disease severity (Figure 4A).
 OTUs associated with *Akkermansia*, *Bacteroides*, *Clostridium sensu stricto*, and *Turicibacter* were detected at higher relative abundances in the mice that became moribund.
-OTUs associated with *Anaerotignum*, *Enterocloster*, and *Murimonas* were more abundant in the non-moribund mice that would develop only low intestinal injury.
+OTUs associated with *Anaerotignum*, *Enterocloster*, and *Murimonas* were more abundant in the non-moribund mice that would develop low intestinal injury.
 To understand the role of toxin activity in disease severity, we applied LEfSe to identify the OTUs at the time of challenge that most likely  explain the differences between communities that had toxin activity detected at anytime point to those that did not (Figure 4B).
@@
 Likewise, OTUs associated with *Enterocloster* and *Murimonas* that were associated with no detected toxin also exhibited greater relative abundance in communities from non-moribund mice with a low histopathologic score.
-Lastly, we tested for correlations between the endpoint (10 dpi) relative abundances of OTUs and the histopathologic summary score (Figure 4C).
+We tested for correlations between the endpoint (10 dpi) relative abundances of OTUs and the histopathologic summary score (Figure 4C).
 The endpoint relative abundance of *Bacteroides*, OTU 17, was positively correlated with histopathologic score, as its day 0 relative abundance did with disease severity (Figure 4A).
-A population of *Bacteroides*, OTU 17, was positively correlated with the histopathologic score and were increased in the group of mice with detectable toxin.
+The population of OTU 17 was also increased in the group of mice with detectable toxin.
 We also tested for correlations between the endpoint relative abundances of OTUs and toxin activity but none were significant.
+Lastly, we tested for associations between temporal changes and disease severity (Figure S4).
+Most groups of bacteria maintained higher relative abundance, relative to the other other outcome groups, from day 0 through the end of the experiment.
 This analysis identified bacterial populations that were associated with the variation in moribundity, histopathologic score, and toxin.
@@
 We trained logistic regression models with bacterial community relative abundance data from the day of colonization at each taxonomic rank to predict toxin, moribundity, and histopathologic summary score.
+We used the highest taxonomic classification rank which performed similar to lower ranks, which suggested the effect is associated with general attributes of the bacterial group as opposed to specific functions of more refined grouping.
 For predicting if detectable toxin would be produced, microbial populations aggregated by genus rank classification performed similarly as models using lower taxonomic ranks (mean AUROC = 0.787, Figure S3).
@@
 Recently, *Helicobacter hepaticus* was shown to be sufficient to cause susceptibility to CDI in IL-10 deficient C57BL/6 mice [@AbernathyClose2021].
-In our experiments, when *Helicobacter* was present, the infection resulted in a high histopathologic score (Figure 4C).
+In our experiments, when *Helicobacter* was present, the infection was more likely to result in a high histopathologic score (Figure 4C, S4).
 While we did not use IL-10 deficient mice, it is possible the bacterial community or host response are similarly modified by *Helicobacter*, allowing *C.
 difficile* infection and host damage.
-These bacteria groups increased in severe outcomes maintained their differences throughout the length of the experiment (Figure S4).
-These results agreed Aside from *Helicobacter*, these groups of bacteria that associated with more severe outcomes did not have a conserved association between their relative abundance and the disease severity across all mice.
+Aside from *Helicobacter*, these groups of bacteria that associated with more severe outcomes did not have a conserved association between their relative abundance and the disease severity across all mice.
 
@@
 difficile*, and the indigenous microbes at the site of infection.
-In the first set of experiments, we used the same host background, C57BL/6 mice, the same infecting microbe, *C.
+In this set of experiments, we used the same host background, C57BL/6 mice, the same infecting microbe, *C.
 difficile* RT027 clinical isolate 431, with different gut bacterial communities.
-The bacterial groups in those communities were often present in both moribund and non-moribund and across the range of histopathologic scores.
+The bacterial groups in those communities were often present in both moribund and non-moribund mice and across the range of histopathologic scores.
 Thus, it was not merely the presence of the bacteria but their activity in response to the other microbes and host which affect the extent of the host damage.
@@
 We tested for differences in individual taxonomic groups that would explain the outcome differences with LEfSe [@Segata2011] in mothur (default parameters, LDA > 4).
-We tested for differences in temporal trends through fitting a linear model to each OTU and testing for differences between histopathological summary scores with LEfSe [@Segata2011] in mothur (default parameters, LDA > 3).
+We tested for differences in temporal trends through fitting a linear regression model to each OTU and tested for differences in regression coefficients by  histopathological summary scores with LEfSe [@Segata2011] in mothur (default parameters, LDA > 3).
 Remaining statistical analysis and data visualization was performed in R (v4.0.5) with the tidyverse package (v1.3.1).
@@
 
-**Code availability.** Scripts necessary to reproduce our analysis and this paper are available in an online repository (https://github.com/SchlossLab/Lesniak_Severity_XXXX_2022).
+**Code availability.** Scripts necessary to reproduce our analysis and this paper are available in an online repository (https://github.com/SchlossLab/Lesniak_Severity_mBio_2022).
 
@@
 Models used all community members but plotted are those members with a mean odds ratio not equal to 1.
-Median (solid points) and interquartile range (lines) of the group relative abundance are plotted.
+Median (solid points) and interquartile range (lines) of the odds ratio are plotted.
 Bacterial groups are ordered by their odds ratio.
 * indicates that the bacterial group was unclassified at lower taxonomic classification ranks.
-(A) Bacterial members grouped by genus predicted which mice would have toxin activity detected at any point throughout the infection (Toxin +, dark purple).
-(B) Bacterial members grouped by order predicted which mice would become moribund (dark blue).
-(C) Bacterial members grouped by OTU predicted if the mice would have a high (score greater than the median score of 5, colored dark green) or low (score less than the median score of 5, colored light green) histopathologic summary score.
+(A) Bacterial members grouped by genus predicted which mice would have toxin activity detected at any point throughout the infection.
+Data with a decreased probability of toxin activity are colored light purple and those with an increased probability of toxin activity are colored dark purple.
+(B) Bacterial members grouped by order predicted which mice would become moribund.
+Data with a decreased probability of moribundity are colored light blue and those with an increased probability of moribundity are colored dark blue.
+(C) Bacterial members grouped by OTU predicted if the mice would have a high (greater than the median score of 5) or low (less than the median score of 5) histopathologic summary score.
+Data with a decreased probability of high histopathologic score are colored light green and those with an increased probability of high histopathologic score are colored dark green.
 
@@
 difficile* challenge (Day 0) that have a significantly different temporal trend by the histopathologic summary score by LEfSe analysis.
-Median (points) and interquartile range (lines) are plotted.
+Median (points) and interquartile range (lines) of relative abundances are plotted.
 Points and lines are colored by infection outcome of moribund (colored black) or non-moribund with either a high histopathologic score (score greater than the median score of 5, colored green) or a low histopathologic summary score (score less than the median score of 5, colored light green).
+
+\hfill\break
+
+**Table S1. Demographic information of subjects whose stool samples used to colonize germ-free mice.**

### 2022-06-15 submission/manuscript.tex
add Table S1 legend to main text
update manuscript with reviewer feedback

@@
 be colonized (Figure 2). Two mice were able to resist \emph{C.
-difficile} colonization, both received their community donor N1, which
-may be attributed to experimental variation since this group also had
-more mice. By colonizing germ-free mice with different human fecal
+difficile} colonization, both received their community from Donor N1,
+which may be attributed to experimental variation since this group also
+had more mice. By colonizing germ-free mice with different human fecal
 communities, we were able to generate diverse gut communities in mice,
@@
 severity level based on moribundity or 10 days post infection (dpi)
-histopathologic score for non-moribund. This analysis revealed bacterial
-operational taxonomic units (OTUs) that were significantly different at
-the time of challenge by the disease severity (Figure 4A). OTUs
-associated with \emph{Akkermansia}, \emph{Bacteroides},
+histopathologic score for non-moribund mice. This analysis revealed
+bacterial operational taxonomic units (OTUs) that were significantly
+different at the time of challenge by the disease severity (Figure 4A).
+OTUs associated with \emph{Akkermansia}, \emph{Bacteroides},
 \emph{Clostridium sensu stricto}, and \emph{Turicibacter} were detected
@@
 \emph{Murimonas} were more abundant in the non-moribund mice that would
-develop only low intestinal injury. To understand the role of toxin
-activity in disease severity, we applied LEfSe to identify the OTUs at
-the time of challenge that most likely explain the differences between
+develop low intestinal injury. To understand the role of toxin activity
+in disease severity, we applied LEfSe to identify the OTUs at the time
+of challenge that most likely explain the differences between
 communities that had toxin activity detected at anytime point to those
@@
 exhibited greater relative abundance in communities from non-moribund
-mice with a low histopathologic score. Lastly, we tested for
-correlations between the endpoint (10 dpi) relative abundances of OTUs
-and the histopathologic summary score (Figure 4C). The endpoint relative
+mice with a low histopathologic score. We tested for correlations
+between the endpoint (10 dpi) relative abundances of OTUs and the
+histopathologic summary score (Figure 4C). The endpoint relative
 abundance of \emph{Bacteroides}, OTU 17, was positively correlated with
 histopathologic score, as its day 0 relative abundance did with disease
-severity (Figure 4A). A population of \emph{Bacteroides}, OTU 17, was
-positively correlated with the histopathologic score and were increased
-in the group of mice with detectable toxin. We also tested for
-correlations between the endpoint relative abundances of OTUs and toxin
-activity but none were significant. This analysis identified bacterial
-populations that were associated with the variation in moribundity,
-histopathologic score, and toxin.
+severity (Figure 4A). The population of OTU 17 was also increased in the
+group of mice with detectable toxin. We also tested for correlations
+between the endpoint relative abundances of OTUs and toxin activity but
+none were significant. Lastly, we tested for associations between
+temporal changes and disease severity (Figure S4). Most groups of
+bacteria maintained higher relative abundance, relative to the other
+other outcome groups, from day 0 through the end of the experiment. This
+analysis identified bacterial populations that were associated with the
+variation in moribundity, histopathologic score, and toxin.
 
@@
 abundance data from the day of colonization at each taxonomic rank to
-predict toxin, moribundity, and histopathologic summary score. For
-predicting if detectable toxin would be produced, microbial populations
-aggregated by genus rank classification performed similarly as models
-using lower taxonomic ranks (mean AUROC = 0.787, Figure S3). \emph{C.
-difficile} increased odds of producing detectable toxin when the
-community infected had less abundant populations of \emph{Monoglobus},
-\emph{Akkermansia}, \emph{Extibacter}, \emph{Intestinimonas} and
-\emph{Holdemania} and had more abundant populations of
-\emph{Lachnospiraceae} (Figure 5A). Next, we assessed the ability of the
-community to predict moribundity. Bacteria grouped by order rank
-classification was sufficient to predict which mice would succumb to the
-infection before the end of the experiment (mean AUROC = 0.9205, Figure
-S3). Many populations contributed to an increase odds of moribundity
-(Figure 5B). Populations related to \emph{Bifidobacteriales} and
-\emph{Clostridia} decreased the odds of a moribund outcome. Lastly, the
-relative abundances of OTUs were able to predict a high or low
+predict toxin, moribundity, and histopathologic summary score. We used
+the highest taxonomic classification rank which performed similar to
+lower ranks, which suggested the effect is associated with general
+attributes of the bacterial group as opposed to specific functions of
+more refined grouping. For predicting if detectable toxin would be
+produced, microbial populations aggregated by genus rank classification
+performed similarly as models using lower taxonomic ranks (mean AUROC =
+0.787, Figure S3). \emph{C. difficile} increased odds of producing
+detectable toxin when the community infected had less abundant
+populations of \emph{Monoglobus}, \emph{Akkermansia}, \emph{Extibacter},
+\emph{Intestinimonas} and \emph{Holdemania} and had more abundant
+populations of \emph{Lachnospiraceae} (Figure 5A). Next, we assessed the
+ability of the community to predict moribundity. Bacteria grouped by
+order rank classification was sufficient to predict which mice would
+succumb to the infection before the end of the experiment (mean AUROC =
+0.9205, Figure S3). Many populations contributed to an increase odds of
+moribundity (Figure 5B). Populations related to \emph{Bifidobacteriales}
+and \emph{Clostridia} decreased the odds of a moribund outcome. Lastly,
+the relative abundances of OTUs were able to predict a high or low
 histopathologic score 10 dpi (histopathologic scores were dichotomized
@@
 mice (73). In our experiments, when \emph{Helicobacter} was present, the
-infection resulted in a high histopathologic score (Figure 4C). While we
-did not use IL-10 deficient mice, it is possible the bacterial community
-or host response are similarly modified by \emph{Helicobacter}, allowing
-\emph{C. difficile} infection and host damage. These bacteria groups
-increased in severe outcomes maintained their differences throughout the
-length of the experiment (Figure S4). These results agreed Aside from
-\emph{Helicobacter}, these groups of bacteria that associated with more
-severe outcomes did not have a conserved association between their
-relative abundance and the disease severity across all mice.
+infection was more likely to result in a high histopathologic score
+(Figure 4C, S4). While we did not use IL-10 deficient mice, it is
+possible the bacterial community or host response are similarly modified
+by \emph{Helicobacter}, allowing \emph{C. difficile} infection and host
+damage. Aside from \emph{Helicobacter}, these groups of bacteria that
+associated with more severe outcomes did not have a conserved
+association between their relative abundance and the disease severity
+across all mice.
 
@@
 microbe, \emph{C. difficile}, and the indigenous microbes at the site of
-infection. In the first set of experiments, we used the same host
-background, C57BL/6 mice, the same infecting microbe, \emph{C.
-difficile} RT027 clinical isolate 431, with different gut bacterial
-communities. The bacterial groups in those communities were often
-present in both moribund and non-moribund and across the range of
-histopathologic scores. Thus, it was not merely the presence of the
-bacteria but their activity in response to the other microbes and host
-which affect the extent of the host damage. Additionally, while each
-mouse and \emph{C. difficile} population had the same genetic
-background, they too were reacting to the specific microbial community.
-Different gut microbial communities can also have different effects on
-the host immune responses (76). Disease severity is driven by the
-cumulative effect of the host immune response and the activity of
-\emph{C. difficile} and the gut bacteria. \emph{C. difficile} drives
-host damage through the production of toxin. The gut microbiota can
-modulate host damage through the balance of metabolic and competitive
-interactions with \emph{C. difficile}, such as bacteriocin production or
-mucin degradation, and interactions with the host, such as host mucus
-glycosylation or intestinal IL-33 expression (15, 77). For example, low
-levels of mucin degradation can provide nutrients to other community
-members producing a diverse non-damaging community (78). However, if
-mucin degradation becomes too great it reduces the protective function
-of the mucin layer and exposes the epithelial cells. This
-over-harvesting can contribute to the host damage due to other members
-producing toxin. Thus, the resultant intestinal damage is the balance of
-all activities in the gut environment. Host damage is the emergent
-property of numerous damage-response curves, such as one for host immune
-response, one for \emph{C. difficile} activity and another for
-microbiome community activity, each of which are a composite curve of
-the individual activities from each group, such as antibody production,
-neutrophil infiltration, toxin production, sporulation, fiber and mucin
+infection. In this set of experiments, we used the same host background,
+C57BL/6 mice, the same infecting microbe, \emph{C. difficile} RT027
+clinical isolate 431, with different gut bacterial communities. The
+bacterial groups in those communities were often present in both
+moribund and non-moribund mice and across the range of histopathologic
+scores. Thus, it was not merely the presence of the bacteria but their
+activity in response to the other microbes and host which affect the
+extent of the host damage. Additionally, while each mouse and \emph{C.
+difficile} population had the same genetic background, they too were
+reacting to the specific microbial community. Different gut microbial
+communities can also have different effects on the host immune responses
+(76). Disease severity is driven by the cumulative effect of the host
+immune response and the activity of \emph{C. difficile} and the gut
+bacteria. \emph{C. difficile} drives host damage through the production
+of toxin. The gut microbiota can modulate host damage through the
+balance of metabolic and competitive interactions with \emph{C.
+difficile}, such as bacteriocin production or mucin degradation, and
+interactions with the host, such as host mucus glycosylation or
+intestinal IL-33 expression (15, 77). For example, low levels of mucin
+degradation can provide nutrients to other community members producing a
+diverse non-damaging community (78). However, if mucin degradation
+becomes too great it reduces the protective function of the mucin layer
+and exposes the epithelial cells. This over-harvesting can contribute to
+the host damage due to other members producing toxin. Thus, the
+resultant intestinal damage is the balance of all activities in the gut
+environment. Host damage is the emergent property of numerous
+damage-response curves, such as one for host immune response, one for
+\emph{C. difficile} activity and another for microbiome community
+activity, each of which are a composite curve of the individual
+activities from each group, such as antibody production, neutrophil
+infiltration, toxin production, sporulation, fiber and mucin
 degradation. Therefore, while we have identified populations of
@@
 \textgreater{} 4). We tested for differences in temporal trends through
-fitting a linear model to each OTU and testing for differences between
-histopathological summary scores with LEfSe (85) in mothur (default
-parameters, LDA \textgreater{} 3). Remaining statistical analysis and
-data visualization was performed in R (v4.0.5) with the tidyverse
-package (v1.3.1). We tested for significant differences in
-\(\beta\)-diversity (\(\theta\)\textsubscript{YC}), histopathological
-scores, and toxin activity using the Wilcoxon rank sum test,
-non-unimodality to non-moribund histopathological summary score using
-Hartigans' dip test, and toxin detection in mice using the Pearson's
-Chi-square test. We used Spearman's correlation to identify which OTUs
-that had a correlation between their relative abundance and the
-histopathologic summary score. \emph{P} values were then corrected for
-multiple comparisons with a Benjamini and Hochberg adjustment for a type
-I error rate of 0.05 (86). We built L2 logistic regression models using
-the mikropml package (87). Sequence counts were summed by taxonomic
-ranks from day 0 samples, normalized by centering to the feature mean
-and scaling by the standard deviation, and features positively or
-negatively correlated were collapsed into a single feature. For each L2
-logistic regression model, we ran 100 random iterations using values of
-1e-0, 1e1, 1e2, 2e2, 3e2, 4e2, 5e2, 6e2, 7e2, 8e2, 9e2, 1e3, 1e4 for the
-L2 regularization penalty with a split of 80\% of the data for training
-and 20\% of the data for testing. Lastly, we did not compare murine
-communities to donor community or clinical data because germ-free mice
-colonized with non-murine fecal communities have been shown to more
-closely resemble the murine communities than the donor species community
-(88). Furthermore, it is not our intention to make any inferences
-regarding human associated bacteria and their relationship with human
-CDI outcome.
+fitting a linear regression model to each OTU and tested for differences
+in regression coefficients by histopathological summary scores with
+LEfSe (85) in mothur (default parameters, LDA \textgreater{} 3).
+Remaining statistical analysis and data visualization was performed in R
+(v4.0.5) with the tidyverse package (v1.3.1). We tested for significant
+differences in \(\beta\)-diversity (\(\theta\)\textsubscript{YC}),
+histopathological scores, and toxin activity using the Wilcoxon rank sum
+test, non-unimodality to non-moribund histopathological summary score
+using Hartigans' dip test, and toxin detection in mice using the
+Pearson's Chi-square test. We used Spearman's correlation to identify
+which OTUs that had a correlation between their relative abundance and
+the histopathologic summary score. \emph{P} values were then corrected
+for multiple comparisons with a Benjamini and Hochberg adjustment for a
+type I error rate of 0.05 (86). We built L2 logistic regression models
+using the mikropml package (87). Sequence counts were summed by
+taxonomic ranks from day 0 samples, normalized by centering to the
+feature mean and scaling by the standard deviation, and features
+positively or negatively correlated were collapsed into a single
+feature. For each L2 logistic regression model, we ran 100 random
+iterations using values of 1e-0, 1e1, 1e2, 2e2, 3e2, 4e2, 5e2, 6e2, 7e2,
+8e2, 9e2, 1e3, 1e4 for the L2 regularization penalty with a split of
+80\% of the data for training and 20\% of the data for testing. Lastly,
+we did not compare murine communities to donor community or clinical
+data because germ-free mice colonized with non-murine fecal communities
+have been shown to more closely resemble the murine communities than the
+donor species community (88). Furthermore, it is not our intention to
+make any inferences regarding human associated bacteria and their
+relationship with human CDI outcome.
 
@@
 and this paper are available in an online repository
-(\url{https://github.com/SchlossLab/Lesniak_Severity_XXXX_2022}).
+(\url{https://github.com/SchlossLab/Lesniak_Severity_mBio_2022}).
 
@@
 odds ratio not equal to 1. Median (solid points) and interquartile range
-(lines) of the group relative abundance are plotted. Bacterial groups
-are ordered by their odds ratio. * indicates that the bacterial group
-was unclassified at lower taxonomic classification ranks. (A) Bacterial
-members grouped by genus predicted which mice would have toxin activity
-detected at any point throughout the infection (Toxin +, dark purple).
-(B) Bacterial members grouped by order predicted which mice would become
-moribund (dark blue). (C) Bacterial members grouped by OTU predicted if
-the mice would have a high (score greater than the median score of 5,
-colored dark green) or low (score less than the median score of 5,
-colored light green) histopathologic summary score.
+(lines) of the odds ratio are plotted. Bacterial groups are ordered by
+their odds ratio. * indicates that the bacterial group was unclassified
+at lower taxonomic classification ranks. (A) Bacterial members grouped
+by genus predicted which mice would have toxin activity detected at any
+point throughout the infection. Data with a decreased probability of
+toxin activity are colored light purple and those with an increased
+probability of toxin activity are colored dark purple. (B) Bacterial
+members grouped by order predicted which mice would become moribund.
+Data with a decreased probability of moribundity are colored light blue
+and those with an increased probability of moribundity are colored dark
+blue. (C) Bacterial members grouped by OTU predicted if the mice would
+have a high (greater than the median score of 5) or low (less than the
+median score of 5) histopathologic summary score. Data with a decreased
+probability of high histopathologic score are colored light green and
+those with an increased probability of high histopathologic score are
+colored dark green.
 
@@
 score by LEfSe analysis. Median (points) and interquartile range (lines)
-are plotted. Points and lines are colored by infection outcome of
-moribund (colored black) or non-moribund with either a high
-histopathologic score (score greater than the median score of 5, colored
-green) or a low histopathologic summary score (score less than the
-median score of 5, colored light green).
+of relative abundances are plotted. Points and lines are colored by
+infection outcome of moribund (colored black) or non-moribund with
+either a high histopathologic score (score greater than the median score
+of 5, colored green) or a low histopathologic summary score (score less
+than the median score of 5, colored light green).
+
+\hfill\break
+
+\textbf{Table S1. Demographic information of subjects whose stool
+samples used to colonize germ-free mice.}
 
